### 2017-11-30 paper/livecoms-template-bestpractices.tex
paper template files added

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% LIVECOMS ARTICLE TEMPLATE FOR BEST PRACTICES GUIDE
%%% ADAPTED FROM ELIFE ARTICLE TEMPLATE (8/10/2017)
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% PREAMBLE
\documentclass[9pt,bestpractices]{livecoms}
% Use the 'onehalfspacing' option for 1.5 line spacing
% Use the 'doublespacing' option for 2.0 line spacing
% Use the 'lineno' option for adding line numbers.
% The 'bestpractices' option for indicates that this is a best practices guide.
% Omit the bestpractices option to remove the marking as a LiveCoMS paper.
% Please note that these options may affect formatting.

\usepackage{lipsum} % Required to insert dummy text
\usepackage[version=4]{mhchem}
\usepackage{siunitx}
\DeclareSIUnit\Molar{M}
\usepackage[italic]{mathastext}
\graphicspath{{figures/}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% IMPORTANT USER CONFIGURATION
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand{\versionnumber}{1.3}  % you should update the minor version number in preprints and major version number of submissions.
\newcommand{\githubrepository}{\url{https://github.com/myaccount/homegithubrepository}}  %this should be the main github repository for this article

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{This is the title : v\versionnumber}

\author[1*]{Firstname Middlename Surname}
\author[1,2\authfn{1}\authfn{3}]{Firstname Middlename Familyname}
\author[2\authfn{1}\authfn{4}]{Firstname Initials Surname}
\author[2*]{Firstname Surname}
\affil[1]{Institution 1}
\affil[2]{Institution 2}

\corr{email1@example.com}{FMS}  % Correspondence emails.  FMS and FS are the appropriate authors initials.
\corr{email2@example.com}{FS}

\contrib[\authfn{1}]{These authors contributed equally to this work}
\contrib[\authfn{2}]{These authors also contributed equally to this work}

\presentadd[\authfn{3}]{Department, Institute, Country}
\presentadd[\authfn{4}]{Department, Institute, Country}

\blurb{This LiveCoMS document is maintained online on GitHub at \githubrepository; to provide feedback, suggestions, or help improve it, please visit the GitHub repository and participate via the issue tracker.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE START
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\begin{frontmatter}
\maketitle

\begin{abstract}
This particular document provides a skeleton illustrating key sections for a Best Practices document. Please see the sample \texttt{sample-document.tex} in \url{github.com/livecomsjournal/article_templates/templates} for additional information on and examples of using the LiveCoMS LaTeX class.
Here we also assume familiarity with LaTeX and knowledge of how to include figures, tables, etc.; if you want examples, see the sample just referenced.

In your work, in this particular slot, please provide an abstract of no more than 250 words.
Your abstract should explain the main contributions of your article, and should not contain any material that is not included in the main text.
Please note that your abstract, plus the authorship material following it, must not extend beyond the title page or modifications to the LaTeX class will likely be needed.
\end{abstract}

\end{frontmatter}

\section{Introduction}

Here you would explain what problem you are tackling and briefly motivate your work.

In this particular template, we have removed most of the usage examples which occur in \texttt{sample-document.tex} to provide a minimal template you can modify; however, we retain a couple of examples illustrating more unusual features of our templates/article class, such as the checklists, and information on algorithms and pseudocode.

\section{Prerequisites}

Here you would identify prerequisites/background knowledge that are assumed by your work and your checklist which you view as critical, ideally giving links to good sources on these topics.
Checklists are normally focused on errors made by users with training and experience in molecular simulations, so you can assume a basic familiarity with the fundamentals of molecular simulations.

\section{Checklist}
Here we use a full-page checklist with multiple sections, so it will appear on a separate page of the sample PDF.
Other checklist formats are possible, as shown in the sample \texttt{sample-document.tex} in \url{github.com/livecomsjournal/article_templates/templates}.

Your checklist should include a succinct list of steps that people should follow when carrying out the task in question.
This is provided to ensure certain basic standards are followed and common but critical major errors are avoided.
Note that a checklist is not intended to cover \emph{all} important steps, but rather focus on the most common reasons for failure or incorrect results, or issues which are particularly crucial.

% This provides a checklist which
% - spans a full page
% - consists of multiple sub-checklists
% - exists on a separate page
% This style of checklist will be especially helpful if you want to encourage readers to print and use your checklist in practice, as they
% can easily print it without also printing other material from your manuscript. However, other styles of checklist are also possible (below).
\begin{Checklists*}[p!]

\begin{checklist}{First list}
\textbf{You can easily make full width checklists that take a whole page}
\begin{itemize}
\item Items in your checklist can and should reference sections where the cited issues are discussed in detail, such as Section~\ref{sec:reference_this}
\item Also remember
\item And finally
\end{itemize}
\end{checklist}

\begin{checklist}{Second list}
\textbf{This is some further description.}
\begin{itemize}
\item First thing
\item Also remember
\item And finally
\end{itemize}
\end{checklist}

\begin{checklist}{Third list}
\textbf{This is some further description.}
\begin{itemize}
\item First thing
\item Also remember
\item And finally
\end{itemize}
\end{checklist}

\begin{checklist}{Fourth list}
\textbf{This is some further description.}
\begin{itemize}
\item First thing
\item Also remember
\item And finally
\end{itemize}
\end{checklist}

\end{Checklists*}

\section{Rationale}

Your Rationale section, or sections, can follow or precede your checklist (we expect that often, following the checklist will be preferable) and provide the necessary rationale for the checklist, and act as more complete \emph{best practices} description.
This should include 1) significant detail as to the possible alternative ways to accomplish a given task, 2) description of advantages and disadvantages of the various approaches, and 3) significant literature documentation about reasons for choices.

\subsection{Algorithms and Pseudocode}
\label{sec:reference_this}

The \texttt{algpseudocode} and \texttt{algorithms} packages is loaded by the document class. \ALG{euclid} was taken directly from the package documentation. (Please do not load \texttt{algorithm2e}; it's not compatible with \texttt{algpseudocde} nor \texttt{algorithms}!)

\begin{algorithm}
\caption{Euclid's algorithm}\label{alg:euclid}
\begin{algorithmic}%[1]  %% uncomment to enable line numbers
\Procedure{Euclid}{$a,b$}\Comment{The g.c.d. of a and b}
   \State $r\gets a\bmod b$
   \While{$r\not=0$}\Comment{We have the answer if r is 0}
      \State $a\gets b$
      \State $b\gets r$
      \State $r\gets a\bmod b$
   \EndWhile\label{euclidendwhile}
   \State \textbf{return} $b$\Comment{The gcd is b}
\EndProcedure
\end{algorithmic}
\end{algorithm}

\section{Author Contributions}
%%%%%%%%%%%%%%%%
% This section mustt describe the actual contributions of
% author. Since this is an electronic-only journal, there is
% no length limit when you describe the authors' contributions,
% so we recommend describing what they actually did rather than
% simply categorizing them in a small number of
% predefined roles as might be done in other journals.
%
% See the policies ``Policies on Authorship'' section of https://livecoms.github.io
% for more information on deciding on authorship and author order.
%%%%%%%%%%%%%%%%

(Explain the contributions of the different authors here)

% We suggest you preserve this comment:
For a more detailed description of author contributions,
see the GitHub issue tracking and changelog at \githubrepository.

\section{Other Contributions}
%%%%%%%%%%%%%%%
% You should include all people who have filed issues that were
% accepted into the paper, or that upon discussion altered what was in the paper.
% Multiple significant contributions might mean that the contributor
% should be moved to authorship at the discretion of the a
%
% See the policies ``Policies on Authorship'' section of https://livecoms.github.io for
% more information on deciding on authorship and author order.
%%%%%%%%%%%%%%%

(Explain the contributions of any non-author contributors here)
% We suggest you preserve this comment:
For a more detailed description of contributions from the community and others, see the GitHub issue tracking and changelog at \githubrepository.

\section{Potentially Conflicting Interests}
%%%%%%%
%Declare any potentially competing interests, financial or otherwise
%%%%%%%

Declare any potentially conflicting interests here, whether or not they pose an actual conflict in your view.

\section{Funding Information}
%%%%%%%
% Authors should acknowledge funding sources here. Reference specific grants.
%%%%%%%
FMS acknowledges the support of NSF grant CHE-1111111.

\bibliography{livecoms-sample}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% APPENDICES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\appendix

\end{document}

### 2017-11-30 paper/manuscript.tex
changed name of main mansucript

--- 2017-11-30 paper/livecoms-template-bestpractices.tex
@@
 
-\newcommand{\versionnumber}{1.3}  % you should update the minor version number in preprints and major version number of submissions.
-\newcommand{\githubrepository}{\url{https://github.com/myaccount/homegithubrepository}}  %this should be the main github repository for this article
+\newcommand{\versionnumber}{0.1}  % you should update the minor version number in preprints and major version number of submissions.
+\newcommand{\githubrepository}{\url{https://github.com/michellab/best_practice_afec}  %this should be the main github repository for this article
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{This is the title : v\versionnumber}
+\title{Best Practices for Alchemical Free Energy Calculations : v\versionnumber}
 

### 2017-12-02 paper/manuscript.tex
converted conted of google drive document into the mansucript template and added some vague to dos

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
+\usepackage[colorinlistoftodos]{todonotes}
 \newcommand{\versionnumber}{0.1}  % you should update the minor version number in preprints and major version number of submissions.
-\newcommand{\githubrepository}{\url{https://github.com/michellab/best_practice_afec}  %this should be the main github repository for this article
+\newcommand{\githubrepository}{\url{https://github.com/michellab/best_practice_afec}}  %this should be the main github repository for this article
 
@@
 
-\author[1*]{Firstname Middlename Surname}
-\author[1,2\authfn{1}\authfn{3}]{Firstname Middlename Familyname}
-\author[2\authfn{1}\authfn{4}]{Firstname Initials Surname}
-\author[2*]{Firstname Surname}
-\affil[1]{Institution 1}
-\affil[2]{Institution 2}
+\author[1*]{John D. Chodera}
+\author[2*]{Antonia S. J. S. Mey}
+\author[2*]{Julien Michel}
+\author[3*]{David L. Mobley}
+\author[4*]{Conor Parks}
+\author[5*]{Julia E. Rice}
+\author[6*]{Michael Shirts}
+\affil[1]{Computational and Systems Biology Program, Memorial Sloan Kettering Cancer Center, New York NY 10065}
+\affil[2]{Departments of Pharmaceutical Sciences and Chemistry, University of California, Irvine}
+\affil[3]{EaStCHEM School of Chemistry, David Brewster road, Joseph Black Building, The King's Buildings, Edinburgh, EH9 3FJ, UK}
+\affil[4]{I don't know my affiliation}
+\affil[5]{I don't know my affiliation}
 
-\corr{email1@example.com}{FMS}  % Correspondence emails.  FMS and FS are the appropriate authors initials.
-\corr{email2@example.com}{FS}
+\corr{john.chodera@choderalab.org}{JDC}
+\corr{dmobley@mobleylab.org}{DLM}
+\corr{antonia.mey@ed.ac.uk}{ASJSM}
+\corr{info@julienmichel.net}{JM}
+\corr{ddd@yyy}{CP}
+\corr{fff@ddd}{MS}
 
@@
 \contrib[\authfn{2}]{These authors also contributed equally to this work}
-
-\presentadd[\authfn{3}]{Department, Institute, Country}
-\presentadd[\authfn{4}]{Department, Institute, Country}
 
@@
 \begin{abstract}
+\todo[inline, color={green!20}]{ASJSM: @Volunteer write abstract}
 This particular document provides a skeleton illustrating key sections for a Best Practices document.
@@
 
+\todototoc
+\listoftodos
+
+%%%%%%%%%%%%%%%%%%%%
+%              Introduction                  %
+%%%%%%%%%%%%%%%%%%%%
 \section{Introduction}
+\label{sec:intro}
+\todo[inline, color={green!20}]{ASJSM: @Volunteer write Intro}
+Meeting Members: 
+David Mobley, Conor Parks, Samarjeet, Julia Rice, Toni Mey, John Chodera, Michael Shirts
+[please add your name if you have contributed to the document!]
+(Original Members: John Chodera, Conor Parks, Heather Mayes)
+People to Invite: Clara Christ (drug discovery expertise), Hannes Loeffler (FESetup expertise)
 
-Here you would explain what problem you are tackling and briefly motivate your work.
-
-In this particular template, we have removed most of the usage examples which occur in \texttt{sample-document.tex} to provide a minimal template you can modify; however, we retain a couple of examples illustrating more unusual features of our templates/article class, such as the checklists, and information on algorithms and pseudocode.
-
+%%%%%%%%%%%%%%%%%%%%
+%              Prerequesites                %
+%%%%%%%%%%%%%%%%%%%%
 \section{Prerequisites}
-
+\label{sec:pre}
 Here you would identify prerequisites/background knowledge that are assumed by your work and your checklist which you view as critical, ideally giving links to good sources on these topics.
@@
 
+\section{Scope and Goals}
+\label{sec:scope}
+\begin{itemize}
+\item Preparation (focused on aspects unique to alchemical free energies), execution, and analysis of:
+\begin{itemize}
+\item Transfer free energies (hydration free energies, partition coefficients, etc.)
+\item Binding free energies
+\end{itemize}
+\item Not in scope:
+\begin{itemize}
+\item Advanced ligand binding topics such as:
+\item Covalent inhibition
+\item Association which is not 1:1
+\item Endpoint free energy methods
+\item PMF binding free energy methods
+\item Force field choice: Should be a separate document (review)
+\end{itemize}
+\end{itemize}
+
+\textbf{Goals:}
+\begin{itemize}
+\item Checklist and background for new practitioners of relative and absolute alchemical free energy calculations: What should you pay attention to in setting up and running calculations in common codes and why.
+\item Provide guidance to authors as to what should be reported about their protocols in a Methods section, either conforming to standard or reporting where they did not conform and why.
+Useful to reviewers as well.
+\end{itemize}
+
 \section{Checklist}
-Here we use a full-page checklist with multiple sections, so it will appear on a separate page of the sample PDF.
-Other checklist formats are possible, as shown in the sample \texttt{sample-document.tex} in \url{github.com/livecomsjournal/article_templates/templates}.
-
-Your checklist should include a succinct list of steps that people should follow when carrying out the task in question.
-This is provided to ensure certain basic standards are followed and common but critical major errors are avoided.
-Note that a checklist is not intended to cover \emph{all} important steps, but rather focus on the most common reasons for failure or incorrect results, or issues which are particularly crucial.
+\label{sec:checklist}
+\todo[inline, color={green!20}]{ASJSM: @Volunteer This needs to be revised and expanded, some initial thoughts were just thrown in.}
+An attempt at identifying most important checklist items.
 
@@
 
-\begin{checklist}{First list}
-\textbf{You can easily make full width checklists that take a whole page}
+\begin{checklist}{Step 0 -- Know what you want to simulate }
+\textbf{What are the first questions that need addressing before setting up a molecular dynamics simulation}\\
+Extensive explanation for the checklist questions can be found in section~\ref{sec:step0}.
 \begin{itemize}
-\item Items in your checklist can and should reference sections where the cited issues are discussed in detail, such as Section~\ref{sec:reference_this}
-\item Also remember
+\item Can I get the required accuracy with the simulation I want to carry out
+\item 
 \item And finally
@@
 
-\begin{checklist}{Second list}
-\textbf{This is some further description.}
+\begin{checklist}{Simulation preparation}
+\textbf{How do I get started setting up an alchemical free energy calculation}
+Extensive explanation for the checklist questions can be found in section~\ref{sec:step1}.
+\begin{itemize}
+\item Have I followed the Best practices for biomolecular simulation set up?
+\item In a relative simulation, will I run into problems with clashing geometries in the ligand transformation or crystal waters?
+\end{itemize}
+\end{checklist}
+\end{Checklists*}
+
+\begin{Checklists*}[p!]
+\begin{checklist}{Absolute simulations}
+\textbf{What are the main things I need to consider for an absolute alchemical free energy calculation?}
+Extensive explanation for the checklist questions can be found in section~\ref{sec:step2}.
+\begin{itemize}
+\item Topology
+\item Restraints
+\item Standard state handling
+\end{itemize}
+\end{checklist}
+
+\begin{checklist}{Relative simulations}
+\textbf{What are the main things I need to consider for an relative alchemical free energy calculation?}
+Extensive explanation for the checklist questions can be found in section~\ref{sec:step2}.
 \begin{itemize}
@@
 
-\begin{checklist}{Third list}
-\textbf{This is some further description.}
+\begin{checklist}{Analysis}
+\textbf{This is all about analysis of the simulation}
+Extensive explanation for the checklist questions can be found in section~\ref{sec:step4}.
 \begin{itemize}
-\item First thing
-\item Also remember
-\item And finally
-\end{itemize}
-\end{checklist}
-
-\begin{checklist}{Fourth list}
-\textbf{This is some further description.}
-\begin{itemize}
-\item First thing
-\item Also remember
-\item And finally
+\item Are my simulations converged enough?
+\item Am I using the right analysis techniques?
 \end{itemize}
@@
 \end{Checklists*}
+\clearpage
 
-\section{Rationale}
+\section{Step 0 -- What can be expected from alchemical simulations?}
+\label{sec:step0}
+\begin{itemize}
+\item What level of accuracy you can expect? 
+\item What timescales and how many transformations can you address given available computational resources? 
+\item Can you even hope to tackle the problem you are attempting? 
+\end{itemize}
 
-Your Rationale section, or sections, can follow or precede your checklist (we expect that often, following the checklist will be preferable) and provide the necessary rationale for the checklist, and act as more complete \emph{best practices} description.
-This should include 1) significant detail as to the possible alternative ways to accomplish a given task, 2) description of advantages and disadvantages of the various approaches, and 3) significant literature documentation about reasons for choices.
+\section{Step 1 -- Simulation prerequesites}
+\label{sec:step1}
+\begin{enumerate}
+\item Generate geometry of initial state: Reference biomolecular simulation preparation best practices
+\item Relative: Generate geometry of final state (Mey)
+\begin{itemize}
+\item e.g. ligand ideally shouldn’t clash with receptor, etc.; satisfy constraints that might be imposed by protocol such as overlapping atoms, etc.
+\item reference biomolecular preparation setup practices for placing ligand into binding site, but elaborate on constraints that must be considered for relative free energy calculations
+\end{itemize}
+\end{enumerate}
 
-\subsection{Algorithms and Pseudocode}
-\label{sec:reference_this}
+\section{Step 2 -- Simulation protocol selection}
+\label{sec:step2}
+\begin{enumerate}
+\item Relative: Select a topology and produce an atom mapping for transformation (Mobley)
+\begin{itemize}
+\item Watching out for constraints to bonds for hydrogen: these cannot be allowed to change without including Jacobian terms
+\item Share atoms between initial and final ligands if possible, otherwise colocalize pair of compounds and exclude their interactions with one another
+\item Ring breaking/forming: Special are is needed; cite references
+\item Tools: enlist Hannes Loeffler (developer of FESetup)?
+\item Clarify terminology: Dual-topology, single-topology, etc.
+\end{itemize}
 
-The \texttt{algpseudocode} and \texttt{algorithms} packages is loaded by the document class.
-\ALG{euclid} was taken directly from the package documentation.
-(Please do not load \texttt{algorithm2e}; it's not compatible with \texttt{algpseudocde} nor \texttt{algorithms}!)
+\item Absolute: Identify \textbf{restraints and standard state handling}
+\begin{itemize}
+\item Practical use of Boresch restraints/other forms of restraints. How to choose atoms involved. etc. (e.g. see Heinzelmann/Gilson BRD4 work; Chodera also has stuff to add)
+\item Levi Naden: Problems with analytical approximation to standard-state correction for Boresch restraints
+\item Harmonic and flat-bottom restraints
+\item Clarify terminology: Double decoupling, etc. See Feature Box below.
+\end{itemize}
+\item Identify whether \textbf{net charge is changing }and how this will be handled:
+Issues that still need to be resolved: Charge corrections vs alchemically modify counterions
 
-\begin{algorithm}
-\caption{Euclid's algorithm}\label{alg:euclid}
-\begin{algorithmic}%[1]  %% uncomment to enable line numbers
-\Procedure{Euclid}{$a,b$}\Comment{The g.c.d. of a and b}
-   \State $r\gets a\bmod b$
-   \While{$r\not=0$}\Comment{We have the answer if r is 0}
-      \State $a\gets b$
-      \State $b\gets r$
-      \State $r\gets a\bmod b$
-   \EndWhile\label{euclidendwhile}
-   \State \textbf{return} $b$\Comment{The gcd is b}
-\EndProcedure
-\end{algorithmic}
-\end{algorithm}
+\item Select an \textbf{alchemical pathway}
+\begin{enumerate}
+\item Choice of alchemical Hamiltonian
+	\begin{itemize}
+	\item Softcore potentials are always recommended, but might not ALWAYS be necessary, e.g.:
+	\item Pure changes in parameters of atoms that don’t insert/delete atoms (turn into dummies/from dummies)
+	\item If roughly isosteric (e.g. lambda dynamics work from C. Brooks)
+	\end{itemize}
+\item Choice of discrete alchemical protocol (Shirts, Mey, Chodera)
+	\begin{itemize}
+	\item Many options: Adaptive scheme, Chebyshev polynomials, linear spacing, “choose your next lambda from data at this lambda”, optimal thermodynamic length approaches (separately: Shirts, Sivak, Huafeng Xu).
+	\item Levi Naden had paper with lambda protocol which worked for all cases -- methane solvation, host-guest (including disappearing host)
+	\end{itemize}
+\item Relative
+Three-stage protocol (discharge unique initial atoms, transform LJ, charge unique final atoms) vs softcore electrostatics/LJ
 
-\section{Author Contributions}
+\item Absolute
+	\begin{itemize}
+	\item Select a \textbf{common alchemically-eliminated end state}
+	\item Decoupled vs annihilated for electrostatics and LJ
+	\item Sequential electrostatics and LJ versus simultaneous (recommend sequential)
+
+\end{itemize}
+\item Concerns:
+Part of AMBER still can’t run at endpoints (lambda = 0 or 1); SANDER cannot but PMEMD can.
+
+\end{enumerate}
+
+\item Determine whether you need to \textbf{handle multiple binding modes}: (Mobley)
+\begin{itemize}
+\item In absolute: Confine-and-release, BLUES
+\item In relative (ugh!!)
+\end{itemize}
+\item Determine \textbf{stopping conditions}
+Uncertainty-directed stopping criteria can ensure target uncertainty is achieved
+
+\item Select which \textbf{data should be saved and with which frequency}
+\begin{itemize}
+\item What data to save: dU/dlambda, Delta E’s between neighbor for BAR, between further for MBAR, …
+\item BAR captures most of info with well-optimized lambda protocol, but MBAR when perhaps not, except when there are way too many lambda values.
+\item Recommend against solely relying on TI when possible 
+\item Recommend cross-comparing methods (TI (spline, trapezoid, etc.), BAR, MBAR) as diagnosis of trouble
+\end{itemize}
+
+\end{enumerate}
+
+\section{Step 3 -- Overview of available analysis techniques}
+\label{sec:step3}
+\begin{enumerate}
+\item Detecting boundary between equilibrated and production regions (Chodera: \url{http://dx.doi.org/10.1021/acs.jctc.5b00784})
+\item Decorrelating samples for analysis
+\begin{itemize}
+\item Subsample different lambdas based on correlation times
+\item Ensure all simulations at least 50x correlation time
+\end{itemize}
+\item Examining output data for common problems with discussions of what exactly to plot or look at; examples of typical curves for dV/dlambda and free energy versus lambda, for example
+\begin{itemize}
+\item Make sure ligand doesn’t tumble out of binding site (Mey has observed this)
+\item Significant discrepancies between different free energy estimators (TI, BAR, MBAR)
+\item Poor replica mixing (for replica-exchange)
+\item Correlation time as a function of lambda as it would be expected to be a smooth
+\item Dependence on initial conformation
+\item Torsional analysis: Is it stuck in specific states? Only very rarely transitions?
+\item More “usual suspects”
+\end{itemize}
+\item Estimators for free energies
+\begin{itemize}
+\item MBAR recommended if all energy differences are available
+\item BAR just as good for highly optimized lambda values
+\item TI should be roughly concordant, but quadrature error hard to quantify
+\item Other variants useful in special circumstances (e.g. Z. Tan stochastic version)
+\end{itemize}
+\item Computing and reporting uncertainties on free energies
+correlated bootstrap v. timeseries analysis
+\item Other considerations for many transformations
+Cycle closure error
+
+\end{enumerate}
+
+%%%%%%%%%%%%%%%%%%%%
+%              Terminology                  %
+%%%%%%%%%%%%%%%%%%%%
+\section{Terminology and abbreviations}
+\label{sec:tem-abbrev}
+\begin{itemize}
+\item Feature Box covering major technical terms and abbreviations
+\item Examples:
+\begin{itemize}
+\item EXP, BAR, MBAR
+\item Double decoupling, single-topology, dual-topology, hybrid-topology, coupled-topology
+\item FEP (free energy peturbation), alchemical, AFE (alchemical free energy)
+\end{itemize}
+\end{itemize}
+%%%%%%%%%%%%%%%%%%%%
+%                Software                    %
+%%%%%%%%%%%%%%%%%%%%
+\section{Available software -- a summary}
+\label{sec:software}
+\begin{itemize}
+\item Commercial:
+   \begin{itemize}
+    \item FEP+
+    \end{itemize}
+\item Free or low-cost for academics / commercial for industry:
+	\begin{itemize}
+	\item CHARMM / DOMDEC / CHARMM-OPENMM
+	\item TIES and AMBER FEW? (Peter Coveney)
+	\item AMBER / PMEMD
+	\end{itemize}
+\item Free (libre) open source:
+	\begin{itemize}
+	\item SIRE
+	\item YANK
+	\item gromacs 
+	\item pmx for mutations
+	\end{itemize}
+\item Setup tools
+	\begin{itemize}
+	\item FESetup: AMBER, gromacs, Sire
+	\item Lomap/Lomap2 : Relative alchemical transformation graph planning
+	\end{itemize}
+\item Analysis tools:
+	\begin{itemize}
+	\item Free Energy Workflows: Sire-specific free energy map analysis using weighted path averages
+	\url{https://github.com/michellab/freenrgworkflows}
+	\item Alchemlyb: Multipackage free energy analysis
+	\url{https://github.com/alchemistry/alchemlyb}
+	\item pymbar: MBAR implementation, but have to roll your own analysis wrapper
+	\url{https://github.com/choderalab/pymbar}
+	\end{itemize}
+\end{itemize}
+
+\section{Online resources}
+\begin{itemize}
+\item \url{http://www.ks.uiuc.edu/Training/Workshop/Urbana_2010A/lectures/TCBG-2010.pdf}
+\item Basic Ingredients of Free Energy Calculations: A Review (\url{DOI: 10.1002/jcc.21450})
+\item Good Practices in Free-Energy Calculations (\url{DOI: 10.1021/jp102971x})
+\item Alchemical Free Energy Methods for Drug Discovery: Progress and Challenges (\url{doi: 10.1016/j.sbi.2011.01.011})
+\item Alchemistry wiki: \url{http://www.alchemistry.org/wiki/Best_Practices}
+\end{itemize}
+
+\section*{Author Contributions}
 %%%%%%%%%%%%%%%%
@@
 
-\section{Other Contributions}
+\section*{Other Contributions}
 %%%%%%%%%%%%%%%
@@
 
-\section{Potentially Conflicting Interests}
+\section*{Potentially Conflicting Interests}
 %%%%%%%
@@
 
-\section{Funding Information}
+\section*{Funding Information}
 %%%%%%%

### 2019-05-09 paper/figures2/OverlapMatrices/plot_matrix_bad.py
added figure drafts

@@ -1,0 +1,56 @@
+import seaborn as sns
+import numpy as np
+import matplotlib.pyplot as plt 
+
+#A helper function to read the overlap matrix from file
+def get_overlap_matrix(filename, n_states):
+    fh = open (filename, 'r')
+    lines = fh.readlines()
+    fh.close()
+    count = 0
+    matrix = []
+    for line in lines:
+        if line.startswith('#Overlap'):
+            matrix = lines[(count+1):(count+1+n_states)]
+            break
+        count = count+1 
+    for i in range(len(matrix)):
+        temp = matrix[i].strip().split(' ')
+        float_temp = [float(j) for j in temp]
+        matrix[i] = float_temp
+    matrix =np.array(matrix)
+    return matrix
+
+# load datasets:
+overlap_good = get_overlap_matrix('11-lambda-overlap_good.dat', n_states=11)
+overlap_bad = get_overlap_matrix('11-lambda-overlap_bad.dat', n_states=11)
+
+
+
+#####Plotting the overlap matrices
+# seperately because seaborn subplotting is a bit cumbersome with labels etc
+sns.set_context("paper", font_scale=1.8) 
+
+plt.figure(figsize=(8,12))
+
+labels = list(map(str, list(np.arange(1, 12, 1))))
+
+sns.heatmap(overlap_bad, annot=True, 
+    fmt='.2f', 
+    linewidths=.5, 
+    annot_kws={"size": 12},
+    xticklabels=labels,
+    yticklabels=labels,
+
+    )
+plt.text(13.85, 4.25, "Degree of overlap",rotation=-90)
+plt.xlabel(r'$\lambda$ window')
+plt.ylabel(r'$\lambda$ window')
+plt.yticks(rotation=0)
+
+
+plt.savefig("overlap_matrix_bad.png" ,dpi=350)
+
+plt.show()
+
+

### 2019-11-29 paper/figures/fig2/softcore/plot3d_softcore.py
expanding softcore and figure2

@@ -1,0 +1,53 @@
+from mpl_toolkits.mplot3d import Axes3D
+
+import matplotlib.pyplot as plt
+from matplotlib import cm
+from matplotlib.ticker import LinearLocator, FormatStrFormatter
+import numpy as np
+
+colormap = cm.cividis # Not sure the best one to use.  We want more distincttion lower.  
+fig = plt.figure()
+ax = fig.gca(projection='3d')
+
+rmin = 0.1
+rmax = 9.0
+ngrid = 100
+alpha = 0.5
+sigma = 3.0
+eps = 1 
+zmax = 8
+
+r = np.linspace(rmin, rmax, ngrid)
+lam = np.linspace(0,1,ngrid)
+r,lam = np.meshgrid(r,lam)
+lamr = 1.0/(alpha*(1.0-lam) + (r/sigma)**6)
+u = 4*eps*lam*(lamr**2 - lamr)
+u[u>zmax] = zmax
+
+surf = ax.plot_surface(r, lam, u, cmap=cm.cividis,
+                       linewidth=0, antialiased=False)
+
+# Customize the z axis.
+ax.set_zlim(-eps, zmax)
+ax.zaxis.set_major_locator(LinearLocator(10))
+ax.zaxis.set_major_formatter(FormatStrFormatter('%.02f'))
+
+# Add a color bar which maps values to colors.
+fig.colorbar(surf, shrink=0.5, aspect=5)
+
+ax.set_title('The softcore interaction potential')
+ax.set_xlabel('Particle distance')
+ax.set_ylabel(r'$\lambda$')
+ax.set_zlabel('Energy')
+
+# add lines at lam = 0 and lam = 1 to show the physical endpoints
+rrange = np.linspace(rmin,rmax,ngrid) 
+ax.plot(rrange,np.zeros(ngrid),np.zeros(ngrid),color='r',linewidth=3)
+lj = 4*eps*((rrange/sigma)**(-12) - (rrange/sigma)**(-6))
+lj[lj>zmax] = np.nan
+ax.plot(rrange,np.ones(ngrid),lj,color='r',linewidth=3)
+
+ax.view_init(elev=20., azim=-65)
+plt.savefig("lj_softcore.pdf")
+
+#plt.show()

### 2019-12-09 paper/figures/fig9_equib_detection/equilibration.py
updating accidently deleted figure

@@ -1,0 +1,351 @@
+#!/usr/bin/env python
+
+"""
+Utilities for automatically detecting equilibrated region of molecular simulations.
+
+John D. Chodera <john.chodera@choderalab.org>
+Sloan Kettering Institute
+Memorial Sloan Kettering Cancer Center
+
+LICENSE
+
+This is free software: you can redistribute it and/or modify
+it under the terms of the GNU Lesser General Public License as
+published by the Free Software Foundation, either version 2.1
+of the License, or (at your option) any later version.
+
+This library is distributed in the hope that it will be useful,
+but WITHOUT ANY WARRANTY; without even the implied warranty of
+MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+GNU Lesser General Public License for more details.
+
+You should have received a copy of the GNU Lesser General Public
+License along with this software. If not, see <http://www.gnu.org/licenses/>.
+
+"""
+
+import numpy as np
+
+def statisticalInefficiency_multiscale(A_n, B_n=None, fast=True, mintime=3):
+    """
+    Compute the (cross) statistical inefficiency of (two) timeseries using multiscale method from Chodera.
+
+    Parameters
+    ----------
+    A_n : np.ndarray, float
+        A_n[n] is nth value of timeseries A.  Length is deduced from vector.
+    B_n : np.ndarray, float, optional, default=None
+        B_n[n] is nth value of timeseries B.  Length is deduced from vector.
+        If supplied, the cross-correlation of timeseries A and B will be estimated instead of the
+        autocorrelation of timeseries A.
+    fast : bool, optional, default=False
+        f True, will use faster (but less accurate) method to estimate correlation
+        time, described in Ref. [1] (default: False).
+    mintime : int, optional, default=3
+        minimum amount of correlation function to compute (default: 3)
+        The algorithm terminates after computing the correlation time out to mintime when the
+        correlation function first goes negative.  Note that this time may need to be increased
+        if there is a strong initial negative peak in the correlation function.
+
+    Returns
+    -------
+    g : np.ndarray,
[…]

### 2019-12-12 paper/figures/AnalysisPractices/gen_plots.py
added good practice in plotting fig

@@ -1,0 +1,140 @@
+#!/usr/bin/env python
+# -*- coding: utf-8 -*-
+
+"""
+Script for generating weakly and strongly correlating example plots for 
+explaining good practices in data representation in AFE calculations.
+"""
+
+import matplotlib.pyplot as plt 
+import seaborn as sns 
+import numpy as np 
+import pandas as pd 
+import scipy
+from sklearn.metrics import mean_absolute_error
+
+###############################################
+def interpolate_data(data):
+	"""
+	1-dimensional linear interpolation to set correlated data to range -15<->-5:
+	"""
+
+	return np.interp(data, (data.min(), data.max()), (-12, -8))
+
+
+# generate toy datasets:
+num_points = 50
+
+xx = np.array([-0.51, 51.2])
+yy = np.array([0.33, 51.6])
+means = [xx.mean(), yy.mean()]  
+stds = [xx.std() / 3, yy.std() / 3]
+corr = 0.85
+covs = [
+		[stds[0]**2, stds[0]*stds[1]*corr], 
+        [stds[0]*stds[1]*corr, stds[1]**2]
+        ] 
+m = np.random.multivariate_normal(means, covs, num_points).T
+
+# rescale data to a realistic range, also add two outliers:
+x_data_raw = interpolate_data(m[0])
+x_data = np.append(x_data_raw, [-11, -12])
+
+y_data_raw = interpolate_data(m[1])
+y_data = np.append(y_data_raw, [-8.3, -9])
+
+###############################################
+# figure plotting:
+
+sns.set(font='sans-serif', style="ticks")
+sns.scatterplot(x_data, y_data, linewidth=0)
+
+# plot kcal bounds:
+
+plt.fill_between(
+				x=[-17, -3], 
+				y2=[-16.5,-2.5],
+				y1=[-17.5, -3.5],
+				lw=0, 
+				zorder=-10,
+				alpha=0.5,
+				color="darkorange")
+# upper bound:
+plt.fill_between(
+				x=[-17, -3], 
+				y2=[-16,-2],
+				y1=[-16.5, -2.5],
+				lw=0, 
+				zorder=-10,
+				color="darkorange", 
+				alpha=0.2)
+# lower bound:
+plt.fill_between(
+				x=[-17, -3], 
+				y2=[-17.5,-3.5],
[…]

### 2019-12-12 paper/figures/AnalysisPractices/gen_plots.py
updated fig

@@ -47,6 +47,9 @@ y_data = np.append(y_data_raw, [-8.3, -9])
 # figure plotting:
 
 sns.set(font='sans-serif', style="ticks")
+plt.tick_params(labelsize=18)
+
+
 sns.scatterplot(x_data, y_data, linewidth=0)
 
 # plot kcal bounds:
@@ -125,14 +128,14 @@ plt.table(cellText=df.values, colLabels=df.columns,
 sns.despine()
 plt.xlim(-12.9, -7.1)
 plt.ylim(plt.xlim())
-plt.xlabel(r"Experimental $\Delta$G [kcal$\cdot$mol$^{-1}$]")
-plt.ylabel(r"AFE-predicted $\Delta$G [kcal$\cdot$mol$^{-1}$]")
+plt.xlabel(r"Experimental $\Delta$G [kcal$\cdot$mol$^{-1}$]", fontsize=18)
+plt.ylabel(r"AFE-predicted $\Delta$G [kcal$\cdot$mol$^{-1}$]", fontsize=18)
 plt.tight_layout()
 
+plt.savefig("Figure.pdf", dpi=300)
 
-plt.savefig("data_depictions.png", dpi=300)
-plt.show()
 
+#plt.show()
 
 
 

### 2020-01-08 paper/figures/fig13_analysis_practices/gen_plots.py
updated plot

@@ -135,7 +135,7 @@ plt.tight_layout()
 plt.savefig("Figure.pdf", dpi=300)
 
 
-#plt.show()
+plt.show()
 
 
 

### 2020-01-08 paper/figures/fig13_analysis_practices/gen_plots.py
adjusted script to output png

@@ -132,8 +132,11 @@ plt.xlabel(r"Experimental $\Delta$G [kcal$\cdot$mol$^{-1}$]", fontsize=18)
 plt.ylabel(r"AFE-predicted $\Delta$G [kcal$\cdot$mol$^{-1}$]", fontsize=18)
 plt.tight_layout()
 
-plt.savefig("Figure.pdf", dpi=300)
 
+# we need the image as PDF but seaborn has an issue with negative axis ticks in PDF.
+# instead, save as PNG and convert to PDF manually.
+# (see https://github.com/mwaskom/seaborn/issues/1107)
+plt.savefig("Figure.png", dpi=300)
 
 plt.show()
 

### 2020-01-10 paper/manuscript.tex
refelct repository name change

--- 2017-11-30 paper/manuscript.tex
@@
 \usepackage{siunitx}
+\usepackage{url}
 \DeclareSIUnit\Molar{M}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\usepackage[colorinlistoftodos]{todonotes}
+\usepackage[most]{tcolorbox}
+\usepackage{enumitem,amssymb}
+\usepackage{textgreek}
+\usepackage{changepage}
 
 \newcommand{\versionnumber}{0.1}  % you should update the minor version number in preprints and major version number of submissions.
-\newcommand{\githubrepository}{\url{https://github.com/michellab/best_practice_afec}  %this should be the main github repository for this article
-
+\newcommand{\githubrepository}{\url{https://github.com/michellab/alchemical-best-practices}}  %this should be the main github repository for this article
+\newcommand{\expect}[1]{\left\langle{#1}\right\rangle}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{Best Practices for Alchemical Free Energy Calculations : v\versionnumber}
-
-\author[1*]{Firstname Middlename Surname}
-\author[1,2\authfn{1}\authfn{3}]{Firstname Middlename Familyname}
-\author[2\authfn{1}\authfn{4}]{Firstname Initials Surname}
-\author[2*]{Firstname Surname}
-\affil[1]{Institution 1}
-\affil[2]{Institution 2}
-
-\corr{email1@example.com}{FMS}  % Correspondence emails.  FMS and FS are the appropriate authors initials.
-\corr{email2@example.com}{FS}
-
-\contrib[\authfn{1}]{These authors contributed equally to this work}
-\contrib[\authfn{2}]{These authors also contributed equally to this work}
-
-\presentadd[\authfn{3}]{Department, Institute, Country}
-\presentadd[\authfn{4}]{Department, Institute, Country}
-
+\title{Best Practices for Alchemical Free Energy Calculations: v\versionnumber}
+\author[1*]{Antonia S. J. S. Mey}
+\author[7]{Bryce K. Allen}
+\author[2*]{John D. Chodera}
+\author[1,10]{Maximilian Kuhn}
+\author[1*]{Julien Michel}
+\author[3*]{David L. Mobley}
+\author[1]{Levi N. Naden}
+\author[4]{Samarjeet Prasad}
+\author[5]{Julia E. Rice}
+\author[2,8]{Andrea Rizzi}
+\author[1]{Jenke Scheen}
+\author[6*]{Michael R. Shirts}
+\author[9]{Gary Tresadern}
+\author[7]{Huafeng Xu}
+%
+\affil[1]{EaStCHEM School of Chemistry, David Brewster Road, Joseph Black Building, The King's Buildings, Edinburgh, EH9 3FJ, UK}
+\affil[2]{Computational and Systems Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York NY, USA}
+\affil[3]{Departments of Pharmaceutical Sciences and Chemistry, University of California, Irvine, USA}
+\affil[4]{National Institutes of Health, Bethesda, MD, USA}
+\affil[5]{IBM Alamden Research Center, Almaden, CA, USA}
+\affil[6]{University of Colorado Boulder, Boulder, CO, USA}
+\affil[7]{Silicon Therapeutics, Boston, MA, USA}
+\affil[8]{Tri-Institutional Training Program in Computational Biology and Medicine, New York, NY, USA}
+\affil[9]{Computational Chemistry, Janssen Research \& Development, Turnhoutseweg 30, Beerse B-2340,Belgium}
+\affil[10]{Cresset, Cambridgeshire, UK}
+%
+\corr{john.chodera@choderalab.org}{JDC}
+\corr{dmobley@mobleylab.org}{DLM}
+\corr{antonia.mey@ed.ac.uk}{ASJSM}
+\corr{mail@julienmichel.net}{JM}
+\corr{michael.shirts@colorado.edu}{MRS}
+%
 \blurb{This LiveCoMS document is maintained online on GitHub at \githubrepository; to provide feedback, suggestions, or help improve it, please visit the GitHub repository and participate via the issue tracker.}
-
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%% PUBLICATION INFORMATION
+%%% Fill out these parameters when available
+%%% These are used when the "pubversion" option is invoked
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\pubDOI{10.XXXX/YYYYYYY}
+\pubvolume{<volume>}
+\pubyear{<year>}
+\articlenum{<number>}
+\datereceived{Month, Day, Year}
+\dateaccepted{Month, Day, Year}
+%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \begin{document}
-
+%
 \begin{frontmatter}
 \maketitle
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%% Abstract
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{abstract}
+Alchemical free energy calculations are a useful tool for predicting free energy differences associated with the transfer of small molecules from one environment to another.
+The hallmark of these methods is the use of "bridging" potential energy functions representing \emph{alchemical} intermediate states that do not exist in as real chemical species.
+The data collected from these bridging alchemical thermodynamic states allows the computation of transfer free energies (or differences in transfer free energies) with orders of magnitude less simulation time than simulating the transfer process directly.
+%
+While these methods are highly flexible, care must be taken in avoiding common pitfalls to ensure that computed free energy differences can be robust and reproducible for the chosen force field, and that appropriate corrections are included to permit direct comparison with experimental data.
+%
+In this paper, we review current best practices for several popular application domains of alchemical free energy calculations, including relative and absolute small molecule binding free energy calculations to biomolecular targets.
+\todo[inline, color=green!20]{JDC: Should we migrate this document to \url{https://github.com/alchemistry}? ASJSM: Let's migrate with the submitted version, but leave it here for now?}
+\end{abstract}
+\end{frontmatter}
+%
+\todototoc
+\listoftodos
+%
+%%%%%%%%%%%%%%%%%%%%
+%  Introduction    %
+%%%%%%%%%%%%%%%%%%%%
+\todo[inline]{}The current Overleaf document is \url{https://www.overleaf.com/5136551543vgjqbkzpmqkp}
+\section{What are alchemical free energy methods?}
+\label{sec:intro}
+\todo[inline, color={green!20}]{@everyone: Any further input in this?}
+Alchemical free energy (AFE) calculations compute free energy differences associated with transfer processes, such as the binding of a small molecule to a receptor or the transfer of a small molecule from an aqueous to apolar phase~\cite{zwanzig1954hightemperature}.
+The calculations use non-physical intermediate states where atoms are changed from one type to another or can be removed or inserted by turning on or off the interactions with their environment. 
+%
+Fig.~\ref{fig:fig_what_is_alchemy} illustrates common free energy changes that may be difficult to compute with conventional computational methods, but are more readily tractable with AFE methods.
+In these types of simulations so-called \textit{alchemical states} permit converged calculations for large changes, such as the entire removal of a ligand from a binding site, as seen in fig.~\ref{fig:fig_what_is_alchemy} (E).
+This type of simulation is known as an absolute AFE calculation.
+Other typical problems tackled with AFE calculations can be as simple as conformational changes of a molecule that exhibit a high barrier between these (fig.~\ref{fig:fig_what_is_alchemy} (A)); computation of partition coefficients, such as $\log P$ or $\log D$ (fig.~\ref{fig:fig_what_is_alchemy} (B))~\cite{rustenburg2016measuring, bosisio2016blinded}; and membrane insertions (fig.~\ref{fig:fig_what_is_alchemy} (C))~\cite{}.
+Then there are common types of AFE that estimate free energies of binding: the ligand stays the same, but a protein residue is alchemically mutated and the effect on the binding affinity is probed (fig.~\ref{fig:fig_what_is_alchemy} (D))\cite{hauser2018predicting,aldeghi2018accurate}, changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}, absolute binding free energies (fig.~\ref{fig:fig_what_is_alchemy} (E))~\cite{}, and relative free energies of binding where the difference of binding free energy between to similar ligands is evaluated (fig.~\ref{fig:fig_what_is_alchemy} (F))~\cite{}.
+%
+In order to obtain a free energy from these AFE simulations, historically two common approaches were used for the analysis: free energy perturbation (FEP) using the Zwanzig relation~\cite{zwanzig1954hightemperature} or Bennett's acceptance ratio~\cite{bennett1976efficient} and thermodynamic integration (TI) for which the theory dates back decades but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamics, merz1989free}.
+Important studies continued during the 2000's often leading to improved implementation of the methods in common simulation software~\cite{vanderspoel2005gromacs, mermelstein2018fast, wang2015accurate, hedges2019biosimspace}.
+This foundational work, combined with the methodological, technological, and hardware improvements of the last 5-10 years have led to an explosion of interest and direct commercial application of the technologies~\cite{wang2015accurate, fratev2019improved, schindler2020largescale, cournia2017relative, sherborne2016collaborating}.
+%
+As the field of molecular simulation moves towards millisecond timescales, accurate AFE calculations on even more challenging problems will become reasonable to perform.
+In the meantime, today's users may find it hard to get started with these complex calculations whilst also keeping up with the fast pace of change.
+This best practice guide aims to serve as a go-to reference for current recommendations and tips for users of all experience.
+%
+\begin{figure}
+    \includegraphics[width=0.95\linewidth]{figures/fig1_what_is_alchemy/Figure.pdf}   
+    \caption{\textbf{Illustration of common types of free energies differences that can be calculated using alchemical free energy methods.} A: Change in free energy due to a conformational change of the molecule.
+B: Partition coefficient such as $\log P$ or $\log D$ can be computed using alchemical free energy methods.
+Here, as an example the partition coefficient between methanol and water is shown.
+C: Free energy difference associated with the insertion of a molecule into a membrane.
+D: Effect of mutations of protein or host residues on free energies of binding.
+E: Absolute free energies of binding of a small molecule to a host (e.g.
+protein), F: Relative free energy of binding of one molecule with respect to another, here toluene and benzyl alcohol.
+    \label{fig:fig_what_is_alchemy}
+    }
+\end{figure}
+%
+%%%%%%%%%%%%%%%%%%%%
+% Prerequesites    %
+%%%%%%%%%%%%%%%%%%%%
+\section{Prerequisites and Scope}
+\label{sec:pre}
+\todo[inline, color=green!20]{ASJSM: @everyone This section is up for review to also check flow with intro and rest of document.}
+This guide is meant to serve as a comprehensive starting point for new practitioners for free energy calculations.
+It assumes moderate experience with molecular simulation packages and concepts, however.
+It is also intended to serve as a reference guide with a convenient checklist (Section~\ref{sec:checklist}) for more seasoned practitioners, to help ensure their simulation meet standard alchemical simulation and analysis practices.
+It can also serve as a set of best practices to ensure simulation robustness and reproducibility which reviewers may wish to consider as they evaluate papers.
+%
+This guide aims to answer the following questions:
+\begin{itemize}
+    \item Is my problem suitable for an alchemical free energy calculation? 
+    \item How do I choose the right free energy protocol and run it? 
+    \item How do I analyze my simulations accurately? 
+    \item What software tools are available to perform alchemical free energy calculations? 
+\end{itemize}
+%
+We assume the reader possesses a basic familiarity with the principles of molecular mechanics, molecular dynamics simulations, statistical mechanics, and the biophysics of protein-ligand association.
+If you feel unfamiliar with some of these concepts please good starting points can be found in these references~\cite{braun2019best, grossfield2018best, klimovich2015guidelines, shirts2012best}.
+Oftentimes, docking calculations are required to generated initial binding pose for a small molecule for an AFE calculation, where binding poses are not known, thus some basic familiarty on how to run docking calculations is also expected~\cite{grinter2014challenges}.
+%
+As some of the theoretical background can seem daunting, we provide an essentials guide to the theory behind alchemical free energy calculations in Section~\ref{sec:theory}.
+In the remainder of this paper, we will cover topics that are key to the preparation~(\ref{sec:step1}), choice and use of correct protocols~(\ref{sec:step2}), and finally the best practices that should be used for the analysis of AFE calculations~(\ref{sec:step3}).
+Particular focus will be given to aspects of the molecular simulations which are unique to alchemical calculations---these include the calculation of transfer free energies (hydration free energies, partition coefficients, etc.), and binding free energies (absolute and relative).
+%
+While we try to address as many methods and practices as possible, the field of free energy calculations is broad, and there are many advanced topics that are left to future best practices documents focusing on specific issues.
+Below, we provide a non-exhaustive list of topics we have not addressed with some references to provide starting points on these more advanced topics:
+\begin{itemize}
+\item Covalent inhibition.
+\item Free energies of mutation of protein side chains
+\item nonspecific binding or multiple binding sites.
+\item Complex association stoichiometry (beyond simple 1:1 association)~\cite{awesome reference}.
+\item Endpoint free energy methods such as MM-PBSA~\cite{genheden2015mm} and LIE~\cite{gutierrez-de-teran2012linear}
+\item Free energy methods that extract the ligand using geometric order parameters and potential of mean force methods~\cite{heinzelmann2017attachpullrelease}
+\item Forcefield dependence for protein, ligand, ions, co-solvents, and co-factors.
+Different studies have looked at the influence of force fields and it is assumed the user has made an adequate choice for the system under study.~\cite{loeffler2018reproducibility, vassetti2019assessment, lopes2015current}
+\end{itemize}
+%
+For convenience we have also compiled a list of common acronyms, common symbols used, and some essential terminology.
+\begin{tcolorbox}[title=Acronyms, colback=blue!10!white]
+    \textit{CPU} - Central Processing Unit\\
+    \textit{AFE} - Alchemical Free Energy\\
+    \textit{BAR} - Bennett Acceptance Ratio\\
+    \textit{FEP} - Free Energy Perturbation\\
+    \textit{GPCR} - G-Protein Coupled Receptor\\
+    \textit{GPU} - Graphics Processing Unit\\
+    \textit{MBAR} - Multistate Bennett Acceptance Ratio\\
+    \textit{MCSS} - Maximum Common Substructure\\
+    \textit{MD} - Molecular Dynamics\\
+    \textit{RBFE} - Relative Binding Free Energy\\
+    \textit{RMSE} - Root Mean Square Error\\
+    \textit{SAR} - Structure-Activity Relationships\\
+    \textit{TI} - Thermodynamic Integration
+\end{tcolorbox}
+%\label{sec:abbreviations}
+\begin{tcolorbox}[title=List of Symbols, colback=green!10!white]
+U - Potential energy\\
+\end{tcolorbox}
+\begin{tcolorbox}[title=Common terminology, colback=green!10!white]
+    ligand - small molecular binding to a larger host molecule such as a protein.\\ 
+    lead optimization - process of designing a drug candidate from an initial lead compound, considering its affinity and various pharmacokinetic parameters.\\
+    thermodynamic cycle - sequence of thermodynamic processes (like changes in temperature, volume or pressure) in which the initial and final state are equal.\\
+\end{tcolorbox}
+%
+%%%%%%%%%%%%%%%%%%%%
+% Theory basics    %
+%%%%%%%%%%%%%%%%%%%%
+\section{Statistical mechanics can tell us why alchemical free energy calculations work}
+\label{sec:theory}
+\todo[inline, color=green!20]{ASJSM: This section needs to be expanded and give a good introduction to the theoretical background without going into too much detail.}
+Why would you want to run an alchemical free energy calculation and why do they work?
+Suppose you want to compute the binding affinity, or free energy of binding, of a ligand $L$ to a protein $P$, given by:
+\begin{equation}
+P+L\leftrightharpoons PL
+\end{equation}
+The equilibrium rate of binding is given by the ratio of concentrations of product $[PL]$ and reactants $[P]$, $[L]$:
+\begin{equation}
+ K_b^{\circ} = c^{\circ}\frac{[PL]}{[L][P]}.
+\end{equation}
+In order to define a reference state, the standard state is usually defined as $1 \mathrm{mol}/\mathrm{L}$ and $c^\circ$ is the standard state concentration. Thus, the binding free energy is given by:
+\begin{equation}
+    \Delta G_{\mathrm{bind}} = -RT\ln K_b^{\circ}.
+\end{equation}
+One way to estimate the binding rate is by computing the probability of finding the molecular system in the bound state or the unbound state, making use of partition function ratios, i.e.
 
-\begin{abstract}
-This particular document provides a skeleton illustrating key sections for a Best Practices document.
-Please see the sample \texttt{sample-document.tex} in \url{github.com/livecomsjournal/article_templates/templates} for additional information on and examples of using the LiveCoMS LaTeX class.
-Here we also assume familiarity with LaTeX and knowledge of how to include figures, tables, etc.; if you want examples, see the sample just referenced.
+\begin{equation}
+    \Delta G_{\mathrm{bind}} = -RT\ln\frac{\int_\mathrm{bound} \exp(-\beta U(\mathbf{x}))d\mathbf{x}}{\int_\mathrm{unbound} \exp(-\beta U(\mathbf{x}))d\mathbf{x}}
+\end{equation}
+Or more generally in a simplified form the free energy difference between the bound and unbound states is:
+\begin{equation}
+\Delta f \equiv f(\mathrm{bound}) - f(\mathrm{unbound}) = -\ln\frac{Z(\mathrm{unbound})}{Z({\mathrm{bound}})},
+\end{equation}
+where $Z = \int_\mathrm{bound} \exp(-\beta U(\mathbf{x}))d\mathbf{x}$ represents the configurational integral. 
+How can we compute $\Delta G_{\mathrm{bind}}$ from a molecular dynamics simulation?
+Running a direct simulation of the binding events is too computationally costly, meaning we cannot easily estimate the configurational integrals of the bound and unbound state from the simulation.
+This is where alchemical free simulations come in.
+We can write down a thermodynamic cycle as seen in ~\ref{fig:fig_binding_thermodynamic_cycle}, for a relative free energy simulation.
+Based on this cycle we can compute the relative free energy of binding between two compounds $A$ and $B$, by going along the arrow of $\Delta G_{\mathrm{bound}}$, where we introduce a second compound that is similar to the first one.
+Now $\Delta G_{\mathrm{bound}}$ can be expressed as a ratio of partition functions of toluene (A) and benzyl alcohol (B).
+The defining characteristic of alchemical free energy calculations is the use of a series of alchemically-modified potential functions $U(x; \lambda)$ in which an alchemical parameter $\lambda$ modulates interactions in a manner that cannot occur in real chemical systems.
+One or more simulations are used to collect data from a multitude of alchemical states to compute a free energy difference between a chemical state ($\lambda_0$) and another chemical or alchemical reference state ($\lambda_1$),
+\begin{eqnarray}
+\Delta f &\equiv& f(\lambda_1) - f(\lambda_0) = - \ln \frac{Z(\lambda_1)}{Z(\lambda_0)} , \label{equation:dimensionless-free-energy-difference}
+\end{eqnarray}
+where the dimensionless free energy $f(\lambda) \equiv \beta F(\lambda)$ is given in terms of partition functions $Z(\lambda)$,
+\begin{eqnarray}
+Z(\lambda) &=& \int dx \, e^{-u(x; \lambda)} .
+\label{equation:partition-function-definition}
+\end{eqnarray}
+Here, the inverse thermal energy $\beta \equiv (k_B T)^{-1}$ where $k_B$ is the Boltzmann constant and $T$ is the absolute temperature, and the \emph{reduced potential} $u(x; \lambda)$~\cite{shirts2008statistically} is generally given by a trace over thermodynamic parameters with their conjugate dynamical variables,
+\begin{eqnarray}
+u(x;\lambda) &\equiv& \beta \left[ U(x;\lambda) + p \, V(x) + \sum_{i=1}^N \mu_i \, N_i(x) + \cdots \right] . \label{equation:reduced-potential}
+\end{eqnarray}
+Here, the collection of thermodynamic and alchemical parameters $\theta \equiv \{\beta, \lambda, p, \mu, \ldots\}$ defines a \emph{thermodynamic state}.
+The physical transformation of interest---such as a protein-ligand association process, or transfer free energy among phases---may involve a thermodynamic cycle that requires several free energy differences $\Delta f$ to be computed in order to produce an estimate of the overall desired quantity.
 
-In your work, in this particular slot, please provide an abstract of no more than 250 words.
-Your abstract should explain the main contributions of your article, and should not contain any material that is not included in the main text.
-Please note that your abstract, plus the authorship material following it, must not extend beyond the title page or modifications to the LaTeX class will likely be needed.
-\end{abstract}
+\begin{figure}
+    \includegraphics[width=0.95\linewidth]{figures/fig2_therm_cyc/Figure.pdf}
+    \caption{{\bf Thermodynamic cycle for computing the relative free energy of binding ($\Delta \Delta G$) between two related small molecules to a supramolecular host.}
+    The relative binding free energy difference between two small molecules, $\Delta \Delta G_{\mathrm{bind}, A \rightarrow B} \equiv \Delta G_{\mathrm{bind}, B} - \Delta G_{\mathrm{bind}, A}$---here benzyl alcohol (top) to toluene (bottom)---can be computed as a difference between two alchemical transformations, $\Delta G_\mathrm{bound} - \Delta G_\mathrm{solvated}$, where $\Delta G_\mathrm{bound}$ represents the free energy change of transforming $A \rightarrow B$ in complex and $\Delta G_\mathrm{solvated}$ the free energy change of transforming $A \rightarrow B$ in solvent.}
+    \label{fig:fig_binding_thermodynamic_cycle}
+\end{figure}
 
-\end{frontmatter}
+%%%%%%%%%%%%
+% Step 0   %
+%%%%%%%%%%%%
+\section{What can be expected from alchemical simulations?}
+\label{sec:step0}
+\todo[inline, color={green!20}]{ASJSM: @everyone please review this section}
+When starting an alchemical free energy project you need to determine if an alchemical free energy calculation is the appropriate strategy by initially counting the cost of your
+project: Can you even hope to tackle the problem with available resources and, if successful, will it
+be worth it?
+\subsection{How accurate are alchemical free energy calculations?}
+\label{subsec:accuracy}
+%
+Current alchemical free energy calculations seem to achieve, in favorable cases, RMS errors around 1-2 kcal/mol depending on force field, system, and a variety of other factors such as simulation time, sampling method, and whether the calculations employed are absolute or relative.
+A small selection of benchmark datasets and case studies can be found in section~\ref{sec:benchmark} at the end of this document.
+However, the domain of applicability is a significant concern~\cite{sherborne2016collaborating, cournia2017relative}, especially for relative calculations, which typically require a bound structure of a closely
+related ligand as a starting point.
+Additional factors such as slow protein or ligand rearrangements, uncertainties in ligand binding mode, or charged ligands can make these calculations far less reliable and more of a research effort.
+%
+It is worth noting that the accuracy of free energy calculations is highly variable across different protein targets, and likely across different ligand chemotypes as well.
+For instance, FEP+ with OPLS3 achieves an RMSE of 0.62 kcal/mol for a set of 21 compounds binding to JNK1 kinase, yet only an RMSE of 1.05 kcal/mol for a set of 34 compounds binding to P38$\alpha$ kinase~\cite{harder2016opls3}.
+Furthermore, perturbations for the same chemotype in different pockets of the BACE enzyme gave varied errors~\cite{keranen2017acylguanidine}
+\todo[inline,color={green!20}]{JDC: It's important to be clear on what error is being reported: $\Delta G$ after shifting by a constant to minimize the RMSE, unshifted $\Delta G$, $\Delta \Delta G$ of computed edges, or $\Delta \Delta G$ of all edges.
+Note that computed edge $\Delta \Delta G$ RMSE significantly underestimates the error of all edge RMSE: \url{https://github.com/jchodera/jacs-dataset-analysis} ASJSM: Can we incorporate your analysis in this section as an example?
+Maybe this is something that needs to be addressed properly in the analysis section and only referenced here?}
+We thus recommend that retrospective study for a particular target and a particular chemical series be performed to establish the relevant accuracy of free energy calculations for each particular application case.
+%
+\subsection{How reproducible are alchemical free energy calculations?}
+\label{subsec:reproducible}
+Finite computing resources necessarily limit the generated number of uncorrelated samples of potential energy surfaces, and therefore alchemical free energy calculations only give free energy estimates to within finite precision.
+An important consideration is how reproducible alchemical free energy calculations are in practice.
+In simple cases such as absolute hydration free energies of small organic molecules, or relative hydration free energy calculations between structurally similar small organic molecules, it should be possible to obtain highly precise estimates with a given software package (with a sample standard deviation under 0.01 kcal/mol).
+For more complex use cases such as protein-ligand binding free energies the repeatability is often substantially worse.
+A good practice is to perform two or three runs of the same perturbation to assess repeatability with a given protocol.
+The sample standard deviation will give a crude estimate of the reliability of the estimates, and whether the precision is sufficient for the problem at hand.
+Where this is practical, a more stringent test is to use different input coordinates for each repeat run.
+%
+Note that these issues concern calculations carried out with a single software package, but simulation package variations can introduce additional issues.
+Such issues of reproducibility of free energy calculations across different simulation packages have attracted attention recently.
+Greater variability is expected due to methodological differences such as integrators, thermostats, barostats, treatment of long-range electrostatics, and potentially other factors.
+For absolute and relative hydration free energies of small organic molecules a variability of ca.
+0.2 kcal/mol between popular simulation packages has been reported.
+\cite{REF Reproducibility of hydration free energies Loefler et al.
+JCTC 2018} In the recent SAMPL6 SAMPLing challenge a larger variability of 0.3 to 1.0 kcal/mol was noted in the computed absolute binding free energies of host/guest systems even though similar input and simulation parameters were used \cite{Ref Rizzi et al SAMPLing biorxiv} and, in many cases, single-point energies were identical or nearly so.
+Further work is needed to ensure reproducibility of alchemical free energy calculations across implementations to guarantee that force-field development efforts lead to transferable potential energy functions.
+%
+\subsection{Is my problem suitable for alchemical free energy calculations?}
+\label{subsec:suitability}
+Before even planning free energy calculations to study binding to a
+particular target, it is important to assess what is known about the
+system and its timescales and its suitability for free energy
+calculations, as well as the \emph{purpose} of the calculations and
+the amount of available computer resources. In some cases, predicting
+accurate binding free energies for a particular target might be a
+\emph{more} challenging effort than simply measuring them! This is
+often the case when dealing with database screening problems, where
+compounds might be easily and quickly available commercially for
+testing and free energy calculations could consume a much larger
+amount of resources. Free energy calculations thus typically only
+appeal when (slow or costly) synthesis would be required to do
+experiments or experiments are otherwise cost-prohibitive.
+%
+Sometimes free energy calculations can provide answers that are not
+readily available from experiments.  For example, type II kinase
+inhibitors selectively bind to different kinases in the so-called
+DFG-out conformations (cite Kuriyan).  The selectivity of such
+inhibitors may be attributed either to their differential binding to
+different kinases in the DFG-out conformations, or to different
+stability of the DFG-out conformations of different kinases.  Let
+$K_C$ be the equilibrium constant between DFG-in and DFG-out
+conformations of one kinase, and $K_D^\ast$ be the dissociation
+constant of a type II inhibitor against this kinase, the apparent
+binding constant of this inhibitor against this kinase is then
+\begin{equation}
+  K_D = K_D^\ast \frac{1 + K_C}{K_C}
+  \label{eqn:conformational-binding}
+\end{equation}
+%
+Since binding experiments cannot resolve $K_D^\ast$ and $K_C$ individually, such experiments cannot address the basis of selectivity of the type II inhibitors.
+Absolute binding free energy calculations, in contrast, can take advantage of the slow kinetics of DFG-in/out conversion, and estimate the conformation-specific binding constant $K_D^\ast$, thus yielding clues as to the source of selectivity.
+%
+\subsection{Is the expected accuracy of the computation sufficient?}
+\label{subsec:accuracy}
+The requisite level of accuracy is another important consideration ---if the
+goal is to guide lead optimization when many compounds will be
+synthesized, free energy calculations can be appealing even with
+accuracy in the 1--2 kcal/mol range~\cite{mobley2012perspective}, but if the number of compounds to be synthesized is very small, this accuracy may not be enough to provide
+much value.
+%
+Here we include a simple estimate of the value provided by alchemical
+free energy calculations in lead optimization.  Let $P(\Delta\Delta
+G)$ be the probability distribution of the changes in the binding free
+energies of a new set of molecules during one round of lead
+optimization, and let $P(\Delta\Delta G^\dagger|\Delta\Delta G)$ be the
+conditional probability of the binding free energy change computed by
+the free energy calculations, $\Delta\Delta G^\dagger$, given the actual
+change $\Delta\Delta G$.  The latter conditional probability can be modeled
+by a normal distribution
+\begin{equation}
+  P(\Delta\Delta G^\dagger|\Delta\Delta G) = \frac{1}{\sqrt{2\pi\sigma^2}}
+  \exp\left(-\frac{(\Delta\Delta G^\dagger - \Delta\Delta G)^2}{2\sigma^2}\right),
+  \label{eqn:free-energy-distribution}
+\end{equation}
+where $\sigma$ signifies the accuracy of free energy calculations.
+Here we assume that there is no systematic bias in the free energy
+calculations, i.e., on average, the free energy change computed by
+free energy calculations agrees with the actual free energy change.
+%
+In lead optimization guided by free energy calculations, we only
+synthesize and experimentally test molecules that are predicted to
+have favorable free energy changes.  We are thus interested in how
+often that a molecule predicted to bind stronger actually turns out to
+bind stronger.  In other words, we are interested in the conditional
+probability:
+\begin{equation}
+  P(\Delta\Delta G<0|\Delta\Delta G^\dagger<0).
+  \label{eqn:true-positive}
+\end{equation}
+%
+For illustrative purposes, we assume that the actual changes in the
+binding free energies for a set of new molecules also follow normal
+distribution, that the standard deviation in the changes is $RT\ln 5$
+(corresponding to a 5-fold change in the binding affinities), and that
+1 in 10 new molecules have increased binding affinity ($\Delta\Delta G
+\leq 0$).  Under such assumptions, the conditional probability in
+Eq.~\ref{eqn:true-positive} can be easily computed.  If the accuracy
+of free energy calculations is $\sigma = 1$ kcal/mol, $P(\Delta\Delta
+G<0|\Delta\Delta G^\dagger<0) = 0.35$, which means that out of every
+10 molecules selected for predicted favorable free energy change, on
+average 3.5 molecules will have actual favorable free energy change.
+In other words, selection by free energy calculations yields 3.5 times
+more molecules of improved affinities than selection without free
+energy calculations.
+%  
+Available computational resources and timescales of motion also factor
+into this initial analysis. An individual free energy calculation
+involves simulations at many different intermediate states (perhaps
+20-40 or more) and each of these must typically be long enough to
+capture the relevant motions in the system. If such motions are
+microsecond events or longer, the computational cost of running 20-40
+microsecond or longer simulations for each of $N$ ligands may become
+prohibitive, or at least should be carefully considered. 
+%
+\subsection{Can I afford the calculation?}
+\label{subsec:affordability}
+Furthermore, are available computational resources sufficient that throughput will be
+reasonable compared to needs of experimental collaborators working on
+this system? How many ligands ($N$) can you afford to handle given
+your computational resources? As cloud computing becomes more available, in-house GPU clusters may not be necessary if calculations are not run on a regular basis.
+This analysis should be done up front, part of ``counting the cost''
+of involvement in a particular project. In some cases, the analysis may conclude that free energy calculations will not be feasible for the proposed problem.
+Here, by ``cost'', we refer not just to financial cost of the calculations relative to experiments, but also time -- can the calculations can be run faster than experiments are done?
+How will the relevant resource and opportunity costs factor in?
+Both computation and experiment require human time, supplies (of different sorts), and equipment.
+In the extreme limit, for example, it would not make sense to spend a month running a binding free energy calculation if the equivalent experiment could be done in a day with resources already on hand.
+Such issues should be considered before deciding to conduct binding free energy calculations.
+%
+\subsection{Is an exploratory study what I want?}
+\label{subsec:exploration}
+An additional consideration is how much is known about your particular
+target, ligand binding modes in the target, and any relevant motions
+-- essentially, has it been studied enough to know whether it might be
+suitable for free energy calculations?
+It is important to know if the system has hardly been studied, because should the initial calculations perform poorly, the effort may turn into an attempt to understand the relevant sampling, force field, or system preparation problems.
+%
+If you are unsure whether your project is feasible, one option may be
+to conduct a short exploratory study to assess tractability for just a very small
+number of ligands. This can be sufficient to get an initial
+idea of feasibility and accuracy of the calculations for the
+proposed target.
+%
+\section{How should alchemical simulations be applied to drug discovery?}
+\label{sec:drugdiscovery}
+\todo[inline, color={green!20}]{ASJSM: Merge/tidy of what can I expect from my simulation, drug discovery aspects and simulation prerequisites, there are some duplications there at the moment, happy for restructuring ideas.
+}
+%
+It is expected that AFE methods can provide a valuable tool for drug discovery, capable of outperforming alternative approaches~\cite{kuhn2017prospective}.
+It has been stressed that successful application in industry requires more knowledge of the "domain of applicability", in other words, a greater understanding of the success and failure~\cite{sherborne2016collaborating}, which in turn requires robust protocols for preparing, submitting and analysing AFE calculations.
+In this regard, the issues mentioned in the previous section such as understanding the suitability and timescales to capture the structure activity relationships (SAR), and performing up-front tests of performance are all relevant to drug discovery applications.
+Without venturing too far into details of system setup, which is beyond the scope of this article, there are some points to be stressed.
+%
+\subsection{Capturing the experimental conditions}
+The calculations aim to capture the alchemical change from one ligand to another as accurately as possible.
+Therefore, it is necessary to consider details of the experimental setup, such as pH.
+Biological assays are usually run at neutral pH but this is not always the case, and protein and ligand preparation protocols should reflect this.
+In more detail, the formal charge, or tautomeric state of the small molecules can change within a series of analogs.
+The change of pKa is a common medicinal chemistry strategy to modify drug properties and would normally require explicit effort to incorporate into alchemical calculations.
+Further concerns about the experimental setup require understanding what protein construct is used in the bioassay.
+For instance, does the X-ray that is to be used for the calculations match the construct used for screening (i.e.
+catalytic vs full length, monomer vs dimer etc) (CITE https://www.nature.com/articles/s41598-018-23039-5)?
+Also, were certain co-factors or partner proteins required in the bioassay?
+%
+\subsection{Is my binding mode accurate?}
+As also mentioned, good performance of alchemical FE calculations requires an accurate representation of the ligand binding mode, normally solved via X-ray crystallography.
+If more than one structure is available, the modeler should pay attention to choose the most suitable.
+The quality of the structure can be a concern, and the reader is referred to work of Warren et al.
+for a detailed discussion of choosing optimal structures for structure-based modeling~\cite{warren2012essential}.
+It is also useful to study the structure activity relationship and understand the expected impact in the binding site, is side chain movement required and is there evidence of this in any alternative X-ray structures of the same protein?
+Normally, only one protein and water configuration is used for a series of alchemical FE calculations, so this needs to be capable of accommodating the smallest through to largest ligands in a way that allows stable and well behaved simulations.
+This can be a practical limitation for the alchemical changes that are feasible, a simple work-around can be to separate compounds into sub-series for different calculations.
+If multiple structures are available there is some evidence the higher affinity complex can give better performance~\cite{perez-benito2019predicting}.
+%
+\subsection{Input setup and scale of calculations}
+In a drug discovery setting it is normal to consider dozens (or more) of ligands and it is necessary to align them in the binding site.
+There is no detailed study of how different alignment approaches may affect results, but the user should be aware of some practical considerations.
+Tools are available to compare the ligands and build the hybrid topologies that define the changes between one ligand and another~\cite{loeffler2015fesetup,hedges2019biosimspace,gapsys2015pmx}.
+In simple terms, providing poor alignment to these tools will make this job harder.
+Therefore, docking ligands with no restraints is not usually the best method for defining the ligand inputs, instead a core restrained docking can provide satisfactory outcomes, but in this case one still should pay careful attention to consistency of alignment for identical substituents.
+Another alternative is to manually edit the same core with the changing substituents.
+This provides assurances that coordinates for the non-perturbed structures remain identical and aromatic substituents for instance obey consistent dihedral angles but is not feasible for many compounds and therefore automation can be desirable.
+Finally, the role of water in ligand binding is now well understood and it can be crucial to capture the changes in binding site solvation during ligand binding.
+Can crystallographic waters be retained?
+Do they clash with some of the larger ligands used in the FEP?
+If so, how will the water be incorporated for smaller ligands that require the water – consider pre-solvation methods such as GCMC steps~\cite{michel2010prediction}.
+Other tools such as watermap or opensource equivalents (SSTmap, others) can be used to define water structure for systems with no experimental evidence of water sites.
+Before launching large numbers of alchemical free energy calculations it is recommended to test the system using classical MD simulations and limited numbers of alchemical perturbations.
+Metrics such as ligand and protein RMSD and RMSF can be inspected, along with visual inspection of simulations, to ensure the system is stable and likely to be suitable for FE calculations.
+%
+Running binding FE calculations in a drug discovery application will typically require the use of software or tools to facilitate the calculations of many compounds.
+The commercial implementation from Schr\"{o}dinger, FEP+, uses a graphical user interface and once the user has calculated any missing force field parameters, calculations can be prepared and submitted within minutes.
+The software builds the network of perturbations, connecting similar compounds based on LOMAP methodology~\cite{liu2013lead}, all inputs for the Desmond simulations are built automatically and the job can be saved or run immediately on available GPUs.
+The cost of this software is often prohibitive but alternative tools are becoming available to allow relatively straight forward running of alchemical FE calculations with other software~\cite{gapsys2015pmx, loeffler2015fesetup, song2019using, gapsys2020large,  jespers2019qligfep}.
+One consideration at this point is to define the perturbations suitable for the application.
+For instance, if the goal is to accurately assess the relative binding energy of a small number of compounds, possibly with challenging synthesis, the map of perturbations may contain as many connections between compounds as affordable.
+However, when running calculations on hundreds of compounds a so called ‘star-map’ can be used that just contains one connection per compound: perturbing every compound to a central ligand, typically the crystal structure ligand~\cite{konze2019reactionbased}.
+In this way the top-ranking examples can be readily identified and submitted to additional calculations in a second round.
+%
+\subsection{Making predictions, understanding errors}
+\label{subsec:predictions}
+For prospective drug discovery applications there are several other considerations including understanding likely errors and taking selection bias into account.
+It is crucial when proposing compounds for synthesis to have some idea of the underlying error or uncertainty in the predictions.
+A retrospective assessment can give an indication of prospective performance for similar molecules~\cite{ciordia2016application}.
+Beyond this, several parameters provide useful indicators of performance, results suggest MBAR errors for perturbations greater than 0.3 kcal/mol can be indicative of inaccuracy~\cite{perez-benito2019predicting}.
+In addition, the hysteresis can be checked by comparing the difference for forward and reverse perturbations of the same structural change, or in the case of large maps of perturbations cycle closure hysteresis~\cite{wang2013modeling} indicates problematic perturbations involved in cycles connecting many compounds.
+Selection bias, as described above, can be problematic in prospective applications as there is normally an aim to focus on only a narrow range of predicted activity (i.e.
+the most active).
+Thus, correlation between prediction and subsequent experiment may lead to disappointment but corrections can be applied based on previous recommendations~\cite{abel2017critical}.
+%
+In summary, the successful use of alchemical calculations in drug discovery requires working in the domain of applicability, using a high quality X-ray structure of the target compounds, and testing the approach retrospectively to ensure the system setup is well-behaved.
+Always assess the confidence in the predictions and communicate this when discussing with experimentalists.
+Consider performing repeat calculations for at least some of the perturbations in the study.
+In a drug discovery setting it is inevitable to consider larger virtual libraries, in these cases again evaluate if shorter simulation time per $\lambda$ window can be sufficient for a ‘virtual screening’ mode of application.
+It is expected that future tools may allow efficient exploration of large virtual libraries.
+There are many accounts of success of alchemical FE calculations, the methods show good performance towards the goal of binding energy prediction.
+However, it is important to have realistic expectations.
+Structure based drug design projects are often capable of improving potency relatively quickly and the range of activity narrows to just two-to-three log units.
+It may seem hard to impact with substantially different, more potent, stand-out compounds in this scenario, but binding energy prediction can still be extremely useful for ensuring activity is maintained as other properties are optimized.
+An interesting cost benefit analysis has shown the value of activity prediction, see discussion above and articles such as~\cite{mobley2012perspectiv}.
+From a drug discovery point of view, FEP is expanding its domain of applicability, there are reports of success using homology models and GPCRs for instance, as well as enabling charge change and scaffold hopping, but these systems are undoubtedly more difficult.
+In the meantime, the use cases are expanding to resistance prediction, disease mutation prediction, solubility prediction – an exciting future for alchemical FE calculations.
+%
+%
+%
+%%%%%%%%%%%%
+% Step 1   %
+%%%%%%%%%%%%
+%
+\section{Simulation prerequisites}
+\label{sec:step1}
+%
+The alchemical free energy protocols, that are discussed in Section~\ref{sec:step2}, are defined for a specific type of free energy calculation, i.e.
+a free energy of binding or a free energy of hydration.
+Depending on the type of simulations different considerations have to be made for ligands, solvent, and host molecules (in the case of the estimation of free energies of binding).
+\subsection{Free energies of binding}
+\label{subsec:binding}
+In principle in the limit of sufficient conformational sampling, the free energy changes estimated from an alchemical free energy calculation should be independent of the system's initial coordinates.
+However, in practice because simulations of thermodynamic states are of finite duration (typically currently 1-100 ns per state) this is only true for certain classes of alchemical free energy calculations such as relative or absolute free energies of hydration of small and relatively rigid organic molecules.
+Host-guest or protein-ligand complexes typically exhibit slowly relaxing degrees of freedom that exceed significantly the duration of an alchemical free energy calculation.
+It is therefore generally important to carefully select input coordinates to obtain satisfactory results.
+The following questions may be relevant before diving into the simulation setup.
+% 
+\begin{itemize}
+    \item Do I have one or multiple good host structures? (e.g. a good resolution X-ray crystal of the protein target)
+    \item Do I have information on one or all of the ligands binding sites (e.g. a X-ray structure)
+    \item Should I include buried waters, or other small molecules that can be found in an X-ray structure.
+    \item Are my ligands part of a congeneric series? (simple R group substitutions around the same scaffold)
+    \end{itemize}
+%
+\paragraph{Are there good X-ray structures available?}
+As for any simulation, care should be taken in selecting available X-ray structures in the protein databank and potentially choosing multiple structures initially, to account for variability in receptor conformations as well as accuracy of available X-ray structures.
+Typically, clustering of receptor structures can be used to identify different receptor conformations near the binding site, as well as assessing relevant side chain placements from the X-ray structure, see for example~\cite{mey2016blinded}.
+In terms of set up and other choices following general best practice guidelines is advisable~\cite{braun2018best}.
+%
+For binding free energy calculations using a congeneric set of ligands and therefore likely to be most suited to a relative free energy protocol (see~\ref{sec:step2}).
+Other than the choice of forcefield for the parameterization of the ligands, some care has to be taken selecting binding poses for these ligands.
+Generally a common assumption for a congeneric series is that the binding mode is conserved.
+Therefore, if an X-ray structure of one of the ligands is available, this should be used to position the ligands in the putative binding site in an energetically reasonable conformation that entails no steric or electrostatic mismatch with the receptor.
+Checking the X-ray structure versus the experimental electron densities is important, as the resolution of part of the ligand or important sidechains may be based on the interpretation of the crystallographer rather than the available electron density.
+For example, looking at a cyclohexane ring density, a chair conformation is vastly more likely than that of a boat and may just be poorly assigned.
+%
+\paragraph{How to deal with different binding modes?}
+Generally, binding modes within congeneric series are conserved~\cite{}, however, exceptions exist~\cite{}, and weaker binding compounds are more likely to adopt different binding modes (REFS).
+There may also be ambiguity about the relative orientation of substituents.
+A known issue involves a 180 degree flip in the dihedral angle of a 6-membered aromatic ring leading to a different spatial position of ortho- or meta- substituents that otherwise should overlap within a series.
+The 180 degree flip of the ring may not occur enough during simulations to overcome the starting configuration bias.
+Another scenario may be equatorial and axial substituted saturated rings (e.g.
+cyclohexane derivatives).
+This situation may be addressed by explicitly modelling different binding modes of the same ligand and combining later computed free energy differences for different binding modes into a relative free energies of binding~\cite{kaus2015how}.
+%
+\paragraph{Have you considered sterioisomers and enationmers?}
+Congeneric series can contain stereoisomers or enantiomers, making sure that you model the correct one is important.
+For racemates, the relative abundance of each stereoisomer is normally not known.
+Therefore, the experimental activity associated with just one conformation is more uncertain.
+However, the modeling typically uses just the bioactive conformation that best fits the active site.
+Clearly this introduces the chance for higher error versus experiment.
+Nevertheless, if all compounds in the congeneric series are racemic, originating from similar synthetic procedures with an expected similar abundance of stereoisomers, then the differences may cancel and the trend in calculated and observed binding energies may be robust.
+Nevertheless, we can see that care and further testing is needed in this scenario, and the quality of the predictions may suffer.
+%
+\paragraph{Conserved binding site waters can play an important role in binding free energies}
+Another aspect to look out for are binding site water molecules that may form water mediated protein-ligand interactions, particularly in situations where exchange with bulk water may be slow on the simulation timescales.
+This happens typically in buried binding site.
+Overlaying multiple protein X-ray structures can identify conserved or just additional water molecules that can be useful to include.
+In cases where water molecules are known to play an important role in the binding, software implementations that use water sampling facilitated by Grand Canonical Monte Carlo methods may be useful.
+For examples see HSP90 as formed part of the D3R grand challenge 2015~\cite{mey2016blinded},A2A~\cite{brucemacdonald2018ligand}, MUP~\cite{ross2015water}, and others~\cite{michel2009energetics}.
+%
+\paragraph{How to align congeneric series}
+Input coordinates for a congeneric series may be generated by docking calculations, or by ligand alignment using maximum common substructure algorithms.
+The latter tends to produce alignments that are more conserved and more consistent free energy changes across a dataset, but will struggle to yield reasonable results for relative binding free energy calculations that involve a significant binding mode rearrangement.
+This may also lead to steric clashes with the receptor coordinates of the reference ligand if structural rearrangements are needed to accommodate different members of the congeneric series.
+Small steric clashes may be resolved during subsequent simulation equilibration prior to data collection but there is a risk that the complex relaxes to an alternative metastable state.
+%
+An additional consideration arises for single topology relative free energy calculations.
+In this class of alchemical free energy calculations it is necessary to generate a molecular topology that may describe the initial and final states of the perturbation (see different topologies as seen in fig.~\ref{fig:fig_topology}).
+In cases where the end-states have high topological similarity and high structural overlap this is relatively straightforward and typically handled by use of maximum common substructure calculations.
+In situations where the end-state topologies differ significantly, or there is relatively little spatial overlap between the two end-states some user intervention may be necessary to produce a satisfactory input topology.
+%
+If the binding site location is uncertain but the structure of the receptor is well defined and plausible binding sites are identified, it may be more useful to chose an absolute free energy protocol to compute the standard free energy of binding of the ligand to a set of binding sites.
+This requires the user to prepare input files describing the bound conformation in different putative binding sites (REF).
+The apparent binding free energy of the ligand may be obtained by combining the individual binding site free energies, which also indicate where the ligand is more likely to bind.
+In this case a docking program can generate initial structures.
+Different commercial and none commercial tools are available, such as: rDock, Autodock Vina, Glide, or Flare to name a few~\cite{ruiz-carmona2014rdock, trott2010autodock, friesner2004glide, cheeseright2006molecular}.
+If the putative binding sites are not apparent, for instance due to significant induced-fit effects, it may be challenging to obtain meaningful free energies of binding.
+One may have to account for the free energy change for forming a binding site in the target receptor which may not be feasible in FE simulation timescales.
+Another key application for absolute binding free energies is to study chemically different ligands, that do not share a common scaffold, but bind at the same site or the same site of similar proteins~\cite{aldeghi2015accurate}.
+%
+\subsection{Free energies of hydration or partition coefficients}
+\label{subsec:hydration}
+Preliminary considerations necessary for using free energy methods to compute partition coefficients are generally more straight forward.
+For example, a 3D minimised structure of the solute can be generated with a simple tool such as babel and solvated to prepare the input  to compute a free energy of hydration.
+However, in these cases a careful choice of forcefield, as well as water models or organic solvents is essential.
+See for example~\cite{bosisio2016blinded,rustenburg2016measuring} for a good discussion on these choices.
+%
+%
+%
+%%%%%%%%%%%%
+% Step 1   %
+%%%%%%%%%%%%
+%
+\section{What simulation protocol should I choose? -- Step 2}
+\label{sec:step2}
+Alchemical free energy calculations can be grouped into two main categories, ``absolute'' (\ref{fig:fig_absolute_thermodynamic_cycle}) and ``relative'' \footnote{The distinction is a bit of a misnomer, since both compute ratios of partition functions relative to another state, and neither computes an absolute free energy.}(\ref{fig:fig_binding_thermodynamic_cycle}), which differ in whether they compute properties for a single molecule (absolute) or compare properties of different, usually closely related, molecules (relative).
+To use binding as a concrete example, in absolute binding free energy calculations, one computes the binding free energy of a ligand to an individual receptor relative to a standard reference concentration.
+In contrast, in relative binding free energy calculations, one compares the binding free energy of two related inhibitors to determine the potency difference.
+\subsection{Absolute and relative free energy calculations have some differences}
+Many of the issues of setup and protocol for alchemical calculations are common, but some are different between absolute and relative calculations, so before treating the common elements we consider the protocol differences.
 
-\section{Introduction}
+\subsubsection{Choices unique to relative free energy calculations}
+\label{sec:relative-fe-protocol}
 
-Here you would explain what problem you are tackling and briefly motivate your work.
+\paragraph{Topologies} A critical first step in relative calculations is to select an approach to these calculations, determining whether to use a \emph{dual topology}, \emph{single topology}, or \emph{hybrid topology} approach to relative calculations.
 
-In this particular template, we have removed most of the usage examples which occur in \texttt{sample-document.tex} to provide a minimal template you can modify; however, we retain a couple of examples illustrating more unusual features of our templates/article class, such as the checklists, and information on algorithms and pseudocode.
+The distinction between these can be illustrated by considering a hypothetical transformation from molecule A to molecule B, where both atoms share a common substructure but differ in their substituents; e.g.
+consider a transformation of benzene (A) to benzyl alcohol (B) fig.~\ref{fig:fig4_topology}.
+In this case the common substructure is the benzene ring, though perhaps may be larger depending on how it is defined, as we discuss below.
+In single topology calculations, the overall transformation is set up to involve as few additional atoms as possible, so benzene would be typically changed into benzyl alcohol by changing one of the hydrogens into a carbon.
+To this new carbon you also find non-interacting atoms called ``dummy atoms'' (retaining their bonded interactions but not interacting with the rest of the system), which will become the two hydrogens attached to the new carbon and the OH group.
+Bond parameters as well as partial charges are adjusted accordingly.
+Thus, in a single topology calculation, atoms may change their type so relatively few dummy atoms are created. This is illustrated in fig.~\ref{fig:fig4_topology} (a). 
+In contrast, in a dual topology free energy calculation, no atoms are allowed to change type~\cite{shirts2012best}.
+This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxy methyl group, then passing through an intermediate state where atoms which are becoming dummy atoms or ceasing being dummy atoms are partially interacting (this state may or may not be well defined~\cite{mobley2014blind}  and culminating in a state where benzyl alcohol is present along with the additional dummy atoms which was previously a corresponding hydrogen of the benzene.
+Fig.~\ref{fig:fig4_topology} (b) depicts how such a dual topology works.
+\begin{figure}[h!]
+    \includegraphics[width=0.95\columnwidth]{figures/fig3_topol/Figure.pdf}
+    \caption{\textbf{Two common topologies for AFE calculations}left: A single topology uses softcore potentials to convert from one type of atom to an other.
+Dummy atoms (Du) are used when there is no corresponding maximum common substructure match between the two molecules for certain atoms.
+right: The dual topology does not convert one species to another, but only converts between Du atoms and an interacting species, but usually uses softcore potentials for this.
+The 'mixed' intermediate atoms are used in both dual and single topology approaches.
+Only the way the transformation occurs and the end states differ.
+Following the arrow along the left and right illustrate the differences.
+Figure adapted from \url{http://www.alchemistry.org/wiki/Constructing_a_Pathway_of_Intermediate_States}}
+    \label{fig:fig_topology}
+\end{figure} 
+Hybrid topology calculations have seen much less use but essentially consist of two absolute free energy calculations in opposite directions at the same time (turning one molecule off while turning the other on)~\cite{jiang2019computing}.
+At present, the most widely used approaches, such as in Schr\"{o}dinger's FEP+~\cite{wang2019protein} and in FESetup~\cite{loeffler2015fesetup} (for which calculations may be planned with Lead Optimization Mapper (LOMAP)~\cite{liu2013lead} seem to use single topology approaches, though some codes only support dual topology.
+To our knowledge efficiency differences have not been thoroughly explored, though conventional wisdom suggests that fewer dummy atoms are better~\cite{liu2013lead,mobley2012perspective}.
+%
+\paragraph{Atom mapping}
+Once a particular approach to the topology is selected, a crucial next step is to identify the common atoms which will not be perturbed.
+Rigorously, this process essentially comprises a maximal common substructure (MCSS) search of the molecules involved to identify the common substructure -- though the parameters of the MCSS search will differ depending on whether single or dual topology calculations are planned.
+Specifically, with a single topology approach in mind, atom types are allowed to change, so a permissive MCSS search can be done, whereas with dual topology a more strict search is required.
+\begin{figure}[h!]
+    \includegraphics[width=0.95\linewidth]{figures/fig4_mcs/Figure.pdf}
+    \caption{\textbf{Illustration of maximum common substructure matches} MCSS is shown in green for when A: a restrictive MCSS match is used and in B: ring breaking is allowed.}
+    \label{fig:fig_mcss}
+\end{figure} 
+%
+There are different tools that allow the generation of MCSS matches as well as single topology input.
+A large number of software can compute MCSS matches, some rely on RDKit, such as LOMAP or FESetup and partially BioSimSpace, others such as fkckombu are standalone tools~\cite{rdkit2019Dec, loeffler2015fesetup, liu2013lead, hedges2019biosimspace, kawabata20143d}.
+Schr\"{o}dinger's FEP+ planning tool is based on a version of LOMAP, and it also uses MCSS matching as well as 3D considerations to plan the network of single topology calculations between molecules~\cite{wang2015accurate}.
+%
+MCSS searches can be relatively time consuming, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial assessment and then use MCSS only to identify final mappings for relative calculations.
+%
+The MCSS approach, though relatively standard, takes into account only topological similarity.
+It is possible that changes in binding mode could actually require a different choice of mapping, so in some cases mappings may need to be planned differently depending on 3D positioning of atoms in space.
+%
+Single topology relative calculations, and calculations based on substructure searches, only work if in fact the ligands share a common substructure, e.g. are part of a congeneric series.
+If no common substructure is shared, then alternative dual or hybrid topology free energy calculations are needed, where one would co-localize a pair of compounds in a binding site, exclude their interactions with one another, and compute the relative binding free energy by turning one molecule on from being dummy atoms while turning the other off.
+To our knowledge no general pipeline for such calculations yet exists and this would likely remain a research problem.
+Using an absolute style free energy calculation instead seems a more promising approach in such a case.
+%
+\paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic, in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle~\cite{}, which may not be true whenever dummy atoms are involved in rings.
+Some approaches have attempted to address this~\cite{clark2019relative} but a general solution is not yet in mainstream use.
+%
+\paragraph{Perturbation maps}
+Based on the input ligand series a perturbation map or network can be planned.
+Recent heuristics have shown the more connected the perturbation network the better, however, there is a way to optimise network structure while minimising the number of perturbations that need to be computed minimising the resulting computational cost~\cite{yang2019optimal}.
+Sometimes the introduction of intermediates that are not part of the original congeneric series are essential to avoid ring breaking, or dealing with perturbations that would otherwise result in large numbers of atoms to be grown or disappearing.
+Some commercial tools have underlying good heuristic but may fail with complicated input, needing user validation in particular when dealing with chiral compounds.
+\begin{figure}[h!]
+    \includegraphics[width=0.95\columnwidth]{figures/fig6_types_of_networks/Figure.pdf}
+    \caption{\textbf{Examples of perturbation networks} A: Star shaped network with the crystal structure in the center.
+B: Network with cycle closures.
+Arrows indicate the direction of the perturbation.
+Green cycles indicate good cycle closure, red poor cycle closure.
+The red arrow indicates a poorly converged simulation that would give rise to bad cycle closures.
+The diamond indicates the use of a crystal structure binding mode.}
+    \label{fig:fig_types_of_networks}
+\end{figure} 
+In some cases, during the lead optimization stage, or if perturbations would be rather large a star shaped network as seen in fig.~\ref{fig:fig_types_of_networks} (A) is used..
+However, adding redundancy into the network means that a better error analysis can be carried out, by looking at cycle closure errors as discussed in sec.~\ref{subsec:predictions}, with an example given in fig.~\ref{fig:fig_types_of_networks}.
 
-\section{Prerequisites}
+Methods in experimental design have been applied to the construction of the perturbation maps.
+Yang et al.~\cite{yang2019optimal} optimized the perturbation map by selecting a fixed number of calculations from the pairwise perturbations so that the resulting set of calculations minimize the total variance.
+Xu~\cite{xu2019diffnet} optimized the perturbation map by allocating different amounts of simulation time to different pairwise perturbations so as to minimize the total variance, given the total simulation time of all the perturbation calculations.
+Both approaches lead to substantial reduction in the statistical error of the estimated free energies.
+%
+\paragraph{Constraints and relative free energy calculations.}
+One issue which requires particular care is the use of constraints.
+Commonly, bonds involving hydrogen are constrained to a fixed length using algorithms such as SHAKE or LINCS, allowing the use of longer timesteps~\cite{}.
+However, in single topology relative free energy calculations, the atoms involved might be mutated to other atom types -- for example, in a mutation of methane to methanol, one hydrogen might become an oxygen atom.
+Typical molecular dynamics engines are not set up to recognize this change, or at least not to correctly include contributions to the free energy from changing constraints/constraint length, so results for a transformation would usually be erroneous.
+At present the most general solution to this problem is simply to avoid the use of constraints (and thus use a smaller timestep if necessary, usually of around 1 fs) in any relative free energy calculation involving a transformation of a constrained bond.
+%
+\subsubsection{Absolute free energy calculations must handle the standard state and use restraints}
+\label{sec:standardstate-restraints}
+%
+\begin{figure}
+    \includegraphics[width=0.95\linewidth]{figures/fig5_thermo_cyc_abs/Figure.pdf}
+    \caption{\textbf{Absolute thermodynamic cycle}}
+    \label{fig:fig_absolute_thermodynamic_cycle}
+\end{figure}
+%
+Absolute free energy calculations involve completely removing the interactions between the ligand or solute and its environment, taking it to a non-interacting state that may or may not retain intramolecular non-bonded interactions.
+This non-interacting state can then be shifted between environments (from the protein to water, or from one solution to another) without changing its free energy, and then interactions can be restored.
+%
+Absolute free energies are typically reported with respect to a specific reference or standard state, which effectively determines the arbitrary point at which the free energy is 0.
+The role of the standard state is particularly evident with binding free energies, in which having a reference state allows us to obtain a well-defined partial molar free energy for the reaction
+\begin{equation*}
+\ce{AB <=> A + B}
+\end{equation*}
+through the well-known expression derived from the law of mass action
+\begin{equation} \label{eq:DGfromKAB}
+\Delta G = -RT ~ \ln \left( c^{\circ} K_{AB} \right)  = -RT ~ \ln\left( \frac{c^{\circ} C_{AB}}{C_A C_B} \right) ,
+\end{equation}
+where $R$ is the gas, $T$ is the temperature, $C_X$ is the equilibrium concentration of the chemical species $X$ in the reaction solvent, and the reference state concentration $c^{\circ}$ converts the binding constant $K_{AB}$ into a dimensionless quantity expressed in reference concentration units.
+It should be noted that ignoring the term $c^{\circ}$ is equivalent to assuming a reference concentration of 1~D$^{-1}$, where D are the units used to express $K_{AB}$, and would thus cause the value of $\Delta G$ to vary with the choice of the units.
+Typically, it is convenient to define a standard state at a constant pressure of 1~atm and where each chemical species (i.e., A, B, and AB) in the reaction solvent has a concentration of $c^{\circ}$~=~1~M~=~1~molecule/1660~\r{A}$^3$ but do not interact with other molecules of A, B, or AB.
+%
+\paragraph{Handling the standard state in absolute free energy calculations.}
+For solvation free energy calculations, handling the standard state is typically straightforward, and treating it correctly simply means ensuring that the non-interacting solute is taken to the same (or equivalent) final reference state in both environments, e.g.
+that the transformation involves a 1M to 1M equivalent transfer free energy (where the non-interacting solute still occupies essentially the same volume as the solute in the interacting system).
+So typically in such cases no special care is required to ensure the correct standard state, as long as the \emph{experimental} data being analyzed uses the same standard state and if it does not, a simple entropic correction is needed.
+%
+However, for binding the situation is much more complex and requires special care.
+Experimental absolute binding free energies are reported relative to a specific reference state -- a 1 M standard state -- which must also be used in calculations.
+In practice, this has implications for how the calculations are done, as the reference concentration must enter the thermodynamic cycle employed.
+%
+Typically, to deal with both practical sampling issues and the standard state issue, restraints are employed in absolute binding free energy calculations to keep the ligand in a well defined volume as its interactions with the system are removed~\cite{gilson1997statisticalthermodynamic}.
+This solves two problems.
+First, if the ligand were not kept in a well-defined region, as its interactions were removed it might wander the system, perhaps quite slowly, and only inadequately sample the non-interacting or weakly interacting state -- yet adequate sampling of these states might be required for convergence.
+So for practical purposes, the use of restraints can dramatically improve sampling as interactions are weakened and removed.
+Second, if the ligand is not kept in a well-defined region then it is hard to determine how to link a computed binding free energy to the correct 1 M standard state.
+In contrast, with restraints, the free energy of releasing the restrained ligand to a 1 M standard state can be computed analytically or numerically by solving the relevant integral, allowing the standard state to enter the thermodynamic cycle.
+%
+\paragraph{Several choices of restraints are possible.}
+\todo[inline, color={blue!20}]{@author: Check references}
+In practice, a variety of types of restraints are common, from simple harmonic distance restraints between the ligand and the protein~\cite{mobley2006use}, to flat-bottom restraints which work similarly but only exert a force if the ligand leaves a specific region~\cite{}.
+%
+Alternatively, a set of restraints proposed by Boresch have also commonly been employed, where all six rigid-body degrees of freedom governing the orientation of the ligand relative to the receptor are restrained~\cite{leitgeb2005alchemical}.
+Further restraints, such as on the overall ligand RMSD have also been used~\cite{woo2005calculation}.
+%
+In principle, all of these forms will yield correct binding free energies in the limit of adequate sampling (if their effects and connection to the standard state are correctly handled) but they have different strengths and weaknesses.
+For example, with more involved restraints, sampling at intermediate lambda values will usually not need to be as extensive but more computational effort must go to computing the restraining free energy.
+Additionally, such restraints would typically keep the ligand from exploring alternative binding modes, which may be undesirable with Hamiltonian lambda exchange or expanded ensemble techniques where allowing the ligand to exchange binding modes when it is non-interacting could provide sampling benefits~\cite{wang2013identifying}.
+Concretely, flat bottom restraints might allow a ligand to explore multiple binding sites, harmonic restraints multiple binding modes within a site, and Boresch restraints a single binding mode within a single site.
+See additional discussion of the possibility of multiple binding modes below~\ref{sec:multiple_binding_modes}.
+%
+Many choices of restraints involve selecting reference atoms.
+Again, in principle this choice is unimportant given adequate simulation time but practical considerations may be important.
+The choice is likely especially important with Boresch-style restraints, where some relative placements of reference atoms are likely to be numerically unstable; additionally, ligand reference atoms should likely be in a part of the molecule which defines the binding orientation well, rather than in a floppy solvent-exposed tail, for example.
+\todo[inline, color={blue!20}]{DLM: Get input from JDC on what they've learned about these.}
+\todo[inline, color={blue!20}]{DLM: Clarify terminology: Double decoupling, etc. See Feature Box below.}
+%
+\subsection{Absolute and relative calculations must deal with some of the same issues}
+\subsubsection{Structural definition of the bound state and weak binders}
+\todo[inline, color={green!20}]{ASJM: This a bit waffly. What are the recommendations, of how to setup these restraints?}
+In binding free energy calculations, extra care should be taken when simulating the bound state, especially when dealing with weak binders and absolute free energy calculations in combination with enhanced sampling techniques.
+In principle, only configurations of the receptor-ligand complex that we consider "bound" should be sampled from the bound state, and, in atomistic simulations, this requires to establish a structural definition of the bound state.
+In practice, the simulation of the bound state starts with the ligand already placed in the binding site and relies on kinetic trapping to maintain a bound complex.
+However, this strategy may not be sufficient if the complex dissociation rate is high or if methodologies such as Hamiltonian replica exchange~\cite{bitetti-putzer2003generalized} and expanded ensemble~\cite{chodera2011replica} are employed in absolute free energy calculations to enhance sampling since, in both cases, the ligand may find a way out of the binding site on timescales that are achievable by modern MD simulations.
+A solution commonly adopted in these cases is the use of one or more restraints making the unbound configurations energetically unfavorable through the addition of extra terms in the potential function.
+In practice, even with restraints, it is not always trivial to force a molecular dynamics simulation to explore a restricted region of the configurational space with a complicated geometry such as a binding site.
+If the ligand is a tight binder\todo[inline, color={red!40}]{AR: Should we give an order of magnitude to define a tight and a weak binder?}, it is usually safe to employ a restraint that allows some of the unbound configurations to be sampled since they generally contribute negligibly to the partition function (i.e., they have a relatively small Boltzmann weight) as long as the sampled volume is not so large that their cumulative contribution becomes significant.
+However, this is more problematic when dealing with weak binders as the unbound configurations can have a non-negligible Boltzmann weight, and their binding affinity can exhibit a significant dependency on the restraint type and parameters that are used to determine the sampled volume~\cite{}.
+Moreover, it should be noted that the additional potential energy terms used to model the restraints can introduce a bias in the predicted free energy.
+The bias can be removed at the analysis stage through reweighting techniques, but this procedure can increase the statistical uncertainty of the binding free energy estimate when the restraints are so strong that the overlap with the reweighted state is diminished~\cite{}.
+%
+Finally, it is useful to keep in mind that, for a meaningful comparison between computational predictions and experimental measurements, the definition of the bound state should be consistent.
+In particular, this means that the signal used by the experimental methodology to determine the fraction of bound complexes in solution should in principle reflect the population of complexes in the bound state as defined in the calculation.
+%
+\subsubsection{Changes in net charge can be challenging/problematic.}
+If the net charge of the system will change as the alchemical calculation progresses, this can pose major challenges.
+Specifically, finite-size effects can introduce profound artifacts into computed binding free energies~\cite{}, in part because typical schemes for long-range electrostatics (including PME and reaction field) do not handle free energy contributions from such changes effectively or as they would be handled in a hypothetical macroscopic bulk solution~\cite{}.
+%
+There are two main potential solutions to avoid artifacts due to changes in net charge: Correcting for the introduced artifacts, or avoiding changing the net charge.
+%
+Many relative free energy planning tools have been set up to avoid changing the net charge of the systems considered, including LOMAP~\cite{} and Schr\"{o}dinger's FEP+~\cite{chen2018accurate}.
+Absolute free energy calculations can also potentially avoid changing the charge of the system by making a charge perturbation of equal and opposite sign elsewhere in the system; for example, as a charged ligand is removed, a charged counterion of opposite sign could also be removed, or one of the same sign could be inserted.
+This is sometimes referred to as an "alchemical ion" approach for dealing with the needed charge change, and is also employed by the Yank free energy package~\cite{wang2013identifying}.
+Charge corrections have also been explored, and are potentially a viable solution to this problem~\cite{} where artifacts introduced by finite-size effects are corrected numerically.
+~\cite{chen2018accurate}. However, application of such corrections typically remains less common than the use of a co-alchemical ion.
+%
+When free energy calculations \emph{do} need to change the charge of a ligand or solute, the literature does not yet seem to indicate what approach should be preferable, so considerable care should be taken.
+We are not yet aware of a careful comparison of charge corrections versus other approaches such as decoupling an ion at the same time, so in our view the issue of proper handling of charge mutations in the context of alchemical calculations remains a research problem.
+%
+\subsubsection{The importance of the alchemical pathway\label{sec:important_path}}
+\todo[inline, color={red!40}]{LNN: write common principles alchemical path choice. MRS: mostly wrote this, LNN can comment on it. LNN: LGTM}
+\todo[inline, color={red!40}]{AR: write absolute-specific section on  alchemical path choice}
+\todo[inline, color={red!40}]{BA: write relative-specific section on  alchemical path choice}
 
-Here you would identify prerequisites/background knowledge that are assumed by your work and your checklist which you view as critical, ideally giving links to good sources on these topics.
-Checklists are normally focused on errors made by users with training and experience in molecular simulations, so you can assume a basic familiarity with the fundamentals of molecular simulations.
+\begin{figure}
+    \includegraphics[width=0.95\linewidth]{figures/fig7_what_is_lam/Figure.pdf}
+    \caption{Alchemical intermediates are createtd by making the potential energy depend on a an additional variable $\lambda$ that interpolates between the chemical endpoints.
+In \textbf{A}, at $\lambda=0$ the molecule is a fully interacting phenol and at $\lambda=1$,  a fully interacting benzene.
+\textbf{B} shows an illustration of the probability distribution of the potential energies as the switching function takes values of $\lambda=0$ to $\lambda=1$, intermediates are required for a sufficient overlap in potential energies to estimate a free energy difference between $\lambda=0$ and $\lambda=1$.
+    One of the most efficient families of intermediate pathways  are softcore potentials, whose shape is illustrated in \textbf{C} with dependence on $\lambda$ and \textbf{D} with dependence on the potential energy.
+Note how when $\lambda$ approaches 0, the energy smoothly approaches zero at all $r$, a necessary requirement for efficient calculations.
+}
+    \label{fig:fig_what_is_lambda}
+\end{figure}
+%
+Both absolute and relative calculations must choose an alchemical pathway connecting initial and final states.
+In principle, because of the path independence of the free energy, any arbitrary pathway will give the correct free energy change, but the choice of pathway will greatly affect the efficiency of the calculations.
+Some choices are particularly crucial -- for example, transformations involving insertions or deletions of atoms should employ a path soft-core potential path for Lennard-Jones or other interactions with repulsive interactions that go to infinite energy at small radius~\cite{}.
+%
+The key consideration for choosing alchemical pathways is that the intermediate states that a given pathway produces should sample configurational ensembles that change as slowly as possible as $\lambda$ changes, while still managing to go from the initial state to the final state as $\lambda$ goes from 0 to 1.
+%
+Another way of stating this is that intermediate states should sample molecular configurations that are as similar as possible to their neighboring states.
+The more similar the configurations are between intermediate states, the lower the statistical uncertainty is in the estimate of free energy between intervals.
+This can be proven directly from the BAR and MBAR formulas~\cite{bennett1976efficient,klimovich2015guidelines}, though the exact same principles apply for TI.
+%
+The simplest possible alchemical pathway is a \textit{linear} pathway:
+%
+\begin{equation}
+U(\vec{q},\lambda) = \lambda U_0(\vec{q}) + (1-\lambda)U_1(\vec{q})    
+\end{equation}
+%
+so-called because the dependence on $\lambda$ is linear.
+This clearly satisfies the basic requirement that it gives the initial endpoint potential energy $U_0(\vec{q})$ when $\lambda=0$ and final endpoint energy $U_1(\vec{q})$ when $\lambda=1$.
+%
+For many energy terms, \textit{as long as a repulsive core remains on}, this is a very good approach.
+For example, it can be shown that if van der Waals repulsions are left on, then the linear approach is very nearly the optimal path  possible for changing, removing, or inserting the electrostatic energy terms, with the path being within about 10--20\% of the minimum possible uncertainty~\cite{naden2015linear} for a fixed amount of simulation, as well as being nearly optimally efficient for van der Waals attractive terms with repulsion terms turned on~\cite{naden2014linear}.
+Although we are not aware of any quantitative tests for dipolar or higher multipole terms, theoretically it should behave equally well for those systems.
+%
+However, this approach ends up being terrible for removing or adding repulsive potentials that go to infinity quickly at or near the origin.
+One way to look at this is to examine how low $\lambda$ must go to reduce the energy at $0.5\sigma$ (the atomic size parameter) down to 1 $k_BT$, where thermal fluctuations make it possible for other atomic sites to penetrate routinely that deep.
+If we are trying go from a particle being present, and desire to make it disappear alchemically, then if the repulsive terms is of the form $\epsilon\frac{\sigma}{r}^{12}$, then if $\epsilon$ was 1 $K_BT$ at the temperature of interest, and we start with the particle present, then solving for $(1-\lambda)(1 kB_T)\frac{\sigma}{\frac{1}{2}\sigma}) = 1 k_B T$ we get $\lambda = 1-2^{-12} \sim 0.999976$.
+So we have gone virtually all the way to the end of the transformation, there's still an impenetrable post in the middle of our simulation!
+This is not very much like the desired final state of no interactions between the particle and its environment.
+We can play around with a few ways of modifying this, like simulating many more intermediate states near $\lambda=1$.
+Various analyses have shown that this is not a very good strategy, however~\cite{pham2011identifying, beutler1994avoiding, zacharias1994separationshifted, blondel2004ensemble}
+%
+What we need instead is some sort of function that smoothly gets rid of this infinity.
+A large number of scheme have been tried~\cite{beutler1994avoiding, zacharias1994separationshifted, blondel2004ensemble, pham2011identifying, pham2012optimal, naden2014linear, donnini2005molecular}, but the most common strategy that appears to be the best practices is to use a "soft-core" potential, of the form:
+%
+\begin{equation}
+    U(\vec{r_{ij}},\lambda) = 4\epsilon_{ij} \lambda \left(\frac{1}{(\alpha(1-\lambda) + (r_{ij}/\sigma_{ij})^6)^2} +  \frac{1}{\alpha(1-\lambda) + (r_{ij}/\sigma_{ij})^6}\right)\label{eq:softcore}
+\end{equation}
+Where $r_{ij}$ is the distance between two particles $i$ and $j$, $\epsilon_{ij}$ and $\sigma_{ij}$ are the Lennard-Jones parameters corresponding to the interaction between particles $i$ and $j$, and $\alpha$ is a constant (0.5 is optimal for the functional form shown above).
+This functional form has exactly the property we are looking for: it recovers the Lennard-Jones potential when $\lambda=1$, and the other endpoint ($\lambda=0$), it is exactly zero for all $r_{ij}$ everywhere, and as $\lambda$ goes to zero, the  $\alpha(1-\lambda)$ term lowers the infinite energy in the core.
+There are several different variants of the same functional form~\cite{zacharias1994separationshifted, beutler1994avoiding,pham2011identifying}, but the one in given in eq.~\ref{eq:softcore} is easy to understand and implement and fairly numerically stable.
+This functional form is shown in parts C and D of figure~\ref{fig:fig_what_is_lambda}.
+It has been shown that more complicated forms are not significantly more efficient than eq.~\cite{pham2012optimal}.
+We recommend using the softcore potential given in eq.~\cite{eq:softcore}, unless there is a compelling reason otherwise.
+Using a similar equation to eq.~\cite{eq:softcore} may be acceptable in most circumstances if that is what is supported in your chosen software.
+However, if you are inserting or removing entire atomic sites, we heavily recommend against using the linear approach; it will be very difficult to get correct results.
+%
+We have discussed optimal ways of disappearing or appearing Lennard-Jones interaction sites and turning on and off electrostatics terms.
+What about performing both transformations at the same time?
+We can't turn off the electrostatics linearly at the same time we turn off the Lennard-Jones terms, as it would leave infinitely large attractive and repulsive electrostatic terms "bare" at small $\lambda$, resulting in the simulation crashing.
+It \textit{is} possible to apply the same soft core approach to the Coulomb interaction, but it turns out to be somewhat difficult to turn off two soft-core terms  terms simultaneously~\cite{Steinbrecher:JCC:2011}.
+We therefore recommend performing the transformations in sequence; first, turning off all electrostatics for atoms that must be removed, inserting and removing Lennard-Jones sites (both the insertion and removal can be done simultaneously), and then turning electrostatics for the introduced particles on.
+However, If there are no removals or introduction to atomic sites, then it is reasonable to change all interactions linearly.
+%
+Other issues, such as whether absolute calculations should retain or remove intramolecular nonbonded interactions, may be less critical and recommendations differ among studies in the literature~\cite{}, and reasonable efficiency can be often obtained with either choice even if some are somewhat better or worse than others.
+Our recommendation is to leave the intramolecular interactions on during the transformation for simplicity if there are no other known issues with this approach.
+The key thing to watch out for is whether the total potential energy, and therefore the intermediate ensembles sampled, change smoothly from beginning to end.
+If a configurational phase transition is noted in the process (like substantial dihedral rearrangement), then other steps must be taken, perhaps including modifying intramolecular interactions~\cite{???}.
+These problems can be diagnosed by noticing lack of configuration space overlap between different simulations (see section~\ref{subsec:})"
+%
+Relative calculations introduce additional choices, including whether to define explicit intermediate states~\cite{} or leave these implicitly defined by the code~\cite{}.
+Typically in single topology relative calculations it proves most efficient to first remove electrostatic interactions of any atoms which will be deleted, then modify other nonbonded interactions, then restore electrostatic interactions of any atoms which are being inserted.
+Other schemes, such as simultaneously changing electrostatic and Lennard-Jones interactions, even with electrostatic ``soft core'' potentials~\cite{steinbrecher2007nonlinear}, in our experience typically introduce errors and/or instabilities or are at least unreliable.
+It is expected that similar considerations will apply to dual topology calculations, and the principle of first removing electrostatics and then removing steric interactions will likely serve well.
+%
+A key additional consideration in choosing the alchemical pathway is the choice of spacing of intermediate states.
+The spacing depends to some extent on the choice of analysis method, though states should essentially be spaced equidistant in the relevant thermodynamic length~\cite{}.
+For BAR/MBAR techniques this means that spacings should typically be equal in [what, variance? ref].
+Some schemes to adaptively optimize the spacing of intermediate states based on initial exploratory simulations have been proposed~\cite{}.
+%
+An example of this approach for relative binding free energy calculations includes saving configurations from short 10 picosecond simulations starting at $\lambda_0$, and evaluating changes in the standard error of mean (SEM) in $\frac{dU}{d\lambda}$ as reevaluated at new values of $\lambda$, starting with a step size of 0.01 up to a predetermined threshold.
+This threshold can be empirically defined as the number of atoms being inserted or deleted in the simulation / 100.
+If the absolute value of the difference in SEM is below the defined threshold, then the change in $\lambda$ for re-evaluation of SEM in $\frac{dU}{d\lambda}$ is defined by the current difference in SEM from the threshold divided by the derivative of the SEM with respect to $\lambda$.
+%
+\subsubsection{Multiple or uncertain binding modes may require considerable care}
+\label{sec:multiple_binding_modes}
+In a discovery setting, new ligands typically have unknown or at least uncertain binding modes~\cite{kaus2015how, plountprice2000analysis,mobley2009binding,calabro2016elucidation},  complicating binding free energy estimation.
+This uncertainty is because it is usually not desirable to estimate a binding affinity for a ligand which already has an available bound structure, since such a compound has already been tested.
+To deal with prospective ligands with unknown binding modes, discovery projects commonly assume that modifications of functional groups on a common scaffold result in a consistent binding mode across all members of a series.
+This is not necessarily always the case~\cite{kaus2015how}, as reviewed elsewhere~cite{mobley2009binding} and in some cases unexpected binding mode changes can be the origin of apparent non-additivity in structure-activity relationships~\cite{calabro2016elucidation}.
+Binding modes also tend to be particularly variable in the case of fragments, which often may have multiple relevant binding modes~\cite{}.
+
+Absolute free energy calculations for dissimilar ligands can have particular challenges because the (potentially incorrect) assumption of consistent binding modes across a series of similar ligands is likely to be even less robust than the in the case of relatively calculations.
+This means that researchers performing absolute binding free energy calculations will have to pay particular attention to generating reasonable putative binding modes.
+
+In some cases, it is tempting to simply use docking techniques to generate initial bound structures for starting molecular dynamics simulations.
+However, timescales for binding mode interconversion are usually slow compared to MD/free energy timescales, meaning that simulations started from different potential binding modes are likely to yield disparate computed binding free energies~\cite{mobley2006use, palma2012computation, mobley2012perspective, gill2018binding} .
+And docking techniques are good at identifying sterically reasonable potential binding modes, but still perform relatively poorly at identifying a single dominant binding mode \emph{a priori}.~\cite{}
+
+It is worth highlighting a recent SAMPL blind challenge on HIV integrase as an illustration of this. 
+Many submissions, using state-of-the-art methods, had difficulty even predicting which \emph{binding site} ligands would bind in (most submissions placed more than half of the ligands into the incorrect binding site), and even given correct binding sites, the binding mode within each site was also quite difficult to predict~\cite{mobley2014blind}.
+The best performing submission for predicting binding modes actually ended up being a human expert (aided by computational tools) with more than 10 years of experience on the particular target~\cite{voet2014combining}, rather than a fully automated approach.
+While free energy calculations on this set had some success, many of the failures actually ended up being cases where the binding mode selected as input for free energy calculations was later found to be incorrect~\cite{gallicchio2014virtual}, highlighting the importance of these issues.
+
+One approach which has shown some success is to retain diverse potential binding modes from docking, perform short MD simulations of these to identify distinct stable binding modes, and then consider these in subsequent calculations~\cite{gallicchio2014virtual, mobley2006use,rocklin2013blind, boyce2009predicting, mobley2007predicting}.
+
+Routes to handle multiple potential binding modes are different depending on whether absolute or relative calculations are selected, unless a method is available to estimate the relative populations of different stable binding modes in advance (e.g.
+such as the BLUES approach currently in development~\cite{gill2018binding}), in which case this approach could be applied to assist both types of calculations.
+
+\paragraph{Handling multiple potential binding modes within absolute calculations.}
+Within absolute binding free energy calculations, multiple potential binding modes can be handled by two main strategies: Consider each binding mode separately (a separation of states strategy) or sample all binding modes within a single simulation~\cite{mobley2012perspective}.
+This couples to the choice of restraints selected, as some restraints will allow transitions between binding modes and even binding sites (Section~\ref{sec:standardstate-restraints}), and others do not.
+
+Sampling all binding modes within a single free energy calculation is usually impractical without some form of enhanced sampling or at least Hamiltonian replica exchange~\cite{wang2013identifying} because barriers for binding mode interconversion result in kinetics which are too slow compared to simulation timescales~\cite{mobley2006use, palma2012computation,mobley2012perspective, gill2018binding}.
+Hamiltonian exchange, coupled with appropriate restraints, can allow the ligand to relatively rapidly exchange between potential binding modes when non-interacting, accelerating sampling of binding modes~\cite{wang2013identifying}.
+\todo[inline, color={yellow!40}]{DLM probably need more background refs on Hamiltonian lambda exchange here.}
+
+Separation of states provides a simple though potentially expensive alternative, where each stable binding mode is considered separately with a binding free energy calculation restricted to that binding mode, and then (as long as the binding modes are non-overlapping) the resulting component binding free energies can be combined into a total~\cite{mobley2006use, mobley2012perspective}.
+This approach necessitates a separate binding free energy calculation for each potential binding mode, however, so it can be computationally quite costly.
+If relative populations of different stable binding modes were available from some other technique, it could make this separation of states approach considerably more efficient~\cite{mobley2012perspective, gill2018binding}.
+
+\paragraph{Handling multiple potential binding modes within relative calculations.}
+
+Multiple potential binding modes pose particular problems for relative free energy calculations, as having multiple starting structures for these calculations could yield substantially different calculated relative binding free energies for the same transformation due to kinetic trapping, and, without additional information (specifically, the free energy of binding mode interconversion or, equivalently, the relative populations of different binding modes) it becomes impossible to sort out which of the multiple answers is in fact the correct relative binding free energy~\cite{}.
+
+To deal with this, some practitioners have actually computed relative binding free energies of different binding modes of the same ligand~\cite{palma2012computation} [ref Jorgensen (?) etc.].
+For example, a mutation which adds a methyl to an aromatic ring of a larger ligand might yield one result if the methyl points in one direction, and a different value if it points in the other due to slow ring motions.
+[e.g.
+get ref from Sukanya]
+One could compute the free energy of turning off the methyl group in one orientation and turning it back on in the other orientation to obtain the free energy difference between the two potential binding modes.
+While this approach has precedent, it is relatively difficult to automate at present and requires considerable care.
+
+Overall, this likely means that relative free energy calculations will be susceptible to problems resulting from uncertainty in ligand binding modes until more robust approaches are available to determine dominant binding modes, or the relative populations of different potential binding modes, in advance.
+
+\subsubsection{What different sampling schemes exist and which one should I use?}
+With all AFE simulations having the requirement of sampling from different sets of $\lambda$ states in common different approaches have been designed.
+Fig.~\ref{fig:fig_sampling_scheme} illustrates the four most common schemes used.
+The simplest approach is defining a set of fixed $\lambda$ values and then run independent simulations at each of the intermediate $\lambda$ values, see fig.~\ref{fig:fig_sampling_scheme}(A).
+This type of scheme is currently used for AMBER TI calculations~\cite{song2019} and for BioSimSpace~\cite{hedges2019biosimspace}.
+A slightly more complex approach is allowing replica mixing, meaning that each independent $\lambda$ simulation is attempted to be exchanged with a neighbouring $\lambda$ replica.
+This is based on ideas of Monte Carlo simulations of spin glasses by Swendsen and Wang~\cite{wendsen1986replica}.
+Using a Metropolis Hastings based acceptance criterion for exchanges means that the generated ensemble of all replicas is still sampling from the Boltzmann distribution and has been used in many different contexts for molecular simulations~\cite{sugita2000multidimensionala,sugita1999replicaexchangea, woods2003developmenta, jiang2010free}.
+The basic idea of the replica exchange scheme is shown in fig.~\ref{fig:fig_sampling_scheme}(B).
+It is supported in various software packages that provide AFE implementations, such as GROMACS~\cite{aldeghi2015accurate}, FEP+~\cite{wang2015accurate}, and NAMD~\cite{jiang2019computing}.
+A third approach borrows ideas from simulated tempering~\cite{marinari1992simulateda}.
+In this scheme a single replica rapidly explores all of $\lambda$ space by optimally working out weights that allow switching between different intermediate $\lambdas$, as seen in fig.~\ref{fig:fig_sampling_scheme}(C) and also referred to as self adjusted mixture sampling~\cite{lyubartsev1992newa, li2007simulated, tan2017optimally} and while promising, has so far only been supported in~\cite{andrearizzi2019choderalab}.
+The last approach available in conjunction with GROMACS makes use of non-equilibrium simulations~\cite{aldeghi2018accurate}.
+In this approach, only end state $\lambda$ replicas ($\lambda$=0, $\lambda=1$) are simulated at equilibrium intermediate information is generated from non-equilibrium simulations.
+A schematic of this approach is shown in fig.~\ref{fig:fig_sampling_scheme}.
+\ref{fig:fig_sampling_scheme} A:\cite{hedges2019biosimspace, song2019using } B\cite{aldeghi2015accurate, wang2015accurate,jiang2019computing} C:\cite{li2007simulated, tan2017optimally} D:\cite{aldeghi2018accurate}
+\begin{figure}
+    \includegraphics[width=0.95\columnwidth]{figures/fig8_sampl_scheme/Figure.pdf}
+    \caption{\textbf{Four sampling strategies available in different software packages.} \textbf{A}: Multiple replicas in parallel at different lambda states.
+\textbf{B}: Hamiltonian replica exchange scheme.
+\textbf{C}: Single replica scheme sampling from all $\lambda$ states.
+\textbf{D}: Non-equilibrium sampling scheme.}
+    \label{fig:fig_sampling_scheme}
+\end{figure} 
+The best practice recommendation, is to use replica exchange type sampling schemes (fig.~\ref{fig:fig_sampling_scheme}(B)), or in absence of their availability and when looking at systems with fast mixing times parallel replica schemes are acceptable (fig.~\ref{fig:fig_sampling_scheme}(A)).
+Single replica schemes and non-equilibrium schemes are not as established yet, but are very promising.
+
+\subsubsection{How long should I run my simulation for and what information should be saved?}
+Before launching alchemical free energy calculations it is wise to consider how convergence and completion will be assessed.
+Different conditions on when to stop alchemical free energy calculations should be determined, and this may require several iterative checks and therefore modifications to the calculation protocol.
+One useful metric to use for termination is the expected or desired uncertainty of a desired free energy estimate, though care must be exercised should the uncertainty estimate prove unreliable.
+In particular, if the rate of change in the free energy estimate is significant when this condition is met, the simulation may not be locally converged, and more sampling may be necessary to determine a stable free energy estimate which is no longer changing significantly over time.
+However, this is not the only metric which should be used, as the uncertainty only captures the information about the sampled phase space, not necessarily the entirety of the phase space.  
+For example, 
+convergence of relative free energy calculations in predictive simulations where the entire phase space is not known in advance, requires sampling the different kinetically stable states~\cite{mobley2012perspective}.
+This highlights the importance of choosing the correct thermodynamic path to ensure you sample the required thermodynamic states as discussed in Section~\ref{sec:important_path}.
+%
+The condition of minimizing the statistical uncertainty of different free energy estimators below a sufficient threshold should be one metric monitored over the simulation.
+This can be done through the uncertainty estimator built into certain analysis tools such as MBAR, or can be done though more general statistical tools like Bootstrap Sampling.
+A target statistical uncertainty should be chosen at the onset of the simulation to avoid excessively long simulations, or falling into the trap of running until the free energy estimate is "good enough," which is subjective and has no defined criteria.
+This could be a fixed value such as $0.20 \mathrm{kcal/mol}$, or a functional quantity such as "below $0.5 \mathrm{kcal/mol}$ and $10\%$ of the free energy estimate." The user does not need to monitor this information in real-time and can choose to run simulations for fixed duration (either time or number of samples) and run analysis on the data collected thus far.
+If more samples are needed, the simulations can be resumed, or, started again in different initial conditions.
+
+Convergence in other alchemical observables should also be monitored to determine if the defined phase space has been sufficiently sampled and enough decorrelated samples have been drawn.
+These additional observables include, but are not
+limited to, the variance in $\frac{dU}{d\lambda}$ across all $\lambda$ values, calculating the variance in free energy using bootstrap analysis, and comparing differences in free energies calculated using different percentages of the simulation in both the forward and reverse directions (figure \ref{fig:convergence_forward_reverse}).
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.95\linewidth]{figures/fig10_forward_reverse/Figure.pdf}
+    \caption{\textbf{Free energy (in k$_{B}$T) for two different relative binding free energy (RBFE) perturbations.} 
+    Each plot shows the estimated free energy change using a varying fraction of total simulation time (up to 5 ns total). 
+    The top panel shows the three-step protocol for a perturbation involving 5 perturbed atoms, while the bottom panel shows the same protocol for a perturbation involving 10 perturbed atoms.
+The difference in energy between the forward (blue) and reverse (red) free energy calculations at the midpoint of the simulation time gives an indication of the overall convergence of the simulation, with differences over 1 k$_{B}$T indicating poor convergence.}
+    \label{fig:convergence_forward_reverse}
+\end{figure}
+
+Each of these metrics have demonstrated promising results for diagnosing when a simulation has a convergence issue beyond simple convergence of uncertainty estimate. 
+Results obtained from calculations with convergence issues should be checked for errors or run for longer before any confidence should be placed in conclusions drawn from their analysis.
+In relative calculations that share similar binding modes, for example, and do not induce large conformational changes when in complex with protein, the need to sample exhaustively to converge estimates in free energy differences is often not necessary due to the locality of sampling changes in the molecular topology and shared phase space of the core atoms.
+However, even subtly induced changes in protein binding configuration will require more sampling or cause local convergence to a free energy estimate that has high error.
+Many enhanced sampling techniques have been proposed to try and overcome barriers in free energy calculations and mitigate convergence to erroneous estimates~\cite{}. 
+The confidence a user should have in a free energy estimate is significantly improved when both the uncertainty of the free energy estimate is low, and when other observables have reached a convergence.
+
+A simple, but effective, way of checking for bias is to compute the same property by multiple estimators. 
+The uncertainty in the free energy, for example, has multiple ways to be estimated, e.g.
+through standard error propagation methods (including MBAR's estimator, which is based on the same principles as standard error propagation), through bootstrap methods, through multiple independent runs, etc.
+Independent of how the property is estimated, its important to remember that they are \textit{estimations of the property}, not the true underlying property itself. 
+These estimators are usually consistent estimators, meaning they will converge to the true answer in the limit of sufficient sampling, not necessarily unbiased ones though.
+As such, it is a good idea to subject different estimators to the same data to see if they yield either the same estimate (within error and bias), or if they fluctuate wildly. 
+This is not a perfect method as some estimators, such as exponential averaging, will converge significantly more slowely, relative to more accurate estimators like MBAR. 
+Therefore, it is a good idea to apply the estimators to different fractions of the data to see if the main estimator of free energy you have chosen is stable.
+
+Each method requries different data about the simulation be collected.
+If, for instance, the free energy estimator selected is thermodynamic integration, then values of $\frac{dU}{d\lambda}$ at uncorrelated data points must be collected,  Once a combination of knowing what type of simulation you will run, which alchemical topology you will simulate, what alchemical path you will simulation along, and what your stopping conditions are, then you are ready to enumerate the information you should capture.
+Below is a sample of the minimal information you need for a set of common estimators:
+
+\begin{itemize}
+    \item Thermodynamic Integration (TI) requires $\frac{\partial U(\vec{x})}{\partial\lambda}$.
+    \item Exponential Averaging (EXP) needs \textit{either} $\Delta U_{k,k+1}(\vec{x})$ or $\Delta U_{k,k-1}(\vec{x})$, depending on the direction its being evaluated in.
+    \item Bennett Acceptance Ratio (BAR) needs \textit{both} $\Delta U_{k,k+1}(\vec{x})$ and $\Delta U_{k,k-1}(\vec{x})$.
+    \item Weighted Histogram Analysis Method (WHAM) and Multistate Bennett Acceptance Ratio (MBAR) both need the complete set of $\Delta U_{k,j} \, \forall \, j=\{1...K\}$.
+WHAM must have this information binned.
+\end{itemize}
+
+The potential derivative required for TI should generally be calculated during the simulation; only under every rare circumstances~\cite{Naden:JCTC:2015} can it postprocessed by a code that does not evaluate the derivatives.
+Many codes already have options for doing this.
+If that option is unavailable, you can estimate it through finite difference (if sufficient information is collected), but this  will introduce significant error, and is generally not a best practice.
+Tthe BAR estimator may be a better, and simpler choice at that point as you will have at least the same level of information.
+The potential energy differences required for EXP, BAR, MBAR, and WHAM can be calculated either during the simulation or in post-processing.
+It is recommended to calculate the potential differences in code when possible to avoid extra overhead and possible errors produced by running the simulation twice, and to reduce the amount of stored information.
+Although TI must be usually be calculated in code, as it requires the derivative, there is one condition under which it actually has the fastest computation time. 
+If the alchemical path you have chosen is a linear alchemical path, then you get $\frac{dU}{d\lambda} = U_0(\vec{x}) - U_1(\vec{x})$, which is the difference between the initial and final states. 
+However, because of the problems with linear paths already discussed in this paper, this simplification is rarely that useful.
+
+Free energy information should generally be saved more frequently than coordinate data, approximately at the rate that uncorrelated samples are produced.  
+The on-disk size of the data for free energy estimation is often significantly smaller than full atomic coordinates, so the information should be collected frequently. 
+However, the information should not be collected \textit{every} time step, as most free energy techniques are operated at equilibrium, and need equilibrated \textit{and decorrelated} samples for an unbiased estimate.
+A sample collected every time step will likely result in most samples being discarded due to decorrelation routines in the analysis.
+However, if it is computationally cheap and disk space is plentiful, do save often.
+One may safely assume that the correlation time is greater than 100-200 fs even for relatively simple systems such as small molecules in solvent, so saving no more frequently than every 50-100 steps is recommended.
+How decorrelation impacts calculations, and how to compute it is discussed in other sections.  [section ref]
+
+\section{Step 3 -- Data analysis}
+\label{sec:step3}
+
+Once data has been collected from alchemical intermediates, it must be analyzed to produce an estimate of the free energy change (and its associated statistical uncertainty) for each leg of the thermodynamic cycle.
+While a number of different estimators are available that will give consistent results under optimal circumstances, some approaches are recommended over others due to their robustness and ability to provide information on poor convergence.
+
+\subsection{Detecting the boundary between equilibrated and production regions}
+\label{sec:automatic-equilibration-detection}
+
+Much of the infrastructure for analyzing alchemical free energy calculations relies on the concept of asymptotically unbiased estimators, which produce unbiased estimates of the free energy when fed very long simulations~\cite{shirts2005comparison}.
+In reality, free energy calculations are often initiated from highly atypical initial conditions (such as a protein-ligand geometry obtained from docking and subjected to a heuristic solvent placement scheme), and simulations are of a finite length dictated by available computational resources and computing demands.
+As a result, these estimators can produce significantly biased estimates if fed the entirety of simulation data generated without further processing~\cite{chodera2016simple}.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.95\linewidth]{figures/fig11_equib_detection/Figure.pdf}
+    \caption{{\bf Automatic partitioning into equilibration and production regions.}
+    \textbf{A} The average (black line) standard deviation (shaded region) of the reduced potential $u^*$ over many independent replicate simulations started from the same initial conditions show a significant initial transient before relaxing to the true average potential energy\textbf{B} A cumulative average (red) of the entire simulation data demonstrates simulation bias not seen when initial simulation data is omitted (blue).
+Using an automated approach to detect equilibration of the boundary $t_0$ using statistical inefficiency $g$ (\textbf{C})  for an effective simulation interval (\textbf{D}) \textbf{E} The optimal equilibration boundary $t_0$ is selected to maximize the number of uncorrelated samples.
+    \emph{Figure adapted from~\cite{chodera2016simple}.}
+    }
+    \label{fig:automatic-equilibration-detection}
+\end{figure}
+
+To minimize this effect, an initial portion of the simulation is often discarded to \emph{equilibration}~\cite{braun2019best}, with the idea of removing the most heavily biased initial portion of simulation data but retaining the unbiased \emph{production} region that represents a stationary Markov chain process sampling from the desired equilibrium target distribution.
+Because the simulation time required for the atypical initial sampler state to relax toward equilibrium is a property of the specific system being simulated and the specific initial conditions selected, it is simplest to collect data for the whole process and use an automated algorithm to select how much data should be discarded to equilibration in a post-processing step.
+
+A simple approach to automatically partitioning simulation data into equilibration and production regions is described in~\cite{chodera2016simple} (illustrated in Figure~\ref{fig:automatic-equilibration-detection}).
+Suppose we have a simulation of length $T$ consisting of correlated data.
+Here, the goal of the post-processing step is to select the equilibration boundary $t_0 \in [0, T]$ so as to \emph{maximize} the number of effectively uncorrelated samples remaining in the production region $N_{[t_0,T]}$, which is defined as
+\begin{eqnarray}
+N_{[t_0,T]} &=& \frac{T - t_0}{g_{[t_0,T]}}
+\end{eqnarray}
+where $g_{[t_0,T]}$ is the \emph{statistical inefficiency} of a timeseries $a_t$, described in more detail below.
+Conveniently, this procedure also produces the information necessary to decorrelate the simulation data for estimating the free energy differences, a requisite next step in analysis.
+This approach is implemented within the MBAR~\cite{kylebeauchamp2019choderalab} and alchemlyb~\cite{daviddotson2019alchemistry} packages, and is highly recommended for standard practice.
+
+For additional discussion of working with correlated data and autocorrelation analysis, please refer to the work on Best Practices for Quantification of Uncertainty and Sampling Quality in Molecular Simulations~\cite{grossfield2018best}.
+
+\paragraph{Computing the timeseries for equilibration detection}
+Typically, the timeseries of note $a_t$ analyzed in automated equilibration detection is the negative logarithm of the probability density sampled by the MCMC algorithm (up to an irrelevant additive constant).
+For simple independent simulations that sample $x_t \sim \pi(x ; \lambda)$, this is given by the reduced potential
+\begin{eqnarray}
+a_t &\equiv& - \ln \pi(x_t; \lambda) + c = u(x_t; \lambda) .
+\end{eqnarray}
+
+Note that the use of the effective reduced potential is not guaranteed to pick up on all slow relaxation processes that may be coupled to the alchemical free energy, but the simplicity of its computation means it is generally appropriate for most cases.
+
+\paragraph{Cautions in automating equilibration detection}
+For simulations that are simply not long enough to contain a large number of samples from true equilibrium either because they are very short or contain slow processes, this procedure cannot completely remove the bias.
+In such cases, this approach simply selects the final portion of the the simulation, which may be contained in a single substate of conformational space, which may itself lead to biased estimates. 
+This situation can be detected if the equilibration boundary $t_0$ is a significant fraction of the total simulation length $T$, with a good rule of thumb being that $T \gtrsim 20 t_0$.
+If this is not possible, advanced analysis techniques that assume only local equilibrium (rather than global equilibrium) such as the TRAM estimators~\cite{mey2014xtram,wu2016multiensemble,nuske2017markov} may be more appropriate.
+
+\subsection{Decorrelating samples for analysis}
+\label{sec:decorrelating-samples}
+
+\paragraph{Computing the statistical inefficiency}
+Most estimators require an uncorrelated set of samples from the equilibrium distribution to produce (relatively) unbiased estimates of the free energy difference and its statistical uncertainty.
+To do this, the production region of the simulation is generally \emph{subsampled} with an interval approximately equal to or greater than the \emph{statistical inefficiency} $g \ge 1$ to produce a set of uncorrelated samples that can be fed to the estimator machinery~\cite{chodera2016simple},
+\begin{eqnarray}
+g &\equiv& 1 + 2 \tau_\mathrm{eq} \label{eq:statistical-inefficiency-definition}
+\end{eqnarray}
+where $\tau_\mathrm{eq}$ is the integrated autocorrelation time, formally defined as
+\begin{eqnarray}
+\tau_\mathrm{eq} &\equiv& \sum_{t=1}^{T-1} \left(1 - \frac{t}{T}\right) C_t \label{eq:integrated-autocorrelation-time-definition} , 
+\end{eqnarray}
+with the discrete-time normalized fluctuation autocorrelation function $C_t$ defined as
+\begin{eqnarray}
+C_t &\equiv& \frac{\expect{a_n a_{n+t}} - \expect{a_n}^2}{\expect{a_n^2} - \expect{a_n}^2} . \label{equation:autocorrelation-definition}
+\end{eqnarray}
+The basic concept is that $\tau_\mathrm{eq}$ corresponds to the single-exponential decay time for the autocorrelation process that generates samples, so the statistical inefficiency $g$ measures the approximate temporal separation between two effectively uncorrelated samples (where two exponential relaxation times are presumed to be sufficient).
+
+Robust estimation of $C_t$ for $t \sim T$ is difficult due to growth in statistical error, so common estimators of $g$ make use of several additional properties of $C_t$ to provide useful estimates (see \emph{Practical Computation of Statistical Inefficiencies} in~\cite{chodera2016simple} for a detailed discussion).
+\todo[inline]{JDC: Describe recommended statistical inefficiency computation algorithm.}
+We recommend using the robust statistical inefficinecy computation routines available within the MBAR~\cite{kylebeauchamp2019choderalab} and alchemlyb~\cite{daviddotson2019alchemistry} packages.
+
+\paragraph{Subsampling data to generate uncorrelated samples}
+Once the statistical inefficiency $g$ has been estimated, it is straightforward to subsample the correlated timeseries simulation data to produce effectively uncorrelated data that can be fed to estimators of the free energy.
+Suppose the correlated timeseries is $\{a_t\}_{t=1}^T$; we can form a new timeseries of $N_{\mathrm{eff}} \approx T / g$ effectively uncorrelated samples by selecting a subset of indices $\{ \: t = \mathrm{round}((n-1) \, g) \: | \: n \in \mathrm{range}(1,\ldots ,N) \: \}$ where $\mathrm{round(x)}$ denotes rounding to the nearest integer.
+
+If independent simulations are used, the alchemical state $\lambda$ may have a significant impact on the correlation time, and these simulations should be subsampled independently using a separate estimate of the statistical inefficinecy $g$ for each alchemical state.
+If coupled simulations are used (such as a Hamiltonian replica exchange simulation), the replicas should undergo equivalent random walks in alchemical space, and the replicas can be can be subsampled with the same $g$ to generate an equal number of uncorrelated samples at each alchemical state.
+Conveniently, the approach described above for automated equilibration detection produces an appropriate estimate of $g$ over the production region for automating this process.
+
+\paragraph{Cautions and considerations}
+Reliable estimation of the statistical inefficiency is difficult, and estimates will not generally be as precise (in a relative error sense) as averages.
+To ensure there is sufficient data available for reliable decorrelation and estimation of free energy differences, it is recommended that the effective number of uncorrelated samples $N_{\mathrm{eff}} \ge \approx 50$ if the BAR or MBAR estimators (discussed below) are used; the number may need to be much higher with alternate estimators.
+
+\subsection{Estimators for free energy differences}
+
+Free energy differences between two different states differing in the energy function are directly related to the
+ratio of probabilities of those states.
+As can be noted, the partition functions in Eq.~\ref{equation:dimensionless-free-energy-difference} are simply the total accumulated probabilites for all possible configurations of the system.
+Virtually all of the ways to estimate this free energy are based in converting this ratio of integrals to something that can be measured in one (or several) simulations.
+
+\paragraph{The Zwanzig relationship (EXP)}
+
+The simplest method for calculating free energy
+differences from simulations is the so-called \textit{Zwanzig
+relationship}~\cite{zwanzig1954hightemperature}, also called one-sided exponential reweighting (EXP), or simply free energy perturbation, thought this final term is sometimes used to encompass all ways of calculating free energy differences.
+
+The (reduced) free energy difference $\Delta f_{01}$ between an initial state 0 and a final state 1 defined by two different potential energy functions 
+$U_0(\vec{q})$ and $U_1(\vec{q})$ over coordinate space $\vec{q}$ can be calculated as:
+\begin{eqnarray}
+\Delta f_{01} & = & \ln \expect{e^{-(U_1(\vec{q}) - u_0(\vec{q}))}}_0 =  \ln \expect{e^{-\Delta u(\vec{q})} }_0
+\end{eqnarray}\label{eqn.zwanzig}
+and the average is over all samples from the simulation performed with $u_0$.
+In the case of NVT (canonical) sampling and assuming the masses do not change, then $u$ is just $U/k_BT$, and $f$ is $A/k_BT$, but it can be generalized to other ensembles with the proper definition of $f$ and $u$.
+Described in words, we take the samples generated during our run with the potential energy function $U_0(\vec{q})$, recalculate what the energy difference would be if we switched to potential energy function $U_1(\vec{q})$, and average the exponential of the -1 times the energy difference to get -1 times the exponential of the free energy difference.
+This almost requires no extra code functionality to perform; one need only to save a full precision trajectory, and run an unmodified molecular simulation code using the $U_1$ in order to calculate the new energies of stored snapshots.
+The analysis can be written in a line of code.
+We note that this method is even more general, in that the instantaneous work to change the potential energy function from $U_0$ to $U_1$ can be replaced by the non-reversible work $W$ to make the same change under the same equilibrium conditions at either end state~\cite{jarzynski1997nonequilibrium,jarzynski1998equilibrium,crooks2000pathensemble}, though we do not go into all of the details of nonequilibrum transformations here, and refer the reader to more advanced treatments~\cite{maragakis2008bayesian,oberhofer2005biased,procacci2015unbiased,shirts2003equilibriuma,ytreberg2004singleensemble}
+
+Although the Zwanzig equation is formally correct (as long as the two states considered sample the same phase space volume, which is true for standard molecular models), it has some very important numerical issues that mean that it often performs badly for standard free energy calculations, even for small molecules.~\cite{shirts2005comparison,lu2003appropriate}  One can show that if the standard deviation of the difference $\Delta U(\vec{q}) = U_1(vec{q})-U_2(\vec{q})$ over all sampled $\vec{q}$ is large (which in this case, means only several times $k_BT)$, then very few samples contribute to the average, and the answer will be both biased and extremely noisy~\cite{xx}.
+Essentially, the method is dominated by contributions of rare snapshots\cite{jarzynski2006rare, wu2005phasespaceb, wu2005phasespacec}.
+There are a number of applications where it this expression is nevertheless useful, but those usually involve investigating the dependence of the free energy on small changes in parameters, not changing molecules or even single atoms in molecules~\cite{xx}
+
+\paragraph{The Bennett Acceptance Ratio (BAR)}
+
+If we have the differences in the potential energy sampled from the distribution defined by $U_0$ to the state defined by $U_1$, and we also have the differences in potential energies from the distribution sampled by $U_1$ to the state defined by $U_0$, we can obtain a significantly improved estimate of the
+free energy difference compared to that obtained by EXP.  
+This estimate was first derived by Bennett and is hence generally called the Bennett Acceptance Ratio (BAR).
+It is solved by finding the reduced free energy $f_{ij}$ that satisfied the following implicit equation:
+\begin{eqnarray}
+ \sum_{i=1}^{n_i} \frac{1}{1 + \exp[\ln(\frac{n_i}{n_j}) + u_{ij}(\vec{q}) - f_{ij})
+ ]} \nonumber \\
+ =\sum_{i=1}^{n_j} \frac{1}{1 + \exp[\ln(\frac{n_i}{n_j}) - U_{ij}(\vec{q}) + f_ij)]}
+\end{eqnarray}
+where $n_i$ and $n_j$ are the number of samples from each state. We
+will refer to this method as BAR. More recent derivations show that this formula is the maximum likelihood estimate of the free energy difference
+given sets of samples from the two states~\cite{shirts.bennett}.
+
+Many studies have demonstrated both the
+theoretical and practical superiority of BAR over EXP in molecular
+simulations~\cite{shirts2005comparison,lu2003appropriate}, and BAR converges to EXP in the limit that all samples are from a
+single state~\cite{bennett1976efficient,shirts.bennett}. BAR also requires significantly less
+overlap between the configurational space of each state to converge than EXP, though some overlap must still exist.
+
+The Bennett acceptance ratio is only defined between two states.
+Usually, the endpoints of interest in a free energy calculation are sufficiently different that we will need a chain of states that gradually change the potential energy function from $U_0$to $U_1$, as discussed in Section~\ref{}.
+One can simply carry out BAR between each pair of states $\Delta G_{1 \rightarrow N} = \Delta {G_{1\rightarrow 2}} + \Delta {G_{2\rightarrow 3}} +  \ldots + \Delta G_{N-1\rightarrow N}$.
+
+There is one important thing to note about the uncertainty estimates when summing multiple free energies together to calculate an overall free energy estimate.
+Although BAR itself gives a free energy estimate that is asymptotically correct in and is much less biased than the uncertainty estimate for EXP, the uncertainties in $\Delta {G_{i-1\rightarrow i}}$ and $\Delta {G_{i\rightarrow i+1}}$ are not uncorrelated, because they both involve the energies $U_i(\vec{q})$.
+The variances of each of the free energies will \textit{not} propagate as variances usually do (in quadrature) to the variance of the overall free energy.
+Instead, some other method for propagating the uncertainty, such as bootstrapping~\cite{Grossfield:L:2018} must be used.
+
+\paragraph{Thermodynamic integration (TI)}
+
+By taking the derivative of the free energy with respect to the
+variable $\lambda$, we find that:
+\begin{equation}
+\frac{df}{d\lambda} = \frac{d}{d\lambda} \ln \int \exp{-u(\lambda,\vec{q})} d\vec{q} = \expect{\frac{du(\lambda,\vec{q})}{d\lambda}\
+}_{\lambda} 
+\end{equation}
+
+And than we can then numerically integrate $df/d\lambda$ over an alchemical transformation, using a range of different well-established techniques, to obtain:
+\begin{equation}
+\Delta f    = \int_{0}^{1} \expect{\frac{du(\lambda,\vec{q})}{d\lambda}}_{\lambda}  d\lambda    
+\end{equation}
+This approach to calculating the free energy is called thermodynamic integration (TI). Averaging over $\expect{\frac{du}{d\lambda}}$ requires
+fewer uncorrelated samples to reach a given level of relative error
+than averaging $e^{-u(\vec{q})}$, as the range is usually
+narrower with a more Gaussian distribution.
+Rather than being limited by overlap, as in the case of BAR and MBAR, we are instead limited by the bias in the numerical quadrature, which must be minimized sufficiently to be beneath the level of statistical noise.
+
+Various numerical integration schemes are possible, but the trapezoid
+rule provides a simple and robust scheme.  All types of
+numerical integration can be written as:
+\[ \Delta f \approx \sum_{k=1}^{K} w_k
+\expect{\frac{du(\lambda,\vec{q})}{d\lambda}}_{k} \] where the weights
+$w_k$ correspond to a particular choice of numerical integration.
+Researchers have tried a large number of different integration
+schemes~\cite{resat1993studies,jorge2010effect,shyu2009reducing}. However, many other integration choices require specific choices of $\lambda$
+to minimize bias, which makes them unsuitable when the intermediates
+have widely-varying levels of uncertainty.
+For example, integrating a cubic spline interpolation provided negligble benefits~\cite{paliwal2011benchmark}.
+For starting researchers, we therefore recommend a simple trapezoid rule scheme, as it allows for maximal flexibility in which values of $\lambda$ are simulated.
+Because fitting to higher
+order polynomials can have numerical instabilities for some energy functions, and because alternate functional forms might only be appropriate
+with some types of transformations, expertise and experience is
+required to perform such numerical integration modifications.
+In practice, adding 2-3 more intermediate states is typically sufficient to match the performance of these more complicated numerical quadrature schemes.
+
+One drawback of TI is that it requires derivatives with respect to $\lambda$ to be calculated directly in the code. Unfortunately, many problems of interest
+require using pathways (such as the soft-core pathways, for removing repulsive interactions) that are not linear, as we discuss, making this more complex.
+Still, if the code of interest does compute $\frac{du}{d\lambda}$, then TI is perhaps the simplest method to use, as it involves a very little post-processing,
+
+\paragraph{The multistate Bennett acceptance ratio (MBAR)}
+
+One can generalize Bennett's logic from two states to multiple states to obtain a free energy estimator that uses energy differences between all intermediate states.
+MBAR gives a system of implicit equations for the free energies $f_i$:
+\begin{equation}
+f_i = - \ln \sum_{n=1}^{N} \frac{\exp(-u_i(\vec{q}_n))}{\sum_{k=1}^K N_k \exp(f_k-u_k(\vec{q}_n))}
+\end{equation}
+where there are $N_k$ samples from each of $K$ states, with $\sum_k N_k=N$ the total number of samples.
+Thus, we need to evaluate the energy function $U_i$ for all samples obtained at all states in the transformation.
+The equations can be solved by a number of different standard routines.
+We note that there are only $K-1$ independent equations, so only K-1 of the free energies are independent variables, and one of the $f_i$ must be specified (usually, without loss of generality, setting it to zero).
+
+MBAR is provably the lowest variance asymptotically unbiased estimator of the free energies given the energies of the samples~\cite{tan2004likelihood}, which means thaht BAR is also for the free energy difference between only 2 states, as it is mathematically exactly the same as MBAR i this case.
+MBAR also provides an uncertainty estimate, derived from standard error propagation methods, which has been shown to be accurate as long as there are sufficient samples at each state~\cite{paliwal2011benchmark}.
+
+MBAR can also be thought of as the Zwanzig estimator of the free energy to state $i$ where the sampled distribution is the \textit{mixture distribution} of all the other samples thrown together in one ``pot'', defined by $p_m(\vec{q}) = N^{-1} \sum_k  N_k \exp(f_k-u_k\vec{q})$, the weighted average of all the individual normalized probability distributions that are performed.~\cite{Shirts:AC:2017}.
+
+\paragraph{Recommendations}
+
+\begin{itemize}
+\item We recommend MBAR if all energy differences are available. It is provably the lowest variance free energy estimate given samples from multiple states.
+\item BAR is essentially just as good as MBAR for highly optimized $\lambda$ intermediates.
+Specifically, if the $\lambda$s are chosen such that intermediate states have moderate overlap with their neighbors (i.e.
+between $i$ and $i+1$ and between $i$ and $i-1$, they will \textit{not} have significant overlap with their next nearest neighbors $i+2$ and $i-2$.
+Thus MBAR doesn't actually get significant information from these energy differences, so one might as well not even calculate them, and just perform BAR between nearest neighbors.~\cite{paliwal2011benchmark}
+\item TI usually gives similar values as MBAR implemented with sufficient numbers of intermediates, but quadrature error hard to estimate beforehand.~\cite{paliwal2011benchmark}
+\item WHAM is an approximation to MBAR, and there are no compelling reasons it should be used. 
+\item Other variants, especially ones that adaptively determine the free energies can be  useful in certain circumstances but beyond the scope of a Best Practices article.
+\end{itemize}
+
+\subsubsection{Uncertainty estimation}
+Analysis of independent samples of energy data from a series of equilibrium simulations allows for the estimation of the free energy and error associated with its estimate. 
+Quantification of different error metrics depends on the form of analysis and method being used, such as those outlined above.
+
+The computation of free energies using TI (section \ref{sec:step3}) is straightforward and the trapezoidal rule is often recommended since it allows unequal spacing of lambda states, which is required to minimize the variance in the free energy estimate.
+The determination of regions of high curvature when estimating the integral is helpful to determine regions of phase space where more sampling and/or more lambda states are necessary to obtain the best approximation of the integral.
+Additionally, computation of the overall variance of TI requires the calculation of the overall variance of integration, rather than each individual $\Delta$G$_{i,i+1}$ and assuming variances add independently.
+Therefore, $\mathrm{var}$($\Delta$f) = $\sum_{i=1}^{K}w_{k}^2 \mathrm{var}(\frac{du}{d\lambda})_{k}$.
+
+For alchemical changes that result in smooth, low curvature sets of $\expect{\frac{dU}{d\lambda}}$, a relatively small number of lambda states is necessary for sufficient accuracy and low variance in the free energy estimate.
+However, with increasing numbers of perturbed atoms, the bias introduced by discretization of the integral can become large due to increased curvature, and more lambda intermediate states become necessary to reduce error.
+It is recommended that researchers verify that a sufficient number of states are included such that the free energy is essentially invariant to the number of lambda intermediate states chosen.
+
+Compared with TI, the MBAR method (section \ref{sec:step3}) discussed above provides uncertainty estimation directly from solving a set of linear equations to compute the variances between all states.
+The number of states and amount of sampling should be optimized to minimize the uncertainty in the MBAR free energy estimate, while balancing other key considerations such as computational expense.
+
+In some cases, several analysis methods can be employed to analyze the same set of simulations, providing an opportunity for synergy.
+Specifically, agreement in the free energy across different estimators can be compared and discrepancy between estimates used to assess convergence.
+For example, differences between free energy estimates computed with BAR, MBAR, and TI correlated with larger errors in a set of MCL1 inhibitors.
+Those perturbations where free energies across different estimators shared high agreement with each other (variance less than 0.1 kcal/mol) had a r$^2$ of 0.623 and mean unsigned error (MUE) of 0.787 kcal/mol (N=17) compared with a r$^2$ of 0.295 and MUE of 1.49 kcal/mol (N=23) (unpublished results).
+
+Uncertainty can also be assessed for a particular perturbation by repeating calculations with slight changes in initial configurations, forcefield parameters, and different random seeds in the MD engine.
+The assessment of variability in free energy calculations due to repeating simulations has been previously reported~\cite{aldeghi2019accurate,paliwal2011benchmark}, and large variance in free energies estimated from simulations with different random seeds should be flagged as issues with convergence or problems with system set up.
+System set up errors include poor ligand or protein preparation, atom mapping, or forcefield parameters as described above.
+
+For relative binding free energy calculations, additional sensitivity analysis can be performed by changing the initial configurations of non-core regions of the perturbation topology and determining if this change in configurations results in a large differences in the computed relative free energy, indicating poor sampling of ligand configuration.
+The proposed changes in configuration are increasingly relevant if no experimental evidence is available to reduce uncertainty in where the changing atoms should be positioned.
+
+In addition to statistical uncertainty and sampling, a variety of other factors can impact results from binding free energy calculations.
+In addition to the choice of initial configuration, results can depend on the choice of force field for the protein/receptor, water, and small molecuel(s), so rerunning calculations with different choices of force field can also be used to assess how sensitive results and conclusions are to these particular choices.
+Other factors, like system preparation (choice of protonation state, tautomer, counterion presence, salt concentration, etc.) can also substantially impact results~\cite{mobley2017predicting, mobley2017predictingb}, so unless modelers are confident they have these factors correct, sensitivity to these choices may also need to be examined.
+
+\subsubsection{Are my simulations any good?}
+\label{sec:are-they-good}
+One way of assessing reliability of the calculations is checking the phase space overlap between neighboring $\lambda$-windows~\cite{wu2005phasespace,wu2005phasespacea}.
+For this purpose, a so-called overlap matrix can be used.
+This is a $K\times K$ matrix, where $K$ is the number of simulated states, i.e.
+$\lambda$s.
+Sufficient overlap is important for reweighting estimators such as BAR or MBAR, but cannot help assess reliability of estimates, when using TI.
+These matrices are graphical representations of the phase space overlap, i.e.
+the average probability that a sample generated at state $j$ can be observed at state $i$.
+As this probability is computed considering the samples from all states (and not just the adjacent states), the values in each row and column add up to 1.
+In this analysis, the goal is to ensure every state has overlap with its neighbors in both directions -- so that off-diagonal elements are significantly larger than zero.
+Usually this means the matrix should be at least tridiagonal.
+
+Details on the calculation and properties of these matrices can be found elsewhere~\cite{klimovich2015guidelines}.
+In an overlap matrix $\mathbf{O}$, the off-diagonal values (${O}_{i,j,i\ne j}$) are negatively correlated with the variance of the free energy difference.
+Accordingly, the uncertainty of the free energy difference between the states i and j will be smaller when ${O}_{i,j,i\ne j}$ is larger (and thus the values in the main diagonal (${O}_{i,j,i=j}$) are smaller).
+In order to obtain a reliable estimate of the free energy all neighbouring states must be connected, i.e.
+there must be sufficient overlap between the samples of these states (general description: ${O}_{i,j,i\ne j}\ge$ threshold).
+However, due to the mathematical derivation it is difficult to explicitly describe the relation of the overlap matrix and the variance by formulae.
+Consequently, the threshold has to be derived empirically.
+It has been proposed that the values of the first off-diagonals (i.e.
+the diagonals above and below the main diagonal) should at least be 0.03 to obtain a reliable free energy estimate~\cite{klimovich2015guidelines}.
+Smaller values should be considered as a warning sign (see Figure~\ref{fig:overlap}), as the variance tends to be underestimated in case of poor overlap.
+
+\begin{figure}
+\includegraphics[width=0.95\columnwidth]{figures/fig12_overlap/Figure.pdf}
+\caption{\label{fig:overlap}Overlap matrices showing (A) good, (B) mediocre, and (C) poor overlap.}
+\end{figure}
+
+Figure~\ref{fig:overlap} shows examples of good, mediocre and poor overlap in ~\ref{fig:overlap} A, B, and C respectively.
+For (A), the probability to find a sample from state $i$ in its neighbouring state $j$ is about 0.2 for all states adjacent to the main diagonal, and hence the overall connectivity is good.
+In the case of (B), the overlap is strongly diminishing in the lower right corner, raising concerns regarding the reliability of the free energy estimate obtained.
+For (C), the state at $\lambda$ index = 6 is connected to neither of its neighbouring states.
+While this does not necessarily imply that the result for this perturbation is wrong, the energy estimate must at least be considered as highly unreliable.
+In order to overcome the issue of poor overlap in this example, additional sampling should be performed by introducing additional states, i.e. $\lambda$ windows.
+
+Interestingly, as the variance is inversely correlated with the number of states\cite{klimovich2015guidelines}, it can in principle be reduced below any arbitrary threshold with enough simulation time and a large enough number of lambda windows.
+However, decreasing the variance to a value close to 0 is not feasible, as this approach would strongly increase the calculation time.
+While variance can be decreased both by increasing simulation length and by increasing the number of lambda values, if overlap is poor, the latter approach often has a larger immediate impact.
+Specifically, often the goal is to adjust the $\lambda$ spacing to perform additional sampling in the areas of poor overlap, rather than using arbitrarily spaced windows.
+Different approaches are described in section \ref{part_of_section_six}) and more details can be found in the literature~\cite{dakka2018concurrent, hahn2019alchemical}.
+
+On a final note, it has been argued that overlap matrices are only a qualitative estimator for the spacing of windows.
+Another estimator named Time Derivative of overall Variance (TDV)~\cite{sun2017barbased} has been suggested for this task, which is an explicit function of the overall variance.
+However, it could not be proven that better suggestions for the $\lambda$ spacing can be obtained by this approach compared to overlap matrices.
+Hence, it should rather be seen as a complementary approach.
+
+In a nutshell, it must be ensured that the values of an overlap matrix in the first diagonals above and below the main diagonal are at least 0.03 in order to yield reliable results.
+
+There are several other parameters which should be investigated.
+Basic analysis of the trajectories should be performed to ensure that the ligand does not tumble out of the binding site by measuring its RMSD.
+Furthermore, distributions of the ligand's dihedral angles can be obtained and compared to experimental values.
+Thus, poorly sampled conformations can be identified and the issue can be adequately addressed, e.g.
+by prolonging the simulation time or choosing a different starting conformation.
+In general, running the same perturbation using different ligand poses may offer insights into whether the calculated free energies are dependent on the initial conformation.
+Another easy check can be performed by comparing the results obtained by different free energy estimators (e.g.
+TI and MBAR).
+In principal, the outcomes of both estimators should be the same (within a certain range).
+Hence, if the outcomes deviate by more than a predefined threshold (e.g.
+1.0 kcal/mol), the free energies may be unreliable and further calculations are necessary.
+\todo[inline]{MK: @ASJSM: Please review and add a few sentences on replica mixing and correlation time}
+\begin{itemize}
+\item Examining output data for common problems with discussions of what exactly to plot or look at; examples of typical curves for dV/dlambda and free energy versus lambda, for example
+\begin{itemize}
+\item Make sure ligand doesn’t tumble out of binding site (Mey has observed this)
+\item Significant discrepancies between different free energy estimators (TI, BAR, MBAR)
+\item Poor replica mixing (for replica-exchange)
+\item Correlation time as a function of lambda as it would be expected to be a smooth
+\item Dependence on initial conformation
+\item Torsional analysis: Is it stuck in specific states? Only very rarely transitions?
+\item More “usual suspects”
+\end{itemize}
+\item Other considerations for many transformations
+\end{itemize}
+
+\begin{figure}
+    \includegraphics[width=0.95\linewidth]{figures/fig9_convergence/Figure.pdf}
+    \caption{Average binding free energy of 5 replicate Hamiltonian replica exchange calculations as a function of total simulation time (i.e.
+the sum of the simulation time of all replicas) for the two host-guest systems CB8-G3 and OA-G3.
+Shaded areas represent 95\% confidence intervals around the mean computed from the 5 replicates data.
+The horizontal dashed lines show the final binding free energy prediction of the two calculations after a total of ~5230 ns for OA-G3 and ~6650 ns for CB8-G3.
+Longer correlation times in CB8-G3 cause the calculation to converge more slowly.
+The original data used to generate the plot can be found at \url{https://github.com/MobleyLab/SAMPL6/blob/master/host_guest/Analysis/SAMPLing/Data/reference_free_energies.csv}.
+}
+    \label{fig:freeenergytrajectories}
+\end{figure}
+
+\paragraph{Cycle closure error}
+Relative free energy calculations, which compute the change in free energy on making a change to a molecule (e.g.
+adding a functional group to a ligand) may provide an additional opportunity for error/consistency checking.
+Particularly, such calculations are often done to span a graph or tree of free energy calculations~\cite{xu2019diffnet,wang2013modeling,liu2013lead} and in some cases the free energy change to go between molecules A and B can be obtained via multiple transformation pathways.
+This allows a type of consistency checking where we assess how much the free energy change for that transformation in practice differs from equivalence.
+Significant deviations from this typically indicate insufficient configurational sampling along the lambda schedule of one or more of the transformations involved.
+
+This approach may be generalised to sets of connected transformations given the requirement that the sum of free energy changes along edges of a closed cycle should be zero.
+This analysis is called ``cycle closure''.
+In practice, such thermodynamic cycles do not actually sum to zero,  and deviations become increasingly large as the size of the cycle increases owing to propagation of error.
+Though no firm guidelines have emerged, it may be judicious to perform additional configurational sampling along edges of a network that are involved in cycles closing poorly.
+This may be done by extending simulations duration, or by averaging free energy changes over multiple repeats.
+The latter approach may yield more reproducible free energy changes, but at the expense of a stronger bias on the estimated free energies due to repeated use of the same input coordinates.
+
+A scheme to reduce cycle closure errors is used in FEP+ whereby calculated free energy changes along the nodes of the network are re-sampled assuming estimates of the calculated free energy change along a node may be obtained from a Gaussian distribution centered on the estimated free energy change and with a standard deviation equal to the estimated standard deviation of the free energy change.
+The procedure then uses a Maximum likelihood method to find new sets of free energy changes that minimize cycle closure errors~\cite{wang2013modeling}.
+An alternative approach computes the free energy change between a target and reference compound as a weighted average over all unique paths in the network, with the weights derived from the propagated uncertainties of each node (REF).
+
+\paragraph{Reversible binding simulations}
+An even more stringent test of the correctness of binding free energy calculations is to compare the results to the equilibrium binding constants derived from long timescale reversible binding simulations~\cite{pan2017quantitative}.
+For small ligands with millimolar affinities, repeated binding to and unbinding from the protein can occur for a large number of times in a sufficiently long unbiased MD simulation (10-100$\mu$s), and the equilibrium binding constants can be straightforwardly computed from the ratio of bound to unbound fractions of the simulation time.
+The agreement between the binding free energy calculations and the reversible binding simulations--given the same system preparation and the same force field parameters--will strongly support the correctness of both calculations, as the same results are arrived at by two independent methods, and any discrepancy will suggest some systematic error in one, or both, of the two methods.
+This test should be applied to all free energy calculation codes in the future.
+
+\subsection{Best practices for data representation}
+
+Practitioners in the field are encouraged to follow best practices on analysis of results and data depictions.
+The following points should be considered when plotting alchemical free energy prediction results.
+\todo[inline]{JS: @everyone I suspect people will have strong opinions on this subject, feel free to include your own suggestions}
+\paragraph{Statistics to include}
+As with any modelling technique, misuse of statistical analysis can skew the perception of how well models perform in FE predictions.
+First and foremost, error estimates should always be included on your predictions in whatever form you present your data (scatterplots, barplots, etc; see next paragraph).
+It is recommended to perform triplicates of your predictions at minimum to ensure some measure of reliability in your data, in part because analytical uncertainty estimates often underestimate the true statistical uncertainty.
+
+Because of the strong drug discovery context and the relevance of alchemical FE predictions to SAR projects, both the overall correlation of your predictions relative to experiment and the ability of your models to detect the best-scoring compounds (i.e.
+rankings) should be computed.
+Conventionally, this means including an R\textsuperscript{2} (or Pearson r) and a Kendall \texttau{} metric in your results.
+Additionally, practitioners may choose to include a Spearman \textrho{} as well.
+Practitioners should take particular care with correlation based metrics, however, when  the dataset dynamic range is small, or only a small number of compounds appear at extrema of the dynamic range.
+Bootstrapping analysis -- including variation in the dataset -- can help with this by providing useful error bars on performance metrics.
+
+Mean unsigned error (MUE, also called mean absolute error/MAE) is the another key statistic to include in your results.
+Even though some models' near-perfect correlation and ranking statistics might suggest excellent accuracy, MUE values can still be multiple kcal/mol, providing important additional insight into performance.
+Furthermore, MUE allows for unbiased comparisons between predictive models as it is less sensitive to dataset size.
+Even though this section may serve as an introduction to the main statistical practices, the authors recommend further reading on evaluation of computational models \cite{cite Nicholls pt1 and pt2 + Jain&Nicholls "Recommendations for.."}.
+
+\paragraph{Presenting your data}
+Seeing as essentially all alchemical FE prediction schemes are regression problems, the preferred type of plot is a scatter plot (see Figure \ref{fig:scatterplot_analysis}).
+Most alchemical FE projects will concern 10-50 ligands; any study with \textless10 ligands is more suitable for bar plots (with inclusion of error bars).
+Any study with \textgreater50 ligands should be trellised to not overestimate statistics.
+In that respect, it is bad practice to place multiple datasets on the same plot because this can suggest high model accuracy even though the individual models perform less well \cite{cite blogpost Pat Walters}.
+
+Because we are interested mainly in the linear relationship between the alchemical FE predictions and the experimentally-determined affinity values, plots should be depicted with the same range on both axes (i.e.
+$x=y$).
+Furthermore, bounds should be depicted for the 1- and 2-kcal/mol confidence regions.
+These regions can serve as tools to communicate your model performance: any perturbation inside the 1 kcal/mol region can be seen as highly reliable, any perturbation inside the 2 kcal/mol region should be seen as somewhat reliable, and any perturbation outside the confidence regions should be expected to be unreliable (and handled as outliers).
+In a drug discovery context, this type of data depiction may suggest the reliability of alchemical FE predictions in the project, and can give an idea of how trustworthy predictions can be for synthesis ideas.
+\begin{figure}
+    \includegraphics[width=0.95\linewidth]{figures/fig13_analysis_practices/Figure.pdf}
+    \caption{Example of a well-depicted alchemical free energy prediction showing the relation between predicted and experimentally-determined Gibbs free energy in kcal per mol with standard errors as error bars.
+The two outliers (underestimations) are shown in red; the dark and light-orange regions depict the 1- and 2-kcal per mol confidence bounds.}
+    \label{fig:scatterplot_analysis}
+\end{figure}
+
+\section{Conclusions}
+\todo[inline, color={green!20}]{ASJSM:Write conclusions}
+%%%%%%%%%%%%%%%%%%%%
+%              Terminology                  %
+%%%%%%%%%%%%%%%%%%%%
+\section{Terminology and abbreviations}
+\label{sec:tem-abbrev}
+\begin{itemize}
+\item Feature Box covering major technical terms and abbreviations
+\item Examples:
+\begin{itemize}
+\item EXP, BAR, MBAR
+\item Double decoupling, single-topology, dual-topology, hybrid-topology, coupled-topology
+\item FEP (free energy perturbation), alchemical, AFE (alchemical free energy)
+\end{itemize}
+\end{itemize}
+\todo[inline]{JS: should this be combined with the acronyms section at the beginning of the paper?}
+%%%%%%%%%%%%%%%%%%%%
+%                Software                    %
+%%%%%%%%%%%%%%%%%%%%
+\section{Available software -- a summary}
+\todo[inline]{ASJSM: Expand and add more references. }
+\label{sec:software}
+There are several softwares availble for carrying out alchemical free energy calculations. In this section we discuss some of the popular commercial and free tools along with their features. 
+\begin{itemize}
+\item Commercial:
+   \begin{itemize}
+    \item FEP+ is a tool offered by Schr\"{o}dinger Inc.
+under a commercial license.
+It has an intuitive GUI which makes it easier for non-experts to run alchemical free energy calculations and analyze the results.
+It runs DESMOND MD package under the hood and hence parallelizes very well on the GPUs.
+    \item Flare is a commercial structure-based drug design software offered by Cresset.
+Similar to FEP+ it has an easily accessible user interface and strives to facilitate free energy calculations for non-experts while offering advanced users full control via a Python API.
+It only runs on GPUs, using CUDA or OpenCL.
+    \item The molecular operating environment (MOE) offered by the Chemical Computing Group (CCG) has a tool for performing free energy calculations. It is build on AMBER-TI (cf. below).
+    \end{itemize}
+\item Free or low-cost for academics / commercial for industry:
+	\begin{itemize}
+	\item CHARMM has a variety of tools developed over the years.
+PERT module can be used to define initial and final states and define the intermediate lambda points.
+FREN and BAR modules can be used to analyze the data after the MD run.
+Lambda-dynamics based free energy calulation can be carried out using the BLOCK module.
+	\item TIES and AMBER FEW? (Peter Coveney)
+	\item AMBER, including its new pmemd.cuda version support free energy calculation. 
+	\end{itemize}
+\item Free (libre) open source:
+	\begin{itemize}
+	\item PLUMED is an open source tool which enables the usage of a variety of MD engines.
+It is designed as a plugin for MD packages such that it analyzes the trajectory on the fly.
+It also offers a VMD based plugin for the computation of collective variables.
+	\item SIRE  is a free, open source, multiscale molecular simulation framework, written to allow computational modellers to quickly prototype and develop new algorithms for molecular simulation and molecular design.
+	\item YANK is a tool developed by John Chodera and group on the top of OpenMM MD package. It allows the users to write their inputs in easy-to-use YAML format.
+	\item GROMACS is a molecular simulation package with a significant number of free energy methods implementations.
+The LiveCOMS GROMACS tutorial has an example free energy calculation~\cite{lemkul2018From}.
+	\item pmx for mutations
+	\end{itemize}
+\item Setup tools
+	\begin{itemize}
+	\item FESetup: AMBER, gromacs, Sire
+	\item Lomap/Lomap2 : Relative alchemical transformation graph planning
+	\item CHARMM-GUI is a web based tool for setting up a variety of MD simulations. It can be used to generate CHARMM scripts for solvation and ligand-binding free energy calculations.
+	\end{itemize}
+\item Analysis tools:
+	\begin{itemize}
+	\item Free Energy Workflows: Sire-specific free energy map analysis using weighted path averages
+	\url{https://github.com/michellab/freenrgworkflows}
+	\item Alchemlyb: Multipackage free energy analysis
+	\url{https://github.com/alchemistry/alchemlyb}
+	\item pymbar: MBAR implementation, but have to roll your own analysis wrapper
+	\url{https://github.com/choderalab/pymbar}
+	\end{itemize}
+\end{itemize}
+
+\section{Online resources}
+\begin{itemize}
+\item \url{http://www.ks.uiuc.edu/Training/Workshop/Urbana_2010A/lectures/TCBG-2010.pdf}
+\item Basic Ingredients of Free Energy Calculations: A Review (\url{DOI: 10.1002/jcc.21450})
+\item Good Practices in Free-Energy Calculations (\url{DOI: 10.1021/jp102971x})
+\item Alchemical Free Energy Methods for Drug Discovery: Progress and Challenges (\url{doi: 10.1016/j.sbi.2011.01.011})
+\item Alchemistry wiki: \url{http://www.alchemistry.org/wiki/Best_Practices}
+\end{itemize}
+
+\clearpage
+\section{Benchmark Datasets}
+
+\label{sec:benchmark}
+\vspace{5mm}
+
+\begin{table}
+\caption{Overview of benchmark datasets}
+\begin{tabular}{cccc}
+
+\textbf{Publication} & \textbf{Targets} & \textbf{Ligands} & \textbf{Force Field} \\
+\hline
+\multicolumn{4}{|c|}{D3R Grand Challenges~\cite{D3R}} \\
+\hline
+GC3~\cite{gaieb2019d3r} & 6 & 266 & various \\
+GC2~\cite{gaieb2018d3r} & 1 & 102 & various \\
+GC2015~\cite{gathiaka2016d3r} & 2 & 215 & various \\
+\hline
+\multicolumn{4}{|c|}{SAMPL Challenges~\cite{SAMPL}} \\
+\hline
+SAMPL6~\cite{rizzi2018overview} & 3 & 21 & various \\
+SAMPL5~\cite{yin2017overview} & 3 & 22 & various \\
+SAMPL4~\cite{muddana2014sampl4} & 2 & 23 & various \\
+\hline
+\multicolumn{4}{|c|}{Schrödinger Datasets} \\
+\hline
+FEP+ Dataset~\cite{wang2015accurate} & 8 & 199 & OPLS2.1 \\
+FEP+ Dataset~\cite{harder2016opls3} & 8 & 199 & OPLS3 \\
+FEP+ Dataset~\cite{roos2019opls3e} & 8 & 199 & OPLS3e \\
+FEP+ Dataset~\cite{song2019using} & 8 & 199 & AMBER18 \\
+Fragment Optimization~\cite{steinbrecher2015accurate} & 8 & 96 & OPLS2.1 \\
+Scaffold Hopping~\cite{wang2017accurate} & 6 & 21 & OPLS3 \\
+Macrocycles~\cite{yu2017accurate} & 7 & 33 & OPLS3 \\
+\hline
+\multicolumn{4}{|c|}{Further Suggested Datasets} \\
+\hline
+Cucurbit[7]uril (CB7)~\cite{mobley2017predicting} & 1 & 15 & NA \\
+Deep cavity cavitand~\cite{mobley2017predicting} & 2 & 19 & NA \\
+T4 Lysozyme~\cite{mobley2017predicting} & 2 & 20 & NA \\
+Merck benchmarking set~\cite{MCompChem2019Sep} & 5 & 169 & OPSL3 \\
+
+\hline
+\label{tab:benchmarks}
+\end{tabular}
+\end{table}
+
+Several other case studies of FEP calculations can be found in the review by Williams-Noonan et al.~\cite{williams-noonan2018free}.
+For comparison of FEP+ and Gromacs (using the AMBER99SB-ILDN and GAFF2 force field), cf.
+the recently published study by Pérez-Benito et al.~\cite{perez-benito2019predicting}.
+An overview of further suggested benchmark sets can be found in the review by Mobley and Gilson~\cite{mobley2017predicting} or on \url{alchemistry.org}~\cite{alchemistry}.
+These include cyclodextrins, the Cytochrome C peroxidase (CCP) protein model binding site, thrombin and bromodomains as well as solvation benchmark sets~\cite{paliwal2011benchmark}.
+Further work on shared protein-ligand benchmark sets is ongoing at \url{https://github.com/openforcefield/PLBenchmarks}.
 
 \section{Checklist}
-Here we use a full-page checklist with multiple sections, so it will appear on a separate page of the sample PDF.
-Other checklist formats are possible, as shown in the sample \texttt{sample-document.tex} in \url{github.com/livecomsjournal/article_templates/templates}.
-
-Your checklist should include a succinct list of steps that people should follow when carrying out the task in question.
-This is provided to ensure certain basic standards are followed and common but critical major errors are avoided.
-Note that a checklist is not intended to cover \emph{all} important steps, but rather focus on the most common reasons for failure or incorrect results, or issues which are particularly crucial.
+\label{sec:checklist}
+\todo[inline, color={green!20}]{JS: @everyone Please input your own suggestions, I made some points off the top of my head. Feel free to add helpful citations to any of the bullet points}
+\todo[inline, color={green!20}]{JS: @self,ASJSM do some final formatting when text is complete.}
 
@@
 % can easily print it without also printing other material from your manuscript. However, other styles of checklist are also possible (below).
-\begin{Checklists*}[p!]
+% \begin{Checklists*}[p!]
 
-\begin{checklist}{First list}
-\textbf{You can easily make full width checklists that take a whole page}
-\begin{itemize}
-\item Items in your checklist can and should reference sections where the cited issues are discussed in detail, such as Section~\ref{sec:reference_this}
-\item Also remember
-\item And finally
-\end{itemize}
-\end{checklist}
+% \begin{checklist}{Step 0 -- Know what you want to simulate }
+% \textbf{What are the first questions that need addressing before setting up a molecular dynamics simulation}\\
+% Extensive explanation for the checklist questions can be found in Section~\ref{sec:step0}.
+% \begin{itemize}
+% \item Can I get the required accuracy with the simulation I want to carry out?
+% \item Have I properly prepared my protein and ligand systems?
+% \item Does my system contain any groups that require custom parameters?
+% \item What simulation protocol will provide the most evidence to answer my hypothesis?
 
-\begin{checklist}{Second list}
-\textbf{This is some further description.}
-\begin{itemize}
-\item First thing
-\item Also remember
-\item And finally
-\end{itemize}
-\end{checklist}
+% \item And finally
+% \end{itemize}
+% \end{checklist}
 
-\begin{checklist}{Third list}
-\textbf{This is some further description.}
-\begin{itemize}
-\item First thing
-\item Also remember
-\item And finally
-\end{itemize}
-\end{checklist}
+% \begin{checklist}{Simulation preparation}
+% \textbf{How do I get started setting up an alchemical free energy calculation}
+% Extensive explanation for the checklist questions can be found in Section~\ref{sec:step1}.
+% \begin{itemize}
+% \item Have I followed the Best practices for biomolecular simulation set up?
+% \item In a relative simulation, will I run into problems with clashing geometries in the ligand transformation or crystal waters?
+% \end{itemize}
+% \end{checklist}
+% \end{Checklists*}
 
-\begin{checklist}{Fourth list}
-\textbf{This is some further description.}
-\begin{itemize}
-\item First thing
-\item Also remember
-\item And finally
-\end{itemize}
-\end{checklist}
+% \begin{Checklists*}[p!]
+% \begin{checklist}{Absolute simulations}
+% \textbf{What are the main things I need to consider for an absolute alchemical free energy calculation?}
+% Extensive explanation for the checklist questions can be found in Section~\ref{sec:step2}.
+% \begin{itemize}
+% \item Topology
+% \item Restraints
+% \item Standard state handling
+% \end{itemize}
+% \end{checklist}
 
-\end{Checklists*}
+% \begin{checklist}{Relative simulations}
+% \textbf{What are the main things I need to consider for an relative alchemical free energy calculation?}
+% Extensive explanation for the checklist questions can be found in Section~\ref{sec:step2}.
+% \begin{itemize}
+% \item First thing
+% \item Also remember
+% \item And finally
+% \end{itemize}
+% \end{checklist}
 
-\section{Rationale}
+% \begin{checklist}{Analysis}
+% \textbf{This is all about analysis of the simulation}
+% Extensive explanation for the checklist questions can be found in Section~\ref{sec:step4}.
+% \begin{itemize}
+% \item Are my simulations converged enough?
+% \item Am I using the right analysis techniques?
+% \end{itemize}
+% \end{checklist}
 
-Your Rationale section, or sections, can follow or precede your checklist (we expect that often, following the checklist will be preferable) and provide the necessary rationale for the checklist, and act as more complete \emph{best practices} description.
-This should include 1) significant detail as to the possible alternative ways to accomplish a given task, 2) description of advantages and disadvantages of the various approaches, and 3) significant literature documentation about reasons for choices.
+% \end{Checklists*}
 
-\subsection{Algorithms and Pseudocode}
-\label{sec:reference_this}
+%Checklist:
+\clearpage
+%%%%%%%%%%%%%%%% 
+%% Checklist-specific precommands:
 
-The \texttt{algpseudocode} and \texttt{algorithms} packages is loaded by the document class.
-\ALG{euclid} was taken directly from the package documentation.
-(Please do not load \texttt{algorithm2e}; it's not compatible with \texttt{algpseudocde} nor \texttt{algorithms}!)
+\clearpage
 
-\begin{algorithm}
-\caption{Euclid's algorithm}\label{alg:euclid}
-\begin{algorithmic}%[1]  %% uncomment to enable line numbers
-\Procedure{Euclid}{$a,b$}\Comment{The g.c.d. of a and b}
-   \State $r\gets a\bmod b$
-   \While{$r\not=0$}\Comment{We have the answer if r is 0}
-      \State $a\gets b$
-      \State $b\gets r$
-      \State $r\gets a\bmod b$
-   \EndWhile\label{euclidendwhile}
-   \State \textbf{return} $b$\Comment{The gcd is b}
-\EndProcedure
-\end{algorithmic}
-\end{algorithm}
-
-\section{Author Contributions}
+\section*{Author Contributions}
 %%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%
-
+\todo[inline, color={green!20}]{ASJSM: @everyone, can you check your contribution is acknowledge and if not add the appropriate information. }
 (Explain the contributions of the different authors here)
+ASJSM: Coordinated the document, contributed to most sections, and co-designed figures ~\ref{} and created figures \ref{fig:fig_what_is_lambda},~\ref{fig:fig1_what_is_alchemy},~\ref{fig:overlap} and replotted~\ref{fig:automatic-equilibration-detection} and improved ~\ref{fig:fig_types_of_networks}.
+JS: created figures \ref{fig:fig_what_is_alchemy},~\ref{fig:fig_binding_thermodynamic_cycle},~\ref{fig:fig_topology},~\ref{fig:fig_mcss},~\ref{fig:fig_types_of_networks},~\ref{fig:fig_absolute_thermodynamic_cycle},~\ref{fig:scatterplot_analysis}, the data representation section, the checklist and contributed to general formatting discussions and editing.
+MK: Contributed to the Data Analysis section, provided the data for figure~\ref{fig:overlap} and helped edit the paper.
+DLM: Contributed to the outline, drafted some of the sections, gave ideas on figures, and helped edit the paper.
+GT: Contributed to introduction and drug discovery sections, and helped edit the paper.
+AR: Created figure~\ref{fig:freeenergytrajectories}, contributed to section \ref{sec:step2}, and helped edit the paper.
+MRS: Helped create figure~\ref{fig:fig_what_is_lambda}, wrote sections describing choices for alchemical pathways and analysis for free energy calculations.
+LN: \todo[inline, color={red!20}]{@Levi insert info here}
+BA: \todo[inline, color={red!20}]{@Bryce insert info here}
+JM: \todo[inline, color={red!20}]{@Julien insert info here}
+JDC: \todo[inline, color={red!20}]{@JDC insert info here}
+HX: Contributed section~\ref{subsec:accuracy}, to section~\ref{sec:relative-fe-protocol}, and to section~\ref{sec:are-they-good}.
 
@@
 Declare any potentially conflicting interests here, whether or not they pose an actual conflict in your view.
+MK is employed by Cresset who recently released an commercial tool for performing alchemical free energy calculations.
 
-\section{Funding Information}
+\section*{Funding Information}
 %%%%%%%
@@
 %%%%%%%
-FMS acknowledges the support of NSF grant CHE-1111111.
+ASJSM and JM acknowledge funding through an EPSRC flagship software grant: EP/P022138/1
+MK acknowledges funding through Innovate UK by KTP partnership 011120.
+AR acknowledges partial support from the Tri-Institutional Program in Computational Biology and Medicine and the Sloan Kettering Institute.
 
-\bibliography{livecoms-sample}
+\bibliography{alchemical1,manual}
 

### 2020-01-10 paper/manuscript.tex
Updates from Overleaf

@@
 \includegraphics[width=0.95\columnwidth]{figures/fig12_overlap/Figure.pdf}
-\caption{\label{fig:overlap}Overlap matrices showing (A) good, (B) mediocre, and (C) poor overlap.}
+\caption{\label{fig:overlap} \textbf{Overlap matrices:} Visualising overlap matrices can help with assessing the quality of simulation data.
+(A) shows good overlap with all first off-diagonal entries well above 0.03, the suggested threshold, (B) is an example of mediocre overlap with good overlap at lower $\lambda$ values and poor overlap at high $\lambda$ values.
+(C)shows poor overlap resulting in disconnected simulations that should not be used for an MBAR estimation of free energies.}
 \end{figure}
 
-Figure~\ref{fig:overlap} shows examples of good, mediocre and poor overlap in ~\ref{fig:overlap} A, B, and C respectively.
+Figure~\ref{fig:overlap} shows examples of good, mediocre and poor overlap in ~\ref{fig:overlap} (A), (B), and (C) respectively.
 For (A), the probability to find a sample from state $i$ in its neighbouring state $j$ is about 0.2 for all states adjacent to the main diagonal, and hence the overall connectivity is good.
@@
 In order to overcome the issue of poor overlap in this example, additional sampling should be performed by introducing additional states, i.e. $\lambda$ windows.
-
+%
 Interestingly, as the variance is inversely correlated with the number of states\cite{klimovich2015guidelines}, it can in principle be reduced below any arbitrary threshold with enough simulation time and a large enough number of lambda windows.
@@
 Different approaches are described in section \ref{part_of_section_six}) and more details can be found in the literature~\cite{dakka2018concurrent, hahn2019alchemical}.
-
-On a final note, it has been argued that overlap matrices are only a qualitative estimator for the spacing of windows.
-Another estimator named Time Derivative of overall Variance (TDV)~\cite{sun2017barbased} has been suggested for this task, which is an explicit function of the overall variance.
-However, it could not be proven that better suggestions for the $\lambda$ spacing can be obtained by this approach compared to overlap matrices.
-Hence, it should rather be seen as a complementary approach.
-
+%
 In a nutshell, it must be ensured that the values of an overlap matrix in the first diagonals above and below the main diagonal are at least 0.03 in order to yield reliable results. 
-
+%
 There are several other parameters which should be investigated.
@@
 Hence, if the outcomes deviate by more than a predefined threshold (e.g.
-1.0 kcal/mol), the free energies may be unreliable and further calculations are necessary.
-\todo[inline]{MK: @ASJSM: Please review and add a few sentences on replica mixing and correlation time}
+1.0 kcal/mol), the free energies are likely unreliable and further calculations are necessary.
+
+\begin{figure}
+    \includegraphics[width=0.95\linewidth]{figures/fig9_convergence/Figure.pdf}
+    \caption{Average binding free energy of 5 replicate Hamiltonian replica exchange calculations as a function of total simulation time (i.e.
+the sum of the simulation time of all replicas) for the two host-guest systems CB8-G3 and OA-G3.
+Shaded areas represent 95\% confidence intervals around the mean computed from the 5 replicates data.
+The horizontal dashed lines show the final binding free energy prediction of the two calculations after a total of ~5230 ns for OA-G3 and ~6650 ns for CB8-G3.
+Longer correlation times in CB8-G3 cause the calculation to converge more slowly.
+The original data used to generate the plot can be found at \url{https://github.com/MobleyLab/SAMPL6/blob/master/host_guest/Analysis/SAMPLing/Data/reference_free_energies.csv}.
+}
+    \label{fig:freeenergytrajectories}
+\end{figure}
+
+\paragraph{Cycle closure error}
+Relative free energy calculations, which compute the change in free energy on making a change to a molecule (e.g.
+adding a functional group to a ligand) may provide an additional opportunity for error/consistency checking.
+Particularly, such calculations are often done to span a graph or tree of free energy calculations~\cite{xu2019diffnet,wang2013modeling,liu2013lead} and in some cases the free energy change to go between molecules A and B can be obtained via multiple transformation pathways.
+This allows a type of consistency checking where we assess how much the free energy change for that transformation in practice differs from equivalence.
+Significant deviations from this typically indicate insufficient configurational sampling along the lambda schedule of one or more of the transformations involved.
+
+This approach may be generalised to sets of connected transformations given the requirement that the sum of free energy changes along edges of a closed cycle should be zero.
+This analysis is called ``cycle closure''.
+In practice, such thermodynamic cycles do not actually sum to zero,  and deviations become increasingly large as the size of the cycle increases owing to propagation of error.
+Though no firm guidelines have emerged, it may be judicious to perform additional configurational sampling along edges of a network that are involved in cycles closing poorly.
+This may be done by extending simulations duration, or by averaging free energy changes over multiple repeats.
+The latter approach may yield more reproducible free energy changes, but at the expense of a stronger bias on the estimated free energies due to repeated use of the same input coordinates.
+
+A scheme to reduce cycle closure errors is used in FEP+ whereby calculated free energy changes along the nodes of the network are re-sampled assuming estimates of the calculated free energy change along a node may be obtained from a Gaussian distribution centered on the estimated free energy change and with a standard deviation equal to the estimated standard deviation of the free energy change.
+The procedure then uses a Maximum likelihood method to find new sets of free energy changes that minimize cycle closure errors~\cite{wang2013modeling}.
+An alternative approach computes the free energy change between a target and reference compound as a weighted average over all unique paths in the network, with the weights derived from the propagated uncertainties of each node (REF).
+
+\paragraph{Reversible binding simulations}
+An even more stringent test of the correctness of binding free energy calculations is to compare the results to the equilibrium binding constants derived from long timescale reversible binding simulations~\cite{pan2017quantitative}.
+For small ligands with millimolar affinities, repeated binding to and unbinding from the protein can occur for a large number of times in a sufficiently long unbiased MD simulation (10-100$\mu$s), and the equilibrium binding constants can be straightforwardly computed from the ratio of bound to unbound fractions of the simulation time.
+The agreement between the binding free energy calculations and the reversible binding simulations--given the same system preparation and the same force field parameters--will strongly support the correctness of both calculations, as the same results are arrived at by two independent methods, and any discrepancy will suggest some systematic error in one, or both, of the two methods.
+This test should be applied to all free energy calculation codes in the future.
+
+\subsection{Recommendations}
+\todo[inline]{ASJSM:Maybe here we have a quick summary on what things should be checked for to make sure simulations are ok?}
 \begin{itemize}
@@
 \end{itemize}
-
-\begin{figure}
-    \includegraphics[width=0.95\linewidth]{figures/fig9_convergence/Figure.pdf}
-    \caption{Average binding free energy of 5 replicate Hamiltonian replica exchange calculations as a function of total simulation time (i.e.
-the sum of the simulation time of all replicas) for the two host-guest systems CB8-G3 and OA-G3.
-Shaded areas represent 95\% confidence intervals around the mean computed from the 5 replicates data.
-The horizontal dashed lines show the final binding free energy prediction of the two calculations after a total of ~5230 ns for OA-G3 and ~6650 ns for CB8-G3.
-Longer correlation times in CB8-G3 cause the calculation to converge more slowly.
-The original data used to generate the plot can be found at \url{https://github.com/MobleyLab/SAMPL6/blob/master/host_guest/Analysis/SAMPLing/Data/reference_free_energies.csv}.
-}
-    \label{fig:freeenergytrajectories}
-\end{figure}
-
-\paragraph{Cycle closure error}
-Relative free energy calculations, which compute the change in free energy on making a change to a molecule (e.g.
-adding a functional group to a ligand) may provide an additional opportunity for error/consistency checking.
-Particularly, such calculations are often done to span a graph or tree of free energy calculations~\cite{xu2019diffnet,wang2013modeling,liu2013lead} and in some cases the free energy change to go between molecules A and B can be obtained via multiple transformation pathways.
-This allows a type of consistency checking where we assess how much the free energy change for that transformation in practice differs from equivalence.
-Significant deviations from this typically indicate insufficient configurational sampling along the lambda schedule of one or more of the transformations involved.
-
-This approach may be generalised to sets of connected transformations given the requirement that the sum of free energy changes along edges of a closed cycle should be zero.
-This analysis is called ``cycle closure''.
-In practice, such thermodynamic cycles do not actually sum to zero,  and deviations become increasingly large as the size of the cycle increases owing to propagation of error.
-Though no firm guidelines have emerged, it may be judicious to perform additional configurational sampling along edges of a network that are involved in cycles closing poorly.
-This may be done by extending simulations duration, or by averaging free energy changes over multiple repeats.
-The latter approach may yield more reproducible free energy changes, but at the expense of a stronger bias on the estimated free energies due to repeated use of the same input coordinates.
-
-A scheme to reduce cycle closure errors is used in FEP+ whereby calculated free energy changes along the nodes of the network are re-sampled assuming estimates of the calculated free energy change along a node may be obtained from a Gaussian distribution centered on the estimated free energy change and with a standard deviation equal to the estimated standard deviation of the free energy change.
-The procedure then uses a Maximum likelihood method to find new sets of free energy changes that minimize cycle closure errors~\cite{wang2013modeling}.
-An alternative approach computes the free energy change between a target and reference compound as a weighted average over all unique paths in the network, with the weights derived from the propagated uncertainties of each node (REF).
-
-\paragraph{Reversible binding simulations}
-An even more stringent test of the correctness of binding free energy calculations is to compare the results to the equilibrium binding constants derived from long timescale reversible binding simulations~\cite{pan2017quantitative}.
-For small ligands with millimolar affinities, repeated binding to and unbinding from the protein can occur for a large number of times in a sufficiently long unbiased MD simulation (10-100$\mu$s), and the equilibrium binding constants can be straightforwardly computed from the ratio of bound to unbound fractions of the simulation time.
-The agreement between the binding free energy calculations and the reversible binding simulations--given the same system preparation and the same force field parameters--will strongly support the correctness of both calculations, as the same results are arrived at by two independent methods, and any discrepancy will suggest some systematic error in one, or both, of the two methods.
-This test should be applied to all free energy calculation codes in the future.
 

### 2020-01-10 paper/manuscript.tex
undoing some silly contribution clashes

@@
 \author[3*]{David L. Mobley}
-\author[1]{Levi N. Naden}
+\author[11]{Levi N. Naden}
 \author[4]{Samarjeet Prasad}
@@
 \affil[10]{Cresset, Cambridgeshire, UK}
+\affil[11]{Molecular Sciences Software Institute, Blacksburg VA, USA}
 %
@@
 MRS: Helped create figure~\ref{fig:fig_what_is_lambda}, wrote sections describing choices for alchemical pathways and analysis for free energy calculations.
-LN: \todo[inline, color={red!20}]{@Levi insert info here}
+LNN: Helped write the simulation length, stopping conditions, and information saving section.
 BA: \todo[inline, color={red!20}]{@Bryce insert info here}
-JM: \todo[inline, color={red!20}]{@Julien insert info here}
+JM: Contributed to sections~\ref{subsec:reproducible},~\ref{sec:step1},~\ref{} 4.2, 5.3, 6, 8.3.1, 8.3.2\
 JDC: \todo[inline, color={red!20}]{@JDC insert info here}

### 2020-01-10 paper/manuscript.tex
Updates from Overleaf

@@
 The distinction between these can be illustrated by considering a hypothetical transformation from molecule A to molecule B, where both atoms share a common substructure but differ in their substituents; e.g.
-consider a transformation of benzene (A) to benzyl alcohol (B) fig.~\ref{fig:fig4_topology}.
+consider a transformation of benzene (A) to benzyl alcohol (B) fig.~\ref{fig:topology}.
 In this case the common substructure is the benzene ring, though perhaps may be larger depending on how it is defined, as we discuss below.
@@
 Bond parameters as well as partial charges are adjusted accordingly.
-Thus, in a single topology calculation, atoms may change their type so relatively few dummy atoms are created. This is illustrated in fig.~\ref{fig:fig4_topology} (a). 
+Thus, in a single topology calculation, atoms may change their type so relatively few dummy atoms are created. This is illustrated in left arm of fig.~\ref{fig:fig_topology}. 
 In contrast, in a dual topology free energy calculation, no atoms are allowed to change type~\cite{shirts2012best}.
 This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxy methyl group, then passing through an intermediate state where atoms which are becoming dummy atoms or ceasing being dummy atoms are partially interacting (this state may or may not be well defined~\cite{mobley2014blind}  and culminating in a state where benzyl alcohol is present along with the additional dummy atoms which was previously a corresponding hydrogen of the benzene.
-Fig.~\ref{fig:fig4_topology} (b) depicts how such a dual topology works.
+Fig.~\ref{fig:fig_topology}'s right branch depicts how such a dual topology works.
 \begin{figure}[h!]
     \includegraphics[width=0.95\columnwidth]{figures/fig3_topol/Figure.pdf}
-    \caption{\textbf{Two common topologies for AFE calculations}left: A single topology uses softcore potentials to convert from one type of atom to an other.
+    \caption{\textbf{Two common topologies for AFE calculations} left: A single topology uses softcore potentials to convert from one type of atom to an other.
 Dummy atoms (Du) are used when there is no corresponding maximum common substructure match between the two molecules for certain atoms.

### 2020-01-11 paper/manuscript.tex
Edits to beginning of the document

@@
 Alchemical free energy calculations are a useful tool for predicting free energy differences associated with the transfer of small molecules from one environment to another.
-The hallmark of these methods is the use of "bridging" potential energy functions representing \emph{alchemical} intermediate states that do not exist in as real chemical species.
-The data collected from these bridging alchemical thermodynamic states allows the computation of transfer free energies (or differences in transfer free energies) with orders of magnitude less simulation time than simulating the transfer process directly.
+The hallmark of these methods is the use of "bridging" potential energy functions representing \emph{alchemical} intermediate states that cannot exist in as real chemical species.
+The data collected from these bridging alchemical thermodynamic states allows the efficient computation of transfer free energies (or differences in transfer free energies) with orders of magnitude less simulation time than simulating the transfer process directly.
 %
@@
 In this paper, we review current best practices for several popular application domains of alchemical free energy calculations, including relative and absolute small molecule binding free energy calculations to biomolecular targets.
-\todo[inline, color=green!20]{JDC: Should we migrate this document to \url{https://github.com/alchemistry}? ASJSM: Let's migrate with the submitted version, but leave it here for now?}
+%\todo[inline, color=green!20]{JDC: Should we migrate this document to \url{https://github.com/alchemistry}? ASJSM: Let's migrate with the submitted version, but leave it here for now?}
 \end{abstract}
@@
 Alchemical free energy (AFE) calculations compute free energy differences associated with transfer processes, such as the binding of a small molecule to a receptor or the transfer of a small molecule from an aqueous to apolar phase~\cite{zwanzig1954hightemperature}.
-The calculations use non-physical intermediate states where atoms are changed from one type to another or can be removed or inserted by turning on or off the interactions with their environment. 
+\todo[inline, color={yellow!20}]{JDC: Do we really want to introduce the abbreviation "AFE" when we can just use the word "alchemical" wherever it might be needed?}
+These calculations use non-physical\footnote{Here, the non-physical nature of the transformation is referred to as "alchemical", a term coined by Tembre and McCammon in Ref.~\cite{tembre1984ligand}.} intermediate states in which the chemical identity of some portion of the system (such as a small molecule ligand or receptor sidechain) is changed by modifying the potential governing the interaction of the atoms being modified, inserted, or deleted with their environment.
 %
 Fig.~\ref{fig:fig_what_is_alchemy} illustrates common free energy changes that may be difficult to compute with conventional computational methods, but are more readily tractable with AFE methods. 
-In these types of simulations so-called \textit{alchemical states} permit converged calculations for large changes, such as the entire removal of a ligand from a binding site, as seen in fig.~\ref{fig:fig_what_is_alchemy} (E).
-This type of simulation is known as an absolute AFE calculation.
-Other typical problems tackled with AFE calculations can be as simple as conformational changes of a molecule that exhibit a high barrier between these (fig.~\ref{fig:fig_what_is_alchemy} (A)); computation of partition coefficients, such as $\log P$ or $\log D$ (fig.~\ref{fig:fig_what_is_alchemy} (B))~\cite{rustenburg2016measuring, bosisio2016blinded}; and membrane insertions (fig.~\ref{fig:fig_what_is_alchemy} (C))~\cite{}.
-Then there are common types of AFE that estimate free energies of binding: the ligand stays the same, but a protein residue is alchemically mutated and the effect on the binding affinity is probed (fig.~\ref{fig:fig_what_is_alchemy} (D))\cite{hauser2018predicting,aldeghi2018accurate}, changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}, absolute binding free energies (fig.~\ref{fig:fig_what_is_alchemy} (E))~\cite{}, and relative free energies of binding where the difference of binding free energy between to similar ligands is evaluated (fig.~\ref{fig:fig_what_is_alchemy} (F))~\cite{}.
-%
-In order to obtain a free energy from these AFE simulations, historically two common approaches were used for the analysis: free energy perturbation (FEP) using the Zwanzig relation~\cite{zwanzig1954hightemperature} or Bennett's acceptance ratio~\cite{bennett1976efficient} and thermodynamic integration (TI) for which the theory dates back decades but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamics, merz1989free}.
+\todo[inline, color={yellow!20}]{JDC: Not sure what is meant by "conventional computational methods". Perhaps we can be precise and say "unbiased molecular dynamics simulations"?}
+In AFE simulations, the introduction of intermediate \textit{alchemical states} that bridge the high-probability regions of configuration space between the two physical endstates of interest permits the free energy of large transformations to be robustly computed.
+%such as the entire removal of a ligand from a binding site (and \emph{absolute AFE calculation}), as illustrated in Fig.~\ref{fig:fig_what_is_alchemy} (E). 
+AFE calculations can be used in a variety of scenarios, such as computing the free energy of a conformational change for a molecule with a high barrier to interconversion (Fig.~\ref{fig:fig_what_is_alchemy} (A));
+\todo[inline, color={yellow!20}]{JDC: Do you intend these to be representative of BLUES type simulations?}
+computing partition ($\log P$) or distribution ($\log D$) coefficients between environments (Fig.~\ref{fig:fig_what_is_alchemy} (B))~\cite{rustenburg2016measuring, bosisio2016blinded} or compartments or membranes (Fig.~\ref{fig:fig_what_is_alchemy} (C))~\cite{}.
+AFE calculations are frequently used to estimate changes in free energies upon modifying a ligand or protein: a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy} (D))\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}; the entire ligand can be alchemical transferred from protein to solvent in an absolute binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (E))~\cite{mobley2007predicting,aldeghi2016accurate,aldeghi2017predictions}; and small alchemical modifications can be made between chemically related ligands to estimate differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (F))~\cite{wang2015accurate,mey2016blinded,gapsys2020large}.
+
+Once an AFE simulation as been performed, the data must be analyzed to estimate the free energy for the transformation of interest.
+Early work used simple (but suboptimal) estimators for this: free energy perturbation (FEP) using the Zwanzig relation~\cite{zwanzig1954hightemperature} or thermodynamic integration (TI), for which the theory dates back decades but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamics, merz1989free}.
+More recent developments have seen new, highly efficient statistical estimators that make better use of all the data, often building on the Bennett acceptance ratio (BAR)~\cite{bennett1976efficient}, producing multistate generalizations~\cite{mbar} or removing the need for global equilibrium~\cite{tram}
+\todo[inline, color={yellow!20}]{JDC: The key pivotal moment was the Shirts paper on extremely precise free energy calculations, which demonstrated that a combination of technologies can produce robust estimates with useful precision.}
 Important studies continued during the 2000's often leading to improved implementation of the methods in common simulation software~\cite{vanderspoel2005gromacs, mermelstein2018fast, wang2015accurate, hedges2019biosimspace}.
-This foundational work, combined with the methodological, technological, and hardware improvements of the last 5-10 years have led to an explosion of interest and direct commercial application of the technologies~\cite{wang2015accurate, fratev2019improved, schindler2020largescale, cournia2017relative, sherborne2016collaborating}.
+This foundational work, combined with the methodological, technological, and hardware improvements of the last 5--10 years, have led to an explosion of interest and direct commercial application of the technologies~\cite{wang2015accurate, fratev2019improved, schindler2020largescale, cournia2017relative, sherborne2016collaborating}.
 %
@@
 An additional consideration arises for single topology relative free energy calculations.
-In this class of alchemical free energy calculations it is necessary to generate a molecular topology that may describe the initial and final states of the perturbation (see different topologies as seen in fig.~\ref{fig:fig_topology}).
+In this class of alchemical free energy calculations it is necessary to generate a molecular topology that may describe the initial and final states of the perturbation (see different topologies as seen in Fig.~\ref{fig:fig_topology}).
 In cases where the end-states have high topological similarity and high structural overlap this is relatively straightforward and typically handled by use of maximum common substructure calculations.
@@
 The distinction between these can be illustrated by considering a hypothetical transformation from molecule A to molecule B, where both atoms share a common substructure but differ in their substituents; e.g.
-consider a transformation of benzene (A) to benzyl alcohol (B) fig.~\ref{fig:topology}.
+consider a transformation of benzene (A) to benzyl alcohol (B) Fig.~\ref{fig:topology}.
 In this case the common substructure is the benzene ring, though perhaps may be larger depending on how it is defined, as we discuss below.
@@
 Bond parameters as well as partial charges are adjusted accordingly.
-Thus, in a single topology calculation, atoms may change their type so relatively few dummy atoms are created. This is illustrated in left arm of fig.~\ref{fig:fig_topology}. 
+Thus, in a single topology calculation, atoms may change their type so relatively few dummy atoms are created. This is illustrated in left arm of Fig.~\ref{fig:fig_topology}. 
 In contrast, in a dual topology free energy calculation, no atoms are allowed to change type~\cite{shirts2012best}.
@@
 \end{figure} 
-In some cases, during the lead optimization stage, or if perturbations would be rather large a star shaped network as seen in fig.~\ref{fig:fig_types_of_networks} (A) is used..
-However, adding redundancy into the network means that a better error analysis can be carried out, by looking at cycle closure errors as discussed in sec.~\ref{subsec:predictions}, with an example given in fig.~\ref{fig:fig_types_of_networks}.
+In some cases, during the lead optimization stage, or if perturbations would be rather large a star shaped network as seen in Fig.~\ref{fig:fig_types_of_networks} (A) is used..
+However, adding redundancy into the network means that a better error analysis can be carried out, by looking at cycle closure errors as discussed in sec.~\ref{subsec:predictions}, with an example given in Fig.~\ref{fig:fig_types_of_networks}.
 
@@
 Fig.~\ref{fig:fig_sampling_scheme} illustrates the four most common schemes used.
-The simplest approach is defining a set of fixed $\lambda$ values and then run independent simulations at each of the intermediate $\lambda$ values, see fig.~\ref{fig:fig_sampling_scheme}(A).
+The simplest approach is defining a set of fixed $\lambda$ values and then run independent simulations at each of the intermediate $\lambda$ values, see Fig.~\ref{fig:fig_sampling_scheme}(A).
 This type of scheme is currently used for AMBER TI calculations~\cite{song2019} and for BioSimSpace~\cite{hedges2019biosimspace}.
@@
 Using a Metropolis Hastings based acceptance criterion for exchanges means that the generated ensemble of all replicas is still sampling from the Boltzmann distribution and has been used in many different contexts for molecular simulations~\cite{sugita2000multidimensionala,sugita1999replicaexchangea, woods2003developmenta, jiang2010free}.
-The basic idea of the replica exchange scheme is shown in fig.~\ref{fig:fig_sampling_scheme}(B).
+The basic idea of the replica exchange scheme is shown in Fig.~\ref{fig:fig_sampling_scheme}(B).
 It is supported in various software packages that provide AFE implementations, such as GROMACS~\cite{aldeghi2015accurate}, FEP+~\cite{wang2015accurate}, and NAMD~\cite{jiang2019computing}.
 A third approach borrows ideas from simulated tempering~\cite{marinari1992simulateda}.
-In this scheme a single replica rapidly explores all of $\lambda$ space by optimally working out weights that allow switching between different intermediate $\lambdas$, as seen in fig.~\ref{fig:fig_sampling_scheme}(C) and also referred to as self adjusted mixture sampling~\cite{lyubartsev1992newa, li2007simulated, tan2017optimally} and while promising, has so far only been supported in~\cite{andrearizzi2019choderalab}.
+In this scheme a single replica rapidly explores all of $\lambda$ space by optimally working out weights that allow switching between different intermediate $\lambda$ values, as seen in Fig.~\ref{fig:fig_sampling_scheme}(C) and also referred to as self adjusted mixture sampling~\cite{lyubartsev1992newa, li2007simulated, tan2017optimally} and while promising, has so far only been supported in~\cite{andrearizzi2019choderalab}.
 The last approach available in conjunction with GROMACS makes use of non-equilibrium simulations~\cite{aldeghi2018accurate}.
 In this approach, only end state $\lambda$ replicas ($\lambda$=0, $\lambda=1$) are simulated at equilibrium intermediate information is generated from non-equilibrium simulations.
-A schematic of this approach is shown in fig.~\ref{fig:fig_sampling_scheme}.
+A schematic of this approach is shown in Fig.~\ref{fig:fig_sampling_scheme}.
 \ref{fig:fig_sampling_scheme} A:\cite{hedges2019biosimspace, song2019using } B\cite{aldeghi2015accurate, wang2015accurate,jiang2019computing} C:\cite{li2007simulated, tan2017optimally} D:\cite{aldeghi2018accurate}
@@
 \end{figure} 
-The best practice recommendation, is to use replica exchange type sampling schemes (fig.~\ref{fig:fig_sampling_scheme}(B)), or in absence of their availability and when looking at systems with fast mixing times parallel replica schemes are acceptable (fig.~\ref{fig:fig_sampling_scheme}(A)).
+The best practice recommendation, is to use replica exchange type sampling schemes (Fig.~\ref{fig:fig_sampling_scheme}(B)), or in absence of their availability and when looking at systems with fast mixing times parallel replica schemes are acceptable (Fig.~\ref{fig:fig_sampling_scheme}(A)).
 Single replica schemes and non-equilibrium schemes are not as established yet, but are very promising.

### 2020-01-14 paper/manuscript.tex
adding changes to cleanup of sec 4

--- 2020-01-10 paper/manuscript.tex
@@
 Running a direct simulation of the binding events is too computationally costly, meaning we cannot easily estimate the configurational integrals of the bound and unbound state from the simulation.
-This is where alchemical free simulations come in.
+This is where alchemical free energy simulations come in.
 We can write down a thermodynamic cycle as seen in ~\ref{fig:fig_binding_thermodynamic_cycle}, for a relative free energy simulation.
@@
 %
-It is expected that AFE methods can provide a valuable tool for drug discovery, capable of outperforming alternative approaches~\cite{kuhn2017prospective}.
-It has been stressed that successful application in industry requires more knowledge of the "domain of applicability", in other words, a greater understanding of the success and failure~\cite{sherborne2016collaborating}, which in turn requires robust protocols for preparing, submitting and analysing AFE calculations.
+Many practitioners expect AFE methods to provide valuable guidance for drug discovery, and to exhibit accuracy superior to most  alternative approaches for suitable targets~\cite{kuhn2017prospective}.
+It has been stressed that successful application in industry may require considerable knowledge of the "domain of applicability" of free energy calculations -- where they work well and where they will not ~\cite{sherborne2016collaborating}.
+Successful application also requires robust protocols for preparing, submitting and analysing AFE calculations.
 In this regard, the issues mentioned in the previous section such as understanding the suitability and timescales to capture the structure activity relationships (SAR), and performing up-front tests of performance are all relevant to drug discovery applications.
-Without venturing too far into details of system setup, which is beyond the scope of this article, there are some points to be stressed.
+Without venturing too far into details of system setup, which is beyond the scope of this article, we highlight some critical factors affecting accuracy and successful application.
 %
@@
 The change of pKa is a common medicinal chemistry strategy to modify drug properties and would normally require explicit effort to incorporate into alchemical calculations.
-Further concerns about the experimental setup require understanding what protein construct is used in the bioassay.
+Also, to ensure modeling matches experiment, we need to ensure that we simulate the same system -- which requires understanding what protein construct is used in the bioassay.
 For instance, does the X-ray that is to be used for the calculations match the construct used for screening (i.e.
@@
 \subsection{Is my binding mode accurate?}
-As also mentioned, good performance of alchemical FE calculations requires an accurate representation of the ligand binding mode, normally solved via X-ray crystallography.
+As also mentioned, good performance of alchemical FE calculations requires an accurate representation of the ligand binding mode, which often is estimated based on X-ray crystallography for a related ligand.
 If more than one structure is available, the modeler should pay attention to choose the most suitable.
@@
 for a detailed discussion of choosing optimal structures for structure-based modeling~\cite{warren2012essential}.
-It is also useful to study the structure activity relationship and understand the expected impact in the binding site, is side chain movement required and is there evidence of this in any alternative X-ray structures of the same protein?
-Normally, only one protein and water configuration is used for a series of alchemical FE calculations, so this needs to be capable of accommodating the smallest through to largest ligands in a way that allows stable and well behaved simulations.
-This can be a practical limitation for the alchemical changes that are feasible, a simple work-around can be to separate compounds into sub-series for different calculations.
-If multiple structures are available there is some evidence the higher affinity complex can give better performance~\cite{perez-benito2019predicting}.
-%
+It is also useful to study the structure activity relationship and understand the expected impact in the binding site, whether side chain movement in the protein will be required, and whether there evidence of this in any alternative X-ray structures of the same protein.
+Often, only one protein and water configuration is used for a series of alchemical FE calculations, so this needs to be capable of accommodating the smallest through to largest ligands in a way that allows stable and well behaved simulations.
+This can be a practical limitation for the alchemical changes that are feasible, though a simple work-around can be to separate compounds into sub-series for different calculations.
+If multiple structures are available there is some evidence the higher affinity complex can give better performance~\cite{perez-benito2019predicting}, at least in some cases.
+However, ligands and proteins can also undergo unexpected changes in binding mode for related ligands, which can make these issues more complex to deal with.
+
 \subsection{Input setup and scale of calculations}
@@
 This provides assurances that coordinates for the non-perturbed structures remain identical and aromatic substituents for instance obey consistent dihedral angles but is not feasible for many compounds and therefore automation can be desirable.
+
 Finally, the role of water in ligand binding is now well understood and it can be crucial to capture the changes in binding site solvation during ligand binding.
@@
 If so, how will the water be incorporated for smaller ligands that require the water – consider pre-solvation methods such as GCMC steps~\cite{michel2010prediction}.
-Other tools such as watermap or opensource equivalents (SSTmap, others) can be used to define water structure for systems with no experimental evidence of water sites.
+Other tools such as WaterMap or open source equivalents (SSTmap, others) can be used to define water structure for systems with no experimental evidence of water sites.
 Before launching large numbers of alchemical free energy calculations it is recommended to test the system using classical MD simulations and limited numbers of alchemical perturbations.
 Metrics such as ligand and protein RMSD and RMSF can be inspected, along with visual inspection of simulations, to ensure the system is stable and likely to be suitable for FE calculations.
-%
+
 Running binding FE calculations in a drug discovery application will typically require the use of software or tools to facilitate the calculations of many compounds.
@@
 A retrospective assessment can give an indication of prospective performance for similar molecules~\cite{ciordia2016application}.
-Beyond this, several parameters provide useful indicators of performance, results suggest MBAR errors for perturbations greater than 0.3 kcal/mol can be indicative of inaccuracy~\cite{perez-benito2019predicting}.
-In addition, the hysteresis can be checked by comparing the difference for forward and reverse perturbations of the same structural change, or in the case of large maps of perturbations cycle closure hysteresis~\cite{wang2013modeling} indicates problematic perturbations involved in cycles connecting many compounds.
+Beyond this, several parameters provide useful indicators of performance.
+Results suggest MBAR errors greater than 0.3 kcal/mol for perturbations can be indicative of inaccuracy~\cite{perez-benito2019predicting}.
+In addition, the hysteresis can be checked by comparing the difference for forward and reverse perturbations of the same structural change, or in the case of large maps of perturbations cycle closure errors~\cite{wang2013modeling} indicate problematic perturbations involved in cycles connecting many compounds.
 Selection bias, as described above, can be problematic in prospective applications as there is normally an aim to focus on only a narrow range of predicted activity (i.e.
@@
 Thus, correlation between prediction and subsequent experiment may lead to disappointment but corrections can be applied based on previous recommendations~\cite{abel2017critical}.
-%
+
 In summary, the successful use of alchemical calculations in drug discovery requires working in the domain of applicability, using a high quality X-ray structure of the target compounds, and testing the approach retrospectively to ensure the system setup is well-behaved.
-Always assess the confidence in the predictions and communicate this when discussing with experimentalists.
+Always assess your confidence in the resulting predictions and communicate this when discussing with experimentalists.
 Consider performing repeat calculations for at least some of the perturbations in the study.
-In a drug discovery setting it is inevitable to consider larger virtual libraries, in these cases again evaluate if shorter simulation time per $\lambda$ window can be sufficient for a ‘virtual screening’ mode of application.
+In a drug discovery setting it is inevitable to consider larger virtual libraries.
+In these cases again evaluate if shorter simulation time per $\lambda$ window can be sufficient for a ‘virtual screening’ mode of application.
 It is expected that future tools may allow efficient exploration of large virtual libraries.
@@
 However, it is important to have realistic expectations.
-Structure based drug design projects are often capable of improving potency relatively quickly and the range of activity narrows to just two-to-three log units.
-It may seem hard to impact with substantially different, more potent, stand-out compounds in this scenario, but binding energy prediction can still be extremely useful for ensuring activity is maintained as other properties are optimized.
+
+Structure based drug design projects are often capable of improving potency relatively quickly, even with only limited application of computational approaches and the range of activity narrows to just two-to-three log units.
+It may seem hard to have impact with substantially different, more potent, stand-out compounds in this scenario, but binding energy prediction can still be extremely useful for ensuring activity is maintained as other properties are optimized.
 An interesting cost benefit analysis has shown the value of activity prediction, see discussion above and articles such as~\cite{mobley2012perspectiv}.
-From a drug discovery point of view, FEP is expanding its domain of applicability, there are reports of success using homology models and GPCRs for instance, as well as enabling charge change and scaffold hopping, but these systems are undoubtedly more difficult.
+From a drug discovery point of view, FEP is expanding its domain of applicability, and there are reports of success using homology models and GPCRs for instance, as well as enabling charge change and scaffold hopping, but these systems are undoubtedly more difficult.
 In the meantime, the use cases are expanding to resistance prediction, disease mutation prediction, solubility prediction – an exciting future for alchemical FE calculations.
@@
 LNN: Helped write the simulation length, stopping conditions, and information saving section.
-BA: \todo[inline, color={red!20}]{@Bryce insert info here}
+BA: Helped write the uncertainty estimation, stopping conditions, and output analysis sections and created figure ~\ref{fig:convergence_forward_reverse}.
 JM: Contributed to sections~\ref{subsec:reproducible},~\ref{sec:step1},~\ref{} 4.2, 5.3, 6, 8.3.1, 8.3.2\

### 2020-01-14 paper/manuscript.tex
changing afe to alchemical

@@
 \todo[inline, color={green!20}]{@everyone: Any further input in this?}
-Alchemical free energy (AFE) calculations compute free energy differences associated with transfer processes, such as the binding of a small molecule to a receptor or the transfer of a small molecule from an aqueous to apolar phase~\cite{zwanzig1954hightemperature}.
-\todo[inline, color={yellow!20}]{JDC: Do we really want to introduce the abbreviation "AFE" when we can just use the word "alchemical" wherever it might be needed?}
+Alchemical free energy calculations compute free energy differences associated with transfer processes, such as the binding of a small molecule to a receptor or the transfer of a small molecule from an aqueous to apolar phase~\cite{zwanzig1954hightemperature}.
 These calculations use non-physical\footnote{Here, the non-physical nature of the transformation is referred to as "alchemical", a term coined by Tembre and McCammon in Ref.~\cite{tembre1984ligand}.} intermediate states in which the chemical identity of some portion of the system (such as a small molecule ligand or receptor sidechain) is changed by modifying the potential governing the interaction of the atoms being modified, inserted, or deleted with their environment.
 %
-Fig.~\ref{fig:fig_what_is_alchemy} illustrates common free energy changes that may be difficult to compute with conventional computational methods, but are more readily tractable with AFE methods. 
+Fig.~\ref{fig:fig_what_is_alchemy} illustrates common free energy changes that may be difficult to compute with conventional computational methods, but are more readily tractable with alchemical methods.
 \todo[inline, color={yellow!20}]{JDC: Not sure what is meant by "conventional computational methods". Perhaps we can be precise and say "unbiased molecular dynamics simulations"?}
-In AFE simulations, the introduction of intermediate \textit{alchemical states} that bridge the high-probability regions of configuration space between the two physical endstates of interest permits the free energy of large transformations to be robustly computed.
+In alchemical simulations, the introduction of intermediate \textit{alchemical states} that bridge the high-probability regions of configuration space between the two physical endstates of interest permits the free energy of large transformations to be robustly computed.
 %such as the entire removal of a ligand from a binding site (and \emph{absolute AFE calculation}), as illustrated in Fig.~\ref{fig:fig_what_is_alchemy} (E). 
-AFE calculations can be used in a variety of scenarios, such as computing the free energy of a conformational change for a molecule with a high barrier to interconversion (Fig.~\ref{fig:fig_what_is_alchemy} (A));
+alchemical calculations can be used in a variety of scenarios, such as computing the free energy of a conformational change for a molecule with a high barrier to interconversion (Fig.~\ref{fig:fig_what_is_alchemy} (A));
 \todo[inline, color={yellow!20}]{JDC: Do you intend these to be representative of BLUES type simulations?}
 computing partition ($\log P$) or distribution ($\log D$) coefficients between environments (Fig.~\ref{fig:fig_what_is_alchemy} (B))~\cite{rustenburg2016measuring, bosisio2016blinded} or compartments or membranes (Fig.~\ref{fig:fig_what_is_alchemy} (C))~\cite{}.
-AFE calculations are frequently used to estimate changes in free energies upon modifying a ligand or protein: a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy} (D))\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}; the entire ligand can be alchemical transferred from protein to solvent in an absolute binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (E))~\cite{mobley2007predicting,aldeghi2016accurate,aldeghi2017predictions}; and small alchemical modifications can be made between chemically related ligands to estimate differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (F))~\cite{wang2015accurate,mey2016blinded,gapsys2020large}.
-
-Once an AFE simulation as been performed, the data must be analyzed to estimate the free energy for the transformation of interest.
+alchemical calculations are frequently used to estimate changes in free energies upon modifying a ligand or protein: a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy} (D))\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}; the entire ligand can be alchemical transferred from protein to solvent in an absolute binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (E))~\cite{mobley2007predicting,aldeghi2016accurate,aldeghi2017predictions}; and small alchemical modifications can be made between chemically related ligands to estimate differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (F))~\cite{wang2015accurate,mey2016blinded,gapsys2020large}.
+
+Once an alchemical simulation as been performed, the data must be analyzed to estimate the free energy for the transformation of interest.
 Early work used simple (but suboptimal) estimators for this: free energy perturbation (FEP) using the Zwanzig relation~\cite{zwanzig1954hightemperature} or thermodynamic integration (TI), for which the theory dates back decades but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamics, merz1989free}.
@@
 %
-As the field of molecular simulation moves towards millisecond timescales, accurate AFE calculations on even more challenging problems will become reasonable to perform.
+As the field of molecular simulation moves towards millisecond timescales, accurate alchemical calculations on even more challenging problems will become reasonable to perform.
 In the meantime, today's users may find it hard to get started with these complex calculations whilst also keeping up with the fast pace of change.
@@
 If you feel unfamiliar with some of these concepts please good starting points can be found in these references~\cite{braun2019best, grossfield2018best, klimovich2015guidelines, shirts2012best}.
-Oftentimes, docking calculations are required to generated initial binding pose for a small molecule for an AFE calculation, where binding poses are not known, thus some basic familiarty on how to run docking calculations is also expected~\cite{grinter2014challenges}.
+Oftentimes, docking calculations are required to generated initial binding pose for a small molecule for an alchemical calculation, where binding poses are not known, thus some basic familiarty on how to run docking calculations is also expected~\cite{grinter2014challenges}.
 %
 As some of the theoretical background can seem daunting, we provide an essentials guide to the theory behind alchemical free energy calculations in Section~\ref{sec:theory}.
-In the remainder of this paper, we will cover topics that are key to the preparation~(\ref{sec:step1}), choice and use of correct protocols~(\ref{sec:step2}), and finally the best practices that should be used for the analysis of AFE calculations~(\ref{sec:step3}).
+In the remainder of this paper, we will cover topics that are key to the preparation~(\ref{sec:step1}), choice and use of correct protocols~(\ref{sec:step2}), and finally the best practices that should be used for the analysis of alchemical calculations~(\ref{sec:step3}).
 Particular focus will be given to aspects of the molecular simulations which are unique to alchemical calculations---these include the calculation of transfer free energies (hydration free energies, partition coefficients, etc.), and binding free energies (absolute and relative).
@@
     \textit{CPU} - Central Processing Unit\\
-    \textit{AFE} - Alchemical Free Energy\\
     \textit{BAR} - Bennett Acceptance Ratio\\
@@
 %
-Many practitioners expect AFE methods to provide valuable guidance for drug discovery, and to exhibit accuracy superior to most  alternative approaches for suitable targets~\cite{kuhn2017prospective}.
+Many practitioners expect alchemical methods to provide valuable guidance for drug discovery, and to exhibit accuracy superior to most  alternative approaches for suitable targets~\cite{kuhn2017prospective}.
 It has been stressed that successful application in industry may require considerable knowledge of the "domain of applicability" of free energy calculations -- where they work well and where they will not ~\cite{sherborne2016collaborating}.
-Successful application also requires robust protocols for preparing, submitting and analysing AFE calculations.
+Successful application also requires robust protocols for preparing, submitting and analysing alchemical calculations.
 In this regard, the issues mentioned in the previous section such as understanding the suitability and timescales to capture the structure activity relationships (SAR), and performing up-front tests of performance are all relevant to drug discovery applications.
@@
     \includegraphics[width=0.95\columnwidth]{figures/fig3_topol/Figure.pdf}
-    \caption{\textbf{Two common topologies for AFE calculations} left: A single topology uses softcore potentials to convert from one type of atom to an other.
+    \caption{\textbf{Two common topologies for alchemical calculations} left: A single topology uses softcore potentials to convert from one type of atom to an other.
 Dummy atoms (Du) are used when there is no corresponding maximum common substructure match between the two molecules for certain atoms.
@@
 \subsubsection{What different sampling schemes exist and which one should I use?}
-With all AFE simulations having the requirement of sampling from different sets of $\lambda$ states in common different approaches have been designed.
+With all alchemical simulations having the requirement of sampling from different sets of $\lambda$ states in common different approaches have been designed.
 Fig.~\ref{fig:fig_sampling_scheme} illustrates the four most common schemes used.
@@
 The basic idea of the replica exchange scheme is shown in Fig.~\ref{fig:fig_sampling_scheme}(B).
-It is supported in various software packages that provide AFE implementations, such as GROMACS~\cite{aldeghi2015accurate}, FEP+~\cite{wang2015accurate}, and NAMD~\cite{jiang2019computing}.
+It is supported in various software packages that provide alchemical implementations, such as GROMACS~\cite{aldeghi2015accurate}, FEP+~\cite{wang2015accurate}, and NAMD~\cite{jiang2019computing}.
 A third approach borrows ideas from simulated tempering~\cite{marinari1992simulateda}.

### 2020-01-14 paper/manuscript.tex
Updates from Overleaf

@@
 Alchemical free energy calculations compute free energy differences associated with transfer processes, such as the binding of a small molecule to a receptor or the transfer of a small molecule from an aqueous to apolar phase~\cite{zwanzig1954hightemperature}.
-These calculations use non-physical\footnote{Here, the non-physical nature of the transformation is referred to as "alchemical", a term coined by Tembre and McCammon in Ref.~\cite{tembre1984ligand}.} intermediate states in which the chemical identity of some portion of the system (such as a small molecule ligand or receptor sidechain) is changed by modifying the potential governing the interaction of the atoms being modified, inserted, or deleted with their environment.
+These calculations use non-physical\footnote{Here, the non-physical nature of the transformation is referred to as "alchemical", a term coined by Tembre and McCammon in Ref.~\cite{tembre1984ligandreceptor}.} intermediate states in which the chemical identity of some portion of the system (such as a small molecule ligand or receptor sidechain) is changed by modifying the potential governing the interaction of the atoms being modified, inserted, or deleted with their environment.
 %
@@
 computing partition ($\log P$) or distribution ($\log D$) coefficients between environments (Fig.~\ref{fig:fig_what_is_alchemy} (B))~\cite{rustenburg2016measuring, bosisio2016blinded} or compartments or membranes (Fig.~\ref{fig:fig_what_is_alchemy} (C))~\cite{}.
-alchemical calculations are frequently used to estimate changes in free energies upon modifying a ligand or protein: a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy} (D))\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}; the entire ligand can be alchemical transferred from protein to solvent in an absolute binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (E))~\cite{mobley2007predicting,aldeghi2016accurate,aldeghi2017predictions}; and small alchemical modifications can be made between chemically related ligands to estimate differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (F))~\cite{wang2015accurate,mey2016blinded,gapsys2020large}.
+alchemical calculations are frequently used to estimate changes in free energies upon modifying a ligand or protein: a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy} (D))\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}; the entire ligand can be alchemical transferred from protein to solvent in an absolute binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (E))~\cite{mobley2007predicting,aldeghi2015accurate,aldeghi2017predictions}; and small alchemical modifications can be made between chemically related ligands to estimate differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (F))~\cite{wang2015accurate,mey2016blinded,gapsys2020large}.
 
@@
 
+\newpage
 \section{Checklist}
@@
 \todo[inline, color={green!20}]{JS: @self,ASJSM do some final formatting when text is complete.}
-
+%Checklist:
+%%%%%%%%%%%%%%%% 
+%% Checklist-specific precommands:
+\newenvironment{checklist_font}{\fontfamily{lmss}\small\selectfont}{\par}
+\newlist{todolist}{itemize}{2}
+\setlist[todolist]{label=$\square$}
+%% start checklist:
+\begin{checklist_font}
+\begin{itemize}
+\begin{todolist}
+
+%%%%%%%%%%%%%%%%%
+%%%%%% step 0
+\begin{tcolorbox}[width=15cm, center]
+    {\large {\color{teal}STEP 0: KNOW WHAT YOU WANT TO SIMULATE}} \newline
+    \textbf{Initial questions you should ask before you set up an alchemical free energy calculation using molecular dynamics simulations}
+    \item Do I understand the biology, chemistry and physics of my system?
+    \item Have I properly prepared my protein and ligand systems?
+    \item Does my system contain any structures that require custom parameters?
+    \item What simulation protocol will provide the most evidence to answer my hypothesis?
+    \item Are the projected computational expense and runtime realistic for my goals?
+    \item Will my protocol be reproducible? 
+    \item Will my statistics be reliable? If not, would more replicates solve the problem? Three replicates is the minimal number of replicates that should be run
+    \item Can I open-source my data?
+\end{tcolorbox}
+
+%%%%%% step 1
+\begin{tcolorbox}[width=15cm, center]
+    {\large {\color{teal}STEP 1: PREPARING YOUR SIMULATIONS}} \newline
+    \textbf{Steps to getting started setting up your alchemical free energy calculation}
+    \item Make sure you know why you have picked your (combination of) force field(s)
+    \item Minimize your system
+    \item Equilibrate your system with your choice of thermodynamic ensemble
+    \item Check the stability of your system and whether it behaves the way you want it to
+\end{tcolorbox}
+
+%%%%%% step 2
+\begin{tcbraster}[raster columns=2,raster equal height, raster width=18cm]
+\begin{adjustwidth}{-4.3cm}{}           % workaround to make the dual box pass the margin
+\begin{todolist}
+    \begin{tcolorbox}
+    % absolute
+        {\large {\color{teal}STEP 2: RUNNING ABSOLUTE SIMULATIONS}} \newline
+        \textbf{Steps to running your absolute alchemical free energy calculations}
+
+        \item Your ligands should have the same, biologically correct binding pose
+        \item Make sure your \textlambda-scheduling is set correctly
+        \item Check if your ligands are discharging and vanishing correctly
+        \item Set up your restraints correctly
+        \item Make sure you subsample in your MBAR protocol
+        \item Apply the right correction terms
+    \end{tcolorbox}
+    \begin{tcolorbox}
+    % relative
+        {\large {\color{teal}STEP 2: RUNNING RELATIVE SIMULATIONS}} \newline
+        \textbf{Steps to running your relative alchemical free energy calculations}
+        \item           % circumvents bug in tcbraster/todolist
+        \item Your ligands should have the same, biologically correct binding pose
+        \item Make sure your λ-scheduling is set correctly
+        \item Make sure your molecular transformations are realistic (1-5 heavy atoms for reliable computations)
+        \item Generate a perturbation network by your method of choice; check whether you have enough cycle closures (preferably, every ligand should be in a cycle)
+        \item Check whether dummy atoms were assigned correctly
+        \item Make sure you subsample in your MBAR protocol
+    \end{tcolorbox}
+
+\end{todolist}
+\end{adjustwidth}
+\end{tcbraster}
+
+%%%%%% step 3
+\begin{tcolorbox}[width=15cm, center]
+    {\large {\color{teal}STEP 3: HOW DO I KNOW WHICH SIMULATIONS ARE NOT RELIABLE?}} \newline
+    \textbf{Situations suggesting your relative alchemical free energy calculations have not run properly (assuming absence of experimental affinities)}
+    \item Standard error (\textsigma) should be \textless1 kcal·mol-1 
+    \item Simulated systems have not converged 
+    \newline\newline\textit{Relative:}
+    \item If you observe hysteresis in perturbations and incorrect cycle closures
+    \item Energy differences \textgreater15 kcal·mol-1  are unreliable
+    \newline\newline\textit{Absolute:}
+    \item Energies below -15 kcal·mol-1  are unreliable
+    \item The ligand has not sampled most of the intended region after the vanishing step
+    \item The ligand is drifting out of the intended region after the vanishing step
+\end{tcolorbox}
+
+%%%%%% step 4
+\begin{tcolorbox}[width=15cm, center]
+    {\large {\color{teal}STEP 4: WHY ARE THEY NOT RELIABLE?}} \newline
+    \textbf{Suggestions for finding out why your alchemical free energy calculations may not be reliable}
+    \item Check again whether dummy atoms were assigned correctly
+    \item Inspect the trajectories across the λ-schedule (particularly the endpoints)
+    \item Inspect the overlap matrices
+\end{tcolorbox}
+
+%%%%%% step 
+\begin{tcolorbox}[width=15cm, center]
+    {\large {\color{teal}STEP 5: DATA ANALYSIS}} \newline
+    \textbf{Steps to analyzing your output data correctly}
+    \item Make sure you have run enough replicates to ensure statistical reliability (\textgreater3)
+    \item Compute both correlation and ranking coefficients and ranking statistics (e.g. r, \textrho, MUE and \texttau)
+    \item You should be able to explain large energy differences based on your system (trajectory)
+\end{tcolorbox}
+
+%%%%%%%%%%%%%%%% 
+% end checklist:
+\end{todolist}
+\end{itemize}
+\end{checklist_font}
+%%%%%%%%%%%%%%%%%
 % This provides a checklist which

### 2020-01-14 paper/manuscript.tex
corrected columning for the checklist

@@
 \todo[inline, color={green!20}]{ASJSM:Write conclusions}
-%%%%%%%%%%%%%%%%%%%%
-%              Terminology                  %
-%%%%%%%%%%%%%%%%%%%%
-\section{Terminology and abbreviations}
-\label{sec:tem-abbrev}
-\begin{itemize}
-\item Feature Box covering major technical terms and abbreviations
-\item Examples:
-\begin{itemize}
-\item EXP, BAR, MBAR
-\item Double decoupling, single-topology, dual-topology, hybrid-topology, coupled-topology
-\item FEP (free energy perturbation), alchemical, AFE (alchemical free energy)
-\end{itemize}
-\end{itemize}
-\todo[inline]{JS: should this be combined with the acronyms section at the beginning of the paper?}
+Alchemical free energy calculations have seen a vast increase in popularity also in the pharmaceutical industry for structure based drug discovery~\cite{schind}.
+Commercial products such as FEP+ and Flare,
+
 %%%%%%%%%%%%%%%%%%%%
@@
 
-\newpage
 \section{Checklist}
@@
 \todo[inline, color={green!20}]{JS: @self,ASJSM do some final formatting when text is complete.}
+\onecolumn
 %Checklist:
@@
 %%%%%%%%%%%%%%%%%
+\twocolumn
 % This provides a checklist which

### 2020-01-16 paper/checklist.tex
working backwards with fixes

--- 2020-01-14 paper/manuscript.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%% LIVECOMS ARTICLE TEMPLATE FOR BEST PRACTICES GUIDE
-%%% ADAPTED FROM ELIFE ARTICLE TEMPLATE (8/10/2017)
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%% PREAMBLE
-\documentclass[9pt,bestpractices]{livecoms}
-% Use the 'onehalfspacing' option for 1.5 line spacing
-% Use the 'doublespacing' option for 2.0 line spacing
-% Use the 'lineno' option for adding line numbers.
-% The 'bestpractices' option for indicates that this is a best practices guide.
-% Omit the bestpractices option to remove the marking as a LiveCoMS paper.
-% Please note that these options may affect formatting.
-
-\usepackage{lipsum} % Required to insert dummy text
-\usepackage[version=4]{mhchem}
-\usepackage{siunitx}
-\usepackage{url}
-\DeclareSIUnit\Molar{M}
-\usepackage[italic]{mathastext}
-\graphicspath{{figures/}}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%% IMPORTANT USER CONFIGURATION
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\usepackage[colorinlistoftodos]{todonotes}
-\usepackage[most]{tcolorbox}
-\usepackage{enumitem,amssymb}
-\usepackage{textgreek}
-\usepackage{changepage}
-
-\newcommand{\versionnumber}{0.1}  % you should update the minor version number in preprints and major version number of submissions.
-\newcommand{\githubrepository}{\url{https://github.com/michellab/alchemical-best-practices}}  %this should be the main github repository for this article
-\newcommand{\expect}[1]{\left\langle{#1}\right\rangle}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%% ARTICLE SETUP
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{Best Practices for Alchemical Free Energy Calculations: v\versionnumber}
-\author[1*]{Antonia S. J. S. Mey}
-\author[7]{Bryce K. Allen}
-\author[2*]{John D. Chodera}
-\author[1,10]{Maximilian Kuhn}
-\author[1*]{Julien Michel}
-\author[3*]{David L. Mobley}
-\author[11]{Levi N. Naden}
-\author[4]{Samarjeet Prasad}
-\author[5]{Julia E. Rice}
-\author[2,8]{Andrea Rizzi}
-\author[1]{Jenke Scheen}
-\author[6*]{Michael R. Shirts}
-\author[9]{Gary Tresadern}
-\author[7]{Huafeng Xu}
-%
-\affil[1]{EaStCHEM School of Chemistry, David Brewster Road, Joseph Black Building, The King's Buildings, Edinburgh, EH9 3FJ, UK}
-\affil[2]{Computational and Systems Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York NY, USA}
-\affil[3]{Departments of Pharmaceutical Sciences and Chemistry, University of California, Irvine, USA}
-\affil[4]{National Institutes of Health, Bethesda, MD, USA}
-\affil[5]{IBM Alamden Research Center, Almaden, CA, USA}
-\affil[6]{University of Colorado Boulder, Boulder, CO, USA}
-\affil[7]{Silicon Therapeutics, Boston, MA, USA}
-\affil[8]{Tri-Institutional Training Program in Computational Biology and Medicine, New York, NY, USA}
-\affil[9]{Computational Chemistry, Janssen Research \& Development, Turnhoutseweg 30, Beerse B-2340,Belgium}
-\affil[10]{Cresset, Cambridgeshire, UK}
-\affil[11]{Molecular Sciences Software Institute, Blacksburg VA, USA}
-%
-\corr{john.chodera@choderalab.org}{JDC}
-\corr{dmobley@mobleylab.org}{DLM}
-\corr{antonia.mey@ed.ac.uk}{ASJSM}
-\corr{mail@julienmichel.net}{JM}
-\corr{michael.shirts@colorado.edu}{MRS}
-%
-\blurb{This LiveCoMS document is maintained online on GitHub at \githubrepository; to provide feedback, suggestions, or help improve it, please visit the GitHub repository and participate via the issue tracker.}
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%% PUBLICATION INFORMATION
-%%% Fill out these parameters when available
-%%% These are used when the "pubversion" option is invoked
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\pubDOI{10.XXXX/YYYYYYY}
-\pubvolume{<volume>}
-\pubyear{<year>}
-\articlenum{<number>}
-\datereceived{Month, Day, Year}
-\dateaccepted{Month, Day, Year}
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%% ARTICLE START
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{document}
-%
-\begin{frontmatter}
-\maketitle
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%% Abstract
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{abstract}
-Alchemical free energy calculations are a useful tool for predicting free energy differences associated with the transfer of small molecules from one environment to another.
-The hallmark of these methods is the use of "bridging" potential energy functions representing \emph{alchemical} intermediate states that cannot exist in as real chemical species.
-The data collected from these bridging alchemical thermodynamic states allows the efficient computation of transfer free energies (or differences in transfer free energies) with orders of magnitude less simulation time than simulating the transfer process directly.
-%
-While these methods are highly flexible, care must be taken in avoiding common pitfalls to ensure that computed free energy differences can be robust and reproducible for the chosen force field, and that appropriate corrections are included to permit direct comparison with experimental data.
-%
-In this paper, we review current best practices for several popular application domains of alchemical free energy calculations, including relative and absolute small molecule binding free energy calculations to biomolecular targets.
-%\todo[inline, color=green!20]{JDC: Should we migrate this document to \url{https://github.com/alchemistry}? ASJSM: Let's migrate with the submitted version, but leave it here for now?}
-\end{abstract}
-\end{frontmatter}
-%
-\todototoc
-\listoftodos
-%
-%%%%%%%%%%%%%%%%%%%%
-%  Introduction    %
-%%%%%%%%%%%%%%%%%%%%
-\todo[inline]{}The current Overleaf document is \url{https://www.overleaf.com/5136551543vgjqbkzpmqkp}
-\section{What are alchemical free energy methods?}
-\label{sec:intro}
-\todo[inline, color={green!20}]{@everyone: Any further input in this?}
-Alchemical free energy calculations compute free energy differences associated with transfer processes, such as the binding of a small molecule to a receptor or the transfer of a small molecule from an aqueous to apolar phase~\cite{zwanzig1954hightemperature}.
-These calculations use non-physical\footnote{Here, the non-physical nature of the transformation is referred to as "alchemical", a term coined by Tembre and McCammon in Ref.~\cite{tembre1984ligandreceptor}.} intermediate states in which the chemical identity of some portion of the system (such as a small molecule ligand or receptor sidechain) is changed by modifying the potential governing the interaction of the atoms being modified, inserted, or deleted with their environment.
-%
-Fig.~\ref{fig:fig_what_is_alchemy} illustrates common free energy changes that may be difficult to compute with conventional computational methods, but are more readily tractable with alchemical methods.
-\todo[inline, color={yellow!20}]{JDC: Not sure what is meant by "conventional computational methods". Perhaps we can be precise and say "unbiased molecular dynamics simulations"?}
-In alchemical simulations, the introduction of intermediate \textit{alchemical states} that bridge the high-probability regions of configuration space between the two physical endstates of interest permits the free energy of large transformations to be robustly computed.
-%such as the entire removal of a ligand from a binding site (and \emph{absolute AFE calculation}), as illustrated in Fig.~\ref{fig:fig_what_is_alchemy} (E). 
-alchemical calculations can be used in a variety of scenarios, such as computing the free energy of a conformational change for a molecule with a high barrier to interconversion (Fig.~\ref{fig:fig_what_is_alchemy} (A));
-\todo[inline, color={yellow!20}]{JDC: Do you intend these to be representative of BLUES type simulations?}
-computing partition ($\log P$) or distribution ($\log D$) coefficients between environments (Fig.~\ref{fig:fig_what_is_alchemy} (B))~\cite{rustenburg2016measuring, bosisio2016blinded} or compartments or membranes (Fig.~\ref{fig:fig_what_is_alchemy} (C))~\cite{}.
-alchemical calculations are frequently used to estimate changes in free energies upon modifying a ligand or protein: a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy} (D))\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}; the entire ligand can be alchemical transferred from protein to solvent in an absolute binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (E))~\cite{mobley2007predicting,aldeghi2015accurate,aldeghi2017predictions}; and small alchemical modifications can be made between chemically related ligands to estimate differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (F))~\cite{wang2015accurate,mey2016blinded,gapsys2020large}.
-
-Once an alchemical simulation as been performed, the data must be analyzed to estimate the free energy for the transformation of interest.
-Early work used simple (but suboptimal) estimators for this: free energy perturbation (FEP) using the Zwanzig relation~\cite{zwanzig1954hightemperature} or thermodynamic integration (TI), for which the theory dates back decades but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamics, merz1989free}.
-More recent developments have seen new, highly efficient statistical estimators that make better use of all the data, often building on the Bennett acceptance ratio (BAR)~\cite{bennett1976efficient}, producing multistate generalizations~\cite{mbar} or removing the need for global equilibrium~\cite{tram}
-\todo[inline, color={yellow!20}]{JDC: The key pivotal moment was the Shirts paper on extremely precise free energy calculations, which demonstrated that a combination of technologies can produce robust estimates with useful precision.}
-Important studies continued during the 2000's often leading to improved implementation of the methods in common simulation software~\cite{vanderspoel2005gromacs, mermelstein2018fast, wang2015accurate, hedges2019biosimspace}.
-This foundational work, combined with the methodological, technological, and hardware improvements of the last 5--10 years, have led to an explosion of interest and direct commercial application of the technologies~\cite{wang2015accurate, fratev2019improved, schindler2020largescale, cournia2017relative, sherborne2016collaborating}.
-%
-As the field of molecular simulation moves towards millisecond timescales, accurate alchemical calculations on even more challenging problems will become reasonable to perform.
-In the meantime, today's users may find it hard to get started with these complex calculations whilst also keeping up with the fast pace of change.
-This best practice guide aims to serve as a go-to reference for current recommendations and tips for users of all experience.
-%
-\begin{figure}
-    \includegraphics[width=0.95\linewidth]{figures/fig1_what_is_alchemy/Figure.pdf}   
-    \caption{\textbf{Illustration of common types of free energies differences that can be calculated using alchemical free energy methods.} A: Change in free energy due to a conformational change of the molecule.
-B: Partition coefficient such as $\log P$ or $\log D$ can be computed using alchemical free energy methods.
-Here, as an example the partition coefficient between methanol and water is shown.
-C: Free energy difference associated with the insertion of a molecule into a membrane.
-D: Effect of mutations of protein or host residues on free energies of binding.
-E: Absolute free energies of binding of a small molecule to a host (e.g.
-protein), F: Relative free energy of binding of one molecule with respect to another, here toluene and benzyl alcohol.
-    \label{fig:fig_what_is_alchemy}
-    }
-\end{figure}
-%
-%%%%%%%%%%%%%%%%%%%%
-% Prerequesites    %
-%%%%%%%%%%%%%%%%%%%%
-\section{Prerequisites and Scope}
-\label{sec:pre}
-\todo[inline, color=green!20]{ASJSM: @everyone This section is up for review to also check flow with intro and rest of document.}
-This guide is meant to serve as a comprehensive starting point for new practitioners for free energy calculations.
-It assumes moderate experience with molecular simulation packages and concepts, however.
-It is also intended to serve as a reference guide with a convenient checklist (Section~\ref{sec:checklist}) for more seasoned practitioners, to help ensure their simulation meet standard alchemical simulation and analysis practices.
-It can also serve as a set of best practices to ensure simulation robustness and reproducibility which reviewers may wish to consider as they evaluate papers.
-%
-This guide aims to answer the following questions:
-\begin{itemize}
-    \item Is my problem suitable for an alchemical free energy calculation? 
-    \item How do I choose the right free energy protocol and run it? 
-    \item How do I analyze my simulations accurately? 
-    \item What software tools are available to perform alchemical free energy calculations? 
-\end{itemize}
-%
-We assume the reader possesses a basic familiarity with the principles of molecular mechanics, molecular dynamics simulations, statistical mechanics, and the biophysics of protein-ligand association.
-If you feel unfamiliar with some of these concepts please good starting points can be found in these references~\cite{braun2019best, grossfield2018best, klimovich2015guidelines, shirts2012best}.
-Oftentimes, docking calculations are required to generated initial binding pose for a small molecule for an alchemical calculation, where binding poses are not known, thus some basic familiarty on how to run docking calculations is also expected~\cite{grinter2014challenges}.
-%
-As some of the theoretical background can seem daunting, we provide an essentials guide to the theory behind alchemical free energy calculations in Section~\ref{sec:theory}.
-In the remainder of this paper, we will cover topics that are key to the preparation~(\ref{sec:step1}), choice and use of correct protocols~(\ref{sec:step2}), and finally the best practices that should be used for the analysis of alchemical calculations~(\ref{sec:step3}).
-Particular focus will be given to aspects of the molecular simulations which are unique to alchemical calculations---these include the calculation of transfer free energies (hydration free energies, partition coefficients, etc.), and binding free energies (absolute and relative).
-%
-While we try to address as many methods and practices as possible, the field of free energy calculations is broad, and there are many advanced topics that are left to future best practices documents focusing on specific issues.
-Below, we provide a non-exhaustive list of topics we have not addressed with some references to provide starting points on these more advanced topics:
-\begin{itemize}
-\item Covalent inhibition.
-\item Free energies of mutation of protein side chains
-\item nonspecific binding or multiple binding sites.
-\item Complex association stoichiometry (beyond simple 1:1 association)~\cite{awesome reference}.
-\item Endpoint free energy methods such as MM-PBSA~\cite{genheden2015mm} and LIE~\cite{gutierrez-de-teran2012linear}
-\item Free energy methods that extract the ligand using geometric order parameters and potential of mean force methods~\cite{heinzelmann2017attachpullrelease}
-\item Forcefield dependence for protein, ligand, ions, co-solvents, and co-factors.
-Different studies have looked at the influence of force fields and it is assumed the user has made an adequate choice for the system under study.~\cite{loeffler2018reproducibility, vassetti2019assessment, lopes2015current}
-\end{itemize}
-%
-For convenience we have also compiled a list of common acronyms, common symbols used, and some essential terminology.
-\begin{tcolorbox}[title=Acronyms, colback=blue!10!white]
-    \textit{CPU} - Central Processing Unit\\
-    \textit{BAR} - Bennett Acceptance Ratio\\
-    \textit{FEP} - Free Energy Perturbation\\
-    \textit{GPCR} - G-Protein Coupled Receptor\\
-    \textit{GPU} - Graphics Processing Unit\\
-    \textit{MBAR} - Multistate Bennett Acceptance Ratio\\
-    \textit{MCSS} - Maximum Common Substructure\\
-    \textit{MD} - Molecular Dynamics\\
-    \textit{RBFE} - Relative Binding Free Energy\\
-    \textit{RMSE} - Root Mean Square Error\\
-    \textit{SAR} - Structure-Activity Relationships\\
-    \textit{TI} - Thermodynamic Integration
-\end{tcolorbox}
-%\label{sec:abbreviations}
-\begin{tcolorbox}[title=List of Symbols, colback=green!10!white]
-U - Potential energy\\
-\end{tcolorbox}
-\begin{tcolorbox}[title=Common terminology, colback=green!10!white]
-    ligand - small molecular binding to a larger host molecule such as a protein.\\ 
-    lead optimization - process of designing a drug candidate from an initial lead compound, considering its affinity and various pharmacokinetic parameters.\\
-    thermodynamic cycle - sequence of thermodynamic processes (like changes in temperature, volume or pressure) in which the initial and final state are equal.\\
-\end{tcolorbox}
-%
-%%%%%%%%%%%%%%%%%%%%
-% Theory basics    %
-%%%%%%%%%%%%%%%%%%%%
-\section{Statistical mechanics can tell us why alchemical free energy calculations work}
-\label{sec:theory}
-\todo[inline, color=green!20]{ASJSM: This section needs to be expanded and give a good introduction to the theoretical background without going into too much detail.}
-Why would you want to run an alchemical free energy calculation and why do they work?
-Suppose you want to compute the binding affinity, or free energy of binding, of a ligand $L$ to a protein $P$, given by:
-\begin{equation}
-P+L\leftrightharpoons PL
-\end{equation}
-The equilibrium rate of binding is given by the ratio of concentrations of product $[PL]$ and reactants $[P]$, $[L]$:
-\begin{equation}
- K_b^{\circ} = c^{\circ}\frac{[PL]}{[L][P]}.
-\end{equation}
-In order to define a reference state, the standard state is usually defined as $1 \mathrm{mol}/\mathrm{L}$ and $c^\circ$ is the standard state concentration. Thus, the binding free energy is given by:
-\begin{equation}
-    \Delta G_{\mathrm{bind}} = -RT\ln K_b^{\circ}.
-\end{equation}
-One way to estimate the binding rate is by computing the probability of finding the molecular system in the bound state or the unbound state, making use of partition function ratios, i.e.
-
-\begin{equation}
-    \Delta G_{\mathrm{bind}} = -RT\ln\frac{\int_\mathrm{bound} \exp(-\beta U(\mathbf{x}))d\mathbf{x}}{\int_\mathrm{unbound} \exp(-\beta U(\mathbf{x}))d\mathbf{x}}
-\end{equation}
-Or more generally in a simplified form the free energy difference between the bound and unbound states is:
-\begin{equation}
-\Delta f \equiv f(\mathrm{bound}) - f(\mathrm{unbound}) = -\ln\frac{Z(\mathrm{unbound})}{Z({\mathrm{bound}})},
-\end{equation}
-where $Z = \int_\mathrm{bound} \exp(-\beta U(\mathbf{x}))d\mathbf{x}$ represents the configurational integral. 
-How can we compute $\Delta G_{\mathrm{bind}}$ from a molecular dynamics simulation?
-Running a direct simulation of the binding events is too computationally costly, meaning we cannot easily estimate the configurational integrals of the bound and unbound state from the simulation.
-This is where alchemical free energy simulations come in.
-We can write down a thermodynamic cycle as seen in ~\ref{fig:fig_binding_thermodynamic_cycle}, for a relative free energy simulation.
-Based on this cycle we can compute the relative free energy of binding between two compounds $A$ and $B$, by going along the arrow of $\Delta G_{\mathrm{bound}}$, where we introduce a second compound that is similar to the first one.
-Now $\Delta G_{\mathrm{bound}}$ can be expressed as a ratio of partition functions of toluene (A) and benzyl alcohol (B).
-The defining characteristic of alchemical free energy calculations is the use of a series of alchemically-modified potential functions $U(x; \lambda)$ in which an alchemical parameter $\lambda$ modulates interactions in a manner that cannot occur in real chemical systems.
-One or more simulations are used to collect data from a multitude of alchemical states to compute a free energy difference between a chemical state ($\lambda_0$) and another chemical or alchemical reference state ($\lambda_1$),
-\begin{eqnarray}
-\Delta f &\equiv& f(\lambda_1) - f(\lambda_0) = - \ln \frac{Z(\lambda_1)}{Z(\lambda_0)} , \label{equation:dimensionless-free-energy-difference}
-\end{eqnarray}
-where the dimensionless free energy $f(\lambda) \equiv \beta F(\lambda)$ is given in terms of partition functions $Z(\lambda)$,
-\begin{eqnarray}
-Z(\lambda) &=& \int dx \, e^{-u(x; \lambda)} .
-\label{equation:partition-function-definition}
-\end{eqnarray}
-Here, the inverse thermal energy $\beta \equiv (k_B T)^{-1}$ where $k_B$ is the Boltzmann constant and $T$ is the absolute temperature, and the \emph{reduced potential} $u(x; \lambda)$~\cite{shirts2008statistically} is generally given by a trace over thermodynamic parameters with their conjugate dynamical variables,
-\begin{eqnarray}
-u(x;\lambda) &\equiv& \beta \left[ U(x;\lambda) + p \, V(x) + \sum_{i=1}^N \mu_i \, N_i(x) + \cdots \right] . \label{equation:reduced-potential}
-\end{eqnarray}
-Here, the collection of thermodynamic and alchemical parameters $\theta \equiv \{\beta, \lambda, p, \mu, \ldots\}$ defines a \emph{thermodynamic state}.
-The physical transformation of interest---such as a protein-ligand association process, or transfer free energy among phases---may involve a thermodynamic cycle that requires several free energy differences $\Delta f$ to be computed in order to produce an estimate of the overall desired quantity.
-
-\begin{figure}
-    \includegraphics[width=0.95\linewidth]{figures/fig2_therm_cyc/Figure.pdf}
-    \caption{{\bf Thermodynamic cycle for computing the relative free energy of binding ($\Delta \Delta G$) between two related small molecules to a supramolecular host.}
-    The relative binding free energy difference between two small molecules, $\Delta \Delta G_{\mathrm{bind}, A \rightarrow B} \equiv \Delta G_{\mathrm{bind}, B} - \Delta G_{\mathrm{bind}, A}$---here benzyl alcohol (top) to toluene (bottom)---can be computed as a difference between two alchemical transformations, $\Delta G_\mathrm{bound} - \Delta G_\mathrm{solvated}$, where $\Delta G_\mathrm{bound}$ represents the free energy change of transforming $A \rightarrow B$ in complex and $\Delta G_\mathrm{solvated}$ the free energy change of transforming $A \rightarrow B$ in solvent.}
-    \label{fig:fig_binding_thermodynamic_cycle}
-\end{figure}
-
-%%%%%%%%%%%%
-% Step 0   %
-%%%%%%%%%%%%
-\section{What can be expected from alchemical simulations?}
-\label{sec:step0}
-\todo[inline, color={green!20}]{ASJSM: @everyone please review this section}
-When starting an alchemical free energy project you need to determine if an alchemical free energy calculation is the appropriate strategy by initially counting the cost of your
-project: Can you even hope to tackle the problem with available resources and, if successful, will it
-be worth it?
-\subsection{How accurate are alchemical free energy calculations?}
-\label{subsec:accuracy}
-%
-Current alchemical free energy calculations seem to achieve, in favorable cases, RMS errors around 1-2 kcal/mol depending on force field, system, and a variety of other factors such as simulation time, sampling method, and whether the calculations employed are absolute or relative.
-A small selection of benchmark datasets and case studies can be found in section~\ref{sec:benchmark} at the end of this document.
-However, the domain of applicability is a significant concern~\cite{sherborne2016collaborating, cournia2017relative}, especially for relative calculations, which typically require a bound structure of a closely
-related ligand as a starting point.
-Additional factors such as slow protein or ligand rearrangements, uncertainties in ligand binding mode, or charged ligands can make these calculations far less reliable and more of a research effort.
-%
-It is worth noting that the accuracy of free energy calculations is highly variable across different protein targets, and likely across different ligand chemotypes as well.
-For instance, FEP+ with OPLS3 achieves an RMSE of 0.62 kcal/mol for a set of 21 compounds binding to JNK1 kinase, yet only an RMSE of 1.05 kcal/mol for a set of 34 compounds binding to P38$\alpha$ kinase~\cite{harder2016opls3}.
-Furthermore, perturbations for the same chemotype in different pockets of the BACE enzyme gave varied errors~\cite{keranen2017acylguanidine}
-\todo[inline,color={green!20}]{JDC: It's important to be clear on what error is being reported: $\Delta G$ after shifting by a constant to minimize the RMSE, unshifted $\Delta G$, $\Delta \Delta G$ of computed edges, or $\Delta \Delta G$ of all edges.
-Note that computed edge $\Delta \Delta G$ RMSE significantly underestimates the error of all edge RMSE: \url{https://github.com/jchodera/jacs-dataset-analysis} ASJSM: Can we incorporate your analysis in this section as an example?
-Maybe this is something that needs to be addressed properly in the analysis section and only referenced here?}
-We thus recommend that retrospective study for a particular target and a particular chemical series be performed to establish the relevant accuracy of free energy calculations for each particular application case.
-%
-\subsection{How reproducible are alchemical free energy calculations?}
-\label{subsec:reproducible}
-Finite computing resources necessarily limit the generated number of uncorrelated samples of potential energy surfaces, and therefore alchemical free energy calculations only give free energy estimates to within finite precision.
-An important consideration is how reproducible alchemical free energy calculations are in practice.
-In simple cases such as absolute hydration free energies of small organic molecules, or relative hydration free energy calculations between structurally similar small organic molecules, it should be possible to obtain highly precise estimates with a given software package (with a sample standard deviation under 0.01 kcal/mol).
-For more complex use cases such as protein-ligand binding free energies the repeatability is often substantially worse.
-A good practice is to perform two or three runs of the same perturbation to assess repeatability with a given protocol.
-The sample standard deviation will give a crude estimate of the reliability of the estimates, and whether the precision is sufficient for the problem at hand.
-Where this is practical, a more stringent test is to use different input coordinates for each repeat run.
-%
-Note that these issues concern calculations carried out with a single software package, but simulation package variations can introduce additional issues.
-Such issues of reproducibility of free energy calculations across different simulation packages have attracted attention recently.
-Greater variability is expected due to methodological differences such as integrators, thermostats, barostats, treatment of long-range electrostatics, and potentially other factors.
-For absolute and relative hydration free energies of small organic molecules a variability of ca.
-0.2 kcal/mol between popular simulation packages has been reported.
-\cite{REF Reproducibility of hydration free energies Loefler et al.
-JCTC 2018} In the recent SAMPL6 SAMPLing challenge a larger variability of 0.3 to 1.0 kcal/mol was noted in the computed absolute binding free energies of host/guest systems even though similar input and simulation parameters were used \cite{Ref Rizzi et al SAMPLing biorxiv} and, in many cases, single-point energies were identical or nearly so.
-Further work is needed to ensure reproducibility of alchemical free energy calculations across implementations to guarantee that force-field development efforts lead to transferable potential energy functions.
-%
-\subsection{Is my problem suitable for alchemical free energy calculations?}
-\label{subsec:suitability}
-Before even planning free energy calculations to study binding to a
-particular target, it is important to assess what is known about the
-system and its timescales and its suitability for free energy
-calculations, as well as the \emph{purpose} of the calculations and
-the amount of available computer resources. In some cases, predicting
-accurate binding free energies for a particular target might be a
-\emph{more} challenging effort than simply measuring them! This is
-often the case when dealing with database screening problems, where
-compounds might be easily and quickly available commercially for
-testing and free energy calculations could consume a much larger
-amount of resources. Free energy calculations thus typically only
-appeal when (slow or costly) synthesis would be required to do
-experiments or experiments are otherwise cost-prohibitive.
-%
-Sometimes free energy calculations can provide answers that are not
-readily available from experiments.  For example, type II kinase
-inhibitors selectively bind to different kinases in the so-called
-DFG-out conformations (cite Kuriyan).  The selectivity of such
-inhibitors may be attributed either to their differential binding to
-different kinases in the DFG-out conformations, or to different
-stability of the DFG-out conformations of different kinases.  Let
-$K_C$ be the equilibrium constant between DFG-in and DFG-out
-conformations of one kinase, and $K_D^\ast$ be the dissociation
-constant of a type II inhibitor against this kinase, the apparent
-binding constant of this inhibitor against this kinase is then
-\begin{equation}
-  K_D = K_D^\ast \frac{1 + K_C}{K_C}
-  \label{eqn:conformational-binding}
-\end{equation}
-%
-Since binding experiments cannot resolve $K_D^\ast$ and $K_C$ individually, such experiments cannot address the basis of selectivity of the type II inhibitors.
-Absolute binding free energy calculations, in contrast, can take advantage of the slow kinetics of DFG-in/out conversion, and estimate the conformation-specific binding constant $K_D^\ast$, thus yielding clues as to the source of selectivity.
-%
-\subsection{Is the expected accuracy of the computation sufficient?}
-\label{subsec:accuracy}
-The requisite level of accuracy is another important consideration ---if the
-goal is to guide lead optimization when many compounds will be
-synthesized, free energy calculations can be appealing even with
-accuracy in the 1--2 kcal/mol range~\cite{mobley2012perspective}, but if the number of compounds to be synthesized is very small, this accuracy may not be enough to provide
-much value.
-%
-Here we include a simple estimate of the value provided by alchemical
-free energy calculations in lead optimization.  Let $P(\Delta\Delta
-G)$ be the probability distribution of the changes in the binding free
-energies of a new set of molecules during one round of lead
-optimization, and let $P(\Delta\Delta G^\dagger|\Delta\Delta G)$ be the
-conditional probability of the binding free energy change computed by
-the free energy calculations, $\Delta\Delta G^\dagger$, given the actual
-change $\Delta\Delta G$.  The latter conditional probability can be modeled
-by a normal distribution
-\begin{equation}
-  P(\Delta\Delta G^\dagger|\Delta\Delta G) = \frac{1}{\sqrt{2\pi\sigma^2}}
-  \exp\left(-\frac{(\Delta\Delta G^\dagger - \Delta\Delta G)^2}{2\sigma^2}\right),
-  \label{eqn:free-energy-distribution}
-\end{equation}
-where $\sigma$ signifies the accuracy of free energy calculations.
-Here we assume that there is no systematic bias in the free energy
-calculations, i.e., on average, the free energy change computed by
-free energy calculations agrees with the actual free energy change.
-%
-In lead optimization guided by free energy calculations, we only
-synthesize and experimentally test molecules that are predicted to
-have favorable free energy changes.  We are thus interested in how
-often that a molecule predicted to bind stronger actually turns out to
-bind stronger.  In other words, we are interested in the conditional
-probability:
-\begin{equation}
-  P(\Delta\Delta G<0|\Delta\Delta G^\dagger<0).
-  \label{eqn:true-positive}
-\end{equation}
-%
-For illustrative purposes, we assume that the actual changes in the
-binding free energies for a set of new molecules also follow normal
-distribution, that the standard deviation in the changes is $RT\ln 5$
-(corresponding to a 5-fold change in the binding affinities), and that
-1 in 10 new molecules have increased binding affinity ($\Delta\Delta G
-\leq 0$).  Under such assumptions, the conditional probability in
-Eq.~\ref{eqn:true-positive} can be easily computed.  If the accuracy
-of free energy calculations is $\sigma = 1$ kcal/mol, $P(\Delta\Delta
-G<0|\Delta\Delta G^\dagger<0) = 0.35$, which means that out of every
-10 molecules selected for predicted favorable free energy change, on
-average 3.5 molecules will have actual favorable free energy change.
-In other words, selection by free energy calculations yields 3.5 times
-more molecules of improved affinities than selection without free
-energy calculations.
-%  
-Available computational resources and timescales of motion also factor
-into this initial analysis. An individual free energy calculation
-involves simulations at many different intermediate states (perhaps
-20-40 or more) and each of these must typically be long enough to
-capture the relevant motions in the system. If such motions are
-microsecond events or longer, the computational cost of running 20-40
-microsecond or longer simulations for each of $N$ ligands may become
-prohibitive, or at least should be carefully considered. 
-%
-\subsection{Can I afford the calculation?}
-\label{subsec:affordability}
-Furthermore, are available computational resources sufficient that throughput will be
-reasonable compared to needs of experimental collaborators working on
-this system? How many ligands ($N$) can you afford to handle given
-your computational resources? As cloud computing becomes more available, in-house GPU clusters may not be necessary if calculations are not run on a regular basis.
-This analysis should be done up front, part of ``counting the cost''
-of involvement in a particular project. In some cases, the analysis may conclude that free energy calculations will not be feasible for the proposed problem.
-Here, by ``cost'', we refer not just to financial cost of the calculations relative to experiments, but also time -- can the calculations can be run faster than experiments are done?
-How will the relevant resource and opportunity costs factor in?
-Both computation and experiment require human time, supplies (of different sorts), and equipment.
-In the extreme limit, for example, it would not make sense to spend a month running a binding free energy calculation if the equivalent experiment could be done in a day with resources already on hand.
-Such issues should be considered before deciding to conduct binding free energy calculations.
-%
-\subsection{Is an exploratory study what I want?}
-\label{subsec:exploration}
-An additional consideration is how much is known about your particular
-target, ligand binding modes in the target, and any relevant motions
--- essentially, has it been studied enough to know whether it might be
-suitable for free energy calculations?
-It is important to know if the system has hardly been studied, because should the initial calculations perform poorly, the effort may turn into an attempt to understand the relevant sampling, force field, or system preparation problems.
-%
-If you are unsure whether your project is feasible, one option may be
-to conduct a short exploratory study to assess tractability for just a very small
-number of ligands. This can be sufficient to get an initial
-idea of feasibility and accuracy of the calculations for the
-proposed target.
-%
-\section{How should alchemical simulations be applied to drug discovery?}
-\label{sec:drugdiscovery}
-\todo[inline, color={green!20}]{ASJSM: Merge/tidy of what can I expect from my simulation, drug discovery aspects and simulation prerequisites, there are some duplications there at the moment, happy for restructuring ideas.
-}
-%
-Many practitioners expect alchemical methods to provide valuable guidance for drug discovery, and to exhibit accuracy superior to most  alternative approaches for suitable targets~\cite{kuhn2017prospective}.
-It has been stressed that successful application in industry may require considerable knowledge of the "domain of applicability" of free energy calculations -- where they work well and where they will not ~\cite{sherborne2016collaborating}.
-Successful application also requires robust protocols for preparing, submitting and analysing alchemical calculations.
-In this regard, the issues mentioned in the previous section such as understanding the suitability and timescales to capture the structure activity relationships (SAR), and performing up-front tests of performance are all relevant to drug discovery applications.
-Without venturing too far into details of system setup, which is beyond the scope of this article, we highlight some critical factors affecting accuracy and successful application.
-%
-\subsection{Capturing the experimental conditions}
-The calculations aim to capture the alchemical change from one ligand to another as accurately as possible.
-Therefore, it is necessary to consider details of the experimental setup, such as pH.
-Biological assays are usually run at neutral pH but this is not always the case, and protein and ligand preparation protocols should reflect this.
-In more detail, the formal charge, or tautomeric state of the small molecules can change within a series of analogs.
-The change of pKa is a common medicinal chemistry strategy to modify drug properties and would normally require explicit effort to incorporate into alchemical calculations.
-Also, to ensure modeling matches experiment, we need to ensure that we simulate the same system -- which requires understanding what protein construct is used in the bioassay.
-For instance, does the X-ray that is to be used for the calculations match the construct used for screening (i.e.
-catalytic vs full length, monomer vs dimer etc) (CITE https://www.nature.com/articles/s41598-018-23039-5)?
-Also, were certain co-factors or partner proteins required in the bioassay?
-%
-\subsection{Is my binding mode accurate?}
-As also mentioned, good performance of alchemical FE calculations requires an accurate representation of the ligand binding mode, which often is estimated based on X-ray crystallography for a related ligand.
-If more than one structure is available, the modeler should pay attention to choose the most suitable.
-The quality of the structure can be a concern, and the reader is referred to work of Warren et al.
-for a detailed discussion of choosing optimal structures for structure-based modeling~\cite{warren2012essential}.
-It is also useful to study the structure activity relationship and understand the expected impact in the binding site, whether side chain movement in the protein will be required, and whether there evidence of this in any alternative X-ray structures of the same protein.
-Often, only one protein and water configuration is used for a series of alchemical FE calculations, so this needs to be capable of accommodating the smallest through to largest ligands in a way that allows stable and well behaved simulations.
-This can be a practical limitation for the alchemical changes that are feasible, though a simple work-around can be to separate compounds into sub-series for different calculations.
-If multiple structures are available there is some evidence the higher affinity complex can give better performance~\cite{perez-benito2019predicting}, at least in some cases.
-However, ligands and proteins can also undergo unexpected changes in binding mode for related ligands, which can make these issues more complex to deal with.
-
-\subsection{Input setup and scale of calculations}
-In a drug discovery setting it is normal to consider dozens (or more) of ligands and it is necessary to align them in the binding site.
-There is no detailed study of how different alignment approaches may affect results, but the user should be aware of some practical considerations.
-Tools are available to compare the ligands and build the hybrid topologies that define the changes between one ligand and another~\cite{loeffler2015fesetup,hedges2019biosimspace,gapsys2015pmx}.
-In simple terms, providing poor alignment to these tools will make this job harder.
-Therefore, docking ligands with no restraints is not usually the best method for defining the ligand inputs, instead a core restrained docking can provide satisfactory outcomes, but in this case one still should pay careful attention to consistency of alignment for identical substituents.
-Another alternative is to manually edit the same core with the changing substituents.
-This provides assurances that coordinates for the non-perturbed structures remain identical and aromatic substituents for instance obey consistent dihedral angles but is not feasible for many compounds and therefore automation can be desirable.
-
-Finally, the role of water in ligand binding is now well understood and it can be crucial to capture the changes in binding site solvation during ligand binding.
-Can crystallographic waters be retained?
-Do they clash with some of the larger ligands used in the FEP?
-If so, how will the water be incorporated for smaller ligands that require the water – consider pre-solvation methods such as GCMC steps~\cite{michel2010prediction}.
-Other tools such as WaterMap or open source equivalents (SSTmap, others) can be used to define water structure for systems with no experimental evidence of water sites.
-Before launching large numbers of alchemical free energy calculations it is recommended to test the system using classical MD simulations and limited numbers of alchemical perturbations.
-Metrics such as ligand and protein RMSD and RMSF can be inspected, along with visual inspection of simulations, to ensure the system is stable and likely to be suitable for FE calculations.
-
-Running binding FE calculations in a drug discovery application will typically require the use of software or tools to facilitate the calculations of many compounds.
-The commercial implementation from Schr\"{o}dinger, FEP+, uses a graphical user interface and once the user has calculated any missing force field parameters, calculations can be prepared and submitted within minutes.
-The software builds the network of perturbations, connecting similar compounds based on LOMAP methodology~\cite{liu2013lead}, all inputs for the Desmond simulations are built automatically and the job can be saved or run immediately on available GPUs.
-The cost of this software is often prohibitive but alternative tools are becoming available to allow relatively straight forward running of alchemical FE calculations with other software~\cite{gapsys2015pmx, loeffler2015fesetup, song2019using, gapsys2020large,  jespers2019qligfep}.
-One consideration at this point is to define the perturbations suitable for the application.
-For instance, if the goal is to accurately assess the relative binding energy of a small number of compounds, possibly with challenging synthesis, the map of perturbations may contain as many connections between compounds as affordable.
-However, when running calculations on hundreds of compounds a so called ‘star-map’ can be used that just contains one connection per compound: perturbing every compound to a central ligand, typically the crystal structure ligand~\cite{konze2019reactionbased}.
-In this way the top-ranking examples can be readily identified and submitted to additional calculations in a second round.
-%
-\subsection{Making predictions, understanding errors}
-\label{subsec:predictions}
-For prospective drug discovery applications there are several other considerations including understanding likely errors and taking selection bias into account.
-It is crucial when proposing compounds for synthesis to have some idea of the underlying error or uncertainty in the predictions.
-A retrospective assessment can give an indication of prospective performance for similar molecules~\cite{ciordia2016application}.
-Beyond this, several parameters provide useful indicators of performance.
-Results suggest MBAR errors greater than 0.3 kcal/mol for perturbations can be indicative of inaccuracy~\cite{perez-benito2019predicting}.
-In addition, the hysteresis can be checked by comparing the difference for forward and reverse perturbations of the same structural change, or in the case of large maps of perturbations cycle closure errors~\cite{wang2013modeling} indicate problematic perturbations involved in cycles connecting many compounds.
-Selection bias, as described above, can be problematic in prospective applications as there is normally an aim to focus on only a narrow range of predicted activity (i.e.
-the most active).
-Thus, correlation between prediction and subsequent experiment may lead to disappointment but corrections can be applied based on previous recommendations~\cite{abel2017critical}.
-
-In summary, the successful use of alchemical calculations in drug discovery requires working in the domain of applicability, using a high quality X-ray structure of the target compounds, and testing the approach retrospectively to ensure the system setup is well-behaved.
-Always assess your confidence in the resulting predictions and communicate this when discussing with experimentalists.
-Consider performing repeat calculations for at least some of the perturbations in the study.
-In a drug discovery setting it is inevitable to consider larger virtual libraries.
-In these cases again evaluate if shorter simulation time per $\lambda$ window can be sufficient for a ‘virtual screening’ mode of application.
-It is expected that future tools may allow efficient exploration of large virtual libraries.
-There are many accounts of success of alchemical FE calculations, the methods show good performance towards the goal of binding energy prediction.
-However, it is important to have realistic expectations.
-
-Structure based drug design projects are often capable of improving potency relatively quickly, even with only limited application of computational approaches and the range of activity narrows to just two-to-three log units.
-It may seem hard to have impact with substantially different, more potent, stand-out compounds in this scenario, but binding energy prediction can still be extremely useful for ensuring activity is maintained as other properties are optimized.
-An interesting cost benefit analysis has shown the value of activity prediction, see discussion above and articles such as~\cite{mobley2012perspectiv}.
-From a drug discovery point of view, FEP is expanding its domain of applicability, and there are reports of success using homology models and GPCRs for instance, as well as enabling charge change and scaffold hopping, but these systems are undoubtedly more difficult.
-In the meantime, the use cases are expanding to resistance prediction, disease mutation prediction, solubility prediction – an exciting future for alchemical FE calculations.
-%
-%
-%
-%%%%%%%%%%%%
-% Step 1   %
-%%%%%%%%%%%%
-%
-\section{Simulation prerequisites}
-\label{sec:step1}
-%
-The alchemical free energy protocols, that are discussed in Section~\ref{sec:step2}, are defined for a specific type of free energy calculation, i.e.
-a free energy of binding or a free energy of hydration.
-Depending on the type of simulations different considerations have to be made for ligands, solvent, and host molecules (in the case of the estimation of free energies of binding).
-\subsection{Free energies of binding}
-\label{subsec:binding}
-In principle in the limit of sufficient conformational sampling, the free energy changes estimated from an alchemical free energy calculation should be independent of the system's initial coordinates.
-However, in practice because simulations of thermodynamic states are of finite duration (typically currently 1-100 ns per state) this is only true for certain classes of alchemical free energy calculations such as relative or absolute free energies of hydration of small and relatively rigid organic molecules.
-Host-guest or protein-ligand complexes typically exhibit slowly relaxing degrees of freedom that exceed significantly the duration of an alchemical free energy calculation.
-It is therefore generally important to carefully select input coordinates to obtain satisfactory results.
-The following questions may be relevant before diving into the simulation setup.
-% 
-\begin{itemize}
-    \item Do I have one or multiple good host structures? (e.g. a good resolution X-ray crystal of the protein target)
-    \item Do I have information on one or all of the ligands binding sites (e.g. a X-ray structure)
-    \item Should I include buried waters, or other small molecules that can be found in an X-ray structure.
-    \item Are my ligands part of a congeneric series? (simple R group substitutions around the same scaffold)
-    \end{itemize}
-%
-\paragraph{Are there good X-ray structures available?}
-As for any simulation, care should be taken in selecting available X-ray structures in the protein databank and potentially choosing multiple structures initially, to account for variability in receptor conformations as well as accuracy of available X-ray structures.
-Typically, clustering of receptor structures can be used to identify different receptor conformations near the binding site, as well as assessing relevant side chain placements from the X-ray structure, see for example~\cite{mey2016blinded}.
-In terms of set up and other choices following general best practice guidelines is advisable~\cite{braun2018best}.
-%
-For binding free energy calculations using a congeneric set of ligands and therefore likely to be most suited to a relative free energy protocol (see~\ref{sec:step2}).
-Other than the choice of forcefield for the parameterization of the ligands, some care has to be taken selecting binding poses for these ligands.
-Generally a common assumption for a congeneric series is that the binding mode is conserved.
-Therefore, if an X-ray structure of one of the ligands is available, this should be used to position the ligands in the putative binding site in an energetically reasonable conformation that entails no steric or electrostatic mismatch with the receptor.
-Checking the X-ray structure versus the experimental electron densities is important, as the resolution of part of the ligand or important sidechains may be based on the interpretation of the crystallographer rather than the available electron density.
-For example, looking at a cyclohexane ring density, a chair conformation is vastly more likely than that of a boat and may just be poorly assigned.
-%
-\paragraph{How to deal with different binding modes?}
-Generally, binding modes within congeneric series are conserved~\cite{}, however, exceptions exist~\cite{}, and weaker binding compounds are more likely to adopt different binding modes (REFS).
-There may also be ambiguity about the relative orientation of substituents.
-A known issue involves a 180 degree flip in the dihedral angle of a 6-membered aromatic ring leading to a different spatial position of ortho- or meta- substituents that otherwise should overlap within a series.
-The 180 degree flip of the ring may not occur enough during simulations to overcome the starting configuration bias.
-Another scenario may be equatorial and axial substituted saturated rings (e.g.
-cyclohexane derivatives).
-This situation may be addressed by explicitly modelling different binding modes of the same ligand and combining later computed free energy differences for different binding modes into a relative free energies of binding~\cite{kaus2015how}.
-%
-\paragraph{Have you considered sterioisomers and enationmers?}
-Congeneric series can contain stereoisomers or enantiomers, making sure that you model the correct one is important.
-For racemates, the relative abundance of each stereoisomer is normally not known.
-Therefore, the experimental activity associated with just one conformation is more uncertain.
-However, the modeling typically uses just the bioactive conformation that best fits the active site.
-Clearly this introduces the chance for higher error versus experiment.
-Nevertheless, if all compounds in the congeneric series are racemic, originating from similar synthetic procedures with an expected similar abundance of stereoisomers, then the differences may cancel and the trend in calculated and observed binding energies may be robust.
-Nevertheless, we can see that care and further testing is needed in this scenario, and the quality of the predictions may suffer.
-%
-\paragraph{Conserved binding site waters can play an important role in binding free energies}
-Another aspect to look out for are binding site water molecules that may form water mediated protein-ligand interactions, particularly in situations where exchange with bulk water may be slow on the simulation timescales.
-This happens typically in buried binding site.
-Overlaying multiple protein X-ray structures can identify conserved or just additional water molecules that can be useful to include.
-In cases where water molecules are known to play an important role in the binding, software implementations that use water sampling facilitated by Grand Canonical Monte Carlo methods may be useful.
-For examples see HSP90 as formed part of the D3R grand challenge 2015~\cite{mey2016blinded},A2A~\cite{brucemacdonald2018ligand}, MUP~\cite{ross2015water}, and others~\cite{michel2009energetics}.
-%
-\paragraph{How to align congeneric series}
-Input coordinates for a congeneric series may be generated by docking calculations, or by ligand alignment using maximum common substructure algorithms.
-The latter tends to produce alignments that are more conserved and more consistent free energy changes across a dataset, but will struggle to yield reasonable results for relative binding free energy calculations that involve a significant binding mode rearrangement.
-This may also lead to steric clashes with the receptor coordinates of the reference ligand if structural rearrangements are needed to accommodate different members of the congeneric series.
-Small steric clashes may be resolved during subsequent simulation equilibration prior to data collection but there is a risk that the complex relaxes to an alternative metastable state.
-%
-An additional consideration arises for single topology relative free energy calculations.
-In this class of alchemical free energy calculations it is necessary to generate a molecular topology that may describe the initial and final states of the perturbation (see different topologies as seen in Fig.~\ref{fig:fig_topology}).
-In cases where the end-states have high topological similarity and high structural overlap this is relatively straightforward and typically handled by use of maximum common substructure calculations.
-In situations where the end-state topologies differ significantly, or there is relatively little spatial overlap between the two end-states some user intervention may be necessary to produce a satisfactory input topology.
-%
-If the binding site location is uncertain but the structure of the receptor is well defined and plausible binding sites are identified, it may be more useful to chose an absolute free energy protocol to compute the standard free energy of binding of the ligand to a set of binding sites.
-This requires the user to prepare input files describing the bound conformation in different putative binding sites (REF).
-The apparent binding free energy of the ligand may be obtained by combining the individual binding site free energies, which also indicate where the ligand is more likely to bind.
-In this case a docking program can generate initial structures.
-Different commercial and none commercial tools are available, such as: rDock, Autodock Vina, Glide, or Flare to name a few~\cite{ruiz-carmona2014rdock, trott2010autodock, friesner2004glide, cheeseright2006molecular}.
-If the putative binding sites are not apparent, for instance due to significant induced-fit effects, it may be challenging to obtain meaningful free energies of binding.
-One may have to account for the free energy change for forming a binding site in the target receptor which may not be feasible in FE simulation timescales.
-Another key application for absolute binding free energies is to study chemically different ligands, that do not share a common scaffold, but bind at the same site or the same site of similar proteins~\cite{aldeghi2015accurate}.
-%
-\subsection{Free energies of hydration or partition coefficients}
-\label{subsec:hydration}
-Preliminary considerations necessary for using free energy methods to compute partition coefficients are generally more straight forward.
-For example, a 3D minimised structure of the solute can be generated with a simple tool such as babel and solvated to prepare the input  to compute a free energy of hydration.
-However, in these cases a careful choice of forcefield, as well as water models or organic solvents is essential.
-See for example~\cite{bosisio2016blinded,rustenburg2016measuring} for a good discussion on these choices.
-%
-%
-%
-%%%%%%%%%%%%
-% Step 1   %
-%%%%%%%%%%%%
-%
-\section{What simulation protocol should I choose? -- Step 2}
-\label{sec:step2}
-Alchemical free energy calculations can be grouped into two main categories, ``absolute'' (\ref{fig:fig_absolute_thermodynamic_cycle}) and ``relative'' \footnote{The distinction is a bit of a misnomer, since both compute ratios of partition functions relative to another state, and neither computes an absolute free energy.}(\ref{fig:fig_binding_thermodynamic_cycle}), which differ in whether they compute properties for a single molecule (absolute) or compare properties of different, usually closely related, molecules (relative).
-To use binding as a concrete example, in absolute binding free energy calculations, one computes the binding free energy of a ligand to an individual receptor relative to a standard reference concentration.
-In contrast, in relative binding free energy calculations, one compares the binding free energy of two related inhibitors to determine the potency difference.
-\subsection{Absolute and relative free energy calculations have some differences}
-Many of the issues of setup and protocol for alchemical calculations are common, but some are different between absolute and relative calculations, so before treating the common elements we consider the protocol differences.
-
-\subsubsection{Choices unique to relative free energy calculations}
-\label{sec:relative-fe-protocol}
-
-\paragraph{Topologies} A critical first step in relative calculations is to select an approach to these calculations, determining whether to use a \emph{dual topology}, \emph{single topology}, or \emph{hybrid topology} approach to relative calculations.
-
-The distinction between these can be illustrated by considering a hypothetical transformation from molecule A to molecule B, where both atoms share a common substructure but differ in their substituents; e.g.
-consider a transformation of benzene (A) to benzyl alcohol (B) Fig.~\ref{fig:topology}.
-In this case the common substructure is the benzene ring, though perhaps may be larger depending on how it is defined, as we discuss below.
-In single topology calculations, the overall transformation is set up to involve as few additional atoms as possible, so benzene would be typically changed into benzyl alcohol by changing one of the hydrogens into a carbon.
-To this new carbon you also find non-interacting atoms called ``dummy atoms'' (retaining their bonded interactions but not interacting with the rest of the system), which will become the two hydrogens attached to the new carbon and the OH group.
-Bond parameters as well as partial charges are adjusted accordingly.
-Thus, in a single topology calculation, atoms may change their type so relatively few dummy atoms are created. This is illustrated in left arm of Fig.~\ref{fig:fig_topology}. 
-In contrast, in a dual topology free energy calculation, no atoms are allowed to change type~\cite{shirts2012best}.
-This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxy methyl group, then passing through an intermediate state where atoms which are becoming dummy atoms or ceasing being dummy atoms are partially interacting (this state may or may not be well defined~\cite{mobley2014blind}  and culminating in a state where benzyl alcohol is present along with the additional dummy atoms which was previously a corresponding hydrogen of the benzene.
-Fig.~\ref{fig:fig_topology}'s right branch depicts how such a dual topology works.
-\begin{figure}[h!]
-    \includegraphics[width=0.95\columnwidth]{figures/fig3_topol/Figure.pdf}
-    \caption{\textbf{Two common topologies for alchemical calculations} left: A single topology uses softcore potentials to convert from one type of atom to an other.
-Dummy atoms (Du) are used when there is no corresponding maximum common substructure match between the two molecules for certain atoms.
-right: The dual topology does not convert one species to another, but only converts between Du atoms and an interacting species, but usually uses softcore potentials for this.
-The 'mixed' intermediate atoms are used in both dual and single topology approaches.
-Only the way the transformation occurs and the end states differ.
-Following the arrow along the left and right illustrate the differences.
-Figure adapted from \url{http://www.alchemistry.org/wiki/Constructing_a_Pathway_of_Intermediate_States}}
-    \label{fig:fig_topology}
-\end{figure} 
-Hybrid topology calculations have seen much less use but essentially consist of two absolute free energy calculations in opposite directions at the same time (turning one molecule off while turning the other on)~\cite{jiang2019computing}.
-At present, the most widely used approaches, such as in Schr\"{o}dinger's FEP+~\cite{wang2019protein} and in FESetup~\cite{loeffler2015fesetup} (for which calculations may be planned with Lead Optimization Mapper (LOMAP)~\cite{liu2013lead} seem to use single topology approaches, though some codes only support dual topology.
-To our knowledge efficiency differences have not been thoroughly explored, though conventional wisdom suggests that fewer dummy atoms are better~\cite{liu2013lead,mobley2012perspective}.
-%
-\paragraph{Atom mapping}
-Once a particular approach to the topology is selected, a crucial next step is to identify the common atoms which will not be perturbed.
-Rigorously, this process essentially comprises a maximal common substructure (MCSS) search of the molecules involved to identify the common substructure -- though the parameters of the MCSS search will differ depending on whether single or dual topology calculations are planned.
-Specifically, with a single topology approach in mind, atom types are allowed to change, so a permissive MCSS search can be done, whereas with dual topology a more strict search is required.
-\begin{figure}[h!]
-    \includegraphics[width=0.95\linewidth]{figures/fig4_mcs/Figure.pdf}
-    \caption{\textbf{Illustration of maximum common substructure matches} MCSS is shown in green for when A: a restrictive MCSS match is used and in B: ring breaking is allowed.}
-    \label{fig:fig_mcss}
-\end{figure} 
-%
-There are different tools that allow the generation of MCSS matches as well as single topology input.
-A large number of software can compute MCSS matches, some rely on RDKit, such as LOMAP or FESetup and partially BioSimSpace, others such as fkckombu are standalone tools~\cite{rdkit2019Dec, loeffler2015fesetup, liu2013lead, hedges2019biosimspace, kawabata20143d}.
-Schr\"{o}dinger's FEP+ planning tool is based on a version of LOMAP, and it also uses MCSS matching as well as 3D considerations to plan the network of single topology calculations between molecules~\cite{wang2015accurate}.
-%
-MCSS searches can be relatively time consuming, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial assessment and then use MCSS only to identify final mappings for relative calculations.
-%
-The MCSS approach, though relatively standard, takes into account only topological similarity.
-It is possible that changes in binding mode could actually require a different choice of mapping, so in some cases mappings may need to be planned differently depending on 3D positioning of atoms in space.
-%
-Single topology relative calculations, and calculations based on substructure searches, only work if in fact the ligands share a common substructure, e.g. are part of a congeneric series.
-If no common substructure is shared, then alternative dual or hybrid topology free energy calculations are needed, where one would co-localize a pair of compounds in a binding site, exclude their interactions with one another, and compute the relative binding free energy by turning one molecule on from being dummy atoms while turning the other off.
-To our knowledge no general pipeline for such calculations yet exists and this would likely remain a research problem.
-Using an absolute style free energy calculation instead seems a more promising approach in such a case.
-%
-\paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic, in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle~\cite{}, which may not be true whenever dummy atoms are involved in rings.
-Some approaches have attempted to address this~\cite{clark2019relative} but a general solution is not yet in mainstream use.
-%
-\paragraph{Perturbation maps}
-Based on the input ligand series a perturbation map or network can be planned.
-Recent heuristics have shown the more connected the perturbation network the better, however, there is a way to optimise network structure while minimising the number of perturbations that need to be computed minimising the resulting computational cost~\cite{yang2019optimal}.
-Sometimes the introduction of intermediates that are not part of the original congeneric series are essential to avoid ring breaking, or dealing with perturbations that would otherwise result in large numbers of atoms to be grown or disappearing.
-Some commercial tools have underlying good heuristic but may fail with complicated input, needing user validation in particular when dealing with chiral compounds.
-\begin{figure}[h!]
-    \includegraphics[width=0.95\columnwidth]{figures/fig6_types_of_networks/Figure.pdf}
-    \caption{\textbf{Examples of perturbation networks} A: Star shaped network with the crystal structure in the center.
-B: Network with cycle closures.
-Arrows indicate the direction of the perturbation.
-Green cycles indicate good cycle closure, red poor cycle closure.
-The red arrow indicates a poorly converged simulation that would give rise to bad cycle closures.
-The diamond indicates the use of a crystal structure binding mode.}
-    \label{fig:fig_types_of_networks}
-\end{figure} 
-In some cases, during the lead optimization stage, or if perturbations would be rather large a star shaped network as seen in Fig.~\ref{fig:fig_types_of_networks} (A) is used..
-However, adding redundancy into the network means that a better error analysis can be carried out, by looking at cycle closure errors as discussed in sec.~\ref{subsec:predictions}, with an example given in Fig.~\ref{fig:fig_types_of_networks}.
-
-Methods in experimental design have been applied to the construction of the perturbation maps.
-Yang et al.~\cite{yang2019optimal} optimized the perturbation map by selecting a fixed number of calculations from the pairwise perturbations so that the resulting set of calculations minimize the total variance.
-Xu~\cite{xu2019diffnet} optimized the perturbation map by allocating different amounts of simulation time to different pairwise perturbations so as to minimize the total variance, given the total simulation time of all the perturbation calculations.
-Both approaches lead to substantial reduction in the statistical error of the estimated free energies.
-%
-\paragraph{Constraints and relative free energy calculations.}
-One issue which requires particular care is the use of constraints.
-Commonly, bonds involving hydrogen are constrained to a fixed length using algorithms such as SHAKE or LINCS, allowing the use of longer timesteps~\cite{}.
-However, in single topology relative free energy calculations, the atoms involved might be mutated to other atom types -- for example, in a mutation of methane to methanol, one hydrogen might become an oxygen atom.
-Typical molecular dynamics engines are not set up to recognize this change, or at least not to correctly include contributions to the free energy from changing constraints/constraint length, so results for a transformation would usually be erroneous.
-At present the most general solution to this problem is simply to avoid the use of constraints (and thus use a smaller timestep if necessary, usually of around 1 fs) in any relative free energy calculation involving a transformation of a constrained bond.
-%
-\subsubsection{Absolute free energy calculations must handle the standard state and use restraints}
-\label{sec:standardstate-restraints}
-%
-\begin{figure}
-    \includegraphics[width=0.95\linewidth]{figures/fig5_thermo_cyc_abs/Figure.pdf}
-    \caption{\textbf{Absolute thermodynamic cycle}}
-    \label{fig:fig_absolute_thermodynamic_cycle}
-\end{figure}
-%
-Absolute free energy calculations involve completely removing the interactions between the ligand or solute and its environment, taking it to a non-interacting state that may or may not retain intramolecular non-bonded interactions.
-This non-interacting state can then be shifted between environments (from the protein to water, or from one solution to another) without changing its free energy, and then interactions can be restored.
-%
-Absolute free energies are typically reported with respect to a specific reference or standard state, which effectively determines the arbitrary point at which the free energy is 0.
-The role of the standard state is particularly evident with binding free energies, in which having a reference state allows us to obtain a well-defined partial molar free energy for the reaction
-\begin{equation*}
-\ce{AB <=> A + B}
-\end{equation*}
-through the well-known expression derived from the law of mass action
-\begin{equation} \label{eq:DGfromKAB}
-\Delta G = -RT ~ \ln \left( c^{\circ} K_{AB} \right)  = -RT ~ \ln\left( \frac{c^{\circ} C_{AB}}{C_A C_B} \right) ,
-\end{equation}
-where $R$ is the gas, $T$ is the temperature, $C_X$ is the equilibrium concentration of the chemical species $X$ in the reaction solvent, and the reference state concentration $c^{\circ}$ converts the binding constant $K_{AB}$ into a dimensionless quantity expressed in reference concentration units.
-It should be noted that ignoring the term $c^{\circ}$ is equivalent to assuming a reference concentration of 1~D$^{-1}$, where D are the units used to express $K_{AB}$, and would thus cause the value of $\Delta G$ to vary with the choice of the units.
-Typically, it is convenient to define a standard state at a constant pressure of 1~atm and where each chemical species (i.e., A, B, and AB) in the reaction solvent has a concentration of $c^{\circ}$~=~1~M~=~1~molecule/1660~\r{A}$^3$ but do not interact with other molecules of A, B, or AB.
-%
-\paragraph{Handling the standard state in absolute free energy calculations.}
-For solvation free energy calculations, handling the standard state is typically straightforward, and treating it correctly simply means ensuring that the non-interacting solute is taken to the same (or equivalent) final reference state in both environments, e.g.
-that the transformation involves a 1M to 1M equivalent transfer free energy (where the non-interacting solute still occupies essentially the same volume as the solute in the interacting system).
-So typically in such cases no special care is required to ensure the correct standard state, as long as the \emph{experimental} data being analyzed uses the same standard state and if it does not, a simple entropic correction is needed.
-%
-However, for binding the situation is much more complex and requires special care.
-Experimental absolute binding free energies are reported relative to a specific reference state -- a 1 M standard state -- which must also be used in calculations.
-In practice, this has implications for how the calculations are done, as the reference concentration must enter the thermodynamic cycle employed.
-%
-Typically, to deal with both practical sampling issues and the standard state issue, restraints are employed in absolute binding free energy calculations to keep the ligand in a well defined volume as its interactions with the system are removed~\cite{gilson1997statisticalthermodynamic}.
-This solves two problems.
-First, if the ligand were not kept in a well-defined region, as its interactions were removed it might wander the system, perhaps quite slowly, and only inadequately sample the non-interacting or weakly interacting state -- yet adequate sampling of these states might be required for convergence.
-So for practical purposes, the use of restraints can dramatically improve sampling as interactions are weakened and removed.
-Second, if the ligand is not kept in a well-defined region then it is hard to determine how to link a computed binding free energy to the correct 1 M standard state.
-In contrast, with restraints, the free energy of releasing the restrained ligand to a 1 M standard state can be computed analytically or numerically by solving the relevant integral, allowing the standard state to enter the thermodynamic cycle.
-%
-\paragraph{Several choices of restraints are possible.}
-\todo[inline, color={blue!20}]{@author: Check references}
-In practice, a variety of types of restraints are common, from simple harmonic distance restraints between the ligand and the protein~\cite{mobley2006use}, to flat-bottom restraints which work similarly but only exert a force if the ligand leaves a specific region~\cite{}.
-%
-Alternatively, a set of restraints proposed by Boresch have also commonly been employed, where all six rigid-body degrees of freedom governing the orientation of the ligand relative to the receptor are restrained~\cite{leitgeb2005alchemical}.
-Further restraints, such as on the overall ligand RMSD have also been used~\cite{woo2005calculation}.
-%
-In principle, all of these forms will yield correct binding free energies in the limit of adequate sampling (if their effects and connection to the standard state are correctly handled) but they have different strengths and weaknesses.
-For example, with more involved restraints, sampling at intermediate lambda values will usually not need to be as extensive but more computational effort must go to computing the restraining free energy.
-Additionally, such restraints would typically keep the ligand from exploring alternative binding modes, which may be undesirable with Hamiltonian lambda exchange or expanded ensemble techniques where allowing the ligand to exchange binding modes when it is non-interacting could provide sampling benefits~\cite{wang2013identifying}.
-Concretely, flat bottom restraints might allow a ligand to explore multiple binding sites, harmonic restraints multiple binding modes within a site, and Boresch restraints a single binding mode within a single site.
-See additional discussion of the possibility of multiple binding modes below~\ref{sec:multiple_binding_modes}.
-%
-Many choices of restraints involve selecting reference atoms.
-Again, in principle this choice is unimportant given adequate simulation time but practical considerations may be important.
-The choice is likely especially important with Boresch-style restraints, where some relative placements of reference atoms are likely to be numerically unstable; additionally, ligand reference atoms should likely be in a part of the molecule which defines the binding orientation well, rather than in a floppy solvent-exposed tail, for example.
-\todo[inline, color={blue!20}]{DLM: Get input from JDC on what they've learned about these.}
-\todo[inline, color={blue!20}]{DLM: Clarify terminology: Double decoupling, etc. See Feature Box below.}
-%
-\subsection{Absolute and relative calculations must deal with some of the same issues}
-\subsubsection{Structural definition of the bound state and weak binders}
-\todo[inline, color={green!20}]{ASJM: This a bit waffly. What are the recommendations, of how to setup these restraints?}
-In binding free energy calculations, extra care should be taken when simulating the bound state, especially when dealing with weak binders and absolute free energy calculations in combination with enhanced sampling techniques.
-In principle, only configurations of the receptor-ligand complex that we consider "bound" should be sampled from the bound state, and, in atomistic simulations, this requires to establish a structural definition of the bound state.
-In practice, the simulation of the bound state starts with the ligand already placed in the binding site and relies on kinetic trapping to maintain a bound complex.
-However, this strategy may not be sufficient if the complex dissociation rate is high or if methodologies such as Hamiltonian replica exchange~\cite{bitetti-putzer2003generalized} and expanded ensemble~\cite{chodera2011replica} are employed in absolute free energy calculations to enhance sampling since, in both cases, the ligand may find a way out of the binding site on timescales that are achievable by modern MD simulations.
-A solution commonly adopted in these cases is the use of one or more restraints making the unbound configurations energetically unfavorable through the addition of extra terms in the potential function.
-In practice, even with restraints, it is not always trivial to force a molecular dynamics simulation to explore a restricted region of the configurational space with a complicated geometry such as a binding site.
-If the ligand is a tight binder\todo[inline, color={red!40}]{AR: Should we give an order of magnitude to define a tight and a weak binder?}, it is usually safe to employ a restraint that allows some of the unbound configurations to be sampled since they generally contribute negligibly to the partition function (i.e., they have a relatively small Boltzmann weight) as long as the sampled volume is not so large that their cumulative contribution becomes significant.
-However, this is more problematic when dealing with weak binders as the unbound configurations can have a non-negligible Boltzmann weight, and their binding affinity can exhibit a significant dependency on the restraint type and parameters that are used to determine the sampled volume~\cite{}.
-Moreover, it should be noted that the additional potential energy terms used to model the restraints can introduce a bias in the predicted free energy.
-The bias can be removed at the analysis stage through reweighting techniques, but this procedure can increase the statistical uncertainty of the binding free energy estimate when the restraints are so strong that the overlap with the reweighted state is diminished~\cite{}.
-%
-Finally, it is useful to keep in mind that, for a meaningful comparison between computational predictions and experimental measurements, the definition of the bound state should be consistent.
-In particular, this means that the signal used by the experimental methodology to determine the fraction of bound complexes in solution should in principle reflect the population of complexes in the bound state as defined in the calculation.
-%
-\subsubsection{Changes in net charge can be challenging/problematic.}
-If the net charge of the system will change as the alchemical calculation progresses, this can pose major challenges.
-Specifically, finite-size effects can introduce profound artifacts into computed binding free energies~\cite{}, in part because typical schemes for long-range electrostatics (including PME and reaction field) do not handle free energy contributions from such changes effectively or as they would be handled in a hypothetical macroscopic bulk solution~\cite{}.
-%
-There are two main potential solutions to avoid artifacts due to changes in net charge: Correcting for the introduced artifacts, or avoiding changing the net charge.
-%
-Many relative free energy planning tools have been set up to avoid changing the net charge of the systems considered, including LOMAP~\cite{} and Schr\"{o}dinger's FEP+~\cite{chen2018accurate}.
-Absolute free energy calculations can also potentially avoid changing the charge of the system by making a charge perturbation of equal and opposite sign elsewhere in the system; for example, as a charged ligand is removed, a charged counterion of opposite sign could also be removed, or one of the same sign could be inserted.
-This is sometimes referred to as an "alchemical ion" approach for dealing with the needed charge change, and is also employed by the Yank free energy package~\cite{wang2013identifying}.
-Charge corrections have also been explored, and are potentially a viable solution to this problem~\cite{} where artifacts introduced by finite-size effects are corrected numerically.
-~\cite{chen2018accurate}. However, application of such corrections typically remains less common than the use of a co-alchemical ion.
-%
-When free energy calculations \emph{do} need to change the charge of a ligand or solute, the literature does not yet seem to indicate what approach should be preferable, so considerable care should be taken.
-We are not yet aware of a careful comparison of charge corrections versus other approaches such as decoupling an ion at the same time, so in our view the issue of proper handling of charge mutations in the context of alchemical calculations remains a research problem.
-%
-\subsubsection{The importance of the alchemical pathway\label{sec:important_path}}
-\todo[inline, color={red!40}]{LNN: write common principles alchemical path choice. MRS: mostly wrote this, LNN can comment on it. LNN: LGTM}
-\todo[inline, color={red!40}]{AR: write absolute-specific section on  alchemical path choice}
-\todo[inline, color={red!40}]{BA: write relative-specific section on  alchemical path choice}
-
-\begin{figure}
-    \includegraphics[width=0.95\linewidth]{figures/fig7_what_is_lam/Figure.pdf}
-    \caption{Alchemical intermediates are createtd by making the potential energy depend on a an additional variable $\lambda$ that interpolates between the chemical endpoints.
-In \textbf{A}, at $\lambda=0$ the molecule is a fully interacting phenol and at $\lambda=1$,  a fully interacting benzene.
-\textbf{B} shows an illustration of the probability distribution of the potential energies as the switching function takes values of $\lambda=0$ to $\lambda=1$, intermediates are required for a sufficient overlap in potential energies to estimate a free energy difference between $\lambda=0$ and $\lambda=1$.
-    One of the most efficient families of intermediate pathways  are softcore potentials, whose shape is illustrated in \textbf{C} with dependence on $\lambda$ and \textbf{D} with dependence on the potential energy.
-Note how when $\lambda$ approaches 0, the energy smoothly approaches zero at all $r$, a necessary requirement for efficient calculations.
-}
-    \label{fig:fig_what_is_lambda}
-\end{figure}
-%
-Both absolute and relative calculations must choose an alchemical pathway connecting initial and final states.
-In principle, because of the path independence of the free energy, any arbitrary pathway will give the correct free energy change, but the choice of pathway will greatly affect the efficiency of the calculations.
-Some choices are particularly crucial -- for example, transformations involving insertions or deletions of atoms should employ a path soft-core potential path for Lennard-Jones or other interactions with repulsive interactions that go to infinite energy at small radius~\cite{}.
-%
-The key consideration for choosing alchemical pathways is that the intermediate states that a given pathway produces should sample configurational ensembles that change as slowly as possible as $\lambda$ changes, while still managing to go from the initial state to the final state as $\lambda$ goes from 0 to 1.
-%
-Another way of stating this is that intermediate states should sample molecular configurations that are as similar as possible to their neighboring states.
-The more similar the configurations are between intermediate states, the lower the statistical uncertainty is in the estimate of free energy between intervals.
-This can be proven directly from the BAR and MBAR formulas~\cite{bennett1976efficient,klimovich2015guidelines}, though the exact same principles apply for TI.
-%
-The simplest possible alchemical pathway is a \textit{linear} pathway:
-%
-\begin{equation}
-U(\vec{q},\lambda) = \lambda U_0(\vec{q}) + (1-\lambda)U_1(\vec{q})    
-\end{equation}
-%
-so-called because the dependence on $\lambda$ is linear.
-This clearly satisfies the basic requirement that it gives the initial endpoint potential energy $U_0(\vec{q})$ when $\lambda=0$ and final endpoint energy $U_1(\vec{q})$ when $\lambda=1$.
-%
-For many energy terms, \textit{as long as a repulsive core remains on}, this is a very good approach.
-For example, it can be shown that if van der Waals repulsions are left on, then the linear approach is very nearly the optimal path  possible for changing, removing, or inserting the electrostatic energy terms, with the path being within about 10--20\% of the minimum possible uncertainty~\cite{naden2015linear} for a fixed amount of simulation, as well as being nearly optimally efficient for van der Waals attractive terms with repulsion terms turned on~\cite{naden2014linear}.
-Although we are not aware of any quantitative tests for dipolar or higher multipole terms, theoretically it should behave equally well for those systems.
-%
-However, this approach ends up being terrible for removing or adding repulsive potentials that go to infinity quickly at or near the origin.
-One way to look at this is to examine how low $\lambda$ must go to reduce the energy at $0.5\sigma$ (the atomic size parameter) down to 1 $k_BT$, where thermal fluctuations make it possible for other atomic sites to penetrate routinely that deep.
-If we are trying go from a particle being present, and desire to make it disappear alchemically, then if the repulsive terms is of the form $\epsilon\frac{\sigma}{r}^{12}$, then if $\epsilon$ was 1 $K_BT$ at the temperature of interest, and we start with the particle present, then solving for $(1-\lambda)(1 kB_T)\frac{\sigma}{\frac{1}{2}\sigma}) = 1 k_B T$ we get $\lambda = 1-2^{-12} \sim 0.999976$.
-So we have gone virtually all the way to the end of the transformation, there's still an impenetrable post in the middle of our simulation!
-This is not very much like the desired final state of no interactions between the particle and its environment.
-We can play around with a few ways of modifying this, like simulating many more intermediate states near $\lambda=1$.
-Various analyses have shown that this is not a very good strategy, however~\cite{pham2011identifying, beutler1994avoiding, zacharias1994separationshifted, blondel2004ensemble}
-%
-What we need instead is some sort of function that smoothly gets rid of this infinity.
-A large number of scheme have been tried~\cite{beutler1994avoiding, zacharias1994separationshifted, blondel2004ensemble, pham2011identifying, pham2012optimal, naden2014linear, donnini2005molecular}, but the most common strategy that appears to be the best practices is to use a "soft-core" potential, of the form:
-%
-\begin{equation}
-    U(\vec{r_{ij}},\lambda) = 4\epsilon_{ij} \lambda \left(\frac{1}{(\alpha(1-\lambda) + (r_{ij}/\sigma_{ij})^6)^2} +  \frac{1}{\alpha(1-\lambda) + (r_{ij}/\sigma_{ij})^6}\right)\label{eq:softcore}
-\end{equation}
-Where $r_{ij}$ is the distance between two particles $i$ and $j$, $\epsilon_{ij}$ and $\sigma_{ij}$ are the Lennard-Jones parameters corresponding to the interaction between particles $i$ and $j$, and $\alpha$ is a constant (0.5 is optimal for the functional form shown above).
-This functional form has exactly the property we are looking for: it recovers the Lennard-Jones potential when $\lambda=1$, and the other endpoint ($\lambda=0$), it is exactly zero for all $r_{ij}$ everywhere, and as $\lambda$ goes to zero, the  $\alpha(1-\lambda)$ term lowers the infinite energy in the core.
-There are several different variants of the same functional form~\cite{zacharias1994separationshifted, beutler1994avoiding,pham2011identifying}, but the one in given in eq.~\ref{eq:softcore} is easy to understand and implement and fairly numerically stable.
-This functional form is shown in parts C and D of figure~\ref{fig:fig_what_is_lambda}.
-It has been shown that more complicated forms are not significantly more efficient than eq.~\cite{pham2012optimal}.
-We recommend using the softcore potential given in eq.~\cite{eq:softcore}, unless there is a compelling reason otherwise.
-Using a similar equation to eq.~\cite{eq:softcore} may be acceptable in most circumstances if that is what is supported in your chosen software.
-However, if you are inserting or removing entire atomic sites, we heavily recommend against using the linear approach; it will be very difficult to get correct results.
-%
-We have discussed optimal ways of disappearing or appearing Lennard-Jones interaction sites and turning on and off electrostatics terms.
-What about performing both transformations at the same time?
-We can't turn off the electrostatics linearly at the same time we turn off the Lennard-Jones terms, as it would leave infinitely large attractive and repulsive electrostatic terms "bare" at small $\lambda$, resulting in the simulation crashing.
-It \textit{is} possible to apply the same soft core approach to the Coulomb interaction, but it turns out to be somewhat difficult to turn off two soft-core terms  terms simultaneously~\cite{Steinbrecher:JCC:2011}.
-We therefore recommend performing the transformations in sequence; first, turning off all electrostatics for atoms that must be removed, inserting and removing Lennard-Jones sites (both the insertion and removal can be done simultaneously), and then turning electrostatics for the introduced particles on.
-However, If there are no removals or introduction to atomic sites, then it is reasonable to change all interactions linearly.
-%
-Other issues, such as whether absolute calculations should retain or remove intramolecular nonbonded interactions, may be less critical and recommendations differ among studies in the literature~\cite{}, and reasonable efficiency can be often obtained with either choice even if some are somewhat better or worse than others.
-Our recommendation is to leave the intramolecular interactions on during the transformation for simplicity if there are no other known issues with this approach.
-The key thing to watch out for is whether the total potential energy, and therefore the intermediate ensembles sampled, change smoothly from beginning to end.
-If a configurational phase transition is noted in the process (like substantial dihedral rearrangement), then other steps must be taken, perhaps including modifying intramolecular interactions~\cite{???}.
-These problems can be diagnosed by noticing lack of configuration space overlap between different simulations (see section~\ref{subsec:})"
-%
-Relative calculations introduce additional choices, including whether to define explicit intermediate states~\cite{} or leave these implicitly defined by the code~\cite{}.
-Typically in single topology relative calculations it proves most efficient to first remove electrostatic interactions of any atoms which will be deleted, then modify other nonbonded interactions, then restore electrostatic interactions of any atoms which are being inserted.
-Other schemes, such as simultaneously changing electrostatic and Lennard-Jones interactions, even with electrostatic ``soft core'' potentials~\cite{steinbrecher2007nonlinear}, in our experience typically introduce errors and/or instabilities or are at least unreliable.
-It is expected that similar considerations will apply to dual topology calculations, and the principle of first removing electrostatics and then removing steric interactions will likely serve well.
-%
-A key additional consideration in choosing the alchemical pathway is the choice of spacing of intermediate states.
-The spacing depends to some extent on the choice of analysis method, though states should essentially be spaced equidistant in the relevant thermodynamic length~\cite{}.
-For BAR/MBAR techniques this means that spacings should typically be equal in [what, variance? ref].
-Some schemes to adaptively optimize the spacing of intermediate states based on initial exploratory simulations have been proposed~\cite{}.
-%
-An example of this approach for relative binding free energy calculations includes saving configurations from short 10 picosecond simulations starting at $\lambda_0$, and evaluating changes in the standard error of mean (SEM) in $\frac{dU}{d\lambda}$ as reevaluated at new values of $\lambda$, starting with a step size of 0.01 up to a predetermined threshold.
-This threshold can be empirically defined as the number of atoms being inserted or deleted in the simulation / 100.
-If the absolute value of the difference in SEM is below the defined threshold, then the change in $\lambda$ for re-evaluation of SEM in $\frac{dU}{d\lambda}$ is defined by the current difference in SEM from the threshold divided by the derivative of the SEM with respect to $\lambda$.
-%
-\subsubsection{Multiple or uncertain binding modes may require considerable care}
-\label{sec:multiple_binding_modes}
-In a discovery setting, new ligands typically have unknown or at least uncertain binding modes~\cite{kaus2015how, plountprice2000analysis,mobley2009binding,calabro2016elucidation},  complicating binding free energy estimation.
-This uncertainty is because it is usually not desirable to estimate a binding affinity for a ligand which already has an available bound structure, since such a compound has already been tested.
-To deal with prospective ligands with unknown binding modes, discovery projects commonly assume that modifications of functional groups on a common scaffold result in a consistent binding mode across all members of a series.
-This is not necessarily always the case~\cite{kaus2015how}, as reviewed elsewhere~cite{mobley2009binding} and in some cases unexpected binding mode changes can be the origin of apparent non-additivity in structure-activity relationships~\cite{calabro2016elucidation}.
-Binding modes also tend to be particularly variable in the case of fragments, which often may have multiple relevant binding modes~\cite{}.
-
-Absolute free energy calculations for dissimilar ligands can have particular challenges because the (potentially incorrect) assumption of consistent binding modes across a series of similar ligands is likely to be even less robust than the in the case of relatively calculations.
-This means that researchers performing absolute binding free energy calculations will have to pay particular attention to generating reasonable putative binding modes.
-
-In some cases, it is tempting to simply use docking techniques to generate initial bound structures for starting molecular dynamics simulations.
-However, timescales for binding mode interconversion are usually slow compared to MD/free energy timescales, meaning that simulations started from different potential binding modes are likely to yield disparate computed binding free energies~\cite{mobley2006use, palma2012computation, mobley2012perspective, gill2018binding} .
-And docking techniques are good at identifying sterically reasonable potential binding modes, but still perform relatively poorly at identifying a single dominant binding mode \emph{a priori}.~\cite{}
-
-It is worth highlighting a recent SAMPL blind challenge on HIV integrase as an illustration of this. 
-Many submissions, using state-of-the-art methods, had difficulty even predicting which \emph{binding site} ligands would bind in (most submissions placed more than half of the ligands into the incorrect binding site), and even given correct binding sites, the binding mode within each site was also quite difficult to predict~\cite{mobley2014blind}.
-The best performing submission for predicting binding modes actually ended up being a human expert (aided by computational tools) with more than 10 years of experience on the particular target~\cite{voet2014combining}, rather than a fully automated approach.
-While free energy calculations on this set had some success, many of the failures actually ended up being cases where the binding mode selected as input for free energy calculations was later found to be incorrect~\cite{gallicchio2014virtual}, highlighting the importance of these issues.
-
-One approach which has shown some success is to retain diverse potential binding modes from docking, perform short MD simulations of these to identify distinct stable binding modes, and then consider these in subsequent calculations~\cite{gallicchio2014virtual, mobley2006use,rocklin2013blind, boyce2009predicting, mobley2007predicting}.
-
-Routes to handle multiple potential binding modes are different depending on whether absolute or relative calculations are selected, unless a method is available to estimate the relative populations of different stable binding modes in advance (e.g.
-such as the BLUES approach currently in development~\cite{gill2018binding}), in which case this approach could be applied to assist both types of calculations.
-
-\paragraph{Handling multiple potential binding modes within absolute calculations.}
-Within absolute binding free energy calculations, multiple potential binding modes can be handled by two main strategies: Consider each binding mode separately (a separation of states strategy) or sample all binding modes within a single simulation~\cite{mobley2012perspective}.
-This couples to the choice of restraints selected, as some restraints will allow transitions between binding modes and even binding sites (Section~\ref{sec:standardstate-restraints}), and others do not.
-
-Sampling all binding modes within a single free energy calculation is usually impractical without some form of enhanced sampling or at least Hamiltonian replica exchange~\cite{wang2013identifying} because barriers for binding mode interconversion result in kinetics which are too slow compared to simulation timescales~\cite{mobley2006use, palma2012computation,mobley2012perspective, gill2018binding}.
-Hamiltonian exchange, coupled with appropriate restraints, can allow the ligand to relatively rapidly exchange between potential binding modes when non-interacting, accelerating sampling of binding modes~\cite{wang2013identifying}.
-\todo[inline, color={yellow!40}]{DLM probably need more background refs on Hamiltonian lambda exchange here.}
-
-Separation of states provides a simple though potentially expensive alternative, where each stable binding mode is considered separately with a binding free energy calculation restricted to that binding mode, and then (as long as the binding modes are non-overlapping) the resulting component binding free energies can be combined into a total~\cite{mobley2006use, mobley2012perspective}.
-This approach necessitates a separate binding free energy calculation for each potential binding mode, however, so it can be computationally quite costly.
-If relative populations of different stable binding modes were available from some other technique, it could make this separation of states approach considerably more efficient~\cite{mobley2012perspective, gill2018binding}.
-
-\paragraph{Handling multiple potential binding modes within relative calculations.}
-
-Multiple potential binding modes pose particular problems for relative free energy calculations, as having multiple starting structures for these calculations could yield substantially different calculated relative binding free energies for the same transformation due to kinetic trapping, and, without additional information (specifically, the free energy of binding mode interconversion or, equivalently, the relative populations of different binding modes) it becomes impossible to sort out which of the multiple answers is in fact the correct relative binding free energy~\cite{}.
-
-To deal with this, some practitioners have actually computed relative binding free energies of different binding modes of the same ligand~\cite{palma2012computation} [ref Jorgensen (?) etc.].
-For example, a mutation which adds a methyl to an aromatic ring of a larger ligand might yield one result if the methyl points in one direction, and a different value if it points in the other due to slow ring motions.
-[e.g.
-get ref from Sukanya]
-One could compute the free energy of turning off the methyl group in one orientation and turning it back on in the other orientation to obtain the free energy difference between the two potential binding modes.
-While this approach has precedent, it is relatively difficult to automate at present and requires considerable care.
-
-Overall, this likely means that relative free energy calculations will be susceptible to problems resulting from uncertainty in ligand binding modes until more robust approaches are available to determine dominant binding modes, or the relative populations of different potential binding modes, in advance.
-
-\subsubsection{What different sampling schemes exist and which one should I use?}
-With all alchemical simulations having the requirement of sampling from different sets of $\lambda$ states in common different approaches have been designed.
-Fig.~\ref{fig:fig_sampling_scheme} illustrates the four most common schemes used.
-The simplest approach is defining a set of fixed $\lambda$ values and then run independent simulations at each of the intermediate $\lambda$ values, see Fig.~\ref{fig:fig_sampling_scheme}(A).
-This type of scheme is currently used for AMBER TI calculations~\cite{song2019} and for BioSimSpace~\cite{hedges2019biosimspace}.
-A slightly more complex approach is allowing replica mixing, meaning that each independent $\lambda$ simulation is attempted to be exchanged with a neighbouring $\lambda$ replica.
-This is based on ideas of Monte Carlo simulations of spin glasses by Swendsen and Wang~\cite{wendsen1986replica}.
-Using a Metropolis Hastings based acceptance criterion for exchanges means that the generated ensemble of all replicas is still sampling from the Boltzmann distribution and has been used in many different contexts for molecular simulations~\cite{sugita2000multidimensionala,sugita1999replicaexchangea, woods2003developmenta, jiang2010free}.
-The basic idea of the replica exchange scheme is shown in Fig.~\ref{fig:fig_sampling_scheme}(B).
-It is supported in various software packages that provide alchemical implementations, such as GROMACS~\cite{aldeghi2015accurate}, FEP+~\cite{wang2015accurate}, and NAMD~\cite{jiang2019computing}.
-A third approach borrows ideas from simulated tempering~\cite{marinari1992simulateda}.
-In this scheme a single replica rapidly explores all of $\lambda$ space by optimally working out weights that allow switching between different intermediate $\lambda$ values, as seen in Fig.~\ref{fig:fig_sampling_scheme}(C) and also referred to as self adjusted mixture sampling~\cite{lyubartsev1992newa, li2007simulated, tan2017optimally} and while promising, has so far only been supported in~\cite{andrearizzi2019choderalab}.
-The last approach available in conjunction with GROMACS makes use of non-equilibrium simulations~\cite{aldeghi2018accurate}.
-In this approach, only end state $\lambda$ replicas ($\lambda$=0, $\lambda=1$) are simulated at equilibrium intermediate information is generated from non-equilibrium simulations.
-A schematic of this approach is shown in Fig.~\ref{fig:fig_sampling_scheme}.
-\ref{fig:fig_sampling_scheme} A:\cite{hedges2019biosimspace, song2019using } B\cite{aldeghi2015accurate, wang2015accurate,jiang2019computing} C:\cite{li2007simulated, tan2017optimally} D:\cite{aldeghi2018accurate}
-\begin{figure}
-    \includegraphics[width=0.95\columnwidth]{figures/fig8_sampl_scheme/Figure.pdf}
-    \caption{\textbf{Four sampling strategies available in different software packages.} \textbf{A}: Multiple replicas in parallel at different lambda states.
-\textbf{B}: Hamiltonian replica exchange scheme.
-\textbf{C}: Single replica scheme sampling from all $\lambda$ states.
-\textbf{D}: Non-equilibrium sampling scheme.}
-    \label{fig:fig_sampling_scheme}
-\end{figure} 
-The best practice recommendation, is to use replica exchange type sampling schemes (Fig.~\ref{fig:fig_sampling_scheme}(B)), or in absence of their availability and when looking at systems with fast mixing times parallel replica schemes are acceptable (Fig.~\ref{fig:fig_sampling_scheme}(A)).
-Single replica schemes and non-equilibrium schemes are not as established yet, but are very promising.
-
-\subsubsection{How long should I run my simulation for and what information should be saved?}
-Before launching alchemical free energy calculations it is wise to consider how convergence and completion will be assessed.
-Different conditions on when to stop alchemical free energy calculations should be determined, and this may require several iterative checks and therefore modifications to the calculation protocol.
-One useful metric to use for termination is the expected or desired uncertainty of a desired free energy estimate, though care must be exercised should the uncertainty estimate prove unreliable.
-In particular, if the rate of change in the free energy estimate is significant when this condition is met, the simulation may not be locally converged, and more sampling may be necessary to determine a stable free energy estimate which is no longer changing significantly over time.
-However, this is not the only metric which should be used, as the uncertainty only captures the information about the sampled phase space, not necessarily the entirety of the phase space.  
-For example, 
-convergence of relative free energy calculations in predictive simulations where the entire phase space is not known in advance, requires sampling the different kinetically stable states~\cite{mobley2012perspective}.
-This highlights the importance of choosing the correct thermodynamic path to ensure you sample the required thermodynamic states as discussed in Section~\ref{sec:important_path}.
-%
-The condition of minimizing the statistical uncertainty of different free energy estimators below a sufficient threshold should be one metric monitored over the simulation.
-This can be done through the uncertainty estimator built into certain analysis tools such as MBAR, or can be done though more general statistical tools like Bootstrap Sampling.
-A target statistical uncertainty should be chosen at the onset of the simulation to avoid excessively long simulations, or falling into the trap of running until the free energy estimate is "good enough," which is subjective and has no defined criteria.
-This could be a fixed value such as $0.20 \mathrm{kcal/mol}$, or a functional quantity such as "below $0.5 \mathrm{kcal/mol}$ and $10\%$ of the free energy estimate." The user does not need to monitor this information in real-time and can choose to run simulations for fixed duration (either time or number of samples) and run analysis on the data collected thus far.
-If more samples are needed, the simulations can be resumed, or, started again in different initial conditions.
-
-Convergence in other alchemical observables should also be monitored to determine if the defined phase space has been sufficiently sampled and enough decorrelated samples have been drawn.
-These additional observables include, but are not
-limited to, the variance in $\frac{dU}{d\lambda}$ across all $\lambda$ values, calculating the variance in free energy using bootstrap analysis, and comparing differences in free energies calculated using different percentages of the simulation in both the forward and reverse directions (figure \ref{fig:convergence_forward_reverse}).
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.95\linewidth]{figures/fig10_forward_reverse/Figure.pdf}
-    \caption{\textbf{Free energy (in k$_{B}$T) for two different relative binding free energy (RBFE) perturbations.} 
-    Each plot shows the estimated free energy change using a varying fraction of total simulation time (up to 5 ns total). 
-    The top panel shows the three-step protocol for a perturbation involving 5 perturbed atoms, while the bottom panel shows the same protocol for a perturbation involving 10 perturbed atoms.
-The difference in energy between the forward (blue) and reverse (red) free energy calculations at the midpoint of the simulation time gives an indication of the overall convergence of the simulation, with differences over 1 k$_{B}$T indicating poor convergence.}
-    \label{fig:convergence_forward_reverse}
-\end{figure}
-
-Each of these metrics have demonstrated promising results for diagnosing when a simulation has a convergence issue beyond simple convergence of uncertainty estimate. 
-Results obtained from calculations with convergence issues should be checked for errors or run for longer before any confidence should be placed in conclusions drawn from their analysis.
-In relative calculations that share similar binding modes, for example, and do not induce large conformational changes when in complex with protein, the need to sample exhaustively to converge estimates in free energy differences is often not necessary due to the locality of sampling changes in the molecular topology and shared phase space of the core atoms.
-However, even subtly induced changes in protein binding configuration will require more sampling or cause local convergence to a free energy estimate that has high error.
-Many enhanced sampling techniques have been proposed to try and overcome barriers in free energy calculations and mitigate convergence to erroneous estimates~\cite{}. 
-The confidence a user should have in a free energy estimate is significantly improved when both the uncertainty of the free energy estimate is low, and when other observables have reached a convergence.
-
-A simple, but effective, way of checking for bias is to compute the same property by multiple estimators. 
-The uncertainty in the free energy, for example, has multiple ways to be estimated, e.g.
-through standard error propagation methods (including MBAR's estimator, which is based on the same principles as standard error propagation), through bootstrap methods, through multiple independent runs, etc.
-Independent of how the property is estimated, its important to remember that they are \textit{estimations of the property}, not the true underlying property itself. 
-These estimators are usually consistent estimators, meaning they will converge to the true answer in the limit of sufficient sampling, not necessarily unbiased ones though.
-As such, it is a good idea to subject different estimators to the same data to see if they yield either the same estimate (within error and bias), or if they fluctuate wildly. 
-This is not a perfect method as some estimators, such as exponential averaging, will converge significantly more slowely, relative to more accurate estimators like MBAR. 
-Therefore, it is a good idea to apply the estimators to different fractions of the data to see if the main estimator of free energy you have chosen is stable.
-
-Each method requries different data about the simulation be collected.
-If, for instance, the free energy estimator selected is thermodynamic integration, then values of $\frac{dU}{d\lambda}$ at uncorrelated data points must be collected,  Once a combination of knowing what type of simulation you will run, which alchemical topology you will simulate, what alchemical path you will simulation along, and what your stopping conditions are, then you are ready to enumerate the information you should capture.
-Below is a sample of the minimal information you need for a set of common estimators:
-
-\begin{itemize}
-    \item Thermodynamic Integration (TI) requires $\frac{\partial U(\vec{x})}{\partial\lambda}$.
-    \item Exponential Averaging (EXP) needs \textit{either} $\Delta U_{k,k+1}(\vec{x})$ or $\Delta U_{k,k-1}(\vec{x})$, depending on the direction its being evaluated in.
-    \item Bennett Acceptance Ratio (BAR) needs \textit{both} $\Delta U_{k,k+1}(\vec{x})$ and $\Delta U_{k,k-1}(\vec{x})$.
-    \item Weighted Histogram Analysis Method (WHAM) and Multistate Bennett Acceptance Ratio (MBAR) both need the complete set of $\Delta U_{k,j} \, \forall \, j=\{1...K\}$.
-WHAM must have this information binned.
-\end{itemize}
-
-The potential derivative required for TI should generally be calculated during the simulation; only under every rare circumstances~\cite{Naden:JCTC:2015} can it postprocessed by a code that does not evaluate the derivatives.
-Many codes already have options for doing this.
-If that option is unavailable, you can estimate it through finite difference (if sufficient information is collected), but this  will introduce significant error, and is generally not a best practice.
-Tthe BAR estimator may be a better, and simpler choice at that point as you will have at least the same level of information.
-The potential energy differences required for EXP, BAR, MBAR, and WHAM can be calculated either during the simulation or in post-processing.
-It is recommended to calculate the potential differences in code when possible to avoid extra overhead and possible errors produced by running the simulation twice, and to reduce the amount of stored information.
-Although TI must be usually be calculated in code, as it requires the derivative, there is one condition under which it actually has the fastest computation time. 
-If the alchemical path you have chosen is a linear alchemical path, then you get $\frac{dU}{d\lambda} = U_0(\vec{x}) - U_1(\vec{x})$, which is the difference between the initial and final states. 
-However, because of the problems with linear paths already discussed in this paper, this simplification is rarely that useful.
-
-Free energy information should generally be saved more frequently than coordinate data, approximately at the rate that uncorrelated samples are produced.  
-The on-disk size of the data for free energy estimation is often significantly smaller than full atomic coordinates, so the information should be collected frequently. 
-However, the information should not be collected \textit{every} time step, as most free energy techniques are operated at equilibrium, and need equilibrated \textit{and decorrelated} samples for an unbiased estimate.
-A sample collected every time step will likely result in most samples being discarded due to decorrelation routines in the analysis.
-However, if it is computationally cheap and disk space is plentiful, do save often.
-One may safely assume that the correlation time is greater than 100-200 fs even for relatively simple systems such as small molecules in solvent, so saving no more frequently than every 50-100 steps is recommended.
-How decorrelation impacts calculations, and how to compute it is discussed in other sections.  [section ref]
-
-\section{Step 3 -- Data analysis}
-\label{sec:step3}
-
-Once data has been collected from alchemical intermediates, it must be analyzed to produce an estimate of the free energy change (and its associated statistical uncertainty) for each leg of the thermodynamic cycle.
-While a number of different estimators are available that will give consistent results under optimal circumstances, some approaches are recommended over others due to their robustness and ability to provide information on poor convergence.
-
-\subsection{Detecting the boundary between equilibrated and production regions}
-\label{sec:automatic-equilibration-detection}
-
-Much of the infrastructure for analyzing alchemical free energy calculations relies on the concept of asymptotically unbiased estimators, which produce unbiased estimates of the free energy when fed very long simulations~\cite{shirts2005comparison}.
-In reality, free energy calculations are often initiated from highly atypical initial conditions (such as a protein-ligand geometry obtained from docking and subjected to a heuristic solvent placement scheme), and simulations are of a finite length dictated by available computational resources and computing demands.
-As a result, these estimators can produce significantly biased estimates if fed the entirety of simulation data generated without further processing~\cite{chodera2016simple}.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.95\linewidth]{figures/fig11_equib_detection/Figure.pdf}
-    \caption{{\bf Automatic partitioning into equilibration and production regions.}
-    \textbf{A} The average (black line) standard deviation (shaded region) of the reduced potential $u^*$ over many independent replicate simulations started from the same initial conditions show a significant initial transient before relaxing to the true average potential energy\textbf{B} A cumulative average (red) of the entire simulation data demonstrates simulation bias not seen when initial simulation data is omitted (blue).
-Using an automated approach to detect equilibration of the boundary $t_0$ using statistical inefficiency $g$ (\textbf{C})  for an effective simulation interval (\textbf{D}) \textbf{E} The optimal equilibration boundary $t_0$ is selected to maximize the number of uncorrelated samples.
-    \emph{Figure adapted from~\cite{chodera2016simple}.}
-    }
-    \label{fig:automatic-equilibration-detection}
-\end{figure}
-
-To minimize this effect, an initial portion of the simulation is often discarded to \emph{equilibration}~\cite{braun2019best}, with the idea of removing the most heavily biased initial portion of simulation data but retaining the unbiased \emph{production} region that represents a stationary Markov chain process sampling from the desired equilibrium target distribution.
-Because the simulation time required for the atypical initial sampler state to relax toward equilibrium is a property of the specific system being simulated and the specific initial conditions selected, it is simplest to collect data for the whole process and use an automated algorithm to select how much data should be discarded to equilibration in a post-processing step.
-
-A simple approach to automatically partitioning simulation data into equilibration and production regions is described in~\cite{chodera2016simple} (illustrated in Figure~\ref{fig:automatic-equilibration-detection}).
-Suppose we have a simulation of length $T$ consisting of correlated data.
-Here, the goal of the post-processing step is to select the equilibration boundary $t_0 \in [0, T]$ so as to \emph{maximize} the number of effectively uncorrelated samples remaining in the production region $N_{[t_0,T]}$, which is defined as
-\begin{eqnarray}
-N_{[t_0,T]} &=& \frac{T - t_0}{g_{[t_0,T]}}
-\end{eqnarray}
-where $g_{[t_0,T]}$ is the \emph{statistical inefficiency} of a timeseries $a_t$, described in more detail below.
-Conveniently, this procedure also produces the information necessary to decorrelate the simulation data for estimating the free energy differences, a requisite next step in analysis.
-This approach is implemented within the MBAR~\cite{kylebeauchamp2019choderalab} and alchemlyb~\cite{daviddotson2019alchemistry} packages, and is highly recommended for standard practice.
-
-For additional discussion of working with correlated data and autocorrelation analysis, please refer to the work on Best Practices for Quantification of Uncertainty and Sampling Quality in Molecular Simulations~\cite{grossfield2018best}.
-
-\paragraph{Computing the timeseries for equilibration detection}
-Typically, the timeseries of note $a_t$ analyzed in automated equilibration detection is the negative logarithm of the probability density sampled by the MCMC algorithm (up to an irrelevant additive constant).
-For simple independent simulations that sample $x_t \sim \pi(x ; \lambda)$, this is given by the reduced potential
-\begin{eqnarray}
-a_t &\equiv& - \ln \pi(x_t; \lambda) + c = u(x_t; \lambda) .
-\end{eqnarray}
-
-Note that the use of the effective reduced potential is not guaranteed to pick up on all slow relaxation processes that may be coupled to the alchemical free energy, but the simplicity of its computation means it is generally appropriate for most cases.
-
-\paragraph{Cautions in automating equilibration detection}
-For simulations that are simply not long enough to contain a large number of samples from true equilibrium either because they are very short or contain slow processes, this procedure cannot completely remove the bias.
-In such cases, this approach simply selects the final portion of the the simulation, which may be contained in a single substate of conformational space, which may itself lead to biased estimates. 
-This situation can be detected if the equilibration boundary $t_0$ is a significant fraction of the total simulation length $T$, with a good rule of thumb being that $T \gtrsim 20 t_0$.
-If this is not possible, advanced analysis techniques that assume only local equilibrium (rather than global equilibrium) such as the TRAM estimators~\cite{mey2014xtram,wu2016multiensemble,nuske2017markov} may be more appropriate.
-
-\subsection{Decorrelating samples for analysis}
-\label{sec:decorrelating-samples}
-
-\paragraph{Computing the statistical inefficiency}
-Most estimators require an uncorrelated set of samples from the equilibrium distribution to produce (relatively) unbiased estimates of the free energy difference and its statistical uncertainty.
-To do this, the production region of the simulation is generally \emph{subsampled} with an interval approximately equal to or greater than the \emph{statistical inefficiency} $g \ge 1$ to produce a set of uncorrelated samples that can be fed to the estimator machinery~\cite{chodera2016simple},
-\begin{eqnarray}
-g &\equiv& 1 + 2 \tau_\mathrm{eq} \label{eq:statistical-inefficiency-definition}
-\end{eqnarray}
-where $\tau_\mathrm{eq}$ is the integrated autocorrelation time, formally defined as
-\begin{eqnarray}
-\tau_\mathrm{eq} &\equiv& \sum_{t=1}^{T-1} \left(1 - \frac{t}{T}\right) C_t \label{eq:integrated-autocorrelation-time-definition} , 
-\end{eqnarray}
-with the discrete-time normalized fluctuation autocorrelation function $C_t$ defined as
-\begin{eqnarray}
-C_t &\equiv& \frac{\expect{a_n a_{n+t}} - \expect{a_n}^2}{\expect{a_n^2} - \expect{a_n}^2} . \label{equation:autocorrelation-definition}
-\end{eqnarray}
-The basic concept is that $\tau_\mathrm{eq}$ corresponds to the single-exponential decay time for the autocorrelation process that generates samples, so the statistical inefficiency $g$ measures the approximate temporal separation between two effectively uncorrelated samples (where two exponential relaxation times are presumed to be sufficient).
-
-Robust estimation of $C_t$ for $t \sim T$ is difficult due to growth in statistical error, so common estimators of $g$ make use of several additional properties of $C_t$ to provide useful estimates (see \emph{Practical Computation of Statistical Inefficiencies} in~\cite{chodera2016simple} for a detailed discussion).
-\todo[inline]{JDC: Describe recommended statistical inefficiency computation algorithm.}
-We recommend using the robust statistical inefficinecy computation routines available within the MBAR~\cite{kylebeauchamp2019choderalab} and alchemlyb~\cite{daviddotson2019alchemistry} packages.
-
-\paragraph{Subsampling data to generate uncorrelated samples}
-Once the statistical inefficiency $g$ has been estimated, it is straightforward to subsample the correlated timeseries simulation data to produce effectively uncorrelated data that can be fed to estimators of the free energy.
-Suppose the correlated timeseries is $\{a_t\}_{t=1}^T$; we can form a new timeseries of $N_{\mathrm{eff}} \approx T / g$ effectively uncorrelated samples by selecting a subset of indices $\{ \: t = \mathrm{round}((n-1) \, g) \: | \: n \in \mathrm{range}(1,\ldots ,N) \: \}$ where $\mathrm{round(x)}$ denotes rounding to the nearest integer.
-
-If independent simulations are used, the alchemical state $\lambda$ may have a significant impact on the correlation time, and these simulations should be subsampled independently using a separate estimate of the statistical inefficinecy $g$ for each alchemical state.
-If coupled simulations are used (such as a Hamiltonian replica exchange simulation), the replicas should undergo equivalent random walks in alchemical space, and the replicas can be can be subsampled with the same $g$ to generate an equal number of uncorrelated samples at each alchemical state.
-Conveniently, the approach described above for automated equilibration detection produces an appropriate estimate of $g$ over the production region for automating this process.
-
-\paragraph{Cautions and considerations}
-Reliable estimation of the statistical inefficiency is difficult, and estimates will not generally be as precise (in a relative error sense) as averages.
-To ensure there is sufficient data available for reliable decorrelation and estimation of free energy differences, it is recommended that the effective number of uncorrelated samples $N_{\mathrm{eff}} \ge \approx 50$ if the BAR or MBAR estimators (discussed below) are used; the number may need to be much higher with alternate estimators.
-
-\subsection{Estimators for free energy differences}
-
-Free energy differences between two different states differing in the energy function are directly related to the
-ratio of probabilities of those states.
-As can be noted, the partition functions in Eq.~\ref{equation:dimensionless-free-energy-difference} are simply the total accumulated probabilites for all possible configurations of the system.
-Virtually all of the ways to estimate this free energy are based in converting this ratio of integrals to something that can be measured in one (or several) simulations.
-
-\paragraph{The Zwanzig relationship (EXP)}
-
-The simplest method for calculating free energy
-differences from simulations is the so-called \textit{Zwanzig
-relationship}~\cite{zwanzig1954hightemperature}, also called one-sided exponential reweighting (EXP), or simply free energy perturbation, thought this final term is sometimes used to encompass all ways of calculating free energy differences.
-
-The (reduced) free energy difference $\Delta f_{01}$ between an initial state 0 and a final state 1 defined by two different potential energy functions 
-$U_0(\vec{q})$ and $U_1(\vec{q})$ over coordinate space $\vec{q}$ can be calculated as:
-\begin{eqnarray}
-\Delta f_{01} & = & \ln \expect{e^{-(U_1(\vec{q}) - u_0(\vec{q}))}}_0 =  \ln \expect{e^{-\Delta u(\vec{q})} }_0
-\end{eqnarray}\label{eqn.zwanzig}
-and the average is over all samples from the simulation performed with $u_0$.
-In the case of NVT (canonical) sampling and assuming the masses do not change, then $u$ is just $U/k_BT$, and $f$ is $A/k_BT$, but it can be generalized to other ensembles with the proper definition of $f$ and $u$.
-Described in words, we take the samples generated during our run with the potential energy function $U_0(\vec{q})$, recalculate what the energy difference would be if we switched to potential energy function $U_1(\vec{q})$, and average the exponential of the -1 times the energy difference to get -1 times the exponential of the free energy difference.
-This almost requires no extra code functionality to perform; one need only to save a full precision trajectory, and run an unmodified molecular simulation code using the $U_1$ in order to calculate the new energies of stored snapshots.
-The analysis can be written in a line of code.
-We note that this method is even more general, in that the instantaneous work to change the potential energy function from $U_0$ to $U_1$ can be replaced by the non-reversible work $W$ to make the same change under the same equilibrium conditions at either end state~\cite{jarzynski1997nonequilibrium,jarzynski1998equilibrium,crooks2000pathensemble}, though we do not go into all of the details of nonequilibrum transformations here, and refer the reader to more advanced treatments~\cite{maragakis2008bayesian,oberhofer2005biased,procacci2015unbiased,shirts2003equilibriuma,ytreberg2004singleensemble}
-
-Although the Zwanzig equation is formally correct (as long as the two states considered sample the same phase space volume, which is true for standard molecular models), it has some very important numerical issues that mean that it often performs badly for standard free energy calculations, even for small molecules.~\cite{shirts2005comparison,lu2003appropriate}  One can show that if the standard deviation of the difference $\Delta U(\vec{q}) = U_1(vec{q})-U_2(\vec{q})$ over all sampled $\vec{q}$ is large (which in this case, means only several times $k_BT)$, then very few samples contribute to the average, and the answer will be both biased and extremely noisy~\cite{xx}.
-Essentially, the method is dominated by contributions of rare snapshots\cite{jarzynski2006rare, wu2005phasespaceb, wu2005phasespacec}.
-There are a number of applications where it this expression is nevertheless useful, but those usually involve investigating the dependence of the free energy on small changes in parameters, not changing molecules or even single atoms in molecules~\cite{xx}
-
-\paragraph{The Bennett Acceptance Ratio (BAR)}
-
-If we have the differences in the potential energy sampled from the distribution defined by $U_0$ to the state defined by $U_1$, and we also have the differences in potential energies from the distribution sampled by $U_1$ to the state defined by $U_0$, we can obtain a significantly improved estimate of the
-free energy difference compared to that obtained by EXP.  
-This estimate was first derived by Bennett and is hence generally called the Bennett Acceptance Ratio (BAR).
-It is solved by finding the reduced free energy $f_{ij}$ that satisfied the following implicit equation:
-\begin{eqnarray}
- \sum_{i=1}^{n_i} \frac{1}{1 + \exp[\ln(\frac{n_i}{n_j}) + u_{ij}(\vec{q}) - f_{ij})
- ]} \nonumber \\
- =\sum_{i=1}^{n_j} \frac{1}{1 + \exp[\ln(\frac{n_i}{n_j}) - U_{ij}(\vec{q}) + f_ij)]}
-\end{eqnarray}
-where $n_i$ and $n_j$ are the number of samples from each state. We
-will refer to this method as BAR. More recent derivations show that this formula is the maximum likelihood estimate of the free energy difference
-given sets of samples from the two states~\cite{shirts.bennett}.
-
-Many studies have demonstrated both the
-theoretical and practical superiority of BAR over EXP in molecular
-simulations~\cite{shirts2005comparison,lu2003appropriate}, and BAR converges to EXP in the limit that all samples are from a
-single state~\cite{bennett1976efficient,shirts.bennett}. BAR also requires significantly less
-overlap between the configurational space of each state to converge than EXP, though some overlap must still exist.
-
-The Bennett acceptance ratio is only defined between two states.
-Usually, the endpoints of interest in a free energy calculation are sufficiently different that we will need a chain of states that gradually change the potential energy function from $U_0$to $U_1$, as discussed in Section~\ref{}.
-One can simply carry out BAR between each pair of states $\Delta G_{1 \rightarrow N} = \Delta {G_{1\rightarrow 2}} + \Delta {G_{2\rightarrow 3}} +  \ldots + \Delta G_{N-1\rightarrow N}$.
-
-There is one important thing to note about the uncertainty estimates when summing multiple free energies together to calculate an overall free energy estimate.
-Although BAR itself gives a free energy estimate that is asymptotically correct in and is much less biased than the uncertainty estimate for EXP, the uncertainties in $\Delta {G_{i-1\rightarrow i}}$ and $\Delta {G_{i\rightarrow i+1}}$ are not uncorrelated, because they both involve the energies $U_i(\vec{q})$.
-The variances of each of the free energies will \textit{not} propagate as variances usually do (in quadrature) to the variance of the overall free energy.
-Instead, some other method for propagating the uncertainty, such as bootstrapping~\cite{Grossfield:L:2018} must be used.
-
-\paragraph{Thermodynamic integration (TI)}
-
-By taking the derivative of the free energy with respect to the
-variable $\lambda$, we find that:
-\begin{equation}
-\frac{df}{d\lambda} = \frac{d}{d\lambda} \ln \int \exp{-u(\lambda,\vec{q})} d\vec{q} = \expect{\frac{du(\lambda,\vec{q})}{d\lambda}\
-}_{\lambda} 
-\end{equation}
-
-And than we can then numerically integrate $df/d\lambda$ over an alchemical transformation, using a range of different well-established techniques, to obtain:
-\begin{equation}
-\Delta f    = \int_{0}^{1} \expect{\frac{du(\lambda,\vec{q})}{d\lambda}}_{\lambda}  d\lambda    
-\end{equation}
-This approach to calculating the free energy is called thermodynamic integration (TI). Averaging over $\expect{\frac{du}{d\lambda}}$ requires
-fewer uncorrelated samples to reach a given level of relative error
-than averaging $e^{-u(\vec{q})}$, as the range is usually
-narrower with a more Gaussian distribution.
-Rather than being limited by overlap, as in the case of BAR and MBAR, we are instead limited by the bias in the numerical quadrature, which must be minimized sufficiently to be beneath the level of statistical noise.
-
-Various numerical integration schemes are possible, but the trapezoid
-rule provides a simple and robust scheme.  All types of
-numerical integration can be written as:
-\[ \Delta f \approx \sum_{k=1}^{K} w_k
-\expect{\frac{du(\lambda,\vec{q})}{d\lambda}}_{k} \] where the weights
-$w_k$ correspond to a particular choice of numerical integration.
-Researchers have tried a large number of different integration
-schemes~\cite{resat1993studies,jorge2010effect,shyu2009reducing}. However, many other integration choices require specific choices of $\lambda$
-to minimize bias, which makes them unsuitable when the intermediates
-have widely-varying levels of uncertainty.
-For example, integrating a cubic spline interpolation provided negligble benefits~\cite{paliwal2011benchmark}.
-For starting researchers, we therefore recommend a simple trapezoid rule scheme, as it allows for maximal flexibility in which values of $\lambda$ are simulated.
-Because fitting to higher
-order polynomials can have numerical instabilities for some energy functions, and because alternate functional forms might only be appropriate
-with some types of transformations, expertise and experience is
-required to perform such numerical integration modifications.
-In practice, adding 2-3 more intermediate states is typically sufficient to match the performance of these more complicated numerical quadrature schemes.
-
-One drawback of TI is that it requires derivatives with respect to $\lambda$ to be calculated directly in the code. Unfortunately, many problems of interest
-require using pathways (such as the soft-core pathways, for removing repulsive interactions) that are not linear, as we discuss, making this more complex.
-Still, if the code of interest does compute $\frac{du}{d\lambda}$, then TI is perhaps the simplest method to use, as it involves a very little post-processing,
-
-\paragraph{The multistate Bennett acceptance ratio (MBAR)}
-
-One can generalize Bennett's logic from two states to multiple states to obtain a free energy estimator that uses energy differences between all intermediate states.
-MBAR gives a system of implicit equations for the free energies $f_i$:
-\begin{equation}
-f_i = - \ln \sum_{n=1}^{N} \frac{\exp(-u_i(\vec{q}_n))}{\sum_{k=1}^K N_k \exp(f_k-u_k(\vec{q}_n))}
-\end{equation}
-where there are $N_k$ samples from each of $K$ states, with $\sum_k N_k=N$ the total number of samples.
-Thus, we need to evaluate the energy function $U_i$ for all samples obtained at all states in the transformation.
-The equations can be solved by a number of different standard routines.
-We note that there are only $K-1$ independent equations, so only K-1 of the free energies are independent variables, and one of the $f_i$ must be specified (usually, without loss of generality, setting it to zero).
-
-MBAR is provably the lowest variance asymptotically unbiased estimator of the free energies given the energies of the samples~\cite{tan2004likelihood}, which means thaht BAR is also for the free energy difference between only 2 states, as it is mathematically exactly the same as MBAR i this case.
-MBAR also provides an uncertainty estimate, derived from standard error propagation methods, which has been shown to be accurate as long as there are sufficient samples at each state~\cite{paliwal2011benchmark}.
-
-MBAR can also be thought of as the Zwanzig estimator of the free energy to state $i$ where the sampled distribution is the \textit{mixture distribution} of all the other samples thrown together in one ``pot'', defined by $p_m(\vec{q}) = N^{-1} \sum_k  N_k \exp(f_k-u_k\vec{q})$, the weighted average of all the individual normalized probability distributions that are performed.~\cite{Shirts:AC:2017}.
-
-\paragraph{Recommendations}
-
-\begin{itemize}
-\item We recommend MBAR if all energy differences are available. It is provably the lowest variance free energy estimate given samples from multiple states.
-\item BAR is essentially just as good as MBAR for highly optimized $\lambda$ intermediates.
-Specifically, if the $\lambda$s are chosen such that intermediate states have moderate overlap with their neighbors (i.e.
-between $i$ and $i+1$ and between $i$ and $i-1$, they will \textit{not} have significant overlap with their next nearest neighbors $i+2$ and $i-2$.
-Thus MBAR doesn't actually get significant information from these energy differences, so one might as well not even calculate them, and just perform BAR between nearest neighbors.~\cite{paliwal2011benchmark}
-\item TI usually gives similar values as MBAR implemented with sufficient numbers of intermediates, but quadrature error hard to estimate beforehand.~\cite{paliwal2011benchmark}
-\item WHAM is an approximation to MBAR, and there are no compelling reasons it should be used. 
-\item Other variants, especially ones that adaptively determine the free energies can be  useful in certain circumstances but beyond the scope of a Best Practices article.
-\end{itemize}
-
-\subsubsection{Uncertainty estimation}
-Analysis of independent samples of energy data from a series of equilibrium simulations allows for the estimation of the free energy and error associated with its estimate. 
-Quantification of different error metrics depends on the form of analysis and method being used, such as those outlined above.
-
-The computation of free energies using TI (section \ref{sec:step3}) is straightforward and the trapezoidal rule is often recommended since it allows unequal spacing of lambda states, which is required to minimize the variance in the free energy estimate.
-The determination of regions of high curvature when estimating the integral is helpful to determine regions of phase space where more sampling and/or more lambda states are necessary to obtain the best approximation of the integral.
-Additionally, computation of the overall variance of TI requires the calculation of the overall variance of integration, rather than each individual $\Delta$G$_{i,i+1}$ and assuming variances add independently.
-Therefore, $\mathrm{var}$($\Delta$f) = $\sum_{i=1}^{K}w_{k}^2 \mathrm{var}(\frac{du}{d\lambda})_{k}$.
-
-For alchemical changes that result in smooth, low curvature sets of $\expect{\frac{dU}{d\lambda}}$, a relatively small number of lambda states is necessary for sufficient accuracy and low variance in the free energy estimate.
-However, with increasing numbers of perturbed atoms, the bias introduced by discretization of the integral can become large due to increased curvature, and more lambda intermediate states become necessary to reduce error.
-It is recommended that researchers verify that a sufficient number of states are included such that the free energy is essentially invariant to the number of lambda intermediate states chosen.
-
-Compared with TI, the MBAR method (section \ref{sec:step3}) discussed above provides uncertainty estimation directly from solving a set of linear equations to compute the variances between all states.
-The number of states and amount of sampling should be optimized to minimize the uncertainty in the MBAR free energy estimate, while balancing other key considerations such as computational expense.
-
-In some cases, several analysis methods can be employed to analyze the same set of simulations, providing an opportunity for synergy.
-Specifically, agreement in the free energy across different estimators can be compared and discrepancy between estimates used to assess convergence.
-For example, differences between free energy estimates computed with BAR, MBAR, and TI correlated with larger errors in a set of MCL1 inhibitors.
-Those perturbations where free energies across different estimators shared high agreement with each other (variance less than 0.1 kcal/mol) had a r$^2$ of 0.623 and mean unsigned error (MUE) of 0.787 kcal/mol (N=17) compared with a r$^2$ of 0.295 and MUE of 1.49 kcal/mol (N=23) (unpublished results).
-
-Uncertainty can also be assessed for a particular perturbation by repeating calculations with slight changes in initial configurations, forcefield parameters, and different random seeds in the MD engine.
-The assessment of variability in free energy calculations due to repeating simulations has been previously reported~\cite{aldeghi2019accurate,paliwal2011benchmark}, and large variance in free energies estimated from simulations with different random seeds should be flagged as issues with convergence or problems with system set up.
-System set up errors include poor ligand or protein preparation, atom mapping, or forcefield parameters as described above.
-
-For relative binding free energy calculations, additional sensitivity analysis can be performed by changing the initial configurations of non-core regions of the perturbation topology and determining if this change in configurations results in a large differences in the computed relative free energy, indicating poor sampling of ligand configuration.
-The proposed changes in configuration are increasingly relevant if no experimental evidence is available to reduce uncertainty in where the changing atoms should be positioned.
-
-In addition to statistical uncertainty and sampling, a variety of other factors can impact results from binding free energy calculations.
-In addition to the choice of initial configuration, results can depend on the choice of force field for the protein/receptor, water, and small molecuel(s), so rerunning calculations with different choices of force field can also be used to assess how sensitive results and conclusions are to these particular choices.
-Other factors, like system preparation (choice of protonation state, tautomer, counterion presence, salt concentration, etc.) can also substantially impact results~\cite{mobley2017predicting, mobley2017predictingb}, so unless modelers are confident they have these factors correct, sensitivity to these choices may also need to be examined.
-
-\subsubsection{Are my simulations any good?}
-\label{sec:are-they-good}
-One way of assessing reliability of the calculations is checking the phase space overlap between neighboring $\lambda$-windows~\cite{wu2005phasespace,wu2005phasespacea}.
-For this purpose, a so-called overlap matrix can be used.
-This is a $K\times K$ matrix, where $K$ is the number of simulated states, i.e.
-$\lambda$s.
-Sufficient overlap is important for reweighting estimators such as BAR or MBAR, but cannot help assess reliability of estimates, when using TI.
-These matrices are graphical representations of the phase space overlap, i.e.
-the average probability that a sample generated at state $j$ can be observed at state $i$.
-As this probability is computed considering the samples from all states (and not just the adjacent states), the values in each row and column add up to 1.
-In this analysis, the goal is to ensure every state has overlap with its neighbors in both directions -- so that off-diagonal elements are significantly larger than zero.
-Usually this means the matrix should be at least tridiagonal.
-
-Details on the calculation and properties of these matrices can be found elsewhere~\cite{klimovich2015guidelines}.
-In an overlap matrix $\mathbf{O}$, the off-diagonal values (${O}_{i,j,i\ne j}$) are negatively correlated with the variance of the free energy difference.
-Accordingly, the uncertainty of the free energy difference between the states i and j will be smaller when ${O}_{i,j,i\ne j}$ is larger (and thus the values in the main diagonal (${O}_{i,j,i=j}$) are smaller).
-In order to obtain a reliable estimate of the free energy all neighbouring states must be connected, i.e.
-there must be sufficient overlap between the samples of these states (general description: ${O}_{i,j,i\ne j}\ge$ threshold).
-However, due to the mathematical derivation it is difficult to explicitly describe the relation of the overlap matrix and the variance by formulae.
-Consequently, the threshold has to be derived empirically.
-It has been proposed that the values of the first off-diagonals (i.e.
-the diagonals above and below the main diagonal) should at least be 0.03 to obtain a reliable free energy estimate~\cite{klimovich2015guidelines}.
-Smaller values should be considered as a warning sign (see Figure~\ref{fig:overlap}), as the variance tends to be underestimated in case of poor overlap.
-
-\begin{figure}
-\includegraphics[width=0.95\columnwidth]{figures/fig12_overlap/Figure.pdf}
-\caption{\label{fig:overlap} \textbf{Overlap matrices:} Visualising overlap matrices can help with assessing the quality of simulation data.
-(A) shows good overlap with all first off-diagonal entries well above 0.03, the suggested threshold, (B) is an example of mediocre overlap with good overlap at lower $\lambda$ values and poor overlap at high $\lambda$ values.
-(C)shows poor overlap resulting in disconnected simulations that should not be used for an MBAR estimation of free energies.}
-\end{figure}
-
-Figure~\ref{fig:overlap} shows examples of good, mediocre and poor overlap in ~\ref{fig:overlap} (A), (B), and (C) respectively.
-For (A), the probability to find a sample from state $i$ in its neighbouring state $j$ is about 0.2 for all states adjacent to the main diagonal, and hence the overall connectivity is good.
-In the case of (B), the overlap is strongly diminishing in the lower right corner, raising concerns regarding the reliability of the free energy estimate obtained.
-For (C), the state at $\lambda$ index = 6 is connected to neither of its neighbouring states.
-While this does not necessarily imply that the result for this perturbation is wrong, the energy estimate must at least be considered as highly unreliable.
-In order to overcome the issue of poor overlap in this example, additional sampling should be performed by introducing additional states, i.e. $\lambda$ windows.
-%
-Interestingly, as the variance is inversely correlated with the number of states\cite{klimovich2015guidelines}, it can in principle be reduced below any arbitrary threshold with enough simulation time and a large enough number of lambda windows.
-However, decreasing the variance to a value close to 0 is not feasible, as this approach would strongly increase the calculation time.
-While variance can be decreased both by increasing simulation length and by increasing the number of lambda values, if overlap is poor, the latter approach often has a larger immediate impact.
-Specifically, often the goal is to adjust the $\lambda$ spacing to perform additional sampling in the areas of poor overlap, rather than using arbitrarily spaced windows.
-Different approaches are described in section \ref{part_of_section_six}) and more details can be found in the literature~\cite{dakka2018concurrent, hahn2019alchemical}.
-%
-In a nutshell, it must be ensured that the values of an overlap matrix in the first diagonals above and below the main diagonal are at least 0.03 in order to yield reliable results. 
-%
-There are several other parameters which should be investigated.
-Basic analysis of the trajectories should be performed to ensure that the ligand does not tumble out of the binding site by measuring its RMSD.
-Furthermore, distributions of the ligand's dihedral angles can be obtained and compared to experimental values.
-Thus, poorly sampled conformations can be identified and the issue can be adequately addressed, e.g.
-by prolonging the simulation time or choosing a different starting conformation.
-In general, running the same perturbation using different ligand poses may offer insights into whether the calculated free energies are dependent on the initial conformation.
-Another easy check can be performed by comparing the results obtained by different free energy estimators (e.g.
-TI and MBAR).
-In principal, the outcomes of both estimators should be the same (within a certain range).
-Hence, if the outcomes deviate by more than a predefined threshold (e.g.
-1.0 kcal/mol), the free energies are likely unreliable and further calculations are necessary.
-
-\begin{figure}
-    \includegraphics[width=0.95\linewidth]{figures/fig9_convergence/Figure.pdf}
-    \caption{Average binding free energy of 5 replicate Hamiltonian replica exchange calculations as a function of total simulation time (i.e.
-the sum of the simulation time of all replicas) for the two host-guest systems CB8-G3 and OA-G3.
-Shaded areas represent 95\% confidence intervals around the mean computed from the 5 replicates data.
-The horizontal dashed lines show the final binding free energy prediction of the two calculations after a total of ~5230 ns for OA-G3 and ~6650 ns for CB8-G3.
-Longer correlation times in CB8-G3 cause the calculation to converge more slowly.
-The original data used to generate the plot can be found at \url{https://github.com/MobleyLab/SAMPL6/blob/master/host_guest/Analysis/SAMPLing/Data/reference_free_energies.csv}.
-}
-    \label{fig:freeenergytrajectories}
-\end{figure}
-
-\paragraph{Cycle closure error}
-Relative free energy calculations, which compute the change in free energy on making a change to a molecule (e.g.
-adding a functional group to a ligand) may provide an additional opportunity for error/consistency checking.
-Particularly, such calculations are often done to span a graph or tree of free energy calculations~\cite{xu2019diffnet,wang2013modeling,liu2013lead} and in some cases the free energy change to go between molecules A and B can be obtained via multiple transformation pathways.
-This allows a type of consistency checking where we assess how much the free energy change for that transformation in practice differs from equivalence.
-Significant deviations from this typically indicate insufficient configurational sampling along the lambda schedule of one or more of the transformations involved.
-
-This approach may be generalised to sets of connected transformations given the requirement that the sum of free energy changes along edges of a closed cycle should be zero.
-This analysis is called ``cycle closure''.
-In practice, such thermodynamic cycles do not actually sum to zero,  and deviations become increasingly large as the size of the cycle increases owing to propagation of error.
-Though no firm guidelines have emerged, it may be judicious to perform additional configurational sampling along edges of a network that are involved in cycles closing poorly.
-This may be done by extending simulations duration, or by averaging free energy changes over multiple repeats.
-The latter approach may yield more reproducible free energy changes, but at the expense of a stronger bias on the estimated free energies due to repeated use of the same input coordinates.
-
-A scheme to reduce cycle closure errors is used in FEP+ whereby calculated free energy changes along the nodes of the network are re-sampled assuming estimates of the calculated free energy change along a node may be obtained from a Gaussian distribution centered on the estimated free energy change and with a standard deviation equal to the estimated standard deviation of the free energy change.
-The procedure then uses a Maximum likelihood method to find new sets of free energy changes that minimize cycle closure errors~\cite{wang2013modeling}.
-An alternative approach computes the free energy change between a target and reference compound as a weighted average over all unique paths in the network, with the weights derived from the propagated uncertainties of each node (REF).
-
-\paragraph{Reversible binding simulations}
-An even more stringent test of the correctness of binding free energy calculations is to compare the results to the equilibrium binding constants derived from long timescale reversible binding simulations~\cite{pan2017quantitative}.
-For small ligands with millimolar affinities, repeated binding to and unbinding from the protein can occur for a large number of times in a sufficiently long unbiased MD simulation (10-100$\mu$s), and the equilibrium binding constants can be straightforwardly computed from the ratio of bound to unbound fractions of the simulation time.
-The agreement between the binding free energy calculations and the reversible binding simulations--given the same system preparation and the same force field parameters--will strongly support the correctness of both calculations, as the same results are arrived at by two independent methods, and any discrepancy will suggest some systematic error in one, or both, of the two methods.
-This test should be applied to all free energy calculation codes in the future.
-
-\subsection{Recommendations}
-\todo[inline]{ASJSM:Maybe here we have a quick summary on what things should be checked for to make sure simulations are ok?}
-\begin{itemize}
-\item Examining output data for common problems with discussions of what exactly to plot or look at; examples of typical curves for dV/dlambda and free energy versus lambda, for example
-\begin{itemize}
-\item Make sure ligand doesn’t tumble out of binding site (Mey has observed this)
-\item Significant discrepancies between different free energy estimators (TI, BAR, MBAR)
-\item Poor replica mixing (for replica-exchange)
-\item Correlation time as a function of lambda as it would be expected to be a smooth
-\item Dependence on initial conformation
-\item Torsional analysis: Is it stuck in specific states? Only very rarely transitions?
-\item More “usual suspects”
-\end{itemize}
-\item Other considerations for many transformations
-\end{itemize}
-
-\subsection{Best practices for data representation}
-
-Practitioners in the field are encouraged to follow best practices on analysis of results and data depictions.
-The following points should be considered when plotting alchemical free energy prediction results.
-\todo[inline]{JS: @everyone I suspect people will have strong opinions on this subject, feel free to include your own suggestions}
-\paragraph{Statistics to include}
-As with any modelling technique, misuse of statistical analysis can skew the perception of how well models perform in FE predictions.
-First and foremost, error estimates should always be included on your predictions in whatever form you present your data (scatterplots, barplots, etc; see next paragraph).
-It is recommended to perform triplicates of your predictions at minimum to ensure some measure of reliability in your data, in part because analytical uncertainty estimates often underestimate the true statistical uncertainty.
-
-Because of the strong drug discovery context and the relevance of alchemical FE predictions to SAR projects, both the overall correlation of your predictions relative to experiment and the ability of your models to detect the best-scoring compounds (i.e.
-rankings) should be computed.
-Conventionally, this means including an R\textsuperscript{2} (or Pearson r) and a Kendall \texttau{} metric in your results.
-Additionally, practitioners may choose to include a Spearman \textrho{} as well.
-Practitioners should take particular care with correlation based metrics, however, when  the dataset dynamic range is small, or only a small number of compounds appear at extrema of the dynamic range.
-Bootstrapping analysis -- including variation in the dataset -- can help with this by providing useful error bars on performance metrics.
-
-Mean unsigned error (MUE, also called mean absolute error/MAE) is the another key statistic to include in your results.
-Even though some models' near-perfect correlation and ranking statistics might suggest excellent accuracy, MUE values can still be multiple kcal/mol, providing important additional insight into performance.
-Furthermore, MUE allows for unbiased comparisons between predictive models as it is less sensitive to dataset size.
-Even though this section may serve as an introduction to the main statistical practices, the authors recommend further reading on evaluation of computational models \cite{cite Nicholls pt1 and pt2 + Jain&Nicholls "Recommendations for.."}.
-
-\paragraph{Presenting your data}
-Seeing as essentially all alchemical FE prediction schemes are regression problems, the preferred type of plot is a scatter plot (see Figure \ref{fig:scatterplot_analysis}).
-Most alchemical FE projects will concern 10-50 ligands; any study with \textless10 ligands is more suitable for bar plots (with inclusion of error bars).
-Any study with \textgreater50 ligands should be trellised to not overestimate statistics.
-In that respect, it is bad practice to place multiple datasets on the same plot because this can suggest high model accuracy even though the individual models perform less well \cite{cite blogpost Pat Walters}.
-
-Because we are interested mainly in the linear relationship between the alchemical FE predictions and the experimentally-determined affinity values, plots should be depicted with the same range on both axes (i.e.
-$x=y$).
-Furthermore, bounds should be depicted for the 1- and 2-kcal/mol confidence regions.
-These regions can serve as tools to communicate your model performance: any perturbation inside the 1 kcal/mol region can be seen as highly reliable, any perturbation inside the 2 kcal/mol region should be seen as somewhat reliable, and any perturbation outside the confidence regions should be expected to be unreliable (and handled as outliers).
-In a drug discovery context, this type of data depiction may suggest the reliability of alchemical FE predictions in the project, and can give an idea of how trustworthy predictions can be for synthesis ideas.
-\begin{figure}
-    \includegraphics[width=0.95\linewidth]{figures/fig13_analysis_practices/Figure.pdf}
-    \caption{Example of a well-depicted alchemical free energy prediction showing the relation between predicted and experimentally-determined Gibbs free energy in kcal per mol with standard errors as error bars.
-The two outliers (underestimations) are shown in red; the dark and light-orange regions depict the 1- and 2-kcal per mol confidence bounds.}
-    \label{fig:scatterplot_analysis}
-\end{figure}
-
-\section{Conclusions}
-\todo[inline, color={green!20}]{ASJSM:Write conclusions}
-Alchemical free energy calculations have seen a vast increase in popularity also in the pharmaceutical industry for structure based drug discovery~\cite{schind}.
-Commercial products such as FEP+ and Flare,
-
-%%%%%%%%%%%%%%%%%%%%
-%                Software                    %
-%%%%%%%%%%%%%%%%%%%%
-\section{Available software -- a summary}
-\todo[inline]{ASJSM: Expand and add more references. }
-\label{sec:software}
-There are several softwares availble for carrying out alchemical free energy calculations. In this section we discuss some of the popular commercial and free tools along with their features. 
-\begin{itemize}
-\item Commercial:
-   \begin{itemize}
-    \item FEP+ is a tool offered by Schr\"{o}dinger Inc.
-under a commercial license.
-It has an intuitive GUI which makes it easier for non-experts to run alchemical free energy calculations and analyze the results.
-It runs DESMOND MD package under the hood and hence parallelizes very well on the GPUs.
-    \item Flare is a commercial structure-based drug design software offered by Cresset.
-Similar to FEP+ it has an easily accessible user interface and strives to facilitate free energy calculations for non-experts while offering advanced users full control via a Python API.
-It only runs on GPUs, using CUDA or OpenCL.
-    \item The molecular operating environment (MOE) offered by the Chemical Computing Group (CCG) has a tool for performing free energy calculations. It is build on AMBER-TI (cf. below).
-    \end{itemize}
-\item Free or low-cost for academics / commercial for industry:
-	\begin{itemize}
-	\item CHARMM has a variety of tools developed over the years.
-PERT module can be used to define initial and final states and define the intermediate lambda points.
-FREN and BAR modules can be used to analyze the data after the MD run.
-Lambda-dynamics based free energy calulation can be carried out using the BLOCK module.
-	\item TIES and AMBER FEW? (Peter Coveney)
-	\item AMBER, including its new pmemd.cuda version support free energy calculation. 
-	\end{itemize}
-\item Free (libre) open source:
-	\begin{itemize}
-	\item PLUMED is an open source tool which enables the usage of a variety of MD engines.
-It is designed as a plugin for MD packages such that it analyzes the trajectory on the fly.
-It also offers a VMD based plugin for the computation of collective variables.
-	\item SIRE  is a free, open source, multiscale molecular simulation framework, written to allow computational modellers to quickly prototype and develop new algorithms for molecular simulation and molecular design.
-	\item YANK is a tool developed by John Chodera and group on the top of OpenMM MD package. It allows the users to write their inputs in easy-to-use YAML format.
-	\item GROMACS is a molecular simulation package with a significant number of free energy methods implementations.
-The LiveCOMS GROMACS tutorial has an example free energy calculation~\cite{lemkul2018From}.
-	\item pmx for mutations
-	\end{itemize}
-\item Setup tools
-	\begin{itemize}
-	\item FESetup: AMBER, gromacs, Sire
-	\item Lomap/Lomap2 : Relative alchemical transformation graph planning
-	\item CHARMM-GUI is a web based tool for setting up a variety of MD simulations. It can be used to generate CHARMM scripts for solvation and ligand-binding free energy calculations.
-	\end{itemize}
-\item Analysis tools:
-	\begin{itemize}
-	\item Free Energy Workflows: Sire-specific free energy map analysis using weighted path averages
-	\url{https://github.com/michellab/freenrgworkflows}
-	\item Alchemlyb: Multipackage free energy analysis
-	\url{https://github.com/alchemistry/alchemlyb}
-	\item pymbar: MBAR implementation, but have to roll your own analysis wrapper
-	\url{https://github.com/choderalab/pymbar}
-	\end{itemize}
-\end{itemize}
-
-\section{Online resources}
-\begin{itemize}
-\item \url{http://www.ks.uiuc.edu/Training/Workshop/Urbana_2010A/lectures/TCBG-2010.pdf}
-\item Basic Ingredients of Free Energy Calculations: A Review (\url{DOI: 10.1002/jcc.21450})
-\item Good Practices in Free-Energy Calculations (\url{DOI: 10.1021/jp102971x})
-\item Alchemical Free Energy Methods for Drug Discovery: Progress and Challenges (\url{doi: 10.1016/j.sbi.2011.01.011})
-\item Alchemistry wiki: \url{http://www.alchemistry.org/wiki/Best_Practices}
-\end{itemize}
-
-\clearpage
-\section{Benchmark Datasets}
-
-\label{sec:benchmark}
-\vspace{5mm}
-
-\begin{table}
-\caption{Overview of benchmark datasets}
-\begin{tabular}{cccc}
-
-\textbf{Publication} & \textbf{Targets} & \textbf{Ligands} & \textbf{Force Field} \\
-\hline
-\multicolumn{4}{|c|}{D3R Grand Challenges~\cite{D3R}} \\
-\hline
-GC3~\cite{gaieb2019d3r} & 6 & 266 & various \\
-GC2~\cite{gaieb2018d3r} & 1 & 102 & various \\
-GC2015~\cite{gathiaka2016d3r} & 2 & 215 & various \\
-\hline
-\multicolumn{4}{|c|}{SAMPL Challenges~\cite{SAMPL}} \\
-\hline
-SAMPL6~\cite{rizzi2018overview} & 3 & 21 & various \\
-SAMPL5~\cite{yin2017overview} & 3 & 22 & various \\
-SAMPL4~\cite{muddana2014sampl4} & 2 & 23 & various \\
-\hline
-\multicolumn{4}{|c|}{Schrödinger Datasets} \\
-\hline
-FEP+ Dataset~\cite{wang2015accurate} & 8 & 199 & OPLS2.1 \\
-FEP+ Dataset~\cite{harder2016opls3} & 8 & 199 & OPLS3 \\
-FEP+ Dataset~\cite{roos2019opls3e} & 8 & 199 & OPLS3e \\
-FEP+ Dataset~\cite{song2019using} & 8 & 199 & AMBER18 \\
-Fragment Optimization~\cite{steinbrecher2015accurate} & 8 & 96 & OPLS2.1 \\
-Scaffold Hopping~\cite{wang2017accurate} & 6 & 21 & OPLS3 \\
-Macrocycles~\cite{yu2017accurate} & 7 & 33 & OPLS3 \\
-\hline
-\multicolumn{4}{|c|}{Further Suggested Datasets} \\
-\hline
-Cucurbit[7]uril (CB7)~\cite{mobley2017predicting} & 1 & 15 & NA \\
-Deep cavity cavitand~\cite{mobley2017predicting} & 2 & 19 & NA \\
-T4 Lysozyme~\cite{mobley2017predicting} & 2 & 20 & NA \\
-Merck benchmarking set~\cite{MCompChem2019Sep} & 5 & 169 & OPSL3 \\
-
-\hline
-\label{tab:benchmarks}
-\end{tabular}
-\end{table}
-
-Several other case studies of FEP calculations can be found in the review by Williams-Noonan et al.~\cite{williams-noonan2018free}.
-For comparison of FEP+ and Gromacs (using the AMBER99SB-ILDN and GAFF2 force field), cf.
-the recently published study by Pérez-Benito et al.~\cite{perez-benito2019predicting}.
-An overview of further suggested benchmark sets can be found in the review by Mobley and Gilson~\cite{mobley2017predicting} or on \url{alchemistry.org}~\cite{alchemistry}.
-These include cyclodextrins, the Cytochrome C peroxidase (CCP) protein model binding site, thrombin and bromodomains as well as solvation benchmark sets~\cite{paliwal2011benchmark}.
-Further work on shared protein-ligand benchmark sets is ongoing at \url{https://github.com/openforcefield/PLBenchmarks}.
-
-\section{Checklist}
-\label{sec:checklist}
-\todo[inline, color={green!20}]{JS: @everyone Please input your own suggestions, I made some points off the top of my head. Feel free to add helpful citations to any of the bullet points}
-\todo[inline, color={green!20}]{JS: @self,ASJSM do some final formatting when text is complete.}
 \onecolumn
@@
 \twocolumn
-% This provides a checklist which
-% - spans a full page
-% - consists of multiple sub-checklists
-% - exists on a separate page
-% This style of checklist will be especially helpful if you want to encourage readers to print and use your checklist in practice, as they
-% can easily print it without also printing other material from your manuscript. However, other styles of checklist are also possible (below).
-% \begin{Checklists*}[p!]
-
-% \begin{checklist}{Step 0 -- Know what you want to simulate }
-% \textbf{What are the first questions that need addressing before setting up a molecular dynamics simulation}\\
-% Extensive explanation for the checklist questions can be found in Section~\ref{sec:step0}.
-% \begin{itemize}
-% \item Can I get the required accuracy with the simulation I want to carry out?
-% \item Have I properly prepared my protein and ligand systems?
-% \item Does my system contain any groups that require custom parameters?
-% \item What simulation protocol will provide the most evidence to answer my hypothesis?
-
-% \item And finally
-% \end{itemize}
-% \end{checklist}
-
-% \begin{checklist}{Simulation preparation}
-% \textbf{How do I get started setting up an alchemical free energy calculation}
-% Extensive explanation for the checklist questions can be found in Section~\ref{sec:step1}.
-% \begin{itemize}
-% \item Have I followed the Best practices for biomolecular simulation set up?
-% \item In a relative simulation, will I run into problems with clashing geometries in the ligand transformation or crystal waters?
-% \end{itemize}
-% \end{checklist}
-% \end{Checklists*}
-
-% \begin{Checklists*}[p!]
-% \begin{checklist}{Absolute simulations}
-% \textbf{What are the main things I need to consider for an absolute alchemical free energy calculation?}
-% Extensive explanation for the checklist questions can be found in Section~\ref{sec:step2}.
-% \begin{itemize}
-% \item Topology
-% \item Restraints
-% \item Standard state handling
-% \end{itemize}
-% \end{checklist}
-
-% \begin{checklist}{Relative simulations}
-% \textbf{What are the main things I need to consider for an relative alchemical free energy calculation?}
-% Extensive explanation for the checklist questions can be found in Section~\ref{sec:step2}.
-% \begin{itemize}
-% \item First thing
-% \item Also remember
-% \item And finally
-% \end{itemize}
-% \end{checklist}
-
-% \begin{checklist}{Analysis}
-% \textbf{This is all about analysis of the simulation}
-% Extensive explanation for the checklist questions can be found in Section~\ref{sec:step4}.
-% \begin{itemize}
-% \item Are my simulations converged enough?
-% \item Am I using the right analysis techniques?
-% \end{itemize}
-% \end{checklist}
-
-% \end{Checklists*}
-
-%Checklist:
-\clearpage
-%%%%%%%%%%%%%%%% 
-%% Checklist-specific precommands:
-
-\clearpage
-
-\section*{Author Contributions}
-%%%%%%%%%%%%%%%%
-% This section mustt describe the actual contributions of
-% author. Since this is an electronic-only journal, there is
-% no length limit when you describe the authors' contributions,
-% so we recommend describing what they actually did rather than
-% simply categorizing them in a small number of
-% predefined roles as might be done in other journals.
-%
-% See the policies ``Policies on Authorship'' section of https://livecoms.github.io
-% for more information on deciding on authorship and author order.
-%%%%%%%%%%%%%%%%
-\todo[inline, color={green!20}]{ASJSM: @everyone, can you check your contribution is acknowledge and if not add the appropriate information. }
-(Explain the contributions of the different authors here)
-ASJSM: Coordinated the document, contributed to most sections, and co-designed figures ~\ref{} and created figures \ref{fig:fig_what_is_lambda},~\ref{fig:fig1_what_is_alchemy},~\ref{fig:overlap} and replotted~\ref{fig:automatic-equilibration-detection} and improved ~\ref{fig:fig_types_of_networks}.
-JS: created figures \ref{fig:fig_what_is_alchemy},~\ref{fig:fig_binding_thermodynamic_cycle},~\ref{fig:fig_topology},~\ref{fig:fig_mcss},~\ref{fig:fig_types_of_networks},~\ref{fig:fig_absolute_thermodynamic_cycle},~\ref{fig:scatterplot_analysis}, the data representation section, the checklist and contributed to general formatting discussions and editing.
-MK: Contributed to the Data Analysis section, provided the data for figure~\ref{fig:overlap} and helped edit the paper.
-DLM: Contributed to the outline, drafted some of the sections, gave ideas on figures, and helped edit the paper.
-GT: Contributed to introduction and drug discovery sections, and helped edit the paper.
-AR: Created figure~\ref{fig:freeenergytrajectories}, contributed to section \ref{sec:step2}, and helped edit the paper.
-MRS: Helped create figure~\ref{fig:fig_what_is_lambda}, wrote sections describing choices for alchemical pathways and analysis for free energy calculations.
-LNN: Helped write the simulation length, stopping conditions, and information saving section.
-BA: Helped write the uncertainty estimation, stopping conditions, and output analysis sections and created figure ~\ref{fig:convergence_forward_reverse}.
-JM: Contributed to sections~\ref{subsec:reproducible},~\ref{sec:step1},~\ref{} 4.2, 5.3, 6, 8.3.1, 8.3.2\
-JDC: \todo[inline, color={red!20}]{@JDC insert info here}
-HX: Contributed section~\ref{subsec:accuracy}, to section~\ref{sec:relative-fe-protocol}, and to section~\ref{sec:are-they-good}.
-
-% We suggest you preserve this comment:
-For a more detailed description of author contributions,
-see the GitHub issue tracking and changelog at \githubrepository.
-
-\section*{Other Contributions}
-%%%%%%%%%%%%%%%
-% You should include all people who have filed issues that were
-% accepted into the paper, or that upon discussion altered what was in the paper.
-% Multiple significant contributions might mean that the contributor
-% should be moved to authorship at the discretion of the a
-%
-% See the policies ``Policies on Authorship'' section of https://livecoms.github.io for
-% more information on deciding on authorship and author order.
-%%%%%%%%%%%%%%%
-
-(Explain the contributions of any non-author contributors here)
-% We suggest you preserve this comment:
-For a more detailed description of contributions from the community and others, see the GitHub issue tracking and changelog at \githubrepository.
-
-\section*{Potentially Conflicting Interests}
-%%%%%%%
-%Declare any potentially competing interests, financial or otherwise
-%%%%%%%
-
-Declare any potentially conflicting interests here, whether or not they pose an actual conflict in your view.
-MK is employed by Cresset who recently released an commercial tool for performing alchemical free energy calculations.
-
-\section*{Funding Information}
-%%%%%%%
-% Authors should acknowledge funding sources here. Reference specific grants.
-%%%%%%%
-ASJSM and JM acknowledge funding through an EPSRC flagship software grant: EP/P022138/1
-MK acknowledges funding through Innovate UK by KTP partnership 011120.
-AR acknowledges partial support from the Tri-Institutional Program in Computational Biology and Medicine and the Sloan Kettering Institute.
-
-\bibliography{alchemical1,manual}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%% APPENDICES
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\appendix
-
-\end{document}

### 2020-01-16 paper/manuscript.tex
working backwards with fixes

--- 2020-01-14 paper/manuscript.tex
@@
 As some of the theoretical background can seem daunting, we provide an essentials guide to the theory behind alchemical free energy calculations in Section~\ref{sec:theory}.
-In the remainder of this paper, we will cover topics that are key to the preparation~(\ref{sec:step1}), choice and use of correct protocols~(\ref{sec:step2}), and finally the best practices that should be used for the analysis of alchemical calculations~(\ref{sec:step3}).
+In the remainder of this paper, we will cover topics that are key to the preparation~(\ref{sec:step1}), choice and use of correct protocols~(\ref{sec:step2}), and finally the best practices that should be used for the analysis of alchemical calculations~(\ref{sec:data_analysis}).
 Particular focus will be given to aspects of the molecular simulations which are unique to alchemical calculations---these include the calculation of transfer free energies (hydration free energies, partition coefficients, etc.), and binding free energies (absolute and relative).
@@
 \subsection{How accurate are alchemical free energy calculations?}
-\label{subsec:accuracy}
+\label{subsec:expectation}
 %
@@
 Based on the input ligand series a perturbation map or network can be planned.
-Recent heuristics have shown the more connected the perturbation network the better, however, there is a way to optimise network structure while minimising the number of perturbations that need to be computed minimising the resulting computational cost~\cite{yang2019optimal}.
+Recent heuristics have shown the more connected the perturbation network the better, however, there is a way to optimise network structure while minimising the number of perturbations that need to be computed minimising the resulting computational cost~\cite{yang2020optimal}.
 Sometimes the introduction of intermediates that are not part of the original congeneric series are essential to avoid ring breaking, or dealing with perturbations that would otherwise result in large numbers of atoms to be grown or disappearing.
@@
 Methods in experimental design have been applied to the construction of the perturbation maps.
-Yang et al.~\cite{yang2019optimal} optimized the perturbation map by selecting a fixed number of calculations from the pairwise perturbations so that the resulting set of calculations minimize the total variance.
+Yang et al.~\cite{yang2020optimal} optimized the perturbation map by selecting a fixed number of calculations from the pairwise perturbations so that the resulting set of calculations minimize the total variance.
 Xu~\cite{xu2019diffnet} optimized the perturbation map by allocating different amounts of simulation time to different pairwise perturbations so as to minimize the total variance, given the total simulation time of all the perturbation calculations.
@@
 %
-\subsubsection{The importance of the alchemical pathway\label{sec:important_path}}
+\subsubsection{The importance of the alchemical pathway
+\label{sec:important_path}}
 \todo[inline, color={red!40}]{LNN: write common principles alchemical path choice. MRS: mostly wrote this, LNN can comment on it. LNN: LGTM}
@@
 How decorrelation impacts calculations, and how to compute it is discussed in other sections.  [section ref]
-
-\section{Step 3 -- Data analysis}
-\label{sec:step3}
-
+%
+\section{Data analysis}
+\label{sec:data_analysis}
 Once data has been collected from alchemical intermediates, it must be analyzed to produce an estimate of the free energy change (and its associated statistical uncertainty) for each leg of the thermodynamic cycle.
 While a number of different estimators are available that will give consistent results under optimal circumstances, some approaches are recommended over others due to their robustness and ability to provide information on poor convergence.
-
+%
 \subsection{Detecting the boundary between equilibrated and production regions}
 \label{sec:automatic-equilibration-detection}
-
 Much of the infrastructure for analyzing alchemical free energy calculations relies on the concept of asymptotically unbiased estimators, which produce unbiased estimates of the free energy when fed very long simulations~\cite{shirts2005comparison}.
@@
 As a result, these estimators can produce significantly biased estimates if fed the entirety of simulation data generated without further processing~\cite{chodera2016simple}.
-
+%
 \begin{figure}
@@
 If this is not possible, advanced analysis techniques that assume only local equilibrium (rather than global equilibrium) such as the TRAM estimators~\cite{mey2014xtram,wu2016multiensemble,nuske2017markov} may be more appropriate.
-
+%
 \subsection{Decorrelating samples for analysis}
 \label{sec:decorrelating-samples}
-
 \paragraph{Computing the statistical inefficiency}
@@
 We recommend using the robust statistical inefficinecy computation routines available within the MBAR~\cite{kylebeauchamp2019choderalab} and alchemlyb~\cite{daviddotson2019alchemistry} packages.
-
+%
 \paragraph{Subsampling data to generate uncorrelated samples}
@@
 Suppose the correlated timeseries is $\{a_t\}_{t=1}^T$; we can form a new timeseries of $N_{\mathrm{eff}} \approx T / g$ effectively uncorrelated samples by selecting a subset of indices $\{ \: t = \mathrm{round}((n-1) \, g) \: | \: n \in \mathrm{range}(1,\ldots ,N) \: \}$ where $\mathrm{round(x)}$ denotes rounding to the nearest integer.
-
+%
 If independent simulations are used, the alchemical state $\lambda$ may have a significant impact on the correlation time, and these simulations should be subsampled independently using a separate estimate of the statistical inefficinecy $g$ for each alchemical state.
@@
 Conveniently, the approach described above for automated equilibration detection produces an appropriate estimate of $g$ over the production region for automating this process.
-
+%
 \paragraph{Cautions and considerations}
 Reliable estimation of the statistical inefficiency is difficult, and estimates will not generally be as precise (in a relative error sense) as averages.
-To ensure there is sufficient data available for reliable decorrelation and estimation of free energy differences, it is recommended that the effective number of uncorrelated samples $N_{\mathrm{eff}} \ge \approx 50$ if the BAR or MBAR estimators (discussed below) are used; the number may need to be much higher with alternate estimators.
-
+To ensure there is sufficient data available for reliable decorrelation and estimation of free energy differences, it is recommended that the effective number of uncorrelated samples $N_{\mathrm{eff}} \ge \approx 50$ if the BAR or MBAR estimators (discussed below in sec~\ref{subsec:estimators}) are used; the number may need to be much higher with alternate estimators.
+%
 \subsection{Estimators for free energy differences}
-
+\label{subsec:estimators}
 Free energy differences between two different states differing in the energy function are directly related to the
@@
 Virtually all of the ways to estimate this free energy are based in converting this ratio of integrals to something that can be measured in one (or several) simulations.
-
+%
 \paragraph{The Zwanzig relationship (EXP)}
-
-The simplest method for calculating free energy
-differences from simulations is the so-called \textit{Zwanzig
+%
+The simplest method for calculating free energy differences from simulations is the so-called \textit{Zwanzig
 relationship}~\cite{zwanzig1954hightemperature}, also called one-sided exponential reweighting (EXP), or simply free energy perturbation, thought this final term is sometimes used to encompass all ways of calculating free energy differences.
-
+%
 The (reduced) free energy difference $\Delta f_{01}$ between an initial state 0 and a final state 1 defined by two different potential energy functions 
@@
 and the average is over all samples from the simulation performed with $u_0$.
-In the case of NVT (canonical) sampling and assuming the masses do not change, then $u$ is just $U/k_BT$, and $f$ is $A/k_BT$, but it can be generalized to other ensembles with the proper definition of $f$ and $u$.
+In the case of NVT (canonical) sampling and assuming the masses do not change, then $u$ is just $U/k_BT$, and $f$ is $F/k_BT$, but it can be generalized to other ensembles with the proper definition of $f$ and $u$.
 Described in words, we take the samples generated during our run with the potential energy function $U_0(\vec{q})$, recalculate what the energy difference would be if we switched to potential energy function $U_1(\vec{q})$, and average the exponential of the -1 times the energy difference to get -1 times the exponential of the free energy difference.
-This almost requires no extra code functionality to perform; one need only to save a full precision trajectory, and run an unmodified molecular simulation code using the $U_1$ in order to calculate the new energies of stored snapshots.
+The original distributions (P($U_0$) as generated at $\lambda=0$ and P($U_1$) would look like the ones seen in figure~\ref{fig:fig_sampling_scheme} (A)-(C)on the left hand side.
+Reevaluating  requires almost no extra code functionality to perform; one need only to save a full precision trajectory, and run an unmodified molecular simulation code using the $U_1$ in order to calculate the new energies of stored snapshots.
 The analysis can be written in a line of code.
-We note that this method is even more general, in that the instantaneous work to change the potential energy function from $U_0$ to $U_1$ can be replaced by the non-reversible work $W$ to make the same change under the same equilibrium conditions at either end state~\cite{jarzynski1997nonequilibrium,jarzynski1998equilibrium,crooks2000pathensemble}, though we do not go into all of the details of nonequilibrum transformations here, and refer the reader to more advanced treatments~\cite{maragakis2008bayesian,oberhofer2005biased,procacci2015unbiased,shirts2003equilibriuma,ytreberg2004singleensemble}
-
-Although the Zwanzig equation is formally correct (as long as the two states considered sample the same phase space volume, which is true for standard molecular models), it has some very important numerical issues that mean that it often performs badly for standard free energy calculations, even for small molecules.~\cite{shirts2005comparison,lu2003appropriate}  One can show that if the standard deviation of the difference $\Delta U(\vec{q}) = U_1(vec{q})-U_2(\vec{q})$ over all sampled $\vec{q}$ is large (which in this case, means only several times $k_BT)$, then very few samples contribute to the average, and the answer will be both biased and extremely noisy~\cite{xx}.
+We note that this method is even more general, in that the instantaneous work to change the potential energy function from $U_0$ to $U_1$ can be replaced by the non-reversible work $W$ to make the same change under the same equilibrium conditions at either end state~\cite{jarzynski1997nonequilibrium,jarzynski1998equilibrium,crooks2000pathensemble}, though we do not go into all of the details of non-equilibrium transformations here, and refer the reader to more advanced treatments~\cite{maragakis2008bayesian,oberhofer2005biased,procacci2015unbiased,shirts2003equilibriuma,ytreberg2004singleensemble}
+%
+Although the Zwanzig equation is formally correct (as long as the two states considered sample the same phase space volume, which is true for standard molecular models), it has some very important numerical issues that mean that it often performs badly for standard free energy calculations, even for small molecules~\cite{shirts2005comparison,lu2003appropriate}.
+One can show that if the standard deviation of the difference $\Delta U(\vec{q}) = U_1(vec{q})-U_2(\vec{q})$ over all sampled $\vec{q}$ is large (which in this case, means only several times $k_BT)$, then very few samples contribute to the average, and the answer will be both biased and extremely noisy~\cite{lelievre2010free}.
 Essentially, the method is dominated by contributions of rare snapshots\cite{jarzynski2006rare, wu2005phasespaceb, wu2005phasespacec}.
 There are a number of applications where it this expression is nevertheless useful, but those usually involve investigating the dependence of the free energy on small changes in parameters, not changing molecules or even single atoms in molecules~\cite{xx}
-
+%
 \paragraph{The Bennett Acceptance Ratio (BAR)}
-
+%
 If we have the differences in the potential energy sampled from the distribution defined by $U_0$ to the state defined by $U_1$, and we also have the differences in potential energies from the distribution sampled by $U_1$ to the state defined by $U_0$, we can obtain a significantly improved estimate of the
@@
  ]} \nonumber \\
- =\sum_{i=1}^{n_j} \frac{1}{1 + \exp[\ln(\frac{n_i}{n_j}) - U_{ij}(\vec{q}) + f_ij)]}
+ =\sum_{i=1}^{n_j} \frac{1}{1 + \exp[\ln(\frac{n_i}{n_j}) - U_{ij}(\vec{q}) + f_ij)]},
 \end{eqnarray}
 where $n_i$ and $n_j$ are the number of samples from each state. We
-will refer to this method as BAR. More recent derivations show that this formula is the maximum likelihood estimate of the free energy difference
-given sets of samples from the two states~\cite{shirts.bennett}.
-
-Many studies have demonstrated both the
-theoretical and practical superiority of BAR over EXP in molecular
-simulations~\cite{shirts2005comparison,lu2003appropriate}, and BAR converges to EXP in the limit that all samples are from a
-single state~\cite{bennett1976efficient,shirts.bennett}. BAR also requires significantly less
-overlap between the configurational space of each state to converge than EXP, though some overlap must still exist.
-
+will refer to this method as BAR.
+More recent derivations show that this formula is the maximum likelihood estimate of the free energy difference given sets of samples from the two states~\cite{bennett1976efficient,shirts2003equilibriuma }.
+%
+Many studies have demonstrated both the theoretical and practical superiority of BAR over EXP in molecular
+simulations~\cite{shirts2005comparison,lu2003appropriate}, and BAR converges to EXP in the limit that all samples are from a single state~\cite{bennett1976efficient,bennett1976efficient,shirts2003equilibriuma}.
+BAR also requires significantly less overlap between the configurational space of each state to converge than EXP, though some overlap must still exist.
+%
 The Bennett acceptance ratio is only defined between two states.
-Usually, the endpoints of interest in a free energy calculation are sufficiently different that we will need a chain of states that gradually change the potential energy function from $U_0$to $U_1$, as discussed in Section~\ref{}.
-One can simply carry out BAR between each pair of states $\Delta G_{1 \rightarrow N} = \Delta {G_{1\rightarrow 2}} + \Delta {G_{2\rightarrow 3}} +  \ldots + \Delta G_{N-1\rightarrow N}$.
-
+Usually, the endpoints of interest in a free energy calculation are sufficiently different that we will need a chain of states that gradually change the potential energy function from $U_0$to $U_1$, as discussed in Section~\ref{sec:important_path}.
+You can simply carry out BAR between each pair of states $\Delta G_{1 \rightarrow N} = \Delta {G_{1\rightarrow 2}} + \Delta {G_{2\rightarrow 3}} +  \ldots + \Delta G_{N-1\rightarrow N}$.
+%
 There is one important thing to note about the uncertainty estimates when summing multiple free energies together to calculate an overall free energy estimate.
@@
 The variances of each of the free energies will \textit{not} propagate as variances usually do (in quadrature) to the variance of the overall free energy.
-Instead, some other method for propagating the uncertainty, such as bootstrapping~\cite{Grossfield:L:2018} must be used.
-
+Instead, some other method for propagating the uncertainty, such as bootstrapping~\cite{grossfield2018best} must be used.
+%
 \paragraph{Thermodynamic integration (TI)}
-
 By taking the derivative of the free energy with respect to the
@@
 \end{equation}
-
 And than we can then numerically integrate $df/d\lambda$ over an alchemical transformation, using a range of different well-established techniques, to obtain:
@@
 \end{equation}
-This approach to calculating the free energy is called thermodynamic integration (TI). Averaging over $\expect{\frac{du}{d\lambda}}$ requires
-fewer uncorrelated samples to reach a given level of relative error
-than averaging $e^{-u(\vec{q})}$, as the range is usually
-narrower with a more Gaussian distribution.
-Rather than being limited by overlap, as in the case of BAR and MBAR, we are instead limited by the bias in the numerical quadrature, which must be minimized sufficiently to be beneath the level of statistical noise.
-
+This approach to calculating the free energy is called thermodynamic integration (TI).
+Averaging over $\expect{\frac{du}{d\lambda}}$ requires fewer uncorrelated samples to reach a given level of relative error
+than averaging $e^{-u(\vec{q})}$, as the range is usually narrower with a more Gaussian distribution.
+Rather than being limited by overlap, as in the case of BAR and MBAR (see below), we are instead limited by the bias in the numerical quadrature, which must be minimized sufficiently to be beneath the level of statistical noise.
+%
 Various numerical integration schemes are possible, but the trapezoid
-rule provides a simple and robust scheme.  All types of
-numerical integration can be written as:
+rule provides a simple and robust scheme.  All types of numerical integration can be written as:
 \[ \Delta f \approx \sum_{k=1}^{K} w_k
-\expect{\frac{du(\lambda,\vec{q})}{d\lambda}}_{k} \] where the weights
-$w_k$ correspond to a particular choice of numerical integration.
-Researchers have tried a large number of different integration
-schemes~\cite{resat1993studies,jorge2010effect,shyu2009reducing}. However, many other integration choices require specific choices of $\lambda$
-to minimize bias, which makes them unsuitable when the intermediates
+\expect{\frac{du(\lambda,\vec{q})}{d\lambda}}_{k} \] where the weights $w_k$ correspond to a particular choice of numerical integration.
+Researchers have tried a large number of different integration schemes~\cite{resat1993studies,jorge2010effect,shyu2009reducing}.
+However, many other integration choices require specific choices of $\lambda$ to minimize bias, which makes them unsuitable when the intermediates
 have widely-varying levels of uncertainty.
-For example, integrating a cubic spline interpolation provided negligble benefits~\cite{paliwal2011benchmark}.
+For example, integrating a cubic spline interpolation provided negligible benefits~\cite{paliwal2011benchmark}.
 For starting researchers, we therefore recommend a simple trapezoid rule scheme, as it allows for maximal flexibility in which values of $\lambda$ are simulated.
-Because fitting to higher
-order polynomials can have numerical instabilities for some energy functions, and because alternate functional forms might only be appropriate
-with some types of transformations, expertise and experience is
-required to perform such numerical integration modifications.
+Because fitting to higher order polynomials can have numerical instabilities for some energy functions, and because alternate functional forms might only be appropriate with some types of transformations, expertise and experience is required to perform such numerical integration modifications.
 In practice, adding 2-3 more intermediate states is typically sufficient to match the performance of these more complicated numerical quadrature schemes.
-
-One drawback of TI is that it requires derivatives with respect to $\lambda$ to be calculated directly in the code. Unfortunately, many problems of interest
-require using pathways (such as the soft-core pathways, for removing repulsive interactions) that are not linear, as we discuss, making this more complex.
-Still, if the code of interest does compute $\frac{du}{d\lambda}$, then TI is perhaps the simplest method to use, as it involves a very little post-processing,
-
+%
+One drawback of TI is that it requires derivatives with respect to $\lambda$ to be calculated directly in the code.
+Unfortunately, many problems of interest require using pathways (such as the soft-core pathways, for removing repulsive interactions) that are not linear, as we discuss, making this more complex.
+Still, if the code of interest does compute $\frac{du}{d\lambda}$, then TI is perhaps the simplest method to use, as it involves a very little post-processing.
+%
 \paragraph{The multistate Bennett acceptance ratio (MBAR)}
-
 One can generalize Bennett's logic from two states to multiple states to obtain a free energy estimator that uses energy differences between all intermediate states.
@@
 We note that there are only $K-1$ independent equations, so only K-1 of the free energies are independent variables, and one of the $f_i$ must be specified (usually, without loss of generality, setting it to zero).
-
-MBAR is provably the lowest variance asymptotically unbiased estimator of the free energies given the energies of the samples~\cite{tan2004likelihood}, which means thaht BAR is also for the free energy difference between only 2 states, as it is mathematically exactly the same as MBAR i this case.
+%
+MBAR is provably the lowest variance asymptotically unbiased estimator of the free energies given the energies of the samples~\cite{tan2004likelihood}, which means that BAR is also for the free energy difference between only 2 states, as it is mathematically exactly the same as MBAR in this case.
 MBAR also provides an uncertainty estimate, derived from standard error propagation methods, which has been shown to be accurate as long as there are sufficient samples at each state~\cite{paliwal2011benchmark}.
-
-MBAR can also be thought of as the Zwanzig estimator of the free energy to state $i$ where the sampled distribution is the \textit{mixture distribution} of all the other samples thrown together in one ``pot'', defined by $p_m(\vec{q}) = N^{-1} \sum_k  N_k \exp(f_k-u_k\vec{q})$, the weighted average of all the individual normalized probability distributions that are performed.~\cite{Shirts:AC:2017}.
-
+%
+MBAR can also be thought of as the Zwanzig estimator of the free energy to state $i$ where the sampled distribution is the \textit{mixture distribution} of all the other samples thrown together in one ``pot'', defined by $p_m(\vec{q}) = N^{-1} \sum_k  N_k \exp(f_k-u_k\vec{q})$, the weighted average of all the individual normalized probability distributions that are performed.~\cite{shirts2017reweighting}.
+%
 \paragraph{Recommendations}
-
 \begin{itemize}
@@
 between $i$ and $i+1$ and between $i$ and $i-1$, they will \textit{not} have significant overlap with their next nearest neighbors $i+2$ and $i-2$.
-Thus MBAR doesn't actually get significant information from these energy differences, so one might as well not even calculate them, and just perform BAR between nearest neighbors.~\cite{paliwal2011benchmark}
-\item TI usually gives similar values as MBAR implemented with sufficient numbers of intermediates, but quadrature error hard to estimate beforehand.~\cite{paliwal2011benchmark}
+Thus MBAR does not actually get significant information from these energy differences, so one might as well not even calculate them, and just perform BAR between nearest neighbors.~\cite{paliwal2011benchmark}
+\item TI usually gives similar values as MBAR implemented with sufficient numbers of intermediates, but quadrature errors are hard to estimate beforehand.~\cite{paliwal2011benchmark}
 \item WHAM is an approximation to MBAR, and there are no compelling reasons it should be used. 
@@
 \end{itemize}
-
-\subsubsection{Uncertainty estimation}
-Analysis of independent samples of energy data from a series of equilibrium simulations allows for the estimation of the free energy and error associated with its estimate. 
-Quantification of different error metrics depends on the form of analysis and method being used, such as those outlined above.
-
-The computation of free energies using TI (section \ref{sec:step3}) is straightforward and the trapezoidal rule is often recommended since it allows unequal spacing of lambda states, which is required to minimize the variance in the free energy estimate.
-The determination of regions of high curvature when estimating the integral is helpful to determine regions of phase space where more sampling and/or more lambda states are necessary to obtain the best approximation of the integral.
+%
+\subsection{Uncertainty estimation}
+\label{subsec:uncertainty}
+It is important to consider the variation in your computed free energies from your equilibrium simulations and in this way obtaining an estimate on the certainty, or uncertainty of the obtained value for the free energies you are interested in.
+A recent best practices paper by Grossfiled et.
+al~\cite{grossfield2018best} goes into great detail on estimating uncertainties from molecular simulations and is a good starting point for this topic.
+In general the quantification of different error metrics depends on both data generation and analysis methods  used from the ones discussed above. 
+%
+The computation of free energies using TI (section~\ref{subsec:estimators}) is straightforward and the trapezoidal rule is often recommended since it allows unequal spacing of $\lambda$ states, which is required to minimize the variance in the free energy estimate, but in principle any good numerical integration method can be used.
+The determination of regions of high curvature when estimating the integral is helpful to determine regions of phase space where more sampling and/or more $\lambda$ states are necessary to obtain the best approximation of the integral.
+Generally a plot of $\lambda$ with respect to the gradients at each of the $\lambda$ values can be helpful here.
 Additionally, computation of the overall variance of TI requires the calculation of the overall variance of integration, rather than each individual $\Delta$G$_{i,i+1}$ and assuming variances add independently.
 Therefore, $\mathrm{var}$($\Delta$f) = $\sum_{i=1}^{K}w_{k}^2 \mathrm{var}(\frac{du}{d\lambda})_{k}$.
-
-For alchemical changes that result in smooth, low curvature sets of $\expect{\frac{dU}{d\lambda}}$, a relatively small number of lambda states is necessary for sufficient accuracy and low variance in the free energy estimate.
-However, with increasing numbers of perturbed atoms, the bias introduced by discretization of the integral can become large due to increased curvature, and more lambda intermediate states become necessary to reduce error.
+%
+For alchemical changes that result in smooth, low curvature sets of $\expect{\frac{dU}{d\lambda}}$, a relatively small number of $\lambda$ states is necessary for sufficient accuracy and low variance in the free energy estimate.
+However, with increasing numbers of perturbed atoms, the bias introduced by discretization of the integral can become large due to increased curvature, and more $\lambda$ intermediate states become necessary to reduce error.
 It is recommended that researchers verify that a sufficient number of states are included such that the free energy is essentially invariant to the number of lambda intermediate states chosen.
-
-Compared with TI, the MBAR method (section \ref{sec:step3}) discussed above provides uncertainty estimation directly from solving a set of linear equations to compute the variances between all states.
+%
+Compared with TI, the MBAR method (section~\ref{subsec:estimators}) discussed above provides uncertainty estimation directly from solving a set of linear equations to compute the variances between all states.
 The number of states and amount of sampling should be optimized to minimize the uncertainty in the MBAR free energy estimate, while balancing other key considerations such as computational expense. 
-
+%
 In some cases, several analysis methods can be employed to analyze the same set of simulations, providing an opportunity for synergy.
@@
 Those perturbations where free energies across different estimators shared high agreement with each other (variance less than 0.1 kcal/mol) had a r$^2$ of 0.623 and mean unsigned error (MUE) of 0.787 kcal/mol (N=17) compared with a r$^2$ of 0.295 and MUE of 1.49 kcal/mol (N=23) (unpublished results).
-
+%
 Uncertainty can also be assessed for a particular perturbation by repeating calculations with slight changes in initial configurations, forcefield parameters, and different random seeds in the MD engine.
-The assessment of variability in free energy calculations due to repeating simulations has been previously reported~\cite{aldeghi2019accurate,paliwal2011benchmark}, and large variance in free energies estimated from simulations with different random seeds should be flagged as issues with convergence or problems with system set up.
-System set up errors include poor ligand or protein preparation, atom mapping, or forcefield parameters as described above.
-
+The assessment of variability in free energy calculations due to repeating simulations has been previously reported~\cite{aldeghi2019accurate,paliwal2011benchmark,mey2016blinded,mey2018impact}, and large variance in free energies estimated from simulations with different random seeds should be flagged as issues with convergence.
+Other sources that may show up as suspicious uncertainties may in fact be system set up errors.
+Common issues include poor ligand or protein preparation, atom mapping for relative calculations, or forcefield parameters as described above.
+%
 For relative binding free energy calculations, additional sensitivity analysis can be performed by changing the initial configurations of non-core regions of the perturbation topology and determining if this change in configurations results in a large differences in the computed relative free energy, indicating poor sampling of ligand configuration.
 The proposed changes in configuration are increasingly relevant if no experimental evidence is available to reduce uncertainty in where the changing atoms should be positioned.
-
+%
 In addition to statistical uncertainty and sampling, a variety of other factors can impact results from binding free energy calculations.
@@
 Other factors, like system preparation (choice of protonation state, tautomer, counterion presence, salt concentration, etc.) can also substantially impact results~\cite{mobley2017predicting, mobley2017predictingb}, so unless modelers are confident they have these factors correct, sensitivity to these choices may also need to be examined.
-
-\subsubsection{Are my simulations any good?}
+%
+\subsection{Are my simulations any good?}
 \label{sec:are-they-good}
+There are different easily measurable indicators that can be looked at to test how well converged simulations are and if all alchemical states have been sufficiently sampled for a rigorous analysis.
+Furthermore, once you have established that individula perturbations are well behaved there are some tricks to ensure the overall perturbation network gives reliable results.
+%
+\paragraph{Convergence of simulations}
+Fig.~\ref{fig:freeenergytrajectories} illustrates how looking at the convergence of your data may be important.
+In this example the same guest G3 shows different convergence behaviour for two different hosts.
+CB8 with guest G3 has a longer correlation time than the octa acid (OA) host.
+In some cases slow correlation time may not be expected and therefore not a feature known in advance.
+To this end you should always look at all simulation data available check convergence behaviour for each free energy estimate and if need be extend existing simulations or try an approach that requires simulations in two separate binding modes where they interconvert at very slwo timescales.
+\begin{figure}
+    \includegraphics[width=0.90\linewidth]{figures/fig9_convergence/Figure.pdf}
+    \caption{Average binding free energy of 5 replicate Hamiltonian replica exchange calculations as a function of total simulation time (i.e.
+the sum of the simulation time of all replicas) for the two host-guest systems CB8-G3 and OA-G3.
+Shaded areas represent 95\% confidence intervals around the mean computed from the 5 replicates data.
+The horizontal dashed lines show the final binding free energy prediction of the two calculations after a total of ~5230 ns for OA-G3 and ~6650 ns for CB8-G3.
+Longer correlation times in CB8-G3 cause the calculation to converge more slowly.
+The original data used to generate the plot can be found at \url{https://github.com/MobleyLab/SAMPL6/blob/master/host_guest/Analysis/SAMPLing/Data/reference_free_energies.csv}.
+}
+    \label{fig:freeenergytrajectories}
+\end{figure}
+%
+\paragraph{Overlap matrix}
 One way of assessing reliability of the calculations is checking the phase space overlap between neighboring $\lambda$-windows~\cite{wu2005phasespace,wu2005phasespacea}.
@@
 Usually this means the matrix should be at least tridiagonal.
-
+%
 Details on the calculation and properties of these matrices can be found elsewhere~\cite{klimovich2015guidelines}.
 In an overlap matrix $\mathbf{O}$, the off-diagonal values (${O}_{i,j,i\ne j}$) are negatively correlated with the variance of the free energy difference.
-Accordingly, the uncertainty of the free energy difference between the states i and j will be smaller when ${O}_{i,j,i\ne j}$ is larger (and thus the values in the main diagonal (${O}_{i,j,i=j}$) are smaller).
+Accordingly, the uncertainty of the free energy difference between the states $i$ and $j$ will be smaller when ${O}_{i,j,i\ne j}$ is larger (and thus the values in the main diagonal (${O}_{i,j,i=j}$) are smaller).
 In order to obtain a reliable estimate of the free energy all neighbouring states must be connected, i.e.
@@
 the diagonals above and below the main diagonal) should at least be 0.03 to obtain a reliable free energy estimate~\cite{klimovich2015guidelines}.
-Smaller values should be considered as a warning sign (see Figure~\ref{fig:overlap}), as the variance tends to be underestimated in case of poor overlap.
+Smaller values should be considered as a warning sign (see Fig.~\ref{fig:overlap}(C)), as the variance tends to be underestimated in case of poor overlap.
 
@@
 
-Figure~\ref{fig:overlap} shows examples of good, mediocre and poor overlap in ~\ref{fig:overlap} (A), (B), and (C) respectively.
-For (A), the probability to find a sample from state $i$ in its neighbouring state $j$ is about 0.2 for all states adjacent to the main diagonal, and hence the overall connectivity is good.
-In the case of (B), the overlap is strongly diminishing in the lower right corner, raising concerns regarding the reliability of the free energy estimate obtained.
-For (C), the state at $\lambda$ index = 6 is connected to neither of its neighbouring states.
+Fig.~\ref{fig:overlap} shows examples of good, mediocre and poor overlap (A), (B), and (C) respectively.
+For Fig.~\ref{fig:overlap}(A), the probability to find a sample from state $i$ in its neighbouring state $j$ is about 0.2 for all states adjacent to the main diagonal, and hence the overall connectivity is good.
+In the case of Fig.~\ref{fig:overlap}(B), the overlap is strongly diminishing in the lower right corner, raising concerns regarding the reliability of the free energy estimate obtained.
+For Fig.~\ref{fig:overlap}(C), the state at $\lambda$ index = 6 is connected to neither of its neighbouring states.
 While this does not necessarily imply that the result for this perturbation is wrong, the energy estimate must at least be considered as highly unreliable.
@@
 %
-Interestingly, as the variance is inversely correlated with the number of states\cite{klimovich2015guidelines}, it can in principle be reduced below any arbitrary threshold with enough simulation time and a large enough number of lambda windows.
+Interestingly, as the variance is inversely correlated with the number of states\cite{klimovich2015guidelines}, it can in principle be reduced below any arbitrary threshold with enough simulation time and a large enough number of $\lambda$ windows.
 However, decreasing the variance to a value close to 0 is not feasible, as this approach would strongly increase the calculation time.
-While variance can be decreased both by increasing simulation length and by increasing the number of lambda values, if overlap is poor, the latter approach often has a larger immediate impact.
+While variance can be decreased both by increasing simulation length and by increasing the number of $\lambda$ values, if overlap is poor, the latter approach often has a larger immediate impact.
 Specifically, often the goal is to adjust the $\lambda$ spacing to perform additional sampling in the areas of poor overlap, rather than using arbitrarily spaced windows.
-Different approaches are described in section \ref{part_of_section_six}) and more details can be found in the literature~\cite{dakka2018concurrent, hahn2019alchemical}.
+Different approaches are described in section~\ref{}) and more details can be found in the literature~\cite{dakka2018concurrent, hahn2019alchemical}.
 %
@@
 1.0 kcal/mol), the free energies are likely unreliable and further calculations are necessary.
-
-\begin{figure}
-    \includegraphics[width=0.95\linewidth]{figures/fig9_convergence/Figure.pdf}
-    \caption{Average binding free energy of 5 replicate Hamiltonian replica exchange calculations as a function of total simulation time (i.e.
-the sum of the simulation time of all replicas) for the two host-guest systems CB8-G3 and OA-G3.
-Shaded areas represent 95\% confidence intervals around the mean computed from the 5 replicates data.
-The horizontal dashed lines show the final binding free energy prediction of the two calculations after a total of ~5230 ns for OA-G3 and ~6650 ns for CB8-G3.
-Longer correlation times in CB8-G3 cause the calculation to converge more slowly.
-The original data used to generate the plot can be found at \url{https://github.com/MobleyLab/SAMPL6/blob/master/host_guest/Analysis/SAMPLing/Data/reference_free_energies.csv}.
-}
-    \label{fig:freeenergytrajectories}
-\end{figure}
-
+%
 \paragraph{Cycle closure error}
@@
 Significant deviations from this typically indicate insufficient configurational sampling along the lambda schedule of one or more of the transformations involved.
-
+%
 This approach may be generalised to sets of connected transformations given the requirement that the sum of free energy changes along edges of a closed cycle should be zero.
@@
 The latter approach may yield more reproducible free energy changes, but at the expense of a stronger bias on the estimated free energies due to repeated use of the same input coordinates.
-
+%
 A scheme to reduce cycle closure errors is used in FEP+ whereby calculated free energy changes along the nodes of the network are re-sampled assuming estimates of the calculated free energy change along a node may be obtained from a Gaussian distribution centered on the estimated free energy change and with a standard deviation equal to the estimated standard deviation of the free energy change.
 The procedure then uses a Maximum likelihood method to find new sets of free energy changes that minimize cycle closure errors~\cite{wang2013modeling}.
-An alternative approach computes the free energy change between a target and reference compound as a weighted average over all unique paths in the network, with the weights derived from the propagated uncertainties of each node (REF).
-
+An alternative approach computes the free energy change between a target and reference compound as a weighted average over all unique paths in the network, with the weights derived from the propagated uncertainties of each node~\cite{mey2016blinded}.
+Approaches as illustrated by Yang et al.
+for perturbation map design can also be used to compute relative free energies between target and reference compounds~\cite{yang2020optimal}.
+%
 \paragraph{Reversible binding simulations}
@@
 The agreement between the binding free energy calculations and the reversible binding simulations--given the same system preparation and the same force field parameters--will strongly support the correctness of both calculations, as the same results are arrived at by two independent methods, and any discrepancy will suggest some systematic error in one, or both, of the two methods.
-This test should be applied to all free energy calculation codes in the future.
-
+As part of validation testing of alchemical free energy codes a benchmark set to compare alchemical and direct computation of equilibrium binding constant should become standard in future.
+%
 \subsection{Recommendations}
@@
 \end{itemize}
-
+%
 \subsection{Best practices for data representation}
-
-Practitioners in the field are encouraged to follow best practices on analysis of results and data depictions.
-The following points should be considered when plotting alchemical free energy prediction results.
+\label{sec:plot_data}
+Data generation and most adequate analysis isn't everything.
+As a practitioner of alchemical free energy simulations you are also encouraged to adhered to best practice for representing and plotting your results.
+We would encourage the following standard set of analyses and ways to represent data.
 \todo[inline]{JS: @everyone I suspect people will have strong opinions on this subject, feel free to include your own suggestions}
 \paragraph{Statistics to include}
-As with any modelling technique, misuse of statistical analysis can skew the perception of how well models perform in FE predictions.
+As with any modelling technique, misuse of statistical analysis can skew the perception of how well models perform in free energy predictions.
 First and foremost, error estimates should always be included on your predictions in whatever form you present your data (scatterplots, barplots, etc; see next paragraph).
-It is recommended to perform triplicates of your predictions at minimum to ensure some measure of reliability in your data, in part because analytical uncertainty estimates often underestimate the true statistical uncertainty.
-
-Because of the strong drug discovery context and the relevance of alchemical FE predictions to SAR projects, both the overall correlation of your predictions relative to experiment and the ability of your models to detect the best-scoring compounds (i.e.
+It is recommended to perform triplicates of your predictions at minimum to ensure some measure of reliability in your data, in part because analytical uncertainty estimates often underestimate the true statistical uncertainty, where this is not possible lower error bounds such as given by the MBAR estimator should be included.
+%
+Because of the strong drug discovery context and the relevance of alchemical free energy predictions to structure activity relationship (SAR) projects, both the overall correlation of your predictions relative to experiment and the ability of your models to detect the best-scoring compounds (i.e.
 rankings) should be computed.
-Conventionally, this means including an R\textsuperscript{2} (or Pearson r) and a Kendall \texttau{} metric in your results.
+Conventionally, this means including an R\textsuperscript{2} (or Pearson R -- 1 high correlation 0 no correlation -1 high anti-correlation) and a Kendall \texttau{} (with perfect ranking agreement when \texttau=1 and perfect disagreement when \texttau=-1) metric in your results.
 Additionally, practitioners may choose to include a Spearman \textrho{} as well.
-Practitioners should take particular care with correlation based metrics, however, when  the dataset dynamic range is small, or only a small number of compounds appear at extrema of the dynamic range.
-Bootstrapping analysis -- including variation in the dataset -- can help with this by providing useful error bars on performance metrics.
-
+Brown et al.~\cite{brown2009healthy} have provided a useful analysis in terms of upper bounds of expected possible correlations between experiment and computation with a given potency range for the compounds.
+For example for potency ranges of 2 log units it would be impossible to get a higher correlation in R than 0.8 because of experimental uncertainties~\cite{brown2009healthy}.
+What often is neglected to include is an error analysis on correlation statistics that arise from the errors of both experimental and computed data.
+One way to include such error analysis for correlation metrics is using bootstrapping on the datasets.
+The D3R community challenges follows best practices on their data evaluation with readily available python scripts online~\cite{2018drugdata}, based on work by Pat Walters~\cite{walters2013what}.
+Other analysis software also provide similar functionality for bootstrapping datasets~\cite{antonia2019michellab}.
+%
 Mean unsigned error (MUE, also called mean absolute error/MAE) is the another key statistic to include in your results.
-Even though some models' near-perfect correlation and ranking statistics might suggest excellent accuracy, MUE values can still be multiple kcal/mol, providing important additional insight into performance.
+Even though some models' near-perfect correlation and ranking statistics might suggest excellent accuracy, MUE values can still be off multiple kcal/mol, providing important additional insight into performance.
 Furthermore, MUE allows for unbiased comparisons between predictive models as it is less sensitive to dataset size.
-Even though this section may serve as an introduction to the main statistical practices, the authors recommend further reading on evaluation of computational models \cite{cite Nicholls pt1 and pt2 + Jain&Nicholls "Recommendations for.."}.
-
+Even though this section may serve as an introduction to the main statistical practices, the authors recommend further reading on evaluation of computational models \cite{jain2008recommendations, walters2013what, brown2009healthy, walterthoughts}.
+%
 \paragraph{Presenting your data}
-Seeing as essentially all alchemical FE prediction schemes are regression problems, the preferred type of plot is a scatter plot (see Figure \ref{fig:scatterplot_analysis}).
+Seeing as essentially all alchemical free energy prediction schemes are regression problems, the preferred type of plot is a scatter plot (see Figure \ref{fig:scatterplot_analysis}).
 Most alchemical FE projects will concern 10-50 ligands; any study with \textless10 ligands is more suitable for bar plots (with inclusion of error bars).
 Any study with \textgreater50 ligands should be trellised to not overestimate statistics.
-In that respect, it is bad practice to place multiple datasets on the same plot because this can suggest high model accuracy even though the individual models perform less well \cite{cite blogpost Pat Walters}.
-
+In that respect, it is bad practice to place multiple datasets on the same plot because this can suggest high model accuracy even though the individual models perform less well \cite{walterthoughts}.
+%
 Because we are interested mainly in the linear relationship between the alchemical FE predictions and the experimentally-determined affinity values, plots should be depicted with the same range on both axes (i.e.
@@
 Furthermore, bounds should be depicted for the 1- and 2-kcal/mol confidence regions.
-These regions can serve as tools to communicate your model performance: any perturbation inside the 1 kcal/mol region can be seen as highly reliable, any perturbation inside the 2 kcal/mol region should be seen as somewhat reliable, and any perturbation outside the confidence regions should be expected to be unreliable (and handled as outliers).
+These regions can serve as tools to communicate your model performance: any predictions inside the 1 kcal/mol region can be seen as highly reliable, any predictions inside the 2 kcal/mol region should be seen as somewhat reliable, and any predictions outside the confidence regions should be expected to be unreliable (and handled as outliers).
 In a drug discovery context, this type of data depiction may suggest the reliability of alchemical FE predictions in the project, and can give an idea of how trustworthy predictions can be for synthesis ideas.
@@
     \includegraphics[width=0.95\linewidth]{figures/fig13_analysis_practices/Figure.pdf}
-    \caption{Example of a well-depicted alchemical free energy prediction showing the relation between predicted and experimentally-determined Gibbs free energy in kcal per mol with standard errors as error bars.
+    \caption{Example of a well-depicted alchemical free energy prediction showing the relation between predicted and experimentally-determined Gibbs free energy in kcal/mol with standard errors as error bars.
 The two outliers (underestimations) are shown in red; the dark and light-orange regions depict the 1- and 2-kcal per mol confidence bounds.}
@@
 \end{figure}
-
+An example of a best practice scatter comparison between computed and experimental values is shown in~\ref{fig:scatterplot_analysis}, highlighting outliers, error bars and confidence intervals.
+The data for this plot is artificially generated for illustration purposes.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Conclusions                 %
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Conclusions}
-\todo[inline, color={green!20}]{ASJSM:Write conclusions}
-%%%%%%%%%%%%%%%%%%%%
-%              Terminology                  %
-%%%%%%%%%%%%%%%%%%%%
-\section{Terminology and abbreviations}
-\label{sec:tem-abbrev}
-\begin{itemize}
-\item Feature Box covering major technical terms and abbreviations
-\item Examples:
-\begin{itemize}
-\item EXP, BAR, MBAR
-\item Double decoupling, single-topology, dual-topology, hybrid-topology, coupled-topology
-\item FEP (free energy perturbation), alchemical, AFE (alchemical free energy)
-\end{itemize}
-\end{itemize}
-\todo[inline]{JS: should this be combined with the acronyms section at the beginning of the paper?}
-%%%%%%%%%%%%%%%%%%%%
+\label{sec:conclusion}
+Alchemical free energy calculations have seen a vast increase in popularity also in the pharmaceutical industry for structure based drug discovery~\cite{schindler2020largescale, sherborne2016collaborating, wagner2017computational}.
+Commercial products such as FEP+ and Flare, which provide a convenient user interface make the setup and use of these methods a lot easier~\cite{wang2015accurate}.
+Prospective prediction challenges such as the drug design data resource grand challenges provide a community driven platform to evaluate different free energy protocols against each other on unknown targets~\cite{gaieb2018d3r, gaieb2019d3r}.
+What has come out of these in the past is that SChrödinger's FEP+, performs similar when run by different participant groups~\cite{gaieb2018d3r}, but may not give the desired accuracy with RMSE errors larger than 1 kcal/mol.
+Other alchemical protocols performed similarly well.
+We hope that the best practice guide provides a set of tools that allow a better understanding of how to setup, run and interpret alchemical free energy calculations.
+The choice of the underlying simulation software is left to the user, but hopefully with this guide it is easier to assess whether the generated data is trustworthy or not.
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % Software                    %
-%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Available software -- a summary}
@@
 JM: Contributed to sections~\ref{subsec:reproducible},~\ref{sec:step1},~\ref{} 4.2, 5.3, 6, 8.3.1, 8.3.2\
-JDC: \todo[inline, color={red!20}]{@JDC insert info here}
+JDC: Wrote section~\ref{sec:decorrelating-samples} and~\ref{sec:automatic-equilibration-detection} discussed structure and design of the whole document, suggested Figs.~\ref{fig:fig_what_is_alchemy} and~\ref{fig:fig_sampling_scheme} \todo[inline, color={red!20}]{@JDC check info}
 HX: Contributed section~\ref{subsec:accuracy}, to section~\ref{sec:relative-fe-protocol}, and to section~\ref{sec:are-they-good}.

### 2020-01-16 paper/manuscript.tex
Updates from Overleaf

@@
 One may safely assume that the correlation time is greater than 100-200 fs even for relatively simple systems such as small molecules in solvent, so saving no more frequently than every 50-100 steps is recommended.
-How decorrelation impacts calculations, and how to compute it is discussed in other sections.  [section ref]
-%
+How decorrelation impacts calculations, and how to compute it is discussed in other section~\ref{sec:decorrelating-samples}. 
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%  Data Analysis               %
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Data analysis}
@@
     \caption{{\bf Automatic partitioning into equilibration and production regions.}
-    \textbf{A} The average (black line) standard deviation (shaded region) of the reduced potential $u^*$ over many independent replicate simulations started from the same initial conditions show a significant initial transient before relaxing to the true average potential energy\textbf{B} A cumulative average (red) of the entire simulation data demonstrates simulation bias not seen when initial simulation data is omitted (blue).
-Using an automated approach to detect equilibration of the boundary $t_0$ using statistical inefficiency $g$ (\textbf{C})  for an effective simulation interval (\textbf{D}) \textbf{E} The optimal equilibration boundary $t_0$ is selected to maximize the number of uncorrelated samples.
+    \textbf{A} The average (black line) standard deviation (shaded region) of the reduced potential $u^*$ over many independent replicate simulations started from the same initial conditions show a significant initial transient change before relaxing to the true average potential energy \textbf{B}.
+A cumulative average (red) of the entire simulation data demonstrates simulation bias not seen when initial simulation data is omitted (blue).
+Using an automated approach to detect equilibration of the boundary $t_0$ using statistical inefficiency $g$ (\textbf{C})  for an effective simulation interval (\textbf{D}).
+\textbf{E} The optimal equilibration boundary $t_0$ is selected to maximize the number of uncorrelated samples.
     \emph{Figure adapted from~\cite{chodera2016simple}.}
@@
 \end{figure}
-
+%
 To minimize this effect, an initial portion of the simulation is often discarded to \emph{equilibration}~\cite{braun2019best}, with the idea of removing the most heavily biased initial portion of simulation data but retaining the unbiased \emph{production} region that represents a stationary Markov chain process sampling from the desired equilibrium target distribution.
@@
 This approach is implemented within the MBAR~\cite{kylebeauchamp2019choderalab} and alchemlyb~\cite{daviddotson2019alchemistry} packages, and is highly recommended for standard practice.
-
+%
 For additional discussion of working with correlated data and autocorrelation analysis, please refer to the work on Best Practices for Quantification of Uncertainty and Sampling Quality in Molecular Simulations~\cite{grossfield2018best}.
-
+%
 \paragraph{Computing the timeseries for equilibration detection}
@@
 \end{eqnarray}
-
+%
 Note that the use of the effective reduced potential is not guaranteed to pick up on all slow relaxation processes that may be coupled to the alchemical free energy, but the simplicity of its computation means it is generally appropriate for most cases.
-
+%
 \paragraph{Cautions in automating equilibration detection}
@@
 This situation can be detected if the equilibration boundary $t_0$ is a significant fraction of the total simulation length $T$, with a good rule of thumb being that $T \gtrsim 20 t_0$.
-If this is not possible, advanced analysis techniques that assume only local equilibrium (rather than global equilibrium) such as the TRAM estimators~\cite{mey2014xtram,wu2016multiensemble,nuske2017markov} may be more appropriate.
+If this is not possible, advanced analysis techniques that assume only local equilibrium (rather than global equilibrium) such as the TRAM estimators~\cite{mey2014xtram,wu2016multiensemble,nuske2017markov} may be more appropriate, but are beyond the scope of this paper.
 %
@@
 The basic concept is that $\tau_\mathrm{eq}$ corresponds to the single-exponential decay time for the autocorrelation process that generates samples, so the statistical inefficiency $g$ measures the approximate temporal separation between two effectively uncorrelated samples (where two exponential relaxation times are presumed to be sufficient).
-
+%
 Robust estimation of $C_t$ for $t \sim T$ is difficult due to growth in statistical error, so common estimators of $g$ make use of several additional properties of $C_t$ to provide useful estimates (see \emph{Practical Computation of Statistical Inefficiencies} in~\cite{chodera2016simple} for a detailed discussion).

### 2020-01-16 paper/manuscript.tex
Working my way through section 7, section 8 is done.

@@
 As some of the theoretical background can seem daunting, we provide an essentials guide to the theory behind alchemical free energy calculations in Section~\ref{sec:theory}.
-In the remainder of this paper, we will cover topics that are key to the preparation~(\ref{sec:step1}), choice and use of correct protocols~(\ref{sec:step2}), and finally the best practices that should be used for the analysis of alchemical calculations~(\ref{sec:data_analysis}).
+In the remainder of this paper, we will cover topics that are key to the preparation~(\ref{sec:step1}), choice and use of correct protocols~(\ref{sec:simulation_protocol_choice}), and finally the best practices that should be used for the analysis of alchemical calculations~(\ref{sec:data_analysis}).
 Particular focus will be given to aspects of the molecular simulations which are unique to alchemical calculations---these include the calculation of transfer free energies (hydration free energies, partition coefficients, etc.), and binding free energies (absolute and relative).
@@
 %
-The alchemical free energy protocols, that are discussed in Section~\ref{sec:step2}, are defined for a specific type of free energy calculation, i.e.
+The alchemical free energy protocols, that are discussed in Section~\ref{sec:simulation_protocol_choice}, are defined for a specific type of free energy calculation, i.e.
 a free energy of binding or a free energy of hydration.
@@
 %
-For binding free energy calculations using a congeneric set of ligands and therefore likely to be most suited to a relative free energy protocol (see~\ref{sec:step2}).
+For binding free energy calculations using a congeneric set of ligands and therefore likely to be most suited to a relative free energy protocol (see~\ref{sec:simulation_protocol_choice}).
 Other than the choice of forcefield for the parameterization of the ligands, some care has to be taken selecting binding poses for these ligands.
@@
 %
-%
-%
-%%%%%%%%%%%%
-% Step 1   %
-%%%%%%%%%%%%
-%
-\section{What simulation protocol should I choose? -- Step 2}
-\label{sec:step2}
-Alchemical free energy calculations can be grouped into two main categories, ``absolute'' (\ref{fig:fig_absolute_thermodynamic_cycle}) and ``relative'' \footnote{The distinction is a bit of a misnomer, since both compute ratios of partition functions relative to another state, and neither computes an absolute free energy.}(\ref{fig:fig_binding_thermodynamic_cycle}), which differ in whether they compute properties for a single molecule (absolute) or compare properties of different, usually closely related, molecules (relative).
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% What simulation protocol should I choose   %
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{What simulation protocol should I choose?}
+\label{sec:simulation_protocol_choice}
+Alchemical free energy calculations can be grouped into two main categories, ``absolute'' (see Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}) and ``relative'' \footnote{The distinction is a bit of a misnomer, since both compute ratios of partition functions relative to another state, and neither computes an absolute free energy.}(see Fig.~\ref{fig:fig_binding_thermodynamic_cycle}), which differ in whether they compute properties for a single molecule (absolute) or compare properties of different, usually closely related, molecules (relative).
 To use binding as a concrete example, in absolute binding free energy calculations, one computes the binding free energy of a ligand to an individual receptor relative to a standard reference concentration.
+%
 In contrast, in relative binding free energy calculations, one compares the binding free energy of two related inhibitors to determine the potency difference.
+%
 \subsection{Absolute and relative free energy calculations have some differences}
-Many of the issues of setup and protocol for alchemical calculations are common, but some are different between absolute and relative calculations, so before treating the common elements we consider the protocol differences.
-
+Many of the issues around simualtion setup and protocol choice for alchemical calculations are common, but there are some differences between absolute and relative calculations.
+First we will consider protocol differences  so before treating the common elements.
+%
 \subsubsection{Choices unique to relative free energy calculations}
 \label{sec:relative-fe-protocol}
-
-\paragraph{Topologies} A critical first step in relative calculations is to select an approach to these calculations, determining whether to use a \emph{dual topology}, \emph{single topology}, or \emph{hybrid topology} approach to relative calculations.
+%
+\paragraph{Topologies} 
+A critical first step in relative calculations is to select an approach to these calculations, determining whether to use a \emph{dual topology}, \emph{single topology}, or \emph{hybrid topology} approach to relative calculations.
 
@@
 In single topology calculations, the overall transformation is set up to involve as few additional atoms as possible, so benzene would be typically changed into benzyl alcohol by changing one of the hydrogens into a carbon.
-To this new carbon you also find non-interacting atoms called ``dummy atoms'' (retaining their bonded interactions but not interacting with the rest of the system), which will become the two hydrogens attached to the new carbon and the OH group.
+On this new carbon you will also find non-interacting atoms called ``dummy atoms'' (retaining their bonded interactions but not interacting with the rest of the system), which will become the two hydrogens attached to the new carbon and the OH group after the transformation.
 Bond parameters as well as partial charges are adjusted accordingly.
-Thus, in a single topology calculation, atoms may change their type so relatively few dummy atoms are created. This is illustrated in left arm of Fig.~\ref{fig:fig_topology}. 
-In contrast, in a dual topology free energy calculation, no atoms are allowed to change type~\cite{shirts2012best}.
+Thus, in a single topology calculation, atoms may change their type ensuring only relatively few dummy atoms are created. This is illustrated in left arm of Fig.~\ref{fig:fig_topology}. 
+In contrast, in a dual topology alchemical free energy calculation, no atoms are allowed to change type~\cite{shirts2012best}.
 This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxy methyl group, then passing through an intermediate state where atoms which are becoming dummy atoms or ceasing being dummy atoms are partially interacting (this state may or may not be well defined~\cite{mobley2014blind}  and culminating in a state where benzyl alcohol is present along with the additional dummy atoms which was previously a corresponding hydrogen of the benzene.
@@
     \includegraphics[width=0.95\columnwidth]{figures/fig3_topol/Figure.pdf}
-    \caption{\textbf{Two common topologies for alchemical calculations} left: A single topology uses softcore potentials to convert from one type of atom to an other.
+    \caption{\textbf{Two common topologies for alchemical calculations: single and dual topology} \textbf{left}: A single topology uses softcore potentials to convert from one type of atom to an other.
 Dummy atoms (Du) are used when there is no corresponding maximum common substructure match between the two molecules for certain atoms.
-right: The dual topology does not convert one species to another, but only converts between Du atoms and an interacting species, but usually uses softcore potentials for this.
+\textbf{right}: The dual topology does not convert one species to another, but only converts between Du atoms and an interacting species, but usually uses softcore potentials for this.
 The 'mixed' intermediate atoms are used in both dual and single topology approaches.
@@
 Hybrid topology calculations have seen much less use but essentially consist of two absolute free energy calculations in opposite directions at the same time (turning one molecule off while turning the other on)~\cite{jiang2019computing}.
-At present, the most widely used approaches, such as in Schr\"{o}dinger's FEP+~\cite{wang2019protein} and in FESetup~\cite{loeffler2015fesetup} (for which calculations may be planned with Lead Optimization Mapper (LOMAP)~\cite{liu2013lead} seem to use single topology approaches, though some codes only support dual topology.
+At present, the most widely used approaches, such as in Schr\"{o}dinger's FEP+~\cite{wang2015accurate,wang2019protein} and in BioSimSpace~\cite{hedges2019biosimspace} (for which calculations may be planned with Lead Optimization Mapper (LOMAP)~\cite{liu2013lead} seem to use single topology approaches, though some codes only support dual topology.
 To our knowledge efficiency differences have not been thoroughly explored, though conventional wisdom suggests that fewer dummy atoms are better~\cite{liu2013lead,mobley2012perspective}.
@@
 Once a particular approach to the topology is selected, a crucial next step is to identify the common atoms which will not be perturbed.
-Rigorously, this process essentially comprises a maximal common substructure (MCSS) search of the molecules involved to identify the common substructure -- though the parameters of the MCSS search will differ depending on whether single or dual topology calculations are planned.
+Rigorously, this process comprises of a maximal common substructure (MCSS) search of the molecules involved to identify the common substructure -- though the parameters of the MCSS search will differ depending on whether single or dual topology calculations are planned.
 Specifically, with a single topology approach in mind, atom types are allowed to change, so a permissive MCSS search can be done, whereas with dual topology a more strict search is required.
@@
     \includegraphics[width=0.95\linewidth]{figures/fig4_mcs/Figure.pdf}
-    \caption{\textbf{Illustration of maximum common substructure matches} MCSS is shown in green for when A: a restrictive MCSS match is used and in B: ring breaking is allowed.}
+    \caption{\textbf{Illustration of maximum common substructure matches} MCSS is shown in green for when (\textbf{A}) a restrictive MCSS match is used and in (\textbf{B}) ring breaking is allowed.}
     \label{fig:fig_mcss}
@@
 There are different tools that allow the generation of MCSS matches as well as single topology input.
-A large number of software can compute MCSS matches, some rely on RDKit, such as LOMAP or FESetup and partially BioSimSpace, others such as fkckombu are standalone tools~\cite{rdkit2019Dec, loeffler2015fesetup, liu2013lead, hedges2019biosimspace, kawabata20143d}.
+A large number of software can compute MCSS matches, some rely on RDKit~\cite{rdkit2019Dec}, such as LOMAP~\cite{liu2013lead} or FESetup~\cite{loeffler2015fesetup} and partially BioSimSpace~\cite{hedges2019biosimspace}, others such as fkckombu~\cite{kawabata20143d} are standalone tools.
 Schr\"{o}dinger's FEP+ planning tool is based on a version of LOMAP, and it also uses MCSS matching as well as 3D considerations to plan the network of single topology calculations between molecules~\cite{wang2015accurate}.
 %
-MCSS searches can be relatively time consuming, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial assessment and then use MCSS only to identify final mappings for relative calculations.
+MCSS searches can be relatively time consuming, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial assessment and then use MCSS only to identify final mappings for relative calculations~\cite{raymond2002maximum,klabunde2012mars,jones2009elucidating}.
 %
@@
 %
-Single topology relative calculations, and calculations based on substructure searches, only work if in fact the ligands share a common substructure, e.g. are part of a congeneric series.
+Single topology relative calculations, and calculations based on substructure searches, only work if in fact the ligands share a common substructure, e.g.
+are part of a congeneric series, see Fig.~\ref{fig:fig_mcss}
 If no common substructure is shared, then alternative dual or hybrid topology free energy calculations are needed, where one would co-localize a pair of compounds in a binding site, exclude their interactions with one another, and compute the relative binding free energy by turning one molecule on from being dummy atoms while turning the other off.
@@
 %
-\paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic, in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle~\cite{}, which may not be true whenever dummy atoms are involved in rings.
+\paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss} \textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings.
 Some approaches have attempted to address this~\cite{clark2019relative} but a general solution is not yet in mainstream use.
@@
 Yang et al.~\cite{yang2020optimal} optimized the perturbation map by selecting a fixed number of calculations from the pairwise perturbations so that the resulting set of calculations minimize the total variance.
-Xu~\cite{xu2019diffnet} optimized the perturbation map by allocating different amounts of simulation time to different pairwise perturbations so as to minimize the total variance, given the total simulation time of all the perturbation calculations.
+Xu~\cite{xu2019optimal} optimized the perturbation map by allocating different amounts of simulation time to different pairwise perturbations so as to minimize the total variance, given the total simulation time of all the perturbation calculations.
 Both approaches lead to substantial reduction in the statistical error of the estimated free energies.
@@
 %
-\begin{figure}
+\begin{figure}[h!]
     \includegraphics[width=0.95\linewidth]{figures/fig5_thermo_cyc_abs/Figure.pdf}
@@
 % \textbf{What are the main things I need to consider for an absolute alchemical free energy calculation?}
-% Extensive explanation for the checklist questions can be found in Section~\ref{sec:step2}.
+% Extensive explanation for the checklist questions can be found in Section~\ref{sec:simulation_protocol_choice}.
 % \begin{itemize}
@@
 % \textbf{What are the main things I need to consider for an relative alchemical free energy calculation?}
-% Extensive explanation for the checklist questions can be found in Section~\ref{sec:step2}.
+% Extensive explanation for the checklist questions can be found in Section~\ref{sec:simulation_protocol_choice}.
 % \begin{itemize}
@@
 GT: Contributed to introduction and drug discovery sections, and helped edit the paper.
-AR: Created figure~\ref{fig:freeenergytrajectories}, contributed to section \ref{sec:step2}, and helped edit the paper.
+AR: Created figure~\ref{fig:freeenergytrajectories}, contributed to section \ref{sec:simulation_protocol_choice}, and helped edit the paper.
 MRS: Helped create figure~\ref{fig:fig_what_is_lambda}, wrote sections describing choices for alchemical pathways and analysis for free energy calculations.

### 2020-01-17 paper/manuscript.tex
minor change

@@
 \subsection{Absolute and relative free energy calculations have some differences}
-Many of the issues around simualtion setup and protocol choice for alchemical calculations are common, but there are some differences between absolute and relative calculations.
+Many of the issues around simulation setup and protocol choice for alchemical calculations are common, but there are some differences between absolute and relative calculations.
 First we will consider protocol differences  so before treating the common elements.

### 2020-01-17 paper/manuscript.tex
Fixes in perturbation map section

@@
     \includegraphics[width=0.95\columnwidth]{figures/fig6_types_of_networks/Figure.pdf}
-    \caption{\textbf{Examples of perturbation networks} A: Star shaped network with the crystal structure in the center.
-B: Network with cycle closures.
+    \caption{\textbf{Examples of perturbation networks} \textbf{A} Star shaped network with the crystal structure in the center.
+\textbf{B} Network with cycle closures (see more on this in Sec.~\ref{sec:are-they-good}).
 Arrows indicate the direction of the perturbation.
@@
 \end{figure} 
-In some cases, during the lead optimization stage, or if perturbations would be rather large a star shaped network as seen in Fig.~\ref{fig:fig_types_of_networks} (A) is used..
-However, adding redundancy into the network means that a better error analysis can be carried out, by looking at cycle closure errors as discussed in sec.~\ref{subsec:predictions}, with an example given in Fig.~\ref{fig:fig_types_of_networks}.
-
+In some cases, during the lead optimization stage, or if perturbations would be rather large a star shaped network as seen in Fig.~\ref{fig:fig_types_of_networks} (\textbf{A}) is used.
+However, adding redundancy into the network means that a better error analysis can be carried out, by looking at cycle closure errors as discussed in sec.~\ref{sec:are-they-good}, with an example given in Fig.~\ref{fig:fig_types_of_networks} (\textbf{B}).
+%
 Methods in experimental design have been applied to the construction of the perturbation maps.
@@
 %
-\paragraph{Constraints and relative free energy calculations.}
+\paragraph{Constraints and relative free energy calculations}
 One issue which requires particular care is the use of constraints.
-Commonly, bonds involving hydrogen are constrained to a fixed length using algorithms such as SHAKE or LINCS, allowing the use of longer timesteps~\cite{}.
+Commonly, bonds involving hydrogen are constrained to a fixed length using algorithms such as SHAKE or LINCS, allowing the use of longer timesteps~\cite{krautler2001fast}.
 However, in single topology relative free energy calculations, the atoms involved might be mutated to other atom types -- for example, in a mutation of methane to methanol, one hydrogen might become an oxygen atom.
 Typical molecular dynamics engines are not set up to recognize this change, or at least not to correctly include contributions to the free energy from changing constraints/constraint length, so results for a transformation would usually be erroneous.
-At present the most general solution to this problem is simply to avoid the use of constraints (and thus use a smaller timestep if necessary, usually of around 1 fs) in any relative free energy calculation involving a transformation of a constrained bond.
+At present the most general solution to this problem is simply to avoid the use of constraints (and thus use a smaller timestep if necessary, usually of around 1 fs) in any relative free energy calculation involving a transformation of a constrained bond, as done by GROMACS.
 %
@@
     \includegraphics[width=0.95\linewidth]{figures/fig5_thermo_cyc_abs/Figure.pdf}
-    \caption{\textbf{Absolute thermodynamic cycle}}
+    \caption{\textbf{Thermodynamic cycle required for an aboslute free energy calculation}}
     \label{fig:fig_absolute_thermodynamic_cycle}

### 2020-01-17 paper/manuscript.tex
Updates from Overleaf

@@
     \includegraphics[width=0.95\linewidth]{figures/fig5_thermo_cyc_abs/Figure.pdf}
-    \caption{\textbf{Thermodynamic cycle required for an aboslute free energy calculation}}
+    \caption{\textbf{Thermodynamic cycle required for an aboslute free energy calculation.} The fully interacting ligand in water (\textbf{A}), has its charges turned off to pass to \textbf{B} followed by turning of van der Waals terms, resulting in a non-interacting ligand in water in \textbf{C}.
+Restraints are used on the fully interacting ligand in the binding site of a protein or host molecule \textbf{D}.
+The next step is to turn off the charges again (\textbf{E}) followed by the van der Waals interactions resulting in a non-interacting complex state (\textbf{F}).
+Free energyes can be computed as $\Delta G_{bind} = (\Delta G^{\mathrm{elec}}_{\mathrm{solv}}+ \Delta G^{\mathrm{VdW}}_{\mathrm{solv}})-(\Delta G^{\mathrm{elec}}_{\mathrm{bound}}+ \Delta G^{\mathrm{VdW}}_{\mathrm{bound}}$)
+    }
     \label{fig:fig_absolute_thermodynamic_cycle}

### 2020-01-17 paper/manuscript.tex
Section 7 reviewed.

--- 2020-01-17 paper/manuscript.tex
@@
     \includegraphics[width=0.95\linewidth]{figures/fig5_thermo_cyc_abs/Figure.pdf}
-    \caption{\textbf{Thermodynamic cycle required for an aboslute free energy calculation}}
+    \caption{\textbf{Thermodynamic cycle required for an aboslute free energy calculation -- absolute free energy of binding example} The fully interacting ligand in water (\textbf{A}), has its charges turned off to pass to \textbf{B} followed by turning of van der Waals terms, resulting in a non-interacting ligand in water in \textbf{C}.
+Restraints are used on the fully interacting ligand in the binding site of a protein or host molecule \textbf{D}.
+The next step is to turn off the charges again (\textbf{E}) followed by the van der Waals interactions resulting in a non-interacting complex state (\textbf{F}).
+Free energyes can be computed as $\Delta G_{bind} = (\Delta G^{\mathrm{elec}}_{\mathrm{solv}}+ \Delta G^{\mathrm{VdW}}_{\mathrm{solv}})-(\Delta G^{\mathrm{elec}}_{\mathrm{bound}}+ \Delta G^{\mathrm{VdW}}_{\mathrm{bound}}$).
+    }
     \label{fig:fig_absolute_thermodynamic_cycle}
@@
 For solvation free energy calculations, handling the standard state is typically straightforward, and treating it correctly simply means ensuring that the non-interacting solute is taken to the same (or equivalent) final reference state in both environments, e.g.
-that the transformation involves a 1M to 1M equivalent transfer free energy (where the non-interacting solute still occupies essentially the same volume as the solute in the interacting system).
+that the transformation involves a 1 M to 1 M equivalent transfer free energy (where the non-interacting solute still occupies essentially the same volume as the solute in the interacting system).
 So typically in such cases no special care is required to ensure the correct standard state, as long as the \emph{experimental} data being analyzed uses the same standard state and if it does not, a simple entropic correction is needed.
@@
 Experimental absolute binding free energies are reported relative to a specific reference state -- a 1 M standard state -- which must also be used in calculations.
-In practice, this has implications for how the calculations are done, as the reference concentration must enter the thermodynamic cycle employed.
+In practice, this has implications for how the calculations are done, as the reference concentration must enter the thermodynamic cycle employed, see an example of such a thermodynamic cycle in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}.
 %
@@
 \paragraph{Several choices of restraints are possible.}
-\todo[inline, color={blue!20}]{@author: Check references}
-In practice, a variety of types of restraints are common, from simple harmonic distance restraints between the ligand and the protein~\cite{mobley2006use}, to flat-bottom restraints which work similarly but only exert a force if the ligand leaves a specific region~\cite{}.
-%
-Alternatively, a set of restraints proposed by Boresch have also commonly been employed, where all six rigid-body degrees of freedom governing the orientation of the ligand relative to the receptor are restrained~\cite{leitgeb2005alchemical}.
+In practice, a variety of types of restraints are common, from simple harmonic distance restraints between the ligand and the protein~\cite{mobley2006use}, to flat-bottom restraints which work similarly but only exert a force if the ligand leaves a specific region~\cite{chen2007can}.
+%
+Alternatively, a set of restraints proposed by Boresch have also commonly been employed, where all six rigid-body degrees of freedom governing the orientation of the ligand relative to the receptor are restrained~\cite{boresch2003absolutea, leitgeb2005alchemical}.
 Further restraints, such as on the overall ligand RMSD have also been used~\cite{woo2005calculation}.
@@
 In principle, all of these forms will yield correct binding free energies in the limit of adequate sampling (if their effects and connection to the standard state are correctly handled) but they have different strengths and weaknesses.
-For example, with more involved restraints, sampling at intermediate lambda values will usually not need to be as extensive but more computational effort must go to computing the restraining free energy.
-Additionally, such restraints would typically keep the ligand from exploring alternative binding modes, which may be undesirable with Hamiltonian lambda exchange or expanded ensemble techniques where allowing the ligand to exchange binding modes when it is non-interacting could provide sampling benefits~\cite{wang2013identifying}.
-Concretely, flat bottom restraints might allow a ligand to explore multiple binding sites, harmonic restraints multiple binding modes within a site, and Boresch restraints a single binding mode within a single site.
-See additional discussion of the possibility of multiple binding modes below~\ref{sec:multiple_binding_modes}.
+For example, with more involved restraints, sampling at intermediate $\lambda$ values will usually not need to be as extensive but more computational effort must go to computing the restraining free energy.
+Additionally, such restraints would typically keep the ligand from exploring alternative binding modes, which may be undesirable with Hamiltonian $\lambda$ exchange or expanded ensemble techniques where allowing the ligand to exchange binding modes when it is non-interacting could provide sampling benefits~\cite{wang2013identifying}.
+Concretely, flat-bottom restraints might allow a ligand to explore multiple binding sites, harmonic restraints multiple binding modes within a site, and Boresch restraints a single binding mode within a single site.
+See additional discussion of the possibility of multiple binding modes in Sec.~\ref{sec:multiple_binding_modes} below.
 %
@@
 \todo[inline, color={blue!20}]{DLM: Clarify terminology: Double decoupling, etc. See Feature Box below.}
+%
 %
@@
 In practice, even with restraints, it is not always trivial to force a molecular dynamics simulation to explore a restricted region of the configurational space with a complicated geometry such as a binding site.
-If the ligand is a tight binder\todo[inline, color={red!40}]{AR: Should we give an order of magnitude to define a tight and a weak binder?}, it is usually safe to employ a restraint that allows some of the unbound configurations to be sampled since they generally contribute negligibly to the partition function (i.e., they have a relatively small Boltzmann weight) as long as the sampled volume is not so large that their cumulative contribution becomes significant.
+If the ligand is a tight binder\todo[inline, color={red!40}]{AR: Should we give an order of magnitude to define a tight and a weak binder?
+ASJSM: Yes!}, it is usually safe to employ a restraint that allows some of the unbound configurations to be sampled since they generally contribute negligibly to the partition function (i.e., they have a relatively small Boltzmann weight) as long as the sampled volume is not so large that their cumulative contribution becomes significant.
 However, this is more problematic when dealing with weak binders as the unbound configurations can have a non-negligible Boltzmann weight, and their binding affinity can exhibit a significant dependency on the restraint type and parameters that are used to determine the sampled volume~\cite{}.
@@
 %
-Many relative free energy planning tools have been set up to avoid changing the net charge of the systems considered, including LOMAP~\cite{} and Schr\"{o}dinger's FEP+~\cite{chen2018accurate}.
+Many relative free energy planning tools have been set up to avoid changing the net charge of the systems considered, including LOMAP~\cite{liu2013lead} and Schr\"{o}dinger's FEP+~\cite{wang2015accurate}.
 Absolute free energy calculations can also potentially avoid changing the charge of the system by making a charge perturbation of equal and opposite sign elsewhere in the system; for example, as a charged ligand is removed, a charged counterion of opposite sign could also be removed, or one of the same sign could be inserted.
 This is sometimes referred to as an "alchemical ion" approach for dealing with the needed charge change, and is also employed by the Yank free energy package~\cite{wang2013identifying}.
-Charge corrections have also been explored, and are potentially a viable solution to this problem~\cite{} where artifacts introduced by finite-size effects are corrected numerically.
+Charge corrections have also been explored, and are potentially a viable solution to this problem~\cite{mey2018impact} where artifacts introduced by finite-size effects are corrected numerically.
 ~\cite{chen2018accurate}. However, application of such corrections typically remains less common than the use of a co-alchemical ion.
@@
 In principle, because of the path independence of the free energy, any arbitrary pathway will give the correct free energy change, but the choice of pathway will greatly affect the efficiency of the calculations.
-Some choices are particularly crucial -- for example, transformations involving insertions or deletions of atoms should employ a path soft-core potential path for Lennard-Jones or other interactions with repulsive interactions that go to infinite energy at small radius~\cite{}.
+Some choices are particularly crucial -- for example, transformations involving insertions or deletions of atoms should employ a soft-core potential path for Lennard-Jones or other interactions with repulsive interactions that go to infinite energy at small radius~\cite{beutler1994avoiding, beutler1994molecular,gapsys2012new}.
 %
@@
 This can be proven directly from the BAR and MBAR formulas~\cite{bennett1976efficient,klimovich2015guidelines}, though the exact same principles apply for TI.
-%
-The simplest possible alchemical pathway is a \textit{linear} pathway:
+For a 'good' path to work sufficient overlap between pontential energies is required.
+Fig.~\ref{fig:fig_what_is_lambda} \textbf{A} and \textb{B} try to illustrate this.
+Fig.~\ref{fig:fig_what_is_lambda} \textbf{A} shows in a pictorial way a soft-core potential can be applied across different $\lamda$s.
+Fig.~\ref{fig:fig_what_is_lambda} \textbf{B} illustrates the potential energy distributions at the different $\lambda$ intermediates, with sufficient overlap between neighboring $\lambda$ states to ensure that reweighting estimators such as MBAR can be used for analysis (see Sec.\ref{subsec:estimators}).
+The actual transformation is best handled with soft-core potentials of the form shown in Fig.\ref{fig:fig_what_is_lambda} \textbf{C} and \textbf{B}, with more details given below.
+%
+So how do we adjust the potentials between the two end states based on $\lambda$? The simplest possible alchemical pathway is a \textit{linear} pathway:
 %
@@
 However, this approach ends up being terrible for removing or adding repulsive potentials that go to infinity quickly at or near the origin.
-One way to look at this is to examine how low $\lambda$ must go to reduce the energy at $0.5\sigma$ (the atomic size parameter) down to 1 $k_BT$, where thermal fluctuations make it possible for other atomic sites to penetrate routinely that deep.
+One way to look at this is to examine how low $\lambda$ values must go to reduce the energy at $0.5\sigma$ (the atomic size parameter) down to 1 $k_BT$, where thermal fluctuations make it possible for other atomic sites to penetrate routinely that deep.
 If we are trying go from a particle being present, and desire to make it disappear alchemically, then if the repulsive terms is of the form $\epsilon\frac{\sigma}{r}^{12}$, then if $\epsilon$ was 1 $K_BT$ at the temperature of interest, and we start with the particle present, then solving for $(1-\lambda)(1 kB_T)\frac{\sigma}{\frac{1}{2}\sigma}) = 1 k_B T$ we get $\lambda = 1-2^{-12} \sim 0.999976$.
@@
 We can play around with a few ways of modifying this, like simulating many more intermediate states near $\lambda=1$.
-Various analyses have shown that this is not a very good strategy, however~\cite{pham2011identifying, beutler1994avoiding, zacharias1994separationshifted, blondel2004ensemble}
+Various analyses have shown that this is not a very good strategy, however~\cite{pham2011identifying, beutler1994avoiding, zacharias1994separationshifted, blondel2004ensemble, gapsys2012new}.
 %
@@
 There are several different variants of the same functional form~\cite{zacharias1994separationshifted, beutler1994avoiding,pham2011identifying}, but the one in given in eq.~\ref{eq:softcore} is easy to understand and implement and fairly numerically stable.
-This functional form is shown in parts C and D of figure~\ref{fig:fig_what_is_lambda}.
+This functional form is shown in parts \textbf{C} and \textbf{D} of Fig.~\ref{fig:fig_what_is_lambda}.
 It has been shown that more complicated forms are not significantly more efficient than eq.~\cite{pham2012optimal}.
@@
 We can't turn off the electrostatics linearly at the same time we turn off the Lennard-Jones terms, as it would leave infinitely large attractive and repulsive electrostatic terms "bare" at small $\lambda$, resulting in the simulation crashing.
-It \textit{is} possible to apply the same soft core approach to the Coulomb interaction, but it turns out to be somewhat difficult to turn off two soft-core terms  terms simultaneously~\cite{Steinbrecher:JCC:2011}.
+It \textit{is} possible to apply the same soft core approach to the Coulomb interaction, but it turns out to be somewhat difficult to turn off two soft-core terms simultaneously~\cite{steinbrecher2011softcore}.
 We therefore recommend performing the transformations in sequence; first, turning off all electrostatics for atoms that must be removed, inserting and removing Lennard-Jones sites (both the insertion and removal can be done simultaneously), and then turning electrostatics for the introduced particles on.
@@
 %
-Other issues, such as whether absolute calculations should retain or remove intramolecular nonbonded interactions, may be less critical and recommendations differ among studies in the literature~\cite{}, and reasonable efficiency can be often obtained with either choice even if some are somewhat better or worse than others.
+Other issues, such as whether absolute calculations should retain or remove intramolecular non-bonded interactions, may be less critical and recommendations differ among studies in the literature~\cite{}, and reasonable efficiency can be often obtained with either choice even if some are somewhat better or worse than others.
 Our recommendation is to leave the intramolecular interactions on during the transformation for simplicity if there are no other known issues with this approach.
@@
 If a configurational phase transition is noted in the process (like substantial dihedral rearrangement), then other steps must be taken, perhaps including modifying intramolecular interactions~\cite{???}.
-These problems can be diagnosed by noticing lack of configuration space overlap between different simulations (see section~\ref{subsec:})"
+These problems can be diagnosed by noticing lack of configuration space overlap between different simulations (see section~\ref{sec:are-they-good})"
 %
 Relative calculations introduce additional choices, including whether to define explicit intermediate states~\cite{} or leave these implicitly defined by the code~\cite{}.
-Typically in single topology relative calculations it proves most efficient to first remove electrostatic interactions of any atoms which will be deleted, then modify other nonbonded interactions, then restore electrostatic interactions of any atoms which are being inserted.
-Other schemes, such as simultaneously changing electrostatic and Lennard-Jones interactions, even with electrostatic ``soft core'' potentials~\cite{steinbrecher2007nonlinear}, in our experience typically introduce errors and/or instabilities or are at least unreliable.
-It is expected that similar considerations will apply to dual topology calculations, and the principle of first removing electrostatics and then removing steric interactions will likely serve well.
+Typically in single topology relative calculations it proves most efficient to first remove electrostatic interactions of any atoms which will be deleted, then modify other non-bonded interactions, then restore electrostatic interactions of any atoms which are being inserted.
+Other schemes, such as simultaneously changing electrostatic and Lennard-Jones interactions, even with electrostatic ``soft-core'' potentials~\cite{steinbrecher2007nonlinear}, in our experience typically introduce errors and/or instabilities or are at least unreliable.
+It is expected that similar considerations will apply to dualtopology calculations, and the principle of first removing electrostatics and then removing steric interactions will likely serve well.
 %
 A key additional consideration in choosing the alchemical pathway is the choice of spacing of intermediate states.
-The spacing depends to some extent on the choice of analysis method, though states should essentially be spaced equidistant in the relevant thermodynamic length~\cite{}.
+The spacing depends to some extent on the choice of analysis method, though states should essentially be spaced equidistant in the relevant thermodynamic length~\cite{crooks2007measuringa, sivak2012thermodynamic}.
 For BAR/MBAR techniques this means that spacings should typically be equal in [what, variance? ref].
@@
 %
-\subsubsection{Multiple or uncertain binding modes may require considerable care}
-\label{sec:multiple_binding_modes}
-In a discovery setting, new ligands typically have unknown or at least uncertain binding modes~\cite{kaus2015how, plountprice2000analysis,mobley2009binding,calabro2016elucidation},  complicating binding free energy estimation.
-This uncertainty is because it is usually not desirable to estimate a binding affinity for a ligand which already has an available bound structure, since such a compound has already been tested.
-To deal with prospective ligands with unknown binding modes, discovery projects commonly assume that modifications of functional groups on a common scaffold result in a consistent binding mode across all members of a series.
-This is not necessarily always the case~\cite{kaus2015how}, as reviewed elsewhere~cite{mobley2009binding} and in some cases unexpected binding mode changes can be the origin of apparent non-additivity in structure-activity relationships~\cite{calabro2016elucidation}.
-Binding modes also tend to be particularly variable in the case of fragments, which often may have multiple relevant binding modes~\cite{}.
-
-Absolute free energy calculations for dissimilar ligands can have particular challenges because the (potentially incorrect) assumption of consistent binding modes across a series of similar ligands is likely to be even less robust than the in the case of relatively calculations.
-This means that researchers performing absolute binding free energy calculations will have to pay particular attention to generating reasonable putative binding modes.
-
-In some cases, it is tempting to simply use docking techniques to generate initial bound structures for starting molecular dynamics simulations.
-However, timescales for binding mode interconversion are usually slow compared to MD/free energy timescales, meaning that simulations started from different potential binding modes are likely to yield disparate computed binding free energies~\cite{mobley2006use, palma2012computation, mobley2012perspective, gill2018binding} .
-And docking techniques are good at identifying sterically reasonable potential binding modes, but still perform relatively poorly at identifying a single dominant binding mode \emph{a priori}.~\cite{}
-
-It is worth highlighting a recent SAMPL blind challenge on HIV integrase as an illustration of this. 
-Many submissions, using state-of-the-art methods, had difficulty even predicting which \emph{binding site} ligands would bind in (most submissions placed more than half of the ligands into the incorrect binding site), and even given correct binding sites, the binding mode within each site was also quite difficult to predict~\cite{mobley2014blind}.
-The best performing submission for predicting binding modes actually ended up being a human expert (aided by computational tools) with more than 10 years of experience on the particular target~\cite{voet2014combining}, rather than a fully automated approach.
-While free energy calculations on this set had some success, many of the failures actually ended up being cases where the binding mode selected as input for free energy calculations was later found to be incorrect~\cite{gallicchio2014virtual}, highlighting the importance of these issues.
-
-One approach which has shown some success is to retain diverse potential binding modes from docking, perform short MD simulations of these to identify distinct stable binding modes, and then consider these in subsequent calculations~\cite{gallicchio2014virtual, mobley2006use,rocklin2013blind, boyce2009predicting, mobley2007predicting}.
-
-Routes to handle multiple potential binding modes are different depending on whether absolute or relative calculations are selected, unless a method is available to estimate the relative populations of different stable binding modes in advance (e.g.
-such as the BLUES approach currently in development~\cite{gill2018binding}), in which case this approach could be applied to assist both types of calculations.
-
-\paragraph{Handling multiple potential binding modes within absolute calculations.}
-Within absolute binding free energy calculations, multiple potential binding modes can be handled by two main strategies: Consider each binding mode separately (a separation of states strategy) or sample all binding modes within a single simulation~\cite{mobley2012perspective}.
-This couples to the choice of restraints selected, as some restraints will allow transitions between binding modes and even binding sites (Section~\ref{sec:standardstate-restraints}), and others do not.
-
-Sampling all binding modes within a single free energy calculation is usually impractical without some form of enhanced sampling or at least Hamiltonian replica exchange~\cite{wang2013identifying} because barriers for binding mode interconversion result in kinetics which are too slow compared to simulation timescales~\cite{mobley2006use, palma2012computation,mobley2012perspective, gill2018binding}.
-Hamiltonian exchange, coupled with appropriate restraints, can allow the ligand to relatively rapidly exchange between potential binding modes when non-interacting, accelerating sampling of binding modes~\cite{wang2013identifying}.
-\todo[inline, color={yellow!40}]{DLM probably need more background refs on Hamiltonian lambda exchange here.}
-
-Separation of states provides a simple though potentially expensive alternative, where each stable binding mode is considered separately with a binding free energy calculation restricted to that binding mode, and then (as long as the binding modes are non-overlapping) the resulting component binding free energies can be combined into a total~\cite{mobley2006use, mobley2012perspective}.
-This approach necessitates a separate binding free energy calculation for each potential binding mode, however, so it can be computationally quite costly.
-If relative populations of different stable binding modes were available from some other technique, it could make this separation of states approach considerably more efficient~\cite{mobley2012perspective, gill2018binding}.
-
-\paragraph{Handling multiple potential binding modes within relative calculations.}
-
-Multiple potential binding modes pose particular problems for relative free energy calculations, as having multiple starting structures for these calculations could yield substantially different calculated relative binding free energies for the same transformation due to kinetic trapping, and, without additional information (specifically, the free energy of binding mode interconversion or, equivalently, the relative populations of different binding modes) it becomes impossible to sort out which of the multiple answers is in fact the correct relative binding free energy~\cite{}.
-
-To deal with this, some practitioners have actually computed relative binding free energies of different binding modes of the same ligand~\cite{palma2012computation} [ref Jorgensen (?) etc.].
-For example, a mutation which adds a methyl to an aromatic ring of a larger ligand might yield one result if the methyl points in one direction, and a different value if it points in the other due to slow ring motions.
-[e.g.
-get ref from Sukanya]
-One could compute the free energy of turning off the methyl group in one orientation and turning it back on in the other orientation to obtain the free energy difference between the two potential binding modes.
-While this approach has precedent, it is relatively difficult to automate at present and requires considerable care.
-
-Overall, this likely means that relative free energy calculations will be susceptible to problems resulting from uncertainty in ligand binding modes until more robust approaches are available to determine dominant binding modes, or the relative populations of different potential binding modes, in advance.
-
 \subsubsection{What different sampling schemes exist and which one should I use?}
-With all alchemical simulations having the requirement of sampling from different sets of $\lambda$ states in common different approaches have been designed.
+\label{sec:sampling_schemes}
+With all alchemical simulations having the requirement of sampling from different sets of $\lambda$ states in common, different approaches have been designed to achieve this.
 Fig.~\ref{fig:fig_sampling_scheme} illustrates the four most common schemes used.
-The simplest approach is defining a set of fixed $\lambda$ values and then run independent simulations at each of the intermediate $\lambda$ values, see Fig.~\ref{fig:fig_sampling_scheme}(A).
-This type of scheme is currently used for AMBER TI calculations~\cite{song2019} and for BioSimSpace~\cite{hedges2019biosimspace}.
-A slightly more complex approach is allowing replica mixing, meaning that each independent $\lambda$ simulation is attempted to be exchanged with a neighbouring $\lambda$ replica.
-This is based on ideas of Monte Carlo simulations of spin glasses by Swendsen and Wang~\cite{wendsen1986replica}.
+The simplest approach is defining a set of fixed replicas each representing a different $\lambda$ value that run independent simulations at each of the intermediate $\lambda$ values, see Fig.~\ref{fig:fig_sampling_scheme}(\textbf{A}).
+This type of scheme is currently used for AMBER TI calculations~\cite{song2019} and for Siremol/Openmm~\cite{}.
+A slightly more complex approach is allowing replica mixing, meaning that each independent $\lambda$ simulation is attempted to be exchanged with a neighboring $\lambda$ replica.
+This is based on ideas of Monte Carlo simulations of spin glasses by Swendsen and Wang~\cite{swendsen1986replica}.
 Using a Metropolis Hastings based acceptance criterion for exchanges means that the generated ensemble of all replicas is still sampling from the Boltzmann distribution and has been used in many different contexts for molecular simulations~\cite{sugita2000multidimensionala,sugita1999replicaexchangea, woods2003developmenta, jiang2010free}.
-The basic idea of the replica exchange scheme is shown in Fig.~\ref{fig:fig_sampling_scheme}(B).
+The basic idea of the replica exchange scheme is shown in Fig.~\ref{fig:fig_sampling_scheme}(\textbf{B}).
 It is supported in various software packages that provide alchemical implementations, such as GROMACS~\cite{aldeghi2015accurate}, FEP+~\cite{wang2015accurate}, and NAMD~\cite{jiang2019computing}.
 A third approach borrows ideas from simulated tempering~\cite{marinari1992simulateda}.
-In this scheme a single replica rapidly explores all of $\lambda$ space by optimally working out weights that allow switching between different intermediate $\lambda$ values, as seen in Fig.~\ref{fig:fig_sampling_scheme}(C) and also referred to as self adjusted mixture sampling~\cite{lyubartsev1992newa, li2007simulated, tan2017optimally} and while promising, has so far only been supported in~\cite{andrearizzi2019choderalab}.
+In this scheme a single replica rapidly explores all of $\lambda$ space by optimally working out weights that allow switching between different intermediate $\lambda$ values, as seen in Fig.~\ref{fig:fig_sampling_scheme}(\textbf{C}) and also referred to as self adjusted mixture sampling~\cite{lyubartsev1992newa, li2007simulated, tan2017optimally} and while promising, has so far only been supported in OpenMM Tools~\cite{andrearizzi2019choderalab}.
 The last approach available in conjunction with GROMACS makes use of non-equilibrium simulations~\cite{aldeghi2018accurate}.
 In this approach, only end state $\lambda$ replicas ($\lambda$=0, $\lambda=1$) are simulated at equilibrium intermediate information is generated from non-equilibrium simulations.
-A schematic of this approach is shown in Fig.~\ref{fig:fig_sampling_scheme}.
-\ref{fig:fig_sampling_scheme} A:\cite{hedges2019biosimspace, song2019using } B\cite{aldeghi2015accurate, wang2015accurate,jiang2019computing} C:\cite{li2007simulated, tan2017optimally} D:\cite{aldeghi2018accurate}
+A schematic of this approach is shown in Fig.~\ref{fig:fig_sampling_scheme} \textbf{D}.
+
 \begin{figure}
     \includegraphics[width=0.95\columnwidth]{figures/fig8_sampl_scheme/Figure.pdf}
-    \caption{\textbf{Four sampling strategies available in different software packages.} \textbf{A}: Multiple replicas in parallel at different lambda states.
+    \caption{\textbf{Four most common sampling strategies are available in different popular software packages.} \textbf{A}: Multiple replicas in parallel at different lambda states.
 \textbf{B}: Hamiltonian replica exchange scheme.
@@
 \end{figure} 
-The best practice recommendation, is to use replica exchange type sampling schemes (Fig.~\ref{fig:fig_sampling_scheme}(B)), or in absence of their availability and when looking at systems with fast mixing times parallel replica schemes are acceptable (Fig.~\ref{fig:fig_sampling_scheme}(A)).
+The best practice recommendation, is to use replica exchange type sampling schemes (Fig.~\ref{fig:fig_sampling_scheme}(\textbf{B})), or in absence of their availability and when looking at systems with fast mixing times parallel replica schemes are acceptable (Fig.~\ref{fig:fig_sampling_scheme}(\textbf{A})).
 Single replica schemes and non-equilibrium schemes are not as established yet, but are very promising.
@@
 \subsubsection{How long should I run my simulation for and what information should be saved?}
+\label{sec:sim_length_information_kept}
 Before launching alchemical free energy calculations it is wise to consider how convergence and completion will be assessed.
@@
 However, this is not the only metric which should be used, as the uncertainty only captures the information about the sampled phase space, not necessarily the entirety of the phase space.  
-For example, 
-convergence of relative free energy calculations in predictive simulations where the entire phase space is not known in advance, requires sampling the different kinetically stable states~\cite{mobley2012perspective}.
-This highlights the importance of choosing the correct thermodynamic path to ensure you sample the required thermodynamic states as discussed in Section~\ref{sec:important_path}.
+For example, convergence of relative free energy calculations in predictive simulations where the entire phase space is not known in advance, requires sampling the different kinetically stable states~\cite{mobley2012perspective}.
+This highlights the importance of choosing the correct thermodynamic path to ensure you sample the required thermodynamic states as discussed in Sec.~\ref{sec:important_path}.
 %
 The condition of minimizing the statistical uncertainty of different free energy estimators below a sufficient threshold should be one metric monitored over the simulation.
-This can be done through the uncertainty estimator built into certain analysis tools such as MBAR, or can be done though more general statistical tools like Bootstrap Sampling.
+This can be done through the uncertainty estimator built into certain analysis tools such as MBAR, or can be done though more general statistical tools like bootstrap sampling.
 A target statistical uncertainty should be chosen at the onset of the simulation to avoid excessively long simulations, or falling into the trap of running until the free energy estimate is "good enough," which is subjective and has no defined criteria.
@@
 If more samples are needed, the simulations can be resumed, or, started again in different initial conditions.
-
+%
 Convergence in other alchemical observables should also be monitored to determine if the defined phase space has been sufficiently sampled and enough decorrelated samples have been drawn.
-These additional observables include, but are not
-limited to, the variance in $\frac{dU}{d\lambda}$ across all $\lambda$ values, calculating the variance in free energy using bootstrap analysis, and comparing differences in free energies calculated using different percentages of the simulation in both the forward and reverse directions (figure \ref{fig:convergence_forward_reverse}).
-
-\begin{figure}
+These additional observables include, but are not limited to, the variance in $\frac{dU}{d\lambda}$ across all $\lambda$ values, calculating the variance in free energy using bootstrap analysis, and comparing differences in free energies calculated using different percentages of the simulation in both the forward and reverse directions ( see Fig.
+\ref{fig:convergence_forward_reverse}).
+%
+\begin{figure}[h!]
     \centering
@@
 \end{figure}
-
+%
 Each of these metrics have demonstrated promising results for diagnosing when a simulation has a convergence issue beyond simple convergence of uncertainty estimate. 
@@
 The confidence a user should have in a free energy estimate is significantly improved when both the uncertainty of the free energy estimate is low, and when other observables have reached a convergence.
-
-A simple, but effective, way of checking for bias is to compute the same property by multiple estimators. 
+%
+Another aspect that can be monitored is bias in the simulation.
+A simple, but effective, way of checking for bias is to compute the same property of interest by multiple estimators (e.g.
+free energy difference).
 The uncertainty in the free energy, for example, has multiple ways to be estimated, e.g.
@@
 As such, it is a good idea to subject different estimators to the same data to see if they yield either the same estimate (within error and bias), or if they fluctuate wildly. 
-This is not a perfect method as some estimators, such as exponential averaging, will converge significantly more slowely, relative to more accurate estimators like MBAR. 
+This is not a perfect method as some estimators, such as exponential averaging, will converge significantly more slowly, relative to more accurate estimators like MBAR. 
 Therefore, it is a good idea to apply the estimators to different fractions of the data to see if the main estimator of free energy you have chosen is stable.
-
-Each method requries different data about the simulation be collected.
-If, for instance, the free energy estimator selected is thermodynamic integration, then values of $\frac{dU}{d\lambda}$ at uncorrelated data points must be collected,  Once a combination of knowing what type of simulation you will run, which alchemical topology you will simulate, what alchemical path you will simulation along, and what your stopping conditions are, then you are ready to enumerate the information you should capture.
-Below is a sample of the minimal information you need for a set of common estimators:
-
+%
+Each method requires different data from the simulation be collected.
+If, for instance, the free energy estimator selected is thermodynamic integration, then values of $\frac{dU}{d\lambda}$ at uncorrelated data points must be collected.
+Once a combination of knowing what type of simulation you will run, which alchemical topology you will simulate, what alchemical path you will simulation along, and what your stopping conditions are, then you are ready to enumerate the information you should capture.
+Below is a sample of the minimal information you need for a set of common estimators (discussed in more detail in Sec.~\ref{subsec:estimators}):
+%
 \begin{itemize}
@@
 \end{itemize}
-
-The potential derivative required for TI should generally be calculated during the simulation; only under every rare circumstances~\cite{Naden:JCTC:2015} can it postprocessed by a code that does not evaluate the derivatives.
+%
+The potential derivative required for TI should generally be calculated during the simulation; only under every rare circumstances~\cite{naden2015linear} can it postprocessed by a code that does not evaluate the derivatives.
 Many codes already have options for doing this.
 If that option is unavailable, you can estimate it through finite difference (if sufficient information is collected), but this  will introduce significant error, and is generally not a best practice.
-Tthe BAR estimator may be a better, and simpler choice at that point as you will have at least the same level of information.
+The BAR estimator may be a better, and simpler choice at that point as you will have at least the same level of information.
 The potential energy differences required for EXP, BAR, MBAR, and WHAM can be calculated either during the simulation or in post-processing.
@@
 However, because of the problems with linear paths already discussed in this paper, this simplification is rarely that useful.
-
+%
 Free energy information should generally be saved more frequently than coordinate data, approximately at the rate that uncorrelated samples are produced.  
@@
 How decorrelation impacts calculations, and how to compute it is discussed in other section~\ref{sec:decorrelating-samples}. 
+%
+\subsubsection{Multiple or uncertain binding modes may require considerable care}
+\label{sec:multiple_binding_modes}
+In a discovery setting, new ligands typically have unknown or at least uncertain binding modes~\cite{kaus2015how, plountprice2000analysis,mobley2009binding,calabro2016elucidation},  complicating binding free energy estimation.
+This uncertainty is because it is usually not desirable to estimate a binding affinity for a ligand which already has an available bound structure, since such a compound has already been tested.
+To deal with prospective ligands with unknown binding modes, discovery projects commonly assume that modifications of functional groups on a common scaffold result in a consistent binding mode across all members of a series.
+This is not necessarily always the case~\cite{kaus2015how}, as reviewed elsewhere~cite{mobley2009binding} and in some cases unexpected binding mode changes can be the origin of apparent non-additivity in structure-activity relationships~\cite{calabro2016elucidation}.
+Binding modes also tend to be particularly variable in the case of fragments, which often may have multiple relevant binding modes~\cite{}.
+%
+Absolute free energy calculations for dissimilar ligands can have particular challenges because the (potentially incorrect) assumption of consistent binding modes across a series of similar ligands is likely to be even less robust than the in the case of relatively calculations.
+This means that researchers performing absolute binding free energy calculations will have to pay particular attention to generating reasonable putative binding modes.
+%
+In some cases, it is tempting to simply use docking techniques to generate initial bound structures for starting molecular dynamics simulations.
+However, timescales for binding mode interconversion are usually slow compared to MD/free energy timescales, meaning that simulations started from different potential binding modes are likely to yield disparate computed binding free energies~\cite{mobley2006use, palma2012computation, mobley2012perspective, gill2018binding} .
+And docking techniques are good at identifying sterically reasonable potential binding modes, but still perform relatively poorly at identifying a single dominant binding mode \emph{a priori}.~\cite{}
+%
+It is worth highlighting a recent SAMPL blind challenge on HIV integrase as an illustration of this. 
+Many submissions, using state-of-the-art methods, had difficulty even predicting which \emph{binding site} ligands would bind in (most submissions placed more than half of the ligands into the incorrect binding site), and even given correct binding sites, the binding mode within each site was also quite difficult to predict~\cite{mobley2014blind}.
+The best performing submission for predicting binding modes actually ended up being a human expert (aided by computational tools) with more than 10 years of experience on the particular target~\cite{voet2014combining}, rather than a fully automated approach.
+While free energy calculations on this set had some success, many of the failures actually ended up being cases where the binding mode selected as input for free energy calculations was later found to be incorrect~\cite{gallicchio2014virtual}, highlighting the importance of these issues.
+%
+One approach which has shown some success is to retain diverse potential binding modes from docking, perform short MD simulations of these to identify distinct stable binding modes, and then consider these in subsequent calculations~\cite{gallicchio2014virtual, mobley2006use,rocklin2013blind, boyce2009predicting, mobley2007predicting}.
+%
+Routes to handle multiple potential binding modes are different depending on whether absolute or relative calculations are selected, unless a method is available to estimate the relative populations of different stable binding modes in advance (e.g.
+such as the BLUES approach currently in development~\cite{gill2018binding}), in which case this approach could be applied to assist both types of calculations.
+%
+\paragraph{Handling multiple potential binding modes within absolute calculations.}
+Within absolute binding free energy calculations, multiple potential binding modes can be handled by two main strategies: Consider each binding mode separately (a separation of states strategy) or sample all binding modes within a single simulation~\cite{mobley2012perspective}.
+This couples to the choice of restraints selected, as some restraints will allow transitions between binding modes and even binding sites (Section~\ref{sec:standardstate-restraints}), and others do not.
+%
+Sampling all binding modes within a single free energy calculation is usually impractical without some form of enhanced sampling or at least Hamiltonian replica exchange~\cite{wang2013identifying} because barriers for binding mode interconversion result in kinetics which are too slow compared to simulation timescales~\cite{mobley2006use, palma2012computation,mobley2012perspective, gill2018binding}.
+Hamiltonian exchange, coupled with appropriate restraints, can allow the ligand to relatively rapidly exchange between potential binding modes when non-interacting, accelerating sampling of binding modes~\cite{wang2013identifying}.
+\todo[inline, color={yellow!40}]{DLM probably need more background refs on Hamiltonian lambda exchange here.}
+%
+Separation of states provides a simple though potentially expensive alternative, where each stable binding mode is considered separately with a binding free energy calculation restricted to that binding mode, and then (as long as the binding modes are non-overlapping) the resulting component binding free energies can be combined into a total~\cite{mobley2006use, mobley2012perspective}.
+This approach necessitates a separate binding free energy calculation for each potential binding mode, however, so it can be computationally quite costly.
+If relative populations of different stable binding modes were available from some other technique, it could make this separation of states approach considerably more efficient~\cite{mobley2012perspective, gill2018binding}.
+%
+\paragraph{Handling multiple potential binding modes within relative calculations.}
+Multiple potential binding modes pose particular problems for relative free energy calculations, as having multiple starting structures for these calculations could yield substantially different calculated relative binding free energies for the same transformation due to kinetic trapping, and, without additional information (specifically, the free energy of binding mode interconversion or, equivalently, the relative populations of different binding modes) it becomes impossible to sort out which of the multiple answers is in fact the correct relative binding free energy~\cite{}.
+%
+To deal with this, some practitioners have actually computed relative binding free energies of different binding modes of the same ligand~\cite{palma2012computation}.
+For example, a perturbation which adds a methyl to an aromatic ring of a larger ligand might yield one result if the methyl points in one direction, and a different value if it points in the other due to slow ring motions.
+[e.g.
+get ref from Sukanya]
+One could compute the free energy of turning off the methyl group in one orientation and turning it back on in the other orientation to obtain the free energy difference between the two potential binding modes.
+While this approach has precedent, it is relatively difficult to automate at present and requires considerable care.
+%
+Overall, this likely means that relative free energy calculations will be susceptible to problems resulting from uncertainty in ligand binding modes until more robust approaches are available to determine dominant binding modes, or the relative populations of different potential binding modes, in advance.
 %

### 2020-01-19 paper/manuscript.tex
fixes to sec 1 2 and 3

@@
 \todo[inline, color={green!20}]{@everyone: Any further input in this?}
-Alchemical free energy calculations compute free energy differences associated with transfer processes, such as the binding of a small molecule to a receptor or the transfer of a small molecule from an aqueous to apolar phase~\cite{zwanzig1954hightemperature}.
+Alchemical free energy calculations compute free energy differences associated with transfer processes, such as the binding of a small molecule to a receptor, the transfer of a small molecule from an aqueous to apolar phase~\cite{zwanzig1954hightemperature}, or effects of protein side chain mutations on e.g.
+binding affinities..
 These calculations use non-physical\footnote{Here, the non-physical nature of the transformation is referred to as "alchemical", a term coined by Tembre and McCammon in Ref.~\cite{tembre1984ligandreceptor}.} intermediate states in which the chemical identity of some portion of the system (such as a small molecule ligand or receptor sidechain) is changed by modifying the potential governing the interaction of the atoms being modified, inserted, or deleted with their environment.
 %
-Fig.~\ref{fig:fig_what_is_alchemy} illustrates common free energy changes that may be difficult to compute with conventional computational methods, but are more readily tractable with alchemical methods.
-\todo[inline, color={yellow!20}]{JDC: Not sure what is meant by "conventional computational methods". Perhaps we can be precise and say "unbiased molecular dynamics simulations"?}
+Fig.~\ref{fig:fig_what_is_alchemy} illustrates common free energy changes that may be difficult to compute with unbiased molecular dynamics methods, but are more readily tractable with alchemical methods.
 In alchemical simulations, the introduction of intermediate \textit{alchemical states} that bridge the high-probability regions of configuration space between the two physical endstates of interest permits the free energy of large transformations to be robustly computed.
-%such as the entire removal of a ligand from a binding site (and \emph{absolute AFE calculation}), as illustrated in Fig.~\ref{fig:fig_what_is_alchemy} (E). 
-alchemical calculations can be used in a variety of scenarios, such as computing the free energy of a conformational change for a molecule with a high barrier to interconversion (Fig.~\ref{fig:fig_what_is_alchemy} (A));
-\todo[inline, color={yellow!20}]{JDC: Do you intend these to be representative of BLUES type simulations?}
-computing partition ($\log P$) or distribution ($\log D$) coefficients between environments (Fig.~\ref{fig:fig_what_is_alchemy} (B))~\cite{rustenburg2016measuring, bosisio2016blinded} or compartments or membranes (Fig.~\ref{fig:fig_what_is_alchemy} (C))~\cite{}.
-alchemical calculations are frequently used to estimate changes in free energies upon modifying a ligand or protein: a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy} (D))\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}; the entire ligand can be alchemical transferred from protein to solvent in an absolute binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (E))~\cite{mobley2007predicting,aldeghi2015accurate,aldeghi2017predictions}; and small alchemical modifications can be made between chemically related ligands to estimate differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (F))~\cite{wang2015accurate,mey2016blinded,gapsys2020large}.
-
+Alchemical calculations can be used in a variety of scenarios, such as computing the free energy of a conformational change for a molecule with a high barrier to interconversion (Fig.~\ref{fig:fig_what_is_alchemy} (A));
+\todo[inline, color={yellow!20}]{JDC: Do you intend these to be representative of BLUES type simulations? -- yes}
+computing partition ($\log P$) or distribution ($\log D$) coefficients between environments (Fig.~\ref{fig:fig_what_is_alchemy} (B))~\cite{rustenburg2016measuring, bosisio2016blinded} or compartments or membranes (Fig.~\ref{fig:fig_what_is_alchemy} (C))~\cite{corey2019insights}.
+Furthermore, alchemical calculations are frequently used to estimate changes in free energies upon modifying a ligand or protein: a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy} (D))\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}; the entire ligand can be alchemical transferred from protein to solvent in an absolute binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (E))~\cite{mobley2007predicting,aldeghi2015accurate,aldeghi2017predictions}; and small alchemical modifications can be made between chemically related ligands to estimate differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (F))~\cite{wang2015accurate,mey2016blinded,gapsys2020large}.
+%
 Once an alchemical simulation as been performed, the data must be analyzed to estimate the free energy for the transformation of interest.
 Early work used simple (but suboptimal) estimators for this: free energy perturbation (FEP) using the Zwanzig relation~\cite{zwanzig1954hightemperature} or thermodynamic integration (TI), for which the theory dates back decades but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamics, merz1989free}.
-More recent developments have seen new, highly efficient statistical estimators that make better use of all the data, often building on the Bennett acceptance ratio (BAR)~\cite{bennett1976efficient}, producing multistate generalizations~\cite{mbar} or removing the need for global equilibrium~\cite{tram}
+%
+More recent developments have seen new, highly efficient statistical estimators that make better use of all the data, often building on the Bennett acceptance ratio (BAR)~\cite{bennett1976efficient}, producing multistate generalizations~\cite{shirts2008statisticallya} or removing the need for global equilibrium~\cite{wu2016multiensemble, mey2014xtram, wu2014statistically}
 \todo[inline, color={yellow!20}]{JDC: The key pivotal moment was the Shirts paper on extremely precise free energy calculations, which demonstrated that a combination of technologies can produce robust estimates with useful precision.}
@@
 In the meantime, today's users may find it hard to get started with these complex calculations whilst also keeping up with the fast pace of change.
-This best practice guide aims to serve as a go-to reference for current recommendations and tips for users of all experience.
+This Best Practice guide aims to serve as a go-to reference for current recommendations and tips for users of all experience.
 %
@@
     \includegraphics[width=0.95\linewidth]{figures/fig1_what_is_alchemy/Figure.pdf}   
-    \caption{\textbf{Illustration of common types of free energies differences that can be calculated using alchemical free energy methods.} A: Change in free energy due to a conformational change of the molecule.
-B: Partition coefficient such as $\log P$ or $\log D$ can be computed using alchemical free energy methods.
-Here, as an example the partition coefficient between methanol and water is shown.
-C: Free energy difference associated with the insertion of a molecule into a membrane.
-D: Effect of mutations of protein or host residues on free energies of binding.
-E: Absolute free energies of binding of a small molecule to a host (e.g.
-protein), F: Relative free energy of binding of one molecule with respect to another, here toluene and benzyl alcohol.
+    \caption{\textbf{Illustration of common types of free energies differences that can be calculated using alchemical free energy methods.} \textbf{A}: Change in free energy due to a conformational change of the molecule.
+\textbf{B}: Partition coefficient such as $\log P$ or $\log D$ depend on a change in free energy between different phases; here, as an example the partition coefficient between methanol and water is shown.
+\textbf{C}: Free energy difference associated with the insertion of a molecule into a membrane.
+\textbf{D]: Effect of mutations of protein or host residues on free energies of binding.
+\textbf{E}: Absolute free energies of binding of a small molecule to a host (e.g.
+protein), \textbf{F}: Relative free energy of binding of one molecule with respect to another, here toluene and benzyl alcohol.
     \label{fig:fig_what_is_alchemy}
@@
 As some of the theoretical background can seem daunting, we provide an essentials guide to the theory behind alchemical free energy calculations in Section~\ref{sec:theory}.
-In the remainder of this paper, we will cover topics that are key to the preparation~(\ref{sec:step1}), choice and use of correct protocols~(\ref{sec:simulation_protocol_choice}), and finally the best practices that should be used for the analysis of alchemical calculations~(\ref{sec:data_analysis}).
+In the remainder of this paper, we will cover topics that are key to the preparation~(\ref{sec:prerequisites}), choice and use of correct protocols~(\ref{sec:simulation_protocol_choice}), and finally the best practices that should be used for the analysis of alchemical calculations~(\ref{sec:data_analysis}).
 Particular focus will be given to aspects of the molecular simulations which are unique to alchemical calculations---these include the calculation of transfer free energies (hydration free energies, partition coefficients, etc.), and binding free energies (absolute and relative).
@@
 \begin{tcolorbox}[title=Acronyms, colback=blue!10!white]
-    \textit{CPU} - Central Processing Unit\\
-    \textit{BAR} - Bennett Acceptance Ratio\\
-    \textit{FEP} - Free Energy Perturbation\\
-    \textit{GPCR} - G-Protein Coupled Receptor\\
-    \textit{GPU} - Graphics Processing Unit\\
-    \textit{MBAR} - Multistate Bennett Acceptance Ratio\\
-    \textit{MCSS} - Maximum Common Substructure\\
-    \textit{MD} - Molecular Dynamics\\
-    \textit{RBFE} - Relative Binding Free Energy\\
-    \textit{RMSE} - Root Mean Square Error\\
-    \textit{SAR} - Structure-Activity Relationships\\
-    \textit{TI} - Thermodynamic Integration
+    \textit{CPU} -- Central Processing Unit\\
+    \textit{BAR} -- Bennett Acceptance Ratio\\
+    \textit{FEP} -- Free Energy Perturbation\\
+    \textit{GPCR} -- G-Protein Coupled Receptor\\
+    \textit{GPU} -- Graphics Processing Unit\\
+    \textit{MBAR} -- Multistate Bennett Acceptance Ratio\\
+    \textit{MCSS} -- Maximum Common Substructure\\
+    \textit{MD} -- Molecular Dynamics\\
+    \textit{RMSE} -- Root Mean Square Error\\
+    \textit{MUE} -- Mean unsigned error\\
+    \textit{SAR} -- Structure-Activity Relationships\\
+    \textit{TI} -- Thermodynamic Integration
 \end{tcolorbox}
@@
 \begin{tcolorbox}[title=List of Symbols, colback=green!10!white]
-U - Potential energy\\
+$U$ -- potential energy\\
+$u$ -- reduced potential energy \\
+$\Delta G$ -- free energy \\
+$\Delta f$-- reduced free energy \\
+$k_B$-- Boltzmann constant \\
+$T$ -- temperature
+$k_b$ -- rate of binding \\
+$K_D$ -- dissociation constant \\
+$\lambda$ -- parameter for regulation of the alchemcial free energy path \\
+$g$ -- statistical inefficiency
 \end{tcolorbox}
-\begin{tcolorbox}[title=Common terminology, colback=green!10!white]
-    ligand - small molecular binding to a larger host molecule such as a protein.\\ 
+\begin{tcolorbox}[title=Common terminology, colback=yellow!10!white]
+    ligand - small molecular binding to a larger host molecule such as a protein (recpetor).\\ 
     lead optimization - process of designing a drug candidate from an initial lead compound, considering its affinity and various pharmacokinetic parameters.\\
@@
 %
-%%%%%%%%%%%%
-% Step 1   %
-%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Simulation prerequisites  %
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %
 \section{Simulation prerequisites}
-\label{sec:step1}
-%
+\label{sec:prerequisites}
 The alchemical free energy protocols, that are discussed in Section~\ref{sec:simulation_protocol_choice}, are defined for a specific type of free energy calculation, i.e.
@@
 \paragraph{How to deal with different binding modes?}
-Generally, binding modes within congeneric series are conserved~\cite{}, however, exceptions exist~\cite{}, and weaker binding compounds are more likely to adopt different binding modes (REFS).
+Generally, binding modes within congeneric series are conserved~\cite{wacker2010conserved}, however, exceptions exist~\cite{brandt2011congeneric,nazare2005probing}, and weaker binding compounds are more likely to adopt different binding modes (REFS).
 There may also be ambiguity about the relative orientation of substituents.
@@
 This may also lead to steric clashes with the receptor coordinates of the reference ligand if structural rearrangements are needed to accommodate different members of the congeneric series.
-Small steric clashes may be resolved during subsequent simulation equilibration prior to data collection but there is a risk that the complex relaxes to an alternative metastable state.
+Small steric clashes may be resolved during subsequent simulation equilibration prior to data collection, but there is a risk that the complex relaxes to an alternative metastable state.
 %
@@
 Preliminary considerations necessary for using free energy methods to compute partition coefficients are generally more straight forward.
-For example, a 3D minimised structure of the solute can be generated with a simple tool such as babel and solvated to prepare the input  to compute a free energy of hydration.
+For example, a 3D minimised structure of the solute can be generated with a simple tool such as openBabel and solvated to prepare the input  to compute a free energy of hydration~\cite{oboyle2011open}.
 However, in these cases a careful choice of forcefield, as well as water models or organic solvents is essential.
@@
 
-Several other case studies of FEP calculations can be found in the review by Williams-Noonan et al.~\cite{williams-noonan2018free}.
+Several other case studies of alchemical calculations can be found in the review by Williams-Noonan et al.~\cite{williams-noonan2018free}.
 For comparison of FEP+ and Gromacs (using the AMBER99SB-ILDN and GAFF2 force field), cf.
@@
 % \textbf{How do I get started setting up an alchemical free energy calculation}
-% Extensive explanation for the checklist questions can be found in Section~\ref{sec:step1}.
+% Extensive explanation for the checklist questions can be found in Section~\ref{sec:prerequisites}.
 % \begin{itemize}
@@
 BA: Helped write the uncertainty estimation, stopping conditions, and output analysis sections and created figure ~\ref{fig:convergence_forward_reverse}.
-JM: Contributed to sections~\ref{subsec:reproducible},~\ref{sec:step1},~\ref{} 4.2, 5.3, 6, 8.3.1, 8.3.2\
+JM: Contributed to sections~\ref{subsec:reproducible},~\ref{sec:prerequisites},~\ref{} 4.2, 5.3, 6, 8.3.1, 8.3.2\
 JDC: Wrote section~\ref{sec:decorrelating-samples} and~\ref{sec:automatic-equilibration-detection} discussed structure and design of the whole document, suggested Figs.~\ref{fig:fig_what_is_alchemy} and~\ref{fig:fig_sampling_scheme} \todo[inline, color={red!20}]{@JDC check info}

### 2020-01-19 paper/manuscript.tex
small fixes

@@
 For instance, does the X-ray that is to be used for the calculations match the construct used for screening (i.e.
-catalytic vs full length, monomer vs dimer etc) (CITE https://www.nature.com/articles/s41598-018-23039-5)?
+catalytic vs full length, monomer vs dimer etc)~\cite{perez-benito2018predicting}?
 Also, were certain co-factors or partner proteins required in the bioassay?
@@
 The software builds the network of perturbations, connecting similar compounds based on LOMAP methodology~\cite{liu2013lead}, all inputs for the Desmond simulations are built automatically and the job can be saved or run immediately on available GPUs.
-The cost of this software is often prohibitive but alternative tools are becoming available to allow relatively straight forward running of alchemical FE calculations with other software~\cite{gapsys2015pmx, loeffler2015fesetup, song2019using, gapsys2020large,  jespers2019qligfep}.
+Commercial tools can be expensive in some cases, but alternative tools are becoming available to allow relatively straight forward running of alchemical FE calculations with other software~\cite{gapsys2015pmx, loeffler2015fesetup, song2019using, gapsys2020large,  jespers2019qligfep}.
 One consideration at this point is to define the perturbations suitable for the application.
@@
 \todo[inline, color={red!40}]{LNN: write common principles alchemical path choice. MRS: mostly wrote this, LNN can comment on it. LNN: LGTM}
-\todo[inline, color={red!40}]{AR: write absolute-specific section on  alchemical path choice}
-\todo[inline, color={red!40}]{BA: write relative-specific section on  alchemical path choice}
-
 \begin{figure}

### 2020-01-20 paper/manuscript.tex
Updates from Overleaf

@@
 In alchemical simulations, the introduction of intermediate \textit{alchemical states} that bridge the high-probability regions of configuration space between the two physical endstates of interest permits the free energy of large transformations to be robustly computed.
-Alchemical calculations can be used in a variety of scenarios, such as computing the free energy of a conformational change for a molecule with a high barrier to interconversion (Fig.~\ref{fig:fig_what_is_alchemy} (A));
+Alchemical calculations can be used in a variety of scenarios, such as computing the free energy of a conformational change for a molecule with a high barrier to interconversion (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{A}));
 \todo[inline, color={yellow!20}]{JDC: Do you intend these to be representative of BLUES type simulations? -- yes}
-computing partition ($\log P$) or distribution ($\log D$) coefficients between environments (Fig.~\ref{fig:fig_what_is_alchemy} (B))~\cite{rustenburg2016measuring, bosisio2016blinded} or compartments or membranes (Fig.~\ref{fig:fig_what_is_alchemy} (C))~\cite{corey2019insights}.
-Furthermore, alchemical calculations are frequently used to estimate changes in free energies upon modifying a ligand or protein: a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy} (D))\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}; the entire ligand can be alchemical transferred from protein to solvent in an absolute binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (E))~\cite{mobley2007predicting,aldeghi2015accurate,aldeghi2017predictions}; and small alchemical modifications can be made between chemically related ligands to estimate differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (F))~\cite{wang2015accurate,mey2016blinded,gapsys2020large}.
+computing partition ($\log P$) or distribution ($\log D$) coefficients between environments (Fig.~\ref{fig:fig_what_is_alchemy} (\texbf{B}))~\cite{rustenburg2016measuring, bosisio2016blinded} or compartments or membranes (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{C}))~\cite{corey2019insights}.
+Furthermore, alchemical calculations are frequently used to estimate changes in free energies upon modifying a ligand or protein: a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{D}))\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}; the entire ligand can be alchemical transferred from protein to solvent in an absolute binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{E}))~\cite{mobley2007predicting,aldeghi2015accurate,aldeghi2017predictions}; and small alchemical modifications can be made between chemically related ligands to estimate differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (F))~\cite{wang2015accurate,mey2016blinded,gapsys2020large}.
 %
@@
 \textbf{C}: Free energy difference associated with the insertion of a molecule into a membrane.
-\textbf{D]: Effect of mutations of protein or host residues on free energies of binding.
+\textbf{D}: Effect of mutations of protein or host residues on free energies of binding.
 \textbf{E}: Absolute free energies of binding of a small molecule to a host (e.g.
@@
 \label{sec:theory}
-\todo[inline, color=green!20]{ASJSM: This section needs to be expanded and give a good introduction to the theoretical background without going into too much detail.}
 Why would you want to run an alchemical free energy calculation and why do they work?
@@
 \end{equation}
-The equilibrium rate of binding is given by the ratio of concentrations of product $[PL]$ and reactants $[P]$, $[L]$:
+The equilibrium constant  ($K_b^{\circ}$) is given by the ratio of concentrations of product $[PL]$ and reactants $[P]$, $[L]$:
 \begin{equation}
@@
 \end{equation}
-In order to define a reference state, the standard state is usually defined as $1 \mathrm{mol}/\mathrm{L}$ and $c^\circ$ is the standard state concentration. Thus, the binding free energy is given by:
+In order to define a reference state, the standard state is usually defined as $1 \mathrm{mol}/\mathrm{L}$ and $c^\circ$ is the standard state concentration.
+Thus, the Gibbs free energy of binding $\Delta G_{\mathrm{bind}}$ is given by:
 \begin{equation}
-    \Delta G_{\mathrm{bind}} = -RT\ln K_b^{\circ}.
+    \Delta G_{\mathrm{bind}} = -k_BT\ln K_b^{\circ},
 \end{equation}
-One way to estimate the binding rate is by computing the probability of finding the molecular system in the bound state or the unbound state, making use of partition function ratios, i.e.
+where $k_B$ is the Boltzmann constant and $T$ the temperature of the system. 
+One way to estimate the equilibrium constant is by computing the probability of finding the molecular system in the bound state ($PL$) or the unbound state ($P+L$), making use of partition function ratios, and assuming that the contribution of water compressibility at 1 atm has a negligible effect on the free energy of binding~\cite{shirts2012best, gilson1997statisticalthermodynamic} i.e.
 
 \begin{equation}
-    \Delta G_{\mathrm{bind}} = -RT\ln\frac{\int_\mathrm{bound} \exp(-\beta U(\mathbf{x}))d\mathbf{x}}{\int_\mathrm{unbound} \exp(-\beta U(\mathbf{x}))d\mathbf{x}}
+    \Delta G_{\mathrm{bind}} \cong \Delta F= -k_BT\ln\frac{\int_\mathrm{bound} \exp(-\beta U(\mathbf{x}))d\mathbf{x}}{\int_\mathrm{unbound} \exp(-\beta U(\mathbf{x}))d\mathbf{x}},
 \end{equation}
-Or more generally in a simplified form the free energy difference between the bound and unbound states is:
+where $\beta = \frac{1}{k_BT}$ and $F$ is the Helmholtz free energy. 
+Or more generally in a simplified form the free energy difference ($\Delta f$) between the bound and unbound states is:
 \begin{equation}
@@
 where $Z = \int_\mathrm{bound} \exp(-\beta U(\mathbf{x}))d\mathbf{x}$ represents the configurational integral. 
-How can we compute $\Delta G_{\mathrm{bind}}$ from a molecular dynamics simulation?
-Running a direct simulation of the binding events is too computationally costly, meaning we cannot easily estimate the configurational integrals of the bound and unbound state from the simulation.
-This is where alchemical free energy simulations come in.
+Unfortuantely it is not possible to evaluate the integral required for the configurational integral exactly, but can be approximated using molecular dynamics simulations. 
+So, how can we compute $\Delta G_{\mathrm{bind}}$ from a molecular dynamics simulation?
+Running a direct simulation of the binding events is too computationally costly, meaning we cannot easily estimate the configurational integrals of the bound and unbound state from the simulation, as typically binding events happen at a $\mu$s scale, and simulations reach around 1 $\mu$s/day.
+For a reliable estimate of a single free energy of binding many days of simulations would be required, making these type of calculations uninteresting at a pharamaceutical timescale.
+Instead, we can run alchemical free energy simulations, by making use of the fact that free energy is a state function.
 We can write down a thermodynamic cycle as seen in ~\ref{fig:fig_binding_thermodynamic_cycle}, for a relative free energy simulation.
@@
 \begin{eqnarray}
-Z(\lambda) &=& \int dx \, e^{-u(x; \lambda)} .
+Z(\lambda) &=& \int_{\Gamma} d\vec{q} \, e^{-u(\vec{q}; \lambda)} ,
 \label{equation:partition-function-definition}
 \end{eqnarray}
-Here, the inverse thermal energy $\beta \equiv (k_B T)^{-1}$ where $k_B$ is the Boltzmann constant and $T$ is the absolute temperature, and the \emph{reduced potential} $u(x; \lambda)$~\cite{shirts2008statistically} is generally given by a trace over thermodynamic parameters with their conjugate dynamical variables,
+, where $\Gamma$ signifies the phase space volume over which to integrate. 
+The \emph{reduced potential} $u(\vec{q}; \lambda)$~\cite{shirts2008statisticallya} is generally given by a trace over thermodynamic parameters with their conjugate dynamical variables,
 \begin{eqnarray}
-u(x;\lambda) &\equiv& \beta \left[ U(x;\lambda) + p \, V(x) + \sum_{i=1}^N \mu_i \, N_i(x) + \cdots \right] . \label{equation:reduced-potential}
+u(\vec{q};\lambda) &\equiv& \beta \left[ U(\vec{q};\lambda) + p \, V(\vec{q}) + \sum_{i=1}^N \mu_i \, N_i(\vec{q}) + \cdots \right] . \label{equation:reduced-potential}
 \end{eqnarray}
 Here, the collection of thermodynamic and alchemical parameters $\theta \equiv \{\beta, \lambda, p, \mu, \ldots\}$ defines a \emph{thermodynamic state}.
-The physical transformation of interest---such as a protein-ligand association process, or transfer free energy among phases---may involve a thermodynamic cycle that requires several free energy differences $\Delta f$ to be computed in order to produce an estimate of the overall desired quantity.
+The physical transformation of interest---such as a protein-ligand association process, or transfer free energy among phases---may involve a thermodynamic cycle that requires several free energy differences $\Delta f$ to be computed in order to produce an estimate of the overall desired quantity, see for example the thermodynamic cycle for an absolute free energy calculation as seen in~\ref{fig:fig_absolute_thermodynamic_cycle}.
 
@@
     \caption{{\bf Thermodynamic cycle for computing the relative free energy of binding ($\Delta \Delta G$) between two related small molecules to a supramolecular host.}
-    The relative binding free energy difference between two small molecules, $\Delta \Delta G_{\mathrm{bind}, A \rightarrow B} \equiv \Delta G_{\mathrm{bind}, B} - \Delta G_{\mathrm{bind}, A}$---here benzyl alcohol (top) to toluene (bottom)---can be computed as a difference between two alchemical transformations, $\Delta G_\mathrm{bound} - \Delta G_\mathrm{solvated}$, where $\Delta G_\mathrm{bound}$ represents the free energy change of transforming $A \rightarrow B$ in complex and $\Delta G_\mathrm{solvated}$ the free energy change of transforming $A \rightarrow B$ in solvent.}
+    The relative binding free energy difference between two small molecules, $\Delta \Delta G_{\mathrm{bind}, A \rightarrow B} \equiv \Delta G_{\mathrm{bind}, B} - \Delta G_{\mathrm{bind}, A}$---here benzyl alcohol (top) to toluene (bottom)---can be computed as a difference between two alchemical transformations, $\Delta G_\mathrm{bound} - \Delta G_\mathrm{solvated}$, where $\Delta G_\mathrm{bound}$ represents the free energy change of transforming $A \rightarrow B$ in complex, i.e.
+bound to a host molecule, and $\Delta G_\mathrm{solvated}$ the free energy change of transforming $A \rightarrow B$ in solvent, typically water.}
     \label{fig:fig_binding_thermodynamic_cycle}
@@
     \includegraphics[width=0.95\columnwidth]{figures/fig6_types_of_networks/Figure.pdf}
-    \caption{\textbf{Examples of perturbation networks} \textbf{A} Star shaped network with the crystal structure in the center.
-\textbf{B} Network with cycle closures (see more on this in Sec.~\ref{sec:are-they-good}).
+    \caption{\textbf{Examples of perturbation networks} (\textbf{A}) Star shaped network with the crystal structure in the center.
+(\textbf{B}) Network with cycle closures (see more on this in Sec.~\ref{sec:are-they-good}).
 Arrows indicate the direction of the perturbation.
@@
 \begin{figure}[h!]
-    \includegraphics[width=0.95\linewidth]{figures/fig5_thermo_cyc_abs/Figure.pdf}
-    \caption{\textbf{Thermodynamic cycle required for an aboslute free energy calculation -- absolute free energy of binding example} The fully interacting ligand in water (\textbf{A}), has its charges turned off to pass to \textbf{B} followed by turning of van der Waals terms, resulting in a non-interacting ligand in water in \textbf{C}.
-Restraints are used on the fully interacting ligand in the binding site of a protein or host molecule \textbf{D}.
+    \includegraphics[width=0.95\linewidth]{paper/figures/fig5_thermo_cyc_abs/Figure.pdf}
+    \caption{\textbf{Thermodynamic cycle required for an aboslute free energy calculation -- absolute free energy of binding example} The fully interacting ligand in water (\textbf{A}), has its charges turned off to pass to (\textbf{B}) followed by turning of van der Waals terms, resulting in a non-interacting ligand in water in (\textbf{C}).
+Restraints are used on the fully interacting ligand in the binding site of a protein or host molecule (\textbf{D}).
 The next step is to turn off the charges again (\textbf{E}) followed by the van der Waals interactions resulting in a non-interacting complex state (\textbf{F}).
@@
     \caption{Alchemical intermediates are createtd by making the potential energy depend on a an additional variable $\lambda$ that interpolates between the chemical endpoints.
-In \textbf{A}, at $\lambda=0$ the molecule is a fully interacting phenol and at $\lambda=1$,  a fully interacting benzene.
-\textbf{B} shows an illustration of the probability distribution of the potential energies as the switching function takes values of $\lambda=0$ to $\lambda=1$, intermediates are required for a sufficient overlap in potential energies to estimate a free energy difference between $\lambda=0$ and $\lambda=1$.
-    One of the most efficient families of intermediate pathways  are softcore potentials, whose shape is illustrated in \textbf{C} with dependence on $\lambda$ and \textbf{D} with dependence on the potential energy.
+In (\textbf{A}), at $\lambda=0$ the molecule is a fully interacting phenol and at $\lambda=1$,  a fully interacting benzene.
+(\textbf{B}) shows an illustration of the probability distribution of the potential energies as the switching function takes values of $\lambda=0$ to $\lambda=1$, intermediates are required for a sufficient overlap in potential energies to estimate a free energy difference between $\lambda=0$ and $\lambda=1$.
+    One of the most efficient families of intermediate pathways  are softcore potentials, whose shape is illustrated in (\textbf{C}) with dependence on $\lambda$ and (\textbf{D}) with dependence on the potential energy.
 Note how when $\lambda$ approaches 0, the energy smoothly approaches zero at all $r$, a necessary requirement for efficient calculations.
@@
 For a 'good' path to work sufficient overlap between pontential energies is required.
-Fig.~\ref{fig:fig_what_is_lambda} \textbf{A} and \textb{B} try to illustrate this.
-Fig.~\ref{fig:fig_what_is_lambda} \textbf{A} shows in a pictorial way a soft-core potential can be applied across different $\lamda$s.
-Fig.~\ref{fig:fig_what_is_lambda} \textbf{B} illustrates the potential energy distributions at the different $\lambda$ intermediates, with sufficient overlap between neighboring $\lambda$ states to ensure that reweighting estimators such as MBAR can be used for analysis (see Sec.\ref{subsec:estimators}).
-The actual transformation is best handled with soft-core potentials of the form shown in Fig.\ref{fig:fig_what_is_lambda} \textbf{C} and \textbf{B}, with more details given below.
+Fig.~\ref{fig:fig_what_is_lambda} (\textbf{A}) and (\textb{B}) try to illustrate this.
+Fig.~\ref{fig:fig_what_is_lambda} (\textbf{A}) shows in a pictorial way a soft-core potential can be applied across different $\lamda$s.
+Fig.~\ref{fig:fig_what_is_lambda} (\textbf{B}) illustrates the potential energy distributions at the different $\lambda$ intermediates, with sufficient overlap between neighboring $\lambda$ states to ensure that reweighting estimators such as MBAR can be used for analysis (see Sec.\ref{subsec:estimators}).
+The actual transformation is best handled with soft-core potentials of the form shown in Fig.\ref{fig:fig_what_is_lambda} (\textbf{C}) and (\textbf{B}), with more details given below.
 %
@@
 \begin{equation}
-    U(\vec{r_{ij}},\lambda) = 4\epsilon_{ij} \lambda \left(\frac{1}{(\alpha(1-\lambda) + (r_{ij}/\sigma_{ij})^6)^2} +  \frac{1}{\alpha(1-\lambda) + (r_{ij}/\sigma_{ij})^6}\right)\label{eq:softcore}
+    U(\vec{r_{ij}},\lambda) = 4\epsilon_{ij} \lambda \left(\frac{1}{(\alpha(1-\lambda) + (r_{ij}/\sigma_{ij})^6)^2} +  \frac{1}{\alpha(1-\lambda) + (r_{ij}/\sigma_{ij})^6}\right)
+    \label{eq:softcore}
 \end{equation}
@@
 There are several different variants of the same functional form~\cite{zacharias1994separationshifted, beutler1994avoiding,pham2011identifying}, but the one in given in eq.~\ref{eq:softcore} is easy to understand and implement and fairly numerically stable.
-This functional form is shown in parts \textbf{C} and \textbf{D} of Fig.~\ref{fig:fig_what_is_lambda}.
+This functional form is shown in parts (\textbf{C}) and (\textbf{D}) of Fig.~\ref{fig:fig_what_is_lambda}.
 It has been shown that more complicated forms are not significantly more efficient than eq.~\cite{pham2012optimal}.
-We recommend using the softcore potential given in eq.~\cite{eq:softcore}, unless there is a compelling reason otherwise.
-Using a similar equation to eq.~\cite{eq:softcore} may be acceptable in most circumstances if that is what is supported in your chosen software.
+We recommend using the softcore potential given in eq.~\ref{eq:softcore}, unless there is a compelling reason otherwise.
+Using a similar equation to eq.~\ref{eq:softcore} may be acceptable in most circumstances if that is what is supported in your chosen software.
 However, if you are inserting or removing entire atomic sites, we heavily recommend against using the linear approach; it will be very difficult to get correct results.

### 2020-01-20 paper/manuscript.tex
uniform use of vectors

@@
 \begin{equation}
-    \Delta G_{\mathrm{bind}} \cong \Delta F= -k_BT\ln\frac{\int_\mathrm{bound} \exp(-\beta U(\mathbf{x}))d\mathbf{x}}{\int_\mathrm{unbound} \exp(-\beta U(\mathbf{x}))d\mathbf{x}},
+    \Delta G_{\mathrm{bind}} \cong \Delta F= -k_BT\ln\frac{\int_\mathrm{bound} \exp(-\beta U(\vec{q}))d\vec{q}}{\int_\mathrm{unbound} \exp(-\beta U(\vec{q}))d\vec{q}},
 \end{equation}
@@
 \end{equation}
-where $Z = \int_\mathrm{bound} \exp(-\beta U(\mathbf{x}))d\mathbf{x}$ represents the configurational integral. 
+where $Z = \int_\mathrm{bound} \exp(-\beta U(\vec{q}))d\vec{q}$ represents the configurational integral. 
 Unfortuantely it is not possible to evaluate the integral required for the configurational integral exactly, but can be approximated using molecular dynamics simulations. 

### 2020-01-20 paper/checklist.tex
Lots of fiddling about with tables and checklist

@@
-\onecolumn
-%Checklist:
-%%%%%%%%%%%%%%%% 
-%% Checklist-specific precommands:
-\newenvironment{checklist_font}{\fontfamily{lmss}\small\selectfont}{\par}
-\newlist{todolist}{itemize}{2}
-\setlist[todolist]{label=$\square$}
-%% start checklist:
-\begin{checklist_font}
+\clearpage
+
+\section{Checklist}
+\label{sec:checklist}
+\begin{Checklists*}[p!]
+\begin{checklist}{ Know what you want to simulate}
+    \textbf{Initial questions you should ask before you set up an alchemical free energy calculation using molecular dynamics simulations}
 \begin{itemize}
-\begin{todolist}
-
-%%%%%%%%%%%%%%%%%
-%%%%%% step 0
-\begin{tcolorbox}[width=15cm, center]
-    {\large {\color{teal}STEP 0: KNOW WHAT YOU WANT TO SIMULATE}} \newline
-    \textbf{Initial questions you should ask before you set up an alchemical free energy calculation using molecular dynamics simulations}
     \item Do I understand the biology, chemistry and physics of my system?
@@
     \item Can I open-source my data?
-\end{tcolorbox}
+\end{itemize}
+\end{checklist}
 
-%%%%%% step 1
-\begin{tcolorbox}[width=15cm, center]
-    {\large {\color{teal}STEP 1: PREPARING YOUR SIMULATIONS}} \newline
+\begin{checklist}{Preparing your simulations}
 \textbf{Steps to getting started setting up your alchemical free energy calculation}
+\begin{itemize}
     \item Make sure you know why you have picked your (combination of) force field(s)
@@
     \item Check the stability of your system and whether it behaves the way you want it to
-\end{tcolorbox}
+\end{itemize}
+\end{checklist}
 
-%%%%%% step 2
-\begin{tcbraster}[raster columns=2,raster equal height, raster width=18cm]
-\begin{adjustwidth}{-4.3cm}{}           % workaround to make the dual box pass the margin
-\begin{todolist}
-    \begin{tcolorbox}
-    % absolute
-        {\large {\color{teal}STEP 2: RUNNING ABSOLUTE SIMULATIONS}} \newline
+\begin{checklist}{Running absolute simultions}
         \textbf{Steps to running your absolute alchemical free energy calculations}
-
+\begin{itemize}
  \item Your ligands should have the same, biologically correct binding pose
@@
         \item Apply the right correction terms
-    \end{tcolorbox}
-    \begin{tcolorbox}
-    % relative
-        {\large {\color{teal}STEP 2: RUNNING RELATIVE SIMULATIONS}} \newline
+\end{itemize}
+\end{checklist}
+
+\begin{checklist}{Running relative simulations}
         \textbf{Steps to running your relative alchemical free energy calculations}
-        \item           % circumvents bug in tcbraster/todolist
+\begin{itemize}
    \item Your ligands should have the same, biologically correct binding pose
@@
         \item Make sure you subsample in your MBAR protocol
-    \end{tcolorbox}
+\end{itemize}
+\end{checklist}
 
-\end{todolist}
-\end{adjustwidth}
-\end{tcbraster}
-
-%%%%%% step 3
-\begin{tcolorbox}[width=15cm, center]
-    {\large {\color{teal}STEP 3: HOW DO I KNOW WHICH SIMULATIONS ARE NOT RELIABLE?}} \newline
+\begin{checklist}{How do I know which simulations are not reliable}
     \textbf{Situations suggesting your relative alchemical free energy calculations have not run properly (assuming absence of experimental affinities)}
+        \begin{itemize}
                 \item Standard error (\textsigma) should be \textless1 kcal·mol-1 
@@
     \item The ligand is drifting out of the intended region after the vanishing step
-\end{tcolorbox}
+        \end{itemize}
+\end{checklist}
 
-%%%%%% step 4
-\begin{tcolorbox}[width=15cm, center]
-    {\large {\color{teal}STEP 4: WHY ARE THEY NOT RELIABLE?}} \newline
+\begin{checklist}{Why are they not reliable}
     \textbf{Suggestions for finding out why your alchemical free energy calculations may not be reliable}
+\begin{itemize}
     \item Check again whether dummy atoms were assigned correctly
@@
     \item Inspect the overlap matrices
-\end{tcolorbox}
+\end{itemize}
+\end{checklist}
 
-%%%%%% step 
-\begin{tcolorbox}[width=15cm, center]
-    {\large {\color{teal}STEP 5: DATA ANALYSIS}} \newline
+\begin{checklist}{Data Analysis}
     \textbf{Steps to analyzing your output data correctly}
+\begin{itemize}
     \item Make sure you have run enough replicates to ensure statistical reliability (\textgreater3)
@@
     \item You should be able to explain large energy differences based on your system (trajectory)
-\end{tcolorbox}
-
-%%%%%%%%%%%%%%%% 
-% end checklist:
-\end{todolist}
 \end{itemize}
-\end{checklist_font}
-%%%%%%%%%%%%%%%%%
-\twocolumn
+\end{checklist}
+\end{Checklists*}

### 2020-01-20 paper/manuscript.tex
Lots of fiddling about with tables and checklist

@@
 \todo[inline, color={yellow!20}]{JDC: Do you intend these to be representative of BLUES type simulations? -- yes}
-computing partition ($\log P$) or distribution ($\log D$) coefficients between environments (Fig.~\ref{fig:fig_what_is_alchemy} (\texbf{B}))~\cite{rustenburg2016measuring, bosisio2016blinded} or compartments or membranes (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{C}))~\cite{corey2019insights}.
-Furthermore, alchemical calculations are frequently used to estimate changes in free energies upon modifying a ligand or protein: a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{D}))\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}; the entire ligand can be alchemical transferred from protein to solvent in an absolute binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{E}))~\cite{mobley2007predicting,aldeghi2015accurate,aldeghi2017predictions}; and small alchemical modifications can be made between chemically related ligands to estimate differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (F))~\cite{wang2015accurate,mey2016blinded,gapsys2020large}.
+computing partition ($\log P$) or distribution ($\log D$) coefficients between environments (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{B}))~\cite{rustenburg2016measuring, bosisio2016blinded} or compartments or membranes (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{C}))~\cite{corey2019insights}.
+Furthermore, alchemical calculations are frequently used to estimate changes in free energies upon modifying a ligand or protein: a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{D}))\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}; the entire ligand can be alchemical transferred from protein to solvent in an absolute binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{E}))~\cite{mobley2007predicting,aldeghi2015accurate,aldeghi2017predictions}; and small alchemical modifications can be made between chemically related ligands to estimate differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{F}))~\cite{wang2015accurate,mey2016blinded,gapsys2020large}.
 %
@@
 \item nonspecific binding or multiple binding sites.
-\item Complex association stoichiometry (beyond simple 1:1 association)~\cite{awesome reference}.
+\item Complex association stoichiometry (beyond simple 1:1 association).
 \item Endpoint free energy methods such as MM-PBSA~\cite{genheden2015mm} and LIE~\cite{gutierrez-de-teran2012linear}
@@
 \begin{eqnarray}
-Z(\lambda) &=& \int_{\Gamma} d\vec{q} \, e^{-u(\vec{q}; \lambda)} ,
+Z(\lambda) &=& \int_{\Gamma} d\vec{q} \, \exp({-u(\vec{q}; \lambda)}) ,
 \label{equation:partition-function-definition}
@@
 0.2 kcal/mol between popular simulation packages has been reported.
-\cite{REF Reproducibility of hydration free energies Loefler et al.
-JCTC 2018} In the recent SAMPL6 SAMPLing challenge a larger variability of 0.3 to 1.0 kcal/mol was noted in the computed absolute binding free energies of host/guest systems even though similar input and simulation parameters were used \cite{Ref Rizzi et al SAMPLing biorxiv} and, in many cases, single-point energies were identical or nearly so.
+\cite{loeffler2018reproducibility} In the recent SAMPL6 SAMPLing challenge a larger variability of 0.3 to 1.0 kcal/mol was noted in the computed absolute binding free energies of host/guest systems even though similar input and simulation parameters were used \cite{rizzi2019sampl6} and, in many cases, single-point energies were identical or nearly so.
 Further work is needed to ensure reproducibility of alchemical free energy calculations across implementations to guarantee that force-field development efforts lead to transferable potential energy functions.
@@
 It may seem hard to have impact with substantially different, more potent, stand-out compounds in this scenario, but binding energy prediction can still be extremely useful for ensuring activity is maintained as other properties are optimized.
-An interesting cost benefit analysis has shown the value of activity prediction, see discussion above and articles such as~\cite{mobley2012perspectiv}.
+An interesting cost benefit analysis has shown the value of activity prediction, see discussion above and articles such as~\cite{mobley2012perspective}.
 From a drug discovery point of view, FEP is expanding its domain of applicability, and there are reports of success using homology models and GPCRs for instance, as well as enabling charge change and scaffold hopping, but these systems are undoubtedly more difficult.
@@
 The distinction between these can be illustrated by considering a hypothetical transformation from molecule A to molecule B, where both atoms share a common substructure but differ in their substituents; e.g.
-consider a transformation of benzene (A) to benzyl alcohol (B) Fig.~\ref{fig:topology}.
+consider a transformation of benzene to benzyl alcohol Fig.~\ref{fig:fig_topology}.
 In this case the common substructure is the benzene ring, though perhaps may be larger depending on how it is defined, as we discuss below.
@@
 Fig.~\ref{fig:fig_topology}'s right branch depicts how such a dual topology works.
-\begin{figure}[h!]
+\begin{figure}
     \includegraphics[width=0.95\columnwidth]{figures/fig3_topol/Figure.pdf}
-    \caption{\textbf{Two common topologies for alchemical calculations: single and dual topology} \textbf{left}: A single topology uses softcore potentials to convert from one type of atom to an other.
+    \caption{\textbf{Two common topologies for alchemical calculations: single and dual topology.} \textbf{Left}: A single topology uses softcore potentials to convert from one type of atom to an other.
 Dummy atoms (Du) are used when there is no corresponding maximum common substructure match between the two molecules for certain atoms.
-\textbf{right}: The dual topology does not convert one species to another, but only converts between Du atoms and an interacting species, but usually uses softcore potentials for this.
+\textbf{Right}: The dual topology does not convert one species to another, but only converts between Du atoms and an interacting species, but usually uses softcore potentials for this.
 The 'mixed' intermediate atoms are used in both dual and single topology approaches.
@@
 Specifically, with a single topology approach in mind, atom types are allowed to change, so a permissive MCSS search can be done, whereas with dual topology a more strict search is required.
-\begin{figure}[h!]
+\begin{figure}
     \includegraphics[width=0.95\linewidth]{figures/fig4_mcs/Figure.pdf}
@@
 %
-\paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss} \textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings.
+\paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss} (\textbf{B})), in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings.
 Some approaches have attempted to address this~\cite{clark2019relative} but a general solution is not yet in mainstream use.
@@
 Some commercial tools have underlying good heuristic but may fail with complicated input, needing user validation in particular when dealing with chiral compounds.
-\begin{figure}[h!]
+\begin{figure}
     \includegraphics[width=0.95\columnwidth]{figures/fig6_types_of_networks/Figure.pdf}
@@
 %
-\begin{figure}[h!]
+\begin{figure}
     \includegraphics[width=0.95\linewidth]{paper/figures/fig5_thermo_cyc_abs/Figure.pdf}
@@
 %
-Alternatively, a set of restraints proposed by Boresch have also commonly been employed, where all six rigid-body degrees of freedom governing the orientation of the ligand relative to the receptor are restrained~\cite{boresch2003absolutea, leitgeb2005alchemical}.
-Further restraints, such as on the overall ligand RMSD have also been used~\cite{woo2005calculation}.
+Alternatively, a set of restraints proposed by Boresch have also commonly been employed, where all six rigid-body degrees of freedom governing the orientation of the ligand relative to the receptor are restrained~\cite{boresch2003absolutea, leitgeb2005alchemicala}.
+Further restraints, such as on the overall ligand RMSD have also been used~\cite{woo2005calculationa}.
 %
@@
 In practice, the simulation of the bound state starts with the ligand already placed in the binding site and relies on kinetic trapping to maintain a bound complex.
-However, this strategy may not be sufficient if the complex dissociation rate is high or if methodologies such as Hamiltonian replica exchange~\cite{bitetti-putzer2003generalized} and expanded ensemble~\cite{chodera2011replica} are employed in absolute free energy calculations to enhance sampling since, in both cases, the ligand may find a way out of the binding site on timescales that are achievable by modern MD simulations.
+However, this strategy may not be sufficient if the complex dissociation rate is high or if methodologies such as Hamiltonian replica exchange~\cite{bitetti-putzer2003generalizeda} and expanded ensemble~\cite{chodera2011replica} are employed in absolute free energy calculations to enhance sampling since, in both cases, the ligand may find a way out of the binding site on timescales that are achievable by modern MD simulations.
 A solution commonly adopted in these cases is the use of one or more restraints making the unbound configurations energetically unfavorable through the addition of extra terms in the potential function.
@@
 This is sometimes referred to as an "alchemical ion" approach for dealing with the needed charge change, and is also employed by the Yank free energy package~\cite{wang2013identifying}.
-Charge corrections have also been explored, and are potentially a viable solution to this problem~\cite{mey2018impact} where artifacts introduced by finite-size effects are corrected numerically.
-~\cite{chen2018accurate}. However, application of such corrections typically remains less common than the use of a co-alchemical ion.
+Charge corrections have also been explored, and are potentially a viable solution to this problem~\cite{mey2018impact} where artifacts introduced by finite-size effects are corrected numerically~\cite{chen2018accurate}.
+However, application of such corrections typically remains less common than the use of a co-alchemical ion.
 %
@@
 The simplest approach is defining a set of fixed replicas each representing a different $\lambda$ value that run independent simulations at each of the intermediate $\lambda$ values, see Fig.~\ref{fig:fig_sampling_scheme}(\textbf{A}).
-This type of scheme is currently used for AMBER TI calculations~\cite{song2019} and for Siremol/Openmm~\cite{}.
+This type of scheme is currently used for AMBER TI calculations~\cite{song2019using} and for Siremol/Openmm~\cite{hedges2019biosimspace}.
 A slightly more complex approach is allowing replica mixing, meaning that each independent $\lambda$ simulation is attempted to be exchanged with a neighboring $\lambda$ replica.
@@
 In this approach, only end state $\lambda$ replicas ($\lambda$=0, $\lambda=1$) are simulated at equilibrium intermediate information is generated from non-equilibrium simulations.
-A schematic of this approach is shown in Fig.~\ref{fig:fig_sampling_scheme} \textbf{D}.
-
+A schematic of this approach is shown in Fig.~\ref{fig:fig_sampling_scheme}(\textbf{D}).
+%
 \begin{figure}
@@
 Single replica schemes and non-equilibrium schemes are not as established yet, but are very promising.
-
+%
 \subsubsection{How long should I run my simulation for and what information should be saved?}
@@
 These additional observables include, but are not limited to, the variance in $\frac{dU}{d\lambda}$ across all $\lambda$ values, calculating the variance in free energy using bootstrap analysis, and comparing differences in free energies calculated using different percentages of the simulation in both the forward and reverse directions ( see Fig.
-\ref{fig:convergence_forward_reverse}).
-%
-\begin{figure}[h!]
+~\ref{fig:convergence_forward_reverse}).
+%
+\begin{figure}
     \centering
@@
 To deal with this, some practitioners have actually computed relative binding free energies of different binding modes of the same ligand~\cite{palma2012computation}.
-For example, a perturbation which adds a methyl to an aromatic ring of a larger ligand might yield one result if the methyl points in one direction, and a different value if it points in the other due to slow ring motions.
-[e.g.
-get ref from Sukanya]
+For example, a perturbation which adds a methyl to an aromatic ring of a larger ligand might yield one result if the methyl points in one direction, and a different value if it points in the other due to slow ring motions~\cite{lincoff2016comparing}.
 One could compute the free energy of turning off the methyl group in one orientation and turning it back on in the other orientation to obtain the free energy difference between the two potential binding modes.
@@
 As a result, these estimators can produce significantly biased estimates if fed the entirety of simulation data generated without further processing~\cite{chodera2016simple}.
-%
 \begin{figure}
@@
     \caption{{\bf Automatic partitioning into equilibration and production regions.}
-    \textbf{A} The average (black line) standard deviation (shaded region) of the reduced potential $u^*$ over many independent replicate simulations started from the same initial conditions show a significant initial transient change before relaxing to the true average potential energy \textbf{B}.
+    (\textbf{A}) The average (black line) standard deviation (shaded region) of the reduced potential $u^*$ over many independent replicate simulations started from the same initial conditions show a significant initial transient change before relaxing to the true average potential energy (\textbf{B}).
 A cumulative average (red) of the entire simulation data demonstrates simulation bias not seen when initial simulation data is omitted (blue).
 Using an automated approach to detect equilibration of the boundary $t_0$ using statistical inefficiency $g$ (\textbf{C})  for an effective simulation interval (\textbf{D}).
-\textbf{E} The optimal equilibration boundary $t_0$ is selected to maximize the number of uncorrelated samples.
+(\textbf{E}) The optimal equilibration boundary $t_0$ is selected to maximize the number of uncorrelated samples.
     \emph{Figure adapted from~\cite{chodera2016simple}.}
@@
 \paragraph{Overlap matrix}
-One way of assessing reliability of the calculations is checking the phase space overlap between neighboring $\lambda$-windows~\cite{wu2005phasespace,wu2005phasespacea}.
-For this purpose, a so-called overlap matrix can be used.
-This is a $K\times K$ matrix, where $K$ is the number of simulated states, i.e.
+One way of assessing reliability of the calculations is checking the phase space overlap between neighboring $\lambda$-windows~\cite{wu2005phasespaceb, wu2005phasespacec}.
+For this purpose, a so-called overlap matrix $\mathcal{O}$ can be used.
+$\mathcal{O} \in \mathbb{R}^{K\cross K}$, with $K$ being the number of simulated states, i.e.
 $\lambda$s.
@@
 Details on the calculation and properties of these matrices can be found elsewhere~\cite{klimovich2015guidelines}.
-In an overlap matrix $\mathbf{O}$, the off-diagonal values (${O}_{i,j,i\ne j}$) are negatively correlated with the variance of the free energy difference.
+In an overlap matrix $\mathcal{O}$, the off-diagonal values (${O}_{i,j,i\ne j}$) are negatively correlated with the variance of the free energy difference.
 Accordingly, the uncertainty of the free energy difference between the states $i$ and $j$ will be smaller when ${O}_{i,j,i\ne j}$ is larger (and thus the values in the main diagonal (${O}_{i,j,i=j}$) are smaller).
@@
 the diagonals above and below the main diagonal) should at least be 0.03 to obtain a reliable free energy estimate~\cite{klimovich2015guidelines}.
-Smaller values should be considered as a warning sign (see Fig.~\ref{fig:overlap}(C)), as the variance tends to be underestimated in case of poor overlap.
-
-\begin{figure}
+Smaller values should be considered as a warning sign (see Fig.~\ref{fig:overlap}(\textbf{C})), as the variance tends to be underestimated in case of poor overlap.
+%
+\begin{figure}[ht!]
 \includegraphics[width=0.95\columnwidth]{figures/fig12_overlap/Figure.pdf}
 \caption{\label{fig:overlap} \textbf{Overlap matrices:} Visualising overlap matrices can help with assessing the quality of simulation data.
-(A) shows good overlap with all first off-diagonal entries well above 0.03, the suggested threshold, (B) is an example of mediocre overlap with good overlap at lower $\lambda$ values and poor overlap at high $\lambda$ values.
-(C)shows poor overlap resulting in disconnected simulations that should not be used for an MBAR estimation of free energies.}
+(\textbf{A}) shows good overlap with all first off-diagonal entries well above 0.03, the suggested threshold, (\textbf{B}) is an example of mediocre overlap with good overlap at lower $\lambda$ values and poor overlap at high $\lambda$ values.
+(\textbf{C})shows poor overlap resulting in disconnected simulations that should not be used for an MBAR estimation of free energies.}
 \end{figure}
-
-Fig.~\ref{fig:overlap} shows examples of good, mediocre and poor overlap (A), (B), and (C) respectively.
-For Fig.~\ref{fig:overlap}(A), the probability to find a sample from state $i$ in its neighbouring state $j$ is about 0.2 for all states adjacent to the main diagonal, and hence the overall connectivity is good.
-In the case of Fig.~\ref{fig:overlap}(B), the overlap is strongly diminishing in the lower right corner, raising concerns regarding the reliability of the free energy estimate obtained.
-For Fig.~\ref{fig:overlap}(C), the state at $\lambda$ index = 6 is connected to neither of its neighbouring states.
+%
+Fig.~\ref{fig:overlap} shows examples of good, mediocre and poor overlap (\textbf{A}), (\textbf{B}), and (\textbf{C}) respectively.
+For Fig.~\ref{fig:overlap}(\textbf{A}), the probability to find a sample from state $i$ in its neighbouring state $j$ is about 0.2 for all states adjacent to the main diagonal, and hence the overall connectivity is good.
+In the case of Fig.~\ref{fig:overlap}(\textbf{B}), the overlap is strongly diminishing in the lower right corner, raising concerns regarding the reliability of the free energy estimate obtained.
+For Fig.~\ref{fig:overlap}(\textbf{C}), the state at $\lambda$ index = 6 is connected to neither of its neighbouring states.
 While this does not necessarily imply that the result for this perturbation is wrong, the energy estimate must at least be considered as highly unreliable.
@@
 Specifically, often the goal is to adjust the $\lambda$ spacing to perform additional sampling in the areas of poor overlap, rather than using arbitrarily spaced windows.
-Different approaches are described in section~\ref{}) and more details can be found in the literature~\cite{dakka2018concurrent, hahn2019alchemical}.
+Different approaches are described in section~\ref{sec:simulation_protocol_choice}) and more details can be found in the literature~\cite{dakka2018concurrent, hahn2019alchemical}.
 %
@@
 adding a functional group to a ligand) may provide an additional opportunity for error/consistency checking.
-Particularly, such calculations are often done to span a graph or tree of free energy calculations~\cite{xu2019diffnet,wang2013modeling,liu2013lead} and in some cases the free energy change to go between molecules A and B can be obtained via multiple transformation pathways.
+Particularly, such calculations are often done to span a graph or tree of free energy calculations~\cite{xu2019optimal,wang2013modeling,liu2013lead} and in some cases the free energy change to go between molecules A and B can be obtained via multiple transformation pathways.
 This allows a type of consistency checking where we assess how much the free energy change for that transformation in practice differs from equivalence.
@@
 We would encourage the following standard set of analyses and ways to represent data.
-\todo[inline]{JS: @everyone I suspect people will have strong opinions on this subject, feel free to include your own suggestions}
 \paragraph{Statistics to include}
@@
 \end{figure}
-An example of a best practice scatter comparison between computed and experimental values is shown in~\ref{fig:scatterplot_analysis}, highlighting outliers, error bars and confidence intervals.
+An example of a best practice scatter comparison between computed and experimental values is shown in Fig.~\ref{fig:scatterplot_analysis}, highlighting outliers, error bars and confidence intervals.
 The data for this plot is artificially generated for illustration purposes.
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%
+%  Online resources    %
+%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Online resources}
+\todo[inline]{Do we really need this section? Everything has been referenced}
 \begin{itemize}
@@
 \clearpage
+%%%%%%%%%%%%%%%%%%%%%%%%
+%  Benchmark information    %
+%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Benchmark Datasets}
-
+Several other case studies of alchemical calculations can be found in the review by Williams-Noonan et al.~\cite{williams-noonan2018free}.
+For comparison of FEP+ and Gromacs (using the AMBER99SB-ILDN and GAFF2 force field), cf.
+the recently published study by Pérez-Benito et al.~\cite{perez-benito2019predicting}.
+An overview of further suggested benchmark sets can be found in the review by Mobley and Gilson~\cite{mobley2017predicting} or on \url{alchemistry.org}~\cite{alchemistry}.
+These include cyclodextrins, the Cytochrome C peroxidase (CCP) protein model binding site, thrombin and bromodomains as well as solvation benchmark sets~\cite{paliwal2011benchmark}.
+Further work on shared protein-ligand benchmark sets is ongoing at \url{https://github.com/openforcefield/PLBenchmarks}.
 \label{sec:benchmark}
 \vspace{5mm}
-
 \begin{table}
@@
 \begin{tabular}{cccc}
-
 \textbf{Publication} & \textbf{Targets} & \textbf{Ligands} & \textbf{Force Field} \\
@@
 FEP+ Dataset~\cite{song2019using} & 8 & 199 & AMBER18 \\
-Fragment Optimization~\cite{steinbrecher2015accurate} & 8 & 96 & OPLS2.1 \\
+Fragments ~\cite{steinbrecher2015accurate} & 8 & 96 & OPLS2.1 \\
 Scaffold Hopping~\cite{wang2017accurate} & 6 & 21 & OPLS3 \\
@@
 T4 Lysozyme~\cite{mobley2017predicting} & 2 & 20 & NA \\
-Merck benchmarking set~\cite{MCompChem2019Sep} & 5 & 169 & OPSL3 \\
+Merck  set~\cite{MCompChem2019Sep} & 5 & 169 & OPSL3 \\
 
@@
 \end{table}
-
-Several other case studies of alchemical calculations can be found in the review by Williams-Noonan et al.~\cite{williams-noonan2018free}.
-For comparison of FEP+ and Gromacs (using the AMBER99SB-ILDN and GAFF2 force field), cf.
-the recently published study by Pérez-Benito et al.~\cite{perez-benito2019predicting}.
-An overview of further suggested benchmark sets can be found in the review by Mobley and Gilson~\cite{mobley2017predicting} or on \url{alchemistry.org}~\cite{alchemistry}.
-These include cyclodextrins, the Cytochrome C peroxidase (CCP) protein model binding site, thrombin and bromodomains as well as solvation benchmark sets~\cite{paliwal2011benchmark}.
-Further work on shared protein-ligand benchmark sets is ongoing at \url{https://github.com/openforcefield/PLBenchmarks}.
-
-\section{Checklist}
-\label{sec:checklist}
-\todo[inline, color={green!20}]{JS: @everyone Please input your own suggestions, I made some points off the top of my head. Feel free to add helpful citations to any of the bullet points}
-\todo[inline, color={green!20}]{JS: @self,ASJSM do some final formatting when text is complete.}
-
-% This provides a checklist which
-% - spans a full page
-% - consists of multiple sub-checklists
-% - exists on a separate page
-% This style of checklist will be especially helpful if you want to encourage readers to print and use your checklist in practice, as they
-% can easily print it without also printing other material from your manuscript. However, other styles of checklist are also possible (below).
-% \begin{Checklists*}[p!]
-
-% \begin{checklist}{Step 0 -- Know what you want to simulate }
-% \textbf{What are the first questions that need addressing before setting up a molecular dynamics simulation}\\
-% Extensive explanation for the checklist questions can be found in Section~\ref{sec:step0}.
-% \begin{itemize}
-% \item Can I get the required accuracy with the simulation I want to carry out?
-% \item Have I properly prepared my protein and ligand systems?
-% \item Does my system contain any groups that require custom parameters?
-% \item What simulation protocol will provide the most evidence to answer my hypothesis?
-
-% \item And finally
-% \end{itemize}
-% \end{checklist}
-
-% \begin{checklist}{Simulation preparation}
-% \textbf{How do I get started setting up an alchemical free energy calculation}
-% Extensive explanation for the checklist questions can be found in Section~\ref{sec:prerequisites}.
-% \begin{itemize}
-% \item Have I followed the Best practices for biomolecular simulation set up?
-% \item In a relative simulation, will I run into problems with clashing geometries in the ligand transformation or crystal waters?
-% \end{itemize}
-% \end{checklist}
-% \end{Checklists*}
-
-% \begin{Checklists*}[p!]
-% \begin{checklist}{Absolute simulations}
-% \textbf{What are the main things I need to consider for an absolute alchemical free energy calculation?}
-% Extensive explanation for the checklist questions can be found in Section~\ref{sec:simulation_protocol_choice}.
-% \begin{itemize}
-% \item Topology
-% \item Restraints
-% \item Standard state handling
-% \end{itemize}
-% \end{checklist}
-
-% \begin{checklist}{Relative simulations}
-% \textbf{What are the main things I need to consider for an relative alchemical free energy calculation?}
-% Extensive explanation for the checklist questions can be found in Section~\ref{sec:simulation_protocol_choice}.
-% \begin{itemize}
-% \item First thing
-% \item Also remember
-% \item And finally
-% \end{itemize}
-% \end{checklist}
-
-% \begin{checklist}{Analysis}
-% \textbf{This is all about analysis of the simulation}
-% Extensive explanation for the checklist questions can be found in Section~\ref{sec:step4}.
-% \begin{itemize}
-% \item Are my simulations converged enough?
-% \item Am I using the right analysis techniques?
-% \end{itemize}
-% \end{checklist}
-
-% \end{Checklists*}
-
-%Checklist:
-\clearpage
-%%%%%%%%%%%%%%%% 
-%% Checklist-specific precommands:
 
@@
 (Explain the contributions of the different authors here)
-ASJSM: Coordinated the document, contributed to most sections, and co-designed figures ~\ref{} and created figures \ref{fig:fig_what_is_lambda},~\ref{fig:fig1_what_is_alchemy},~\ref{fig:overlap} and replotted~\ref{fig:automatic-equilibration-detection} and improved ~\ref{fig:fig_types_of_networks}.
-JS: created figures \ref{fig:fig_what_is_alchemy},~\ref{fig:fig_binding_thermodynamic_cycle},~\ref{fig:fig_topology},~\ref{fig:fig_mcss},~\ref{fig:fig_types_of_networks},~\ref{fig:fig_absolute_thermodynamic_cycle},~\ref{fig:scatterplot_analysis}, the data representation section, the checklist and contributed to general formatting discussions and editing.
-MK: Contributed to the Data Analysis section, provided the data for figure~\ref{fig:overlap} and helped edit the paper.
-DLM: Contributed to the outline, drafted some of the sections, gave ideas on figures, and helped edit the paper.
-GT: Contributed to introduction and drug discovery sections, and helped edit the paper.
-AR: Created figure~\ref{fig:freeenergytrajectories}, contributed to section \ref{sec:simulation_protocol_choice}, and helped edit the paper.
-MRS: Helped create figure~\ref{fig:fig_what_is_lambda}, wrote sections describing choices for alchemical pathways and analysis for free energy calculations.
-LNN: Helped write the simulation length, stopping conditions, and information saving section.
-BA: Helped write the uncertainty estimation, stopping conditions, and output analysis sections and created figure ~\ref{fig:convergence_forward_reverse}.
-JM: Contributed to sections~\ref{subsec:reproducible},~\ref{sec:prerequisites},~\ref{} 4.2, 5.3, 6, 8.3.1, 8.3.2\
-JDC: Wrote section~\ref{sec:decorrelating-samples} and~\ref{sec:automatic-equilibration-detection} discussed structure and design of the whole document, suggested Figs.~\ref{fig:fig_what_is_alchemy} and~\ref{fig:fig_sampling_scheme} \todo[inline, color={red!20}]{@JDC check info}
-HX: Contributed section~\ref{subsec:accuracy}, to section~\ref{sec:relative-fe-protocol}, and to section~\ref{sec:are-they-good}.
+\textbf{ASJSM}: Coordinated the document, contributed to most sections, and co-designed figures ~\ref{} and created figures \ref{fig:fig_what_is_lambda},~\ref{fig:fig1_what_is_alchemy},~\ref{fig:overlap} and replotted~\ref{fig:automatic-equilibration-detection} and improved ~\ref{fig:fig_types_of_networks}.\\
+\textbf{JS}: created figures \ref{fig:fig_what_is_alchemy},~\ref{fig:fig_binding_thermodynamic_cycle},~\ref{fig:fig_topology},~\ref{fig:fig_mcss},~\ref{fig:fig_types_of_networks},~\ref{fig:fig_absolute_thermodynamic_cycle},~\ref{fig:scatterplot_analysis}, the data representation section, the checklist and contributed to general formatting discussions and editing.\\
+\textbf{MK}: Contributed to the Data Analysis section, provided the data for figure~\ref{fig:overlap} and helped edit the paper.\\
+DLM: Contributed to the outline, drafted some of the sections, gave ideas on figures, and helped edit the paper.\\
+\textbf{GT}: Contributed to introduction and drug discovery sections, and helped edit the paper.\\
+\textbf{AR}: Created figure~\ref{fig:freeenergytrajectories}, contributed to section \ref{sec:simulation_protocol_choice}, and helped edit the paper.\\
+\textbf{MRS}: Helped create figure~\ref{fig:fig_what_is_lambda}, wrote sections describing choices for alchemical pathways and analysis for free energy calculations.\\
+\textbf{LNN}: Helped write the simulation length, stopping conditions, and information saving section.\\
+\textbf{BA}: Helped write the uncertainty estimation, stopping conditions, and output analysis sections and created figure ~\ref{fig:convergence_forward_reverse}.
+\textbf{JM}: Contributed to sections~\ref{subsec:reproducible},~\ref{sec:prerequisites},~\ref{} 4.2, 5.3, 6, 8.3.1, 8.3.2\\
+\textbf{JDC}: Wrote section~\ref{sec:decorrelating-samples} and~\ref{sec:automatic-equilibration-detection} discussed structure and design of the whole document, suggested Figs.~\ref{fig:fig_what_is_alchemy} and~\ref{fig:fig_sampling_scheme} \todo[inline, color={red!20}]{@JDC check info}\\
+\textbf{HX}: Contributed section~\ref{subsec:accuracy}, to section~\ref{sec:relative-fe-protocol}, and to section~\ref{sec:are-they-good}.
 

### 2020-01-20 paper/manuscript.tex
trying to fiddle about with figure placement and checklist

@@
 \begin{figure}
-    \includegraphics[width=0.95\linewidth]{paper/figures/fig5_thermo_cyc_abs/Figure.pdf}
+    \includegraphics[width=0.95\linewidth]{figures/fig5_thermo_cyc_abs/Figure.pdf}
     \caption{\textbf{Thermodynamic cycle required for an aboslute free energy calculation -- absolute free energy of binding example} The fully interacting ligand in water (\textbf{A}), has its charges turned off to pass to (\textbf{B}) followed by turning of van der Waals terms, resulting in a non-interacting ligand in water in (\textbf{C}).
@@
 For a 'good' path to work sufficient overlap between pontential energies is required.
-Fig.~\ref{fig:fig_what_is_lambda} (\textbf{A}) and (\textb{B}) try to illustrate this.
-Fig.~\ref{fig:fig_what_is_lambda} (\textbf{A}) shows in a pictorial way a soft-core potential can be applied across different $\lamda$s.
+Fig.~\ref{fig:fig_what_is_lambda} (\textbf{A}) and (\textbf{B}) try to illustrate this.
+Fig.~\ref{fig:fig_what_is_lambda} (\textbf{A}) shows in a pictorial way a soft-core potential can be applied across different $\lambda$s.
 Fig.~\ref{fig:fig_what_is_lambda} (\textbf{B}) illustrates the potential energy distributions at the different $\lambda$ intermediates, with sufficient overlap between neighboring $\lambda$ states to ensure that reweighting estimators such as MBAR can be used for analysis (see Sec.\ref{subsec:estimators}).
@@
 For this purpose, a so-called overlap matrix $\mathcal{O}$ can be used.
-$\mathcal{O} \in \mathbb{R}^{K\cross K}$, with $K$ being the number of simulated states, i.e.
+$\mathcal{O} \in \mathbb{R}^{K\times K}$, with $K$ being the number of simulated states, i.e.
 $\lambda$s.
@@
 %
-Interestingly, as the variance is inversely correlated with the number of states\cite{klimovich2015guidelines}, it can in principle be reduced below any arbitrary threshold with enough simulation time and a large enough number of $\lambda$ windows.
+Interestingly, as the variance is inversely correlated with the number of states~\cite{klimovich2015guidelines}, it can in principle be reduced below any arbitrary threshold with enough simulation time and a large enough number of $\lambda$ windows.
 However, decreasing the variance to a value close to 0 is not feasible, as this approach would strongly increase the calculation time.
@@
 
+\clearpage
+
+\section{Checklist}
+\label{sec:checklist}
+\begin{Checklists*}[p!]
+\begin{checklist}{ Know what you want to simulate}
+    \textbf{Initial questions you should ask before you set up an alchemical free energy calculation using molecular dynamics simulations}
+\begin{itemize}
+    \item Do I understand the biology, chemistry and physics of my system?
+    \item Have I properly prepared my protein and ligand systems?
+    \item Does my system contain any structures that require custom parameters?
+    \item What simulation protocol will provide the most evidence to answer my hypothesis?
+    \item Are the projected computational expense and runtime realistic for my goals?
+    \item Will my protocol be reproducible? 
+    \item Will my statistics be reliable? If not, would more replicates solve the problem? Three replicates is the minimal number of replicates that should be run
+    \item Can I open-source my data?
+\end{itemize}
+\end{checklist}
+
+\begin{checklist}{Preparing your simulations}
+\textbf{Steps to getting started setting up your alchemical free energy calculation}
+\begin{itemize}
+    \item Make sure you know why you have picked your (combination of) force field(s)
+    \item Minimize your system
+    \item Equilibrate your system with your choice of thermodynamic ensemble
+    \item Check the stability of your system and whether it behaves the way you want it to
+\end{itemize}
+\end{checklist}
+
+\begin{checklist}{Running absolute simultions}
+        \textbf{Steps to running your absolute alchemical free energy calculations}
+\begin{itemize}
+ \item Your ligands should have the same, biologically correct binding pose
+        \item Make sure your \textlambda-scheduling is set correctly
+        \item Check if your ligands are discharging and vanishing correctly
+        \item Set up your restraints correctly
+        \item Make sure you subsample in your MBAR protocol
+        \item Apply the right correction terms
+\end{itemize}
+\end{checklist}
+
+\begin{checklist}{Running relative simulations}
+        \textbf{Steps to running your relative alchemical free energy calculations}
+\begin{itemize}
+   \item Your ligands should have the same, biologically correct binding pose
+        \item Make sure your λ-scheduling is set correctly
+        \item Make sure your molecular transformations are realistic (1-5 heavy atoms for reliable computations)
+        \item Generate a perturbation network by your method of choice; check whether you have enough cycle closures (preferably, every ligand should be in a cycle)
+        \item Check whether dummy atoms were assigned correctly
+        \item Make sure you subsample in your MBAR protocol
+\end{itemize}
+\end{checklist}
+
+\begin{checklist}{How do I know which simulations are not reliable}
+    \textbf{Situations suggesting your relative alchemical free energy calculations have not run properly (assuming absence of experimental affinities)}
+        \begin{itemize}
+                \item Standard error (\textsigma) should be \textless1 kcal·mol-1 
+    \item Simulated systems have not converged 
+    \newline\newline\textit{Relative:}
+    \item If you observe hysteresis in perturbations and incorrect cycle closures
+    \item Energy differences \textgreater15 kcal·mol-1  are unreliable
+    \newline\newline\textit{Absolute:}
+    \item Energies below -15 kcal·mol-1  are unreliable
+    \item The ligand has not sampled most of the intended region after the vanishing step
+    \item The ligand is drifting out of the intended region after the vanishing step
+        \end{itemize}
+\end{checklist}
+
+\begin{checklist}{Why are they not reliable}
+    \textbf{Suggestions for finding out why your alchemical free energy calculations may not be reliable}
+\begin{itemize}
+    \item Check again whether dummy atoms were assigned correctly
+    \item Inspect the trajectories across the λ-schedule (particularly the endpoints)
+    \item Inspect the overlap matrices
+\end{itemize}
+\end{checklist}
+
+\begin{checklist}{Data Analysis}
+    \textbf{Steps to analyzing your output data correctly}
+\begin{itemize}
+    \item Make sure you have run enough replicates to ensure statistical reliability (\textgreater3)
+    \item Compute both correlation and ranking coefficients and ranking statistics (e.g. r, \textrho, MUE and \texttau)
+    \item You should be able to explain large energy differences based on your system (trajectory)
+\end{itemize}
+\end{checklist}
+\end{Checklists*}
+\clearpage
+
 \section*{Author Contributions}
@@
 \textbf{MK}: Contributed to the Data Analysis section, provided the data for figure~\ref{fig:overlap} and helped edit the paper.\\
-DLM: Contributed to the outline, drafted some of the sections, gave ideas on figures, and helped edit the paper.\\
+\textbf{DLM}: Contributed to the outline, drafted some of the sections, gave ideas on figures, and helped edit the paper.\\
 \textbf{GT}: Contributed to introduction and drug discovery sections, and helped edit the paper.\\
@@
 \textbf{LNN}: Helped write the simulation length, stopping conditions, and information saving section.\\
-\textbf{BA}: Helped write the uncertainty estimation, stopping conditions, and output analysis sections and created figure ~\ref{fig:convergence_forward_reverse}.
+\textbf{BA}: Helped write the uncertainty estimation, stopping conditions, and output analysis sections and created figure ~\ref{fig:convergence_forward_reverse}.\\
 \textbf{JM}: Contributed to sections~\ref{subsec:reproducible},~\ref{sec:prerequisites},~\ref{} 4.2, 5.3, 6, 8.3.1, 8.3.2\\
-\textbf{JDC}: Wrote section~\ref{sec:decorrelating-samples} and~\ref{sec:automatic-equilibration-detection} discussed structure and design of the whole document, suggested Figs.~\ref{fig:fig_what_is_alchemy} and~\ref{fig:fig_sampling_scheme} \todo[inline, color={red!20}]{@JDC check info}\\
-\textbf{HX}: Contributed section~\ref{subsec:accuracy}, to section~\ref{sec:relative-fe-protocol}, and to section~\ref{sec:are-they-good}.
-
+\textbf{JDC}: Wrote section~\ref{sec:decorrelating-samples} and~\ref{sec:automatic-equilibration-detection} discussed structure and design of the whole document, suggested Figs.~\ref{fig:fig_what_is_alchemy} and~\ref{fig:fig_sampling_scheme} \\
+\textbf{HX}: Contributed section~\ref{subsec:accuracy}, to section~\ref{sec:relative-fe-protocol}, and to section~\ref{sec:are-they-good}.\\
 % We suggest you preserve this comment:

### 2020-01-20 paper/manuscript.tex
small changes

@@
 \begin{figure}[ht!]
-\includegraphics[width=0.95\columnwidth]{figures/fig12_overlap/Figure.pdf}
+\includegraphics[width=0.90\columnwidth]{figures/fig12_overlap/Figure.pdf}
 \caption{\label{fig:overlap} \textbf{Overlap matrices:} Visualising overlap matrices can help with assessing the quality of simulation data.
 (\textbf{A}) shows good overlap with all first off-diagonal entries well above 0.03, the suggested threshold, (\textbf{B}) is an example of mediocre overlap with good overlap at lower $\lambda$ values and poor overlap at high $\lambda$ values.
-(\textbf{C})shows poor overlap resulting in disconnected simulations that should not be used for an MBAR estimation of free energies.}
+(\textbf{C}) shows poor overlap resulting in disconnected simulations with unreliable MBAR estimates.}
 \end{figure}

### 2020-01-21 paper/manuscript.tex
typos

@@
     \includegraphics[width=0.95\linewidth]{figures/fig7_what_is_lam/Figure.pdf}
-    \caption{Alchemical intermediates are createtd by making the potential energy depend on a an additional variable $\lambda$ that interpolates between the chemical endpoints.
+    \caption{Alchemical intermediates are created by making the potential energy depend on a an additional variable $\lambda$ that interpolates between the chemical endpoints.
 In (\textbf{A}), at $\lambda=0$ the molecule is a fully interacting phenol and at $\lambda=1$,  a fully interacting benzene.

### 2020-01-23 paper/manuscript.tex
Updating figure 9

@@
 Arrows indicate the direction of the perturbation.
-Green cycles indicate good cycle closure, red poor cycle closure.
+Fully converged binding free energy calculations yield binding free energy changes which sum to zero around any closed cycle.
+However, in practice errors may not sum to zero around closed cycles, providing a way to look for potential sampling problems.
+Here in (B), green cycles indicate cycles with hypothetically good cycle closure, red those with poor cycle closure.
 The red arrow indicates a poorly converged simulation that would give rise to bad cycle closures.
-The diamond indicates the use of a crystal structure binding mode.}
+The diamond indicates the use of a crystallographic binding mode.}
     \label{fig:fig_types_of_networks}
@@
     \includegraphics[width=0.95\linewidth]{figures/fig7_what_is_lam/Figure.pdf}
-    \caption{Alchemical intermediates are created by making the potential energy depend on a an additional variable $\lambda$ that interpolates between the chemical endpoints.
+    \caption{Alchemical intermediates are created by making the potential energy depend on an additional variable $\lambda$ that interpolates between the chemical endpoints.
 In (\textbf{A}), at $\lambda=0$ the molecule is a fully interacting phenol and at $\lambda=1$,  a fully interacting benzene.
-(\textbf{B}) shows an illustration of the probability distribution of the potential energies as the switching function takes values of $\lambda=0$ to $\lambda=1$, intermediates are required for a sufficient overlap in potential energies to estimate a free energy difference between $\lambda=0$ and $\lambda=1$.
-    One of the most efficient families of intermediate pathways  are softcore potentials, whose shape is illustrated in (\textbf{C}) with dependence on $\lambda$ and (\textbf{D}) with dependence on the potential energy.
-Note how when $\lambda$ approaches 0, the energy smoothly approaches zero at all $r$, a necessary requirement for efficient calculations.
+(\textbf{B}) shows an illustration of the probability distribution of the potential energies as the switching function takes values of $\lambda=0$ to $\lambda=1$.
+Intermediates states are required for a sufficient overlap in potential energies to estimate a free energy difference between $\lambda=0$ and $\lambda=1$.
+    Softcore potentials provide one of the most efficient families of intermediate pathways, with a $\lambda$ dependence illustrated in (\textbf{C}) and a potential energy dependence shown in (\textbf{D}).
+Note how as $\lambda$ approaches 0, the energy smoothly approaches zero at all $r$, a necessary requirement for efficient calculations.
 }
@@
 \begin{equation}
-    U(\vec{r_{ij}},\lambda) = 4\epsilon_{ij} \lambda \left(\frac{1}{(\alpha(1-\lambda) + (r_{ij}/\sigma_{ij})^6)^2} +  \frac{1}{\alpha(1-\lambda) + (r_{ij}/\sigma_{ij})^6}\right)
+    U(\vec{r_{ij}},\lambda) = 4\epsilon_{ij} \lambda \left(\frac{1}{(\alpha(1-\lambda) + (r_{ij}/\sigma_{ij})^6)^2} -  \frac{1}{\alpha(1-\lambda) + (r_{ij}/\sigma_{ij})^6}\right)
     \label{eq:softcore}
@@
 We can't turn off the electrostatics linearly at the same time we turn off the Lennard-Jones terms, as it would leave infinitely large attractive and repulsive electrostatic terms "bare" at small $\lambda$, resulting in the simulation crashing.
-It \textit{is} possible to apply the same soft core approach to the Coulomb interaction, but it turns out to be somewhat difficult to turn off two soft-core terms simultaneously~\cite{steinbrecher2011softcore}.
-We therefore recommend performing the transformations in sequence; first, turning off all electrostatics for atoms that must be removed, inserting and removing Lennard-Jones sites (both the insertion and removal can be done simultaneously), and then turning electrostatics for the introduced particles on.
-However, If there are no removals or introduction to atomic sites, then it is reasonable to change all interactions linearly.
-%
-Other issues, such as whether absolute calculations should retain or remove intramolecular non-bonded interactions, may be less critical and recommendations differ among studies in the literature~\cite{}, and reasonable efficiency can be often obtained with either choice even if some are somewhat better or worse than others.
+It \textit{is} possible to apply the same soft core approach to the Coulomb interaction and this is indeed done in a number of implementations, in which case it is important that the Coulomb interaction is softened more rapidly than the Lennard-Jones interaction to avoid charge penetration issues, this can be tricky to ensure for all types of perturbations ~\cite{steinbrecher2011softcore}.
+A safe but potentially more computationally expensive approach is to perform the transformations in sequence; first, turning off all electrostatics for atoms that must be removed, inserting and removing Lennard-Jones sites (both the insertion and removal can be done simultaneously), and then turning electrostatics for the introduced particles on.
+Again, If there are no removals or introduction to atomic sites, then it is reasonable to change all interactions linearly.
+%
+Other issues, such as whether absolute calculations should retain or remove intramolecular non-bonded interactions
+through either annihilation~\cite{hermans1997inclusion, mann2000modeling, boresch2003absolute, wang2006absoulte, mobley2006use} or decoupling~\cite{fujitani2005direct, mobley2006use} must be considered.
+Reasonable efficiency can be often obtained with either choice even if some are somewhat better or worse than others although there is not consensus on which is better in all situations.
 Our recommendation is to leave the intramolecular interactions on during the transformation for simplicity if there are no other known issues with this approach.
 The key thing to watch out for is whether the total potential energy, and therefore the intermediate ensembles sampled, change smoothly from beginning to end.
-If a configurational phase transition is noted in the process (like substantial dihedral rearrangement), then other steps must be taken, perhaps including modifying intramolecular interactions~\cite{???}.
 These problems can be diagnosed by noticing lack of configuration space overlap between different simulations (see section~\ref{sec:are-they-good})"
 %
-Relative calculations introduce additional choices, including whether to define explicit intermediate states~\cite{} or leave these implicitly defined by the code~\cite{}.
-Typically in single topology relative calculations it proves most efficient to first remove electrostatic interactions of any atoms which will be deleted, then modify other non-bonded interactions, then restore electrostatic interactions of any atoms which are being inserted.
-Other schemes, such as simultaneously changing electrostatic and Lennard-Jones interactions, even with electrostatic ``soft-core'' potentials~\cite{steinbrecher2007nonlinear}, in our experience typically introduce errors and/or instabilities or are at least unreliable.
-It is expected that similar considerations will apply to dualtopology calculations, and the principle of first removing electrostatics and then removing steric interactions will likely serve well.
+Relative calculations introduce additional choices, such as the order in which to modify nonbonded interactions.
+A common process in single topology relative calculations is to first remove electrostatic interactions of any atoms which will be deleted, then modify other non-bonded interactions, then restore electrostatic interactions of any atoms which are being inserted.
+Although this is a simpler path to understand cognitively and can take advantage of the soft-core potential from eq~\ref{eq:softcore}, this can lead to more intermediate steps and thus be more computationally expensive.
+Other schemes, such as simultaneously changing electrostatic and Lennard-Jones interactions with electrostatic ``soft-core'' potentials~\cite{steinbrecher2007nonlinear} may be implemented with fewer intermediate but could require fine-tuning of electrostatic and Lennard-Jones softcore parameters to avoid numerical instabilities.
+At the time of writting, there has not been conclusive evidence to suggest one approach is better in general than the other, so discretion should be left up to the user as to what is viable from both hardware resources, and what the simulation software supports.
 %
@@
 The spacing depends to some extent on the choice of analysis method, though states should essentially be spaced equidistant in the relevant thermodynamic length~\cite{crooks2007measuringa, sivak2012thermodynamic}.
-For BAR/MBAR techniques this means that spacings should typically be equal in [what, variance? ref].
-Some schemes to adaptively optimize the spacing of intermediate states based on initial exploratory simulations have been proposed~\cite{}.
-%
-An example of this approach for relative binding free energy calculations includes saving configurations from short 10 picosecond simulations starting at $\lambda_0$, and evaluating changes in the standard error of mean (SEM) in $\frac{dU}{d\lambda}$ as reevaluated at new values of $\lambda$, starting with a step size of 0.01 up to a predetermined threshold.
-This threshold can be empirically defined as the number of atoms being inserted or deleted in the simulation / 100.
-If the absolute value of the difference in SEM is below the defined threshold, then the change in $\lambda$ for re-evaluation of SEM in $\frac{dU}{d\lambda}$ is defined by the current difference in SEM from the threshold divided by the derivative of the SEM with respect to $\lambda$.
+For BAR/MBAR techniques this means that state spacing should typically be equal in variance; along a minimal variance path, this is equadistance in $\lambda$, but most paths are not truly minimal variance.~\ref{pham2012optimal, shenfeld2009minimizing}
+Some schemes to adaptively optimize the spacing of intermediate states based on initial exploratory simulations have been proposed~\cite{hayes2017adaptive}.
+
 %
@@
 There are different easily measurable indicators that can be looked at to test how well converged simulations are and if all alchemical states have been sufficiently sampled for a rigorous analysis.
-Furthermore, once you have established that individula perturbations are well behaved there are some tricks to ensure the overall perturbation network gives reliable results.
+Furthermore, once you have established that individual perturbations are well behaved, there are some tricks to ensure the overall perturbation network gives reliable results.
 %
@@
 Fig.~\ref{fig:freeenergytrajectories} illustrates how looking at the convergence of your data may be important.
-In this example the same guest G3 shows different convergence behaviour for two different hosts.
-CB8 with guest G3 has a longer correlation time than the octa acid (OA) host.
-In some cases slow correlation time may not be expected and therefore not a feature known in advance.
-To this end you should always look at all simulation data available check convergence behaviour for each free energy estimate and if need be extend existing simulations or try an approach that requires simulations in two separate binding modes where they interconvert at very slwo timescales.
+In this example, the guest G3 shows different convergence behaviour for two different hosts.
+The CB8 host with guest G3 has a longer correlation time than the octa acid (OA) host.
+In some cases, slow correlation time may not be expected and therefore not a feature known in advance.
+To this end, you should always look at all simulation data available check convergence behaviour for each free energy estimate and if need be extend existing simulations or try an approach that requires simulations in two separate binding modes where they interconvert at very slow timescales.
 \begin{figure}
@@
 Shaded areas represent 95\% confidence intervals around the mean computed from the 5 replicates data.
-The horizontal dashed lines show the final binding free energy prediction of the two calculations after a total of ~5230 ns for OA-G3 and ~6650 ns for CB8-G3.
+The horizontal dashed lines show the final binding free energy prediction of the two calculations after a total of 5230 ns for OA-G3 and 6650 ns for CB8-G3.
 Longer correlation times in CB8-G3 cause the calculation to converge more slowly.
@@
 $\lambda$s.
-Sufficient overlap is important for reweighting estimators such as BAR or MBAR, but cannot help assess reliability of estimates, when using TI.
+Sufficient overlap is important for reweighting estimators such as BAR or MBAR, but cannot help assess reliability of estimates when using TI.
 These matrices are graphical representations of the phase space overlap, i.e.
@@
 Specifically, often the goal is to adjust the $\lambda$ spacing to perform additional sampling in the areas of poor overlap, rather than using arbitrarily spaced windows.
-Different approaches are described in section~\ref{sec:simulation_protocol_choice}) and more details can be found in the literature~\cite{dakka2018concurrent, hahn2019alchemical}.
+Different approaches are described in section~\ref{sec:simulation_protocol_choice} and more details can be found in the literature~\cite{dakka2018concurrent, hahn2019alchemical}.
 %
@@
 Though no firm guidelines have emerged, it may be judicious to perform additional configurational sampling along edges of a network that are involved in cycles closing poorly.
-This may be done by extending simulations duration, or by averaging free energy changes over multiple repeats.
+This may be done by extending the duration of simulations, or by averaging free energy changes over multiple repeats.
 The latter approach may yield more reproducible free energy changes, but at the expense of a stronger bias on the estimated free energies due to repeated use of the same input coordinates.
@@
 A scheme to reduce cycle closure errors is used in FEP+ whereby calculated free energy changes along the nodes of the network are re-sampled assuming estimates of the calculated free energy change along a node may be obtained from a Gaussian distribution centered on the estimated free energy change and with a standard deviation equal to the estimated standard deviation of the free energy change.
-The procedure then uses a Maximum likelihood method to find new sets of free energy changes that minimize cycle closure errors~\cite{wang2013modeling}.
+The procedure then uses a maximum likelihood method to find new sets of free energy changes that minimize cycle closure errors~\cite{wang2013modeling}.
 An alternative approach computes the free energy change between a target and reference compound as a weighted average over all unique paths in the network, with the weights derived from the propagated uncertainties of each node~\cite{mey2016blinded}.
@@
 Alchemical free energy calculations have seen a vast increase in popularity also in the pharmaceutical industry for structure based drug discovery~\cite{schindler2020largescale, sherborne2016collaborating, wagner2017computational}.
-Commercial products such as FEP+ and Flare, which provide a convenient user interface make the setup and use of these methods a lot easier~\cite{wang2015accurate}.
-Prospective prediction challenges such as the drug design data resource grand challenges provide a community driven platform to evaluate different free energy protocols against each other on unknown targets~\cite{gaieb2018d3r, gaieb2019d3r}.
-What has come out of these in the past is that SChrödinger's FEP+, performs similar when run by different participant groups~\cite{gaieb2018d3r}, but may not give the desired accuracy with RMSE errors larger than 1 kcal/mol.
-Other alchemical protocols performed similarly well.
-We hope that the best practice guide provides a set of tools that allow a better understanding of how to setup, run and interpret alchemical free energy calculations.
-The choice of the underlying simulation software is left to the user, but hopefully with this guide it is easier to assess whether the generated data is trustworthy or not.
+Commercial products such as FEP+ and Flare, which provide a convenient user interface make the setup and use of these methods a lot easier~\cite{wang2015accurate}, but this convenience comes with less flexibility on the simulation protocols.
+It is also important to understand the current limitations of the methodology to recognise when automated worflow tools can be used effectively for a given protein target.
+Prospective prediction challenges such as the drug design data resource grand challenges provide a community driven platform to evaluate different free energy protocols against each other on unknown targets~\cite{gaieb2018d3r, gaieb2019d3r}.Such efforts have highlighted that selection of seemingly identical or similar potential energy function or simulation package does not guarantee production of similar free energies owing to differences in simulation protocols.
+(SUGGEST REMOVING THIS SENTENCE What has come out of these in the past is that SChrödinger's FEP+, performs similar when run by different participant groups~\cite{gaieb2018d3r}, but may not give the desired accuracy with RMSE errors larger than 1 kcal/mol.
+Other alchemical protocols performed similarly well.) We hope that the best practice guide provides a set of tools that allow a better understanding of how to setup, run and interpret reliably alchemical free energy calculations.
 %
@@
 \textbf{MRS}: Helped create figure~\ref{fig:fig_what_is_lambda}, wrote sections describing choices for alchemical pathways and analysis for free energy calculations.\\
-\textbf{LNN}: Helped write the simulation length, stopping conditions, and information saving section.\\
+\textbf{LNN}: Helped write the simulation length, stopping conditions, and information saving section. Edited and reviewed alchemical path section.\\
 \textbf{BA}: Helped write the uncertainty estimation, stopping conditions, and output analysis sections and created figure ~\ref{fig:convergence_forward_reverse}.\\
-\textbf{JM}: Contributed to sections~\ref{subsec:reproducible},~\ref{sec:prerequisites},~\ref{} 4.2, 5.3, 6, 8.3.1, 8.3.2\\
+\textbf{JM}: Contributed to sections~\ref{subsec:reproducible},~\ref{sec:prerequisites},~\ref{} 4.2, 5.3, 6, 7.2.3, 8.3.1, 8.3.2, 9\\
 \textbf{JDC}: Wrote section~\ref{sec:decorrelating-samples} and~\ref{sec:automatic-equilibration-detection} discussed structure and design of the whole document, suggested Figs.~\ref{fig:fig_what_is_alchemy} and~\ref{fig:fig_sampling_scheme} \\
@@
 Declare any potentially conflicting interests here, whether or not they pose an actual conflict in your view.
-MK is employed by Cresset who recently released an commercial tool for performing alchemical free energy calculations.
+MK is employed by Cresset who recently released a commercial tool for performing alchemical free energy calculations.
 
@@
 ASJSM and JM acknowledge funding through an EPSRC flagship software grant: EP/P022138/1
-MK acknowledges funding through Innovate UK by KTP partnership 011120.
+MK and JM acknowledges funding through Innovate UK by KTP partnership 011120.
 AR acknowledges partial support from the Tri-Institutional Program in Computational Biology and Medicine and the Sloan Kettering Institute.

### 2020-01-23 paper/manuscript.tex
updatng figure for convergence

@@
     Each plot shows the estimated free energy change using a varying fraction of total simulation time (up to 5 ns total). 
-    The top panel shows the three-step protocol for a perturbation involving 5 perturbed atoms, while the bottom panel shows the same protocol for a perturbation involving 10 perturbed atoms.
+    Subplots \textbf{A}, \textbf{C}, and \textbf{E} show a three step protocol  for a perturbation involving 3 perturbed atoms, while \textbf{B}, \textbf{D}, and \textbf{F} shows the same protocol for a perturbation involving 10 perturbed atoms.
+The first step of the protocol is the recharging then removing van der Waals interactions and then recharging.
 The difference in energy between the forward (blue) and reverse (red) free energy calculations at the midpoint of the simulation time gives an indication of the overall convergence of the simulation, with differences over 1 k$_{B}$T indicating poor convergence.}

### 2020-01-27 paper/manuscript.tex
typo fix

@@
 Alchemical free energy calculations compute free energy differences associated with transfer processes, such as the binding of a small molecule to a receptor, the transfer of a small molecule from an aqueous to apolar phase~\cite{zwanzig1954hightemperature}, or effects of protein side chain mutations on e.g.
-binding affinities..
+binding affinities.
 These calculations use non-physical\footnote{Here, the non-physical nature of the transformation is referred to as "alchemical", a term coined by Tembre and McCammon in Ref.~\cite{tembre1984ligandreceptor}.} intermediate states in which the chemical identity of some portion of the system (such as a small molecule ligand or receptor sidechain) is changed by modifying the potential governing the interaction of the atoms being modified, inserted, or deleted with their environment.
@@
 In alchemical simulations, the introduction of intermediate \textit{alchemical states} that bridge the high-probability regions of configuration space between the two physical endstates of interest permits the free energy of large transformations to be robustly computed.
-Alchemical calculations can be used in a variety of scenarios, such as computing the free energy of a conformational change for a molecule with a high barrier to interconversion (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{A}));
+Alchemical calculations can be used in a variety of scenarios, such as: 
+\begin{itemize}
+\item computing the free energy of a conformational change for a molecule with a high barrier to interconversion (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{A}));
 \todo[inline, color={yellow!20}]{JDC: Do you intend these to be representative of BLUES type simulations? -- yes}
-computing partition ($\log P$) or distribution ($\log D$) coefficients between environments (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{B}))~\cite{rustenburg2016measuring, bosisio2016blinded} or compartments or membranes (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{C}))~\cite{corey2019insights}.
-Furthermore, alchemical calculations are frequently used to estimate changes in free energies upon modifying a ligand or protein: a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{D}))\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}; the entire ligand can be alchemical transferred from protein to solvent in an absolute binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{E}))~\cite{mobley2007predicting,aldeghi2015accurate,aldeghi2017predictions}; and small alchemical modifications can be made between chemically related ligands to estimate differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{F}))~\cite{wang2015accurate,mey2016blinded,gapsys2020large}.
-%
-Once an alchemical simulation as been performed, the data must be analyzed to estimate the free energy for the transformation of interest.
-Early work used simple (but suboptimal) estimators for this: free energy perturbation (FEP) using the Zwanzig relation~\cite{zwanzig1954hightemperature} or thermodynamic integration (TI), for which the theory dates back decades but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamics, merz1989free}.
+\item computing partition ($\log P$) or distribution ($\log D$) coefficients between environments (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{B}))~\cite{rustenburg2016measuring, bosisio2016blinded} 
+\item or between compartments or membranes (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{C}))~\cite{corey2019insights}. 
+\end{itemize}
+
+Furthermore, alchemical calculations are frequently used to estimate changes in free energies upon modifying a ligand or protein: 
+\begin{itemize}
+\item a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{D}))\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate};
+\item the entire ligand can be alchemical transferred from protein to solvent in an absolute binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{E}))~\cite{mobley2007predicting,aldeghi2015accurate,aldeghi2017predictions};
+\item and small alchemical modifications can be made between chemically related ligands to estimate differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{F}))~\cite{wang2015accurate,mey2016blinded,gapsys2020large}.
+\end{itemize}
+%
+Once an alchemical simulation procedure is performed (which usually involves more than one simulation), the data must be analyzed to estimate the free energy for the transformation of interest.
+Early work used simple but statistically suboptimal estimators for this: free energy perturbation (FEP) using the Zwanzig relation~\cite{zwanzig1954hightemperature} or thermodynamic integration (TI), for which the theory dates back decades but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamics, merz1989free}.
 %
 More recent developments have seen new, highly efficient statistical estimators that make better use of all the data, often building on the Bennett acceptance ratio (BAR)~\cite{bennett1976efficient}, producing multistate generalizations~\cite{shirts2008statisticallya} or removing the need for global equilibrium~\cite{wu2016multiensemble, mey2014xtram, wu2014statistically}
-\todo[inline, color={yellow!20}]{JDC: The key pivotal moment was the Shirts paper on extremely precise free energy calculations, which demonstrated that a combination of technologies can produce robust estimates with useful precision.}
-Important studies continued during the 2000's often leading to improved implementation of the methods in common simulation software~\cite{vanderspoel2005gromacs, mermelstein2018fast, wang2015accurate, hedges2019biosimspace}.
+\todo[inline, color={yellow!20}]{JDC: The key pivotal moment was the Shirts paper on extremely precise free energy calculations, which demonstrated that a combination of technologies can produce robust estimates with useful precision.
+MRS: aw, shucks!
+added.}
+Important studies continued during the 2000's often leading to improved implementation of the methods in common simulation software~\cite{shirts2003extremely,shirts2005solvation,vanderspoel2005gromacs, mermelstein2018fast, wang2015accurate, hedges2019biosimspace}.
 This foundational work, combined with the methodological, technological, and hardware improvements of the last 5--10 years, have led to an explosion of interest and direct commercial application of the technologies~\cite{wang2015accurate, fratev2019improved, schindler2020largescale, cournia2017relative, sherborne2016collaborating}.
@@
 In the meantime, today's users may find it hard to get started with these complex calculations whilst also keeping up with the fast pace of change.
-This Best Practice guide aims to serve as a go-to reference for current recommendations and tips for users of all experience.
+This Best Practice guide provides current recommendations and tips for users of all experience.
 %
@@
     \includegraphics[width=0.95\linewidth]{figures/fig1_what_is_alchemy/Figure.pdf}   
-    \caption{\textbf{Illustration of common types of free energies differences that can be calculated using alchemical free energy methods.} \textbf{A}: Change in free energy due to a conformational change of the molecule.
+    \caption{\textbf{Illustration of common types of free energies differences that can be calculated using alchemical free energy methods.} \textbf{A}: Change in free energy due to a conformational change of the molecule across a high barrier.
 \textbf{B}: Partition coefficient such as $\log P$ or $\log D$ depend on a change in free energy between different phases; here, as an example the partition coefficient between methanol and water is shown.
@@
 \todo[inline, color=green!20]{ASJSM: @everyone This section is up for review to also check flow with intro and rest of document.}
-This guide is meant to serve as a comprehensive starting point for new practitioners for free energy calculations.
-It assumes moderate experience with molecular simulation packages and concepts, however.
-It is also intended to serve as a reference guide with a convenient checklist (Section~\ref{sec:checklist}) for more seasoned practitioners, to help ensure their simulation meet standard alchemical simulation and analysis practices.
+This guide is meant to serve as a comprehensive starting point for new practitioners of free energy calculations.
+However, it assumes moderate experience with molecular simulation packages and concepts.
+It is also intended to serve as a reference guide with a convenient checklist (Section~\ref{sec:checklist}) for more seasoned practitioners, to help ensure their calculations meet standard alchemical simulation and analysis practices.
 It can also serve as a set of best practices to ensure simulation robustness and reproducibility which reviewers may wish to consider as they evaluate papers.
-%
+Importantly, it is not, a review of all free energy calculation methods at the cutting edge of current research.
+
 This guide aims to answer the following questions:
@@
 We assume the reader possesses a basic familiarity with the principles of molecular mechanics, molecular dynamics simulations, statistical mechanics, and the biophysics of protein-ligand association.
-If you feel unfamiliar with some of these concepts please good starting points can be found in these references~\cite{braun2019best, grossfield2018best, klimovich2015guidelines, shirts2012best}.
-Oftentimes, docking calculations are required to generated initial binding pose for a small molecule for an alchemical calculation, where binding poses are not known, thus some basic familiarty on how to run docking calculations is also expected~\cite{grinter2014challenges}.
-%
-As some of the theoretical background can seem daunting, we provide an essentials guide to the theory behind alchemical free energy calculations in Section~\ref{sec:theory}.
+If you feel unfamiliar with some of these concepts, good starting points can be found in these references~\cite{braun2019best, grossfield2018best, klimovich2015guidelines, shirts2012best}.
+
+Other background information may be needed depending on the project.
+For example, often, if binding poses are not known, docking calculations can be used to generate an initial small molecule binding pose to start alchemical calculations, thus some basic familiarity on how to perform docking is also necessary for such projects~\cite{grinter2014challenges}.
+%
+As some of the theoretical background can seem daunting, in this best practices description, we do provide a guide to the essential theory behind alchemical free energy calculations in Section~\ref{sec:theory}.
+
 In the remainder of this paper, we will cover topics that are key to the preparation~(\ref{sec:prerequisites}), choice and use of correct protocols~(\ref{sec:simulation_protocol_choice}), and finally the best practices that should be used for the analysis of alchemical calculations~(\ref{sec:data_analysis}).
@@
 %
+
 While we try to address as many methods and practices as possible, the field of free energy calculations is broad, and there are many advanced topics that are left to future best practices documents focusing on specific issues.
@@
 \item Complex association stoichiometry (beyond simple 1:1 association).
-\item Endpoint free energy methods such as MM-PBSA~\cite{genheden2015mm} and LIE~\cite{gutierrez-de-teran2012linear}
+\item Approximate endpoint free energy methods such as MM-PBSA~\cite{genheden2015mm} and LIE~\cite{gutierrez-de-teran2012linear}
 \item Free energy methods that extract the ligand using geometric order parameters and potential of mean force methods~\cite{heinzelmann2017attachpullrelease}
 \item Forcefield dependence for protein, ligand, ions, co-solvents, and co-factors.
-Different studies have looked at the influence of force fields and it is assumed the user has made an adequate choice for the system under study.~\cite{loeffler2018reproducibility, vassetti2019assessment, lopes2015current}
+A number of different studies have looked at the influence of force fields and it is assumed the user has made an adequate choice for the system under study.~\cite{loeffler2018reproducibility, vassetti2019assessment, lopes2015current}
 \end{itemize}
@@
 $\Delta f$-- reduced free energy \\
+$K_b$ -- binding or association constant \\
 $k_B$-- Boltzmann constant \\
-$T$ -- temperature
+$T$ -- temperature \\
+$\beta$ -- (k_BT)^{-1}
 $k_b$ -- rate of binding \\
 $K_D$ -- dissociation constant \\
-$\lambda$ -- parameter for regulation of the alchemcial free energy path \\
+$\lambda$ -- parameter for progress of the alchemical free energy path \\
 $g$ -- statistical inefficiency
@@
 \begin{tcolorbox}[title=Common terminology, colback=yellow!10!white]
-    ligand - small molecular binding to a larger host molecule such as a protein (recpetor).\\ 
-    lead optimization - process of designing a drug candidate from an initial lead compound, considering its affinity and various pharmacokinetic parameters.\\
-    thermodynamic cycle - sequence of thermodynamic processes (like changes in temperature, volume or pressure) in which the initial and final state are equal.\\
+    \emph{ligand} - small molecule binding to a larger host molecule such as a protein (receptor).\\ 
+    \emph{lead optimization} - process of designing a drug candidate from an initial lead compound, usually involving optimizing its affinity, possibly selectivity, and various pharmacokinetic parameters.\\
+    \emph{thermodynamic cycle} - sequence of thermodynamic processes (like changes in temperature, volume or pressure) in which the initial and final state are equal.\\
 \end{tcolorbox}
@@
 %%%%%%%%%%%%%%%%%%%%
-\section{Statistical mechanics can tell us why alchemical free energy calculations work}
+\section{Statistical mechanics demonstrates why alchemical free energy calculations work}
 \label{sec:theory}
 Why would you want to run an alchemical free energy calculation and why do they work?
-Suppose you want to compute the binding affinity, or free energy of binding, of a ligand $L$ to a protein $P$, given by:
+In this section, we use the example of relative free energy calculation to sketch the theory of alchemical calculations and illustrate their utility.
+The emphasis here is placed on bridging theoretical foundations and intuition.
+A rigorous derivation of the standard (absolute) free energy of binding using the principles of statistical mechanics can be found in Gilson's classic work~\cite{gilson1997statisticalthermodynamic}.
+
+\subsection{Simulating binding events of receptor-drug systems can be computationally expensive}
+
+Suppose you want to compute the binding affinity, or free energy of binding, of a ligand $L$ to a receptor $R$, given by:
 \begin{equation}
-P+L\leftrightharpoons PL
-\end{equation}
-The equilibrium constant  ($K_b^{\circ}$) is given by the ratio of concentrations of product $[PL]$ and reactants $[P]$, $[L]$:
+R+L \leftrightharpoons RL
+\end{equation}
+The binding constant  ($K_b^{\circ}$) is given by the law of mass action as the ratio of concentrations of product $[RL]$ and reactants $[R]$, $[L]$:
 \begin{equation}
- K_b^{\circ} = c^{\circ}\frac{[PL]}{[L][P]}.
-\end{equation}
-In order to define a reference state, the standard state is usually defined as $1 \mathrm{mol}/\mathrm{L}$ and $c^\circ$ is the standard state concentration.
+ K_b^{\circ} = c^{\circ}\frac{[RL]}{[L][R]}.
+\end{equation}
+The standard state concentration $c^\circ$ depends on the reference state, but it is usually set to $1\,\mathrm{mol}/\mathrm{L}$ assuming a constant pressure of $1\,\mathrm{atm}$ (see also Section~\ref{sec:standardstate-restraints}).
 Thus, the Gibbs free energy of binding $\Delta G_{\mathrm{bind}}$ is given by:
 \begin{equation}
-    \Delta G_{\mathrm{bind}} = -k_BT\ln K_b^{\circ},
+    \Delta G_{\mathrm{bind},L} = -k_BT\ln K_b^{\circ},
 \end{equation}
 where $k_B$ is the Boltzmann constant and $T$ the temperature of the system.
-One way to estimate the equilibrium constant is by computing the probability of finding the molecular system in the bound state ($PL$) or the unbound state ($P+L$), making use of partition function ratios, and assuming that the contribution of water compressibility at 1 atm has a negligible effect on the free energy of binding~\cite{shirts2012best, gilson1997statisticalthermodynamic} i.e.
-
+
+\paragraph{The free energy of binding can be expressed as a ratio of partition functions}
+A natural though generally very computational expensive way to estimate the equilibrium constant is by directly simulating several binding and unbinding events and computing the probability of finding the receptor-ligand system in the bound state, $P(RL)$, or the unbound state, $P(R+L)$.
+Assuming the volume change upon binding to be negligible, which is often the case at $1~\mathrm{atm}$ due to the incompressibility of water, then the Gibbs free energy $\Delta G_{\mathrm{bind}, L}$ is approximately equal to the Helmholtz free energy $\Delta A_{\mathrm{bind}, L}$, and we can simulate the system in a box of volume $V$ to obtain~\cite{jong_determining_2011}
+\begin{equation}\label{eq:binding-free-energy-from-bound-unbound-probability-ratio}
+    \Delta G_{\mathrm{bind},L} \approx \Delta A_{\mathrm{bind}, L} = - k_B T  \left( \mathrm{ln} \frac{P(RL)}{P(R+L)} + \ln \left( c^\circ N_{\mathrm{Av}} V \right) \right) \, ,
+\end{equation}
+where $N_{\mathrm{Av}}$ is the Avogadro number, and the last term corrects for the simulated concentration being different than the standard concentration.
+Let $\Gamma_{\mathrm{bound}}$ and $\Gamma_{\mathrm{unbound}}$ be the set of receptor-ligand conformations $\vec{q}$ that we consider bound and unbound respectively.
+The probability of a conformation $\vec{q}$ is given by the Boltzmann probability density function
 \begin{equation}
-    \Delta G_{\mathrm{bind}} \cong \Delta F= -k_BT\ln\frac{\int_\mathrm{bound} \exp(-\beta U(\vec{q}))d\vec{q}}{\int_\mathrm{unbound} \exp(-\beta U(\vec{q}))d\vec{q}},
-\end{equation}
-where $\beta = \frac{1}{k_BT}$ and $F$ is the Helmholtz free energy. 
-Or more generally in a simplified form the free energy difference ($\Delta f$) between the bound and unbound states is:
+    P(\vec{q}) = \frac{\exp\left( \beta U(\vec{q}) \right)}{\int_{\Gamma} \exp\left( \beta U(\vec{q}) \right) \, d\vec{q}} \, ,
+\end{equation}
+where $\beta = (k_B T)^{-1}$ is the inverse temperature, $U(\vec{q})$ is the potential energy of conformation $\vec{q}$, and the integration is over the set of all possible conformations accessible in the simulation box volume $\Gamma$, with $\Gamma_{\mathrm{bound}},\Gamma_{\mathrm{unbound}} \subset \Gamma$.
+If the simulation is long enough, we expect the fraction of conformations $\vec{q}$ found in the bound state to converge to
 \begin{equation}
-\Delta f \equiv f(\mathrm{bound}) - f(\mathrm{unbound}) = -\ln\frac{Z(\mathrm{unbound})}{Z({\mathrm{bound}})},
-\end{equation}
-where $Z = \int_\mathrm{bound} \exp(-\beta U(\vec{q}))d\vec{q}$ represents the configurational integral. 
-Unfortuantely it is not possible to evaluate the integral required for the configurational integral exactly, but can be approximated using molecular dynamics simulations. 
-So, how can we compute $\Delta G_{\mathrm{bind}}$ from a molecular dynamics simulation?
-Running a direct simulation of the binding events is too computationally costly, meaning we cannot easily estimate the configurational integrals of the bound and unbound state from the simulation, as typically binding events happen at a $\mu$s scale, and simulations reach around 1 $\mu$s/day.
-For a reliable estimate of a single free energy of binding many days of simulations would be required, making these type of calculations uninteresting at a pharamaceutical timescale.
-Instead, we can run alchemical free energy simulations, by making use of the fact that free energy is a state function.
-We can write down a thermodynamic cycle as seen in ~\ref{fig:fig_binding_thermodynamic_cycle}, for a relative free energy simulation.
-Based on this cycle we can compute the relative free energy of binding between two compounds $A$ and $B$, by going along the arrow of $\Delta G_{\mathrm{bound}}$, where we introduce a second compound that is similar to the first one.
-Now $\Delta G_{\mathrm{bound}}$ can be expressed as a ratio of partition functions of toluene (A) and benzyl alcohol (B).
-The defining characteristic of alchemical free energy calculations is the use of a series of alchemically-modified potential functions $U(x; \lambda)$ in which an alchemical parameter $\lambda$ modulates interactions in a manner that cannot occur in real chemical systems.
-One or more simulations are used to collect data from a multitude of alchemical states to compute a free energy difference between a chemical state ($\lambda_0$) and another chemical or alchemical reference state ($\lambda_1$),
-\begin{eqnarray}
-\Delta f &\equiv& f(\lambda_1) - f(\lambda_0) = - \ln \frac{Z(\lambda_1)}{Z(\lambda_0)} , \label{equation:dimensionless-free-energy-difference}
-\end{eqnarray}
-where the dimensionless free energy $f(\lambda) \equiv \beta F(\lambda)$ is given in terms of partition functions $Z(\lambda)$,
-\begin{eqnarray}
-Z(\lambda) &=& \int_{\Gamma} d\vec{q} \, \exp({-u(\vec{q}; \lambda)}) ,
-\label{equation:partition-function-definition}
-\end{eqnarray}
-, where $\Gamma$ signifies the phase space volume over which to integrate. 
-The \emph{reduced potential} $u(\vec{q}; \lambda)$~\cite{shirts2008statisticallya} is generally given by a trace over thermodynamic parameters with their conjugate dynamical variables,
-\begin{eqnarray}
-u(\vec{q};\lambda) &\equiv& \beta \left[ U(\vec{q};\lambda) + p \, V(\vec{q}) + \sum_{i=1}^N \mu_i \, N_i(\vec{q}) + \cdots \right] . \label{equation:reduced-potential}
-\end{eqnarray}
-Here, the collection of thermodynamic and alchemical parameters $\theta \equiv \{\beta, \lambda, p, \mu, \ldots\}$ defines a \emph{thermodynamic state}.
-The physical transformation of interest---such as a protein-ligand association process, or transfer free energy among phases---may involve a thermodynamic cycle that requires several free energy differences $\Delta f$ to be computed in order to produce an estimate of the overall desired quantity, see for example the thermodynamic cycle for an absolute free energy calculation as seen in~\ref{fig:fig_absolute_thermodynamic_cycle}.
+    P(RL) = \int_{\Gamma_{\mathrm{bound}}} P(\vec{q}) \, d\vec{q} = \frac{\int_{\Gamma_{\mathrm{bound}}} \exp\left( \beta U(\vec{q}) \right) \, d\vec{q}}{\int_{\Gamma} \exp\left( \beta U(\vec{q}) \right) \, d\vec{q}} \, .
+\end{equation}
+After similar considerations for $P(R+L)$, we find that the ratio of visited bound and unbound conformations, in the limit of long simulations, should converge to
+\begin{equation}\label{eq:bound-unbound-probability-ratio}
+    \frac{P(RL)}{P(R+L)} = \frac{\int_{\Gamma_{\mathrm{bound}}} \exp\left( -\beta U(\vec{q}) \right) \, d\vec{q}}{\int_{\Gamma_{\mathrm{unbound}}} \exp\left( -\beta U(\vec{q}) \right) \, d\vec{q}} = \frac{Z(RL)}{Z(R+L)} \, ,
+\end{equation}
+where we have defined the \textit{configurational integral} or \textit{partition function} $Z(\mathrm{state}) \equiv \int_{\Gamma_{\mathrm{state}}} \exp\left(-\beta U(\vec{q})\right) \, d\vec{q}$.
+
+\paragraph{Simulating binding events is computationally expensive}
+While simulating binding events has been used to estimate binding affinities~\cite{jong_determining_2011,pan2017quantitative} or to get insights into the binding pathways and kinetics of receptor-ligand systems~\cite{teo_adaptive_2016,votapka_seekr_2017,doerr_--fly_2014,plattner_protein_2015,dixon_predicting_2018}, the computational cost of these calculations is usually dominated by the rate of dissociation, which can be on the $\mu$s scale even for millimolar binders~\cite{pan2017quantitative} and reaches the ms or s scale for a typical drug~\cite{basavapathruni_conformational_2012,hyre_cooperative_2006}.
+Depending on system size and simulation settings, common molecular dynamics software packages can reach a few hundreds of ns/day using currently available high-end GPUs~\cite{eastman_openmm_2017,kutzner_more_2019}, making these type of calculations uninteresting at pharmaceutical drug discovery timescales.
+Other methods compute the free energy of binding by building potential of mean force profiles along a reaction coordinate~\cite{woo2005calculationa,velezvega_overcoming_2013,limongelli_funnel_2013,heinzelmann2017attachpullrelease}, but these methods require prior knowledge of a high-probability binding pathway, which is not easily available, especially in the prospective scenarios typical of the drug development process.
+
+\subsection{Alchemical free energy calculations yield predictions that do not require direct simulation of binding events}
+
+In many cases, the quantity of interest is the change in binding affinity when a compound $A$ is mutated to compound $B$ (e.g., by modifying one the drug scaffold's substituents), which, by using Eq.~\ref{eq:binding-free-energy-from-bound-unbound-probability-ratio} and \ref{eq:bound-unbound-probability-ratio} is given by
+\begin{equation}\label{eq:delta-delta-G-physical}
+\begin{split}
+    \Delta \Delta G_{\mathrm{bind}, AB} &= \Delta G_{\mathrm{bind}, B} - \Delta G_{\mathrm{bind}, A} \\
+    &\approx -k_BT \left( \ln \frac{Z(RB)}{Z(R+B)} - \ln \frac{Z(RA)}{Z(R+A)} \right) \, .
+\end{split}
+\end{equation}
+Note that the terms involving the standard concentration cancel out when we assume that the volume is identical for $A$ and $B$.
+Predictions of $\Delta \Delta G_{\mathrm{bind}, AB}$ with non-alchemical methods generally require long simulations of both ligands, possibly through different binding pathways.
+Alchemical relative free energy calculations avoid the need to simulate binding and unbinding events by making use of the fact that the free energy is a state function and exploiting the thermodynamic cycle illustrated in Figure~\ref{fig:fig_binding_thermodynamic_cycle}.
+This is apparent after rewriting Eq.~\ref{eq:delta-delta-G-physical} as
+\begin{equation}\label{eq:delta-delta-G-alchemical}
+\begin{split}
+    \Delta \Delta G_{\mathrm{bind}, AB} &\approx -k_BT \left( \ln \frac{Z(RB)}{Z(RA)} - \ln \frac{Z(R+B)}{Z(R+A)} \right) \\
+    &= -k_BT \left( \ln \frac{Z(RB)}{Z(RA)} - \ln \frac{Z(B)}{Z(A)} \right) \\
+    &= \Delta G_{\mathrm{bound}} - \Delta G_{\mathrm{unbound}} \, ,
+\end{split}
+\end{equation}
+where $\Delta G_{\mathrm{bound/unbound}}$ is the free energy of mutating $A$ to $B$ in the bound/unbound state.
+Eq.~\ref{eq:delta-delta-G-alchemical} and Figure~\ref{fig:fig_binding_thermodynamic_cycle} tell us that the difference in free energy of binding between toluene ($A$) and benzyl alcohol ($B$) can be computed by running two independent calculations estimating the free energy cost of mutating $A$ into $B$ in the binding pocket ($\Delta G_{\mathrm{bound}}$) and in solvent ($\Delta G_{\mathrm{unbound}}$), saving us the need to simulate the physical binding process of the two compounds.
+In particular, the second line of Eq.~\ref{eq:delta-delta-G-alchemical} is a consequence of $\Delta G_{\mathrm{unbound}}$ being independent of the presence of the receptor in the simulation box as the definition of the unbound state assumes receptor and ligand to be at a sufficient distance for them to have no energetic interactions.
+Note that, when $A$ and $B$ have different number of atoms, Eq.~\ref{eq:delta-delta-G-alchemical} implies the presence of a factor having units of volume entering both logarithms, which require unitless arguments.
+The factor's value is, however, inconsequential as it cancels out in practice.
 
@@
 
+\paragraph{How are alchemical transformations performed in practice?}
+
+In practice, the mutation of $A$ to $B$ in carried out by introducing one or more parameters $\vec{\lambda}$ controlling the potential energy function $U(\vec{q};\vec{\lambda})$ such that the potential of compounds $A$ and $B$ is recovered at two particular values $\vec{\lambda}_A$ and $\vec{\lambda}_B$.
+Briefly, this is achieved by simulating a ``chimera'' molecule composed by enough atoms to represent both $A$ and $B$.
+A subset of the energetic terms in $U(\vec{q};\vec{\lambda})$ is then modulated by $\vec{\lambda}$ so that at $\vec{\lambda}_A$, the atoms that form molecule $A$ are activated and those belonging exclusively to $B$ are non-interacting ``dummy atoms'', while the opposite occur at $\lambda_B$ (see Section~\ref{sec:relative-fe-protocol} for details).
+
+We can rigorously account for fluctuations in other thermodynamic parameters such as changes in volume $V$ when simulating at constant pressure $p$ or changes in number of molecules $N_i$ of species $i$ at constant chemical potential $\mu_i$ (e.g., number of waters or ions) by introducing the \textit{reduced potential}~\cite{shirts2008statisticallya}
+\begin{equation}\label{eq:reduced-potential}
+u(\vec{q};\vec{\lambda}) \equiv \beta \left[ U(\vec{q};\vec{\lambda}) + p \, V(\vec{q}) + \sum_i \mu_i \, N_i(\vec{q}) + \cdots \right] \, .
+\end{equation}
+Here, the collection of thermodynamic and alchemical parameters $\{\beta, \vec{\lambda}, p, \mu, \ldots\}$ defines a \emph{thermodynamic state}.
+In the context of alchemical calculations, in which the thermodynamic states vary only in their value of $\vec{\lambda}$, these are also referred to as \emph{alchemical states}.
+The free energy of mutating $A$ to $B$ in any environment (e.g., binding site, solvent) can then be computed as
+\begin{equation}\label{eq:delta-G-env}
+    \Delta G_{\mathrm{env}} = - k_BT \ln \frac{Z(\vec{\lambda}_B)}{Z(\vec{\lambda}_A)} = - k_BT \ln \frac{\int_{\Gamma_{\mathrm{env}}} \exp\left( u(\vec{q}; \vec{\lambda}_B) \right) \, d\vec{q}}{\int_{\Gamma_{\mathrm{env}}} \exp\left( u(\vec{q}; \vec{\lambda}_A) \right) \, d\vec{q}} \, .
+\end{equation}
+While it is generally not feasible to compute the two partition functions $Z(\vec{\lambda})$, several estimators have been devised to robustly estimate the ratio of partition functions in Eq.~\ref{eq:delta-G-env} (see Section~\ref{subsec:estimators}) from a set of conformations usually collected with MD simulations from the thermodynamic states defined at $\vec{\lambda}_A$ and $\vec{\lambda}_B$.
+
+\paragraph{Why do alchemical calculations need unphysical intermediate states?}
+While it is theoretically possible to estimate the ratio of partition functions from samples collected only at states $\vec{\lambda}_A$ and $\vec{\lambda}_B$, the efficiency of the free energy estimators rapidly decreases with the phase-space overlap between the two states~\cite{wu2005phasespaceb,wu2005phasespacec}.
+Roughly, the phase-space overlap between two thermodynamic states measures the degree to which high-probability conformations (i.e., those with very negative potential energy) in one state are high-probability conformations in the other state as well (see Section~\ref{sec:are-they-good}).
+
+To solve the of having poor overlap between the states of interest, multiple intermediate alchemical states are introduced at values $\vec{\lambda}_A = \vec{\lambda}_0, \vec{\lambda}_1, \cdots, \vec{\lambda}_K = \vec{\lambda}_B$ so that each pair of consecutive states $\vec{\lambda}_k, \vec{\lambda}_{k+1}$ share good overlap.
+Each intermediate state models a ligand that is neither $A$ nor $B$ but a mix of the two.
+Many estimators (e.g., exponential reweighting (EXP)~\cite{zwanzig1954hightemperature} and Bennet acceptance ratio (BAR)~\cite{bennett1976efficient,shirts2003equilibriuma}) can then be used to compute the free energy as
+\begin{equation}
+    \Delta G_{\mathrm{env}} = k_BT \sum_{k=0}^{K-1} \Delta f(\vec{\lambda}_k, \vec{\lambda}_{k+1})
+\end{equation}
+from samples collected at all the alchemical states $\{\vec{\lambda}_k \}$, where $\Delta f$ is the \emph{unitless free energy difference}
+\begin{equation}
+    \Delta f(\vec{\lambda}_k, \vec{\lambda}_{k+1}) = f(\vec{\lambda}_{k+1}) - f(\vec{\lambda}_k) = - \ln \frac{Z(\vec{\lambda}_{k+1})}{Z(\vec{\lambda}_k)} \, .
+\end{equation}
+While this strategy usually results in sampling thermodynamic states whose Boltzmann distributions is very similar, thus collecting information that is to some degree redundant, some estimators such as the Multistate Bennett acceptance ratio (MBAR)~\cite{shirts2008statisticallya} can exploit similarities between states to improve the precision of the estimates by using the conformations sampled in all alchemical states $\{\vec{\lambda}_k \}$ to compute the free energy difference $\Delta f(\vec{\lambda}_i, \vec{\lambda}_{j})$ between any pair of states $i,j$  (Section~\ref{subsec:estimators}).
+
+\paragraph{How do absolute free energy calculations differ from relative?}
+
+While absolute and relative free energy calculations have subtle differences in their practical applications (e.g., use of restraints, handling of the standard state), the fundamental ideas and concepts of relative free energy approaches remain unaltered in other types of alchemical calculations.
+Absolute binding, hydration, and partition free energy still use thermodynamic cycles that enable computing transfer free energies without actually simulating the physical transfer from one environment to another.
+
+The main difference in these approaches lies instead in the thermodynamic cycle to which this strategy is applied.
+For example, a typical thermodynamic cycle for an alchemical absolute binding free energy calculation is represented in Figure~\ref{fig:fig_absolute_thermodynamic_cycle}.
+In this case, two independent calculations compute the free energy of removing the interactions between the ligand and its environment in solvent or in the binding site respectively through a series of intermediate states in which the energy terms are only partially deactivated.
+
 %%%%%%%%%%%%
@@
 \todo[inline, color={green!20}]{ASJSM: @everyone please review this section}
-When starting an alchemical free energy project you need to determine if an alchemical free energy calculation is the appropriate strategy by initially counting the cost of your
+When starting an alchemical free energy project, a key first step is to decide whether free energy calcualtiosn are really the right tool. Particularly, count the cost of your
 project: Can you even hope to tackle the problem with available resources and, if successful, will it
-be worth it?
+be worth it in terms of human and computational cost?
 \subsection{How accurate are alchemical free energy calculations?}
@@
 %
-Current alchemical free energy calculations seem to achieve, in favorable cases, RMS errors around 1-2 kcal/mol depending on force field, system, and a variety of other factors such as simulation time, sampling method, and whether the calculations employed are absolute or relative.
+Current alchemical free energy calculations involving small molecules seem to achieve, in favorable cases, RMS errors around 1-2 kcal/mol depending on force field, system, and a variety of other factors such as simulation time, sampling method, and whether the calculations employed are absolute or relative.
 A small selection of benchmark datasets and case studies can be found in section~\ref{sec:benchmark} at the end of this document.
-However, the domain of applicability is a significant concern~\cite{sherborne2016collaborating, cournia2017relative}, especially for relative calculations, which typically require a bound structure of a closely
+However, the domain of applicability is a significant concern~\cite{sherborne2016collaborating, cournia2017relative}, especially for relative calculations, which typically require a high quality and usually experimental bound structure of a closely
 related ligand as a starting point.
@@
 %
+
 It is worth noting that the accuracy of free energy calculations is highly variable across different protein targets, and likely across different ligand chemotypes as well.
-For instance, FEP+ with OPLS3 achieves an RMSE of 0.62 kcal/mol for a set of 21 compounds binding to JNK1 kinase, yet only an RMSE of 1.05 kcal/mol for a set of 34 compounds binding to P38$\alpha$ kinase~\cite{harder2016opls3}.
-Furthermore, perturbations for the same chemotype in different pockets of the BACE enzyme gave varied errors~\cite{keranen2017acylguanidine}
-\todo[inline,color={green!20}]{JDC: It's important to be clear on what error is being reported: $\Delta G$ after shifting by a constant to minimize the RMSE, unshifted $\Delta G$, $\Delta \Delta G$ of computed edges, or $\Delta \Delta G$ of all edges.
-Note that computed edge $\Delta \Delta G$ RMSE significantly underestimates the error of all edge RMSE: \url{https://github.com/jchodera/jacs-dataset-analysis} ASJSM: Can we incorporate your analysis in this section as an example?
-Maybe this is something that needs to be addressed properly in the analysis section and only referenced here?}
-We thus recommend that retrospective study for a particular target and a particular chemical series be performed to establish the relevant accuracy of free energy calculations for each particular application case.
+For instance, FEP+ with OPLS3 achieves an RMSE of 0.62 kcal/mol for a set of 21 compounds binding to JNK1 kinase, but an RMSE of 1.05 kcal/mol for a set of 34 compounds binding to P38$\alpha$ kinase~\cite{harder2016opls3}.
+Furthermore, perturbations for the same chemotype in different pockets of the BACE enzyme gave varied errors~\cite{keranen2017acylguanidine}.
+Here the errors refer to the difference in $\Delta G$ after fitting to the mean experimental binding energy.
+This prompts us to consider another important aspect.
+It's important to be clear on what error is being reported: $\Delta G$ after shifting by a constant to minimize the RMSE, unshifted $\Delta G$, $\Delta \Delta G$ of computed edges, or $\Delta \Delta G$ of all edges.
+In general, the computed edge $\Delta \Delta G$ RMSE significantly underestimates the error of all edge RMSE: \url{https://github.com/jchodera/jacs-dataset-analysis}.
+Given the need to understand the performance of the system with alchemical free energy calculations, we recommend that retrospective studies for a particular target and a particular chemical series be performed for each particular application case.
 %
@@
 In simple cases such as absolute hydration free energies of small organic molecules, or relative hydration free energy calculations between structurally similar small organic molecules, it should be possible to obtain highly precise estimates with a given software package (with a sample standard deviation under 0.01 kcal/mol).
+
 For more complex use cases such as protein-ligand binding free energies the repeatability is often substantially worse.
@@
 Where this is practical, a more stringent test is to use different input coordinates for each repeat run.
-%
+
 Note that these issues concern calculations carried out with a single software package, but simulation package variations can introduce additional issues.
@@
 For absolute and relative hydration free energies of small organic molecules a variability of ca.
-0.2 kcal/mol between popular simulation packages has been reported.
-\cite{loeffler2018reproducibility} In the recent SAMPL6 SAMPLing challenge a larger variability of 0.3 to 1.0 kcal/mol was noted in the computed absolute binding free energies of host/guest systems even though similar input and simulation parameters were used \cite{rizzi2019sampl6} and, in many cases, single-point energies were identical or nearly so.
+0.2 kcal/mol between popular simulation packages has been reported, even after fixes to the underlying simulation packages.\cite{loeffler2018reproducibility} In the recent SAMPL6 SAMPLing challenge a larger variability of 0.3 to 1.0 kcal/mol was noted in the computed absolute binding free energies of host/guest systems even though similar input and simulation parameters were used \cite{rizzi2019sampl6} and, in many cases, single-point energies were identical or nearly so.
 Further work is needed to ensure reproducibility of alchemical free energy calculations across implementations to guarantee that force-field development efforts lead to transferable potential energy functions.
@@
 calculations, as well as the \emph{purpose} of the calculations and
-the amount of available computer resources. In some cases, predicting
-accurate binding free energies for a particular target might be a
-\emph{more} challenging effort than simply measuring them! This is
+the amount of available computer resources. In some cases, predicting accurate binding free energies for a particular target might be
+\emph{more} challenging than simply measuring them! This is
 often the case when dealing with database screening problems, where
 compounds might be easily and quickly available commercially for
-testing and free energy calculations could consume a much larger
-amount of resources. Free energy calculations thus typically only
-appeal when (slow or costly) synthesis would be required to do
-experiments or experiments are otherwise cost-prohibitive.
-%
+testing and free energy calculations could consume far more resources. Free energy calculations thus typically only
+appeal when (slow or costly) synthesis would be required or experiments are otherwise cost-prohibitive.
+
 Sometimes free energy calculations can provide answers that are not
@@
 much value.
-%
-Here we include a simple estimate of the value provided by alchemical
+
+Here we provide a simple estimate of the value provided by alchemical
 free energy calculations in lead optimization.  Let $P(\Delta\Delta
@@
 free energy calculations agrees with the actual free energy change.
-%
-In lead optimization guided by free energy calculations, we only
+
+In lead optimization guided by free energy calculations, we will likely only
 synthesize and experimentally test molecules that are predicted to
@@
 \end{equation}
-%
-For illustrative purposes, we assume that the actual changes in the
-binding free energies for a set of new molecules also follow normal
+
+For illustrative purposes, assume that the actual changes in the
+binding free energies for a set of new molecules also follows a normal
 distribution, that the standard deviation in the changes is $RT\ln 5$
@@
 \leq 0$).  Under such assumptions, the conditional probability in
-Eq.~\ref{eqn:true-positive} can be easily computed.  If the accuracy
+Eq.~\ref{eqn:true-positive} can be easily computed.
+
+If the accuracy
 of free energy calculations is $\sigma = 1$ kcal/mol, $P(\Delta\Delta
@@
 more molecules of improved affinities than selection without free
-energy calculations.
-%  
+energy calculations under these assumptions.
+
 Available computational resources and timescales of motion also factor
@@
 microsecond or longer simulations for each of $N$ ligands may become
-prohibitive, or at least should be carefully considered. 
+prohibitive, or at least should be carefully considered. Alternatively, if key motions are fast and minimal, only much shorter simulations may be necessary. 
 %
@@
 of involvement in a particular project. In some cases, the analysis may conclude that free energy calculations will not be feasible for the proposed problem.
-Here, by ``cost'', we refer not just to financial cost of the calculations relative to experiments, but also time -- can the calculations can be run faster than experiments are done?
+Here, by ``cost'', we refer not just to financial cost of the calculations relative to experiments, but also time -- can the calculations be run faster than experiments are done?
 How will the relevant resource and opportunity costs factor in?
@@
 It is important to know if the system has hardly been studied, because should the initial calculations perform poorly, the effort may turn into an attempt to understand the relevant sampling, force field, or system preparation problems.
-%
-If you are unsure whether your project is feasible, one option may be
-to conduct a short exploratory study to assess tractability for just a very small
+
+If you are unsure whether your project is feasible, as mentioned above one recommended option is to conduct a short exploratory study to assess tractability for a small
 number of ligands. This can be sufficient to get an initial
@@
 \label{sec:drugdiscovery}
-\todo[inline, color={green!20}]{ASJSM: Merge/tidy of what can I expect from my simulation, drug discovery aspects and simulation prerequisites, there are some duplications there at the moment, happy for restructuring ideas.
-}
 %
 Many practitioners expect alchemical methods to provide valuable guidance for drug discovery, and to exhibit accuracy superior to most  alternative approaches for suitable targets~\cite{kuhn2017prospective}.
-It has been stressed that successful application in industry may require considerable knowledge of the "domain of applicability" of free energy calculations -- where they work well and where they will not ~\cite{sherborne2016collaborating}.
+Successful application in industry may require considerable knowledge of the "domain of applicability" of free energy calculations -- where they work well and where they will not~\cite{sherborne2016collaborating}.
 Successful application also requires robust protocols for preparing, submitting and analysing alchemical calculations.
@@
 %
-\subsection{Capturing the experimental conditions}
+\subsection{Capturing experimental conditions}
 The calculations aim to capture the alchemical change from one ligand to another as accurately as possible.
 Therefore, it is necessary to consider details of the experimental setup, such as pH.
-Biological assays are usually run at neutral pH but this is not always the case, and protein and ligand preparation protocols should reflect this.
-In more detail, the formal charge, or tautomeric state of the small molecules can change within a series of analogs.
-The change of pKa is a common medicinal chemistry strategy to modify drug properties and would normally require explicit effort to incorporate into alchemical calculations.
-Also, to ensure modeling matches experiment, we need to ensure that we simulate the same system -- which requires understanding what protein construct is used in the bioassay.
-For instance, does the X-ray that is to be used for the calculations match the construct used for screening (i.e.
+Biological assays are usually run at neutral pH but this is not always the case, and protein and ligand preparation protocols should reflect experimental pH.
+
+The formal charge and/or tautomeric state of the small molecules can change within a series of analogs, necessitating care in treatment.
+Additionally, medicinal chemistry efforts might deliberately modify the pKa of a series to modify drug properties, requiring explicit efforts to incorporate these changes into alchemical calculations.
+
+To ensure modeling matches experiment, we also need to ensure that we simulate the same system -- which requires understanding what protein construct is used in the bioassay.
+For instance, does the X-ray structure that is to be used for the calculations match the construct used for screening (i.e.
 catalytic vs full length, monomer vs dimer etc)~\cite{perez-benito2018predicting}?
@@
 \subsection{Is my binding mode accurate?}
-As also mentioned, good performance of alchemical FE calculations requires an accurate representation of the ligand binding mode, which often is estimated based on X-ray crystallography for a related ligand.
+As also mentioned, good performance of alchemical FE calculations requires an accurate representation of the ligand binding mode, which often is estimated based on X-ray crystallography.
 If more than one structure is available, the modeler should pay attention to choose the most suitable.
@@
 for a detailed discussion of choosing optimal structures for structure-based modeling~\cite{warren2012essential}.
+
 It is also useful to study the structure activity relationship and understand the expected impact in the binding site, whether side chain movement in the protein will be required, and whether there evidence of this in any alternative X-ray structures of the same protein.
 Often, only one protein and water configuration is used for a series of alchemical FE calculations, so this needs to be capable of accommodating the smallest through to largest ligands in a way that allows stable and well behaved simulations.
-This can be a practical limitation for the alchemical changes that are feasible, though a simple work-around can be to separate compounds into sub-series for different calculations.
+This can provide a practical limit on the alchemical changes that are feasible, though a simple work-around can be to separate compounds into sub-series for different calculations.
+
 If multiple structures are available there is some evidence the higher affinity complex can give better performance~\cite{perez-benito2019predicting}, at least in some cases.
@@
 Therefore, docking ligands with no restraints is not usually the best method for defining the ligand inputs, instead a core restrained docking can provide satisfactory outcomes, but in this case one still should pay careful attention to consistency of alignment for identical substituents.
-Another alternative is to manually edit the same core with the changing substituents.
-This provides assurances that coordinates for the non-perturbed structures remain identical and aromatic substituents for instance obey consistent dihedral angles but is not feasible for many compounds and therefore automation can be desirable.
-
-Finally, the role of water in ligand binding is now well understood and it can be crucial to capture the changes in binding site solvation during ligand binding.
+Another alternative is to manually edit the same core and add/modify the changing substituents.
+This provides assurances that coordinates for the non-perturbed portion of the structure  remain identical and aromatic substituents for instance have consistent dihedral angles but is not feasible for many compounds and therefore automation can be desirable.
+
+Finally, the role of water in ligand binding is not always well understood and it can be crucial to capture the changes in binding site solvation during ligand binding.
 Can crystallographic waters be retained?
@@
 
-Running binding FE calculations in a drug discovery application will typically require the use of software or tools to facilitate the calculations of many compounds.
+Running binding FE calculations in a drug discovery application will typically require the use of software or tools to facilitate the large number of calculations.
 The commercial implementation from Schr\"{o}dinger, FEP+, uses a graphical user interface and once the user has calculated any missing force field parameters, calculations can be prepared and submitted within minutes.
 The software builds the network of perturbations, connecting similar compounds based on LOMAP methodology~\cite{liu2013lead}, all inputs for the Desmond simulations are built automatically and the job can be saved or run immediately on available GPUs.
-Commercial tools can be expensive in some cases, but alternative tools are becoming available to allow relatively straight forward running of alchemical FE calculations with other software~\cite{gapsys2015pmx, loeffler2015fesetup, song2019using, gapsys2020large,  jespers2019qligfep}.
-One consideration at this point is to define the perturbations suitable for the application.
-For instance, if the goal is to accurately assess the relative binding energy of a small number of compounds, possibly with challenging synthesis, the map of perturbations may contain as many connections between compounds as affordable.
+Commercial tools can be expensive in some cases, but alternative tools are becoming available to allow relatively straightforward alchemical FE calculations with other software~\cite{gapsys2015pmx, loeffler2015fesetup, song2019using, gapsys2020large,  jespers2019qligfep}.
+
+For relative free energy calculations, various graph topologies or maps of calculations are possible, and choices may depend on the target application.
+For instance, if the goal is to accurately assess the relative binding energy of a small number of compounds, possibly with challenging synthesis, the map of perturbations should contain as many connections between compounds as affordable.
 However, when running calculations on hundreds of compounds a so called ‘star-map’ can be used that just contains one connection per compound: perturbing every compound to a central ligand, typically the crystal structure ligand~\cite{konze2019reactionbased}.
 In this way the top-ranking examples can be readily identified and submitted to additional calculations in a second round.
+Alternatively, if the goal is to achieve the smallest possible error with minimal computational expense, certain graph topologies provide benefits~\cite{https://doi.org/10.1002/jcc.26095, https://doi.org/10.1021/acs.jcim.9b00528}
+
 %
@@
 For prospective drug discovery applications there are several other considerations including understanding likely errors and taking selection bias into account.
+
 It is crucial when proposing compounds for synthesis to have some idea of the underlying error or uncertainty in the predictions.
@@
 In addition, the hysteresis can be checked by comparing the difference for forward and reverse perturbations of the same structural change, or in the case of large maps of perturbations cycle closure errors~\cite{wang2013modeling} indicate problematic perturbations involved in cycles connecting many compounds.
-Selection bias, as described above, can be problematic in prospective applications as there is normally an aim to focus on only a narrow range of predicted activity (i.e.
-the most active).
+Selection bias, as described above, can be problematic in prospective applications as projects normallly seek to focus only on the most potent compounds.
 Thus, correlation between prediction and subsequent experiment may lead to disappointment but corrections can be applied based on previous recommendations~\cite{abel2017critical}.
 
-In summary, the successful use of alchemical calculations in drug discovery requires working in the domain of applicability, using a high quality X-ray structure of the target compounds, and testing the approach retrospectively to ensure the system setup is well-behaved.
+In summary, the successful use of alchemical calculations in drug discovery requires working in the domain of applicability, using a high quality X-ray structure of the target bound to compounds in the series, and testing the approach retrospectively to ensure the system setup is well-behaved.
 Always assess your confidence in the resulting predictions and communicate this when discussing with experimentalists.
 Consider performing repeat calculations for at least some of the perturbations in the study.
+
 In a drug discovery setting it is inevitable to consider larger virtual libraries.
@@
 \label{sec:prerequisites}
-The alchemical free energy protocols, that are discussed in Section~\ref{sec:simulation_protocol_choice}, are defined for a specific type of free energy calculation, i.e.
+Alchemical free energy protocols as discussed below (Section~\ref{sec:simulation_protocol_choice}) are defined for a specific type of free energy calculation, i.e.
 a free energy of binding or a free energy of hydration.
-Depending on the type of simulations different considerations have to be made for ligands, solvent, and host molecules (in the case of the estimation of free energies of binding).
+Different types of simulations require different considerations for ligands, solvent, and host molecules (in the case of the estimation of free energies of binding).
+
 \subsection{Free energies of binding}
@@
 In principle in the limit of sufficient conformational sampling, the free energy changes estimated from an alchemical free energy calculation should be independent of the system's initial coordinates.
-However, in practice because simulations of thermodynamic states are of finite duration (typically currently 1-100 ns per state) this is only true for certain classes of alchemical free energy calculations such as relative or absolute free energies of hydration of small and relatively rigid organic molecules.
-Host-guest or protein-ligand complexes typically exhibit slowly relaxing degrees of freedom that exceed significantly the duration of an alchemical free energy calculation.
+However, in practice because simulations are of finite duration (typically 1-100 ns per state at present) this is only true for certain classes of alchemical free energy calculations such as relative or absolute free energies of hydration of small and relatively rigid organic molecules.
+Protein-ligand complexes typically exhibit slowly relaxing degrees of freedom that exceed significantly the duration of an alchemical free energy calculation, and host-guest calculations can be susceptible to these issues as well, depending on timescale and system.
 It is therefore generally important to carefully select input coordinates to obtain satisfactory results.
@@
 \begin{itemize}
-    \item Do I have one or multiple good host structures? (e.g. a good resolution X-ray crystal of the protein target)
-    \item Do I have information on one or all of the ligands binding sites (e.g. a X-ray structure)
-    \item Should I include buried waters, or other small molecules that can be found in an X-ray structure.
-    \item Are my ligands part of a congeneric series? (simple R group substitutions around the same scaffold)
+    \item Do I have one or multiple good receptor structures? (e.g. a good resolution X-ray crystal of the protein target)
+    \item Do I have information on one or all of the ligand binding sites (e.g. a X-ray structure)
+    \item Should I include buried waters, or other small molecules that can be found in an X-ray structure?
+    \item Are my ligands part of a congeneric series? (i.e. simple R group substitutions around the same scaffold)
     \end{itemize}
@@
 \paragraph{Are there good X-ray structures available?}
-As for any simulation, care should be taken in selecting available X-ray structures in the protein databank and potentially choosing multiple structures initially, to account for variability in receptor conformations as well as accuracy of available X-ray structures.
+As with any simulation, care should be taken in selecting available X-ray structures in the protein databank.
+In some cases it may be wise to choose multiple starting structures to account for variability in receptor conformations as well as the accuracy of available X-ray structures.
 Typically, clustering of receptor structures can be used to identify different receptor conformations near the binding site, as well as assessing relevant side chain placements from the X-ray structure, see for example~\cite{mey2016blinded}.
-In terms of set up and other choices following general best practice guidelines is advisable~\cite{braun2018best}.
-%
-For binding free energy calculations using a congeneric set of ligands and therefore likely to be most suited to a relative free energy protocol (see~\ref{sec:simulation_protocol_choice}).
-Other than the choice of forcefield for the parameterization of the ligands, some care has to be taken selecting binding poses for these ligands.
+In terms of set up and other choices, following general best practice guidelines is advisable~\cite{braun2018best}.
+
+Many free energy calculations focus on a congeneric series of ligands, which can make these calculations suitable for relative free energy protocols (see~\ref{sec:simulation_protocol_choice}).
+For relative calculations, some care has to be taken selecting binding poses for these ligands.
 Generally a common assumption for a congeneric series is that the binding mode is conserved.
-Therefore, if an X-ray structure of one of the ligands is available, this should be used to position the ligands in the putative binding site in an energetically reasonable conformation that entails no steric or electrostatic mismatch with the receptor.
-Checking the X-ray structure versus the experimental electron densities is important, as the resolution of part of the ligand or important sidechains may be based on the interpretation of the crystallographer rather than the available electron density.
-For example, looking at a cyclohexane ring density, a chair conformation is vastly more likely than that of a boat and may just be poorly assigned.
-%
+Therefore, if an X-ray structure of one of the ligands is available, this should be used to position the ligands in the putative binding site in an energetically reasonable conformation without steric or electrostatic mismatch with the receptor.
+Checking the X-ray structure versus the experimental electron densities is important, as the position of part of the ligand or important sidechains may be based on the interpretation of the crystallographer rather than the available electron density, especially in cases of missing density.
+For example, looking at a cyclohexane ring density, a chair conformation is vastly more likely than that of a boat and, if a boat conformation is present in the structure, it may be worth inspecting the density to ensure it adequately supports this choice.
+
 \paragraph{How to deal with different binding modes?}
-Generally, binding modes within congeneric series are conserved~\cite{wacker2010conserved}, however, exceptions exist~\cite{brandt2011congeneric,nazare2005probing}, and weaker binding compounds are more likely to adopt different binding modes (REFS).
-There may also be ambiguity about the relative orientation of substituents.
-A known issue involves a 180 degree flip in the dihedral angle of a 6-membered aromatic ring leading to a different spatial position of ortho- or meta- substituents that otherwise should overlap within a series.
-The 180 degree flip of the ring may not occur enough during simulations to overcome the starting configuration bias.
+Generally, binding modes within congeneric series are conserved~\cite{wacker2010conserved}, however, exceptions exist~\cite{brandt2011congeneric,nazare2005probing}, as discussed in more detail in Section~\ref{sec:multiple_binding_modes}.
+Certain  functional groups may be particularly prone to this due to symmetries or near symmetries.
+One such issue involves a 180 degree flip in the dihedral angle of a 6-membered aromatic ring leading to a different spatial position of ortho- or meta- substituents that otherwise should overlap within a series.
+The 180 degree flip of the ring may not occur enough during simulations to overcome bias due to the starting configuration.
 Another scenario may be equatorial and axial substituted saturated rings (e.g.
@@
 \paragraph{Have you considered sterioisomers and enationmers?}
-Congeneric series can contain stereoisomers or enantiomers, making sure that you model the correct one is important.
+Congeneric series can contain stereoisomers or enantiomers which can bind very differently, resulting in large errors if treated incorrectly.
 For racemates, the relative abundance of each stereoisomer is normally not known.
-Therefore, the experimental activity associated with just one conformation is more uncertain.
+Therefore, the experimental activity associated with just one stereoisomer/enantiomer is more uncertain.
 However, the modeling typically uses just the bioactive conformation that best fits the active site.
-Clearly this introduces the chance for higher error versus experiment.
+Clearly this introduces potential for larger errors compared to experiment.
 Nevertheless, if all compounds in the congeneric series are racemic, originating from similar synthetic procedures with an expected similar abundance of stereoisomers, then the differences may cancel and the trend in calculated and observed binding energies may be robust.
 Nevertheless, we can see that care and further testing is needed in this scenario, and the quality of the predictions may suffer.
+Additionally, unexpected changes in what stereoisomer binds experimentally, if they occur, could pose significant challenges for modelling efforts.
 %
 \paragraph{Conserved binding site waters can play an important role in binding free energies}
-Another aspect to look out for are binding site water molecules that may form water mediated protein-ligand interactions, particularly in situations where exchange with bulk water may be slow on the simulation timescales.
-This happens typically in buried binding site.
-Overlaying multiple protein X-ray structures can identify conserved or just additional water molecules that can be useful to include.
+Binding site water molecules may form water mediated protein-ligand interactions which can pose challenges whenever exchange with bulk water is slow compared to simulation timescales.
+This happens typically in buried binding sites.
+Overlaying multiple protein X-ray structures can identify conserved or additional water molecules that can be useful to include in calculations.
 In cases where water molecules are known to play an important role in the binding, software implementations that use water sampling facilitated by Grand Canonical Monte Carlo methods may be useful.
-For examples see HSP90 as formed part of the D3R grand challenge 2015~\cite{mey2016blinded},A2A~\cite{brucemacdonald2018ligand}, MUP~\cite{ross2015water}, and others~\cite{michel2009energetics}.
+For examples, see HSP90 which formed part of the D3R grand challenge 2015~\cite{mey2016blinded}, A2A~\cite{brucemacdonald2018ligand}, MUP~\cite{ross2015water}, and others~\cite{michel2009energetics}.
 %
@@
 Small steric clashes may be resolved during subsequent simulation equilibration prior to data collection, but there is a risk that the complex relaxes to an alternative metastable state.
-%
+
 An additional consideration arises for single topology relative free energy calculations.
-In this class of alchemical free energy calculations it is necessary to generate a molecular topology that may describe the initial and final states of the perturbation (see different topologies as seen in Fig.~\ref{fig:fig_topology}).
+In this class of alchemical free energy calculations it is necessary to generate a molecular topology that may describe the initial and final states of the perturbation (see Fig.~\ref{fig:fig_topology}).
 In cases where the end-states have high topological similarity and high structural overlap this is relatively straightforward and typically handled by use of maximum common substructure calculations.
 In situations where the end-state topologies differ significantly, or there is relatively little spatial overlap between the two end-states some user intervention may be necessary to produce a satisfactory input topology.
-%
+
 If the binding site location is uncertain but the structure of the receptor is well defined and plausible binding sites are identified, it may be more useful to chose an absolute free energy protocol to compute the standard free energy of binding of the ligand to a set of binding sites.
-This requires the user to prepare input files describing the bound conformation in different putative binding sites (REF).
+This requires the user to prepare input files describing the bound conformation in different putative binding sites~\cite{10.1039/C6CP05680F} (REF).
 The apparent binding free energy of the ligand may be obtained by combining the individual binding site free energies, which also indicate where the ligand is more likely to bind.
 In this case a docking program can generate initial structures.
-Different commercial and none commercial tools are available, such as: rDock, Autodock Vina, Glide, or Flare to name a few~\cite{ruiz-carmona2014rdock, trott2010autodock, friesner2004glide, cheeseright2006molecular}.
+Different commercial and none commercial tools are available, such as rDock, Autodock Vina, Glide, or Flare, to name a few~\cite{ruiz-carmona2014rdock, trott2010autodock, friesner2004glide, cheeseright2006molecular}.
+
 If the putative binding sites are not apparent, for instance due to significant induced-fit effects, it may be challenging to obtain meaningful free energies of binding.
-One may have to account for the free energy change for forming a binding site in the target receptor which may not be feasible in FE simulation timescales.
+One may have to account for the free energy change for forming a binding site in the target receptor which may not be feasible on FE simulation timescales.
 Another key application for absolute binding free energies is to study chemically different ligands, that do not share a common scaffold, but bind at the same site or the same site of similar proteins~\cite{aldeghi2015accurate}.
-%
+
 \subsection{Free energies of hydration or partition coefficients}
@@
 However, in these cases a careful choice of forcefield, as well as water models or organic solvents is essential.
-See for example~\cite{bosisio2016blinded,rustenburg2016measuring} for a good discussion on these choices.
+See for example~\cite{bosisio2016blinded,rustenburg2016measuring} for a good discussion of these choices.
 %
@@
 Many of the issues around simulation setup and protocol choice for alchemical calculations are common, but there are some differences between absolute and relative calculations.
-First we will consider protocol differences  so before treating the common elements.
+We will consider protocol differences before treating the common elements.
 %
@@
 Bond parameters as well as partial charges are adjusted accordingly.
-Thus, in a single topology calculation, atoms may change their type ensuring only relatively few dummy atoms are created. This is illustrated in left arm of Fig.~\ref{fig:fig_topology}. 
+Thus, in a single topology calculation, atoms may change their type ensuring only relatively few dummy atoms are created. This is illustrated in the left arm of Fig.~\ref{fig:fig_topology}. 
 In contrast, in a dual topology alchemical free energy calculation, no atoms are allowed to change type~\cite{shirts2012best}.
-This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxy methyl group, then passing through an intermediate state where atoms which are becoming dummy atoms or ceasing being dummy atoms are partially interacting (this state may or may not be well defined~\cite{mobley2014blind}  and culminating in a state where benzyl alcohol is present along with the additional dummy atoms which was previously a corresponding hydrogen of the benzene.
+This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxy methyl group, then passing through an intermediate state where atoms which are becoming dummy atoms or ceasing being dummy atoms are partially interacting (this state may or may not be well defined~\cite{mobley2014blind}  and culminating in a state where benzyl alcohol is present along with the additional dummy atom which was previously a corresponding hydrogen of the benzene.
 Fig.~\ref{fig:fig_topology}'s right branch depicts how such a dual topology works.
@@
 To our knowledge no general pipeline for such calculations yet exists and this would likely remain a research problem.
-Using an absolute style free energy calculation instead seems a more promising approach in such a case.
+Using an absolute free energy approach  instead seems more promising in such a case.
 %
@@
 Based on the input ligand series a perturbation map or network can be planned.
-Recent heuristics have shown the more connected the perturbation network the better, however, there is a way to optimise network structure while minimising the number of perturbations that need to be computed minimising the resulting computational cost~\cite{yang2020optimal}.
+Recent heuristics have shown the more connected the perturbation network the better, however, there is a way to optimise network structure while minimising the number of perturbations that need to be computed reducing the resulting computational cost~\cite{yang2020optimal}.
 Sometimes the introduction of intermediates that are not part of the original congeneric series are essential to avoid ring breaking, or dealing with perturbations that would otherwise result in large numbers of atoms to be grown or disappearing.
-Some commercial tools have underlying good heuristic but may fail with complicated input, needing user validation in particular when dealing with chiral compounds.
+Some commercial tools have good underlying heuristics but may fail with complicated input, needing user validation in particular when dealing with chiral compounds.
 \begin{figure}
@@
 \end{figure} 
-In some cases, during the lead optimization stage, or if perturbations would be rather large a star shaped network as seen in Fig.~\ref{fig:fig_types_of_networks} (\textbf{A}) is used.
+In some cases, during the lead optimization stage, or for very large datasets that would benefit from rougher initial FE ranking, or in cases where perturbations would be rather large a star shaped network as seen in Fig.~\ref{fig:fig_types_of_networks} (\textbf{A}) is used.
 However, adding redundancy into the network means that a better error analysis can be carried out, by looking at cycle closure errors as discussed in sec.~\ref{sec:are-they-good}, with an example given in Fig.~\ref{fig:fig_types_of_networks} (\textbf{B}).
@@
 Absolute free energies are typically reported with respect to a specific reference or standard state, which effectively determines the arbitrary point at which the free energy is 0.
-The role of the standard state is particularly evident with binding free energies, in which having a reference state allows us to obtain a well-defined partial molar free energy for the reaction
-\begin{equation*}
-\ce{AB <=> A + B}
-\end{equation*}
-through the well-known expression derived from the law of mass action
+The role of the standard state is particularly evident from the expression of the binding free energy between a receptor $R$ and ligand $L$
 \begin{equation} \label{eq:DGfromKAB}
-\Delta G = -RT ~ \ln \left( c^{\circ} K_{AB} \right)  = -RT ~ \ln\left( \frac{c^{\circ} C_{AB}}{C_A C_B} \right) ,
-\end{equation}
-where $R$ is the gas, $T$ is the temperature, $C_X$ is the equilibrium concentration of the chemical species $X$ in the reaction solvent, and the reference state concentration $c^{\circ}$ converts the binding constant $K_{AB}$ into a dimensionless quantity expressed in reference concentration units.
-It should be noted that ignoring the term $c^{\circ}$ is equivalent to assuming a reference concentration of 1~D$^{-1}$, where D are the units used to express $K_{AB}$, and would thus cause the value of $\Delta G$ to vary with the choice of the units.
-Typically, it is convenient to define a standard state at a constant pressure of 1~atm and where each chemical species (i.e., A, B, and AB) in the reaction solvent has a concentration of $c^{\circ}$~=~1~M~=~1~molecule/1660~\r{A}$^3$ but do not interact with other molecules of A, B, or AB.
+    \Delta G = -k_BT ~ \ln \left( c^{\circ} K_b \right)  = -k_BT ~ \ln\left( c^{\circ} \frac{[RL]}{[L][R]} \right) \, .
+\end{equation}
+Here, the reference state concentration $c^{\circ}$ converts the binding constant $K_b$ into a dimensionless quantity expressed in reference concentration units.
+It should be noted that ignoring the term $c^{\circ}$ is equivalent to assuming a reference concentration of 1~D$^{-1}$, where D are the units used to express $K_b$, and would thus cause the value of $\Delta G$ to vary with the choice of the units.
+Typically, it is convenient to define a standard state at a constant pressure of 1~atm and where each chemical species (i.e., A, B, and AB) in the reaction solvent has a concentration of $c^{\circ}$~=~1~M~=~1~molecule/1660~\r{A}$^3$ but do not interact with other molecules of $R$, $B$, or $RL$.
 %
@@
 The choice is likely especially important with Boresch-style restraints, where some relative placements of reference atoms are likely to be numerically unstable; additionally, ligand reference atoms should likely be in a part of the molecule which defines the binding orientation well, rather than in a floppy solvent-exposed tail, for example.
-\todo[inline, color={blue!20}]{DLM: Get input from JDC on what they've learned about these.}
-\todo[inline, color={blue!20}]{DLM: Clarify terminology: Double decoupling, etc. See Feature Box below.}
-%
-%
+
 \subsection{Absolute and relative calculations must deal with some of the same issues}
 \subsubsection{Structural definition of the bound state and weak binders}
-\todo[inline, color={green!20}]{ASJM: This a bit waffly. What are the recommendations, of how to setup these restraints?}
+\todo[inline, color={green!20}]{ASJM: This a bit waffly. What are the recommendations, of how to setup these restraints? AR: I'm on it!}
 In binding free energy calculations, extra care should be taken when simulating the bound state, especially when dealing with weak binders and absolute free energy calculations in combination with enhanced sampling techniques.
@@
 In principle, because of the path independence of the free energy, any arbitrary pathway will give the correct free energy change, but the choice of pathway will greatly affect the efficiency of the calculations.
+
 Some choices are particularly crucial -- for example, transformations involving insertions or deletions of atoms should employ a soft-core potential path for Lennard-Jones or other interactions with repulsive interactions that go to infinite energy at small radius~\cite{beutler1994avoiding, beutler1994molecular,gapsys2012new}.
-%
+
 The key consideration for choosing alchemical pathways is that the intermediate states that a given pathway produces should sample configurational ensembles that change as slowly as possible as $\lambda$ changes, while still managing to go from the initial state to the final state as $\lambda$ goes from 0 to 1.
-%
+
 Another way of stating this is that intermediate states should sample molecular configurations that are as similar as possible to their neighboring states.
@@
 The actual transformation is best handled with soft-core potentials of the form shown in Fig.\ref{fig:fig_what_is_lambda} (\textbf{C}) and (\textbf{B}), with more details given below.
-%
+
 So how do we adjust the potentials between the two end states based on $\lambda$? The simplest possible alchemical pathway is a \textit{linear} pathway:
-%
 \begin{equation}
@@
 This clearly satisfies the basic requirement that it gives the initial endpoint potential energy $U_0(\vec{q})$ when $\lambda=0$ and final endpoint energy $U_1(\vec{q})$ when $\lambda=1$.
-%
+
 For many energy terms, \textit{as long as a repulsive core remains on}, this is a very good approach.
@@
 Although we are not aware of any quantitative tests for dipolar or higher multipole terms, theoretically it should behave equally well for those systems.
-%
+
 However, this approach ends up being terrible for removing or adding repulsive potentials that go to infinity quickly at or near the origin.
 One way to look at this is to examine how low $\lambda$ values must go to reduce the energy at $0.5\sigma$ (the atomic size parameter) down to 1 $k_BT$, where thermal fluctuations make it possible for other atomic sites to penetrate routinely that deep.
-If we are trying go from a particle being present, and desire to make it disappear alchemically, then if the repulsive terms is of the form $\epsilon\frac{\sigma}{r}^{12}$, then if $\epsilon$ was 1 $K_BT$ at the temperature of interest, and we start with the particle present, then solving for $(1-\lambda)(1 kB_T)\frac{\sigma}{\frac{1}{2}\sigma}) = 1 k_B T$ we get $\lambda = 1-2^{-12} \sim 0.999976$.
+If we are trying go from a particle being present, and desire to make it disappear alchemically, then if the repulsive terms is of the form $\epsilon\frac{\sigma}{r}^{12}$, then if $\epsilon$ was 1 $k_BT$ at the temperature of interest, and we start with the particle present, then solving for $(1-\lambda)(1 kB_T)\frac{\sigma}{\frac{1}{2}\sigma}) = 1 k_B T$ we get $\lambda = 1-2^{-12} \sim 0.999976$.
 So we have gone virtually all the way to the end of the transformation, there's still an impenetrable post in the middle of our simulation!
@@
 Various analyses have shown that this is not a very good strategy, however~\cite{pham2011identifying, beutler1994avoiding, zacharias1994separationshifted, blondel2004ensemble, gapsys2012new}.
-%
+
 What we need instead is some sort of function that smoothly gets rid of this infinity.
@@
 \end{equation}
+%
 Where $r_{ij}$ is the distance between two particles $i$ and $j$, $\epsilon_{ij}$ and $\sigma_{ij}$ are the Lennard-Jones parameters corresponding to the interaction between particles $i$ and $j$, and $\alpha$ is a constant (0.5 is optimal for the functional form shown above).
@@
 This functional form is shown in parts (\textbf{C}) and (\textbf{D}) of Fig.~\ref{fig:fig_what_is_lambda}.
+%
 It has been shown that more complicated forms are not significantly more efficient than eq.~\cite{pham2012optimal}.
@@
 However, if you are inserting or removing entire atomic sites, we heavily recommend against using the linear approach; it will be very difficult to get correct results.
-%
+
 We have discussed optimal ways of disappearing or appearing Lennard-Jones interaction sites and turning on and off electrostatics terms.
@@
 It \textit{is} possible to apply the same soft core approach to the Coulomb interaction and this is indeed done in a number of implementations, in which case it is important that the Coulomb interaction is softened more rapidly than the Lennard-Jones interaction to avoid charge penetration issues, this can be tricky to ensure for all types of perturbations ~\cite{steinbrecher2011softcore}.
+
 A safe but potentially more computationally expensive approach is to perform the transformations in sequence; first, turning off all electrostatics for atoms that must be removed, inserting and removing Lennard-Jones sites (both the insertion and removal can be done simultaneously), and then turning electrostatics for the introduced particles on.
 Again, If there are no removals or introduction to atomic sites, then it is reasonable to change all interactions linearly.
-%
+
 Other issues, such as whether absolute calculations should retain or remove intramolecular non-bonded interactions
@@
 These problems can be diagnosed by noticing lack of configuration space overlap between different simulations (see section~\ref{sec:are-they-good})"
-%
+
 Relative calculations introduce additional choices, such as the order in which to modify nonbonded interactions.
@@
 At the time of writting, there has not been conclusive evidence to suggest one approach is better in general than the other, so discretion should be left up to the user as to what is viable from both hardware resources, and what the simulation software supports.
-%
+
 A key additional consideration in choosing the alchemical pathway is the choice of spacing of intermediate states.
 The spacing depends to some extent on the choice of analysis method, though states should essentially be spaced equidistant in the relevant thermodynamic length~\cite{crooks2007measuringa, sivak2012thermodynamic}.
-For BAR/MBAR techniques this means that state spacing should typically be equal in variance; along a minimal variance path, this is equadistance in $\lambda$, but most paths are not truly minimal variance.~\ref{pham2012optimal, shenfeld2009minimizing}
+For BAR/MBAR techniques this means that states should be spaced so that the statistical uncertainties between neighboring states spacing should typically be equal;~\ref{pham2012optimal, shenfeld2009minimizing}
 Some schemes to adaptively optimize the spacing of intermediate states based on initial exploratory simulations have been proposed~\cite{hayes2017adaptive}.
-
-%
-\subsubsection{What different sampling schemes exist and which one should I use?}
+For molecules changing in dense solvent, then the best path is roughly independent of molecule size and shape, so what works for one molecular transformation is likely to be relatively efficient for another~\cite{monroe2004converging}
+
+\subsubsection{Which sampling scheme will work best for my problem?}
 \label{sec:sampling_schemes}
-With all alchemical simulations having the requirement of sampling from different sets of $\lambda$ states in common, different approaches have been designed to achieve this.
-Fig.~\ref{fig:fig_sampling_scheme} illustrates the four most common schemes used.
-The simplest approach is defining a set of fixed replicas each representing a different $\lambda$ value that run independent simulations at each of the intermediate $\lambda$ values, see Fig.~\ref{fig:fig_sampling_scheme}(\textbf{A}).
+Though all alchemical simulations must the sample from multiple $\lambda$ states, different approaches can be used to achieve this.
+Fig.~\ref{fig:fig_sampling_scheme} illustrates the four most common schemes.
+The simplest approach involves running an independent simulation at each of the reuqisite $\lambda$ values (see Fig.~\ref{fig:fig_sampling_scheme}(\textbf{A})).
 This type of scheme is currently used for AMBER TI calculations~\cite{song2019using} and for Siremol/Openmm~\cite{hedges2019biosimspace}.
-A slightly more complex approach is allowing replica mixing, meaning that each independent $\lambda$ simulation is attempted to be exchanged with a neighboring $\lambda$ replica.
-This is based on ideas of Monte Carlo simulations of spin glasses by Swendsen and Wang~\cite{swendsen1986replica}.
-Using a Metropolis Hastings based acceptance criterion for exchanges means that the generated ensemble of all replicas is still sampling from the Boltzmann distribution and has been used in many different contexts for molecular simulations~\cite{sugita2000multidimensionala,sugita1999replicaexchangea, woods2003developmenta, jiang2010free}.
+However, if these simulations can be run simultaneously with communication between them, a simple extension allows mixing between these replicas.
+In this approach, the simulation at each $\lambda$ can undergo periodic exchanges with neighboring $\lambda$ values.
+This form of replica exchange (Hamiltonian replica exchange) is based on ideas developed fro Monte Carlo simulations of spin glasses by Swendsen and Wang~\cite{swendsen1986replica}.
+With the Metropolis-Hastings acceptance criterion for exchanges, the generated ensemble of all replicas still samples from the Boltzmann distribution, thus this approach has been used in many different contexts for molecular simulations~\cite{sugita2000multidimensionala,sugita1999replicaexchangea, woods2003developmenta, jiang2010free}.
 The basic idea of the replica exchange scheme is shown in Fig.~\ref{fig:fig_sampling_scheme}(\textbf{B}).
@@
 A third approach borrows ideas from simulated tempering~\cite{marinari1992simulateda}.
-In this scheme a single replica rapidly explores all of $\lambda$ space by optimally working out weights that allow switching between different intermediate $\lambda$ values, as seen in Fig.~\ref{fig:fig_sampling_scheme}(\textbf{C}) and also referred to as self adjusted mixture sampling~\cite{lyubartsev1992newa, li2007simulated, tan2017optimally} and while promising, has so far only been supported in OpenMM Tools~\cite{andrearizzi2019choderalab}.
-The last approach available in conjunction with GROMACS makes use of non-equilibrium simulations~\cite{aldeghi2018accurate}.
-In this approach, only end state $\lambda$ replicas ($\lambda$=0, $\lambda=1$) are simulated at equilibrium intermediate information is generated from non-equilibrium simulations.
+In this scheme a single replica rapidly explores all of $\lambda$ space by working out optimal weights that allow switching between different intermediate $\lambda$ values, as seen in Fig.~\ref{fig:fig_sampling_scheme}(\textbf{C}).
+This approach is also referred to as self adjusted mixture sampling~\cite{lyubartsev1992newa, li2007simulated, tan2017optimally} and while promising, has so far only been supported in OpenMM Tools~\cite{andrearizzi2019choderalab}, and so far has not yet seen published applications for binding free energy calculations.
+The last approach makes use of non-equilibrium simulations~\cite{aldeghi2018accurate}.
+In this approach, only end state $\lambda$ replicas ($\lambda$=0, $\lambda=1$) are simulated at equilibrium; intermediate information is generated from non-equilibrium simulations that rapidly transition between endstates.
+This approach is available in GROMACS and appears to be coming online in several other packages.
 A schematic of this approach is shown in Fig.~\ref{fig:fig_sampling_scheme}(\textbf{D}).
@@
 \end{figure}
-The best practice recommendation, is to use replica exchange type sampling schemes (Fig.~\ref{fig:fig_sampling_scheme}(\textbf{B})), or in absence of their availability and when looking at systems with fast mixing times parallel replica schemes are acceptable (Fig.~\ref{fig:fig_sampling_scheme}(\textbf{A})).
+
+Currently, we recommend using Hamiltonian replica exchange type sampling schemes (Fig.~\ref{fig:fig_sampling_scheme}(\textbf{B})).
+If these are not available in the code of choice, running independent simulations at different $\lambda$ values can be acceptable, especially when conformational sampling is fast (Fig.~\ref{fig:fig_sampling_scheme}(\textbf{A})).
 Single replica schemes and non-equilibrium schemes are not as established yet, but are very promising.
-%
+
 \subsubsection{How long should I run my simulation for and what information should be saved?}
@@
 To deal with prospective ligands with unknown binding modes, discovery projects commonly assume that modifications of functional groups on a common scaffold result in a consistent binding mode across all members of a series.
-This is not necessarily always the case~\cite{kaus2015how}, as reviewed elsewhere~cite{mobley2009binding} and in some cases unexpected binding mode changes can be the origin of apparent non-additivity in structure-activity relationships~\cite{calabro2016elucidation}.
+This is not necessarily always the case~\cite{kaus2015how}, as reviewed elsewhere~\cite{mobley2009binding} and in some cases unexpected binding mode changes can be the origin of apparent non-additivity in structure-activity relationships~\cite{calabro2016elucidation}.
 Binding modes also tend to be particularly variable in the case of fragments, which often may have multiple relevant binding modes~\cite{}.
@@
 Hamiltonian exchange, coupled with appropriate restraints, can allow the ligand to relatively rapidly exchange between potential binding modes when non-interacting, accelerating sampling of binding modes~\cite{wang2013identifying}.
-\todo[inline, color={yellow!40}]{DLM probably need more background refs on Hamiltonian lambda exchange here.}
-%
+
 Separation of states provides a simple though potentially expensive alternative, where each stable binding mode is considered separately with a binding free energy calculation restricted to that binding mode, and then (as long as the binding modes are non-overlapping) the resulting component binding free energies can be combined into a total~\cite{mobley2006use, mobley2012perspective}.
@@
 While this approach has precedent, it is relatively difficult to automate at present and requires considerable care.
-%
+
 Overall, this likely means that relative free energy calculations will be susceptible to problems resulting from uncertainty in ligand binding modes until more robust approaches are available to determine dominant binding modes, or the relative populations of different potential binding modes, in advance.
@@
 Robust estimation of $C_t$ for $t \sim T$ is difficult due to growth in statistical error, so common estimators of $g$ make use of several additional properties of $C_t$ to provide useful estimates (see \emph{Practical Computation of Statistical Inefficiencies} in~\cite{chodera2016simple} for a detailed discussion).
-\todo[inline]{JDC: Describe recommended statistical inefficiency computation algorithm.}
-We recommend using the robust statistical inefficinecy computation routines available within the MBAR~\cite{kylebeauchamp2019choderalab} and alchemlyb~\cite{daviddotson2019alchemistry} packages.
+
+We recommend using the robust statistical inefficiency computation routines available within the MBAR~\cite{kylebeauchamp2019choderalab} and alchemlyb~\cite{daviddotson2019alchemistry} packages.
 %
@@
 will refer to this method as BAR.
-More recent derivations show that this formula is the maximum likelihood estimate of the free energy difference given sets of samples from the two states~\cite{bennett1976efficient,shirts2003equilibriuma }.
-%
+More recent derivations show that this formula is the maximum likelihood estimate of the free energy difference given sets of samples from the two states~\cite{bennett1976efficient,shirts2003equilibriuma}.
+
 Many studies have demonstrated both the theoretical and practical superiority of BAR over EXP in molecular
@@
 %
-Mean unsigned error (MUE, also called mean absolute error/MAE) is the another key statistic to include in your results.
+Mean unsigned error (MUE, also called mean absolute error/MAE) is another key statistic to include in your results.
 Even though some models' near-perfect correlation and ranking statistics might suggest excellent accuracy, MUE values can still be off multiple kcal/mol, providing important additional insight into performance.
@@
 \section{Online resources}
-\todo[inline]{Do we really need this section? Everything has been referenced}
+\todo[inline]{Do we really need this section?
+Everything has been referenced.
+Seems like these could be listed at the beginning.
+If this stays here, it would be more useful if they had some description of what they are.}
 \begin{itemize}
@@
     \item Equilibrate your system with your choice of thermodynamic ensemble
-    \item Check the stability of your system and whether it behaves the way you want it to
+    \item Check the stability of your system and whether it behaves the way you believe it should
 \end{itemize}
@@
         \item Set up your restraints correctly
-        \item Make sure you subsample in your MBAR protocol
-        \item Apply the right correction terms
+        \item Make sure you subsample the data in your free energy estimation protocol
+        \item Apply the appropriate correction terms
 \end{itemize}
@@
         \item Make sure your molecular transformations are realistic (1-5 heavy atoms for reliable computations)
-        \item Generate a perturbation network by your method of choice; check whether you have enough cycle closures (preferably, every ligand should be in a cycle)
+        \item Generate a perturbation network by your method of choice; check whether you have enough cycle closures to check consistency in the results (preferably, every ligand should be in a cycle)
         \item Check whether dummy atoms were assigned correctly
-        \item Make sure you subsample in your MBAR protocol
+        \item Make sure you subsample the data in your free energy estimation protocol
+        \item Apply the 
+        appropriate correction terms
 \end{itemize}
@@
     \item If you observe hysteresis in perturbations and incorrect cycle closures
-    \item Energy differences \textgreater15 kcal·mol-1  are unreliable
+    \item Energy differences \textgreater15 kcal$\cdot$mol$^{-1}$  are unreliable
     \newline\newline\textit{Absolute:}
-    \item Energies below -15 kcal·mol-1  are unreliable
+    \item Energies below -15 kcal$\cdot$mol$^{-1}$  are unreliable
     \item The ligand has not sampled most of the intended region after the vanishing step
@@
 \textbf{GT}: Contributed to introduction and drug discovery sections, and helped edit the paper.\\
-\textbf{AR}: Created figure~\ref{fig:freeenergytrajectories}, contributed to section \ref{sec:simulation_protocol_choice}, and helped edit the paper.\\
-\textbf{MRS}: Helped create figure~\ref{fig:fig_what_is_lambda}, wrote sections describing choices for alchemical pathways and analysis for free energy calculations.\\
+\textbf{AR}: Created figure~\ref{fig:freeenergytrajectories}, contributed to sections~\ref{sec:theory} and \ref{sec:simulation_protocol_choice}, and helped edit the paper.\\
+\textbf{MRS}: Helped create figure~\ref{fig:fig_what_is_lambda}, wrote sections describing choices for alchemical pathways and analysis for free energy calculations.
+Reviewed and edited text throughout.\\
 \textbf{LNN}: Helped write the simulation length, stopping conditions, and information saving section. Edited and reviewed alchemical path section.\\
@@
 %%%%%%%%%%%%%%%
-
-(Explain the contributions of any non-author contributors here)
+\todo[inline, color={green!20}]{Explain the contributions of any non-author contributors here}
+()
 % We suggest you preserve this comment:
@@
 ASJSM and JM acknowledge funding through an EPSRC flagship software grant: EP/P022138/1
-MK and JM acknowledges funding through Innovate UK by KTP partnership 011120.
+MK and JM acknowledge funding through Innovate UK by KTP partnership 011120.
 AR acknowledges partial support from the Tri-Institutional Program in Computational Biology and Medicine and the Sloan Kettering Institute.

### 2020-02-03 paper/manuscript.tex
pushing latest changes just to sync.

@@
 \author[4]{Samarjeet Prasad}
-\author[5]{Julia E. Rice}
 \author[2,8]{Andrea Rizzi}
@@
 \affil[4]{National Institutes of Health, Bethesda, MD, USA}
-\affil[5]{IBM Alamden Research Center, Almaden, CA, USA}
 \affil[6]{University of Colorado Boulder, Boulder, CO, USA}
@@
 \begin{itemize}
-\item computing the free energy of a conformational change for a molecule with a high barrier to interconversion (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{A}));
-\todo[inline, color={yellow!20}]{JDC: Do you intend these to be representative of BLUES type simulations? -- yes}
-\item computing partition ($\log P$) or distribution ($\log D$) coefficients between environments (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{B}))~\cite{rustenburg2016measuring, bosisio2016blinded} 
-\item or between compartments or membranes (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{C}))~\cite{corey2019insights}. 
+\item computing the free energy of a conformational change for a molecule with a high barrier to interconversion (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{A});
+\item computing partition ($\log P$) or distribution ($\log D$) coefficients between environments (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{B})~\cite{rustenburg2016measuring, bosisio2016blinded} 
+\item or between compartments or membranes (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{C})~\cite{corey2019insights}. 
 \end{itemize}
@@
 \begin{itemize}
-\item a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{D}))\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate};
-\item the entire ligand can be alchemical transferred from protein to solvent in an absolute binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{E}))~\cite{mobley2007predicting,aldeghi2015accurate,aldeghi2017predictions};
-\item and small alchemical modifications can be made between chemically related ligands to estimate differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (\textbf{F}))~\cite{wang2015accurate,mey2016blinded,gapsys2020large}.
-\end{itemize}
-%
+\item a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{D})\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate};
+\item the entire ligand can be alchemical transferred from protein to solvent in an absolute binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{E})~\cite{mobley2007predicting,aldeghi2015accurate,aldeghi2017predictions};
+\item and small alchemical modifications can be made between chemically related ligands to estimate differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{F})~\cite{wang2015accurate,mey2016blinded,gapsys2020large}.
+\end{itemize}
+%
+
 Once an alchemical simulation procedure is performed (which usually involves more than one simulation), the data must be analyzed to estimate the free energy for the transformation of interest.
@@
 %
-More recent developments have seen new, highly efficient statistical estimators that make better use of all the data, often building on the Bennett acceptance ratio (BAR)~\cite{bennett1976efficient}, producing multistate generalizations~\cite{shirts2008statisticallya} or removing the need for global equilibrium~\cite{wu2016multiensemble, mey2014xtram, wu2014statistically}
+More recent developments have seen new, highly efficient statistical estimators that make better use of all the data, often building on the Bennett acceptance ratio (BAR)~\cite{bennett1976efficient}, producing multistate generalizations~\cite{shirts2008statisticallya} or removing the need for global equilibrium~\cite{wu2016multiensemble, mey2014xtram, wu2014statistically}.
 \todo[inline, color={yellow!20}]{JDC: The key pivotal moment was the Shirts paper on extremely precise free energy calculations, which demonstrated that a combination of technologies can produce robust estimates with useful precision.
@@
 %
+
 As the field of molecular simulation moves towards millisecond timescales, accurate alchemical calculations on even more challenging problems will become reasonable to perform.
@@
 \begin{itemize}
-\item Covalent inhibition.
-\item Free energies of mutation of protein side chains
+\item covalent inhibition
+\item free energies of mutation of protein side chains
 \item nonspecific binding or multiple binding sites.
-\item Complex association stoichiometry (beyond simple 1:1 association).
-\item Approximate endpoint free energy methods such as MM-PBSA~\cite{genheden2015mm} and LIE~\cite{gutierrez-de-teran2012linear}
+\item complex association stoichiometry (beyond simple 1:1 association).
+\item approximate endpoint free energy methods such as MM-PBSA~\cite{genheden2015mm} and LIE~\cite{gutierrez-de-teran2012linear}
 \item Free energy methods that extract the ligand using geometric order parameters and potential of mean force methods~\cite{heinzelmann2017attachpullrelease}
-\item Forcefield dependence for protein, ligand, ions, co-solvents, and co-factors.
-A number of different studies have looked at the influence of force fields and it is assumed the user has made an adequate choice for the system under study.~\cite{loeffler2018reproducibility, vassetti2019assessment, lopes2015current}
-\end{itemize}
-%
+\item forcefield dependence for protein, ligand, ions, co-solvents, and co-factors.
+A number of different studies have looked at the influence of force fields and it is assumed the user has made an adequate choice for the system under study~\cite{loeffler2018reproducibility, vassetti2019assessment, lopes2015current}.
+\end{itemize}
+%
+
 For convenience we have also compiled a list of common acronyms, common symbols used, and some essential terminology.
@@
 $\Delta G$ -- free energy \\
-$\Delta f$-- reduced free energy \\
+$\Delta f$-- reduced (dimensionless) free energy \\
 $K_b$ -- binding or association constant \\
@@
 $T$ -- temperature \\
-$\beta$ -- (k_BT)^{-1}
-$k_b$ -- rate of binding \\
+$\beta \equiv (k_B T)^{-1}$ -- inverse thermal energy \\
 $K_D$ -- dissociation constant \\
-$\lambda$ -- parameter for progress of the alchemical free energy path \\
+$\lambda$ -- alchemical progress parameter \\
+$\Gamma$ -- conformation space accessible by simulations \\
+$\vec{q}$ -- single configuration\\
 $g$ -- statistical inefficiency
@@
 \subsection{Simulating binding events of receptor-drug systems can be computationally expensive}
-
 Suppose you want to compute the binding affinity, or free energy of binding, of a ligand $L$ to a receptor $R$, given by:
@@
 
-\subsection{Alchemical free energy calculations yield predictions that do not require direct simulation of binding events}
+\subsection{Alchemical free energy calculations yield predictions that do not require direct simulation of binding/unbinding events}
 
@@
 \end{equation}
-While this strategy usually results in sampling thermodynamic states whose Boltzmann distributions is very similar, thus collecting information that is to some degree redundant, some estimators such as the Multistate Bennett acceptance ratio (MBAR)~\cite{shirts2008statisticallya} can exploit similarities between states to improve the precision of the estimates by using the conformations sampled in all alchemical states $\{\vec{\lambda}_k \}$ to compute the free energy difference $\Delta f(\vec{\lambda}_i, \vec{\lambda}_{j})$ between any pair of states $i,j$  (Section~\ref{subsec:estimators}).
+While this strategy usually results in sampling thermodynamic states whose Boltzmann distributions is very similar, thus collecting information that is to some degree redundant, some estimators such as the Multistate Bennett acceptance ratio (MBAR)~\cite{shirts2008statisticallya} can exploit similarities between states to improve the precision of the estimates by using the conformations sampled in all alchemical states $\{\vec{\lambda}_k \}$ to compute the free energy difference $\Delta f(\vec{\lambda}_i, \vec{\lambda}_{j})$ between any pair of states $i,j$  (see Section~\ref{subsec:estimators}).
 
@@
 \todo[inline, color={green!20}]{ASJSM: @everyone please review this section}
-When starting an alchemical free energy project, a key first step is to decide whether free energy calcualtiosn are really the right tool. Particularly, count the cost of your
+When starting an alchemical free energy project, a key first step is to decide whether free energy calculations are really the right tool. Particularly, count the cost of your
 project: Can you even hope to tackle the problem with available resources and, if successful, will it
@@
 In addition, the hysteresis can be checked by comparing the difference for forward and reverse perturbations of the same structural change, or in the case of large maps of perturbations cycle closure errors~\cite{wang2013modeling} indicate problematic perturbations involved in cycles connecting many compounds.
-Selection bias, as described above, can be problematic in prospective applications as projects normallly seek to focus only on the most potent compounds.
+Selection bias, as described above, can be problematic in prospective applications as projects normally seek to focus only on the most potent compounds.
 Thus, correlation between prediction and subsequent experiment may lead to disappointment but corrections can be applied based on previous recommendations~\cite{abel2017critical}.
@@
 %
-\paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss} (\textbf{B})), in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings.
+\paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss} \textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings.
 Some approaches have attempted to address this~\cite{clark2019relative} but a general solution is not yet in mainstream use.
@@
 \end{figure} 
-In some cases, during the lead optimization stage, or for very large datasets that would benefit from rougher initial FE ranking, or in cases where perturbations would be rather large a star shaped network as seen in Fig.~\ref{fig:fig_types_of_networks} (\textbf{A}) is used.
-However, adding redundancy into the network means that a better error analysis can be carried out, by looking at cycle closure errors as discussed in sec.~\ref{sec:are-they-good}, with an example given in Fig.~\ref{fig:fig_types_of_networks} (\textbf{B}).
+In some cases, during the lead optimization stage, or for very large datasets that would benefit from rougher initial FE ranking, or in cases where perturbations would be rather large a star shaped network as seen in Fig.~\ref{fig:fig_types_of_networks} \textbf{A} is used.
+However, adding redundancy into the network means that a better error analysis can be carried out, by looking at cycle closure errors as discussed in sec.~\ref{sec:are-they-good}, with an example given in Fig.~\ref{fig:fig_types_of_networks} \textbf{B}.
+
 %
@@
 Both approaches lead to substantial reduction in the statistical error of the estimated free energies.
+
 %
@@
 At present the most general solution to this problem is simply to avoid the use of constraints (and thus use a smaller timestep if necessary, usually of around 1 fs) in any relative free energy calculation involving a transformation of a constrained bond, as done by GROMACS.
+
 %
@@
 %
-Absolute free energies are typically reported with respect to a specific reference or standard state, which effectively determines the arbitrary point at which the free energy is 0.
+Absolute free energies are by definition  reported with respect to a specific reference or standard state, which effectively determines the arbitrary point at which the free energy is 0.
 The role of the standard state is particularly evident from the expression of the binding free energy between a receptor $R$ and ligand $L$
@@
 It should be noted that ignoring the term $c^{\circ}$ is equivalent to assuming a reference concentration of 1~D$^{-1}$, where D are the units used to express $K_b$, and would thus cause the value of $\Delta G$ to vary with the choice of the units.
-Typically, it is convenient to define a standard state at a constant pressure of 1~atm and where each chemical species (i.e., A, B, and AB) in the reaction solvent has a concentration of $c^{\circ}$~=~1~M~=~1~molecule/1660~\r{A}$^3$ but do not interact with other molecules of $R$, $B$, or $RL$.
+It is convenient to define a standard state at a constant pressure of 1~atm and where each chemical species (i.e., A, B, and AB) in the reaction solvent has a concentration of $c^{\circ}$~=~1~M~=~1~molecule/1660~\r{A}$^3$ but do not interact with other molecules of $R$, $B$, or $RL$.
 %
@@
 This solves two problems.
-First, if the ligand were not kept in a well-defined region, as its interactions were removed it might wander the system, perhaps quite slowly, and only inadequately sample the non-interacting or weakly interacting state -- yet adequate sampling of these states might be required for convergence.
-So for practical purposes, the use of restraints can dramatically improve sampling as interactions are weakened and removed.
+First, if the ligand were not kept in a well-defined region, as its interactions were removed it might wander the system, perhaps quite slowly, and only inadequately sample the non-interacting or weakly interacting state.
+However, adequate sampling of these states might be required for convergence.
+Tthe use of restraints can therefore dramatically improve sampling as interactions are weakened and removed.
 Second, if the ligand is not kept in a well-defined region then it is hard to determine how to link a computed binding free energy to the correct 1 M standard state.
@@
 \subsection{Absolute and relative calculations must deal with some of the same issues}
-\subsubsection{Structural definition of the bound state and weak binders}
-\todo[inline, color={green!20}]{ASJM: This a bit waffly. What are the recommendations, of how to setup these restraints? AR: I'm on it!}
-In binding free energy calculations, extra care should be taken when simulating the bound state, especially when dealing with weak binders and absolute free energy calculations in combination with enhanced sampling techniques.
-In principle, only configurations of the receptor-ligand complex that we consider "bound" should be sampled from the bound state, and, in atomistic simulations, this requires to establish a structural definition of the bound state.
-In practice, the simulation of the bound state starts with the ligand already placed in the binding site and relies on kinetic trapping to maintain a bound complex.
-However, this strategy may not be sufficient if the complex dissociation rate is high or if methodologies such as Hamiltonian replica exchange~\cite{bitetti-putzer2003generalizeda} and expanded ensemble~\cite{chodera2011replica} are employed in absolute free energy calculations to enhance sampling since, in both cases, the ligand may find a way out of the binding site on timescales that are achievable by modern MD simulations.
-A solution commonly adopted in these cases is the use of one or more restraints making the unbound configurations energetically unfavorable through the addition of extra terms in the potential function.
-In practice, even with restraints, it is not always trivial to force a molecular dynamics simulation to explore a restricted region of the configurational space with a complicated geometry such as a binding site.
-If the ligand is a tight binder\todo[inline, color={red!40}]{AR: Should we give an order of magnitude to define a tight and a weak binder?
-ASJSM: Yes!}, it is usually safe to employ a restraint that allows some of the unbound configurations to be sampled since they generally contribute negligibly to the partition function (i.e., they have a relatively small Boltzmann weight) as long as the sampled volume is not so large that their cumulative contribution becomes significant.
-However, this is more problematic when dealing with weak binders as the unbound configurations can have a non-negligible Boltzmann weight, and their binding affinity can exhibit a significant dependency on the restraint type and parameters that are used to determine the sampled volume~\cite{}.
-Moreover, it should be noted that the additional potential energy terms used to model the restraints can introduce a bias in the predicted free energy.
-The bias can be removed at the analysis stage through reweighting techniques, but this procedure can increase the statistical uncertainty of the binding free energy estimate when the restraints are so strong that the overlap with the reweighted state is diminished~\cite{}.
-%
-Finally, it is useful to keep in mind that, for a meaningful comparison between computational predictions and experimental measurements, the definition of the bound state should be consistent.
-In particular, this means that the signal used by the experimental methodology to determine the fraction of bound complexes in solution should in principle reflect the population of complexes in the bound state as defined in the calculation.
+\subsubsection{Handling weak binders and high dissociation rates}
+\todo[inline, color={green!20}]{ASJM: This a bit waffly. What are the recommendations, of how to setup these restraints?}
+\todo[inline, color={red!40}]{AR: Took a pass at this.}
+In binding free energy calculations, only the conformations in which the receptor and ligand form a bound complex should be sampled from the bound states (Section~\ref{sec:theory}).
+In absolute calculations, this applies to the fully interacting state in the complex leg of the thermodynamic cycle (top-right state in Figure~\ref{fig:absolute-thermodynamic-cycle}), while in relative calculations this must be true at both end states of the complex leg (top- and bottom-right states in Figure~\ref{fig:binding-thermodynamic-cycle}).
+In principle, this requires defining which conformations are considered to be bound before running the calculation, but it is common practice to start the simulation with the ligand already placed in the binding site and rely on kinetic trapping to maintain the bound complex.
+This strategy, however, can fail when the dissociation rate of the ligand has the same or smaller order of magnitude than the length of the simulation.
+This is typical of weak binders such as fragments binding shallow pockets with $\mu$M-mM affinities~\cite{georgiou_pushing_2017,pan2017quantitative}.
+In this case, using a flat-bottom or harmonic restraint between receptor and ligand in the bound state(s) can prevent dissociations~\cite{georgiou_pushing_2017,rizzi2019sampl6}.
+We stress that this is normally avoided as it generally introduces bias in the free energy estimate, which is why the restraint is usually activated only in the intermediate states in absolute calculations.
+The bias can be corrected through reweighting schemes~\cite{rizzi2019sampl6}, but this post-processing step can be avoided if a flat-bottom restraint is used and the ligand never hits the potential wall during the simulation in the bound state.
+It is important to note that the spring constant and/or radius parameters of the restraint effectively determine which conformations are considered to be bound.
+As a consequence, these parameters must be tuned to the system so that only the binding site is accessible to the ligand.
+This step is particularly important for weak binders as their free energy of binding is known to be more sensitive to the definition of the binding site~\cite{gilson1997statisticalthermodynamic}.
+
+In absolute calculations, this restraint can substitute or be added to the restraint used to handle the standard state correction (Section~\ref{sec:standardstate-restraints}).
+In the latter case, to compute the standard state correction analytically, the bound-state restraint must be turned off in the decoupled state.
+Alternatively, a flat-bottom restraint can be activate also in the decoupled state as long as the second restraint (e.g., a harmonic or Boresch restraint) prevent the ligand to hit the wall of the flat-bottom potential~\cite{rizzi2019sampl6}.
+Finally, even for tight binders, dissociation events can be enhanced by methods such as Hamiltonian replica exchange~\ref{sugita2000multidimensionala,chodera2011replica,wang2013identifying} and expanded ensemble~\ref{lyubartsev1992newa,li2007simulated}, especially in absolute free energy calculations using harmonic or flat-bottom restraints.
+In the latter case, dissociations can be averted simply by increasing the spring constant and/or reducing the radius of the restraint potential to prevent the exploration of ligand conformations outside the binding site in the decoupled state (bottom-right state in Figure~\ref{fig:absolute-thermodynamic-cycle}) that could be propagated to the bound state.
 %
@@
 This can be proven directly from the BAR and MBAR formulas~\cite{bennett1976efficient,klimovich2015guidelines}, though the exact same principles apply for TI.
-For a 'good' path to work sufficient overlap between pontential energies is required.
-Fig.~\ref{fig:fig_what_is_lambda} (\textbf{A}) and (\textbf{B}) try to illustrate this.
-Fig.~\ref{fig:fig_what_is_lambda} (\textbf{A}) shows in a pictorial way a soft-core potential can be applied across different $\lambda$s.
-Fig.~\ref{fig:fig_what_is_lambda} (\textbf{B}) illustrates the potential energy distributions at the different $\lambda$ intermediates, with sufficient overlap between neighboring $\lambda$ states to ensure that reweighting estimators such as MBAR can be used for analysis (see Sec.\ref{subsec:estimators}).
-The actual transformation is best handled with soft-core potentials of the form shown in Fig.\ref{fig:fig_what_is_lambda} (\textbf{C}) and (\textbf{B}), with more details given below.
+For a 'good' path to work sufficient overlap between potential energies is required.
+Fig.~\ref{fig:fig_what_is_lambda} \textbf{A} and \textbf{B} try to illustrate this.
+Fig.~\ref{fig:fig_what_is_lambda} \textbf{A} shows in a pictorial way a soft-core potential can be applied across different $\lambda$s.
+Fig.~\ref{fig:fig_what_is_lambda} \textbf{B} illustrates the potential energy distributions at the different $\lambda$ intermediates, with sufficient overlap between neighboring $\lambda$ states to ensure that reweighting estimators such as MBAR can be used for analysis (see Sec.~\ref{subsec:estimators}).
+The actual transformation is best handled with soft-core potentials of the form shown in Fig.~\ref{fig:fig_what_is_lambda} \textbf{C} and \textbf{B}, with more details given below.
 
@@
 There are several different variants of the same functional form~\cite{zacharias1994separationshifted, beutler1994avoiding,pham2011identifying}, but the one in given in eq.~\ref{eq:softcore} is easy to understand and implement and fairly numerically stable.
-This functional form is shown in parts (\textbf{C}) and (\textbf{D}) of Fig.~\ref{fig:fig_what_is_lambda}.
-%
+This functional form is shown in parts in\textbf{C} and \textbf{D} of Fig.~\ref{fig:fig_what_is_lambda}.
+
 It has been shown that more complicated forms are not significantly more efficient than eq.~\cite{pham2012optimal}.
@@
 Fig.~\ref{fig:fig_sampling_scheme} illustrates the four most common schemes.
-The simplest approach involves running an independent simulation at each of the reuqisite $\lambda$ values (see Fig.~\ref{fig:fig_sampling_scheme}(\textbf{A})).
+The simplest approach involves running an independent simulation at each of the reuqisite $\lambda$ values (see Fig.~\ref{fig:fig_sampling_scheme}\textbf{A}).
 This type of scheme is currently used for AMBER TI calculations~\cite{song2019using} and for Siremol/Openmm~\cite{hedges2019biosimspace}.
@@
 With the Metropolis-Hastings acceptance criterion for exchanges, the generated ensemble of all replicas still samples from the Boltzmann distribution, thus this approach has been used in many different contexts for molecular simulations~\cite{sugita2000multidimensionala,sugita1999replicaexchangea, woods2003developmenta, jiang2010free}.
-The basic idea of the replica exchange scheme is shown in Fig.~\ref{fig:fig_sampling_scheme}(\textbf{B}).
+The basic idea of the replica exchange scheme is shown in Fig.~\ref{fig:fig_sampling_scheme} \textbf{B}.
 It is supported in various software packages that provide alchemical implementations, such as GROMACS~\cite{aldeghi2015accurate}, FEP+~\cite{wang2015accurate}, and NAMD~\cite{jiang2019computing}.
@@
 This approach is available in GROMACS and appears to be coming online in several other packages.
-A schematic of this approach is shown in Fig.~\ref{fig:fig_sampling_scheme}(\textbf{D}).
+A schematic of this approach is shown in Fig.~\ref{fig:fig_sampling_scheme} \textbf{D}.
 %
@@
 
-Currently, we recommend using Hamiltonian replica exchange type sampling schemes (Fig.~\ref{fig:fig_sampling_scheme}(\textbf{B})).
-If these are not available in the code of choice, running independent simulations at different $\lambda$ values can be acceptable, especially when conformational sampling is fast (Fig.~\ref{fig:fig_sampling_scheme}(\textbf{A})).
+Currently, we recommend using Hamiltonian replica exchange type sampling schemes (Fig.~\ref{fig:fig_sampling_scheme}\textbf{B}).
+If these are not available in the code of choice, running independent simulations at different $\lambda$ values can be acceptable, especially when conformational sampling is fast (Fig.~\ref{fig:fig_sampling_scheme} \textbf{A}).
 Single replica schemes and non-equilibrium schemes are not as established yet, but are very promising.
@@
 \begin{equation}
-\frac{df}{d\lambda} = \frac{d}{d\lambda} \ln \int \exp{-u(\lambda,\vec{q})} d\vec{q} = \expect{\frac{du(\lambda,\vec{q})}{d\lambda}\
+\frac{df}{d\lambda} = \frac{d}{d\lambda} \left[-\ln \int \frac{\exp{-u(\lambda,\vec{q})}}{Q(\lambda)} d\vec{q}\right] = \expect{\frac{du(\lambda,\vec{q})}{d\lambda}\
 }_{\lambda} 
@@
 \label{subsec:uncertainty}
-It is important to consider the variation in your computed free energies from your equilibrium simulations and in this way obtaining an estimate on the certainty, or uncertainty of the obtained value for the free energies you are interested in.
-A recent best practices paper by Grossfiled et.
-al~\cite{grossfield2018best} goes into great detail on estimating uncertainties from molecular simulations and is a good starting point for this topic.
+It is important to consider the variation in your computed free energies from your equilibrium simulations, in order  to  obtaining an estimate of  uncertainty of the obtained value for the free energies of interest.
+A recent best practices paper by Grossfield et.
+al~\cite{grossfield2018best} provides substantial detail on how to estimate uncertainties from molecular simulations and is a good starting point for this topic.
 In general the quantification of different error metrics depends on both data generation and analysis methods  used from the ones discussed above.
-%
+
 The computation of free energies using TI (section~\ref{subsec:estimators}) is straightforward and the trapezoidal rule is often recommended since it allows unequal spacing of $\lambda$ states, which is required to minimize the variance in the free energy estimate, but in principle any good numerical integration method can be used.
@@
 For this purpose, a so-called overlap matrix $\mathcal{O}$ can be used.
-$\mathcal{O} \in \mathbb{R}^{K\times K}$, with $K$ being the number of simulated states, i.e.
+$\mathcal{O}$ is a $K\times K$ matrix, with $K$ being the number of simulated states, i.e.
 $\lambda$s.
@@
 As this probability is computed considering the samples from all states (and not just the adjacent states), the values in each row and column add up to 1.
-In this analysis, the goal is to ensure every state has overlap with its neighbors in both directions -- so that off-diagonal elements are significantly larger than zero.
-Usually this means the matrix should be at least tridiagonal.
-%
+In this analysis, the goal is to ensure every state has overlap with its neighbors in both directions -- so that off-diagonal elements are sufficiently larger than zero.
+For accurate calculations, the matrix should be at least tridiagonal.
+
 Details on the calculation and properties of these matrices can be found elsewhere~\cite{klimovich2015guidelines}.
@@
 the diagonals above and below the main diagonal) should at least be 0.03 to obtain a reliable free energy estimate~\cite{klimovich2015guidelines}.
-Smaller values should be considered as a warning sign (see Fig.~\ref{fig:overlap}(\textbf{C})), as the variance tends to be underestimated in case of poor overlap.
+Smaller values should be considered as a warning sign (see Fig.~\ref{fig:overlap} \textbf{C}), as the variance tends to be underestimated in case of poor overlap.
 %
@@
 %
-Fig.~\ref{fig:overlap} shows examples of good, mediocre and poor overlap (\textbf{A}), (\textbf{B}), and (\textbf{C}) respectively.
+Fig.~\ref{fig:overlap} \textbf{A}, \textbf{B}, and \textbf{C} shows examples of good, mediocre and poor overlap respectively.
 For Fig.~\ref{fig:overlap}(\textbf{A}), the probability to find a sample from state $i$ in its neighbouring state $j$ is about 0.2 for all states adjacent to the main diagonal, and hence the overall connectivity is good.
@@
 %
-In a nutshell, it must be ensured that the values of an overlap matrix in the first diagonals above and below the main diagonal are at least 0.03 in order to yield reliable results. 
-%
+
 There are several other parameters which should be investigated.
-Basic analysis of the trajectories should be performed to ensure that the ligand does not tumble out of the binding site by measuring its RMSD.
+Basic analysis of the trajectories should be performed to ensure that the ligand does not tumble out of the binding site, such as measuring its RMSD to the starting position.
 Furthermore, distributions of the ligand's dihedral angles can be obtained and compared to experimental values.
@@
 %
+
 \paragraph{Cycle closure error}
@@
 %
+
 A scheme to reduce cycle closure errors is used in FEP+ whereby calculated free energy changes along the nodes of the network are re-sampled assuming estimates of the calculated free energy change along a node may be obtained from a Gaussian distribution centered on the estimated free energy change and with a standard deviation equal to the estimated standard deviation of the free energy change.
@@
 for perturbation map design can also be used to compute relative free energies between target and reference compounds~\cite{yang2020optimal}.
+
 %
@@
 \begin{itemize}
-\item Make sure ligand doesn’t tumble out of binding site (Mey has observed this)
-\item Significant discrepancies between different free energy estimators (TI, BAR, MBAR)
+\item Make sure ligand doesn’t tumble out of binding site. 
+\item Significant discrepancies between free energies calculated with different free energy estimators (TI, BAR, MBAR)
 \item Poor replica mixing (for replica-exchange)
-\item Correlation time as a function of lambda as it would be expected to be a smooth
-\item Dependence on initial conformation
-\item Torsional analysis: Is it stuck in specific states? Only very rarely transitions?
+\item Correlation time as a function of lambda that does not vary relatively smoothly
+\item Dependence of the free energy on initial conformation
+\item Torsions that have multiple low-energy minima that are visited rarely or not at all. 
 \item More “usual suspects”
@@
 %
-\subsection{Best practices for data representation}
+\subsection{Best practices for reporting data }
 \label{sec:plot_data}
-Data generation and most adequate analysis isn't everything.
-As a practitioner of alchemical free energy simulations you are also encouraged to adhered to best practice for representing and plotting your results.
-We would encourage the following standard set of analyses and ways to represent data.
+Data generation and adequate analysis is not sufficient to finish a study.
+As a practitioner of alchemical free energy simulations you should use best practice for reporting and plotting your results.
+We  encourage the following standard set of analyses and ways to represent data.
 \paragraph{Statistics to include}
 As with any modelling technique, misuse of statistical analysis can skew the perception of how well models perform in free energy predictions.
-First and foremost, error estimates should always be included on your predictions in whatever form you present your data (scatterplots, barplots, etc; see next paragraph).
-It is recommended to perform triplicates of your predictions at minimum to ensure some measure of reliability in your data, in part because analytical uncertainty estimates often underestimate the true statistical uncertainty, where this is not possible lower error bounds such as given by the MBAR estimator should be included.
-%
+First, error estimates should always be included on your predictions in whatever form you present your data (scatterplots, barplots, etc; see next paragraph).
+We recommend performing triplicates of your predictions at minimum to ensure some measure of reliability in your data.
+It may seem exessive, but  uncertainty estimates often underestimate the true statistical uncertainty.
+Where performing multiple replicas of the simulation is not possible, lower error bounds such as given by the MBAR estimator should be included.
+
 Because of the strong drug discovery context and the relevance of alchemical free energy predictions to structure activity relationship (SAR) projects, both the overall correlation of your predictions relative to experiment and the ability of your models to detect the best-scoring compounds (i.e.
@@
 %
+
 \paragraph{Presenting your data}
@@
 %
+
 Because we are interested mainly in the linear relationship between the alchemical FE predictions and the experimentally-determined affinity values, plots should be depicted with the same range on both axes (i.e.
@@
 Seems like these could be listed at the beginning.
-If this stays here, it would be more useful if they had some description of what they are.}
+If this stays here, it would be more useful if they had some description of what the different references contain.}
 \begin{itemize}
@@
 %%%%%%%%%%%%%%%
-\todo[inline, color={green!20}]{Explain the contributions of any non-author contributors here}
-()
+Julia E. Rice participated in the original discussion of the document at the Best Practices in Molecular Simulation Workshop Hosted by at NIST, Gaithersburg, MD, August 24th-25th, 2017
+
 % We suggest you preserve this comment:

### 2020-02-03 paper/manuscript.tex

@@
 \item a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{D})\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate};
-\item the entire ligand can be alchemical transferred from protein to solvent in an absolute binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{E})~\cite{mobley2007predicting,aldeghi2015accurate,aldeghi2017predictions};
-\item and small alchemical modifications can be made between chemically related ligands to estimate differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{F})~\cite{wang2015accurate,mey2016blinded,gapsys2020large}.
+\item the entire ligand can be alchemical transferred from protein to solvent in an absolute binding free energy calculation (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{E})~\cite{mobley2007predicting,aldeghi2015accurate,aldeghi2017predictions};
+\item and small alchemical modifications can be made between chemically related ligands to estimate relative differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{F})~\cite{wang2015accurate,mey2016blinded,gapsys2020large}.
 \end{itemize}
@@
 Once an alchemical simulation procedure is performed (which usually involves more than one simulation), the data must be analyzed to estimate the free energy for the transformation of interest.
-Early work used simple but statistically suboptimal estimators for this: free energy perturbation (FEP) using the Zwanzig relation~\cite{zwanzig1954hightemperature} or thermodynamic integration (TI), for which the theory dates back decades but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamics, merz1989free}.
+Early work used simple but statistically suboptimal estimators for this: free energy perturbation (FEP) using the Zwanzig relation~\cite{zwanzig1954hightemperature} or thermodynamic integration (TI), for which the theory dates back decades but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamicsa, merz1989free}.
 %
 More recent developments have seen new, highly efficient statistical estimators that make better use of all the data, often building on the Bennett acceptance ratio (BAR)~\cite{bennett1976efficient}, producing multistate generalizations~\cite{shirts2008statisticallya} or removing the need for global equilibrium~\cite{wu2016multiensemble, mey2014xtram, wu2014statistically}.
-\todo[inline, color={yellow!20}]{JDC: The key pivotal moment was the Shirts paper on extremely precise free energy calculations, which demonstrated that a combination of technologies can produce robust estimates with useful precision.
-MRS: aw, shucks!
-added.}
+
 Important studies continued during the 2000's often leading to improved implementation of the methods in common simulation software~\cite{shirts2003extremely,shirts2005solvation,vanderspoel2005gromacs, mermelstein2018fast, wang2015accurate, hedges2019biosimspace}.

### 2020-02-03 paper/manuscript.tex
Fixes to intro and prerequisites.

@@
 \end{figure}
-%
+
 %%%%%%%%%%%%%%%%%%%%
@@
 \label{sec:pre}
-\todo[inline, color=green!20]{ASJSM: @everyone This section is up for review to also check flow with intro and rest of document.}
-This guide is meant to serve as a comprehensive starting point for new practitioners of free energy calculations.
-However, it assumes moderate experience with molecular simulation packages and concepts.
-It is also intended to serve as a reference guide with a convenient checklist (Section~\ref{sec:checklist}) for more seasoned practitioners, to help ensure their calculations meet standard alchemical simulation and analysis practices.
+What can be expected from this Best Practice guide is a good starting point for new practitioners and a reference for experience practitioners.
+For this propose we provide a convenient checklist (Section~\ref{sec:checklist}) to help ensure all calculations meet standard alchemical simulation and analysis practices.
 It can also serve as a set of best practices to ensure simulation robustness and reproducibility which reviewers may wish to consider as they evaluate papers.
-Importantly, it is not, a review of all free energy calculation methods at the cutting edge of current research.
-
-This guide aims to answer the following questions:
+What we assume from a novice practitioner is a moderate experience with molecular simulation packages and concepts.
+Furthermore basic familiarity with the principles of molecular mechanics, molecular dynamics simulations, statistical mechanics, and the biophysics of protein-ligand association are essential.
+If you feel unfamiliar with some of these concepts, good starting points can be found in these references~\cite{braun2019best, grossfield2018best, klimovich2015guidelines, shirts2012best}.
+While reading this Best Practice document, it is important to bare in mind \emph{this is not a review} of all free energy calculation methods at the cutting edge of current research.
+Instead the guide aims to answer the following questions:
 \begin{itemize}
@@
 %
-We assume the reader possesses a basic familiarity with the principles of molecular mechanics, molecular dynamics simulations, statistical mechanics, and the biophysics of protein-ligand association.
-If you feel unfamiliar with some of these concepts, good starting points can be found in these references~\cite{braun2019best, grossfield2018best, klimovich2015guidelines, shirts2012best}.
-
-Other background information may be needed depending on the project.
-For example, often, if binding poses are not known, docking calculations can be used to generate an initial small molecule binding pose to start alchemical calculations, thus some basic familiarity on how to perform docking is also necessary for such projects~\cite{grinter2014challenges}.
-%
-As some of the theoretical background can seem daunting, in this best practices description, we do provide a guide to the essential theory behind alchemical free energy calculations in Section~\ref{sec:theory}.
-
-In the remainder of this paper, we will cover topics that are key to the preparation~(\ref{sec:prerequisites}), choice and use of correct protocols~(\ref{sec:simulation_protocol_choice}), and finally the best practices that should be used for the analysis of alchemical calculations~(\ref{sec:data_analysis}).
+
+Some other background information may be needed depending on the nature of the alchemical project.
+For example, often, if binding poses are not known, docking calculations can be used to generate an initial small molecule binding pose to start alchemical calculations.
+This will require some basic familiarity on how to perform docking calculations to generate reasonable simulation starting points~\cite{grinter2014challenges}.
+%
+As some of the theoretical background can seem daunting, we do, however, provide a guide to the essential theory behind alchemical free energy calculations in Section~\ref{sec:theory}.
+
+In the remainder of this paper, we will cover topics that are key to the preparation~(\ref{sec:prerequisites}), choice and use of correct protocols~(\ref{sec:simulation_protocol_choice}), and finally the best practices that should be used in the analysis of alchemical calculations~(\ref{sec:data_analysis}).
 Particular focus will be given to aspects of the molecular simulations which are unique to alchemical calculations---these include the calculation of transfer free energies (hydration free energies, partition coefficients, etc.), and binding free energies (absolute and relative).
@@
 
-While we try to address as many methods and practices as possible, the field of free energy calculations is broad, and there are many advanced topics that are left to future best practices documents focusing on specific issues.
+While we try to address as many methods and practices as possible, the field of free energy calculations is broad, and there are many advanced topics that are left to future Best Practices documents focusing on specific issues.
 Below, we provide a non-exhaustive list of topics we have not addressed with some references to provide starting points on these more advanced topics:
 \begin{itemize}
-\item covalent inhibition
-\item free energies of mutation of protein side chains
-\item nonspecific binding or multiple binding sites.
+\item covalent inhibition~\cite{lameira2019predicting}
+\item free energies of mutation of protein side chains~\cite{gapsys2016accurate, aldeghi2018accurate}
+\item nonspecific binding or multiple binding sites~\cite{gill2018binding}
 \item complex association stoichiometry (beyond simple 1:1 association).
-\item approximate endpoint free energy methods such as MM-PBSA~\cite{genheden2015mm} and LIE~\cite{gutierrez-de-teran2012linear}
+\item approximate and often less accurate endpoint free energy methods such as MM-PBSA~\cite{genheden2015mm} and LIE~\cite{gutierrez-de-teran2012linear}
 \item Free energy methods that extract the ligand using geometric order parameters and potential of mean force methods~\cite{heinzelmann2017attachpullrelease}
@@
 \end{tcolorbox}
-%\label{sec:abbreviations}
+
 \begin{tcolorbox}[title=List of Symbols, colback=green!10!white]
@@
 \end{tcolorbox}
+
 \begin{tcolorbox}[title=Common terminology, colback=yellow!10!white]
@@
 \end{tcolorbox}
-%
+
 %%%%%%%%%%%%%%%%%%%%

### 2020-02-04 paper/manuscript.tex
Gary walk through

@@
 \item a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{D})\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate};
-\item the entire ligand can be alchemical transferred from protein to solvent in an absolute binding free energy calculation (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{E})~\cite{mobley2007predicting,aldeghi2015accurate,aldeghi2017predictions};
+\item the entire ligand can be alchemically transferred from protein to solvent in an absolute binding free energy calculation (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{E})~\cite{mobley2007predicting,aldeghi2015accurate,aldeghi2017predictions};
 \item and small alchemical modifications can be made between chemically related ligands to estimate relative differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{F})~\cite{wang2015accurate,mey2016blinded,gapsys2020large}.
@@
 \label{sec:pre}
-What can be expected from this Best Practice guide is a good starting point for new practitioners and a reference for experience practitioners.
+What can be expected from this Best Practice guide is a good starting point for new practitioners and a reference for experienced practitioners.
 For this propose we provide a convenient checklist (Section~\ref{sec:checklist}) to help ensure all calculations meet standard alchemical simulation and analysis practices.
@@
 What we assume from a novice practitioner is a moderate experience with molecular simulation packages and concepts.
-Furthermore basic familiarity with the principles of molecular mechanics, molecular dynamics simulations, statistical mechanics, and the biophysics of protein-ligand association are essential.
+Furthermore, basic familiarity with the principles of molecular mechanics, molecular dynamics simulations, statistical mechanics, and the biophysics of protein-ligand association are essential.
 If you feel unfamiliar with some of these concepts, good starting points can be found in these references~\cite{braun2019best, grossfield2018best, klimovich2015guidelines, shirts2012best}.
@@
 \paragraph{Simulating binding events is computationally expensive}
-While simulating binding events has been used to estimate binding affinities~\cite{jong_determining_2011,pan2017quantitative} or to get insights into the binding pathways and kinetics of receptor-ligand systems~\cite{teo_adaptive_2016,votapka_seekr_2017,doerr_--fly_2014,plattner_protein_2015,dixon_predicting_2018}, the computational cost of these calculations is usually dominated by the rate of dissociation, which can be on the $\mu$s scale even for millimolar binders~\cite{pan2017quantitative} and reaches the ms or s scale for a typical drug~\cite{basavapathruni_conformational_2012,hyre_cooperative_2006}.
-Depending on system size and simulation settings, common molecular dynamics software packages can reach a few hundreds of ns/day using currently available high-end GPUs~\cite{eastman_openmm_2017,kutzner_more_2019}, making these type of calculations uninteresting at pharmaceutical drug discovery timescales.
-Other methods compute the free energy of binding by building potential of mean force profiles along a reaction coordinate~\cite{woo2005calculationa,velezvega_overcoming_2013,limongelli_funnel_2013,heinzelmann2017attachpullrelease}, but these methods require prior knowledge of a high-probability binding pathway, which is not easily available, especially in the prospective scenarios typical of the drug development process.
+While simulating binding events has been used to estimate binding affinities~\cite{jong2011determining,pan2017quantitative} or to get insights into the binding pathways and kinetics of receptor-ligand systems~\cite{teo2016adaptive,votapka2017seekr,doerr2014onthefly,plattner2015protein,dixon2018predicting}, the computational cost of these calculations is usually dominated by the rate of dissociation, which can be on the {\textmu}s scale even for millimolar binders~\cite{pan2017quantitative} and reaches the ms or s scale for a typical drug~\cite{basavapathruni2012conformational,hyre2006cooperative}.
+Depending on system size and simulation settings, common molecular dynamics software packages can reach a few hundreds of ns/day using currently available high-end GPUs~\cite{eastman2017openmm,kutzner2019more}, making these type of calculations uninteresting at pharmaceutical drug discovery timescales.
+Other methods compute the free energy of binding by building potential of mean force profiles along a reaction coordinate~\cite{woo2005calculationa,velez-vega2013overcoming,limongelli2013funnel,heinzelmann2017attachpullrelease}, but these methods require prior knowledge of a high-probability binding pathway, which is not easily available, especially in the prospective scenarios typical of the drug development process.
 
@@
 
-In many cases, the quantity of interest is the change in binding affinity when a compound $A$ is mutated to compound $B$ (e.g., by modifying one the drug scaffold's substituents), which, by using Eq.~\ref{eq:binding-free-energy-from-bound-unbound-probability-ratio} and \ref{eq:bound-unbound-probability-ratio} is given by
+In many cases, the quantity of interest is the change in binding affinity when a compound $A$ is mutated to compound $B$ (e.g., by modifying one the drug scaffold's substituents, see Fig.~\ref{fig:fig_what_is_alchemy} \textbf{F}), which, by using Eq.~\ref{eq:binding-free-energy-from-bound-unbound-probability-ratio} and \ref{eq:bound-unbound-probability-ratio} is given by
 \begin{equation}\label{eq:delta-delta-G-physical}
@@
 In practice, the mutation of $A$ to $B$ in carried out by introducing one or more parameters $\vec{\lambda}$ controlling the potential energy function $U(\vec{q};\vec{\lambda})$ such that the potential of compounds $A$ and $B$ is recovered at two particular values $\vec{\lambda}_A$ and $\vec{\lambda}_B$.
-Briefly, this is achieved by simulating a ``chimera'' molecule composed by enough atoms to represent both $A$ and $B$.
+Briefly, this is achieved by simulating a ``chimera'' molecule composed of enough atoms to represent both $A$ and $B$.
 A subset of the energetic terms in $U(\vec{q};\vec{\lambda})$ is then modulated by $\vec{\lambda}$ so that at $\vec{\lambda}_A$, the atoms that form molecule $A$ are activated and those belonging exclusively to $B$ are non-interacting ``dummy atoms'', while the opposite occur at $\lambda_B$ (see Section~\ref{sec:relative-fe-protocol} for details).
@@
 \end{equation}
-While it is generally not feasible to compute the two partition functions $Z(\vec{\lambda})$, several estimators have been devised to robustly estimate the ratio of partition functions in Eq.~\ref{eq:delta-G-env} (see Section~\ref{subsec:estimators}) from a set of conformations usually collected with MD simulations from the thermodynamic states defined at $\vec{\lambda}_A$ and $\vec{\lambda}_B$.
+While it is generally not feasible to compute the two partition functions $Z(\vec{\lambda})$, several estimators have been devised to robustly estimate the ratio of partition functions in Eq.~\ref{eq:delta-G-env} (see Section~\ref{subsec:estimators}) from a set of conformations usually collected with MD simulations from the thermodynamic states defined at $\vec{\lambda}_A$ and $\vec{\lambda}_B$ and intermediates thereof.
 
 \paragraph{Why do alchemical calculations need unphysical intermediate states?}
-While it is theoretically possible to estimate the ratio of partition functions from samples collected only at states $\vec{\lambda}_A$ and $\vec{\lambda}_B$, the efficiency of the free energy estimators rapidly decreases with the phase-space overlap between the two states~\cite{wu2005phasespaceb,wu2005phasespacec}.
-Roughly, the phase-space overlap between two thermodynamic states measures the degree to which high-probability conformations (i.e., those with very negative potential energy) in one state are high-probability conformations in the other state as well (see Section~\ref{sec:are-they-good}).
-
-To solve the of having poor overlap between the states of interest, multiple intermediate alchemical states are introduced at values $\vec{\lambda}_A = \vec{\lambda}_0, \vec{\lambda}_1, \cdots, \vec{\lambda}_K = \vec{\lambda}_B$ so that each pair of consecutive states $\vec{\lambda}_k, \vec{\lambda}_{k+1}$ share good overlap.
+While it is theoretically possible to estimate the ratio of partition functions from samples collected only at states $\vec{\lambda}_A$ and $\vec{\lambda}_B$, the efficiency of the free energy estimators rapidly decreases as the phase-space overlap between the two states also decreases~\cite{wu2005phasespaceb,wu2005phasespacec}.
+Roughly, the phase-space overlap between two thermodynamic states measures the degree to which high-probability conformations (i.e., those with very negative potential energy) in one state are high-probability conformations in the other state as well (see Section~\ref{sec:are-they-good} and Fig.~\ref{fig:fig_what_is_lambda}).
+
+To solve the problem of having poor overlap between the states of interest, multiple intermediate alchemical states are introduced at values $\vec{\lambda}_A = \vec{\lambda}_0, \vec{\lambda}_1, \cdots, \vec{\lambda}_K = \vec{\lambda}_B$ so that each pair of consecutive states $\vec{\lambda}_k, \vec{\lambda}_{k+1}$ share good overlap.
 Each intermediate state models a ligand that is neither $A$ nor $B$ but a mix of the two.
-Many estimators (e.g., exponential reweighting (EXP)~\cite{zwanzig1954hightemperature} and Bennet acceptance ratio (BAR)~\cite{bennett1976efficient,shirts2003equilibriuma}) can then be used to compute the free energy as
+Many estimators (e.g., exponential reweighting (EXP)~\cite{zwanzig1954hightemperature} and Bennet's acceptance ratio (BAR)~\cite{bennett1976efficient,shirts2003equilibriuma}) can then be used to compute the free energy as
 \begin{equation}
@@
 
-To ensure modeling matches experiment, we also need to ensure that we simulate the same system -- which requires understanding what protein construct is used in the bioassay.
+To ensure modeling matches experiment, we also need to accurately model and  simulate the same system -- which requires understanding what protein construct is used in the bioassay.
 For instance, does the X-ray structure that is to be used for the calculations match the construct used for screening (i.e.
@@
 \subsection{Is my binding mode accurate?}
-As also mentioned, good performance of alchemical FE calculations requires an accurate representation of the ligand binding mode, which often is estimated based on X-ray crystallography.
+As also mentioned, good performance of alchemical FE calculations requires an accurate representation of the ligand binding mode, usually from a high quality X-ray crystal structure.
 If more than one structure is available, the modeler should pay attention to choose the most suitable.
@@
 
-It is also useful to study the structure activity relationship and understand the expected impact in the binding site, whether side chain movement in the protein will be required, and whether there evidence of this in any alternative X-ray structures of the same protein.
+It is also useful to study the structure activity relationship and understand the expected impact in the binding site, whether side chain movement in the protein will be required, and whether there is evidence of this in any alternative X-ray structures of the same protein.
 Often, only one protein and water configuration is used for a series of alchemical FE calculations, so this needs to be capable of accommodating the smallest through to largest ligands in a way that allows stable and well behaved simulations.
@@
 Do they clash with some of the larger ligands used in the FEP?
-If so, how will the water be incorporated for smaller ligands that require the water – consider pre-solvation methods such as GCMC steps~\cite{michel2010prediction}.
+If so, how will the solvent be incorporated for smaller ligands that require the water – consider pre-solvation methods such as GCMC steps~\cite{michel2010prediction}.
 Other tools such as WaterMap or open source equivalents (SSTmap, others) can be used to define water structure for systems with no experimental evidence of water sites.
@@
 Nevertheless, if all compounds in the congeneric series are racemic, originating from similar synthetic procedures with an expected similar abundance of stereoisomers, then the differences may cancel and the trend in calculated and observed binding energies may be robust.
-Nevertheless, we can see that care and further testing is needed in this scenario, and the quality of the predictions may suffer.
+Despite this, we can see that care and further testing is needed in this scenario, and the quality of the predictions may suffer.
 Additionally, unexpected changes in what stereoisomer binds experimentally, if they occur, could pose significant challenges for modelling efforts.
@@
 consider a transformation of benzene to benzyl alcohol Fig.~\ref{fig:fig_topology}.
-In this case the common substructure is the benzene ring, though perhaps may be larger depending on how it is defined, as we discuss below.
+In this case the common substructure is the benzene ring, though may be larger depending on how it is defined, as we discuss below.
 In single topology calculations, the overall transformation is set up to involve as few additional atoms as possible, so benzene would be typically changed into benzyl alcohol by changing one of the hydrogens into a carbon.
@@
 Once a particular approach to the topology is selected, a crucial next step is to identify the common atoms which will not be perturbed.
-Rigorously, this process comprises of a maximal common substructure (MCSS) search of the molecules involved to identify the common substructure -- though the parameters of the MCSS search will differ depending on whether single or dual topology calculations are planned.
+Rigorously, this process comprises a maximal common substructure (MCSS) search of the molecules involved to identify the common substructure -- though the parameters of the MCSS search will differ depending on whether single or dual topology calculations are planned.
 Specifically, with a single topology approach in mind, atom types are allowed to change, so a permissive MCSS search can be done, whereas with dual topology a more strict search is required.
@@
 Some approaches have attempted to address this~\cite{clark2019relative} but a general solution is not yet in mainstream use.
+(at end of first sentence of this pgraph I would also cite this work from Liu, Wang and Mobley, https://pubs.acs.org/doi/abs/10.1021/acs.jcim.5b00057)
 %
@@
 However, adequate sampling of these states might be required for convergence.
-Tthe use of restraints can therefore dramatically improve sampling as interactions are weakened and removed.
+The use of restraints can therefore dramatically improve sampling as interactions are weakened and removed.
 Second, if the ligand is not kept in a well-defined region then it is hard to determine how to link a computed binding free energy to the correct 1 M standard state.
@@
 As a consequence, these parameters must be tuned to the system so that only the binding site is accessible to the ligand.
-This step is particularly important for weak binders as their free energy of binding is known to be more sensitive to the definition of the binding site~\cite{gilson1997statisticalthermodynamic}.
+Again, this step is particularly important for weak binders as their free energy of binding is known to be more sensitive to the definition of the binding site~\cite{gilson1997statisticalthermodynamic}.
 
@@
 In the latter case, to compute the standard state correction analytically, the bound-state restraint must be turned off in the decoupled state.
-Alternatively, a flat-bottom restraint can be activate also in the decoupled state as long as the second restraint (e.g., a harmonic or Boresch restraint) prevent the ligand to hit the wall of the flat-bottom potential~\cite{rizzi2019sampl6}.
+Alternatively, a flat-bottom restraint can be activated also in the decoupled state as long as the second restraint (e.g., a harmonic or Boresch restraint) prevents the ligand to hit the wall of the flat-bottom potential~\cite{rizzi2019sampl6}.
 Finally, even for tight binders, dissociation events can be enhanced by methods such as Hamiltonian replica exchange~\ref{sugita2000multidimensionala,chodera2011replica,wang2013identifying} and expanded ensemble~\ref{lyubartsev1992newa,li2007simulated}, especially in absolute free energy calculations using harmonic or flat-bottom restraints.
@@
 If the net charge of the system will change as the alchemical calculation progresses, this can pose major challenges.
-Specifically, finite-size effects can introduce profound artifacts into computed binding free energies~\cite{}, in part because typical schemes for long-range electrostatics (including PME and reaction field) do not handle free energy contributions from such changes effectively or as they would be handled in a hypothetical macroscopic bulk solution~\cite{}.
+Specifically, finite-size effects can introduce profound artifacts into computed binding free energies~\cite{}MISSING citation, in part because typical schemes for long-range electrostatics (including PME and reaction field) do not handle free energy contributions from such changes effectively or as they would be handled in a hypothetical macroscopic bulk solution~\cite{}.
 %
@@
     Softcore potentials provide one of the most efficient families of intermediate pathways, with a $\lambda$ dependence illustrated in (\textbf{C}) and a potential energy dependence shown in (\textbf{D}).
-Note how as $\lambda$ approaches 0, the energy smoothly approaches zero at all $r$, a necessary requirement for efficient calculations.
+Note how as $\lambda$ approaches 0, the energy smoothly approaches zero at all $r$, a necessary requirement for efficient and stable calculations.
 }
@@
 This can be proven directly from the BAR and MBAR formulas~\cite{bennett1976efficient,klimovich2015guidelines}, though the exact same principles apply for TI.
-For a 'good' path to work sufficient overlap between potential energies is required.
-Fig.~\ref{fig:fig_what_is_lambda} \textbf{A} and \textbf{B} try to illustrate this.
+For a 'good' path to work, sufficient overlap between potential energies is required.
+Fig.~\ref{fig:fig_what_is_lambda} \textbf{A} and \textbf{B} illustrate this.
 Fig.~\ref{fig:fig_what_is_lambda} \textbf{A} shows in a pictorial way a soft-core potential can be applied across different $\lambda$s.
@@
 One way to look at this is to examine how low $\lambda$ values must go to reduce the energy at $0.5\sigma$ (the atomic size parameter) down to 1 $k_BT$, where thermal fluctuations make it possible for other atomic sites to penetrate routinely that deep.
-If we are trying go from a particle being present, and desire to make it disappear alchemically, then if the repulsive terms is of the form $\epsilon\frac{\sigma}{r}^{12}$, then if $\epsilon$ was 1 $k_BT$ at the temperature of interest, and we start with the particle present, then solving for $(1-\lambda)(1 kB_T)\frac{\sigma}{\frac{1}{2}\sigma}) = 1 k_B T$ we get $\lambda = 1-2^{-12} \sim 0.999976$.
-So we have gone virtually all the way to the end of the transformation, there's still an impenetrable post in the middle of our simulation!
+If we are trying to go from a particle being present, and desire to make it disappear alchemically, then if the repulsive terms are of the form $\epsilon\frac{\sigma}{r}^{12}$, then if $\epsilon$ was 1 $k_BT$ at the temperature of interest, and we start with the particle present, then solving for $(1-\lambda)(1 kB_T)\frac{\sigma}{\frac{1}{2}\sigma}) = 1 k_B T$ we get $\lambda = 1-2^{-12} \sim 0.999976$.
+So we have gone virtually all the way to the end of the transformation, but there is still an impenetrable post in the middle of our simulation!
 This is not very much like the desired final state of no interactions between the particle and its environment.
 We can play around with a few ways of modifying this, like simulating many more intermediate states near $\lambda=1$.
-Various analyses have shown that this is not a very good strategy, however~\cite{pham2011identifying, beutler1994avoiding, zacharias1994separationshifted, blondel2004ensemble, gapsys2012new}.
-
-What we need instead is some sort of function that smoothly gets rid of this infinity.
-A large number of scheme have been tried~\cite{beutler1994avoiding, zacharias1994separationshifted, blondel2004ensemble, pham2011identifying, pham2012optimal, naden2014linear, donnini2005molecular}, but the most common strategy that appears to be the best practices is to use a "soft-core" potential, of the form:
+However, various analyses have shown that this is not a very good strategy~\cite{pham2011identifying, beutler1994avoiding, zacharias1994separationshifted, blondel2004ensemble, gapsys2012new}.
+
+What we need instead is a function that smoothly gets rid of this infinity.
+A large number of schemes have been tried~\cite{beutler1994avoiding, zacharias1994separationshifted, blondel2004ensemble, pham2011identifying, pham2012optimal, naden2014linear, donnini2005molecular}, but the most common strategy that appears to be the best practice is to use a "soft-core" potential, of the form:
 %
@@
 This functional form has exactly the property we are looking for: it recovers the Lennard-Jones potential when $\lambda=1$, and the other endpoint ($\lambda=0$), it is exactly zero for all $r_{ij}$ everywhere, and as $\lambda$ goes to zero, the  $\alpha(1-\lambda)$ term lowers the infinite energy in the core.
-There are several different variants of the same functional form~\cite{zacharias1994separationshifted, beutler1994avoiding,pham2011identifying}, but the one in given in eq.~\ref{eq:softcore} is easy to understand and implement and fairly numerically stable.
-This functional form is shown in parts in\textbf{C} and \textbf{D} of Fig.~\ref{fig:fig_what_is_lambda}.
+There are several different variants of the same functional form~\cite{zacharias1994separationshifted, beutler1994avoiding,pham2011identifying}, but the one  given in eq.~\ref{eq:softcore} is easy to understand and implement and fairly numerically stable.
+This functional form is shown in \textbf{C} and \textbf{D} of Fig.~\ref{fig:fig_what_is_lambda}.
 
@@
 What about performing both transformations at the same time?
-We can't turn off the electrostatics linearly at the same time we turn off the Lennard-Jones terms, as it would leave infinitely large attractive and repulsive electrostatic terms "bare" at small $\lambda$, resulting in the simulation crashing.
+We can not turn off the electrostatics linearly at the same time we turn off the Lennard-Jones terms, as it would leave infinitely large attractive and repulsive electrostatic terms "bare" at small $\lambda$, resulting in the simulation crashing.
 It \textit{is} possible to apply the same soft core approach to the Coulomb interaction and this is indeed done in a number of implementations, in which case it is important that the Coulomb interaction is softened more rapidly than the Lennard-Jones interaction to avoid charge penetration issues, this can be tricky to ensure for all types of perturbations ~\cite{steinbrecher2011softcore}.
@@
 The spacing depends to some extent on the choice of analysis method, though states should essentially be spaced equidistant in the relevant thermodynamic length~\cite{crooks2007measuringa, sivak2012thermodynamic}.
-For BAR/MBAR techniques this means that states should be spaced so that the statistical uncertainties between neighboring states spacing should typically be equal;~\ref{pham2012optimal, shenfeld2009minimizing}
+For BAR/MBAR techniques this means that states should be spaced so that the statistical uncertainties between neighboring states be equal;~\ref{pham2012optimal, shenfeld2009minimizing}
 Some schemes to adaptively optimize the spacing of intermediate states based on initial exploratory simulations have been proposed~\cite{hayes2017adaptive}.
@@
 \label{sec:sampling_schemes}
-Though all alchemical simulations must the sample from multiple $\lambda$ states, different approaches can be used to achieve this.
+Though all alchemical simulations must sample from multiple $\lambda$ states, different approaches can be used to achieve this.
 Fig.~\ref{fig:fig_sampling_scheme} illustrates the four most common schemes.
@@
 In this approach, the simulation at each $\lambda$ can undergo periodic exchanges with neighboring $\lambda$ values.
-This form of replica exchange (Hamiltonian replica exchange) is based on ideas developed fro Monte Carlo simulations of spin glasses by Swendsen and Wang~\cite{swendsen1986replica}.
+This form of replica exchange (Hamiltonian replica exchange) is based on ideas developed from Monte Carlo simulations of spin glasses by Swendsen and Wang~\cite{swendsen1986replica}.
 With the Metropolis-Hastings acceptance criterion for exchanges, the generated ensemble of all replicas still samples from the Boltzmann distribution, thus this approach has been used in many different contexts for molecular simulations~\cite{sugita2000multidimensionala,sugita1999replicaexchangea, woods2003developmenta, jiang2010free}.
@@
 In this scheme a single replica rapidly explores all of $\lambda$ space by working out optimal weights that allow switching between different intermediate $\lambda$ values, as seen in Fig.~\ref{fig:fig_sampling_scheme}(\textbf{C}).
-This approach is also referred to as self adjusted mixture sampling~\cite{lyubartsev1992newa, li2007simulated, tan2017optimally} and while promising, has so far only been supported in OpenMM Tools~\cite{andrearizzi2019choderalab}, and so far has not yet seen published applications for binding free energy calculations.
+This approach is also referred to as self adjusted mixture sampling~\cite{lyubartsev1992newa, li2007simulated, tan2017optimally} and while promising, has so far only been supported in OpenMM Tools~\cite{andrearizzi2019choderalab}, and to date has not seen published applications for binding free energy calculations.
 The last approach makes use of non-equilibrium simulations~\cite{aldeghi2018accurate}.
@@
 However, even subtly induced changes in protein binding configuration will require more sampling or cause local convergence to a free energy estimate that has high error.
-Many enhanced sampling techniques have been proposed to try and overcome barriers in free energy calculations and mitigate convergence to erroneous estimates~\cite{}. 
+Many enhanced sampling techniques have been proposed to try and overcome barriers in free energy calculations and mitigate convergence to erroneous estimates~\cite{}MISSING citation. 
 The confidence a user should have in a free energy estimate is significantly improved when both the uncertainty of the free energy estimate is low, and when other observables have reached a convergence.
@@
 %
-The potential derivative required for TI should generally be calculated during the simulation; only under every rare circumstances~\cite{naden2015linear} can it postprocessed by a code that does not evaluate the derivatives.
+The potential derivative required for TI should generally be calculated during the simulation; only under very rare circumstances~\cite{naden2015linear} can it post-processed by a code that does not evaluate the derivatives.
 Many codes already have options for doing this.
@@
 It is recommended to calculate the potential differences in code when possible to avoid extra overhead and possible errors produced by running the simulation twice, and to reduce the amount of stored information.
-Although TI must be usually be calculated in code, as it requires the derivative, there is one condition under which it actually has the fastest computation time. 
+Although TI must usually be calculated in code, as it requires the derivative, there is one condition under which it actually has the fastest computation time. 
 If the alchemical path you have chosen is a linear alchemical path, then you get $\frac{dU}{d\lambda} = U_0(\vec{x}) - U_1(\vec{x})$, which is the difference between the initial and final states. 
@@
 %
+
 Free energy information should generally be saved more frequently than coordinate data, approximately at the rate that uncorrelated samples are produced.  
@@
 \label{sec:multiple_binding_modes}
-In a discovery setting, new ligands typically have unknown or at least uncertain binding modes~\cite{kaus2015how, plountprice2000analysis,mobley2009binding,calabro2016elucidation},  complicating binding free energy estimation.
+In a discovery setting, new ligands can have unknown or at least uncertain binding modes~\cite{kaus2015how, plountprice2000analysis,mobley2009binding,calabro2016elucidation},  complicating binding free energy estimation.
 This uncertainty is because it is usually not desirable to estimate a binding affinity for a ligand which already has an available bound structure, since such a compound has already been tested.
@@
 %
+
 Absolute free energy calculations for dissimilar ligands can have particular challenges because the (potentially incorrect) assumption of consistent binding modes across a series of similar ligands is likely to be even less robust than the in the case of relatively calculations.
@@
 %
+
 It is worth highlighting a recent SAMPL blind challenge on HIV integrase as an illustration of this. 
@@
 %
+
 Routes to handle multiple potential binding modes are different depending on whether absolute or relative calculations are selected, unless a method is available to estimate the relative populations of different stable binding modes in advance (e.g.
@@
 For simulations that are simply not long enough to contain a large number of samples from true equilibrium either because they are very short or contain slow processes, this procedure cannot completely remove the bias.
-In such cases, this approach simply selects the final portion of the the simulation, which may be contained in a single substate of conformational space, which may itself lead to biased estimates. 
+In such cases, this approach simply selects the final portion of the the simulation, which may be contained in a single substate of conformational space, and may itself lead to biased estimates. 
 This situation can be detected if the equilibration boundary $t_0$ is a significant fraction of the total simulation length $T$, with a good rule of thumb being that $T \gtrsim 20 t_0$.
@@
 The simplest method for calculating free energy differences from simulations is the so-called \textit{Zwanzig
-relationship}~\cite{zwanzig1954hightemperature}, also called one-sided exponential reweighting (EXP), or simply free energy perturbation, thought this final term is sometimes used to encompass all ways of calculating free energy differences.
+relationship}~\cite{zwanzig1954hightemperature}, also called one-sided exponential re-weighting (EXP), or simply free energy perturbation, though this final term is sometimes used to encompass all ways of calculating free energy differences.
 %
@@
 Essentially, the method is dominated by contributions of rare snapshots\cite{jarzynski2006rare, wu2005phasespaceb, wu2005phasespacec}.
-There are a number of applications where it this expression is nevertheless useful, but those usually involve investigating the dependence of the free energy on small changes in parameters, not changing molecules or even single atoms in molecules~\cite{xx}
+There are a number of applications where this expression is nevertheless useful, but those usually involve investigating the dependence of the free energy on small changes in parameters, not changing molecules or even single atoms in molecules~\cite{xx}
 %
@@
 \begin{itemize}
-\item We recommend MBAR if all energy differences are available. It is provably the lowest variance free energy estimate given samples from multiple states.
+\item We recommend MBAR if all energy differences are available. It is probably the lowest variance free energy estimate given samples from multiple states.
 \item BAR is essentially just as good as MBAR for highly optimized $\lambda$ intermediates.
@@
 \label{subsec:uncertainty}
-It is important to consider the variation in your computed free energies from your equilibrium simulations, in order  to  obtaining an estimate of  uncertainty of the obtained value for the free energies of interest.
+It is important to consider the variation in your computed free energies from your equilibrium simulations, in order  to  obtain an estimate of  uncertainty of the obtained value for the free energies of interest.
 A recent best practices paper by Grossfield et.
@@
 \label{sec:are-they-good}
-There are different easily measurable indicators that can be looked at to test how well converged simulations are and if all alchemical states have been sufficiently sampled for a rigorous analysis.
+There are different easily measurable indicators that can test how well converged simulations are, and if all alchemical states have been sufficiently sampled for a rigorous analysis.
 Furthermore, once you have established that individual perturbations are well behaved, there are some tricks to ensure the overall perturbation network gives reliable results.
@@
 \begin{itemize}
-\item Make sure ligand doesn’t tumble out of binding site. 
+\item Make sure ligand sampling in the binding site is reasonable for it's expected potency and fit, and that it does not depart out of binding site. 
 \item Significant discrepancies between free energies calculated with different free energy estimators (TI, BAR, MBAR)
@@
 \item Dependence of the free energy on initial conformation
-\item Torsions that have multiple low-energy minima that are visited rarely or not at all. 
+\item Torsions that have multiple low-energy minima that are visited rarely or not at all. Also, comparing torsional behavior of ligands in solvent and complex simulations 
+\item The convergence of the free energy change as a function of lambda
 \item More “usual suspects”
@@
 Brown et al.~\cite{brown2009healthy} have provided a useful analysis in terms of upper bounds of expected possible correlations between experiment and computation with a given potency range for the compounds.
-For example for potency ranges of 2 log units it would be impossible to get a higher correlation in R than 0.8 because of experimental uncertainties~\cite{brown2009healthy}.
+For example, for potency ranges of 2 log units it would be impossible to get a higher correlation in R than 0.8 because of experimental uncertainties~\cite{brown2009healthy}.
 What often is neglected to include is an error analysis on correlation statistics that arise from the errors of both experimental and computed data.
@@
 Mean unsigned error (MUE, also called mean absolute error/MAE) is another key statistic to include in your results.
-Even though some models' near-perfect correlation and ranking statistics might suggest excellent accuracy, MUE values can still be off multiple kcal/mol, providing important additional insight into performance.
+Even though some models' near-perfect correlation and ranking statistics might suggest excellent accuracy, MUE values can still show multiple kcal/mol of error, providing important additional insight into performance.
 Furthermore, MUE allows for unbiased comparisons between predictive models as it is less sensitive to dataset size.
@@
 \paragraph{Presenting your data}
-Seeing as essentially all alchemical free energy prediction schemes are regression problems, the preferred type of plot is a scatter plot (see Figure \ref{fig:scatterplot_analysis}).
+As essentially all alchemical free energy prediction schemes are regression problems, the preferred type of plot is a scatter plot (see Figure \ref{fig:scatterplot_analysis}).
 Most alchemical FE projects will concern 10-50 ligands; any study with \textless10 ligands is more suitable for bar plots (with inclusion of error bars).
@@
 Prospective prediction challenges such as the drug design data resource grand challenges provide a community driven platform to evaluate different free energy protocols against each other on unknown targets~\cite{gaieb2018d3r, gaieb2019d3r}.Such efforts have highlighted that selection of seemingly identical or similar potential energy function or simulation package does not guarantee production of similar free energies owing to differences in simulation protocols.
-(SUGGEST REMOVING THIS SENTENCE What has come out of these in the past is that SChrödinger's FEP+, performs similar when run by different participant groups~\cite{gaieb2018d3r}, but may not give the desired accuracy with RMSE errors larger than 1 kcal/mol.
-Other alchemical protocols performed similarly well.) We hope that the best practice guide provides a set of tools that allow a better understanding of how to setup, run and interpret reliably alchemical free energy calculations.
+(SUGGEST REMOVING THIS SENTENCE What has come out of these in the past is that Schrödinger's FEP+, performs similar when run by different participant groups~\cite{gaieb2018d3r}, but may not give the desired accuracy with RMSE errors larger than 1 kcal/mol.
+Other alchemical protocols performed similarly well.(CITE GAPSYS PMX paper for large FEP+ comparison) We hope that the best practice guide provides a set of tools that allow a better understanding of how to setup, run and interpret reliably alchemical free energy calculations.
 %
@@
 It only runs on GPUs, using CUDA or OpenCL.
-    \item The molecular operating environment (MOE) offered by the Chemical Computing Group (CCG) has a tool for performing free energy calculations. It is build on AMBER-TI (cf. below).
+    \item The molecular operating environment (MOE) offered by the Chemical Computing Group (CCG) has a tool for performing free energy calculations. It is built on AMBER-TI (cf. below).
     \end{itemize}
@@
 	\begin{itemize}
-	\item FESetup: AMBER, gromacs, Sire
+	\item FESetup: AMBER, GROMACS, Sire
+	\item PMX: GROMACS \url{https://github.com/deGrootLab/pmx}
 	\item Lomap/Lomap2 : Relative alchemical transformation graph planning

### 2020-02-05 paper/manuscript.tex
manuscript updates for theory section

@@
 \begin{tcolorbox}[title=List of Symbols, colback=green!10!white]
+$L$ and $R$ -- generic names for ligand and receptor\\
+$K_b^{\circ}$ -- binding  constant \\
+$c^{\circ}$ -- standard state concentration \\
 $U$ -- potential energy\\
-$u$ -- reduced potential energy \\
-$\Delta G$ -- free energy \\
+$u$ -- reduced potential describing a thermodynamic state \\
+$\Delta G$ -- Gibbs free energy (isothermal isobaric ensemble)\\
+$\Delta A$ -- Helmholtz free energy (canonical ensemble)\\
 $\Delta f$-- reduced (dimensionless) free energy \\
-$K_b$ -- binding or association constant \\
-$k_B$-- Boltzmann constant \\
+$\Delta \hat{f}$ -- estimate from an estimator for the reduced free energy\\
+$\Gamma$ -- conformation space accessible by simulations \\
+$\vec{q}$ -- vector of a single configuration, i.e. $x$, $y$, $z$ coordinates of the simulation system\\
+$k_B$ -- Boltzmann constant \\
+$Z$ -- partition function \\
+$p$ -- pressure \\
+$\mu$ -- chemical potential (grand canonical ensemble)\\
 $T$ -- temperature \\
 $\beta \equiv (k_B T)^{-1}$ -- inverse thermal energy \\
-$K_D$ -- dissociation constant \\
-$\lambda$ -- alchemical progress parameter \\
-$\Gamma$ -- conformation space accessible by simulations \\
-$\vec{q}$ -- single configuration\\
+$\vec{\lambda}$ -- alchemical progress parameter \\
 $g$ -- statistical inefficiency
@@
     \emph{lead optimization} - process of designing a drug candidate from an initial lead compound, usually involving optimizing its affinity, possibly selectivity, and various pharmacokinetic parameters.\\
+    \emph{docking} -- Method of generate simulation starting structures for ligand receptor systems, when no crystal structure is available\\
     \emph{thermodynamic cycle} - sequence of thermodynamic processes (like changes in temperature, volume or pressure) in which the initial and final state are equal.\\
+    \emph{alchemical} -- states or simulations that contain non-physical intermediates \\
 \end{tcolorbox}
@@
 Briefly, this is achieved by simulating a ``chimera'' molecule composed of enough atoms to represent both $A$ and $B$.
-A subset of the energetic terms in $U(\vec{q};\vec{\lambda})$ is then modulated by $\vec{\lambda}$ so that at $\vec{\lambda}_A$, the atoms that form molecule $A$ are activated and those belonging exclusively to $B$ are non-interacting ``dummy atoms'', while the opposite occur at $\lambda_B$ (see Section~\ref{sec:relative-fe-protocol} for details).
+A subset of the energetic terms in $U(\vec{q};\vec{\lambda})$ is then modulated by $\vec{\lambda}$ so that at $\vec{\lambda}_A$, the atoms that form molecule $A$ are activated and those belonging exclusively to $B$ are non-interacting ``dummy atoms'', while the opposite occur at $\vec{\lambda}_B$ (see Section~\ref{sec:relative-fe-protocol} for details).
 
@@
 \end{equation}
-While this strategy usually results in sampling thermodynamic states whose Boltzmann distributions is very similar, thus collecting information that is to some degree redundant, some estimators such as the Multistate Bennett acceptance ratio (MBAR)~\cite{shirts2008statisticallya} can exploit similarities between states to improve the precision of the estimates by using the conformations sampled in all alchemical states $\{\vec{\lambda}_k \}$ to compute the free energy difference $\Delta f(\vec{\lambda}_i, \vec{\lambda}_{j})$ between any pair of states $i,j$  (see Section~\ref{subsec:estimators}).
+While this strategy usually results in sampling thermodynamic states whose Boltzmann distributions is very similar, thus collecting information that is to some degree redundant, some estimators such as the Multistate Bennett acceptance ratio (MBAR)~\cite{shirts2008statisticallya} can exploit similarities between states to improve the precision of the estimates by using the conformations sampled at all alchemical states $\{\vec{\lambda}_k \}$ to compute the free energy difference $\Delta f(\vec{\lambda}_i, \vec{\lambda}_{j})$ between any pair of states $i,j$  (see Section~\ref{subsec:estimators}).
 
@@
 \label{sec:step0}
-\todo[inline, color={green!20}]{ASJSM: @everyone please review this section}
 When starting an alchemical free energy project, a key first step is to decide whether free energy calculations are really the right tool. Particularly, count the cost of your
@@
 be worth it in terms of human and computational cost?
+
 \subsection{How accurate are alchemical free energy calculations?}
 \label{subsec:expectation}
-%
-Current alchemical free energy calculations involving small molecules seem to achieve, in favorable cases, RMS errors around 1-2 kcal/mol depending on force field, system, and a variety of other factors such as simulation time, sampling method, and whether the calculations employed are absolute or relative.
+Current alchemical free energy calculations involving small molecules seem to achieve, in favorable cases, root mean square (RMS) errors around 1-2 kcal/mol depending on force field, system, and a variety of other factors such as simulation time, sampling method, and whether the calculations employed are absolute or relative.
 A small selection of benchmark datasets and case studies can be found in section~\ref{sec:benchmark} at the end of this document.
@@
 Additional factors such as slow protein or ligand rearrangements, uncertainties in ligand binding mode, or charged ligands can make these calculations far less reliable and more of a research effort.
-%
 
@@
 This prompts us to consider another important aspect.
-It's important to be clear on what error is being reported: $\Delta G$ after shifting by a constant to minimize the RMSE, unshifted $\Delta G$, $\Delta \Delta G$ of computed edges, or $\Delta \Delta G$ of all edges.
-In general, the computed edge $\Delta \Delta G$ RMSE significantly underestimates the error of all edge RMSE: \url{https://github.com/jchodera/jacs-dataset-analysis}.
+It is important to be clear on what error to report: $\Delta G$ after shifting by a constant to minimize the RMSE, unshifted $\Delta G$, $\Delta \Delta G$ of computed edges, or $\Delta \Delta G$ of all edges.
+(See also the section recommendations for best practices of reporting data Sec.~\ref{sec:plot_data}) In general, the computed edge $\Delta \Delta G$ RMSEs significantly underestimates the error of all edge RMSEs: \url{https://github.com/jchodera/jacs-dataset-analysis}.
 Given the need to understand the performance of the system with alchemical free energy calculations, we recommend that retrospective studies for a particular target and a particular chemical series be performed for each particular application case.
-%
+
 \subsection{How reproducible are alchemical free energy calculations?}
@@
 In simple cases such as absolute hydration free energies of small organic molecules, or relative hydration free energy calculations between structurally similar small organic molecules, it should be possible to obtain highly precise estimates with a given software package (with a sample standard deviation under 0.01 kcal/mol).
+[DO WE HAVE A CIATION FOR THIS?]
 
@@
 For absolute and relative hydration free energies of small organic molecules a variability of ca.
-0.2 kcal/mol between popular simulation packages has been reported, even after fixes to the underlying simulation packages.\cite{loeffler2018reproducibility} In the recent SAMPL6 SAMPLing challenge a larger variability of 0.3 to 1.0 kcal/mol was noted in the computed absolute binding free energies of host/guest systems even though similar input and simulation parameters were used \cite{rizzi2019sampl6} and, in many cases, single-point energies were identical or nearly so.
+0.2 kcal/mol between popular simulation packages has been reported, even after fixes to the underlying simulation packages~\cite{loeffler2018reproducibility}.
+In the recent SAMPL6 SAMPLing challenge a larger variability of 0.3 to 1.0 kcal/mol was noted in the computed absolute binding free energies of host/guest systems even though similar input and simulation parameters were used~\cite{rizzi2019sampl6} and, in many cases, single-point energies were identical or nearly so.
 Further work is needed to ensure reproducibility of alchemical free energy calculations across implementations to guarantee that force-field development efforts lead to transferable potential energy functions.
-%
+
 \subsection{Is my problem suitable for alchemical free energy calculations?}
@@
 In principle, all of these forms will yield correct binding free energies in the limit of adequate sampling (if their effects and connection to the standard state are correctly handled) but they have different strengths and weaknesses.
-For example, with more involved restraints, sampling at intermediate $\lambda$ values will usually not need to be as extensive but more computational effort must go to computing the restraining free energy.
+For example, with more involved restraints, sampling at intermediate $\vec{\lambda}$ values will usually not need to be as extensive but more computational effort must go to computing the restraining free energy.
 Additionally, such restraints would typically keep the ligand from exploring alternative binding modes, which may be undesirable with Hamiltonian $\lambda$ exchange or expanded ensemble techniques where allowing the ligand to exchange binding modes when it is non-interacting could provide sampling benefits~\cite{wang2013identifying}.
@@
     \includegraphics[width=0.95\linewidth]{figures/fig7_what_is_lam/Figure.pdf}
-    \caption{Alchemical intermediates are created by making the potential energy depend on an additional variable $\lambda$ that interpolates between the chemical endpoints.
-In (\textbf{A}), at $\lambda=0$ the molecule is a fully interacting phenol and at $\lambda=1$,  a fully interacting benzene.
-(\textbf{B}) shows an illustration of the probability distribution of the potential energies as the switching function takes values of $\lambda=0$ to $\lambda=1$.
-Intermediates states are required for a sufficient overlap in potential energies to estimate a free energy difference between $\lambda=0$ and $\lambda=1$.
-    Softcore potentials provide one of the most efficient families of intermediate pathways, with a $\lambda$ dependence illustrated in (\textbf{C}) and a potential energy dependence shown in (\textbf{D}).
-Note how as $\lambda$ approaches 0, the energy smoothly approaches zero at all $r$, a necessary requirement for efficient and stable calculations.
+    \caption{Alchemical intermediates are created by making the potential energy depend on an additional variable $\vec{\lambda}$ that interpolates between the chemical endpoints.
+In (\textbf{A}), at $\vec{\lambda}=0$ the molecule is a fully interacting phenol and at $\vec{\lambda}=1$,  a fully interacting benzene.
+(\textbf{B}) shows an illustration of the probability distribution of the potential energies as the switching function takes values of $\vec{\lambda}=0$ to $\vec{\lambda}=1$.
+Intermediates states are required for a sufficient overlap in potential energies to estimate a free energy difference between $\vec{\lambda}=0$ and $\vec{\lambda}=1$.
+    Softcore potentials provide one of the most efficient families of intermediate pathways, with a $\vec{\lambda}$ dependence illustrated in (\textbf{C}) and a potential energy dependence shown in (\textbf{D}).
+Note how as $\vec{\lambda}$ approaches 0, the energy smoothly approaches zero at all $r$, a necessary requirement for efficient and stable calculations.
 }
@@
 
-The key consideration for choosing alchemical pathways is that the intermediate states that a given pathway produces should sample configurational ensembles that change as slowly as possible as $\lambda$ changes, while still managing to go from the initial state to the final state as $\lambda$ goes from 0 to 1.
+The key consideration for choosing alchemical pathways is that the intermediate states that a given pathway produces should sample configurational ensembles that change as slowly as possible as $\vec{\lambda}$ changes, while still managing to go from the initial state to the final state as $\vec{\lambda}$ goes from 0 to 1.
 
@@
 Fig.~\ref{fig:fig_what_is_lambda} \textbf{A} and \textbf{B} illustrate this.
-Fig.~\ref{fig:fig_what_is_lambda} \textbf{A} shows in a pictorial way a soft-core potential can be applied across different $\lambda$s.
-Fig.~\ref{fig:fig_what_is_lambda} \textbf{B} illustrates the potential energy distributions at the different $\lambda$ intermediates, with sufficient overlap between neighboring $\lambda$ states to ensure that reweighting estimators such as MBAR can be used for analysis (see Sec.~\ref{subsec:estimators}).
+Fig.~\ref{fig:fig_what_is_lambda} \textbf{A} shows in a pictorial way a soft-core potential can be applied across different $\vec{\lambda}$s.
+Fig.~\ref{fig:fig_what_is_lambda} \textbf{B} illustrates the potential energy distributions at the different $\vec{\lambda}$ intermediates, with sufficient overlap between neighboring $\vec{\lambda}$ states to ensure that reweighting estimators such as MBAR can be used for analysis (see Sec.~\ref{subsec:estimators}).
 The actual transformation is best handled with soft-core potentials of the form shown in Fig.~\ref{fig:fig_what_is_lambda} \textbf{C} and \textbf{B}, with more details given below.
 
-So how do we adjust the potentials between the two end states based on $\lambda$? The simplest possible alchemical pathway is a \textit{linear} pathway:
+So how do we adjust the potentials between the two end states based on $\vec{\lambda}$? The simplest possible alchemical pathway is a \textit{linear} pathway:
 \begin{equation}
-U(\vec{q},\lambda) = \lambda U_0(\vec{q}) + (1-\lambda)U_1(\vec{q})   
-\end{equation}
-%
-so-called because the dependence on $\lambda$ is linear.
-This clearly satisfies the basic requirement that it gives the initial endpoint potential energy $U_0(\vec{q})$ when $\lambda=0$ and final endpoint energy $U_1(\vec{q})$ when $\lambda=1$.
+U(\vec{q},\vec{\lambda}) = \vec{\lambda} U_0(\vec{q}) + (1-\vec{\lambda})U_1(\vec{q})   
+\end{equation}
+%
+so-called because the dependence on $\vec{\lambda}$ is linear.
+This clearly satisfies the basic requirement that it gives the initial endpoint potential energy $U_0(\vec{q})$ when $\vec{\lambda}=0$ and final endpoint energy $U_1(\vec{q})$ when $\vec{\lambda}=1$.
 
@@
 However, this approach ends up being terrible for removing or adding repulsive potentials that go to infinity quickly at or near the origin.
-One way to look at this is to examine how low $\lambda$ values must go to reduce the energy at $0.5\sigma$ (the atomic size parameter) down to 1 $k_BT$, where thermal fluctuations make it possible for other atomic sites to penetrate routinely that deep.
-If we are trying to go from a particle being present, and desire to make it disappear alchemically, then if the repulsive terms are of the form $\epsilon\frac{\sigma}{r}^{12}$, then if $\epsilon$ was 1 $k_BT$ at the temperature of interest, and we start with the particle present, then solving for $(1-\lambda)(1 kB_T)\frac{\sigma}{\frac{1}{2}\sigma}) = 1 k_B T$ we get $\lambda = 1-2^{-12} \sim 0.999976$.
+One way to look at this is to examine how low $\vec{\lambda}$ values must go to reduce the energy at $0.5\sigma$ (the atomic size parameter) down to 1 $k_BT$, where thermal fluctuations make it possible for other atomic sites to penetrate routinely that deep.
+If we are trying to go from a particle being present, and desire to make it disappear alchemically, then if the repulsive terms are of the form $\epsilon\frac{\sigma}{r}^{12}$, then if $\epsilon$ was 1 $k_BT$ at the temperature of interest, and we start with the particle present, then solving for $(1-\vec{\lambda})(1 kB_T)\frac{\sigma}{\frac{1}{2}\sigma}) = 1 k_B T$ we get $\vec{\lambda} = 1-2^{-12} \sim 0.999976$.
 So we have gone virtually all the way to the end of the transformation, but there is still an impenetrable post in the middle of our simulation!
 This is not very much like the desired final state of no interactions between the particle and its environment.
-We can play around with a few ways of modifying this, like simulating many more intermediate states near $\lambda=1$.
+We can play around with a few ways of modifying this, like simulating many more intermediate states near $\vec{\lambda}=1$.
 However, various analyses have shown that this is not a very good strategy~\cite{pham2011identifying, beutler1994avoiding, zacharias1994separationshifted, blondel2004ensemble, gapsys2012new}.
@@
 \begin{equation}
-    U(\vec{r_{ij}},\lambda) = 4\epsilon_{ij} \lambda \left(\frac{1}{(\alpha(1-\lambda) + (r_{ij}/\sigma_{ij})^6)^2} -  \frac{1}{\alpha(1-\lambda) + (r_{ij}/\sigma_{ij})^6}\right)
+    U(\vec{r_{ij}},\vec{\lambda}) = 4\epsilon_{ij} \vec{\lambda} \left(\frac{1}{(\alpha(1-\vec{\lambda}) + (r_{ij}/\sigma_{ij})^6)^2} -  \frac{1}{\alpha(1-\vec{\lambda}) + (r_{ij}/\sigma_{ij})^6}\right)
     \label{eq:softcore}
@@
 Where $r_{ij}$ is the distance between two particles $i$ and $j$, $\epsilon_{ij}$ and $\sigma_{ij}$ are the Lennard-Jones parameters corresponding to the interaction between particles $i$ and $j$, and $\alpha$ is a constant (0.5 is optimal for the functional form shown above).
-This functional form has exactly the property we are looking for: it recovers the Lennard-Jones potential when $\lambda=1$, and the other endpoint ($\lambda=0$), it is exactly zero for all $r_{ij}$ everywhere, and as $\lambda$ goes to zero, the  $\alpha(1-\lambda)$ term lowers the infinite energy in the core.
+This functional form has exactly the property we are looking for: it recovers the Lennard-Jones potential when $\vec{\lambda}=1$, and the other endpoint ($\vec{\lambda}=0$), it is exactly zero for all $r_{ij}$ everywhere, and as $\vec{\lambda}$ goes to zero, the  $\alpha(1-\vec{\lambda})$ term lowers the infinite energy in the core.
 There are several different variants of the same functional form~\cite{zacharias1994separationshifted, beutler1994avoiding,pham2011identifying}, but the one  given in eq.~\ref{eq:softcore} is easy to understand and implement and fairly numerically stable.
@@
 What about performing both transformations at the same time?
-We can not turn off the electrostatics linearly at the same time we turn off the Lennard-Jones terms, as it would leave infinitely large attractive and repulsive electrostatic terms "bare" at small $\lambda$, resulting in the simulation crashing.
+We can not turn off the electrostatics linearly at the same time we turn off the Lennard-Jones terms, as it would leave infinitely large attractive and repulsive electrostatic terms "bare" at small $\vec{\lambda}$, resulting in the simulation crashing.
 It \textit{is} possible to apply the same soft core approach to the Coulomb interaction and this is indeed done in a number of implementations, in which case it is important that the Coulomb interaction is softened more rapidly than the Lennard-Jones interaction to avoid charge penetration issues, this can be tricky to ensure for all types of perturbations ~\cite{steinbrecher2011softcore}.
@@
 \label{sec:sampling_schemes}
-Though all alchemical simulations must sample from multiple $\lambda$ states, different approaches can be used to achieve this.
+Though all alchemical simulations must sample from multiple $\vec{\lambda}$ states, different approaches can be used to achieve this.
 Fig.~\ref{fig:fig_sampling_scheme} illustrates the four most common schemes.
-The simplest approach involves running an independent simulation at each of the reuqisite $\lambda$ values (see Fig.~\ref{fig:fig_sampling_scheme}\textbf{A}).
+The simplest approach involves running an independent simulation at each of the predefined $\vec{\lambda}$ values (see Fig.~\ref{fig:fig_sampling_scheme}\textbf{A}).
 This type of scheme is currently used for AMBER TI calculations~\cite{song2019using} and for Siremol/Openmm~\cite{hedges2019biosimspace}.
 However, if these simulations can be run simultaneously with communication between them, a simple extension allows mixing between these replicas.
-In this approach, the simulation at each $\lambda$ can undergo periodic exchanges with neighboring $\lambda$ values.
+In this approach, the simulation at each $\vec{\lambda}$ can undergo periodic exchanges with neighboring $\vec{\lambda}$ values.
 This form of replica exchange (Hamiltonian replica exchange) is based on ideas developed from Monte Carlo simulations of spin glasses by Swendsen and Wang~\cite{swendsen1986replica}.
@@
 A third approach borrows ideas from simulated tempering~\cite{marinari1992simulateda}.
-In this scheme a single replica rapidly explores all of $\lambda$ space by working out optimal weights that allow switching between different intermediate $\lambda$ values, as seen in Fig.~\ref{fig:fig_sampling_scheme}(\textbf{C}).
+In this scheme a single replica rapidly explores all of $\vec{\lambda}$ space by working out optimal weights that allow switching between different intermediate $\vec{\lambda}$ values, as seen in Fig.~\ref{fig:fig_sampling_scheme}(\textbf{C}).
 This approach is also referred to as self adjusted mixture sampling~\cite{lyubartsev1992newa, li2007simulated, tan2017optimally} and while promising, has so far only been supported in OpenMM Tools~\cite{andrearizzi2019choderalab}, and to date has not seen published applications for binding free energy calculations.
 The last approach makes use of non-equilibrium simulations~\cite{aldeghi2018accurate}.
-In this approach, only end state $\lambda$ replicas ($\lambda$=0, $\lambda=1$) are simulated at equilibrium; intermediate information is generated from non-equilibrium simulations that rapidly transition between endstates.
+In this approach, only end state $\vec{\lambda}$ replicas ($\vec{\lambda}$=0, $\vec{\lambda}=1$) are simulated at equilibrium; intermediate information is generated from non-equilibrium simulations that rapidly transition between endstates.
 This approach is available in GROMACS and appears to be coming online in several other packages.
@@
 \textbf{B}: Hamiltonian replica exchange scheme.
-\textbf{C}: Single replica scheme sampling from all $\lambda$ states.
+\textbf{C}: Single replica scheme sampling from all $\vec{\lambda}$ states.
 \textbf{D}: Non-equilibrium sampling scheme.}
@@
 Currently, we recommend using Hamiltonian replica exchange type sampling schemes (Fig.~\ref{fig:fig_sampling_scheme}\textbf{B}).
-If these are not available in the code of choice, running independent simulations at different $\lambda$ values can be acceptable, especially when conformational sampling is fast (Fig.~\ref{fig:fig_sampling_scheme} \textbf{A}).
+If these are not available in the code of choice, running independent simulations at different $\vec{\lambda}$ values can be acceptable, especially when conformational sampling is fast (Fig.~\ref{fig:fig_sampling_scheme} \textbf{A}).
 Single replica schemes and non-equilibrium schemes are not as established yet, but are very promising.
@@
 Convergence in other alchemical observables should also be monitored to determine if the defined phase space has been sufficiently sampled and enough decorrelated samples have been drawn.
-These additional observables include, but are not limited to, the variance in $\frac{dU}{d\lambda}$ across all $\lambda$ values, calculating the variance in free energy using bootstrap analysis, and comparing differences in free energies calculated using different percentages of the simulation in both the forward and reverse directions ( see Fig.
+These additional observables include, but are not limited to, the variance in $\frac{dU}{d\vec{\lambda}}$ across all $\vec{\lambda}$ values, calculating the variance in free energy using bootstrap analysis, and comparing differences in free energies calculated using different percentages of the simulation in both the forward and reverse directions ( see Fig.
 ~\ref{fig:convergence_forward_reverse}).
@@
 Each method requires different data from the simulation be collected.
-If, for instance, the free energy estimator selected is thermodynamic integration, then values of $\frac{dU}{d\lambda}$ at uncorrelated data points must be collected.
+If, for instance, the free energy estimator selected is thermodynamic integration, then values of $\frac{dU}{d\vec{\lambda}}$ at uncorrelated data points must be collected.
 Once a combination of knowing what type of simulation you will run, which alchemical topology you will simulate, what alchemical path you will simulation along, and what your stopping conditions are, then you are ready to enumerate the information you should capture.
@@
 \begin{itemize}
-    \item Thermodynamic Integration (TI) requires $\frac{\partial U(\vec{x})}{\partial\lambda}$.
+    \item Thermodynamic Integration (TI) requires $\frac{\partial U(\vec{x})}{\partial\vec{\lambda}}$.
     \item Exponential Averaging (EXP) needs \textit{either} $\Delta U_{k,k+1}(\vec{x})$ or $\Delta U_{k,k-1}(\vec{x})$, depending on the direction its being evaluated in.
@@
 Although TI must usually be calculated in code, as it requires the derivative, there is one condition under which it actually has the fastest computation time. 
-If the alchemical path you have chosen is a linear alchemical path, then you get $\frac{dU}{d\lambda} = U_0(\vec{x}) - U_1(\vec{x})$, which is the difference between the initial and final states. 
+If the alchemical path you have chosen is a linear alchemical path, then you get $\frac{dU}{d\vec{\lambda}} = U_0(\vec{x}) - U_1(\vec{x})$, which is the difference between the initial and final states.
 However, because of the problems with linear paths already discussed in this paper, this simplification is rarely that useful.
@@
 Typically, the timeseries of note $a_t$ analyzed in automated equilibration detection is the negative logarithm of the probability density sampled by the MCMC algorithm (up to an irrelevant additive constant).
-For simple independent simulations that sample $x_t \sim \pi(x ; \lambda)$, this is given by the reduced potential
+For simple independent simulations that sample $x_t \sim \pi(x ; \vec{\lambda})$, this is given by the reduced potential
 \begin{eqnarray}
-a_t &\equiv& - \ln \pi(x_t; \lambda) + c = u(x_t; \lambda) .
+a_t &\equiv& - \ln \pi(x_t; \vec{\lambda}) + c = u(x_t; \vec{\lambda}) .
 \end{eqnarray}
@@
 %
-If independent simulations are used, the alchemical state $\lambda$ may have a significant impact on the correlation time, and these simulations should be subsampled independently using a separate estimate of the statistical inefficinecy $g$ for each alchemical state.
+If independent simulations are used, the alchemical state $\vec{\lambda}$ may have a significant impact on the correlation time, and these simulations should be subsampled independently using a separate estimate of the statistical inefficinecy $g$ for each alchemical state.
 If coupled simulations are used (such as a Hamiltonian replica exchange simulation), the replicas should undergo equivalent random walks in alchemical space, and the replicas can be can be subsampled with the same $g$ to generate an equal number of uncorrelated samples at each alchemical state.
@@
 Described in words, we take the samples generated during our run with the potential energy function $U_0(\vec{q})$, recalculate what the energy difference would be if we switched to potential energy function $U_1(\vec{q})$, and average the exponential of the -1 times the energy difference to get -1 times the exponential of the free energy difference.
-The original distributions (P($U_0$) as generated at $\lambda=0$ and P($U_1$) would look like the ones seen in figure~\ref{fig:fig_sampling_scheme} (A)-(C)on the left hand side.
+The original distributions (P($U_0$) as generated at $\vec{\lambda}=0$ and P($U_1$) would look like the ones seen in figure~\ref{fig:fig_sampling_scheme} (A)-(C)on the left hand side.
 Reevaluating  requires almost no extra code functionality to perform; one need only to save a full precision trajectory, and run an unmodified molecular simulation code using the $U_1$ in order to calculate the new energies of stored snapshots.
@@
 By taking the derivative of the free energy with respect to the
-variable $\lambda$, we find that:
+variable $\vec{\lambda}$, we find that:
 \begin{equation}
-\frac{df}{d\lambda} = \frac{d}{d\lambda} \left[-\ln \int \frac{\exp{-u(\lambda,\vec{q})}}{Q(\lambda)} d\vec{q}\right] = \expect{\frac{du(\lambda,\vec{q})}{d\lambda}\
-}_{\lambda} 
-\end{equation}
-And than we can then numerically integrate $df/d\lambda$ over an alchemical transformation, using a range of different well-established techniques, to obtain:
+\frac{df}{d\vec{\lambda}} = \frac{d}{d\vec{\lambda}} \left[-\ln \int \frac{\exp{-u(\vec{\lambda},\vec{q})}}{Q(\vec{\lambda})} d\vec{q}\right] = \expect{\frac{du(\vec{\lambda},\vec{q})}{d\vec{\lambda}}\
+}_{\vec{\lambda}} 
+\end{equation}
+And than we can then numerically integrate $df/d\vec{\lambda}$ over an alchemical transformation, using a range of different well-established techniques, to obtain:
 \begin{equation}
-\Delta f    = \int_{0}^{1} \expect{\frac{du(\lambda,\vec{q})}{d\lambda}}_{\lambda}  d\lambda    
+\Delta f    = \int_{0}^{1} \expect{\frac{du(\vec{\lambda},\vec{q})}{d\vec{\lambda}}}_{\vec{\lambda}}  d\vec{\lambda}    
 \end{equation}
 This approach to calculating the free energy is called thermodynamic integration (TI).
-Averaging over $\expect{\frac{du}{d\lambda}}$ requires fewer uncorrelated samples to reach a given level of relative error
+Averaging over $\expect{\frac{du}{d\vec{\lambda}}}$ requires fewer uncorrelated samples to reach a given level of relative error
 than averaging $e^{-u(\vec{q})}$, as the range is usually narrower with a more Gaussian distribution.
@@
 \[ \Delta f \approx \sum_{k=1}^{K} w_k
-\expect{\frac{du(\lambda,\vec{q})}{d\lambda}}_{k} \] where the weights $w_k$ correspond to a particular choice of numerical integration.
+\expect{\frac{du(\vec{\lambda},\vec{q})}{d\vec{\lambda}}}_{k} \] where the weights $w_k$ correspond to a particular choice of numerical integration.
 Researchers have tried a large number of different integration schemes~\cite{resat1993studies,jorge2010effect,shyu2009reducing}.
-However, many other integration choices require specific choices of $\lambda$ to minimize bias, which makes them unsuitable when the intermediates
+However, many other integration choices require specific choices of $\vec{\lambda}$ to minimize bias, which makes them unsuitable when the intermediates
 have widely-varying levels of uncertainty.
 For example, integrating a cubic spline interpolation provided negligible benefits~\cite{paliwal2011benchmark}.
-For starting researchers, we therefore recommend a simple trapezoid rule scheme, as it allows for maximal flexibility in which values of $\lambda$ are simulated.
+For starting researchers, we therefore recommend a simple trapezoid rule scheme, as it allows for maximal flexibility in which values of $\vec{\lambda}$ are simulated.
 Because fitting to higher order polynomials can have numerical instabilities for some energy functions, and because alternate functional forms might only be appropriate with some types of transformations, expertise and experience is required to perform such numerical integration modifications.
@@
 %
-One drawback of TI is that it requires derivatives with respect to $\lambda$ to be calculated directly in the code.
+One drawback of TI is that it requires derivatives with respect to $\vec{\lambda}$ to be calculated directly in the code.
 Unfortunately, many problems of interest require using pathways (such as the soft-core pathways, for removing repulsive interactions) that are not linear, as we discuss, making this more complex.
-Still, if the code of interest does compute $\frac{du}{d\lambda}$, then TI is perhaps the simplest method to use, as it involves a very little post-processing.
+Still, if the code of interest does compute $\frac{du}{d\vec{\lambda}}$, then TI is perhaps the simplest method to use, as it involves a very little post-processing.
 %
@@
 \item We recommend MBAR if all energy differences are available. It is probably the lowest variance free energy estimate given samples from multiple states.
-\item BAR is essentially just as good as MBAR for highly optimized $\lambda$ intermediates.
-Specifically, if the $\lambda$s are chosen such that intermediate states have moderate overlap with their neighbors (i.e.
+\item BAR is essentially just as good as MBAR for highly optimized $\vec{\lambda}$ intermediates.
+Specifically, if the $\vec{\lambda}$s are chosen such that intermediate states have moderate overlap with their neighbors (i.e.
 between $i$ and $i+1$ and between $i$ and $i-1$, they will \textit{not} have significant overlap with their next nearest neighbors $i+2$ and $i-2$.
@@
 
-The computation of free energies using TI (section~\ref{subsec:estimators}) is straightforward and the trapezoidal rule is often recommended since it allows unequal spacing of $\lambda$ states, which is required to minimize the variance in the free energy estimate, but in principle any good numerical integration method can be used.
-The determination of regions of high curvature when estimating the integral is helpful to determine regions of phase space where more sampling and/or more $\lambda$ states are necessary to obtain the best approximation of the integral.
-Generally a plot of $\lambda$ with respect to the gradients at each of the $\lambda$ values can be helpful here.
+The computation of free energies using TI (section~\ref{subsec:estimators}) is straightforward and the trapezoidal rule is often recommended since it allows unequal spacing of $\vec{\lambda}$ states, which is required to minimize the variance in the free energy estimate, but in principle any good numerical integration method can be used.
+The determination of regions of high curvature when estimating the integral is helpful to determine regions of phase space where more sampling and/or more $\vec{\lambda}$ states are necessary to obtain the best approximation of the integral.
+Generally a plot of $\vec{\lambda}$ with respect to the gradients at each of the $\vec{\lambda}$ values can be helpful here.
 Additionally, computation of the overall variance of TI requires the calculation of the overall variance of integration, rather than each individual $\Delta$G$_{i,i+1}$ and assuming variances add independently.
-Therefore, $\mathrm{var}$($\Delta$f) = $\sum_{i=1}^{K}w_{k}^2 \mathrm{var}(\frac{du}{d\lambda})_{k}$.
-%
-For alchemical changes that result in smooth, low curvature sets of $\expect{\frac{dU}{d\lambda}}$, a relatively small number of $\lambda$ states is necessary for sufficient accuracy and low variance in the free energy estimate.
-However, with increasing numbers of perturbed atoms, the bias introduced by discretization of the integral can become large due to increased curvature, and more $\lambda$ intermediate states become necessary to reduce error.
+Therefore, $\mathrm{var}$($\Delta$f) = $\sum_{i=1}^{K}w_{k}^2 \mathrm{var}(\frac{du}{d\vec{\lambda}})_{k}$.
+%
+For alchemical changes that result in smooth, low curvature sets of $\expect{\frac{dU}{d\vec{\lambda}}}$, a relatively small number of $\vec{\lambda}$ states is necessary for sufficient accuracy and low variance in the free energy estimate.
+However, with increasing numbers of perturbed atoms, the bias introduced by discretization of the integral can become large due to increased curvature, and more $\vec{\lambda}$ intermediate states become necessary to reduce error.
 It is recommended that researchers verify that a sufficient number of states are included such that the free energy is essentially invariant to the number of lambda intermediate states chosen.
@@
 \paragraph{Overlap matrix}
-One way of assessing reliability of the calculations is checking the phase space overlap between neighboring $\lambda$-windows~\cite{wu2005phasespaceb, wu2005phasespacec}.
+One way of assessing reliability of the calculations is checking the phase space overlap between neighboring $\vec{\lambda}$-windows~\cite{wu2005phasespaceb, wu2005phasespacec}.
 For this purpose, a so-called overlap matrix $\mathcal{O}$ can be used.
 $\mathcal{O}$ is a $K\times K$ matrix, with $K$ being the number of simulated states, i.e.
-$\lambda$s.
+$\vec{\lambda}$s.
 Sufficient overlap is important for reweighting estimators such as BAR or MBAR, but cannot help assess reliability of estimates when using TI.
@@
 \caption{\label{fig:overlap} \textbf{Overlap matrices:} Visualising overlap matrices can help with assessing the quality of simulation data.
-(\textbf{A}) shows good overlap with all first off-diagonal entries well above 0.03, the suggested threshold, (\textbf{B}) is an example of mediocre overlap with good overlap at lower $\lambda$ values and poor overlap at high $\lambda$ values.
+(\textbf{A}) shows good overlap with all first off-diagonal entries well above 0.03, the suggested threshold, (\textbf{B}) is an example of mediocre overlap with good overlap at lower $\vec{\lambda}$ values and poor overlap at high $\vec{\lambda}$ values.
 (\textbf{C}) shows poor overlap resulting in disconnected simulations with unreliable MBAR estimates.}
@@
 In the case of Fig.~\ref{fig:overlap}(\textbf{B}), the overlap is strongly diminishing in the lower right corner, raising concerns regarding the reliability of the free energy estimate obtained.
-For Fig.~\ref{fig:overlap}(\textbf{C}), the state at $\lambda$ index = 6 is connected to neither of its neighbouring states.
+For Fig.~\ref{fig:overlap}(\textbf{C}), the state at $\vec{\lambda}$ index = 6 is connected to neither of its neighbouring states.
 While this does not necessarily imply that the result for this perturbation is wrong, the energy estimate must at least be considered as highly unreliable.
-In order to overcome the issue of poor overlap in this example, additional sampling should be performed by introducing additional states, i.e. $\lambda$ windows.
-%
-Interestingly, as the variance is inversely correlated with the number of states~\cite{klimovich2015guidelines}, it can in principle be reduced below any arbitrary threshold with enough simulation time and a large enough number of $\lambda$ windows.
+In order to overcome the issue of poor overlap in this example, additional sampling should be performed by introducing additional states, i.e. $\vec{\lambda}$ windows.
+%
+Interestingly, as the variance is inversely correlated with the number of states~\cite{klimovich2015guidelines}, it can in principle be reduced below any arbitrary threshold with enough simulation time and a large enough number of $\vec{\lambda}$ windows.
 However, decreasing the variance to a value close to 0 is not feasible, as this approach would strongly increase the calculation time.
-While variance can be decreased both by increasing simulation length and by increasing the number of $\lambda$ values, if overlap is poor, the latter approach often has a larger immediate impact.
-Specifically, often the goal is to adjust the $\lambda$ spacing to perform additional sampling in the areas of poor overlap, rather than using arbitrarily spaced windows.
+While variance can be decreased both by increasing simulation length and by increasing the number of $\vec{\lambda}$ values, if overlap is poor, the latter approach often has a larger immediate impact.
+Specifically, often the goal is to adjust the $\vec{\lambda}$ spacing to perform additional sampling in the areas of poor overlap, rather than using arbitrarily spaced windows.
 Different approaches are described in section~\ref{sec:simulation_protocol_choice} and more details can be found in the literature~\cite{dakka2018concurrent, hahn2019alchemical}.

### 2020-02-05 paper/manuscript.tex
Section 4 and 5 clean up

@@
 An important consideration is how reproducible alchemical free energy calculations are in practice.
-In simple cases such as absolute hydration free energies of small organic molecules, or relative hydration free energy calculations between structurally similar small organic molecules, it should be possible to obtain highly precise estimates with a given software package (with a sample standard deviation under 0.01 kcal/mol).
-[DO WE HAVE A CIATION FOR THIS?]
-
-For more complex use cases such as protein-ligand binding free energies the repeatability is often substantially worse.
+In simple cases such as absolute hydration free energies of small organic molecules, or relative hydration free energy calculations between structurally similar small organic molecules, it should be possible to obtain highly precise estimates with a given software package (with a sample standard deviation under 0.01 kcal/mol)~\cite{rizzi2019sampl6}.
+
+For more complex use cases such as protein-ligand binding free energies the repeatability is often substantially worse~\cite{rizzi2019sampl6}.
 A good practice is to perform two or three runs of the same perturbation to assess repeatability with a given protocol.
@@
 
+\todo[inline, color={green!20}]{ASJSM: @JDC, Can you check the issues below? }
 Sometimes free energy calculations can provide answers that are not
@@
 \end{equation}
-%
+
 Since binding experiments cannot resolve $K_D^\ast$ and $K_C$ individually, such experiments cannot address the basis of selectivity of the type II inhibitors.
 Absolute binding free energy calculations, in contrast, can take advantage of the slow kinetics of DFG-in/out conversion, and estimate the conformation-specific binding constant $K_D^\ast$, thus yielding clues as to the source of selectivity.
-%
+
 \subsection{Is the expected accuracy of the computation sufficient?}
@@
 synthesized, free energy calculations can be appealing even with
-accuracy in the 1--2 kcal/mol range~\cite{mobley2012perspective}, but if the number of compounds to be synthesized is very small, this accuracy may not be enough to provide
-much value.
+accuracies in the 1--2 kcal/mol range~\cite{mobley2012perspective}, but if the number of compounds to be synthesized is very small, this accuracy may not be enough to provide much value.
 
@@
 
-If the accuracy
-of free energy calculations is $\sigma = 1$ kcal/mol, $P(\Delta\Delta
+If the accuracy of free energy calculations is $\sigma = 1$ kcal/mol, $P(\Delta\Delta
 G<0|\Delta\Delta G^\dagger<0) = 0.35$, which means that out of every
@@
 prohibitive, or at least should be carefully considered. Alternatively, if key motions are fast and minimal, only much shorter simulations may be necessary.
-%
+
 \subsection{Can I afford the calculation?}
 \label{subsec:affordability}
-Furthermore, are available computational resources sufficient that throughput will be
-reasonable compared to needs of experimental collaborators working on
+Furthermore, are available computational resources sufficient that throughput will be reasonable compared to needs of experimental collaborators working on
 this system? How many ligands ($N$) can you afford to handle given
 your computational resources? As cloud computing becomes more available, in-house GPU clusters may not be necessary if calculations are not run on a regular basis.
-This analysis should be done up front, part of ``counting the cost''
+This analysis should be done up front as part of ``counting the cost''
 of involvement in a particular project. In some cases, the analysis may conclude that free energy calculations will not be feasible for the proposed problem.
@@
 Such issues should be considered before deciding to conduct binding free energy calculations.
-%
+
 \subsection{Is an exploratory study what I want?}
@@
 \label{sec:drugdiscovery}
-%
 Many practitioners expect alchemical methods to provide valuable guidance for drug discovery, and to exhibit accuracy superior to most alternative approaches for suitable targets~\cite{kuhn2017prospective}.
-Successful application in industry may require considerable knowledge of the "domain of applicability" of free energy calculations -- where they work well and where they will not~\cite{sherborne2016collaborating}.
+Successful application in industry may require considerable knowledge of the ``domain of applicabilit'' of free energy calculations -- where they work well and where they will not~\cite{sherborne2016collaborating}.
 Successful application also requires robust protocols for preparing, submitting and analysing alchemical calculations.
@@
 
-To ensure modeling matches experiment, we also need to accurately model and  simulate the same system -- which requires understanding what protein construct is used in the bioassay.
+To ensure modeling matches experiment, we also need to accurately prepare and  simulate the same system -- which requires understanding what protein construct is used in the bioassay.
 For instance, does the X-ray structure that is to be used for the calculations match the construct used for screening (i.e.
@@
 If multiple structures are available there is some evidence the higher affinity complex can give better performance~\cite{perez-benito2019predicting}, at least in some cases.
-However, ligands and proteins can also undergo unexpected changes in binding mode for related ligands, which can make these issues more complex to deal with.
+However, ligands and proteins can also undergo unexpected changes in binding mode for related ligands, which can make these issues more complex to deal with~\cite{mey2016blinded}.
 
@@
 Can crystallographic waters be retained?
-Do they clash with some of the larger ligands used in the FEP?
-If so, how will the solvent be incorporated for smaller ligands that require the water – consider pre-solvation methods such as GCMC steps~\cite{michel2010prediction}.
+Do they clash with some of the larger ligands used in the alchemical perturbation?
+If so, how will the solvent be incorporated for smaller ligands that require the water -- consider pre-solvation methods such as GCMC steps~\cite{michel2010prediction}.
 Other tools such as WaterMap or open source equivalents (SSTmap, others) can be used to define water structure for systems with no experimental evidence of water sites.
 Before launching large numbers of alchemical free energy calculations it is recommended to test the system using classical MD simulations and limited numbers of alchemical perturbations.
-Metrics such as ligand and protein RMSD and RMSF can be inspected, along with visual inspection of simulations, to ensure the system is stable and likely to be suitable for FE calculations.
-
-Running binding FE calculations in a drug discovery application will typically require the use of software or tools to facilitate the large number of calculations.
+Metrics such as ligand and protein RMSD and RMSF can be inspected, along with visual inspection of simulations, to ensure the system is stable and likely to be suitable for alchemical calculations.
+
+Running free energy of binding calculations in a drug discovery application will typically require the use of software or tools to facilitate the large number of calculations.
 The commercial implementation from Schr\"{o}dinger, FEP+, uses a graphical user interface and once the user has calculated any missing force field parameters, calculations can be prepared and submitted within minutes.

### 2020-02-05 paper/manuscript.tex
Working my way through section 5/6

@@
 The software builds the network of perturbations, connecting similar compounds based on LOMAP methodology~\cite{liu2013lead}, all inputs for the Desmond simulations are built automatically and the job can be saved or run immediately on available GPUs.
-Commercial tools can be expensive in some cases, but alternative tools are becoming available to allow relatively straightforward alchemical FE calculations with other software~\cite{gapsys2015pmx, loeffler2015fesetup, song2019using, gapsys2020large,  jespers2019qligfep}.
+Commercial tools can be expensive in some cases, but non-comercial tools are becoming more straight forward to use to run alchemical free energy calculations~\cite{gapsys2015pmx, loeffler2015fesetup, song2019using, gapsys2020large,  jespers2019qligfep, hedges2019biosimspace}.
 
@@
 For instance, if the goal is to accurately assess the relative binding energy of a small number of compounds, possibly with challenging synthesis, the map of perturbations should contain as many connections between compounds as affordable.
-However, when running calculations on hundreds of compounds a so called ‘star-map’ can be used that just contains one connection per compound: perturbing every compound to a central ligand, typically the crystal structure ligand~\cite{konze2019reactionbased}.
+However, when running calculations on hundreds of compounds a so called \emph{star-map} (see Fig.~\ref{fig:fig_types_of_networks} \textbf{A}) can be used that just contains one connection per compound: perturbing every compound to a central ligand, typically the crystal structure ligand~\cite{konze2019reactionbased}.
 In this way the top-ranking examples can be readily identified and submitted to additional calculations in a second round.
@@
 
-%
 \subsection{Making predictions, understanding errors}
@@
 Beyond this, several parameters provide useful indicators of performance.
-Results suggest MBAR errors greater than 0.3 kcal/mol for perturbations can be indicative of inaccuracy~\cite{perez-benito2019predicting}.
-In addition, the hysteresis can be checked by comparing the difference for forward and reverse perturbations of the same structural change, or in the case of large maps of perturbations cycle closure errors~\cite{wang2013modeling} indicate problematic perturbations involved in cycles connecting many compounds.
+For example error estimates provided by free energy estimators that are too large can highlight poorly converged simulations~\cite{perez-benito2019predicting}.
+Hysteresis, within cycles in the perturbation network or between forward and backward perturbations can be checked~\cite{wang2013modeling} indicate problematic perturbations involved in cycles connecting many compounds (See also Secs.~\ref{sec:relative-fe-protocol} and~\ref{sec:are-they-good}).
 Selection bias, as described above, can be problematic in prospective applications as projects normally seek to focus only on the most potent compounds.
 Thus, correlation between prediction and subsequent experiment may lead to disappointment but corrections can be applied based on previous recommendations~\cite{abel2017critical}.
-
-In summary, the successful use of alchemical calculations in drug discovery requires working in the domain of applicability, using a high quality X-ray structure of the target bound to compounds in the series, and testing the approach retrospectively to ensure the system setup is well-behaved.
+For a more detailed discussion on checking the robsustness of your alchemcial free energy calculation see also Section~\ref{sec:are-they-good}.
+
+In summary, the successful use of alchemical calculations not only in, but particularly for drug discovery requires working in the domain of applicability, using a high quality X-ray structure of the target bound to compounds in the series, and testing the approach retrospectively to ensure the system setup is well-behaved.
 Always assess your confidence in the resulting predictions and communicate this when discussing with experimentalists.
 Consider performing repeat calculations for at least some of the perturbations in the study.
-
-In a drug discovery setting it is inevitable to consider larger virtual libraries.
-In these cases again evaluate if shorter simulation time per $\lambda$ window can be sufficient for a ‘virtual screening’ mode of application.
-It is expected that future tools may allow efficient exploration of large virtual libraries.
-There are many accounts of success of alchemical FE calculations, the methods show good performance towards the goal of binding energy prediction.
+ There are many accounts of success of alchemical calculations, the methods show good performance towards the goal of binding energy prediction.
 However, it is important to have realistic expectations.
@@
 In the meantime, the use cases are expanding to resistance prediction, disease mutation prediction, solubility prediction – an exciting future for alchemical FE calculations.
-%
-%
+
 %
@@
 a free energy of binding or a free energy of hydration.
-Different types of simulations require different considerations for ligands, solvent, and host molecules (in the case of the estimation of free energies of binding).
+Different types of simulations require different choices for ligands, solvent, and host molecules (in the case of the estimation of free energies of binding).
 
@@
 For examples, see HSP90 which formed part of the D3R grand challenge 2015~\cite{mey2016blinded}, A2A~\cite{brucemacdonald2018ligand}, MUP~\cite{ross2015water}, and others~\cite{michel2009energetics}.
-%
+
 \paragraph{How to align congeneric series}
@@
 If the binding site location is uncertain but the structure of the receptor is well defined and plausible binding sites are identified, it may be more useful to chose an absolute free energy protocol to compute the standard free energy of binding of the ligand to a set of binding sites.
-This requires the user to prepare input files describing the bound conformation in different putative binding sites~\cite{10.1039/C6CP05680F} (REF).
+This requires the user to prepare input files describing the bound conformation in different putative binding sites~\cite{evoli2016multiple}.
 The apparent binding free energy of the ligand may be obtained by combining the individual binding site free energies, which also indicate where the ligand is more likely to bind.

### 2020-02-09 paper/manuscript.tex
Some edits to introduction

@@
 \todo[inline, color={green!20}]{@everyone: Any further input in this?}
-Alchemical free energy calculations compute free energy differences associated with transfer processes, such as the binding of a small molecule to a receptor, the transfer of a small molecule from an aqueous to apolar phase~\cite{zwanzig1954hightemperature}, or effects of protein side chain mutations on e.g.
-binding affinities.
-These calculations use non-physical\footnote{Here, the non-physical nature of the transformation is referred to as "alchemical", a term coined by Tembre and McCammon in Ref.~\cite{tembre1984ligandreceptor}.} intermediate states in which the chemical identity of some portion of the system (such as a small molecule ligand or receptor sidechain) is changed by modifying the potential governing the interaction of the atoms being modified, inserted, or deleted with their environment.
-%
+Alchemical free energy calculations compute free energy differences associated with transfer processes, such as the binding of a small molecule to a receptor, the transfer of a small molecule from an aqueous to apolar phase~\cite{zwanzig1954hightemperature}, or the effects of protein side chain mutations on binding affinities or thermostabilities.
+These calculations use non-physical\footnote{Here, the non-physical nature of the transformation is referred to as "alchemical", a term coined by Tembre and McCammon in Ref.~\cite{tembre1984ligandreceptor}.} intermediate states in which the chemical identity of some portion of the system (such as a small molecule ligand or protein sidechain) is changed by modifying the potential governing the interactions with the environment for the atoms being modified, inserted, or deleted.
+
 Fig.~\ref{fig:fig_what_is_alchemy} illustrates common free energy changes that may be difficult to compute with unbiased molecular dynamics methods, but are more readily tractable with alchemical methods.
@@
 \end{itemize}
-
+%
 Furthermore, alchemical calculations are frequently used to estimate changes in free energies upon modifying a ligand or protein: 
@@
 
-Once an alchemical simulation procedure is performed (which usually involves more than one simulation), the data must be analyzed to estimate the free energy for the transformation of interest.
-Early work used simple but statistically suboptimal estimators for this: free energy perturbation (FEP) using the Zwanzig relation~\cite{zwanzig1954hightemperature} or thermodynamic integration (TI), for which the theory dates back decades but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamicsa, merz1989free}.
-%
-More recent developments have seen new, highly efficient statistical estimators that make better use of all the data, often building on the Bennett acceptance ratio (BAR)~\cite{bennett1976efficient}, producing multistate generalizations~\cite{shirts2008statisticallya} or removing the need for global equilibrium~\cite{wu2016multiensemble, mey2014xtram, wu2014statistically}.
-
-Important studies continued during the 2000's often leading to improved implementation of the methods in common simulation software~\cite{shirts2003extremely,shirts2005solvation,vanderspoel2005gromacs, mermelstein2018fast, wang2015accurate, hedges2019biosimspace}.
+After an alchemical calculation is performed (which generally involves multiple simulations at a variety of alchemical states), the data must be analyzed to compute an estimate the free energy for the transformation of interest.
+Early work used simple but statistically suboptimal estimators for this: free energy perturbation (FEP) used a simple (but highly biased) estimator based on the Zwanzig relation~\cite{zwanzig1954hightemperature} or numerical quadrature via thermodynamic integration (TI), for which the theory dates back decades but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamicsa, merz1989free}.
+%
+More recent developments have seen new, highly efficient statistical estimators that make better use of all the data, often building on the more efficient and less biased Bennett acceptance ratio (BAR)~\cite{bennett1976efficient}, producing multistate generalizations~\cite{shirts2008statisticallya} or removing the need for global equilibrium~\cite{wu2016multiensemble, mey2014xtram, wu2014statistically}.
+
+Subsequent work in the 2000s led to improved implementations of alchemical methods in popular biomolecular simulation packages~\cite{shirts2003extremely,shirts2005solvation,vanderspoel2005gromacs, mermelstein2018fast, wang2015accurate, hedges2019biosimspace}.
 This foundational work, combined with the methodological, technological, and hardware improvements of the last 5--10 years, have led to an explosion of interest and direct commercial application of the technologies~\cite{wang2015accurate, fratev2019improved, schindler2020largescale, cournia2017relative, sherborne2016collaborating}.
@@
 
-As the field of molecular simulation moves towards millisecond timescales, accurate alchemical calculations on even more challenging problems will become reasonable to perform.
-In the meantime, today's users may find it hard to get started with these complex calculations whilst also keeping up with the fast pace of change.
-This Best Practice guide provides current recommendations and tips for users of all experience.
+As the field of molecular simulation can now routinely access microsecond timescales with the aid of GPUs~\cite{salomon2013routine} and millisecond timescales appear to soon be within reach, accurate alchemical calculations on even more challenging problems will become reasonable to perform.
+In the meantime, today's users may find it difficult to get started with these complex calculations whilst also keeping up with the fast pace of change. 
+This Best Practices guide provides current recommendations and tips for users of all experience.  
 %
@@
 \label{sec:pre}
-What can be expected from this Best Practice guide is a good starting point for new practitioners and a reference for experienced practitioners.
-For this propose we provide a convenient checklist (Section~\ref{sec:checklist}) to help ensure all calculations meet standard alchemical simulation and analysis practices.
+What can be expected from this Best Practices guide is a good starting point for new practitioners and a reference for experienced practitioners. 
+For this propose we provide a convenient checklist (Section~\ref{sec:checklist}) to help ensure all calculations comply with currently-understood best practices for alchemical simulation and analysis.
+Where the best practices are currently unsettled, we highlight areas where further research is needed to identify an unambiguous recommendation.
 It can also serve as a set of best practices to ensure simulation robustness and reproducibility which reviewers may wish to consider as they evaluate papers.
+
 What we assume from a novice practitioner is a moderate experience with molecular simulation packages and concepts. 
@@
 If you feel unfamiliar with some of these concepts, good starting points can be found in these references~\cite{braun2019best, grossfield2018best, klimovich2015guidelines, shirts2012best}.
-While reading this Best Practice document, it is important to bare in mind \emph{this is not a review} of all free energy calculation methods at the cutting edge of current research.
+
+While reading this Best Practices guide, it is important to bear in mind \emph{this is not a review} of all free energy calculation methods at the cutting edge of current research.
 Instead the guide aims to answer the following questions:
 \begin{itemize}
-    \item Is my problem suitable for an alchemical free energy calculation? 
-    \item How do I choose the right free energy protocol and run it? 
-    \item How do I analyze my simulations accurately? 
-    \item What software tools are available to perform alchemical free energy calculations? 
-\end{itemize}
-%
-
-Some other background information may be needed depending on the nature of the alchemical project.
-For example, often, if binding poses are not known, docking calculations can be used to generate an initial small molecule binding pose to start alchemical calculations.
-This will require some basic familiarity on how to perform docking calculations to generate reasonable simulation starting points~\cite{grinter2014challenges}.
+    \item Is my problem suitable for an alchemical calculation? 
+    \item How do I select an appropriate alchemical protocol? 
+    \item What software tools are available to perform alchemical calculations? 
+    \item How should I analyze my data and report uncertainties? 
+\end{itemize}
+%
+
+Some other background information may be required depending on the nature of the alchemical project. 
+For example, if binding poses are not known, docking calculations can be used to generate an initial small molecule binding pose to start alchemical calculations. 
+This will require some basic familiarity with how to perform docking calculations to generate reasonable simulation starting points~\cite{grinter2014challenges}. 
 %
@@
 While we try to address as many methods and practices as possible, the field of free energy calculations is broad, and there are many advanced topics that are left to future Best Practices documents focusing on specific issues.
-Below, we provide a non-exhaustive list of topics we have not addressed with some references to provide starting points on these more advanced topics:
+Below, we provide a non-exhaustive list of topics we have \emph{not} addressed, along with some references to provide starting points on these more advanced topics:
 \begin{itemize}
@@
 
-For convenience we have also compiled a list of common acronyms, common symbols used, and some essential terminology.
+For convenience, we have also compiled a list of common acronyms, common symbols used, and some essential terminology.
 \begin{tcolorbox}[title=Acronyms, colback=blue!10!white]
-    \textit{CPU} -- Central Processing Unit\\
-    \textit{BAR} -- Bennett Acceptance Ratio\\
-    \textit{FEP} -- Free Energy Perturbation\\
-    \textit{GPCR} -- G-Protein Coupled Receptor\\
-    \textit{GPU} -- Graphics Processing Unit\\
-    \textit{MBAR} -- Multistate Bennett Acceptance Ratio\\
-    \textit{MCSS} -- Maximum Common Substructure\\
-    \textit{MD} -- Molecular Dynamics\\
-    \textit{RMSE} -- Root Mean Square Error\\
-    \textit{MUE} -- Mean unsigned error\\
-    \textit{SAR} -- Structure-Activity Relationships\\
-    \textit{TI} -- Thermodynamic Integration
+    {\bf CPU} -- Central Processing Unit\\
+    {\bf BAR} -- Bennett Acceptance Ratio\\
+    {\bf FEP} -- Free Energy Perturbation\\
+    {\bf GPCR} -- G-Protein Coupled Receptor\\
+    {\bf GPU} -- Graphics Processing Unit\\
+    {\bf MBAR} -- Multistate Bennett Acceptance Ratio\\
+    {\bf MCSS} -- Maximum Common Substructure\\
+    {\bf MD} -- Molecular Dynamics\\
+    {\bf RMSE} -- Root Mean Square Error\\
+    {\bf MUE} -- Mean unsigned error\\
+    {\bf SAR} -- Structure-Activity Relationships\\
+    {\bf TI} -- Thermodynamic Integration
 \end{tcolorbox}
@@
 \begin{tcolorbox}[title=Common terminology, colback=yellow!10!white]
-    \emph{ligand} - small molecule binding to a larger host molecule such as a protein (receptor).\\ 
-    \emph{lead optimization} - process of designing a drug candidate from an initial lead compound, usually involving optimizing its affinity, possibly selectivity, and various pharmacokinetic parameters.\\
-    \emph{docking} -- Method of generate simulation starting structures for ligand receptor systems, when no crystal structure is available\\
-    \emph{thermodynamic cycle} - sequence of thermodynamic processes (like changes in temperature, volume or pressure) in which the initial and final state are equal.\\
-    \emph{alchemical} -- states or simulations that contain non-physical intermediates \\
+    {\bf ligand} -- small molecule binding to a larger host molecule such as a protein (receptor).\\ 
+    {\bf lead optimization} -- process of designing a drug candidate from an initial lead compound, usually involving optimizing its affinity, possibly selectivity, and various pharmacokinetic parameters.\\
+    {\bf docking} -- Method of generate simulation starting structures for ligand receptor systems, when no crystal structure is available\\
+    {\bf thermodynamic cycle} - sequence of thermodynamic processes (like changes in temperature, volume or pressure) in which the initial and final state are equal.\\
+    {\bf alchemical} -- states or simulations that contain non-physical intermediates \\
 \end{tcolorbox}

### 2020-05-29 paper/manuscript.tex

--- 2020-02-05 paper/manuscript.tex
@@
 \item the entire ligand can be alchemically transferred from protein to solvent in an absolute binding free energy calculation (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{E})~\cite{mobley2007predicting,aldeghi2015accurate,aldeghi2017predictions};
-\item and small alchemical modifications can be made between chemically related ligands to estimate relative differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{F})~\cite{wang2015accurate,mey2016blinded,gapsys2020large}.
+\item and small alchemical modifications can be made between chemically related ligands to estimate relative differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{F})~\cite{wang2015accurate,mey2016blinded,song2019using,gapsys2020large,kuhn2020automated}.
 \end{itemize}
@@
 The software builds the network of perturbations, connecting similar compounds based on LOMAP methodology~\cite{liu2013lead}, all inputs for the Desmond simulations are built automatically and the job can be saved or run immediately on available GPUs.
-Commercial tools can be expensive in some cases, but non-comercial tools are becoming more straight forward to use to run alchemical free energy calculations~\cite{gapsys2015pmx, loeffler2015fesetup, song2019using, gapsys2020large,  jespers2019qligfep, hedges2019biosimspace}.
+Commercial tools can be expensive in some cases, but non-comercial tools are becoming more straight forward to use to run alchemical free energy calculations~\cite{gapsys2015pmx, loeffler2015fesetup, song2019using, gapsys2020large,  jespers2019qligfep, hedges2019biosimspace, kuhn2020automated}.
 
@@
 This situation may be addressed by explicitly modelling different binding modes of the same ligand and combining later computed free energy differences for different binding modes into a relative free energies of binding~\cite{kaus2015how}.
-%
+
 \paragraph{Have you considered sterioisomers and enationmers?}
@@
 Additionally, unexpected changes in what stereoisomer binds experimentally, if they occur, could pose significant challenges for modelling efforts.
-%
+
 \paragraph{Conserved binding site waters can play an important role in binding free energies}
@@
 
+\paragraph{Protonation states depend on the pH of the experimental assay}
+I think we should have a tiny bit on protonation states. This is something I always easily forget as a non-chemist.
+
 \paragraph{How to align congeneric series}
@@
 In this case a docking program can generate initial structures.
-Different commercial and none commercial tools are available, such as rDock, Autodock Vina, Glide, or Flare, to name a few~\cite{ruiz-carmona2014rdock, trott2010autodock, friesner2004glide, cheeseright2006molecular}.
+Different commercial and none commercial tools are available, such as rDock, Autodock Vina, Glide, or Flare, to name a few~\cite{ruiz-carmona2014rdock, trott2010autodock, friesner2004glide, kuhn2020automated}.
 
 If the putative binding sites are not apparent, for instance due to significant induced-fit effects, it may be challenging to obtain meaningful free energies of binding.
-One may have to account for the free energy change for forming a binding site in the target receptor which may not be feasible on FE simulation timescales.
-Another key application for absolute binding free energies is to study chemically different ligands, that do not share a common scaffold, but bind at the same site or the same site of similar proteins~\cite{aldeghi2015accurate}.
+One may have to account for the free energy change for forming a binding site in the target receptor which may not be feasible on alchemcial simulation timescales.
 
@@
 See for example~\cite{bosisio2016blinded,rustenburg2016measuring} for a good discussion of these choices.
-%
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 Alchemical free energy calculations can be grouped into two main categories, ``absolute'' (see Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}) and ``relative'' \footnote{The distinction is a bit of a misnomer, since both compute ratios of partition functions relative to another state, and neither computes an absolute free energy.}(see Fig.~\ref{fig:fig_binding_thermodynamic_cycle}), which differ in whether they compute properties for a single molecule (absolute) or compare properties of different, usually closely related, molecules (relative).
-To use binding as a concrete example, in absolute binding free energy calculations, one computes the binding free energy of a ligand to an individual receptor relative to a standard reference concentration.
-%
-In contrast, in relative binding free energy calculations, one compares the binding free energy of two related inhibitors to determine the potency difference.
+To use binding as a concrete example, in absolute binding free energy calculations, we compute the binding free energy of a ligand to an individual receptor relative to a standard reference concentration.
+
+In contrast, in relative binding free energy calculations, we compare the binding free energy of two related inhibitors to determine the potency difference.
 %
@@
 \label{sec:relative-fe-protocol}
-%
+
 \paragraph{Topologies} 
@@
     \includegraphics[width=0.95\linewidth]{figures/fig5_thermo_cyc_abs/Figure.pdf}
-    \caption{\textbf{Thermodynamic cycle required for an aboslute free energy calculation -- absolute free energy of binding example} The fully interacting ligand in water (\textbf{A}), has its charges turned off to pass to (\textbf{B}) followed by turning of van der Waals terms, resulting in a non-interacting ligand in water in (\textbf{C}).
+    \caption{\textbf{Thermodynamic cycle required for an absolute free energy calculation -- absolute free energy of binding example} The fully interacting ligand in water (\textbf{A}), has its charges turned off to pass to (\textbf{B}) followed by turning of van der Waals terms, resulting in a non-interacting ligand in water in (\textbf{C}).
 Restraints are used on the fully interacting ligand in the binding site of a protein or host molecule (\textbf{D}).
@@
 \item Significant discrepancies between free energies calculated with different free energy estimators (TI, BAR, MBAR)
-\item Poor replica mixing (for replica-exchange)
+\item Poor replica mixing (for replica-exchange) or lambda-space sampling for single-replica methods
 \item Correlation time as a function of lambda that does not vary relatively smoothly
@@
 \item Torsions that have multiple low-energy minima that are visited rarely or not at all. Also, comparing torsional behavior of ligands in solvent and complex simulations 
-\item The convergence of the free energy change as a function of lambda
+\item The convergence of the free energy change as a function of lambda, or of simulation time (Figure 11).
+\item Diagnostics for non-equilibrium methods, robustness for switching time etc.
 \item More “usual suspects”
@@
 We recommend performing triplicates of your predictions at minimum to ensure some measure of reliability in your data.
-It may seem exessive, but  uncertainty estimates often underestimate the true statistical uncertainty.
+It may seem excessive, but  uncertainty estimates often underestimate the true statistical uncertainty.
 Where performing multiple replicas of the simulation is not possible, lower error bounds such as given by the MBAR estimator should be included.
@@
 rankings) should be computed.
-Conventionally, this means including an R\textsuperscript{2} (or Pearson R -- 1 high correlation 0 no correlation -1 high anti-correlation) and a Kendall \texttau{} (with perfect ranking agreement when \texttau=1 and perfect disagreement when \texttau=-1) metric in your results.
+Conventionally, this means including an R\textsuperscript{2} (or Pearson's R: +1 high correlation 0 no correlation -1 high anti-correlation) and a Kendall \texttau{} (with perfect ranking agreement when \texttau=1 and perfect disagreement when \texttau=-1) metric in your results.
 Additionally, practitioners may choose to include a Spearman \textrho{} as well.
@@
 Furthermore, MUE allows for unbiased comparisons between predictive models as it is less sensitive to dataset size.
+Other metrics such as Gaussian Random Affinity Model (GRAM)~\cite{cui2019gram}, Predictive Interval (PI) Relative Absolute Error (RAE), that attempt to correct for the inherent potency range of a dataset, which can aid in comparing success between different targets.
 Even though this section may serve as an introduction to the main statistical practices, the authors recommend further reading on evaluation of computational models \cite{jain2008recommendations, walters2013what, brown2009healthy, walterthoughts}.
+
+Reporting the results of relative free energy calculations require an additional level of care.
+As shown in Figure 5, relative free energies can be performed in two arbitrary directions, thus relative free energies may be reported as either positively or negatively valued, depending on the choice of what is forward.
+The consequence of the two possible signs for relative free energies is that correlation statistics (such as Pearson's R and Kendall \texttau{}) can be skewed depending on which sign is analysed.
+The issue of this inconsistency can be circumvented by either plotting all datapoints within a consistent quadrant\cite{perez-benito2019predicting}, or by avoiding the use of correlation statistics for assessment of relative free energy calculations and instead measuring accuracy using RMSE and MUE which are unaffected by choice of sign.
 %
@@
 
-Because we are interested mainly in the linear relationship between the alchemical FE predictions and the experimentally-determined affinity values, plots should be depicted with the same range on both axes (i.e.
-$x=y$).
+Because we are interested mainly in the linear relationship between the alchemical FE predictions and the experimentally-determined affinity values,  and if units are consistent for both axis, plots should be depicted with the same range on both axes (i.e.
+$x=y$) with a 1:1 aspect ratio.
 Furthermore, bounds should be depicted for the 1- and 2-kcal/mol confidence regions.
@@
 Alchemical free energy calculations have seen a vast increase in popularity also in the pharmaceutical industry for structure based drug discovery~\cite{schindler2020largescale, sherborne2016collaborating, wagner2017computational}.
-Commercial products such as FEP+ and Flare, which provide a convenient user interface make the setup and use of these methods a lot easier~\cite{wang2015accurate}, but this convenience comes with less flexibility on the simulation protocols.
-It is also important to understand the current limitations of the methodology to recognise when automated worflow tools can be used effectively for a given protein target.
+Commercial products such as FEP+ and Flare, which provide a convenient user interface make the setup and use of these methods a lot easier~\cite{wang2015accurate, kuhn2020automated}, but this convenience comes with less flexibility on the simulation protocols.
+It is also important to understand the current limitations of the methodology to recognise when automated workflow tools can be used effectively for a given protein target.
 Prospective prediction challenges such as the drug design data resource grand challenges provide a community driven platform to evaluate different free energy protocols against each other on unknown targets~\cite{gaieb2018d3r, gaieb2019d3r}.Such efforts have highlighted that selection of seemingly identical or similar potential energy function or simulation package does not guarantee production of similar free energies owing to differences in simulation protocols.
@@
    \begin{itemize}
-    \item FEP+ is a tool offered by Schr\"{o}dinger Inc.
+    \item \href{https://www.schrodinger.com/fep}{FEP+} is a tool offered by Schr\"{o}dinger Inc.
 under a commercial license.
 It has an intuitive GUI which makes it easier for non-experts to run alchemical free energy calculations and analyze the results.
-It runs DESMOND MD package under the hood and hence parallelizes very well on the GPUs.
-    \item Flare is a commercial structure-based drug design software offered by Cresset.
+It runs the DESMOND MD package under the hood and hence parallelizes well on GPUs.
+    \item \href{https://www.cresset-group.com/software/flare/}{Flare} is a commercial structure-based drug design software offered by Cresset.
 Similar to FEP+ it has an easily accessible user interface and strives to facilitate free energy calculations for non-experts while offering advanced users full control via a Python API.
 It only runs on GPUs, using CUDA or OpenCL.
-    \item The molecular operating environment (MOE) offered by the Chemical Computing Group (CCG) has a tool for performing free energy calculations. It is built on AMBER-TI (cf. below).
+    \item The molecular operating environment (\href{https://www.chemcomp.com/Products.htm}{MOE}) offered by the Chemical Computing Group (CCG) has a tool for performing free energy calculations.
+It is built on AMBER-TI (cf.
+below).
     \end{itemize}
@@
 	\begin{itemize}
-	\item CHARMM has a variety of tools developed over the years.
-PERT module can be used to define initial and final states and define the intermediate lambda points.
+	\item \href{https://www.charmm.org/}{CHARMM} has a variety of tools developed over the years.
+The PERT module can be used to define initial and final states and define the intermediate lambda points.
 FREN and BAR modules can be used to analyze the data after the MD run.
-Lambda-dynamics based free energy calulation can be carried out using the BLOCK module.
-	\item TIES and AMBER FEW? (Peter Coveney)
-	\item AMBER, including its new pmemd.cuda version support free energy calculation. 
+Lambda-dynamics-based free energy calculation can be carried out using the BLOCK module.
+	\item \href{https://ambermd.org/}{AMBER}, including its new pmemd.cuda version supports free energy calculations. 
 	\end{itemize}
@@
 	\begin{itemize}
-	\item PLUMED is an open source tool which enables the usage of a variety of MD engines.
+	\item \href{https://www.plumed.org/}{PLUMED} is an open source tool which enables the usage of a variety of MD engines.
 It is designed as a plugin for MD packages such that it analyzes the trajectory on the fly.
 It also offers a VMD based plugin for the computation of collective variables.
-	\item SIRE  is a free, open source, multiscale molecular simulation framework, written to allow computational modellers to quickly prototype and develop new algorithms for molecular simulation and molecular design.
-	\item YANK is a tool developed by John Chodera and group on the top of OpenMM MD package. It allows the users to write their inputs in easy-to-use YAML format.
-	\item GROMACS is a molecular simulation package with a significant number of free energy methods implementations.
+	\item \href{https://siremol.org/}{SIRE}  is a free, open source, multiscale molecular simulation framework, written to allow computational modellers to quickly prototype and develop new algorithms for molecular simulation and molecular design.
+	\item \href{http://getyank.org/latest/index.html}{YANK} is a tool developed by John Chodera and group on the top of OpenMM MD package.
+It allows the users to write their inputs in easy-to-use YAML format.
+	\item \href{http://www.gromacs.org/}{GROMACS} is a molecular simulation package with a significant number of free energy methods implementations.
 The LiveCOMS GROMACS tutorial has an example free energy calculation~\cite{lemkul2018From}.
-	\item pmx for mutations
+	\item \href{http://pmx.mpibpc.mpg.de/instructions.html}{PMX}, an add-on to GROMACS, offers a mutation free energy calculation module.
+	\item \href{https://github.com/qusers/Q6}{Q} is MD code for performing FEP calculations using a variety of force fields. 
 	\end{itemize}
@@
 	\begin{itemize}
-	\item FESetup: AMBER, GROMACS, Sire
-	\item PMX: GROMACS \url{https://github.com/deGrootLab/pmx}
-	\item Lomap/Lomap2 : Relative alchemical transformation graph planning
-	\item CHARMM-GUI is a web based tool for setting up a variety of MD simulations. It can be used to generate CHARMM scripts for solvation and ligand-binding free energy calculations.
+	\item \href{https://fesetup.readthedocs.io/en/latest/introduction.html}{FESetup} has been developed primarily to setup calculations in AMBER, GROMACS and SIRE.
+	\item href{http://pmx.mpibpc.mpg.de/instructions.html}{PMX}: GROMACS \url{https://github.com/deGrootLab/pmx}.
+	\item \href{https://github.com/MobleyLab/Lomap}{Lomap/Lomap2} : Relative alchemical transformation graph planning for setting up perturbation networks.
+	\item \href{http://www.charmm-gui.org/}{CHARMM-GUI} is a web-based tool for setting up a variety of MD simulations.
+It can be used to generate CHARMM scripts for solvation and ligand-binding free energy calculations.
+	\item \href{https://github.com/qusers/qligfep}{QligFEP} offers robust and fast setup of FEP calculations for the software package Q
 	\end{itemize}
@@
 	\item Free Energy Workflows: Sire-specific free energy map analysis using weighted path averages
-	\url{https://github.com/michellab/freenrgworkflows}
+	\url{https://github.com/michellab/freenrgworkflows}.
 	\item Alchemlyb: Multipackage free energy analysis
-	\url{https://github.com/alchemistry/alchemlyb}
+	\url{https://github.com/alchemistry/alchemlyb}.
 	\item pymbar: MBAR implementation, but have to roll your own analysis wrapper
-	\url{https://github.com/choderalab/pymbar}
-	\end{itemize}
+	\url{https://github.com/choderalab/pymbar}.
+	\end{itemize}
+\item[] Generally, commercial software will offer more complete pipelines in which standalone analysis applications are not necessarily needed; free and open source packages often require manual analysis.
 \end{itemize}
@@
 FEP+ Dataset~\cite{roos2019opls3e} & 8 & 199 & OPLS3e \\
-FEP+ Dataset~\cite{song2019using} & 8 & 199 & AMBER18 \\
+FEP+ Dataset~\cite{song2019using} & 8 & 199 & GAFF 1.8 \\
+FEP+ Dataset~\cite{gapsys2020large} & 8 & 199 & various \\
+FEP+ Dataset~\cite{kuhn2020automated} & 8 & 199 & GAFF2.1 \\
 Fragments ~\cite{steinbrecher2015accurate} & 8 & 96 & OPLS2.1 \\
 Scaffold Hopping~\cite{wang2017accurate} & 6 & 21 & OPLS3 \\
+Scaffold Hopping~\cite{kuhn2020automated} & 6 & 21 & GAFF2.1 \\
 Macrocycles~\cite{yu2017accurate} & 7 & 33 & OPLS3 \\
@@
 Declare any potentially conflicting interests here, whether or not they pose an actual conflict in your view.
-MK is employed by Cresset who recently released a commercial tool for performing alchemical free energy calculations.
+MK is employed by Cresset who commercially distribute a software for performing alchemical free energy calculations.
 

### 2020-07-13 paper/manuscript.tex
Updates from Overleaf

@@
 
-\subsection{Absolute and relative calculations must deal with some of the same issues}
+\subsection{Absolute and relative calculations deal with some of the same issues}
 \subsubsection{Handling weak binders and high dissociation rates}
@@
 \todo[inline]{ASJSM:Maybe here we have a quick summary on what things should be checked for to make sure simulations are ok?}
-\begin{itemize}
-\item Examining output data for common problems with discussions of what exactly to plot or look at; examples of typical curves for dV/dlambda and free energy versus lambda, for example
-\begin{itemize}
-\item Make sure ligand sampling in the binding site is reasonable for it's expected potency and fit, and that it does not depart out of binding site. 
-\item Significant discrepancies between free energies calculated with different free energy estimators (TI, BAR, MBAR)
-\item Poor replica mixing (for replica-exchange) or lambda-space sampling for single-replica methods
-\item Correlation time as a function of lambda that does not vary relatively smoothly
-\item Dependence of the free energy on initial conformation
-\item Torsions that have multiple low-energy minima that are visited rarely or not at all. Also, comparing torsional behavior of ligands in solvent and complex simulations 
-\item The convergence of the free energy change as a function of lambda, or of simulation time (Figure 11).
-\item Diagnostics for non-equilibrium methods, robustness for switching time etc.
-\item More “usual suspects”
-\end{itemize}
-\item Other considerations for many transformations
-\end{itemize}
-%
+
+\item It is important to carefully examine output data for common problems.  Some of the most important things to check for rae:
+\begin{itemize}
+\item Make sure the ligand samples the binding site reasonably tightly for it's expected potency and fit, and that it does not depart out of binding site in the coupled end state if it is a moderate to strong binder.
+\item Significant discrepancies (further outside the error estimates than would be plausible) between free energies calculated with different free energy estimators such as TI, BAR, and MBAR.
+All of these estimators converge to the same results with sufficient sampling.
+Differences between them indicate poor overlap or errors in processing.
+\item Poor replica mixing (for replica-exchange) or lambda-space sampling for single-replica methods.
+If the system is not mixing between states, then  the states are insufficiently close for mixing, or else there are bottlenecks in the configurational sampling that limit the accuracy.
+\item Correlation time  that does not vary relatively smoothly as a function of $\lambda$.
+Discontinuities indicate that the system is sampling significantly different configurations when the Hamiltonian changes only a bit, which usually indicates sampling problems.
+\item Dependence of the free energy on initial conformation of the system. Clearly, ensemble average properties should not depend on the starting point.
+\item Torsions with multiple low-energy minima where some of these minima are that are visited rarely or not at all.
+Which torsions have low energy minima can best be found by comparing to the simulation in the solvent.
+There should be clear physical reasons that  simulation in the complex has different torsional distributions that the ligand in the solvent.
+\item The free energy should vary relatively smoothly with $\lambda$.
+If it varies drastically, then either there need to be finer sampling in $\lambda$ in this region, or there are sampling problems there.
+\item The free energy should clearly converge as a function or of simulation time (Figure $\ref{fig:convergence_forward_reverse}$).
+\item If using nonequilbirum methods, is the result independent of the speed at which the nonequilibrium change is performed?
+Nonequilibrium methods are in theory independent of the switching time in the limit of good sampling unless the switching time is simply too short.
+%\item Other Diagnostics for non-equilibrium methods, robustness for switching time etc.?
+\item In general, look at the output data such as energies and see if it matches your expectations.and also visualize the simulation trajectories themselves to see if it matched expectations.
+There are many errors that are caught by only one of the two approaches
+%\item More “usual suspects”
+\end{itemize}
+%\item Other considerations for many transformations
+
 \subsection{Best practices for reporting data }
 \label{sec:plot_data}
-Data generation and adequate analysis is not sufficient to finish a study.
-As a practitioner of alchemical free energy simulations you should use best practice for reporting and plotting your results.
+Carrying out good data generation and proper analysis is not sufficient to finish a study.
+As a practitioner of alchemical free energy simulations you also  should use best practices for reporting and plotting your results.
 We  encourage the following standard set of analyses and ways to represent data.
@@
 First, error estimates should always be included on your predictions in whatever form you present your data (scatterplots, barplots, etc; see next paragraph).
-We recommend performing triplicates of your predictions at minimum to ensure some measure of reliability in your data.
-It may seem excessive, but  uncertainty estimates often underestimate the true statistical uncertainty.
+We recommend performing triplicates of your predictions at minimum, with starting points that are expected to be uncorrelated, to ensure some measure of reliability in your data.
+This replication may seem excessive, but  uncertainty estimates often underestimate the true statistical uncertainty.
 Where performing multiple replicas of the simulation is not possible, lower error bounds such as given by the MBAR estimator should be included.
 
-Because of the strong drug discovery context and the relevance of alchemical free energy predictions to structure activity relationship (SAR) projects, both the overall correlation of your predictions relative to experiment and the ability of your models to detect the best-scoring compounds (i.e.
+Because of the recent importance of alchemical free energy predictions in the  drug discovery context, and the relevance of these quantities to structure activity relationship (SAR), both the overall correlation of your predictions relative to experiment and the ability of your models to detect the best-scoring compounds (i.e.
 rankings) should be computed.
@@
 Mean unsigned error (MUE, also called mean absolute error/MAE) is another key statistic to include in your results.
-Even though some models' near-perfect correlation and ranking statistics might suggest excellent accuracy, MUE values can still show multiple kcal/mol of error, providing important additional insight into performance.
+Even though some models' near-perfect correlation and ranking statistics might suggest excellent accuracy, MUE values can still have errors of show multiple kcal/mol of error, providing important additional insight into performance.
 Furthermore, MUE allows for unbiased comparisons between predictive models as it is less sensitive to dataset size.
 Other metrics such as Gaussian Random Affinity Model (GRAM)~\cite{cui2019gram}, Predictive Interval (PI) Relative Absolute Error (RAE), that attempt to correct for the inherent potency range of a dataset, which can aid in comparing success between different targets.
-Even though this section may serve as an introduction to the main statistical practices, the authors recommend further reading on evaluation of computational models \cite{jain2008recommendations, walters2013what, brown2009healthy, walterthoughts}.
-
-Reporting the results of relative free energy calculations require an additional level of care.
-As shown in Figure 5, relative free energies can be performed in two arbitrary directions, thus relative free energies may be reported as either positively or negatively valued, depending on the choice of what is forward.
+We recommend further reading on evaluation of computational models \cite{jain2008recommendations, walters2013what, brown2009healthy, walterthoughts}.
+
+Reporting the results of relative free energy calculations requires care.
+As shown in Figure~\ref{fig:5}, relative free energies can be performed arbitrarily as a forward or a reverse process, and thus relative free energies may be reported as either positively or negatively valued.
 The consequence of the two possible signs for relative free energies is that correlation statistics (such as Pearson's R and Kendall \texttau{}) can be skewed depending on which sign is analysed.
@@
 Any study with \textgreater50 ligands should be trellised to not overestimate statistics.
-In that respect, it is bad practice to place multiple datasets on the same plot because this can suggest high model accuracy even though the individual models perform less well \cite{walterthoughts}.
-%
-
-Because we are interested mainly in the linear relationship between the alchemical FE predictions and the experimentally-determined affinity values,  and if units are consistent for both axis, plots should be depicted with the same range on both axes (i.e.
-$x=y$) with a 1:1 aspect ratio.
+Because of this, it is bad practice to place multiple datasets on the same plot, because this can suggest high model accuracy even though the individual models perform less well \cite{walterthoughts}.
+%
+
+Because we are interested mainly in the linear relationship between the alchemical FE predictions and the experimentally-determined affinity values, plots should be depicted with the same range on both axes (i.e.
+$x=y$) with a 1:1 aspect ratio, with units for both experiment and simulation converted to be the same.
 Furthermore, bounds should be depicted for the 1- and 2-kcal/mol confidence regions.

### 2020-07-13 paper/manuscript.tex
Fixing section 9, 10, 11, 12

@@
 As essentially all alchemical free energy prediction schemes are regression problems, the preferred type of plot is a scatter plot (see Figure \ref{fig:scatterplot_analysis}).
-Most alchemical FE projects will concern 10-50 ligands; any study with \textless10 ligands is more suitable for bar plots (with inclusion of error bars).
+Most alchemical free energy projects will look at 10-50 ligands; any study with \textless10 ligands is more suitable for bar plots (with inclusion of error bars), and will unlikely provide meaningful statistics.
 Any study with \textgreater50 ligands should be trellised to not overestimate statistics.
-Because of this, it is bad practice to place multiple datasets on the same plot, because this can suggest high model accuracy even though the individual models perform less well \cite{walterthoughts}.
-%
-
-Because we are interested mainly in the linear relationship between the alchemical FE predictions and the experimentally-determined affinity values, plots should be depicted with the same range on both axes (i.e.
+Because of this, it is bad practice to place multiple datasets on the same plot, because this can suggest high model accuracy even though the individual models perform less well~\cite{walterthoughts}.
+%
+
+Because we are interested mainly in the linear relationship between the alchemical free energy predictions and the experimentally-determined affinity values, plots should be depicted with the same range on both axes (i.e.
 $x=y$) with a 1:1 aspect ratio, with units for both experiment and simulation converted to be the same.
+If this is skews the plot to a point where it is difficult to read of information, using the same dimensions, such that e.g.
+1 cm is 1 kcal/mol is acceptable.
 Furthermore, bounds should be depicted for the 1- and 2-kcal/mol confidence regions.
@@
 In a drug discovery context, this type of data depiction may suggest the reliability of alchemical FE predictions in the project, and can give an idea of how trustworthy predictions can be for synthesis ideas.
+
 \begin{figure}
     \includegraphics[width=0.95\linewidth]{figures/fig13_analysis_practices/Figure.pdf}
-    \caption{Example of a well-depicted alchemical free energy prediction showing the relation between predicted and experimentally-determined Gibbs free energy in kcal/mol with standard errors as error bars.
-The two outliers (underestimations) are shown in red; the dark and light-orange regions depict the 1- and 2-kcal per mol confidence bounds.}
+    \caption{\textbf{An example of recommended practices for graphing alchemical free energy predictions.} This figure shows the relation between predicted and experimentally-determined Gibbs free energy in kcal/mol with standard errors as error bars.
+The dark and light-orange regions depict the 1- and 2-kcal per mol confidence bounds.
+Extra care should be taken when investigating potential outliers further.}
     \label{fig:scatterplot_analysis}
 \end{figure}
+
 An example of a best practice scatter comparison between computed and experimental values is shown in Fig.~\ref{fig:scatterplot_analysis}, highlighting outliers, error bars and confidence intervals.
 The data for this plot is artificially generated for illustration purposes.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \label{sec:conclusion}
-Alchemical free energy calculations have seen a vast increase in popularity also in the pharmaceutical industry for structure based drug discovery~\cite{schindler2020largescale, sherborne2016collaborating, wagner2017computational}.
-Commercial products such as FEP+ and Flare, which provide a convenient user interface make the setup and use of these methods a lot easier~\cite{wang2015accurate, kuhn2020automated}, but this convenience comes with less flexibility on the simulation protocols.
-It is also important to understand the current limitations of the methodology to recognise when automated workflow tools can be used effectively for a given protein target.
-Prospective prediction challenges such as the drug design data resource grand challenges provide a community driven platform to evaluate different free energy protocols against each other on unknown targets~\cite{gaieb2018d3r, gaieb2019d3r}.Such efforts have highlighted that selection of seemingly identical or similar potential energy function or simulation package does not guarantee production of similar free energies owing to differences in simulation protocols.
-(SUGGEST REMOVING THIS SENTENCE What has come out of these in the past is that Schrödinger's FEP+, performs similar when run by different participant groups~\cite{gaieb2018d3r}, but may not give the desired accuracy with RMSE errors larger than 1 kcal/mol.
-Other alchemical protocols performed similarly well.(CITE GAPSYS PMX paper for large FEP+ comparison) We hope that the best practice guide provides a set of tools that allow a better understanding of how to setup, run and interpret reliably alchemical free energy calculations.
-%
+Alchemical free energy calculations have seen a vast increase in popularity both in academic research as well as  pharmaceutical industry applications in structure based drug discovery~\cite{schindler2020largescale, sherborne2016collaborating, wagner2017computational}.
+Commercial products such as FEP+ and Flare, which provide a convenient user interface make the setup and use of these methods a lot easier~\cite{wang2015accurate, kuhn2020automateda}, but this convenience comes with less flexibility in terms of choice of simulation protocols.
+It is also important to understand the current limitations of the methodology to recognise when automated workflow tools can be used effectively for a given protein target and when they are likely to fail still.
+Prospective prediction challenges such as the Drug Design Data resource grand challenges provide a community driven platform to evaluate different free energy protocols against each other on blinded targets~\cite{gaieb2018d3r, gaieb2019d3r}.
+Such efforts have highlighted that selection of seemingly identical or similar potential energy function or simulation package does not guarantee production of similar free energies owing to differences in simulation protocols.
+We hope that the best practice guide provides a set of tools that allow a better understanding of how to setup, run, and reliably interpret alchemical free energy calculations.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Available software -- a summary}
-\todo[inline]{ASJSM: Expand and add more references. }
+\section{Non-exhaustive overview of available software packages}
+
 \label{sec:software}
-There are several softwares availble for carrying out alchemical free energy calculations. In this section we discuss some of the popular commercial and free tools along with their features. 
-\begin{itemize}
-\item Commercial:
+There are many different software solutions available for the setup, running, and analysis of alchemcial free energy calculations.
+These will vary in customizability and ways in which they are ran, e.g.
+graphical user interface versus command line tool or python script.
+The following provides a non-exhaustive list of commercial and noncommercial tools available for conducting alchemical free energy calculations.
+\begin{itemize}
+
+\item [] \textbf{Simulation software: Commercial}
    \begin{itemize}
@@
 It has an intuitive GUI which makes it easier for non-experts to run alchemical free energy calculations and analyze the results.
-It runs the DESMOND MD package under the hood and hence parallelizes well on GPUs.
+It runs the DESMOND MD package under the hood and hence parallelizes well on GPUs~\cite{wang2015accurate}.
     \item \href{https://www.cresset-group.com/software/flare/}{Flare} is a commercial structure-based drug design software offered by Cresset.
-Similar to FEP+ it has an easily accessible user interface and strives to facilitate free energy calculations for non-experts while offering advanced users full control via a Python API.
-It only runs on GPUs, using CUDA or OpenCL.
+Similar to FEP+ it has an easily accessible graphical user interface and strives to facilitate free energy calculations for non-experts while offering advanced users full control via a Python API.
+It only runs on GPUs, using CUDA or OpenCL~\cite{kuhn2020assessmenta}.
     \item The molecular operating environment (\href{https://www.chemcomp.com/Products.htm}{MOE}) offered by the Chemical Computing Group (CCG) has a tool for performing free energy calculations.
@@
     \end{itemize}
-\item Free or low-cost for academics / commercial for industry:
+    \item[]All the above tools also provide a convenient setup and analysis suite and are really a one in all product. 
+\item [] \textbf{Simulation software: Free/low cost academic and Commercial}
 	\begin{itemize}
@@
 Lambda-dynamics-based free energy calculation can be carried out using the BLOCK module.
-	\item \href{https://ambermd.org/}{AMBER}, including its new pmemd.cuda version supports free energy calculations. 
-	\end{itemize}
-\item Free (libre) open source:
+	\item \href{https://ambermd.org/}{AMBER}, including its new pmemd.cuda version supports free energy calculations~\cite{salomon-ferrer2013overview}. 
+	\end{itemize}
+\item [] \textbf{Simulation software: Open Source}
 	\begin{itemize}
@@
 It is designed as a plugin for MD packages such that it analyzes the trajectory on the fly.
-It also offers a VMD based plugin for the computation of collective variables.
-	\item \href{https://siremol.org/}{SIRE}  is a free, open source, multiscale molecular simulation framework, written to allow computational modellers to quickly prototype and develop new algorithms for molecular simulation and molecular design.
+It also offers a VMD based plugin for the computation of collective variables~\cite{bonomi2019promoting}.
+	\item \href{https://biosimspace.org/}{BioSimSpace}  is a free, open source, multiscale molecular simulation framework, written to allow computational modellers to quickly prototype and develop new algorithms for molecular simulation and molecular design~\cite{hedges2019biosimspace}.
 	\item \href{http://getyank.org/latest/index.html}{YANK} is a tool developed by John Chodera and group on the top of OpenMM MD package.
-It allows the users to write their inputs in easy-to-use YAML format.
+It allows the users to write their inputs in easy-to-use YAML format\cite{}.
 	\item \href{http://www.gromacs.org/}{GROMACS} is a molecular simulation package with a significant number of free energy methods implementations.
 The LiveCOMS GROMACS tutorial has an example free energy calculation~\cite{lemkul2018From}.
-	\item \href{http://pmx.mpibpc.mpg.de/instructions.html}{PMX}, an add-on to GROMACS, offers a mutation free energy calculation module.
-	\item \href{https://github.com/qusers/Q6}{Q} is MD code for performing FEP calculations using a variety of force fields. 
-	\end{itemize}
-\item Setup tools
-	\begin{itemize}
-	\item \href{https://fesetup.readthedocs.io/en/latest/introduction.html}{FESetup} has been developed primarily to setup calculations in AMBER, GROMACS and SIRE.
+	\item \href{http://pmx.mpibpc.mpg.de/instructions.html}{PMX}, an add-on to GROMACS, offers a mutation free energy calculation module\cite{abraham2015gromacs}.
+	\item \href{https://github.com/qusers/Q6}{Q} is MD code for performing FEP calculations using a variety of force fields~\cite{aqvistjohan2017q6}. 
+	\end{itemize}
+\item[] \textbf{Setup tools:}
+	\begin{itemize}
 	\item href{http://pmx.mpibpc.mpg.de/instructions.html}{PMX}: GROMACS \url{https://github.com/deGrootLab/pmx}.
-	\item \href{https://github.com/MobleyLab/Lomap}{Lomap/Lomap2} : Relative alchemical transformation graph planning for setting up perturbation networks.
+	\item \href{https://github.com/MobleyLab/Lomap}{Lomap/Lomap2} : Relative alchemical transformation graph planning for setting up perturbation networks~\cite{liu2013lead}.
 	\item \href{http://www.charmm-gui.org/}{CHARMM-GUI} is a web-based tool for setting up a variety of MD simulations.
-It can be used to generate CHARMM scripts for solvation and ligand-binding free energy calculations.
-	\item \href{https://github.com/qusers/qligfep}{QligFEP} offers robust and fast setup of FEP calculations for the software package Q
-	\end{itemize}
-\item Analysis tools:
-	\begin{itemize}
-	\item Free Energy Workflows: Sire-specific free energy map analysis using weighted path averages
-	\url{https://github.com/michellab/freenrgworkflows}.
+It can be used to generate CHARMM scripts for solvation and ligand-binding free energy calculations~\cite{jo2008charmmgui}.
+	\item \href{https://github.com/qusers/qligfep}{QligFEP} offers robust and fast setup of FEP calculations for the software package Q~\cite{jespers2019qligfep}.
+	\item ProtoCaller, a setup tool for the automation of Gromacs free energy calculations. \href{https://github.com/protocaller/ProtoCaller}{ProtoCaller}~\cite{suruzhon2020protocaller}
+	\item \href{https://fesetup.readthedocs.io/en/latest/introduction.html}{FESetup} has been developed primarily to setup calculations in AMBER, GROMACS and SIRE~\cite{loeffler2015fesetup}
+	\end{itemize}
+\item []\textbf{Analysis tools:}
+	\begin{itemize}
+
 	\item Alchemlyb: Multipackage free energy analysis
-	\url{https://github.com/alchemistry/alchemlyb}.
+	\url{https://github.com/alchemistry/alchemlyb}~\cite{daviddotson2020alchemistry}.
 	\item pymbar: MBAR implementation, but have to roll your own analysis wrapper
-	\url{https://github.com/choderalab/pymbar}.
+	\url{https://github.com/choderalab/pymbar}~\cite{shirts2008statistically}.
+	\item Beryllium: Standardising alchemical free energy analysis \url{https://github.com/openforcefield/beryllium}
+	\item Free Energy Workflows: Sire-specific free energy map analysis using weighted path averages \url{https://github.com/michellab/freenrgworkflows}.
 	\end{itemize}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%
-%  Online resources    %
-%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Online resources}
-\todo[inline]{Do we really need this section?
-Everything has been referenced.
-Seems like these could be listed at the beginning.
-If this stays here, it would be more useful if they had some description of what the different references contain.}
-\begin{itemize}
-\item \url{http://www.ks.uiuc.edu/Training/Workshop/Urbana_2010A/lectures/TCBG-2010.pdf}
-\item Basic Ingredients of Free Energy Calculations: A Review (\url{DOI: 10.1002/jcc.21450})
-\item Good Practices in Free-Energy Calculations (\url{DOI: 10.1021/jp102971x})
-\item Alchemical Free Energy Methods for Drug Discovery: Progress and Challenges (\url{doi: 10.1016/j.sbi.2011.01.011})
-\item Alchemistry wiki: \url{http://www.alchemistry.org/wiki/Best_Practices}
-\end{itemize}
-
-\clearpage
-%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %  Benchmark information  %
-%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Benchmark Datasets}
-Several other case studies of alchemical calculations can be found in the review by Williams-Noonan et al.~\cite{williams-noonan2018free}.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Alchemical free energy datasets: an overview}
+\label{sec:benchmark}
+The following contains a non-exhaustive summary of alchemical free energy datasets that can serve as a starting point to review approaches or test new implementations.
+The field is moving towards a more standardised way of generating protein-ligand benchmark datasets and the progress of these efforts can be tracked here: \url{https://github.com/openforcefield/FE-Benchmarks-Best-Practices}.
+Currently lacking an exhaustive set of benchmark datasets, the review by Williams-Noonan et al.~\cite{williams-noonan2018free} contains an overview of recently published alchemical free energy studies.
 For comparison of FEP+ and Gromacs (using the AMBER99SB-ILDN and GAFF2 force field), cf.
@@
 These include cyclodextrins, the Cytochrome C peroxidase (CCP) protein model binding site, thrombin and bromodomains as well as solvation benchmark sets~\cite{paliwal2011benchmark}.
-Further work on shared protein-ligand benchmark sets is ongoing at \url{https://github.com/openforcefield/PLBenchmarks}.
-\label{sec:benchmark}
-\vspace{5mm}
+Please refere to ~\ref{tab:benchmarks}, for a small overview of datasets, what forcefields they used, and what the original study was it came from.
 \begin{table}
@@
     \item Will my protocol be reproducible? 
-    \item Will my statistics be reliable? If not, would more replicates solve the problem? Three replicates is the minimal number of replicates that should be run
+    \item Will my statistics be reliable? If not, would more replicates solve the problem? 
     \item Can I open-source my data?
@@
         \item Make sure your molecular transformations are realistic (1-5 heavy atoms for reliable computations)
-        \item Generate a perturbation network by your method of choice; check whether you have enough cycle closures to check consistency in the results (preferably, every ligand should be in a cycle)
+        \item Generate a perturbation network by your method of choice; check whether you have enough cycle closures to check consistency in the results
         \item Check whether dummy atoms were assigned correctly
@@
 %%%%%%%%%%%%%%%%
-\todo[inline, color={green!20}]{ASJSM: @everyone, can you check your contribution is acknowledge and if not add the appropriate information. }
-(Explain the contributions of the different authors here)
-\textbf{ASJSM}: Coordinated the document, contributed to most sections, and co-designed figures ~\ref{} and created figures \ref{fig:fig_what_is_lambda},~\ref{fig:fig1_what_is_alchemy},~\ref{fig:overlap} and replotted~\ref{fig:automatic-equilibration-detection} and improved ~\ref{fig:fig_types_of_networks}.\\
+\textbf{ASJSM}: Coordinated the document, contributed to most sections, and co-designed figures~\ref{fig:fig_binding_thermodynamic_cycle},~\ref{fig:fig_topology},~\ref{fig:fig_mcss},~\ref{fig:fig_types_of_networks},~\ref{fig:fig_absolute_thermodynamic_cycle},~\ref{fig:scatterplot_analysis},and created figures \ref{fig:fig_what_is_lambda},~\ref{fig:fig1_what_is_alchemy},~\ref{fig:overlap} and replotted~\ref{fig:automatic-equilibration-detection} and improved ~\ref{fig:fig_types_of_networks}.\\
 \textbf{JS}: created figures \ref{fig:fig_what_is_alchemy},~\ref{fig:fig_binding_thermodynamic_cycle},~\ref{fig:fig_topology},~\ref{fig:fig_mcss},~\ref{fig:fig_types_of_networks},~\ref{fig:fig_absolute_thermodynamic_cycle},~\ref{fig:scatterplot_analysis}, the data representation section, the checklist and contributed to general formatting discussions and editing.\\

### 2020-07-13 paper/figures/fig13_analysis_practices/gen_plots.py
removing outlies in fig13

@@ -88,8 +88,8 @@ plt.fill_between(
 yerr = np.random.uniform(low=0.1, high=0.4, size=(num_points,))
 # append high errors for the outliers and plot them in a different colour:
 yerr = np.append(yerr, [0.7, 0.9])
-plt.scatter(x=-11, y=-8.3, color="crimson")
-plt.scatter(x=-12, y=-9, color="crimson")
+plt.scatter(x=-11, y=-8.3, color='#1f77b4')
+plt.scatter(x=-12, y=-9, color='#1f77b4')
 
 
 plt.errorbar(x_data, y_data, 
@@ -169,7 +169,7 @@ sns.despine()
 plt.xlim(-12.9, -7.1)
 plt.ylim(plt.xlim())
 plt.xlabel(r"Experimental $\Delta$G [kcal$\cdot$mol$^{-1}$]", fontsize=18)
-plt.ylabel(r"AFE-predicted $\Delta$G [kcal$\cdot$mol$^{-1}$]", fontsize=18)
+plt.ylabel(r"Computed $\Delta$G [kcal$\cdot$mol$^{-1}$]", fontsize=18)
 plt.tight_layout()
 
 

### 2020-07-15 paper/manuscript.tex

@@
 \author[1,10]{Maximilian Kuhn}
+\author[2]{Hannah E. Bruce Macdonald}
 \author[1*]{Julien Michel}
@@
 Essentially, the method is dominated by contributions of rare snapshots\cite{jarzynski2006rare, wu2005phasespaceb, wu2005phasespacec}.
-There are a number of applications where this expression is nevertheless useful, but those usually involve investigating the dependence of the free energy on small changes in parameters, not changing molecules or even single atoms in molecules~\cite{xx}
-%
+
 \paragraph{The Bennett Acceptance Ratio (BAR)}
-%
 If we have the differences in the potential energy sampled from the distribution defined by $U_0$ to the state defined by $U_1$, and we also have the differences in potential energies from the distribution sampled by $U_1$ to the state defined by $U_0$, we can obtain a significantly improved estimate of the
@@
 will refer to this method as BAR.
-More recent derivations show that this formula is the maximum likelihood estimate of the free energy difference given sets of samples from the two states~\cite{bennett1976efficient,shirts2003equilibriuma}.
+More recent derivations show that this formula is the maximum likelihood estimate of the free energy difference given sets of samples from the two states~\cite{shirts2003equilibriuma}.
 
@@
 BAR also requires significantly less overlap between the configurational space of each state to converge than EXP, though some overlap must still exist.
-%
+
 The Bennett acceptance ratio is only defined between two states.
@@
 You can simply carry out BAR between each pair of states $\Delta G_{1 \rightarrow N} = \Delta {G_{1\rightarrow 2}} + \Delta {G_{2\rightarrow 3}} +  \ldots + \Delta G_{N-1\rightarrow N}$.
-%
+
 There is one important thing to note about the uncertainty estimates when summing multiple free energies together to calculate an overall free energy estimate.
@@
 Instead, some other method for propagating the uncertainty, such as bootstrapping~\cite{grossfield2018best} must be used.
-%
+
 \paragraph{Thermodynamic integration (TI)}
@@
 Rather than being limited by overlap, as in the case of BAR and MBAR (see below), we are instead limited by the bias in the numerical quadrature, which must be minimized sufficiently to be beneath the level of statistical noise.
-%
+
 Various numerical integration schemes are possible, but the trapezoid
@@
 In practice, adding 2-3 more intermediate states is typically sufficient to match the performance of these more complicated numerical quadrature schemes.
-%
+
 One drawback of TI is that it requires derivatives with respect to $\vec{\lambda}$ to be calculated directly in the code.
@@
 Still, if the code of interest does compute $\frac{du}{d\vec{\lambda}}$, then TI is perhaps the simplest method to use, as it involves a very little post-processing.
-%
+
 \paragraph{The multistate Bennett acceptance ratio (MBAR)}
@@
 We note that there are only $K-1$ independent equations, so only K-1 of the free energies are independent variables, and one of the $f_i$ must be specified (usually, without loss of generality, setting it to zero).
-%
+
 MBAR is provably the lowest variance asymptotically unbiased estimator of the free energies given the energies of the samples~\cite{tan2004likelihood}, which means that BAR is also for the free energy difference between only 2 states, as it is mathematically exactly the same as MBAR in this case.
 MBAR also provides an uncertainty estimate, derived from standard error propagation methods, which has been shown to be accurate as long as there are sufficient samples at each state~\cite{paliwal2011benchmark}.
-%
+
 MBAR can also be thought of as the Zwanzig estimator of the free energy to state $i$ where the sampled distribution is the \textit{mixture distribution} of all the other samples thrown together in one ``pot'', defined by $p_m(\vec{q}) = N^{-1} \sum_k  N_k \exp(f_k-u_k\vec{q})$, the weighted average of all the individual normalized probability distributions that are performed.~\cite{shirts2017reweighting}.
-%
+
 \paragraph{Recommendations}
@@
 \end{itemize}
-%
+
 \subsection{Uncertainty estimation}
@@
 Therefore, $\mathrm{var}$($\Delta$f) = $\sum_{i=1}^{K}w_{k}^2 \mathrm{var}(\frac{du}{d\vec{\lambda}})_{k}$.
-%
+
 For alchemical changes that result in smooth, low curvature sets of $\expect{\frac{dU}{d\vec{\lambda}}}$, a relatively small number of $\vec{\lambda}$ states is necessary for sufficient accuracy and low variance in the free energy estimate.
@@
 It is recommended that researchers verify that a sufficient number of states are included such that the free energy is essentially invariant to the number of lambda intermediate states chosen.
-%
+
 Compared with TI, the MBAR method (section~\ref{subsec:estimators}) discussed above provides uncertainty estimation directly from solving a set of linear equations to compute the variances between all states.
 The number of states and amount of sampling should be optimized to minimize the uncertainty in the MBAR free energy estimate, while balancing other key considerations such as computational expense.
-%
+
 In some cases, several analysis methods can be employed to analyze the same set of simulations, providing an opportunity for synergy.
@@
 Those perturbations where free energies across different estimators shared high agreement with each other (variance less than 0.1 kcal/mol) had a r$^2$ of 0.623 and mean unsigned error (MUE) of 0.787 kcal/mol (N=17) compared with a r$^2$ of 0.295 and MUE of 1.49 kcal/mol (N=23) (unpublished results).
-%
+
 Uncertainty can also be assessed for a particular perturbation by repeating calculations with slight changes in initial configurations, forcefield parameters, and different random seeds in the MD engine.
@@
 Common issues include poor ligand or protein preparation, atom mapping for relative calculations, or forcefield parameters as described above.
-%
+
 For relative binding free energy calculations, additional sensitivity analysis can be performed by changing the initial configurations of non-core regions of the perturbation topology and determining if this change in configurations results in a large differences in the computed relative free energy, indicating poor sampling of ligand configuration.
@@
 \begin{figure}
-    \includegraphics[width=0.95\linewidth]{figures/fig13_analysis_practices/Figure.pdf}
+    \includegraphics[width=0.95\linewidth]{figures/fig13_analysis_practices/FigureHBMedit.png}
     \caption{\textbf{An example of recommended practices for graphing alchemical free energy predictions.} This figure shows the relation between predicted and experimentally-determined Gibbs free energy in kcal/mol with standard errors as error bars.
@@
 \textbf{HX}: Contributed section~\ref{subsec:accuracy}, to section~\ref{sec:relative-fe-protocol}, and to section~\ref{sec:are-they-good}.\\
+\textbf{HBM} Contributed to section \ref{sec:plot_data}and Figure \ref{fig:scatterplot_analysis}.\\
 % We suggest you preserve this comment:

### 2020-07-24 paper/manuscript.tex

@@
 \author[7]{Bryce K. Allen}
+\author[2]{Hannah E. Bruce Macdonald}
 \author[2*]{John D. Chodera}
 \author[1,10]{Maximilian Kuhn}
-\author[2]{Hannah E. Bruce Macdonald}
 \author[1*]{Julien Michel}
@@
 \dateaccepted{Month, Day, Year}
-%
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \begin{abstract}
-Alchemical free energy calculations are a useful tool for predicting free energy differences associated with the transfer of small molecules from one environment to another.
+Alchemical free energy calculations are a useful tool for predicting free energy differences associated with the transfer of molecules from one environment to another.
 The hallmark of these methods is the use of "bridging" potential energy functions representing \emph{alchemical} intermediate states that cannot exist in as real chemical species.
 The data collected from these bridging alchemical thermodynamic states allows the efficient computation of transfer free energies (or differences in transfer free energies) with orders of magnitude less simulation time than simulating the transfer process directly.
-%
+
 While these methods are highly flexible, care must be taken in avoiding common pitfalls to ensure that computed free energy differences can be robust and reproducible for the chosen force field, and that appropriate corrections are included to permit direct comparison with experimental data.
-%
+
 In this paper, we review current best practices for several popular application domains of alchemical free energy calculations, including relative and absolute small molecule binding free energy calculations to biomolecular targets.
@@
 \end{frontmatter}
-%
+
 \todototoc
 \listoftodos
-%
+
 %%%%%%%%%%%%%%%%%%%%
@@
 \label{sec:intro}
-\todo[inline, color={green!20}]{@everyone: Any further input in this?}
-Alchemical free energy calculations compute free energy differences associated with transfer processes, such as the binding of a small molecule to a receptor, the transfer of a small molecule from an aqueous to apolar phase~\cite{zwanzig1954hightemperature}, or effects of protein side chain mutations on e.g.
-binding affinities.
-These calculations use non-physical\footnote{Here, the non-physical nature of the transformation is referred to as "alchemical", a term coined by Tembre and McCammon in Ref.~\cite{tembre1984ligandreceptor}.} intermediate states in which the chemical identity of some portion of the system (such as a small molecule ligand or receptor sidechain) is changed by modifying the potential governing the interaction of the atoms being modified, inserted, or deleted with their environment.
-%
+Alchemical free energy calculations compute free energy differences associated with transfer processes, such as the binding of a small molecule to a receptor, the transfer of a small molecule from an aqueous to apolar phase~\cite{zwanzig1954hightemperature}, or the effects of protein side chain mutations on binding affinities or thermostabilities.
+ These calculations use non-physical\footnote{Here, the non-physical nature of the transformation is referred to as "alchemical", a term coined by Tembre and McCammon in Ref.~\cite{tembre1984ligandreceptor}.} intermediate states in which the chemical identity of some portion of the system (such as a small molecule ligand or protein sidechain) is changed by modifying the potential governing the interactions with the environment for the atoms being modified, inserted, or deleted.
+
 Fig.~\ref{fig:fig_what_is_alchemy} illustrates common free energy changes that may be difficult to compute with unbiased molecular dynamics methods, but are more readily tractable with alchemical methods.
@@
 \end{itemize}
-%
-
-Once an alchemical simulation procedure is performed (which usually involves more than one simulation), the data must be analyzed to estimate the free energy for the transformation of interest.
-Early work used simple but statistically suboptimal estimators for this: free energy perturbation (FEP) using the Zwanzig relation~\cite{zwanzig1954hightemperature} or thermodynamic integration (TI), for which the theory dates back decades but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamicsa, merz1989free}.
-%
-More recent developments have seen new, highly efficient statistical estimators that make better use of all the data, often building on the Bennett acceptance ratio (BAR)~\cite{bennett1976efficient}, producing multistate generalizations~\cite{shirts2008statisticallya} or removing the need for global equilibrium~\cite{wu2016multiensemble, mey2014xtram, wu2014statistically}.
-
-Important studies continued during the 2000's often leading to improved implementation of the methods in common simulation software~\cite{shirts2003extremely,shirts2005solvation,vanderspoel2005gromacs, mermelstein2018fast, wang2015accurate, hedges2019biosimspace}.
+
+After an alchemical calculation is performed (which generally involves multiple simulations at a variety of alchemical states), the data must be analyzed to compute an estimate the free energy for the transformation of interest.
+ Early work used simple but statistically suboptimal estimators for this: free energy perturbation (FEP) used a simple (but highly biased) estimator based on the Zwanzig relation~\cite{zwanzig1954hightemperature} or numerical quadrature via thermodynamic integration (TI), for which the theory dates back decades but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamicsa, merz1989free}.
+%
+ More recent developments have seen new, highly efficient statistical estimators that make better use of all the data, often building on the more efficient and less biased Bennett acceptance ratio (BAR)~\cite{bennett1976efficient}, producing multistate generalizations~\cite{shirts2008statisticallya} or removing the need for global equilibrium~\cite{wu2016multiensemble, mey2014xtram, wu2014statistically}.
+
+Subsequent work in the 2000s led to improved implementations of alchemical methods in popular biomolecular simulation packages~\cite{shirts2003extremely,shirts2005solvation,vanderspoel2005gromacs, mermelstein2018fast, wang2015accurate, hedges2019biosimspace}.
  This foundational work, combined with the methodological, technological, and hardware improvements of the last 5--10 years, have led to an explosion of interest and direct commercial application of the technologies~\cite{wang2015accurate, fratev2019improved, schindler2020largescale, cournia2017relative, sherborne2016collaborating}.
-%
-
-As the field of molecular simulation moves towards millisecond timescales, accurate alchemical calculations on even more challenging problems will become reasonable to perform.
-In the meantime, today's users may find it hard to get started with these complex calculations whilst also keeping up with the fast pace of change.
-This Best Practice guide provides current recommendations and tips for users of all experience.
-%
+
+As the field of molecular simulation can now routinely access microsecond timescales with the aid of GPUs~\cite{salomon-ferrer2013routine} and millisecond timescales appear to soon be within reach, accurate alchemical calculations on even more challenging problems will become reasonable to perform.
+In the meantime, today's users may find it difficult to get started with these complex calculations whilst also keeping up with the fast pace of change. 
+This Best Practices guide provides current recommendations and tips for users of all experience.
+
 \begin{figure}
@@
 \label{sec:pre}
-What can be expected from this Best Practice guide is a good starting point for new practitioners and a reference for experienced practitioners.
-For this propose we provide a convenient checklist (Section~\ref{sec:checklist}) to help ensure all calculations meet standard alchemical simulation and analysis practices.
+What can be expected from this Best Practices guide is a good starting point for new practitioners and a reference for experienced practitioners. 
+ For this propose we provide a convenient checklist (Section~\ref{sec:checklist}) to help ensure all calculations comply with currently-understood best practices for alchemical simulation and analysis.
+ Where the best practices are currently unsettled, we highlight areas where further research is needed to identify an unambiguous recommendation.
  It can also serve as a set of best practices to ensure simulation robustness and reproducibility which reviewers may wish to consider as they evaluate papers.
+
  What we assume from a novice practitioner is a moderate experience with molecular simulation packages and concepts. 
@@
 If you feel unfamiliar with some of these concepts, good starting points can be found in these references~\cite{braun2019best, grossfield2018best, klimovich2015guidelines, shirts2012best}.
-While reading this Best Practice document, it is important to bare in mind \emph{this is not a review} of all free energy calculation methods at the cutting edge of current research.
+
+ While reading this Best Practices guide, it is important to bear in mind \emph{this is not a review} of all free energy calculation methods at the cutting edge of current research.
 Instead the guide aims to answer the following questions:
 \begin{itemize}
-    \item Is my problem suitable for an alchemical free energy calculation? 
-    \item How do I choose the right free energy protocol and run it? 
-    \item How do I analyze my simulations accurately? 
-    \item What software tools are available to perform alchemical free energy calculations? 
-\end{itemize}
-%
+    \item Is my problem suitable for an alchemical calculation? 
+     \item How do I select an appropriate alchemical protocol? 
+     \item What software tools are available to perform alchemical calculations? 
+     \item How should I analyze my data and report uncertainties? 
+\end{itemize}
 
@@
 This will require some basic familiarity on how to perform docking calculations to generate reasonable simulation starting points~\cite{grinter2014challenges}.
-%
+
 As some of the theoretical background can seem daunting, we do, however, provide a guide to the essential theory behind alchemical free energy calculations in Section~\ref{sec:theory}.
@@
 Particular focus will be given to aspects of the molecular simulations which are unique to alchemical calculations---these include the calculation of transfer free energies (hydration free energies, partition coefficients, etc.), and binding free energies (absolute and relative).
-%
 
 While we try to address as many methods and practices as possible, the field of free energy calculations is broad, and there are many advanced topics that are left to future Best Practices documents focusing on specific issues.
-Below, we provide a non-exhaustive list of topics we have not addressed with some references to provide starting points on these more advanced topics:
+Below, we provide a non-exhaustive list of topics we have \emph{not} addressed, along with some references to provide starting points on these more advanced topics:
 \begin{itemize}
@@
 \item nonspecific binding or multiple binding sites~\cite{gill2018binding}
-\item complex association stoichiometry (beyond simple 1:1 association).
 \item approximate and often less accurate endpoint free energy methods such as MM-PBSA~\cite{genheden2015mm} and LIE~\cite{gutierrez-de-teran2012linear}
@@
 \end{itemize}
-%
 
@@
 \begin{tcolorbox}[title=Acronyms, colback=blue!10!white]
-    \textit{CPU} -- Central Processing Unit\\
-    \textit{BAR} -- Bennett Acceptance Ratio\\
-    \textit{FEP} -- Free Energy Perturbation\\
-    \textit{GPCR} -- G-Protein Coupled Receptor\\
-    \textit{GPU} -- Graphics Processing Unit\\
-    \textit{MBAR} -- Multistate Bennett Acceptance Ratio\\
-    \textit{MCSS} -- Maximum Common Substructure\\
-    \textit{MD} -- Molecular Dynamics\\
-    \textit{RMSE} -- Root Mean Square Error\\
-    \textit{MUE} -- Mean unsigned error\\
-    \textit{SAR} -- Structure-Activity Relationships\\
-    \textit{TI} -- Thermodynamic Integration
+    {\bf CPU} -- Central Processing Unit\\
+     {\bf BAR} -- Bennett Acceptance Ratio\\
+     {\bf FEP} -- Free Energy Perturbation\\
+     {\bf GPCR} -- G-Protein Coupled Receptor\\
+     {\bf GPU} -- Graphics Processing Unit\\
+     {\bf MBAR} -- Multistate Bennett Acceptance Ratio\\
+     {\bf MCSS} -- Maximum Common Substructure\\
+     {\bf MD} -- Molecular Dynamics\\
+     {\bf RMSE} -- Root Mean Square Error\\
+     {\bf MUE} -- Mean unsigned error\\
+     {\bf SAR} -- Structure-Activity Relationships\\
+     {\bf TI} -- Thermodynamic Integration
 \end{tcolorbox}
@@
 $\vec{\lambda}$ -- alchemical progress parameter \\
-$g$ -- statistical inefficiency
-\end{tcolorbox}
-
-\begin{tcolorbox}[title=Common terminology, colback=yellow!10!white]
-    \emph{ligand} - small molecule binding to a larger host molecule such as a protein (receptor).\\ 
-    \emph{lead optimization} - process of designing a drug candidate from an initial lead compound, usually involving optimizing its affinity, possibly selectivity, and various pharmacokinetic parameters.\\
-    \emph{docking} -- Method of generate simulation starting structures for ligand receptor systems, when no crystal structure is available\\
-    \emph{thermodynamic cycle} - sequence of thermodynamic processes (like changes in temperature, volume or pressure) in which the initial and final state are equal.\\
-    \emph{alchemical} -- states or simulations that contain non-physical intermediates \\
+$g$ -- statistical inefficiency\\
+$\omega$ - overlap matrix\\
+$C_t$- discrete-time-normalized fluctuation auto-correlation function\\
+$\tau _{eq}$ - integrated auto-correlation time\\
+$t_0$ - equilibration time
 \end{tcolorbox}
@@
 A natural though generally very computational expensive way to estimate the equilibrium constant is by directly simulating several binding and unbinding events and computing the probability of finding the receptor-ligand system in the bound state, $P(RL)$, or the unbound state, $P(R+L)$.
-Assuming the volume change upon binding to be negligible, which is often the case at $1~\mathrm{atm}$ due to the incompressibility of water, then the Gibbs free energy $\Delta G_{\mathrm{bind}, L}$ is approximately equal to the Helmholtz free energy $\Delta A_{\mathrm{bind}, L}$, and we can simulate the system in a box of volume $V$ to obtain~\cite{jong_determining_2011}
+Assuming the volume change upon binding to be negligible, which is often the case at $1~\mathrm{atm}$ due to the incompressibility of water, then the Gibbs free energy $\Delta G_{\mathrm{bind}, L}$ is approximately equal to the Helmholtz free energy $\Delta A_{\mathrm{bind}, L}$, and we can simulate the system in a box of volume $V$ to obtain~\cite{jong2011determining}
 \begin{equation}\label{eq:binding-free-energy-from-bound-unbound-probability-ratio}
@@
     P(\vec{q}) = \frac{\exp\left( \beta U(\vec{q}) \right)}{\int_{\Gamma} \exp\left( \beta U(\vec{q}) \right) \, d\vec{q}} \, ,
+    \label{eq:conf_probability}
 \end{equation}
@@
 Predictions of $\Delta \Delta G_{\mathrm{bind}, AB}$ with non-alchemical methods generally require long simulations of both ligands, possibly through different binding pathways.
-Alchemical relative free energy calculations avoid the need to simulate binding and unbinding events by making use of the fact that the free energy is a state function and exploiting the thermodynamic cycle illustrated in Figure~\ref{fig:fig_binding_thermodynamic_cycle}.
+Alchemical relative free energy calculations avoid the need to simulate binding and unbinding events by making use of the fact that the free energy is a state function and exploiting the thermodynamic cycle illustrated in Fig.~\ref{fig:fig_binding_thermodynamic_cycle}.
 This is apparent after rewriting Eq.~\ref{eq:delta-delta-G-physical} as
@@
 where $\Delta G_{\mathrm{bound/unbound}}$ is the free energy of mutating $A$ to $B$ in the bound/unbound state.
-Eq.~\ref{eq:delta-delta-G-alchemical} and Figure~\ref{fig:fig_binding_thermodynamic_cycle} tell us that the difference in free energy of binding between toluene ($A$) and benzyl alcohol ($B$) can be computed by running two independent calculations estimating the free energy cost of mutating $A$ into $B$ in the binding pocket ($\Delta G_{\mathrm{bound}}$) and in solvent ($\Delta G_{\mathrm{unbound}}$), saving us the need to simulate the physical binding process of the two compounds.
+Eq.~\ref{eq:delta-delta-G-alchemical} and Fig.~\ref{fig:fig_binding_thermodynamic_cycle} tell us that the difference in free energy of binding between toluene ($A$) and benzyl alcohol ($B$) can be computed by running two independent calculations estimating the free energy cost of mutating $A$ into $B$ in the binding pocket ($\Delta G_{\mathrm{bound}}$) and in solvent ($\Delta G_{\mathrm{unbound}}$), saving us the need to simulate the physical binding process of the two compounds.
 In particular, the second line of Eq.~\ref{eq:delta-delta-G-alchemical} is a consequence of $\Delta G_{\mathrm{unbound}}$ being independent of the presence of the receptor in the simulation box as the definition of the unbound state assumes receptor and ligand to be at a sufficient distance for them to have no energetic interactions.
 Note that, when $A$ and $B$ have different number of atoms, Eq.~\ref{eq:delta-delta-G-alchemical} implies the presence of a factor having units of volume entering both logarithms, which require unitless arguments.
-The factor's value is, however, inconsequential as it cancels out in practice.
+However, the value of this factor is inconsequential as it cancels out in practice.
 
@@
 While it is theoretically possible to estimate the ratio of partition functions from samples collected only at states $\vec{\lambda}_A$ and $\vec{\lambda}_B$, the efficiency of the free energy estimators rapidly decreases as the phase-space overlap between the two states also decreases~\cite{wu2005phasespaceb,wu2005phasespacec}.
-Roughly, the phase-space overlap between two thermodynamic states measures the degree to which high-probability conformations (i.e., those with very negative potential energy) in one state are high-probability conformations in the other state as well (see Section~\ref{sec:are-they-good} and Fig.~\ref{fig:fig_what_is_lambda}).
+Roughly, the phase-space overlap between two thermodynamic states measures the degree to which high-probability conformations (i.e., those with very negative potential energy) in one state are also high-probability conformations in the other state(see Section~\ref{sec:are-they-good} and Fig.~\ref{fig:fig_what_is_lambda}).
 
@@
 The main difference in these approaches lies instead in the thermodynamic cycle to which this strategy is applied.
-For example, a typical thermodynamic cycle for an alchemical absolute binding free energy calculation is represented in Figure~\ref{fig:fig_absolute_thermodynamic_cycle}.
+For example, a typical thermodynamic cycle for an alchemical absolute binding free energy calculation is represented in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}.
 In this case, two independent calculations compute the free energy of removing the interactions between the ligand and its environment in solvent or in the binding site respectively through a series of intermediate states in which the energy terms are only partially deactivated.
@@
 inhibitors selectively bind to different kinases in the so-called
-DFG-out conformations (cite Kuriyan).  The selectivity of such
+DFG-out conformations~\cite{schindler2000structural}.  The selectivity of such
 inhibitors may be attributed either to their differential binding to
@@
 If so, how will the solvent be incorporated for smaller ligands that require the water -- consider pre-solvation methods such as GCMC steps~\cite{michel2010prediction}.
-Other tools such as WaterMap or open source equivalents (SSTmap, others) can be used to define water structure for systems with no experimental evidence of water sites.
+Other tools such as WaterMap or open source equivalents (SSTmap, others) can be used to define water structure for systems with no experimental evidence of water sites~\cite{wang2011ligand}.
 Before launching large numbers of alchemical free energy calculations it is recommended to test the system using classical MD simulations and limited numbers of alchemical perturbations.
@@
 In this way the top-ranking examples can be readily identified and submitted to additional calculations in a second round.
-Alternatively, if the goal is to achieve the smallest possible error with minimal computational expense, certain graph topologies provide benefits~\cite{https://doi.org/10.1002/jcc.26095, https://doi.org/10.1021/acs.jcim.9b00528}
+Alternatively, if the goal is to achieve the smallest possible error with minimal computational expense, certain graph topologies provide benefits~\cite{yang2020optimal, xu2019optimal}
 
@@
 The following questions may be relevant before diving into the simulation setup.
-% 
+
 \begin{itemize}
@@
     \end{itemize}
-%
+
 \paragraph{Are there good X-ray structures available?}
@@
 Typically, clustering of receptor structures can be used to identify different receptor conformations near the binding site, as well as assessing relevant side chain placements from the X-ray structure, see for example~\cite{mey2016blinded}.
-In terms of set up and other choices, following general best practice guidelines is advisable~\cite{braun2018best}.
+In terms of set up and other choices, following general best practice guidelines is advisable~\cite{braun2019best}.
 
@@
 \paragraph{Protonation states depend on the pH of the experimental assay}
-I think we should have a tiny bit on protonation states. This is something I always easily forget as a non-chemist.
+Care should be taken when preparing ligands and proteins to match the experimental assay conditions in terms of pH, if they are known.
+The pH of the assay will determine the protonation states of the proteins and ligands.
+Since the pKa of amino acid residues is known different tools are available for proteins, such as the H++ server, or Maestro are good starting points~\cite{anandakrishnan2012automating, 2020schrodinger}.
+For ligands often the pKa needs to be deterimined if not known experimentally.
+There are many different available tools for this purpose, but common choices may be propKa~\cite{olsson2011propka3,sondergaard2011improved}, Chemicalize (\url{https://chemicalize.com/welcome}), or Maestro~\cite{2020schrodinger}.
+While mostly it can be assumed that protonation states will remain the same for an e.g.
+ligand protein binding calculation, some care needs to be taken with systems where the protonation state may change upon binding~\cite{onufriev2013protonation}.
 
@@
 \end{figure}
-%
+
 There are different tools that allow the generation of MCSS matches as well as single topology input.
@@
 Schr\"{o}dinger's FEP+ planning tool is based on a version of LOMAP, and it also uses MCSS matching as well as 3D considerations to plan the network of single topology calculations between molecules~\cite{wang2015accurate}.
-%
+
 MCSS searches can be relatively time consuming, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial assessment and then use MCSS only to identify final mappings for relative calculations~\cite{raymond2002maximum,klabunde2012mars,jones2009elucidating}.
-%
+
 The MCSS approach, though relatively standard, takes into account only topological similarity.
 It is possible that changes in binding mode could actually require a different choice of mapping, so in some cases mappings may need to be planned differently depending on 3D positioning of atoms in space.
-%
+
 Single topology relative calculations, and calculations based on substructure searches, only work if in fact the ligands share a common substructure, e.g.
@@
 Using an absolute free energy approach  instead seems more promising in such a case.
-%
+
 \paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss} \textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings.
@@
 (at end of first sentence of this pgraph I would also cite this work from Liu, Wang and Mobley, https://pubs.acs.org/doi/abs/10.1021/acs.jcim.5b00057)
-%
+
 \paragraph{Perturbation maps}
@@
 
-%
 Methods in experimental design have been applied to the construction of the perturbation maps.
@@
 
-%
 \paragraph{Constraints and relative free energy calculations}
@@
 It is convenient to define a standard state at a constant pressure of 1~atm and where each chemical species (i.e., A, B, and AB) in the reaction solvent has a concentration of $c^{\circ}$~=~1~M~=~1~molecule/1660~\r{A}$^3$ but do not interact with other molecules of $R$, $B$, or $RL$.
-%
+
 \paragraph{Handling the standard state in absolute free energy calculations.}
@@
 So typically in such cases no special care is required to ensure the correct standard state, as long as the \emph{experimental} data being analyzed uses the same standard state and if it does not, a simple entropic correction is needed.
-%
+
 However, for binding the situation is much more complex and requires special care.
@@
 In practice, this has implications for how the calculations are done, as the reference concentration must enter the thermodynamic cycle employed, see an example of such a thermodynamic cycle in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}.
-%
+
 Typically, to deal with both practical sampling issues and the standard state issue, restraints are employed in absolute binding free energy calculations to keep the ligand in a well defined volume as its interactions with the system are removed~\cite{gilson1997statisticalthermodynamic}.
@@
 In contrast, with restraints, the free energy of releasing the restrained ligand to a 1 M standard state can be computed analytically or numerically by solving the relevant integral, allowing the standard state to enter the thermodynamic cycle.
-%
+
 \paragraph{Several choices of restraints are possible.}
@@
 \subsubsection{Handling weak binders and high dissociation rates}
-\todo[inline, color={green!20}]{ASJM: This a bit waffly. What are the recommendations, of how to setup these restraints?}
-\todo[inline, color={red!40}]{AR: Took a pass at this.}
 In binding free energy calculations, only the conformations in which the receptor and ligand form a bound complex should be sampled from the bound states (Section~\ref{sec:theory}).
-In absolute calculations, this applies to the fully interacting state in the complex leg of the thermodynamic cycle (top-right state in Figure~\ref{fig:absolute-thermodynamic-cycle}), while in relative calculations this must be true at both end states of the complex leg (top- and bottom-right states in Figure~\ref{fig:binding-thermodynamic-cycle}).
+In absolute calculations, this applies to the fully interacting state in the complex leg of the thermodynamic cycle (top-right state in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}), while in relative calculations this must be true at both end states of the complex leg (top- and bottom-right states in Fig.~\ref{fig:fig_binding_thermodynamic_cycle}).
 In principle, this requires defining which conformations are considered to be bound before running the calculation, but it is common practice to start the simulation with the ligand already placed in the binding site and rely on kinetic trapping to maintain the bound complex.
 This strategy, however, can fail when the dissociation rate of the ligand has the same or smaller order of magnitude than the length of the simulation.
-This is typical of weak binders such as fragments binding shallow pockets with $\mu$M-mM affinities~\cite{georgiou_pushing_2017,pan2017quantitative}.
-In this case, using a flat-bottom or harmonic restraint between receptor and ligand in the bound state(s) can prevent dissociations~\cite{georgiou_pushing_2017,rizzi2019sampl6}.
+This is typical of weak binders such as fragments binding shallow pockets with $\mu$M-mM affinities~\cite{georgiou2017pushing,pan2017quantitative}.
+In this case, using a flat-bottom or harmonic restraint between receptor and ligand in the bound state(s) can prevent dissociations~\cite{georgiou2017pushing,rizzi2019sampl6}.
 We stress that this is normally avoided as it generally introduces bias in the free energy estimate, which is why the restraint is usually activated only in the intermediate states in absolute calculations.
@@
 Alternatively, a flat-bottom restraint can be activated also in the decoupled state as long as the second restraint (e.g., a harmonic or Boresch restraint) prevents the ligand to hit the wall of the flat-bottom potential~\cite{rizzi2019sampl6}.
-Finally, even for tight binders, dissociation events can be enhanced by methods such as Hamiltonian replica exchange~\ref{sugita2000multidimensionala,chodera2011replica,wang2013identifying} and expanded ensemble~\ref{lyubartsev1992newa,li2007simulated}, especially in absolute free energy calculations using harmonic or flat-bottom restraints.
-In the latter case, dissociations can be averted simply by increasing the spring constant and/or reducing the radius of the restraint potential to prevent the exploration of ligand conformations outside the binding site in the decoupled state (bottom-right state in Figure~\ref{fig:absolute-thermodynamic-cycle}) that could be propagated to the bound state.
-%
+Finally, even for tight binders, dissociation events can be enhanced by methods such as Hamiltonian replica exchange~\cite{sugita2000multidimensionala,chodera2011replica,wang2013identifying} and expanded ensemble~\cite{lyubartsev1992newa,li2007simulated}, especially in absolute free energy calculations using harmonic or flat-bottom restraints.
+In the latter case, dissociations can be averted simply by increasing the spring constant and/or reducing the radius of the restraint potential to prevent the exploration of ligand conformations outside the binding site in the decoupled state (bottom-right state in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}) that could be propagated to the bound state.
+
 \subsubsection{Changes in net charge can be challenging/problematic.}
@@
 Specifically, finite-size effects can introduce profound artifacts into computed binding free energies~\cite{}MISSING citation, in part because typical schemes for long-range electrostatics (including PME and reaction field) do not handle free energy contributions from such changes effectively or as they would be handled in a hypothetical macroscopic bulk solution~\cite{}.
-%
+
 There are two main potential solutions to avoid artifacts due to changes in net charge: Correcting for the introduced artifacts, or avoiding changing the net charge.
-%
+
 Many relative free energy planning tools have been set up to avoid changing the net charge of the systems considered, including LOMAP~\cite{liu2013lead} and Schr\"{o}dinger's FEP+~\cite{wang2015accurate}.
@@
 However, application of such corrections typically remains less common than the use of a co-alchemical ion.
-%
+
 When free energy calculations \emph{do} need to change the charge of a ligand or solute, the literature does not yet seem to indicate what approach should be preferable, so considerable care should be taken.
 We are not yet aware of a careful comparison of charge corrections versus other approaches such as decoupling an ion at the same time, so in our view the issue of proper handling of charge mutations in the context of alchemical calculations remains a research problem.
-%
+
 \subsubsection{The importance of the alchemical pathway
 \label{sec:important_path}}
-\todo[inline, color={red!40}]{LNN: write common principles alchemical path choice. MRS: mostly wrote this, LNN can comment on it. LNN: LGTM}
 \begin{figure}
@@
 \end{figure}
-%
+
 Both absolute and relative calculations must choose an alchemical pathway connecting initial and final states.
@@
 \begin{equation}
-U(\vec{q},\vec{\lambda}) = \vec{\lambda} U_0(\vec{q}) + (1-\vec{\lambda})U_1(\vec{q})   
-\end{equation}
-%
+U(\vec{q},\vec{\lambda}) = \vec{\lambda} U_0(\vec{q}) + (1-\vec{\lambda})U_1(\vec{q}) \end{equation}
+
 so-called because the dependence on $\vec{\lambda}$ is linear.
@@
 A large number of schemes have been tried~\cite{beutler1994avoiding, zacharias1994separationshifted, blondel2004ensemble, pham2011identifying, pham2012optimal, naden2014linear, donnini2005molecular}, but the most common strategy that appears to be the best practice is to use a "soft-core" potential, of the form:
-%
+
 \begin{equation}
@@
 \end{equation}
-%
+
 Where $r_{ij}$ is the distance between two particles $i$ and $j$, $\epsilon_{ij}$ and $\sigma_{ij}$ are the Lennard-Jones parameters corresponding to the interaction between particles $i$ and $j$, and $\alpha$ is a constant (0.5 is optimal for the functional form shown above).
@@
 Other issues, such as whether absolute calculations should retain or remove intramolecular non-bonded interactions
-through either annihilation~\cite{hermans1997inclusion, mann2000modeling, boresch2003absolute, wang2006absoulte, mobley2006use} or decoupling~\cite{fujitani2005direct, mobley2006use} must be considered.
+through either annihilation~\cite{hermans1997inclusiona, mann2000modelinga, boresch2003absolutea, wang2006absolutea, mobley2006use} or decoupling~\cite{fujitani2005directa, mobley2006use} must be considered.
 Reasonable efficiency can be often obtained with either choice even if some are somewhat better or worse than others although there is not consensus on which is better in all situations.
@@
 The spacing depends to some extent on the choice of analysis method, though states should essentially be spaced equidistant in the relevant thermodynamic length~\cite{crooks2007measuringa, sivak2012thermodynamic}.
-For BAR/MBAR techniques this means that states should be spaced so that the statistical uncertainties between neighboring states be equal;~\ref{pham2012optimal, shenfeld2009minimizing}
+For BAR/MBAR techniques this means that states should be spaced so that the statistical uncertainties between neighboring states be equal;~\cite{pham2012optimal, shenfeld2009minimizing}
 Some schemes to adaptively optimize the spacing of intermediate states based on initial exploratory simulations have been proposed~\cite{hayes2017adaptive}.
@@
 A third approach borrows ideas from simulated tempering~\cite{marinari1992simulateda}.
-In this scheme a single replica rapidly explores all of $\vec{\lambda}$ space by working out optimal weights that allow switching between different intermediate $\vec{\lambda}$ values, as seen in Fig.~\ref{fig:fig_sampling_scheme}(\textbf{C}).
+In this scheme a single replica rapidly explores all of $\vec{\lambda}$ space by working out optimal weights that allow switching between different intermediate $\vec{\lambda}$ values, as seen in Fig.~\ref{fig:fig_sampling_scheme} \textbf{C} .
 This approach is also referred to as self adjusted mixture sampling~\cite{lyubartsev1992newa, li2007simulated, tan2017optimally} and while promising, has so far only been supported in OpenMM Tools~\cite{andrearizzi2019choderalab}, and to date has not seen published applications for binding free energy calculations.
@@
 A schematic of this approach is shown in Fig.~\ref{fig:fig_sampling_scheme} \textbf{D}.
-%
+
 \begin{figure}
@@
 
-Currently, we recommend using Hamiltonian replica exchange type sampling schemes (Fig.~\ref{fig:fig_sampling_scheme}\textbf{B}).
+Currently, we recommend using Hamiltonian replica exchange type sampling schemes (Fig.~\ref{fig:fig_sampling_scheme} \textbf{B}).
 If these are not available in the code of choice, running independent simulations at different $\vec{\lambda}$ values can be acceptable, especially when conformational sampling is fast (Fig.~\ref{fig:fig_sampling_scheme} \textbf{A}).
@@
 This highlights the importance of choosing the correct thermodynamic path to ensure you sample the required thermodynamic states as discussed in Sec.~\ref{sec:important_path}.
-%
+
 The condition of minimizing the statistical uncertainty of different free energy estimators below a sufficient threshold should be one metric monitored over the simulation.
@@
 If more samples are needed, the simulations can be resumed, or, started again in different initial conditions.
-%
+
 Convergence in other alchemical observables should also be monitored to determine if the defined phase space has been sufficiently sampled and enough decorrelated samples have been drawn.
@@
 ~\ref{fig:convergence_forward_reverse}).
-%
+
 \begin{figure}
@@
 \end{figure}
-%
+
 Each of these metrics have demonstrated promising results for diagnosing when a simulation has a convergence issue beyond simple convergence of uncertainty estimate. 
@@
 The confidence a user should have in a free energy estimate is significantly improved when both the uncertainty of the free energy estimate is low, and when other observables have reached a convergence.
-%
+
 Another aspect that can be monitored is bias in the simulation.
@@
 Therefore, it is a good idea to apply the estimators to different fractions of the data to see if the main estimator of free energy you have chosen is stable.
-%
+
 Each method requires different data from the simulation be collected.
@@
 Below is a sample of the minimal information you need for a set of common estimators (discussed in more detail in Sec.~\ref{subsec:estimators}):
-%
-\begin{itemize}
-    \item Thermodynamic Integration (TI) requires $\frac{\partial U(\vec{x})}{\partial\vec{\lambda}}$.
-    \item Exponential Averaging (EXP) needs \textit{either} $\Delta U_{k,k+1}(\vec{x})$ or $\Delta U_{k,k-1}(\vec{x})$, depending on the direction its being evaluated in.
-    \item Bennett Acceptance Ratio (BAR) needs \textit{both} $\Delta U_{k,k+1}(\vec{x})$ and $\Delta U_{k,k-1}(\vec{x})$.
-    \item Weighted Histogram Analysis Method (WHAM) and Multistate Bennett Acceptance Ratio (MBAR) both need the complete set of $\Delta U_{k,j} \, \forall \, j=\{1...K\}$.
+
+\begin{itemize}
+    \item Thermodynamic Integration (TI) requires $\frac{\partial u(\vec{q})}{\partial\vec{\lambda}}$.
+    \item Exponential Averaging (EXP) needs \textit{either} $\Delta u_{k,k+1}(\vec{q})$ or $\Delta u_{k,k-1}(\vec{q})$, depending on the direction its being evaluated in.
+    \item Bennett Acceptance Ratio (BAR) needs \textit{both} $\Delta u_{k,k+1}(\vec{q})$ and $\Delta u_{k,k-1}(\vec{q})$.
+    \item Weighted Histogram Analysis Method (WHAM) and Multistate Bennett Acceptance Ratio (MBAR) both need the complete set of $\Delta u_{k,j} \, \forall \, j=\{1...K\}$.
 WHAM must have this information binned.
 \end{itemize}
-%
+
 The potential derivative required for TI should generally be calculated during the simulation; only under very rare circumstances~\cite{naden2015linear} can it post-processed by a code that does not evaluate the derivatives.
@@
 Although TI must usually be calculated in code, as it requires the derivative, there is one condition under which it actually has the fastest computation time. 
-If the alchemical path you have chosen is a linear alchemical path, then you get $\frac{dU}{d\vec{\lambda}} = U_0(\vec{x}) - U_1(\vec{x})$, which is the difference between the initial and final states.
+If the alchemical path you have chosen is a linear alchemical path, then you get $\frac{du}{d\vec{\lambda}} = u_0(\vec{q}) - u_1(\vec{q})$, which is the difference between the initial and final states.
 However, because of the problems with linear paths already discussed in this paper, this simplification is rarely that useful.
-%
 
@@
 How decorrelation impacts calculations, and how to compute it is discussed in other section~\ref{sec:decorrelating-samples}.
-%
+
 \subsubsection{Multiple or uncertain binding modes may require considerable care}
@@
 Binding modes also tend to be particularly variable in the case of fragments, which often may have multiple relevant binding modes~\cite{}.
-%
 
@@
 This means that researchers performing absolute binding free energy calculations will have to pay particular attention to generating reasonable putative binding modes.
-%
+
 In some cases, it is tempting to simply use docking techniques to generate initial bound structures for starting molecular dynamics simulations.
@@
 And docking techniques are good at identifying sterically reasonable potential binding modes, but still perform relatively poorly at identifying a single dominant binding mode \emph{a priori}.~\cite{}
-%
 
@@
 While free energy calculations on this set had some success, many of the failures actually ended up being cases where the binding mode selected as input for free energy calculations was later found to be incorrect~\cite{gallicchio2014virtual}, highlighting the importance of these issues.
-%
+
 One approach which has shown some success is to retain diverse potential binding modes from docking, perform short MD simulations of these to identify distinct stable binding modes, and then consider these in subsequent calculations~\cite{gallicchio2014virtual, mobley2006use,rocklin2013blind, boyce2009predicting, mobley2007predicting}.
-%
 
@@
 such as the BLUES approach currently in development~\cite{gill2018binding}), in which case this approach could be applied to assist both types of calculations.
-%
+
 \paragraph{Handling multiple potential binding modes within absolute calculations.}
@@
 This couples to the choice of restraints selected, as some restraints will allow transitions between binding modes and even binding sites (Section~\ref{sec:standardstate-restraints}), and others do not.
-%
+
 Sampling all binding modes within a single free energy calculation is usually impractical without some form of enhanced sampling or at least Hamiltonian replica exchange~\cite{wang2013identifying} because barriers for binding mode interconversion result in kinetics which are too slow compared to simulation timescales~\cite{mobley2006use, palma2012computation,mobley2012perspective, gill2018binding}.
@@
 If relative populations of different stable binding modes were available from some other technique, it could make this separation of states approach considerably more efficient~\cite{mobley2012perspective, gill2018binding}.
-%
+
 \paragraph{Handling multiple potential binding modes within relative calculations.}
 Multiple potential binding modes pose particular problems for relative free energy calculations, as having multiple starting structures for these calculations could yield substantially different calculated relative binding free energies for the same transformation due to kinetic trapping, and, without additional information (specifically, the free energy of binding mode interconversion or, equivalently, the relative populations of different binding modes) it becomes impossible to sort out which of the multiple answers is in fact the correct relative binding free energy~\cite{}.
-%
+
 To deal with this, some practitioners have actually computed relative binding free energies of different binding modes of the same ligand~\cite{palma2012computation}.
@@
 Overall, this likely means that relative free energy calculations will be susceptible to problems resulting from uncertainty in ligand binding modes until more robust approaches are available to determine dominant binding modes, or the relative populations of different potential binding modes, in advance.
-%
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 While a number of different estimators are available that will give consistent results under optimal circumstances, some approaches are recommended over others due to their robustness and ability to provide information on poor convergence.
-%
+
 \subsection{Detecting the boundary between equilibrated and production regions}
@@
 \end{figure}
-%
+
 To minimize this effect, an initial portion of the simulation is often discarded to \emph{equilibration}~\cite{braun2019best}, with the idea of removing the most heavily biased initial portion of simulation data but retaining the unbiased \emph{production} region that represents a stationary Markov chain process sampling from the desired equilibrium target distribution.
@@
 
-A simple approach to automatically partitioning simulation data into equilibration and production regions is described in~\cite{chodera2016simple} (illustrated in Figure~\ref{fig:automatic-equilibration-detection}).
+A simple approach to automatically partitioning simulation data into equilibration and production regions is described in~\cite{chodera2016simple} (illustrated in Fig.~\ref{fig:automatic-equilibration-detection}).
 Suppose we have a simulation of length $T$ consisting of correlated data.
@@
 This approach is implemented within the MBAR~\cite{kylebeauchamp2019choderalab} and alchemlyb~\cite{daviddotson2019alchemistry} packages, and is highly recommended for standard practice.
-%
+
 For additional discussion of working with correlated data and autocorrelation analysis, please refer to the work on Best Practices for Quantification of Uncertainty and Sampling Quality in Molecular Simulations~\cite{grossfield2018best}.
-%
+
 \paragraph{Computing the timeseries for equilibration detection}
@@
 \end{eqnarray}
-%
+
 Note that the use of the effective reduced potential is not guaranteed to pick up on all slow relaxation processes that may be coupled to the alchemical free energy, but the simplicity of its computation means it is generally appropriate for most cases.
-%
+
 \paragraph{Cautions in automating equilibration detection}
@@
 If this is not possible, advanced analysis techniques that assume only local equilibrium (rather than global equilibrium) such as the TRAM estimators~\cite{mey2014xtram,wu2016multiensemble,nuske2017markov} may be more appropriate, but are beyond the scope of this paper.
-%
+
 \subsection{Decorrelating samples for analysis}
@@
 \end{eqnarray}
-The basic concept is that $\tau_\mathrm{eq}$ corresponds to the single-exponential decay time for the autocorrelation process that generates samples, so the statistical inefficiency $g$ measures the approximate temporal separation between two effectively uncorrelated samples (where two exponential relaxation times are presumed to be sufficient).
-%
+The basic concept is that $\tau_\mathrm{eq}$ corresponds to the single- exponential decay time for the autocorrelation process that generates samples, so the statistical inefficiency $g$ measures the approximate temporal separation between two effectively uncorrelated samples (where two exponential relaxation times are presumed to be sufficient).
+
 Robust estimation of $C_t$ for $t \sim T$ is difficult due to growth in statistical error, so common estimators of $g$ make use of several additional properties of $C_t$ to provide useful estimates (see \emph{Practical Computation of Statistical Inefficiencies} in~\cite{chodera2016simple} for a detailed discussion).
@@
 We recommend using the robust statistical inefficiency computation routines available within the MBAR~\cite{kylebeauchamp2019choderalab} and alchemlyb~\cite{daviddotson2019alchemistry} packages.
-%
+
 \paragraph{Subsampling data to generate uncorrelated samples}
-Once the statistical inefficiency $g$ has been estimated, it is straightforward to subsample the correlated timeseries simulation data to produce effectively uncorrelated data that can be fed to estimators of the free energy.
+Once the statistical inefficiency $g$ has been estimated, it is straightforward to subsample the correlated timeseries simulation data to produce effectively uncorrelated data that can be fed to the free energy estimators.
 Suppose the correlated timeseries is $\{a_t\}_{t=1}^T$; we can form a new timeseries of $N_{\mathrm{eff}} \approx T / g$ effectively uncorrelated samples by selecting a subset of indices $\{ \: t = \mathrm{round}((n-1) \, g) \: | \: n \in \mathrm{range}(1,\ldots ,N) \: \}$ where $\mathrm{round(x)}$ denotes rounding to the nearest integer.
-%
+
 If independent simulations are used, the alchemical state $\vec{\lambda}$ may have a significant impact on the correlation time, and these simulations should be subsampled independently using a separate estimate of the statistical inefficinecy $g$ for each alchemical state.
@@
 Conveniently, the approach described above for automated equilibration detection produces an appropriate estimate of $g$ over the production region for automating this process.
-%
+
 \paragraph{Cautions and considerations}
@@
 To ensure there is sufficient data available for reliable decorrelation and estimation of free energy differences, it is recommended that the effective number of uncorrelated samples $N_{\mathrm{eff}} \ge \approx 50$ if the BAR or MBAR estimators (discussed below in sec~\ref{subsec:estimators}) are used; the number may need to be much higher with alternate estimators.
-%
+
 \subsection{Estimators for free energy differences}
@@
 ratio of probabilities of those states.
-As can be noted, the partition functions in Eq.~\ref{equation:dimensionless-free-energy-difference} are simply the total accumulated probabilites for all possible configurations of the system.
+As can be noted, the partition functions in Eq.~\ref{eq:conf_probability} are simply the total accumulated probabilities for all possible configurations of the system.
 Virtually all of the ways to estimate this free energy are based in converting this ratio of integrals to something that can be measured in one (or several) simulations.
-%
+
 \paragraph{The Zwanzig relationship (EXP)}
-%
 The simplest method for calculating free energy differences from simulations is the so-called \textit{Zwanzig
 relationship}~\cite{zwanzig1954hightemperature}, also called one-sided exponential re-weighting (EXP), or simply free energy perturbation, though this final term is sometimes used to encompass all ways of calculating free energy differences.
-%
+
 The (reduced) free energy difference $\Delta f_{01}$ between an initial state 0 and a final state 1 defined by two different potential energy functions 
-$U_0(\vec{q})$ and $U_1(\vec{q})$ over coordinate space $\vec{q}$ can be calculated as:
+$u_0(\vec{q})$ and $u_1(\vec{q})$ over coordinate space $\vec{q}$ can be calculated as:
 \begin{eqnarray}
-\Delta f_{01} & = & \ln \expect{e^{-(U_1(\vec{q}) - u_0(\vec{q}))}}_0 =  \ln \expect{e^{-\Delta u(\vec{q})} }_0
+\Delta f_{01} & = & \ln \expect{e^{-(u_1(\vec{q}) - u_0(\vec{q}))}}_0 =  \ln \expect{e^{-\Delta u(\vec{q})} }_0
 \end{eqnarray}\label{eqn.zwanzig}
 and the average is over all samples from the simulation performed with $u_0$.
-In the case of NVT (canonical) sampling and assuming the masses do not change, then $u$ is just $U/k_BT$, and $f$ is $F/k_BT$, but it can be generalized to other ensembles with the proper definition of $f$ and $u$.
-Described in words, we take the samples generated during our run with the potential energy function $U_0(\vec{q})$, recalculate what the energy difference would be if we switched to potential energy function $U_1(\vec{q})$, and average the exponential of the -1 times the energy difference to get -1 times the exponential of the free energy difference.
-The original distributions (P($U_0$) as generated at $\vec{\lambda}=0$ and P($U_1$) would look like the ones seen in figure~\ref{fig:fig_sampling_scheme} (A)-(C)on the left hand side.
-Reevaluating  requires almost no extra code functionality to perform; one need only to save a full precision trajectory, and run an unmodified molecular simulation code using the $U_1$ in order to calculate the new energies of stored snapshots.
+In the case of NVT (canonical) sampling and assuming the masses do not change, then $u$ is simply $U/k_BT$, and $f$ is $F/k_BT$, but it can be generalized to other ensembles with the proper definition of $f$ and $u$.
+Described in words, we take the samples generated during our run with the potential energy function $u_0(\vec{q})$ and calculate what the difference in energy would be if we switched to potential energy function $u_1(\vec{q})$, and average the exponential of the negative energy difference to get the negative of the exponential of the free energy difference.
+The original distributions, P($u_0$) as generated at $\vec{\lambda}=0$ and P($u_1$) would look like those seen in Fig.~\ref{fig:fig_sampling_scheme} (\textbf{A})-(\textbf{C}) on the left hand side.
+Reevaluating requires almost no extra code functionality to perform; one need only to save a full precision trajectory, and run an unmodified molecular simulation code using the $u_1$ in order to calculate the new energies of stored snapshots.
 The analysis can be written in a line of code.
-We note that this method is even more general, in that the instantaneous work to change the potential energy function from $U_0$ to $U_1$ can be replaced by the non-reversible work $W$ to make the same change under the same equilibrium conditions at either end state~\cite{jarzynski1997nonequilibrium,jarzynski1998equilibrium,crooks2000pathensemble}, though we do not go into all of the details of non-equilibrium transformations here, and refer the reader to more advanced treatments~\cite{maragakis2008bayesian,oberhofer2005biased,procacci2015unbiased,shirts2003equilibriuma,ytreberg2004singleensemble}
-%
+We note that this method is even more general, in that the instantaneous work to change the potential energy function from $u_0$ to $u_1$ can be replaced by the non-reversible work $W$ to make the same change under the same equilibrium conditions at either end state~\cite{jarzynski1997nonequilibrium,jarzynski1998equilibrium,crooks2000pathensemble}, though we do not go into all of the details of non-equilibrium transformations here, and refer the reader to more advanced treatments~\cite{maragakis2008bayesian,oberhofer2005biased,procacci2015unbiased,shirts2003equilibriuma,ytreberg2004singleensemble}.
+
 Although the Zwanzig equation is formally correct (as long as the two states considered sample the same phase space volume, which is true for standard molecular models), it has some very important numerical issues that mean that it often performs badly for standard free energy calculations, even for small molecules~\cite{shirts2005comparison,lu2003appropriate}.
-One can show that if the standard deviation of the difference $\Delta U(\vec{q}) = U_1(vec{q})-U_2(\vec{q})$ over all sampled $\vec{q}$ is large (which in this case, means only several times $k_BT)$, then very few samples contribute to the average, and the answer will be both biased and extremely noisy~\cite{lelievre2010free}.
+One can show that if the standard deviation of the difference $\Delta u(\vec{q}) = u_1(\vec{q})-u_2(\vec{q})$ over all sampled $\vec{q}$ is large (which in this case, means only several times $k_BT)$, then very few samples contribute to the average, and the answer will be both biased and extremely noisy~\cite{lelievre2010free}.
 Essentially, the method is dominated by contributions of rare snapshots\cite{jarzynski2006rare, wu2005phasespaceb, wu2005phasespacec}.
@@
 \paragraph{The Bennett Acceptance Ratio (BAR)}
-If we have the differences in the potential energy sampled from the distribution defined by $U_0$ to the state defined by $U_1$, and we also have the differences in potential energies from the distribution sampled by $U_1$ to the state defined by $U_0$, we can obtain a significantly improved estimate of the
+If we have the differences in the potential energy sampled from the distribution defined by $u_0$ to the state defined by $u_1$, and we also have the differences in potential energies from the distribution sampled by $u_1$ to the state defined by $u_0$, we can obtain a significantly improved estimate of the
 free energy difference compared to that obtained by EXP. 
@@
  ]} \nonumber \\
- =\sum_{i=1}^{n_j} \frac{1}{1 + \exp[\ln(\frac{n_i}{n_j}) - U_{ij}(\vec{q}) + f_ij)]},
+ =\sum_{i=1}^{n_j} \frac{1}{1 + \exp[\ln(\frac{n_i}{n_j}) - u_{ij}(\vec{q}) + f_ij)]},
 \end{eqnarray}
-where $n_i$ and $n_j$ are the number of samples from each state. We
-will refer to this method as BAR.
+where $n_i$ and $n_j$ are the number of samples from each state.
 More recent derivations show that this formula is the maximum likelihood estimate of the free energy difference given sets of samples from the two states~\cite{shirts2003equilibriuma}.
@@
 The Bennett acceptance ratio is only defined between two states.
-Usually, the endpoints of interest in a free energy calculation are sufficiently different that we will need a chain of states that gradually change the potential energy function from $U_0$to $U_1$, as discussed in Section~\ref{sec:important_path}.
-You can simply carry out BAR between each pair of states $\Delta G_{1 \rightarrow N} = \Delta {G_{1\rightarrow 2}} + \Delta {G_{2\rightarrow 3}} +  \ldots + \Delta G_{N-1\rightarrow N}$.
+Usually, the endpoints of interest in a free energy calculation are sufficiently different that we will need a chain of states that gradually change the potential energy function from $u_0$ to $u_1$, as discussed in Section~\ref{sec:important_path}.
+You can simply carry out BAR between each pair of states $\Delta f_{1 \rightarrow N} = \Delta {f_{1\rightarrow 2}} + \Delta {f_{2\rightarrow 3}} +  \ldots + \Delta f_{N-1\rightarrow N}$.
 
 There is one important thing to note about the uncertainty estimates when summing multiple free energies together to calculate an overall free energy estimate.
-Although BAR itself gives a free energy estimate that is asymptotically correct in and is much less biased than the uncertainty estimate for EXP, the uncertainties in $\Delta {G_{i-1\rightarrow i}}$ and $\Delta {G_{i\rightarrow i+1}}$ are not uncorrelated, because they both involve the energies $U_i(\vec{q})$.
+Although BAR itself gives a free energy estimate that is asymptotically correct in and is much less biased than the uncertainty estimate for EXP, the uncertainties in $\Delta {f_{i-1\rightarrow i}}$ and $\Delta {f_{i\rightarrow i+1}}$ are not uncorrelated, because they both involve the energies $u_i(\vec{q})$.
 The variances of each of the free energies will \textit{not} propagate as variances usually do (in quadrature) to the variance of the overall free energy.
@@
 \begin{equation}
-\frac{df}{d\vec{\lambda}} = \frac{d}{d\vec{\lambda}} \left[-\ln \int \frac{\exp{-u(\vec{\lambda},\vec{q})}}{Q(\vec{\lambda})} d\vec{q}\right] = \expect{\frac{du(\vec{\lambda},\vec{q})}{d\vec{\lambda}}\
+\frac{df}{d\vec{\lambda}} = \frac{d}{d\vec{\lambda}} \left[-\ln \int \frac{\exp{-u(\vec{\lambda},\vec{q})}}{Z(\vec{\lambda})} d\vec{q}\right] = \expect{\frac{du(\vec{\lambda},\vec{q})}{d\vec{\lambda}}\
 }_{\vec{\lambda}} 
@@
 For starting researchers, we therefore recommend a simple trapezoid rule scheme, as it allows for maximal flexibility in which values of $\vec{\lambda}$ are simulated.
-Because fitting to higher order polynomials can have numerical instabilities for some energy functions, and because alternate functional forms might only be appropriate with some types of transformations, expertise and experience is required to perform such numerical integration modifications.
+As fitting to higher order polynomials can have numerical instabilities for some energy functions, and because alternate functional forms might only be appropriate with some types of transformations, expertise and experience is required to perform such numerical integration modifications.
 In practice, adding 2-3 more intermediate states is typically sufficient to match the performance of these more complicated numerical quadrature schemes.
@@
 where there are $N_k$ samples from each of $K$ states, with $\sum_k N_k=N$ the total number of samples.
-Thus, we need to evaluate the energy function $U_i$ for all samples obtained at all states in the transformation.
+Thus, we need to evaluate the energy function $u_i$ for all samples obtained at all states in the transformation.
 The equations can be solved by a number of different standard routines.
@@
 
-MBAR is provably the lowest variance asymptotically unbiased estimator of the free energies given the energies of the samples~\cite{tan2004likelihood}, which means that BAR is also for the free energy difference between only 2 states, as it is mathematically exactly the same as MBAR in this case.
+MBAR is probably the lowest variance asymptotically unbiased estimator of the free energy given the energies of the samples~\cite{tan2004likelihood}, which means that BAR is also for the free energy difference between only 2 states, as it is mathematically exactly the same as MBAR in this case.
 MBAR also provides an uncertainty estimate, derived from standard error propagation methods, which has been shown to be accurate as long as there are sufficient samples at each state~\cite{paliwal2011benchmark}.
@@
 \begin{itemize}
-\item We recommend MBAR if all energy differences are available. It is probably the lowest variance free energy estimate given samples from multiple states.
+\item We recommend MBAR if all energy differences are available. It is likely the lowest variance free energy estimate given samples from multiple states.
 \item BAR is essentially just as good as MBAR for highly optimized $\vec{\lambda}$ intermediates.
@@
 It is important to consider the variation in your computed free energies from your equilibrium simulations, in order to  obtain an estimate of uncertainty of the obtained value for the free energies of interest.
-A recent best practices paper by Grossfield et.
-al~\cite{grossfield2018best} provides substantial detail on how to estimate uncertainties from molecular simulations and is a good starting point for this topic.
-In general the quantification of different error metrics depends on both data generation and analysis methods  used from the ones discussed above.
+A recent best practices paper by Grossfield et al,~\cite{grossfield2018best} provides substantial detail on how to estimate uncertainties from molecular simulations and is a good starting point for this topic.
+In general, the quantification of different error metrics depends on both data generation and analysis methods used from the ones discussed above.
 
@@
 The determination of regions of high curvature when estimating the integral is helpful to determine regions of phase space where more sampling and/or more $\vec{\lambda}$ states are necessary to obtain the best approximation of the integral.
-Generally a plot of $\vec{\lambda}$ with respect to the gradients at each of the $\vec{\lambda}$ values can be helpful here.
+Plotting $\vec{\lambda}$ with respect to the gradients at each of the $\vec{\lambda}$ values can be be a helpful diagnostic.
 Additionally, computation of the overall variance of TI requires the calculation of the overall variance of integration, rather than each individual $\Delta$G$_{i,i+1}$ and assuming variances add independently.
@@
 For alchemical changes that result in smooth, low curvature sets of $\expect{\frac{dU}{d\vec{\lambda}}}$, a relatively small number of $\vec{\lambda}$ states is necessary for sufficient accuracy and low variance in the free energy estimate.
-However, with increasing numbers of perturbed atoms, the bias introduced by discretization of the integral can become large due to increased curvature, and more $\vec{\lambda}$ intermediate states become necessary to reduce error.
+Depending on the 'difficulty' of the perturbation, the bias introduced by discretization of the integral can become large due to increased curvature, and more $\vec{\lambda}$ intermediate states become necessary to reduce error.
 It is recommended that researchers verify that a sufficient number of states are included such that the free energy is essentially invariant to the number of lambda intermediate states chosen.
+Good heuristics or measures to assess the 'difficulty' of a given perturbation is still an ongoing research topic.
 
@@
 
-In some cases, several analysis methods can be employed to analyze the same set of simulations, providing an opportunity for synergy.
-Specifically, agreement in the free energy across different estimators can be compared and discrepancy between estimates used to assess convergence.
-For example, differences between free energy estimates computed with BAR, MBAR, and TI correlated with larger errors in a set of MCL1 inhibitors.
-Those perturbations where free energies across different estimators shared high agreement with each other (variance less than 0.1 kcal/mol) had a r$^2$ of 0.623 and mean unsigned error (MUE) of 0.787 kcal/mol (N=17) compared with a r$^2$ of 0.295 and MUE of 1.49 kcal/mol (N=23) (unpublished results).
+If possible, it is advisable to  analyze the same set of simulations with different estimators, providing an opportunity for synergy.
+If different estimators agree the free energy estimate is more reliable than if there are differences between methods that are larger than 1 kcal/mol and would indicate poor convergence.
 
@@
 The assessment of variability in free energy calculations due to repeating simulations has been previously reported~\cite{aldeghi2019accurate,paliwal2011benchmark,mey2016blinded,mey2018impact}, and large variance in free energies estimated from simulations with different random seeds should be flagged as issues with convergence.
-Other sources that may show up as suspicious uncertainties may in fact be system set up errors.
-Common issues include poor ligand or protein preparation, atom mapping for relative calculations, or forcefield parameters as described above.
 
@@
 The proposed changes in configuration are increasingly relevant if no experimental evidence is available to reduce uncertainty in where the changing atoms should be positioned.
-%
+
 In addition to statistical uncertainty and sampling, a variety of other factors can impact results from binding free energy calculations.
-In addition to the choice of initial configuration, results can depend on the choice of force field for the protein/receptor, water, and small molecuel(s), so rerunning calculations with different choices of force field can also be used to assess how sensitive results and conclusions are to these particular choices.
+In addition to the choice of initial configuration, results can depend on the choice of force field for the protein/receptor, water, and small molecule(s), so rerunning calculations with different choices of force field can also be used to assess how sensitive results and conclusions are to these particular choices.
 Other factors, like system preparation (choice of protonation state, tautomer, counterion presence, salt concentration, etc.) can also substantially impact results~\cite{mobley2017predicting, mobley2017predictingb}, so unless modelers are confident they have these factors correct, sensitivity to these choices may also need to be examined.
-%
+
 \subsection{Are my simulations any good?}
@@
 Furthermore, once you have established that individual perturbations are well behaved, there are some tricks to ensure the overall perturbation network gives reliable results.
-%
+
 \paragraph{Convergence of simulations}
@@
 \end{figure}
-%
+
 \paragraph{Overlap matrix}
@@
 These matrices are graphical representations of the phase space overlap, i.e.
-the average probability that a sample generated at state $j$ can be observed at state $i$.
+the average probability that a sample generated at state $\vec{\lambda}_{j}$ can be observed at state $\vec{\lambda_{i}}$.
 As this probability is computed considering the samples from all states (and not just the adjacent states), the values in each row and column add up to 1.
@@
 Smaller values should be considered as a warning sign (see Fig.~\ref{fig:overlap} \textbf{C}), as the variance tends to be underestimated in case of poor overlap.
-%
-\begin{figure}[ht!]
+
+\begin{figure}
 \includegraphics[width=0.90\columnwidth]{figures/fig12_overlap/Figure.pdf}
@@
 \end{figure}
-%
-Fig.~\ref{fig:overlap} \textbf{A}, \textbf{B}, and \textbf{C} shows examples of good, mediocre and poor overlap respectively.
+
+Fig.~\ref{fig:overlap} \textbf{A}, \textbf{B}, and \textbf{C} shows examples of good, mediocre, and poor overlap respectively.
 For Fig.~\ref{fig:overlap}(\textbf{A}), the probability to find a sample from state $i$ in its neighbouring state $j$ is about 0.2 for all states adjacent to the main diagonal, and hence the overall connectivity is good.
@@
 In order to overcome the issue of poor overlap in this example, additional sampling should be performed by introducing additional states, i.e. $\vec{\lambda}$ windows.
-%
+
 Interestingly, as the variance is inversely correlated with the number of states~\cite{klimovich2015guidelines}, it can in principle be reduced below any arbitrary threshold with enough simulation time and a large enough number of $\vec{\lambda}$ windows.
-However, decreasing the variance to a value close to 0 is not feasible, as this approach would strongly increase the calculation time.
-While variance can be decreased both by increasing simulation length and by increasing the number of $\vec{\lambda}$ values, if overlap is poor, the latter approach often has a larger immediate impact.
-Specifically, often the goal is to adjust the $\vec{\lambda}$ spacing to perform additional sampling in the areas of poor overlap, rather than using arbitrarily spaced windows.
+However, decreasing the variance to a value close to 0 is not feasible, as this approach would significantly increase the calculation time.
+While variance can be decreased by increasing simulation length, if the overlap between states is known to be poor, increasing the number of $\vec{\lambda}$ values, or adjusting the spacing of those values to better cover regions of poor overlap will likely provide a larger immediate impact.
 Different approaches are described in section~\ref{sec:simulation_protocol_choice} and more details can be found in the literature~\cite{dakka2018concurrent, hahn2019alchemical}.
-%
-
-There are several other parameters which should be investigated.
-Basic analysis of the trajectories should be performed to ensure that the ligand does not tumble out of the binding site, such as measuring its RMSD to the starting position.
-Furthermore, distributions of the ligand's dihedral angles can be obtained and compared to experimental values.
-Thus, poorly sampled conformations can be identified and the issue can be adequately addressed, e.g.
-by prolonging the simulation time or choosing a different starting conformation.
-In general, running the same perturbation using different ligand poses may offer insights into whether the calculated free energies are dependent on the initial conformation.
-Another easy check can be performed by comparing the results obtained by different free energy estimators (e.g.
-TI and MBAR).
-In principal, the outcomes of both estimators should be the same (within a certain range).
-Hence, if the outcomes deviate by more than a predefined threshold (e.g.
-1.0 kcal/mol), the free energies are likely unreliable and further calculations are necessary.
-%
 
@@
 adding a functional group to a ligand) may provide an additional opportunity for error/consistency checking.
-Particularly, such calculations are often done to span a graph or tree of free energy calculations~\cite{xu2019optimal,wang2013modeling,liu2013lead} and in some cases the free energy change to go between molecules A and B can be obtained via multiple transformation pathways.
+Particularly, such calculations are often done to span a graph or tree of free energy calculations~\cite{xu2019optimal,wang2013modeling,liu2013lead}.
+In some cases the free energy change to go between molecules A and B can be obtained via multiple transformation pathways.
 This allows a type of consistency checking where we assess how much the free energy change for that transformation in practice differs from equivalence.
+
 Significant deviations from this typically indicate insufficient configurational sampling along the lambda schedule of one or more of the transformations involved.
-%
 This approach may be generalised to sets of connected transformations given the requirement that the sum of free energy changes along edges of a closed cycle should be zero.
@@
 The latter approach may yield more reproducible free energy changes, but at the expense of a stronger bias on the estimated free energies due to repeated use of the same input coordinates.
-%
 
@@
 
-%
 \paragraph{Reversible binding simulations}
 An even more stringent test of the correctness of binding free energy calculations is to compare the results to the equilibrium binding constants derived from long timescale reversible binding simulations~\cite{pan2017quantitative}.
-For small ligands with millimolar affinities, repeated binding to and unbinding from the protein can occur for a large number of times in a sufficiently long unbiased MD simulation (10-100$\mu$s), and the equilibrium binding constants can be straightforwardly computed from the ratio of bound to unbound fractions of the simulation time.
-The agreement between the binding free energy calculations and the reversible binding simulations--given the same system preparation and the same force field parameters--will strongly support the correctness of both calculations, as the same results are arrived at by two independent methods, and any discrepancy will suggest some systematic error in one, or both, of the two methods.
+For small ligands with millimolar affinities, repeated binding to and unbinding from the protein can occur for a large number of times in a sufficiently long unbiased MD simulation (10-100 $\mu$s), and the equilibrium binding constants can be  computed from the ratio of bound to unbound fractions of the simulation time.
+The agreement between the binding free energy calculations and the reversible binding simulations---given the same system preparation and the same force field parameters---will strongly support the correctness of both calculations, as the same results are arrived at by two independent methods, and any discrepancy will suggest some systematic error in one, or both, of the two methods.
 As part of validation testing of alchemical free energy codes a benchmark set to compare alchemical and direct computation of equilibrium binding constant should become standard in future.
-%
-\subsection{Recommendations}
-\todo[inline]{ASJSM:Maybe here we have a quick summary on what things should be checked for to make sure simulations are ok?}
-
-\item It is important to carefully examine output data for common problems.  Some of the most important things to check for rae:
-\begin{itemize}
-\item Make sure the ligand samples the binding site reasonably tightly for it's expected potency and fit, and that it does not depart out of binding site in the coupled end state if it is a moderate to strong binder.
-\item Significant discrepancies (further outside the error estimates than would be plausible) between free energies calculated with different free energy estimators such as TI, BAR, and MBAR.
+
+\subsection{Common issues to watch out for during analysis}
+
+It is important to carefully examine output data for common problems.  Some of the most important things to check for are:
+\begin{itemize}
+\item \textbf{Sampling of the binding site by the ligand:} Make sure the ligand samples the binding site reasonably tightly for its expected potency and fit, and that it does not depart out of binding site in the coupled end state if it is a moderate to strong binder.
+\item \textbf{Consistency of free energy estimates across different estimators} Significant discrepancies (further outside the error estimates than would be plausible) between free energies calculated with different free energy estimators such as TI, BAR, and MBAR.
 All of these estimators converge to the same results with sufficient sampling.
 Differences between them indicate poor overlap or errors in processing.
-\item Poor replica mixing (for replica-exchange) or lambda-space sampling for single-replica methods.
+\item \textbf{Have replicas mixed well?} Poor replica mixing (for replica-exchange) or $\lambda$-space sampling for single-replica methods.
 If the system is not mixing between states, then  the states are insufficiently close for mixing, or else there are bottlenecks in the configurational sampling that limit the accuracy.
-\item Correlation time  that does not vary relatively smoothly as a function of $\lambda$.
-Discontinuities indicate that the system is sampling significantly different configurations when the Hamiltonian changes only a bit, which usually indicates sampling problems.
-\item Dependence of the free energy on initial conformation of the system. Clearly, ensemble average properties should not depend on the starting point.
-\item Torsions with multiple low-energy minima where some of these minima are that are visited rarely or not at all.
+\item \textbf{Behaviour of correlation times:} Correlation time  that does not vary relatively smoothly as a function of $\vec{\lambda}$.
+Discontinuities indicate that the system is sampling significantly different configurations with only small changes to the Hamiltonian changes.
+This usually indicates sampling problems.
+\item \textbf{Dependence of the free energy on initial conformation of the system.} Ensemble average properties should not depend on the starting point.
+\item \textbf{Torsional sampling} Torsions with multiple low-energy minima where some of these minima are that are visited rarely or not at all.
 Which torsions have low energy minima can best be found by comparing to the simulation in the solvent.
 There should be clear physical reasons that  simulation in the complex has different torsional distributions that the ligand in the solvent.
-\item The free energy should vary relatively smoothly with $\lambda$.
-If it varies drastically, then either there need to be finer sampling in $\lambda$ in this region, or there are sampling problems there.
-\item The free energy should clearly converge as a function or of simulation time (Figure $\ref{fig:convergence_forward_reverse}$).
+\item \textbf{Free energy dependence on $\vec{\lambda}$} The free energy should vary relatively smoothly with $\vec{\lambda}$.
+If it varies drastically, then either there need to be finer sampling in $\vec{\lambda}$ in this region, or there are sampling problems there.
+\item \textbf{Convergence of free energy} The free energy should clearly converge as a function or of simulation time (Fig. $\ref{fig:convergence_forward_reverse}$).
 \item If using nonequilbirum methods, is the result independent of the speed at which the nonequilibrium change is performed?
@@
 %\item Other Diagnostics for non-equilibrium methods, robustness for switching time etc.?
-\item In general, look at the output data such as energies and see if it matches your expectations.and also visualize the simulation trajectories themselves to see if it matched expectations.
-There are many errors that are caught by only one of the two approaches
-%\item More “usual suspects”
-\end{itemize}
-%\item Other considerations for many transformations
+\item \textbf{Visualisation of data} In general, inspect output data such as energies and visualize the simulation trajectories and assess if they match your expectations.
+Many issues can be spotted by a straight forward visualiztion.
+\end{itemize}
 
@@
 \label{sec:plot_data}
-Carrying out good data generation and proper analysis is not sufficient to finish a study.
+Following best practices for data generation and their analysis does not mean that data is reported in the optimal way.
 As a practitioner of alchemical free energy simulations you also  should use best practices for reporting and plotting your results.
@@
 This replication may seem excessive, but  uncertainty estimates often underestimate the true statistical uncertainty.
-Where performing multiple replicas of the simulation is not possible, lower error bounds such as given by the MBAR estimator should be included.
-
-Because of the recent importance of alchemical free energy predictions in the  drug discovery context, and the relevance of these quantities to structure activity relationship (SAR), both the overall correlation of your predictions relative to experiment and the ability of your models to detect the best-scoring compounds (i.e.
-rankings) should be computed.
-Conventionally, this means including an R\textsuperscript{2} (or Pearson's R: +1 high correlation 0 no correlation -1 high anti-correlation) and a Kendall \texttau{} (with perfect ranking agreement when \texttau=1 and perfect disagreement when \texttau=-1) metric in your results.
+Where performing multiple replicas of the simulation is not possible, an error estimate from e.g.
+MBAR can be used, though bearing in mind this is likely an underestimated error.
+
+As alchemical free energy methods are used in drug discovery to quantify and rationalise structure activity relationships (SAR), the models ability to (a) correlate well with experiment and (b) rank-order the molecules by affinity, should both be computed.
+Conventionally, this means including an R\textsuperscript{2} (or Pearson's R), where $R=+1$ means high correlation, $R=0$ means no correlation, and $R=-1$ means high anti-correlation) and a Kendall \texttau{} (with perfect ranking agreement when \texttau=1 and perfect disagreement when \texttau=-1) metric in your results.
 Additionally, practitioners may choose to include a Spearman \textrho{} as well.
@@
 Other analysis software also provide similar functionality for bootstrapping datasets~\cite{antonia2019michellab}.
-%
+
 Mean unsigned error (MUE, also called mean absolute error/MAE) is another key statistic to include in your results.
@@
 Furthermore, MUE allows for unbiased comparisons between predictive models as it is less sensitive to dataset size.
-Other metrics such as Gaussian Random Affinity Model (GRAM)~\cite{cui2019gram}, Predictive Interval (PI) Relative Absolute Error (RAE), that attempt to correct for the inherent potency range of a dataset, which can aid in comparing success between different targets.
-We recommend further reading on evaluation of computational models \cite{jain2008recommendations, walters2013what, brown2009healthy, walterthoughts}.
+Other metrics such as Gaussian Random Affinity Model (GRAM)~\cite{cui2020gram}, Predictive Interval (PI) and  Relative Absolute Error (RAE), attempt to correct for the inherent potency range of a dataset, which can aid in comparing success between different targets.
+We recommend further reading on evaluation of computational models~\cite{jain2008recommendations, walters2013what, brown2009healthy, walterthoughts}.
 
 Reporting the results of relative free energy calculations requires care.
-As shown in Figure~\ref{fig:5}, relative free energies can be performed arbitrarily as a forward or a reverse process, and thus relative free energies may be reported as either positively or negatively valued.
+As shown in Fig.~\ref{fig:fig_types_of_networks}, relative free energies can be performed arbitrarily as a forward or a reverse process, and thus relative free energies may be reported as either positively or negatively valued.
 The consequence of the two possible signs for relative free energies is that correlation statistics (such as Pearson's R and Kendall \texttau{}) can be skewed depending on which sign is analysed.
-The issue of this inconsistency can be circumvented by either plotting all datapoints within a consistent quadrant\cite{perez-benito2019predicting}, or by avoiding the use of correlation statistics for assessment of relative free energy calculations and instead measuring accuracy using RMSE and MUE which are unaffected by choice of sign.
-%
+The issue of this inconsistency can be circumvented by either plotting all datapoints within a consistent quadrant~\cite{perez-benito2019predicting}, or by avoiding the use of correlation statistics for assessment of relative free energy calculations and instead measuring accuracy using RMSE and MUE which are unaffected by choice of sign.
 
 \paragraph{Presenting your data}
-As essentially all alchemical free energy prediction schemes are regression problems, the preferred type of plot is a scatter plot (see Figure \ref{fig:scatterplot_analysis}).
+As essentially all alchemical free energy prediction schemes are regression problems, the preferred type of plot is a scatter plot (see Fig.~\ref{fig:scatterplot_analysis}).
 Most alchemical free energy projects will look at 10-50 ligands; any study with \textless10 ligands is more suitable for bar plots (with inclusion of error bars), and will unlikely provide meaningful statistics.
-Any study with \textgreater50 ligands should be trellised to not overestimate statistics.
-Because of this, it is bad practice to place multiple datasets on the same plot, because this can suggest high model accuracy even though the individual models perform less well~\cite{walterthoughts}.
-%
-
-Because we are interested mainly in the linear relationship between the alchemical free energy predictions and the experimentally-determined affinity values, plots should be depicted with the same range on both axes (i.e.
+Any study with \textgreater50 ligands often contains multiple protein targets to which alchemical free energies may perform better on some targets than others.
+Because of this, it is bad practice to place multiple datasets on the same plot as this can suggest high model accuracy even though the individual models perform less well~\cite{walterthoughts}.
+
+As we are interested mainly in the linear relationship between the alchemical free energy predictions and the experimentally-determined affinity values, plots should be depicted with the same range on both axes (i.e.
 $x=y$) with a 1:1 aspect ratio, with units for both experiment and simulation converted to be the same.
-If this is skews the plot to a point where it is difficult to read of information, using the same dimensions, such that e.g.
+If this skews the plot to a point where it is difficult to read of information, using the same dimensions, such that e.g.
 1 cm is 1 kcal/mol is acceptable.
@@
 \begin{figure}
-    \includegraphics[width=0.95\linewidth]{figures/fig13_analysis_practices/FigureHBMedit.png}
+    \includegraphics[width=0.95\linewidth]{paper/figures/fig13_analysis_practices/Figure.png}
     \caption{\textbf{An example of recommended practices for graphing alchemical free energy predictions.} This figure shows the relation between predicted and experimentally-determined Gibbs free energy in kcal/mol with standard errors as error bars.
-The dark and light-orange regions depict the 1- and 2-kcal per mol confidence bounds.
+The dark and light-orange regions depict the 1- and 2-kcal/mol confidence bounds.
+Statistical metrics for the data are reported, with 95\% confidence intervals determined by bootstrapping analysis.
 Extra care should be taken when investigating potential outliers further.}
@@
 Alchemical free energy calculations have seen a vast increase in popularity both in academic research as well as  pharmaceutical industry applications in structure based drug discovery~\cite{schindler2020largescale, sherborne2016collaborating, wagner2017computational}.
-Commercial products such as FEP+ and Flare, which provide a convenient user interface make the setup and use of these methods a lot easier~\cite{wang2015accurate, kuhn2020automateda}, but this convenience comes with less flexibility in terms of choice of simulation protocols.
+Commercial products such as FEP+ and Flare, which provide a convenient user interface make the setup and use of these methods a lot easier~\cite{wang2015accurate, kuhn2020assessment}, but this convenience comes with less flexibility in terms of choice of simulation protocols.
 It is also important to understand the current limitations of the methodology to recognise when automated workflow tools can be used effectively for a given protein target and when they are likely to fail still.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Non-exhaustive overview of available software packages}
+\section{Selection of available software packages}
 
@@
 Similar to FEP+ it has an easily accessible graphical user interface and strives to facilitate free energy calculations for non-experts while offering advanced users full control via a Python API.
-It only runs on GPUs, using CUDA or OpenCL~\cite{kuhn2020assessmenta}.
+It only runs on GPUs, using CUDA or OpenCL~\cite{kuhn2020assessment}.
     \item The molecular operating environment (\href{https://www.chemcomp.com/Products.htm}{MOE}) offered by the Chemical Computing Group (CCG) has a tool for performing free energy calculations.
@@
 	\item pymbar: MBAR implementation, but have to roll your own analysis wrapper
-	\url{https://github.com/choderalab/pymbar}~\cite{shirts2008statistically}.
+	\url{https://github.com/choderalab/pymbar}~\cite{shirts2008statisticallya}.
 	\item Beryllium: Standardising alchemical free energy analysis \url{https://github.com/openforcefield/beryllium}
@@
 \begin{table}
-\caption{Overview of benchmark datasets}
+\caption{Selection of example datasets}
 \begin{tabular}{cccc}
@@
 
-\clearpage
-
 \section{Checklist}
@@
         \begin{itemize}
-                \item Standard error (\textsigma) should be \textless1 kcal·mol-1 
+                \item Standard error (\textsigma) should be \textless1 kcal·mol$^{-1}$ 
     \item Simulated systems have not converged 
@@
 %%%%%%%%%%%%%%%%
-\textbf{ASJSM}: Coordinated the document, contributed to most sections, and co-designed figures~\ref{fig:fig_binding_thermodynamic_cycle},~\ref{fig:fig_topology},~\ref{fig:fig_mcss},~\ref{fig:fig_types_of_networks},~\ref{fig:fig_absolute_thermodynamic_cycle},~\ref{fig:scatterplot_analysis},and created figures \ref{fig:fig_what_is_lambda},~\ref{fig:fig1_what_is_alchemy},~\ref{fig:overlap} and replotted~\ref{fig:automatic-equilibration-detection} and improved ~\ref{fig:fig_types_of_networks}.\\
+\textbf{ASJSM}: Coordinated the document, contributed to most sections, and co-designed figures~\ref{fig:fig_binding_thermodynamic_cycle},~\ref{fig:fig_topology},~\ref{fig:fig_mcss},~\ref{fig:fig_types_of_networks},~\ref{fig:fig_absolute_thermodynamic_cycle},~\ref{fig:scatterplot_analysis},and created figures \ref{fig:fig_what_is_lambda},~\ref{fig:fig_what_is_alchemy},~\ref{fig:overlap} and replotted~\ref{fig:automatic-equilibration-detection} and improved ~\ref{fig:fig_types_of_networks}.\\
 \textbf{JS}: created figures \ref{fig:fig_what_is_alchemy},~\ref{fig:fig_binding_thermodynamic_cycle},~\ref{fig:fig_topology},~\ref{fig:fig_mcss},~\ref{fig:fig_types_of_networks},~\ref{fig:fig_absolute_thermodynamic_cycle},~\ref{fig:scatterplot_analysis}, the data representation section, the checklist and contributed to general formatting discussions and editing.\\
@@
 \textbf{BA}: Helped write the uncertainty estimation, stopping conditions, and output analysis sections and created figure ~\ref{fig:convergence_forward_reverse}.\\
-\textbf{JM}: Contributed to sections~\ref{subsec:reproducible},~\ref{sec:prerequisites},~\ref{} 4.2, 5.3, 6, 7.2.3, 8.3.1, 8.3.2, 9\\
+\textbf{JM}: Contributed to sections~\ref{subsec:reproducible},~\ref{sec:prerequisites},4.2, 5.3, 6, 7.2.3, 8.3.1, 8.3.2, 9\\
 \textbf{JDC}: Wrote section~\ref{sec:decorrelating-samples} and~\ref{sec:automatic-equilibration-detection} discussed structure and design of the whole document, suggested Figs.~\ref{fig:fig_what_is_alchemy} and~\ref{fig:fig_sampling_scheme} \\
 \textbf{HX}: Contributed section~\ref{subsec:accuracy}, to section~\ref{sec:relative-fe-protocol}, and to section~\ref{sec:are-they-good}.\\
-\textbf{HBM} Contributed to section \ref{sec:plot_data}and Figure \ref{fig:scatterplot_analysis}.\\
+\textbf{HBM} Contributed to section \ref{sec:plot_data} and figure \ref{fig:scatterplot_analysis} and helped edit the paper.\\
 % We suggest you preserve this comment:
@@
 % We suggest you preserve this comment:
-For a more detailed description of contributions from the community and others, see the GitHub issue tracking and changelog at \githubrepository.
+For a more detailed description of contributions from the community and others, 
+see the GitHub issue tracking and changelog at \githubrepository.
 
@@
 
-\bibliography{alchemical1,manual}
+\bibliography{alchemical,manual}
 

### 2020-07-26 paper/manuscript.tex
last check sec 1,2,3

@@
 %%%%%%%%%%%%%%%%%%%%
-\todo[inline]{}The current Overleaf document is \url{https://www.overleaf.com/5136551543vgjqbkzpmqkp}
+The current Overleaf document is \url{https://www.overleaf.com/5136551543vgjqbkzpmqkp}
 \section{What are alchemical free energy methods?}
@@
 Fig.~\ref{fig:fig_what_is_alchemy} illustrates common free energy changes that may be difficult to compute with unbiased molecular dynamics methods, but are more readily tractable with alchemical methods.
-In alchemical simulations, the introduction of intermediate \textit{alchemical states} that bridge the high-probability regions of configuration space between the two physical endstates of interest permits the free energy of large transformations to be robustly computed.
+In alchemical simulations, the introduction of intermediate \textit{alchemical states} that bridge the high-probability regions of configuration space between two physical endstates of interest, permits the robust computation of free energy for large transformations.
 Alchemical calculations can be used in a variety of scenarios, such as: 
@@
 Subsequent work in the 2000s led to improved implementations of alchemical methods in popular biomolecular simulation packages~\cite{shirts2003extremely,shirts2005solvation,vanderspoel2005gromacs, mermelstein2018fast, wang2015accurate, hedges2019biosimspace}.
- This foundational work, combined with the methodological, technological, and hardware improvements of the last 5--10 years, have led to an explosion of interest and direct commercial application of the technologies~\cite{wang2015accurate, fratev2019improved, schindler2020largescale, cournia2017relative, sherborne2016collaborating}.
+ This foundational work, combined with the methodological, technological, and hardware improvements of the last 5--10 years, has led to an explosion of interest and direct commercial application of these technologies~\cite{wang2015accurate, fratev2019improved, schindler2020largescale, cournia2017relative, sherborne2016collaborating, kuhn2020automated}.
 
@@
 \label{sec:pre}
-What can be expected from this Best Practices guide is a good starting point for new practitioners and a reference for experienced practitioners. 
+What can be expected from this Best Practices guide is: a good starting point for new practitioners and a reference for experienced practitioners. 
  For this propose we provide a convenient checklist (Section~\ref{sec:checklist}) to help ensure all calculations comply with currently-understood best practices for alchemical simulation and analysis.
@@
 
- What we assume from a novice practitioner is a moderate experience with molecular simulation packages and concepts. 
+ What we assume from a novice practitioner is a moderate experience with molecular simulation concepts and use of simulation packages. 
  Furthermore, basic familiarity with the principles of molecular mechanics, molecular dynamics simulations, statistical mechanics, and the biophysics of protein-ligand association are essential.
@@
 As some of the theoretical background can seem daunting, we do, however, provide a guide to the essential theory behind alchemical free energy calculations in Section~\ref{sec:theory}.
-
-In the remainder of this paper, we will cover topics that are key to the preparation~(\ref{sec:prerequisites}), choice and use of correct protocols~(\ref{sec:simulation_protocol_choice}), and finally the best practices that should be used in the analysis of alchemical calculations~(\ref{sec:data_analysis}).
+In the remainder of this paper, we will cover topics that are key to the preparation( Section~\ref{sec:prerequisites}), choice and use of correct protocols (Section~\ref{sec:simulation_protocol_choice}), and finally the best practices that should be used in the analysis of alchemical calculations (Section~\ref{sec:data_analysis}).
 Particular focus will be given to aspects of the molecular simulations which are unique to alchemical calculations---these include the calculation of transfer free energies (hydration free energies, partition coefficients, etc.), and binding free energies (absolute and relative).
@@
 
-For convenience we have also compiled a list of common acronyms, common symbols used, and some essential terminology.
+For convenience we have also compiled a list of common acronyms and common symbols used throughout this paper.
 \begin{tcolorbox}[title=Acronyms, colback=blue!10!white]
-    {\bf CPU} -- Central Processing Unit\\
-     {\bf BAR} -- Bennett Acceptance Ratio\\
-     {\bf FEP} -- Free Energy Perturbation\\
-     {\bf GPCR} -- G-Protein Coupled Receptor\\
-     {\bf GPU} -- Graphics Processing Unit\\
-     {\bf MBAR} -- Multistate Bennett Acceptance Ratio\\
-     {\bf MCSS} -- Maximum Common Substructure\\
-     {\bf MD} -- Molecular Dynamics\\
-     {\bf RMSE} -- Root Mean Square Error\\
-     {\bf MUE} -- Mean unsigned error\\
-     {\bf SAR} -- Structure-Activity Relationships\\
-     {\bf TI} -- Thermodynamic Integration
+    {\bf CPU} --- Central Processing Unit\\
+     {\bf BAR} --- Bennett Acceptance Ratio\\
+     {\bf FEP} --- Free Energy Perturbation\\
+     {\bf GPCR} --- G-Protein Coupled Receptor\\
+     {\bf GPU} --- Graphics Processing Unit\\
+     {\bf MBAR} --- Multistate Bennett Acceptance Ratio\\
+     {\bf MCSS} --- Maximum Common Substructure\\
+     {\bf MD} --- Molecular Dynamics\\
+     {\bf RMSE} --- Root Mean Square Error\\
+     {\bf MUE} --- Mean unsigned error\\
+     {\bf SAR} --- Structure-Activity Relationships\\
+     {\bf TI} --- Thermodynamic Integration
 \end{tcolorbox}
@@
 \begin{tcolorbox}[title=List of Symbols, colback=green!10!white]
-$L$ and $R$ -- generic names for ligand and receptor\\
-$K_b^{\circ}$ -- binding  constant \\
-$c^{\circ}$ -- standard state concentration \\
-$U$ -- potential energy\\
-$u$ -- reduced potential describing a thermodynamic state \\
-$\Delta G$ -- Gibbs free energy (isothermal isobaric ensemble)\\
-$\Delta A$ -- Helmholtz free energy (canonical ensemble)\\
-$\Delta f$-- reduced (dimensionless) free energy \\
-$\Delta \hat{f}$ -- estimate from an estimator for the reduced free energy\\
-$\Gamma$ -- conformation space accessible by simulations \\
-$\vec{q}$ -- vector of a single configuration, i.e. $x$, $y$, $z$ coordinates of the simulation system\\
-$k_B$ -- Boltzmann constant \\
-$Z$ -- partition function \\
-$p$ -- pressure \\
-$\mu$ -- chemical potential (grand canonical ensemble)\\
-$T$ -- temperature \\
-$\beta \equiv (k_B T)^{-1}$ -- inverse thermal energy \\
-$\vec{\lambda}$ -- alchemical progress parameter \\
-$g$ -- statistical inefficiency\\
-$\omega$ - overlap matrix\\
-$C_t$- discrete-time-normalized fluctuation auto-correlation function\\
-$\tau _{eq}$ - integrated auto-correlation time\\
-$t_0$ - equilibration time
+$L$ and $R$ --- generic names for ligand and receptor\\
+$K_b^{\circ}$ --- binding  constant \\
+$c^{\circ}$ --- standard state concentration \\
+$U$ --- potential energy\\
+$u$ --- reduced potential describing a thermodynamic state \\
+$\Delta G$ --- Gibbs free energy (isothermal isobaric ensemble)\\
+$\Delta A$ --- Helmholtz free energy (canonical ensemble)\\
+$\Delta f$ --- reduced (dimensionless) free energy \\
+$\Delta \hat{f}$ --- estimate from an estimator for the reduced free energy\\
+$\Gamma$ --- conformation space accessible by simulations \\
+$\vec{q}$ --- vector of a single configuration, i.e. $x$, $y$, $z$ coordinates of the simulation system\\
+$k_B$ --- Boltzmann constant \\
+$Z$ --- partition function \\
+$p$ --- pressure \\
+$\mu$ --- chemical potential (grand canonical ensemble)\\
+$T$ --- temperature \\
+$\beta \equiv (k_B T)^{-1}$ --- inverse thermal energy \\
+$\vec{\lambda}$ --- alchemical progress parameter \\
+$g$ --- statistical inefficiency\\
+$\mathcal{O}$ --- overlap matrix\\
+$C_t$ --- discrete-time-normalized fluctuation auto-correlation function\\
+$\tau _{eq}$ --- integrated auto-correlation time\\
+$t_0$ --- equilibration time
 \end{tcolorbox}
@@
 \begin{equation}
-R+L \leftrightharpoons RL
+R+L \leftrightharpoons RL.
 \end{equation}
@@
 \end{equation}
-where we have defined the \textit{configurational integral} or \textit{partition function} $Z(\mathrm{state}) \equiv \int_{\Gamma_{\mathrm{state}}} \exp\left(-\beta U(\vec{q})\right) \, d\vec{q}$.
+where we have defined the \textit{configurational integral} or \textit{configurational partition function} as $Z(\mathrm{state}) \equiv \int_{\Gamma_{\mathrm{state}}} \exp\left(-\beta U(\vec{q})\right) \, d\vec{q}$.
 
 \paragraph{Simulating binding events is computationally expensive}
-While simulating binding events has been used to estimate binding affinities~\cite{jong2011determining,pan2017quantitative} or to get insights into the binding pathways and kinetics of receptor-ligand systems~\cite{teo2016adaptive,votapka2017seekr,doerr2014onthefly,plattner2015protein,dixon2018predicting}, the computational cost of these calculations is usually dominated by the rate of dissociation, which can be on the {\textmu}s scale even for millimolar binders~\cite{pan2017quantitative} and reaches the ms or s scale for a typical drug~\cite{basavapathruni2012conformational,hyre2006cooperative}.
+While simulating binding events has been used to estimate binding affinities~\cite{jong2011determining,pan2017quantitative} or to get insights into the binding pathways and kinetics of receptor-ligand systems~\cite{teo2016adaptive,votapka2017seekr,doerr2014onthefly,plattner2015protein,dixon2018predicting}, the computational cost of these calculations is usually dominated by the rate of dissociation, which can be on the {\textmu}s scale even for millimolar binders~\cite{pan2017quantitative} and reaches the micosecond to second scale for a typical drug~\cite{basavapathruni2012conformational,hyre2006cooperative}.
 Depending on system size and simulation settings, common molecular dynamics software packages can reach a few hundreds of ns/day using currently available high-end GPUs~\cite{eastman2017openmm,kutzner2019more}, making these type of calculations uninteresting at pharmaceutical drug discovery timescales.
@@
 
-In many cases, the quantity of interest is the change in binding affinity when a compound $A$ is mutated to compound $B$ (e.g., by modifying one the drug scaffold's substituents, see Fig.~\ref{fig:fig_what_is_alchemy} \textbf{F}), which, by using Eq.~\ref{eq:binding-free-energy-from-bound-unbound-probability-ratio} and \ref{eq:bound-unbound-probability-ratio} is given by
+In many cases, the quantity of interest is the change in binding affinity between a compound $A$ and a related compound $B$ (e.g., by modifying one the drug scaffold's substituents, see (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{F}), which, by using Eq.~\ref{eq:binding-free-energy-from-bound-unbound-probability-ratio} and \ref{eq:bound-unbound-probability-ratio} is given by
 \begin{equation}\label{eq:delta-delta-G-physical}
@@
 
-\todo[inline, color={green!20}]{ASJSM: @JDC, Can you check the issues below? }
 Sometimes free energy calculations can provide answers that are not
@@
 proposed target.
-%
+
 \section{How should alchemical simulations be applied to drug discovery?}
@@
 Without venturing too far into details of system setup, which is beyond the scope of this article, we highlight some critical factors affecting accuracy and successful application.
-%
+
 \subsection{Capturing experimental conditions}
@@
 
-%
 \subsubsection{Absolute free energy calculations must handle the standard state and use restraints}
 \label{sec:standardstate-restraints}
-%
+
 \begin{figure}
@@
 \end{figure}
-%
+
 Absolute free energy calculations involve completely removing the interactions between the ligand or solute and its environment, taking it to a non-interacting state that may or may not retain intramolecular non-bonded interactions.
 This non-interacting state can then be shifted between environments (from the protein to water, or from one solution to another) without changing its free energy, and then interactions can be restored.
-%
+
 Absolute free energies are by definition  reported with respect to a specific reference or standard state, which effectively determines the arbitrary point at which the free energy is 0.
@@
 In practice, a variety of types of restraints are common, from simple harmonic distance restraints between the ligand and the protein~\cite{mobley2006use}, to flat-bottom restraints which work similarly but only exert a force if the ligand leaves a specific region~\cite{chen2007can}.
-%
+
 Alternatively, a set of restraints proposed by Boresch have also commonly been employed, where all six rigid-body degrees of freedom governing the orientation of the ligand relative to the receptor are restrained~\cite{boresch2003absolutea, leitgeb2005alchemicala}.
 Further restraints, such as on the overall ligand RMSD have also been used~\cite{woo2005calculationa}.
-%
+
 In principle, all of these forms will yield correct binding free energies in the limit of adequate sampling (if their effects and connection to the standard state are correctly handled) but they have different strengths and weaknesses.
@@
 See additional discussion of the possibility of multiple binding modes in Sec.~\ref{sec:multiple_binding_modes} below.
-%
+
 Many choices of restraints involve selecting reference atoms.
@@
 Fig.~\ref{fig:fig_sampling_scheme} illustrates the four most common schemes.
-The simplest approach involves running an independent simulation at each of the predefined $\vec{\lambda}$ values (see Fig.~\ref{fig:fig_sampling_scheme}\textbf{A}).
+The simplest approach involves running an independent simulation at each of the predefined $\vec{\lambda}$ values (see Fig.~\ref{fig:fig_sampling_scheme} \textbf{A}).
 This type of scheme is currently used for AMBER TI calculations~\cite{song2019using} and for Siremol/Openmm~\cite{hedges2019biosimspace}.
@@
     \includegraphics[width=0.95\columnwidth]{figures/fig8_sampl_scheme/Figure.pdf}
-    \caption{\textbf{Four most common sampling strategies are available in different popular software packages.} \textbf{A}: Multiple replicas in parallel at different lambda states.
-\textbf{B}: Hamiltonian replica exchange scheme.
-\textbf{C}: Single replica scheme sampling from all $\vec{\lambda}$ states.
-\textbf{D}: Non-equilibrium sampling scheme.}
+    \caption{\textbf{Four most common sampling strategies are available in different popular software packages.} (\textbf{A}): Multiple replicas in parallel at different lambda states.
+(\textbf{B}): Hamiltonian replica exchange scheme.
+(\textbf{C}): Single replica scheme sampling from all $\vec{\lambda}$ states.
+(\textbf{D}): Non-equilibrium sampling scheme.}
     \label{fig:fig_sampling_scheme}
@@
     Each plot shows the estimated free energy change using a varying fraction of total simulation time (up to 5 ns total). 
-    Subplots \textbf{A}, \textbf{C}, and \textbf{E} show a three step protocol  for a perturbation involving 3 perturbed atoms, while \textbf{B}, \textbf{D}, and \textbf{F} shows the same protocol for a perturbation involving 10 perturbed atoms.
+    Subplots (\textbf{A}), (\textbf{C}), and (\textbf{E}) show a three step protocol  for a perturbation involving 3 perturbed atoms, while (\textbf{B}), (\textbf{D}), and (\textbf{F}) shows the same protocol for a perturbation involving 10 perturbed atoms.
 The first step of the protocol is the recharging then removing van der Waals interactions and then recharging.
@@
 free energy difference).
+See for example the potential mean force with respect to $\vec{\lambda}$ estimated from a bound simulation of a Tyk2 ligand pair of Wang et al.~\cite{wang2015accurate}, as seen in Fig.~\ref{fig:pmf}.
+\begin{figure}
+    \centering
+    \includegraphics[width=0.95\linewidth]{paper/figures/fig_pmf/pmf.png}
+    \caption{{\bf Potential of mean force with respect to $\vec{\lambda}$ for TI and MBAR}
+    The estimated PMF for a bound calculation of a Tyk2 ligand pair of the Wang et al.~\cite{wang2015accurate} with respect to $\vec{\lambda}$ estimated from TI and MBAR and showing agreement within errorbars.
+    }
+    \label{fig:pmf}
+\end{figure}
+
 The uncertainty in the free energy, for example, has multiple ways to be estimated, e.g.
@@
 However, timescales for binding mode interconversion are usually slow compared to MD/free energy timescales, meaning that simulations started from different potential binding modes are likely to yield disparate computed binding free energies~\cite{mobley2006use, palma2012computation, mobley2012perspective, gill2018binding} .
-And docking techniques are good at identifying sterically reasonable potential binding modes, but still perform relatively poorly at identifying a single dominant binding mode \emph{a priori}.~\cite{}
+And docking techniques are good at identifying sterically reasonable potential binding modes, but still perform relatively poorly at identifying a single dominant binding mode \emph{a priori}.
 
@@
 \end{eqnarray}
-The basic concept is that $\tau_\mathrm{eq}$ corresponds to the single- exponential decay time for the autocorrelation process that generates samples, so the statistical inefficiency $g$ measures the approximate temporal separation between two effectively uncorrelated samples (where two exponential relaxation times are presumed to be sufficient).
+The basic concept is that $\tau_\mathrm{eq}$ corresponds to the single-exponential decay time for the autocorrelation process that generates samples, so the statistical inefficiency $g$ measures the approximate temporal separation between two effectively uncorrelated samples (where two exponential relaxation times are presumed to be sufficient).
 
@@
 Described in words, we take the samples generated during our run with the potential energy function $u_0(\vec{q})$ and calculate what the difference in energy would be if we switched to potential energy function $u_1(\vec{q})$, and average the exponential of the negative energy difference to get the negative of the exponential of the free energy difference.
-The original distributions, P($u_0$) as generated at $\vec{\lambda}=0$ and P($u_1$) would look like those seen in Fig.~\ref{fig:fig_sampling_scheme} (\textbf{A})-(\textbf{C}) on the left hand side.
+The original distributions, P($u_0$) as generated at $\vec{\lambda}=0$ and P($u_1$) would look like those seen in Fig.~\ref{fig:fig_sampling_scheme} \textbf{A}-\textbf{C} on the left hand side.
 Reevaluating requires almost no extra code functionality to perform; one need only to save a full precision trajectory, and run an unmodified molecular simulation code using the $u_1$ in order to calculate the new energies of stored snapshots.
@@
 It is important to consider the variation in your computed free energies from your equilibrium simulations, in order to  obtain an estimate of uncertainty of the obtained value for the free energies of interest.
-A recent best practices paper by Grossfield et al,~\cite{grossfield2018best} provides substantial detail on how to estimate uncertainties from molecular simulations and is a good starting point for this topic.
+A recent best practices paper by Grossfield et al.,~\cite{grossfield2018best} provides substantial detail on how to estimate uncertainties from molecular simulations and is a good starting point for this topic.
 In general, the quantification of different error metrics depends on both data generation and analysis methods used from the ones discussed above.
@@
 Fig.~\ref{fig:overlap} \textbf{A}, \textbf{B}, and \textbf{C} shows examples of good, mediocre, and poor overlap respectively.
-For Fig.~\ref{fig:overlap}(\textbf{A}), the probability to find a sample from state $i$ in its neighbouring state $j$ is about 0.2 for all states adjacent to the main diagonal, and hence the overall connectivity is good.
-In the case of Fig.~\ref{fig:overlap}(\textbf{B}), the overlap is strongly diminishing in the lower right corner, raising concerns regarding the reliability of the free energy estimate obtained.
-For Fig.~\ref{fig:overlap}(\textbf{C}), the state at $\vec{\lambda}$ index = 6 is connected to neither of its neighbouring states.
+For Fig.~\ref{fig:overlap} \textbf{A}, the probability to find a sample from state $i$ in its neighbouring state $j$ is about 0.2 for all states adjacent to the main diagonal, and hence the overall connectivity is good.
+In the case of Fig.~\ref{fig:overlap} \textbf{B}, the overlap is strongly diminishing in the lower right corner, raising concerns regarding the reliability of the free energy estimate obtained.
+For Fig.~\ref{fig:overlap} \textbf{C}, the state at $\vec{\lambda}$ index = 6 is connected to neither of its neighbouring states.
 While this does not necessarily imply that the result for this perturbation is wrong, the energy estimate must at least be considered as highly unreliable.
@@
 	\item \href{http://getyank.org/latest/index.html}{YANK} is a tool developed by John Chodera and group on the top of OpenMM MD package.
-It allows the users to write their inputs in easy-to-use YAML format\cite{}.
+It allows the users to write their inputs in easy-to-use YAML format.
 	\item \href{http://www.gromacs.org/}{GROMACS} is a molecular simulation package with a significant number of free energy methods implementations.

### 2020-07-27 paper/manuscript.tex

@@
 \item the entire ligand can be alchemically transferred from protein to solvent in an absolute binding free energy calculation (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{E})~\cite{mobley2007predicting,aldeghi2015accurate,aldeghi2017predictions};
-\item and small alchemical modifications can be made between chemically related ligands to estimate relative differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{F})~\cite{wang2015accurate,mey2016blinded,song2019using,gapsys2020large,kuhn2020automated}.
+\item and small alchemical modifications can be made between chemically related ligands to estimate relative differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{F})~\cite{wang2015accurate,mey2016blinded,song2019using,gapsys2020large,kuhn2020assessment}.
 \end{itemize}
@@
 Subsequent work in the 2000s led to improved implementations of alchemical methods in popular biomolecular simulation packages~\cite{shirts2003extremely,shirts2005solvation,vanderspoel2005gromacs, mermelstein2018fast, wang2015accurate, hedges2019biosimspace}.
- This foundational work, combined with the methodological, technological, and hardware improvements of the last 5--10 years, has led to an explosion of interest and direct commercial application of these technologies~\cite{wang2015accurate, fratev2019improved, schindler2020largescale, cournia2017relative, sherborne2016collaborating, kuhn2020automated}.
+ This foundational work, combined with the methodological, technological, and hardware improvements of the last 5--10 years, has led to an explosion of interest and direct commercial application of these technologies~\cite{wang2015accurate, fratev2019improved, schindler2020largescale, cournia2017relative, sherborne2016collaborating, kuhn2020assessment}.
 
@@
 Current alchemical free energy calculations involving small molecules seem to achieve, in favorable cases, root mean square (RMS) errors around 1-2 kcal/mol depending on force field, system, and a variety of other factors such as simulation time, sampling method, and whether the calculations employed are absolute or relative.
-A small selection of benchmark datasets and case studies can be found in section~\ref{sec:benchmark} at the end of this document.
+A small selection of example datasets and case studies can be found in section~\ref{sec:benchmark} at the end of this document.
 However, the domain of applicability is a significant concern~\cite{sherborne2016collaborating, cournia2017relative}, especially for relative calculations, which typically require a high quality and usually experimental bound structure of a closely
@@
 Furthermore, perturbations for the same chemotype in different pockets of the BACE enzyme gave varied errors~\cite{keranen2017acylguanidine}.
-Here the errors refer to the difference in $\Delta G$ after fitting to the mean experimental binding energy.
+Here the errors refer to the difference in $\Delta G$ after doing a linear regression fit to observed experimental binding free energies.
 This prompts us to consider another important aspect.
@@
 In simple cases such as absolute hydration free energies of small organic molecules, or relative hydration free energy calculations between structurally similar small organic molecules, it should be possible to obtain highly precise estimates with a given software package (with a sample standard deviation under 0.01 kcal/mol)~\cite{rizzi2019sampl6}.
-
 For more complex use cases such as protein-ligand binding free energies the repeatability is often substantially worse~\cite{rizzi2019sampl6}.
@@
 For absolute and relative hydration free energies of small organic molecules a variability of ca.
-0.2 kcal/mol between popular simulation packages has been reported, even after fixes to the underlying simulation packages~\cite{loeffler2018reproducibility}.
+0.2 kcal/mol between popular simulation packages has been reported~\cite{loeffler2018reproducibility}.
 In the recent SAMPL6 SAMPLing challenge a larger variability of 0.3 to 1.0 kcal/mol was noted in the computed absolute binding free energies of host/guest systems even though similar input and simulation parameters were used~\cite{rizzi2019sampl6} and, in many cases, single-point energies were identical or nearly so.
-Further work is needed to ensure reproducibility of alchemical free energy calculations across implementations to guarantee that force-field development efforts lead to transferable potential energy functions.
+Further work is needed to ensure reproducibility of alchemical free energy calculations across different software implementations to guarantee that force-field development efforts lead to transferable potential energy functions.
 
@@
 Also, were certain co-factors or partner proteins required in the bioassay?
-%
+
 \subsection{Is my binding mode accurate?}
-As also mentioned, good performance of alchemical FE calculations requires an accurate representation of the ligand binding mode, usually from a high quality X-ray crystal structure.
+As also mentioned, good performance of alchemical calculations requires an accurate representation of the ligand binding mode, usually from a high quality X-ray crystal structure.
 If more than one structure is available, the modeler should pay attention to choose the most suitable.
@@
 It is also useful to study the structure activity relationship and understand the expected impact in the binding site, whether side chain movement in the protein will be required, and whether there is evidence of this in any alternative X-ray structures of the same protein.
-Often, only one protein and water configuration is used for a series of alchemical FE calculations, so this needs to be capable of accommodating the smallest through to largest ligands in a way that allows stable and well behaved simulations.
+Often, only one protein and water configuration is used for a series of alchemical calculations, so this needs to be capable of accommodating the smallest through to largest ligands in a way that allows stable and well behaved simulations.
 This can provide a practical limit on the alchemical changes that are feasible, though a simple work-around can be to separate compounds into sub-series for different calculations.
@@
 Do they clash with some of the larger ligands used in the alchemical perturbation?
-If so, how will the solvent be incorporated for smaller ligands that require the water -- consider pre-solvation methods such as GCMC steps~\cite{michel2010prediction}.
-Other tools such as WaterMap or open source equivalents (SSTmap, others) can be used to define water structure for systems with no experimental evidence of water sites~\cite{wang2011ligand}.
-Before launching large numbers of alchemical free energy calculations it is recommended to test the system using classical MD simulations and limited numbers of alchemical perturbations.
+See Sec.~\ref{subsec:binding}  for different strategies that can be applied to dealing with waters.
+Generally, before launching large numbers of alchemical free energy calculations it is always recommended to test the system using classical MD simulations and limited numbers of alchemical perturbations.
 Metrics such as ligand and protein RMSD and RMSF can be inspected, along with visual inspection of simulations, to ensure the system is stable and likely to be suitable for alchemical calculations.
@@
 Running free energy of binding calculations in a drug discovery application will typically require the use of software or tools to facilitate the large number of calculations.
-The commercial implementation from Schr\"{o}dinger, FEP+, uses a graphical user interface and once the user has calculated any missing force field parameters, calculations can be prepared and submitted within minutes.
-The software builds the network of perturbations, connecting similar compounds based on LOMAP methodology~\cite{liu2013lead}, all inputs for the Desmond simulations are built automatically and the job can be saved or run immediately on available GPUs.
-Commercial tools can be expensive in some cases, but non-comercial tools are becoming more straight forward to use to run alchemical free energy calculations~\cite{gapsys2015pmx, loeffler2015fesetup, song2019using, gapsys2020large,  jespers2019qligfep, hedges2019biosimspace, kuhn2020automated}.
+The commercial implementation such as FEP+, OpenEye Tools, or Flare allow for a fast setup and deployment to GPU hardware in minutes, but will have limited ability to customize calculations~\cite{wang2015accurate,kuhn2020assessment}.
+Commercial tools can be expensive in some cases, but non-comercial tools are becoming more straight forward to use to run alchemical free energy calculations~\cite{gapsys2015pmx, loeffler2015fesetup, song2019using, gapsys2020large,  jespers2019qligfep, hedges2019biosimspace, kuhn2020assessment}.
 
@@
 An interesting cost benefit analysis has shown the value of activity prediction, see discussion above and articles such as~\cite{mobley2012perspective}.
-From a drug discovery point of view, FEP is expanding its domain of applicability, and there are reports of success using homology models and GPCRs for instance, as well as enabling charge change and scaffold hopping, but these systems are undoubtedly more difficult.
-In the meantime, the use cases are expanding to resistance prediction, disease mutation prediction, solubility prediction – an exciting future for alchemical FE calculations.
+From a drug discovery point of view, alchemical calculations are expanding their domain of applicability, and there are reports of success using homology models and GPCRs for instance, as well as enabling charge change and scaffold hopping, but these systems are undoubtedly more difficult.
+In the meantime, the use cases are expanding to resistance prediction, disease mutation prediction, solubility prediction – an exciting future for alchemical calculations.
 
@@
 However, in practice because simulations are of finite duration (typically 1-100 ns per state at present) this is only true for certain classes of alchemical free energy calculations such as relative or absolute free energies of hydration of small and relatively rigid organic molecules.
-Protein-ligand complexes typically exhibit slowly relaxing degrees of freedom that exceed significantly the duration of an alchemical free energy calculation, and host-guest calculations can be susceptible to these issues as well, depending on timescale and system.
+Protein-ligand complexes typically exhibit slowly relaxing degrees of freedom that significantly exceed the duration of an alchemical free energy calculation, and host-guest calculations can be susceptible to these issues as well, depending on timescale and system.
 It is therefore generally important to carefully select input coordinates to obtain satisfactory results.
@@
 For relative calculations, some care has to be taken selecting binding poses for these ligands.
-Generally a common assumption for a congeneric series is that the binding mode is conserved.
+Generally, a common assumption for a congeneric series is that the binding mode is conserved.
 Therefore, if an X-ray structure of one of the ligands is available, this should be used to position the ligands in the putative binding site in an energetically reasonable conformation without steric or electrostatic mismatch with the receptor.
@@
 Overlaying multiple protein X-ray structures can identify conserved or additional water molecules that can be useful to include in calculations.
-In cases where water molecules are known to play an important role in the binding, software implementations that use water sampling facilitated by Grand Canonical Monte Carlo methods may be useful.
-For examples, see HSP90 which formed part of the D3R grand challenge 2015~\cite{mey2016blinded}, A2A~\cite{brucemacdonald2018ligand}, MUP~\cite{ross2015water}, and others~\cite{michel2009energetics}.
+In cases where water molecules are known to play an important role in the binding, software implementations that use water sampling facilitated by Grand Canonical Monte Carlo methods may be useful, i.e.
+consider pre-solvation methods such as GCMC steps~\cite{michel2010prediction}.
+Other tools such as WaterMap or open source equivalents (SSTmap, others) can be used to define water structure for systems with no experimental evidence of water sites~\cite{wang2011ligand}.
+Well known protein systems with water mediated ligand interactions are for example: HSP90 which formed part of the D3R grand challenge 2015~\cite{mey2016blinded}, A2A~\cite{brucemacdonald2018ligand}, MUP~\cite{ross2015water}, and others~\cite{michel2009energetics}.
 
@@
 \paragraph{How to align congeneric series}
-Input coordinates for a congeneric series may be generated by docking calculations, or by ligand alignment using maximum common substructure algorithms.
+Input coordinates for a congeneric series may be generated by docking calculations, or by ligand alignment using maximum common substructure (MCSS) algorithms.
 The latter tends to produce alignments that are more conserved and more consistent free energy changes across a dataset, but will struggle to yield reasonable results for relative binding free energy calculations that involve a significant binding mode rearrangement.
@@
 In this class of alchemical free energy calculations it is necessary to generate a molecular topology that may describe the initial and final states of the perturbation (see Fig.~\ref{fig:fig_topology}).
-In cases where the end-states have high topological similarity and high structural overlap this is relatively straightforward and typically handled by use of maximum common substructure calculations.
+In cases where the end-states have high topological similarity and high structural overlap this is relatively straightforward and typically handled by use of MCSS calculations.
 In situations where the end-state topologies differ significantly, or there is relatively little spatial overlap between the two end-states some user intervention may be necessary to produce a satisfactory input topology.
@@
 In this case a docking program can generate initial structures.
-Different commercial and none commercial tools are available, such as rDock, Autodock Vina, Glide, or Flare, to name a few~\cite{ruiz-carmona2014rdock, trott2010autodock, friesner2004glide, kuhn2020automated}.
+Different commercial and none commercial tools are available, such as rDock, Autodock Vina, Glide, or Flare, to name a few~\cite{ruiz-carmona2014rdock, trott2010autodock, friesner2004glide, kuhn2020assessment}.
 
@@
 \label{sec:simulation_protocol_choice}
-Alchemical free energy calculations can be grouped into two main categories, ``absolute'' (see Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}) and ``relative'' \footnote{The distinction is a bit of a misnomer, since both compute ratios of partition functions relative to another state, and neither computes an absolute free energy.}(see Fig.~\ref{fig:fig_binding_thermodynamic_cycle}), which differ in whether they compute properties for a single molecule (absolute) or compare properties of different, usually closely related, molecules (relative).
+Alchemical free energy calculations can be grouped into two main categories, ``absolute'' (see Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}) and ``relative''\footnote{The distinction is a bit of a misnomer, since both compute ratios of partition functions relative to another state, and neither computes an absolute free energy.} (see Fig.~\ref{fig:fig_binding_thermodynamic_cycle}), which differ in whether they compute properties for a single molecule (absolute) or compare properties of different, usually closely related, molecules (relative).
 To use binding as a concrete example, in absolute binding free energy calculations, we compute the binding free energy of a ligand to an individual receptor relative to a standard reference concentration.
@@
 Hybrid topology calculations have seen much less use but essentially consist of two absolute free energy calculations in opposite directions at the same time (turning one molecule off while turning the other on)~\cite{jiang2019computing}.
-At present, the most widely used approaches, such as in Schr\"{o}dinger's FEP+~\cite{wang2015accurate,wang2019protein} and in BioSimSpace~\cite{hedges2019biosimspace} (for which calculations may be planned with Lead Optimization Mapper (LOMAP)~\cite{liu2013lead} seem to use single topology approaches, though some codes only support dual topology.
-To our knowledge efficiency differences have not been thoroughly explored, though conventional wisdom suggests that fewer dummy atoms are better~\cite{liu2013lead,mobley2012perspective}.
-%
+Various software implementations currently mostly rely on single or dual topology approaches; for example AMBER TI uses a dual topology approach, while Gromacs and BioSimSpace use single topology approaches.
+Please make sure to check what approach is used with your software package of choice.
+To our knowledge efficiency differences have not been thoroughly explored, though conventional wisdom suggests that fewer dummy atoms are better~\cite{liu2013lead, mobley2012perspective}.
+
 \paragraph{Atom mapping}
@@
 MCSS searches can be relatively time consuming, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial assessment and then use MCSS only to identify final mappings for relative calculations~\cite{raymond2002maximum,klabunde2012mars,jones2009elucidating}.
-
 The MCSS approach, though relatively standard, takes into account only topological similarity.
@@
 Single topology relative calculations, and calculations based on substructure searches, only work if in fact the ligands share a common substructure, e.g.
-are part of a congeneric series, see Fig.~\ref{fig:fig_mcss}
+are part of a congeneric series, see Fig.~\ref{fig:fig_mcss}.
 If no common substructure is shared, then alternative dual or hybrid topology free energy calculations are needed, where one would co-localize a pair of compounds in a binding site, exclude their interactions with one another, and compute the relative binding free energy by turning one molecule on from being dummy atoms while turning the other off.
@@
 
-\paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss} \textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings.
+\paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss} \textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
 Some approaches have attempted to address this~\cite{clark2019relative} but a general solution is not yet in mainstream use.
-(at end of first sentence of this pgraph I would also cite this work from Liu, Wang and Mobley, https://pubs.acs.org/doi/abs/10.1021/acs.jcim.5b00057)
 
@@
 Based on the input ligand series a perturbation map or network can be planned.
-Recent heuristics have shown the more connected the perturbation network the better, however, there is a way to optimise network structure while minimising the number of perturbations that need to be computed reducing the resulting computational cost~\cite{yang2020optimal}.
-Sometimes the introduction of intermediates that are not part of the original congeneric series are essential to avoid ring breaking, or dealing with perturbations that would otherwise result in large numbers of atoms to be grown or disappearing.
+Recent heuristics have shown the more connected the perturbation network the better, however, there is a way to optimize network structure while minimizing the number of perturbations that need to be computed reducing the resulting computational cost~\cite{yang2020optimal}.
+Sometimes the introduction of intermediates that are not part of the original congeneric series are essential to avoid ring breaking, or deal with perturbations that would otherwise result in large numbers of atoms to be grown or disappearing.
 Some commercial tools have good underlying heuristics but may fail with complicated input, needing user validation in particular when dealing with chiral compounds.
+
 \begin{figure}
@@
 \end{figure}
-In some cases, during the lead optimization stage, or for very large datasets that would benefit from rougher initial FE ranking, or in cases where perturbations would be rather large a star shaped network as seen in Fig.~\ref{fig:fig_types_of_networks} \textbf{A} is used.
+
+In some cases, during the lead optimization stage, or for very large datasets that would benefit from rougher initial free evnergy ranking, or in cases where perturbations would be rather large a star shaped network as seen in Fig.~\ref{fig:fig_types_of_networks} \textbf{A} is used.
 However, adding redundancy into the network means that a better error analysis can be carried out, by looking at cycle closure errors as discussed in sec.~\ref{sec:are-they-good}, with an example given in Fig.~\ref{fig:fig_types_of_networks} \textbf{B}.
@@
 FEP+ Dataset~\cite{gapsys2020large} & 8 & 199 & various \\
-FEP+ Dataset~\cite{kuhn2020automated} & 8 & 199 & GAFF2.1 \\
+FEP+ Dataset~\cite{kuhn2020assessment} & 8 & 199 & GAFF2.1 \\
 Fragments ~\cite{steinbrecher2015accurate} & 8 & 96 & OPLS2.1 \\
 Scaffold Hopping~\cite{wang2017accurate} & 6 & 21 & OPLS3 \\
-Scaffold Hopping~\cite{kuhn2020automated} & 6 & 21 & GAFF2.1 \\
+Scaffold Hopping~\cite{kuhn2020assessment} & 6 & 21 & GAFF2.1 \\
 Macrocycles~\cite{yu2017accurate} & 7 & 33 & OPLS3 \\

### 2020-07-27 paper/manuscript.tex
stylistic fixes

@@
 \begin{figure}
-    \includegraphics[width=0.95\linewidth]{figures/fig5_thermo_cyc_abs/Figure.pdf}
+    \includegraphics[width=0.95\linewidth]{figures/fig5_thermo_cyc_abs/Figure_1.pdf}
     \caption{\textbf{Thermodynamic cycle required for an absolute free energy calculation -- absolute free energy of binding example} The fully interacting ligand in water (\textbf{A}), has its charges turned off to pass to (\textbf{B}) followed by turning of van der Waals terms, resulting in a non-interacting ligand in water in (\textbf{C}).
@@
 
-The key consideration for choosing alchemical pathways is that the intermediate states that a given pathway produces should sample configurational ensembles that change as slowly as possible as $\vec{\lambda}$ changes, while still managing to go from the initial state to the final state as $\vec{\lambda}$ goes from 0 to 1.
+The key consideration for choosing alchemical pathways is that the intermediate states that a given pathway produces, should sample configurational ensembles that change as slowly as possible as $\vec{\lambda}$ changes, while still managing to go from the initial state to the final state as $\vec{\lambda}$ goes from 0 to 1.
 
@@
     U(\vec{r_{ij}},\vec{\lambda}) = 4\epsilon_{ij} \vec{\lambda} \left(\frac{1}{(\alpha(1-\vec{\lambda}) + (r_{ij}/\sigma_{ij})^6)^2} -  \frac{1}{\alpha(1-\vec{\lambda}) + (r_{ij}/\sigma_{ij})^6}\right)
-    \label{eq:softcore}
-\end{equation}
-
-Where $r_{ij}$ is the distance between two particles $i$ and $j$, $\epsilon_{ij}$ and $\sigma_{ij}$ are the Lennard-Jones parameters corresponding to the interaction between particles $i$ and $j$, and $\alpha$ is a constant (0.5 is optimal for the functional form shown above).
+    \label{eq:softcore},
+\end{equation}
+
+where $r_{ij}$ is the distance between two particles $i$ and $j$, $\epsilon_{ij}$ and $\sigma_{ij}$ are the Lennard-Jones parameters corresponding to the interaction between particles $i$ and $j$, and $\alpha$ is a constant (0.5 is optimal for the functional form shown above).
 This functional form has exactly the property we are looking for: it recovers the Lennard-Jones potential when $\vec{\lambda}=1$, and the other endpoint ($\vec{\lambda}=0$), it is exactly zero for all $r_{ij}$ everywhere, and as $\vec{\lambda}$ goes to zero, the $\alpha(1-\vec{\lambda})$ term lowers the infinite energy in the core.
@@
 Other schemes, such as simultaneously changing electrostatic and Lennard-Jones interactions with electrostatic ``soft-core'' potentials~\cite{steinbrecher2007nonlinear} may be implemented with fewer intermediate but could require fine-tuning of electrostatic and Lennard-Jones softcore parameters to avoid numerical instabilities.
-At the time of writting, there has not been conclusive evidence to suggest one approach is better in general than the other, so discretion should be left up to the user as to what is viable from both hardware resources, and what the simulation software supports.
+At the time of writing, there has not been conclusive evidence to suggest one approach is better in general than the other, so discretion should be left up to the user as to what is viable from both hardware resources, and what the simulation software supports.
 
@@
 Some schemes to adaptively optimize the spacing of intermediate states based on initial exploratory simulations have been proposed~\cite{hayes2017adaptive}.
-For molecules changing in dense solvent, then the best path is roughly independent of molecule size and shape, so what works for one molecular transformation is likely to be relatively efficient for another~\cite{monroe2004converging}
+For molecules changing in dense solvent, then the best path is roughly independent of molecule size and shape, so what works for one molecular transformation is likely to be relatively efficient for another~\cite{monroe2004converging}.
 
@@
 The simplest approach involves running an independent simulation at each of the predefined $\vec{\lambda}$ values (see Fig.~\ref{fig:fig_sampling_scheme} \textbf{A}).
-This type of scheme is currently used for AMBER TI calculations~\cite{song2019using} and for Siremol/Openmm~\cite{hedges2019biosimspace}.
+This type of scheme is currently used for AMBER TI calculations~\cite{song2019using} and for Sire as implemented in BioSimSpace~\cite{hedges2019biosimspace}.
 However, if these simulations can be run simultaneously with communication between them, a simple extension allows mixing between these replicas.
@@
 In this scheme a single replica rapidly explores all of $\vec{\lambda}$ space by working out optimal weights that allow switching between different intermediate $\vec{\lambda}$ values, as seen in Fig.~\ref{fig:fig_sampling_scheme} \textbf{C} .
-This approach is also referred to as self adjusted mixture sampling~\cite{lyubartsev1992newa, li2007simulated, tan2017optimally} and while promising, has so far only been supported in OpenMM Tools~\cite{andrearizzi2019choderalab}, and to date has not seen published applications for binding free energy calculations.
+This approach is also referred to as self adjusted mixture sampling~\cite{lyubartsev1992newa, li2007simulated, tan2017optimally} and while promising, has so far only been supported in OpenMM Tools~\cite{andrearizzi2019choderalab}.
 The last approach makes use of non-equilibrium simulations~\cite{aldeghi2018accurate}.
-In this approach, only end state $\vec{\lambda}$ replicas ($\vec{\lambda}$=0, $\vec{\lambda}=1$) are simulated at equilibrium; intermediate information is generated from non-equilibrium simulations that rapidly transition between endstates.
+In this approach, only end state $\vec{\lambda}$ replicas ($\vec{\lambda}$=0, $\vec{\lambda}=1$) are simulated at equilibrium; intermediate information is generated from non-equilibrium simulations that rapidly transition between end-states.
 This approach is available in GROMACS and appears to be coming online in several other packages.
@@
 \begin{figure}
-    \includegraphics[width=0.95\columnwidth]{figures/fig8_sampl_scheme/Figure.pdf}
-    \caption{\textbf{Four most common sampling strategies are available in different popular software packages.} (\textbf{A}): Multiple replicas in parallel at different lambda states.
+    \includegraphics[width=0.88\columnwidth]{figures/fig8_sampl_scheme/Figure.pdf}
+    \caption{\textbf{Four most common sampling strategies.} (\textbf{A}): Multiple replicas in parallel at different lambda states.
+Each arrow symbolises an independent $\vec{\lambda}$ simulation.
 (\textbf{B}): Hamiltonian replica exchange scheme.
+Each arrow represents a short simulation interval before an exchange through Metropolis Hastings acceptance (dice) is attempted.
+A tick means an accepted exchange a cross a rejected exchange.
 (\textbf{C}): Single replica scheme sampling from all $\vec{\lambda}$ states.
-(\textbf{D}): Non-equilibrium sampling scheme.}
+After a short simulation time symbolised by the arrow, the lambda-state is attempted to change until all N lambda states will be sampled.
+(\textbf{D}): Non-equilibrium sampling scheme, where two equilibrium simulations at the end-states are run as indicated by the blue and pink arrow.
+Non-equilibrium simulations are attempted at intervals to switch between the two end-states.}
     \label{fig:fig_sampling_scheme}
@@
     \includegraphics[width=0.95\linewidth]{figures/fig10_forward_reverse/Figure.pdf}
-    \caption{\textbf{Free energy (in k$_{B}$T) for two different relative binding free energy (RBFE) perturbations.} 
+    \caption{\textbf{Free energy (in k$_{B}$T) for two different relative binding free energy perturbations.} 
     Each plot shows the estimated free energy change using a varying fraction of total simulation time (up to 5 ns total). 
     Subplots (\textbf{A}), (\textbf{C}), and (\textbf{E}) show a three step protocol for a perturbation involving 3 perturbed atoms, while (\textbf{B}), (\textbf{D}), and (\textbf{F}) shows the same protocol for a perturbation involving 10 perturbed atoms.
-The first step of the protocol is the recharging then removing van der Waals interactions and then recharging.
+The first step of the protocol is the decharging then removing van der Waals interactions and then recharging.
 The difference in energy between the forward (blue) and reverse (red) free energy calculations at the midpoint of the simulation time gives an indication of the overall convergence of the simulation, with differences over 1 k$_{B}$T indicating poor convergence.}
@@
 However, even subtly induced changes in protein binding configuration will require more sampling or cause local convergence to a free energy estimate that has high error.
-Many enhanced sampling techniques have been proposed to try and overcome barriers in free energy calculations and mitigate convergence to erroneous estimates~\cite{}MISSING citation. 
+Many enhanced sampling techniques have been proposed to try and overcome barriers in free energy calculations and mitigate convergence to erroneous estimates~\cite{xx}. 
 The confidence a user should have in a free energy estimate is significantly improved when both the uncertainty of the free energy estimate is low, and when other observables have reached a convergence.
 
-Another aspect that can be monitored is bias in the simulation.
-A simple, but effective, way of checking for bias is to compute the same property of interest by multiple estimators (e.g.
-free energy difference).
-See for example the potential mean force with respect to $\vec{\lambda}$ estimated from a bound simulation of a Tyk2 ligand pair of Wang et al.~\cite{wang2015accurate}, as seen in Fig.~\ref{fig:pmf}.
 \begin{figure}
@@
 As such, it is a good idea to subject different estimators to the same data to see if they yield either the same estimate (within error and bias), or if they fluctuate wildly.
+See for example the potential of mean force with respect to $\vec{\lambda}$ estimated from a bound simulation of a Tyk2 ligand pair of Wang et al.~\cite{wang2015accurate} for both the MBAR and TI estimators, as seen in Fig.~\ref{fig:pmf}.
 This is not a perfect method as some estimators, such as exponential averaging, will converge significantly more slowly, relative to more accurate estimators like MBAR. 
@@
 If, for instance, the free energy estimator selected is thermodynamic integration, then values of $\frac{dU}{d\vec{\lambda}}$ at uncorrelated data points must be collected.
-Once a combination of knowing what type of simulation you will run, which alchemical topology you will simulate, what alchemical path you will simulation along, and what your stopping conditions are, then you are ready to enumerate the information you should capture.
+Once a combination of knowing what type of simulation you will run, which alchemical topology you will simulate, what alchemical path you will simulate along, and what your stopping conditions are, then you are ready to enumerate the information you should capture.
 Below is a sample of the minimal information you need for a set of common estimators (discussed in more detail in Sec.~\ref{subsec:estimators}):
@@
 One may safely assume that the correlation time is greater than 100-200 fs even for relatively simple systems such as small molecules in solvent, so saving no more frequently than every 50-100 steps is recommended.
-How decorrelation impacts calculations, and how to compute it is discussed in other section~\ref{sec:decorrelating-samples}.
+How decorrelation impacts calculations, and how to compute it is discussed in Sec.~\ref{sec:decorrelating-samples}.
 
@@
 To deal with this, some practitioners have actually computed relative binding free energies of different binding modes of the same ligand~\cite{palma2012computation}.
-For example, a perturbation which adds a methyl to an aromatic ring of a larger ligand might yield one result if the methyl points in one direction, and a different value if it points in the other due to slow ring motions~\cite{lincoff2016comparing}.
+For example, a perturbation which adds a methyl to an aromatic ring of a larger ligand might yield one result if the methyl points in one direction, and a different value if it points in the other due to slow ring motions~\cite{lincoff2016comparing, sasmal2020sampling}.
 One could compute the free energy of turning off the methyl group in one orientation and turning it back on in the other orientation to obtain the free energy difference between the two potential binding modes.
@@
 \paragraph{Computing the timeseries for equilibration detection}
-Typically, the timeseries of note $a_t$ analyzed in automated equilibration detection is the negative logarithm of the probability density sampled by the MCMC algorithm (up to an irrelevant additive constant).
+Typically, the timeseries of note $a_t$ analyzed in automated equilibration detection is the negative logarithm of the probability density ($\pi(x_t; \vec{\lambda}$)) sampled by the MCMC algorithm (up to an irrelevant additive constant).
 For simple independent simulations that sample $x_t \sim \pi(x ; \vec{\lambda})$, this is given by the reduced potential
@@
 \frac{df}{d\vec{\lambda}} = \frac{d}{d\vec{\lambda}} \left[-\ln \int \frac{\exp{-u(\vec{\lambda},\vec{q})}}{Z(\vec{\lambda})} d\vec{q}\right] = \expect{\frac{du(\vec{\lambda},\vec{q})}{d\vec{\lambda}}\
-}_{\vec{\lambda}} 
+}_{\vec{\lambda}} .
 \end{equation}
@@
 \begin{equation}
-\Delta f    = \int_{0}^{1} \expect{\frac{du(\vec{\lambda},\vec{q})}{d\vec{\lambda}}}_{\vec{\lambda}}  d\vec{\lambda}    
+\Delta f    = \int_{0}^{1} \expect{\frac{du(\vec{\lambda},\vec{q})}{d\vec{\lambda}}}_{\vec{\lambda}}  d\vec{\lambda}.    
 \end{equation}
@@
 \[ \Delta f \approx \sum_{k=1}^{K} w_k
-\expect{\frac{du(\vec{\lambda},\vec{q})}{d\vec{\lambda}}}_{k} \] where the weights $w_k$ correspond to a particular choice of numerical integration.
+\expect{\frac{du(\vec{\lambda},\vec{q})}{d\vec{\lambda}}}_{k} \], where the weights $w_k$ correspond to a particular choice of numerical integration.
 Researchers have tried a large number of different integration schemes~\cite{resat1993studies,jorge2010effect,shyu2009reducing}.
@@
 \begin{equation}
-f_i = - \ln \sum_{n=1}^{N} \frac{\exp(-u_i(\vec{q}_n))}{\sum_{k=1}^K N_k \exp(f_k-u_k(\vec{q}_n))}
+f_i = - \ln \sum_{n=1}^{N} \frac{\exp(-u_i(\vec{q}_n))}{\sum_{k=1}^K N_k \exp(f_k-u_k(\vec{q}_n))},
 \end{equation}
@@
 If it varies drastically, then either there need to be finer sampling in $\vec{\lambda}$ in this region, or there are sampling problems there.
-\item \textbf{Convergence of free energy} The free energy should clearly converge as a function or of simulation time (Fig. $\ref{fig:convergence_forward_reverse}$).
+\item \textbf{Convergence of free energy} The free energy should clearly converge as a function or of simulation time (Fig. ~\ref{fig:convergence_forward_reverse}).
 \item If using nonequilbirum methods, is the result independent of the speed at which the nonequilibrium change is performed?
@@
 %\item Other Diagnostics for non-equilibrium methods, robustness for switching time etc.?
-\item \textbf{Visualisation of data} In general, inspect output data such as energies and visualize the simulation trajectories and assess if they match your expectations.
+\item \textbf{Visualization of data} In general, inspect output data such as energies and visualize the simulation trajectories and assess if they match your expectations.
 Many issues can be spotted by a straight forward visualiztion.
@@
 \label{sec:checklist}
-\begin{Checklists*}[p!]
+\begin{Checklists*}
 \begin{checklist}{ Know what you want to simulate}
@@
 \end{checklist}
-
+\end{Checklists*}
+
+%Analyis checklist
+\begin{Checklists*}
 \begin{checklist}{How do I know which simulations are not reliable}

### 2020-07-27 paper/manuscript.tex
fixing references and contributions

@@
 %%%%%%%%%%%%%%%%
-\textbf{ASJSM}: Coordinated the document, contributed to most sections, and co-designed figures~\ref{fig:fig_binding_thermodynamic_cycle},~\ref{fig:fig_topology},~\ref{fig:fig_mcss},~\ref{fig:fig_types_of_networks},~\ref{fig:fig_absolute_thermodynamic_cycle},~\ref{fig:scatterplot_analysis},and created figures \ref{fig:fig_what_is_lambda},~\ref{fig:fig_what_is_alchemy},~\ref{fig:overlap} and replotted~\ref{fig:automatic-equilibration-detection} and improved ~\ref{fig:fig_types_of_networks}.\\
-\textbf{JS}: created figures \ref{fig:fig_what_is_alchemy},~\ref{fig:fig_binding_thermodynamic_cycle},~\ref{fig:fig_topology},~\ref{fig:fig_mcss},~\ref{fig:fig_types_of_networks},~\ref{fig:fig_absolute_thermodynamic_cycle},~\ref{fig:scatterplot_analysis}, the data representation section, the checklist and contributed to general formatting discussions and editing.\\
-\textbf{MK}: Contributed to the Data Analysis section, provided the data for figure~\ref{fig:overlap} and helped edit the paper.\\
+\textbf{ASJSM}: Coordinated the document, contributed to most sections, and co-designed Figs.~\ref{fig:fig_binding_thermodynamic_cycle},~\ref{fig:fig_topology},~\ref{fig:fig_mcss},~\ref{fig:fig_types_of_networks},~\ref{fig:fig_absolute_thermodynamic_cycle},~\ref{fig:scatterplot_analysis},and created Figs.~\ref{fig:fig_what_is_lambda},~\ref{fig:fig_what_is_alchemy},~\ref{fig:overlap},~\ref[fig:pmf} and replotted~\ref{fig:automatic-equilibration-detection} and~\ref{fig:fig_types_of_networks}.\\
+\textbf{JS}: Created Figs~ \ref{fig:fig_what_is_alchemy},~\ref{fig:fig_binding_thermodynamic_cycle},~\ref{fig:fig_topology},~\ref{fig:fig_mcss},~\ref{fig:fig_absolute_thermodynamic_cycle},~\ref{fig:scatterplot_analysis}, and an initial draft of~\ref{fig:fig_types_of_networks}.
+Wrote Sec.~\ref{sec:plot_data}, the checklist Sec.~\ref{sec:checklist}, and contributed to general formatting discussions and editing.\\
+\textbf{MK}: Contributed to Sec.~\ref{sec:data_analysis}, provided the data for figure~\ref{fig:overlap}, compiled the dataset for Sec.~\ref{sec:benchmark} and helped edit the paper.\\
 \textbf{DLM}: Contributed to the outline, drafted some of the sections, gave ideas on figures, and helped edit the paper.\\
-\textbf{GT}: Contributed to introduction and drug discovery sections, and helped edit the paper.\\
-\textbf{AR}: Created figure~\ref{fig:freeenergytrajectories}, contributed to sections~\ref{sec:theory} and \ref{sec:simulation_protocol_choice}, and helped edit the paper.\\
-\textbf{MRS}: Helped create figure~\ref{fig:fig_what_is_lambda}, wrote sections describing choices for alchemical pathways and analysis for free energy calculations.
+\textbf{GT}: Contributed to Sec.~\ref{sec:intro} and~\ref{sec:drugdiscovery}, and helped edit the paper.\\
+\textbf{AR}: Created figure~\ref{fig:freeenergytrajectories}, contributed to sections~\ref{sec:theory} and~\ref{sec:simulation_protocol_choice}, and helped edit the paper.\\
+\textbf{MRS}: Helped create figure~\ref{fig:fig_what_is_lambda}, wrote Sec.~\ref{sec:important_path} describing choices for alchemical pathways and parts of~\ref{sec:data_analysis} on the analysis for free energy calculations.
 Reviewed and edited text throughout.\\
@@
 \textbf{BA}: Helped write the uncertainty estimation, stopping conditions, and output analysis sections and created figure ~\ref{fig:convergence_forward_reverse}.\\
-\textbf{JM}: Contributed to sections~\ref{subsec:reproducible},~\ref{sec:prerequisites},4.2, 5.3, 6, 7.2.3, 8.3.1, 8.3.2, 9\\
-\textbf{JDC}: Wrote section~\ref{sec:decorrelating-samples} and~\ref{sec:automatic-equilibration-detection} discussed structure and design of the whole document, suggested Figs.~\ref{fig:fig_what_is_alchemy} and~\ref{fig:fig_sampling_scheme} \\
-\textbf{HX}: Contributed section~\ref{subsec:accuracy}, to section~\ref{sec:relative-fe-protocol}, and to section~\ref{sec:are-they-good}.\\
-\textbf{HBM} Contributed to section \ref{sec:plot_data} and figure \ref{fig:scatterplot_analysis} and helped edit the paper.\\
+\textbf{JM}: Contributed to Sec.~\ref{subsec:reproducible},~\ref{sec:prerequisites},~\ref{sec:important_path},~\ref{subsec:estimators},~\ref{subsec:uncertainty}, and~\ref{sec:conclusion}\\
+\textbf{JDC}: Wrote Sec.~\ref{sec:decorrelating-samples} and~\ref{sec:automatic-equilibration-detection} discussed structure and design of the whole document, suggested Figs.~\ref{fig:fig_what_is_alchemy} and~\ref{fig:fig_sampling_scheme} \\
+\textbf{HX}: Contributed Sec.~\ref{subsec:accuracy}, to Sec.~\ref{sec:relative-fe-protocol}, and to Sec.~\ref{sec:are-they-good}.\\
+\textbf{HBM} Contributed to Sec.~\ref{sec:plot_data} and Fig.~ \ref{fig:scatterplot_analysis} and helped edit the paper.\\
 % We suggest you preserve this comment:

### 2020-07-27 paper/manuscript.tex
last changes

@@
 \begin{document}
-%
+
 \begin{frontmatter}
@@
 
-As we are interested mainly in the linear relationship between the alchemical free energy predictions and the experimentally-determined affinity values, plots should be depicted with the same range on both axes (i.e.
-$x=y$) with a 1:1 aspect ratio, with units for both experiment and simulation converted to be the same.
-If this skews the plot to a point where it is difficult to read of information, using the same dimensions, such that e.g.
-1 cm is 1 kcal/mol is acceptable.
-Furthermore, bounds should be depicted for the 1- and 2-kcal/mol confidence regions.
-These regions can serve as tools to communicate your model performance: any predictions inside the 1 kcal/mol region can be seen as highly reliable, any predictions inside the 2 kcal/mol region should be seen as somewhat reliable, and any predictions outside the confidence regions should be expected to be unreliable (and handled as outliers).
-In a drug discovery context, this type of data depiction may suggest the reliability of alchemical FE predictions in the project, and can give an idea of how trustworthy predictions can be for synthesis ideas.
-
 \begin{figure}
@@
 \end{figure}
+
+As we are interested mainly in the linear relationship between the alchemical free energy predictions and the experimentally-determined affinity values, plots should be depicted with the same range on both axes (i.e.
+$x=y$) with a 1:1 aspect ratio, with units for both experiment and simulation converted to be the same.
+If this skews the plot to a point where it is difficult to read of information, using the same dimensions, such that e.g.
+1 cm is 1 kcal/mol is acceptable.
+Furthermore, bounds should be depicted for the 1- and 2-kcal/mol confidence regions.
+These regions can serve as tools to communicate your model performance: any predictions inside the 1 kcal/mol region can be seen as highly reliable, any predictions inside the 2 kcal/mol region should be seen as somewhat reliable, and any predictions outside the confidence regions should be expected to be unreliable (and handled as outliers).
+In a drug discovery context, this type of data depiction may suggest the reliability of alchemical FE predictions in the project, and can give an idea of how trustworthy predictions can be for synthesis ideas.
 
@@
 Please refere to ~\ref{tab:benchmarks}, for a small overview of datasets, what forcefields they used, and what the original study was it came from.
+
 \begin{table}
@@
 %%%%%%%%%%%%%%%%
-\textbf{ASJSM}: Coordinated the document, contributed to most sections, and co-designed Figs.~\ref{fig:fig_binding_thermodynamic_cycle},~\ref{fig:fig_topology},~\ref{fig:fig_mcss},~\ref{fig:fig_types_of_networks},~\ref{fig:fig_absolute_thermodynamic_cycle},~\ref{fig:scatterplot_analysis},and created Figs.~\ref{fig:fig_what_is_lambda},~\ref{fig:fig_what_is_alchemy},~\ref{fig:overlap},~\ref[fig:pmf} and replotted~\ref{fig:automatic-equilibration-detection} and~\ref{fig:fig_types_of_networks}.\\
+\textbf{ASJSM}: Coordinated the document, contributed to most sections, and co-designed Figs.~\ref{fig:fig_binding_thermodynamic_cycle},~\ref{fig:fig_topology},~\ref{fig:fig_mcss},~\ref{fig:fig_types_of_networks},~\ref{fig:fig_absolute_thermodynamic_cycle},~\ref{fig:scatterplot_analysis},and created Figs.~\ref{fig:fig_what_is_lambda},~\ref{fig:fig_what_is_alchemy},~\ref{fig:overlap},~\ref{fig:pmf} and replotted~\ref{fig:automatic-equilibration-detection} and~\ref{fig:fig_types_of_networks}.\\
 \textbf{JS}: Created Figs~ \ref{fig:fig_what_is_alchemy},~\ref{fig:fig_binding_thermodynamic_cycle},~\ref{fig:fig_topology},~\ref{fig:fig_mcss},~\ref{fig:fig_absolute_thermodynamic_cycle},~\ref{fig:scatterplot_analysis}, and an initial draft of~\ref{fig:fig_types_of_networks}.

### 2020-08-04 paper/manuscript.tex
last changes

@@
 Alchemical free energy calculations are a useful tool for predicting free energy differences associated with the transfer of molecules from one environment to another.
-The hallmark of these methods is the use of "bridging" potential energy functions representing \emph{alchemical} intermediate states that cannot exist in as real chemical species.
+The hallmark of these methods is the use of "bridging" potential energy functions representing \emph{alchemical} intermediate states that cannot exist as real chemical species.
 The data collected from these bridging alchemical thermodynamic states allows the efficient computation of transfer free energies (or differences in transfer free energies) with orders of magnitude less simulation time than simulating the transfer process directly.
@@
 
-Fig.~\ref{fig:fig_what_is_alchemy} illustrates common free energy changes that may be difficult to compute with unbiased molecular dynamics methods, but are more readily tractable with alchemical methods.
+Fig.~\ref{fig:fig_what_is_alchemy} illustrates common free energy changes that may be difficult to compute with unbiased molecular dynamics methods, but are more tractable with alchemical methods.
 In alchemical simulations, the introduction of intermediate \textit{alchemical states} that bridge the high-probability regions of configuration space between two physical endstates of interest, permits the robust computation of free energy for large transformations.
@@
 \item computing partition ($\log P$) or distribution ($\log D$) coefficients between environments (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{B})~\cite{rustenburg2016measuring, bosisio2016blinded} 
-\item or between compartments or membranes (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{C})~\cite{corey2019insights}. 
+\item determining partitioning between compartments into membranes (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{C})~\cite{corey2019insights}. 
 \end{itemize}
@@
 \item the entire ligand can be alchemically transferred from protein to solvent in an absolute binding free energy calculation (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{E})~\cite{mobley2007predicting,aldeghi2015accurate,aldeghi2017predictions};
-\item and small alchemical modifications can be made between chemically related ligands to estimate relative differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{F})~\cite{wang2015accurate,mey2016blinded,song2019using,gapsys2020large,kuhn2020assessment}.
-\end{itemize}
-
-After an alchemical calculation is performed (which generally involves multiple simulations at a variety of alchemical states), the data must be analyzed to compute an estimate the free energy for the transformation of interest.
- Early work used simple but statistically suboptimal estimators for this: free energy perturbation (FEP) used a simple (but highly biased) estimator based on the Zwanzig relation~\cite{zwanzig1954hightemperature} or numerical quadrature via thermodynamic integration (TI), for which the theory dates back decades but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamicsa, merz1989free}.
+\item small alchemical modifications can be made between chemically related ligands to estimate relative differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{F})~\cite{wang2015accurate,mey2016blinded,song2019using,gapsys2020large,kuhn2020assessment}.
+\end{itemize}
+
+After an alchemical calculation is performed (which generally involves multiple simulations at a variety of alchemical states), the data must be analyzed to compute an estimate of the free energy for the transformation of interest.
+ Early work used simple but statistically suboptimal estimators for this: free energy perturbation (FEP) used a simple (but highly biased) estimator based on the Zwanzig relation~\cite{zwanzig1954hightemperature} or numerical quadrature via thermodynamic integration (TI), for which the theory dates back the better part of a century but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamicsa, merz1989free}.
 %
@@
 
-As the field of molecular simulation can now routinely access microsecond timescales with the aid of GPUs~\cite{salomon-ferrer2013routine} and millisecond timescales appear to soon be within reach, accurate alchemical calculations on even more challenging problems will become reasonable to perform.
+As the field of molecular simulation can now routinely access microsecond timescales with the aid of GPUs~\cite{salomon-ferrer2013routine}, and millisecond timescales appear to soon be within reach, accurate alchemical calculations on even more challenging problems will become reasonable to perform.
 In the meantime, today's users may find it difficult to get started with these complex calculations whilst also keeping up with the fast pace of change. 
-This Best Practices guide provides current recommendations and tips for users of all experience.
+This Best Practices guide provides current recommendations and tips for users of all experience. Updates and suggestions are welcomed via our GitHub repository.
 
@@
 \label{sec:pre}
-What can be expected from this Best Practices guide is: a good starting point for new practitioners and a reference for experienced practitioners. 
+This Best Practices guide focuses on providing a good starting point for new practitioners and a reference for experienced practitioners. 
  For this propose we provide a convenient checklist (Section~\ref{sec:checklist}) to help ensure all calculations comply with currently-understood best practices for alchemical simulation and analysis.
-Where the best practices are currently unsettled, we highlight areas where further research is needed to identify an unambiguous recommendation.
+Where the best practices are currently not certain, we highlight areas where further research is needed to identify an unambiguous recommendation.
  It can also serve as a set of best practices to ensure simulation robustness and reproducibility which reviewers may wish to consider as they evaluate papers.
 
- What we assume from a novice practitioner is a moderate experience with molecular simulation concepts and use of simulation packages. 
+ We assume that novice practitioners have at least moderate experience with molecular simulation concepts and use of simulation packages. 
  Furthermore, basic familiarity with the principles of molecular mechanics, molecular dynamics simulations, statistical mechanics, and the biophysics of protein-ligand association are essential.
@@
  While reading this Best Practices guide, it is important to bear in mind \emph{this is not a review} of all free energy calculation methods at the cutting edge of current research.
-Instead the guide aims to answer the following questions:
+Instead this guide aims to answer the following questions:
 \begin{itemize}
@@
 Some other background information may be needed depending on the nature of the alchemical project.
-For example, often, if binding poses are not known, docking calculations can be used to generate an initial small molecule binding pose to start alchemical calculations.
-This will require some basic familiarity on how to perform docking calculations to generate reasonable simulation starting points~\cite{grinter2014challenges}.
+For example, often, if binding poses are not known, docking calculations can be used to generate an initial small molecule binding pose to start alchemical simulations.
+This will require some basic familiarity on how to perform docking to generate reasonable simulation starting points~\cite{grinter2014challenges}.
 
@@
 $\beta \equiv (k_B T)^{-1}$ --- inverse thermal energy \\
-$\vec{\lambda}$ --- alchemical progress parameter \\
+$\vec{\lambda}$ --- alchemical progress parameter, which may be multidimensional \\
 $g$ --- statistical inefficiency\\
@@
 Why would you want to run an alchemical free energy calculation and why do they work?
-In this section, we use the example of relative free energy calculation to sketch the theory of alchemical calculations and illustrate their utility.
+In this section, we use the example of relative free energy calculations to sketch the theory of alchemical simulations and illustrate their utility.
 The emphasis here is placed on bridging theoretical foundations and intuition.
@@
 \paragraph{The free energy of binding can be expressed as a ratio of partition functions}
-A natural though generally very computational expensive way to estimate the equilibrium constant is by directly simulating several binding and unbinding events and computing the probability of finding the receptor-ligand system in the bound state, $P(RL)$, or the unbound state, $P(R+L)$.
+A natural, though generally very computationally expensive, way to estimate the equilibrium constant is by directly simulating several binding and unbinding events and computing the probability of finding the receptor-ligand system in the bound state, $P(RL)$, or the unbound state, $P(R+L)$.
 Assuming the volume change upon binding to be negligible, which is often the case at $1~\mathrm{atm}$ due to the incompressibility of water, then the Gibbs free energy $\Delta G_{\mathrm{bind}, L}$ is approximately equal to the Helmholtz free energy $\Delta A_{\mathrm{bind}, L}$, and we can simulate the system in a box of volume $V$ to obtain~\cite{jong2011determining}
@@
 \paragraph{Simulating binding events is computationally expensive}
-While simulating binding events has been used to estimate binding affinities~\cite{jong2011determining,pan2017quantitative} or to get insights into the binding pathways and kinetics of receptor-ligand systems~\cite{teo2016adaptive,votapka2017seekr,doerr2014onthefly,plattner2015protein,dixon2018predicting}, the computational cost of these calculations is usually dominated by the rate of dissociation, which can be on the {\textmu}s scale even for millimolar binders~\cite{pan2017quantitative} and reaches the micosecond to second scale for a typical drug~\cite{basavapathruni2012conformational,hyre2006cooperative}.
-Depending on system size and simulation settings, common molecular dynamics software packages can reach a few hundreds of ns/day using currently available high-end GPUs~\cite{eastman2017openmm,kutzner2019more}, making these type of calculations uninteresting at pharmaceutical drug discovery timescales.
+While simulating binding events has been used to estimate binding affinities~\cite{jong2011determining,pan2017quantitative} or to get insights into the binding pathways and kinetics of receptor-ligand systems~\cite{teo2016adaptive,votapka2017seekr,doerr2014onthefly,plattner2015protein,dixon2018predicting}, the computational cost of these calculations is usually dominated by the rate of dissociation, which can be on the microsecond timescale even for millimolar binders~\cite{pan2017quantitative} and reaches the millisecond to second timescale for a typical drug~\cite{basavapathruni2012conformational,hyre2006cooperative}.
+Depending on system size and simulation settings, common molecular dynamics software packages can reach a few hundreds of ns/day using currently available high-end GPUs~\cite{eastman2017openmm,kutzner2019more}, making these type of calculations unappealing and irrelevant on a pharmaceutical drug discovery timescale.
 Other methods compute the free energy of binding by building potential of mean force profiles along a reaction coordinate~\cite{woo2005calculationa,velez-vega2013overcoming,limongelli2013funnel,heinzelmann2017attachpullrelease}, but these methods require prior knowledge of a high-probability binding pathway, which is not easily available, especially in the prospective scenarios typical of the drug development process.
@@
     \includegraphics[width=0.95\linewidth]{figures/fig2_therm_cyc/Figure.pdf}
-    \caption{{\bf Thermodynamic cycle for computing the relative free energy of binding ($\Delta \Delta G$) between two related small molecules to a supramolecular host.}
+    \caption{{\bf Thermodynamic cycle for computing the relative free energy of binding ($\Delta \Delta G$) between two related small molecules to a supramolecular host or a rigid receptor.}
     The relative binding free energy difference between two small molecules, $\Delta \Delta G_{\mathrm{bind}, A \rightarrow B} \equiv \Delta G_{\mathrm{bind}, B} - \Delta G_{\mathrm{bind}, A}$---here benzyl alcohol (top) to toluene (bottom)---can be computed as a difference between two alchemical transformations, $\Delta G_\mathrm{bound} - \Delta G_\mathrm{solvated}$, where $\Delta G_\mathrm{bound}$ represents the free energy change of transforming $A \rightarrow B$ in complex, i.e.
@@
 
-In practice, the mutation of $A$ to $B$ in carried out by introducing one or more parameters $\vec{\lambda}$ controlling the potential energy function $U(\vec{q};\vec{\lambda})$ such that the potential of compounds $A$ and $B$ is recovered at two particular values $\vec{\lambda}_A$ and $\vec{\lambda}_B$.
-Briefly, this is achieved by simulating a ``chimera'' molecule composed of enough atoms to represent both $A$ and $B$.
+In practice, the mutation of $A$ to $B$ is carried out by introducing one or more parameters $\vec{\lambda}$ controlling the potential energy function $U(\vec{q};\vec{\lambda})$ such that the potential of compounds $A$ and $B$ is recovered at two particular values $\vec{\lambda}_A$ and $\vec{\lambda}_B$.
+Briefly, this is achieved by simulating a ``chimeric'' molecule composed of enough atoms to represent both $A$ and $B$.
 A subset of the energetic terms in $U(\vec{q};\vec{\lambda})$ is then modulated by $\vec{\lambda}$ so that at $\vec{\lambda}_A$, the atoms that form molecule $A$ are activated and those belonging exclusively to $B$ are non-interacting ``dummy atoms'', while the opposite occur at $\vec{\lambda}_B$ (see Section~\ref{sec:relative-fe-protocol} for details).
@@
 While it is theoretically possible to estimate the ratio of partition functions from samples collected only at states $\vec{\lambda}_A$ and $\vec{\lambda}_B$, the efficiency of the free energy estimators rapidly decreases as the phase-space overlap between the two states also decreases~\cite{wu2005phasespaceb,wu2005phasespacec}.
-Roughly, the phase-space overlap between two thermodynamic states measures the degree to which high-probability conformations (i.e., those with very negative potential energy) in one state are also high-probability conformations in the other state(see Section~\ref{sec:are-they-good} and Fig.~\ref{fig:fig_what_is_lambda}).
+Roughly, the phase-space overlap between two thermodynamic states measures the degree to which high-probability conformations (i.e., those with very negative potential energy) in one state are also high-probability conformations in the other state (see Section~\ref{sec:are-they-good} and Fig.~\ref{fig:fig_what_is_lambda}).
 
@@
 \end{equation}
-While this strategy usually results in sampling thermodynamic states whose Boltzmann distributions is very similar, thus collecting information that is to some degree redundant, some estimators such as the Multistate Bennett acceptance ratio (MBAR)~\cite{shirts2008statisticallya} can exploit similarities between states to improve the precision of the estimates by using the conformations sampled at all alchemical states $\{\vec{\lambda}_k \}$ to compute the free energy difference $\Delta f(\vec{\lambda}_i, \vec{\lambda}_{j})$ between any pair of states $i,j$ (see Section~\ref{subsec:estimators}).
+While this strategy usually results in sampling thermodynamic states whose Boltzmann distributions are very similar, thus collecting information that is to some degree redundant, some estimators such as the Multistate Bennett acceptance ratio (MBAR)~\cite{shirts2008statisticallya} can exploit similarities between states to improve the precision of the estimates.
+This is achieved by using the conformations sampled at all alchemical states $\{\vec{\lambda}_k \}$ to compute the free energy difference $\Delta f(\vec{\lambda}_i, \vec{\lambda}_{j})$ between any pair of states $i,j$ (see Section~\ref{subsec:estimators}).
 
@@
 While absolute and relative free energy calculations have subtle differences in their practical applications (e.g., use of restraints, handling of the standard state), the fundamental ideas and concepts of relative free energy approaches remain unaltered in other types of alchemical calculations.
-Absolute binding, hydration, and partition free energy still use thermodynamic cycles that enable computing transfer free energies without actually simulating the physical transfer from one environment to another.
+Absolute binding, hydration, and partition free energies still use thermodynamic cycles that enable computing transfer free energies without actually simulating the physical transfer from one environment to another.
 
@@
 Current alchemical free energy calculations involving small molecules seem to achieve, in favorable cases, root mean square (RMS) errors around 1-2 kcal/mol depending on force field, system, and a variety of other factors such as simulation time, sampling method, and whether the calculations employed are absolute or relative.
-A small selection of example datasets and case studies can be found in section~\ref{sec:benchmark} at the end of this document.
+A small selection of example datasets and case studies can be found in Section~\ref{sec:benchmark} at the end of this document.
 However, the domain of applicability is a significant concern~\cite{sherborne2016collaborating, cournia2017relative}, especially for relative calculations, which typically require a high quality and usually experimental bound structure of a closely
@@
 It is important to be clear on what error to report: $\Delta G$ after shifting by a constant to minimize the RMSE, unshifted $\Delta G$, $\Delta \Delta G$ of computed edges, or $\Delta \Delta G$ of all edges.
-(See also the section recommendations for best practices of reporting data Sec.~\ref{sec:plot_data}) In general, the computed edge $\Delta \Delta G$ RMSEs significantly underestimates the error of all edge RMSEs: \url{https://github.com/jchodera/jacs-dataset-analysis}.
-Given the need to understand the performance of the system with alchemical free energy calculations, we recommend that retrospective studies for a particular target and a particular chemical series be performed for each particular application case.
+(See recommendations for reporting best practices, Sec.~\ref{sec:plot_data}) In general, the computed edge $\Delta \Delta G$ RMSEs significantly underestimates the error of all edge RMSEs: \url{https://github.com/jchodera/jacs-dataset-analysis}.
+Given the need to understand the performance of the system with alchemical free energy calculations, we recommend that retrospective studies for a particular target and a particular chemical series be performed for each application case.
 
@@
 For more complex use cases such as protein-ligand binding free energies the repeatability is often substantially worse~\cite{rizzi2019sampl6}.
-A good practice is to perform two or three runs of the same perturbation to assess repeatability with a given protocol.
+A good practice is to perform two or three runs of the same perturbation to assess reproducibility with a given protocol.
 The sample standard deviation will give a crude estimate of the reliability of the estimates, and whether the precision is sufficient for the problem at hand.
-Where this is practical, a more stringent test is to use different input coordinates for each repeat run.
+When practical, a more stringent test is to use different input coordinates for each repeat run.
 
@@
 0.2 kcal/mol between popular simulation packages has been reported~\cite{loeffler2018reproducibility}.
-In the recent SAMPL6 SAMPLing challenge a larger variability of 0.3 to 1.0 kcal/mol was noted in the computed absolute binding free energies of host/guest systems even though similar input and simulation parameters were used~\cite{rizzi2019sampl6} and, in many cases, single-point energies were identical or nearly so.
+In the recent SAMPL6 SAMPLing challenge a larger variability of 0.3 to 1.0 kcal/mol was noted in the computed absolute binding free energies of host/guest systems even though the study sought to use identical input and simulation parameters~\cite{rizzi2019sampl6} and, in many cases, single-point energies were identical or nearly so.
 Further work is needed to ensure reproducibility of alchemical free energy calculations across different software implementations to guarantee that force-field development efforts lead to transferable potential energy functions.
@@
 different kinases in the DFG-out conformations, or to different
-stability of the DFG-out conformations of different kinases. Let
+stability of the DFG-out conformations of different kinases.
+
+Let
 $K_C$ be the equilibrium constant between DFG-in and DFG-out
@@
 \label{subsec:accuracy}
-The requisite level of accuracy is another important consideration ---if the
+The requisite level of accuracy is another important consideration. If the
 goal is to guide lead optimization when many compounds will be
@@
 microsecond events or longer, the computational cost of running 20-40
-microsecond or longer simulations for each of $N$ ligands may become
-prohibitive, or at least should be carefully considered. Alternatively, if key motions are fast and minimal, only much shorter simulations may be necessary.
+microsecond or longer simulations for each of $N$ ligands will likely be
+prohibitive for most users with today's hardware. Alternatively, if key motions are fast and minimal (as is often assumed in practice), only much shorter simulations may be necessary.
 
@@
 
-If you are unsure whether your project is feasible, as mentioned above one recommended option is to conduct a short exploratory study to assess tractability for a small
+If you are unsure whether your project is feasible, as mentioned above, one recommended option is to conduct a short exploratory study to assess tractability for a small
 number of ligands. This can be sufficient to get an initial
@@
 Many practitioners expect alchemical methods to provide valuable guidance for drug discovery, and to exhibit accuracy superior to most alternative approaches for suitable targets~\cite{kuhn2017prospective}.
-Successful application in industry may require considerable knowledge of the ``domain of applicabilit'' of free energy calculations -- where they work well and where they will not~\cite{sherborne2016collaborating}.
+Successful application in industry may require considerable knowledge of the ``domain of applicability'' of free energy calculations -- where they work well and where they will not~\cite{sherborne2016collaborating}.
 Successful application also requires robust protocols for preparing, submitting and analysing alchemical calculations.
@@
 Therefore, it is necessary to consider details of the experimental setup, such as pH.
-Biological assays are usually run at neutral pH but this is not always the case, and protein and ligand preparation protocols should reflect experimental pH.
+Biological assays are usually run at neutral pH but this is not always the case.
+For example, some enzymes exhibit pH-dependent activity and assays may thus be done in conditions other from neutral pH.
+Therefore, computational protein and ligand preparation protocols should reflect experimental pH.
 
@@
 For instance, does the X-ray structure that is to be used for the calculations match the construct used for screening (i.e.
-catalytic vs full length, monomer vs dimer etc)~\cite{perez-benito2018predicting}?
+only the catalytic domain vs.
+full length, monomer vs.
+dimer, etc.)~\cite{perez-benito2018predicting}?
 Also, were certain co-factors or partner proteins required in the bioassay?
@@
 
-It is also useful to study the structure activity relationship and understand the expected impact in the binding site, whether side chain movement in the protein will be required, and whether there is evidence of this in any alternative X-ray structures of the same protein.
+It is also useful to study the structure activity relationship and understand the expected impact of any mutations on the binding site, such as whether side chain movement in the protein will be required, and whether there is evidence of this in any alternative X-ray structures of the same protein.
 Often, only one protein and water configuration is used for a series of alchemical calculations, so this needs to be capable of accommodating the smallest through to largest ligands in a way that allows stable and well behaved simulations.
@@
 In simple terms, providing poor alignment to these tools will make this job harder.
-Therefore, docking ligands with no restraints is not usually the best method for defining the ligand inputs, instead a core restrained docking can provide satisfactory outcomes, but in this case one still should pay careful attention to consistency of alignment for identical substituents.
+Docking with restraints is often beneficial in this regard.
+Particularly, fixing the 3D spatial position of the scaffold using maximal common substructure (MCSS) restrained docking can help provide well aligned input for the topology generation.
+Nevertheless, in this case careful attention is still needed to ensure consistency of alignment for identical substituents.
 Another alternative is to manually edit the same core and add/modify the changing substituents.
-This provides assurances that coordinates for the non-perturbed portion of the structure remain identical and aromatic substituents for instance have consistent dihedral angles but is not feasible for many compounds and therefore automation can be desirable.
+This provides assurances that coordinates for the non-perturbed portion of the structure remain identical and aromatic substituents, for instance, have consistent dihedral angles.
+However, it is not feasible for many compounds and therefore automation is desirable.
 
@@
 
-Running free energy of binding calculations in a drug discovery application will typically require the use of software or tools to facilitate the large number of calculations.
-The commercial implementation such as FEP+, OpenEye Tools, or Flare allow for a fast setup and deployment to GPU hardware in minutes, but will have limited ability to customize calculations~\cite{wang2015accurate,kuhn2020assessment}.
-Commercial tools can be expensive in some cases, but non-comercial tools are becoming more straight forward to use to run alchemical free energy calculations~\cite{gapsys2015pmx, loeffler2015fesetup, song2019using, gapsys2020large, jespers2019qligfep, hedges2019biosimspace, kuhn2020assessment}.
+Running binding free energy calculations in a drug discovery application will typically require the use of software or tools to facilitate the large number of calculations.
+Commercial implementations such as FEP+, OpenEye Tools, or Flare allow for a fast setup and deployment to GPU hardware in minutes, but may have limited ability to customize calculations~\cite{wang2015accurate,kuhn2020assessment}.
+Commercial tools can be expensive in some cases, but non-commercial tools are becoming more straight forward to use to run alchemical free energy calculations~\cite{gapsys2015pmx, loeffler2015fesetup, song2019using, gapsys2020large, jespers2019qligfep, hedges2019biosimspace, kuhn2020assessment}.
 
@@
 For example error estimates provided by free energy estimators that are too large can highlight poorly converged simulations~\cite{perez-benito2019predicting}.
-Hysteresis, within cycles in the perturbation network or between forward and backward perturbations can be checked~\cite{wang2013modeling} indicate problematic perturbations involved in cycles connecting many compounds (See also Secs.~\ref{sec:relative-fe-protocol} and~\ref{sec:are-they-good}).
-Selection bias, as described above, can be problematic in prospective applications as projects normally seek to focus only on the most potent compounds.
-Thus, correlation between prediction and subsequent experiment may lead to disappointment but corrections can be applied based on previous recommendations~\cite{abel2017critical}.
-For a more detailed discussion on checking the robsustness of your alchemcial free energy calculation see also Section~\ref{sec:are-they-good}.
+Hysteresis, within cycles in the perturbation network or between forward and backward perturbations can be checked~\cite{wang2013modeling} to indicate problematic perturbations involved in cycles connecting many compounds (See also Secs.~\ref{sec:relative-fe-protocol} and~\ref{sec:are-they-good}).
+Once synthesis and testing of compounds is complete it would be normal to look back at how the calculations performed.
+In this regard it is important to consider upfront the issue of selection bias.
+It is tempting to only synthesize the compounds predicted to be most active, thus a narrow range of calculated activity is tested that imposes limits on the statistical assessment of performance, ideally example molecules from across the range of predicted activity can be assessed or corrections can be applied based on previous recommendations~\cite{abel2017critical}.
+For a more detailed discussion on checking the robustness of your alchemcial free energy calculation see also Section~\ref{sec:are-they-good}.
 
@@
 \label{subsec:binding}
-In principle in the limit of sufficient conformational sampling, the free energy changes estimated from an alchemical free energy calculation should be independent of the system's initial coordinates.
-However, in practice because simulations are of finite duration (typically 1-100 ns per state at present) this is only true for certain classes of alchemical free energy calculations such as relative or absolute free energies of hydration of small and relatively rigid organic molecules.
+In principle, in the limit of sufficient conformational sampling, the free energy changes estimated from an alchemical free energy calculation should be independent of the system's initial coordinates.
+However, in practice, because simulations are of finite duration (typically 1-100 ns per state at present), this is only true for certain classes of alchemical free energy calculations such as relative or absolute free energies of hydration of small and relatively rigid organic molecules.
 Protein-ligand complexes typically exhibit slowly relaxing degrees of freedom that significantly exceed the duration of an alchemical free energy calculation, and host-guest calculations can be susceptible to these issues as well, depending on timescale and system.
@@
 \paragraph{Are there good X-ray structures available?}
-As with any simulation, care should be taken in selecting available X-ray structures in the protein databank.
+As with any simulation, care should be taken in selecting available X-ray structures in the Protein DataBank.
 In some cases it may be wise to choose multiple starting structures to account for variability in receptor conformations as well as the accuracy of available X-ray structures.
@@
 
-Many free energy calculations focus on a congeneric series of ligands, which can make these calculations suitable for relative free energy protocols (see~\ref{sec:simulation_protocol_choice}).
+Many free energy calculations focus on a congeneric series of ligands, which can make these calculations suitable for relative free energy protocols (see Section~\ref{sec:simulation_protocol_choice}).
 For relative calculations, some care has to be taken selecting binding poses for these ligands.
@@
 
-\paragraph{How to deal with different binding modes?}
+\paragraph{Are you prepared to deal with any binding mode challenges?}
 Generally, binding modes within congeneric series are conserved~\cite{wacker2010conserved}, however, exceptions exist~\cite{brandt2011congeneric,nazare2005probing}, as discussed in more detail in Section~\ref{sec:multiple_binding_modes}.
 Certain functional groups may be particularly prone to this due to symmetries or near symmetries.
-One such issue involves a 180 degree flip in the dihedral angle of a 6-membered aromatic ring leading to a different spatial position of ortho- or meta- substituents that otherwise should overlap within a series.
-The 180 degree flip of the ring may not occur enough during simulations to overcome bias due to the starting configuration.
-Another scenario may be equatorial and axial substituted saturated rings (e.g.
+One such issue involves a 180 degree flip in the dihedral angle of an aromatic ring leading to a different spatial position of ortho- or meta- substituents that otherwise should overlap within a series.
+The 180 degree flip of the ring may not occur enough during simulations (due to steric obstructions) to overcome bias due to the starting configuration.
+Another scenario may be equatorial and axially substituted saturated rings (e.g.
 cyclohexane derivatives).
@@
 
-\paragraph{Have you considered sterioisomers and enationmers?}
+\paragraph{Have you considered sterioisomers and enantiomers?}
 Congeneric series can contain stereoisomers or enantiomers which can bind very differently, resulting in large errors if treated incorrectly.
@@
 consider pre-solvation methods such as GCMC steps~\cite{michel2010prediction}.
-Other tools such as WaterMap or open source equivalents (SSTmap, others) can be used to define water structure for systems with no experimental evidence of water sites~\cite{wang2011ligand}.
-Well known protein systems with water mediated ligand interactions are for example: HSP90 which formed part of the D3R grand challenge 2015~\cite{mey2016blinded}, A2A~\cite{brucemacdonald2018ligand}, MUP~\cite{ross2015water}, and others~\cite{michel2009energetics}.
+Other tools such as WaterMap or open source equivalents (SSTMap, GIST, and others) can be used to define water structure for systems with no experimental evidence of water sites~\cite{wang2011ligand}.
+Well known protein systems with water mediated ligand interactions are for example: HSP90 which formed part of the D3R grand challenge 2015~\cite{mey2016blinded}, A2A~\cite{brucemacdonald2018ligand}, MUP~\cite{ross2015water}, GPCRs, and others~\cite{michel2009energetics}.
 
 \paragraph{Protonation states depend on the pH of the experimental assay}
-Care should be taken when preparing ligands and proteins to match the experimental assay conditions in terms of pH, if they are known.
-The pH of the assay will determine the protonation states of the proteins and ligands.
-Since the pKa of amino acid residues is known different tools are available for proteins, such as the H++ server, or Maestro are good starting points~\cite{anandakrishnan2012automating, 2020schrodinger}.
-For ligands often the pKa needs to be deterimined if not known experimentally.
+Care should be taken when preparing ligands and proteins to match the pH of the experimental assay, if known.
+As mentioned above in Section 5.1, the pH of the assay can differ from neutral pH and will determine the protonation states of the proteins and ligands.
+Since the pKa of reference amino acid sidechain residues is known, but can vary in the protein environment, many different tools have emerged for predicting sidechain pKa in proteins, such as the H++ server, ProPKa, APBS, and Maestro ~\cite{anandakrishnan2012automating, 2020schrodinger}.
+Strongly acidic (Glu, Asp) or basic (Arg, Lys) sidechains can reliably predicted to be ionised, but care is still needed as the local environment can modify expected ionization states (for instance the catalytic Asp dyad in proteases).
+Histidine is notoriously more difficult to predict as its pKa suggests it ionizes closer to the experimental pH range.
+For ligands often the pKa needs to be determined, if not known experimentally.
 There are many different available tools for this purpose, but common choices may be propKa~\cite{olsson2011propka3,sondergaard2011improved}, Chemicalize (\url{https://chemicalize.com/welcome}), or Maestro~\cite{2020schrodinger}.
-While mostly it can be assumed that protonation states will remain the same for an e.g.
-ligand protein binding calculation, some care needs to be taken with systems where the protonation state may change upon binding~\cite{onufriev2013protonation}.
-
-\paragraph{How to align congeneric series}
-Input coordinates for a congeneric series may be generated by docking calculations, or by ligand alignment using maximum common substructure (MCSS) algorithms.
+Still, accurate pKa prediction for small molecules remains a challenging problem, even with dedicated tools.
+While often it can be assumed that the protonation state of a ligand and protein will remain the same foras a ligand binds, some care needs to be taken with systems where the protonation state may change upon binding~\cite{onufriev2013protonation}.
+BACE, for example, famously undergoes a protonation state change on ligand binding.
+
+\paragraph{Congeneric series often need alignment}
+Input coordinates for a congeneric series may be generated by docking calculations, or by ligand alignment using MCSS algorithms.
 The latter tends to produce alignments that are more conserved and more consistent free energy changes across a dataset, but will struggle to yield reasonable results for relative binding free energy calculations that involve a significant binding mode rearrangement.
@@
 
-If the binding site location is uncertain but the structure of the receptor is well defined and plausible binding sites are identified, it may be more useful to chose an absolute free energy protocol to compute the standard free energy of binding of the ligand to a set of binding sites.
+If the binding site location is uncertain but the structure of the receptor is well defined and plausible binding sites are identified, it may be more useful to choose an absolute free energy protocol to compute the standard free energy of binding of the ligand to a set of binding sites.
 This requires the user to prepare input files describing the bound conformation in different putative binding sites~\cite{evoli2016multiple}.
@@
 If the putative binding sites are not apparent, for instance due to significant induced-fit effects, it may be challenging to obtain meaningful free energies of binding.
-One may have to account for the free energy change for forming a binding site in the target receptor which may not be feasible on alchemcial simulation timescales.
+One may have to account for the free energy cost of forming a binding site in the target receptor which may not be feasible on alchemical simulation timescales.
 
@@
 See for example~\cite{bosisio2016blinded,rustenburg2016measuring} for a good discussion of these choices.
+And, while sampling problems might seem to be a non-issue for small molecules, this is not always the case; e.g.
+even the hydroxyl orientation on neutral carboxylic acids can occasionally pose a challenge~\cite{}.
 
@@
 \label{sec:simulation_protocol_choice}
-Alchemical free energy calculations can be grouped into two main categories, ``absolute'' (see Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}) and ``relative''\footnote{The distinction is a bit of a misnomer, since both compute ratios of partition functions relative to another state, and neither computes an absolute free energy.} (see Fig.~\ref{fig:fig_binding_thermodynamic_cycle}), which differ in whether they compute properties for a single molecule (absolute) or compare properties of different, usually closely related, molecules (relative).
+Alchemical free energy calculations can be grouped into two main categories, ``absolute'' (see Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}) and ``relative''\footnote{The distinction is a bit of a misnomer, since both compute ratios of partition functions relative to another state and in that sense are relative, while neither computes an absolute free energy.} (see Fig.~\ref{fig:fig_binding_thermodynamic_cycle}), which differ in whether they compute properties for a single molecule (absolute) or compare properties of different, usually closely related, molecules (relative).
 To use binding as a concrete example, in absolute binding free energy calculations, we compute the binding free energy of a ligand to an individual receptor relative to a standard reference concentration.
@@
 consider a transformation of benzene to benzyl alcohol Fig.~\ref{fig:fig_topology}.
-In this case the common substructure is the benzene ring, though may be larger depending on how it is defined, as we discuss below.
+In this case the common substructure is the benzene ring, though the substructure may be larger depending on how it is defined, as we discuss below.
+
 In single topology calculations, the overall transformation is set up to involve as few additional atoms as possible, so benzene would be typically changed into benzyl alcohol by changing one of the hydrogens into a carbon.
-On this new carbon you will also find non-interacting atoms called ``dummy atoms'' (retaining their bonded interactions but not interacting with the rest of the system), which will become the two hydrogens attached to the new carbon and the OH group after the transformation.
-Bond parameters as well as partial charges are adjusted accordingly.
-Thus, in a single topology calculation, atoms may change their type ensuring only relatively few dummy atoms are created. This is illustrated in the left arm of Fig.~\ref{fig:fig_topology}. 
+This site will also be the future home of two additional hydrogen atoms bound to the new carbon, so these must initially be present as non-interacting atoms called ``dummy atoms'', which retain their bonded interactions but do not interact with the rest of the system.
+Bond parameters as well as partial charges between the changing atoms are adjusted accordingly between the initial and final state.
+Thus, in a single topology calculation, atoms may change their type, ensuring minimal  dummy atoms are created. This is illustrated in the left arm of Fig.~\ref{fig:fig_topology}.
+
 In contrast, in a dual topology alchemical free energy calculation, no atoms are allowed to change type~\cite{shirts2012best}.
-This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxy methyl group, then passing through an intermediate state where atoms which are becoming dummy atoms or ceasing being dummy atoms are partially interacting (this state may or may not be well defined~\cite{mobley2014blind} and culminating in a state where benzyl alcohol is present along with the additional dummy atom which was previously a corresponding hydrogen of the benzene.
+This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxy methyl group, then passing through an intermediate state where some atoms are partially interacting--- particularly, those atoms which are becoming dummy atoms or ceasing to be dummy atoms (this state may or may not be well-defined~\cite{mobley2014blind}).
+The transformation finally culminates in a state where benzyl alcohol is present along with the additional dummy atom which was previously a corresponding hydrogen of the benzene.
 Fig.~\ref{fig:fig_topology}'s right branch depicts how such a dual topology works.
@@
 \end{figure}
-Hybrid topology calculations have seen much less use but essentially consist of two absolute free energy calculations in opposite directions at the same time (turning one molecule off while turning the other on)~\cite{jiang2019computing}.
-Various software implementations currently mostly rely on single or dual topology approaches; for example AMBER TI uses a dual topology approach, while Gromacs and BioSimSpace use single topology approaches.
-Please make sure to check what approach is used with your software package of choice.
-To our knowledge efficiency differences have not been thoroughly explored, though conventional wisdom suggests that fewer dummy atoms are better~\cite{liu2013lead, mobley2012perspective}.
+
+Hybrid topology calculations have seen much less use but essentially consist of two absolute free energy calculations in opposite directions at the same time, turning one molecule's interactions with the environment off, while turning the other molecule's interaction on~\cite{jiang2019computing}.
+
+Most existing software implementations currently use single or dual topology approaches; for example, AMBER TI uses a dual topology approach, while BioSimSpace uses a single topology approach.
+Please make sure to check what approach is used with your software package of choice, or whether it supports your choice of approach (GROMACS, for example, supports both).
+To our knowledge efficiency differences have not been thoroughly explored, though conventional wisdom suggests that fewer dummy atoms are better, as introducing or removing atomic sites is usually more difficult, requireing more intermediate steps ~\cite{liu2013lead, mobley2012perspective}.
 
@@
 Once a particular approach to the topology is selected, a crucial next step is to identify the common atoms which will not be perturbed.
-Rigorously, this process comprises a maximal common substructure (MCSS) search of the molecules involved to identify the common substructure -- though the parameters of the MCSS search will differ depending on whether single or dual topology calculations are planned.
+Rigorously, this process typically comprises a MCSS search of the molecules involved to identify the common substructure---though the parameters of the MCSS search will differ depending on whether single or dual topology calculations are planned.
 Specifically, with a single topology approach in mind, atom types are allowed to change, so a permissive MCSS search can be done, whereas with dual topology a more strict search is required.
@@
 There are different tools that allow the generation of MCSS matches as well as single topology input.
-A large number of software can compute MCSS matches, some rely on RDKit~\cite{rdkit2019Dec}, such as LOMAP~\cite{liu2013lead} or FESetup~\cite{loeffler2015fesetup} and partially BioSimSpace~\cite{hedges2019biosimspace}, others such as fkckombu~\cite{kawabata20143d} are standalone tools.
-Schr\"{o}dinger's FEP+ planning tool is based on a version of LOMAP, and it also uses MCSS matching as well as 3D considerations to plan the network of single topology calculations between molecules~\cite{wang2015accurate}.
-
-MCSS searches can be relatively time consuming, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial assessment and then use MCSS only to identify final mappings for relative calculations~\cite{raymond2002maximum,klabunde2012mars,jones2009elucidating}.
+A large number of software tools can compute MCSS matches using different cheminformatics packages.
+Some rely on RDKit~\cite{rdkit2019Dec}, such as LOMAP~\cite{liu2013lead}, FESetup~\cite{loeffler2015fesetup} and partially BioSimSpace~\cite{hedges2019biosimspace}, while others such as fkckombu~\cite{kawabata20143d} are standalone tools.
+Schr\"{o}dinger's FEP+ planning tool was originally based on a version of LOMAP, and it also uses MCSS matching as well as 3D considerations to plan the network of single topology calculations between molecules~\cite{wang2015accurate}.
+
+MCSS searches can be relatively time consuming, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial similarity assessment and then use MCSS only to identify final mappings for relative calculations~\cite{raymond2002maximum,klabunde2012mars,jones2009elucidating}.
 The MCSS approach, though relatively standard, takes into account only topological similarity.
@@
 \paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss} \textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
-Some approaches have attempted to address this~\cite{clark2019relative} but a general solution is not yet in mainstream use.
+Some approaches have attempted to address this~\cite{clark2019relative} but a general solution is not yet in mainstream use, though FEP+ implements one solution.
 
 \paragraph{Perturbation maps}
-Based on the input ligand series a perturbation map or network can be planned.
+Based on the input ligand series, a perturbation map or network can be planned.
 Recent heuristics have shown the more connected the perturbation network the better, however, there is a way to optimize network structure while minimizing the number of perturbations that need to be computed reducing the resulting computational cost~\cite{yang2020optimal}.
-Sometimes the introduction of intermediates that are not part of the original congeneric series are essential to avoid ring breaking, or deal with perturbations that would otherwise result in large numbers of atoms to be grown or disappearing.
+Sometimes the introduction of intermediates that are not part of the original congeneric series are essential to avoid ring breaking, or deal with perturbations that would otherwise result in large numbers of atoms being inserted or deleted.
 Some commercial tools have good underlying heuristics but may fail with complicated input, needing user validation in particular when dealing with chiral compounds.
@@
 
-In some cases, during the lead optimization stage, or for very large datasets that would benefit from rougher initial free evnergy ranking, or in cases where perturbations would be rather large a star shaped network as seen in Fig.~\ref{fig:fig_types_of_networks} \textbf{A} is used.
+In some cases, during the lead optimization stage, or for very large datasets that would benefit from rougher initial free energy ranking, or in cases where perturbations would be rather large a star shaped network as seen in Fig.~\ref{fig:fig_types_of_networks} \textbf{A} is used.
 However, adding redundancy into the network means that a better error analysis can be carried out, by looking at cycle closure errors as discussed in sec.~\ref{sec:are-they-good}, with an example given in Fig.~\ref{fig:fig_types_of_networks} \textbf{B}.
@@
 Absolute free energy calculations involve completely removing the interactions between the ligand or solute and its environment, taking it to a non-interacting state that may or may not retain intramolecular non-bonded interactions.
-This non-interacting state can then be shifted between environments (from the protein to water, or from one solution to another) without changing its free energy, and then interactions can be restored.
+This non-interacting state can then be shifted between environments (from the protein to water, or from one solution to another) without changing its free energy other than that due to the changing volume of the simulations, and then interactions can be restored in the new environment.
 
@@
 that the transformation involves a 1 M to 1 M equivalent transfer free energy (where the non-interacting solute still occupies essentially the same volume as the solute in the interacting system).
-So typically in such cases no special care is required to ensure the correct standard state, as long as the \emph{experimental} data being analyzed uses the same standard state and if it does not, a simple entropic correction is needed.
-
-However, for binding the situation is much more complex and requires special care.
-Experimental absolute binding free energies are reported relative to a specific reference state -- a 1 M standard state -- which must also be used in calculations.
-In practice, this has implications for how the calculations are done, as the reference concentration must enter the thermodynamic cycle employed, see an example of such a thermodynamic cycle in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}.
-
-Typically, to deal with both practical sampling issues and the standard state issue, restraints are employed in absolute binding free energy calculations to keep the ligand in a well defined volume as its interactions with the system are removed~\cite{gilson1997statisticalthermodynamic}.
-This solves two problems.
-First, if the ligand were not kept in a well-defined region, as its interactions were removed it might wander the system, perhaps quite slowly, and only inadequately sample the non-interacting or weakly interacting state.
-However, adequate sampling of these states might be required for convergence.
-The use of restraints can therefore dramatically improve sampling as interactions are weakened and removed.
-Second, if the ligand is not kept in a well-defined region then it is hard to determine how to link a computed binding free energy to the correct 1 M standard state.
-In contrast, with restraints, the free energy of releasing the restrained ligand to a 1 M standard state can be computed analytically or numerically by solving the relevant integral, allowing the standard state to enter the thermodynamic cycle.
+So typically in such cases no special care is required to ensure the correct standard state, as long as the \emph{experimental} data being analyzed uses the same standard state.
+If this is not the case, a simple entropic correction is needed.
+
+For binding, however, the situation is more complex and requires special care.
+Because the simulations are typically performed using restraints and at concentrations that are different from 1 M, the expression of the free energy requires the following correction~\cite{gilson1997statisticalthermodynamic} (see an example of such a thermodynamic cycle in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle})
+\begin{equation}\label{eq:restraint-correction}
+    \Delta G^{\circ}_{\mathrm{restr}} = -k_BT ~ \ln \left( \frac{c^{\circ}}{V_L} \right) -k_BT ~ \ln \left( \frac{\xi_L}{8 \pi^2} \right)
+\end{equation}
+where $V_L$ and $\xi_L$ are respectively the volume of the translational and rotational degrees of freedom of the non-interacting ligand in the simulation box.
+When no restraints are used, the non-interacting ligand is free to translate and rotate in the simulation box (i.e., $V_L = V_{\mathrm{box}}$ and $\xi_L = 8\pi^2$), and the rotational term is zero.
+A sufficiently thorough exploration of the simulation box by the non-interacting ligand is, however, required for the formula to be valid.
+This is typically hard to achieve as the exploration process is governed by diffusion.
+The addition of a restraint limits the volume available to the non-interacting ligand, thus speeding the convergence of the sampling.
+In addition, when enhanced sampling methods such as Hamiltonian $\lambda$ exchange are used (see Section~\ref{sec:sampling_schemes}), the use of a restraint is typically necessary as it keeps the ligand in the binding site in the interacting state (see also Section~\ref{sec:weak-binders}) and generally reduce the round-trip time of replicas.
+When restraint are employed, the values of $V_L$ and $\xi_L$ are restraint-dependent, but for commonly employed restraints, these can be usually easily computed analytically or numerically by solving the relevant integral.
 
@@
 In practice, a variety of types of restraints are common, from simple harmonic distance restraints between the ligand and the protein~\cite{mobley2006use}, to flat-bottom restraints which work similarly but only exert a force if the ligand leaves a specific region~\cite{chen2007can}.
+Because these restraints do not limit the rotational degrees of freedom of the ligand, the rotational term entering the correction in Eq.~\ref{eq:restraint-correction} is zero.
 
@@
 
-In principle, all of these forms will yield correct binding free energies in the limit of adequate sampling (if their effects and connection to the standard state are correctly handled) but they have different strengths and weaknesses.
+In principle, all of these forms will yield correct binding free energies in the limit of adequate sampling if their effects and connection to the standard state are correctly handled, but they have different strengths and weaknesses.
 For example, with more involved restraints, sampling at intermediate $\vec{\lambda}$ values will usually not need to be as extensive but more computational effort must go to computing the restraining free energy.
-Additionally, such restraints would typically keep the ligand from exploring alternative binding modes, which may be undesirable with Hamiltonian $\lambda$ exchange or expanded ensemble techniques where allowing the ligand to exchange binding modes when it is non-interacting could provide sampling benefits~\cite{wang2013identifying}.
-Concretely, flat-bottom restraints might allow a ligand to explore multiple binding sites, harmonic restraints multiple binding modes within a site, and Boresch restraints a single binding mode within a single site.
+Additionally, such restraints would typically keep the ligand from exploring alternative binding modes.
+This  restriction may be undesirable when using Hamiltonian $\lambda$ exchange or expanded ensemble techniques where allowing the ligand to exchange binding modes when it is non-interacting could provide sampling benefits~\cite{wang2013identifying}.
+More specifically, flat-bottom restraints might allow a ligand to explore multiple binding sites, harmonic restraints multiple binding modes within a site, while Boresch restraints a single binding mode within a single site.
 See additional discussion of the possibility of multiple binding modes in Sec.~\ref{sec:multiple_binding_modes} below.
@@
 \subsection{Absolute and relative calculations deal with some of the same issues}
-\subsubsection{Handling weak binders and high dissociation rates}
+\subsubsection{Handling weak binders and high dissociation rates}\label{sec:weak-binders}
 In binding free energy calculations, only the conformations in which the receptor and ligand form a bound complex should be sampled from the bound states (Section~\ref{sec:theory}).
@@
 \subsubsection{Changes in net charge can be challenging/problematic.}
-If the net charge of the system will change as the alchemical calculation progresses, this can pose major challenges.
-Specifically, finite-size effects can introduce profound artifacts into computed binding free energies~\cite{}MISSING citation, in part because typical schemes for long-range electrostatics (including PME and reaction field) do not handle free energy contributions from such changes effectively or as they would be handled in a hypothetical macroscopic bulk solution~\cite{}.
+If the net charge of the system changes as the alchemical variable changes during the  calculation, this can pose major challenges.
+Specifically, finite-size effects can introduce significant charge-dependent artifacts into computed binding free energies~\cite{}MISSING citation, in part because typical schemes for long-range electrostatics (including PME and reaction field) do not handle free energy contributions from such changes effectively or as they would be handled in a hypothetical macroscopic bulk solution~\cite{}.
 
@@
 Intermediates states are required for a sufficient overlap in potential energies to estimate a free energy difference between $\vec{\lambda}=0$ and $\vec{\lambda}=1$.
-    Softcore potentials provide one of the most efficient families of intermediate pathways, with a $\vec{\lambda}$ dependence illustrated in (\textbf{C}) and a potential energy dependence shown in (\textbf{D}).
+    Soft-core potentials provide one of the most efficient families of intermediate pathways, with a $\vec{\lambda}$ dependence illustrated in (\textbf{C}) and a potential energy dependence shown in (\textbf{D}).
 Note how as $\vec{\lambda}$ approaches 0, the energy smoothly approaches zero at all $r$, a necessary requirement for efficient and stable calculations.
@@
 
-Some choices are particularly crucial -- for example, transformations involving insertions or deletions of atoms should employ a soft-core potential path for Lennard-Jones or other interactions with repulsive interactions that go to infinite energy at small radius~\cite{beutler1994avoiding, beutler1994molecular,gapsys2012new}.
-
-The key consideration for choosing alchemical pathways is that the intermediate states that a given pathway produces, should sample configurational ensembles that change as slowly as possible as $\vec{\lambda}$ changes, while still managing to go from the initial state to the final state as $\vec{\lambda}$ goes from 0 to 1.
+Some choices are particularly crucial---for example, transformations involving insertions or deletions of atoms should employ a soft-core potential path for Lennard-Jones or other interactions with repulsive interactions that go to infinite energy at small radius~\cite{beutler1994avoiding, beutler1994molecular,gapsys2012new}.
+
+The key consideration for choosing alchemical pathways is that the intermediate states that a given pathway produces should sample configurational ensembles that change as slowly as possible as $\vec{\lambda}$ changes, while still managing to go from the initial state to the final state as $\vec{\lambda}$ goes from 0 to 1.
 
@@
 This can be proven directly from the BAR and MBAR formulas~\cite{bennett1976efficient,klimovich2015guidelines}, though the exact same principles apply for TI.
-For a 'good' path to work, sufficient overlap between potential energies is required.
+For a 'good' path to work and give a sequence of states with maximally similar configurations, sufficient similarity in potential energies is required.
 Fig.~\ref{fig:fig_what_is_lambda} \textbf{A} and \textbf{B} illustrate this.
@@
 
-So how do we adjust the potentials between the two end states based on $\vec{\lambda}$? The simplest possible alchemical pathway is a \textit{linear} pathway:
+So what are the options to adjust the potentials between the two end states based on $\vec{\lambda}$? The simplest possible alchemical pathway is a \textit{linear} pathway:
 \begin{equation}
@@
 One way to look at this is to examine how low $\vec{\lambda}$ values must go to reduce the energy at $0.5\sigma$ (the atomic size parameter) down to 1 $k_BT$, where thermal fluctuations make it possible for other atomic sites to penetrate routinely that deep.
-If we are trying to go from a particle being present, and desire to make it disappear alchemically, then if the repulsive terms are of the form $\epsilon\frac{\sigma}{r}^{12}$, then if $\epsilon$ was 1 $k_BT$ at the temperature of interest, and we start with the particle present, then solving for $(1-\vec{\lambda})(1 kB_T)\frac{\sigma}{\frac{1}{2}\sigma}) = 1 k_B T$ we get $\vec{\lambda} = 1-2^{-12} \sim 0.999976$.
-So we have gone virtually all the way to the end of the transformation, but there is still an impenetrable post in the middle of our simulation!
+If we are trying to go from a particle being present, and desire to make it disappear alchemically, then if the repulsive terms are of the form $\epsilon\frac{\sigma}{r}^{12}$, then if $\epsilon$ was 1 $k_BT$ at the temperature of interest, and we start with the particle present, then solving for $(1-\vec{\lambda})(1 kB_T)\left(\frac{\sigma}{\frac{1}{2}\sigma}\right)^{12} = 1 k_B T$ we get $\vec{\lambda} = 1-2^{-12} \sim 0.999976$.
+At this point,  we have gone virtually all the way to the end of the transformation, but there is still an impenetrable post in the middle of our simulation!
 This is not very much like the desired final state of no interactions between the particle and its environment.
@@
 
-where $r_{ij}$ is the distance between two particles $i$ and $j$, $\epsilon_{ij}$ and $\sigma_{ij}$ are the Lennard-Jones parameters corresponding to the interaction between particles $i$ and $j$, and $\alpha$ is a constant (0.5 is optimal for the functional form shown above).
+where $r_{ij}$ is the distance between two particles $i$ and $j$, $\epsilon_{ij}$ and $\sigma_{ij}$ are the Lennard-Jones parameters corresponding to the interaction between particles $i$ and $j$, and $\alpha$ is a constant (0.5 is optimal for the specific functional form shown above).
 This functional form has exactly the property we are looking for: it recovers the Lennard-Jones potential when $\vec{\lambda}=1$, and the other endpoint ($\vec{\lambda}=0$), it is exactly zero for all $r_{ij}$ everywhere, and as $\vec{\lambda}$ goes to zero, the $\alpha(1-\vec{\lambda})$ term lowers the infinite energy in the core.
@@
 
-It has been shown that more complicated forms are not significantly more efficient than eq.~\cite{pham2012optimal}.
-We recommend using the softcore potential given in eq.~\ref{eq:softcore}, unless there is a compelling reason otherwise.
+It has been shown that more complicated forms are not significantly more efficient than eq.~\ref{eq:softcore}~\cite{pham2012optimal}.
+We therefore recommend using the softcore potential given in eq.~\ref{eq:softcore}, unless there is a compelling reason otherwise.
 Using a similar equation to eq.~\ref{eq:softcore} may be acceptable in most circumstances if that is what is supported in your chosen software.
-However, if you are inserting or removing entire atomic sites, we heavily recommend against using the linear approach; it will be very difficult to get correct results.
-
-We have discussed optimal ways of disappearing or appearing Lennard-Jones interaction sites and turning on and off electrostatics terms.
+However, if you are inserting or removing entire atomic sites, we heavily recommend against using the linear approach; it will be very difficult to get correct and converged results.
+
+So far in this section, we have discussed optimal ways of disappearing or appearing Lennard-Jones interaction sites and turning on and off electrostatics terms.
 What about performing both transformations at the same time?
 We can not turn off the electrostatics linearly at the same time we turn off the Lennard-Jones terms, as it would leave infinitely large attractive and repulsive electrostatic terms "bare" at small $\vec{\lambda}$, resulting in the simulation crashing.
-It \textit{is} possible to apply the same soft core approach to the Coulomb interaction and this is indeed done in a number of implementations, in which case it is important that the Coulomb interaction is softened more rapidly than the Lennard-Jones interaction to avoid charge penetration issues, this can be tricky to ensure for all types of perturbations ~\cite{steinbrecher2011softcore}.
+It \textit{is} possible to apply the same soft core approach to the Coulomb interaction and this is indeed done in a number of implementations, in which case it is important that the Coulomb interaction is softened more rapidly than the Lennard-Jones interaction to avoid charge penetration issues, which  can be tricky to ensure for all types of perturbations ~\cite{steinbrecher2011softcore}.
 
 A safe but potentially more computationally expensive approach is to perform the transformations in sequence; first, turning off all electrostatics for atoms that must be removed, inserting and removing Lennard-Jones sites (both the insertion and removal can be done simultaneously), and then turning electrostatics for the introduced particles on.
-Again, If there are no removals or introduction to atomic sites, then it is reasonable to change all interactions linearly.
+Again, If there are no removals or introduction to atomic sites, then it is reasonable to change the interactions in the first and third steps  linearly.
 
@@
 through either annihilation~\cite{hermans1997inclusiona, mann2000modelinga, boresch2003absolutea, wang2006absolutea, mobley2006use} or decoupling~\cite{fujitani2005directa, mobley2006use} must be considered.
-Reasonable efficiency can be often obtained with either choice even if some are somewhat better or worse than others although there is not consensus on which is better in all situations.
+Reasonable efficiency can be often obtained with either choice even if some are somewhat better or worse than others, and there is no consensus on which is better in most given situations.
 Our recommendation is to leave the intramolecular interactions on during the transformation for simplicity if there are no other known issues with this approach.
 The key thing to watch out for is whether the total potential energy, and therefore the intermediate ensembles sampled, change smoothly from beginning to end.
-These problems can be diagnosed by noticing lack of configuration space overlap between different simulations (see section~\ref{sec:are-they-good})"
+These problems can be diagnosed by noticing lack of configuration space overlap between different simulations (see section~\ref{sec:are-they-good}).
 
@@
 
-Absolute free energy calculations for dissimilar ligands can have particular challenges because the (potentially incorrect) assumption of consistent binding modes across a series of similar ligands is likely to be even less robust than the in the case of relatively calculations.
+Absolute free energy calculations for dissimilar ligands can have particular challenges because the (potentially incorrect) assumption of consistent binding modes across a series of similar ligands is likely to be even less robust than the in the case of relative calculations.
 This means that researchers performing absolute binding free energy calculations will have to pay particular attention to generating reasonable putative binding modes.
@@
 
-Sampling all binding modes within a single free energy calculation is usually impractical without some form of enhanced sampling or at least Hamiltonian replica exchange~\cite{wang2013identifying} because barriers for binding mode interconversion result in kinetics which are too slow compared to simulation timescales~\cite{mobley2006use, palma2012computation,mobley2012perspective, gill2018binding}.
+Sampling all potential lignad binding modes within a single free energy calculation is usually impractical without some form of enhanced sampling or at least Hamiltonian replica exchange~\cite{wang2013identifying} because barriers for binding mode interconversion result in kinetics which are too slow compared to simulation timescales~\cite{mobley2006use, palma2012computation,mobley2012perspective, gill2018binding}.
 Hamiltonian exchange, coupled with appropriate restraints, can allow the ligand to relatively rapidly exchange between potential binding modes when non-interacting, accelerating sampling of binding modes~\cite{wang2013identifying}.
+However, it is not always clear that this is desirable, since this also increases the size of the configuration space which must be sampled even if the binding mode is known.
 
@@
 Although BAR itself gives a free energy estimate that is asymptotically correct in and is much less biased than the uncertainty estimate for EXP, the uncertainties in $\Delta {f_{i-1\rightarrow i}}$ and $\Delta {f_{i\rightarrow i+1}}$ are not uncorrelated, because they both involve the energies $u_i(\vec{q})$.
-The variances of each of the free energies will \textit{not} propagate as variances usually do (in quadrature) to the variance of the overall free energy.
+The variances of each of the free energies will \textit{not} propagate as variances usually do (in quadrature) into the variance of the overall free energy.
 Instead, some other method for propagating the uncertainty, such as bootstrapping~\cite{grossfield2018best} must be used.
@@
 Averaging over $\expect{\frac{du}{d\vec{\lambda}}}$ requires fewer uncorrelated samples to reach a given level of relative error
-than averaging $e^{-u(\vec{q})}$, as the range is usually narrower with a more Gaussian distribution.
+than averaging $e^{-u(\vec{q})}$, as the distribution of values is usually narrower, with a more Gaussian shape to the distribution.
 Rather than being limited by overlap, as in the case of BAR and MBAR (see below), we are instead limited by the bias in the numerical quadrature, which must be minimized sufficiently to be beneath the level of statistical noise.
@@
 have widely-varying levels of uncertainty.
-For example, integrating a cubic spline interpolation provided negligible benefits~\cite{paliwal2011benchmark}.
-For starting researchers, we therefore recommend a simple trapezoid rule scheme, as it allows for maximal flexibility in which values of $\vec{\lambda}$ are simulated.
+For example, integrating a cubic spline interpolation provided negligible benefits over a simple trapezoid rule~\cite{paliwal2011benchmark}.
+For starting researchers, we therefore recommend the simple trapezoid rule scheme, as it allows for maximal flexibility in which values of $\vec{\lambda}$ are simulated.
 As fitting to higher order polynomials can have numerical instabilities for some energy functions, and because alternate functional forms might only be appropriate with some types of transformations, expertise and experience is required to perform such numerical integration modifications.
@@
 Unfortunately, many problems of interest require using pathways (such as the soft-core pathways, for removing repulsive interactions) that are not linear, as we discuss, making this more complex.
-Still, if the code of interest does compute $\frac{du}{d\vec{\lambda}}$, then TI is perhaps the simplest method to use, as it involves a very little post-processing.
+Still, if the code of interest does compute $\frac{du}{d\vec{\lambda}}$, then TI is perhaps the simplest method to use, as it involves a very little post-processing effort.
 
 \paragraph{The multistate Bennett acceptance ratio (MBAR)}
-One can generalize Bennett's logic from two states to multiple states to obtain a free energy estimator that uses energy differences between all intermediate states.
+One can generalize Bennett's logic from two states to multiple states to obtain a free energy estimator that uses energy differences between configurations at all intermediate states to compute free energy differences between all states.
 MBAR gives a system of implicit equations for the free energies $f_i$:
@@
 The equations can be solved by a number of different standard routines.
-We note that there are only $K-1$ independent equations, so only K-1 of the free energies are independent variables, and one of the $f_i$ must be specified (usually, without loss of generality, setting it to zero).
-
-MBAR is probably the lowest variance asymptotically unbiased estimator of the free energy given the energies of the samples~\cite{tan2004likelihood}, which means that BAR is also for the free energy difference between only 2 states, as it is mathematically exactly the same as MBAR in this case.
-MBAR also provides an uncertainty estimate, derived from standard error propagation methods, which has been shown to be accurate as long as there are sufficient samples at each state~\cite{paliwal2011benchmark}.
-
-MBAR can also be thought of as the Zwanzig estimator of the free energy to state $i$ where the sampled distribution is the \textit{mixture distribution} of all the other samples thrown together in one ``pot'', defined by $p_m(\vec{q}) = N^{-1} \sum_k N_k \exp(f_k-u_k\vec{q})$, the weighted average of all the individual normalized probability distributions that are performed.~\cite{shirts2017reweighting}.
+We note that there are only $K-1$ independent equations, so only $K-1$ of the free energies are independent variables, and one of the $f_i$ must be specified (usually, without loss of generality, setting it to zero).
+
+MBAR is provably the lowest variance asymptotically unbiased estimator of the free energy given the energies of the samples~\cite{tan2004likelihood}, which means that BAR is also the lowest variance estimator for the free energy difference between only 2 states, as it is mathematically exactly the same as MBAR in this case.
+MBAR also provides an uncertainty estimate, derived from standard error propagation methods for implicit functions, which has been shown to be highly accurate as long as there are sufficient samples at each state~\cite{paliwal2011benchmark}.
+
+MBAR can also be thought of as the Zwanzig estimator of the free energy to state $i$ where the sampled distribution is the \textit{mixture distribution} of all the other samples thrown together in one ``pot'', defined by $p_m(\vec{q}) = N^{-1} \sum_k N_k \exp(f_k-u_k\vec{q})$, which is the weighted average of all the individual normalized probability distributions from the simulations that are performed.~\cite{shirts2017reweighting}.
 
@@
 \begin{itemize}
-\item We recommend MBAR if all energy differences are available. It is likely the lowest variance free energy estimate given samples from multiple states.
+\item We recommend MBAR if all energy differences are available. It is the lowest variance unbiased free energy estimate given samples from multiple states.
 \item BAR is essentially just as good as MBAR for highly optimized $\vec{\lambda}$ intermediates.
@@
 Thus MBAR does not actually get significant information from these energy differences, so one might as well not even calculate them, and just perform BAR between nearest neighbors.~\cite{paliwal2011benchmark}
-\item TI usually gives similar values as MBAR implemented with sufficient numbers of intermediates, but quadrature errors are hard to estimate beforehand.~\cite{paliwal2011benchmark}
+\item TI usually gives similar values as MBAR implemented with sufficient numbers of intermediates, but quadrature errors are hard to estimate beforehand  can occur if one is not careful.~\cite{paliwal2011benchmark}
 \item WHAM is an approximation to MBAR, and there are no compelling reasons it should be used.
+If careful, it is not necessarily much worse than the other methods, but it always introduces some degree if binning error.
 \item Other variants, especially ones that adaptively determine the free energies can be useful in certain circumstances but beyond the scope of a Best Practices article.
@@
 For alchemical changes that result in smooth, low curvature sets of $\expect{\frac{dU}{d\vec{\lambda}}}$, a relatively small number of $\vec{\lambda}$ states is necessary for sufficient accuracy and low variance in the free energy estimate.
-Depending on the 'difficulty' of the perturbation, the bias introduced by discretization of the integral can become large due to increased curvature, and more $\vec{\lambda}$ intermediate states become necessary to reduce error.
+Depending on the difficulty of the perturbation, the bias introduced by discretization of the integral can become large due to increased curvature, and more $\vec{\lambda}$ intermediate states become necessary to reduce error.
 It is recommended that researchers verify that a sufficient number of states are included such that the free energy is essentially invariant to the number of lambda intermediate states chosen.
@@
 Longer correlation times in CB8-G3 cause the calculation to converge more slowly.
-The original data used to generate the plot can be found at \url{https://github.com/MobleyLab/SAMPL6/blob/master/host_guest/Analysis/SAMPLing/Data/reference_free_energies.csv}.
+The original data used to generate the plot can be found at \url{https://github.com/samplchallenges/SAMPL6/blob/master/host_guest/Analysis/SAMPLing/Data/reference_free_energies.csv}.
 }
@@
 $\mathcal{O}$ is a $K\times K$ matrix, with $K$ being the number of simulated states, i.e.
-$\vec{\lambda}$s.
+values of $\vec{\lambda}$.
 Sufficient overlap is important for reweighting estimators such as BAR or MBAR, but cannot help assess reliability of estimates when using TI.
@@
 While this does not necessarily imply that the result for this perturbation is wrong, the energy estimate must at least be considered as highly unreliable.
-In order to overcome the issue of poor overlap in this example, additional sampling should be performed by introducing additional states, i.e. $\vec{\lambda}$ windows.
+In order to overcome the issue of poor overlap in this example, additional sampling should be performed by introducing additional states, i.e. $\vec{\lambda}$ values.
 
@@
 
-As we are interested mainly in the linear relationship between the alchemical free energy predictions and the experimentally-determined affinity values, plots should be depicted with the same range on both axes (i.e.
+As we are interested mainly in the linear relationship between the alchemical free energy predictions and the experimentally -determined affinity values, plots should be depicted with the same range on both axes (i.e.
 $x=y$) with a 1:1 aspect ratio, with units for both experiment and simulation converted to be the same.
@@
 It only runs on GPUs, using CUDA or OpenCL~\cite{kuhn2020assessment}.
+It is build on top of the open source software packages Sire and BioSimSpace (cf.
+below).
     \item The molecular operating environment (\href{https://www.chemcomp.com/Products.htm}{MOE}) offered by the Chemical Computing Group (CCG) has a tool for performing free energy calculations.
@@
 	\item \href{https://biosimspace.org/}{BioSimSpace} is a free, open source, multiscale molecular simulation framework, written to allow computational modellers to quickly prototype and develop new algorithms for molecular simulation and molecular design~\cite{hedges2019biosimspace}.
+	\item \href{https://siremol.org/}{Sire} is a multiscale, molecular simulation framework that provides several applications, including SOMD, an MD/MC code for performing FEP calculations via an interface to OpenMM.
 	\item \href{http://getyank.org/latest/index.html}{YANK} is a tool developed by John Chodera and group on the top of OpenMM MD package.
@@
 	\item \href{http://www.gromacs.org/}{GROMACS} is a molecular simulation package with a significant number of free energy methods implementations.
-The LiveCOMS GROMACS tutorial has an example free energy calculation~\cite{lemkul2018From}.
+The LiveCOMS GROMACS tutorial includes an example free energy calculation~\cite{lemkul2018From}.
 	\item \href{http://pmx.mpibpc.mpg.de/instructions.html}{PMX}, an add-on to GROMACS, offers a mutation free energy calculation module\cite{abraham2015gromacs}.
@@
 	\begin{itemize}
-	\item href{http://pmx.mpibpc.mpg.de/instructions.html}{PMX}: GROMACS \url{https://github.com/deGrootLab/pmx}.
+	\item \href{http://pmx.mpibpc.mpg.de/instructions.html}{PMX}: GROMACS \url{https://github.com/deGrootLab/pmx}.
 	\item \href{https://github.com/MobleyLab/Lomap}{Lomap/Lomap2} : Relative alchemical transformation graph planning for setting up perturbation networks~\cite{liu2013lead}.
@@
 \begin{itemize}
- \item Your ligands should have the same, biologically correct binding pose
-        \item Make sure your \textlambda-scheduling is set correctly
+ \item Check your ligands have the same, biologically correct binding pose
+        \item Make sure your \textlambda-scheduling is set appropriately
         \item Check if your ligands are discharging and vanishing correctly
@@
 \begin{itemize}
-   \item Your ligands should have the same, biologically correct binding pose
+   \item Check your ligands  have the same, biologically correct binding pose
         \item Make sure your λ-scheduling is set correctly
@@
 \begin{Checklists*}
-\begin{checklist}{How do I know which simulations are not reliable}
+\begin{checklist}{How do I know which simulations are unreliable?}
     \textbf{Situations suggesting your relative alchemical free energy calculations have not run properly (assuming absence of experimental affinities)}
@@
 
-\begin{checklist}{Why are they not reliable}
+\begin{checklist}{Why are they not reliable?}
     \textbf{Suggestions for finding out why your alchemical free energy calculations may not be reliable}
@@
     \item Check again whether dummy atoms were assigned correctly
-    \item Inspect the trajectories across the λ-schedule (particularly the endpoints)
-    \item Inspect the overlap matrices
+    \item Inspect the trajectories across the λ-schedule (particularly the endpoints) for problems described in the text
+    \item Inspect the overlap matrices for lack of overlap
 \end{itemize}
@@
 
-Declare any potentially conflicting interests here, whether or not they pose an actual conflict in your view.
-MK is employed by Cresset who commercially distribute a software for performing alchemical free energy calculations.
+Declare any potentially conflicting interests here, whether or not they pose an actual conflict in your view. JM is a current member of the Scientific Advisory Board of Cresset. 
+MK is employed by Cresset who commercially distribute a software for performing alchemical free energy calculations. MRS is a Open Science Fellow and consultant for Silicon Therapeutics.
 
@@
 AR acknowledges partial support from the Tri-Institutional Program in Computational Biology and Medicine and the Sloan Kettering Institute.
+HEBM acknowledges support from a Molecular Sciences Software Institute Investment Fellowship and Relay Therapeutics.
 

### 2020-08-05 paper/manuscript.tex
approaching final v0.1

@@
 
-\todototoc
-\listoftodos
+%\todototoc
+%\listoftodos
 
@@
 %%%%%%%%%%%%%%%%%%%%
-The current Overleaf document is \url{https://www.overleaf.com/5136551543vgjqbkzpmqkp}
+
 \section{What are alchemical free energy methods?}
@@
 This Best Practices guide focuses on providing a good starting point for new practitioners and a reference for experienced practitioners. 
- For this propose we provide a convenient checklist (Section~\ref{sec:checklist}) to help ensure all calculations comply with currently-understood best practices for alchemical simulation and analysis.
+ For this propose we provide a convenient checklist (Sec.~\ref{sec:checklist}) to help ensure all calculations comply with currently-understood best practices for alchemical simulation and analysis.
 Where the best practices are currently not certain, we highlight areas where further research is needed to identify an unambiguous recommendation.
@@
 
-As some of the theoretical background can seem daunting, we do, however, provide a guide to the essential theory behind alchemical free energy calculations in Section~\ref{sec:theory}.
-In the remainder of this paper, we will cover topics that are key to the preparation( Section~\ref{sec:prerequisites}), choice and use of correct protocols (Section~\ref{sec:simulation_protocol_choice}), and finally the best practices that should be used in the analysis of alchemical calculations (Section~\ref{sec:data_analysis}).
+As some of the theoretical background can seem daunting, we do, however, provide a guide to the essential theory behind alchemical free energy calculations in Sec.~\ref{sec:theory}.
+In the remainder of this paper, we will cover topics that are key to the preparation( Sec.~\ref{sec:prerequisites}), choice and use of correct protocols (Sec.~\ref{sec:simulation_protocol_choice}), and finally the best practices that should be used in the analysis of alchemical calculations (Sec.~\ref{sec:data_analysis}).
 Particular focus will be given to aspects of the molecular simulations which are unique to alchemical calculations---these include the calculation of transfer free energies (hydration free energies, partition coefficients, etc.), and binding free energies (absolute and relative).
@@
 \end{equation}
-The standard state concentration $c^\circ$ depends on the reference state, but it is usually set to $1\,\mathrm{mol}/\mathrm{L}$ assuming a constant pressure of $1\,\mathrm{atm}$ (see also Section~\ref{sec:standardstate-restraints}).
+The standard state concentration $c^\circ$ depends on the reference state, but it is usually set to $1\,\mathrm{mol}/\mathrm{L}$ assuming a constant pressure of $1\,\mathrm{atm}$ (see also Sec.~\ref{sec:standardstate-restraints}).
 Thus, the Gibbs free energy of binding $\Delta G_{\mathrm{bind}}$ is given by:
@@
 \paragraph{Simulating binding events is computationally expensive}
-While simulating binding events has been used to estimate binding affinities~\cite{jong2011determining,pan2017quantitative} or to get insights into the binding pathways and kinetics of receptor-ligand systems~\cite{teo2016adaptive,votapka2017seekr,doerr2014onthefly,plattner2015protein,dixon2018predicting}, the computational cost of these calculations is usually dominated by the rate of dissociation, which can be on the microsecond timescale even for millimolar binders~\cite{pan2017quantitative} and reaches the millisecond to second timescale for a typical drug~\cite{basavapathruni2012conformational,hyre2006cooperative}.
+While simulating binding events has been used to estimate binding affinities~\cite{jong2011determining,pan2017quantitative} or to get insights into the binding pathways and kinetics of receptor-ligand systems~\cite{teo2016adaptive,votapka2017seekr,doerr2014onthefly,plattner2015protein,dixon2018predicting}, the computational cost of these calculations is usually dominated by the rate of dissociation, which can be on the microsecond timescale even for millimolar binders~\cite{pan2017quantitative} and reaches the micosecond to second timescale for a typical drug~\cite{basavapathruni2012conformational,hyre2006cooperative}.
 Depending on system size and simulation settings, common molecular dynamics software packages can reach a few hundreds of ns/day using currently available high-end GPUs~\cite{eastman2017openmm,kutzner2019more}, making these type of calculations unappealing and irrelevant on a pharmaceutical drug discovery timescale.
@@
 Briefly, this is achieved by simulating a ``chimeric'' molecule composed of enough atoms to represent both $A$ and $B$.
-A subset of the energetic terms in $U(\vec{q};\vec{\lambda})$ is then modulated by $\vec{\lambda}$ so that at $\vec{\lambda}_A$, the atoms that form molecule $A$ are activated and those belonging exclusively to $B$ are non-interacting ``dummy atoms'', while the opposite occur at $\vec{\lambda}_B$ (see Section~\ref{sec:relative-fe-protocol} for details).
+A subset of the energetic terms in $U(\vec{q};\vec{\lambda})$ is then modulated by $\vec{\lambda}$ so that at $\vec{\lambda}_A$, the atoms that form molecule $A$ are activated and those belonging exclusively to $B$ are non-interacting ``dummy atoms'', while the opposite occur at $\vec{\lambda}_B$ (see Sec.~\ref{sec:relative-fe-protocol} for details).
 
@@
 \end{equation}
-While it is generally not feasible to compute the two partition functions $Z(\vec{\lambda})$, several estimators have been devised to robustly estimate the ratio of partition functions in Eq.~\ref{eq:delta-G-env} (see Section~\ref{subsec:estimators}) from a set of conformations usually collected with MD simulations from the thermodynamic states defined at $\vec{\lambda}_A$ and $\vec{\lambda}_B$ and intermediates thereof.
+While it is generally not feasible to compute the two partition functions $Z(\vec{\lambda})$, several estimators have been devised to robustly estimate the ratio of partition functions in Eq.~\ref{eq:delta-G-env} (see Sec.~\ref{subsec:estimators}) from a set of conformations usually collected with MD simulations from the thermodynamic states defined at $\vec{\lambda}_A$ and $\vec{\lambda}_B$ and intermediates thereof.
 
@@
 While it is theoretically possible to estimate the ratio of partition functions from samples collected only at states $\vec{\lambda}_A$ and $\vec{\lambda}_B$, the efficiency of the free energy estimators rapidly decreases as the phase-space overlap between the two states also decreases~\cite{wu2005phasespaceb,wu2005phasespacec}.
-Roughly, the phase-space overlap between two thermodynamic states measures the degree to which high-probability conformations (i.e., those with very negative potential energy) in one state are also high-probability conformations in the other state (see Section~\ref{sec:are-they-good} and Fig.~\ref{fig:fig_what_is_lambda}).
+Roughly, the phase-space overlap between two thermodynamic states measures the degree to which high-probability conformations (i.e., those with very negative potential energy) in one state are also high-probability conformations in the other state (see Sec.~\ref{sec:are-they-good} and Fig.~\ref{fig:fig_what_is_lambda}).
 
@@
 While this strategy usually results in sampling thermodynamic states whose Boltzmann distributions are very similar, thus collecting information that is to some degree redundant, some estimators such as the Multistate Bennett acceptance ratio (MBAR)~\cite{shirts2008statisticallya} can exploit similarities between states to improve the precision of the estimates.
-This is achieved by using the conformations sampled at all alchemical states $\{\vec{\lambda}_k \}$ to compute the free energy difference $\Delta f(\vec{\lambda}_i, \vec{\lambda}_{j})$ between any pair of states $i,j$ (see Section~\ref{subsec:estimators}).
+This is achieved by using the conformations sampled at all alchemical states $\{\vec{\lambda}_k \}$ to compute the free energy difference $\Delta f(\vec{\lambda}_i, \vec{\lambda}_{j})$ between any pair of states $i,j$ (see Sec.~\ref{subsec:estimators}).
 
@@
 Current alchemical free energy calculations involving small molecules seem to achieve, in favorable cases, root mean square (RMS) errors around 1-2 kcal/mol depending on force field, system, and a variety of other factors such as simulation time, sampling method, and whether the calculations employed are absolute or relative.
-A small selection of example datasets and case studies can be found in Section~\ref{sec:benchmark} at the end of this document.
+A small selection of example datasets and case studies can be found in Sec.~\ref{sec:benchmark} at the end of this document.
 However, the domain of applicability is a significant concern~\cite{sherborne2016collaborating, cournia2017relative}, especially for relative calculations, which typically require a high quality and usually experimental bound structure of a closely
@@
 Furthermore, perturbations for the same chemotype in different pockets of the BACE enzyme gave varied errors~\cite{keranen2017acylguanidine}.
-Here the errors refer to the difference in $\Delta G$ after doing a linear regression fit to observed experimental binding free energies.
+Here the errors refer to the difference in $\Delta G$ derived by fitting the $\Delta \Delta G$'s to the known experimental binding free energies~\cite{wang2015accurate}.
 This prompts us to consider another important aspect.
 It is important to be clear on what error to report: $\Delta G$ after shifting by a constant to minimize the RMSE, unshifted $\Delta G$, $\Delta \Delta G$ of computed edges, or $\Delta \Delta G$ of all edges.
-(See recommendations for reporting best practices, Sec.~\ref{sec:plot_data}) In general, the computed edge $\Delta \Delta G$ RMSEs significantly underestimates the error of all edge RMSEs: \url{https://github.com/jchodera/jacs-dataset-analysis}.
+(See recommendations for reporting best practices, Sec.~\ref{sec:plot_data}.) Additionally, as it is possible to perform calculations on a set of ligands using different pairwise comparisons of molecules, the performance of the method may be biased based on which pairs of comparisons are performed.
+Additionally, it is possible that the error associated to the relative free energy between a two ligands that was not directly computed (however can be deduced using thermodynamic paths involving other ligands) will likely be more uncertain \url{https://github.com/jchodera/jacs-dataset-analysis}.
 Given the need to understand the performance of the system with alchemical free energy calculations, we recommend that retrospective studies for a particular target and a particular chemical series be performed for each application case.
@@
 For more complex use cases such as protein-ligand binding free energies the repeatability is often substantially worse~\cite{rizzi2019sampl6}.
-A good practice is to perform two or three runs of the same perturbation to assess reproducibility with a given protocol.
+A good practice is to perform two or three runs of the same perturbation to assess precision with a given protocol.
 The sample standard deviation will give a crude estimate of the reliability of the estimates, and whether the precision is sufficient for the problem at hand.
@@
 
-For illustrative purposes, assume that the actual changes in the
-binding free energies for a set of new molecules also follows a normal
-distribution, that the standard deviation in the changes is $RT\ln 5$
+For illustrative purposes, consider a proposed set of new molecules, and assume that the changes proposed in these molecules yield a set of relative binding free energies that follow a normal
+distribution. That is, assume that the standard deviation in the relative binding free enrgies for the changes represented is $RT\ln 5$
 (corresponding to a 5-fold change in the binding affinities), and that
@@
 idea of feasibility and accuracy of the calculations for the
-proposed target.
+proposed target~\cite{schindler2020largescale}.
 
@@
 \subsection{Capturing experimental conditions}
+\label{subsec:exp_condition}
 The calculations aim to capture the alchemical change from one ligand to another as accurately as possible.
@@
 Hysteresis, within cycles in the perturbation network or between forward and backward perturbations can be checked~\cite{wang2013modeling} to indicate problematic perturbations involved in cycles connecting many compounds (See also Secs.~\ref{sec:relative-fe-protocol} and~\ref{sec:are-they-good}).
-Once synthesis and testing of compounds is complete it would be normal to look back at how the calculations performed.
-In this regard it is important to consider upfront the issue of selection bias.
+Once synthesis and testing of compounds is complete a standard strategy is to look back at how the calculations performed.
+In this regard it is important to consider the issue of selection bias upfront.
 It is tempting to only synthesize the compounds predicted to be most active, thus a narrow range of calculated activity is tested that imposes limits on the statistical assessment of performance, ideally example molecules from across the range of predicted activity can be assessed or corrections can be applied based on previous recommendations~\cite{abel2017critical}.
-For a more detailed discussion on checking the robustness of your alchemcial free energy calculation see also Section~\ref{sec:are-they-good}.
+For a more detailed discussion on checking the robustness of your alchemcial free energy calculation see also Sec.~\ref{sec:are-they-good}.
 
@@
 An interesting cost benefit analysis has shown the value of activity prediction, see discussion above and articles such as~\cite{mobley2012perspective}.
-From a drug discovery point of view, alchemical calculations are expanding their domain of applicability, and there are reports of success using homology models and GPCRs for instance, as well as enabling charge change and scaffold hopping, but these systems are undoubtedly more difficult.
-In the meantime, the use cases are expanding to resistance prediction, disease mutation prediction, solubility prediction – an exciting future for alchemical calculations.
+From a drug discovery point of view, alchemical calculations are expanding their domain of applicability, and there are reports of success using homology models~\cite{cappel2016relative} and GPCRs~\cite{deflorian2020accurate,lenselink2016predicting} for instance, as well as enabling charge change and scaffold hopping~\cite{chen2018accurate, wang2017accurate}, but these systems are undoubtedly more difficult.
+In the meantime, the use cases are expanding to resistance prediction, selectivity prediction , solubility prediction – an exciting future for alchemical calculations~\cite{hauser2018predicting, albanese2020structure, mondal2019free}.
 
@@
 \label{sec:prerequisites}
-Alchemical free energy protocols as discussed below (Section~\ref{sec:simulation_protocol_choice}) are defined for a specific type of free energy calculation, i.e.
+Alchemical free energy protocols as discussed below (Sec.~\ref{sec:simulation_protocol_choice}) are defined for a specific type of free energy calculation, i.e.
 a free energy of binding or a free energy of hydration.
@@
 \paragraph{Are there good X-ray structures available?}
-As with any simulation, care should be taken in selecting available X-ray structures in the Protein DataBank.
+As with any simulation, care should be taken in selecting available X-ray structures in the Protein DataBank~\cite{berman2003announcing}.
 In some cases it may be wise to choose multiple starting structures to account for variability in receptor conformations as well as the accuracy of available X-ray structures.
@@
 
-Many free energy calculations focus on a congeneric series of ligands, which can make these calculations suitable for relative free energy protocols (see Section~\ref{sec:simulation_protocol_choice}).
+Many free energy calculations focus on a congeneric series of ligands, which can make these calculations suitable for relative free energy protocols (see Sec.~\ref{sec:simulation_protocol_choice}).
 For relative calculations, some care has to be taken selecting binding poses for these ligands.
@@
 \paragraph{Are you prepared to deal with any binding mode challenges?}
-Generally, binding modes within congeneric series are conserved~\cite{wacker2010conserved}, however, exceptions exist~\cite{brandt2011congeneric,nazare2005probing}, as discussed in more detail in Section~\ref{sec:multiple_binding_modes}.
+Generally, binding modes within congeneric series are conserved~\cite{wacker2010conserved}, however, exceptions exist~\cite{brandt2011congeneric,nazare2005probing}, as discussed in more detail in Sec.~\ref{sec:multiple_binding_modes}.
 Certain functional groups may be particularly prone to this due to symmetries or near symmetries.
-One such issue involves a 180 degree flip in the dihedral angle of an aromatic ring leading to a different spatial position of ortho- or meta- substituents that otherwise should overlap within a series.
+One such issue involves a 180 degree flip in the dihedral angle of an aromatic ring, or five-membered ring leading to a different spatial position of ortho- or meta- substituents that otherwise should overlap within a series.
 The 180 degree flip of the ring may not occur enough during simulations (due to steric obstructions) to overcome bias due to the starting configuration.
@@
 Other tools such as WaterMap or open source equivalents (SSTMap, GIST, and others) can be used to define water structure for systems with no experimental evidence of water sites~\cite{wang2011ligand}.
-Well known protein systems with water mediated ligand interactions are for example: HSP90 which formed part of the D3R grand challenge 2015~\cite{mey2016blinded}, A2A~\cite{brucemacdonald2018ligand}, MUP~\cite{ross2015water}, GPCRs, and others~\cite{michel2009energetics}.
+Well known protein systems with water mediated ligand interactions are for example: HSP90 which formed part of the D3R grand challenge 2015~\cite{mey2016blinded}, A2A~\cite{brucemacdonald2018ligand}, MUP~\cite{ross2015water}, ~\cite{deflorian2020accurate}, and others~\cite{michel2009energetics}.
 
@@
 Care should be taken when preparing ligands and proteins to match the pH of the experimental assay, if known.
-As mentioned above in Section 5.1, the pH of the assay can differ from neutral pH and will determine the protonation states of the proteins and ligands.
-Since the pKa of reference amino acid sidechain residues is known, but can vary in the protein environment, many different tools have emerged for predicting sidechain pKa in proteins, such as the H++ server, ProPKa, APBS, and Maestro ~\cite{anandakrishnan2012automating, 2020schrodinger}.
+As mentioned above in Sec.~\ref{subsec:exp_condition}, the pH of the assay can differ from neutral pH and will determine the protonation states of the proteins and ligands.
+Since the pKa of reference amino acid sidechain residues is known, but can vary in the protein environment, many different tools have emerged for predicting sidechain pKa in proteins, such as the H++ server, ProPKa, APBS, and Maestro ~\cite{anandakrishnan2012automating, sondergaard2011improved, jurrus2018improving, 2020schrodinger}.
 Strongly acidic (Glu, Asp) or basic (Arg, Lys) sidechains can reliably predicted to be ionised, but care is still needed as the local environment can modify expected ionization states (for instance the catalytic Asp dyad in proteases).
@@
 There are many different available tools for this purpose, but common choices may be propKa~\cite{olsson2011propka3,sondergaard2011improved}, Chemicalize (\url{https://chemicalize.com/welcome}), or Maestro~\cite{2020schrodinger}.
-Still, accurate pKa prediction for small molecules remains a challenging problem, even with dedicated tools.
-While often it can be assumed that the protonation state of a ligand and protein will remain the same foras a ligand binds, some care needs to be taken with systems where the protonation state may change upon binding~\cite{onufriev2013protonation}.
+Still, accurate pKa prediction for small molecules remains a challenging problem, even with dedicated tools~\cite{isik2018pka}.
+While often it can be assumed that the protonation state of a ligand and protein will remain the same as a ligand binds, some care needs to be taken with systems where the protonation state may change upon binding~\cite{onufriev2013protonation}.
 BACE, for example, famously undergoes a protonation state change on ligand binding.
@@
 And, while sampling problems might seem to be a non-issue for small molecules, this is not always the case; e.g.
-even the hydroxyl orientation on neutral carboxylic acids can occasionally pose a challenge~\cite{}.
+even the hydroxyl orientation on neutral carboxylic acids can occasionally pose a challenge~\cite{klimovich2010predicting, lim2019assessing}.
 
@@
 In contrast, in relative binding free energy calculations, we compare the binding free energy of two related inhibitors to determine the potency difference.
-%
+
 \subsection{Absolute and relative free energy calculations have some differences}
@@
 We will consider protocol differences before treating the common elements.
-%
+
 \subsubsection{Choices unique to relative free energy calculations}
@@
 In contrast, in a dual topology alchemical free energy calculation, no atoms are allowed to change type~\cite{shirts2012best}.
-This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxy methyl group, then passing through an intermediate state where some atoms are partially interacting--- particularly, those atoms which are becoming dummy atoms or ceasing to be dummy atoms (this state may or may not be well-defined~\cite{mobley2014blind}).
+This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxy methyl group, then passing through an intermediate state where some atoms are partially interacting--- particularly, those atoms which are becoming dummy atoms or ceasing to be dummy atoms~\cite{mobley2014blind}.
 The transformation finally culminates in a state where benzyl alcohol is present along with the additional dummy atom which was previously a corresponding hydrogen of the benzene.
@@
 
-Hybrid topology calculations have seen much less use but essentially consist of two absolute free energy calculations in opposite directions at the same time, turning one molecule's interactions with the environment off, while turning the other molecule's interaction on~\cite{jiang2019computing}.
+Hybrid topology calculations have seen much less use but essentially consist of two absolute free energy calculations in opposite directions at the same time, turning one molecule's interactions with the environment off, while turning the other molecule's interaction on~\cite{jiang2019computing, rocklin2013separated}.
 
@@
 Please make sure to check what approach is used with your software package of choice, or whether it supports your choice of approach (GROMACS, for example, supports both).
-To our knowledge efficiency differences have not been thoroughly explored, though conventional wisdom suggests that fewer dummy atoms are better, as introducing or removing atomic sites is usually more difficult, requireing more intermediate steps ~\cite{liu2013lead, mobley2012perspective}.
+To our knowledge efficiency differences have not been thoroughly explored, though conventional wisdom suggests that fewer dummy atoms are better, as introducing or removing atomic sites is usually more difficult, requiring more intermediate steps ~\cite{liu2013lead, mobley2012perspective}.
 
@@
 Based on the input ligand series, a perturbation map or network can be planned.
-Recent heuristics have shown the more connected the perturbation network the better, however, there is a way to optimize network structure while minimizing the number of perturbations that need to be computed reducing the resulting computational cost~\cite{yang2020optimal}.
+Recent heuristics have shown the more connected the perturbation network the better, however, there is a way to optimize network structure while minimizing the number of perturbations that need to be computed reducing the resulting computational cost~\cite{yang2020optimal,xu2019optimal}.
 Sometimes the introduction of intermediates that are not part of the original congeneric series are essential to avoid ring breaking, or deal with perturbations that would otherwise result in large numbers of atoms being inserted or deleted.
@@
 The addition of a restraint limits the volume available to the non-interacting ligand, thus speeding the convergence of the sampling.
-In addition, when enhanced sampling methods such as Hamiltonian $\lambda$ exchange are used (see Section~\ref{sec:sampling_schemes}), the use of a restraint is typically necessary as it keeps the ligand in the binding site in the interacting state (see also Section~\ref{sec:weak-binders}) and generally reduce the round-trip time of replicas.
+In addition, when enhanced sampling methods such as Hamiltonian $\lambda$ exchange are used (see Sec.~\ref{sec:sampling_schemes}), the use of a restraint is typically necessary as it keeps the ligand in the binding site in the interacting state (see also Sec.~\ref{sec:weak-binders}) and generally reduce the round-trip time of replicas.
 When restraint are employed, the values of $V_L$ and $\xi_L$ are restraint-dependent, but for commonly employed restraints, these can be usually easily computed analytically or numerically by solving the relevant integral.
@@
 \subsubsection{Handling weak binders and high dissociation rates}\label{sec:weak-binders}
-In binding free energy calculations, only the conformations in which the receptor and ligand form a bound complex should be sampled from the bound states (Section~\ref{sec:theory}).
+In binding free energy calculations, only the conformations in which the receptor and ligand form a bound complex should be sampled from the bound states (Sec.~\ref{sec:theory}).
+Determining what the bound states actually are can be challenging for weakly bound ligands.
+For tightly bound ligands, virtually all reasonable definitions of the bound state will lead to be equivalent free energies, since the partition function will be dominated by a relatively small number of low-energy poses.
+For weak binders, this simplification breaks down.
+In fact, the correct bound state may depend on the type of experiment performed.
+For example, isothermal titration calorimetry (ITC) or surface plasmon resonance (SPR) measurements effectively define a binding state that includes all ligand comformations that are complexed with the protein, regardless of where on the protein they bind.
+In contrast, fluorescence polarization competition assays measure binding to only a single location, where the ligand of interest displaces a competing binder.
+Therefore, care must be taken to ensure that a reasonable definition of the binding site is used~\cite{wang2013identifying}.
 In absolute calculations, this applies to the fully interacting state in the complex leg of the thermodynamic cycle (top-right state in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}), while in relative calculations this must be true at both end states of the complex leg (top- and bottom-right states in Fig.~\ref{fig:fig_binding_thermodynamic_cycle}).
@@
 
-In absolute calculations, this restraint can substitute or be added to the restraint used to handle the standard state correction (Section~\ref{sec:standardstate-restraints}).
+In absolute calculations, this restraint can substitute or be added to the restraint used to handle the standard state correction (Sec.~\ref{sec:standardstate-restraints}).
 In the latter case, to compute the standard state correction analytically, the bound-state restraint must be turned off in the decoupled state.
@@
 If the net charge of the system changes as the alchemical variable changes during the  calculation, this can pose major challenges.
-Specifically, finite-size effects can introduce significant charge-dependent artifacts into computed binding free energies~\cite{}MISSING citation, in part because typical schemes for long-range electrostatics (including PME and reaction field) do not handle free energy contributions from such changes effectively or as they would be handled in a hypothetical macroscopic bulk solution~\cite{}.
+Specifically, finite-size effects can introduce significant charge-dependent artifacts into computed binding free energies, in part because typical schemes for long-range electrostatics (including PME and reaction field) do not handle free energy contributions from such changes effectively or as they would be handled in a hypothetical macroscopic bulk solution~\cite{lin2014overview, ohlknecht2020correcting, rocklin2013calculating}.
 
@@
 This is sometimes referred to as an "alchemical ion" approach for dealing with the needed charge change, and is also employed by the Yank free energy package~\cite{wang2013identifying}.
-Charge corrections have also been explored, and are potentially a viable solution to this problem~\cite{mey2018impact} where artifacts introduced by finite-size effects are corrected numerically~\cite{chen2018accurate}.
+Charge corrections have also been explored, and are potentially a viable solution to this problem~\cite{mey2018impact} where artifacts introduced by finite-size effects are corrected numerically~\cite{chen2018accurate, ohlknecht2020correcting}.
 However, application of such corrections typically remains less common than the use of a co-alchemical ion.
@@
 Intermediates states are required for a sufficient overlap in potential energies to estimate a free energy difference between $\vec{\lambda}=0$ and $\vec{\lambda}=1$.
-    Soft-core potentials provide one of the most efficient families of intermediate pathways, with a $\vec{\lambda}$ dependence illustrated in (\textbf{C}) and a potential energy dependence shown in (\textbf{D}).
+    Soft-core potentials provide one of the most efficient families of intermediate pathways, with a $\vec{\lambda}$ dependence.
+In (\textbf{C}) the potential energy surface is coloured according to $\vec{\lambda}$ with blue being $\vec{\lambda}=0$ and $\vec{\lambda}=1$ orange.
+In (\textbf{D}) the potential is coloured according to the potential energy.
 Note how as $\vec{\lambda}$ approaches 0, the energy smoothly approaches zero at all $r$, a necessary requirement for efficient and stable calculations.
@@
 The key thing to watch out for is whether the total potential energy, and therefore the intermediate ensembles sampled, change smoothly from beginning to end.
-These problems can be diagnosed by noticing lack of configuration space overlap between different simulations (see section~\ref{sec:are-they-good}).
+These problems can be diagnosed by noticing lack of configuration space overlap between different simulations (see Sec.~\ref{sec:are-they-good}).
 
@@
 Some schemes to adaptively optimize the spacing of intermediate states based on initial exploratory simulations have been proposed~\cite{hayes2017adaptive}.
-For molecules changing in dense solvent, then the best path is roughly independent of molecule size and shape, so what works for one molecular transformation is likely to be relatively efficient for another~\cite{monroe2004converging}.
+For molecules changing in dense solvent, then the best path is roughly independent of molecule size and shape, so what works for one molecular transformation is likely to be relatively efficient for another~\cite{monroe2014converging}.
 
@@
 However, even subtly induced changes in protein binding configuration will require more sampling or cause local convergence to a free energy estimate that has high error.
-Many enhanced sampling techniques have been proposed to try and overcome barriers in free energy calculations and mitigate convergence to erroneous estimates~\cite{xx}. 
 The confidence a user should have in a free energy estimate is significantly improved when both the uncertainty of the free energy estimate is low, and when other observables have reached a convergence.
@@
     \centering
-    \includegraphics[width=0.95\linewidth]{paper/figures/fig_pmf/pmf.png}
+    \includegraphics[width=0.8\linewidth]{paper/figures/fig_pmf/pmf.png}
     \caption{{\bf Potential of mean force with respect to $\vec{\lambda}$ for TI and MBAR}
@@
 This is not necessarily always the case~\cite{kaus2015how}, as reviewed elsewhere~\cite{mobley2009binding} and in some cases unexpected binding mode changes can be the origin of apparent non-additivity in structure-activity relationships~\cite{calabro2016elucidation}.
-Binding modes also tend to be particularly variable in the case of fragments, which often may have multiple relevant binding modes~\cite{}.
+Binding modes also tend to be particularly variable in the case of fragments, which often may have multiple relevant binding modes~\cite{steinbrecher2015accurate}.
 
@@
 Within absolute binding free energy calculations, multiple potential binding modes can be handled by two main strategies: Consider each binding mode separately (a separation of states strategy) or sample all binding modes within a single simulation~\cite{mobley2012perspective}.
-This couples to the choice of restraints selected, as some restraints will allow transitions between binding modes and even binding sites (Section~\ref{sec:standardstate-restraints}), and others do not.
+This couples to the choice of restraints selected, as some restraints will allow transitions between binding modes and even binding sites (Sec.~\ref{sec:standardstate-restraints}), and others do not.
 
@@
 One can show that if the standard deviation of the difference $\Delta u(\vec{q}) = u_1(\vec{q})-u_2(\vec{q})$ over all sampled $\vec{q}$ is large (which in this case, means only several times $k_BT)$, then very few samples contribute to the average, and the answer will be both biased and extremely noisy~\cite{lelievre2010free}.
-Essentially, the method is dominated by contributions of rare snapshots\cite{jarzynski2006rare, wu2005phasespaceb, wu2005phasespacec}.
+Essentially, the method is dominated by contributions of rare snapshots~\cite{jarzynski2006rare, wu2005phasespaceb, wu2005phasespacec}.
 
@@
 The Bennett acceptance ratio is only defined between two states.
-Usually, the endpoints of interest in a free energy calculation are sufficiently different that we will need a chain of states that gradually change the potential energy function from $u_0$ to $u_1$, as discussed in Section~\ref{sec:important_path}.
+Usually, the endpoints of interest in a free energy calculation are sufficiently different that we will need a chain of states that gradually change the potential energy function from $u_0$ to $u_1$, as discussed in Sec.~\ref{sec:important_path}.
 You can simply carry out BAR between each pair of states $\Delta f_{1 \rightarrow N} = \Delta {f_{1\rightarrow 2}} + \Delta {f_{2\rightarrow 3}} +  \ldots + \Delta f_{N-1\rightarrow N}$.
@@
 \[ \Delta f \approx \sum_{k=1}^{K} w_k
-\expect{\frac{du(\vec{\lambda},\vec{q})}{d\vec{\lambda}}}_{k} \], where the weights $w_k$ correspond to a particular choice of numerical integration.
+\expect{\frac{du(\vec{\lambda},\vec{q})}{d\vec{\lambda}}}_{k}, \] where the weights $w_k$ correspond to a particular choice of numerical integration.
 Researchers have tried a large number of different integration schemes~\cite{resat1993studies,jorge2010effect,shyu2009reducing}.
@@
 
-The computation of free energies using TI (section~\ref{subsec:estimators}) is straightforward and the trapezoidal rule is often recommended since it allows unequal spacing of $\vec{\lambda}$ states, which is required to minimize the variance in the free energy estimate, but in principle any good numerical integration method can be used.
+The computation of free energies using TI (Sec.n~\ref{subsec:estimators}) is straightforward and the trapezoidal rule is often recommended since it allows unequal spacing of $\vec{\lambda}$ states, which is required to minimize the variance in the free energy estimate, but in principle any good numerical integration method can be used.
 The determination of regions of high curvature when estimating the integral is helpful to determine regions of phase space where more sampling and/or more $\vec{\lambda}$ states are necessary to obtain the best approximation of the integral.
@@
 
-Compared with TI, the MBAR method (section~\ref{subsec:estimators}) discussed above provides uncertainty estimation directly from solving a set of linear equations to compute the variances between all states.
+Compared with TI, the MBAR method (Sec.~\ref{subsec:estimators}) discussed above provides uncertainty estimation directly from solving a set of linear equations to compute the variances between all states.
 The number of states and amount of sampling should be optimized to minimize the uncertainty in the MBAR free energy estimate, while balancing other key considerations such as computational expense.
@@
 While variance can be decreased by increasing simulation length, if the overlap between states is known to be poor, increasing the number of $\vec{\lambda}$ values, or adjusting the spacing of those values to better cover regions of poor overlap will likely provide a larger immediate impact.
-Different approaches are described in section~\ref{sec:simulation_protocol_choice} and more details can be found in the literature~\cite{dakka2018concurrent, hahn2019alchemical}.
+Different approaches are described in Sec.~\ref{sec:simulation_protocol_choice} and more details can be found in the literature~\cite{dakka2018concurrent, hahn2019alchemical}.
 
@@
 	\item pymbar: MBAR implementation, but have to roll your own analysis wrapper      
-	\url{https://github.com/choderalab/pymbar}~\cite{shirts2008statisticallya}.
-	\item Beryllium: Standardising alchemical free energy analysis \url{https://github.com/openforcefield/beryllium}
+	\url{https://github.com/choderalab/pymbar} \cite{shirts2008statisticallya}.
+	\item Arsenic: Standardising alchemical free energy analysis \url{https://github.com/openforcefield/Arsenic}
 	\item Free Energy Workflows: Sire-specific free energy map analysis using weighted path averages \url{https://github.com/michellab/freenrgworkflows}.
@@
 These include cyclodextrins, the Cytochrome C peroxidase (CCP) protein model binding site, thrombin and bromodomains as well as solvation benchmark sets~\cite{paliwal2011benchmark}.
-Please refere to ~\ref{tab:benchmarks}, for a small overview of datasets, what forcefields they used, and what the original study was it came from.
+Please refer to table~\ref{tab:benchmarks}, for a small overview of datasets, what forcefields they used, and what the original study was it came from.
 
@@
         \begin{itemize}
-                \item Standard error (\textsigma) should be \textless1 kcal·mol$^{-1}$ 
-    \item Simulated systems have not converged 
+                \item Standard error (\textsigma) should not be \textgreater1 kcal·mol$^{-1}$ 
+    \item Simulated systems have not converged - this
     \newline\newline\textit{Relative:}
@@
 \textbf{ASJSM}: Coordinated the document, contributed to most sections, and co-designed Figs.~\ref{fig:fig_binding_thermodynamic_cycle},~\ref{fig:fig_topology},~\ref{fig:fig_mcss},~\ref{fig:fig_types_of_networks},~\ref{fig:fig_absolute_thermodynamic_cycle},~\ref{fig:scatterplot_analysis},and created Figs.~\ref{fig:fig_what_is_lambda},~\ref{fig:fig_what_is_alchemy},~\ref{fig:overlap},~\ref{fig:pmf} and replotted~\ref{fig:automatic-equilibration-detection} and~\ref{fig:fig_types_of_networks}.\\
-\textbf{JS}: Created Figs~ \ref{fig:fig_what_is_alchemy},~\ref{fig:fig_binding_thermodynamic_cycle},~\ref{fig:fig_topology},~\ref{fig:fig_mcss},~\ref{fig:fig_absolute_thermodynamic_cycle},~\ref{fig:scatterplot_analysis}, and an initial draft of~\ref{fig:fig_types_of_networks}.
+\textbf{JS}: Created Figs.~ \ref{fig:fig_what_is_alchemy},~\ref{fig:fig_binding_thermodynamic_cycle},~\ref{fig:fig_topology},~\ref{fig:fig_mcss},~\ref{fig:fig_absolute_thermodynamic_cycle},~\ref{fig:scatterplot_analysis}, and an initial draft of~\ref{fig:fig_types_of_networks}.
 Wrote Sec.~\ref{sec:plot_data}, the checklist Sec.~\ref{sec:checklist}, and contributed to general formatting discussions and editing.\\
@@
 HEBM acknowledges support from a Molecular Sciences Software Institute Investment Fellowship and Relay Therapeutics.
+JDC is a current member of the Scientific Advisory Board of OpenEye Scientific Software and a consultant to Foresite Laboratories.
+A complete funding history for the Chodera lab can be found at http://choderalab.org/funding.
+DLM appreciates financial support from the National Institutes of Health (1R01GM108889-01, 1R01GM124270-01A1 and R01GM132386), and the National Science Foundation (CHE 1352608).
 

### 2020-08-05 paper/manuscript.tex
Submission version.

@@
 
-\newcommand{\versionnumber}{0.1} % you should update the minor version number in preprints and major version number of submissions.
+\newcommand{\versionnumber}{1.0} % you should update the minor version number in preprints and major version number of submissions.
 \newcommand{\githubrepository}{\url{https://github.com/michellab/alchemical-best-practices}} %this should be the main github repository for this article
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{Best Practices for Alchemical Free Energy Calculations: v\versionnumber}
+\title{Best Practices for Alchemical Free Energy Calculations [Article v \versionnumber]}
 \author[1*]{Antonia S. J. S. Mey}
@@
 \author[1,10]{Maximilian Kuhn}
-\author[1*]{Julien Michel}
+\author[1]{Julien Michel}
 \author[3*]{David L. Mobley}
@@
 %
+\corr{antonia.mey@ed.ac.uk}{ASJSM}
 \corr{john.chodera@choderalab.org}{JDC}
 \corr{dmobley@mobleylab.org}{DLM}
-\corr{antonia.mey@ed.ac.uk}{ASJSM}
-\corr{mail@julienmichel.net}{JM}
 \corr{michael.shirts@colorado.edu}{MRS}
-%
+
+\orcid{Antonia S. J. S. Mey}{0000-0001-7512-5252}
+\orcid{Bryce Allen}{0000-0002-0804-8127}
+\orcid{Hannah E. Bruce Macdonald}{0000-0002-5562-6866}
+\orcid{John D. Chodera}{0000-0003-0542-119X}
+\orcid{Maximilian Kuhn}{0000-0002-2811-3934}
+\orcid{Julien Michel}{0000-0003-0360-1760}
+\orcid{David L. Mobley}{0000-0002-1083-5533}
+\orcid{Levi N. Naden}{0000-0002-3692-5027}
+\orcid{Samarjeet Prasad}{0000-0001-8320-6482}
+\orcid{Andrea Rizzi}{0000-0001-7693-2013}
+\orcid{Jenke Scheen}{0000-0001-9781-0445}
+\orcid{Michael R. Shirts}{0000-0003-3249-1097}
+\orcid{Gary Tresadern}{0000-0002-4801-1644}
+\orcid{Huafeng Xu}{0000-0001-5447-0452}
+
 \blurb{This LiveCoMS document is maintained online on GitHub at \githubrepository; to provide feedback, suggestions, or help improve it, please visit the GitHub repository and participate via the issue tracker.}
@@
 As mentioned above in Sec.~\ref{subsec:exp_condition}, the pH of the assay can differ from neutral pH and will determine the protonation states of the proteins and ligands.
-Since the pKa of reference amino acid sidechain residues is known, but can vary in the protein environment, many different tools have emerged for predicting sidechain pKa in proteins, such as the H++ server, ProPKa, APBS, and Maestro ~\cite{anandakrishnan2012automating, sondergaard2011improved, jurrus2018improving, 2020schrodinger}.
+Since the pKa of reference amino acid sidechain residues is known, but can vary in the protein environment, many different tools have emerged for predicting sidechain pKa in proteins, such as the H++ server, ProPKa, APBS, and Maestro ~\cite{anandakrishnan2012automating, sondergaard2011improved, jurrus2018improvements, 2020schrodinger}.
 Strongly acidic (Glu, Asp) or basic (Arg, Lys) sidechains can reliably predicted to be ionised, but care is still needed as the local environment can modify expected ionization states (for instance the catalytic Asp dyad in proteases).
@@
 \begin{equation}\label{eq:restraint-correction}
-    \Delta G^{\circ}_{\mathrm{restr}} = -k_BT ~ \ln \left( \frac{c^{\circ}}{V_L} \right) -k_BT ~ \ln \left( \frac{\xi_L}{8 \pi^2} \right)
+    \Delta G^{\circ}_{\mathrm{restr}} = -k_BT ~ \ln \left( \frac{c^{\circ}}{V_L} \right) -k_BT ~ \ln \left( \frac{\xi_L}{8 \pi^2} \right),
 \end{equation}
@@
 \paragraph{Handling multiple potential binding modes within relative calculations.}
-Multiple potential binding modes pose particular problems for relative free energy calculations, as having multiple starting structures for these calculations could yield substantially different calculated relative binding free energies for the same transformation due to kinetic trapping, and, without additional information (specifically, the free energy of binding mode interconversion or, equivalently, the relative populations of different binding modes) it becomes impossible to sort out which of the multiple answers is in fact the correct relative binding free energy~\cite{}.
+Multiple potential binding modes pose particular problems for relative free energy calculations, as having multiple starting structures for these calculations could yield substantially different calculated relative binding free energies for the same transformation due to kinetic trapping, and, without additional information (specifically, the free energy of binding mode interconversion or, equivalently, the relative populations of different binding modes) it becomes impossible to sort out which of the multiple answers is in fact the correct relative binding free energy.
 
@@
 In a drug discovery context, this type of data depiction may suggest the reliability of alchemical FE predictions in the project, and can give an idea of how trustworthy predictions can be for synthesis ideas.
+It is also recommended to included experimental error bars in all plots.
 
@@
 
-\begin{checklist}{Running absolute simultions}
+\begin{checklist}{Running absolute simulations}
         \textbf{Steps to running your absolute alchemical free energy calculations}
@@
         \item Make sure your \textlambda-scheduling is set appropriately
-        \item Check if your ligands are discharging and vanishing correctly
+        \item Check if your ligands are discharging and decoupling correctly
         \item Set up your restraints correctly
@@
         \item Check whether dummy atoms were assigned correctly
-        \item Make sure you subsample the data in your free energy estimation protocol
+        \item Consider subsampling the data in your free energy estimation protocol
         \item Apply the 
@@
                 \item Standard error (\textsigma) should not be \textgreater1 kcal·mol$^{-1}$ 
-    \item Simulated systems have not converged - this
+    \item Simulated systems have not converged - trajectories should be manually checked for consistency; other methods such as generating RMSD plots are also recommended
     \newline\newline\textit{Relative:}
     \item If you observe hysteresis in perturbations and incorrect cycle closures
-    \item Energy differences \textgreater15 kcal$\cdot$mol$^{-1}$  are unreliable
+    \item Energy differences \textgreater$\sim$15 kcal$\cdot$mol$^{-1}$  are likely unreliable
     \newline\newline\textit{Absolute:}
-    \item Energies below -15 kcal$\cdot$mol$^{-1}$  are unreliable
-    \item The ligand has not sampled most of the intended region after the vanishing step
-    \item The ligand is drifting out of the intended region after the vanishing step
+    \item Energies \textless$\sim$-15 kcal$\cdot$mol$^{-1}$  are likely unreliable
+    \item The ligand has not sampled most of the intended region after the decoupling step
+    \item The ligand is drifting out of the intended region after the decoupling step
         \end{itemize}
@@
     \item Compute both correlation and ranking coefficients and ranking statistics (e.g. r, \textrho, MUE and \texttau)
-    \item You should be able to explain large energy differences based on your system (trajectory)
+    \item Include error bars in all your visual analyses
 \end{itemize}
@@
 JDC is a current member of the Scientific Advisory Board of OpenEye Scientific Software and a consultant to Foresite Laboratories.
-A complete funding history for the Chodera lab can be found at http://choderalab.org/funding.
+A complete funding history for the Chodera lab can be found at \url{http://choderalab.org/funding}.
 DLM appreciates financial support from the National Institutes of Health (1R01GM108889-01, 1R01GM124270-01A1 and R01GM132386), and the National Science Foundation (CHE 1352608).
 
+\section*{Author Information}
+\makeorcid
 \bibliography{alchemical,manual}

### 2020-08-14 paper/manuscript.tex
minor change
fixing issue #90

@@
 Typical molecular dynamics engines are not set up to recognize this change, or at least not to correctly include contributions to the free energy from changing constraints/constraint length, so results for a transformation would usually be erroneous.
-At present the most general solution to this problem is simply to avoid the use of constraints (and thus use a smaller timestep if necessary, usually of around 1 fs) in any relative free energy calculation involving a transformation of a constrained bond, as done by GROMACS.
+At present the most general solution to this problem is simply to avoid the use of constraints (and thus use a smaller timestep if necessary, usually of around 1 fs) in any relative free energy calculation involving a transformation of a constrained bond.
+When using TI in GROMACS $\frac{dH}{d\lambda}$ calculations have contribution from the LINCS constrained bonds, thus transforming constrained bonds in GROMACS is possible.
 
@@
 %%%%%%%
-
-Declare any potentially conflicting interests here, whether or not they pose an actual conflict in your view. JM is a current member of the Scientific Advisory Board of Cresset. 
+JM is a current member of the Scientific Advisory Board of Cresset. 
 MK is employed by Cresset who commercially distribute a software for performing alchemical free energy calculations. MRS is a Open Science Fellow and consultant for Silicon Therapeutics.

### 2020-08-30 paper/manuscript.tex

@@
 A complete funding history for the Chodera lab can be found at \url{http://choderalab.org/funding}.
-DLM appreciates financial support from the National Institutes of Health (1R01GM108889-01, 1R01GM124270-01A1 and R01GM132386), and the National Science Foundation (CHE 1352608).
+DLM acknowledges support from the National Institutes of Health (R01GM108889, R01GM124270, and R01GM132386), and the National Science Foundation (CHE 1352608).
+JDC acknowledges support from the National Institutes of Health (NIH P30 CA008748, NIH R01 GM121505, R01 GM132386).
+The content is solely the responsibility of the authors and does not necessarily represent the official views of the National Institutes of Health.
 

### 2020-10-16 review_response/Responses.tex
adding review responses merge from broken overleaf

\documentclass[10pt,stdletter]{newlfm}
\usepackage{charter}
\usepackage{xcolor}

\widowpenalty=1000
\clubpenalty=1000

\newsavebox{\Luiuc}
\sbox{\Luiuc}{%
    \parbox[b]{1.75in}{%
        \vspace{0.5in}%
        \includegraphics[scale=0.2,ext=.pdf]
        {logo.pdf}%
    }%
}%

\makeletterhead{Uiuc}{\Lheader{\usebox{\Luiuc}}}

\newlfmP{headermarginskip=5pt}
\newlfmP{sigsize=5pt}
\newlfmP{dateskipafter=20pt}
\newlfmP{addrfromphone}
\newlfmP{addrfromemail}
\PhrPhone{Phone}
\PhrEmail{Email}

\lthUiuc

\namefrom{Antonia Mey}
\addrfrom{%
    University of Edinburgh\\
    David Brewster Road\\
    Edinburgh\\
    EH9 3FJ, UK\\ 
}

\phonefrom{ +44 131 651 3309}
\emailfrom{antonia.mey@ed.ac.uk}

\addrto{%
Living Journal of Computational Molecular Science}

\greetto{Dear Prof Oostenbrink,}
\closeline{Sincerely,}
\begin{document}
\begin{newlfm}

[fix me]

\textbf{Reviewer 1 comments:}
The best practices guide “Best Practices for Alchemical Free Energy Calculations” by Mey et al. gives an extensive introduction to the field of alchemical free energy calculations. After a general introduction, the reader is introduced to the statistical mechanics foundations needed to understand the concepts introduced later, followed by general discussions on applicability and prerequisites of free energy calculations. The authors then continue to discuss practical methodologies and issues and guide the reader towards a general understanding on how such calculations can be performed, including proper data analysis. It is then concluded that it is important to understand the functionalities and limitations of today’s often highly automated free energy workflows and that blind prediction challenges help to understand the used algorithms and issues related to reproducibility better.

The manuscript is very well-written and concepts of alchemical free energy calculations are clearly presented. The provided checklist can be a useful tool and can serve as initial orientation, especially for scientists who are new to the field.

However, I would like to make some comments to the manuscript:

Figure 1: For me there is no intrinsic difference between the free energy calculations shown in panel D and F, both transformations should be performed in the bound and unbound states to obtain meaningful results, and both should be termed relative calculations (ΔΔG). Maybe it could be more clearly indicated in general if a whole molecule is decoupled and if only a part of a molecule is transformed?
\textcolor{red}{JS: swapped dG to ddG; swapped E and F, made 'Host' and 'membrane' text larger, added interaction schematic (green) in D}

Section 4.1, line 3: RMS errors to experiment? This should be clearly stated.

Section 4.1, line 30: it is not entirely clear to me what the authors mean by “edges”. This should be better explained in the text.

Section 4.1, end: There is a github URL in the text which is not referenced or explained directly. What can the reader expect to find on this git?

Equations 14 and 2: I would suggest using either KD or Kb consistently to improve readability.

Section 5.4, last paragraph: I would call it “binding free energy prediction”, not “binding energy prediction”

Constraints and relative free energy calculations: for me, a timestep of 1 fs is more an upper limit for MD simulations involving non-constrained bonds, I would rather suggest a value of 0.5 fs.
\textcolor{MRS: This is a valuable point to raise. Certainly 1.0 fs is near the limit of stability without any constraints with standard fields, and there are studies showing there are nonnegligible statistical mechanical errors with timesteps of 1.0 fs. However, many studies with free energy calculations have been done with these parameters and it's not clear they have statistically significant deviation from simulations without simulations. We have added a discussion of timestep to section XXXX with a tentative recommendation that 1.0 fs may be sufficient, but some warnings there may be potentially be issues}

Section 7.2.1: Using a flat-bottom restraining potential in which the ligand never hits the potential wall is similar to using no restraining potential at all. I would not suggest skipping a reweighting step; if the energetic contribution of the restraint is little, the reweighting will not introduce much noise and is technically easy to perform.

\textcolor{red}{AR: We changed the wording to clarify this point. The reweighting can be skipped if the ligand is never \textit{sampled} while hitting the wall because, in this case, the numerical correction is exactly zero. Using the restraint is different from using no restraining potential at all because during the MD simulation the ligand might still very occasionally hit the wall (thus limiting its volume to the binding site) even if we these events are not observed after subsampling the trajectory.}

Section 7.2.5 and Figure 9: The concept of evaluating both forward and reverse cumulative averages is nicely described in reference 42. I would cite the reference in this section.

\textcolor{red}{AR: We added the reference to the text.}

Section 8.5 and Figure 12: Here I would again mention the use of both forward and reverse cumulative averages and plot them explicitly in Figure 12.

\textcolor{red}{AR: Both the text in the section and the figure were updated to include the forward/reverse convergence test.}

Section 10: GROMOS is a simulation package with very extensive (alchemical) free energy calculation capabilities and free for academic use. I think it should be mentioned here.

\textbf{Reviewer 2 Comments}
-Much of the discussion is related to alchemical free-energy calculations in drug design. I think that the success of such calculations is also very dependent on the force field. Although not a topic of the present manuscript, the authors may want to issue a recommendation/warning regarding the validation of automatically generated ligand topologies. I guess, many practitioners just take such topologies for granted.

\color[red]{We have added explicit comments in section 4.1 with respect to the need for using reasonable parameter/topology input files that have been inspected to make sure the molecular description has no obvious errors, or lack of chemical coverage in the method used to generate them}

-In the list of more advanced topics on page 3 the application to QM/MM simulations may be mentioned as well, since this seems to be an area of growing importance

\color[red]{MRS: we have added QM/MM explicitly in that list as an advanced topic beyond the scope of this document.}

-At the end of section 7.1.1 on pages 15/16 the following statement appears: “Typical molecular dynamics engines are not set up to recognize this change, or at least not to correctly include contributions to the free energy from changing constraints/constraint length, so results for a transformation would usually be erroneous.” This statement is a bit surprising since methods to calculate the contribution of the work done by constraint forces have been described a few times, e.g.: 
-Van Gunsteren, W.F., In van Gunsteren W.F. and Weiner, P.K. (Eds.) Computer Simulation of Biomolecular Systems: Theoretical and Experimental Applications, ESCOM, Leiden, 1989, pp. 1-26.

-Pearlman, D.A., Kollman, P.A., The overlooked bond-stretching contribution in free energy perturbation calculations, J. Chem. Phys. 1991, 94(6), 4532-4545.

-Straatsma, T.P., Zacharias, M., McCammon, J.A., Holonomic constraint contribution to free energy differences from thermodynamic integration molecular dynamics simulations, Chem. Phys. Lett., 1992, 196(3-4), 297-302.

-Pearlman, D.A., Determining the contributions of constraints in free energy calculations: Development, characterization, and recommendations, J. Chem. Phys. 1993, 98(11), 8964-8957.

One implementation, based on integrating over the constraint forces, can for example be found in the MD code GROMOS. I suggest to mention the availability of both the theory and also practical implementations in the manuscript.

\color[red]{We have revised this text in 7.1, making it clear that the corrections are known, that different programs vary in their implementation of the corrections, and that users should check the user manuals if they have bond lengths that both change and are constrainted}

-In section 7.2.2, when discussing changes in net charge the “Double-system/single-box” setup described by Gapsys et al. may be mentioned. It is related to the alchemical ion approach but can be mentioned here for completeness and to help newcomers to the field in sorting the various approaches. Gapsys, V., Michielssens, S., Peters, J. H., DeGroot, B. L., Leonov, H. Calculation of Binding Free Energies, In Andreas Kukol (ed.), Molecular Modeling of Proteins, Methods in Molecular Biology, vol. 1215, doi: 10.1007/978-1-4939-1465-4\_9, Springer Science+Business Media New York 2015.

SOME TECHNICAL ISSUES

-In the list of symbols on page 4, the quantity $\Delta G$ is denoted as Gibbs free energy while $\Delta A$ is denoted as Helmholtz free energy. However, according to IUPAC or IUPAP these quantities are referred to as Gibbs energy or Helmholtz energy (see \url{https://doi.org/10.1039/9781847557889} or \url{https://doi.org/10.1016/0378-4371(78)90209-1}) or sometimes also as “Gibbs function” or “Helmholtz function” or as “free enthalpy” or “free energy” respectively. Of course, the nomenclature used by authors is very common in the field. Therefore, one solution would be to use the IUPAC nomenclature in the List of Symbols along with a footnote explaining that here a different nomenclature is used (or the other way round).

\color[red]{MRS: we have added the less-commonly used but also correct terms as alternate names for these quantities in the list of symbols.}

-Eq. (2) holds for dilute solutions if thermodynamic activities can be approximated by concentrations. This is also stated in the classical Gilson paper but I guess repeating it here, could be useful.

-Page 5: In statistical mechanics nomenclature the term “configuration” is usually used for q ⃑. In chemistry and biology, the term “conformation” is probably more common. In the context of the present manuscript “configuration” fits better in my opinion.

-The quantity $\Delta G_{solvated}$ in Figure 2 is called $\Delta G_{unbound}$ in Eq. (9).

-In the first paragraph of section 4.2 it is mentioned that a good practice is to perform two or three runs of the same perturbation. Here one might add that the initial conditions (i.e. velocities) should be different.

-Page 17, line 9 under Eq. (17): A, B, and AB may be replaced by R, L, and RL.

-In the paragraph of author contributions, the description of the contribution of Samarjeet Prasad seems to be missing.

TYPOS OR OTHER EDITORIAL ISSUES

-Page 3, right column, line 5 in third paragraph: preparation( Sec. 6) -> preparation (Sec. 6)

\textcolor{red}{JS: Fixed.}

-Page 4, left column, list of acronyms: “unsigned error” should be written in upper case to be consistent with the other entries.

\textcolor{red}{JS: Fixed.}

-Page 5, left column: In the nominator and denominator of Eqs (5) and (6) the minus sign is missing in the exponentials.

\textcolor{red}{JS: Fixed.}

-Page 5, right column, first line: “one of the …”

\textcolor{red}{JS: Fixed.}

-Page 6, right column, two lines above Eq. 12: Bennet’s -> Bennett’s

\textcolor{red}{JS: Fixed.}

-Page 8, right column, fifth line below Eq. 16: enrgies -> energies

\textcolor{red}{JS: Fixed.}

-Page 8, right column, 10th line below Eq 16: Perhaps the line break in the unit can be avoided.

\textcolor{red}{JS: Fixed by changing line to ..If the accuracy of \textit{a collection of} free energy calculations is..}

-Page 11, left column, paragraph 5.4: there are two occurrences of the term “binding energy prediction”. Is “binding free energy prediction” meant?

\textcolor{red}{JS: Fixed.}

-Page 12, left column: “stereoisomers” is misspelled.

\textcolor{red}{JS: Fixed.}

-Page 15, Figure 4: the right structure in the lower part has no green area. Is that intended?

\textcolor{green}{JS: Yes, as this structure does not match the MCS. Could perhaps be explained more clearly in the text?(subsection "ring breaking and forming"}

-Page 17, Eq. 18: the ratio of $c^o/V_L$ seems not correct. It should be either $V^o/V_L$ where $V^o$ is the volume corresponding to $c^o$, or, when sticking to $c^o$: $c^o * V_L$

\textcolor{green}{JS: leaving this up to Toni"}

-Page 17, line 14: “restraints” (s was missing)

\textcolor{red}{JS: Fixed.}

-Page 19, Figure 7: The axis labels are hard to read.

\textcolor{red}{JS: Fixed in PR.}

-Page 19, Eq. 19: I think it should be $(1-\lambda)U_0 + \lambda U_1$ to fit to the text below the equation

\textcolor{green}{JS: leaving this up to Toni"}

-Page 20, left column, line 14: σ/r should be written in parenthesis.

\textcolor{red}{JS: Fixed.}

-Page 20, left column, line 16: B in $k_BT$ should be in subscript and T not.

\textcolor{red}{JS: Fixed.}

-Page 20, left column, line 10 from the bottom: is there an “at” missing in “and at the other endpoint …”?

\textcolor{red}{JS: Yes, fixed.}

-Page 22, left column, caption of Figure 8 (line 6): should there be a comma after “exchange”?

\textcolor{green}{JS: I don't see it?}

-Page 22, right column, line 2: though -> through

\textcolor{red}{JS: Fixed.}

-Page 22, right column, lines 8 and 9: Please add a blank between value and unit.

\textcolor{red}{JS: Seems to have been fixed by DM.}

-Page 22, right column, last line of second paragraph: Please remove blank between “(“ and “see”

\textcolor{red}{JS: Seems to have been fixed by DLM.}

-Page 23, left column, caption of Figure 9 (first and last lines): k and T should be written in italic.

\textcolor{red}{JS: Fixed.}

-Page 24, right column, line 4 in second paragraph: There seems to be a non-fitting “the” in the sentence.

\textcolor{red}{JS: Fixed.}

-Page 24, right column, line 7 in third paragraph: Please remove blank between “]” and “.”

\textcolor{red}{JS: Fixed.}

-Page 25, left column, first line in second paragraph of section “Handling multiple …”: lignad -> ligand

\textcolor{red}{JS: Fixed.}

-Page 26, right column, figure 11: The “u” on the y-axis in the second panel might also be written in italics.

\textcolor{green}{JS: Should be fixed by the creator of the figure.}

-Page 27, right column, line 8 in second paragraph of section “Subsampling data to …”: “can be” appears twice.

\textcolor{red}{JS: Fixed.}

-Page 28, left column, line 21: should it be the “right hand side” instead of “left hand side”?

\textcolor{green}{JS: Should be fixed by JDC/JM/DLM.}

-Page 28, right column, Eq. (27): the $j$ in $f_ij$ should be subscript.

\textcolor{red}{JS: Fixed.}

-Page 28, right column, line 4 in paragraph 4: “that is asymptotically correct in …”. Is the “in” correct?

\textcolor{red}{JS: Fixed.}

-Page 28, right column, second last line from bottom: “than” should be “then”

\textcolor{red}{JS: Fixed.}

-Page 29, right column, line 4 from the bottom: is there a “that” missing in “… errors that are hard to estimate …”

\textcolor{red}{JS: Fixed.}

-Page 31, left column, line 2: is there a comma missing after “available”?

\textcolor{red}{JS: Fixed.}

-Page 33, left column, line 8 from bottom: “… minima are that are visited …” needs revision

\textcolor{red}{JS: Fixed.}

-Page 33, right column, line 4: The “or” does not fit here.

\textcolor{red}{JS: Fixed.}

-Page 34, left column, line 5 in second paragraph: “… have errors of show multiple …” needs revision.

\textcolor{red}{JS: Fixed.}

-Page 35, right column, line 6 from bottom: Please add blank between “module” and “[“.

\textcolor{red}{JS: Fixed.}

-Page 39, left column: line 3: Please add blank after “14,”

\textcolor{red}{JS: Fixed.}

-Page 40, ref [10]: Please abbreviate the journal name.

\textcolor{green}{JS: leaving refs up to Toni"}

-Page 45, ref [131]: Please abbreviate the journal name.

-Page 45, ref [155]: This paper has the title: “Incorporating the effect of ionic strength in free energy calculations using explicit ions”.

-Page 47, ref [189]: Please abbreviate the journal name.

-Page 48, ref [216]: Please provide a link here.

-Page 48, ref [224]: Please use abbreviations for the author names.

\textbf{Reviewer 3 Comments}

This is a comprehensive, very systematic, and well-written best practice guide that will be of immense help both new and advanced practitioners. I found the checklist for novice practitioners particularly appealing. It definitely should be published, and I only have a few remarks that may help clarification or avoid misconceptions. I also added a few suggestions of issues I personally would find also helpful to new practitioners, but I would leave it at the discretion of the authors whether they want to include those or not.

For the amusement of the authors I should mention that while reading the following occurred to me four times in a row: at some point I thought: Hey, the following issue should be mentioned, took a note, did read on, and found precisely this issue discussed, and thus I deleted my note. I think that shows that the manuscript is really very well structured, always fulfilling the reader’s expectation what should come next!

(1) In Chap 3.1 Eq suggests that the law of mass action can be applied to calculate K_b and Delta G_B from MD simulations. I am aware the authors just mean to provide the conceptual background here, but I am similarly aware of a few groups that actually have used the LMA. This is very dangerous. As shown in [54], the LMA does not strictly hold for systems with very few particles, and it contradicts in fact the more fundamental and general eq (7). E.g., if a simulation system just contains one L and one R molecule, and these bind and unbind many times during the simulation, eq (7) would (correctly) predict that $\Delta G = -kT log Z(RL)/Z(R+L) \approx -kT log T(RL)/T(R+L)$, where the latter means the time fraction of the trajectory where the system is bound / unbound. In contrast, the LMA would (wrongly) claim that $\Delta G = -kT log [RL]/[R][L] \approx -kT log T(RL)/(T(R+L))^2$. The authors might want to add a word of caution to this not so well-known fact.

\textcolor{red}{AR: The referee raises a valid concern, and we have made the warning explicit in Section 3.2.}

(2) Eq (4) is a bit confusing, at least to a physicist. Naively, I would say it would be correct without the concentration term: $\Delta G$ then simply means the free energy difference for the concentration used in the simulation. Apparently, this is not meant here, but then I would refer to Delta G as the standard free energy, $\Delta G^0$.

\textcolor{red}{AR: The standard free energy is indeed what we are referring to throughout the paper. We have adopted the convention of using \Delta G (without the ^o) to refer to the standard free energy. This is common practice in the alchemical calculations literature as the standard free energy is typically the only quantity of interest and the one reported and compared to experiments. We agree this might create some confusion, and we have thus added a comment in Section 3.1 to clarify the notation.}

(3) In Sec 3.2, the issue of a Delta G (Delta Delta G) between states with different number of atoms is touched upon. This issue is more subtle than the text suggests, and the authors might want to spend a few more sentences to explain under which conditions the ‘unit-of-volume-factor’ actually cancels out and when not. There’s chances for pitfalls for new practitioners!

(4) I agree with the notes in 7.2.2. that no agreed upon way to handle changes in total charges is available. The authors might consider adding (if they agree) that whenever there is a change of total charge, direct comparison to experiment is meaningless, because in experiments the total charge is always conserved (otherwise the test tube would explode). One might even go so far as to say that any calculation with change in total charge is meaningless…which kind of solves the problem from an unexpected angle.

(5) Can the authors give some rules of thumb how fast the accuracy increases with simulation length? Is it always as simple as /sqrt(N)?

\color[red]{MRS: This is a point to raise. The analytical estimators for free energy calculations do have variance formulae that do scale with $1/N$ (See Shirts and Pande, 2005, among others), where $N$ is the number of uncorrelated samples, so the uncertainty will indeed scale as $N^{-1/2}$ as the simulation length increases, as long as it has already sampled several uncorrelated samples. We have added a bit more discussion and context in section 7.2.5}.

(6) Non-equilibrium (Crooks) based calculations have been shown to be quite efficient, and IMHO should also be included (I did see it briefly mentioned in 7.2.4).

Our feeling is that these simulation techniques, while they have the potential to be very efficient, can be tricky in implementation that keep them from being best practice yet (likely a chance for a future LiveCoMS paper!); this is not intended as comprehensive review.  We have:

\begin{itemize}
\item Added word "equilibrium" to abstract to indicate that's our focus
\item Added two mentions in Scope (Section 2) that we are NOT covering this here, along with one reference to Gapsys et al.
\item Added a mention in 3.2 that specifies we're talking about equilibrium calculations.
Added sentence at end of 3.2 mentioning these
\item, Clarify mention of them in 7.2.4 
\end{itemize}

(7) What impact does the way of morphing from state A to B have? Rules of thumb what should be avoided, and what helps? I saw sec 7.1.1., but found the focus on dual topology, single topology, or hybrid topology a bit narrow. I do understand that these are the established ones that stood the test of time, but practitioners should also be aware that there is much more out there to choose from, such as enveloping potentials, or variational morphing, which the authors might want to add to their discussion in 7.2.3.

\color[red]{MRS: The section issues with soft core and $\lambda$ examines much of the clear lessons on simulation endpoints and pathways.  We now bring up EDS and enveloping distribution approaches as alternatives in section 7.2, but at least in this paper choose not to recommend them as best practices not because they don't work, but because they at the present time there are issues with kinetic trapping and determining scaling parameters that mean they are not the recommended way to start doing these calculations for non-experts or maximal reliability.}

(8) In 7.2.5, concerning uncertainty estimates. I found it helpful for many students to realize that free energy error estimates often suffer from exactly the same sampling problems as the free energy estimate itself. Specifically, if the free energy estimate is off because an important region in configuration space has not been sampled, the error estimate has no way to indicate this. The authors might want to make this point super-clear.

(10) Use larger fonts in Fig 2 &7, generally try to have equally-sized fonts throughout all Figs.

\textcolor{red}{JS: PR includes increased font sizes in figs 2 and 7. Additionally, slightly reformatted pane 7B to have a cleaner horizontal line}

\end{newlfm}

\end{document}

### 2020-11-03 review_response/Responses.tex
last bibliography update

@@
-\documentclass[10pt,stdletter]{newlfm}
-\usepackage{charter}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Professional Formal Letter
+% LaTeX Template
+% Version 1.0 (28/12/13)
+%
+% This template has been downloaded from:
+% http://www.LaTeXTemplates.com
+%
+% Original author:
+% Brian Moses (http://www.ms.uky.edu/~math/Resources/Templates/LaTeX/)
+% with extensive modifications by Vel (vel@latextemplates.com)
+%
+% Further modifications including UoY specific changes by George Tsoulas 
+% (george.tsoulas@gmail.com}
+%
+% A simple letter template with the University of York Logo mostly meant for % personal rather than departmental letters.
+%
+% License:
+% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%---------------------------------------------
+% PACKAGES AND OTHER DOCUMENT CONFIGURATIONS |
+%---------------------------------------------
+
+\documentclass[11pt,a4paper]{letter} % Specify the font size (10pt, 11pt and 12pt) and paper size (letterpaper, a4paper, etc)
+
+\usepackage{graphicx} % Required for including pictures
+\usepackage{microtype} % Improves typography
+\usepackage[T1]{fontenc} % Required for accented characters
+\usepackage{fbb} %Use the free version of the Bembo font Family
 \usepackage{xcolor}
+\usepackage{hyperref}
+%\usepackage{gfsdidot} % Use the GFS Didot font: http://www.tug.dk/FontCatalogue/gfsdidot/
+%\usepackage{lipsum}
+% Create a new command for the horizontal rule in the document which allows thickness specification
+\makeatletter
+\def\vhrulefill#1{\leavevmode\leaders\hrule\@height#1\hfill \kern\z@}
+\makeatother
 
-\widowpenalty=1000
-\clubpenalty=1000
+%-------------------
+% DOCUMENT MARGINS |
+%-------------------
 
-\newsavebox{\Luiuc}
-\sbox{\Luiuc}{%
-    \parbox[b]{1.75in}{%
-        \vspace{0.5in}%
-        \includegraphics[scale=0.2,ext=.pdf]
-        {logo.pdf}%
-    }%
-}%
+\usepackage[left=1.2in,
+top=1.25in,
+right=1.25in,
+bottom=1.25in,
+headsep=0.25in,
+headheight=20pt,
+heightrounded]
+{geometry}
 
-\makeletterhead{Uiuc}{\Lheader{\usebox{\Luiuc}}}
+%---------------------
+% SENDER INFORMATION |
+%---------------------
 
-\newlfmP{headermarginskip=5pt}
-\newlfmP{sigsize=5pt}
-\newlfmP{dateskipafter=20pt}
-\newlfmP{addrfromphone}
-\newlfmP{addrfromemail}
-\PhrPhone{Phone}
-\PhrEmail{Email}
+\def\Who{Antonia Mey} % Your name
+\def\What{Dr } % Your title
+\def\Where{School of Chemistry} % Your department/institution
+\def\Address{University of Edinburgh} % Your address
+\def\Address{David Brewster Road, EH9 3FJ} % Your address
+\def\CityZip{} % Your city, zip code, country, etc
+\def\Email{antonia.mey@ed.ac.uk} % Your email address
+\def\URL{http://www.mey-research.org} % Your URL
 
-\lthUiuc
+%------------------------------------
+% HEADER AND FROM ADDRESS STRUCTURE |
+%------------------------------------
 
-\namefrom{Antonia Mey}
-\addrfrom{%
-    University of Edinburgh\\
-    David Brewster Road\\
-    Edinburgh\\
-    EH9 3FJ, UK\\ 
+\address{\vspace*{-2.3cm}
+  \includegraphics[width=1in]{Ylogo.png} %Include the logo of your institution
+\hspace*{5in} % Position of the institution logo, increase to move left, decrease to move right
+\hspace{-3in} \hfill \normalsize % Second line of institution name, adjust hspace if your logo is wide
+\makebox[0ex][r]{\bf \What \Who }\hspace{0.08in} % Print your name and title with a little whitespace to the right
+~\\[-0.03in] % Reduce the whitespace above the horizontal rule
+\vspace*{-.08in}\hspace{.1in}\vhrulefill{1pt} \\ % Horizontal rule, adjust hspace if your logo is wide and \vhrulefill for the thickness of the rule
+\hspace{\fill}\parbox[t]{2.85in}{ % Create a box for your details underneath the horizontal rule on the right
+\footnotesize % Use a smaller font size for the details
+% \Who \\ Your name if you want it again there
+\em % all text after this will be italicized
+\Where\\ % Your department
+\Address\\ % Your address
+%\CityZip\\ % Your city and zip code
+\TEL\\ % Your phone number
+\Email\\ % Your email address
+\URL % Your URL
+}
+\hspace{-1.4in} % Horizontal position of this block, increase to move left, decrease to move right
+\vspace{-1in} % Move the letter content up for a more compact look
 }
 
-\phonefrom{ +44 131 651 3309}
-\emailfrom{antonia.mey@ed.ac.uk}
+%-----------------------
+% TO ADDRESS STRUCTURE |
+%-----------------------
 
-\addrto{%
-Living Journal of Computational Molecular Science}
+\def\opening#1{\thispagestyle{empty}
+{\centering\fromaddress \vspace{1.2in} \\ % Print the header and from address here, add whitespace to move date down
+\hspace*{}\today\hspace*{\fill}\par} % Print today's date, remove \today to not display it
+{\raggedright \toname \\ \toaddress \par} % Print the to name and address
+\vspace{0.4in} % White space after the to address
+\noindent #1 % Print the opening line
+% Uncomment the 4 lines below to print a footnote with custom text
+%\def\thefootnote{}
+%\def\footnoterule{\hrule}
+%\footnotetext{\hspace*{\fill}{\footnotesize\em Footnote text}}
+%\def\thefootnote{\arabic{footnote}}
+}
 
-\greetto{Dear Prof Oostenbrink,}
-\closeline{Sincerely,}
+%----------------------
+% SIGNATURE STRUCTURE |
+%----------------------
+
+\signature{\What \Who} % The signature is a combination of your name and title
+
+\long\def\closing#1{
+\vspace{0.1in} % Some whitespace after the letter content and before the signature
+\noindent % Stop paragraph indentation
+\hspace*{\longindentation} % Move the signature right
+\parbox{\indentedwidth}{\raggedright
+#1 % Print the signature text
+\vskip 0.65in % Whitespace between the signature text and your name
+\fromsig}} % Print your name and title
+
+%--------------------------------------------------------------------------
+
 \begin{document}
-\begin{newlfm}
 
-[fix me]
+%-------------
+% TO ADDRESS |
+%-------------
 
-\textbf{Reviewer 1 comments:}
-The best practices guide “Best Practices for Alchemical Free Energy Calculations” by Mey et al.
-gives an extensive introduction to the field of alchemical free energy calculations.
-After a general introduction, the reader is introduced to the statistical mechanics foundations needed to understand the concepts introduced later, followed by general discussions on applicability and prerequisites of free energy calculations.
-The authors then continue to discuss practical methodologies and issues and guide the reader towards a general understanding on how such calculations can be performed, including proper data analysis.
-It is then concluded that it is important to understand the functionalities and limitations of today’s often highly automated free energy workflows and that blind prediction challenges help to understand the used algorithms and issues related to reproducibility better.
+\begin{letter}
+{Living Journal of Computational Molecular Sciences\\
+}
+%% Add more lines as you wish
 
-The manuscript is very well-written and concepts of alchemical free energy calculations are clearly presented.
-The provided checklist can be a useful tool and can serve as initial orientation, especially for scientists who are new to the field.
+%
+%-----------------
+% LETTER CONTENT |
+%-----------------
+
+\opening{Dear Professor Oostenbrink,}
+
+We would like to thank the reviewers for having taken the time to read the manuscript and to give such excellent feedback.
+Below we endeavor to address each of the comments in detail.
+All the suggestions and references from you, the editor, have also been included now, giving Gromos equally prominent space as other software packages in the manuscript.
+
+\textbf{Reviewer 1 comments:}\\
+(...)
 
@@
 
-Figure 1: For me there is no intrinsic difference between the free energy calculations shown in panel D and F, both transformations should be performed in the bound and unbound states to obtain meaningful results, and both should be termed relative calculations (ΔΔG).
+Figure 1: For me there is no intrinsic difference between the free energy calculations shown in panel D and F, both transformations should be performed in the bound and unbound states to obtain meaningful results, and both should be termed relative calculations ($\Delta \Delta$ G).
 Maybe it could be more clearly indicated in general if a whole molecule is decoupled and if only a part of a molecule is transformed?
-\textcolor{red}{JS: swapped dG to ddG; swapped E and F, made 'Host' and 'membrane' text larger, added interaction schematic (green) in D}
+
+\textcolor{red}{We used  $\Delta\Delta G$ instead of  $\Delta$ G; swapped E and F, made 'Host' and 'membrane' text larger and added interaction schematic (green) in D.}
 
@@
 
-Section 4.1, line 30: it is not entirely clear to me what the authors mean by “edges”. This should be better explained in the text.
+\textcolor{red}{We added a clarification to this in Sec. 4.1.}
+
+Section 4.1, line 30: it is not entirely clear to me what the authors mean by "edges". This should be better explained in the text.
+
+\textcolor{red}{We added a sentence in Sec. 4.1 to clarify this.}
 
@@
 
+\textcolor{red}{We removed this link, as it didn't really add anything useful.}
+
 Equations 14 and 2: I would suggest using either KD or Kb consistently to improve readability.
 
-Section 5.4, last paragraph: I would call it “binding free energy prediction”, not “binding energy prediction”
+\textcolor{red}{Using $K_b$ consistently now.}
+
+Section 5.4, last paragraph: I would call it "binding free energy prediction", not "binding energy prediction"
+
+\textcolor{red}{Fixed.}
 
 Constraints and relative free energy calculations: for me, a timestep of 1 fs is more an upper limit for MD simulations involving non-constrained bonds, I would rather suggest a value of 0.5 fs.
-\textcolor{MRS: This is a valuable point to raise.
-Certainly 1.0 fs is near the limit of stability without any constraints with standard fields, and there are studies showing there are nonnegligible statistical mechanical errors with timesteps of 1.0 fs.
-However, many studies with free energy calculations have been done with these parameters and it's not clear they have statistically significant deviation from simulations without simulations.
-We have added a discussion of timestep to section XXXX with a tentative recommendation that 1.0 fs may be sufficient, but some warnings there may be potentially be issues}
+
+\textcolor{red}{ This is a valuable point to raise.
+Certainly 1.0 fs is near the limit of stability without any constraints with standard fields, and there are studies showing there are non-negligible statistical mechanical errors with timesteps of 1.0 fs.
+However, many studies with free energy calculations have been done with these parameters and it's not clear they have statistically significant deviation from simulations with smaller timesteps.
+We have added a discussion on timesteps to section 7.3 with a tentative recommendation that 1.0 fs may be acceptable.}
 
@@
 
-\textcolor{red}{AR: We changed the wording to clarify this point.
+\textcolor{red}{We changed the wording to clarify this point.
 The reweighting can be skipped if the ligand is never \textit{sampled} while hitting the wall because, in this case, the numerical correction is exactly zero.
@@
 
-\textcolor{red}{AR: We added the reference to the text.}
+\textcolor{red}{We added the reference to the text.}
 
@@
 
-\textcolor{red}{AR: Both the text in the section and the figure were updated to include the forward/reverse convergence test.}
+\textcolor{red}{Both the text in the section and the figure were updated to include the forward/reverse convergence test.}
 
@@
 
-\textbf{Reviewer 2 Comments}
--Much of the discussion is related to alchemical free-energy calculations in drug design.
+\textcolor{red}{We have added Gromos to the software section and also added appropriate references to it throughout the manuscript.}
+
+\textbf{Reviewer 2 Comments:}\\
+Much of the discussion is related to alchemical free-energy calculations in drug design.
 I think that the success of such calculations is also very dependent on the force field.
@@
 
-\color[red]{We have added explicit comments in section 4.1 with respect to the need for using reasonable parameter/topology input files that have been inspected to make sure the molecular description has no obvious errors, or lack of chemical coverage in the method used to generate them}
+\textcolor{red}{We have added explicit comments in section 4.1 with respect to the need for using reasonable parameter/topology input files that have been inspected to make sure the molecular description has no obvious errors, or lack of chemical coverage in the method used to generate them}
 
--In the list of more advanced topics on page 3 the application to QM/MM simulations may be mentioned as well, since this seems to be an area of growing importance
+In the list of more advanced topics on page 3 the application to QM/MM simulations may be mentioned as well, since this seems to be an area of growing importance.
 
-\color[red]{MRS: we have added QM/MM explicitly in that list as an advanced topic beyond the scope of this document.}
+\textcolor{red}{We have added QM/MM and ML topics explicitly in that list and added some starting point references.}
 
--At the end of section 7.1.1 on pages 15/16 the following statement appears: “Typical molecular dynamics engines are not set up to recognize this change, or at least not to correctly include contributions to the free energy from changing constraints/constraint length, so results for a transformation would usually be erroneous.” This statement is a bit surprising since methods to calculate the contribution of the work done by constraint forces have been described a few times, e.g.:
--Van Gunsteren, W.F., In van Gunsteren W.F. and Weiner, P.K. (Eds.) Computer Simulation of Biomolecular Systems: Theoretical and Experimental Applications, ESCOM, Leiden, 1989, pp. 1-26.
+At the end of section 7.1.1 on pages 15/16 the following statement appears: "Typical molecular dynamics engines are not set up to recognize this change, or at least not to correctly include contributions to the free energy from changing constraints/constraint length, so results for a transformation would usually be erroneous." This statement is a bit surprising since methods to calculate the contribution of the work done by constraint forces have been described a few times, e.g.:
+\begin{itemize}
 
--Pearlman, D.A., Kollman, P.A., The overlooked bond-stretching contribution in free energy perturbation calculations, J. Chem. Phys. 1991, 94(6), 4532-4545.
+    \item Van Gunsteren, W.F., In van Gunsteren W.F. and Weiner, P.K. (Eds.) Computer Simulation of Biomolecular Systems: Theoretical and Experimental Applications, ESCOM, Leiden, 1989, pp. 1-26.
 
--Straatsma, T.P., Zacharias, M., McCammon, J.A., Holonomic constraint contribution to free energy differences from thermodynamic integration molecular dynamics simulations, Chem.
+    \item Pearlman, D.A., Kollman, P.A., The overlooked bond-stretching contribution in free energy perturbation calculations, J. Chem. Phys. 1991, 94(6), 4532-4545.
+
+    \item Straatsma, T.P., Zacharias, M., McCammon, J.A., Holonomic constraint contribution to free energy differences from thermodynamic integration molecular dynamics simulations, Chem.
 Phys.
@@
 
--Pearlman, D.A., Determining the contributions of constraints in free energy calculations: Development, characterization, and recommendations, J. Chem. Phys. 1993, 98(11), 8964-8957.
+    \item Pearlman, D.A., Determining the contributions of constraints in free energy calculations: Development, characterization, and recommendations, J. Chem. Phys. 1993, 98(11), 8964-8957.
 
+\end{itemize}
 One implementation, based on integrating over the constraint forces, can for example be found in the MD code GROMOS.
@@
 
-\color[red]{We have revised this text in 7.1, making it clear that the corrections are known, that different programs vary in their implementation of the corrections, and that users should check the user manuals if they have bond lengths that both change and are constrainted}
+\textcolor{red}{We have revised this text in 7.1, making it clear that the corrections are known, that different programs vary in their implementation of the corrections, and that users should check the user manuals if they have bond lengths that both change and are constrainted.
+We also added the suggested references.}
 
--In section 7.2.2, when discussing changes in net charge the “Double-system/single-box” setup described by Gapsys et al.
+In section 7.2.2, when discussing changes in net charge the "Double-system/single-box" setup described by Gapsys et al.
 may be mentioned.
@@
 
+\textcolor{red}{We made appropriate corrections to this. }
+
 SOME TECHNICAL ISSUES
 
--In the list of symbols on page 4, the quantity $\Delta G$ is denoted as Gibbs free energy while $\Delta A$ is denoted as Helmholtz free energy.
-However, according to IUPAC or IUPAP these quantities are referred to as Gibbs energy or Helmholtz energy (see \url{https://doi.org/10.1039/9781847557889} or \url{https://doi.org/10.1016/0378-4371(78)90209-1}) or sometimes also as “Gibbs function” or “Helmholtz function” or as “free enthalpy” or “free energy” respectively.
+In the list of symbols on page 4, the quantity $\Delta G$ is denoted as Gibbs free energy while $\Delta A$ is denoted as Helmholtz free energy.
+However, according to IUPAC or IUPAP these quantities are referred to as Gibbs energy or Helmholtz energy (see \url{https://doi.org/10.1039/9781847557889} or \url{https://doi.org/10.1016/0378-4371(78)90209-1}) or sometimes also as "Gibbs function" or "Helmholtz function" or as "free enthalpy" or "free energy" respectively.
 Of course, the nomenclature used by authors is very common in the field.
@@
 
-\color[red]{MRS: we have added the less-commonly used but also correct terms as alternate names for these quantities in the list of symbols.}
+\textcolor{red}{We have added the less-commonly used but also correct terms as alternate names for these quantities in the list of symbols.}
 
--Eq. (2) holds for dilute solutions if thermodynamic activities can be approximated by concentrations. This is also stated in the classical Gilson paper but I guess repeating it here, could be useful.
+Eq. (2) holds for dilute solutions if thermodynamic activities can be approximated by concentrations. This is also stated in the classical Gilson paper but I guess repeating it here, could be useful.
 
--Page 5: In statistical mechanics nomenclature the term “configuration” is usually used for q ⃑.
-In chemistry and biology, the term “conformation” is probably more common.
-In the context of the present manuscript “configuration” fits better in my opinion.
+\textcolor{red}{We now state this explicitly referencing the Gilson paper.}
 
--The quantity $\Delta G_{solvated}$ in Figure 2 is called $\Delta G_{unbound}$ in Eq. (9).
+Page 5: In statistical mechanics nomenclature the term "configuration" is usually used for $\vec{q}$.
+In chemistry and biology, the term "conformation" is probably more common.
+In the context of the present manuscript "configuration" fits better in my opinion.
 
--In the first paragraph of section 4.2 it is mentioned that a good practice is to perform two or three runs of the same perturbation.
+\textcolor{red}{We updated this to use configuration consistently and explained that a conformation can be a set of configurations that can be grouped together into a metastable state.}
+
+The quantity $\Delta G_{solvated}$ in Figure 2 is called $\Delta G_{unbound}$ in Eq. (9).
+
+\textcolor{red}{We updated the labelling to make it consistent. }
+
+In the first paragraph of section 4.2 it is mentioned that a good practice is to perform two or three runs of the same perturbation.
 Here one might add that the initial conditions (i.e.
@@
 
--Page 17, line 9 under Eq. (17): A, B, and AB may be replaced by R, L, and RL.
+\textcolor{red} {We have added this.}
 
--In the paragraph of author contributions, the description of the contribution of Samarjeet Prasad seems to be missing.
+Page 17, line 9 under Eq. (17): A, B, and AB may be replaced by R, L, and RL.
+
+\textcolor{red} {We have fixed this.}
+
+In the paragraph of author contributions, the description of the contribution of Samarjeet Prasad seems to be missing.
+
+\textcolor{red} {The contributions were added.}
 
 TYPOS OR OTHER EDITORIAL ISSUES
+\begin{itemize}
 
--Page 3, right column, line 5 in third paragraph: preparation( Sec. 6) -> preparation (Sec. 6)
+   \item Page 3, right column, line 5 in third paragraph: preparation( Sec. 6) $\rightarrow$ preparation (Sec. 6)    
+   \item Page 4, left column, list of acronyms: "unsigned error" should be written in upper case to be consistent with the other entries.
 
-\textcolor{red}{JS: Fixed.}
+   \item Page 5, left column: In the nominator and denominator of Eqs (5) and (6) the minus sign is missing in the exponentials.
 
--Page 4, left column, list of acronyms: “unsigned error” should be written in upper case to be consistent with the other entries.
+    \item Page 5, right column, first line: "one of the ..."
 
-\textcolor{red}{JS: Fixed.}
+   \item Page 6, right column, two lines above Eq. 12: Bennet's $\rightarrow$ Bennett's
 
--Page 5, left column: In the nominator and denominator of Eqs (5) and (6) the minus sign is missing in the exponentials.
+   \item Page 8, right column, fifth line below Eq. 16: enrgies  $\rightarrow$ energies
+\end{itemize}
+\textcolor{red}{All above have been fixed.}
 
-\textcolor{red}{JS: Fixed.}
+Page 8, right column, 10th line below Eq 16: Perhaps the line break in the unit can be avoided.
 
--Page 5, right column, first line: “one of the …”
+\textcolor{red}{Fixed by changing line to ..If the accuracy of \textit{a collection of} free energy calculations is..}
 
-\textcolor{red}{JS: Fixed.}
+Page 11, left column, paragraph 5.4: there are two occurrences of the term "binding energy prediction". Is "binding free energy prediction" meant?
 
--Page 6, right column, two lines above Eq. 12: Bennet’s -> Bennett’s
+\textcolor{red}{We have fixed these.}
 
-\textcolor{red}{JS: Fixed.}
+Page 12, left column: "stereoisomers" is misspelled.
 
--Page 8, right column, fifth line below Eq. 16: enrgies -> energies
+\textcolor{red}{This has been fixed.}
 
-\textcolor{red}{JS: Fixed.}
+Page 15, Figure 4: the right structure in the lower part has no green area. Is that intended?
 
--Page 8, right column, 10th line below Eq 16: Perhaps the line break in the unit can be avoided.
+\textcolor{red}{We updated the caption to explain this further. }
 
-\textcolor{red}{JS: Fixed by changing line to ..If the accuracy of \textit{a collection of} free energy calculations is..}
+Page 17, Eq. 18: the ratio of $c^o/V_L$ seems not correct. It should be either $V^o/V_L$ where $V^o$ is the volume corresponding to $c^o$, or, when sticking to $c^o$: $c^o * V_L$
 
--Page 11, left column, paragraph 5.4: there are two occurrences of the term “binding energy prediction”. Is “binding free energy prediction” meant?
+\textcolor{red}{This has been fixed.}
 
-\textcolor{red}{JS: Fixed.}
+Page 17, line 14: "restraints" (s was missing)
 
--Page 12, left column: “stereoisomers” is misspelled.
+\textcolor{red}{This has been fixed.}
 
-\textcolor{red}{JS: Fixed.}
+Page 19, Figure 7: The axis labels are hard to read.
 
--Page 15, Figure 4: the right structure in the lower part has no green area. Is that intended?
+\textcolor{red}{This has been fixed.}
 
-\textcolor{green}{JS: Yes, as this structure does not match the MCS. Could perhaps be explained more clearly in the text?(subsection "ring breaking and forming"}
+Page 19, Eq. 19: I think it should be $(1-\lambda)U_0 + \lambda U_1$ to fit to the text below the equation
 
--Page 17, Eq. 18: the ratio of $c^o/V_L$ seems not correct. It should be either $V^o/V_L$ where $V^o$ is the volume corresponding to $c^o$, or, when sticking to $c^o$: $c^o * V_L$
+\textcolor{red}{Correct. This has been fixed.}
 
-\textcolor{green}{JS: leaving this up to Toni"}
+Page 20, left column, line 14: $\sigma/r$ should be written in parenthesis.
 
--Page 17, line 14: “restraints” (s was missing)
+\textcolor{red}{This has been fixed.}
 
-\textcolor{red}{JS: Fixed.}
+Page 20, left column, line 16: B in $k_{B}T$ should be in subscript and T not.
 
--Page 19, Figure 7: The axis labels are hard to read.
+\textcolor{red}{This has been fixed.}
 
-\textcolor{red}{JS: Fixed in PR.}
+Page 20, left column, line 10 from the bottom: is there an "at" missing in "and at the other endpoint ..."?
 
--Page 19, Eq. 19: I think it should be $(1-\lambda)U_0 + \lambda U_1$ to fit to the text below the equation
+\textcolor{red}{ Yes, this has been fixed.}
 
-\textcolor{green}{JS: leaving this up to Toni"}
+Page 22, left column, caption of Figure 8 (line 6): should there be a comma after "exchang"?
 
--Page 20, left column, line 14: σ/r should be written in parenthesis.
+\textcolor{red}{this has been fixed.}
 
-\textcolor{red}{JS: Fixed.}
+Page 22, right column, line 2: though $\rightarrow$ through
 
--Page 20, left column, line 16: B in $k_BT$ should be in subscript and T not.
+\textcolor{red}{This has been fixed.}
 
-\textcolor{red}{JS: Fixed.}
+Page 22, right column, lines 8 and 9: Please add a blank between value and unit.
 
--Page 20, left column, line 10 from the bottom: is there an “at” missing in “and at the other endpoint …”?
+\textcolor{red}{This has been fixed.}
 
-\textcolor{red}{JS: Yes, fixed.}
+Page 22, right column, last line of second paragraph: Please remove blank between "(" and "see"
 
--Page 22, left column, caption of Figure 8 (line 6): should there be a comma after “exchange”?
+\textcolor{red}{This has been fixed.}
 
-\textcolor{green}{JS: I don't see it?}
+Page 23, left column, caption of Figure 9 (first and last lines): k and T should be written in italic.
 
--Page 22, right column, line 2: though -> through
+\textcolor{red}{This has been fixed.}
 
-\textcolor{red}{JS: Fixed.}
+Page 24, right column, line 4 in second paragraph: There seems to be a non-fitting "the" in the sentence.
 
--Page 22, right column, lines 8 and 9: Please add a blank between value and unit.
+\textcolor{red}{This has been fixed.}
 
-\textcolor{red}{JS: Seems to have been fixed by DM.}
+Page 24, right column, line 7 in third paragraph: Please remove blank between "]" and "."
 
--Page 22, right column, last line of second paragraph: Please remove blank between “(“ and “see”
+\textcolor{red}{This has been fixed.}
 
-\textcolor{red}{JS: Seems to have been fixed by DLM.}
+Page 25, left column, first line in second paragraph of section "Handling multiple ...": lignad  $\rightarrow$ ligand
 
--Page 23, left column, caption of Figure 9 (first and last lines): k and T should be written in italic.
+\textcolor{red}{This has been fixed.}
 
-\textcolor{red}{JS: Fixed.}
+Page 26, right column, figure 11: The "u" on the y-axis in the second panel might also be written in italics.
 
--Page 24, right column, line 4 in second paragraph: There seems to be a non-fitting “the” in the sentence.
+\textcolor{red}{This has been fixed.}
 
-\textcolor{red}{JS: Fixed.}
+Page 27, right column, line 8 in second paragraph of section "Subsampling data to ...": "can be" appears twice.
 
--Page 24, right column, line 7 in third paragraph: Please remove blank between “]” and “.”
+\textcolor{red}{This has been fixed.}
 
-\textcolor{red}{JS: Fixed.}
+Page 28, left column, line 21: should it be the "right hand side" instead of "left hand side"?
 
--Page 25, left column, first line in second paragraph of section “Handling multiple …”: lignad -> ligand
+\textcolor{red}{This has been fixed.}
 
-\textcolor{red}{JS: Fixed.}
+Page 28, right column, Eq. (27): the $j$ in $f_ij$ should be subscript.
 
--Page 26, right column, figure 11: The “u” on the y-axis in the second panel might also be written in italics.
+\textcolor{red}{This has been fixed.}
 
-\textcolor{green}{JS: Should be fixed by the creator of the figure.}
+Page 28, right column, line 4 in paragraph 4: "that is asymptotically correct in ...". Is the "in" correct?
 
--Page 27, right column, line 8 in second paragraph of section “Subsampling data to …”: “can be” appears twice.
+\textcolor{red}{This has been fixed.}
 
-\textcolor{red}{JS: Fixed.}
+Page 28, right column, second last line from bottom: "than" should be "then"
 
--Page 28, left column, line 21: should it be the “right hand side” instead of “left hand side”?
+\textcolor{red}{This has been fixed.}
 
-\textcolor{green}{JS: Should be fixed by JDC/JM/DLM.}
+Page 29, right column, line 4 from the bottom: is there a "that" missing in "... errors that are hard to estimate ..."
 
--Page 28, right column, Eq. (27): the $j$ in $f_ij$ should be subscript.
+\textcolor{red}{This has been fixed.}
 
-\textcolor{red}{JS: Fixed.}
+Page 31, left column, line 2: is there a comma missing after "available"?
 
--Page 28, right column, line 4 in paragraph 4: “that is asymptotically correct in …”. Is the “in” correct?
+\textcolor{red}{This has been fixed.}
 
-\textcolor{red}{JS: Fixed.}
+Page 33, left column, line 8 from bottom: "... minima are that are visited ..." needs revision
 
--Page 28, right column, second last line from bottom: “than” should be “then”
+\textcolor{red}{This has been fixed.}
 
-\textcolor{red}{JS: Fixed.}
+Page 33, right column, line 4: The "or" does not fit here.
 
--Page 29, right column, line 4 from the bottom: is there a “that” missing in “… errors that are hard to estimate …”
+\textcolor{red}{This has been fixed.}
 
-\textcolor{red}{JS: Fixed.}
+Page 34, left column, line 5 in second paragraph: "... have errors of show multiple ..." needs revision.
 
--Page 31, left column, line 2: is there a comma missing after “available”?
+\textcolor{red}{This has been fixed.}
 
-\textcolor{red}{JS: Fixed.}
+Page 35, right column, line 6 from bottom: Please add blank between "module" and "[".
 
--Page 33, left column, line 8 from bottom: “… minima are that are visited …” needs revision
+\textcolor{red}{This has been fixed.}
 
-\textcolor{red}{JS: Fixed.}
+Page 39, left column: line 3: Please add blank after "14,"
 
--Page 33, right column, line 4: The “or” does not fit here.
+\textcolor{red}{This has been fixed.}
 
-\textcolor{red}{JS: Fixed.}
+Page 40, ref [10]: Please abbreviate the journal name.
 
--Page 34, left column, line 5 in second paragraph: “… have errors of show multiple …” needs revision.
+\textcolor{red}{This has been fixed.}
 
-\textcolor{red}{JS: Fixed.}
+Page 45, ref [131]: Please abbreviate the journal name.
 
--Page 35, right column, line 6 from bottom: Please add blank between “module” and “[“.
+\textcolor{red}{This has been fixed.}
 
-\textcolor{red}{JS: Fixed.}
+Page 45, ref [155]: This paper has the title: "Incorporating the effect of ionic strength in free energy calculations using explicit ions".
 
--Page 39, left column: line 3: Please add blank after “14,”
+\textcolor{red}{This has been fixed.}
 
-\textcolor{red}{JS: Fixed.}
+Page 47, ref [189]: Please abbreviate the journal name.
 
--Page 40, ref [10]: Please abbreviate the journal name.
+\textcolor{red}{This has been fixed.}
 
-\textcolor{green}{JS: leaving refs up to Toni"}
+Page 48, ref [216]: Please provide a link here.
 
--Page 45, ref [131]: Please abbreviate the journal name.
+\textcolor{red}{This has been fixed.}
 
--Page 45, ref [155]: This paper has the title: “Incorporating the effect of ionic strength in free energy calculations using explicit ions”.
+Page 48, ref [224]: Please use abbreviations for the author names.
 
--Page 47, ref [189]: Please abbreviate the journal name.
+\textcolor{red}{This has been fixed.}
 
--Page 48, ref [216]: Please provide a link here.
-
--Page 48, ref [224]: Please use abbreviations for the author names.
-
-\textbf{Reviewer 3 Comments}
+\textbf{Reviewer 3 Comments:}
 
@@
 For the amusement of the authors I should mention that while reading the following occurred to me four times in a row: at some point I thought: Hey, the following issue should be mentioned, took a note, did read on, and found precisely this issue discussed, and thus I deleted my note.
-I think that shows that the manuscript is really very well structured, always fulfilling the reader’s expectation what should come next!
+I think that shows that the manuscript is really very well structured, always fulfilling the reader's expectation what should come next!
 
-(1) In Chap 3.1 Eq suggests that the law of mass action can be applied to calculate K_b and Delta G_B from MD simulations.
+(1) In Chap 3.1 Eq suggests that the law of mass action can be applied to calculate $K_b$ and Delta $G_B$ from MD simulations.
 I am aware the authors just mean to provide the conceptual background here, but I am similarly aware of a few groups that actually have used the LMA.
@@
 As shown in [54], the LMA does not strictly hold for systems with very few particles, and it contradicts in fact the more fundamental and general eq (7).
-E.g., if a simulation system just contains one L and one R molecule, and these bind and unbind many times during the simulation, eq (7) would (correctly) predict that $\Delta G = -kT log Z(RL)/Z(R+L) \approx -kT log T(RL)/T(R+L)$, where the latter means the time fraction of the trajectory where the system is bound / unbound.
+E.g., if a simulation system just contains one L and one R molecule, and these bind and unbind many times during the simulation, eq (7) would (correctly) predict that $\Delta G = -kT \log Z(RL)/Z(R+L) \approx -kT log T(RL)/T(R+L)$, where the latter means the time fraction of the trajectory where the system is bound / unbound.
 In contrast, the LMA would (wrongly) claim that $\Delta G = -kT log [RL]/[R][L] \approx -kT log T(RL)/(T(R+L))^2$.
@@
 
-\textcolor{red}{AR: The referee raises a valid concern, and we have made the warning explicit in Section 3.2.}
+\textcolor{red}{The referee raises a valid concern, and we have made the warning explicit in Section 3.2.}
 
@@
 
-\textcolor{red}{AR: The standard free energy is indeed what we are referring to throughout the paper.
-We have adopted the convention of using \Delta G (without the ^o) to refer to the standard free energy.
+\textcolor{red}{The standard free energy is indeed what we are referring to throughout the paper.
+We have adopted the convention of using $\Delta G$ (without the $^o$) to refer to the standard free energy.
 This is common practice in the alchemical calculations literature as the standard free energy is typically the only quantity of interest and the one reported and compared to experiments.
@@
 
-(3) In Sec 3.2, the issue of a Delta G (Delta Delta G) between states with different number of atoms is touched upon.
-This issue is more subtle than the text suggests, and the authors might want to spend a few more sentences to explain under which conditions the ‘unit-of-volume-factor’ actually cancels out and when not.
-There’s chances for pitfalls for new practitioners!
+(3) In Sec 3.2, the issue of a $\Delta G$ ($\Delta \Delta G$) between states with different number of atoms is touched upon.
+This issue is more subtle than the text suggests, and the authors might want to spend a few more sentences to explain under which conditions the 'unit-of-volume-factor' actually cancels out and when not.
+There's chances for pitfalls for new practitioners!
+
+\textcolor{red}{We addressed this in an updated version of Sec 3.2., clarifying under what conditions the equality holds.}
 
@@
 The authors might consider adding (if they agree) that whenever there is a change of total charge, direct comparison to experiment is meaningless, because in experiments the total charge is always conserved (otherwise the test tube would explode).
-One might even go so far as to say that any calculation with change in total charge is meaningless…which kind of solves the problem from an unexpected angle.
+One might even go so far as to say that any calculation with change in total charge is meaningless...which kind of solves the problem from an unexpected angle.
+
+\textcolor{red}{We have rephrased parts of this section. }
 
@@
 
-\color[red]{MRS: This is a point to raise.
+\textcolor{red}{This is a point to raise.
 The analytical estimators for free energy calculations do have variance formulae that do scale with $1/N$ (See Shirts and Pande, 2005, among others), where $N$ is the number of uncorrelated samples, so the uncertainty will indeed scale as $N^{-1/2}$ as the simulation length increases, as long as it has already sampled several uncorrelated samples.
-We have added a bit more discussion and context in section 7.2.5}.
+We have added a bit more discussion and context in section 7.2.5.}
 
@@
 
+\textcolor{red}{
 Our feeling is that these simulation techniques, while they have the potential to be very efficient, can be tricky in implementation that keep them from being best practice yet (likely a chance for a future LiveCoMS paper!); this is not intended as comprehensive review.
 We have:
-
 \begin{itemize}
@@
 \item Added a mention in 3.2 that specifies we're talking about equilibrium calculations.
-Added sentence at end of 3.2 mentioning these
-\item, Clarify mention of them in 7.2.4 
+\item Added sentence at end of 3.2 mentioning these
+\item Clarify mention of them in 7.2.4 
 \end{itemize}
+}
 
@@
 
-\color[red]{MRS: The section issues with soft core and $\lambda$ examines much of the clear lessons on simulation endpoints and pathways.
-We now bring up EDS and enveloping distribution approaches as alternatives in section 7.2, but at least in this paper choose not to recommend them as best practices not because they don't work, but because they at the present time there are issues with kinetic trapping and determining scaling parameters that mean they are not the recommended way to start doing these calculations for non-experts or maximal reliability.}
+\textcolor{red}{The section issues with soft core and $\lambda$ examines much of the clear lessons on simulation endpoints and pathways.
+We now bring up EDS and enveloping distribution approaches as alternatives in section 7.2, but at least in this paper, choose not to recommend them as best practices not because they don't work, but because at present there are issues with kinetic trapping and determining scaling parameters that mean they are not the recommended way to start doing these calculations for non-experts or maximal reliability.}
 
@@
 
-(10) Use larger fonts in Fig 2 &7, generally try to have equally-sized fonts throughout all Figs.
+\textcolor{red}{We added a couple of sentences to highlight this in section 7.2.5 and 8.4}
 
-\textcolor{red}{JS: PR includes increased font sizes in figs 2 and 7. Additionally, slightly reformatted pane 7B to have a cleaner horizontal line}
+(10) Use larger fonts in Fig 2 and 7, generally try to have equally-sized fonts throughout all Figs.
 
-\end{newlfm}
+\textcolor{red}{We increased the font sizes in figs 2 and 7. Additionally, slightly reformatted pane 7B to have a cleaner horizontal line}
 
+\closing{Sincerely,\\[2mm]
+% \hspace*{-1cm}\includegraphics[scale=.7]{Your signature}\\ \vspace*{-1.2cm}
+% The command above fits my signature, adjust scaling and spacing depending  on %the size of yours.
+}
+
+%-----------------------------------------------------------------------------
+
+\end{letter}
 \end{document}

### 2020-11-03 paper/manuscript.tex
adding version 1.0

@@
 \newcommand{\versionnumber}{1.0} % you should update the minor version number in preprints and major version number of submissions.
-\newcommand{\githubrepository}{\url{https://github.com/michellab/alchemical-best-practices}} %this should be the main github repository for this article
+\newcommand{\githubrepository}{\url{https://github.com/alchemistry/alchemical-best-practices}} %this should be the main github repository for this article
 \newcommand{\expect}[1]{\left\langle{#1}\right\rangle}
@@
 \author[1*]{Antonia S. J. S. Mey}
-\author[7]{Bryce K. Allen}
-\author[2]{Hannah E. Bruce Macdonald}
-\author[2*]{John D. Chodera}
+\author[2]{Bryce K. Allen}
+\author[3]{Hannah E. Bruce Macdonald}
+\author[3*]{John D. Chodera}
+\author[9]{David F. Hahn}
 \author[1,10]{Maximilian Kuhn}
 \author[1]{Julien Michel}
-\author[3*]{David L. Mobley}
-\author[11]{Levi N. Naden}
-\author[4]{Samarjeet Prasad}
-\author[2,8]{Andrea Rizzi}
+\author[4*]{David L. Mobley}
+\author[5]{Levi N. Naden}
+\author[6]{Samarjeet Prasad}
+\author[2,7]{Andrea Rizzi}
 \author[1]{Jenke Scheen}
-\author[6*]{Michael R. Shirts}
+\author[8*]{Michael R. Shirts}
 \author[9]{Gary Tresadern}
-\author[7]{Huafeng Xu}
+\author[2]{Huafeng Xu}
 %
 \affil[1]{EaStCHEM School of Chemistry, David Brewster Road, Joseph Black Building, The King's Buildings, Edinburgh, EH9 3FJ, UK}
-\affil[2]{Computational and Systems Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York NY, USA}
-\affil[3]{Departments of Pharmaceutical Sciences and Chemistry, University of California, Irvine, USA}
-\affil[4]{National Institutes of Health, Bethesda, MD, USA}
-\affil[6]{University of Colorado Boulder, Boulder, CO, USA}
-\affil[7]{Silicon Therapeutics, Boston, MA, USA}
-\affil[8]{Tri-Institutional Training Program in Computational Biology and Medicine, New York, NY, USA}
+\affil[2]{Silicon Therapeutics, Boston, MA, USA}
+\affil[3]{Computational and Systems Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York NY, USA}
+\affil[4]{Departments of Pharmaceutical Sciences and Chemistry, University of California, Irvine, USA}
+\affil[5]{Molecular Sciences Software Institute, Blacksburg VA, USA}
+\affil[6]{National Institutes of Health, Bethesda, MD, USA}
+\affil[7]{Tri-Institutional Training Program in Computational Biology and Medicine, New York, NY, USA}
+\affil[8]{University of Colorado Boulder, Boulder, CO, USA}
 \affil[9]{Computational Chemistry, Janssen Research \& Development, Turnhoutseweg 30, Beerse B-2340,Belgium}
 \affil[10]{Cresset, Cambridgeshire, UK}
-\affil[11]{Molecular Sciences Software Institute, Blacksburg VA, USA}
+
 %
@@
 \orcid{Huafeng Xu}{0000-0001-5447-0452}
+\orcid{David F. Hahn}{0000-0003-2830-6880}
 
 \blurb{This LiveCoMS document is maintained online on GitHub at \githubrepository; to provide feedback, suggestions, or help improve it, please visit the GitHub repository and participate via the issue tracker.}
-%
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-In this paper, we review current best practices for several popular application domains of alchemical free energy calculations, including relative and absolute small molecule binding free energy calculations to biomolecular targets.
+In this paper, we review current best practices for several popular application domains of equilibrium alchemical free energy calculations, including relative and absolute small molecule binding free energy calculations to biomolecular targets.
 %\todo[inline, color=green!20]{JDC: Should we migrate this document to \url{https://github.com/alchemistry}? ASJSM: Let's migrate with the submitted version, but leave it here for now?}
@@
 
-Subsequent work in the 2000s led to improved implementations of alchemical methods in popular biomolecular simulation packages~\cite{shirts2003extremely,shirts2005solvation,vanderspoel2005gromacs, mermelstein2018fast, wang2015accurate, hedges2019biosimspace}.
+Subsequent work in the 2000s led to improved implementations of alchemical methods in popular biomolecular simulation packages~\cite{shirts2003extremely,shirts2005solvation,vanderspoel2005gromacs, mermelstein2018fast, wang2015accurate, hedges2019biosimspace, riniker2011calculation}.
  This foundational work, combined with the methodological, technological, and hardware improvements of the last 5--10 years, has led to an explosion of interest and direct commercial application of these technologies~\cite{wang2015accurate, fratev2019improved, schindler2020largescale, cournia2017relative, sherborne2016collaborating, kuhn2020assessment}.
@@
 As some of the theoretical background can seem daunting, we do, however, provide a guide to the essential theory behind alchemical free energy calculations in Sec.~\ref{sec:theory}.
-In the remainder of this paper, we will cover topics that are key to the preparation( Sec.~\ref{sec:prerequisites}), choice and use of correct protocols (Sec.~\ref{sec:simulation_protocol_choice}), and finally the best practices that should be used in the analysis of alchemical calculations (Sec.~\ref{sec:data_analysis}).
+In the remainder of this paper, we will cover topics that are key to the preparation (Sec.~\ref{sec:prerequisites}), choice and use of correct protocols (Sec.~\ref{sec:simulation_protocol_choice}), and finally the best practices that should be used in the analysis of alchemical calculations (Sec.~\ref{sec:data_analysis}).
 Particular focus will be given to aspects of the molecular simulations which are unique to alchemical calculations---these include the calculation of transfer free energies (hydration free energies, partition coefficients, etc.), and binding free energies (absolute and relative).
+We primarily focus on \emph{equilibrium} free energy calculations in this best practices guide, as best practices for these are more developed, and nonequilibrium techniques may warrant their own guide as such practices evolve.
 
@@
 A number of different studies have looked at the influence of force fields and it is assumed the user has made an adequate choice for the system under study~\cite{loeffler2018reproducibility, vassetti2019assessment, lopes2015current}.
+\item nonequilibrium free energy calculations~\cite{gapsys2020large}
+\item Free energy calculations using QM/MM methods~\cite{beierlein2011simple,dybeck2016comparison,cave-ayland2015direct}.
+\item Free energy calculations using Machine learning methods~\cite{rufa2020chemical, scheen2020hybrid, cole2020machine}
 \end{itemize}
@@
      {\bf RMSE} --- Root Mean Square Error\\
-     {\bf MUE} --- Mean unsigned error\\
+     {\bf MUE} --- Mean Unsigned Error\\
      {\bf SAR} --- Structure-Activity Relationships\\
@@
 $u$ --- reduced potential describing a thermodynamic state \\
-$\Delta G$ --- Gibbs free energy (isothermal isobaric ensemble)\\
-$\Delta A$ --- Helmholtz free energy (canonical ensemble)\\
+$\Delta G$ --- Gibbs free energy (isothermal isobaric ensemble), Gibbs function, or free enthalpy, though the most common term Gibbs free energy is used in the text\\
+$\Delta A$ --- Helmholtz free energy (canonical ensemble) or Helmholtz function, with Helmholtz free energy used in the text.\\
 $\Delta f$ --- reduced (dimensionless) free energy \\
 $\Delta \hat{f}$ --- estimate from an estimator for the reduced free energy\\
-$\Gamma$ --- conformation space accessible by simulations \\
+$\Gamma$ --- configurational space accessible by simulations \\
 $\vec{q}$ --- vector of a single configuration, i.e. $x$, $y$, $z$ coordinates of the simulation system\\
@@
 The binding constant ($K_b^{\circ}$) is given by the law of mass action as the ratio of concentrations of product $[RL]$ and reactants $[R]$, $[L]$:
-\begin{equation}
+\begin{equation}\label{eq:law-mass-action}
  K_b^{\circ} = c^{\circ}\frac{[RL]}{[L][R]}.
 \end{equation}
-The standard state concentration $c^\circ$ depends on the reference state, but it is usually set to $1\,\mathrm{mol}/\mathrm{L}$ assuming a constant pressure of $1\,\mathrm{atm}$ (see also Sec.~\ref{sec:standardstate-restraints}).
-Thus, the Gibbs free energy of binding $\Delta G_{\mathrm{bind}}$ is given by:
+The standard state concentration $c^\circ$ depends on the reference state, but it is usually set to $1\,\mathrm{mol}/\mathrm{L}$ assuming a constant pressure of $1\,\mathrm{atm}$ (see also Sec.~\ref{sec:standardstate-restraints}), but also holds for dilute solutions if thermodynamic activities can be approximated by concentrations~\cite{gilson1997statisticalthermodynamic}.
+Thus, the standard Gibbs free energy of binding $\Delta G_{\mathrm{bind}}$ is given by:
 \begin{equation}
@@
 where $k_B$ is the Boltzmann constant and $T$ the temperature of the system.
+Note that, unless otherwise specified, we use $\Delta G$ throughout the paper to refer to the \textit{standard} free energy of binding (or solvation) (see also Sec.~\ref{sec:standardstate-restraints}), which is often indicated in other works with $\Delta G^{\circ}$ to differentiate it from non-standard free energies.
+This is in line with current literature on alchemical calculations, where the standard free energy is normally the only quantity of interest.
+Furthermore, we will use the term configuration for a single set of position vectors and occasionally use the term conformation to refer to a set of configurations that represent a metastable state.
 
 \paragraph{The free energy of binding can be expressed as a ratio of partition functions}
-A natural, though generally very computationally expensive, way to estimate the equilibrium constant is by directly simulating several binding and unbinding events and computing the probability of finding the receptor-ligand system in the bound state, $P(RL)$, or the unbound state, $P(R+L)$.
+The law of mass action in Eq.~\ref{eq:law-mass-action} is not directly applicable to typical molecular simulations as they normally include a single receptor/ligand in a small box (i.e., using large concentrations)~\cite{jong2011determining}.
+Instead, a natural, though generally very computationally expensive, way to estimate the equilibrium constant is by directly simulating several binding and unbinding events and computing the probability of finding the receptor-ligand system in the bound state, $P(RL)$, or the unbound state, $P(R+L)$.
 Assuming the volume change upon binding to be negligible, which is often the case at $1~\mathrm{atm}$ due to the incompressibility of water, then the Gibbs free energy $\Delta G_{\mathrm{bind}, L}$ is approximately equal to the Helmholtz free energy $\Delta A_{\mathrm{bind}, L}$, and we can simulate the system in a box of volume $V$ to obtain~\cite{jong2011determining}
@@
 where $N_{\mathrm{Av}}$ is the Avogadro number, and the last term corrects for the simulated concentration being different than the standard concentration.
-Let $\Gamma_{\mathrm{bound}}$ and $\Gamma_{\mathrm{unbound}}$ be the set of receptor-ligand conformations $\vec{q}$ that we consider bound and unbound respectively.
-The probability of a conformation $\vec{q}$ is given by the Boltzmann probability density function
+Let $\Gamma_{\mathrm{bound}}$ and $\Gamma_{\mathrm{unbound}}$ be the set of receptor-ligand configurations $\vec{q}$ that we consider bound and unbound respectively.
+The probability of a configuration $\vec{q}$ is given by the Boltzmann probability density function
 \begin{equation}
-    P(\vec{q}) = \frac{\exp\left( \beta U(\vec{q}) \right)}{\int_{\Gamma} \exp\left( \beta U(\vec{q}) \right) \, d\vec{q}} \, ,
+    P(\vec{q}) = \frac{\exp\left( -\beta U(\vec{q}) \right)}{\int_{\Gamma} \exp\left( -\beta U(\vec{q}) \right) \, d\vec{q}} \, ,
     \label{eq:conf_probability}
 \end{equation}
-where $\beta = (k_B T)^{-1}$ is the inverse temperature, $U(\vec{q})$ is the potential energy of conformation $\vec{q}$, and the integration is over the set of all possible conformations accessible in the simulation box volume $\Gamma$, with $\Gamma_{\mathrm{bound}},\Gamma_{\mathrm{unbound}} \subset \Gamma$.
-If the simulation is long enough, we expect the fraction of conformations $\vec{q}$ found in the bound state to converge to
+where $\beta = (k_B T)^{-1}$ is the inverse temperature, $U(\vec{q})$ is the potential energy of configuration $\vec{q}$, and the integration is over the set of all possible configurations accessible in the simulation box volume $\Gamma$, with $\Gamma_{\mathrm{bound}},\Gamma_{\mathrm{unbound}} \subset \Gamma$.
+If the simulation is long enough, we expect the fraction of configurations $\vec{q}$ found in the bound state to converge to
 \begin{equation}
-    P(RL) = \int_{\Gamma_{\mathrm{bound}}} P(\vec{q}) \, d\vec{q} = \frac{\int_{\Gamma_{\mathrm{bound}}} \exp\left( \beta U(\vec{q}) \right) \, d\vec{q}}{\int_{\Gamma} \exp\left( \beta U(\vec{q}) \right) \, d\vec{q}} \, .
+    P(RL) = \int_{\Gamma_{\mathrm{bound}}} P(\vec{q}) \, d\vec{q} = \frac{\int_{\Gamma_{\mathrm{bound}}} \exp\left( -\beta U(\vec{q}) \right) \, d\vec{q}}{\int_{\Gamma} \exp\left( -\beta U(\vec{q}) \right) \, d\vec{q}} \, .
 \end{equation}
@@
 \paragraph{Simulating binding events is computationally expensive}
-While simulating binding events has been used to estimate binding affinities~\cite{jong2011determining,pan2017quantitative} or to get insights into the binding pathways and kinetics of receptor-ligand systems~\cite{teo2016adaptive,votapka2017seekr,doerr2014onthefly,plattner2015protein,dixon2018predicting}, the computational cost of these calculations is usually dominated by the rate of dissociation, which can be on the microsecond timescale even for millimolar binders~\cite{pan2017quantitative} and reaches the micosecond to second timescale for a typical drug~\cite{basavapathruni2012conformational,hyre2006cooperative}.
+While simulating binding events has been used to estimate binding affinities~\cite{jong2011determining,pan2017quantitative} or to get insights into the binding pathways and kinetics of receptor-ligand systems~\cite{teo2016adaptive,votapka2017seekr,doerr2014onthefly,plattner2015protein,dixon2018predicting}, the computational cost of these calculations is usually dominated by the rate of dissociation, which can be on the microsecond timescale even for millimolar binders~\cite{pan2017quantitative} and reaches the microsecond to second timescale for a typical drug~\cite{basavapathruni2012conformational,hyre2006cooperative}.
 Depending on system size and simulation settings, common molecular dynamics software packages can reach a few hundreds of ns/day using currently available high-end GPUs~\cite{eastman2017openmm,kutzner2019more}, making these type of calculations unappealing and irrelevant on a pharmaceutical drug discovery timescale.
@@
 
-In many cases, the quantity of interest is the change in binding affinity between a compound $A$ and a related compound $B$ (e.g., by modifying one the drug scaffold's substituents, see (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{F}), which, by using Eq.~\ref{eq:binding-free-energy-from-bound-unbound-probability-ratio} and \ref{eq:bound-unbound-probability-ratio} is given by
+In many cases, the quantity of interest is the change in binding affinity between a compound $A$ and a related compound $B$ (e.g., by modifying one of the drug scaffold's substituents, see (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{F}), which, by using Eq.~\ref{eq:binding-free-energy-from-bound-unbound-probability-ratio} and \ref{eq:bound-unbound-probability-ratio} is given by
 \begin{equation}\label{eq:delta-delta-G-physical}
@@
 In particular, the second line of Eq.~\ref{eq:delta-delta-G-alchemical} is a consequence of $\Delta G_{\mathrm{unbound}}$ being independent of the presence of the receptor in the simulation box as the definition of the unbound state assumes receptor and ligand to be at a sufficient distance for them to have no energetic interactions.
-Note that, when $A$ and $B$ have different number of atoms, Eq.~\ref{eq:delta-delta-G-alchemical} implies the presence of a factor having units of volume entering both logarithms, which require unitless arguments.
-However, the value of this factor is inconsequential as it cancels out in practice.
+Note that, when $A$ and $B$ have different number of atoms, the factors $ \ln \frac{Z(RB)}{Z(RA)}$ and $\ln \frac{Z(B)}{Z(A)}$ in   Eq.~\ref{eq:delta-delta-G-alchemical} appear both to have factors with units of volume in the logarithms, but these factors exactly cancel between the terms.
 
@@
     The relative binding free energy difference between two small molecules, $\Delta \Delta G_{\mathrm{bind}, A \rightarrow B} \equiv \Delta G_{\mathrm{bind}, B} - \Delta G_{\mathrm{bind}, A}$---here benzyl alcohol (top) to toluene (bottom)---can be computed as a difference between two alchemical transformations, $\Delta G_\mathrm{bound} - \Delta G_\mathrm{solvated}$, where $\Delta G_\mathrm{bound}$ represents the free energy change of transforming $A \rightarrow B$ in complex, i.e.
-bound to a host molecule, and $\Delta G_\mathrm{solvated}$ the free energy change of transforming $A \rightarrow B$ in solvent, typically water.}
+bound to a host molecule, and $\Delta G_\mathrm{unbound}$ the free energy change of transforming $A \rightarrow B$ in solvent, typically water.}
     \label{fig:fig_binding_thermodynamic_cycle}
@@
 \end{equation}
-While it is generally not feasible to compute the two partition functions $Z(\vec{\lambda})$, several estimators have been devised to robustly estimate the ratio of partition functions in Eq.~\ref{eq:delta-G-env} (see Sec.~\ref{subsec:estimators}) from a set of conformations usually collected with MD simulations from the thermodynamic states defined at $\vec{\lambda}_A$ and $\vec{\lambda}_B$ and intermediates thereof.
+While it is generally not feasible to compute the two partition functions $Z(\vec{\lambda})$, several estimators have been devised to robustly estimate the ratio of partition functions in Eq.~\ref{eq:delta-G-env} (see Sec.~\ref{subsec:estimators}) from a set of configurations usually collected with MD simulations from the thermodynamic states defined at $\vec{\lambda}_A$ and $\vec{\lambda}_B$ and intermediates thereof.
 
@@
 While it is theoretically possible to estimate the ratio of partition functions from samples collected only at states $\vec{\lambda}_A$ and $\vec{\lambda}_B$, the efficiency of the free energy estimators rapidly decreases as the phase-space overlap between the two states also decreases~\cite{wu2005phasespaceb,wu2005phasespacec}.
-Roughly, the phase-space overlap between two thermodynamic states measures the degree to which high-probability conformations (i.e., those with very negative potential energy) in one state are also high-probability conformations in the other state (see Sec.~\ref{sec:are-they-good} and Fig.~\ref{fig:fig_what_is_lambda}).
-
-To solve the problem of having poor overlap between the states of interest, multiple intermediate alchemical states are introduced at values $\vec{\lambda}_A = \vec{\lambda}_0, \vec{\lambda}_1, \cdots, \vec{\lambda}_K = \vec{\lambda}_B$ so that each pair of consecutive states $\vec{\lambda}_k, \vec{\lambda}_{k+1}$ share good overlap.
+Roughly, the phase-space overlap between two thermodynamic states measures the degree to which high-probability configurations (i.e., those with very negative potential energy) in one state are also high-probability configurations in the other state (see Sec.~\ref{sec:are-they-good} and Fig.~\ref{fig:fig_what_is_lambda}).
+
+Equilibrium free energy calculations, our focus here, solve the problem of having poor overlap between the states of interest, multiple intermediate alchemical states are introduced at values $\vec{\lambda}_A = \vec{\lambda}_0, \vec{\lambda}_1, \cdots, \vec{\lambda}_K = \vec{\lambda}_B$ so that each pair of consecutive states $\vec{\lambda}_k, \vec{\lambda}_{k+1}$ share good overlap.
 Each intermediate state models a ligand that is neither $A$ nor $B$ but a mix of the two.
-Many estimators (e.g., exponential reweighting (EXP)~\cite{zwanzig1954hightemperature} and Bennet's acceptance ratio (BAR)~\cite{bennett1976efficient,shirts2003equilibriuma}) can then be used to compute the free energy as
+Many estimators (e.g., exponential reweighting (EXP)~\cite{zwanzig1954hightemperature} and Bennett's acceptance ratio (BAR)~\cite{bennett1976efficient,shirts2003equilibriuma}) can then be used to compute the free energy as
 \begin{equation}
@@
 While this strategy usually results in sampling thermodynamic states whose Boltzmann distributions are very similar, thus collecting information that is to some degree redundant, some estimators such as the Multistate Bennett acceptance ratio (MBAR)~\cite{shirts2008statisticallya} can exploit similarities between states to improve the precision of the estimates.
-This is achieved by using the conformations sampled at all alchemical states $\{\vec{\lambda}_k \}$ to compute the free energy difference $\Delta f(\vec{\lambda}_i, \vec{\lambda}_{j})$ between any pair of states $i,j$ (see Sec.~\ref{subsec:estimators}).
+This is achieved by using the configurations sampled at all alchemical states $\{\vec{\lambda}_k \}$ to compute the free energy difference $\Delta f(\vec{\lambda}_i, \vec{\lambda}_{j})$ between any pair of states $i,j$ (see Sec.~\ref{subsec:estimators}).
+
+Nonequilibrium free energy techniques provide an alternate approach to this problem, driving $\lambda$ between states, but these are not our focus here.~\cite{jarzynski1997nonequilibrium,jarzynski1998equilibrium,crooks2000pathensemble, gapsys2020large}
 
@@
 \label{subsec:expectation}
-Current alchemical free energy calculations involving small molecules seem to achieve, in favorable cases, root mean square (RMS) errors around 1-2 kcal/mol depending on force field, system, and a variety of other factors such as simulation time, sampling method, and whether the calculations employed are absolute or relative.
+We first note that the accuracy of any free energy calculation method will be dependent on the quality of the underlying force field.
+Therefore, any description of the force field for the molecules under study must be carefully checked to be sufficiently accurate to experiment.
+In particular, one must make sure that if using automatically generated molecular descriptions from either one's own workflow or some other computational chemistry program, there are not obvious problems in these files either through errors in the workflow, or lack of chemical coverage in the data used to construct the molecular description.
+Torsional parameters, in particular can be misassigned or improperly parameterized.
+
+Given reasonably accurate, current generation force fields correctly implemented, alchemical free energy calculations involving small molecules seem to achieve, in favorable cases, root mean square (RMS) errors relative to experiment around 1-2 kcal/mol depending on force field, system, and a variety of other factors such as simulation time, sampling method, and whether the calculations employed are absolute or relative.
 A small selection of example datasets and case studies can be found in Sec.~\ref{sec:benchmark} at the end of this document.
@@
 Furthermore, perturbations for the same chemotype in different pockets of the BACE enzyme gave varied errors~\cite{keranen2017acylguanidine}.
-Here the errors refer to the difference in $\Delta G$ derived by fitting the $\Delta \Delta G$'s to the known experimental binding free energies~\cite{wang2015accurate}.
-This prompts us to consider another important aspect.
+Here the errors refer to the difference in $\Delta G$ derived from calculated $\Delta \Delta G$'s while fitting a constant offset to best reproduce the experimental binding free energies for known compounds~\cite{wang2015accurate}.
+Each $\Delta \Delta G$ is associated with a particular free energy calculation or transformation, which can be thought of as an edge in the graph spanning the compound series, see examples of such graphs in Fig.~\ref{fig:fig_types_of_networks}.
+
+The fact that we can analyze both $\Delta G$ values and $\Delta \Delta G$ values raises an important question about analysis -- which calculated values should we assess?
 It is important to be clear on what error to report: $\Delta G$ after shifting by a constant to minimize the RMSE, unshifted $\Delta G$, $\Delta \Delta G$ of computed edges, or $\Delta \Delta G$ of all edges.
 (See recommendations for reporting best practices, Sec.~\ref{sec:plot_data}.) Additionally, as it is possible to perform calculations on a set of ligands using different pairwise comparisons of molecules, the performance of the method may be biased based on which pairs of comparisons are performed.
-Additionally, it is possible that the error associated to the relative free energy between a two ligands that was not directly computed (however can be deduced using thermodynamic paths involving other ligands) will likely be more uncertain \url{https://github.com/jchodera/jacs-dataset-analysis}.
+Additionally, it is possible that the error associated with the relative free energy between a two ligands that was not directly computed (however can be deduced using thermodynamic paths involving other ligands) will likely be more uncertain.
 Given the need to understand the performance of the system with alchemical free energy calculations, we recommend that retrospective studies for a particular target and a particular chemical series be performed for each application case.
@@
 \label{subsec:reproducible}
-Finite computing resources necessarily limit the generated number of uncorrelated samples of potential energy surfaces, and therefore alchemical free energy calculations only give free energy estimates to within finite precision.
+We restrict our analysis here to repetitions of the same calculation performed with precisely the same force field - different force fields used to describe the same molecule can lead in some cases to wide difference in free energies.
+Even within this restriction,
+finite computing resources necessarily limit the generated number of uncorrelated samples of potential energy surfaces, and therefore alchemical free energy calculations only give free energy estimates to within finite precision.
 An important consideration is how reproducible alchemical free energy calculations are in practice.
@@
 For more complex use cases such as protein-ligand binding free energies the repeatability is often substantially worse~\cite{rizzi2019sampl6}.
-A good practice is to perform two or three runs of the same perturbation to assess precision with a given protocol.
+A good practice is to perform two or three runs of the same perturbation to assess precision with a given protocol, using different initial conditions (i.e.
+velocities).
 The sample standard deviation will give a crude estimate of the reliability of the estimates, and whether the precision is sufficient for the problem at hand.
@@
 \begin{equation}
-  K_D = K_D^\ast \frac{1 + K_C}{K_C}
+  K_b = K_D^\ast \frac{1 + K_C}{K_C}
   \label{eqn:conformational-binding}
@@
 For illustrative purposes, consider a proposed set of new molecules, and assume that the changes proposed in these molecules yield a set of relative binding free energies that follow a normal
-distribution. That is, assume that the standard deviation in the relative binding free enrgies for the changes represented is $RT\ln 5$
+distribution. That is, assume that the standard deviation in the relative binding free energies for the changes represented is $RT\ln 5$
 (corresponding to a 5-fold change in the binding affinities), and that
@@
 
-If the accuracy of free energy calculations is $\sigma = 1$ kcal/mol, $P(\Delta\Delta
+If the accuracy of a collection of free energy calculations is $\sigma = 1$ kcal/mol, $P(\Delta\Delta
 G<0|\Delta\Delta G^\dagger<0) = 0.35$, which means that out of every
@@
 There is no detailed study of how different alignment approaches may affect results, but the user should be aware of some practical considerations.
-Tools are available to compare the ligands and build the hybrid topologies that define the changes between one ligand and another~\cite{loeffler2015fesetup,hedges2019biosimspace,gapsys2015pmx}.
+Tools are available to compare the ligands and build the combined topologies that define the changes between one ligand and another~\cite{loeffler2015fesetup,hedges2019biosimspace,gapsys2015pmx}.
 In simple terms, providing poor alignment to these tools will make this job harder.
@@
 It is tempting to only synthesize the compounds predicted to be most active, thus a narrow range of calculated activity is tested that imposes limits on the statistical assessment of performance, ideally example molecules from across the range of predicted activity can be assessed or corrections can be applied based on previous recommendations~\cite{abel2017critical}.
-For a more detailed discussion on checking the robustness of your alchemcial free energy calculation see also Sec.~\ref{sec:are-they-good}.
+For a more detailed discussion on checking the robustness of your alchemical free energy calculation see also Sec.~\ref{sec:are-they-good}.
 
@@
 Consider performing repeat calculations for at least some of the perturbations in the study.
- There are many accounts of success of alchemical calculations, the methods show good performance towards the goal of binding energy prediction.
+ There are many accounts of success of alchemical calculations, the methods show good performance towards the goal of binding free energy prediction.
 However, it is important to have realistic expectations.
@@
 Structure based drug design projects are often capable of improving potency relatively quickly, even with only limited application of computational approaches and the range of activity narrows to just two-to-three log units.
-It may seem hard to have impact with substantially different, more potent, stand-out compounds in this scenario, but binding energy prediction can still be extremely useful for ensuring activity is maintained as other properties are optimized.
+It may seem hard to have impact with substantially different, more potent, stand-out compounds in this scenario, but binding free energy predictions can still be extremely useful for ensuring activity is maintained as other properties are optimized.
 An interesting cost benefit analysis has shown the value of activity prediction, see discussion above and articles such as~\cite{mobley2012perspective}.
@@
 \label{subsec:binding}
-In principle, in the limit of sufficient conformational sampling, the free energy changes estimated from an alchemical free energy calculation should be independent of the system's initial coordinates.
+In principle, in the limit of sufficient configurational sampling, the free energy changes estimated from an alchemical free energy calculation should be independent of the system's initial coordinates.
 However, in practice, because simulations are of finite duration (typically 1-100 ns per state at present), this is only true for certain classes of alchemical free energy calculations such as relative or absolute free energies of hydration of small and relatively rigid organic molecules.
@@
 Checking the X-ray structure versus the experimental electron densities is important, as the position of part of the ligand or important sidechains may be based on the interpretation of the crystallographer rather than the available electron density, especially in cases of missing density.
-For example, looking at a cyclohexane ring density, a chair conformation is vastly more likely than that of a boat and, if a boat conformation is present in the structure, it may be worth inspecting the density to ensure it adequately supports this choice.
+For example, looking at a cyclohexane ring density, a chair configuration is vastly more likely than that of a boat and, if a boat configuration is present in the structure, it may be worth inspecting the density to ensure it adequately supports this choice.
 
@@
 
-\paragraph{Have you considered sterioisomers and enantiomers?}
+\paragraph{Have you considered stereoisomers and enantiomers?}
 Congeneric series can contain stereoisomers or enantiomers which can bind very differently, resulting in large errors if treated incorrectly.
@@
 \paragraph{Topologies} 
-A critical first step in relative calculations is to select an approach to these calculations, determining whether to use a \emph{dual topology}, \emph{single topology}, or \emph{hybrid topology} approach to relative calculations.
-
-The distinction between these can be illustrated by considering a hypothetical transformation from molecule A to molecule B, where both atoms share a common substructure but differ in their substituents; e.g.
+A critical first step in relative calculations is to select the approach to be used for these calculations and determine what this means for atom mapping (though this may sometimes be dictated by the requirements of the chosen simulation package.
+Sometimes this approach is described in terms of a \emph{dual topology} versus \emph{single topology} approach to relative calculations.
+
+The distinction between single and dual topology can be illustrated by considering a hypothetical transformation from molecule A to molecule B, where both atoms share a common substructure but differ in their substituents; e.g.
 consider a transformation of benzene to benzyl alcohol Fig.~\ref{fig:fig_topology}.
@@
 
-In contrast, in a dual topology alchemical free energy calculation, no atoms are allowed to change type~\cite{shirts2012best}.
+In contrast, in a dual topology alchemical free energy calculation, no atoms are allowed to change type~\cite{Boresch:1999:J.Phys.Chem.A, shirts2012best}.
 This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxy methyl group, then passing through an intermediate state where some atoms are partially interacting--- particularly, those atoms which are becoming dummy atoms or ceasing to be dummy atoms~\cite{mobley2014blind}.
@@
 Fig.~\ref{fig:fig_topology}'s right branch depicts how such a dual topology works.
+
 \begin{figure}
@@
 \textbf{Right}: The dual topology does not convert one species to another, but only converts between Du atoms and an interacting species, but usually uses softcore potentials for this.
-The 'mixed' intermediate atoms are used in both dual and single topology approaches.
+The 'mixed' intermediate atoms (M) are used in both dual and single topology approaches.
 Only the way the transformation occurs and the end states differ.
@@
 
-Hybrid topology calculations have seen much less use but essentially consist of two absolute free energy calculations in opposite directions at the same time, turning one molecule's interactions with the environment off, while turning the other molecule's interaction on~\cite{jiang2019computing, rocklin2013separated}.
-
-Most existing software implementations currently use single or dual topology approaches; for example, AMBER TI uses a dual topology approach, while BioSimSpace uses a single topology approach.
-Please make sure to check what approach is used with your software package of choice, or whether it supports your choice of approach (GROMACS, for example, supports both).
+As far as we are aware, the distinction between single and dual topology approaches is made primarily for two main reasons.
+First, this choice affects the alchemical pathway followed, and thus may affect convergence properties -- though we are not yet aware of a study of the relative efficiency and merits of these two approaches.
+Additionally, historically, some simulation packages one approach and not the other, meaning that the distinction was functional.
+
+Some additional terms are also employed.
+Particularly, some studies refer to a ``hybrid'' topology approach to free energy calculations~\cite{gapsys2015pmx, gapsys2016accurate, gapsys2020large}, though this term may not have yet achieved widespread use.
+In this case, "hybrid" seems to indicate that the set up of these free energy calculations involves a hybrid of the two molecules, and much of what is done in these studies uses a single topology approach~\cite{gapsys2020large}.
+
+One final approach, so far in its infancy, has been called ``separated topologies'' and essentially consist of two absolute free energy calculations in opposite directions at the same time, turning one molecule's interactions with the environment off, while turning the other molecule's interaction on~\cite{jiang2019computing, rocklin2013separated}.
+
+Software packages vary in their use single or dual topology approaches; for example, AMBER TI uses a dual topology approach, while BioSimSpace uses a single topology approach.
+Please make sure to check what approach is used with your software package of choice, or whether it supports your choice of approach (GROMACS and GROMOS, for example, support both).
 To our knowledge efficiency differences have not been thoroughly explored, though conventional wisdom suggests that fewer dummy atoms are better, as introducing or removing atomic sites is usually more difficult, requiring more intermediate steps ~\cite{liu2013lead, mobley2012perspective}.
@@
     \includegraphics[width=0.95\linewidth]{figures/fig4_mcs/Figure.pdf}
-    \caption{\textbf{Illustration of maximum common substructure matches} MCSS is shown in green for when (\textbf{A}) a restrictive MCSS match is used and in (\textbf{B}) ring breaking is allowed.}
+    \caption{\textbf{Illustration of maximum common substructure matches} MCSS is shown in green for when (\textbf{A}) a restrictive MCSS match is used and in (\textbf{B}) ring breaking is allowed, meaning there is no MCSS match between the two compounds.}
     \label{fig:fig_mcss}
@@
 A large number of software tools can compute MCSS matches using different cheminformatics packages.
-Some rely on RDKit~\cite{rdkit2019Dec}, such as LOMAP~\cite{liu2013lead}, FESetup~\cite{loeffler2015fesetup} and partially BioSimSpace~\cite{hedges2019biosimspace}, while others such as fkckombu~\cite{kawabata20143d} are standalone tools.
+Some rely on RDKit~\cite{rdkit2019Dec}, such as pmx~\cite{gapsys2015pmx}, LOMAP~\cite{liu2013lead}, FESetup~\cite{loeffler2015fesetup} and partially BioSimSpace~\cite{hedges2019biosimspace}, while others such as fkckombu~\cite{kawabata20143d} are standalone tools.
 Schr\"{o}dinger's FEP+ planning tool was originally based on a version of LOMAP, and it also uses MCSS matching as well as 3D considerations to plan the network of single topology calculations between molecules~\cite{wang2015accurate}.
@@
 It is possible that changes in binding mode could actually require a different choice of mapping, so in some cases mappings may need to be planned differently depending on 3D positioning of atoms in space.
+It is worth visual inspection prior to simulation to ensure that the mapping criteria corresponds to the expected binding mode.
+If the mapping protocol returns simulations that correspond to different binding modes of a ligand within a perturbation map, which can cause large hysteresis.
 
@@
 are part of a congeneric series, see Fig.~\ref{fig:fig_mcss}.
-If no common substructure is shared, then alternative dual or hybrid topology free energy calculations are needed, where one would co-localize a pair of compounds in a binding site, exclude their interactions with one another, and compute the relative binding free energy by turning one molecule on from being dummy atoms while turning the other off.
+If no common substructure is shared, then alternative dual or separated topology free energy calculations are needed, where one would co-localize a pair of compounds in a binding site, exclude their interactions with one another, and compute the relative binding free energy by turning one molecule on from being dummy atoms while turning the other off.
 To our knowledge no general pipeline for such calculations yet exists and this would likely remain a research problem.
@@
 However, in single topology relative free energy calculations, the atoms involved might be mutated to other atom types -- for example, in a mutation of methane to methanol, one hydrogen might become an oxygen atom.
-Typical molecular dynamics engines are not set up to recognize this change, or at least not to correctly include contributions to the free energy from changing constraints/constraint length, so results for a transformation would usually be erroneous.
+Some molecular dynamics engines are not set up to recognize this change, or at least not to correctly include contributions to the free energy from changing constraints/constraint length, so results for a transformation can be erroneous.
 At present the most general solution to this problem is simply to avoid the use of constraints (and thus use a smaller timestep if necessary, usually of around 1 fs) in any relative free energy calculation involving a transformation of a constrained bond.
-When using TI in GROMACS $\frac{dH}{d\lambda}$ calculations have contribution from the LINCS constrained bonds, thus transforming constrained bonds in GROMACS is possible.
+Individual software programs and settings can handle such issues.
+For example, bonds can be transformed in both GROMACS and GROMOS, because contributions of LINCS (GROMACS) and SHAKE (GROMOS) contrained bonds are added to the $\frac{dH}{d\lambda}$ term~\cite{pearlman1993determining, straatsma1992holonomica, pearlman1991overlooked, gunsteren1997computer}.
+However, the constraints are not taken into account when calculating energy differences to other intermediate states (as the effect is entropic, not energetic).
+User manuals should carefully checked for how these effects are included if a constrained bond changes in length.
+We do note that if one performs calculations of the changing constraint in both solution and in the binding site, then the errors often cancel substantially, but the cancellation is not guaranteed.
 
@@
 It should be noted that ignoring the term $c^{\circ}$ is equivalent to assuming a reference concentration of 1~D$^{-1}$, where D are the units used to express $K_b$, and would thus cause the value of $\Delta G$ to vary with the choice of the units.
-It is convenient to define a standard state at a constant pressure of 1~atm and where each chemical species (i.e., A, B, and AB) in the reaction solvent has a concentration of $c^{\circ}$~=~1~M~=~1~molecule/1660~\r{A}$^3$ but do not interact with other molecules of $R$, $B$, or $RL$.
+It is convenient to define a standard state at a constant pressure of 1~atm and where each chemical species (i.e., R, L, and RL) in the reaction solvent has a concentration of $c^{\circ}$~=~1~M~=~1~molecule/1660~\r{A}$^3$ but do not interact with other molecules of $R$, $L$, or $RL$.
 
@@
 \begin{equation}\label{eq:restraint-correction}
-    \Delta G^{\circ}_{\mathrm{restr}} = -k_BT ~ \ln \left( \frac{c^{\circ}}{V_L} \right) -k_BT ~ \ln \left( \frac{\xi_L}{8 \pi^2} \right),
+    \Delta G^{\circ}_{\mathrm{restr}} = -k_BT ~ \ln \left( c^{\circ}{V_L} \right) -k_BT ~ \ln \left( \frac{\xi_L}{8 \pi^2} \right),
 \end{equation}
@@
 In addition, when enhanced sampling methods such as Hamiltonian $\lambda$ exchange are used (see Sec.~\ref{sec:sampling_schemes}), the use of a restraint is typically necessary as it keeps the ligand in the binding site in the interacting state (see also Sec.~\ref{sec:weak-binders}) and generally reduce the round-trip time of replicas.
-When restraint are employed, the values of $V_L$ and $\xi_L$ are restraint-dependent, but for commonly employed restraints, these can be usually easily computed analytically or numerically by solving the relevant integral.
+When restraints are employed, the values of $V_L$ and $\xi_L$ are restraint-dependent, but for commonly employed restraints, these can be usually easily computed analytically or numerically by solving the relevant integral.
 
@@
 We stress that this is normally avoided as it generally introduces bias in the free energy estimate, which is why the restraint is usually activated only in the intermediate states in absolute calculations.
-The bias can be corrected through reweighting schemes~\cite{rizzi2019sampl6}, but this post-processing step can be avoided if a flat-bottom restraint is used and the ligand never hits the potential wall during the simulation in the bound state.
+The bias can be corrected through reweighting schemes~\cite{rizzi2019sampl6}, but this post-processing step can be avoided if a flat-bottom restraint is used and the ligand is never sampled while hitting the potential wall during the simulation in the bound state as the numerical correction will be exactly zero.
 It is important to note that the spring constant and/or radius parameters of the restraint effectively determine which conformations are considered to be bound.
@@
 In absolute calculations, this restraint can substitute or be added to the restraint used to handle the standard state correction (Sec.~\ref{sec:standardstate-restraints}).
-In the latter case, to compute the standard state correction analytically, the bound-state restraint must be turned off in the decoupled state.
-Alternatively, a flat-bottom restraint can be activated also in the decoupled state as long as the second restraint (e.g., a harmonic or Boresch restraint) prevents the ligand to hit the wall of the flat-bottom potential~\cite{rizzi2019sampl6}.
+In the latter case, however, care must be taken when computing the standard state correction.
+When multiple restraining potentials are active in the non-interacting state, the correction can generally be computed only through numerical integration.
+Alternatively, one can adopt a protocol that removes the bound-state restraint in the non-interacting state.
 Finally, even for tight binders, dissociation events can be enhanced by methods such as Hamiltonian replica exchange~\cite{sugita2000multidimensionala,chodera2011replica,wang2013identifying} and expanded ensemble~\cite{lyubartsev1992newa,li2007simulated}, especially in absolute free energy calculations using harmonic or flat-bottom restraints.
@@
 However, application of such corrections typically remains less common than the use of a co-alchemical ion.
+A third approach has been proposed by Gapsys et al.~\cite{gapsys2015calculation} which uses a double-system/single box setup.
 
@@
 \begin{equation}
-U(\vec{q},\vec{\lambda}) = \vec{\lambda} U_0(\vec{q}) + (1-\vec{\lambda})U_1(\vec{q}) \end{equation}
+U(\vec{q},\vec{\lambda}) = (1-\vec{\lambda}) U_0(\vec{q}) + \vec{\lambda}U_1(\vec{q}) \end{equation}
 
@@
 One way to look at this is to examine how low $\vec{\lambda}$ values must go to reduce the energy at $0.5\sigma$ (the atomic size parameter) down to 1 $k_BT$, where thermal fluctuations make it possible for other atomic sites to penetrate routinely that deep.
-If we are trying to go from a particle being present, and desire to make it disappear alchemically, then if the repulsive terms are of the form $\epsilon\frac{\sigma}{r}^{12}$, then if $\epsilon$ was 1 $k_BT$ at the temperature of interest, and we start with the particle present, then solving for $(1-\vec{\lambda})(1 kB_T)\left(\frac{\sigma}{\frac{1}{2}\sigma}\right)^{12} = 1 k_B T$ we get $\vec{\lambda} = 1-2^{-12} \sim 0.999976$.
+Assume we are trying to go from a particle being present, and desire to make it disappear alchemically.
+If the repulsive terms are of the form $\epsilon(\frac{\sigma}{r})^{12}$, and if $\epsilon$ is 1 $k_BT$ at the temperature of interest, and we start with the particle present, we may solve for $(1-\vec{\lambda})(1 k_B T)\left(\frac{\sigma}{\frac{1}{2}\sigma}\right)^{12} = 1 k_B T$.
+This yields $\vec{\lambda} = 1-2^{-12} \sim 0.999976$.
 At this point,  we have gone virtually all the way to the end of the transformation, but there is still an impenetrable post in the middle of our simulation!
@@
 What we need instead is a function that smoothly gets rid of this infinity.
-A large number of schemes have been tried~\cite{beutler1994avoiding, zacharias1994separationshifted, blondel2004ensemble, pham2011identifying, pham2012optimal, naden2014linear, donnini2005molecular}, but the most common strategy that appears to be the best practice is to use a "soft-core" potential, of the form:
+A large number of schemes have been tried~\cite{beutler1994avoiding, zacharias1994separationshifted, blondel2004ensemble, pham2011identifying, pham2012optimal, naden2014linear, donnini2005incorporating}, but the most common strategy that appears to be the best practice is to use a "soft-core" potential, of the form:
 
@@
 where $r_{ij}$ is the distance between two particles $i$ and $j$, $\epsilon_{ij}$ and $\sigma_{ij}$ are the Lennard-Jones parameters corresponding to the interaction between particles $i$ and $j$, and $\alpha$ is a constant (0.5 is optimal for the specific functional form shown above).
-This functional form has exactly the property we are looking for: it recovers the Lennard-Jones potential when $\vec{\lambda}=1$, and the other endpoint ($\vec{\lambda}=0$), it is exactly zero for all $r_{ij}$ everywhere, and as $\vec{\lambda}$ goes to zero, the $\alpha(1-\vec{\lambda})$ term lowers the infinite energy in the core.
+This functional form has exactly the property we are looking for: it recovers the Lennard-Jones potential when $\vec{\lambda}=1$, and the at other endpoint ($\vec{\lambda}=0$), it is exactly zero for all $r_{ij}$ everywhere, and as $\vec{\lambda}$ goes to zero, the $\alpha(1-\vec{\lambda})$ term lowers the infinite energy in the core.
 There are several different variants of the same functional form~\cite{zacharias1994separationshifted, beutler1994avoiding,pham2011identifying}, but the one given in eq.~\ref{eq:softcore} is easy to understand and implement and fairly numerically stable.
@@
 Although this is a simpler path to understand cognitively and can take advantage of the soft-core potential from eq~\ref{eq:softcore}, this can lead to more intermediate steps and thus be more computationally expensive.
-Other schemes, such as simultaneously changing electrostatic and Lennard-Jones interactions with electrostatic ``soft-core'' potentials~\cite{steinbrecher2007nonlinear} may be implemented with fewer intermediate but could require fine-tuning of electrostatic and Lennard-Jones softcore parameters to avoid numerical instabilities.
+Other schemes, such as simultaneously changing electrostatic and Lennard-Jones interactions with electrostatic ``soft-core'' potentials~\cite{steinbrecher2007nonlinear} may be implemented with fewer intermediate states but could require fine-tuning of electrostatic and Lennard-Jones softcore parameters to avoid numerical instabilities.
 At the time of writing, there has not been conclusive evidence to suggest one approach is better in general than the other, so discretion should be left up to the user as to what is viable from both hardware resources, and what the simulation software supports.
@@
 The spacing depends to some extent on the choice of analysis method, though states should essentially be spaced equidistant in the relevant thermodynamic length~\cite{crooks2007measuringa, sivak2012thermodynamic}.
-For BAR/MBAR techniques this means that states should be spaced so that the statistical uncertainties between neighboring states be equal;~\cite{pham2012optimal, shenfeld2009minimizing}
+For BAR/MBAR techniques this means that states should be spaced so that the statistical uncertainties between neighboring states be approximately equal~\cite{pham2012optimal, shenfeld2009minimizing}, where "approximately" is roughly within 30-50\% in magnitude.
 Some schemes to adaptively optimize the spacing of intermediate states based on initial exploratory simulations have been proposed~\cite{hayes2017adaptive}.
 For molecules changing in dense solvent, then the best path is roughly independent of molecule size and shape, so what works for one molecular transformation is likely to be relatively efficient for another~\cite{monroe2014converging}.
+
+Some approaches have attempted to find alternative pathways to improve efficiency or find paths of low thermodynamic length~\cite{naden2014linear,naden2015linear,pham2012optimal}.
+For example, the enveloping distribution sampling (EDS) approach, and its multiple-replica and accelerated variant, attempts to improve efficiency by creating a single artificial intermediate state which simultaneously samples all end state phase spaces~\cite{perthold2018accelerateda,sidler2017efficientb, christ2007envelopinga}.
+When this can be done, it provides an extremely efficient way to calculate the relative free energy difference between multiple ligands from a single simulation.
+However, it can often fail whenever the simulation of this intermediate state ends up trapped in configurations characteristic of only one end state.
+Thus, successful use of EDS can require system-dependent tuning making it difficult to implement in an automated and reliable way, but when successful it can be very efficient.
+
+In our view, there is still some room for further exploration of how to best choose transformation pathways, especially for relative binding calculations or more complex molecules, as most existing studies focused on smaller molecules.
+In principle, any pathway that connects the desired end states is rigorously correct, but as discussed above, different paths may differ dramatically in thermodynamic length and therefore efficiency.
+Additionally, some paths simply may not converge due to issues noted above such as those encountered without soft core potentials.
+However, the recommendations above are reasonable, reliable, and are likely not be that much less efficient than potentially more optimal choices~\cite{naden2014linear,naden2015linear,pham2012optimal}, as the real problems with low efficiency are lack of sampling of slow conformational modes.
+
+It is however important to note that different packages also differ in how they handle implementation of alchemical transformations, making it difficult to give rules of thumb concerning specific efficient transformations which work equally well across simulation packages.
+Thus we are hesitant to recommend best practices within specific software packages at point in time, although any good transformation pathway will conform to the guidelines we have outlined above.
 
@@
 The basic idea of the replica exchange scheme is shown in Fig.~\ref{fig:fig_sampling_scheme} \textbf{B}.
-It is supported in various software packages that provide alchemical implementations, such as GROMACS~\cite{aldeghi2015accurate}, FEP+~\cite{wang2015accurate}, and NAMD~\cite{jiang2019computing}.
+It is supported in various software packages that provide alchemical implementations, such as GROMACS~\cite{aldeghi2015accurate}, GROMOS~\cite{hritz2008hamiltoniana,hritz2007optimizationa}, FEP+~\cite{wang2015accurate}, and NAMD~\cite{jiang2019computing}.
 A third approach borrows ideas from simulated tempering~\cite{marinari1992simulateda}.
@@
 Each arrow represents a short simulation interval before an exchange through Metropolis Hastings acceptance (dice) is attempted.
-A tick means an accepted exchange a cross a rejected exchange.
+A tick means an accepted exchange, a cross a rejected exchange.
 (\textbf{C}): Single replica scheme sampling from all $\vec{\lambda}$ states.
@@
 Currently, we recommend using Hamiltonian replica exchange type sampling schemes (Fig.~\ref{fig:fig_sampling_scheme} \textbf{B}).
-If these are not available in the code of choice, running independent simulations at different $\vec{\lambda}$ values can be acceptable, especially when conformational sampling is fast (Fig.~\ref{fig:fig_sampling_scheme} \textbf{A}).
+If these are not available in the code of choice, running independent simulations at different $\vec{\lambda}$ values can be acceptable, especially when configurational sampling is fast (Fig.~\ref{fig:fig_sampling_scheme} \textbf{A}).
 Single replica schemes and non-equilibrium schemes are not as established yet, but are very promising.
@@
 The condition of minimizing the statistical uncertainty of different free energy estimators below a sufficient threshold should be one metric monitored over the simulation.
-This can be done through the uncertainty estimator built into certain analysis tools such as MBAR, or can be done though more general statistical tools like bootstrap sampling.
+This can be done through the uncertainty estimator built into certain analysis tools such as MBAR, or can be done through more general statistical tools like bootstrap sampling.
+It should be noted however, that uncertainty estimates have the same limitations as other metrics of convergence, as they are only an uncertainty based on the phase space sampled so far in a simulation, and cannot account for states not sampled, and it is worth considering that they will be an underestimation of the true uncertainty.
 A target statistical uncertainty should be chosen at the onset of the simulation to avoid excessively long simulations, or falling into the trap of running until the free energy estimate is "good enough," which is subjective and has no defined criteria.
-This could be a fixed value such as $0.20 \mathrm{kcal/mol}$, or a functional quantity such as "below $0.5 \mathrm{kcal/mol}$ and $10\%$ of the free energy estimate." The user does not need to monitor this information in real-time and can choose to run simulations for fixed duration (either time or number of samples) and run analysis on the data collected thus far.
+This could be a fixed value such as $0.20$ $\mathrm{kcal/mol}$, or a functional quantity such as "below $0.5$ $\mathrm{kcal/mol}$ and $10\%$ of the free energy estimate." The user does not need to monitor this information in real-time and can choose to run simulations for fixed duration (either time or number of samples) and run analysis on the data collected thus far.
 If more samples are needed, the simulations can be resumed, or, started again in different initial conditions.
@@
 Convergence in other alchemical observables should also be monitored to determine if the defined phase space has been sufficiently sampled and enough decorrelated samples have been drawn.
-These additional observables include, but are not limited to, the variance in $\frac{dU}{d\vec{\lambda}}$ across all $\vec{\lambda}$ values, calculating the variance in free energy using bootstrap analysis, and comparing differences in free energies calculated using different percentages of the simulation in both the forward and reverse directions ( see Fig.
+These additional observables include, but are not limited to, the variance in $\frac{dU}{d\vec{\lambda}}$ across all $\vec{\lambda}$ values, calculating the variance in free energy using bootstrap analysis, and comparing differences in free energies calculated using different percentages of the simulation in both the forward and reverse directions~\cite{klimovich2015guidelines} (see Fig.
 ~\ref{fig:convergence_forward_reverse}).
@@
     \includegraphics[width=0.95\linewidth]{figures/fig10_forward_reverse/Figure.pdf}
-    \caption{\textbf{Free energy (in k$_{B}$T) for two different relative binding free energy perturbations.} 
+    \caption{\textbf{Free energy (in $k_{B}T$) for two different relative binding free energy perturbations.} 
     Each plot shows the estimated free energy change using a varying fraction of total simulation time (up to 5 ns total). 
@@
 The first step of the protocol is the decharging then removing van der Waals interactions and then recharging.
-The difference in energy between the forward (blue) and reverse (red) free energy calculations at the midpoint of the simulation time gives an indication of the overall convergence of the simulation, with differences over 1 k$_{B}$T indicating poor convergence.}
+The difference in energy between the forward (blue) and reverse (red) free energy calculations at the midpoint of the simulation time gives an indication of the overall convergence of the simulation, with differences over 1 $k_{B}T$ indicating poor convergence.}
     \label{fig:convergence_forward_reverse}
@@
 
-Each of these metrics have demonstrated promising results for diagnosing when a simulation has a convergence issue beyond simple convergence of uncertainty estimate. 
+Each of these metrics shows some promise for diagnosing when a simulation has a convergence issue beyond simple convergence of uncertainty estimates. 
 Results obtained from calculations with convergence issues should be checked for errors or run for longer before any confidence should be placed in conclusions drawn from their analysis.
@@
     \centering
-    \includegraphics[width=0.8\linewidth]{paper/figures/fig_pmf/pmf.png}
+    \includegraphics[width=0.8\linewidth]{figures/fig_pmf/pmf.png}
     \caption{{\bf Potential of mean force with respect to $\vec{\lambda}$ for TI and MBAR}
@@
 
-The uncertainty in the free energy, for example, has multiple ways to be estimated, e.g.
+The uncertainty in the free energy, for example, can be estimated in multiple ways, e.g.
 through standard error propagation methods (including MBAR's estimator, which is based on the same principles as standard error propagation), through bootstrap methods, through multiple independent runs, etc.
@@
 
+In general, uncertainties can be assumed to decrease as $1/\sqrt(N)$ where $N$ is the number of uncorrelated samples, for all standard free energy calculation methods~\cite{shirts2005comparison}.
+However, this carries the notable caveat that such estimates require accurate estimation of the correlation time which, if important motions are slow compared to simulation timescales, can be difficult.
+Still, this metric provides a good rule of thumb, and as long as conformational transitions are captured by the simulation, increasing the aggregate simulation time by a factor of $T$ will reduce the uncertainty by a factor of approximately $\sqrt(T)$.
+
 \subsubsection{Multiple or uncertain binding modes may require considerable care}
@@
 
-Absolute free energy calculations for dissimilar ligands can have particular challenges because the (potentially incorrect) assumption of consistent binding modes across a series of similar ligands is likely to be even less robust than the in the case of relative calculations.
+Absolute free energy calculations for dissimilar ligands can have particular challenges because the (potentially incorrect) assumption of consistent binding modes across a series of similar ligands is likely to be even less robust than in the case of relative calculations.
 This means that researchers performing absolute binding free energy calculations will have to pay particular attention to generating reasonable putative binding modes.
@@
 In some cases, it is tempting to simply use docking techniques to generate initial bound structures for starting molecular dynamics simulations.
-However, timescales for binding mode interconversion are usually slow compared to MD/free energy timescales, meaning that simulations started from different potential binding modes are likely to yield disparate computed binding free energies~\cite{mobley2006use, palma2012computation, mobley2012perspective, gill2018binding} .
+However, timescales for binding mode interconversion are usually slow compared to MD/free energy timescales, meaning that simulations started from different potential binding modes are likely to yield disparate computed binding free energies~\cite{mobley2006use, palma2012computation, mobley2012perspective, gill2018binding}.
 And docking techniques are good at identifying sterically reasonable potential binding modes, but still perform relatively poorly at identifying a single dominant binding mode \emph{a priori}.
@@
 
-Sampling all potential lignad binding modes within a single free energy calculation is usually impractical without some form of enhanced sampling or at least Hamiltonian replica exchange~\cite{wang2013identifying} because barriers for binding mode interconversion result in kinetics which are too slow compared to simulation timescales~\cite{mobley2006use, palma2012computation,mobley2012perspective, gill2018binding}.
+Sampling all potential ligand binding modes within a single free energy calculation is usually impractical without some form of enhanced sampling or at least Hamiltonian replica exchange~\cite{wang2013identifying} because barriers for binding mode interconversion result in kinetics which are too slow compared to simulation timescales~\cite{mobley2006use, palma2012computation,mobley2012perspective, gill2018binding}.
 Hamiltonian exchange, coupled with appropriate restraints, can allow the ligand to relatively rapidly exchange between potential binding modes when non-interacting, accelerating sampling of binding modes~\cite{wang2013identifying}.
@@
 Overall, this likely means that relative free energy calculations will be susceptible to problems resulting from uncertainty in ligand binding modes until more robust approaches are available to determine dominant binding modes, or the relative populations of different potential binding modes, in advance.
+
+\subsection{General simulation setting choices that could affect free energy calculations.}
+There are many parameter choices that are common to standard MD simulations.
+Optimal choices may depend on the simulation package and it is best to consult the manual of each package to make the right choices.
+One particular parameter we want to draw attention to which is often overlooked is the timestep.
+The choice of integrator can drastically affect the accuracy of a given calculation depending on the timestep~\cite{leimkuhler2016efficient}.
+Using a 1.0 fs timestep near the limit of stability without any constraints can lead to non-negligible statistical mechanical errors.
+However, from various previous free energy studies using 1 fs integration timesteps it is not clear that a significant error would be introduced into a free energy estimate and may warrant some further investigation.
 
@@
 \label{sec:data_analysis}
-Once data has been collected from alchemical intermediates, it must be analyzed to produce an estimate of the free energy change (and its associated statistical uncertainty) for each leg of the thermodynamic cycle.
+Once equilibrium data has been collected from alchemical intermediates, it must be analyzed to produce an estimate of the free energy change (and its associated statistical uncertainty) for each leg of the thermodynamic cycle.
 While a number of different estimators are available that will give consistent results under optimal circumstances, some approaches are recommended over others due to their robustness and ability to provide information on poor convergence.
@@
 For simulations that are simply not long enough to contain a large number of samples from true equilibrium either because they are very short or contain slow processes, this procedure cannot completely remove the bias.
-In such cases, this approach simply selects the final portion of the the simulation, which may be contained in a single substate of conformational space, and may itself lead to biased estimates. 
+In such cases, this approach simply selects the final portion of the the simulation, which may be contained in a single substate of configurationaal space, and may itself lead to biased estimates. 
 This situation can be detected if the equilibration boundary $t_0$ is a significant fraction of the total simulation length $T$, with a good rule of thumb being that $T \gtrsim 20 t_0$.
@@
 If independent simulations are used, the alchemical state $\vec{\lambda}$ may have a significant impact on the correlation time, and these simulations should be subsampled independently using a separate estimate of the statistical inefficinecy $g$ for each alchemical state.
-If coupled simulations are used (such as a Hamiltonian replica exchange simulation), the replicas should undergo equivalent random walks in alchemical space, and the replicas can be can be subsampled with the same $g$ to generate an equal number of uncorrelated samples at each alchemical state.
+If coupled simulations are used (such as a Hamiltonian replica exchange simulation), the replicas should undergo equivalent random walks in alchemical space, and the replicas can be subsampled with the same $g$ to generate an equal number of uncorrelated samples at each alchemical state.
 Conveniently, the approach described above for automated equilibration detection produces an appropriate estimate of $g$ over the production region for automating this process.
@@
 Described in words, we take the samples generated during our run with the potential energy function $u_0(\vec{q})$ and calculate what the difference in energy would be if we switched to potential energy function $u_1(\vec{q})$, and average the exponential of the negative energy difference to get the negative of the exponential of the free energy difference.
-The original distributions, P($u_0$) as generated at $\vec{\lambda}=0$ and P($u_1$) would look like those seen in Fig.~\ref{fig:fig_sampling_scheme} \textbf{A}-\textbf{C} on the left hand side.
+The original distributions, P($u_0$) as generated at $\vec{\lambda}=0$ and P($u_1$) would look like those seen in Fig.~\ref{fig:fig_sampling_scheme} \textbf{A}-\textbf{C} on the right hand side.
 Reevaluating requires almost no extra code functionality to perform; one need only to save a full precision trajectory, and run an unmodified molecular simulation code using the $u_1$ in order to calculate the new energies of stored snapshots.
 The analysis can be written in a line of code.
-We note that this method is even more general, in that the instantaneous work to change the potential energy function from $u_0$ to $u_1$ can be replaced by the non-reversible work $W$ to make the same change under the same equilibrium conditions at either end state~\cite{jarzynski1997nonequilibrium,jarzynski1998equilibrium,crooks2000pathensemble}, though we do not go into all of the details of non-equilibrium transformations here, and refer the reader to more advanced treatments~\cite{maragakis2008bayesian,oberhofer2005biased,procacci2015unbiased,shirts2003equilibriuma,ytreberg2004singleensemble}.
+We note that this method is even more general, in that the instantaneous work to change the potential energy function from $u_0$ to $u_1$ can be replaced by the non-reversible work $W$ to make the same change beginning from the same equilibrium conditions at either end state~\cite{jarzynski1997nonequilibrium,jarzynski1998equilibrium,crooks2000pathensemble}, allowing for non-equilibrium free energy calculations, an alternate approach.
+We do not detail non-equilibrium transformations here, and refer the reader to more advanced treatments~\cite{maragakis2008bayesian,oberhofer2005biased,procacci2015unbiased,shirts2003equilibriuma,ytreberg2004singleensemble, gapsys2020large}, as our focus here is on equilibrium free energy techniques.
 
@@
  ]} \nonumber \\
- =\sum_{i=1}^{n_j} \frac{1}{1 + \exp[\ln(\frac{n_i}{n_j}) - u_{ij}(\vec{q}) + f_ij)]},
+ =\sum_{i=1}^{n_j} \frac{1}{1 + \exp[\ln(\frac{n_i}{n_j}) - u_{ij}(\vec{q}) + f_{ij})]},
 \end{eqnarray}
@@
 There is one important thing to note about the uncertainty estimates when summing multiple free energies together to calculate an overall free energy estimate.
-Although BAR itself gives a free energy estimate that is asymptotically correct in and is much less biased than the uncertainty estimate for EXP, the uncertainties in $\Delta {f_{i-1\rightarrow i}}$ and $\Delta {f_{i\rightarrow i+1}}$ are not uncorrelated, because they both involve the energies $u_i(\vec{q})$.
+Although BAR itself gives a free energy estimate that is asymptotically correct and is much less biased than the uncertainty estimate for EXP, the uncertainties in $\Delta {f_{i-1\rightarrow i}}$ and $\Delta {f_{i\rightarrow i+1}}$ are not uncorrelated, because they both involve the energies $u_i(\vec{q})$.
 The variances of each of the free energies will \textit{not} propagate as variances usually do (in quadrature) into the variance of the overall free energy.
@@
 \end{equation}
-And than we can then numerically integrate $df/d\vec{\lambda}$ over an alchemical transformation, using a range of different well-established techniques, to obtain:
+And then we can numerically integrate $df/d\vec{\lambda}$ over an alchemical transformation, using a range of different well-established techniques, to obtain:
 \begin{equation}
@@
 In practice, adding 2-3 more intermediate states is typically sufficient to match the performance of these more complicated numerical quadrature schemes.
+It is also possible to calculate the $\lambda$-derivatives at non-simulated states from simulated states in a scheme named extended TI~\cite{ruiter2016extended} which reduces the integration error.
 
@@
 Thus MBAR does not actually get significant information from these energy differences, so one might as well not even calculate them, and just perform BAR between nearest neighbors.~\cite{paliwal2011benchmark}
-\item TI usually gives similar values as MBAR implemented with sufficient numbers of intermediates, but quadrature errors are hard to estimate beforehand  can occur if one is not careful.~\cite{paliwal2011benchmark}
+\item TI usually gives similar values as MBAR implemented with sufficient numbers of intermediates, but quadrature errors that are hard to estimate beforehand  can occur if one is not careful.~\cite{paliwal2011benchmark}
 \item WHAM is an approximation to MBAR, and there are no compelling reasons it should be used.
@@
 A recent best practices paper by Grossfield et al.,~\cite{grossfield2018best} provides substantial detail on how to estimate uncertainties from molecular simulations and is a good starting point for this topic.
+The uncertainty of a free energy estimate is limited however, as it is only an estimate of the configurations sampled and cannot contain any information from the phase space not sampled during a simulation.
+As a consequence, the variance in free energy afforded by any estimator will always be an underestimate of the true uncertainty.
 In general, the quantification of different error metrics depends on both data generation and analysis methods used from the ones discussed above.
@@
 In some cases, slow correlation time may not be expected and therefore not a feature known in advance.
-To this end, you should always look at all simulation data available check convergence behaviour for each free energy estimate and if need be extend existing simulations or try an approach that requires simulations in two separate binding modes where they interconvert at very slow timescales.
+To this end, you should always look at all simulation data available and check convergence behaviour for each free energy estimate.
+Comparing the free energy trajectories as a function of the simulation time in the forward and reverse time direction is a useful convergence test~\cite{klimovich2015guidelines}.
+As shown in Fig.~\ref{fig:freeenergytrajectories}, disagreements larger than 1~$k_{B}T$ in the final part of the forward and reverse trajectories can be useful to detect unconverged results (see also Fig.~\ref{fig:convergence_forward_reverse}).
+In this case, one can extend the simulations or try an approach that requires simulations in two separate binding modes where they interconvert at very slow timescales.  
 \begin{figure}
@@
 Shaded areas represent 95\% confidence intervals around the mean computed from the 5 replicates data.
-The horizontal dashed lines show the final binding free energy prediction of the two calculations after a total of 5230 ns for OA-G3 and 6650 ns for CB8-G3.
+The horizontal dash-dot lines show the final binding free energy prediction of the two calculations after a total of 5230 ns for OA-G3 and 6650 ns for CB8-G3.
+Dashed lines are the free energy trajectories computed in the forward (blue) and reverse (red) time direction for a single replicate calculation.
 Longer correlation times in CB8-G3 cause the calculation to converge more slowly.
@@
 This usually indicates sampling problems.
-\item \textbf{Dependence of the free energy on initial conformation of the system.} Ensemble average properties should not depend on the starting point.
-\item \textbf{Torsional sampling} Torsions with multiple low-energy minima where some of these minima are that are visited rarely or not at all.
+\item \textbf{Dependence of the free energy on initial configuration of the system.} Ensemble average properties should not depend on the starting point.
+\item \textbf{Torsional sampling} Torsions with multiple low-energy minima where some of these minima are visited rarely or not at all.
 Which torsions have low energy minima can best be found by comparing to the simulation in the solvent.
@@
 If it varies drastically, then either there need to be finer sampling in $\vec{\lambda}$ in this region, or there are sampling problems there.
-\item \textbf{Convergence of free energy} The free energy should clearly converge as a function or of simulation time (Fig. ~\ref{fig:convergence_forward_reverse}).
+\item \textbf{Convergence of free energy} The free energy should clearly converge as a function of simulation time (Fig. ~\ref{fig:convergence_forward_reverse}).
 \item If using nonequilbirum methods, is the result independent of the speed at which the nonequilibrium change is performed?
@@
 Mean unsigned error (MUE, also called mean absolute error/MAE) is another key statistic to include in your results.
-Even though some models' near-perfect correlation and ranking statistics might suggest excellent accuracy, MUE values can still have errors of show multiple kcal/mol of error, providing important additional insight into performance.
+Even though some models' near-perfect correlation and ranking statistics might suggest excellent accuracy, MUE values can still have errors of multiple kcal/mol, providing important additional insight into performance.
 Furthermore, MUE allows for unbiased comparisons between predictive models as it is less sensitive to dataset size.
@@
 \begin{figure}
-  \includegraphics[width=0.95\linewidth]{paper/figures/fig13_analysis_practices/Figure.png}
+  \includegraphics[width=0.95\linewidth]{figures/fig13_analysis_practices/Figure.png}
   \caption{\textbf{An example of recommended practices for graphing alchemical free energy predictions.} This figure shows the relation between predicted and experimentally-determined Gibbs free energy in kcal/mol with standard errors as error bars.
@@
 \label{sec:software}
-There are many different software solutions available for the setup, running, and analysis of alchemcial free energy calculations.
+There are many different software solutions available for the setup, running, and analysis of alchemical free energy calculations.
 These will vary in customizability and ways in which they are ran, e.g.
@@
 	\item \href{https://ambermd.org/}{AMBER}, including its new pmemd.cuda version supports free energy calculations~\cite{salomon-ferrer2013overview}. 
+	\item \href{http://www.gromos.net/}{GROMOS} offers an extensive and flexible molecular dynamics and simulations analysis suites with free energy calculation functionalities including customizable alchemical paths and various sampling protocols~\cite{schmid2012architecture, kunz2012new, eichenberger2011gromos}.
 	\end{itemize}
@@
 The LiveCOMS GROMACS tutorial includes an example free energy calculation~\cite{lemkul2018From}.
-	\item \href{http://pmx.mpibpc.mpg.de/instructions.html}{PMX}, an add-on to GROMACS, offers a mutation free energy calculation module\cite{abraham2015gromacs}.
-	\item \href{https://github.com/qusers/Q6}{Q} is MD code for performing FEP calculations using a variety of force fields~\cite{aqvistjohan2017q6}. 
+	\item \href{http://pmx.mpibpc.mpg.de/instructions.html}{PMX}, an add-on to GROMACS, offers a mutation free energy calculation module~\cite{abraham2015gromacs}.
+	\item \href{https://github.com/qusers/Q6}{Q} is MD code for performing FEP calculations using a variety of force fields~\cite{aqvistj.2017q6}. 
 	\end{itemize}
@@
 	\begin{itemize}
-	\item \href{http://pmx.mpibpc.mpg.de/instructions.html}{PMX}: GROMACS \url{https://github.com/deGrootLab/pmx}.
+	\item \href{http://pmx.mpibpc.mpg.de/instructions.html}{PMX}: Setup of perturbation maps, perturbed topologies and input coordinates for GROMACS simulations\url{https://github.com/deGrootLab/pmx}.
 	\item \href{https://github.com/MobleyLab/Lomap}{Lomap/Lomap2} : Relative alchemical transformation graph planning for setting up perturbation networks~\cite{liu2013lead}.
@@
    \item Check your ligands  have the same, biologically correct binding pose
-        \item Make sure your λ-scheduling is set correctly
+        \item Make sure your $\lambda$-scheduling is set correctly
         \item Make sure your molecular transformations are realistic (1-5 heavy atoms for reliable computations)
@@
     \item Check again whether dummy atoms were assigned correctly
-    \item Inspect the trajectories across the λ-schedule (particularly the endpoints) for problems described in the text
+    \item Inspect the trajectories across the $\lambda$-schedule (particularly the endpoints) for problems described in the text
     \item Inspect the overlap matrices for lack of overlap
@@
 %%%%%%%%%%%%%%%%
-\textbf{ASJSM}: Coordinated the document, contributed to most sections, and co-designed Figs.~\ref{fig:fig_binding_thermodynamic_cycle},~\ref{fig:fig_topology},~\ref{fig:fig_mcss},~\ref{fig:fig_types_of_networks},~\ref{fig:fig_absolute_thermodynamic_cycle},~\ref{fig:scatterplot_analysis},and created Figs.~\ref{fig:fig_what_is_lambda},~\ref{fig:fig_what_is_alchemy},~\ref{fig:overlap},~\ref{fig:pmf} and replotted~\ref{fig:automatic-equilibration-detection} and~\ref{fig:fig_types_of_networks}.\\
+\textbf{ASJSM}: Coordinated the document, contributed to most sections, and co-designed Figs.~\ref{fig:fig_binding_thermodynamic_cycle},~\ref{fig:fig_topology},~\ref{fig:fig_mcss},~\ref{fig:fig_types_of_networks},~\ref{fig:fig_absolute_thermodynamic_cycle},~\ref{fig:scatterplot_analysis}, and created Figs.~\ref{fig:fig_what_is_lambda},~\ref{fig:fig_what_is_alchemy},~\ref{fig:overlap},~\ref{fig:pmf} and replotted~\ref{fig:automatic-equilibration-detection} and~\ref{fig:fig_types_of_networks}.\\
 \textbf{JS}: Created Figs.~ \ref{fig:fig_what_is_alchemy},~\ref{fig:fig_binding_thermodynamic_cycle},~\ref{fig:fig_topology},~\ref{fig:fig_mcss},~\ref{fig:fig_absolute_thermodynamic_cycle},~\ref{fig:scatterplot_analysis}, and an initial draft of~\ref{fig:fig_types_of_networks}.
@@
 \textbf{HBM} Contributed to Sec.~\ref{sec:plot_data} and Fig.~ \ref{fig:scatterplot_analysis} and helped edit the paper.\\
+\textbf{SP} Wrote Sec.~\ref{sec:software}.
+\textbf{DFH} Contributed to Sec.~\ref{sec:software} and helped edit the paper.
 % We suggest you preserve this comment:
@@
 JDC is a current member of the Scientific Advisory Board of OpenEye Scientific Software and a consultant to Foresite Laboratories.
+A complete funding history for the Chodera lab can be found at \url{http://choderalab.org/funding}..
 A complete funding history for the Chodera lab can be found at \url{http://choderalab.org/funding}.

### 2020-11-03 paper/manuscript.tex
fixing things that broke on merge

--- 2020-08-30 paper/manuscript.tex
@@
 JM is a current member of the Scientific Advisory Board of Cresset. 
-MK is employed by Cresset who commercially distribute a software for performing alchemical free energy calculations. MRS is a Open Science Fellow and consultant for Silicon Therapeutics.
+MK is employed by Cresset who commercially distribute a software for performing alchemical free energy calculations.
+MRS is a Open Science Fellow and consultant for Silicon Therapeutics.
+JDC is a current member of the Scientific Advisory Board of OpenEye Scientific Software and a consultant to Foresite Laboratories.
 
@@
 HEBM acknowledges support from a Molecular Sciences Software Institute Investment Fellowship and Relay Therapeutics.
-JDC is a current member of the Scientific Advisory Board of OpenEye Scientific Software and a consultant to Foresite Laboratories.
 A complete funding history for the Chodera lab can be found at \url{http://choderalab.org/funding}. 

### 2020-11-28 paper/manuscript.tex
version 1.0

--- 2020-11-03 paper/manuscript.tex
@@
 \textbf{ASJSM}: Coordinated the document, contributed to most sections, and co-designed Figs.~\ref{fig:fig_binding_thermodynamic_cycle},~\ref{fig:fig_topology},~\ref{fig:fig_mcss},~\ref{fig:fig_types_of_networks},~\ref{fig:fig_absolute_thermodynamic_cycle},~\ref{fig:scatterplot_analysis}, and created Figs.~\ref{fig:fig_what_is_lambda},~\ref{fig:fig_what_is_alchemy},~\ref{fig:overlap},~\ref{fig:pmf} and replotted~\ref{fig:automatic-equilibration-detection} and~\ref{fig:fig_types_of_networks}.\\
+\textbf{BA}: Helped write the uncertainty estimation, stopping conditions, and output analysis sections and created figure ~\ref{fig:convergence_forward_reverse}.\\
+\textbf{HBM} Contributed to Sec.~\ref{sec:plot_data} and Fig.~ \ref{fig:scatterplot_analysis} and helped edit the paper.\\
+\textbf{JDC}: Wrote Sec.~\ref{sec:decorrelating-samples} and~\ref{sec:automatic-equilibration-detection} discussed structure and design of the whole document, suggested Figs.~\ref{fig:fig_what_is_alchemy} and~\ref{fig:fig_sampling_scheme} \\
+\textbf{DFH} Contributed to Sec.~\ref{sec:software} and helped edit the paper. \\
+\textbf{MK}: Contributed to Sec.~\ref{sec:data_analysis}, provided the data for figure~\ref{fig:overlap}, compiled the dataset for Sec.~\ref{sec:benchmark} and helped edit the paper.\\
+\textbf{JM}: Contributed to Sec.~\ref{subsec:reproducible},~\ref{sec:prerequisites},~\ref{sec:important_path},~\ref{subsec:estimators},~\ref{subsec:uncertainty}, and~\ref{sec:conclusion}\\
+\textbf{DLM}: Contributed to the outline, drafted some of the sections, gave ideas on figures, and helped edit the paper.\\
+\textbf{LNN}: Helped write the simulation length, stopping conditions, and information saving section. Edited and reviewed alchemical path section.\\
+\textbf{SP} Wrote Sec.~\ref{sec:software}. \\
+\textbf{AR}: Created figure~\ref{fig:freeenergytrajectories}, contributed to sections~\ref{sec:theory} and~\ref{sec:simulation_protocol_choice}, and helped edit the paper.\\
 \textbf{JS}: Created Figs.~ \ref{fig:fig_what_is_alchemy},~\ref{fig:fig_binding_thermodynamic_cycle},~\ref{fig:fig_topology},~\ref{fig:fig_mcss},~\ref{fig:fig_absolute_thermodynamic_cycle},~\ref{fig:scatterplot_analysis}, and an initial draft of~\ref{fig:fig_types_of_networks}.
 Wrote Sec.~\ref{sec:plot_data}, the checklist Sec.~\ref{sec:checklist}, and contributed to general formatting discussions and editing.\\
-\textbf{MK}: Contributed to Sec.~\ref{sec:data_analysis}, provided the data for figure~\ref{fig:overlap}, compiled the dataset for Sec.~\ref{sec:benchmark} and helped edit the paper.\\
-\textbf{DLM}: Contributed to the outline, drafted some of the sections, gave ideas on figures, and helped edit the paper.\\
-\textbf{GT}: Contributed to Sec.~\ref{sec:intro} and~\ref{sec:drugdiscovery}, and helped edit the paper.\\
-\textbf{AR}: Created figure~\ref{fig:freeenergytrajectories}, contributed to sections~\ref{sec:theory} and~\ref{sec:simulation_protocol_choice}, and helped edit the paper.\\
 \textbf{MRS}: Helped create figure~\ref{fig:fig_what_is_lambda}, wrote Sec.~\ref{sec:important_path} describing choices for alchemical pathways and parts of~\ref{sec:data_analysis} on the analysis for free energy calculations.
 Reviewed and edited text throughout.\\
-\textbf{LNN}: Helped write the simulation length, stopping conditions, and information saving section. Edited and reviewed alchemical path section.\\
-\textbf{BA}: Helped write the uncertainty estimation, stopping conditions, and output analysis sections and created figure ~\ref{fig:convergence_forward_reverse}.\\
-\textbf{JM}: Contributed to Sec.~\ref{subsec:reproducible},~\ref{sec:prerequisites},~\ref{sec:important_path},~\ref{subsec:estimators},~\ref{subsec:uncertainty}, and~\ref{sec:conclusion}\\
-\textbf{JDC}: Wrote Sec.~\ref{sec:decorrelating-samples} and~\ref{sec:automatic-equilibration-detection} discussed structure and design of the whole document, suggested Figs.~\ref{fig:fig_what_is_alchemy} and~\ref{fig:fig_sampling_scheme} \\
-\textbf{HX}: Contributed Sec.~\ref{subsec:accuracy}, to Sec.~\ref{sec:relative-fe-protocol}, and to Sec.~\ref{sec:are-they-good}.\\
-\textbf{HBM} Contributed to Sec.~\ref{sec:plot_data} and Fig.~ \ref{fig:scatterplot_analysis} and helped edit the paper.\\
-\textbf{SP} Wrote Sec.~\ref{sec:software}.
-\textbf{DFH} Contributed to Sec.~\ref{sec:software} and helped edit the paper.
+\textbf{GT}: Contributed to Sec.~\ref{sec:intro} and~\ref{sec:drugdiscovery}, and helped edit the paper.\\
+\textbf{HX}: Contributed Sec.~\ref{subsec:accuracy}, to Sec.~\ref{sec:relative-fe-protocol}, and to Sec.~\ref{sec:are-they-good}.
 % We suggest you preserve this comment:
@@
 MK is employed by Cresset who commercially distribute a software for performing alchemical free energy calculations. MRS is a Open Science Fellow and consultant for Silicon Therapeutics.
-
+JDC is a current member of the Scientific Advisory Board of OpenEye Scientific Software and a consultant to Foresite Laboratories.
 \section*{Funding Information}
@@
 HEBM acknowledges support from a Molecular Sciences Software Institute Investment Fellowship and Relay Therapeutics.
-JDC is a current member of the Scientific Advisory Board of OpenEye Scientific Software and a consultant to Foresite Laboratories.
-A complete funding history for the Chodera lab can be found at \url{http://choderalab.org/funding}..
-A complete funding history for the Chodera lab can be found at \url{http://choderalab.org/funding}.
 DLM acknowledges support from the National Institutes of Health (R01GM108889, R01GM124270, and R01GM132386), and the National Science Foundation (CHE 1352608).
 JDC acknowledges support from the National Institutes of Health (NIH P30 CA008748, NIH R01 GM121505, R01 GM132386).
+A complete funding history for the Chodera lab can be found at \url{http://choderalab.org/funding}.
 The content is solely the responsibility of the authors and does not necessarily represent the official views of the National Institutes of Health.

### 2020-11-28 paper/manuscript.tex
Updates from Overleaf

@@
 \textbf{D}: Effect of mutations of protein or host residues on free energies of binding.
-\textbf{E}: Absolute free energies of binding of a small molecule to a host (e.g.
-protein), \textbf{F}: Relative free energy of binding of one molecule with respect to another, here toluene and benzyl alcohol.
+\textbf{E}: Relative free energy of binding of one molecule with respect to another, here toluene and benzyl alcohol, \textbf{F}: absolute free energies of binding of a small molecule to a host (e.g.
+protein).
     \label{fig:fig_what_is_alchemy}
@@
 
-In contrast, in a dual topology alchemical free energy calculation, no atoms are allowed to change type~\cite{Boresch:1999:J.Phys.Chem.A, shirts2012best}.
+In contrast, in a dual topology alchemical free energy calculation, no atoms are allowed to change type~\cite{boresch1999rolea, shirts2012best}.
 This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxy methyl group, then passing through an intermediate state where some atoms are partially interacting--- particularly, those atoms which are becoming dummy atoms or ceasing to be dummy atoms~\cite{mobley2014blind}.
@@
 It is however important to note that different packages also differ in how they handle implementation of alchemical transformations, making it difficult to give rules of thumb concerning specific efficient transformations which work equally well across simulation packages.
-Thus we are hesitant to recommend best practices within specific software packages at point in time, although any good transformation pathway will conform to the guidelines we have outlined above.
+Thus we are hesitant to recommend best practices within specific software packages at this point in time, although any good transformation pathway will conform to the guidelines we have outlined above.
 

### 2020-11-28 paper/manuscript.tex

@@
 Individual software programs and settings can handle such issues.
-For example, bonds can be transformed in both GROMACS and GROMOS, because contributions of LINCS (GROMACS) and SHAKE (GROMOS) contrained bonds are added to the $\frac{dH}{d\lambda}$ term~\cite{pearlman1993determining, straatsma1992holonomica, pearlman1991overlooked, gunsteren1997computer}.
+For example, bonds can be transformed in both GROMACS and GROMOS, because contributions of LINCS (GROMACS) and SHAKE (GROMOS) contrained bonds are added to the $\frac{dH}{d\lambda}$ term~\cite{pearlman1993determining, straatsma1992holonomica, pearlman1991overlooked, gunsteren1998computer}.
 However, the constraints are not taken into account when calculating energy differences to other intermediate states (as the effect is entropic, not energetic).

### 2020-11-28 paper/manuscript.tex
Updates from Overleaf

@@
 Individual software programs and settings can handle such issues.
-For example, bonds can be transformed in both GROMACS and GROMOS, because contributions of LINCS (GROMACS) and SHAKE (GROMOS) contrained bonds are added to the $\frac{dH}{d\lambda}$ term~\cite{pearlman1993determining, straatsma1992holonomica, pearlman1991overlooked, gunsteren1998computer}.
+For example, bonds can be transformed in both GROMACS and GROMOS, because contributions of LINCS (GROMACS) and SHAKE (GROMOS) contrained bonds are added to the $\frac{dH}{d\lambda}$ term~\cite{pearlman1993determining, straatsma1992holonomica, pearlman1991overlooked, gunsteren1989computer}.
 However, the constraints are not taken into account when calculating energy differences to other intermediate states (as the effect is entropic, not energetic).

### 2020-12-23 paper/manuscript.tex
Updated for publication.
Updates from Overleaf

@@
 %%% PREAMBLE
-\documentclass[9pt,bestpractices]{livecoms}
+\documentclass[9pt,bestpractices,ASAPversion]{livecoms}
 % Use the 'onehalfspacing' option for 1.5 line spacing
@@
 
-\newcommand{\versionnumber}{1.0} % you should update the minor version number in preprints and major version number of submissions.
+\newcommand{\versionnumber}{0.3} % you should update the minor version number in preprints and major version number of submissions.
 \newcommand{\githubrepository}{\url{https://github.com/alchemistry/alchemical-best-practices}} %this should be the main github repository for this article
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{Best Practices for Alchemical Free Energy Calculations [Article v \versionnumber]}
+\title{Best Practices for Alchemical Free Energy Calculations [Article v\versionnumber]}
 \author[1*]{Antonia S. J. S. Mey}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\pubDOI{10.XXXX/YYYYYYY}
-\pubvolume{<volume>}
-\pubyear{<year>}
-\articlenum{<number>}
-\datereceived{Month, Day, Year}
-\dateaccepted{Month, Day, Year}
+\pubDOI{10.33011/livecoms.2.1.18378}
+\pubvolume{2}
+\pubyear{2020}
+\articlenum{18378}
+\datereceived{5 August 2020}
+\dateaccepted{25 November 2020}
 
@@
 \begin{equation}
-  K_b = K_D^\ast \frac{1 + K_C}{K_C}
+  K_D = K_D^\ast \frac{1 + K_C}{K_C}
   \label{eqn:conformational-binding}

### 2020-12-30 paper/manuscript.tex
michaels comments

@@
 \affil[3]{Computational and Systems Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York NY, USA}
-\affil[4]{Departments of Pharmaceutical Sciences and Chemistry, University of California, Irvine, USA}
+\affil[4]{Departments of Pharmaceutical Sciences and Chemistry, University of California, Irvine, Irvine, USA}
 \affil[5]{Molecular Sciences Software Institute, Blacksburg VA, USA}
@@
 \affil[8]{University of Colorado Boulder, Boulder, CO, USA}
-\affil[9]{Computational Chemistry, Janssen Research \& Development, Turnhoutseweg 30, Beerse B-2340,Belgium}
+\affil[9]{Computational Chemistry, Janssen Research \& Development, Turnhoutseweg 30, Beerse B-2340, Belgium}
 \affil[10]{Cresset, Cambridgeshire, UK}
@@
 The data collected from these bridging alchemical thermodynamic states allows the efficient computation of transfer free energies (or differences in transfer free energies) with orders of magnitude less simulation time than simulating the transfer process directly.
-
 While these methods are highly flexible, care must be taken in avoiding common pitfalls to ensure that computed free energy differences can be robust and reproducible for the chosen force field, and that appropriate corrections are included to permit direct comparison with experimental data.
 
-In this paper, we review current best practices for several popular application domains of equilibrium alchemical free energy calculations, including relative and absolute small molecule binding free energy calculations to biomolecular targets.
+In this paper, we review current best practices for several popular application domains of  alchemical free energy calculations performed with equilibrium simulations, in particular relative and absolute small molecule binding free energy calculations to biomolecular targets.
 %\todo[inline, color=green!20]{JDC: Should we migrate this document to \url{https://github.com/alchemistry}? ASJSM: Let's migrate with the submitted version, but leave it here for now?}
@@
 \begin{itemize}
-\item computing the free energy of a conformational change for a molecule with a high barrier to interconversion (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{A});
-\item computing partition ($\log P$) or distribution ($\log D$) coefficients between environments (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{B})~\cite{rustenburg2016measuring, bosisio2016blinded} 
-\item determining partitioning between compartments into membranes (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{C})~\cite{corey2019insights}. 
+\item computing the free energy of a conformational change for a molecule with a high barrier to interconversion (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{A});
+\item computing partition ($\log P$) or distribution ($\log D$) coefficients between environments (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{B})~\cite{rustenburg2016measuring, bosisio2016blinded} 
+\item determining partitioning between compartments into membranes (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{C})~\cite{corey2019insights}. 
 \end{itemize}
@@
 \begin{itemize}
-\item a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{D})\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate};
-\item the entire ligand can be alchemically transferred from protein to solvent in an absolute binding free energy calculation (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{E})~\cite{mobley2007predicting,aldeghi2015accurate,aldeghi2017predictions};
-\item small alchemical modifications can be made between chemically related ligands to estimate relative differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{F})~\cite{wang2015accurate,mey2016blinded,song2019using,gapsys2020large,kuhn2020assessment}.
+\item a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{D})\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate};
+\item the entire ligand can be alchemically transferred from protein to solvent in an absolute binding free energy calculation (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{E})~\cite{mobley2007predicting,aldeghi2015accurate,aldeghi2017predictions};
+\item small alchemical modifications can be made between chemically related ligands to estimate relative differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{F})~\cite{wang2015accurate,mey2016blinded,song2019using,gapsys2020large,kuhn2020assessment}.
 \end{itemize}
 
-After an alchemical calculation is performed (which generally involves multiple simulations at a variety of alchemical states), the data must be analyzed to compute an estimate of the free energy for the transformation of interest.
+After an alchemical calculation is performed, which generally involves multiple simulations at a variety of alchemical states, the data must be analyzed to compute an estimate of the free energy for the transformation of interest.
  Early work used simple but statistically suboptimal estimators for this: free energy perturbation (FEP) used a simple (but highly biased) estimator based on the Zwanzig relation~\cite{zwanzig1954hightemperature} or numerical quadrature via thermodynamic integration (TI), for which the theory dates back the better part of a century but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamicsa, merz1989free}.
@@
 In the meantime, today's users may find it difficult to get started with these complex calculations whilst also keeping up with the fast pace of change. 
-This Best Practices guide provides current recommendations and tips for users of all experience. Updates and suggestions are welcomed via our GitHub repository.
+This Best Practices guide provides current recommendations and tips for users of all experience.
+Updates and suggestions are welcomed via our GitHub repository at \url{https://github.com/alchemistry/alchemical-best-practices}.
 
@@
 Where the best practices are currently not certain, we highlight areas where further research is needed to identify an unambiguous recommendation.
- It can also serve as a set of best practices to ensure simulation robustness and reproducibility which reviewers may wish to consider as they evaluate papers.
+ This guide can also serve as a set of best practices to ensure simulation robustness and reproducibility which reviewers may wish to consider as they evaluate papers.
 
@@
 Particular focus will be given to aspects of the molecular simulations which are unique to alchemical calculations---these include the calculation of transfer free energies (hydration free energies, partition coefficients, etc.), and binding free energies (absolute and relative).
-We primarily focus on \emph{equilibrium} free energy calculations in this best practices guide, as best practices for these are more developed, and nonequilibrium techniques may warrant their own guide as such practices evolve.
+We primarily focus on free energy calculations using simulations performed at \emph{equilibrium} in this Best Practices guide, as best practices for these are more developed, and non-equilibrium techniques may warrant their own guide as such practices evolve.
 
@@
 \item covalent inhibition~\cite{lameira2019predicting}
-\item free energies of mutation of protein side chains~\cite{gapsys2016accurate, aldeghi2018accurate}
+\item free energies of mutation of protein side chains~\cite{gapsys2016accurate,aldeghi2018accurate}
 \item nonspecific binding or multiple binding sites~\cite{gill2018binding}
@@
 A number of different studies have looked at the influence of force fields and it is assumed the user has made an adequate choice for the system under study~\cite{loeffler2018reproducibility, vassetti2019assessment, lopes2015current}.
-\item nonequilibrium free energy calculations~\cite{gapsys2020large}
+\item non-equilibrium free energy calculations~\cite{gapsys2020large}
 \item Free energy calculations using QM/MM methods~\cite{beierlein2011simple,dybeck2016comparison,cave-ayland2015direct}.
-\item Free energy calculations using Machine learning methods~\cite{rufa2020chemical, scheen2020hybrid, cole2020machine}
+\item Free energy calculations using machine learning methods~\cite{rufa2020chemical, scheen2020hybrid, cole2020machine}
 \end{itemize}
@@
 $U$ --- potential energy\\
-$u$ --- reduced potential describing a thermodynamic state \\
-$\Delta G$ --- Gibbs free energy (isothermal isobaric ensemble), Gibbs function, or free enthalpy, though the most common term Gibbs free energy is used in the text\\
-$\Delta A$ --- Helmholtz free energy (canonical ensemble) or Helmholtz function, with Helmholtz free energy used in the text.\\
-$\Delta f$ --- reduced (dimensionless) free energy \\
-$\Delta \hat{f}$ --- estimate from an estimator for the reduced free energy\\
+$u$ --- reduced (dimensionless) potential describing a thermodynamic state \\
+$G$ --- Gibbs free energy (free energy in the isothermal isobaric ensemble), Gibbs function, or free enthalpy, though the most common term Gibbs free energy is used in the text\\
+$A$ --- Helmholtz free energy (free energy in the canonical ensemble) or Helmholtz function, with Helmholtz free energy used in the text.\\
+$f$ --- reduced (dimensionless) free energy \\
+$\Delta \hat{f}$ --- estimate from an estimator for the reduced free energy difference between two states\\
 $\Gamma$ --- configurational space accessible by simulations \\
@@
 \end{equation}
-The standard state concentration $c^\circ$ depends on the reference state, but it is usually set to $1\,\mathrm{mol}/\mathrm{L}$ assuming a constant pressure of $1\,\mathrm{atm}$ (see also Sec.~\ref{sec:standardstate-restraints}), but also holds for dilute solutions if thermodynamic activities can be approximated by concentrations~\cite{gilson1997statisticalthermodynamic}.
+The standard state concentration $c^\circ$ depends on the reference state, but it is usually set to $1\,\mathrm{mol}/\mathrm{L}$ assuming a constant pressure of $1\,\mathrm{atm}$ (see also Sec.~\ref{sec:standardstate-restraints}).
+Eq.~\ref{eq:law-mass-action} also holds for dilute solutions if thermodynamic activities can be approximated by concentrations~\cite{gilson1997statisticalthermodynamic}.
 Thus, the standard Gibbs free energy of binding $\Delta G_{\mathrm{bind}}$ is given by:
@@
 
-In many cases, the quantity of interest is the change in binding affinity between a compound $A$ and a related compound $B$ (e.g., by modifying one of the drug scaffold's substituents, see (Fig.~\ref{fig:fig_what_is_alchemy} \textbf{F}), which, by using Eq.~\ref{eq:binding-free-energy-from-bound-unbound-probability-ratio} and \ref{eq:bound-unbound-probability-ratio} is given by
+In many cases, the quantity of interest is the change in binding affinity between a compound $A$ and a related compound $B$ (e.g., by modifying one of the drug scaffold's substituents, see (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{F}), which, by using Eq.~\ref{eq:binding-free-energy-from-bound-unbound-probability-ratio} and \ref{eq:bound-unbound-probability-ratio} is given by
 \begin{equation}\label{eq:delta-delta-G-physical}
@@
 
-Equilibrium free energy calculations, our focus here, solve the problem of having poor overlap between the states of interest, multiple intermediate alchemical states are introduced at values $\vec{\lambda}_A = \vec{\lambda}_0, \vec{\lambda}_1, \cdots, \vec{\lambda}_K = \vec{\lambda}_B$ so that each pair of consecutive states $\vec{\lambda}_k, \vec{\lambda}_{k+1}$ share good overlap.
-Each intermediate state models a ligand that is neither $A$ nor $B$ but a mix of the two.
+Equilibrium free energy calculations, our focus here, solve the problem of having poor overlap between the states of interest, by introducing multiple intermediate alchemical states at values $\vec{\lambda}_A = \vec{\lambda}_0, \vec{\lambda}_1, \cdots, \vec{\lambda}_K = \vec{\lambda}_B$ so that each pair of consecutive states $\vec{\lambda}_k, \vec{\lambda}_{k+1}$ share good overlap.
+Each intermediate state models a ligand that is neither $A$ nor $B$ but a interpolation of the two.
 Many estimators (e.g., exponential reweighting (EXP)~\cite{zwanzig1954hightemperature} and Bennett's acceptance ratio (BAR)~\cite{bennett1976efficient,shirts2003equilibriuma}) can then be used to compute the free energy as
@@
 
-Nonequilibrium free energy techniques provide an alternate approach to this problem, driving $\lambda$ between states, but these are not our focus here.~\cite{jarzynski1997nonequilibrium,jarzynski1998equilibrium,crooks2000pathensemble, gapsys2020large}
+Non-equilibrium free energy techniques provide an alternate approach to this problem, driving $\lambda$ between states, but these are not our focus here.~\cite{jarzynski1997nonequilibrium,jarzynski1998equilibrium,crooks2000pathensemble, gapsys2020large}
 
@@
 (See recommendations for reporting best practices, Sec.~\ref{sec:plot_data}.) Additionally, as it is possible to perform calculations on a set of ligands using different pairwise comparisons of molecules, the performance of the method may be biased based on which pairs of comparisons are performed.
-Additionally, it is possible that the error associated with the relative free energy between a two ligands that was not directly computed (however can be deduced using thermodynamic paths involving other ligands) will likely be more uncertain.
+Additionally, it is possible that the error associated with the relative free energy between a two ligands that was not directly computed,  but can be deduced using one or more thermodynamic paths involving other ligands will likely be more uncertain.
 Given the need to understand the performance of the system with alchemical free energy calculations, we recommend that retrospective studies for a particular target and a particular chemical series be performed for each application case.
@@
 \label{subsec:reproducible}
-We restrict our analysis here to repetitions of the same calculation performed with precisely the same force field - different force fields used to describe the same molecule can lead in some cases to wide difference in free energies.
+We restrict our analysis here to repetitions of the same calculation performed with precisely the same force field, as different force fields used to describe the same molecule can lead in some cases to wide difference in free energies.
 Even within this restriction,
@@
 An important consideration is how reproducible alchemical free energy calculations are in practice.
-In simple cases such as absolute hydration free energies of small organic molecules, or relative hydration free energy calculations between structurally similar small organic molecules, it should be possible to obtain highly precise estimates with a given software package (with a sample standard deviation under 0.01 kcal/mol)~\cite{rizzi2019sampl6}.
+In simple cases such as absolute hydration free energies of small organic molecules, or relative hydration free energy calculations between structurally similar small organic molecules, it should be possible to obtain highly precise estimates with a given software package, i.e.with a sample standard deviation under 0.01 kcal/mol~\cite{rizzi2019sampl6}.
 For more complex use cases such as protein-ligand binding free energies the repeatability is often substantially worse~\cite{rizzi2019sampl6}.
-A good practice is to perform two or three runs of the same perturbation to assess precision with a given protocol, using different initial conditions (i.e.
-velocities).
+A good practice is to perform two or three runs of the same perturbation to assess precision with a given protocol, using different initial velocities.
 The sample standard deviation will give a crude estimate of the reliability of the estimates, and whether the precision is sufficient for the problem at hand.
-When practical, a more stringent test is to use different input coordinates for each repeat run.
-
-Note that these issues concern calculations carried out with a single software package, but simulation package variations can introduce additional issues.
-Such issues of reproducibility of free energy calculations across different simulation packages have attracted attention recently.
-Greater variability is expected due to methodological differences such as integrators, thermostats, barostats, treatment of long-range electrostatics, and potentially other factors.
+When practical, a more stringent test is to use different input coordinates for each repeat run as well as different velocities.
+
+Note that these types of statistical differences concern calculations carried out with a single software package, but simulation package variations can introduce additional discrepancies.
+Such issues of reproducibility of free energy calculations across different simulation packages have attracted attention recently~\cite{loeffler2018reproducibility,rizzi2019sampl6}.
+Greater variability is expected between packages due to methodological differences such as integrators, thermostats, barostats, treatment of long-range electrostatics, and potentially other factors.
 For absolute and relative hydration free energies of small organic molecules a variability of ca.
@@
 
-Sometimes free energy calculations can provide answers that are not
+Sometimes, however, free energy calculations can provide answers that are not
 readily available from experiments. For example, type II kinase
@@
 calculations, i.e., on average, the free energy change computed by
-free energy calculations agrees with the actual free energy change.
+free energy calculations agrees with the actual free energy change. Additional analysis of this type is presented in Brown et al.~\cite{shirts2010}
 
@@
 
-If multiple structures are available there is some evidence the higher affinity complex can give better performance~\cite{perez-benito2019predicting}, at least in some cases.
+If multiple structures are available there is some evidence the higher affinity complex can give better match to experiment~\cite{perez-benito2019predicting}, at least in some cases.
 However, ligands and proteins can also undergo unexpected changes in binding mode for related ligands, which can make these issues more complex to deal with~\cite{mey2016blinded}.
@@
 For relative free energy calculations, various graph topologies or maps of calculations are possible, and choices may depend on the target application.
-For instance, if the goal is to accurately assess the relative binding energy of a small number of compounds, possibly with challenging synthesis, the map of perturbations should contain as many connections between compounds as affordable.
-However, when running calculations on hundreds of compounds a so called \emph{star-map} (see Fig.~\ref{fig:fig_types_of_networks} \textbf{A}) can be used that just contains one connection per compound: perturbing every compound to a central ligand, typically the crystal structure ligand~\cite{konze2019reactionbased}.
+For instance, if the goal is to accurately assess the relative binding energy of a small number of compounds, possibly with challenging syntheses, the map of perturbations should contain as many connections between compounds as affordable.
+However, when running calculations on hundreds of compounds a so called \emph{star-map} (see Fig.~\ref{fig:fig_types_of_networks}\textbf{A}) can be used that just contains one connection per compound: perturbing every compound to a central ligand, typically the crystal structure ligand~\cite{konze2019reactionbased}.
 In this way the top-ranking examples can be readily identified and submitted to additional calculations in a second round.
@@
 For example error estimates provided by free energy estimators that are too large can highlight poorly converged simulations~\cite{perez-benito2019predicting}.
-Hysteresis, within cycles in the perturbation network or between forward and backward perturbations can be checked~\cite{wang2013modeling} to indicate problematic perturbations involved in cycles connecting many compounds (See also Secs.~\ref{sec:relative-fe-protocol} and~\ref{sec:are-they-good}).
+Hysteresis, either within cycles in the perturbation network or between forward and backward perturbations can be checked~\cite{wang2013modeling} to indicate problematic perturbations involved in cycles connecting many compounds (See also Secs.~\ref{sec:relative-fe-protocol} and~\ref{sec:are-they-good}).
 Once synthesis and testing of compounds is complete a standard strategy is to look back at how the calculations performed.
@@
     \item Do I have one or multiple good receptor structures? (e.g. a good resolution X-ray crystal of the protein target)
-    \item Do I have information on one or all of the ligand binding sites (e.g. a X-ray structure)
+    \item Do I have information on one or all of the ligand binding sites? (e.g. a X-ray structure)
     \item Should I include buried waters, or other small molecules that can be found in an X-ray structure?
@@
 Overlaying multiple protein X-ray structures can identify conserved or additional water molecules that can be useful to include in calculations.
-In cases where water molecules are known to play an important role in the binding, software implementations that use water sampling facilitated by Grand Canonical Monte Carlo methods may be useful, i.e.
-consider pre-solvation methods such as GCMC steps~\cite{michel2010prediction}.
+In cases where water molecules are known to play an important role in the binding, software implementations that use water sampling facilitated by Grand Canonical Monte Carlo methods may be useful~\cite{michel2010prediction}.
 Other tools such as WaterMap or open source equivalents (SSTMap, GIST, and others) can be used to define water structure for systems with no experimental evidence of water sites~\cite{wang2011ligand}.
-Well known protein systems with water mediated ligand interactions are for example: HSP90 which formed part of the D3R grand challenge 2015~\cite{mey2016blinded}, A2A~\cite{brucemacdonald2018ligand}, MUP~\cite{ross2015water}, ~\cite{deflorian2020accurate}, and others~\cite{michel2009energetics}.
+Well-known protein systems with water mediated ligand interactions are for example: HSP90 which formed part of the D3R grand challenge 2015~\cite{mey2016blinded}, A2A~\cite{brucemacdonald2018ligand}, MUP~\cite{ross2015water}, ~\cite{deflorian2020accurate}, and others~\cite{michel2009energetics}.
 
@@
 Since the pKa of reference amino acid sidechain residues is known, but can vary in the protein environment, many different tools have emerged for predicting sidechain pKa in proteins, such as the H++ server, ProPKa, APBS, and Maestro ~\cite{anandakrishnan2012automating, sondergaard2011improved, jurrus2018improvements, 2020schrodinger}.
-Strongly acidic (Glu, Asp) or basic (Arg, Lys) sidechains can reliably predicted to be ionised, but care is still needed as the local environment can modify expected ionization states (for instance the catalytic Asp dyad in proteases).
+Strongly acidic (Glu, Asp) or basic (Arg, Lys) sidechains can reliably predicted to be ionized, but care is still needed as the local environment can modify expected ionization states, such as  the catalytic Asp dyad in proteases.
 Histidine is notoriously more difficult to predict as its pKa suggests it ionizes closer to the experimental pH range.
@@
 In this class of alchemical free energy calculations it is necessary to generate a molecular topology that may describe the initial and final states of the perturbation (see Fig.~\ref{fig:fig_topology}).
-In cases where the end-states have high topological similarity and high structural overlap this is relatively straightforward and typically handled by use of MCSS calculations.
-In situations where the end-state topologies differ significantly, or there is relatively little spatial overlap between the two end-states some user intervention may be necessary to produce a satisfactory input topology.
+In cases where the end states have high topological similarity and high structural overlap this is relatively straightforward and typically handled by use of MCSS calculations.
+In situations where the end state topologies differ significantly, or there is relatively little spatial overlap between the two end states some user intervention may be necessary to produce a satisfactory input topology.
 
@@
 In this case a docking program can generate initial structures.
-Different commercial and none commercial tools are available, such as rDock, Autodock Vina, Glide, or Flare, to name a few~\cite{ruiz-carmona2014rdock, trott2010autodock, friesner2004glide, kuhn2020assessment}.
+Different commercial and none commercial tools are available, such as rDock~\cite{ruiz-carmona2014rdock}, Autodock Vina~\cite{trott2010autodock}, Glide~\cite{friesner2004glide}, or Flare, to name a few~\cite{kuhn2020assessment}.
 
@@
 To use binding as a concrete example, in absolute binding free energy calculations, we compute the binding free energy of a ligand to an individual receptor relative to a standard reference concentration.
-
 In contrast, in relative binding free energy calculations, we compare the binding free energy of two related inhibitors to determine the potency difference.
 
-\subsection{Absolute and relative free energy calculations have some differences}
+\subsection{Absolute and relative free energy calculations have important differences}
 Many of the issues around simulation setup and protocol choice for alchemical calculations are common, but there are some differences between absolute and relative calculations.
@@
 \paragraph{Topologies} 
-A critical first step in relative calculations is to select the approach to be used for these calculations and determine what this means for atom mapping (though this may sometimes be dictated by the requirements of the chosen simulation package.
-Sometimes this approach is described in terms of a \emph{dual topology} versus \emph{single topology} approach to relative calculations.
-
-The distinction between single and dual topology can be illustrated by considering a hypothetical transformation from molecule A to molecule B, where both atoms share a common substructure but differ in their substituents; e.g.
-consider a transformation of benzene to benzyl alcohol Fig.~\ref{fig:fig_topology}.
-In this case the common substructure is the benzene ring, though the substructure may be larger depending on how it is defined, as we discuss below.
+A critical first step is to determine which approach to use for atom mapping between end state molecules.
+Often this is predetermined by the choice of simulation software.
+Typically it is possible to chose between a \emph{dual topology} versus a \emph{single topology}
+
+The distinction between single and dual topology can be illustrated by considering a hypothetical transformation from molecule A to molecule B, where both atoms share a common substructure but differ in their substituents; in particular, consider a transformation of benzene to benzyl alcohol shown in Fig.~\ref{fig:fig_topology}.
+In this case the common substructure between the two molecules is the benzene ring, though in practice substructure may be selected to be larger depending the mapping chosen, as we discuss below.
 
 In single topology calculations, the overall transformation is set up to involve as few additional atoms as possible, so benzene would be typically changed into benzyl alcohol by changing one of the hydrogens into a carbon.
-This site will also be the future home of two additional hydrogen atoms bound to the new carbon, so these must initially be present as non-interacting atoms called ``dummy atoms'', which retain their bonded interactions but do not interact with the rest of the system.
+The site of this transformation will also be the future home of two additional hydrogen atoms bound to the new carbon, so these must initially be present as non-interacting atoms called ``dummy atoms'', which retain their bonded interactions but do not interact with the rest of the system.
 Bond parameters as well as partial charges between the changing atoms are adjusted accordingly between the initial and final state.
@@
 
-In contrast, in a dual topology alchemical free energy calculation, no atoms are allowed to change type~\cite{boresch1999rolea, shirts2012best}.
-This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxy methyl group, then passing through an intermediate state where some atoms are partially interacting--- particularly, those atoms which are becoming dummy atoms or ceasing to be dummy atoms~\cite{mobley2014blind}.
+In contrast, in a dual topology alchemical free energy calculation, \emph{no atoms are allowed to change type}~\cite{boresch1999rolea, shirts2012best}.
+This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxy methyl group, then passing through an intermediate state where some atoms are partially interacting---particularly, those atoms which are becoming dummy atoms or ceasing to be dummy atoms~\cite{mobley2014blind}.
 The transformation finally culminates in a state where benzyl alcohol is present along with the additional dummy atom which was previously a corresponding hydrogen of the benzene.
@@
     \includegraphics[width=0.95\columnwidth]{figures/fig3_topol/Figure.pdf}
-    \caption{\textbf{Two common topologies for alchemical calculations: single and dual topology.} \textbf{Left}: A single topology uses softcore potentials to convert from one type of atom to an other.
-Dummy atoms (Du) are used when there is no corresponding maximum common substructure match between the two molecules for certain atoms.
-\textbf{Right}: The dual topology does not convert one species to another, but only converts between Du atoms and an interacting species, but usually uses softcore potentials for this.
+    \caption{\textbf{Two common topologies for alchemical calculations: single and dual topology.} \textbf{Left}: A single topology converts from one type of atom to an other.
+Dummy atoms (Du) are used when there is no corresponding maximum common substructure match between the two molecules for certain atoms, using a soft-core interaction to improve overlap between the dummy atoms and the "real" atoms.
+\textbf{Right}: The dual topology does not convert one species to another, but only converts between Du atoms and an interacting species, but usually uses soft-core potentials for this.
 The 'mixed' intermediate atoms (M) are used in both dual and single topology approaches.
@@
 As far as we are aware, the distinction between single and dual topology approaches is made primarily for two main reasons.
-First, this choice affects the alchemical pathway followed, and thus may affect convergence properties -- though we are not yet aware of a study of the relative efficiency and merits of these two approaches.
-Additionally, historically, some simulation packages one approach and not the other, meaning that the distinction was functional.
-
-Some additional terms are also employed.
+First, this choice affects the alchemical pathway followed, and thus may affect convergence properties---though we are not yet aware of a study of the relative efficiency and merits of these two approaches.
+Additionally, historically, some simulation packages implemented only one approach and not the other, meaning that the distinction was functional.
+
+Some additional terms have also been employed to talk about these different intermediate pathways .
 Particularly, some studies refer to a ``hybrid'' topology approach to free energy calculations~\cite{gapsys2015pmx, gapsys2016accurate, gapsys2020large}, though this term may not have yet achieved widespread use.
@@
 Please make sure to check what approach is used with your software package of choice, or whether it supports your choice of approach (GROMACS and GROMOS, for example, support both).
-To our knowledge efficiency differences have not been thoroughly explored, though conventional wisdom suggests that fewer dummy atoms are better, as introducing or removing atomic sites is usually more difficult, requiring more intermediate steps ~\cite{liu2013lead, mobley2012perspective}.
+To our knowledge, efficiency differences have not been thoroughly explored, though conventional wisdom suggests that fewer dummy atoms are better, as introducing or removing atomic sites is usually more difficult, requiring more intermediate steps ~\cite{liu2013lead, mobley2012perspective}.
 
@@
 
-Single topology relative calculations, and calculations based on substructure searches, only work if in fact the ligands share a common substructure, e.g.
+Single topology relative calculations and calculations based on substructure searches only work if in fact the ligands share a common substructure, e.g.
 are part of a congeneric series, see Fig.~\ref{fig:fig_mcss}.
-If no common substructure is shared, then alternative dual or separated topology free energy calculations are needed, where one would co-localize a pair of compounds in a binding site, exclude their interactions with one another, and compute the relative binding free energy by turning one molecule on from being dummy atoms while turning the other off.
+If no common substructure is shared, then alternative dual or separated topology free energy calculations are needed.
+One would co-localize a pair of compounds in a binding site, exclude their interactions with one another, and compute the relative binding free energy by turning one molecule on from being dummy atoms while turning the other off.
 To our knowledge no general pipeline for such calculations yet exists and this would likely remain a research problem.
@@
 
-\paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss} \textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
+\paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
 Some approaches have attempted to address this~\cite{clark2019relative} but a general solution is not yet in mainstream use, though FEP+ implements one solution.
@@
 Based on the input ligand series, a perturbation map or network can be planned.
-Recent heuristics have shown the more connected the perturbation network the better, however, there is a way to optimize network structure while minimizing the number of perturbations that need to be computed reducing the resulting computational cost~\cite{yang2020optimal,xu2019optimal}.
-Sometimes the introduction of intermediates that are not part of the original congeneric series are essential to avoid ring breaking, or deal with perturbations that would otherwise result in large numbers of atoms being inserted or deleted.
+Recent heuristics have shown the more connected the perturbation network the better.
+However, there is a way to optimize network structure while minimizing the number of perturbations that need to be computed reducing the resulting computational cost~\cite{yang2020optimal,xu2019optimal}.
+Sometimes the introduction of intermediates that are not part of the original congeneric series are essential to avoid ring breaking, or to deal with perturbations that would otherwise result in large numbers of atoms being inserted or deleted.
 Some commercial tools have good underlying heuristics but may fail with complicated input, needing user validation in particular when dealing with chiral compounds.
@@
 
-In some cases, during the lead optimization stage, or for very large datasets that would benefit from rougher initial free energy ranking, or in cases where perturbations would be rather large a star shaped network as seen in Fig.~\ref{fig:fig_types_of_networks} \textbf{A} is used.
-However, adding redundancy into the network means that a better error analysis can be carried out, by looking at cycle closure errors as discussed in sec.~\ref{sec:are-they-good}, with an example given in Fig.~\ref{fig:fig_types_of_networks} \textbf{B}.
-
-Methods in experimental design have been applied to the construction of the perturbation maps.
-Yang et al.~\cite{yang2020optimal} optimized the perturbation map by selecting a fixed number of calculations from the pairwise perturbations so that the resulting set of calculations minimize the total variance.
-Xu~\cite{xu2019optimal} optimized the perturbation map by allocating different amounts of simulation time to different pairwise perturbations so as to minimize the total variance, given the total simulation time of all the perturbation calculations.
+In some cases, during the lead optimization stage, or for very large datasets that would benefit from rougher initial free energy ranking, or in cases where perturbations would be rather large, a star shaped network as seen in Fig.~\ref{fig:fig_types_of_networks} \textbf{A} is used.
+However, adding redundancy into the network which means that a better error analysis can be carried out, by looking at cycle closure errors as discussed in sec.~\ref{sec:are-they-good}, with an example given in Fig.~\ref{fig:fig_types_of_networks}\textbf{B}.
+
+Methods from experimental design have been applied to the construction of the perturbation maps.
+Yang et al.~\cite{yang2020optimal} showed how to optimize the perturbation map by selecting a fixed number of calculations from the pairwise perturbations so that the resulting set of calculations minimize the total variance.
+Xu~\cite{xu2019optimal} showed how to optimize the perturbation map by allocating different amounts of simulation time to different pairwise perturbations so as to minimize the total variance, given the total simulation time of all the perturbation calculations.
 Both approaches lead to substantial reduction in the statistical error of the estimated free energies.
@@
 Commonly, bonds involving hydrogen are constrained to a fixed length using algorithms such as SHAKE or LINCS, allowing the use of longer timesteps~\cite{krautler2001fast}.
-However, in single topology relative free energy calculations, the atoms involved might be mutated to other atom types -- for example, in a mutation of methane to methanol, one hydrogen might become an oxygen atom.
+However, in single topology relative free energy calculations, the atoms involved might be mutated to other atom types---for example, in a mutation of methane to methanol, one hydrogen might become an oxygen atom.
+The bonds with such atoms might not have any constraints, or if all bond are constrained, would have constraints of different lengths.
 Some molecular dynamics engines are not set up to recognize this change, or at least not to correctly include contributions to the free energy from changing constraints/constraint length, so results for a transformation can be erroneous.
@@
 For example, bonds can be transformed in both GROMACS and GROMOS, because contributions of LINCS (GROMACS) and SHAKE (GROMOS) contrained bonds are added to the $\frac{dH}{d\lambda}$ term~\cite{pearlman1993determining, straatsma1992holonomica, pearlman1991overlooked, gunsteren1989computer}.
-However, the constraints are not taken into account when calculating energy differences to other intermediate states (as the effect is entropic, not energetic).
+However, the constraints are not taken into account when calculating energy differences to other intermediate states as the effect is entropic, not energetic.
 User manuals should carefully checked for how these effects are included if a constrained bond changes in length.
@@
 Absolute free energy calculations involve completely removing the interactions between the ligand or solute and its environment, taking it to a non-interacting state that may or may not retain intramolecular non-bonded interactions.
-This non-interacting state can then be shifted between environments (from the protein to water, or from one solution to another) without changing its free energy other than that due to the changing volume of the simulations, and then interactions can be restored in the new environment.
+This non-interacting state can then be shifted between environments---from the protein to water, or from one solution to another---without changing its free energy other than that due to the changing volume of the simulations, and then interactions can be restored in the new environment.
 
@@
 \paragraph{Handling the standard state in absolute free energy calculations.}
-For solvation free energy calculations, handling the standard state is typically straightforward, and treating it correctly simply means ensuring that the non-interacting solute is taken to the same (or equivalent) final reference state in both environments, e.g.
-that the transformation involves a 1 M to 1 M equivalent transfer free energy (where the non-interacting solute still occupies essentially the same volume as the solute in the interacting system).
+For solvation free energy calculations, handling the standard state is typically straightforward, and treating it correctly simply means ensuring that the non-interacting solute still occupies essentially the same volume as the solute in the interacting system.
 So typically in such cases no special care is required to ensure the correct standard state, as long as the \emph{experimental} data being analyzed uses the same standard state.
-If this is not the case, a simple entropic correction is needed.
+If this is not the case, a simple entropic correction to the free energy of $k_B T\ln (V_f/V_i) = k_B T \ln (C_i/C_f)$ to the experimental data is needed.
 
@@
 A sufficiently thorough exploration of the simulation box by the non-interacting ligand is, however, required for the formula to be valid.
-This is typically hard to achieve as the exploration process is governed by diffusion.
+This is typically hard to achieve as the exploration process is governed by diffusion, and weak transient nonspecific binding will occur at other sites on the protein. 
 The addition of a restraint limits the volume available to the non-interacting ligand, thus speeding the convergence of the sampling.
@@
 In principle, all of these forms will yield correct binding free energies in the limit of adequate sampling if their effects and connection to the standard state are correctly handled, but they have different strengths and weaknesses.
-For example, with more involved restraints, sampling at intermediate $\vec{\lambda}$ values will usually not need to be as extensive but more computational effort must go to computing the restraining free energy.
+For example, with more involved restraints, sampling at intermediate $\vec{\lambda}$ values will usually not need to be as extensive but more computational effort must go to computing the free energy to turn on the restraints.
 Additionally, such restraints would typically keep the ligand from exploring alternative binding modes.
 This  restriction may be undesirable when using Hamiltonian $\lambda$ exchange or expanded ensemble techniques where allowing the ligand to exchange binding modes when it is non-interacting could provide sampling benefits~\cite{wang2013identifying}.
-More specifically, flat-bottom restraints might allow a ligand to explore multiple binding sites, harmonic restraints multiple binding modes within a site, while Boresch restraints a single binding mode within a single site.
+More specifically, flat-bottom restraints might allow a ligand to explore multiple binding sites, harmonic restraints multiple binding modes within a site, while Boresch restraints only allow a single binding mode within a single site.
 See additional discussion of the possibility of multiple binding modes in Sec.~\ref{sec:multiple_binding_modes} below.
@@
 For tightly bound ligands, virtually all reasonable definitions of the bound state will lead to be equivalent free energies, since the partition function will be dominated by a relatively small number of low-energy poses.
+
 For weak binders, this simplification breaks down.
@@
 Therefore, care must be taken to ensure that a reasonable definition of the binding site is used~\cite{wang2013identifying}.
-In absolute calculations, this applies to the fully interacting state in the complex leg of the thermodynamic cycle (top-right state in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}), while in relative calculations this must be true at both end states of the complex leg (top- and bottom-right states in Fig.~\ref{fig:fig_binding_thermodynamic_cycle}).
+
+In absolute calculations, this need to explicitly define a binding site applies to the fully interacting state in the complex leg of the thermodynamic cycle (top-right state in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}), while in relative calculations the binding site must be defined at both end states of the complex leg (top- and bottom-right states in Fig.~\ref{fig:fig_binding_thermodynamic_cycle}).
 In principle, this requires defining which conformations are considered to be bound before running the calculation, but it is common practice to start the simulation with the ligand already placed in the binding site and rely on kinetic trapping to maintain the bound complex.
-This strategy, however, can fail when the dissociation rate of the ligand has the same or smaller order of magnitude than the length of the simulation.
+
+However, this strategy of using kinetic trapping to maintain the bound complex can fail when the dissociation rate of the ligand has the same or smaller order of magnitude than the length of the simulation.
 This is typical of weak binders such as fragments binding shallow pockets with $\mu$M-mM affinities~\cite{georgiou2017pushing,pan2017quantitative}.
-In this case, using a flat-bottom or harmonic restraint between receptor and ligand in the bound state(s) can prevent dissociations~\cite{georgiou2017pushing,rizzi2019sampl6}.
-We stress that this is normally avoided as it generally introduces bias in the free energy estimate, which is why the restraint is usually activated only in the intermediate states in absolute calculations.
+In the case of weak binders, using a flat-bottom or harmonic restraint between receptor and ligand in the bound state(s) can prevent dissociations~\cite{georgiou2017pushing,rizzi2019sampl6}.
+We stress that this type of restraint is normally avoided as it generally introduces bias in the free energy estimate, which is why the restraint is usually activated only in the intermediate states in absolute calculations.
 The bias can be corrected through reweighting schemes~\cite{rizzi2019sampl6}, but this post-processing step can be avoided if a flat-bottom restraint is used and the ligand is never sampled while hitting the potential wall during the simulation in the bound state as the numerical correction will be exactly zero.
@@
 
-There are two main potential solutions to avoid artifacts due to changes in net charge: Correcting for the introduced artifacts, or avoiding changing the net charge.
+There are two main potential solutions to avoid artifacts due to changes in net charge: avoiding changing the net charge, and correcting for the introduced artifacts.
 
@@
 When free energy calculations \emph{do} need to change the charge of a ligand or solute, the literature does not yet seem to indicate what approach should be preferable, so considerable care should be taken.
-We are not yet aware of a careful comparison of charge corrections versus other approaches such as decoupling an ion at the same time, so in our view the issue of proper handling of charge mutations in the context of alchemical calculations remains a research problem.
+We are not yet aware of a careful comparison of charge corrections versus other approaches such as decoupling an ion at the same time, so in our view the issue of proper handling of charge mutations in the context of alchemical calculations remains a research problem, and several papers give good guides for exploring the problem further~\cite{lin2014overview, ohlknecht2020correcting, rocklin2013calculating}.
 
@@
 In principle, because of the path independence of the free energy, any arbitrary pathway will give the correct free energy change, but the choice of pathway will greatly affect the efficiency of the calculations.
-
 Some choices are particularly crucial---for example, transformations involving insertions or deletions of atoms should employ a soft-core potential path for Lennard-Jones or other interactions with repulsive interactions that go to infinite energy at small radius~\cite{beutler1994avoiding, beutler1994molecular,gapsys2012new}.
@@
 For a 'good' path to work and give a sequence of states with maximally similar configurations, sufficient similarity in potential energies is required.
-Fig.~\ref{fig:fig_what_is_lambda} \textbf{A} and \textbf{B} illustrate this.
-Fig.~\ref{fig:fig_what_is_lambda} \textbf{A} shows in a pictorial way a soft-core potential can be applied across different $\vec{\lambda}$s.
-Fig.~\ref{fig:fig_what_is_lambda} \textbf{B} illustrates the potential energy distributions at the different $\vec{\lambda}$ intermediates, with sufficient overlap between neighboring $\vec{\lambda}$ states to ensure that reweighting estimators such as MBAR can be used for analysis (see Sec.~\ref{subsec:estimators}).
+Fig.~\ref{fig:fig_what_is_lambda}\textbf{A} and \textbf{B} illustrate this.
+Fig.~\ref{fig:fig_what_is_lambda}\textbf{A} shows in a pictorial way a soft-core potential can be applied across different $\vec{\lambda}$s.
+Fig.~\ref{fig:fig_what_is_lambda}\textbf{B} illustrates the potential energy distributions at the different $\vec{\lambda}$ intermediates, with sufficient overlap between neighboring $\vec{\lambda}$ states to ensure that reweighting estimators such as MBAR can be used for analysis (see Sec.~\ref{subsec:estimators}).
 The actual transformation is best handled with soft-core potentials of the form shown in Fig.~\ref{fig:fig_what_is_lambda} \textbf{C} and \textbf{B}, with more details given below.
@@
 \begin{equation}
-U(\vec{q},\vec{\lambda}) = (1-\vec{\lambda}) U_0(\vec{q}) + \vec{\lambda}U_1(\vec{q}) \end{equation}
+U(\vec{q},\vec{\lambda}) = (1-\vec{\lambda}) U_0(\vec{q}) + \vec{\lambda}U_1(\vec{q}) \end{equation},
 
@@
 
-For many energy terms, \textit{as long as a repulsive core remains on}, this is a very good approach.
-For example, it can be shown that if van der Waals repulsions are left on, then the linear approach is very nearly the optimal path possible for changing, removing, or inserting the electrostatic energy terms, with the path being within about 10--20\% of the minimum possible uncertainty~\cite{naden2015linear} for a fixed amount of simulation, as well as being nearly optimally efficient for van der Waals attractive terms with repulsion terms turned on~\cite{naden2014linear}.
+For many energy terms this is a very good approach, \textit{as long as a repulsive core remains on}.
+For example, it can be shown that if van der Waals repulsions are left on, then the linear approach is very nearly the optimal path possible for changing, removing, or inserting the electrostatic energy terms, with the alchemical path being within about 10--20\% of the minimum possible uncertainty~\cite{naden2015linear} for a fixed amount of simulation time, as well as being nearly optimally efficient for van der Waals attractive terms with repulsion terms turned on~\cite{naden2014linear}.
 Although we are not aware of any quantitative tests for dipolar or higher multipole terms, theoretically it should behave equally well for those systems.
@@
 Assume we are trying to go from a particle being present, and desire to make it disappear alchemically.
-If the repulsive terms are of the form $\epsilon(\frac{\sigma}{r})^{12}$, and if $\epsilon$ is 1 $k_BT$ at the temperature of interest, and we start with the particle present, we may solve for $(1-\vec{\lambda})(1 k_B T)\left(\frac{\sigma}{\frac{1}{2}\sigma}\right)^{12} = 1 k_B T$.
+If the repulsive terms are of the form $\epsilon(\frac{\sigma}{r})^{12}$, and if $\epsilon$ is 1 $k_BT$ at the temperature of interest, and we start with the particle present, we may solve for $(1-\vec{\lambda})(1 k_B T)\left(\frac{\sigma}{0.5\sigma}\right)^{12} = 1 k_B T$.
 This yields $\vec{\lambda} = 1-2^{-12} \sim 0.999976$.
@@
 
-where $r_{ij}$ is the distance between two particles $i$ and $j$, $\epsilon_{ij}$ and $\sigma_{ij}$ are the Lennard-Jones parameters corresponding to the interaction between particles $i$ and $j$, and $\alpha$ is a constant (0.5 is optimal for the specific functional form shown above).
+where $r_{ij}$ is the distance between two particles $i$ and $j$, $\epsilon_{ij}$ and $\sigma_{ij}$ are the Lennard-Jones parameters corresponding to the interaction between particles $i$ and $j$, and $\alpha$ is a constant.
+In particular, $\alpha=0.5$ is statistically optimal for the specific functional form shown above.
 This functional form has exactly the property we are looking for: it recovers the Lennard-Jones potential when $\vec{\lambda}=1$, and the at other endpoint ($\vec{\lambda}=0$), it is exactly zero for all $r_{ij}$ everywhere, and as $\vec{\lambda}$ goes to zero, the $\alpha(1-\vec{\lambda})$ term lowers the infinite energy in the core.
@@
 It has been shown that more complicated forms are not significantly more efficient than eq.~\ref{eq:softcore}~\cite{pham2012optimal}.
-We therefore recommend using the softcore potential given in eq.~\ref{eq:softcore}, unless there is a compelling reason otherwise.
+We therefore recommend using the soft-core potential given in eq.~\ref{eq:softcore}, unless there is a compelling reason otherwise.
 Using a similar equation to eq.~\ref{eq:softcore} may be acceptable in most circumstances if that is what is supported in your chosen software.
-However, if you are inserting or removing entire atomic sites, we heavily recommend against using the linear approach; it will be very difficult to get correct and converged results.
+However, if you are inserting or removing entire atomic sites, we heavily recommend against using the linear approach; it will be very difficult to get correct or converged results.
 
@@
 What about performing both transformations at the same time?
-We can not turn off the electrostatics linearly at the same time we turn off the Lennard-Jones terms, as it would leave infinitely large attractive and repulsive electrostatic terms "bare" at small $\vec{\lambda}$, resulting in the simulation crashing.
-It \textit{is} possible to apply the same soft core approach to the Coulomb interaction and this is indeed done in a number of implementations, in which case it is important that the Coulomb interaction is softened more rapidly than the Lennard-Jones interaction to avoid charge penetration issues, which  can be tricky to ensure for all types of perturbations ~\cite{steinbrecher2011softcore}.
+We cannot turn off the electrostatics linearly at the same time we turn off the Lennard-Jones terms, as it would leave infinitely large attractive and repulsive electrostatic terms "bare" at small $\vec{\lambda}$, resulting in the simulation crashing.
+It \textit{is} possible to apply the same soft-core approach to the Coulomb interaction as to the van der Waals interaction, and this is indeed done in a number of implementations.
+In this case, it is important that the Coulomb interaction is softened as rapidly or more rapidly than the Lennard-Jones interaction to avoid charge penetration issues into the repulsive core, which  can be tricky to ensure for multiple types of perturbations simultaneously~\cite{steinbrecher2011softcore}.
 
@@
 Our recommendation is to leave the intramolecular interactions on during the transformation for simplicity if there are no other known issues with this approach.
-The key thing to watch out for is whether the total potential energy, and therefore the intermediate ensembles sampled, change smoothly from beginning to end.
-These problems can be diagnosed by noticing lack of configuration space overlap between different simulations (see Sec.~\ref{sec:are-they-good}).
+The key feature of the simulation to watch out for is whether the total potential energy, and therefore the intermediate ensembles sampled, change smoothly from beginning to end.
+Problems of discontinuous changes of the potential energy can be diagnosed by noticing lack of configuration space overlap between different simulations (see Sec.~\ref{sec:are-they-good}).
 
 Relative calculations introduce additional choices, such as the order in which to modify nonbonded interactions.
-A common process in single topology relative calculations is to first remove electrostatic interactions of any atoms which will be deleted, then modify other non-bonded interactions, then restore electrostatic interactions of any atoms which are being inserted.
-Although this is a simpler path to understand cognitively and can take advantage of the soft-core potential from eq~\ref{eq:softcore}, this can lead to more intermediate steps and thus be more computationally expensive.
-Other schemes, such as simultaneously changing electrostatic and Lennard-Jones interactions with electrostatic ``soft-core'' potentials~\cite{steinbrecher2007nonlinear} may be implemented with fewer intermediate states but could require fine-tuning of electrostatic and Lennard-Jones softcore parameters to avoid numerical instabilities.
-At the time of writing, there has not been conclusive evidence to suggest one approach is better in general than the other, so discretion should be left up to the user as to what is viable from both hardware resources, and what the simulation software supports.
+A common process in single topology relative calculations is, as noted above, to first remove electrostatic interactions of any atoms which will be deleted, then modify other non-bonded interactions, then restore electrostatic interactions of any atoms which are being inserted.
+Although this is a simpler path to understand cognitively and can take advantage of the soft-core potential from Eq.~\ref{eq:softcore}, this can lead to more intermediate steps and thus be more computationally expensive.
+Other schemes, such as simultaneously changing electrostatic and Lennard-Jones interactions with electrostatic soft-core potentials~\cite{steinbrecher2007nonlinear}, as already discussed above, may be implemented with fewer intermediate states but could require fine-tuning of electrostatic and Lennard-Jones soft-core parameters to avoid numerical instabilities.
+At the time of writing, there has not been conclusive evidence to suggest the separate or simultaneous approach is in general better than the other, all factors considered, so discretion should be left up to the user as to what is viable from both hardware resources, and what the simulation software supports.
 
@@
 Some approaches have attempted to find alternative pathways to improve efficiency or find paths of low thermodynamic length~\cite{naden2014linear,naden2015linear,pham2012optimal}.
-For example, the enveloping distribution sampling (EDS) approach, and its multiple-replica and accelerated variant, attempts to improve efficiency by creating a single artificial intermediate state which simultaneously samples all end state phase spaces~\cite{perthold2018accelerateda,sidler2017efficientb, christ2007envelopinga}.
+For example, the enveloping distribution sampling (EDS) approach, and its multiple-replica and accelerated variant, works to improve efficiency by creating a single artificial intermediate state which simultaneously samples all end state phase spaces~\cite{perthold2018accelerateda,sidler2017efficientb, christ2007envelopinga}.
 When this can be done, it provides an extremely efficient way to calculate the relative free energy difference between multiple ligands from a single simulation.
@@
 
-In our view, there is still some room for further exploration of how to best choose transformation pathways, especially for relative binding calculations or more complex molecules, as most existing studies focused on smaller molecules.
-In principle, any pathway that connects the desired end states is rigorously correct, but as discussed above, different paths may differ dramatically in thermodynamic length and therefore efficiency.
+In our view, there is still some room for further exploration of how to best choose transformation pathways, especially for relative binding calculations or more complex molecules, as most existing studies focus on smaller molecules.
+As we have stressed, in principle, any pathway that connects the desired end states is rigorously correct, but as discussed above, different paths may differ dramatically in thermodynamic length and therefore efficiency.
 Additionally, some paths simply may not converge due to issues noted above such as those encountered without soft core potentials.
-However, the recommendations above are reasonable, reliable, and are likely not be that much less efficient than potentially more optimal choices~\cite{naden2014linear,naden2015linear,pham2012optimal}, as the real problems with low efficiency are lack of sampling of slow conformational modes.
+However, the recommendations above are reasonable, reliable, and are likely not be that much less efficient than potentially more optimal choices~\cite{naden2014linear,naden2015linear,pham2012optimal}, as the real problems with the efficiency of calculating free energies are lack of sampling of slow conformational modes, rather than the lack of efficiency of the transformations.
 
@@
 Fig.~\ref{fig:fig_sampling_scheme} illustrates the four most common schemes.
-The simplest approach involves running an independent simulation at each of the predefined $\vec{\lambda}$ values (see Fig.~\ref{fig:fig_sampling_scheme} \textbf{A}).
+The simplest approach involves running an independent simulation at each of the predefined $\vec{\lambda}$ values (see Fig.~\ref{fig:fig_sampling_scheme}\textbf{A}).
 This type of scheme is currently used for AMBER TI calculations~\cite{song2019using} and for Sire as implemented in BioSimSpace~\cite{hedges2019biosimspace}.
@@
 In this approach, the simulation at each $\vec{\lambda}$ can undergo periodic exchanges with neighboring $\vec{\lambda}$ values.
-This form of replica exchange (Hamiltonian replica exchange) is based on ideas developed from Monte Carlo simulations of spin glasses by Swendsen and Wang~\cite{swendsen1986replica}.
-With the Metropolis-Hastings acceptance criterion for exchanges, the generated ensemble of all replicas still samples from the Boltzmann distribution, thus this approach has been used in many different contexts for molecular simulations~\cite{sugita2000multidimensionala,sugita1999replicaexchangea, woods2003developmenta, jiang2010free}.
-The basic idea of the replica exchange scheme is shown in Fig.~\ref{fig:fig_sampling_scheme} \textbf{B}.
+This form of replica exchange, called Hamiltonian replica exchange, is based on ideas developed from Monte Carlo simulations of spin glasses by Swendsen and Wang~\cite{swendsen1986replica}.
+With the Metropolis-Hastings acceptance criterion for exchanges, the generated ensemble of all replicas still samples from the Boltzmann distribution for each replica.
+This approach has been used in many different contexts for molecular simulations~\cite{sugita2000multidimensionala,sugita1999replicaexchangea, woods2003developmenta, jiang2010free}.
+The basic idea of the replica exchange scheme is shown in Fig.~\ref{fig:fig_sampling_scheme}\textbf{B}.
 It is supported in various software packages that provide alchemical implementations, such as GROMACS~\cite{aldeghi2015accurate}, GROMOS~\cite{hritz2008hamiltoniana,hritz2007optimizationa}, FEP+~\cite{wang2015accurate}, and NAMD~\cite{jiang2019computing}.
+
 A third approach borrows ideas from simulated tempering~\cite{marinari1992simulateda}.
-In this scheme a single replica rapidly explores all of $\vec{\lambda}$ space by working out optimal weights that allow switching between different intermediate $\vec{\lambda}$ values, as seen in Fig.~\ref{fig:fig_sampling_scheme} \textbf{C} .
-This approach is also referred to as self adjusted mixture sampling~\cite{lyubartsev1992newa, li2007simulated, tan2017optimally} and while promising, has so far only been supported in OpenMM Tools~\cite{andrearizzi2019choderalab}.
+In this scheme a single replica rapidly explores all of $\vec{\lambda}$ space by working out optimal weights that allow switching between different intermediate $\vec{\lambda}$ values, as seen in Fig.~\ref{fig:fig_sampling_scheme}\textbf{C} .
+This approach is also referred to as self-adjusted mixture sampling~\cite{lyubartsev1992newa, li2007simulated, tan2017optimally} and while promising, has so far only been supported in OpenMM Tools~\cite{andrearizzi2019choderalab} and GROMACS.
+Although this approach allows multiple states to be simulated in a single simulation, the weights do not always converge to the proper equilibrium distribution, and care must be taken that the final results are converged.
+
 The last approach makes use of non-equilibrium simulations~\cite{aldeghi2018accurate}.
-In this approach, only end state $\vec{\lambda}$ replicas ($\vec{\lambda}$=0, $\vec{\lambda}=1$) are simulated at equilibrium; intermediate information is generated from non-equilibrium simulations that rapidly transition between end-states.
+In this approach, only end state $\vec{\lambda}$ replicas ($\vec{\lambda}$=0, $\vec{\lambda}=1$) are simulated at equilibrium; intermediate information is generated from non-equilibrium simulations that rapidly transition between end states.
 This approach is available in GROMACS and appears to be coming online in several other packages.
-A schematic of this approach is shown in Fig.~\ref{fig:fig_sampling_scheme} \textbf{D}.
+A schematic of this approach is shown in Fig.~\ref{fig:fig_sampling_scheme}\textbf{D}.
 
@@
 After a short simulation time symbolised by the arrow, the lambda-state is attempted to change until all N lambda states will be sampled.
-(\textbf{D}): Non-equilibrium sampling scheme, where two equilibrium simulations at the end-states are run as indicated by the blue and pink arrow.
-Non-equilibrium simulations are attempted at intervals to switch between the two end-states.}
+(\textbf{D}): Non-equilibrium sampling scheme, where two equilibrium simulations at the end states are run as indicated by the blue and pink arrow.
+Non-equilibrium simulations are attempted at intervals to switch between the two end states.}
     \label{fig:fig_sampling_scheme}
@@
 If these are not available in the code of choice, running independent simulations at different $\vec{\lambda}$ values can be acceptable, especially when configurational sampling is fast (Fig.~\ref{fig:fig_sampling_scheme} \textbf{A}).
-Single replica schemes and non-equilibrium schemes are not as established yet, but are very promising.
+Single replica schemes and non-equilibrium schemes are not as established yet because of potential failure modes, but are very promising for use in the near future.
 
@@
 In particular, if the rate of change in the free energy estimate is significant when this condition is met, the simulation may not be locally converged, and more sampling may be necessary to determine a stable free energy estimate which is no longer changing significantly over time.
-However, this is not the only metric which should be used, as the uncertainty only captures the information about the sampled phase space, not necessarily the entirety of the phase space.  
+
+However, this is not the only metric which can or should be used, as the uncertainty only captures the information about the sampled phase space, not necessarily the entirety of the phase space.  
 For example, convergence of relative free energy calculations in predictive simulations where the entire phase space is not known in advance, requires sampling the different kinetically stable states~\cite{mobley2012perspective}.
@@
 A target statistical uncertainty should be chosen at the onset of the simulation to avoid excessively long simulations, or falling into the trap of running until the free energy estimate is "good enough," which is subjective and has no defined criteria.
-This could be a fixed value such as $0.20$ $\mathrm{kcal/mol}$, or a functional quantity such as "below $0.5$ $\mathrm{kcal/mol}$ and $10\%$ of the free energy estimate." The user does not need to monitor this information in real-time and can choose to run simulations for fixed duration (either time or number of samples) and run analysis on the data collected thus far.
+This could be a fixed value such as $0.20$ kcal/mol, or a functional quantity such as "below $0.5$ kcal/mol and $10\%$ of the free energy estimate." The user does not need to monitor this information in real-time and can choose to run simulations for fixed duration (either time or number of samples) and run analysis on the data collected thus far.
 If more samples are needed, the simulations can be resumed, or, started again in different initial conditions.
@@
 Convergence in other alchemical observables should also be monitored to determine if the defined phase space has been sufficiently sampled and enough decorrelated samples have been drawn.
-These additional observables include, but are not limited to, the variance in $\frac{dU}{d\vec{\lambda}}$ across all $\vec{\lambda}$ values, calculating the variance in free energy using bootstrap analysis, and comparing differences in free energies calculated using different percentages of the simulation in both the forward and reverse directions~\cite{klimovich2015guidelines} (see Fig.
-~\ref{fig:convergence_forward_reverse}).
+These additional observables include, but are not limited to, the variance in $\frac{dU}{d\vec{\lambda}}$ across all $\vec{\lambda}$ values, calculating the variance in free energy using bootstrap analysis, and comparing differences in free energies calculated using different percentages of the simulation in both the forward and reverse directions~\cite{klimovich2015guidelines} (see Fig.~\ref{fig:convergence_forward_reverse}).
 
@@
 Results obtained from calculations with convergence issues should be checked for errors or run for longer before any confidence should be placed in conclusions drawn from their analysis.
-In relative calculations that share similar binding modes, for example, and do not induce large conformational changes when in complex with protein, the need to sample exhaustively to converge estimates in free energy differences is often not necessary due to the locality of sampling changes in the molecular topology and shared phase space of the core atoms.
+For example, in relative calculations that share similar binding modes and do not induce large conformational changes when in complex with protein, the need to sample exhaustively to converge estimates in free energy differences is often not necessary due to the locality of sampling changes in the molecular topology and shared phase space of the core atoms.
 However, even subtly induced changes in protein binding configuration will require more sampling or cause local convergence to a free energy estimate that has high error.
@@
 The uncertainty in the free energy, for example, can be estimated in multiple ways, e.g.
-through standard error propagation methods (including MBAR's estimator, which is based on the same principles as standard error propagation), through bootstrap methods, through multiple independent runs, etc.
-Independent of how the property is estimated, its important to remember that they are \textit{estimations of the property}, not the true underlying property itself. 
+through standard error propagation methods (including MBAR's estimator, which is based on the same principles as standard error propagation), through bootstrap methods, and through multiple independent runs.
+Independent of how the property is estimated, it is important to remember that results of any free energy analysis are \textit{estimations of the given property}, not the true underlying value of the property itself.
 These estimators are usually consistent estimators, meaning they will converge to the true answer in the limit of sufficient sampling, not necessarily unbiased ones though.
 As such, it is a good idea to subject different estimators to the same data to see if they yield either the same estimate (within error and bias), or if they fluctuate wildly.
-See for example the potential of mean force with respect to $\vec{\lambda}$ estimated from a bound simulation of a Tyk2 ligand pair of Wang et al.~\cite{wang2015accurate} for both the MBAR and TI estimators, as seen in Fig.~\ref{fig:pmf}.
+See, for example, the potential of mean force with respect to $\vec{\lambda}$ estimated from a bound simulation of a Tyk2 ligand pair of Wang et al.~\cite{wang2015accurate} for both the MBAR and TI estimators, as seen in Fig.~\ref{fig:pmf}.
 This is not a perfect method as some estimators, such as exponential averaging, will converge significantly more slowly, relative to more accurate estimators like MBAR. 
@@
 If, for instance, the free energy estimator selected is thermodynamic integration, then values of $\frac{dU}{d\vec{\lambda}}$ at uncorrelated data points must be collected.
-Once a combination of knowing what type of simulation you will run, which alchemical topology you will simulate, what alchemical path you will simulate along, and what your stopping conditions are, then you are ready to enumerate the information you should capture.
+Once you have made a choice of the combination of the type of simulation you will run, which alchemical topology you will simulate, what alchemical path you will simulate along, and what your stopping conditions are, then you are ready to enumerate the information you should capture.
 Below is a sample of the minimal information you need for a set of common estimators (discussed in more detail in Sec.~\ref{subsec:estimators}):
@@
     \item Weighted Histogram Analysis Method (WHAM) and Multistate Bennett Acceptance Ratio (MBAR) both need the complete set of $\Delta u_{k,j} \, \forall \, j=\{1...K\}$.
-WHAM must have this information binned.
+WHAM must have this same information binned with some choice of bin width small enough not to affect the results.
 \end{itemize}
@@
 The BAR estimator may be a better, and simpler choice at that point as you will have at least the same level of information.
+
 The potential energy differences required for EXP, BAR, MBAR, and WHAM can be calculated either during the simulation or in post-processing.
-It is recommended to calculate the potential differences in code when possible to avoid extra overhead and possible errors produced by running the simulation twice, and to reduce the amount of stored information.
-Although TI must usually be calculated in code, as it requires the derivative, there is one condition under which it actually has the fastest computation time. 
-If the alchemical path you have chosen is a linear alchemical path, then you get $\frac{du}{d\vec{\lambda}} = u_0(\vec{q}) - u_1(\vec{q})$, which is the difference between the initial and final states.
+It is recommended to calculate the potential differences in code when possible to avoid extra overhead and possible errors produced by evaluating the energy of the configuration twice, and to reduce the amount of stored information.
+Although potential energy derivatives must usually be calculated in code there is one condition under which they can be easily computed in post-simulation analysis. 
+If the alchemical path you have chosen is a linear alchemical path, then $\frac{du}{d\vec{\lambda}} = u_0(\vec{q}) - u_1(\vec{q})$, which is the difference between the initial and final states, which are already calculated by the simulation and can be recorded easily without additional computational expense.
 However, because of the problems with linear paths already discussed in this paper, this simplification is rarely that useful.
@@
 Free energy information should generally be saved more frequently than coordinate data, approximately at the rate that uncorrelated samples are produced.  
-The on-disk size of the data for free energy estimation is often significantly smaller than full atomic coordinates, so the information should be collected frequently. 
+The on-disk size of the data for free energy estimation is often significantly smaller than full atomic coordinates, so the information can easily be collected frequently. 
 However, the information should not be collected \textit{every} time step, as most free energy techniques are operated at equilibrium, and need equilibrated \textit{and decorrelated} samples for an unbiased estimate.
-A sample collected every time step will likely result in most samples being discarded due to decorrelation routines in the analysis.
+Samples collected every time step will likely result in most samples being discarded due to the detection of correlation in the time series by decorrelation routines in the analysis.
 However, if it is computationally cheap and disk space is plentiful, do save often.
@@
 However, this carries the notable caveat that such estimates require accurate estimation of the correlation time which, if important motions are slow compared to simulation timescales, can be difficult.
-Still, this metric provides a good rule of thumb, and as long as conformational transitions are captured by the simulation, increasing the aggregate simulation time by a factor of $T$ will reduce the uncertainty by a factor of approximately $\sqrt(T)$.
+Still, this metric provides a good rule of thumb, and as long as conformational transitions are captured by the simulation, increasing the aggregate simulation time by a factor of $T$ will reduce the uncertainty by a factor of approximately $\sqrt{T}$.
 
@@
 In a discovery setting, new ligands can have unknown or at least uncertain binding modes~\cite{kaus2015how, plountprice2000analysis,mobley2009binding,calabro2016elucidation}, complicating binding free energy estimation.
-This uncertainty is because it is usually not desirable to estimate a binding affinity for a ligand which already has an available bound structure, since such a compound has already been tested.
+
 To deal with prospective ligands with unknown binding modes, discovery projects commonly assume that modifications of functional groups on a common scaffold result in a consistent binding mode across all members of a series.
@@
 It is worth highlighting a recent SAMPL blind challenge on HIV integrase as an illustration of this. 
-Many submissions, using state-of-the-art methods, had difficulty even predicting which \emph{binding site} ligands would bind in (most submissions placed more than half of the ligands into the incorrect binding site), and even given correct binding sites, the binding mode within each site was also quite difficult to predict~\cite{mobley2014blind}.
+Many submissions, using state-of-the-art methods, had difficulty even predicting which \emph{binding site} ligands would bind in---most submissions placed more than half of the ligands into the incorrect binding site---and even given correct binding sites, the binding mode within each site was also quite difficult to predict~\cite{mobley2014blind}.
 The best performing submission for predicting binding modes actually ended up being a human expert (aided by computational tools) with more than 10 years of experience on the particular target~\cite{voet2014combining}, rather than a fully automated approach.
@@
 
-One approach which has shown some success is to retain diverse potential binding modes from docking, perform short MD simulations of these to identify distinct stable binding modes, and then consider these in subsequent calculations~\cite{gallicchio2014virtual, mobley2006use,rocklin2013blind, boyce2009predicting, mobley2007predicting}.
+One approach which has shown some success in identifying accurate binding modes \emph{de novo} is to retain diverse potential binding modes from docking, perform short MD simulations of these to identify distinct stable binding modes, and then consider only these stable modes in subsequent calculations~\cite{gallicchio2014virtual, mobley2006use,rocklin2013blind, boyce2009predicting, mobley2007predicting}.
 
@@
 \paragraph{Handling multiple potential binding modes within absolute calculations.}
-Within absolute binding free energy calculations, multiple potential binding modes can be handled by two main strategies: Consider each binding mode separately (a separation of states strategy) or sample all binding modes within a single simulation~\cite{mobley2012perspective}.
+Within absolute binding free energy calculations, multiple potential binding modes can be handled by two main strategies: Considering each binding mode separately (a separation of states strategy) or sampling all binding modes within a single simulation~\cite{mobley2012perspective}.
 This couples to the choice of restraints selected, as some restraints will allow transitions between binding modes and even binding sites (Sec.~\ref{sec:standardstate-restraints}), and others do not.
@@
 \begin{eqnarray}
-\Delta f_{01} & = & \ln \expect{e^{-(u_1(\vec{q}) - u_0(\vec{q}))}}_0 =  \ln \expect{e^{-\Delta u(\vec{q})} }_0
+\Delta f_{01} & = & \ln \expect{e^{-(u_1(\vec{q}) - u_0(\vec{q}))}}_0 =  -\ln \expect{e^{-\Delta u(\vec{q})} }_0
 \end{eqnarray}\label{eqn.zwanzig}
@@
 In the case of NVT (canonical) sampling and assuming the masses do not change, then $u$ is simply $U/k_BT$, and $f$ is $F/k_BT$, but it can be generalized to other ensembles with the proper definition of $f$ and $u$.
-Described in words, we take the samples generated during our run with the potential energy function $u_0(\vec{q})$ and calculate what the difference in energy would be if we switched to potential energy function $u_1(\vec{q})$, and average the exponential of the negative energy difference to get the negative of the exponential of the free energy difference.
-The original distributions, P($u_0$) as generated at $\vec{\lambda}=0$ and P($u_1$) would look like those seen in Fig.~\ref{fig:fig_sampling_scheme} \textbf{A}-\textbf{C} on the right hand side.
+Described in words, we take the samples generated during our run with the potential energy function $u_0(\vec{q})$ and calculate what the difference in energy would be if we switched instantaneously to the potential energy function $u_1(\vec{q})$, and average the exponential of the negative energy difference to get the negative of the exponential of the free energy difference.
+The original distributions, P($u_0$) as generated at $\vec{\lambda}=0$ and P($u_1$) would look like those seen in Fig.~\ref{fig:fig_sampling_scheme}\textbf{A}-\textbf{C} on the right hand side.
 Reevaluating requires almost no extra code functionality to perform; one need only to save a full precision trajectory, and run an unmodified molecular simulation code using the $u_1$ in order to calculate the new energies of stored snapshots.
@@
 
-Although the Zwanzig equation is formally correct (as long as the two states considered sample the same phase space volume, which is true for standard molecular models), it has some very important numerical issues that mean that it often performs badly for standard free energy calculations, even for small molecules~\cite{shirts2005comparison,lu2003appropriate}.
-One can show that if the standard deviation of the difference $\Delta u(\vec{q}) = u_1(\vec{q})-u_2(\vec{q})$ over all sampled $\vec{q}$ is large (which in this case, means only several times $k_BT)$, then very few samples contribute to the average, and the answer will be both biased and extremely noisy~\cite{lelievre2010free}.
+Although the Zwanzig equation is formally correct as long as the two states considered sample the same phase space volume, which is true for standard molecular models, it has some very important numerical issues that mean that it often performs badly for standard free energy calculations, even for small molecules~\cite{shirts2005comparison,lu2003appropriate}.
+One can show that if the standard deviation of the difference $\Delta u(\vec{q}) = u_1(\vec{q})-u_2(\vec{q})$ over all sampled $\vec{q}$ is large, which in this case, means only several times $k_BT$, then very few samples contribute to the average, and the answer will be both biased and extremely noisy~\cite{lelievre2010free}.
 Essentially, the method is dominated by contributions of rare snapshots~\cite{jarzynski2006rare, wu2005phasespaceb, wu2005phasespacec}.
@@
 Usually, the endpoints of interest in a free energy calculation are sufficiently different that we will need a chain of states that gradually change the potential energy function from $u_0$ to $u_1$, as discussed in Sec.~\ref{sec:important_path}.
-You can simply carry out BAR between each pair of states $\Delta f_{1 \rightarrow N} = \Delta {f_{1\rightarrow 2}} + \Delta {f_{2\rightarrow 3}} +  \ldots + \Delta f_{N-1\rightarrow N}$.
+You can carry out a BAR estimate between each pair of states $\Delta f_{1 \rightarrow N} = \Delta {f_{1\rightarrow 2}} + \Delta {f_{2\rightarrow 3}} +  \ldots + \Delta f_{N-1\rightarrow N}$.
 
@@
 Researchers have tried a large number of different integration schemes~\cite{resat1993studies,jorge2010effect,shyu2009reducing}.
-However, many other integration choices require specific choices of $\vec{\lambda}$ to minimize bias, which makes them unsuitable when the intermediates
+However, many integration choices require specific choices of $\vec{\lambda}$ to minimize bias, which makes them unsuitable when the intermediates
 have widely-varying levels of uncertainty.
 For example, integrating a cubic spline interpolation provided negligible benefits over a simple trapezoid rule~\cite{paliwal2011benchmark}.
+As fitting to higher order polynomials can have numerical instabilities for some energy functions, and because alternate functional forms might only be appropriate with some types of transformations, expertise and experience is required to perform such numerical integration modifications.
 For starting researchers, we therefore recommend the simple trapezoid rule scheme, as it allows for maximal flexibility in which values of $\vec{\lambda}$ are simulated.
-As fitting to higher order polynomials can have numerical instabilities for some energy functions, and because alternate functional forms might only be appropriate with some types of transformations, expertise and experience is required to perform such numerical integration modifications.
 In practice, adding 2-3 more intermediate states is typically sufficient to match the performance of these more complicated numerical quadrature schemes.
@@
 
-MBAR can also be thought of as the Zwanzig estimator of the free energy to state $i$ where the sampled distribution is the \textit{mixture distribution} of all the other samples thrown together in one ``pot'', defined by $p_m(\vec{q}) = N^{-1} \sum_k N_k \exp(f_k-u_k\vec{q})$, which is the weighted average of all the individual normalized probability distributions from the simulations that are performed.~\cite{shirts2017reweighting}.
+MBAR can also be thought of as the Zwanzig estimator of the free energy to state $i$ where the sampled distribution is the \textit{mixture distribution} of all the other samples thrown together in one ``pot'', defined by $p_m(\vec{q}) = N^{-1} \sum_k N_k \exp(f_k-u_k\vec{q})$, which is the weighted average of all the individual normalized probability distributions from the simulations that are performed~\cite{shirts2017reweighting}.
 
@@
 It is important to consider the variation in your computed free energies from your equilibrium simulations, in order to obtain an estimate of uncertainty of the obtained value for the free energies of interest.
-A recent best practices paper by Grossfield et al.,~\cite{grossfield2018best} provides substantial detail on how to estimate uncertainties from molecular simulations and is a good starting point for this topic.
+A recent Best Practices paper by Grossfield et al.,~\cite{grossfield2018best} provides substantial detail on how to estimate uncertainties from molecular simulations and is a good starting point for this topic.
 The uncertainty of a free energy estimate is limited however, as it is only an estimate of the configurations sampled and cannot contain any information from the phase space not sampled during a simulation.
@@
 the average probability that a sample generated at state $\vec{\lambda}_{j}$ can be observed at state $\vec{\lambda_{i}}$.
-As this probability is computed considering the samples from all states (and not just the adjacent states), the values in each row and column add up to 1.
-In this analysis, the goal is to ensure every state has overlap with its neighbors in both directions -- so that off-diagonal elements are sufficiently larger than zero.
+As this probability is computed considering the samples from all states, and not just the adjacent states, the values in each row and column add up to 1.
+In this analysis, the goal is to ensure every state has overlap with its neighbors in both directions, indicated by off-diagonal elements that are sufficiently larger than zero.
 For accurate calculations, the matrix should be at least tridiagonal.
@@
 In order to obtain a reliable estimate of the free energy all neighbouring states must be connected, i.e.
-there must be sufficient overlap between the samples of these states (general description: ${O}_{i,j,i\ne j}\ge$ threshold).
+there must be sufficient overlap between the samples of these states, such that ${O}_{i,j,i\ne j}\ge$ threshold).
 However, due to the mathematical derivation it is difficult to explicitly describe the relation of the overlap matrix and the variance by formulae.
@@
 the diagonals above and below the main diagonal) should at least be 0.03 to obtain a reliable free energy estimate~\cite{klimovich2015guidelines}.
-Smaller values should be considered as a warning sign (see Fig.~\ref{fig:overlap} \textbf{C}), as the variance tends to be underestimated in case of poor overlap.
+Smaller values should be considered as a warning sign (see Fig.~\ref{fig:overlap}\textbf{C}), as the variance tends to be underestimated in case of poor overlap.
 
@@
 
-Fig.~\ref{fig:overlap} \textbf{A}, \textbf{B}, and \textbf{C} shows examples of good, mediocre, and poor overlap respectively.
+Fig.~\ref{fig:overlap}\textbf{A}, \textbf{B}, and \textbf{C} shows examples of good, mediocre, and poor overlap respectively.
 For Fig.~\ref{fig:overlap} \textbf{A}, the probability to find a sample from state $i$ in its neighbouring state $j$ is about 0.2 for all states adjacent to the main diagonal, and hence the overall connectivity is good.
@@
 
-Significant deviations from this typically indicate insufficient configurational sampling along the lambda schedule of one or more of the transformations involved.
+Significant deviations of agreement from the same transformation by different routes  typically indicate insufficient configurational sampling along the lambda schedule of one or more of the transformations involved.
 This approach may be generalised to sets of connected transformations given the requirement that the sum of free energy changes along edges of a closed cycle should be zero.
@@
 \item \textbf{Sampling of the binding site by the ligand:} Make sure the ligand samples the binding site reasonably tightly for its expected potency and fit, and that it does not depart out of binding site in the coupled end state if it is a moderate to strong binder.
-\item \textbf{Consistency of free energy estimates across different estimators} Significant discrepancies (further outside the error estimates than would be plausible) between free energies calculated with different free energy estimators such as TI, BAR, and MBAR.
+\item \textbf{Consistency of free energy estimates across different estimators} Significant discrepancies, meaning results that further outside the mutual error estimates than would be plausible statistically, between free energies calculated with different free energy estimators such as TI, BAR, and MBAR.
 All of these estimators converge to the same results with sufficient sampling.
@@
 \item \textbf{Behaviour of correlation times:} Correlation time that does not vary relatively smoothly as a function of $\vec{\lambda}$.
-Discontinuities indicate that the system is sampling significantly different configurations with only small changes to the Hamiltonian changes.
+Discontinuities in correlation time with $\vec{\lambda}$ indicate that the system is sampling significantly different configurations with only small changes to the Hamiltonian changes.
 This usually indicates sampling problems.
@@
 There should be clear physical reasons that simulation in the complex has different torsional distributions that the ligand in the solvent.
-\item \textbf{Free energy dependence on $\vec{\lambda}$} The free energy should vary relatively smoothly with $\vec{\lambda}$.
+\item \textbf{Free energy dependence on $\vec{\lambda}$} The free energy difference between states should vary relatively smoothly with $\vec{\lambda}$.
 If it varies drastically, then either there need to be finer sampling in $\vec{\lambda}$ in this region, or there are sampling problems there.
 \item \textbf{Convergence of free energy} The free energy should clearly converge as a function of simulation time (Fig. ~\ref{fig:convergence_forward_reverse}).
-\item If using nonequilbirum methods, is the result independent of the speed at which the nonequilibrium change is performed?
-Nonequilibrium methods are in theory independent of the switching time in the limit of good sampling unless the switching time is simply too short.
+\item If using non-equilibrium methods, is the result independent of the speed at which the non-equilibrium change is performed?
+Non-equilibrium methods are in theory independent of the switching time in the limit of good sampling unless the switching time is simply too short.
 %\item Other Diagnostics for non-equilibrium methods, robustness for switching time etc.?
 \item \textbf{Visualization of data} In general, inspect output data such as energies and visualize the simulation trajectories and assess if they match your expectations.
-Many issues can be spotted by a straight forward visualiztion.
+Many issues can be spotted by a straight forward visualization.
 \end{itemize}
@@
 We encourage the following standard set of analyses and ways to represent data.
-\paragraph{Statistics to include}
+\paragraph{Statistics to include in reporting data}
 As with any modelling technique, misuse of statistical analysis can skew the perception of how well models perform in free energy predictions.
@@
 
-As we are interested mainly in the linear relationship between the alchemical free energy predictions and the experimentally -determined affinity values, plots should be depicted with the same range on both axes (i.e.
+As we are interested mainly in the linear relationship between the alchemical free energy predictions and the experimentally-determined affinity values, plots should be depicted with the same range on both axes (i.e.
 $x=y$) with a 1:1 aspect ratio, with units for both experiment and simulation converted to be the same.
@@
 Furthermore, bounds should be depicted for the 1- and 2-kcal/mol confidence regions.
-These regions can serve as tools to communicate your model performance: any predictions inside the 1 kcal/mol region can be seen as highly reliable, any predictions inside the 2 kcal/mol region should be seen as somewhat reliable, and any predictions outside the confidence regions should be expected to be unreliable (and handled as outliers).
+These regions can serve as tools to communicate your model performance: any predictions inside the 1 kcal/mol region can be seen as highly reliable, any predictions inside the 2 kcal/mol region should be seen as somewhat reliable, and any predictions outside the confidence regions should be expected to be unreliable and handled as outliers.
 In a drug discovery context, this type of data depiction may suggest the reliability of alchemical FE predictions in the project, and can give an idea of how trustworthy predictions can be for synthesis ideas.
@@
 Alchemical free energy calculations have seen a vast increase in popularity both in academic research as well as pharmaceutical industry applications in structure based drug discovery~\cite{schindler2020largescale, sherborne2016collaborating, wagner2017computational}.
-Commercial products such as FEP+ and Flare, which provide a convenient user interface make the setup and use of these methods a lot easier~\cite{wang2015accurate, kuhn2020assessment}, but this convenience comes with less flexibility in terms of choice of simulation protocols.
+Commercial products such as FEP+ and Flare, which provide a convenient user interface make the setup and use of these methods much easier~\cite{wang2015accurate, kuhn2020assessment}, but this convenience comes with less flexibility in terms of choice of simulation protocols.
 It is also important to understand the current limitations of the methodology to recognise when automated workflow tools can be used effectively for a given protein target and when they are likely to fail still.
@@
 	\begin{itemize}
-	\item \href{http://pmx.mpibpc.mpg.de/instructions.html}{PMX}: Setup of perturbation maps, perturbed topologies and input coordinates for GROMACS simulations\url{https://github.com/deGrootLab/pmx}.
+	\item \href{http://pmx.mpibpc.mpg.de/instructions.html}{PMX}: Setup of perturbation maps, perturbed topologies and input coordinates for GROMACS simulations at \url{https://github.com/deGrootLab/pmx}.
 	\item \href{https://github.com/MobleyLab/Lomap}{Lomap/Lomap2} : Relative alchemical transformation graph planning for setting up perturbation networks~\cite{liu2013lead}.
@@
 	\item \href{https://github.com/qusers/qligfep}{QligFEP} offers robust and fast setup of FEP calculations for the software package Q~\cite{jespers2019qligfep}.
-	\item ProtoCaller, a setup tool for the automation of Gromacs free energy calculations. \href{https://github.com/protocaller/ProtoCaller}{ProtoCaller}~\cite{suruzhon2020protocaller}
-	\item \href{https://fesetup.readthedocs.io/en/latest/introduction.html}{FESetup} has been developed primarily to setup calculations in AMBER, GROMACS and SIRE~\cite{loeffler2015fesetup}
+	\item \href{https://github.com/protocaller/ProtoCaller}{ProtoCaller}, a setup tool for the automation of Gromacs free energy calculations~\cite{suruzhon2020protocaller}.
+	\item \href{https://fesetup.readthedocs.io/en/latest/introduction.html}{FESetup} has been developed primarily to setup calculations in AMBER, GROMACS and SIRE~\cite{loeffler2015fesetup}.
 	\end{itemize}
@@
 	\end{itemize}
-\item[] Generally, commercial software will offer more complete pipelines in which standalone analysis applications are not necessarily needed; free and open source packages often require manual analysis.
+\item[] Generally, commercial software will offer more complete pipelines in which standalone analysis applications are not necessarily needed; free and open source packages often require manual analysis but allow more flexibility and modification.
 \end{itemize}
@@
     \item What simulation protocol will provide the most evidence to answer my hypothesis?
-    \item Are the projected computational expense and runtime realistic for my goals?
+    \item Are the projected computational expense and runtime realistic for my scientific goals?
     \item Will my protocol be reproducible? 
@@
     \item Make sure you know why you have picked your (combination of) force field(s)
-    \item Minimize your system
-    \item Equilibrate your system with your choice of thermodynamic ensemble
+    \item Energy minimize your system
+    \item Equilibrate your system properly with your choice of thermodynamic ensemble
     \item Check the stability of your system and whether it behaves the way you believe it should
@@
  \item Check your ligands have the same, biologically correct binding pose
-        \item Make sure your \textlambda-scheduling is set appropriately
+        \item Make sure your \textlambda-scheduling is  appropriate
         \item Check if your ligands are discharging and decoupling correctly
@@
 \textbf{HBM} Contributed to Sec.~\ref{sec:plot_data} and Fig.~ \ref{fig:scatterplot_analysis} and helped edit the paper.\\
-\textbf{JDC}: Wrote Sec.~\ref{sec:decorrelating-samples} and~\ref{sec:automatic-equilibration-detection} discussed structure and design of the whole document, suggested Figs.~\ref{fig:fig_what_is_alchemy} and~\ref{fig:fig_sampling_scheme} \\
+\textbf{JDC}: Wrote Sec.~\ref{sec:decorrelating-samples} and~\ref{sec:automatic-equilibration-detection} discussed structure and design of the whole document, suggested Figs.~\ref{fig:fig_what_is_alchemy} and~\ref{fig:fig_sampling_scheme}.
+\\
 \textbf{DFH} Contributed to Sec.~\ref{sec:software} and helped edit the paper. \\

### 2021-03-16 paper/manuscript.tex
Last fixes.

@@
 This Best Practices guide focuses on providing a good starting point for new practitioners and a reference for experienced practitioners. 
- For this propose we provide a convenient checklist (Sec.~\ref{sec:checklist}) to help ensure all calculations comply with currently-understood best practices for alchemical simulation and analysis.
+ For this purpose we provide a convenient checklist (Sec.~\ref{sec:checklist}) to help ensure all calculations comply with currently-understood best practices for alchemical simulation and analysis.
 Where the best practices are currently not certain, we highlight areas where further research is needed to identify an unambiguous recommendation.
@@
 \item forcefield dependence for protein, ligand, ions, co-solvents, and co-factors.
-A number of different studies have looked at the influence of force fields and it is assumed the user has made an adequate choice for the system under study~\cite{loeffler2018reproducibility, vassetti2019assessment, lopes2015current}.
+A number of different studies have looked at the influence of force fields and it is assumed the user has made an appropriate choice for the system under study~\cite{loeffler2018reproducibility, vassetti2019assessment, lopes2015current}.
 \item non-equilibrium free energy calculations~\cite{gapsys2020large}
@@
 
-In many cases, the quantity of interest is the change in binding affinity between a compound $A$ and a related compound $B$ (e.g., by modifying one of the drug scaffold's substituents, see (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{F}), which, by using Eq.~\ref{eq:binding-free-energy-from-bound-unbound-probability-ratio} and \ref{eq:bound-unbound-probability-ratio} is given by
+In many cases, the quantity of interest is the change in binding affinity between a compound $A$ and a related compound $B$ (e.g., by modifying one of the drug scaffold's substituents, see (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{F})), which, by using Eq.~\ref{eq:binding-free-energy-from-bound-unbound-probability-ratio} and \ref{eq:bound-unbound-probability-ratio} is given by
 \begin{equation}\label{eq:delta-delta-G-physical}
@@
 In particular, the second line of Eq.~\ref{eq:delta-delta-G-alchemical} is a consequence of $\Delta G_{\mathrm{unbound}}$ being independent of the presence of the receptor in the simulation box as the definition of the unbound state assumes receptor and ligand to be at a sufficient distance for them to have no energetic interactions.
-Note that, when $A$ and $B$ have different number of atoms, the factors $ \ln \frac{Z(RB)}{Z(RA)}$ and $\ln \frac{Z(B)}{Z(A)}$ in   Eq.~\ref{eq:delta-delta-G-alchemical} appear both to have factors with units of volume in the logarithms, but these factors exactly cancel between the terms.
+Note that, when $A$ and $B$ have a different number of atoms, the factors $ \ln \frac{Z(RB)}{Z(RA)}$ and $\ln \frac{Z(B)}{Z(A)}$ in   Eq.~\ref{eq:delta-delta-G-alchemical} appear both to have factors with units of volume in the logarithms, but these factors exactly cancel between the terms.
 
@@
 Briefly, this is achieved by simulating a ``chimeric'' molecule composed of enough atoms to represent both $A$ and $B$.
-A subset of the energetic terms in $U(\vec{q};\vec{\lambda})$ is then modulated by $\vec{\lambda}$ so that at $\vec{\lambda}_A$, the atoms that form molecule $A$ are activated and those belonging exclusively to $B$ are non-interacting ``dummy atoms'', while the opposite occur at $\vec{\lambda}_B$ (see Sec.~\ref{sec:relative-fe-protocol} for details).
+A subset of the energetic terms in $U(\vec{q};\vec{\lambda})$ is then modulated by $\vec{\lambda}$ so that at $\vec{\lambda}_A$, the atoms that form molecule $A$ are activated and those belonging exclusively to $B$ are non-interacting ``dummy atoms'', while the opposite occurs at $\vec{\lambda}_B$ (see Sec.~\ref{sec:relative-fe-protocol} for details).
 
@@
 In the context of alchemical calculations, in which the thermodynamic states vary only in their value of $\vec{\lambda}$, these are also referred to as \emph{alchemical states}.
-The free energy of mutating $A$ to $B$ in any environment (e.g., binding site, solvent) can then be computed as
+The free energy of mutating $A$ to $B$ in any environment ($\Delta G_{\mathrm{env}}$ e.g., binding site, solvent) can then be computed as
 \begin{equation}\label{eq:delta-G-env}
-    \Delta G_{\mathrm{env}} = - k_BT \ln \frac{Z(\vec{\lambda}_B)}{Z(\vec{\lambda}_A)} = - k_BT \ln \frac{\int_{\Gamma_{\mathrm{env}}} \exp\left( u(\vec{q}; \vec{\lambda}_B) \right) \, d\vec{q}}{\int_{\Gamma_{\mathrm{env}}} \exp\left( u(\vec{q}; \vec{\lambda}_A) \right) \, d\vec{q}} \, .
-\end{equation}
+    \Delta G_{\mathrm{env}} = - k_BT \ln \frac{Z(\vec{\lambda}_B)}{Z(\vec{\lambda}_A)} = - k_BT \ln \frac{\int_{\Gamma_{\mathrm{env}}} \exp\left( u(\vec{q}; \vec{\lambda}_B) \right) \, d\vec{q}}{\int_{\Gamma_{\mathrm{env}}} \exp\left( u(\vec{q}; \vec{\lambda}_A) \right) \, d\vec{q}} \, ,
+\end{equation}
+over the configurational space of the environment ($\Gamma_{\mathrm{env}}$).
 While it is generally not feasible to compute the two partition functions $Z(\vec{\lambda})$, several estimators have been devised to robustly estimate the ratio of partition functions in Eq.~\ref{eq:delta-G-env} (see Sec.~\ref{subsec:estimators}) from a set of configurations usually collected with MD simulations from the thermodynamic states defined at $\vec{\lambda}_A$ and $\vec{\lambda}_B$ and intermediates thereof.
@@
 
-Equilibrium free energy calculations, our focus here, solve the problem of having poor overlap between the states of interest, by introducing multiple intermediate alchemical states at values $\vec{\lambda}_A = \vec{\lambda}_0, \vec{\lambda}_1, \cdots, \vec{\lambda}_K = \vec{\lambda}_B$ so that each pair of consecutive states $\vec{\lambda}_k, \vec{\lambda}_{k+1}$ share good overlap.
+Equilibrium free energy calculations, our focus here, solve the problem of having poor overlap between the states of interest by introducing multiple intermediate alchemical states at values $\vec{\lambda}_A = \vec{\lambda}_0, \vec{\lambda}_1, \cdots, \vec{\lambda}_K = \vec{\lambda}_B$ so that each pair of consecutive states $\vec{\lambda}_k, \vec{\lambda}_{k+1}$ share good overlap.
 Each intermediate state models a ligand that is neither $A$ nor $B$ but a interpolation of the two.
@@
 \begin{equation}
-    \Delta G_{\mathrm{env}} = k_BT \sum_{k=0}^{K-1} \Delta f(\vec{\lambda}_k, \vec{\lambda}_{k+1})
+    \Delta G_{\mathrm{env}} = k_BT \sum_{k=0}^{K-1} \Delta f(\vec{\lambda}_k, \vec{\lambda}_{k+1}),
 \end{equation}
@@
 \end{equation}
-While this strategy usually results in sampling thermodynamic states whose Boltzmann distributions are very similar, thus collecting information that is to some degree redundant, some estimators such as the Multistate Bennett acceptance ratio (MBAR)~\cite{shirts2008statisticallya} can exploit similarities between states to improve the precision of the estimates.
+While this strategy usually results in sampling thermodynamic states whose Boltzmann distributions are very similar, thus collecting information that is to some degree redundant, some estimators, such as the Multistate Bennett acceptance ratio (MBAR)~\cite{shirts2008statisticallya}, can exploit similarities between states to improve the precision of the estimates.
 This is achieved by using the configurations sampled at all alchemical states $\{\vec{\lambda}_k \}$ to compute the free energy difference $\Delta f(\vec{\lambda}_i, \vec{\lambda}_{j})$ between any pair of states $i,j$ (see Sec.~\ref{subsec:estimators}).
@@
 \label{subsec:expectation}
-We first note that the accuracy of any free energy calculation method will be dependent on the quality of the underlying force field.
+We first note that the accuracy of any free energy calculation method will depend on the quality of the underlying force field.
 Therefore, any description of the force field for the molecules under study must be carefully checked to be sufficiently accurate to experiment.
-In particular, one must make sure that if using automatically generated molecular descriptions from either one's own workflow or some other computational chemistry program, there are not obvious problems in these files either through errors in the workflow, or lack of chemical coverage in the data used to construct the molecular description.
+In particular, one must make sure that if using automatically generated molecular descriptions from either one's own workflow or some other computational chemistry program, there are no obvious problems in these files either through errors in the workflow, or lack of chemical coverage in the data used to construct the molecular description.
 Torsional parameters, in particular can be misassigned or improperly parameterized.
 
-Given reasonably accurate, current generation force fields correctly implemented, alchemical free energy calculations involving small molecules seem to achieve, in favorable cases, root mean square (RMS) errors relative to experiment around 1-2 kcal/mol depending on force field, system, and a variety of other factors such as simulation time, sampling method, and whether the calculations employed are absolute or relative.
+Alchemical free energy calculations involving small molecules seem to achieve, in favorable cases, root mean square (RMS) errors relative to experiment around 1-2 kcal/mol depending on force field, system, and a variety of other factors such as simulation time, sampling method, and whether the calculations employed are absolute or relative.
 A small selection of example datasets and case studies can be found in Sec.~\ref{sec:benchmark} at the end of this document.
@@
 \label{subsec:reproducible}
-We restrict our analysis here to repetitions of the same calculation performed with precisely the same force field, as different force fields used to describe the same molecule can lead in some cases to wide difference in free energies.
+We restrict our analysis here to repetitions of the same calculation performed with precisely the same force field, as different force fields used to describe the same molecule can lead to wide differences in free energies in some cases.
 Even within this restriction,
@@
 have favorable free energy changes. We are thus interested in how
-often that a molecule predicted to bind stronger actually turns out to
+often a molecule predicted to bind stronger actually turns out to
 bind stronger. In other words, we are interested in the conditional
@@
 microsecond or longer simulations for each of $N$ ligands will likely be
-prohibitive for most users with today's hardware. Alternatively, if key motions are fast and minimal (as is often assumed in practice), only much shorter simulations may be necessary.
+prohibitive for most users with today's hardware. On the other hand, if key motions are fast and minimal (as is often assumed in practice), much shorter simulations may be sufficient.
 
@@
 Biological assays are usually run at neutral pH but this is not always the case.
-For example, some enzymes exhibit pH-dependent activity and assays may thus be done in conditions other from neutral pH.
+For example, some enzymes exhibit pH-dependent activity and assays may thus be done in conditions other than neutral pH.
 Therefore, computational protein and ligand preparation protocols should reflect experimental pH.
@@
 
-In summary, the successful use of alchemical calculations not only in, but particularly for drug discovery requires working in the domain of applicability, using a high quality X-ray structure of the target bound to compounds in the series, and testing the approach retrospectively to ensure the system setup is well-behaved.
+In summary, the successful use of alchemical calculations, particularly for drug discovery, requires working in the domain of applicability, using a high quality X-ray structure of the target bound to compounds in the series, and testing the approach retrospectively to ensure the system setup is well-behaved.
 Always assess your confidence in the resulting predictions and communicate this when discussing with experimentalists.
@@
 From a drug discovery point of view, alchemical calculations are expanding their domain of applicability, and there are reports of success using homology models~\cite{cappel2016relative} and GPCRs~\cite{deflorian2020accurate,lenselink2016predicting} for instance, as well as enabling charge change and scaffold hopping~\cite{chen2018accurate, wang2017accurate}, but these systems are undoubtedly more difficult.
-In the meantime, the use cases are expanding to resistance prediction, selectivity prediction , solubility prediction – an exciting future for alchemical calculations~\cite{hauser2018predicting, albanese2020structure, mondal2019free}.
+In the meantime, use cases are expanding to resistance prediction, selectivity prediction , solubility prediction – an exciting future for alchemical calculations~\cite{hauser2018predicting, albanese2020structure, mondal2019free}.
 
@@
     \item Do I have one or multiple good receptor structures? (e.g. a good resolution X-ray crystal of the protein target)
-    \item Do I have information on one or all of the ligand binding sites? (e.g. a X-ray structure)
+    \item Do I have information on one or all of the ligand binding sites? (e.g. an X-ray structure)
     \item Should I include buried waters, or other small molecules that can be found in an X-ray structure?
@@
 Since the pKa of reference amino acid sidechain residues is known, but can vary in the protein environment, many different tools have emerged for predicting sidechain pKa in proteins, such as the H++ server, ProPKa, APBS, and Maestro ~\cite{anandakrishnan2012automating, sondergaard2011improved, jurrus2018improvements, 2020schrodinger}.
-Strongly acidic (Glu, Asp) or basic (Arg, Lys) sidechains can reliably predicted to be ionized, but care is still needed as the local environment can modify expected ionization states, such as  the catalytic Asp dyad in proteases.
+Strongly acidic (Glu, Asp) or basic (Arg, Lys) sidechains can reliably be predicted to be ionized, but care is still needed as the local environment can modify expected ionization states, such as  the catalytic Asp dyad in proteases.
 Histidine is notoriously more difficult to predict as its pKa suggests it ionizes closer to the experimental pH range.
-For ligands often the pKa needs to be determined, if not known experimentally.
+For ligands, often the pKa needs to be determined, if it is not known experimentally.
 There are many different available tools for this purpose, but common choices may be propKa~\cite{olsson2011propka3,sondergaard2011improved}, Chemicalize (\url{https://chemicalize.com/welcome}), or Maestro~\cite{2020schrodinger}.
 Still, accurate pKa prediction for small molecules remains a challenging problem, even with dedicated tools~\cite{isik2018pka}.
-While often it can be assumed that the protonation state of a ligand and protein will remain the same as a ligand binds, some care needs to be taken with systems where the protonation state may change upon binding~\cite{onufriev2013protonation}.
-BACE, for example, famously undergoes a protonation state change on ligand binding.
+While often it can be assumed that the protonation state of a ligand and protein will remain the same as the ligand binds, some care needs to be taken with systems where the protonation state may change upon binding~\cite{onufriev2013protonation}.
+BACE~\cite{kim2015conformational}, for example, famously undergoes a protonation state change on ligand binding.
 
@@
 In cases where the end states have high topological similarity and high structural overlap this is relatively straightforward and typically handled by use of MCSS calculations.
-In situations where the end state topologies differ significantly, or there is relatively little spatial overlap between the two end states some user intervention may be necessary to produce a satisfactory input topology.
+In situations where the end state topologies differ significantly, or where there is relatively little spatial overlap between the two end states, some user intervention may be necessary to produce a satisfactory input topology.
 
@@
 In this case a docking program can generate initial structures.
-Different commercial and none commercial tools are available, such as rDock~\cite{ruiz-carmona2014rdock}, Autodock Vina~\cite{trott2010autodock}, Glide~\cite{friesner2004glide}, or Flare, to name a few~\cite{kuhn2020assessment}.
+Different commercial and non-commercial tools are available, such as rDock~\cite{ruiz-carmona2014rdock}, Autodock Vina~\cite{trott2010autodock}, Glide~\cite{friesner2004glide}, or Flare, to name a few~\cite{kuhn2020assessment}.
 
@@
 Preliminary considerations necessary for using free energy methods to compute partition coefficients are generally more straight forward.
-For example, a 3D minimised structure of the solute can be generated with a simple tool such as openBabel and solvated to prepare the input to compute a free energy of hydration~\cite{oboyle2011open}.
-However, in these cases a careful choice of forcefield, as well as water models or organic solvents is essential.
+For example, a 3D minimised structure of a solute can be generated with a simple tool such as openBabel and solvated to prepare the input to compute a free energy of hydration~\cite{oboyle2011open}.
+However, in these cases a careful choice of forcefield for the organic solvent model, as well as water model is essential.
 See for example~\cite{bosisio2016blinded,rustenburg2016measuring} for a good discussion of these choices.
@@
 To use binding as a concrete example, in absolute binding free energy calculations, we compute the binding free energy of a ligand to an individual receptor relative to a standard reference concentration.
-In contrast, in relative binding free energy calculations, we compare the binding free energy of two related inhibitors to determine the potency difference.
+In contrast, in relative binding free energy calculations, we compare the binding free energy of two related ligands to determine the potency difference.
 
@@
 
-In single topology calculations, the overall transformation is set up to involve as few additional atoms as possible, so benzene would be typically changed into benzyl alcohol by changing one of the hydrogens into a carbon.
+In single topology calculations, the overall transformation is set up to involve as few additional atoms as possible, so benzene would be typically changed into benzyl alcohol by first changing one of the hydrogens into a carbon.
 The site of this transformation will also be the future home of two additional hydrogen atoms bound to the new carbon, so these must initially be present as non-interacting atoms called ``dummy atoms'', which retain their bonded interactions but do not interact with the rest of the system.
@@
 In contrast, in a dual topology alchemical free energy calculation, \emph{no atoms are allowed to change type}~\cite{boresch1999rolea, shirts2012best}.
-This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxy methyl group, then passing through an intermediate state where some atoms are partially interacting---particularly, those atoms which are becoming dummy atoms or ceasing to be dummy atoms~\cite{mobley2014blind}.
+This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxymethyl group, then passing through an intermediate state where some atoms are partially interacting---particularly, those atoms which are becoming dummy atoms or ceasing to be dummy atoms~\cite{mobley2014blind}.
 The transformation finally culminates in a state where benzyl alcohol is present along with the additional dummy atom which was previously a corresponding hydrogen of the benzene.
@@
 Some additional terms have also been employed to talk about these different intermediate pathways .
-Particularly, some studies refer to a ``hybrid'' topology approach to free energy calculations~\cite{gapsys2015pmx, gapsys2016accurate, gapsys2020large}, though this term may not have yet achieved widespread use.
+Particularly, some studies refer to a ``hybrid'' topology approach to free energy calculations~\cite{gapsys2015pmx, gapsys2016accurate, gapsys2020large}, though this term may not yet have achieved widespread use.
 In this case, "hybrid" seems to indicate that the set up of these free energy calculations involves a hybrid of the two molecules, and much of what is done in these studies uses a single topology approach~\cite{gapsys2020large}.
@@
 
-Software packages vary in their use single or dual topology approaches; for example, AMBER TI uses a dual topology approach, while BioSimSpace uses a single topology approach.
+Software packages vary in their use of single or dual topology approaches; for example, AMBER TI uses a dual topology approach, while BioSimSpace uses a single topology approach.
 Please make sure to check what approach is used with your software package of choice, or whether it supports your choice of approach (GROMACS and GROMOS, for example, support both).
@@
 It is possible that changes in binding mode could actually require a different choice of mapping, so in some cases mappings may need to be planned differently depending on 3D positioning of atoms in space.
-It is worth visual inspection prior to simulation to ensure that the mapping criteria corresponds to the expected binding mode.
-If the mapping protocol returns simulations that correspond to different binding modes of a ligand within a perturbation map, which can cause large hysteresis.
+Visual inspection prior to simulation is recommended to ensure that the mapping criteria correspond to the expected binding mode.
+If the mapping protocol returns simulations that correspond to different binding modes of a ligand within a perturbation map, this can cause large hysteresis.
 
@@
 \paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
-Some approaches have attempted to address this~\cite{clark2019relative} but a general solution is not yet in mainstream use, though FEP+ implements one solution.
+Some approaches have attempted to address this~\cite{clark2019relative} but a general solution is not yet in mainstream use. Still, FEP+ implements one solution.
 
@@
 In some cases, during the lead optimization stage, or for very large datasets that would benefit from rougher initial free energy ranking, or in cases where perturbations would be rather large, a star shaped network as seen in Fig.~\ref{fig:fig_types_of_networks} \textbf{A} is used.
-However, adding redundancy into the network which means that a better error analysis can be carried out, by looking at cycle closure errors as discussed in sec.~\ref{sec:are-they-good}, with an example given in Fig.~\ref{fig:fig_types_of_networks}\textbf{B}.
+However, adding redundancy into the network means that a better error analysis can be carried out by looking at cycle closure errors as discussed in sec.~\ref{sec:are-they-good}, with an example given in Fig.~\ref{fig:fig_types_of_networks}\textbf{B}.
 
@@
 Individual software programs and settings can handle such issues.
-For example, bonds can be transformed in both GROMACS and GROMOS, because contributions of LINCS (GROMACS) and SHAKE (GROMOS) contrained bonds are added to the $\frac{dH}{d\lambda}$ term~\cite{pearlman1993determining, straatsma1992holonomica, pearlman1991overlooked, gunsteren1989computer}.
+For example, bonds can be transformed in both GROMACS and GROMOS, because contributions of LINCS (GROMACS) and SHAKE (GROMOS) constrained bonds are added to the $\frac{dH}{d\lambda}$ term~\cite{pearlman1993determining, straatsma1992holonomica, pearlman1991overlooked, gunsteren1989computer}.
 However, the constraints are not taken into account when calculating energy differences to other intermediate states as the effect is entropic, not energetic.
@@
 This  restriction may be undesirable when using Hamiltonian $\lambda$ exchange or expanded ensemble techniques where allowing the ligand to exchange binding modes when it is non-interacting could provide sampling benefits~\cite{wang2013identifying}.
-More specifically, flat-bottom restraints might allow a ligand to explore multiple binding sites, harmonic restraints multiple binding modes within a site, while Boresch restraints only allow a single binding mode within a single site.
+More specifically, flat-bottom restraints might allow a ligand to explore multiple binding sites, and harmonic restraints allow exploration of multiple binding modes within a site, while Boresch restraints only allow a single binding mode within a single site.
 See additional discussion of the possibility of multiple binding modes in Sec.~\ref{sec:multiple_binding_modes} below.
@@
 A safe but potentially more computationally expensive approach is to perform the transformations in sequence; first, turning off all electrostatics for atoms that must be removed, inserting and removing Lennard-Jones sites (both the insertion and removal can be done simultaneously), and then turning electrostatics for the introduced particles on.
-Again, If there are no removals or introduction to atomic sites, then it is reasonable to change the interactions in the first and third steps  linearly.
+Again, if there are no removals or additions to atomic sites, then it is reasonable to change the interactions in the first and third steps  linearly.
 
 Other issues, such as whether absolute calculations should retain or remove intramolecular non-bonded interactions
-through either annihilation~\cite{hermans1997inclusiona, mann2000modelinga, boresch2003absolutea, wang2006absolutea, mobley2006use} or decoupling~\cite{fujitani2005directa, mobley2006use} must be considered.
+through either annihilation~\cite{hermans1997inclusiona, mann2000modelinga, boresch2003absolutea, wang2006absolutea, mobley2006use} or decoupling~\cite{fujitani2005directa, mobley2006use}, must be considered.
 Reasonable efficiency can be often obtained with either choice even if some are somewhat better or worse than others, and there is no consensus on which is better in most given situations.
 Our recommendation is to leave the intramolecular interactions on during the transformation for simplicity if there are no other known issues with this approach.
-The key feature of the simulation to watch out for is whether the total potential energy, and therefore the intermediate ensembles sampled, change smoothly from beginning to end.
+The key feature of the simulation to watch out for is whether the total potential energy, and therefore the intermediate ensembles sampled, changes smoothly from beginning to end.
 Problems of discontinuous changes of the potential energy can be diagnosed by noticing lack of configuration space overlap between different simulations (see Sec.~\ref{sec:are-they-good}).
@@
 However, it can often fail whenever the simulation of this intermediate state ends up trapped in configurations characteristic of only one end state.
-Thus, successful use of EDS can require system-dependent tuning making it difficult to implement in an automated and reliable way, but when successful it can be very efficient.
+Thus, successful use of EDS can require system-dependent tuning, making it difficult to implement in an automated and reliable way.
+However, when successful, it can be very efficient.
 
@@
 Additionally, some paths simply may not converge due to issues noted above such as those encountered without soft core potentials.
-However, the recommendations above are reasonable, reliable, and are likely not be that much less efficient than potentially more optimal choices~\cite{naden2014linear,naden2015linear,pham2012optimal}, as the real problems with the efficiency of calculating free energies are lack of sampling of slow conformational modes, rather than the lack of efficiency of the transformations.
+However, the recommendations above are reasonable, reliable, and are likely not that much less efficient than potentially more optimal choices~\cite{naden2014linear,naden2015linear,pham2012optimal}, as the real problems with the efficiency of calculating free energies are lack of sampling of slow conformational modes, rather than the lack of efficiency of the transformations.
 
 It is however important to note that different packages also differ in how they handle implementation of alchemical transformations, making it difficult to give rules of thumb concerning specific efficient transformations which work equally well across simulation packages.
-Thus we are hesitant to recommend best practices within specific software packages at this point in time, although any good transformation pathway will conform to the guidelines we have outlined above.
+Thus, we are hesitant to recommend best practices within specific software packages at this point in time, although any good transformation pathway will conform to the guidelines we have outlined above.
 
@@
 The condition of minimizing the statistical uncertainty of different free energy estimators below a sufficient threshold should be one metric monitored over the simulation.
-This can be done through the uncertainty estimator built into certain analysis tools such as MBAR, or can be done through more general statistical tools like bootstrap sampling.
+This can be done through the uncertainty estimator built into certain analysis tools such as MBAR, or through more general statistical tools like bootstrap sampling.
 It should be noted however, that uncertainty estimates have the same limitations as other metrics of convergence, as they are only an uncertainty based on the phase space sampled so far in a simulation, and cannot account for states not sampled, and it is worth considering that they will be an underestimation of the true uncertainty.
@@
 Results obtained from calculations with convergence issues should be checked for errors or run for longer before any confidence should be placed in conclusions drawn from their analysis.
-For example, in relative calculations that share similar binding modes and do not induce large conformational changes when in complex with protein, the need to sample exhaustively to converge estimates in free energy differences is often not necessary due to the locality of sampling changes in the molecular topology and shared phase space of the core atoms.
-However, even subtly induced changes in protein binding configuration will require more sampling or cause local convergence to a free energy estimate that has high error.
+For example, in relative calculations ligands that share similar binding modes and do not induce large conformational changes when in complex with protein, the need to sample exhaustively to converge estimates in free energy differences is often minimal due to the locality of sampling changes in the molecular topology and shared phase space of the core atoms.
+However, even subtle induced changes in protein binding configuration will require more sampling or cause local convergence to a free energy estimate that has high error.
 The confidence a user should have in a free energy estimate is significantly improved when both the uncertainty of the free energy estimate is low, and when other observables have reached a convergence.
@@
     \caption{{\bf Potential of mean force with respect to $\vec{\lambda}$ for TI and MBAR}
-    The estimated PMF for a bound calculation of a Tyk2 ligand pair of the Wang et al.~\cite{wang2015accurate} with respect to $\vec{\lambda}$ estimated from TI and MBAR and showing agreement within errorbars.
+    The estimated PMF for a bound calculation of a Tyk2 ligand pair of Wang et al.~\cite{wang2015accurate} with respect to $\vec{\lambda}$ estimated from TI and MBAR and showing agreement within errorbars.
     }
@@
 However, timescales for binding mode interconversion are usually slow compared to MD/free energy timescales, meaning that simulations started from different potential binding modes are likely to yield disparate computed binding free energies~\cite{mobley2006use, palma2012computation, mobley2012perspective, gill2018binding}.
-And docking techniques are good at identifying sterically reasonable potential binding modes, but still perform relatively poorly at identifying a single dominant binding mode \emph{a priori}.
+Moreover, docking techniques are good at identifying sterically reasonable potential binding modes, but still perform relatively poorly at identifying a single dominant binding mode \emph{a priori}.
 
@@
 For simulations that are simply not long enough to contain a large number of samples from true equilibrium either because they are very short or contain slow processes, this procedure cannot completely remove the bias.
-In such cases, this approach simply selects the final portion of the the simulation, which may be contained in a single substate of configurationaal space, and may itself lead to biased estimates. 
+In such cases, this approach simply selects the final portion of the the simulation, which may be contained in a single substate of configurational space, and may itself lead to biased estimates. 
 This situation can be detected if the equilibration boundary $t_0$ is a significant fraction of the total simulation length $T$, with a good rule of thumb being that $T \gtrsim 20 t_0$.
@@
 
-If independent simulations are used, the alchemical state $\vec{\lambda}$ may have a significant impact on the correlation time, and these simulations should be subsampled independently using a separate estimate of the statistical inefficinecy $g$ for each alchemical state.
+If independent simulations are used, the alchemical state $\vec{\lambda}$ may have a significant impact on the correlation time, and these simulations should be subsampled independently using a separate estimate of the statistical inefficiency $g$ for each alchemical state.
 If coupled simulations are used (such as a Hamiltonian replica exchange simulation), the replicas should undergo equivalent random walks in alchemical space, and the replicas can be subsampled with the same $g$ to generate an equal number of uncorrelated samples at each alchemical state.
@@
 \begin{eqnarray}
-\Delta f_{01} & = & \ln \expect{e^{-(u_1(\vec{q}) - u_0(\vec{q}))}}_0 =  -\ln \expect{e^{-\Delta u(\vec{q})} }_0
+\Delta f_{01} & = & -\ln \expect{e^{-(u_1(\vec{q}) - u_0(\vec{q}))}}_0 =  -\ln \expect{e^{-\Delta u(\vec{q})} }_0
 \end{eqnarray}\label{eqn.zwanzig}
@@
 
-MBAR is provably the lowest variance asymptotically unbiased estimator of the free energy given the energies of the samples~\cite{tan2004likelihood}, which means that BAR is also the lowest variance estimator for the free energy difference between only 2 states, as it is mathematically exactly the same as MBAR in this case.
+MBAR is probably the lowest variance asymptotically unbiased estimator of the free energy given the energies of the samples~\cite{tan2004likelihood}, which means that BAR is also the lowest variance estimator for the free energy difference between only 2 states, as it is mathematically exactly the same as MBAR in this case.
 MBAR also provides an uncertainty estimate, derived from standard error propagation methods for implicit functions, which has been shown to be highly accurate as long as there are sufficient samples at each state~\cite{paliwal2011benchmark}.
@@
 \item \textbf{Convergence of free energy} The free energy should clearly converge as a function of simulation time (Fig. ~\ref{fig:convergence_forward_reverse}).
-\item If using non-equilibrium methods, is the result independent of the speed at which the non-equilibrium change is performed?
+\item If using non-equilibrium methods, \textbf{is the result independent of the speed at which the non-equilibrium change is performed}?
 Non-equilibrium methods are in theory independent of the switching time in the limit of good sampling unless the switching time is simply too short.
@@
 The consequence of the two possible signs for relative free energies is that correlation statistics (such as Pearson's R and Kendall \texttau{}) can be skewed depending on which sign is analysed.
-The issue of this inconsistency can be circumvented by either plotting all datapoints within a consistent quadrant~\cite{perez-benito2019predicting}, or by avoiding the use of correlation statistics for assessment of relative free energy calculations and instead measuring accuracy using RMSE and MUE which are unaffected by choice of sign.
+The issue of this inconsistency can be circumvented by either plotting all datapoints within a consistent quadrant~\cite{perez-benito2019predicting}, or by avoiding the use of correlation statistics for assessment of relative free energy calculations and instead measuring accuracy using RMSE and MUE, which are unaffected by choice of sign.
 
@@
 As essentially all alchemical free energy prediction schemes are regression problems, the preferred type of plot is a scatter plot (see Fig.~\ref{fig:scatterplot_analysis}).
-Most alchemical free energy projects will look at 10-50 ligands; any study with \textless10 ligands is more suitable for bar plots (with inclusion of error bars), and will unlikely provide meaningful statistics.
-Any study with \textgreater50 ligands often contains multiple protein targets to which alchemical free energies may perform better on some targets than others.
+Most alchemical free energy projects will look at 10-50 ligands.
+Any study with \textless10 ligands is more suitable for bar plots (with inclusion of error bars), and is unlikely to provide meaningful statistics.
+Any study with \textgreater50 ligands typically contains multiple protein targets to which alchemical free energies may perform better on some targets than others.
 Because of this, it is bad practice to place multiple datasets on the same plot as this can suggest high model accuracy even though the individual models perform less well~\cite{walterthoughts}.
@@
     \end{itemize}
-    \item[]All the above tools also provide a convenient setup and analysis suite and are really a one in all product. 
+    \item[]All the above tools also provide a convenient setup and analysis suite and are really a one in all product. \newline
 \item [] \textbf{Simulation software: Free/low cost academic and Commercial}
@@
 	\begin{itemize}
-	\item \href{https://www.plumed.org/}{PLUMED} is an open source tool which enables the usage of a variety of MD engines.
+	\item \href{https://www.plumed.org/}{PLUMED} is a tool which enables the usage of a variety of MD engines.
 It is designed as a plugin for MD packages such that it analyzes the trajectory on the fly.
 It also offers a VMD based plugin for the computation of collective variables~\cite{bonomi2019promoting}.
-	\item \href{https://biosimspace.org/}{BioSimSpace} is a free, open source, multiscale molecular simulation framework, written to allow computational modellers to quickly prototype and develop new algorithms for molecular simulation and molecular design~\cite{hedges2019biosimspace}.
+	\item \href{https://biosimspace.org/}{BioSimSpace} is a multiscale molecular simulation framework, written to allow computational modellers to quickly prototype and develop new algorithms for molecular simulation and molecular design~\cite{hedges2019biosimspace}.
 	\item \href{https://siremol.org/}{Sire} is a multiscale, molecular simulation framework that provides several applications, including SOMD, an MD/MC code for performing FEP calculations via an interface to OpenMM.
@@
     \item Does my system contain any structures that require custom parameters?
-    \item What simulation protocol will provide the most evidence to answer my hypothesis?
+    \item What simulation protocol will provide the most evidence to verify my hypothesis?
     \item Are the projected computational expense and runtime realistic for my scientific goals?
@@
 %%%%%%%%%%%%%%%
-Julia E. Rice participated in the original discussion of the document at the Best Practices in Molecular Simulation Workshop Hosted by at NIST, Gaithersburg, MD, August 24th-25th, 2017
+Julia E. Rice participated in the original discussion of the document at the Best Practices in Molecular Simulation Workshop Hosted by at NIST, Gaithersburg, MD, August 24th-25th, 2017.
+Marieke Schor proofread the manuscript.
 

### 2021-03-17 paper/manuscript.tex

@@
 After an alchemical calculation is performed, which generally involves multiple simulations at a variety of alchemical states, the data must be analyzed to compute an estimate of the free energy for the transformation of interest.
- Early work used simple but statistically suboptimal estimators for this: free energy perturbation (FEP) used a simple (but highly biased) estimator based on the Zwanzig relation~\cite{zwanzig1954hightemperature} or numerical quadrature via thermodynamic integration (TI), for which the theory dates back the better part of a century but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamicsa, merz1989free}.
+ Early work used simple but statistically suboptimal estimators for this: free energy perturbation (FEP) used a simple (but highly biased) estimator based on the Zwanzig relation~\cite{zwanzig1954hightemperature} or numerical quadrature via thermodynamic integration (TI), for which the theory dates back the better part of a century but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamics, merz1989free}.
 %
@@
 Depending on system size and simulation settings, common molecular dynamics software packages can reach a few hundreds of ns/day using currently available high-end GPUs~\cite{eastman2017openmm,kutzner2019more}, making these type of calculations unappealing and irrelevant on a pharmaceutical drug discovery timescale.
-Other methods compute the free energy of binding by building potential of mean force profiles along a reaction coordinate~\cite{woo2005calculationa,velez-vega2013overcoming,limongelli2013funnel,heinzelmann2017attachpullrelease}, but these methods require prior knowledge of a high-probability binding pathway, which is not easily available, especially in the prospective scenarios typical of the drug development process.
+Other methods compute the free energy of binding by building potential of mean force profiles along a reaction coordinate~\cite{woo2005calculation,velez-vega2013overcoming,limongelli2013funnel,heinzelmann2017attachpullrelease}, but these methods require prior knowledge of a high-probability binding pathway, which is not easily available, especially in the prospective scenarios typical of the drug development process.
 
@@
 \paragraph{Why do alchemical calculations need unphysical intermediate states?}
-While it is theoretically possible to estimate the ratio of partition functions from samples collected only at states $\vec{\lambda}_A$ and $\vec{\lambda}_B$, the efficiency of the free energy estimators rapidly decreases as the phase-space overlap between the two states also decreases~\cite{wu2005phasespaceb,wu2005phasespacec}.
+While it is theoretically possible to estimate the ratio of partition functions from samples collected only at states $\vec{\lambda}_A$ and $\vec{\lambda}_B$, the efficiency of the free energy estimators rapidly decreases as the phase-space overlap between the two states also decreases~\cite{wu2005phasespace, wu2005phasespacea}.
 Roughly, the phase-space overlap between two thermodynamic states measures the degree to which high-probability configurations (i.e., those with very negative potential energy) in one state are also high-probability configurations in the other state (see Sec.~\ref{sec:are-they-good} and Fig.~\ref{fig:fig_what_is_lambda}).
@@
 Each intermediate state models a ligand that is neither $A$ nor $B$ but a interpolation of the two.
-Many estimators (e.g., exponential reweighting (EXP)~\cite{zwanzig1954hightemperature} and Bennett's acceptance ratio (BAR)~\cite{bennett1976efficient,shirts2003equilibriuma}) can then be used to compute the free energy as
+Many estimators (e.g., exponential reweighting (EXP)~\cite{zwanzig1954hightemperature} and Bennett's acceptance ratio (BAR)~\cite{bennett1976efficient,shirts2003equilibrium}) can then be used to compute the free energy as
 \begin{equation}
@@
 calculations, i.e., on average, the free energy change computed by
-free energy calculations agrees with the actual free energy change. Additional analysis of this type is presented in Brown et al.~\cite{shirts2010}
+free energy calculations agrees with the actual free energy change. Additional analysis of this type is presented in Brown et al.~\cite{brown2010freeenergy}
 
@@
 In this regard it is important to consider the issue of selection bias upfront.
-It is tempting to only synthesize the compounds predicted to be most active, thus a narrow range of calculated activity is tested that imposes limits on the statistical assessment of performance, ideally example molecules from across the range of predicted activity can be assessed or corrections can be applied based on previous recommendations~\cite{abel2017critical}.
+It is tempting to only synthesize the compounds predicted to be most active, thus a narrow range of calculated activity is tested that imposes limits on the statistical assessment of performance, ideally example molecules from across the range of predicted activity can be assessed or corrections can be applied based on previous recommendations~\cite{robert2017critical}.
 For a more detailed discussion on checking the robustness of your alchemical free energy calculation see also Sec.~\ref{sec:are-they-good}.
@@
 
-In contrast, in a dual topology alchemical free energy calculation, \emph{no atoms are allowed to change type}~\cite{boresch1999rolea, shirts2012best}.
+In contrast, in a dual topology alchemical free energy calculation, \emph{no atoms are allowed to change type}~\cite{boresch1999role, shirts2012best}.
 This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxymethyl group, then passing through an intermediate state where some atoms are partially interacting---particularly, those atoms which are becoming dummy atoms or ceasing to be dummy atoms~\cite{mobley2014blind}.
@@
 Individual software programs and settings can handle such issues.
-For example, bonds can be transformed in both GROMACS and GROMOS, because contributions of LINCS (GROMACS) and SHAKE (GROMOS) constrained bonds are added to the $\frac{dH}{d\lambda}$ term~\cite{pearlman1993determining, straatsma1992holonomica, pearlman1991overlooked, gunsteren1989computer}.
+For example, bonds can be transformed in both GROMACS and GROMOS, because contributions of LINCS (GROMACS) and SHAKE (GROMOS) constrained bonds are added to the $\frac{dH}{d\lambda}$ term~\cite{pearlman1993determining, straatsma1992holonomic, pearlman1991overlooked, gunsteren1989computer}.
 However, the constraints are not taken into account when calculating energy differences to other intermediate states as the effect is entropic, not energetic.
@@
 
-Alternatively, a set of restraints proposed by Boresch have also commonly been employed, where all six rigid-body degrees of freedom governing the orientation of the ligand relative to the receptor are restrained~\cite{boresch2003absolutea, leitgeb2005alchemicala}.
-Further restraints, such as on the overall ligand RMSD have also been used~\cite{woo2005calculationa}.
+Alternatively, a set of restraints proposed by Boresch have also commonly been employed, where all six rigid-body degrees of freedom governing the orientation of the ligand relative to the receptor are restrained~\cite{boresch2003absolute, leitgeb2005alchemical}.
+Further restraints, such as on the overall ligand RMSD have also been used~\cite{woo2005calculation}.
 
@@
 Alternatively, one can adopt a protocol that removes the bound-state restraint in the non-interacting state.
-Finally, even for tight binders, dissociation events can be enhanced by methods such as Hamiltonian replica exchange~\cite{sugita2000multidimensionala,chodera2011replica,wang2013identifying} and expanded ensemble~\cite{lyubartsev1992newa,li2007simulated}, especially in absolute free energy calculations using harmonic or flat-bottom restraints.
+Finally, even for tight binders, dissociation events can be enhanced by methods such as Hamiltonian replica exchange~\cite{sugita2000multidimensional,chodera2011replica,wang2013identifying} and expanded ensemble~\cite{lyubartsev1992new,li2007simulated}, especially in absolute free energy calculations using harmonic or flat-bottom restraints.
 In the latter case, dissociations can be averted simply by increasing the spring constant and/or reducing the radius of the restraint potential to prevent the exploration of ligand conformations outside the binding site in the decoupled state (bottom-right state in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}) that could be propagated to the bound state.
@@
 Other issues, such as whether absolute calculations should retain or remove intramolecular non-bonded interactions
-through either annihilation~\cite{hermans1997inclusiona, mann2000modelinga, boresch2003absolutea, wang2006absolutea, mobley2006use} or decoupling~\cite{fujitani2005directa, mobley2006use}, must be considered.
+through either annihilation~\cite{hermans1997inclusion, mann2000modeling, boresch2003absolute, wang2006absolute, mobley2006use} or decoupling~\cite{fujitani2005direct, mobley2006use}, must be considered.
 Reasonable efficiency can be often obtained with either choice even if some are somewhat better or worse than others, and there is no consensus on which is better in most given situations.
@@
 A key additional consideration in choosing the alchemical pathway is the choice of spacing of intermediate states.
-The spacing depends to some extent on the choice of analysis method, though states should essentially be spaced equidistant in the relevant thermodynamic length~\cite{crooks2007measuringa, sivak2012thermodynamic}.
+The spacing depends to some extent on the choice of analysis method, though states should essentially be spaced equidistant in the relevant thermodynamic length~\cite{crooks2007measuring, sivak2012thermodynamic}.
 For BAR/MBAR techniques this means that states should be spaced so that the statistical uncertainties between neighboring states be approximately equal~\cite{pham2012optimal, shenfeld2009minimizing}, where "approximately" is roughly within 30-50\% in magnitude.
@@
 Some approaches have attempted to find alternative pathways to improve efficiency or find paths of low thermodynamic length~\cite{naden2014linear,naden2015linear,pham2012optimal}.
-For example, the enveloping distribution sampling (EDS) approach, and its multiple-replica and accelerated variant, works to improve efficiency by creating a single artificial intermediate state which simultaneously samples all end state phase spaces~\cite{perthold2018accelerateda,sidler2017efficientb, christ2007envelopinga}.
+For example, the enveloping distribution sampling (EDS) approach, and its multiple-replica and accelerated variant, works to improve efficiency by creating a single artificial intermediate state which simultaneously samples all end state phase spaces~\cite{perthold2018accelerated,sidler2017efficient, christ2007enveloping}.
 When this can be done, it provides an extremely efficient way to calculate the relative free energy difference between multiple ligands from a single simulation.
@@
 With the Metropolis-Hastings acceptance criterion for exchanges, the generated ensemble of all replicas still samples from the Boltzmann distribution for each replica.
-This approach has been used in many different contexts for molecular simulations~\cite{sugita2000multidimensionala,sugita1999replicaexchangea, woods2003developmenta, jiang2010free}.
+This approach has been used in many different contexts for molecular simulations~\cite{sugita2000multidimensional,sugita1999replicaexchange, woods2003development, jiang2010free}.
 The basic idea of the replica exchange scheme is shown in Fig.~\ref{fig:fig_sampling_scheme}\textbf{B}.
-It is supported in various software packages that provide alchemical implementations, such as GROMACS~\cite{aldeghi2015accurate}, GROMOS~\cite{hritz2008hamiltoniana,hritz2007optimizationa}, FEP+~\cite{wang2015accurate}, and NAMD~\cite{jiang2019computing}.
-
-A third approach borrows ideas from simulated tempering~\cite{marinari1992simulateda}.
+It is supported in various software packages that provide alchemical implementations, such as GROMACS~\cite{aldeghi2015accurate}, GROMOS~\cite{hritz2008hamiltonian,hritz2007optimization}, FEP+~\cite{wang2015accurate}, and NAMD~\cite{jiang2019computing}.
+
+A third approach borrows ideas from simulated tempering~\cite{marinari1992simulated}.
 In this scheme a single replica rapidly explores all of $\vec{\lambda}$ space by working out optimal weights that allow switching between different intermediate $\vec{\lambda}$ values, as seen in Fig.~\ref{fig:fig_sampling_scheme}\textbf{C} .
-This approach is also referred to as self-adjusted mixture sampling~\cite{lyubartsev1992newa, li2007simulated, tan2017optimally} and while promising, has so far only been supported in OpenMM Tools~\cite{andrearizzi2019choderalab} and GROMACS.
+This approach is also referred to as self-adjusted mixture sampling~\cite{lyubartsev1992new, li2007simulated, tan2017optimally} and while promising, has so far only been supported in OpenMM Tools~\cite{andrearizzi2019choderalab} and GROMACS.
 Although this approach allows multiple states to be simulated in a single simulation, the weights do not always converge to the proper equilibrium distribution, and care must be taken that the final results are converged.
@@
 We note that this method is even more general, in that the instantaneous work to change the potential energy function from $u_0$ to $u_1$ can be replaced by the non-reversible work $W$ to make the same change beginning from the same equilibrium conditions at either end state~\cite{jarzynski1997nonequilibrium,jarzynski1998equilibrium,crooks2000pathensemble}, allowing for non-equilibrium free energy calculations, an alternate approach.
-We do not detail non-equilibrium transformations here, and refer the reader to more advanced treatments~\cite{maragakis2008bayesian,oberhofer2005biased,procacci2015unbiased,shirts2003equilibriuma,ytreberg2004singleensemble, gapsys2020large}, as our focus here is on equilibrium free energy techniques.
+We do not detail non-equilibrium transformations here, and refer the reader to more advanced treatments~\cite{maragakis2008bayesian,oberhofer2005biased,procacci2015unbiased,shirts2003equilibrium,ytreberg2004singleensemble, gapsys2020large}, as our focus here is on equilibrium free energy techniques.
 
@@
 One can show that if the standard deviation of the difference $\Delta u(\vec{q}) = u_1(\vec{q})-u_2(\vec{q})$ over all sampled $\vec{q}$ is large, which in this case, means only several times $k_BT$, then very few samples contribute to the average, and the answer will be both biased and extremely noisy~\cite{lelievre2010free}.
-Essentially, the method is dominated by contributions of rare snapshots~\cite{jarzynski2006rare, wu2005phasespaceb, wu2005phasespacec}.
+Essentially, the method is dominated by contributions of rare snapshots~\cite{jarzynski2006rare, wu2005phasespace, wu2005phasespacea}.
 
@@
 where $n_i$ and $n_j$ are the number of samples from each state.
-More recent derivations show that this formula is the maximum likelihood estimate of the free energy difference given sets of samples from the two states~\cite{shirts2003equilibriuma}.
+More recent derivations show that this formula is the maximum likelihood estimate of the free energy difference given sets of samples from the two states~\cite{shirts2003equilibrium}.
 
 Many studies have demonstrated both the theoretical and practical superiority of BAR over EXP in molecular
-simulations~\cite{shirts2005comparison,lu2003appropriate}, and BAR converges to EXP in the limit that all samples are from a single state~\cite{bennett1976efficient,bennett1976efficient,shirts2003equilibriuma}.
+simulations~\cite{shirts2005comparison,lu2003appropriate}, and BAR converges to EXP in the limit that all samples are from a single state~\cite{bennett1976efficient,bennett1976efficient,shirts2003equilibrium}.
 BAR also requires significantly less overlap between the configurational space of each state to converge than EXP, though some overlap must still exist.
@@
 In addition to the choice of initial configuration, results can depend on the choice of force field for the protein/receptor, water, and small molecule(s), so rerunning calculations with different choices of force field can also be used to assess how sensitive results and conclusions are to these particular choices.
-Other factors, like system preparation (choice of protonation state, tautomer, counterion presence, salt concentration, etc.) can also substantially impact results~\cite{mobley2017predicting, mobley2017predictingb}, so unless modelers are confident they have these factors correct, sensitivity to these choices may also need to be examined.
+Other factors, like system preparation (choice of protonation state, tautomer, counterion presence, salt concentration, etc.) can also substantially impact results~\cite{mobley2017predicting, mobley2017predictinga}, so unless modelers are confident they have these factors correct, sensitivity to these choices may also need to be examined.
 
@@
 \paragraph{Overlap matrix}
-One way of assessing reliability of the calculations is checking the phase space overlap between neighboring $\vec{\lambda}$-windows~\cite{wu2005phasespaceb, wu2005phasespacec}.
+One way of assessing reliability of the calculations is checking the phase space overlap between neighboring $\vec{\lambda}$-windows~\cite{wu2005phasespace, wu2005phasespacea}.
 For this purpose, a so-called overlap matrix $\mathcal{O}$ can be used.

### 2021-03-17 paper/manuscript.tex
fixing published version

@@
 %%% PREAMBLE
-\documentclass[9pt,bestpractices,ASAPversion]{livecoms}
+\documentclass[9pt,bestpractices,pubversion]{livecoms}
 % Use the 'onehalfspacing' option for 1.5 line spacing
@@
 
-\newcommand{\versionnumber}{0.3} % you should update the minor version number in preprints and major version number of submissions.
+\newcommand{\versionnumber}{1.0} % you should update the minor version number in preprints and major version number of submissions.
 \newcommand{\githubrepository}{\url{https://github.com/alchemistry/alchemical-best-practices}} %this should be the main github repository for this article
@@
 calculations, i.e., on average, the free energy change computed by
-free energy calculations agrees with the actual free energy change. Additional analysis of this type is presented in Brown et al.~\cite{brown2010freeenergy}
+free energy calculations agrees with the actual free energy change. Additional analysis of this type is presented in Brown et al.~\cite{brown2010free}
 
@@
 Fig.~\ref{fig:freeenergytrajectories} illustrates how looking at the convergence of your data may be important.
-In this example, the guest G3 shows different convergence behaviour for two different hosts.
-The CB8 host with guest G3 has a longer correlation time than the octa acid (OA) host.
+The CB8 host with guest G3 has a longer correlation time than the G6 guest in the octa acid (OA) host.
 In some cases, slow correlation time may not be expected and therefore not a feature known in advance.
@@
     \caption{Average binding free energy of 5 replicate Hamiltonian replica exchange calculations as a function of total simulation time (i.e.
-the sum of the simulation time of all replicas) for the two host-guest systems CB8-G3 and OA-G3.
+the sum of the simulation time of all replicas) for the two host-guest systems CB8-G3 and OA-G6.
 Shaded areas represent 95\% confidence intervals around the mean computed from the 5 replicates data.
-The horizontal dash-dot lines show the final binding free energy prediction of the two calculations after a total of 5230 ns for OA-G3 and 6650 ns for CB8-G3.
+The horizontal dash-dot lines show the final binding free energy prediction of the two calculations after a total of 5230 ns for OA-G6 and 6650 ns for CB8-G3.
 Dashed lines are the free energy trajectories computed in the forward (blue) and reverse (red) time direction for a single replicate calculation.
@@
 	\item \href{http://pmx.mpibpc.mpg.de/instructions.html}{PMX}, an add-on to GROMACS, offers a mutation free energy calculation module~\cite{abraham2015gromacs}.
-	\item \href{https://github.com/qusers/Q6}{Q} is MD code for performing FEP calculations using a variety of force fields~\cite{aqvistj.2017q6}. 
+	\item \href{https://github.com/qusers/Q6}{Q} is MD code for performing FEP calculations using a variety of force fields~\cite{aaquist2017q6}. 
 	\end{itemize}

### 2024-07-12 paper/manuscript.tex
Add paragraph on newer ABFE restraint schemes
Specifically the quaternion-based restraints of Fu et al. and
the DBC restraint from Salari et al..

@@
 
+More recently, restrictive restraint schemes have been proposed that avoid relying on a few critical reference atoms.
+Instead, these aim to simplify restraint selection by using larger sets of atomic positions in the receptor and ligand.
+Fu et al.
+suggested restraining the six relative rigid-body degrees of freedom derived by finding the rotation of the ligand which minimises RMSD to a reference structure of the receptor-ligand complex (after accounting for rotation and translation of the receptor)~\cite{fu2017new}.
+Salari et al.
+proposed restraining the ``distance-from-bound-configuration'' (DBC) variable, which is the RMSD of ligand coordinates in the frame of reference of the binding site~\cite{salari2018streamlined, ebrahimi2022symmetry}.
+This may allow the binding pose to be closely preserved as the ligand intermolecular interactions are removed, but has the disadvantage of coupling the internal and relative external degrees of freedom of the receptor and ligand, preventing easy computation of the standard state correction.
+This necessitates an additional step where the DBC restraints are released to a single harmonic restraint.
+Both restraint schemes are implemented in open-source workflows~\cite{fu2021bfee2, fu2022accurate, santiago-mcrae2023computing}.
+
 \subsection{Absolute and relative calculations deal with some of the same issues}
@@
 The determination of regions of high curvature when estimating the integral is helpful to determine regions of phase space where more sampling and/or more $\vec{\lambda}$ states are necessary to obtain the best approximation of the integral.
-Plotting $\vec{\lambda}$ with respect to the gradients at each of the $\vec{\lambda}$ values can be be a helpful diagnostic.
+Plotting $\vec{\lambda}$ with respect to the gradients at each of the $\vec{\lambda}$ values can be a helpful diagnostic.
 Additionally, computation of the overall variance of TI requires the calculation of the overall variance of integration, rather than each individual $\Delta$G$_{i,i+1}$ and assuming variances add independently.

### 2024-07-12 paper/manuscript.tex

@@
 The choice is likely especially important with Boresch-style restraints, where some relative placements of reference atoms are likely to be numerically unstable; additionally, ligand reference atoms should likely be in a part of the molecule which defines the binding orientation well, rather than in a floppy solvent-exposed tail, for example.
+Several automated methods have been proposed to select anchor points for Boresch restraints, which typically involve analysing a simulation of the fully-interacting receptor-ligand complex~\cite{alibay2022evaluating, alibay2021mdrestraintsgenerator, baumann2023broadening, chen2023enhancing, hedges2023suite}.
 

### 2024-10-15 paper/manuscript.tex
Add required changes for v2.0
As discussed during the steering meeting. Some content has been moved
to the second paper (see
https://github.com/alchemistry/largescale-alchemy-best-practices).

@@
 
-\newcommand{\versionnumber}{1.0} % you should update the minor version number in preprints and major version number of submissions.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%% Added to allow customisation of table of contents
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\usepackage{tocloft}
+\usepackage{xcolor}
+\usepackage{todonotes}
+
+% Define custom commands for colored TOC entries
+\newcommand{\tocgreen}[1]{\textcolor{green}{#1}}
+\newcommand{\tocorange}[1]{\textcolor{orange}{#1}}
+\newcommand{\tocred}[1]{\textcolor{red}{#1}}
+
+% Define custom commands for TOC entries with comments
+\makeatletter
+\newcommand{\tocsectioncomment}[1]{%
+  \addtocontents{toc}{%
+    {\leftskip \cftsecindent\relax
+     \advance\leftskip \cftsecnumwidth\relax
+     \rightskip \@tocrmarg\relax
+     \textit{#1}\protect\par}}%
+  \textit{#1}\par}
+
+% Command for subsection comments in TOC
+\newcommand{\tocsubsectioncomment}[1]{%
+  \addtocontents{toc}{%
+    {\leftskip \cftsubsecindent\relax
+     \advance\leftskip \cftsubsecnumwidth\relax
+     \rightskip \@tocrmarg\relax
+     \textit{#1}\protect\par}}%
+  \textit{#1}\par}
+
+% Command for subsubsection comments in TOC
+\newcommand{\tocsubsubsectioncomment}[1]{%
+  \addtocontents{toc}{%
+    {\leftskip \cftsubsecindent\relax
+     \advance\leftskip \cftsubsubsecnumwidth\relax
+     \rightskip \@tocrmarg\relax
+     \textit{#1}\protect\par}}%
+  \textit{#1}\par}
+
+\makeatother
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\newcommand{\versionnumber}{2.0} % you should update the minor version number in preprints and major version number of submissions.
 \newcommand{\githubrepository}{\url{https://github.com/alchemistry/alchemical-best-practices}} %this should be the main github repository for this article
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{Best Practices for Alchemical Free Energy Calculations [Article v\versionnumber]}
+\title{Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]}
 \author[1*]{Antonia S. J. S. Mey}
@@
 \begin{abstract}
+
+\todo[inline, color=green!20]{
+
+The original best practices guide will be split into two documents: the first will gives best practices for running a \emp{single} calculation (and cover theory in more depth/ detail system-specific setup etc.), while the second will give best practices for running \emp{multiple} calculations (with a focus on automation).
+The first will become version 2.0 of the current guide, while the second will be submitted as a new manuscript.
+
+This document outlines the contents of the first document for single calculations.
+}
+
+\todo[inline, color=green!20]{@Volunteer Update abstract to highlight the split of the document and focus on the single calculations.}
+
 Alchemical free energy calculations are a useful tool for predicting free energy differences associated with the transfer of molecules from one environment to another.
@@
 In this paper, we review current best practices for several popular application domains of  alchemical free energy calculations performed with equilibrium simulations, in particular relative and absolute small molecule binding free energy calculations to biomolecular targets.
-%\todo[inline, color=green!20]{JDC: Should we migrate this document to \url{https://github.com/alchemistry}? ASJSM: Let's migrate with the submitted version, but leave it here for now?}
+
+% \todo[inline, color=green!20]{JDC: Should we migrate this document to \url{https://github.com/alchemistry}? ASJSM: Let's migrate with the submitted version, but leave it here for now?}
 \end{abstract}
@@
 
-%\todototoc
-%\listoftodos
+\newpage
+% Add a comment to the top of the contents page
+\addtocontents{toc}{\vskip2pt\noindent\textcolor{green}{\textbf{Green sections}: Minimal revisions required}\par\vskip2pt}
+\addtocontents{toc}{\vskip2pt\noindent\textcolor{orange}{\textbf{Orange sections}: Substantial revisions required}\par\vskip2pt}
+\addtocontents{toc}{\vskip2pt\noindent\textcolor{red}{\textbf{Red sections}: Needs written from scratch.}\par\vskip2pt}
+
+\tableofcontents
+
+\newpage
 
@@
 
-\section{What are alchemical free energy methods?}
+\section{\tocgreen{What are alchemical free energy methods?}}
+\tocsectioncomment{Steering Meeting: @Jenke Scheen to think about how figure can be made the "central axis" of the paper.
+Possibly update/ create new full-page figure and expand so that it shows a ``dummy workflow'' for an alchemical simluation including all general stages (e.g setup, analysis with different estimators etc.), and guides readers to appropriate sections.}
 \label{sec:intro}
@@
 %%%%%%%%%%%%%%%%%%%%
-\section{Prerequisites and Scope}
+\section{\tocorange{Prerequisites and Scope}}
+\tocsectioncomment{Steering Meeting: @Volunteer to update reflecting new split between single and multiple calculations. @Michael Shirts to update best practices on multiple binding sites.}
 \label{sec:pre}
+
 This Best Practices guide focuses on providing a good starting point for new practitioners and a reference for experienced practitioners. 
@@
 %%%%%%%%%%%%%%%%%%%%
-\section{Statistical mechanics demonstrates why alchemical free energy calculations work}
+\section{\tocorange{Statistical mechanics demonstrates why alchemical free energy calculations work}}
+\tocsectioncomment{Steering Meeting: @Emilio Gallicchio to update theory.
+@Stefan Boresch to comment on dummy atom considerations?
+Will frame it to be as general as possible, while remaining accessible to beginners.
+Will align with Gilson et al.
+1997 paper.
+@Darrin York to review.}
 \label{sec:theory}
+
 Why would you want to run an alchemical free energy calculation and why do they work?
@@
 %%%%%%%%%%%%
-\section{What can be expected from alchemical simulations?}
+\section{\tocorange{What can be expected from alchemical simulations?}}
+\tocsectioncomment{Steering Meeting: @Volunteer Shorten and move most material to second paper.
+Possibly add a walkthrough of a ``dummy calculation'' (which could be a particular realisation of a general workflow laid out in @Jenke Scheen's figure)?
+This could lead on to discussion of particular methods.}
 \label{sec:step0}
@@
 
-\subsection{Is the expected accuracy of the computation sufficient?}
-\label{subsec:accuracy}
-The requisite level of accuracy is another important consideration. If the
-goal is to guide lead optimization when many compounds will be
-synthesized, free energy calculations can be appealing even with
-accuracies in the 1--2 kcal/mol range~\cite{mobley2012perspective}, but if the number of compounds to be synthesized is very small, this accuracy may not be enough to provide much value.
-
-Here we provide a simple estimate of the value provided by alchemical
-free energy calculations in lead optimization. Let $P(\Delta\Delta
-G)$ be the probability distribution of the changes in the binding free
-energies of a new set of molecules during one round of lead
-optimization, and let $P(\Delta\Delta G^\dagger|\Delta\Delta G)$ be the
-conditional probability of the binding free energy change computed by
-the free energy calculations, $\Delta\Delta G^\dagger$, given the actual
-change $\Delta\Delta G$. The latter conditional probability can be modeled
-by a normal distribution
-\begin{equation}
-  P(\Delta\Delta G^\dagger|\Delta\Delta G) = \frac{1}{\sqrt{2\pi\sigma^2}}
-  \exp\left(-\frac{(\Delta\Delta G^\dagger - \Delta\Delta G)^2}{2\sigma^2}\right),
-  \label{eqn:free-energy-distribution}
-\end{equation}
-where $\sigma$ signifies the accuracy of free energy calculations.
-Here we assume that there is no systematic bias in the free energy
-calculations, i.e., on average, the free energy change computed by
-free energy calculations agrees with the actual free energy change. Additional analysis of this type is presented in Brown et al.~\cite{brown2010free}
-
-In lead optimization guided by free energy calculations, we will likely only
-synthesize and experimentally test molecules that are predicted to
-have favorable free energy changes. We are thus interested in how
-often a molecule predicted to bind stronger actually turns out to
-bind stronger. In other words, we are interested in the conditional
-probability:
-\begin{equation}
-  P(\Delta\Delta G<0|\Delta\Delta G^\dagger<0).
-  \label{eqn:true-positive}
-\end{equation}
-
-For illustrative purposes, consider a proposed set of new molecules, and assume that the changes proposed in these molecules yield a set of relative binding free energies that follow a normal
-distribution. That is, assume that the standard deviation in the relative binding free energies for the changes represented is $RT\ln 5$
-(corresponding to a 5-fold change in the binding affinities), and that
-1 in 10 new molecules have increased binding affinity ($\Delta\Delta G
-\leq 0$). Under such assumptions, the conditional probability in
-Eq.~\ref{eqn:true-positive} can be easily computed.
-
-If the accuracy of a collection of free energy calculations is $\sigma = 1$ kcal/mol, $P(\Delta\Delta
-G<0|\Delta\Delta G^\dagger<0) = 0.35$, which means that out of every
-10 molecules selected for predicted favorable free energy change, on
-average 3.5 molecules will have actual favorable free energy change.
-In other words, selection by free energy calculations yields 3.5 times
-more molecules of improved affinities than selection without free
-energy calculations under these assumptions.
-
-Available computational resources and timescales of motion also factor
-into this initial analysis. An individual free energy calculation
-involves simulations at many different intermediate states (perhaps
-20-40 or more) and each of these must typically be long enough to
-capture the relevant motions in the system. If such motions are
-microsecond events or longer, the computational cost of running 20-40
-microsecond or longer simulations for each of $N$ ligands will likely be
-prohibitive for most users with today's hardware. On the other hand, if key motions are fast and minimal (as is often assumed in practice), much shorter simulations may be sufficient.
-
-\subsection{Can I afford the calculation?}
-\label{subsec:affordability}
-Furthermore, are available computational resources sufficient that throughput will be reasonable compared to needs of experimental collaborators working on
-this system? How many ligands ($N$) can you afford to handle given
-your computational resources? As cloud computing becomes more available, in-house GPU clusters may not be necessary if calculations are not run on a regular basis.
-This analysis should be done up front as part of ``counting the cost''
-of involvement in a particular project. In some cases, the analysis may conclude that free energy calculations will not be feasible for the proposed problem.
-Here, by ``cost'', we refer not just to financial cost of the calculations relative to experiments, but also time -- can the calculations be run faster than experiments are done?
-How will the relevant resource and opportunity costs factor in?
-Both computation and experiment require human time, supplies (of different sorts), and equipment.
-In the extreme limit, for example, it would not make sense to spend a month running a binding free energy calculation if the equivalent experiment could be done in a day with resources already on hand.
-Such issues should be considered before deciding to conduct binding free energy calculations.
-
-\subsection{Is an exploratory study what I want?}
-\label{subsec:exploration}
-An additional consideration is how much is known about your particular
-target, ligand binding modes in the target, and any relevant motions
--- essentially, has it been studied enough to know whether it might be
-suitable for free energy calculations?
-It is important to know if the system has hardly been studied, because should the initial calculations perform poorly, the effort may turn into an attempt to understand the relevant sampling, force field, or system preparation problems.
-
-If you are unsure whether your project is feasible, as mentioned above, one recommended option is to conduct a short exploratory study to assess tractability for a small
-number of ligands. This can be sufficient to get an initial
-idea of feasibility and accuracy of the calculations for the
-proposed target~\cite{schindler2020largescale}.
-
-\section{How should alchemical simulations be applied to drug discovery?}
-\label{sec:drugdiscovery}
-Many practitioners expect alchemical methods to provide valuable guidance for drug discovery, and to exhibit accuracy superior to most alternative approaches for suitable targets~\cite{kuhn2017prospective}.
-Successful application in industry may require considerable knowledge of the ``domain of applicability'' of free energy calculations -- where they work well and where they will not~\cite{sherborne2016collaborating}.
-Successful application also requires robust protocols for preparing, submitting and analysing alchemical calculations.
-In this regard, the issues mentioned in the previous section such as understanding the suitability and timescales to capture the structure activity relationships (SAR), and performing up-front tests of performance are all relevant to drug discovery applications.
-Without venturing too far into details of system setup, which is beyond the scope of this article, we highlight some critical factors affecting accuracy and successful application.
-
-\subsection{Capturing experimental conditions}
-\label{subsec:exp_condition}
-The calculations aim to capture the alchemical change from one ligand to another as accurately as possible.
-Therefore, it is necessary to consider details of the experimental setup, such as pH.
-Biological assays are usually run at neutral pH but this is not always the case.
-For example, some enzymes exhibit pH-dependent activity and assays may thus be done in conditions other than neutral pH.
-Therefore, computational protein and ligand preparation protocols should reflect experimental pH.
-
-The formal charge and/or tautomeric state of the small molecules can change within a series of analogs, necessitating care in treatment.
-Additionally, medicinal chemistry efforts might deliberately modify the pKa of a series to modify drug properties, requiring explicit efforts to incorporate these changes into alchemical calculations.
-
-To ensure modeling matches experiment, we also need to accurately prepare and simulate the same system -- which requires understanding what protein construct is used in the bioassay.
-For instance, does the X-ray structure that is to be used for the calculations match the construct used for screening (i.e.
-only the catalytic domain vs.
-full length, monomer vs.
-dimer, etc.)~\cite{perez-benito2018predicting}?
-Also, were certain co-factors or partner proteins required in the bioassay?
-
-\subsection{Is my binding mode accurate?}
-As also mentioned, good performance of alchemical calculations requires an accurate representation of the ligand binding mode, usually from a high quality X-ray crystal structure.
-If more than one structure is available, the modeler should pay attention to choose the most suitable.
-The quality of the structure can be a concern, and the reader is referred to work of Warren et al.
-for a detailed discussion of choosing optimal structures for structure-based modeling~\cite{warren2012essential}.
-
-It is also useful to study the structure activity relationship and understand the expected impact of any mutations on the binding site, such as whether side chain movement in the protein will be required, and whether there is evidence of this in any alternative X-ray structures of the same protein.
-Often, only one protein and water configuration is used for a series of alchemical calculations, so this needs to be capable of accommodating the smallest through to largest ligands in a way that allows stable and well behaved simulations.
-This can provide a practical limit on the alchemical changes that are feasible, though a simple work-around can be to separate compounds into sub-series for different calculations.
-
-If multiple structures are available there is some evidence the higher affinity complex can give better match to experiment~\cite{perez-benito2019predicting}, at least in some cases.
-However, ligands and proteins can also undergo unexpected changes in binding mode for related ligands, which can make these issues more complex to deal with~\cite{mey2016blinded}.
-
-\subsection{Input setup and scale of calculations}
-In a drug discovery setting it is normal to consider dozens (or more) of ligands and it is necessary to align them in the binding site.
-There is no detailed study of how different alignment approaches may affect results, but the user should be aware of some practical considerations.
-Tools are available to compare the ligands and build the combined topologies that define the changes between one ligand and another~\cite{loeffler2015fesetup,hedges2019biosimspace,gapsys2015pmx}.
-In simple terms, providing poor alignment to these tools will make this job harder.
-Docking with restraints is often beneficial in this regard.
-Particularly, fixing the 3D spatial position of the scaffold using maximal common substructure (MCSS) restrained docking can help provide well aligned input for the topology generation.
-Nevertheless, in this case careful attention is still needed to ensure consistency of alignment for identical substituents.
-Another alternative is to manually edit the same core and add/modify the changing substituents.
-This provides assurances that coordinates for the non-perturbed portion of the structure remain identical and aromatic substituents, for instance, have consistent dihedral angles.
-However, it is not feasible for many compounds and therefore automation is desirable.
-
-Finally, the role of water in ligand binding is not always well understood and it can be crucial to capture the changes in binding site solvation during ligand binding.
-Can crystallographic waters be retained?
-Do they clash with some of the larger ligands used in the alchemical perturbation?
-See Sec.~\ref{subsec:binding} for different strategies that can be applied to dealing with waters.
-Generally, before launching large numbers of alchemical free energy calculations it is always recommended to test the system using classical MD simulations and limited numbers of alchemical perturbations.
-Metrics such as ligand and protein RMSD and RMSF can be inspected, along with visual inspection of simulations, to ensure the system is stable and likely to be suitable for alchemical calculations.
-
-Running binding free energy calculations in a drug discovery application will typically require the use of software or tools to facilitate the large number of calculations.
-Commercial implementations such as FEP+, OpenEye Tools, or Flare allow for a fast setup and deployment to GPU hardware in minutes, but may have limited ability to customize calculations~\cite{wang2015accurate,kuhn2020assessment}.
-Commercial tools can be expensive in some cases, but non-commercial tools are becoming more straight forward to use to run alchemical free energy calculations~\cite{gapsys2015pmx, loeffler2015fesetup, song2019using, gapsys2020large, jespers2019qligfep, hedges2019biosimspace, kuhn2020assessment}.
-
-For relative free energy calculations, various graph topologies or maps of calculations are possible, and choices may depend on the target application.
-For instance, if the goal is to accurately assess the relative binding energy of a small number of compounds, possibly with challenging syntheses, the map of perturbations should contain as many connections between compounds as affordable.
-However, when running calculations on hundreds of compounds a so called \emph{star-map} (see Fig.~\ref{fig:fig_types_of_networks}\textbf{A}) can be used that just contains one connection per compound: perturbing every compound to a central ligand, typically the crystal structure ligand~\cite{konze2019reactionbased}.
-In this way the top-ranking examples can be readily identified and submitted to additional calculations in a second round.
-Alternatively, if the goal is to achieve the smallest possible error with minimal computational expense, certain graph topologies provide benefits~\cite{yang2020optimal, xu2019optimal}
-
-\subsection{Making predictions, understanding errors}
-\label{subsec:predictions}
-For prospective drug discovery applications there are several other considerations including understanding likely errors and taking selection bias into account.
-
-It is crucial when proposing compounds for synthesis to have some idea of the underlying error or uncertainty in the predictions.
-A retrospective assessment can give an indication of prospective performance for similar molecules~\cite{ciordia2016application}.
-Beyond this, several parameters provide useful indicators of performance.
-For example error estimates provided by free energy estimators that are too large can highlight poorly converged simulations~\cite{perez-benito2019predicting}.
-Hysteresis, either within cycles in the perturbation network or between forward and backward perturbations can be checked~\cite{wang2013modeling} to indicate problematic perturbations involved in cycles connecting many compounds (See also Secs.~\ref{sec:relative-fe-protocol} and~\ref{sec:are-they-good}).
-Once synthesis and testing of compounds is complete a standard strategy is to look back at how the calculations performed.
-In this regard it is important to consider the issue of selection bias upfront.
-It is tempting to only synthesize the compounds predicted to be most active, thus a narrow range of calculated activity is tested that imposes limits on the statistical assessment of performance, ideally example molecules from across the range of predicted activity can be assessed or corrections can be applied based on previous recommendations~\cite{robert2017critical}.
-For a more detailed discussion on checking the robustness of your alchemical free energy calculation see also Sec.~\ref{sec:are-they-good}.
-
-In summary, the successful use of alchemical calculations, particularly for drug discovery, requires working in the domain of applicability, using a high quality X-ray structure of the target bound to compounds in the series, and testing the approach retrospectively to ensure the system setup is well-behaved.
-Always assess your confidence in the resulting predictions and communicate this when discussing with experimentalists.
-Consider performing repeat calculations for at least some of the perturbations in the study.
- There are many accounts of success of alchemical calculations, the methods show good performance towards the goal of binding free energy prediction.
-However, it is important to have realistic expectations.
-
-Structure based drug design projects are often capable of improving potency relatively quickly, even with only limited application of computational approaches and the range of activity narrows to just two-to-three log units.
-It may seem hard to have impact with substantially different, more potent, stand-out compounds in this scenario, but binding free energy predictions can still be extremely useful for ensuring activity is maintained as other properties are optimized.
-An interesting cost benefit analysis has shown the value of activity prediction, see discussion above and articles such as~\cite{mobley2012perspective}.
-From a drug discovery point of view, alchemical calculations are expanding their domain of applicability, and there are reports of success using homology models~\cite{cappel2016relative} and GPCRs~\cite{deflorian2020accurate,lenselink2016predicting} for instance, as well as enabling charge change and scaffold hopping~\cite{chen2018accurate, wang2017accurate}, but these systems are undoubtedly more difficult.
-In the meantime, use cases are expanding to resistance prediction, selectivity prediction , solubility prediction – an exciting future for alchemical calculations~\cite{hauser2018predicting, albanese2020structure, mondal2019free}.
-
 %
@@
 %
-\section{Simulation prerequisites}
+\section{\tocorange{Simulation prerequisites}}
+\tocsectioncomment{There was substantial interest in expanding this section (see initial suggestions/ discussion).}
+\tocsectioncomment{@Irfan Alibay/ @Volunteer to suggest additional points.}
+\tocsectioncomment{@Irfan Alibay/ @Emilio Gallicchio/ @Volunteer to expand on approaches for explicitly accounting for protonation/ tautomer changes.}
+\tocsectioncomment{@Hannah Bruce Macdonald/ @David Slowchower/ @Volunteer to expand range of systems discussed, e.g. membrane proteins, cofactors, multimers, PPIs.}
 \label{sec:prerequisites}
@@
 \subsection{Free energies of binding}
+\tocsubsectioncomment{{Steering Meeting: @Volunteer move paragraph ``Congeneric series often need alignment'' to second paper if we are assuming a single calculation?}}
 \label{subsec:binding}
@@
 \subsection{Free energies of hydration or partition coefficients}
+\tocsubsectioncomment{Steering Meeting: @Volunteer to expand coverage of free energies of hydration/ partition coefficients to make coverage of applications more equal.}
 \label{subsec:hydration}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{What simulation protocol should I choose?}
+\section{\tocorange{What simulation protocol should I choose?}}
+\tocsectioncomment{Steering group: Possibly expand to include other methods group agrees on.}
+\tocsectioncomment{@Stefan Boresch to restructure as per his suggestion to remove single/dual topology section and begin with a more fundamental discussion which might more naturally include all options e.g.
+ATM.
+@Stefan Boresch to mention best practices for dummy atoms?
+This might benefit from following @Emilio Gallicchio's changes to make the theory section more general.}
+\tocsectioncomment{@Sheenam Khuttan / @Emilio Gallicchio to contribute section on ATM.}
+\tocsectioncomment{@Volunteer to add best practices for non-equilibrium calculations.}
 \label{sec:simulation_protocol_choice}
@@
 \subsubsection{Choices unique to relative free energy calculations}
+\tocsubsubsectioncomment{Steering Group: @Volunteer to mainly move atom mapping section to second paper and assume that the change is small (and hence atom mapping is trivial).
+@Benjamin Ries to add discussion of 3D atom mapping in second paper.}
 \label{sec:relative-fe-protocol}
@@
 Some approaches have attempted to address this~\cite{clark2019relative} but a general solution is not yet in mainstream use. Still, FEP+ implements one solution.
-
-\paragraph{Perturbation maps}
-Based on the input ligand series, a perturbation map or network can be planned.
-Recent heuristics have shown the more connected the perturbation network the better.
-However, there is a way to optimize network structure while minimizing the number of perturbations that need to be computed reducing the resulting computational cost~\cite{yang2020optimal,xu2019optimal}.
-Sometimes the introduction of intermediates that are not part of the original congeneric series are essential to avoid ring breaking, or to deal with perturbations that would otherwise result in large numbers of atoms being inserted or deleted.
-Some commercial tools have good underlying heuristics but may fail with complicated input, needing user validation in particular when dealing with chiral compounds.
-
-\begin{figure}
-    \includegraphics[width=0.95\columnwidth]{figures/fig6_types_of_networks/Figure.pdf}
-    \caption{\textbf{Examples of perturbation networks} (\textbf{A}) Star shaped network with the crystal structure in the center.
-(\textbf{B}) Network with cycle closures (see more on this in Sec.~\ref{sec:are-they-good}).
-Arrows indicate the direction of the perturbation.
-Fully converged binding free energy calculations yield binding free energy changes which sum to zero around any closed cycle.
-However, in practice errors may not sum to zero around closed cycles, providing a way to look for potential sampling problems.
-Here in (B), green cycles indicate cycles with hypothetically good cycle closure, red those with poor cycle closure.
-The red arrow indicates a poorly converged simulation that would give rise to bad cycle closures.
-The diamond indicates the use of a crystallographic binding mode.}
-    \label{fig:fig_types_of_networks}
-\end{figure}
-
-In some cases, during the lead optimization stage, or for very large datasets that would benefit from rougher initial free energy ranking, or in cases where perturbations would be rather large, a star shaped network as seen in Fig.~\ref{fig:fig_types_of_networks} \textbf{A} is used.
-However, adding redundancy into the network means that a better error analysis can be carried out by looking at cycle closure errors as discussed in sec.~\ref{sec:are-they-good}, with an example given in Fig.~\ref{fig:fig_types_of_networks}\textbf{B}.
-
-Methods from experimental design have been applied to the construction of the perturbation maps.
-Yang et al.~\cite{yang2020optimal} showed how to optimize the perturbation map by selecting a fixed number of calculations from the pairwise perturbations so that the resulting set of calculations minimize the total variance.
-Xu~\cite{xu2019optimal} showed how to optimize the perturbation map by allocating different amounts of simulation time to different pairwise perturbations so as to minimize the total variance, given the total simulation time of all the perturbation calculations.
-Both approaches lead to substantial reduction in the statistical error of the estimated free energies.
 
@@
 \subsubsection{Absolute free energy calculations must handle the standard state and use restraints}
+\tocsubsubsectioncomment{@Emilio Gallicchio to comment on possible issues with theoretical discussion and possibly update in the context of revised theory section.
+See PR from @Finlay Clark to improve discussion of restraint schemes and automated restraint selection.
+@Stefan Boresch may like to weigh in on analytical corrections/ general discussion.
+@Andrea Rizzi to reivew?}
 \label{sec:standardstate-restraints}
@@
 \subsubsection{Changes in net charge can be challenging/problematic.}
+\tocsubsubsectioncomment{@Chris Oostenbrink/ Vytas Gapsys to update.}
 If the net charge of the system changes as the alchemical variable changes during the  calculation, this can pose major challenges.
@@
 \subsubsection{How long should I run my simulation for and what information should be saved?}
+\tocsubsubsectioncomment{@Volunteer: Move parts of this to analysis section (in Single/ Multiple papers).}
 \label{sec:sim_length_information_kept}
@@
 \subsubsection{Multiple or uncertain binding modes may require considerable care}
+\tocsubsubsectioncomment{@Michael Shirts to update discussion of treatment of multiple binding poses.}
 \label{sec:multiple_binding_modes}
@@
 \subsection{General simulation setting choices that could affect free energy calculations.}
+\tocsubsectioncomment{@Volunteer: Add short discussion of recommended integrators? BAOAB?}
 There are many parameter choices that are common to standard MD simulations.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Data analysis}
+\section{\tocorange{Data analysis}}
 \label{sec:data_analysis}
@@
 While a number of different estimators are available that will give consistent results under optimal circumstances, some approaches are recommended over others due to their robustness and ability to provide information on poor convergence.
+
+\subsection{Estimators for free energy differences}
+\label{subsec:estimators}
+Free energy differences between two different states differing in the energy function are directly related to the
+ratio of probabilities of those states.
+As can be noted, the partition functions in Eq.~\ref{eq:conf_probability} are simply the total accumulated probabilities for all possible configurations of the system.
+Virtually all of the ways to estimate this free energy are based in converting this ratio of integrals to something that can be measured in one (or several) simulations.
+
+\paragraph{The Zwanzig relationship (EXP)}
+The simplest method for calculating free energy differences from simulations is the so-called \textit{Zwanzig
+relationship}~\cite{zwanzig1954hightemperature}, also called one-sided exponential re-weighting (EXP), or simply free energy perturbation, though this final term is sometimes used to encompass all ways of calculating free energy differences.
+
+The (reduced) free energy difference $\Delta f_{01}$ between an initial state 0 and a final state 1 defined by two different potential energy functions 
+$u_0(\vec{q})$ and $u_1(\vec{q})$ over coordinate space $\vec{q}$ can be calculated as:
+\begin{eqnarray}
+\Delta f_{01} & = & -\ln \expect{e^{-(u_1(\vec{q}) - u_0(\vec{q}))}}_0 =  -\ln \expect{e^{-\Delta u(\vec{q})} }_0
+\end{eqnarray}\label{eqn.zwanzig}
+and the average is over all samples from the simulation performed with $u_0$.
+In the case of NVT (canonical) sampling and assuming the masses do not change, then $u$ is simply $U/k_BT$, and $f$ is $F/k_BT$, but it can be generalized to other ensembles with the proper definition of $f$ and $u$.
+Described in words, we take the samples generated during our run with the potential energy function $u_0(\vec{q})$ and calculate what the difference in energy would be if we switched instantaneously to the potential energy function $u_1(\vec{q})$, and average the exponential of the negative energy difference to get the negative of the exponential of the free energy difference.
+The original distributions, P($u_0$) as generated at $\vec{\lambda}=0$ and P($u_1$) would look like those seen in Fig.~\ref{fig:fig_sampling_scheme}\textbf{A}-\textbf{C} on the right hand side.
+Reevaluating requires almost no extra code functionality to perform; one need only to save a full precision trajectory, and run an unmodified molecular simulation code using the $u_1$ in order to calculate the new energies of stored snapshots.
+The analysis can be written in a line of code.
+We note that this method is even more general, in that the instantaneous work to change the potential energy function from $u_0$ to $u_1$ can be replaced by the non-reversible work $W$ to make the same change beginning from the same equilibrium conditions at either end state~\cite{jarzynski1997nonequilibrium,jarzynski1998equilibrium,crooks2000pathensemble}, allowing for non-equilibrium free energy calculations, an alternate approach.
+We do not detail non-equilibrium transformations here, and refer the reader to more advanced treatments~\cite{maragakis2008bayesian,oberhofer2005biased,procacci2015unbiased,shirts2003equilibrium,ytreberg2004singleensemble, gapsys2020large}, as our focus here is on equilibrium free energy techniques.
+
+Although the Zwanzig equation is formally correct as long as the two states considered sample the same phase space volume, which is true for standard molecular models, it has some very important numerical issues that mean that it often performs badly for standard free energy calculations, even for small molecules~\cite{shirts2005comparison,lu2003appropriate}.
+One can show that if the standard deviation of the difference $\Delta u(\vec{q}) = u_1(\vec{q})-u_2(\vec{q})$ over all sampled $\vec{q}$ is large, which in this case, means only several times $k_BT$, then very few samples contribute to the average, and the answer will be both biased and extremely noisy~\cite{lelievre2010free}.
+Essentially, the method is dominated by contributions of rare snapshots~\cite{jarzynski2006rare, wu2005phasespace, wu2005phasespacea}.
+
+\paragraph{The Bennett Acceptance Ratio (BAR)}
+If we have the differences in the potential energy sampled from the distribution defined by $u_0$ to the state defined by $u_1$, and we also have the differences in potential energies from the distribution sampled by $u_1$ to the state defined by $u_0$, we can obtain a significantly improved estimate of the
+free energy difference compared to that obtained by EXP. 
+This estimate was first derived by Bennett and is hence generally called the Bennett Acceptance Ratio (BAR).
+It is solved by finding the reduced free energy $f_{ij}$ that satisfied the following implicit equation:
+\begin{eqnarray}
+ \sum_{i=1}^{n_i} \frac{1}{1 + \exp[\ln(\frac{n_i}{n_j}) + u_{ij}(\vec{q}) - f_{ij})
+ ]} \nonumber \\
+ =\sum_{i=1}^{n_j} \frac{1}{1 + \exp[\ln(\frac{n_i}{n_j}) - u_{ij}(\vec{q}) + f_{ij})]},
+\end{eqnarray}
+where $n_i$ and $n_j$ are the number of samples from each state.
+More recent derivations show that this formula is the maximum likelihood estimate of the free energy difference given sets of samples from the two states~\cite{shirts2003equilibrium}.
+
+Many studies have demonstrated both the theoretical and practical superiority of BAR over EXP in molecular
+simulations~\cite{shirts2005comparison,lu2003appropriate}, and BAR converges to EXP in the limit that all samples are from a single state~\cite{bennett1976efficient,bennett1976efficient,shirts2003equilibrium}.
+BAR also requires significantly less overlap between the configurational space of each state to converge than EXP, though some overlap must still exist.
+
+The Bennett acceptance ratio is only defined between two states.
+Usually, the endpoints of interest in a free energy calculation are sufficiently different that we will need a chain of states that gradually change the potential energy function from $u_0$ to $u_1$, as discussed in Sec.~\ref{sec:important_path}.
+You can carry out a BAR estimate between each pair of states $\Delta f_{1 \rightarrow N} = \Delta {f_{1\rightarrow 2}} + \Delta {f_{2\rightarrow 3}} +  \ldots + \Delta f_{N-1\rightarrow N}$.
+
+There is one important thing to note about the uncertainty estimates when summing multiple free energies together to calculate an overall free energy estimate.
+Although BAR itself gives a free energy estimate that is asymptotically correct and is much less biased than the uncertainty estimate for EXP, the uncertainties in $\Delta {f_{i-1\rightarrow i}}$ and $\Delta {f_{i\rightarrow i+1}}$ are not uncorrelated, because they both involve the energies $u_i(\vec{q})$.
+The variances of each of the free energies will \textit{not} propagate as variances usually do (in quadrature) into the variance of the overall free energy.
+Instead, some other method for propagating the uncertainty, such as bootstrapping~\cite{grossfield2018best} must be used.
+
+\paragraph{Thermodynamic integration (TI)}
+By taking the derivative of the free energy with respect to the
+variable $\vec{\lambda}$, we find that:
+\begin{equation}
+\frac{df}{d\vec{\lambda}} = \frac{d}{d\vec{\lambda}} \left[-\ln \int \frac{\exp{-u(\vec{\lambda},\vec{q})}}{Z(\vec{\lambda})} d\vec{q}\right] = \expect{\frac{du(\vec{\lambda},\vec{q})}{d\vec{\lambda}}\
+}_{\vec{\lambda}} .
+\end{equation}
+And then we can numerically integrate $df/d\vec{\lambda}$ over an alchemical transformation, using a range of different well-established techniques, to obtain:
+\begin{equation}
+\Delta f    = \int_{0}^{1} \expect{\frac{du(\vec{\lambda},\vec{q})}{d\vec{\lambda}}}_{\vec{\lambda}}  d\vec{\lambda}.    
+\end{equation}
+This approach to calculating the free energy is called thermodynamic integration (TI).
+Averaging over $\expect{\frac{du}{d\vec{\lambda}}}$ requires fewer uncorrelated samples to reach a given level of relative error
+than averaging $e^{-u(\vec{q})}$, as the distribution of values is usually narrower, with a more Gaussian shape to the distribution.
+Rather than being limited by overlap, as in the case of BAR and MBAR (see below), we are instead limited by the bias in the numerical quadrature, which must be minimized sufficiently to be beneath the level of statistical noise.
+
+Various numerical integration schemes are possible, but the trapezoid
+rule provides a simple and robust scheme. All types of numerical integration can be written as:
+\[ \Delta f \approx \sum_{k=1}^{K} w_k
+\expect{\frac{du(\vec{\lambda},\vec{q})}{d\vec{\lambda}}}_{k}, \] where the weights $w_k$ correspond to a particular choice of numerical integration.
+Researchers have tried a large number of different integration schemes~\cite{resat1993studies,jorge2010effect,shyu2009reducing}.
+However, many integration choices require specific choices of $\vec{\lambda}$ to minimize bias, which makes them unsuitable when the intermediates
+have widely-varying levels of uncertainty.
+For example, integrating a cubic spline interpolation provided negligible benefits over a simple trapezoid rule~\cite{paliwal2011benchmark}.
+As fitting to higher order polynomials can have numerical instabilities for some energy functions, and because alternate functional forms might only be appropriate with some types of transformations, expertise and experience is required to perform such numerical integration modifications.
+For starting researchers, we therefore recommend the simple trapezoid rule scheme, as it allows for maximal flexibility in which values of $\vec{\lambda}$ are simulated.
+In practice, adding 2-3 more intermediate states is typically sufficient to match the performance of these more complicated numerical quadrature schemes.
+It is also possible to calculate the $\lambda$-derivatives at non-simulated states from simulated states in a scheme named extended TI~\cite{ruiter2016extended} which reduces the integration error.
+
+One drawback of TI is that it requires derivatives with respect to $\vec{\lambda}$ to be calculated directly in the code.
+Unfortunately, many problems of interest require using pathways (such as the soft-core pathways, for removing repulsive interactions) that are not linear, as we discuss, making this more complex.
+Still, if the code of interest does compute $\frac{du}{d\vec{\lambda}}$, then TI is perhaps the simplest method to use, as it involves a very little post-processing effort.
+
+\paragraph{The multistate Bennett acceptance ratio (MBAR)}
+One can generalize Bennett's logic from two states to multiple states to obtain a free energy estimator that uses energy differences between configurations at all intermediate states to compute free energy differences between all states.
+MBAR gives a system of implicit equations for the free energies $f_i$:
+\begin{equation}
+f_i = - \ln \sum_{n=1}^{N} \frac{\exp(-u_i(\vec{q}_n))}{\sum_{k=1}^K N_k \exp(f_k-u_k(\vec{q}_n))},
+\end{equation}
+where there are $N_k$ samples from each of $K$ states, with $\sum_k N_k=N$ the total number of samples.
+Thus, we need to evaluate the energy function $u_i$ for all samples obtained at all states in the transformation.
+The equations can be solved by a number of different standard routines.
+We note that there are only $K-1$ independent equations, so only $K-1$ of the free energies are independent variables, and one of the $f_i$ must be specified (usually, without loss of generality, setting it to zero).
+
+MBAR is probably the lowest variance asymptotically unbiased estimator of the free energy given the energies of the samples~\cite{tan2004likelihood}, which means that BAR is also the lowest variance estimator for the free energy difference between only 2 states, as it is mathematically exactly the same as MBAR in this case.
+MBAR also provides an uncertainty estimate, derived from standard error propagation methods for implicit functions, which has been shown to be highly accurate as long as there are sufficient samples at each state~\cite{paliwal2011benchmark}.
+
+MBAR can also be thought of as the Zwanzig estimator of the free energy to state $i$ where the sampled distribution is the \textit{mixture distribution} of all the other samples thrown together in one ``pot'', defined by $p_m(\vec{q}) = N^{-1} \sum_k N_k \exp(f_k-u_k\vec{q})$, which is the weighted average of all the individual normalized probability distributions from the simulations that are performed~\cite{shirts2017reweighting}.
+
+\paragraph{Recommendations}
+\begin{itemize}
+\item We recommend MBAR if all energy differences are available. It is the lowest variance unbiased free energy estimate given samples from multiple states.
+\item BAR is essentially just as good as MBAR for highly optimized $\vec{\lambda}$ intermediates.
+Specifically, if the $\vec{\lambda}$s are chosen such that intermediate states have moderate overlap with their neighbors (i.e.
+between $i$ and $i+1$ and between $i$ and $i-1$, they will \textit{not} have significant overlap with their next nearest neighbors $i+2$ and $i-2$.
+Thus MBAR does not actually get significant information from these energy differences, so one might as well not even calculate them, and just perform BAR between nearest neighbors.~\cite{paliwal2011benchmark}
+\item TI usually gives similar values as MBAR implemented with sufficient numbers of intermediates, but quadrature errors that are hard to estimate beforehand  can occur if one is not careful.~\cite{paliwal2011benchmark}
+\item WHAM is an approximation to MBAR, and there are no compelling reasons it should be used.
+If careful, it is not necessarily much worse than the other methods, but it always introduces some degree if binning error.
+\item Other variants, especially ones that adaptively determine the free energies can be useful in certain circumstances but beyond the scope of a Best Practices article.
+\end{itemize}
 
@@
 
-\subsection{Estimators for free energy differences}
-\label{subsec:estimators}
-Free energy differences between two different states differing in the energy function are directly related to the
-ratio of probabilities of those states.
-As can be noted, the partition functions in Eq.~\ref{eq:conf_probability} are simply the total accumulated probabilities for all possible configurations of the system.
-Virtually all of the ways to estimate this free energy are based in converting this ratio of integrals to something that can be measured in one (or several) simulations.
-
-\paragraph{The Zwanzig relationship (EXP)}
-The simplest method for calculating free energy differences from simulations is the so-called \textit{Zwanzig
-relationship}~\cite{zwanzig1954hightemperature}, also called one-sided exponential re-weighting (EXP), or simply free energy perturbation, though this final term is sometimes used to encompass all ways of calculating free energy differences.
-
-The (reduced) free energy difference $\Delta f_{01}$ between an initial state 0 and a final state 1 defined by two different potential energy functions 
-$u_0(\vec{q})$ and $u_1(\vec{q})$ over coordinate space $\vec{q}$ can be calculated as:
-\begin{eqnarray}
-\Delta f_{01} & = & -\ln \expect{e^{-(u_1(\vec{q}) - u_0(\vec{q}))}}_0 =  -\ln \expect{e^{-\Delta u(\vec{q})} }_0
-\end{eqnarray}\label{eqn.zwanzig}
-and the average is over all samples from the simulation performed with $u_0$.
-In the case of NVT (canonical) sampling and assuming the masses do not change, then $u$ is simply $U/k_BT$, and $f$ is $F/k_BT$, but it can be generalized to other ensembles with the proper definition of $f$ and $u$.
-Described in words, we take the samples generated during our run with the potential energy function $u_0(\vec{q})$ and calculate what the difference in energy would be if we switched instantaneously to the potential energy function $u_1(\vec{q})$, and average the exponential of the negative energy difference to get the negative of the exponential of the free energy difference.
-The original distributions, P($u_0$) as generated at $\vec{\lambda}=0$ and P($u_1$) would look like those seen in Fig.~\ref{fig:fig_sampling_scheme}\textbf{A}-\textbf{C} on the right hand side.
-Reevaluating requires almost no extra code functionality to perform; one need only to save a full precision trajectory, and run an unmodified molecular simulation code using the $u_1$ in order to calculate the new energies of stored snapshots.
-The analysis can be written in a line of code.
-We note that this method is even more general, in that the instantaneous work to change the potential energy function from $u_0$ to $u_1$ can be replaced by the non-reversible work $W$ to make the same change beginning from the same equilibrium conditions at either end state~\cite{jarzynski1997nonequilibrium,jarzynski1998equilibrium,crooks2000pathensemble}, allowing for non-equilibrium free energy calculations, an alternate approach.
-We do not detail non-equilibrium transformations here, and refer the reader to more advanced treatments~\cite{maragakis2008bayesian,oberhofer2005biased,procacci2015unbiased,shirts2003equilibrium,ytreberg2004singleensemble, gapsys2020large}, as our focus here is on equilibrium free energy techniques.
-
-Although the Zwanzig equation is formally correct as long as the two states considered sample the same phase space volume, which is true for standard molecular models, it has some very important numerical issues that mean that it often performs badly for standard free energy calculations, even for small molecules~\cite{shirts2005comparison,lu2003appropriate}.
-One can show that if the standard deviation of the difference $\Delta u(\vec{q}) = u_1(\vec{q})-u_2(\vec{q})$ over all sampled $\vec{q}$ is large, which in this case, means only several times $k_BT$, then very few samples contribute to the average, and the answer will be both biased and extremely noisy~\cite{lelievre2010free}.
-Essentially, the method is dominated by contributions of rare snapshots~\cite{jarzynski2006rare, wu2005phasespace, wu2005phasespacea}.
-
-\paragraph{The Bennett Acceptance Ratio (BAR)}
-If we have the differences in the potential energy sampled from the distribution defined by $u_0$ to the state defined by $u_1$, and we also have the differences in potential energies from the distribution sampled by $u_1$ to the state defined by $u_0$, we can obtain a significantly improved estimate of the
-free energy difference compared to that obtained by EXP. 
-This estimate was first derived by Bennett and is hence generally called the Bennett Acceptance Ratio (BAR).
-It is solved by finding the reduced free energy $f_{ij}$ that satisfied the following implicit equation:
-\begin{eqnarray}
- \sum_{i=1}^{n_i} \frac{1}{1 + \exp[\ln(\frac{n_i}{n_j}) + u_{ij}(\vec{q}) - f_{ij})
- ]} \nonumber \\
- =\sum_{i=1}^{n_j} \frac{1}{1 + \exp[\ln(\frac{n_i}{n_j}) - u_{ij}(\vec{q}) + f_{ij})]},
-\end{eqnarray}
-where $n_i$ and $n_j$ are the number of samples from each state.
-More recent derivations show that this formula is the maximum likelihood estimate of the free energy difference given sets of samples from the two states~\cite{shirts2003equilibrium}.
-
-Many studies have demonstrated both the theoretical and practical superiority of BAR over EXP in molecular
-simulations~\cite{shirts2005comparison,lu2003appropriate}, and BAR converges to EXP in the limit that all samples are from a single state~\cite{bennett1976efficient,bennett1976efficient,shirts2003equilibrium}.
-BAR also requires significantly less overlap between the configurational space of each state to converge than EXP, though some overlap must still exist.
-
-The Bennett acceptance ratio is only defined between two states.
-Usually, the endpoints of interest in a free energy calculation are sufficiently different that we will need a chain of states that gradually change the potential energy function from $u_0$ to $u_1$, as discussed in Sec.~\ref{sec:important_path}.
-You can carry out a BAR estimate between each pair of states $\Delta f_{1 \rightarrow N} = \Delta {f_{1\rightarrow 2}} + \Delta {f_{2\rightarrow 3}} +  \ldots + \Delta f_{N-1\rightarrow N}$.
-
-There is one important thing to note about the uncertainty estimates when summing multiple free energies together to calculate an overall free energy estimate.
-Although BAR itself gives a free energy estimate that is asymptotically correct and is much less biased than the uncertainty estimate for EXP, the uncertainties in $\Delta {f_{i-1\rightarrow i}}$ and $\Delta {f_{i\rightarrow i+1}}$ are not uncorrelated, because they both involve the energies $u_i(\vec{q})$.
-The variances of each of the free energies will \textit{not} propagate as variances usually do (in quadrature) into the variance of the overall free energy.
-Instead, some other method for propagating the uncertainty, such as bootstrapping~\cite{grossfield2018best} must be used.
-
-\paragraph{Thermodynamic integration (TI)}
-By taking the derivative of the free energy with respect to the
-variable $\vec{\lambda}$, we find that:
-\begin{equation}
-\frac{df}{d\vec{\lambda}} = \frac{d}{d\vec{\lambda}} \left[-\ln \int \frac{\exp{-u(\vec{\lambda},\vec{q})}}{Z(\vec{\lambda})} d\vec{q}\right] = \expect{\frac{du(\vec{\lambda},\vec{q})}{d\vec{\lambda}}\
-}_{\vec{\lambda}} .
-\end{equation}
-And then we can numerically integrate $df/d\vec{\lambda}$ over an alchemical transformation, using a range of different well-established techniques, to obtain:
-\begin{equation}
-\Delta f    = \int_{0}^{1} \expect{\frac{du(\vec{\lambda},\vec{q})}{d\vec{\lambda}}}_{\vec{\lambda}}  d\vec{\lambda}.    
-\end{equation}
-This approach to calculating the free energy is called thermodynamic integration (TI).
-Averaging over $\expect{\frac{du}{d\vec{\lambda}}}$ requires fewer uncorrelated samples to reach a given level of relative error
-than averaging $e^{-u(\vec{q})}$, as the distribution of values is usually narrower, with a more Gaussian shape to the distribution.
-Rather than being limited by overlap, as in the case of BAR and MBAR (see below), we are instead limited by the bias in the numerical quadrature, which must be minimized sufficiently to be beneath the level of statistical noise.
-
-Various numerical integration schemes are possible, but the trapezoid
-rule provides a simple and robust scheme. All types of numerical integration can be written as:
-\[ \Delta f \approx \sum_{k=1}^{K} w_k
-\expect{\frac{du(\vec{\lambda},\vec{q})}{d\vec{\lambda}}}_{k}, \] where the weights $w_k$ correspond to a particular choice of numerical integration.
-Researchers have tried a large number of different integration schemes~\cite{resat1993studies,jorge2010effect,shyu2009reducing}.
-However, many integration choices require specific choices of $\vec{\lambda}$ to minimize bias, which makes them unsuitable when the intermediates
-have widely-varying levels of uncertainty.
-For example, integrating a cubic spline interpolation provided negligible benefits over a simple trapezoid rule~\cite{paliwal2011benchmark}.
-As fitting to higher order polynomials can have numerical instabilities for some energy functions, and because alternate functional forms might only be appropriate with some types of transformations, expertise and experience is required to perform such numerical integration modifications.
-For starting researchers, we therefore recommend the simple trapezoid rule scheme, as it allows for maximal flexibility in which values of $\vec{\lambda}$ are simulated.
-In practice, adding 2-3 more intermediate states is typically sufficient to match the performance of these more complicated numerical quadrature schemes.
-It is also possible to calculate the $\lambda$-derivatives at non-simulated states from simulated states in a scheme named extended TI~\cite{ruiter2016extended} which reduces the integration error.
-
-One drawback of TI is that it requires derivatives with respect to $\vec{\lambda}$ to be calculated directly in the code.
-Unfortunately, many problems of interest require using pathways (such as the soft-core pathways, for removing repulsive interactions) that are not linear, as we discuss, making this more complex.
-Still, if the code of interest does compute $\frac{du}{d\vec{\lambda}}$, then TI is perhaps the simplest method to use, as it involves a very little post-processing effort.
-
-\paragraph{The multistate Bennett acceptance ratio (MBAR)}
-One can generalize Bennett's logic from two states to multiple states to obtain a free energy estimator that uses energy differences between configurations at all intermediate states to compute free energy differences between all states.
-MBAR gives a system of implicit equations for the free energies $f_i$:
-\begin{equation}
-f_i = - \ln \sum_{n=1}^{N} \frac{\exp(-u_i(\vec{q}_n))}{\sum_{k=1}^K N_k \exp(f_k-u_k(\vec{q}_n))},
-\end{equation}
-where there are $N_k$ samples from each of $K$ states, with $\sum_k N_k=N$ the total number of samples.
-Thus, we need to evaluate the energy function $u_i$ for all samples obtained at all states in the transformation.
-The equations can be solved by a number of different standard routines.
-We note that there are only $K-1$ independent equations, so only $K-1$ of the free energies are independent variables, and one of the $f_i$ must be specified (usually, without loss of generality, setting it to zero).
-
-MBAR is probably the lowest variance asymptotically unbiased estimator of the free energy given the energies of the samples~\cite{tan2004likelihood}, which means that BAR is also the lowest variance estimator for the free energy difference between only 2 states, as it is mathematically exactly the same as MBAR in this case.
-MBAR also provides an uncertainty estimate, derived from standard error propagation methods for implicit functions, which has been shown to be highly accurate as long as there are sufficient samples at each state~\cite{paliwal2011benchmark}.
-
-MBAR can also be thought of as the Zwanzig estimator of the free energy to state $i$ where the sampled distribution is the \textit{mixture distribution} of all the other samples thrown together in one ``pot'', defined by $p_m(\vec{q}) = N^{-1} \sum_k N_k \exp(f_k-u_k\vec{q})$, which is the weighted average of all the individual normalized probability distributions from the simulations that are performed~\cite{shirts2017reweighting}.
-
-\paragraph{Recommendations}
-\begin{itemize}
-\item We recommend MBAR if all energy differences are available. It is the lowest variance unbiased free energy estimate given samples from multiple states.
-\item BAR is essentially just as good as MBAR for highly optimized $\vec{\lambda}$ intermediates.
-Specifically, if the $\vec{\lambda}$s are chosen such that intermediate states have moderate overlap with their neighbors (i.e.
-between $i$ and $i+1$ and between $i$ and $i-1$, they will \textit{not} have significant overlap with their next nearest neighbors $i+2$ and $i-2$.
-Thus MBAR does not actually get significant information from these energy differences, so one might as well not even calculate them, and just perform BAR between nearest neighbors.~\cite{paliwal2011benchmark}
-\item TI usually gives similar values as MBAR implemented with sufficient numbers of intermediates, but quadrature errors that are hard to estimate beforehand  can occur if one is not careful.~\cite{paliwal2011benchmark}
-\item WHAM is an approximation to MBAR, and there are no compelling reasons it should be used.
-If careful, it is not necessarily much worse than the other methods, but it always introduces some degree if binning error.
-\item Other variants, especially ones that adaptively determine the free energies can be useful in certain circumstances but beyond the scope of a Best Practices article.
-\end{itemize}
-
 \subsection{Uncertainty estimation}
+\tocsubsectioncomment{To be split between single/ multiple papers. @Agastya P Bhati/ Peter Coveney to expand. Cycle closure etc. to be moved to second paper.}
+\tocsubsectioncomment{@Volunteer: Add discussion of bootstrapping vs t-based confidence intervals for repeats of single calculations (both bad options)?}
 \label{subsec:uncertainty}
@@
 Different approaches are described in Sec.~\ref{sec:simulation_protocol_choice} and more details can be found in the literature~\cite{dakka2018concurrent, hahn2019alchemical}.
-
-\paragraph{Cycle closure error}
-Relative free energy calculations, which compute the change in free energy on making a change to a molecule (e.g.
-adding a functional group to a ligand) may provide an additional opportunity for error/consistency checking.
-Particularly, such calculations are often done to span a graph or tree of free energy calculations~\cite{xu2019optimal,wang2013modeling,liu2013lead}.
-In some cases the free energy change to go between molecules A and B can be obtained via multiple transformation pathways.
-This allows a type of consistency checking where we assess how much the free energy change for that transformation in practice differs from equivalence.
-
-Significant deviations of agreement from the same transformation by different routes  typically indicate insufficient configurational sampling along the lambda schedule of one or more of the transformations involved.
-This approach may be generalised to sets of connected transformations given the requirement that the sum of free energy changes along edges of a closed cycle should be zero.
-This analysis is called ``cycle closure''.
-In practice, such thermodynamic cycles do not actually sum to zero, and deviations become increasingly large as the size of the cycle increases owing to propagation of error.
-Though no firm guidelines have emerged, it may be judicious to perform additional configurational sampling along edges of a network that are involved in cycles closing poorly.
-This may be done by extending the duration of simulations, or by averaging free energy changes over multiple repeats.
-The latter approach may yield more reproducible free energy changes, but at the expense of a stronger bias on the estimated free energies due to repeated use of the same input coordinates.
-
-A scheme to reduce cycle closure errors is used in FEP+ whereby calculated free energy changes along the nodes of the network are re-sampled assuming estimates of the calculated free energy change along a node may be obtained from a Gaussian distribution centered on the estimated free energy change and with a standard deviation equal to the estimated standard deviation of the free energy change.
-The procedure then uses a maximum likelihood method to find new sets of free energy changes that minimize cycle closure errors~\cite{wang2013modeling}.
-An alternative approach computes the free energy change between a target and reference compound as a weighted average over all unique paths in the network, with the weights derived from the propagated uncertainties of each node~\cite{mey2016blinded}.
-Approaches as illustrated by Yang et al.
-for perturbation map design can also be used to compute relative free energies between target and reference compounds~\cite{yang2020optimal}.
 
@@
 
-\subsection{Best practices for reporting data }
-\label{sec:plot_data}
-Following best practices for data generation and their analysis does not mean that data is reported in the optimal way.
-As a practitioner of alchemical free energy simulations you also should use best practices for reporting and plotting your results.
-We encourage the following standard set of analyses and ways to represent data.
-\paragraph{Statistics to include in reporting data}
-As with any modelling technique, misuse of statistical analysis can skew the perception of how well models perform in free energy predictions.
-First, error estimates should always be included on your predictions in whatever form you present your data (scatterplots, barplots, etc; see next paragraph).
-We recommend performing triplicates of your predictions at minimum, with starting points that are expected to be uncorrelated, to ensure some measure of reliability in your data.
-This replication may seem excessive, but uncertainty estimates often underestimate the true statistical uncertainty.
-Where performing multiple replicas of the simulation is not possible, an error estimate from e.g.
-MBAR can be used, though bearing in mind this is likely an underestimated error.
-
-As alchemical free energy methods are used in drug discovery to quantify and rationalise structure activity relationships (SAR), the models ability to (a) correlate well with experiment and (b) rank-order the molecules by affinity, should both be computed.
-Conventionally, this means including an R\textsuperscript{2} (or Pearson's R), where $R=+1$ means high correlation, $R=0$ means no correlation, and $R=-1$ means high anti-correlation) and a Kendall \texttau{} (with perfect ranking agreement when \texttau=1 and perfect disagreement when \texttau=-1) metric in your results.
-Additionally, practitioners may choose to include a Spearman \textrho{} as well.
-Brown et al.~\cite{brown2009healthy} have provided a useful analysis in terms of upper bounds of expected possible correlations between experiment and computation with a given potency range for the compounds.
-For example, for potency ranges of 2 log units it would be impossible to get a higher correlation in R than 0.8 because of experimental uncertainties~\cite{brown2009healthy}.
-What often is neglected to include is an error analysis on correlation statistics that arise from the errors of both experimental and computed data.
-One way to include such error analysis for correlation metrics is using bootstrapping on the datasets.
-The D3R community challenges follows best practices on their data evaluation with readily available python scripts online~\cite{2018drugdata}, based on work by Pat Walters~\cite{walters2013what}.
-Other analysis software also provide similar functionality for bootstrapping datasets~\cite{antonia2019michellab}.
-
-Mean unsigned error (MUE, also called mean absolute error/MAE) is another key statistic to include in your results.
-Even though some models' near-perfect correlation and ranking statistics might suggest excellent accuracy, MUE values can still have errors of multiple kcal/mol, providing important additional insight into performance.
-Furthermore, MUE allows for unbiased comparisons between predictive models as it is less sensitive to dataset size.
-Other metrics such as Gaussian Random Affinity Model (GRAM)~\cite{cui2020gram}, Predictive Interval (PI) and Relative Absolute Error (RAE), attempt to correct for the inherent potency range of a dataset, which can aid in comparing success between different targets.
-We recommend further reading on evaluation of computational models~\cite{jain2008recommendations, walters2013what, brown2009healthy, walterthoughts}.
-
-Reporting the results of relative free energy calculations requires care.
-As shown in Fig.~\ref{fig:fig_types_of_networks}, relative free energies can be performed arbitrarily as a forward or a reverse process, and thus relative free energies may be reported as either positively or negatively valued.
-The consequence of the two possible signs for relative free energies is that correlation statistics (such as Pearson's R and Kendall \texttau{}) can be skewed depending on which sign is analysed.
-The issue of this inconsistency can be circumvented by either plotting all datapoints within a consistent quadrant~\cite{perez-benito2019predicting}, or by avoiding the use of correlation statistics for assessment of relative free energy calculations and instead measuring accuracy using RMSE and MUE, which are unaffected by choice of sign.
-
-\paragraph{Presenting your data}
-As essentially all alchemical free energy prediction schemes are regression problems, the preferred type of plot is a scatter plot (see Fig.~\ref{fig:scatterplot_analysis}).
-Most alchemical free energy projects will look at 10-50 ligands.
-Any study with \textless10 ligands is more suitable for bar plots (with inclusion of error bars), and is unlikely to provide meaningful statistics.
-Any study with \textgreater50 ligands typically contains multiple protein targets to which alchemical free energies may perform better on some targets than others.
-Because of this, it is bad practice to place multiple datasets on the same plot as this can suggest high model accuracy even though the individual models perform less well~\cite{walterthoughts}.
-
-\begin{figure}
-  \includegraphics[width=0.95\linewidth]{figures/fig13_analysis_practices/Figure.png}
-  \caption{\textbf{An example of recommended practices for graphing alchemical free energy predictions.} This figure shows the relation between predicted and experimentally-determined Gibbs free energy in kcal/mol with standard errors as error bars.
-The dark and light-orange regions depict the 1- and 2-kcal/mol confidence bounds.
-Statistical metrics for the data are reported, with 95\% confidence intervals determined by bootstrapping analysis.
-Extra care should be taken when investigating potential outliers further.}
-   \label{fig:scatterplot_analysis}
-\end{figure}
-
-As we are interested mainly in the linear relationship between the alchemical free energy predictions and the experimentally-determined affinity values, plots should be depicted with the same range on both axes (i.e.
-$x=y$) with a 1:1 aspect ratio, with units for both experiment and simulation converted to be the same.
-If this skews the plot to a point where it is difficult to read of information, using the same dimensions, such that e.g.
-1 cm is 1 kcal/mol is acceptable.
-Furthermore, bounds should be depicted for the 1- and 2-kcal/mol confidence regions.
-These regions can serve as tools to communicate your model performance: any predictions inside the 1 kcal/mol region can be seen as highly reliable, any predictions inside the 2 kcal/mol region should be seen as somewhat reliable, and any predictions outside the confidence regions should be expected to be unreliable and handled as outliers.
-In a drug discovery context, this type of data depiction may suggest the reliability of alchemical FE predictions in the project, and can give an idea of how trustworthy predictions can be for synthesis ideas.
-It is also recommended to included experimental error bars in all plots.
-
-An example of a best practice scatter comparison between computed and experimental values is shown in Fig.~\ref{fig:scatterplot_analysis}, highlighting outliers, error bars and confidence intervals.
-The data for this plot is artificially generated for illustration purposes.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusions}
+\section{\tocred{Unsettled Practices and Current Limitations}}
+\tocsectioncomment{@Everyone to contribute to section on unsettled practices/ current limitiations, as suggested by John Chodera.}
+\label{sec:unsettled}
+
+\subsection{Unsettled practices}
+
+\subsection{Current limitations}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Conclusions                 %
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{\tocorange{Conclusions}}
+\tocsubsectioncomment{@Volunteer to update reflecting focus on single calculations.}
 \label{sec:conclusion}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Selection of available software packages}
+\section{\tocorange{Selection of available software packages}}
+\tocsectioncomment{@Volunteers to update.}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%  Benchmark information  %
-%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Alchemical free energy datasets: an overview}
-\label{sec:benchmark}
-The following contains a non-exhaustive summary of alchemical free energy datasets that can serve as a starting point to review approaches or test new implementations.
-The field is moving towards a more standardised way of generating protein-ligand benchmark datasets and the progress of these efforts can be tracked here: \url{https://github.com/openforcefield/FE-Benchmarks-Best-Practices}.
-Currently lacking an exhaustive set of benchmark datasets, the review by Williams-Noonan et al.~\cite{williams-noonan2018free} contains an overview of recently published alchemical free energy studies.
-For comparison of FEP+ and Gromacs (using the AMBER99SB-ILDN and GAFF2 force field), cf.
-the recently published study by Pérez-Benito et al.~\cite{perez-benito2019predicting}.
-An overview of further suggested benchmark sets can be found in the review by Mobley and Gilson~\cite{mobley2017predicting} or on \url{alchemistry.org}~\cite{alchemistry}.
-These include cyclodextrins, the Cytochrome C peroxidase (CCP) protein model binding site, thrombin and bromodomains as well as solvation benchmark sets~\cite{paliwal2011benchmark}.
-Please refer to table~\ref{tab:benchmarks}, for a small overview of datasets, what forcefields they used, and what the original study was it came from.
-
-\begin{table}
-\caption{Selection of example datasets}
-\begin{tabular}{cccc}
-\textbf{Publication} & \textbf{Targets} & \textbf{Ligands} & \textbf{Force Field} \\
-\hline
-\multicolumn{4}{|c|}{D3R Grand Challenges~\cite{D3R}} \\
-\hline
-GC3~\cite{gaieb2019d3r} & 6 & 266 & various \\
-GC2~\cite{gaieb2018d3r} & 1 & 102 & various \\
-GC2015~\cite{gathiaka2016d3r} & 2 & 215 & various \\
-\hline
-\multicolumn{4}{|c|}{SAMPL Challenges~\cite{SAMPL}} \\
-\hline
-SAMPL6~\cite{rizzi2018overview} & 3 & 21 & various \\
-SAMPL5~\cite{yin2017overview} & 3 & 22 & various \\
-SAMPL4~\cite{muddana2014sampl4} & 2 & 23 & various \\
-\hline
-\multicolumn{4}{|c|}{Schrödinger Datasets} \\
-\hline
-FEP+ Dataset~\cite{wang2015accurate} & 8 & 199 & OPLS2.1 \\
-FEP+ Dataset~\cite{harder2016opls3} & 8 & 199 & OPLS3 \\
-FEP+ Dataset~\cite{roos2019opls3e} & 8 & 199 & OPLS3e \\
-FEP+ Dataset~\cite{song2019using} & 8 & 199 & GAFF 1.8 \\
-FEP+ Dataset~\cite{gapsys2020large} & 8 & 199 & various \\
-FEP+ Dataset~\cite{kuhn2020assessment} & 8 & 199 & GAFF2.1 \\
-Fragments ~\cite{steinbrecher2015accurate} & 8 & 96 & OPLS2.1 \\
-Scaffold Hopping~\cite{wang2017accurate} & 6 & 21 & OPLS3 \\
-Scaffold Hopping~\cite{kuhn2020assessment} & 6 & 21 & GAFF2.1 \\
-Macrocycles~\cite{yu2017accurate} & 7 & 33 & OPLS3 \\
-\hline
-\multicolumn{4}{|c|}{Further Suggested Datasets} \\
-\hline
-Cucurbit[7]uril (CB7)~\cite{mobley2017predicting} & 1 & 15 & NA \\
-Deep cavity cavitand~\cite{mobley2017predicting} & 2 & 19 & NA \\
-T4 Lysozyme~\cite{mobley2017predicting} & 2 & 20 & NA \\
-Merck set~\cite{MCompChem2019Sep} & 5 & 169 & OPSL3 \\
-
-\hline
-\label{tab:benchmarks}
-\end{tabular}
-\end{table}
-
-\section{Checklist}
+\section{\tocorange{Checklist}}
+\tocsectioncomment{@Volunteer to update checklists/ move relevant sections to second paper.}
 \label{sec:checklist}

### 2024-12-13 paper/manuscript.tex
Add steering group feedback on proposed structure
Final changes before general contributions are sought.

@@
 Possibly add a walkthrough of a ``dummy calculation'' (which could be a particular realisation of a general workflow laid out in @Jenke Scheen's figure)?
-This could lead on to discussion of particular methods.}
+This could lead on to discussion of particular methods.
+@Toni Mey I could do this}
 \label{sec:step0}
@@
 \tocsectioncomment{@Irfan Alibay/ @Emilio Gallicchio/ @Volunteer to expand on approaches for explicitly accounting for protonation/ tautomer changes.}
-\tocsectioncomment{@Hannah Bruce Macdonald/ @David Slowchower/ @Volunteer to expand range of systems discussed, e.g. membrane proteins, cofactors, multimers, PPIs.}
+\tocsectioncomment{@Hannah Bruce Macdonald/ @David Slowchower/ @Vytas Gapsys/ @Volunteer to expand range of systems discussed, e.g. membrane proteins, cofactors, multimers, PPIs.}
 \label{sec:prerequisites}
@@
 \tocsectioncomment{@Sheenam Khuttan / @Emilio Gallicchio to contribute section on ATM.}
-\tocsectioncomment{@Volunteer to add best practices for non-equilibrium calculations.}
+\tocsectioncomment{@Vytas Gapsys / @Volunteer to add best practices for non-equilibrium calculations.}
 \label{sec:simulation_protocol_choice}

### 2024-12-26 paper/manuscript.tex
updated theory section

@@
 
-The original best practices guide will be split into two documents: the first will gives best practices for running a \emp{single} calculation (and cover theory in more depth/ detail system-specific setup etc.), while the second will give best practices for running \emp{multiple} calculations (with a focus on automation).
+The original best practices guide will be split into two documents: the first will gives best practices for running a \emph{single} calculation (and cover theory in more depth/ detail system-specific setup etc.), while the second will give best practices for running \emph{multiple} calculations (with a focus on automation).
 The first will become version 2.0 of the current guide, while the second will be submitted as a new manuscript.
@@
 \section{\tocorange{Statistical mechanics demonstrates why alchemical free energy calculations work}}
+%\section{\tocorange{Statistical mechanics theory of solvation, solvent partitioning, and binding}}
 \tocsectioncomment{Steering Meeting: @Emilio Gallicchio to update theory.
@@
 
-Why would you want to run an alchemical free energy calculation and why do they work?
-In this section, we use the example of relative free energy calculations to sketch the theory of alchemical simulations and illustrate their utility.
+In this section, we use a statistical mechanics theory of dilute solutions to derive free energy expressions that form the basis of alchemical computational protocols to model non-covalent molecular binding, solvation, and solvent partitioning equilibria.
 The emphasis here is placed on bridging theoretical foundations and intuition.
-A rigorous derivation of the standard (absolute) free energy of binding using the principles of statistical mechanics can be found in Gilson's classic work~\cite{gilson1997statisticalthermodynamic}.
-
-\subsection{Simulating binding events of receptor-drug systems can be computationally expensive}
-Suppose you want to compute the binding affinity, or free energy of binding, of a ligand $L$ to a receptor $R$, given by:
-\begin{equation}
-R+L \leftrightharpoons RL.
-\end{equation}
-The binding constant ($K_b^{\circ}$) is given by the law of mass action as the ratio of concentrations of product $[RL]$ and reactants $[R]$, $[L]$:
+Since chemical equilibria are regulated by the chemical potentials of the species involved, we will start by considering a statistical mechanics expression of the chemical potential of a solute in a solvent.
+We will then combine chemical potentials to develop models for each process.
+
+\subsection{Chemical potential}
+
+The chemical potential of a solute $u$ in an ideally diluted solution in a solvent $v$ can be written as~\cite{guggenheim1952mixtures504,gilson1997statisticalthermodynamic,gilson2007calculation}
+\begin{equation}
+  \mu_u = -k_B T \ln \frac{q'_u}{C_u \Lambda_u^3}
+  \label{eq:chemical-potential-definition}
+\end{equation}
+where $k_B$ is Boltzmann's constant, $T$ is the absolute temperature, $C_u = N_u/V$ is the number concentration of the solute, $\Lambda_u = h/\sqrt{2 \pi M_u k_B T}$ is the thermal De-Broglie wavelength of the solute, where $h$ is Planck's constant and $M_u$ is the solute's total mass.
+The quantity $q'_u$ in Eq.~(\ref{eq:chemical-potential-definition}) is the molecular partition function of the solute at rest in the solvent
+\begin{equation}
+  q'_u = \frac{\Lambda_u^3}{\prod_i \lambda_i^3} 8 \pi^2 z_u
+  \label{eq:intra-qpu-def}
+\end{equation}
+where $\beta = 1/(k_B T)$, $\lambda_i$ is the thermal De Broglie wavelength of the $i$-th atom of the solute, and $z_u$ is the intramolecular configurational partition function of the solute
+\begin{equation}
+  z_u = \int dx_u e^{-\beta \Psi_v(x_u)}
+  \label{eq:intra-zu-def}
+\end{equation}
+where $x_u$ represents the collection of internal degrees of freedom of the solute (bond lengths, bond angles, and dihedral angles) with the corresponding volume element $dx_u$ assumed to include the appropriate Jacobian factors.
+The function $\Psi_v(x_u)$ in Eq.~\ref{eq:intra-zu-def} is the solvent-averaged potential of mean force of the solute in conformation $x_u$ in solvent $v$\cite{rouxandsimonson1999solventmodels}
+\begin{equation}
+  e^{-\beta \Psi_v(x_u)} = \frac{1}{Z_{N_v}} \int d\vec{r}_v e^{-\beta U(x_u, \vec{r}_v)}
+  \label{eq:solvent-pmf-def}
+\end{equation}
+where $\vec{r}_v$ represents the collection of Cartesian coordinates of the atoms of $N_v$ solvent molecules, $U(x_u, \vec{r}_v)$ is the potential energy of the solvent with one solute molecule in conformation $x_u$ placed in an arbitrary position in a volume $V$.
+When the potential energy function of the solution is separable into an intramolecular potential energy of the solute $U(x_u)$ and a solute-solvent non-bonded interaction energy, the potential of mean force can be written as the sum $\Psi_u(x_u) = U(x_u) + W_v(x_u)$ where $W_u(x_u)$ is the solvent potential of mean force, the solvation free energy of the solute fixed in configuration $x_u$.\cite{gilson2007calculation} However, the potential of mean force definition Eq.~(\ref{eq:solvent-pmf-def}) is valid in general.
+Finally, $Z_{N_v}$ is the configurational partition function of the pure solvent
+\begin{equation}
+  Z_{N_v} = \int d\vec{r}_v e^{-\beta U(\vec{r}_v)} \, .
+  \label{eq:Z-pure-solvent}
+\end{equation}
+
+A few notes are in order.
+First, Eq.~(\ref{eq:chemical-potential-definition}) is derived using classical statistical mechanics theory.~\cite{hill1986statthermo} Classical mechanics is considered an excellent approximation for the room temperature processes studied here involving only changes in intermolecular interactions.
+A quantum mechanical treatment is necessary for processes not covered here, such as covalent binding, that involve breaking and forming chemical bonds.
+Classically, equilibrium thermodynamics properties, such as gas solubilities and binding constants, do not depend on atomic masses.
+Hence, while they are included in Eqs.~\ref{eq:chemical-potential-definition} and Eq.~\ref{eq:intra-qpu-def} for completeness, the thermal De Broglie wavelength terms coming from the partition functions of the momenta cancel in the resulting expressions below and we will no longer consider them.
+
+Eq.~(\ref{eq:chemical-potential-definition}) is derived by writing the classical canonical configurational partition function $Z(N_u = 1, N_v)$ of one solute molecule in a solvent of $N_v$ molecules in a volume $V$ and multiplying and dividing it by configurational partition function of the pure solvent [Eq.~(\ref{eq:Z-pure-solvent})], and separating out the integration over the external degrees of freedom of the solute (translations and rotations) giving an $8 \pi^2 V$ term to yield $Z(N_u = 1, N_v) = Z_{N_v} 8 \pi^2 V z_u$ where $z_u$ is defined by Eqs.~(\ref{eq:intra-zu-def}) and (\ref{eq:solvent-pmf-def}).
+The canonical configurational partition function of $N_u$ independent solute molecules in a solvent is then obtained by considering the product of the solute terms $Z(N_u, N_v) = Z_{N_v} (8 \pi^2 V z_u)^{N_u}/N_u!$,~\cite{guggenheim1952mixtures504,simonson2016physical} where the factorial term takes care of indistinguishability.
+Finally, the canonical partition function $Q(N_u, N_v)$ of the solution is obtained by including the appropriate terms coming from the momenta and differentiating the resulting Helmholtz free energy $A(N_u, N_v) = -k_B T \ln Q(N_u, Nv)$ with respect to $N_u$ to obtain Eq.~(\ref{eq:chemical-potential-definition}).
+
+The form Eq.~(\ref{eq:chemical-potential-definition}) for the solute chemical potential using the potential of mean force description is useful, as we will see, because it separates the solute's translational motion, which leads to the concentration dependence, from the intramolecular contributions.
+The potential of mean force description does not introduce approximations.
+With the exception of the use of classical statistical mechanics and the assumption of sufficient dilution so that the solute molecules can be considered independent, Eq.~(\ref{eq:chemical-potential-definition}) is rigorous.
+Due to rotation-vibration couplings, the separation of the solute's overall rotational motion leading to the $8 \pi^2$ term in Eq.~(\ref{eq:intra-qpu-def}) is not exact.
+However, it is not a strict requirement and the theory can be developed without it.
+The separation of rotational degrees of freedom is considered an excellent approximation at room temperature and a useful simplification, so it is adopted here.
+
+We derived Eq.~(\ref{eq:chemical-potential-definition}) using the canonical (constant volume) partition function of the solution.
+Since the chemical potential can be equivalently obtained from the constant-volume differentiation of the Helmholtz free energy or the constant-pressure differentiation of the Gibbs free energy, Eq.~(\ref{eq:chemical-potential-definition}) is correct.
+However, the volume dependence of the chemical potential has to be considered when modeling the Gibbs free energy change for constant-pressure processes where the system's volume changes significantly.~\cite{gilson1997statisticalthermodynamic} Terms due to volume variations are generally considered small for most processes of solvation, solvent partitioning, and molecular binding at atmospheric pressure and are often ignored in alchemical calculations.
+Hence, we will use Eq.~(\ref{eq:chemical-potential-definition}) to derive expressions for $\Delta G$'s.
+
+The standard chemical potential of a solute in an ideally dilute solution is obtained from Eq.~(\ref{eq:chemical-potential-definition}) by setting the concentration to the standard concentration $C^\circ$, generally $1$ mol/L equivalent to $1/1,661$ particles/\AA$^{3}$,
+\begin{equation}
+  \mu_u^\circ = -k_B T \ln \frac{q'_u}{C^\circ \Lambda_u^3} \, .
+  \label{eq:chemical-potential-standard}
+\end{equation}
+The standard state state of a dilute solution is defined as an ideally dilute solution at concentration $C^\circ$.~\cite{levine2009physicalchemistrybook6ed} Hence, Eq.~(\ref{eq:chemical-potential-standard}) is formally correct for real solutions as well.
+
+\subsection{Solvation}
+
+The standard molar Gibbs free energy of the solvation of a gas species $A$ in a solvent $v$
+\begin{equation}
+A(\mathrm{g})  \leftrightharpoons A(v)
+\end{equation}
+is
+\begin{equation}
+  \Delta G^\circ_{\mathrm{slv}} = \mu^\circ_{A(v)}  - \mu^\circ_{A(g)} = -k_B T \ln \frac{P^\circ}{C^\circ k_B T} - k_B T \ln \frac{z_{A(v)}}{z_{A(\mathrm{g})}}
+  \label{eq:DG0-solvation-def}
+\end{equation}
+where we used Eq.~(\ref{eq:chemical-potential-standard}) for the solution and  Eq.~(\ref{eq:chemical-potential-definition}) for the gas (assumed ideal) at the standard gas concentration $C = P^\circ/k_B T$ where $P^\circ = 1$ bar is the standard pressure, and Eq.~(\ref{eq:intra-qpu-def}) noting that all momentum terms cancel.
+The first term in Eq.~(\ref{eq:DG0-solvation-def}) is an ideal term $\Delta G^\circ_{\mathrm{slv,ideal}}$ that accounts for the difference of standard states for solution and gases.
+The second term is the excess solvation free energy $\Delta G_{\mathrm{slv, exc.}}$ that can be evaluated by numerical alchemical calculations (see below).
+Hence Eq.~(\ref{eq:chemical-potential-standard}) can be used to compare calculated and experimental standard free energies of solvation.
+
+However, equilibrium experimental gas solvation data is most often available in terms of Henry's constants $k_H$.
+Setting $\Delta G_{\mathrm{slv}} = 0$ when the gas partial pressure and the solute's concentrations are equal to their equilibrium values $P_A$ and $C_A$, respectively, we obtain
+\begin{equation}
+  P_A = k_H C_A
+  \label{eq:Henrys-law}
+\end{equation}
+where
+\begin{equation}
+  k_H = k_B T \frac{z_{A(\mathrm{g})}}{z_{A(v)}} \, .
+  \label{eq:Henrys-constant}
+\end{equation}
+The relation above is used to connect alchemical calculations to Henry's constant values, often after converting them to the excess free energy scale~\cite{nicholls2008predicting,mobley2014freesolv}
+\begin{equation}
+\Delta G_{\mathrm{slv, exc.}} = - k_B T \ln \frac{z_{A(v)}}{z_{A(\mathrm{g})}} = k_B T \ln \frac{k_H}{k_B T} \, .
+\label{eq:DG-solvation-excess}
+\end{equation}
+Because it involves only intramolecular partition functions, Eq.~(\ref{eq:DG-solvation-excess}) states that the excess free energy of solvation corresponds to the process of moving the solute from a fixed position and orientation in the gas to a fixed position and orientation in the solvent.
+This is sometimes referred to as the solvation free energy in the Ben-Naim standard state.~\cite{bennaim1984solvation}
+
+The next step is to express the ratio of intramolecular configurational partition functions in Eq.~(\ref{eq:DG-solvation-excess}) in an ensemble average form amenable to calculation by computer simulation.
+First, the potential of mean force in $z_{A(v)}$ [Eq.~(\ref{eq:intra-zu-def})] is replaced with its definition from Eq.~(\ref{eq:solvent-pmf-def}) to explicitly expose the coordinates of the solvent:
+\begin{equation}
+  \frac{z_{A(v)}}{z_{A(\mathrm{g})}} =
+  \frac{
+    \int dx_A d\vec{r}_v e^{-\beta U(x_A, \vec{r}_v)}
+  }{
+    \int dx_A d\vec{r}_v e^{-\beta [U(\vec{r}_v) + U(x_A)] }
+  }
+\end{equation}
+where the integral $Z_{N_v}$ for the pure solvent from Eq.~(\ref{eq:Z-pure-solvent}) has been combined in the denominator with the integral $z_{A(\mathrm{g})}$ of the solute in the gas phase to obtain the configurational partition function of an artificial system where the solute is in the solvent but it is decoupled from it (note the absence of coupling terms between the degrees of freedom of the solvent and those of the solute).
+
+Next, we multiply and divide by the integral $\int d\zeta$ of the six external degrees of freedom of the solute (the center of mass position and the orientation angles).
+These degrees of freedom, together with the internal degrees of freedom $x_A$, allow the recast the integrals in terms of the Cartesian coordinates $\vec{r}_A$ of the $n_A$ solute's atoms which are typically used in computer simulations, without restraining the position and orientation of the solute.
+Finally, we multiply and divide the integrand in the numerator by the integrand in the denominator to obtain
+\begin{equation}
+  \frac{z_{A(v)}}{z_{A(\mathrm{g})}} =
+  \frac{
+    \int d\vec{r}_A  d\vec{r}_v  e^{-\beta \{ U(\vec{r}_A, \vec{r}_v) -  [ U(\vec{r}_v) + U(\vec{r}_A) ] \}} e^{-\beta [U(\vec{r}_v) + U(  \vec{r}_A  )] } 
+  }{
+    \int d\vec{r}_A d\vec{r}_v e^{-\beta [U(\vec{r}_v) + U( \vec{r}_A  )] } 
+  } = \langle e^{-\beta \Delta U} \rangle_0
+  \label{eq:zratio-solv-average}
+\end{equation}
+where $\langle \ldots \rangle_0$ is an average in the decoupled ensemble and $\Delta U =  U(\vec{r}_A , \vec{r}_v) -  [ U(\vec{r}_v) + U(\vec{r}_A ) ]$ the change in the system's potential energy for coupling the solute to the solvent.
+Equivalently, we can consider the reciprocal of Eq.~(\ref{eq:zratio-solv-average}) and express it in terms of an ensemble average $\langle \ldots \rangle_1$ in the coupled ensemble:
+\begin{equation}
+  \frac{z_{A(v)}}{z_{A(\mathrm{g})}} = \frac{1}{\langle e^{\beta \Delta U} \rangle_1}
+\end{equation}
+where $-\Delta U = [ U(\vec{r}_v) + U(\vec{r}_A ) ] - U(\vec{r}_A , \vec{r}_v)$ is a decoupling energy.
+
+Combining Eqs.~(\ref{eq:DG-solvation-excess}) and (\ref{eq:zratio-solv-average}) yields
+\begin{equation}
+  \Delta G_{\mathrm{slv, exc.}} = - k_B T \ln \langle e^{-\beta \Delta U} \rangle_0 = k_B T \ln \langle e^{\beta \Delta U} \rangle_1
+  \label{eq:zratio-solv-dgexc}
+\end{equation}
+The objective of alchemical solvation free energy calculations is to evaluate Eq.~(\ref{eq:zratio-solv-dgexc}), by means of \emph{coupling} or \emph{decoupling} protocols.
+
+\subsection{Solvent partitioning}
+
+The standard molar Gibbs free energy of transfer of a solute species $A$ from a  solvent $s$ to a solvent $v$
+\begin{equation}
+A(s)  \leftrightharpoons A(v)
+\end{equation}
+is
+\begin{equation}
+  \Delta G^\circ_{\mathrm{part}} = -k_B T \ln P_{sv} = \mu^\circ_{A(v)}  - \mu^\circ_{A(s)} = -k_B T \ln \frac{z_{A(v)}}{z_{A(s)}}
+  \label{eq:DG0-partitioning-def}
+\end{equation}
+where $P_{sv} = C_{A(v)}/C_{A(s)}$ is the partition coefficient of $A$ for the two solvents.~\cite{rustenburg2016measuring} Eq.~(\ref{eq:DG0-partitioning-def}) 
+immediately identifies the partition coefficient with the ratio of intramolecular configurational partition functions of $A$ in the two solvents:
+\begin{equation}
+  P_{sv} = \frac{z_{A(v)}}{z_{A(s)}} \, .
+  \label{eq:partition-coefficient}
+\end{equation}
+This relation connects calculated ratios of partition functions to experimental partition coefficients.
+Similarly to Eq.~(\ref{eq:DG-solvation-excess}), Eq.~(\ref{eq:partition-coefficient}) corresponds to the process of transferring the solute from a fixed position and orientation in one solvent to a fixed position and orientation in the other.
+
+The ratio of partition functions in Eq.~(\ref{eq:partition-coefficient}) can be treated in the same way as Eq.~(\ref{eq:zratio-solv-average}) to express it as the ensemble average in solvent $s$ of the change in potential energy $\Delta U$ for moving the solute to the other solvent $v$.
+However, in practice it has been more common to evaluate Eq.~(\ref{eq:DG0-partitioning-def}) as the difference of excess solvation free energies in the two solvents separately evaluated using Eq.~(\ref{eq:zratio-solv-dgexc}).~\cite{bosisio2016blinded} Formally, this approach is derived by multiplying and dividing Eq~(\ref{eq:partition-coefficient}) by $z_{A(\mathrm{g})}$ and expressing each terms as an ensemble average in the decoupled ensemble
+\begin{equation}
+  P_{sv} = \frac{z_{A(v)}}{z_{A(\mathrm{g})}}  \frac{z_{A(\mathrm{g})}}{z_{A(s)}} = \frac{\langle e^{-\beta \Delta U_{(v)}} \rangle_0}{\langle e^{-\beta \Delta U_{(s)} } \rangle_0 }
+  \label{eq:partition-coefficient-coupling}
+\end{equation}
+where $\Delta U_{(v)}$ and $\Delta U_{(s)}$ are the solute's coupling energies to solvent $v$ and $s$, respectively. 
+The implementation of Eq.~(\ref{eq:partition-coefficient-coupling}) in terms of decoupling
+\begin{equation}
+P_{sv} = \frac{\langle e^{\beta \Delta U_{(s)}} \rangle_1}{\langle e^{\beta \Delta U_{(v)} } \rangle_1 }
+\label{eq:partition-coefficient-decoupling}
+\end{equation}
+is an example of a \emph{double-decoupling} alchemical computational protocol.~\cite{hermans1986freeenergy,gilson1997statisticalthermodynamic}
+
+\subsection{Non-covalent molecular binding}
+
+The standard Gibbs free energy of binding between a receptor $R$ and a ligand $L$ to form a complex $RL$ in a solvent $v$
+\begin{equation}
+  R(v) + L(v)  \leftrightharpoons RL(v)
+  \label{eq:binding-reaction}
+\end{equation}
+is
+\begin{equation}
+  \Delta G^\circ_{b} = \mu^\circ_{RL(v)}  - \mu^\circ_{R(v)} - \mu^\circ_{L(v)} = -k_B T \ln Kb 
+  \label{eq:DG0-binding-def}
+\end{equation}
+where~\cite{gilson1997statisticalthermodynamic,gilson2007calculation}
+\begin{equation}
+  Kb = \frac{C^\circ}{8 \pi^2} \frac{z_{RL(v)}}{z_{R(v)} z_{L(v)}}
+  \label{eq:DG0-binding-constant-def}
+\end{equation}
+is the statistical mechanics expression for the binding constant.
+Eq.~(\ref{eq:DG0-binding-constant-def}) follows from the application of  Eq.~(\ref{eq:chemical-potential-standard}) for each of the three species and noting that the contributions of the momenta cancel.
+As discussed below, various alchemical protocols exist to evaluate the ratios of partition functions in Eq.~(\ref{eq:DG0-binding-constant-def}}), providing a connection between the theory and observed binding affinities.
+
+Eq.~(\ref{eq:intra-zu-def}) describes the intramolecular configurational partition functions $z_{R(v)}$ and $z_{L(v)}$ for the receptor $R$ and ligand $L$.
+The intramolecular configurational partition function $z_{RL(v)}$ of the complex is similar but the integration is limited to a chosen range of configurations in which receptor and ligand are considered ``bound''.
+To this end, Gilson et al.~\cite{gilson1997statisticalthermodynamic} introduced an indicator function $I(\zeta)$ equal to one when the six degrees of freedom of the complex that defines the position and orientation of the ligand relative to the receptor, collectively denoted as $\zeta$, are within the specified range of the bound state and equal to zero otherwise:
+\begin{equation}
+  z_{RL(v)} = \int dx_R dx_L d\zeta I(\zeta) e^{-\beta \Psi_v(x_R, x_L, \zeta)} \, .
+  \label{eq:intra-zRL-def}
+\end{equation}
+
+The definition of the bound macrostate of the complex is an essential input of the theory.
+Unlike bond lengths, bond angles, and dihedral angles, the three degrees of freedom of the complex that correspond to the position of the ligand relative to the receptor are unbounded.
+Unless their extent is limited in some fashion, the bound state of the complex cannot be distinguished from the unbound state, and their free energy difference will be undefined.
+For example, it is reasonable to insist that the ligand be close to the receptor in configurations regarded as bound.
+The definition of the target macrostates is a requirement in many areas of computational chemistry and biophysics, such as when specifying the range of backbone angles to measure the relative population of $\alpha$-helical and $\beta$-strand conformations of peptides.
+
+Conversely, the value of the binding free energy depends on the chosen structural definition of the bound complex.
+The relationship between the predictions of this theory and experimental reporters (thermochemical, spectroscopic, inhibition kinetics, and others) of complex formation has been discussed.\cite{gilson1997statisticalthermodynamic,mihailescu2004theory,gallicchio2011recentadv,simonson2016physical,gallicchio2021comppeptsci} The interpretation of experimental reporters of binding is a complex issue in itself.
+Still, the consensus is that this theory gives nearly equivalent results when the binding is strong and specific, and the definition of the bound complex covers the major bound poses while excluding obvious unbound configurations.
+In some cases it is valuable to consider specific definitions of the complex by, for example, restricting the orientation of the ligand to pinpoint the relative contribution of different binding poses.
+
+Eq.~(\ref{eq:DG0-binding-constant-def}) is the basis of a variety of alchemical computational implementations discussed in the following sections that have have been developed to study molecular binding by alchemical means.
+
+\subsection{Simulating binding events is computationally expensive}
+
+The appeal of alchemical approaches to study binding can be best appreciated by considering the alternative of evaluating the binding constant directly from the law of mass action definition
 \begin{equation}\label{eq:law-mass-action}
- K_b^{\circ} = c^{\circ}\frac{[RL]}{[L][R]}.
-\end{equation}
-The standard state concentration $c^\circ$ depends on the reference state, but it is usually set to $1\,\mathrm{mol}/\mathrm{L}$ assuming a constant pressure of $1\,\mathrm{atm}$ (see also Sec.~\ref{sec:standardstate-restraints}).
-Eq.~\ref{eq:law-mass-action} also holds for dilute solutions if thermodynamic activities can be approximated by concentrations~\cite{gilson1997statisticalthermodynamic}.
-Thus, the standard Gibbs free energy of binding $\Delta G_{\mathrm{bind}}$ is given by:
-\begin{equation}
-    \Delta G_{\mathrm{bind},L} = -k_BT\ln K_b^{\circ},
-\end{equation}
-where $k_B$ is the Boltzmann constant and $T$ the temperature of the system.
-Note that, unless otherwise specified, we use $\Delta G$ throughout the paper to refer to the \textit{standard} free energy of binding (or solvation) (see also Sec.~\ref{sec:standardstate-restraints}), which is often indicated in other works with $\Delta G^{\circ}$ to differentiate it from non-standard free energies.
-This is in line with current literature on alchemical calculations, where the standard free energy is normally the only quantity of interest.
-Furthermore, we will use the term configuration for a single set of position vectors and occasionally use the term conformation to refer to a set of configurations that represent a metastable state.
-
-\paragraph{The free energy of binding can be expressed as a ratio of partition functions}
-The law of mass action in Eq.~\ref{eq:law-mass-action} is not directly applicable to typical molecular simulations as they normally include a single receptor/ligand in a small box (i.e., using large concentrations)~\cite{jong2011determining}.
-Instead, a natural, though generally very computationally expensive, way to estimate the equilibrium constant is by directly simulating several binding and unbinding events and computing the probability of finding the receptor-ligand system in the bound state, $P(RL)$, or the unbound state, $P(R+L)$.
-Assuming the volume change upon binding to be negligible, which is often the case at $1~\mathrm{atm}$ due to the incompressibility of water, then the Gibbs free energy $\Delta G_{\mathrm{bind}, L}$ is approximately equal to the Helmholtz free energy $\Delta A_{\mathrm{bind}, L}$, and we can simulate the system in a box of volume $V$ to obtain~\cite{jong2011determining}
-\begin{equation}\label{eq:binding-free-energy-from-bound-unbound-probability-ratio}
-    \Delta G_{\mathrm{bind},L} \approx \Delta A_{\mathrm{bind}, L} = - k_B T  \left( \mathrm{ln} \frac{P(RL)}{P(R+L)} + \ln \left( c^\circ N_{\mathrm{Av}} V \right) \right) \, ,
-\end{equation}
-where $N_{\mathrm{Av}}$ is the Avogadro number, and the last term corrects for the simulated concentration being different than the standard concentration.
-Let $\Gamma_{\mathrm{bound}}$ and $\Gamma_{\mathrm{unbound}}$ be the set of receptor-ligand configurations $\vec{q}$ that we consider bound and unbound respectively.
-The probability of a configuration $\vec{q}$ is given by the Boltzmann probability density function
-\begin{equation}
-    P(\vec{q}) = \frac{\exp\left( -\beta U(\vec{q}) \right)}{\int_{\Gamma} \exp\left( -\beta U(\vec{q}) \right) \, d\vec{q}} \, ,
-    \label{eq:conf_probability}
-\end{equation}
-where $\beta = (k_B T)^{-1}$ is the inverse temperature, $U(\vec{q})$ is the potential energy of configuration $\vec{q}$, and the integration is over the set of all possible configurations accessible in the simulation box volume $\Gamma$, with $\Gamma_{\mathrm{bound}},\Gamma_{\mathrm{unbound}} \subset \Gamma$.
-If the simulation is long enough, we expect the fraction of configurations $\vec{q}$ found in the bound state to converge to
-\begin{equation}
-    P(RL) = \int_{\Gamma_{\mathrm{bound}}} P(\vec{q}) \, d\vec{q} = \frac{\int_{\Gamma_{\mathrm{bound}}} \exp\left( -\beta U(\vec{q}) \right) \, d\vec{q}}{\int_{\Gamma} \exp\left( -\beta U(\vec{q}) \right) \, d\vec{q}} \, .
-\end{equation}
-After similar considerations for $P(R+L)$, we find that the ratio of visited bound and unbound conformations, in the limit of long simulations, should converge to
-\begin{equation}\label{eq:bound-unbound-probability-ratio}
-    \frac{P(RL)}{P(R+L)} = \frac{\int_{\Gamma_{\mathrm{bound}}} \exp\left( -\beta U(\vec{q}) \right) \, d\vec{q}}{\int_{\Gamma_{\mathrm{unbound}}} \exp\left( -\beta U(\vec{q}) \right) \, d\vec{q}} = \frac{Z(RL)}{Z(R+L)} \, ,
-\end{equation}
-where we have defined the \textit{configurational integral} or \textit{configurational partition function} as $Z(\mathrm{state}) \equiv \int_{\Gamma_{\mathrm{state}}} \exp\left(-\beta U(\vec{q})\right) \, d\vec{q}$.
-
-\paragraph{Simulating binding events is computationally expensive}
-While simulating binding events has been used to estimate binding affinities~\cite{jong2011determining,pan2017quantitative} or to get insights into the binding pathways and kinetics of receptor-ligand systems~\cite{teo2016adaptive,votapka2017seekr,doerr2014onthefly,plattner2015protein,dixon2018predicting}, the computational cost of these calculations is usually dominated by the rate of dissociation, which can be on the microsecond timescale even for millimolar binders~\cite{pan2017quantitative} and reaches the microsecond to second timescale for a typical drug~\cite{basavapathruni2012conformational,hyre2006cooperative}.
+ K_b = C^{\circ}\frac{[RL]}{[L][R]},
+\end{equation}
+where $[RL]$, $[L]$, and $[R]$ are the equilibrium concentrations of complex, ligand, and receptor, respectively, by, for example, monitoring the relative population $P(RL)/P(R+L)$ of bound and unbound configurations in molecular dynamics simulations.~\cite{jong2011determining,pan2017quantitative}
+
+The computational protocol based on this idea can be formally derived from Eq.~(\ref{eq:DG0-binding-constant-def}) by multiplying and dividing it by
+\begin{equation}
+z_v = \int dx_R dx_L d\zeta e^{-\beta \Psi_v(x_R, x_L, \zeta)}
+\end{equation}
+that, because, unlike Eq.~(\ref{eq:intra-zRL-def}), the position of the ligand is not limited by the indicator function, corresponds to a system in which the ligand $L$ is freely moving in a box of solvent of volume $V$ containing the receptor $R$.
+The ratio between $z_{RL(v)}$ and $z_v$ is the ensemble average of $I(\zeta)$ or the probability $P(RL)$ that $L$ is in the binding site of the receptor $R$.
+To measure the probability $P(R+L)$ of the $R+L$ unbound state, we consider an indicator function $I_{\mathrm{out}}(\zeta)$ which is one when the ligand, as measured from the center of mass position for example, is far away from the receptor.
+The volume of the far away region is
+\begin{equation}
+8 \pi^2 V_{\mathrm{out}} = \int d\zeta I_{\mathrm{out}}(\zeta)
+\end{equation}
+where the $8 \pi^2$ term comes from the integration of the ligand's orientations not limited by $I_{\mathrm{out}}(\zeta)$.
+Next, multiply the ratio $z_{R(v)} z_{L(v)}/z_v$ by $\int d\zeta I_{\mathrm{out}}(\zeta)$ and divide it by $8 \pi^2 V_{\mathrm{out}}$ yielding
+\begin{equation}
+K_b = C^\circ V_{\mathrm{out}} \frac{P(RL)}{P(R+L)}
+\end{equation}
+where we have used the identity
+\begin{equation}
+P(R+L) = \langle I_{\mathrm{out}}(\zeta) \rangle_v = \frac{\int dx_R dx_L d\zeta I_{\mathrm{out}}(\zeta) e^{-\beta \Psi_v(x_R, x_L, \zeta)}}{\int dx_R dx_L d\zeta e^{-\beta \Psi_v(x_R, x_L, \zeta)}} \, .
+\end{equation}
+
+While simulating explicit binding events has been used to estimate binding constants~\cite{jong2011determining,pan2017quantitative} or to get insights into the binding pathways and kinetics of receptor-ligand systems~\cite{teo2016adaptive,votapka2017seekr,doerr2014onthefly,plattner2015protein,dixon2018predicting}, the computational cost of these calculations is usually dominated by the rate of dissociation, which can be on the microsecond timescale even for millimolar binders~\cite{pan2017quantitative} and reaches the microsecond to second timescale for a typical drug~\cite{basavapathruni2012conformational,hyre2006cooperative}.
 Depending on system size and simulation settings, common molecular dynamics software packages can reach a few hundreds of ns/day using currently available high-end GPUs~\cite{eastman2017openmm,kutzner2019more}, making these type of calculations unappealing and irrelevant on a pharmaceutical drug discovery timescale.
 Other methods compute the free energy of binding by building potential of mean force profiles along a reaction coordinate~\cite{woo2005calculation,velez-vega2013overcoming,limongelli2013funnel,heinzelmann2017attachpullrelease}, but these methods require prior knowledge of a high-probability binding pathway, which is not easily available, especially in the prospective scenarios typical of the drug development process.
+
+Similarly, because of the long time-scale of transition events and the need to observe many events to extract probabilities, it would be very difficult to estimate solvation free energies and solvent partition coefficients by simulating explicitly the motion of solutes in and out of solutions.
 

### 2025-02-11 paper/manuscript.tex
add Widom citation
Updates from Overleaf

--- 2024-12-13 paper/manuscript.tex
@@
 \tocsectioncomment{Steering Meeting: @Emilio Gallicchio to update theory.
-@Stefan Boresch to comment on dummy atom considerations?
 Will frame it to be as general as possible, while remaining accessible to beginners.
@@
 \subsubsection{How long should I run my simulation for and what information should be saved?}
-\tocsubsubsectioncomment{@Volunteer: Move parts of this to analysis section (in Single/ Multiple papers).}
 \label{sec:sim_length_information_kept}
@@
 Overall, this likely means that relative free energy calculations will be susceptible to problems resulting from uncertainty in ligand binding modes until more robust approaches are available to determine dominant binding modes, or the relative populations of different potential binding modes, in advance.
+
+\subsection{Absolute and Relative Binding Free Energies with Multiple Chemical States}
+
+Consider, as an illustration, the binding equilibrium between a receptor $R$ and a ligand $A$ that exists in solution as a mixture of protonated $AH$ and ionized $A^-$ forms.
+It is often useful to consider the observed equilibrium binding constant between $R$ and the $A$ mixture
+\begin{equation}
+  K_b(A) = C^\circ \frac{[RA]}{[R][A]}
+  \label{eqn:observed-binding-constant}
+\end{equation}
+where $[A] = [AH] + [A^-]$ and $[RA] = [RAH] + [RA^-]$ are the total equilibrium concentrations of free $A$ and receptor-ligand complexes.
+The observed equilibrium constant is significant because it corresponds to binding affinity measurements that do not distinguish between the protonation states of the ligand.
+Inserting the definitions of $[A]$ and $[RA]$ in Eq.~(\ref{eqn:observed-binding-constant}), splitting the sum in the numerator, and multiplying and dividing each term by either $[AH]$ or $[A^-]$ we obtain
+\begin{equation}
+  K_b(A) = \frac{[AH]}{[A]}  K_b(AH) +  \frac{[A^-]}{[A]} K_b(A^-) =  p_0(AH) K_b(AH) +  p_0(A-) K_b(A^-)
+  \label{eqn:observed-binding-constant-split}
+\end{equation}
+where
+\begin{equation}
+  K_b(AH) = C^\circ \frac{[RAH]}{[R][AH]}
+  \label{eqn:observed-binding-constant-prot}
+\end{equation}
+is the specific equilibrium binding constant between the receptor and the protonated form of the ligand, and similarly for $K_b(A^-)$.
+The specific $K_b$'s are the binding constants that would be obtained from ABFE calculations for those species.
+Eq.~(\ref{eqn:observed-binding-constant-prot}) states that the observed binding constant is an average of the specific binding constants weighted by the populations, $p_0(AH)$ and  $p_0(A^-)$ of each ligand's chemical form in solution.
+In this specific example, if the $pKa$ of $AH$ is known and $\alpha = [A^-]/[AH] = 10^{pH-pKa} $ is ionization ratio of $A$ at a a given $pH$, the populations of the ionized and protonated forms are $p_0(A^-) = \alpha/(\alpha+1)$ and $p_0(AH) = 1/(\alpha+1)$, respectively.\cite{champion2024multistate,azimi2022application}
+
+Eq.~(\ref{eqn:observed-binding-constant-split}) is a particular case of a fundamental result that extends to an arbitrary number of chemical forms of a ligand in solution.
+In general, the observed binding constant of a ligand present in solution in multiple forms $A_i$ (protomers, tautomers, rotamers, etc.) with populations $p_0(A_i)$ is\cite{khuttan2023taming}
+\begin{equation}
+  K_b(A) = \sum_i p_0(A_i) K_b(A_i)
+  \label{eqn:observed-binding-constant-general}
+\end{equation}
+and the corresponding standard binding free energy is
+\begin{equation}
+ \Delta G^\circ_b(A) = -k_B T \ln K_b(A) = -k_B T \ln \sum_i p_0(A_i) e^{-\Delta G^\circ_b(A_i)/k_B T}
+  \label{eqn:observed-binding-free-energy-general}
+\end{equation}
+where
+\begin{equation}
+\Delta G^\circ_b(A_i) = -k_B T \ln K_b(A_i)
+\end{equation}
+is the standard binding free energy of the specific form $A_i$.
+Eq.~(\ref{eqn:observed-binding-constant-general}) is known as the free energy combination formula, which, in the case of multiple conformational states, can be derived directly from the statistical mechanics definition Eq.~(\ref{eq:DG0-binding-constant-def}).\cite{jayachandran2006parallelized,gallicchio2011recent} It states that the contribution of a chemical or conformational state of the ligand to the observed binding free energy depends on the specific binding free energy of that state and its solution population.
+A state with a strong specific binding affinity may not contribute significantly to the observed binding free energy if its solution concentration is low.
+Conversely, a sparsely populated ligand state in solution could be responsible for most of the observed binding free energy if its binding affinity is sufficiently strong relative to the other states.\cite{azimi2022application}
+
+The result of Eq.~(\ref{eqn:observed-binding-constant-general}), which applies to absolute binding free energy calculations, extends to relative binding free energies or, equivalently, to ratios of observed binding constants for the same receptor of two ligands $A$ and $B$ present in solution in multiple chemical forms:
+\begin{equation}
+  \frac{K_b(B)}{K_b(A)} = \frac{\sum_i p_0(B_i) K_b(B_i)}{\sum_i p_0(A_i) K_b(A_i)}
+  \label{eqn:observed-binding-constant-ratio}
+\end{equation}
+The result above can be expressed in a more computationally tractable form by separating out from the numerator and denominator the specific binding constant of a reference form of each ligand ($A_0$ and $B_0$) to obtain\cite{de2018rigorous}
+\begin{equation}
+ \frac{K_b(B)}{K_b(A)} = \frac{K_b(B_0)}{K_b(A_0)} \frac{
+ p_0(B_0) + \sum_{i=1} p_0(B_i) K_b(B_i)/K_b(B_0)
+ }{
+p_0(A_0) + \sum_{i=1} p_0(A_i) K_b(A_i)/K_b(A_0)
+ }
+\end{equation}
+which involves only quantities obtainable by relative binding free energy calculations--one for the two reference forms of the ligand to get  $K_b(B_0)/K_b(A_0)$ and those of each form of each ligand relative to the reference to get $K_b(B_i)/K_b(B_0)$ and $K_b(A_i)/K_b(A_0)$.
 

### 2025-02-11 paper/manuscript.tex

--- 2024-12-26 paper/manuscript.tex
@@
 %%%%%%%%%%%%%%%%%%%%
-\section{\tocorange{Statistical mechanics demonstrates why alchemical free energy calculations work}}
+\section{\tocgreen{Statistical mechanics demonstrates why alchemical free energy calculations work}}
 %\section{\tocorange{Statistical mechanics theory of solvation, solvent partitioning, and binding}}
-\tocsectioncomment{Steering Meeting: @Emilio Gallicchio to update theory.
-@Stefan Boresch to comment on dummy atom considerations?
-Will frame it to be as general as possible, while remaining accessible to beginners.
-Will align with Gilson et al.
-1997 paper.
-@Darrin York to review.}
 \label{sec:theory}
@@
 
-Combining Eqs.~(\ref{eq:DG-solvation-excess}) and (\ref{eq:zratio-solv-average}) yields
+Combining Eqs.~(\ref{eq:DG-solvation-excess}) and (\ref{eq:zratio-solv-average}) yields\cite{widom1982potential} 
 \begin{equation}

### 2025-02-17 paper/manuscript.tex
Updates from Overleaf

@@
 Eq.~(\ref{eq:DG0-binding-constant-def}) follows from the application of  Eq.~(\ref{eq:chemical-potential-standard}) for each of the three species and noting that the contributions of the momenta cancel.
-As discussed below, various alchemical protocols exist to evaluate the ratios of partition functions in Eq.~(\ref{eq:DG0-binding-constant-def}}), providing a connection between the theory and observed binding affinities.
+As discussed below, various alchemical protocols exist to evaluate the ratios of partition functions in Eq.~(\ref{eq:DG0-binding-constant-def}), providing a connection between the theory and observed binding affinities.
 

### 2025-02-26 paper/manuscript.tex
Start rewrite of relative FES approaches  On branch sboresch
 Changes to be committed:
	modified:   manuscript.tex
	new file:   sb.bib

@@
 
-\paragraph{Topologies} 
-A critical first step is to determine which approach to use for atom mapping between end state molecules.
-Often this is predetermined by the choice of simulation software.
-Typically it is possible to chose between a \emph{dual topology} versus a \emph{single topology}
-
+\paragraph{Topologies} In regular biomolecular simulations one studies physical systems; in particular, all entities in the system, e.g., protein, ligand and waters, represent real molecules.
+For the following, it is helpful to classify such simulation systems as having a {\em chemical} topology.
+In relative free energy calculations, however, one has to transform, e.g., ligand A into ligand B etc.
+(see {\bf XXX}).
+The required alchemical transformations can be realized in several ways, each with distinct advantages and disadvantages.
+In the first relative free energy calculations, these transformations were achieved with the help of chimeric entities, capable of representing both ligand A or ligand B.
+The important point here is that there is a single set of coordinates for this chimeric ligand, but its ``topology'' is flexible and can, for the same set of coordinates, represent either end state.
+We will consider this class of approaches as using an {\em alchemical} topology with a single set of coordinates.
+The construction of these chimeric intermediates quickly gets complicated and is difficult to automate; therefore, approaches have been developed in which both end states (e.g., ligand A and B) are present simultaneously.
+Thus, we have dual coordinates, but chemical topologies for each of the entities.
+In the following, we outline these approaches in more detail, together with additional considerations how the various approaches in practical use differ.
+While we will point out the respective advantages and disadavantages, we stress that the user's option in practice often are limited by the choice of simulation hardware.
+
+\paragraph{Single coordinate approaches --- alchemical topologies}
+
+When operating with a single set of coordinates for the region undergoing the alchemical transformation, an additional distinction between approaches concerns the atom mapping between end state molecules.
+%Often this is predetermined by the choice of simulation software.
+As first suggested by Pearlman, these have been referred to as \emph{dual topology} versus a \emph{single topology} \cite{Pearlman_1994}.
 The distinction between single and dual topology can be illustrated by considering a hypothetical transformation from molecule A to molecule B, where both atoms share a common substructure but differ in their substituents; in particular, consider a transformation of benzene to benzyl alcohol shown in Fig.~\ref{fig:fig_topology}.
@@
 \makeorcid
-\bibliography{alchemical,manual}
+\bibliography{alchemical,manual,sb}
 

### 2025-02-26 paper/manuscript.tex
Add my \sbnote{} macro  On branch sboresch  Changes to be committed: 	modified:   manuscript.tex
Adapted/inserted dual coordinate section from gdrive  Changes to be committed: 	modified:   manuscript.tex 	modified:   sb.bib
NEXT: also mention the other refs Emilio sent in dual coordinate section
THEN: naming disambiguation

@@
   \textit{#1}\par}
+
+% aux. comments Stefan Boresch -> Stefan Boresch
+\newcommand{\sbnote}[1]{%
+  {\bfseries{}[SB: }%
+  {\textcolor{blue}{#1}}{\bfseries{}]}
+  }
 
@@
 
-When operating with a single set of coordinates for the region undergoing the alchemical transformation, an additional distinction between approaches concerns the atom mapping between end state molecules.
+When operating with a single set of coordinates for the region undergoing the alchemical transformation, a key distinction between approaches concerns the atom mapping between end state molecules.
 %Often this is predetermined by the choice of simulation software.
@@
 
-As far as we are aware, the distinction between single and dual topology approaches is made primarily for two main reasons.
+\sbnote{This probably needs to be rewritten\ldots} As far as we are aware, the distinction between single and dual topology approaches is made primarily for two main reasons.
 First, this choice affects the alchemical pathway followed, and thus may affect convergence properties---though we are not yet aware of a study of the relative efficiency and merits of these two approaches.
 Additionally, historically, some simulation packages implemented only one approach and not the other, meaning that the distinction was functional.
+
+\paragraph{Dual coordinate approaches --- dual chemical topologies}
+
+The chimeric constructs needed in single and dual topology approaches in the sense of Pearlman’s definition just described quickly become non-trivial to set up and require care to avoid artifacts \sbnote{Ssee below; where do we place the dummy atom discussion?}.
+Some of these difficulties might be avoided by having the initial and final state present as separate molecules with separate coordinate sets.
+To the best of our knowledge, such an approach was first described by Axelsen and Li \cite{Axelsen_1998}.
+Both initial and final state are present simultaneously; the two copies never see each other and interactions with the environment are turned on/off as a function of $\lambda$.
+Harmonic restraints between corresponding atoms in the initial and final state copy prevent the two molecules (sets of atoms from drifting apart).
+%Unfortunately, Axelsen and Li referred to their ansatz also as dual-topology, though in practice it is quite different from Pearlman’s meaning of the term.
+A recent implementation of this idea is Restraintmaker \cite{Ries_2022}, a tool to automatically assign the distance restraints needed to keep the two molecules / regions (one for each of the physical endstates) on top of each other.
+Somewhat unfortunately, Axelsen and Li, as well as Ries et al refer to their approach as dual topology.
+Ries et al suggested to use the term hybrid topology for (single coordinate) dual topology in the sense of Pearlman.
+We will attempt a disambiguation of the various nomenclatures below.
+
+Another variant of a dual coordinate approach is implemented in NAMD \cite{jiang2019computing}.
+Whereas Ries et al.\  use distance restraints to keep two molecules on top of each other, corresponding atoms in NAMD are held on top of each other by constraints; in practice, this means that these coordinates are forced to be identical.
+AMBER's free energy implementation recently switched to a quite similar approach \cite{Lee_2023}.
+%The authors refer to this approach as hybrid single-dual topology.
+While it technically uses two separate sets of coordinates, the constraint that corresponding atoms have identical coordinates makes it effectively a single coordinate method; hence, I consider NAMD’s approach to be a variant of Pearlman’s (single coordinate) dual topology.
+
+Dual coordinate approaches with chemical topologies encompass a third approach, where essentially two absolute free energy calculations are carried out in opposite directions at the same time.
+The so-called ``separated topologies'' (SepTop) method \cite{rocklin2013separated,Baumann_2023} and the ``alchemical transfer method'' (ATM) \cite{Azimi_2022} belong into this category.
+In SepTop, the two copies are held on top of each other via restraints with respect to the protein/receptor they bind to (as opposed to Restraintmaker, where the molecules are restrained to each other.
+In ATM, a restraint is used to keep the bound ligand 1 and the free ligand 2 always some minimum distance apart; during exchange moves, suitable rigid-body rotations/translations are applied to move the free ligand into a position/orientation as similar as possible to the one of the bound ligand.
+Both copies are present simultaneously, but they do not occupy the same space.
+
+%Other methodologies also use both end states simultaneously; examples are the Separated Topologies (SepTop) approach [Baumann2023] and the ATM method [Azimi2022].
+In the former, the two copies are held on top of each other via restraints with respect to the protein/receptor they bind to.
+Ries et al.
+refer to such an approach as “separated dual topology”, whereas they label their approach as “linked dual topology”.
+In ATM, a restraint is used to keep the bound ligand 1 and the free ligand 2 always some minimum distance apart; during exchange moves, suitable rigid-body rotations/translations are applied to move the free ligand into a position/orientation as similar as possible to the one of the bound ligand.
+Both copies are present simultaneously, but they do not occupy the same space.
+
+%One final approach, so far in its infancy, has been called ``separated topologies'' and essentially consist of two absolute free energy calculations in opposite directions at the same time, turning one molecule's interactions with the environment off, while turning the other molecule's interaction on~\cite{jiang2019computing, rocklin2013separated}.
+
+\paragraph{Summary, comparison of single and dual coordinate methods, orthogonal classifications:}
 
@@
 
-One final approach, so far in its infancy, has been called ``separated topologies'' and essentially consist of two absolute free energy calculations in opposite directions at the same time, turning one molecule's interactions with the environment off, while turning the other molecule's interaction on~\cite{jiang2019computing, rocklin2013separated}.
-
-Software packages vary in their use of single or dual topology approaches; for example, AMBER TI uses a dual topology approach, while BioSimSpace uses a single topology approach.
+\sbnote{I think this should get its own paragraph/subsection; to be expanded!} Software packages vary in their use of single or dual topology approaches; for example, AMBER TI uses a dual topology approach, while BioSimSpace uses a single topology approach.
 Please make sure to check what approach is used with your software package of choice, or whether it supports your choice of approach (GROMACS and GROMOS, for example, support both).

### 2025-03-05 paper/manuscript.tex
Restart writing a bit ..  On branch sboresch
 Changes to be committed:
	modified:   manuscript.tex
	modified:   sb.bib

@@
 Whereas Ries et al.\  use distance restraints to keep two molecules on top of each other, corresponding atoms in NAMD are held on top of each other by constraints; in practice, this means that these coordinates are forced to be identical.
-AMBER's free energy implementation recently switched to a quite similar approach \cite{Lee_2023}.
+AMBER's free energy implementation recently switched to a similar approach \cite{Lee_2023}.
 %The authors refer to this approach as hybrid single-dual topology.
@@
 Both copies are present simultaneously, but they do not occupy the same space.
+Recently, a variant of $\lambda$-dynamics that uses chemical topologies (separate coordinates for each state) has been proposed \cite{Liesen_2024}.
 
@@
 
-Some additional terms have also been employed to talk about these different intermediate pathways .
+%Some additional terms have also been employed to talk about these different intermediate pathways .
 Particularly, some studies refer to a ``hybrid'' topology approach to free energy calculations~\cite{gapsys2015pmx, gapsys2016accurate, gapsys2020large}, though this term may not yet have achieved widespread use.
 In this case, "hybrid" seems to indicate that the set up of these free energy calculations involves a hybrid of the two molecules, and much of what is done in these studies uses a single topology approach~\cite{gapsys2020large}.
+
+Unfortunately, during the past years the use of the terms single and dual topology has proliferated, often with quite different meaning from the original intentions.
+The above presentation makes clear that the primary difference between approaches is the use of separate coordinates for each end state (chemical topology) or the use of an alchemical chimera with a single set of coordinates (alchemical topology).
+We therefore suggest the following classification
+\begin{itemize}
+\item Single coordinate (SC) --- Alchemical topology
+  \begin{itemize}
+  \item Single topology (SCST)
+    \item Dual topology (SCDT)
+  \end{itemize}
+  \item Dual coordinate --- Chemical topology (DC)
+  \end{itemize}
+For completeness, we list some additional terminology that has been used: Some studies refer to a ``hybrid'' topology approach to free energy calculations~\cite{gapsys2015pmx, gapsys2016accurate, gapsys2020large}, though in our view these calculations should be classified as single coordinate single topology \sbnote{Check whether this is correct!}.
+%this term may not yet have achieved widespread use.
+In this case, "hybrid" seems to indicate that the set up of these free energy calculations involves a hybrid of the two molecules, and much of what is done in these studies uses a single topology approach~\cite{gapsys2020large}.
+Ries et al.\ suggested terminology \cite{Ries_2022} in which single topology corresponds to SCST, hybrid topology to SCDT, and dual topology to DC.
 

### 2025-03-05 paper/manuscript.tex
First pass through my intented revisions  On branch sboresch  Changes to be committed: 	modified:   manuscript.tex 	modified:   sb.bib

@@
   }
+\newcommand{\rref}{{\bfseries[REFs]}{}}
 
@@
 
+The practical implementation of the approaches to carry out alchemial transformations above leads to additional, often non-trivial differences within each methodology.
+In the description of DC approaches we already pointed out that the two sets of coordinates can be held on top of each other by constraints \rref or restraints.
+In the latter case, the restraints may couple the two set of coordinates directly \rref, or the restraints may be defined relative to a common reference frame.
+Finally, in ATM the two copies of the coordinates are held separate.
+
+Concerning SC approaches, Pearlman’s usage of the term single topology not only meant a chimeric construct as depicted in Fig.~{\bfseries XX}, but specifically implied that the coupling parameter $\lambda$ was applied to the force field parameters.
+E.g., a bond stretching energy term changing from C-H to C-C has the functional form $U_{bond}(\lambda)=K(\lambda)(r-r_0(\lambda))^2$.
+For $\lambda=0$, $K(\lambda)$ and $r_0(\lambda)$ correspond to the values appropriate for an aliphatic C-H bond, and for $\lambda=1$ to those for an aliphatic C-C bond.
+The other energy terms are treated analogously.
+The single topology paradigm as just summarized was the default mode of alchemical transformations in AMBER and GROMOS at that time.
+More recently, this implementation of single topology has been referred to as parameter interpolation or mixing \cite{Giese_2018}.
+
+The $\lambda$-dependent interpolation can also be accomplished by mixing the energies and forces of the initial ($\lambda=0$) and final state ($\lambda=1$), e.g., linearly according to $U(\lambda)=(1-\lambda) U_0+\lambda\,U_1$.
+Other λ-dependencies can be used, and soft-cores \rref may be used for Lennard-Jones and electrostatic interactions to avoid endpoint problems.
+Giese et al.
+\cite{Giese_2018} refer to this approach as energy mixing (interpolation).%; it requires two separate calls to the energy/force routines, but since only a fraction of interactions is affected by the alchemical changes, various optimizations are possible to keep the computational overhead manageable.
+An early example of energy mixing for SCST is the PERT free energy module of CHARMM (written by Bernie Brooks around 1990 and briefly described in Ref.~\cite{brooks_2009}.
+
+Any SCDT approach requires two energy/force calculations and, thus, uses energy mixing.
+Therefore, approaches that implement energy mixing together with an alchemical topology (a single set of coordinates) often support both SCST and SCDT.
+E.g., CHARMM's PERT module just mentioned could be used to set up either of the SC approaches sketched in Fig.~{\bf XX}.
+Finally, all DC approaches require two energy/force evaluations.
+
+Any energy mixing approach, whether in a SC or DC context, requires two energy/force evaluations.
+While various optimizations keep the computational overhead significantly below a factor of two,  parameter mixing, if applicable, is the compuationally most efficient approach.
+However, it is closely tied to SCST and the more complicated the alchemical transformation, the more challenging a strict SCST approach becomes.
+Also, the mixing of partial charges and the calculation of $\partial U/\partial\lambda$ for the reciprocal space particle-mesh Ewald (PME) summation is tricky.
+Giese et al.\ described how to do this \cite{Giese_2018}, but most energy mixing implementations calculate the PME term twice and mix the resulting energy and forces.
+
 \sbnote{I think this should get its own paragraph/subsection; to be expanded!} Software packages vary in their use of single or dual topology approaches; for example, AMBER TI uses a dual topology approach, while BioSimSpace uses a single topology approach.

### 2025-03-12 paper/manuscript.tex
Some minor fixes  On branch sboresch  Changes to be committed: 	modified:   manuscript.tex 	modified:   sb.bib

@@
 
-\paragraph{Topologies} In regular biomolecular simulations one studies physical systems; in particular, all entities in the system, e.g., protein, ligand and waters, represent real molecules.
+\paragraph{Topologies} In regular biomolecular simulations one studies physical systems or, in other words, all entities in the system, e.g., protein, ligand and waters, represent real molecules.
 For the following, it is helpful to classify such simulation systems as having a {\em chemical} topology.
@@
 Thus, we have dual coordinates, but chemical topologies for each of the entities.
-In the following, we outline these approaches in more detail, together with additional considerations how the various approaches in practical use differ.
-While we will point out the respective advantages and disadavantages, we stress that the user's option in practice often are limited by the choice of simulation hardware.
+In the following, we outline these approaches in more detail, together with additional considerations how the various approaches differ in practical use.
+While we will point out the respective advantages and disadavantages, we stress that the user's option in practice often are limited by the choice of simulation software.
 
@@
 The distinction between single and dual topology can be illustrated by considering a hypothetical transformation from molecule A to molecule B, where both atoms share a common substructure but differ in their substituents; in particular, consider a transformation of benzene to benzyl alcohol shown in Fig.~\ref{fig:fig_topology}.
-In this case the common substructure between the two molecules is the benzene ring, though in practice substructure may be selected to be larger depending the mapping chosen, as we discuss below.
+\sbnote{Fig.~3 needs to be updated for dual coordinate methods and corrected, some bonds for the dummy atoms have the wrong anchor atom!}
+In this case the common substructure between the two molecules is the benzene ring, though in practice the substructure may be selected to be larger depending on the mapping chosen, as we discuss below.
 
@@
 The site of this transformation will also be the future home of two additional hydrogen atoms bound to the new carbon, so these must initially be present as non-interacting atoms called ``dummy atoms'', which retain their bonded interactions but do not interact with the rest of the system.
-Bond parameters as well as partial charges between the changing atoms are adjusted accordingly between the initial and final state.
-Thus, in a single topology calculation, atoms may change their type, ensuring minimal dummy atoms are created. This is illustrated in the left arm of Fig.~\ref{fig:fig_topology}.
+Bond parameters as well as partial charges between the changing atoms are adjusted accordingly between the initial and final state; see below for further details.
+Thus, in a single topology calculation, atoms may change their type, ensuring minimal dummy atoms are created.
+This is illustrated in the left arm of Fig.~\ref{fig:fig_topology}.
 
@@
 Fig.~\ref{fig:fig_topology}'s right branch depicts how such a dual topology works.
+Note that the two groups that are switched on or off never see each other.
 
@@
 Additionally, historically, some simulation packages implemented only one approach and not the other, meaning that the distinction was functional.
+For an in-depth analyis of differences resulting from the use of a  single vs.\ a dual topology approach (single coordinate in both cases) see Ref.~\cite{Boresch_2002}.
 
@@
 
-The chimeric constructs needed in single and dual topology approaches in the sense of Pearlman’s definition just described quickly become non-trivial to set up and require care to avoid artifacts \sbnote{Ssee below; where do we place the dummy atom discussion?}.
+The chimeric constructs needed in single and dual topology approaches in the sense of Pearlman’s definition just described quickly become non-trivial to set up and require care to avoid artifacts \sbnote{See below; where do we place the dummy atom discussion?}.
 Some of these difficulties might be avoided by having the initial and final state present as separate molecules with separate coordinate sets.
@@
 
-Dual coordinate approaches with chemical topologies encompass a third approach, where essentially two absolute free energy calculations are carried out in opposite directions at the same time.
+Dual coordinate approaches with chemical topologies encompass a third group of methods, where essentially two absolute free energy calculations are carried out in opposite directions at the same time.
 The so-called ``separated topologies'' (SepTop) method \cite{rocklin2013separated,Baumann_2023} and the ``alchemical transfer method'' (ATM) \cite{Azimi_2022} belong into this category.
@@
 
-Concerning SC approaches, Pearlman’s usage of the term single topology not only meant a chimeric construct as depicted in Fig.~{\bfseries XX}, but specifically implied that the coupling parameter $\lambda$ was applied to the force field parameters.
+Concerning SC approaches, Pearlman’s usage of the term single topology not only meant a chimeric construct as depicted in Fig.~\ref{fig:fig_topology}, but specifically implied that the coupling parameter $\lambda$ was applied to the force field parameters.
 E.g., a bond stretching energy term changing from C-H to C-C has the functional form $U_{bond}(\lambda)=K(\lambda)(r-r_0(\lambda))^2$.

### 2025-03-12 paper/manuscript.tex
Add final edit  On branch main
 Changes to be committed:
	modified:   manuscript.tex

@@
 E.g., a bond stretching energy term changing from C-H to C-C has the functional form $U_{bond}(\lambda)=K(\lambda)(r-r_0(\lambda))^2$.
-For $\lambda=0$, $K(\lambda)$ and $r_0(\lambda)$ correspond to the values appropriate for an aliphatic C-H bond, and for $\lambda=1$ to those for an aliphatic C-C bond.
+For $\lambda=0$, $K(\lambda)$ and $r_0(\lambda)$ correspond to the values appropriate for a C-H bond, and for $\lambda=1$ to those for a C-C bond.
 The other energy terms are treated analogously.

### 2025-04-01 paper/manuscript.tex
first draft of atom mapping update

--- 2025-03-05 paper/manuscript.tex
@@
   \textit{#1}\par}
-
-% aux. comments Stefan Boresch -> Stefan Boresch
-\newcommand{\sbnote}[1]{%
-  {\bfseries{}[SB: }%
-  {\textcolor{blue}{#1}}{\bfseries{}]}
-  }
-\newcommand{\rref}{{\bfseries[REFs]}{}}
 
@@
 
-\paragraph{Topologies} In regular biomolecular simulations one studies physical systems; in particular, all entities in the system, e.g., protein, ligand and waters, represent real molecules.
-For the following, it is helpful to classify such simulation systems as having a {\em chemical} topology.
-In relative free energy calculations, however, one has to transform, e.g., ligand A into ligand B etc.
-(see {\bf XXX}).
-The required alchemical transformations can be realized in several ways, each with distinct advantages and disadvantages.
-In the first relative free energy calculations, these transformations were achieved with the help of chimeric entities, capable of representing both ligand A or ligand B.
-The important point here is that there is a single set of coordinates for this chimeric ligand, but its ``topology'' is flexible and can, for the same set of coordinates, represent either end state.
-We will consider this class of approaches as using an {\em alchemical} topology with a single set of coordinates.
-The construction of these chimeric intermediates quickly gets complicated and is difficult to automate; therefore, approaches have been developed in which both end states (e.g., ligand A and B) are present simultaneously.
-Thus, we have dual coordinates, but chemical topologies for each of the entities.
-In the following, we outline these approaches in more detail, together with additional considerations how the various approaches in practical use differ.
-While we will point out the respective advantages and disadavantages, we stress that the user's option in practice often are limited by the choice of simulation hardware.
-
-\paragraph{Single coordinate approaches --- alchemical topologies}
-
-When operating with a single set of coordinates for the region undergoing the alchemical transformation, a key distinction between approaches concerns the atom mapping between end state molecules.
-%Often this is predetermined by the choice of simulation software.
-As first suggested by Pearlman, these have been referred to as \emph{dual topology} versus a \emph{single topology} \cite{Pearlman_1994}.
+\paragraph{Topologies} 
+A critical first step is to determine which approach to use for atom mapping between end state molecules.
+Often this is predetermined by the choice of simulation software.
+Typically it is possible to chose between a \emph{dual topology} versus a \emph{single topology}
+
 The distinction between single and dual topology can be illustrated by considering a hypothetical transformation from molecule A to molecule B, where both atoms share a common substructure but differ in their substituents; in particular, consider a transformation of benzene to benzyl alcohol shown in Fig.~\ref{fig:fig_topology}.
@@
 
-\sbnote{This probably needs to be rewritten\ldots} As far as we are aware, the distinction between single and dual topology approaches is made primarily for two main reasons.
+As far as we are aware, the distinction between single and dual topology approaches is made primarily for two main reasons.
 First, this choice affects the alchemical pathway followed, and thus may affect convergence properties---though we are not yet aware of a study of the relative efficiency and merits of these two approaches.
@@
 
-\paragraph{Dual coordinate approaches --- dual chemical topologies}
-
-The chimeric constructs needed in single and dual topology approaches in the sense of Pearlman’s definition just described quickly become non-trivial to set up and require care to avoid artifacts \sbnote{Ssee below; where do we place the dummy atom discussion?}.
-Some of these difficulties might be avoided by having the initial and final state present as separate molecules with separate coordinate sets.
-To the best of our knowledge, such an approach was first described by Axelsen and Li \cite{Axelsen_1998}.
-Both initial and final state are present simultaneously; the two copies never see each other and interactions with the environment are turned on/off as a function of $\lambda$.
-Harmonic restraints between corresponding atoms in the initial and final state copy prevent the two molecules (sets of atoms from drifting apart).
-%Unfortunately, Axelsen and Li referred to their ansatz also as dual-topology, though in practice it is quite different from Pearlman’s meaning of the term.
-A recent implementation of this idea is Restraintmaker \cite{Ries_2022}, a tool to automatically assign the distance restraints needed to keep the two molecules / regions (one for each of the physical endstates) on top of each other.
-Somewhat unfortunately, Axelsen and Li, as well as Ries et al refer to their approach as dual topology.
-Ries et al suggested to use the term hybrid topology for (single coordinate) dual topology in the sense of Pearlman.
-We will attempt a disambiguation of the various nomenclatures below.
-
-Another variant of a dual coordinate approach is implemented in NAMD \cite{jiang2019computing}.
-Whereas Ries et al.\  use distance restraints to keep two molecules on top of each other, corresponding atoms in NAMD are held on top of each other by constraints; in practice, this means that these coordinates are forced to be identical.
-AMBER's free energy implementation recently switched to a similar approach \cite{Lee_2023}.
-%The authors refer to this approach as hybrid single-dual topology.
-While it technically uses two separate sets of coordinates, the constraint that corresponding atoms have identical coordinates makes it effectively a single coordinate method; hence, I consider NAMD’s approach to be a variant of Pearlman’s (single coordinate) dual topology.
-
-Dual coordinate approaches with chemical topologies encompass a third approach, where essentially two absolute free energy calculations are carried out in opposite directions at the same time.
-The so-called ``separated topologies'' (SepTop) method \cite{rocklin2013separated,Baumann_2023} and the ``alchemical transfer method'' (ATM) \cite{Azimi_2022} belong into this category.
-In SepTop, the two copies are held on top of each other via restraints with respect to the protein/receptor they bind to (as opposed to Restraintmaker, where the molecules are restrained to each other.
-In ATM, a restraint is used to keep the bound ligand 1 and the free ligand 2 always some minimum distance apart; during exchange moves, suitable rigid-body rotations/translations are applied to move the free ligand into a position/orientation as similar as possible to the one of the bound ligand.
-Both copies are present simultaneously, but they do not occupy the same space.
-Recently, a variant of $\lambda$-dynamics that uses chemical topologies (separate coordinates for each state) has been proposed \cite{Liesen_2024}.
-
-%Other methodologies also use both end states simultaneously; examples are the Separated Topologies (SepTop) approach [Baumann2023] and the ATM method [Azimi2022].
-In the former, the two copies are held on top of each other via restraints with respect to the protein/receptor they bind to.
-Ries et al.
-refer to such an approach as “separated dual topology”, whereas they label their approach as “linked dual topology”.
-In ATM, a restraint is used to keep the bound ligand 1 and the free ligand 2 always some minimum distance apart; during exchange moves, suitable rigid-body rotations/translations are applied to move the free ligand into a position/orientation as similar as possible to the one of the bound ligand.
-Both copies are present simultaneously, but they do not occupy the same space.
-
-%One final approach, so far in its infancy, has been called ``separated topologies'' and essentially consist of two absolute free energy calculations in opposite directions at the same time, turning one molecule's interactions with the environment off, while turning the other molecule's interaction on~\cite{jiang2019computing, rocklin2013separated}.
-
-\paragraph{Summary, comparison of single and dual coordinate methods, orthogonal classifications:}
-
-%Some additional terms have also been employed to talk about these different intermediate pathways .
+Some additional terms have also been employed to talk about these different intermediate pathways .
 Particularly, some studies refer to a ``hybrid'' topology approach to free energy calculations~\cite{gapsys2015pmx, gapsys2016accurate, gapsys2020large}, though this term may not yet have achieved widespread use.
@@
 
-Unfortunately, during the past years the use of the terms single and dual topology has proliferated, often with quite different meaning from the original intentions.
-The above presentation makes clear that the primary difference between approaches is the use of separate coordinates for each end state (chemical topology) or the use of an alchemical chimera with a single set of coordinates (alchemical topology).
-We therefore suggest the following classification
-\begin{itemize}
-\item Single coordinate (SC) --- Alchemical topology
-  \begin{itemize}
-  \item Single topology (SCST)
-    \item Dual topology (SCDT)
-  \end{itemize}
-  \item Dual coordinate --- Chemical topology (DC)
-  \end{itemize}
-For completeness, we list some additional terminology that has been used: Some studies refer to a ``hybrid'' topology approach to free energy calculations~\cite{gapsys2015pmx, gapsys2016accurate, gapsys2020large}, though in our view these calculations should be classified as single coordinate single topology \sbnote{Check whether this is correct!}.
-%this term may not yet have achieved widespread use.
-In this case, "hybrid" seems to indicate that the set up of these free energy calculations involves a hybrid of the two molecules, and much of what is done in these studies uses a single topology approach~\cite{gapsys2020large}.
-Ries et al.\ suggested terminology \cite{Ries_2022} in which single topology corresponds to SCST, hybrid topology to SCDT, and dual topology to DC.
-
-The practical implementation of the approaches to carry out alchemial transformations above leads to additional, often non-trivial differences within each methodology.
-In the description of DC approaches we already pointed out that the two sets of coordinates can be held on top of each other by constraints \rref or restraints.
-In the latter case, the restraints may couple the two set of coordinates directly \rref, or the restraints may be defined relative to a common reference frame.
-Finally, in ATM the two copies of the coordinates are held separate.
-
-Concerning SC approaches, Pearlman’s usage of the term single topology not only meant a chimeric construct as depicted in Fig.~{\bfseries XX}, but specifically implied that the coupling parameter $\lambda$ was applied to the force field parameters.
-E.g., a bond stretching energy term changing from C-H to C-C has the functional form $U_{bond}(\lambda)=K(\lambda)(r-r_0(\lambda))^2$.
-For $\lambda=0$, $K(\lambda)$ and $r_0(\lambda)$ correspond to the values appropriate for an aliphatic C-H bond, and for $\lambda=1$ to those for an aliphatic C-C bond.
-The other energy terms are treated analogously.
-The single topology paradigm as just summarized was the default mode of alchemical transformations in AMBER and GROMOS at that time.
-More recently, this implementation of single topology has been referred to as parameter interpolation or mixing \cite{Giese_2018}.
-
-The $\lambda$-dependent interpolation can also be accomplished by mixing the energies and forces of the initial ($\lambda=0$) and final state ($\lambda=1$), e.g., linearly according to $U(\lambda)=(1-\lambda) U_0+\lambda\,U_1$.
-Other λ-dependencies can be used, and soft-cores \rref may be used for Lennard-Jones and electrostatic interactions to avoid endpoint problems.
-Giese et al.
-\cite{Giese_2018} refer to this approach as energy mixing (interpolation).%; it requires two separate calls to the energy/force routines, but since only a fraction of interactions is affected by the alchemical changes, various optimizations are possible to keep the computational overhead manageable.
-An early example of energy mixing for SCST is the PERT free energy module of CHARMM (written by Bernie Brooks around 1990 and briefly described in Ref.~\cite{brooks_2009}.
-
-Any SCDT approach requires two energy/force calculations and, thus, uses energy mixing.
-Therefore, approaches that implement energy mixing together with an alchemical topology (a single set of coordinates) often support both SCST and SCDT.
-E.g., CHARMM's PERT module just mentioned could be used to set up either of the SC approaches sketched in Fig.~{\bf XX}.
-Finally, all DC approaches require two energy/force evaluations.
-
-Any energy mixing approach, whether in a SC or DC context, requires two energy/force evaluations.
-While various optimizations keep the computational overhead significantly below a factor of two,  parameter mixing, if applicable, is the compuationally most efficient approach.
-However, it is closely tied to SCST and the more complicated the alchemical transformation, the more challenging a strict SCST approach becomes.
-Also, the mixing of partial charges and the calculation of $\partial U/\partial\lambda$ for the reciprocal space particle-mesh Ewald (PME) summation is tricky.
-Giese et al.\ described how to do this \cite{Giese_2018}, but most energy mixing implementations calculate the PME term twice and mix the resulting energy and forces.
-
-\sbnote{I think this should get its own paragraph/subsection; to be expanded!} Software packages vary in their use of single or dual topology approaches; for example, AMBER TI uses a dual topology approach, while BioSimSpace uses a single topology approach.
+One final approach, so far in its infancy, has been called ``separated topologies'' and essentially consist of two absolute free energy calculations in opposite directions at the same time, turning one molecule's interactions with the environment off, while turning the other molecule's interaction on~\cite{jiang2019computing, rocklin2013separated}.
+
+Software packages vary in their use of single or dual topology approaches; for example, AMBER TI uses a dual topology approach, while BioSimSpace uses a single topology approach.
 Please make sure to check what approach is used with your software package of choice, or whether it supports your choice of approach (GROMACS and GROMOS, for example, support both).
@@
 \paragraph{Atom mapping}
-Once a particular approach to the topology is selected, a crucial next step is to identify the common atoms which will not be perturbed.
-Rigorously, this process typically comprises a MCSS search of the molecules involved to identify the common substructure---though the parameters of the MCSS search will differ depending on whether single or dual topology calculations are planned.
-Specifically, with a single topology approach in mind, atom types are allowed to change, so a permissive MCSS search can be done, whereas with dual topology a more strict search is required.
+Following the choice of system representation and sampling strategy, it might be required to generate a atom mapping.
+This is the case especially for system representations using a single coordinate set (Type 1.
+& 2), however a dual coordinate set approach could use an atom mapping as well, depending on the sampling method, this will be explained in the next steps.
+Once a single coordinate set approach is selected is selected, a crucial next step is to identify the common atoms which will not be perturbedbetween the two molecules (respectivly end-states).
+Rigorously, this process typically comprises a MCSS search of the molecules involved to identify the common substructure.
+Specifically, with a single coordinate set approach in mind, atom types are allowed to change, so a permissive MCSS search can be done.
+This however requires either careful implementation of atom mass changes, bond length etc.
+or accepting the introduction of potential systematic error, that is in many cases negliable due to the cancelation of errors.
+\cite{ries2024Kartograf}
+
 \begin{figure}
@@
 
-There are different tools that allow the generation of MCSS matches as well as single topology input.
-A large number of software tools can compute MCSS matches using different cheminformatics packages.
-Some rely on RDKit~\cite{rdkit2019Dec}, such as pmx~\cite{gapsys2015pmx}, LOMAP~\cite{liu2013lead}, FESetup~\cite{loeffler2015fesetup} and partially BioSimSpace~\cite{hedges2019biosimspace}, while others such as fkckombu~\cite{kawabata20143d} are standalone tools.
+In general there are mutliple ways on how a MCSS can be calculated.
+One of the most prominent approaches is to estimate the MCSS with a 2D subgraph isomorphism solver(SIS) using the 2D topological stuructre information of molecules\cite{raymond2002maximum}.
+A large number of software tools can compute SIS MCSS matches using different cheminformatics packages.
+Some rely on RDKit~\cite{rdkit2019Dec}, such as pmx~\cite{gapsys2015pmx}, LOMAP~\cite{liu2013lead}, FESetup~\cite{loeffler2015fesetup}, ProtoCaller~\cite{suruzhon2020protocaller}, SMArt~\cite{petrov2021perturbation}, and partially BioSimSpace~\cite{hedges2019biosimspace}, while others such as fkckombu~\cite{kawabata20143d} are standalone tools.
 Schr\"{o}dinger's FEP+ planning tool was originally based on a version of LOMAP, and it also uses MCSS matching as well as 3D considerations to plan the network of single topology calculations between molecules~\cite{wang2015accurate}.
 
-MCSS searches can be relatively time consuming, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial similarity assessment and then use MCSS only to identify final mappings for relative calculations~\cite{raymond2002maximum,klabunde2012mars,jones2009elucidating}.
-The MCSS approach, though relatively standard, takes into account only topological similarity.
+SIS based methods, showed to struggle with 3D properties of the molecules, that need to be mapped, like for example stereochemistry.
+Therefore some packages developed a post-mapping step, where they filtered for atom distances, and therefore added 3D information into the mapping.
+Still SIS MCSS searches can be relatively time consuming as the algorithmic complexity is classified as NP-complete, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial similarity assessment and then use MCSS only to identify final mappings for relative calculations~\cite{raymond2002maximum,klabunde2012mars,jones2009elucidating, Duesbury2018Comparison}.
+The SIS MCSS approach, though relatively standard, takes into account only topological similarity.
 It is possible that changes in binding mode could actually require a different choice of mapping, so in some cases mappings may need to be planned differently depending on 3D positioning of atoms in space.
-Visual inspection prior to simulation is recommended to ensure that the mapping criteria correspond to the expected binding mode.
+
+Here an alternative way to estimate the MCSS is to calculate a bipartite graph matching based on the 3D positions of two molecules.
+These approaches tend to be very fast (can have algorithmic complexity of $O(n)=N^3$) and geometric accurate in their mapping (e.g.
+stereochemistry or binding mode differing).
+The geometric approaches additionally allow mappings of differing binding modes of the same molecule, or can be even used for significant larger molecules like proteins, due to the speed increase.
+A disadvantage of these approaches is the strong dependency on the coordinate sets being well aligned.
+Another aspect is taking the topology changes of the molecules not into account, which can be done if necessary in a next step with a filtering approach, like checking for changes in ring hybridization or ring breaking (this problem is discusse below)|.
+Tools that can be used to estimate the MCSS this way are contained for example in pmx \cite{gapsys2015pmx} or kartograf \cite{ries2024kartograf}.
+
+In any case Visual inspection prior to simulation is recommended to ensure that the mapping criteria correspond to the expected binding mode.
 If the mapping protocol returns simulations that correspond to different binding modes of a ligand within a perturbation map, this can cause large hysteresis.
 
-Single topology relative calculations and calculations based on substructure searches only work if in fact the ligands share a common substructure, e.g.
+A next step for atom mappings, would be to support multistate methods (e.g.
+RE-EDS~\cite{sidler2016replica}, A-EDS~\cite{perthold2020toward} or Ladybugs~\cite{robo2022achieving}) with efficient mapping algorithms.
+There are protoype starting point for solving this problem, but to our knowledge, there is only few published work on this aspect.
+\cite{petrov2021Perturbation}
+
+Single coordinate set relative calculations and calculations based on substructure searches only work if in fact the ligands share a common substructure, e.g.
 are part of a congeneric series, see Fig.~\ref{fig:fig_mcss}.
-If no common substructure is shared, then alternative dual or separated topology free energy calculations are needed.
+If no common substructure is shared, then dual coordinate sets methods like linked-dual or separated topology free energy calculations are needed.
 One would co-localize a pair of compounds in a binding site, exclude their interactions with one another, and compute the relative binding free energy by turning one molecule on from being dummy atoms while turning the other off.
-To our knowledge no general pipeline for such calculations yet exists and this would likely remain a research problem.
-Using an absolute free energy approach instead seems more promising in such a case.
+For such approaches it is crucial to find a good set of restraints, that keep the ligands colocalized over the time of the simulation.
+Few descriptions of the approach can be found for such approaches like the one suggested by Baumann et al.~\cite{baumann2023broadening} based on boresch style restraints~\cite{boresch2003absolute} for seperated topology approach, finding , or Restraintmaker~\cite{ries2022restraintmaker} for the linked-dualtopology approaches, which finds a small set of atoms to be linked via distance restraints, with a small distance to each other.
+However the implementation and testing of an fully automatized dual-topology pipeline for RBFE dual topology approaches is still to our knowledge a remaining research problem and was only tested with small datasets.
+\cite{jespers2019qligfep, Rieder2022Lev}
+Alternativley an approach where such a restraint algorithm was included in a pipeline for ABFE calculations is described by the ABFE\_workflow \cite{Ries2024ABFE} using MDRestraintsGenerator\cite{alibay2021mdrestraintsgenerator}.
 
 \paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
-Some approaches have attempted to address this~\cite{clark2019relative} but a general solution is not yet in mainstream use. Still, FEP+ implements one solution.
+Some approaches have attempted to address thisusing a single coordinate sets and with soft bond approaches ~\cite{clark2019relative,wang2017accurate} or dual coordinate set approach \cite{jespers2019qligfep, ries2022relative} but a general solution is not yet in mainstream use.
+Still, FEP+ implements one solution\cite{wang2017accurate}.
 
@@
 \makeorcid
-\bibliography{alchemical,manual,sb}
+\bibliography{alchemical,manual}
 

### 2025-04-04 paper/manuscript.tex
removing abfe part, as is taken up later.
Updated authors

@@
 \author[2]{Huafeng Xu}
+\author[11]{William G. Poole}
 %
@@
 \affil[10]{Cresset, Cambridgeshire, UK}
+\affil[11]{School of Chemistry, University of Southampton, Southampton, SO17 1BJ}
 
@@
 \paragraph{Atom mapping}
-Following the choice of system representation and sampling strategy, it might be required to generate a atom mapping.
-This is the case especially for system representations using a single coordinate set (Type 1.
-& 2), however a dual coordinate set approach could use an atom mapping as well, depending on the sampling method, this will be explained in the next steps.
-Once a single coordinate set approach is selected is selected, a crucial next step is to identify the common atoms which will not be perturbedbetween the two molecules (respectivly end-states).
-Rigorously, this process typically comprises a MCSS search of the molecules involved to identify the common substructure.
-Specifically, with a single coordinate set approach in mind, atom types are allowed to change, so a permissive MCSS search can be done.
-This however requires either careful implementation of atom mass changes, bond length etc.
-or accepting the introduction of potential systematic error, that is in many cases negliable due to the cancelation of errors.
-\cite{ries2024Kartograf}
-
+Once a particular approach to the topology is selected, a crucial next step is to identify the common atoms which will not be perturbed.
+Rigorously, this process typically comprises a MCSS search of the molecules involved to identify the common substructure---though the parameters of the MCSS search will differ depending on whether single or dual topology calculations are planned.
+Specifically, with a single topology approach in mind, atom types are allowed to change, so a permissive MCSS search can be done, whereas with dual topology a more strict search is required.
 \begin{figure}
@@
 
-In general there are mutliple ways on how a MCSS can be calculated.
-One of the most prominent approaches is to estimate the MCSS with a 2D subgraph isomorphism solver(SIS) using the 2D topological stuructre information of molecules\cite{raymond2002maximum}.
-A large number of software tools can compute SIS MCSS matches using different cheminformatics packages.
-Some rely on RDKit~\cite{rdkit2019Dec}, such as pmx~\cite{gapsys2015pmx}, LOMAP~\cite{liu2013lead}, FESetup~\cite{loeffler2015fesetup}, ProtoCaller~\cite{suruzhon2020protocaller}, SMArt~\cite{petrov2021perturbation}, and partially BioSimSpace~\cite{hedges2019biosimspace}, while others such as fkckombu~\cite{kawabata20143d} are standalone tools.
+There are different tools that allow the generation of MCSS matches as well as single topology input.
+A large number of software tools can compute MCSS matches using different cheminformatics packages.
+Some rely on RDKit~\cite{rdkit2019Dec}, such as pmx~\cite{gapsys2015pmx}, LOMAP~\cite{liu2013lead}, FESetup~\cite{loeffler2015fesetup} and partially BioSimSpace~\cite{hedges2019biosimspace}, while others such as fkckombu~\cite{kawabata20143d} are standalone tools.
 Schr\"{o}dinger's FEP+ planning tool was originally based on a version of LOMAP, and it also uses MCSS matching as well as 3D considerations to plan the network of single topology calculations between molecules~\cite{wang2015accurate}.
 
-SIS based methods, showed to struggle with 3D properties of the molecules, that need to be mapped, like for example stereochemistry.
-Therefore some packages developed a post-mapping step, where they filtered for atom distances, and therefore added 3D information into the mapping.
-Still SIS MCSS searches can be relatively time consuming as the algorithmic complexity is classified as NP-complete, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial similarity assessment and then use MCSS only to identify final mappings for relative calculations~\cite{raymond2002maximum,klabunde2012mars,jones2009elucidating, Duesbury2018Comparison}.
-The SIS MCSS approach, though relatively standard, takes into account only topological similarity.
+MCSS searches can be relatively time consuming, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial similarity assessment and then use MCSS only to identify final mappings for relative calculations~\cite{raymond2002maximum,klabunde2012mars,jones2009elucidating}.
+The MCSS approach, though relatively standard, takes into account only topological similarity.
 It is possible that changes in binding mode could actually require a different choice of mapping, so in some cases mappings may need to be planned differently depending on 3D positioning of atoms in space.
-
-Here an alternative way to estimate the MCSS is to calculate a bipartite graph matching based on the 3D positions of two molecules.
-These approaches tend to be very fast (can have algorithmic complexity of $O(n)=N^3$) and geometric accurate in their mapping (e.g.
-stereochemistry or binding mode differing).
-The geometric approaches additionally allow mappings of differing binding modes of the same molecule, or can be even used for significant larger molecules like proteins, due to the speed increase.
-A disadvantage of these approaches is the strong dependency on the coordinate sets being well aligned.
-Another aspect is taking the topology changes of the molecules not into account, which can be done if necessary in a next step with a filtering approach, like checking for changes in ring hybridization or ring breaking (this problem is discusse below)|.
-Tools that can be used to estimate the MCSS this way are contained for example in pmx \cite{gapsys2015pmx} or kartograf \cite{ries2024kartograf}.
-
-In any case Visual inspection prior to simulation is recommended to ensure that the mapping criteria correspond to the expected binding mode.
+Visual inspection prior to simulation is recommended to ensure that the mapping criteria correspond to the expected binding mode.
 If the mapping protocol returns simulations that correspond to different binding modes of a ligand within a perturbation map, this can cause large hysteresis.
 
-A next step for atom mappings, would be to support multistate methods (e.g.
-RE-EDS~\cite{sidler2016replica}, A-EDS~\cite{perthold2020toward} or Ladybugs~\cite{robo2022achieving}) with efficient mapping algorithms.
-There are protoype starting point for solving this problem, but to our knowledge, there is only few published work on this aspect.
-\cite{petrov2021Perturbation}
-
-Single coordinate set relative calculations and calculations based on substructure searches only work if in fact the ligands share a common substructure, e.g.
+Single topology relative calculations and calculations based on substructure searches only work if in fact the ligands share a common substructure, e.g.
 are part of a congeneric series, see Fig.~\ref{fig:fig_mcss}.
-If no common substructure is shared, then dual coordinate sets methods like linked-dual or separated topology free energy calculations are needed.
+If no common substructure is shared, then alternative dual or separated topology free energy calculations are needed.
 One would co-localize a pair of compounds in a binding site, exclude their interactions with one another, and compute the relative binding free energy by turning one molecule on from being dummy atoms while turning the other off.
-For such approaches it is crucial to find a good set of restraints, that keep the ligands colocalized over the time of the simulation.
-Few descriptions of the approach can be found for such approaches like the one suggested by Baumann et al.~\cite{baumann2023broadening} based on boresch style restraints~\cite{boresch2003absolute} for seperated topology approach, finding , or Restraintmaker~\cite{ries2022restraintmaker} for the linked-dualtopology approaches, which finds a small set of atoms to be linked via distance restraints, with a small distance to each other.
-However the implementation and testing of an fully automatized dual-topology pipeline for RBFE dual topology approaches is still to our knowledge a remaining research problem and was only tested with small datasets.
-\cite{jespers2019qligfep, Rieder2022Lev}
-Alternativley an approach where such a restraint algorithm was included in a pipeline for ABFE calculations is described by the ABFE\_workflow \cite{Ries2024ABFE} using MDRestraintsGenerator\cite{alibay2021mdrestraintsgenerator}.
+To our knowledge no general pipeline for such calculations yet exists and this would likely remain a research problem.
+Using an absolute free energy approach instead seems more promising in such a case.
 
 \paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
-Some approaches have attempted to address thisusing a single coordinate sets and with soft bond approaches ~\cite{clark2019relative,wang2017accurate} or dual coordinate set approach \cite{jespers2019qligfep, ries2022relative} but a general solution is not yet in mainstream use.
-Still, FEP+ implements one solution\cite{wang2017accurate}.
+Some approaches have attempted to address this~\cite{clark2019relative} but a general solution is not yet in mainstream use. Still, FEP+ implements one solution.
 

### 2025-04-29 paper/manuscript.tex
Start merging a new dummy atom section  On branch sboresch
 Changes to be committed:
	modified:   manuscript.tex
	modified:   sb.bib

--- 2025-03-12 paper/manuscript.tex
@@
 
+\paragraph{Dummy atoms}
+
+In practically all relative free energy simulations, regardless of the approach, SCST, SCDT, or DCDT, dummy atoms (non-interacting placeholder atoms) are required.
+We stress this point, as sometimes dummy atoms are viewed as something specific to SCST.
+In SC approaches, dummy atoms may even be required if the number of atoms between initial and final state of the system is identical, consider, e.g., tautomeric forms of a molecule.
+
+The need for and role of dummy atoms is clearly visible in Fig.~\ref{fig:fig_topology}: One or more dummy atoms are attached to the physical molecule.
+A crucial requirement is that these atoms, typically connected through bonded energy terms (bond stretching, angle bending, dihedral angles) do not affect equilibrium and dynamical properties of the physical system.
+However, since a physical molecule + dummy atoms attached to it can or must be viewed as a modified physical molecule, the partition function of the molecules with and without dummy atoms will NOT be identical.
+In other words, a single free energy difference between two molecules A and B will be different from that of A(+D) and B(+D), where the (+D) indicates the presence of one or more dummy atoms needed so that the alchemical transformation can be carried out.
+When a dual coordinate approach is used, the noninteracting part of the respective other endstate is tied to the physical one, and will contribute to the partition function.
+By construction, relative free energy simulatins involve two analogous alchemical transformations, e.g., A(+D) -> B(+D) in gas phase and aqueous solution (RSFEs), or in the bound and unbound state (RBFEs).
+If dummy atoms are interacting with the physical system in such a way that $\mathsf{Z = Z(R) \times Z(R\text{-}D) \times Z(D)}$, where Z is the total partition function, Z(R) the partition function of the physical system, Z(D) the partition function contribution resulting from interactions between only the dummy atoms, and Z(R-D) the contribution resulting from the bonded connections between the physical atoms and dummy atoms, then their contributions (Z(D) and Z(R-D)) factor from the total partition function.
+As dummy atoms do not interact with the environment, their contribution, therefore, is identical in the two alchemical transformations required in the two legs of the relative transformation and will cancel exactly from the double free energy difference of interest.
+
+While some early work on the correct treatment of dummy atoms, i.e., how to treat them so that their contribution to the partition function factors, has been available for years \cite{Boresch_2002,Shobana_2000,Wang_2012}, most workers seem to mostly turn off nonbonded interactions for dummy atoms, but leaving all bonded interactions intact \sbnote{To all: While I am quite sure that this is what most of us have been doing most of the time, I wonder whether this has been epxlicitly described anywhere?}.
+Since current force fields use multiple angle and torsion terms for each atom, there are often a large number of bonded energy terms connecting dummy atoms to the physical molecule.
+However, as already described in the early studies and repeated in a more recent study focusing on the correct treatment of dummy atoms \cite{Fleck_2021}, the desired factorization can only be guaranteed if any dummy atom connected to the physical molecule is held by exactly three (non-redundant) degrees of freedom.
+Naive application of the three-degree-of-freedom per dummy atom rule (for dummy atoms having at least one joint bonded term with the physical molecule), however, can result in simulations where the dummy part adopts unphysical configurations/conformations; Fleck et al.
+referred to this as “flapping”.
+Thus, depending on the details of the physical molecule -- dummy atom(s) junction, the three degree of freedoms need to be selected with care, or a carefully chosen additional redundant degree of freedom may be needed.
+In the latter case, the factorization of the partition function is not exact, but the overall error can still be made arbitrarily small.
+For the full details, we refer the reader to Ref.~\cite{Fleck_2021}.
+
+One situation involving dummy atoms that needs to be avoided is the creation of a cyclic structure involving dummy atoms, i.e., whenever the (group of) dummy atom(s) is attached to two different parts of the physical molecule.
+In this case, the factorization of the partition function is never possible \cite{Shobana_2000}.
+It follows that ring opening or closing is extremely challenging to carry out using SCST approaches \cite{liu2015ring}; these difficulties have in part motivated the development and adoption of dual coordinate approaches.
+For more on ring opening/closing, see also below.
+However, DC approaches are also extremely prone to exactly this type of problem: to keep the noninteracting molecule on top of the interacting one, typically multiple bonded restraints (e.g.
+harmonic bonds) are used.
+These effectively, are equivalent to the formation of cyclic structures with potential erroneous contributions to the double free energy difference of interest.
+
+The influence from an incorrect treatment of dummy atoms (e.g., leaving all bonded terms intact) is difficult to estimate, though it may be small in many cases \cite{Fleck2021}.
+Nevertheless, some guidelines should be followed:
+\begin{itemize}
+  \item Obviously, a system with and without dummy atoms should have the same energetic minimum.
+Note that in larger groups of dummy atoms, redundant bonded terms involving only dummy atoms will be essential to avoid flapping, and the energy of the optimized geometry may be different from that of the physical system without dummy atoms.
+Thus, a \underline{necessary} condition is as follows: Optimize the geometry of the physical system (no dummy atoms); the final energy is the reference value.
+Optimize the geometry of the same molecule in the presence of dummy atoms; the final energy may well be different from the reference energy.
+Now delete the dummy atoms and recompute the energy without further minimization.
+The energy after the deletion of the dummy atoms must be identical to the reference energy.
+\item The above is only a \underline{necessary} criterion, and the dummy atoms may still influence the dynamical behavior of the physical molecule.
+Furthermore, overzealous removal of redundant energy terms may result in flapping of the dummy atoms.
+If one suspects this to be the case, then a short simulation in the gas phase should reveal flapping.
+Unfortunately, while straightforward in principle, both checks described are extremely hard to automate, which is essential in large RFEs campaigns.
+\item The correct treatment of dummy atoms can be validated by closing a suitable thermodynamic cycle.
+E.g., Fleck et al.
+calculated ASFEs for each of their endstates, as well as the RSFE between them.
+In all cases when dummy atoms were treated according to best practices, the cycle closure error was well within the statistical uncertainty, which was well below 0.1 kcal/mol.
+A related, earlier study by Loeffler et al.
+also reported ASFEs and RSFEs.
+If one attempts to close the cycle based on the numbers reported there, cycle closure error was nonnegligible in some cases.
+[Loeffler2017]
+\item While solvation free energies can nowadays be computed quite fast, their systematic calculation would nevertheless be too costly.
+The role of solvent can be replaced by some external restraint, distorting the average geometry of the physical molecule.
+If one computes the free energy cost of applying the restraint with and without dummy atoms, the two numbers must agree within statistical significance.
+While much faster than full solvation free energy calculations, applying a restraint in the gas phase requires careful thermalization of the simulation system, and automation is difficult.
+  \end{itemize}
+Further work automating the correct setup and treatment of dummy atoms is necessary.
+
 \paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.

### 2025-04-29 paper/manuscript.tex
Quick and dirty first complete version; now wait for feedback  On branch sboresch
 Changes to be committed:
	modified:   manuscript.tex

@@
 
-The influence from an incorrect treatment of dummy atoms (e.g., leaving all bonded terms intact) is difficult to estimate, though it may be small in many cases \cite{Fleck2021}.
+The influence from an incorrect treatment of dummy atoms (e.g., leaving all bonded terms intact) is difficult to estimate, though it may be small in many cases \cite{Fleck_2021}.
 Nevertheless, some guidelines should be followed:
@@
 Optimize the geometry of the same molecule in the presence of dummy atoms; the final energy may well be different from the reference energy.
-Now delete the dummy atoms and recompute the energy without further minimization.
+Now delete the dummy atoms and recompute the energy \underline{without} further minimization.
 The energy after the deletion of the dummy atoms must be identical to the reference energy.
@@
 Furthermore, overzealous removal of redundant energy terms may result in flapping of the dummy atoms.
-If one suspects this to be the case, then a short simulation in the gas phase should reveal flapping.
-Unfortunately, while straightforward in principle, both checks described are extremely hard to automate, which is essential in large RFEs campaigns.
+If one suspects the latter, then a short simulation in the gas phase should reveal flapping.
+Unfortunately, while straightforward in principle, both checks described are hard to automate, which is essential in large RFEs campaigns.
 \item The correct treatment of dummy atoms can be validated by closing a suitable thermodynamic cycle.
-E.g., Fleck et al.
-calculated ASFEs for each of their endstates, as well as the RSFE between them.
+E.g., Fleck et al.\ calculated ASFEs for each of their endstates, as well as the RSFE between them \cite{Fleck_2021}.
 In all cases when dummy atoms were treated according to best practices, the cycle closure error was well within the statistical uncertainty, which was well below 0.1 kcal/mol.
@@
 also reported ASFEs and RSFEs.
-If one attempts to close the cycle based on the numbers reported there, cycle closure error was nonnegligible in some cases.
-[Loeffler2017]
+If one attempts to close the cycle based on the numbers reported there, cycle closure error was nonnegligible in some cases \cite{loeffler2018reproducibility}.
 \item While solvation free energies can nowadays be computed quite fast, their systematic calculation would nevertheless be too costly.
@@
   \end{itemize}
-Further work automating the correct setup and treatment of dummy atoms is necessary.
+%Further work automating the correct setup and treatment of dummy atoms is necessary.
 

### 2025-05-06 paper/manuscript.tex
Add paragraph on membrane systems

--- 2025-04-04 paper/manuscript.tex
@@
 \author[2]{Huafeng Xu}
-\author[11]{William G. Poole}
 %
@@
 \affil[10]{Cresset, Cambridgeshire, UK}
-\affil[11]{School of Chemistry, University of Southampton, Southampton, SO17 1BJ}
 
@@
 \tocsectioncomment{@Irfan Alibay/ @Emilio Gallicchio/ @Volunteer to expand on approaches for explicitly accounting for protonation/ tautomer changes.}
-\tocsectioncomment{@Hannah Bruce Macdonald/ @David Slowchower/ @Vytas Gapsys/ @Volunteer to expand range of systems discussed, e.g. membrane proteins, cofactors, multimers, PPIs.}
+\tocsectioncomment{@Hannah Bruce Macdonald/ @David Slochower/ @Vytas Gapsys/ @Volunteer to expand range of systems discussed, e.g. membrane proteins, cofactors, multimers, PPIs.}
 \label{sec:prerequisites}
@@
 Well-known protein systems with water mediated ligand interactions are for example: HSP90 which formed part of the D3R grand challenge 2015~\cite{mey2016blinded}, A2A~\cite{brucemacdonald2018ligand}, MUP~\cite{ross2015water}, ~\cite{deflorian2020accurate}, and others~\cite{michel2009energetics}.
+
+\paragraph{Membrane proteins require additional care}
+Integral membrane proteins require embedding in a suitable model membrane to remain stable during simulations.
+Lipids are often not included in protein force fields and, although they may be covered by generic small molecule force fields, have often been parameterized separately (e.g., AMBER's Lipid21 \cite{AMBERLipid21}, CHARMM lipid force field \cite{CHARMMLipidFF}, and OpenFF's upcoming (?) ``biopolymer" force field).
+Membrane systems can be simulated in multiple thermodynamic ensembles, but the semi-isotropic scheme $NP_{z}P_{xy}T$ is frequently employed, allowing the pressure in the bilayer plane ($xy$) to scale independently \cite{LipidBestPracticesCOMS}.
+These systems require more extensive equilibration than soluble proteins.
+Some packages, such as Desmond, include a 7-step protocol to relax membrane systems.
+
+In the case of a lipid-facing binding site, there are two additional challenges.
+First, simulation time should be extended to ensure sufficient sampling of lipid configurational space.
+Second, the free energy of binding for a small molecule may be strongly coupled to the lipid species.
+In some cases, the identity of the lipid may be known.
+But in other cases, it may be difficult to discern the identity of the lipid from X-ray or CryoEM structures due to the flexible nature of lipid head groups and benchmarking is recommended to match experimental data.
 

### 2025-05-11 paper/manuscript.tex
Minor changes

@@
 \tocsectioncomment{@Irfan Alibay/ @Emilio Gallicchio/ @Volunteer to expand on approaches for explicitly accounting for protonation/ tautomer changes.}
-\tocsectioncomment{@Hannah Bruce Macdonald/ @David Slochower/ @Vytas Gapsys/ @Volunteer to expand range of systems discussed, e.g. membrane proteins, cofactors, multimers, PPIs.}
+\tocsectioncomment{@Hannah Bruce Macdonald/ @David Slowchower/ @Vytas Gapsys/ @Volunteer to expand range of systems discussed, e.g. membrane proteins, cofactors, multimers, PPIs.}
 \label{sec:prerequisites}
@@
 Well-known protein systems with water mediated ligand interactions are for example: HSP90 which formed part of the D3R grand challenge 2015~\cite{mey2016blinded}, A2A~\cite{brucemacdonald2018ligand}, MUP~\cite{ross2015water}, ~\cite{deflorian2020accurate}, and others~\cite{michel2009energetics}.
-
-\paragraph{Membrane proteins require additional care}
-Integral membrane proteins require embedding in a suitable model membrane to remain stable during simulations.
-Lipids are often not included in protein force fields and, although they may be covered by generic small molecule force fields, have often been parameterized separately (e.g., AMBER's Lipid21 \cite{AMBERLipid21}, CHARMM lipid force field \cite{CHARMMLipidFF}, and OpenFF's upcoming (?) ``biopolymer" force field).
-Membrane systems can be simulated in multiple thermodynamic ensembles, but the semi-isotropic scheme $NP_{z}P_{xy}T$ is frequently employed, allowing the pressure in the bilayer plane ($xy$) to scale independently \cite{LipidBestPracticesCOMS}.
-These systems require more extensive equilibration than soluble proteins.
-Some packages, such as Desmond, include a 7-step protocol to relax membrane systems.
-
-In the case of a lipid-facing binding site, there are two additional challenges.
-First, simulation time should be extended to ensure sufficient sampling of lipid configurational space.
-Second, the free energy of binding for a small molecule may be strongly coupled to the lipid species.
-In some cases, the identity of the lipid may be known.
-But in other cases, it may be difficult to discern the identity of the lipid from X-ray or CryoEM structures due to the flexible nature of lipid head groups and benchmarking is recommended to match experimental data.
 
@@
 \paragraph{Atom mapping}
-Once a particular approach to the topology is selected, a crucial next step is to identify the common atoms which will not be perturbed.
-Rigorously, this process typically comprises a MCSS search of the molecules involved to identify the common substructure---though the parameters of the MCSS search will differ depending on whether single or dual topology calculations are planned.
-Specifically, with a single topology approach in mind, atom types are allowed to change, so a permissive MCSS search can be done, whereas with dual topology a more strict search is required.
+Following the choice of system representation and sampling strategy, it might be required to generate a atom mapping.
+This is the case especially for system representations using a single coordinate set (Type 1.
+& 2), however a dual coordinate set approach could use an atom mapping as well, depending on the sampling method, this will be explained in the next steps.
+Once a single coordinate set approach is selected is selected, a crucial next step is to identify the common atoms which will not be perturbed between the two molecules (respectivly end-states).
+Rigorously, this process typically comprises a MCSS search of the molecules involved to identify the common substructure.
+Specifically, with a single coordinate set approach in mind, if atom types are allowed to change, a permissive MCSS search can be done.
+Atom-type changes however requires either careful implementation of atom mass changes, and bonded parameters or accepting the introduction of potential systematic error, that is in many cases negliable due to the cancelation of errors.
+\cite{ries2024Kartograf}
+
 \begin{figure}
@@
 
-There are different tools that allow the generation of MCSS matches as well as single topology input.
-A large number of software tools can compute MCSS matches using different cheminformatics packages.
-Some rely on RDKit~\cite{rdkit2019Dec}, such as pmx~\cite{gapsys2015pmx}, LOMAP~\cite{liu2013lead}, FESetup~\cite{loeffler2015fesetup} and partially BioSimSpace~\cite{hedges2019biosimspace}, while others such as fkckombu~\cite{kawabata20143d} are standalone tools.
+In general,  there are mutliple ways on how a MCSS can be calculated.
+One of the most prominent approaches is to estimate the MCSS with a 2D subgraph isomorphism solver (SIS) using the 2D topological stuructre information of molecules\cite{raymond2002maximum}.
+A large number of software tools can compute SIS MCSS matches using different cheminformatics packages.
+Some rely on RDKit~\cite{rdkit2019Dec}, such as pmx~\cite{gapsys2015pmx}, LOMAP~\cite{liu2013lead}, FESetup~\cite{loeffler2015fesetup}, ProtoCaller~\cite{suruzhon2020protocaller}, SMArt~\cite{petrov2021perturbation}, and partially BioSimSpace~\cite{hedges2019biosimspace}, while others such as fkckombu~\cite{kawabata20143d} are standalone tools.
 Schr\"{o}dinger's FEP+ planning tool was originally based on a version of LOMAP, and it also uses MCSS matching as well as 3D considerations to plan the network of single topology calculations between molecules~\cite{wang2015accurate}.
 
-MCSS searches can be relatively time consuming, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial similarity assessment and then use MCSS only to identify final mappings for relative calculations~\cite{raymond2002maximum,klabunde2012mars,jones2009elucidating}.
-The MCSS approach, though relatively standard, takes into account only topological similarity.
+SIS based methods struggle with 3D properties of the molecules that need to be mapped, for example stereochemistry.
+Therefore some packages developed a post-mapping step, where matched atoms are filtered to ensure mapped atom distances are within a cutoff, which therefore adds 3D information into the mapping.
+SIS MCSS searches can still be relatively time consuming as the algorithmic complexity is classified as NP-complete, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial similarity assessment and then use MCSS only to identify final mappings for relative calculations~\cite{raymond2002maximum,klabunde2012mars,jones2009elucidating, Duesbury2018Comparison}.
+The SIS MCSS approach, though relatively standard, takes into account only topological similarity.
 It is possible that changes in binding mode could actually require a different choice of mapping, so in some cases mappings may need to be planned differently depending on 3D positioning of atoms in space.
-Visual inspection prior to simulation is recommended to ensure that the mapping criteria correspond to the expected binding mode.
+
+An alternative way to estimate the MCSS is to calculate a bipartite graph matching based on the 3D positions of two molecules.
+These approaches tend to be very fast (can have algorithmic complexity of $O(n)=N^3$) and are geometrically accurate in their mapping (e.g.
+stereochemistry or binding mode differing).
+The geometric approaches additionally allow mappings of differing binding modes of the same molecule, which are not captured by 2D mappings, or can be even used for significant larger molecules like proteins, due to the speed increase.
+A disadvantage of these approaches is the strong reliance on the coordinate sets being well aligned.
+Additionally, not taking the topology changes of the molecules into account, can introduce errors such as checking for changes in ring hybridization or ring breaking (this problem is discusse below).
+Tools that can be used to estimate the MCSS this way are contained for example in pmx \cite{gapsys2015pmx} or kartograf \cite{ries2024kartograf}.
+
+In any case, visual inspection prior to simulation is recommended to ensure that the mapping criteria correspond to the expected binding mode.
 If the mapping protocol returns simulations that correspond to different binding modes of a ligand within a perturbation map, this can cause large hysteresis.
 
-Single topology relative calculations and calculations based on substructure searches only work if in fact the ligands share a common substructure, e.g.
+A next step for atom mappings, would be to support multistate methods (e.g.
+RE-EDS~\cite{sidler2016replica}, A-EDS~\cite{perthold2020toward} or Ladybugs~\cite{robo2022achieving}) with efficient mapping algorithms.
+There are protoype starting point for solving this problem, but to our knowledge, there is only few published work on this aspect.
+\cite{petrov2021Perturbation}
+
+Single coordinate set relative calculations and calculations based on substructure searches only work if in fact the ligands share a common substructure, e.g.
 are part of a congeneric series, see Fig.~\ref{fig:fig_mcss}.
-If no common substructure is shared, then alternative dual or separated topology free energy calculations are needed.
+If no common substructure is shared, then dual coordinate sets methods like linked-dual or separated topology free energy calculations are needed.
 One would co-localize a pair of compounds in a binding site, exclude their interactions with one another, and compute the relative binding free energy by turning one molecule on from being dummy atoms while turning the other off.
-To our knowledge no general pipeline for such calculations yet exists and this would likely remain a research problem.
-Using an absolute free energy approach instead seems more promising in such a case.
+For such approaches it is crucial to find a good set of restraints, that keep the ligands colocalized over the time of the simulation.
+Few descriptions of the approach can be found for such approaches like the one suggested by Baumann et al.~\cite{baumann2023broadening} based on boresch style restraints~\cite{boresch2003absolute} for seperated topology approach, finding , or Restraintmaker~\cite{ries2022restraintmaker} for the linked-dualtopology approaches, which finds a small set of atoms to be linked via distance restraints, with a small distance to each other.
+However the implementation and testing of an fully automatized dual-topology pipeline for RBFE dual topology approaches is still to our knowledge a remaining research problem and was only tested with small datasets.
+\cite{jespers2019qligfep, Rieder2022Lev}
 
 \paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
-Some approaches have attempted to address this~\cite{clark2019relative} but a general solution is not yet in mainstream use. Still, FEP+ implements one solution.
+Some approaches have attempted to address thisusing a single coordinate sets and with soft bond approaches ~\cite{clark2019relative,wang2017accurate} or dual coordinate set approach \cite{jespers2019qligfep, ries2022relative} but a general solution is not yet in mainstream use.
+Still, FEP+ implements one solution\cite{wang2017accurate}.
 

### 2025-05-12 paper/manuscript.tex
added small sentence part for boresch style approach

@@
 Rigorously, this process typically comprises a MCSS search of the molecules involved to identify the common substructure.
-Specifically, with a single coordinate set approach in mind, if atom types are allowed to change, a permissive MCSS search can be done.
-Atom-type changes however requires either careful implementation of atom mass changes, and bonded parameters or accepting the introduction of potential systematic error, that is in many cases negliable due to the cancelation of errors.
+Specifically, with a single coordinate set approach in mind, atom types are allowed to change, so a permissive MCSS search can be done.
+This however requires either careful implementation of atom mass changes, bond length etc.
+or accepting the introduction of potential systematic error, that is in many cases negliable due to the cancelation of errors.
 \cite{ries2024Kartograf}
@@
 
-In general,  there are mutliple ways on how a MCSS can be calculated.
-One of the most prominent approaches is to estimate the MCSS with a 2D subgraph isomorphism solver (SIS) using the 2D topological stuructre information of molecules\cite{raymond2002maximum}.
+In general there are mutliple ways on how a MCSS can be calculated.
+One of the most prominent approaches is to estimate the MCSS with a 2D subgraph isomorphism solver(SIS) using the 2D topological stuructre information of molecules\cite{raymond2002maximum}.
 A large number of software tools can compute SIS MCSS matches using different cheminformatics packages.
@@
 
-SIS based methods struggle with 3D properties of the molecules that need to be mapped, for example stereochemistry.
-Therefore some packages developed a post-mapping step, where matched atoms are filtered to ensure mapped atom distances are within a cutoff, which therefore adds 3D information into the mapping.
-SIS MCSS searches can still be relatively time consuming as the algorithmic complexity is classified as NP-complete, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial similarity assessment and then use MCSS only to identify final mappings for relative calculations~\cite{raymond2002maximum,klabunde2012mars,jones2009elucidating, Duesbury2018Comparison}.
+SIS based methods, showed to struggle with 3D properties of the molecules, that need to be mapped, like for example stereochemistry.
+Therefore some packages developed a post-mapping step, where they filtered for atom distances, and therefore added 3D information into the mapping.
+Still SIS MCSS searches can be relatively time consuming as the algorithmic complexity is classified as NP-complete, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial similarity assessment and then use MCSS only to identify final mappings for relative calculations~\cite{raymond2002maximum,klabunde2012mars,jones2009elucidating, Duesbury2018Comparison}.
 The SIS MCSS approach, though relatively standard, takes into account only topological similarity.
@@
 
-An alternative way to estimate the MCSS is to calculate a bipartite graph matching based on the 3D positions of two molecules.
-These approaches tend to be very fast (can have algorithmic complexity of $O(n)=N^3$) and are geometrically accurate in their mapping (e.g.
+Here an alternative way to estimate the MCSS is to calculate a bipartite graph matching based on the 3D positions of two molecules.
+These approaches tend to be very fast (can have algorithmic complexity of $O(n)=N^3$) and geometric accurate in their mapping (e.g.
 stereochemistry or binding mode differing).
-The geometric approaches additionally allow mappings of differing binding modes of the same molecule, which are not captured by 2D mappings, or can be even used for significant larger molecules like proteins, due to the speed increase.
-A disadvantage of these approaches is the strong reliance on the coordinate sets being well aligned.
-Additionally, not taking the topology changes of the molecules into account, can introduce errors such as checking for changes in ring hybridization or ring breaking (this problem is discusse below).
-Tools that can be used to estimate the MCSS this way are contained for example in pmx \cite{gapsys2015pmx} or kartograf \cite{ries2024kartograf}.
-
-In any case, visual inspection prior to simulation is recommended to ensure that the mapping criteria correspond to the expected binding mode.
+The geometric approaches additionally allow mappings of differing binding modes of the same molecule, or can be even used for significant larger molecules like proteins, due to the speed increase.
+A disadvantage of these approaches is the strong dependency on the coordinate sets being well aligned.
+Another aspect is taking the topology changes of the molecules not into account, which can be done if necessary in a next step with a filtering approach, like checking for changes in ring hybridization or ring breaking (this problem is discusse below)|.
+Tools that can be used to estimate the MCSS this way are contained for example in pmx \cite{gapsys2015pmx} or Kartograf \cite{ries2024kartograf}.
+
+In any case Visual inspection prior to simulation is recommended to ensure that the mapping criteria correspond to the expected binding mode.
 If the mapping protocol returns simulations that correspond to different binding modes of a ligand within a perturbation map, this can cause large hysteresis.
@@
 For such approaches it is crucial to find a good set of restraints, that keep the ligands colocalized over the time of the simulation.
-Few descriptions of the approach can be found for such approaches like the one suggested by Baumann et al.~\cite{baumann2023broadening} based on boresch style restraints~\cite{boresch2003absolute} for seperated topology approach, finding , or Restraintmaker~\cite{ries2022restraintmaker} for the linked-dualtopology approaches, which finds a small set of atoms to be linked via distance restraints, with a small distance to each other.
+Few descriptions of the approach can be found for such approaches like the one suggested by \textit{Baumann et al.}~\cite{baumann2023broadening} based on boresch style restraints~\cite{boresch2003absolute}, with an elaborate automatic atom selection approach, for the seperated topology approach, or Restraintmaker~\cite{ries2022restraintmaker} for the linked-dualtopology approaches, which finds a small set of atoms to be linked via distance restraints, with a small distance to each other.
 However the implementation and testing of an fully automatized dual-topology pipeline for RBFE dual topology approaches is still to our knowledge a remaining research problem and was only tested with small datasets.
@@
 \paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
-Some approaches have attempted to address thisusing a single coordinate sets and with soft bond approaches ~\cite{clark2019relative,wang2017accurate} or dual coordinate set approach \cite{jespers2019qligfep, ries2022relative} but a general solution is not yet in mainstream use.
+Some approaches have attempted to address this using a single coordinate sets and with soft bond approaches ~\cite{clark2019relative,wang2017accurate} or dual coordinate set approach \cite{jespers2019qligfep, ries2022relative} but a general solution is not yet in mainstream use.
 Still, FEP+ implements one solution\cite{wang2017accurate}.

### 2025-05-17 paper/manuscript.tex
Revise membrane paragraph based on Irfan feedback

--- 2025-05-06 paper/manuscript.tex
@@
 Integral membrane proteins require embedding in a suitable model membrane to remain stable during simulations.
-Lipids are often not included in protein force fields and, although they may be covered by generic small molecule force fields, have often been parameterized separately (e.g., AMBER's Lipid21 \cite{AMBERLipid21}, CHARMM lipid force field \cite{CHARMMLipidFF}, and OpenFF's upcoming (?) ``biopolymer" force field).
+Lipids are usually not included in protein force fields and, although they may be covered by generic small molecule force fields, have often been parameterized separately (e.g., AMBER's Lipid21 \cite{AMBERLipid21} and the CHARMM lipid force field \cite{CHARMMLipidFF}).
 Membrane systems can be simulated in multiple thermodynamic ensembles, but the semi-isotropic scheme $NP_{z}P_{xy}T$ is frequently employed, allowing the pressure in the bilayer plane ($xy$) to scale independently \cite{LipidBestPracticesCOMS}.
@@
 
+Alchemical transformations that involve net charge changes can provide difficulties for membrane systems; phospholipids carry functional groups with nonzero formal charge and cannot be treated as a solvent with a uniform dielectric in the same way as water.
+Several approaches are reviewed in \cite{wu2022correction-51c} and the recommended method is to use a co-alchemical ion to maintain system neutrality.
+
 In the case of a lipid-facing binding site, there are two additional challenges.
 First, simulation time should be extended to ensure sufficient sampling of lipid configurational space.
+Lipids may either bind alongside small molecules, acting as cofactors, or be displaced from the binding site by a ligand.
+If lipids participate in binding through protein-lipid or ligand-lipid interactions, it is necessary to sample these intermolecular interactions through additional $\lambda$ windows, longer simulation times, or different alchemical protocols.
+Displacing a lipid will have its own free energy cost which may take a long time to converge if it requires diffusion and rearrangement of nearby lipids.
 Second, the free energy of binding for a small molecule may be strongly coupled to the lipid species.
 In some cases, the identity of the lipid may be known.
-But in other cases, it may be difficult to discern the identity of the lipid from X-ray or CryoEM structures due to the flexible nature of lipid head groups and benchmarking is recommended to match experimental data.
+But in other cases, it may be difficult to discern the identity of the lipid from X-ray or cryo-EM structures due to the flexible nature of lipid head groups and benchmarking is recommended to match experimental data.
 

### 2025-05-21 paper/manuscript.tex
Water Sampling

@@
 \author[2]{Huafeng Xu}
+\author[11]{Oliver J. Melling}
+\author[11]{William G. Poole}
+\author[11]{Jonathan W. Essex}
+
 %
@@
 Binding site water molecules may form water mediated protein-ligand interactions which can pose challenges whenever exchange with bulk water is slow compared to simulation timescales.
-This happens typically in buried binding sites.
-Overlaying multiple protein X-ray structures can identify conserved or additional water molecules that can be useful to include in calculations.
-In cases where water molecules are known to play an important role in the binding, software implementations that use water sampling facilitated by Grand Canonical Monte Carlo methods may be useful~\cite{michel2010prediction}.
-Other tools such as WaterMap or open source equivalents (SSTMap, GIST, and others) can be used to define water structure for systems with no experimental evidence of water sites~\cite{wang2011ligand}.
+This can happen in buried binding sites~\cite{laage2017water}.
+Overlaying multiple protein X-ray structures can identify conserved water molecules or those that are associated only with certain ligands in a series and are therefore important to include in calculations.
+In cases where water molecules are known to play an important role in ligand binding, for example if the presence of electron density suggests they are tightly bound (indicated by a high EDIA score~\cite{nittinger2015evidence,meyder2017estimating}), software implementations that use water sampling facilitated by Grand Canonical Monte Carlo methods may be useful~\cite{michel2010prediction}.
+Other tools such as WaterMap~\cite{abel2008role,young2007motifs} or open source equivalents (SSTMap~\cite{haider2018solvation}, GIST~\cite{ramsey2016solvation}, and others) can be used to define the hydration states for systems with no experimental evidence of water sites~\cite{wang2011ligand}.
 Well-known protein systems with water mediated ligand interactions are for example: HSP90 which formed part of the D3R grand challenge 2015~\cite{mey2016blinded}, A2A~\cite{brucemacdonald2018ligand}, MUP~\cite{ross2015water}, ~\cite{deflorian2020accurate}, and others~\cite{michel2009energetics}.
 
-\paragraph{Membrane proteins require additional care}
-Integral membrane proteins require embedding in a suitable model membrane to remain stable during simulations.
-Lipids are usually not included in protein force fields and, although they may be covered by generic small molecule force fields, have often been parameterized separately (e.g., AMBER's Lipid21 \cite{AMBERLipid21} and the CHARMM lipid force field \cite{CHARMMLipidFF}).
-Membrane systems can be simulated in multiple thermodynamic ensembles, but the semi-isotropic scheme $NP_{z}P_{xy}T$ is frequently employed, allowing the pressure in the bilayer plane ($xy$) to scale independently \cite{LipidBestPracticesCOMS}.
-These systems require more extensive equilibration than soluble proteins.
-Some packages, such as Desmond, include a 7-step protocol to relax membrane systems.
-
-Alchemical transformations that involve net charge changes can provide difficulties for membrane systems; phospholipids carry functional groups with nonzero formal charge and cannot be treated as a solvent with a uniform dielectric in the same way as water.
-Several approaches are reviewed in \cite{wu2022correction-51c} and the recommended method is to use a co-alchemical ion to maintain system neutrality.
-
-In the case of a lipid-facing binding site, there are two additional challenges.
-First, simulation time should be extended to ensure sufficient sampling of lipid configurational space.
-Lipids may either bind alongside small molecules, acting as cofactors, or be displaced from the binding site by a ligand.
-If lipids participate in binding through protein-lipid or ligand-lipid interactions, it is necessary to sample these intermolecular interactions through additional $\lambda$ windows, longer simulation times, or different alchemical protocols.
-Displacing a lipid will have its own free energy cost which may take a long time to converge if it requires diffusion and rearrangement of nearby lipids.
-Second, the free energy of binding for a small molecule may be strongly coupled to the lipid species.
-In some cases, the identity of the lipid may be known.
-But in other cases, it may be difficult to discern the identity of the lipid from X-ray or cryo-EM structures due to the flexible nature of lipid head groups and benchmarking is recommended to match experimental data.
+In the context of relative binding free energy calculations, determining the correct hydration pattern for the two endpoints may not be sufficient for an accurate and reproducible calculation.
+If the locations of the bound water molecules differs at the two endpoints - i.e.
+if a perturbation to the ligand displaces a bound water molecule - then the network needs to be able to adapt during the perturbation.
+Otherwise, it is likely that the estimated relative binding free energies will be dependent on the direction in which the perturbation is performed~\cite{ross2020enhancing}.
+
+In absolute binding free energy calculations, a ligand is fully decoupled from a binding site cavity which may otherwise be hydrated in the \textit{apo} state.
+In this case, the sampling of water molecules must be sufficient to hydrate the binding pocket as the ligand is decoupled (or vice versa for the reverse).
+When the binding site is exposed and water can move freely between the site and bulk solvent, then sampling is often not an issue.
+However, in the case of more buried sites or sites which form complex water networks, enhanced water sampling may be required to ensure an accurate binding affinity.
 
@@
 We are not yet aware of a careful comparison of charge corrections versus other approaches such as decoupling an ion at the same time, so in our view the issue of proper handling of charge mutations in the context of alchemical calculations remains a research problem, and several papers give good guides for exploring the problem further~\cite{lin2014overview, ohlknecht2020correcting, rocklin2013calculating}.
+
+\subsubsection{Handling poorly sampled water networks}
+The location of any water molecules within the binding site of interest must be determined prior to carrying out a free energy calculation.
+Experimental structures, such as those determined by X-ray and cryo-EM, are often sufficient for this, provided the resolution is low enough.
+However, we recommend using grand canonical sampling methods\cite{melling2023enhanced, samways2020grand} as a means of verifying the experimental water placement and determining any potential differences owing to the classical forcefield used.
+Many other water placement methods are available, all varying in their methodology, computational cost and accuracy.
+Details of these methods can be found in the publication from Samways et al.\cite{samways2021water}.
+
+In relative calculations, if the hydration patterns differ between the two ligand endpoints then careful consideration will need to be given to the adaptation of the water network during the free energy calculation.
+We recommend performing the free energy calculation in the conventional way, i.e.
+with no enhanced water sampling, to determine whether the water network can adapt to the ligand perturbation itself.
+This is likely to happen if the hydration site is not particularly occluded from bulk solvent and water molecules can move freely from bulk into the binding site.
+
+If a conventional sampling method, such as molecular dynamics or Monte Carlo, is unable to rearrange the water network with sufficient efficacy to accommodate the perturbation then an enhanced sampling method will be required to achieve accurate and converged results.
+One such approach is to include an absolute free energy calculation of the water molecule into the thermodynamic cycle\cite{ross2020enhancing, hamelberg2004standard, barillari2007classification, yu2008free, ge2022abfewat} such that the free energies of both the ligand transformation, and the decoupling of any displaced water molecules, are calculated separately and combined to achieve the total free energy change for the process.
+This method can be difficult to use in practice as the water molecule needs to be carefully restrained while being decoupled, and care needs to be taken to ensure water does not diffuse back into the hydration site.
+
+Monte Carlo approaches such as the hybrid MC/MD approach proposed by Bergazin et al.\cite{bergazin2021enhancing, ben-shalom2021fast, ge2022enhancing} allow waters to move between bulk solvent and the binding site with greater frequency by proposing MC moves that translate the water molecules within a spherical region that encompasses both bulk solvent and the binding site.
+This avoids the difficulty of performing a decoupling calculation on the bound waters but can be time-consuming and risk computational time being wasted on move proposals that are not of interest.
+
+Grand canonical (GC) methods propose the addition and removal of water molecules to and from a binding site in accordance with a defined chemical potential.
+Several implementations of GC moves being coupled to RBFE calculations exist, such as in Schrodinger's FEP+ package, introduced by Ross et al.\cite{ross2020enhancing} with moves proposed both within the binding site (to ensure the correct hydration pattern) and also within bulk solvent (to enhance the sampling of the bulk density).
+A cavity-bias technique was used for the binding site moves.
+Grand canonical alchemical perturbation (GCAP)\cite{brucemacdonald2018ligand} is also available in ProtoMS\cite{woods2018protoms}, whereby 2-dimensional titrations are performed in both lambda and chemical potential space.
+However, given ProtoMS only allows MC sampling to be performed, the configurational sampling of the protein is restricted.
+
+More recently, an OpenMM-based implementation allows for GC moves to be carried out during both equilibrium and nonequilibrium RBFE calculations\cite{melling2025developing}.
+These moves are performed in the form of grand canonical nonequilibrium candidate Monte Carlo (GCNCMC) moves\cite{melling2023enhanced} and are run during the equilibration phase at each lambda window in the case of equilibrium FEP calculations, or during the nonequilibrium switch in the NES calculations.
+
+Although often water sampling is less of an issue in ABFE calculations, handling changes to water networks in buried pockets during ligand coupling/decoupling can be handled in a similar fashion to RBFEs with grand canonical methods.
+
+% in ABFE calculations is less established, although in principle, the same rules apply.
+When conventional sampling of water is insufficient such that water does not efficiently bind during ligand decoupling, performing extra ABFE calculations for the removal or addition of water molecules to the thermodynamic cycle is one way of ensuring the contribution of these solvent effects is captured.
+Performing grand canonical water moves along side a more traditional ABFE calculation of a ligand will ensure that any changes to binding pocket hydration states are sampled \textit{in situ}.
+
+Employing the techniques described above with care should allow the water network to adapt during the perturbation of the ligand to ensure more accurate and converged affinity estimations that are independent of the starting configuration and direction in which the calculation is performed.
 

### 2025-05-22 paper/manuscript.tex

--- 2025-05-12 paper/manuscript.tex
@@
 Congeneric series can contain stereoisomers or enantiomers which can bind very differently, resulting in large errors if treated incorrectly.
-For racemates, the relative abundance of each stereoisomer is normally not known.
-Therefore, the experimental activity associated with just one stereoisomer/enantiomer is more uncertain.
-However, the modeling typically uses just the bioactive conformation that best fits the active site.
-Clearly this introduces potential for larger errors compared to experiment.
-Nevertheless, if all compounds in the congeneric series are racemic, originating from similar synthetic procedures with an expected similar abundance of stereoisomers, then the differences may cancel and the trend in calculated and observed binding energies may be robust.
+For racemates, the relative abundance of each stereoisomer is normally not known, however when there is one stereocenter in the molecule it can often be fair to assume that the substance is a 50\%:50\% mixture of the two.
+If one enantiomer is active (eutomer) and the other is inactive (distomer), this corresponds to the effective concentration of your bioactive molecule being half of what is expected in the assay, which corresponds to a ~0.2 pIC50 loss in potency relative to the pure eutomer being tested, which could be captured within the assoicated error calculations.
+If possible, both enantiomers could be considered computationally, however if a series conatins multiple stereocenters, or multiple compounds within the series have stereocenters, this may be computationally restrictive.
+If compounds contain multiple undefined stereoecenters, the error and complication of understanding predictions may become prohibitive.Therefore, the experimental activity associated with just one stereoisomer/enantiomer is more uncertain.
+If enumerating unknown stereocenters is not tractable, modeling using just the conformation that best fits the active site is reasonable, however  this clearly this introduces potential for larger errors compared to experiment.
 Despite this, we can see that care and further testing is needed in this scenario, and the quality of the predictions may suffer.
 Additionally, unexpected changes in what stereoisomer binds experimentally, if they occur, could pose significant challenges for modelling efforts.
+
+\paragraph{Are there cofactors expected to be present?}
+Crystallographic structures can contain cofactors, which may be close to the ligand series of interest, and understanding if they are required or not for simulation can be an interdisciplinary effort between the modeller, crystallographer and bioligists to establish if they are a crystallographic artifact, or if they are likely to be present in the assay.
+In addition to both the crystallographic and assay conditions, what is most relevant _in vivo_ might be a third, equally difficult consideration.
+Following this, if a cofactor is considered to be included in the calculations, it is worth considering the appropriate forcefield for the cofactors.
+Cofactors, in particular metal ions can be challenging\cite{mobley2017predicting}.
+If the congeneric series of ligands is expected to interact with the cofactors differently, such as metal ions in different charge states, or to displace the cofactor for some of the molecules then this is likely beyond current modelling techniques.
+If there is still uncertainty around cofactor binding, running free energy calculations for a set of molecules both with and without the cofactor and comparing the relative rank-ordering of molecules could provide a sanity check.
 
@@
 Rigorously, this process typically comprises a MCSS search of the molecules involved to identify the common substructure.
-Specifically, with a single coordinate set approach in mind, atom types are allowed to change, so a permissive MCSS search can be done.
-This however requires either careful implementation of atom mass changes, bond length etc.
-or accepting the introduction of potential systematic error, that is in many cases negliable due to the cancelation of errors.
+Specifically, with a single coordinate set approach in mind, if atom types are allowed to change, a permissive MCSS search can be done.
+Atom-type changes however requires either careful implementation of atom mass changes, and bonded parameters or accepting the introduction of potential systematic error, that is in many cases negliable due to the cancelation of errors.
 \cite{ries2024Kartograf}
@@
 
-In general there are mutliple ways on how a MCSS can be calculated.
-One of the most prominent approaches is to estimate the MCSS with a 2D subgraph isomorphism solver(SIS) using the 2D topological stuructre information of molecules\cite{raymond2002maximum}.
+In general,  there are mutliple ways on how a MCSS can be calculated.
+One of the most prominent approaches is to estimate the MCSS with a 2D subgraph isomorphism solver (SIS) using the 2D topological stuructre information of molecules\cite{raymond2002maximum}.
 A large number of software tools can compute SIS MCSS matches using different cheminformatics packages.
@@
 
-SIS based methods, showed to struggle with 3D properties of the molecules, that need to be mapped, like for example stereochemistry.
-Therefore some packages developed a post-mapping step, where they filtered for atom distances, and therefore added 3D information into the mapping.
-Still SIS MCSS searches can be relatively time consuming as the algorithmic complexity is classified as NP-complete, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial similarity assessment and then use MCSS only to identify final mappings for relative calculations~\cite{raymond2002maximum,klabunde2012mars,jones2009elucidating, Duesbury2018Comparison}.
+SIS based methods struggle with 3D properties of the molecules that need to be mapped, for example stereochemistry.
+Therefore some packages developed a post-mapping step, where matched atoms are filtered to ensure mapped atom distances are within a cutoff, which therefore adds 3D information into the mapping.
+SIS MCSS searches can still be relatively time consuming as the algorithmic complexity is classified as NP-complete, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial similarity assessment and then use MCSS only to identify final mappings for relative calculations~\cite{raymond2002maximum,klabunde2012mars,jones2009elucidating, Duesbury2018Comparison}.
 The SIS MCSS approach, though relatively standard, takes into account only topological similarity.
@@
 
-Here an alternative way to estimate the MCSS is to calculate a bipartite graph matching based on the 3D positions of two molecules.
-These approaches tend to be very fast (can have algorithmic complexity of $O(n)=N^3$) and geometric accurate in their mapping (e.g.
+An alternative way to estimate the MCSS is to calculate a bipartite graph matching based on the 3D positions of two molecules.
+These approaches tend to be very fast (can have algorithmic complexity of $O(n)=N^3$) and are geometrically accurate in their mapping (e.g.
 stereochemistry or binding mode differing).
-The geometric approaches additionally allow mappings of differing binding modes of the same molecule, or can be even used for significant larger molecules like proteins, due to the speed increase.
-A disadvantage of these approaches is the strong dependency on the coordinate sets being well aligned.
-Another aspect is taking the topology changes of the molecules not into account, which can be done if necessary in a next step with a filtering approach, like checking for changes in ring hybridization or ring breaking (this problem is discusse below)|.
-Tools that can be used to estimate the MCSS this way are contained for example in pmx \cite{gapsys2015pmx} or Kartograf \cite{ries2024kartograf}.
-
-In any case Visual inspection prior to simulation is recommended to ensure that the mapping criteria correspond to the expected binding mode.
+The geometric approaches additionally allow mappings of differing binding modes of the same molecule, which are not captured by 2D mappings, or can be even used for significant larger molecules like proteins, due to the speed increase.
+A disadvantage of these approaches is the strong reliance on the coordinate sets being well aligned.
+Additionally, not taking the topology changes of the molecules into account, can introduce errors such as checking for changes in ring hybridization or ring breaking (this problem is discusse below).
+Tools that can be used to estimate the MCSS this way are contained for example in pmx \cite{gapsys2015pmx} or kartograf \cite{ries2024kartograf}.
+
+In any case, visual inspection prior to simulation is recommended to ensure that the mapping criteria correspond to the expected binding mode.
 If the mapping protocol returns simulations that correspond to different binding modes of a ligand within a perturbation map, this can cause large hysteresis.
@@
 For such approaches it is crucial to find a good set of restraints, that keep the ligands colocalized over the time of the simulation.
-Few descriptions of the approach can be found for such approaches like the one suggested by \textit{Baumann et al.}~\cite{baumann2023broadening} based on boresch style restraints~\cite{boresch2003absolute}, with an elaborate automatic atom selection approach, for the seperated topology approach, or Restraintmaker~\cite{ries2022restraintmaker} for the linked-dualtopology approaches, which finds a small set of atoms to be linked via distance restraints, with a small distance to each other.
+Few descriptions of the approach can be found for such approaches like the one suggested by Baumann et al.~\cite{baumann2023broadening} based on boresch style restraints~\cite{boresch2003absolute} for seperated topology approach, finding , or Restraintmaker~\cite{ries2022restraintmaker} for the linked-dualtopology approaches, which finds a small set of atoms to be linked via distance restraints, with a small distance to each other.
 However the implementation and testing of an fully automatized dual-topology pipeline for RBFE dual topology approaches is still to our knowledge a remaining research problem and was only tested with small datasets.
@@
 \paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
-Some approaches have attempted to address this using a single coordinate sets and with soft bond approaches ~\cite{clark2019relative,wang2017accurate} or dual coordinate set approach \cite{jespers2019qligfep, ries2022relative} but a general solution is not yet in mainstream use.
+Some approaches have attempted to address thisusing a single coordinate sets and with soft bond approaches ~\cite{clark2019relative,wang2017accurate} or dual coordinate set approach \cite{jespers2019qligfep, ries2022relative} but a general solution is not yet in mainstream use.
 Still, FEP+ implements one solution\cite{wang2017accurate}.

### 2025-05-23 paper/manuscript.tex
Clarify meaning of co-alchemical transformation for ions

--- 2025-05-17 paper/manuscript.tex
@@
 Alchemical transformations that involve net charge changes can provide difficulties for membrane systems; phospholipids carry functional groups with nonzero formal charge and cannot be treated as a solvent with a uniform dielectric in the same way as water.
-Several approaches are reviewed in \cite{wu2022correction-51c} and the recommended method is to use a co-alchemical ion to maintain system neutrality.
+Several approaches are reviewed in \cite{wu2022correction-51c}.
+The authors recommend the use of charge transfer by making an ion of equivalent charge appear alongside the disappearing charged molecule to maintain system neutrality.
 

### 2025-05-27 paper/manuscript.tex
Add authorship
Add revised manuscript version  On branch sboresch
 Changes to be committed:
	modified:   manuscript.tex

@@
   \textit{#1}\par}
+
+% aux. comments Stefan Boresch -> Stefan Boresch
+\newcommand{\sbnote}[1]{%
+  {\bfseries{}[SB: }%
+  {\textcolor{blue}{#1}}{\bfseries{}]}
+  }
+\newcommand{\rref}{{\bfseries[REFs]}{}}
 
@@
 Well-known protein systems with water mediated ligand interactions are for example: HSP90 which formed part of the D3R grand challenge 2015~\cite{mey2016blinded}, A2A~\cite{brucemacdonald2018ligand}, MUP~\cite{ross2015water}, ~\cite{deflorian2020accurate}, and others~\cite{michel2009energetics}.
-
-\paragraph{Membrane proteins require additional care}
-Integral membrane proteins require embedding in a suitable model membrane to remain stable during simulations.
-Lipids are usually not included in protein force fields and, although they may be covered by generic small molecule force fields, have often been parameterized separately (e.g., AMBER's Lipid21 \cite{AMBERLipid21} and the CHARMM lipid force field \cite{CHARMMLipidFF}).
-Membrane systems can be simulated in multiple thermodynamic ensembles, but the semi-isotropic scheme $NP_{z}P_{xy}T$ is frequently employed, allowing the pressure in the bilayer plane ($xy$) to scale independently \cite{LipidBestPracticesCOMS}.
-These systems require more extensive equilibration than soluble proteins.
-Some packages, such as Desmond, include a 7-step protocol to relax membrane systems.
-
-Alchemical transformations that involve net charge changes can provide difficulties for membrane systems; phospholipids carry functional groups with nonzero formal charge and cannot be treated as a solvent with a uniform dielectric in the same way as water.
-Several approaches are reviewed in \cite{wu2022correction-51c}.
-The authors recommend the use of charge transfer by making an ion of equivalent charge appear alongside the disappearing charged molecule to maintain system neutrality.
-
-In the case of a lipid-facing binding site, there are two additional challenges.
-First, simulation time should be extended to ensure sufficient sampling of lipid configurational space.
-Lipids may either bind alongside small molecules, acting as cofactors, or be displaced from the binding site by a ligand.
-If lipids participate in binding through protein-lipid or ligand-lipid interactions, it is necessary to sample these intermolecular interactions through additional $\lambda$ windows, longer simulation times, or different alchemical protocols.
-Displacing a lipid will have its own free energy cost which may take a long time to converge if it requires diffusion and rearrangement of nearby lipids.
-Second, the free energy of binding for a small molecule may be strongly coupled to the lipid species.
-In some cases, the identity of the lipid may be known.
-But in other cases, it may be difficult to discern the identity of the lipid from X-ray or cryo-EM structures due to the flexible nature of lipid head groups and benchmarking is recommended to match experimental data.
 
@@
 
-\paragraph{Topologies} 
-A critical first step is to determine which approach to use for atom mapping between end state molecules.
-Often this is predetermined by the choice of simulation software.
-Typically it is possible to chose between a \emph{dual topology} versus a \emph{single topology}
-
+\paragraph{Topologies} In regular biomolecular simulations one studies physical systems or, in other words, all entities in the system, e.g., protein, ligand and waters, represent real molecules.
+For the following, it is helpful to classify such simulation systems as having a {\em chemical} topology.
+In relative free energy calculations, however, one has to transform, e.g., ligand A into ligand B etc.
+(see {\bf XXX}).
+The required alchemical transformations can be realized in several ways, each with distinct advantages and disadvantages.
+In the first relative free energy calculations, these transformations were achieved with the help of chimeric entities, capable of representing both ligand A or ligand B.
+The important point here is that there is a single set of coordinates for this chimeric ligand, but its ``topology'' is flexible and can, for the same set of coordinates, represent either end state.
+We will consider this class of approaches as using an {\em alchemical} topology with a single set of coordinates.
+The construction of these chimeric intermediates quickly gets complicated and is difficult to automate; therefore, approaches have been developed in which both end states (e.g., ligand A and B) are present simultaneously.
+Thus, we have dual coordinates, but chemical topologies for each of the entities.
+In the following, we outline these approaches in more detail, together with additional considerations how the various approaches differ in practical use.
+While we will point out the respective advantages and disadavantages, we stress that the user's option in practice often are limited by the choice of simulation software.
+
+\paragraph{Single coordinate approaches --- alchemical topologies}
+
+When operating with a single set of coordinates for the region undergoing the alchemical transformation, a key distinction between approaches concerns the atom mapping between end state molecules.
+%Often this is predetermined by the choice of simulation software.
+As first suggested by Pearlman, these have been referred to as \emph{dual topology} versus a \emph{single topology} \cite{Pearlman_1994}.
 The distinction between single and dual topology can be illustrated by considering a hypothetical transformation from molecule A to molecule B, where both atoms share a common substructure but differ in their substituents; in particular, consider a transformation of benzene to benzyl alcohol shown in Fig.~\ref{fig:fig_topology}.
-In this case the common substructure between the two molecules is the benzene ring, though in practice substructure may be selected to be larger depending the mapping chosen, as we discuss below.
+The single and dual topology approaches as proposed by Pearlman correspond to the left and middle paths of Fig.~\ref{fig:fig_topology}.
+%\sbnote{Fig.~3 needs to be updated for dual coordinate methods and corrected, some bonds for the dummy atoms have the wrong anchor atom!}
+The common substructure between the two molecules is the benzene ring, though in practice the substructure may be selected to be larger depending on the mapping chosen, as we discuss below.
 
@@
     \includegraphics[width=0.95\columnwidth]{figures/fig3_topol/Figure.pdf}
-    \caption{\textbf{Two common topologies for alchemical calculations: single and dual topology.} \textbf{Left}: A single topology converts from one type of atom to an other.
+    \caption{\textbf{Strategies for alchemical calculations: single coordinate single and dual topology vs.\ dual coordinate dual topology.} \textbf{Left}: A single coordinate single topology converts from one type of atom to an other.
 Dummy atoms (Du) are used when there is no corresponding maximum common substructure match between the two molecules for certain atoms, using a soft-core interaction to improve overlap between the dummy atoms and the "real" atoms.
-\textbf{Right}: The dual topology does not convert one species to another, but only converts between Du atoms and an interacting species, but usually uses soft-core potentials for this.
-The 'mixed' intermediate atoms (M) are used in both dual and single topology approaches.
+\textbf{Middle}: The single coordinate dual topology does not convert one species to another, but only converts between Du atoms and an interacting species, but usually uses soft-core potentials for this.
+The 'mixed' intermediate atoms (M) are used in both single coordinate approaches (single and dual topology).
 Only the way the transformation occurs and the end states differ.
 Following the arrow along the left and right illustrate the differences.
-Figure adapted from \url{http://www.alchemistry.org/wiki/Constructing_a_Pathway_of_Intermediate_States}}
+\textbf{Right}: In the dual coordinate dual topology approach both end states are present as separate molecules simultaneously.
+Initially, benzyl alcohol is not interacting with the environment, at intermediate states the environment sees a mixture of the interactions of the two solutes, and, finally, benzene is not interacting with the environment.
+The two solutes are kept on top of each other by suitable restraints, indicated by the black ``spring'', and  the two molecules never see each other.
+Figure adapted from the one used in V1 of this manuscript; adapted and extended by Sara Tkaczyk.
+      % \url{http://www.alchemistry.org/wiki/Constructing_a_Pathway_of_Intermediate_States}
+    }
     \label{fig:fig_topology}
@@
 
-As far as we are aware, the distinction between single and dual topology approaches is made primarily for two main reasons.
+\sbnote{This probably needs to be rewritten\ldots} As far as we are aware, the distinction between (single coordinate) single and dual topology approaches has been made primarily for two main reasons.
 First, this choice affects the alchemical pathway followed, and thus may affect convergence properties---though we are not yet aware of a study of the relative efficiency and merits of these two approaches.
 Additionally, historically, some simulation packages implemented only one approach and not the other, meaning that the distinction was functional.
-
-Some additional terms have also been employed to talk about these different intermediate pathways .
+For an in-depth analyis of differences resulting from the use of a  single vs.\ a dual topology approach (single coordinate in both cases) see Ref.~\cite{Boresch_2002}.
+
+\paragraph{Dual coordinate approaches --- dual chemical topologies}
+
+The chimeric constructs needed in single and dual topology approaches in the sense of Pearlman’s definition just described quickly become non-trivial to set up and require care to avoid artifacts \sbnote{See below; where do we place the dummy atom discussion?}.
+Some of these difficulties might be avoided by having the initial and final state present as separate molecules with separate coordinate sets.
+To the best of our knowledge, such an approach was first described by Axelsen and Li \cite{Axelsen_1998}.
+Both initial and final state are present simultaneously; the two copies never see each other and interactions with the environment are turned on/off as a function of $\lambda$.
+Harmonic restraints between corresponding atoms in the initial and final state copy prevent the two molecules (sets of atoms from drifting apart).
+%Unfortunately, Axelsen and Li referred to their ansatz also as dual-topology, though in practice it is quite different from Pearlman’s meaning of the term.
+A recent implementation of this idea is Restraintmaker \cite{Ries_2022}, a tool to automatically assign the distance restraints needed to keep the two molecules / regions (one for each of the physical endstates) on top of each other.
+Somewhat unfortunately, Axelsen and Li, as well as Ries et al refer to their approach as dual topology.
+Ries et al suggested to use the term hybrid topology for (single coordinate) dual topology in the sense of Pearlman.
+We will attempt a disambiguation of the various nomenclatures below.
+
+Another variant of a dual coordinate approach is implemented in NAMD \cite{jiang2019computing}.
+Whereas Ries et al.\  use distance restraints to keep two molecules on top of each other, corresponding atoms in NAMD are held on top of each other by constraints; in practice, this means that these coordinates are forced to be identical.
+AMBER's free energy implementation recently switched to a similar approach \cite{Lee_2023}.
+%The authors refer to this approach as hybrid single-dual topology.
+While it technically uses two separate sets of coordinates, the constraint that corresponding atoms have identical coordinates makes it effectively a single coordinate method; hence, I consider NAMD’s approach to be a variant of Pearlman’s (single coordinate) dual topology.
+
+Dual coordinate approaches with chemical topologies encompass a third group of methods, where essentially two absolute free energy calculations are carried out in opposite directions at the same time.
+The so-called ``separated topologies'' (SepTop) method \cite{rocklin2013separated,Baumann_2023} and the ``alchemical transfer method'' (ATM) \cite{Azimi_2022} belong into this category.
+In SepTop, the two copies are held on top of each other via restraints with respect to the protein/receptor they bind to (as opposed to Restraintmaker, where the molecules are restrained to each other.
+In ATM, a restraint is used to keep the bound ligand 1 and the free ligand 2 always some minimum distance apart; during exchange moves, suitable rigid-body rotations/translations are applied to move the free ligand into a position/orientation as similar as possible to the one of the bound ligand.
+Both copies are present simultaneously, but they do not occupy the same space.
+Recently, a variant of $\lambda$-dynamics that uses chemical topologies (separate coordinates for each state) has been proposed \cite{Liesen_2024}.
+
+%Other methodologies also use both end states simultaneously; examples are the Separated Topologies (SepTop) approach [Baumann2023] and the ATM method [Azimi2022].
+In the former, the two copies are held on top of each other via restraints with respect to the protein/receptor they bind to.
+Ries et al.
+refer to such an approach as “separated dual topology”, whereas they label their approach as “linked dual topology”.
+In ATM, a restraint is used to keep the bound ligand 1 and the free ligand 2 always some minimum distance apart; during exchange moves, suitable rigid-body rotations/translations are applied to move the free ligand into a position/orientation as similar as possible to the one of the bound ligand.
+Both copies are present simultaneously, but they do not occupy the same space.
+
+%One final approach, so far in its infancy, has been called ``separated topologies'' and essentially consist of two absolute free energy calculations in opposite directions at the same time, turning one molecule's interactions with the environment off, while turning the other molecule's interaction on~\cite{jiang2019computing, rocklin2013separated}.
+
+\paragraph{Summary, comparison of single and dual coordinate methods, orthogonal classifications:}
+
+%Some additional terms have also been employed to talk about these different intermediate pathways .
 Particularly, some studies refer to a ``hybrid'' topology approach to free energy calculations~\cite{gapsys2015pmx, gapsys2016accurate, gapsys2020large}, though this term may not yet have achieved widespread use.
@@
 
-One final approach, so far in its infancy, has been called ``separated topologies'' and essentially consist of two absolute free energy calculations in opposite directions at the same time, turning one molecule's interactions with the environment off, while turning the other molecule's interaction on~\cite{jiang2019computing, rocklin2013separated}.
-
-Software packages vary in their use of single or dual topology approaches; for example, AMBER TI uses a dual topology approach, while BioSimSpace uses a single topology approach.
+Unfortunately, during the past years the use of the terms single and dual topology has proliferated, often with quite different meaning from the original intentions.
+The above presentation makes clear that the primary difference between approaches is the use of separate coordinates for each end state (chemical topology) or the use of an alchemical chimera with a single set of coordinates (alchemical topology).
+We therefore suggest the following classification
+\begin{itemize}
+\item Single coordinate (SC) --- Alchemical topology
+  \begin{itemize}
+  \item Single topology (SCST)
+    \item Dual topology (SCDT)
+  \end{itemize}
+  \item Dual coordinate --- Chemical topology (DC)
+  \end{itemize}
+For completeness, we list some additional terminology that has been used: Some studies refer to a ``hybrid'' topology approach to free energy calculations~\cite{gapsys2015pmx, gapsys2016accurate, gapsys2020large}, though in our view these calculations should be classified as single coordinate single topology \sbnote{Check whether this is correct!}.
+%this term may not yet have achieved widespread use.
+In this case, "hybrid" seems to indicate that the set up of these free energy calculations involves a hybrid of the two molecules, and much of what is done in these studies uses a single topology approach~\cite{gapsys2020large}.
+Ries et al.\ suggested terminology \cite{Ries_2022} in which single topology corresponds to SCST, hybrid topology to SCDT, and dual topology to DC.
+
+The practical implementation of the approaches to carry out alchemial transformations above leads to additional, often non-trivial differences within each methodology.
+In the description of DC approaches we already pointed out that the two sets of coordinates can be held on top of each other by constraints \cite{jiang2019computing,Lee_2023} or restraints.
+In the latter case, the restraints may couple the two set of coordinates directly \cite{Axelsen_1998,Ries_2022}, or the restraints may be defined relative to a common reference frame \cite{rocklin2013separated,Baumann_2023}.
+Finally, in ATM the two copies of the coordinates are held separate \cite{Azimi_2022}.
+
+Concerning SC approaches, Pearlman’s usage of the term single topology not only meant a chimeric construct as depicted in Fig.~\ref{fig:fig_topology}, but specifically implied that the coupling parameter $\lambda$ was applied to the force field parameters.
+E.g., a bond stretching energy term changing from C-H to C-C has the functional form $U_{bond}(\lambda)=K(\lambda)(r-r_0(\lambda))^2$.
+For $\lambda=0$, $K(\lambda)$ and $r_0(\lambda)$ correspond to the values appropriate for a C-H bond, and for $\lambda=1$ to those for a C-C bond.
+The other energy terms are treated analogously.
+The single topology paradigm as just summarized was the default mode of alchemical transformations in AMBER and GROMOS at that time.
+More recently, this implementation of single topology has been referred to as parameter interpolation or mixing \cite{Giese_2018}.
+
+The $\lambda$-dependent interpolation can also be accomplished by mixing the energies and forces of the initial ($\lambda=0$) and final state ($\lambda=1$), e.g., linearly according to $U(\lambda)=(1-\lambda) U_0+\lambda\,U_1$.
+Other λ-dependencies can be used, and soft-cores \rref may be used for Lennard-Jones and electrostatic interactions to avoid endpoint problems.
+Giese et al.
+\cite{Giese_2018} refer to this approach as energy mixing (interpolation).%; it requires two separate calls to the energy/force routines, but since only a fraction of interactions is affected by the alchemical changes, various optimizations are possible to keep the computational overhead manageable.
+An early example of energy mixing for SCST is the PERT free energy module of CHARMM (written by Bernie Brooks around 1990 and briefly described in Ref.~\cite{brooks_2009}.
+
+Any SCDT approach requires two energy/force calculations and, thus, uses energy mixing.
+Therefore, approaches that implement energy mixing together with an alchemical topology (a single set of coordinates) often support both SCST and SCDT.
+E.g., CHARMM's PERT module just mentioned could be used to set up either of the SC approaches sketched in Fig.~\ref{fig:fig_topology}.
+Finally, all DC approaches require two energy/force evaluations.
+
+Any energy mixing approach, whether in a SC or DC context, requires two energy/force evaluations.
+While various optimizations keep the computational overhead significantly below a factor of two,  parameter mixing, if applicable, is the compuationally most efficient approach.
+However, it is closely tied to SCST and the more complicated the alchemical transformation, the more challenging a strict SCST approach becomes.
+Also, the mixing of partial charges and the calculation of $\partial U/\partial\lambda$ for the reciprocal space particle-mesh Ewald (PME) summation is tricky.
+Giese et al.\ described how to do this \cite{Giese_2018}, but most energy mixing implementations calculate the PME term twice and mix the resulting energy and forces.
+
+\sbnote{I think this should get its own paragraph/subsection; to be expanded!} Software packages vary in their use of single or dual topology approaches; for example, AMBER TI uses a dual topology approach, while BioSimSpace uses a single topology approach.
 Please make sure to check what approach is used with your software package of choice, or whether it supports your choice of approach (GROMACS and GROMOS, for example, support both).
@@
 Using an absolute free energy approach instead seems more promising in such a case.
+
+\paragraph{Dummy atoms}
+
+In practically all relative free energy simulations, regardless of the approach, SCST, SCDT, or DCDT, dummy atoms (non-interacting placeholder atoms) are required.
+We stress this point, as sometimes dummy atoms are viewed as something specific to SCST.
+In SC approaches, dummy atoms may even be required if the number of atoms between initial and final state of the system is identical, consider, e.g., tautomeric forms of a molecule.
+
+The need for and role of dummy atoms is clearly visible in Fig.~\ref{fig:fig_topology}: One or more dummy atoms are attached to the physical molecule.
+A crucial requirement is that these atoms, typically connected through bonded energy terms (bond stretching, angle bending, dihedral angles) do not affect equilibrium and dynamical properties of the physical system.
+However, since a physical molecule + dummy atoms attached to it can or must be viewed as a modified physical molecule, the partition function of the molecules with and without dummy atoms will NOT be identical.
+In other words, a single free energy difference between two molecules A and B will be different from that of A(+D) and B(+D), where the (+D) indicates the presence of one or more dummy atoms needed so that the alchemical transformation can be carried out.
+When a dual coordinate approach is used, the noninteracting part of the respective other endstate is tied to the physical one, and will contribute to the partition function.
+By construction, relative free energy simulatins involve two analogous alchemical transformations, e.g., A(+D) -> B(+D) in gas phase and aqueous solution (RSFEs), or in the bound and unbound state (RBFEs).
+If dummy atoms are interacting with the physical system in such a way that $\mathsf{Z = Z(R) \times Z(R\text{-}D) \times Z(D)}$, where Z is the total partition function, Z(R) the partition function of the physical system, Z(D) the partition function contribution resulting from interactions between only the dummy atoms, and Z(R-D) the contribution resulting from the bonded connections between the physical atoms and dummy atoms, then their contributions (Z(D) and Z(R-D)) factor from the total partition function.
+As dummy atoms do not interact with the environment, their contribution, therefore, is identical in the two alchemical transformations required in the two legs of the relative transformation and will cancel exactly from the double free energy difference of interest.
+
+While some early work on the correct treatment of dummy atoms, i.e., how to treat them so that their contribution to the partition function factors, has been available for years \cite{Boresch_2002,Shobana_2000,Wang_2012}, most workers seem to mostly turn off nonbonded interactions for dummy atoms, but leaving all bonded interactions intact \sbnote{To all: While I am quite sure that this is what most of us have been doing most of the time, I wonder whether this has been epxlicitly described anywhere?}.
+Since current force fields use multiple angle and torsion terms for each atom, there are often a large number of bonded energy terms connecting dummy atoms to the physical molecule.
+However, as already described in the early studies and repeated in a more recent study focusing on the correct treatment of dummy atoms \cite{Fleck_2021}, the desired factorization can only be guaranteed if any dummy atom connected to the physical molecule is held by exactly three (non-redundant) degrees of freedom.
+Naive application of the three-degree-of-freedom per dummy atom rule (for dummy atoms having at least one joint bonded term with the physical molecule), however, can result in simulations where the dummy part adopts unphysical configurations/conformations; Fleck et al.
+referred to this as “flapping”.
+Thus, depending on the details of the physical molecule -- dummy atom(s) junction, the three degree of freedoms need to be selected with care, or a carefully chosen additional redundant degree of freedom may be needed.
+In the latter case, the factorization of the partition function is not exact, but the overall error can still be made arbitrarily small.
+For the full details, we refer the reader to Ref.~\cite{Fleck_2021}.
+
+One situation involving dummy atoms that needs to be avoided is the creation of a cyclic structure involving dummy atoms, i.e., whenever the (group of) dummy atom(s) is attached to two different parts of the physical molecule.
+In this case, the factorization of the partition function is never possible \cite{Shobana_2000}.
+It follows that ring opening or closing is extremely challenging to carry out using SCST approaches \cite{liu2015ring}; these difficulties have in part motivated the development and adoption of dual coordinate approaches.
+For more on ring opening/closing, see also below.
+However, DC approaches are also extremely prone to exactly this type of problem: to keep the noninteracting molecule on top of the interacting one, typically multiple bonded restraints (e.g.
+harmonic bonds) are used.
+These effectively, are equivalent to the formation of cyclic structures with potential erroneous contributions to the double free energy difference of interest.
+
+The influence from an incorrect treatment of dummy atoms (e.g., leaving all bonded terms intact) is difficult to estimate, though it may be small in many cases \cite{Fleck_2021}.
+Nevertheless, some guidelines should be followed:
+\begin{itemize}
+  \item Obviously, a system with and without dummy atoms should have the same energetic minimum.
+Note that in larger groups of dummy atoms, redundant bonded terms involving only dummy atoms will be essential to avoid flapping, and the energy of the optimized geometry may be different from that of the physical system without dummy atoms.
+Thus, a \underline{necessary} condition is as follows: Optimize the geometry of the physical system (no dummy atoms); the final energy is the reference value.
+Optimize the geometry of the same molecule in the presence of dummy atoms; the final energy may well be different from the reference energy.
+Now delete the dummy atoms and recompute the energy \underline{without} further minimization.
+The energy after the deletion of the dummy atoms must be identical to the reference energy.
+\item The above is only a \underline{necessary} criterion, and the dummy atoms may still influence the dynamical behavior of the physical molecule.
+Furthermore, overzealous removal of redundant energy terms may result in flapping of the dummy atoms.
+If one suspects the latter, then a short simulation in the gas phase should reveal flapping.
+Unfortunately, while straightforward in principle, both checks described are hard to automate, which is essential in large RFEs campaigns.
+\item The correct treatment of dummy atoms can be validated by closing a suitable thermodynamic cycle.
+E.g., Fleck et al.\ calculated ASFEs for each of their endstates, as well as the RSFE between them \cite{Fleck_2021}.
+In all cases when dummy atoms were treated according to best practices, the cycle closure error was well within the statistical uncertainty, which was well below 0.1 kcal/mol.
+A related, earlier study by Loeffler et al.
+also reported ASFEs and RSFEs.
+If one attempts to close the cycle based on the numbers reported there, cycle closure error was nonnegligible in some cases \cite{loeffler2018reproducibility}.
+\item While solvation free energies can nowadays be computed quite fast, their systematic calculation would nevertheless be too costly.
+The role of solvent can be replaced by some external restraint, distorting the average geometry of the physical molecule.
+If one computes the free energy cost of applying the restraint with and without dummy atoms, the two numbers must agree within statistical significance.
+While much faster than full solvation free energy calculations, applying a restraint in the gas phase requires careful thermalization of the simulation system, and automation is difficult.
+  \end{itemize}
+%Further work automating the correct setup and treatment of dummy atoms is necessary.
 

### 2025-05-30 paper/manuscript.tex
Adding reference to ABFE part

--- 2025-05-22 paper/manuscript.tex
@@
 Congeneric series can contain stereoisomers or enantiomers which can bind very differently, resulting in large errors if treated incorrectly.
-For racemates, the relative abundance of each stereoisomer is normally not known, however when there is one stereocenter in the molecule it can often be fair to assume that the substance is a 50\%:50\% mixture of the two.
-If one enantiomer is active (eutomer) and the other is inactive (distomer), this corresponds to the effective concentration of your bioactive molecule being half of what is expected in the assay, which corresponds to a ~0.2 pIC50 loss in potency relative to the pure eutomer being tested, which could be captured within the assoicated error calculations.
-If possible, both enantiomers could be considered computationally, however if a series conatins multiple stereocenters, or multiple compounds within the series have stereocenters, this may be computationally restrictive.
-If compounds contain multiple undefined stereoecenters, the error and complication of understanding predictions may become prohibitive.Therefore, the experimental activity associated with just one stereoisomer/enantiomer is more uncertain.
-If enumerating unknown stereocenters is not tractable, modeling using just the conformation that best fits the active site is reasonable, however  this clearly this introduces potential for larger errors compared to experiment.
+For racemates, the relative abundance of each stereoisomer is normally not known.
+Therefore, the experimental activity associated with just one stereoisomer/enantiomer is more uncertain.
+However, the modeling typically uses just the bioactive conformation that best fits the active site.
+Clearly this introduces potential for larger errors compared to experiment.
+Nevertheless, if all compounds in the congeneric series are racemic, originating from similar synthetic procedures with an expected similar abundance of stereoisomers, then the differences may cancel and the trend in calculated and observed binding energies may be robust.
 Despite this, we can see that care and further testing is needed in this scenario, and the quality of the predictions may suffer.
 Additionally, unexpected changes in what stereoisomer binds experimentally, if they occur, could pose significant challenges for modelling efforts.
-
-\paragraph{Are there cofactors expected to be present?}
-Crystallographic structures can contain cofactors, which may be close to the ligand series of interest, and understanding if they are required or not for simulation can be an interdisciplinary effort between the modeller, crystallographer and bioligists to establish if they are a crystallographic artifact, or if they are likely to be present in the assay.
-In addition to both the crystallographic and assay conditions, what is most relevant _in vivo_ might be a third, equally difficult consideration.
-Following this, if a cofactor is considered to be included in the calculations, it is worth considering the appropriate forcefield for the cofactors.
-Cofactors, in particular metal ions can be challenging\cite{mobley2017predicting}.
-If the congeneric series of ligands is expected to interact with the cofactors differently, such as metal ions in different charge states, or to displace the cofactor for some of the molecules then this is likely beyond current modelling techniques.
-If there is still uncertainty around cofactor binding, running free energy calculations for a set of molecules both with and without the cofactor and comparing the relative rank-ordering of molecules could provide a sanity check.
 
@@
 One would co-localize a pair of compounds in a binding site, exclude their interactions with one another, and compute the relative binding free energy by turning one molecule on from being dummy atoms while turning the other off.
-For such approaches it is crucial to find a good set of restraints, that keep the ligands colocalized over the time of the simulation.
-Few descriptions of the approach can be found for such approaches like the one suggested by Baumann et al.~\cite{baumann2023broadening} based on boresch style restraints~\cite{boresch2003absolute} for seperated topology approach, finding , or Restraintmaker~\cite{ries2022restraintmaker} for the linked-dualtopology approaches, which finds a small set of atoms to be linked via distance restraints, with a small distance to each other.
+For such approaches it is crucial to find a good set of restraints, that keep the ligands colocalized over the time of the simulation .
+This problem is similar to the required restraints in ABFE Calculations (see \ref{ABFErest}).
+Few descriptions of possible approaches in RBFE calculations can be found for such approaches like the one suggested by \textit{Baumann et al.}~\cite{baumann2023broadening} based on boresch style restraints~\cite{boresch2003absolute}, with an elaborate automatic atom selection approach, for the seperated topology approach, or Restraintmaker~\cite{ries2022restraintmaker} for the linked-dualtopology approaches, which finds a small set of atoms to be linked via distance restraints, with a small distance to each other.
 However the implementation and testing of an fully automatized dual-topology pipeline for RBFE dual topology approaches is still to our knowledge a remaining research problem and was only tested with small datasets.
@@
 
-\paragraph{Several choices of restraints are possible.}
+\paragraph{Several choices of restraints are possible.}\label{ABFErest}
 In practice, a variety of types of restraints are common, from simple harmonic distance restraints between the ligand and the protein~\cite{mobley2006use}, to flat-bottom restraints which work similarly but only exert a force if the ligand leaves a specific region~\cite{chen2007can}.

### 2025-06-01 paper/manuscript.tex

--- 2025-05-27 paper/manuscript.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{\tocorange{What simulation protocol should I choose?}}
+%\section{\tocorange{What simulation protocol should I choose?}}
+\section{\tocorange{Choice of simulation protocol and implementation of alchemical transformations}}
+  \sbnote{Emilio suggested to make the title more general; this is my suggestion!}
 \tocsectioncomment{Steering group: Possibly expand to include other methods group agrees on.}
@@
 In relative free energy calculations, however, one has to transform, e.g., ligand A into ligand B etc.
-(see {\bf XXX}).
+(see \sbnote{appropriate cross-reference!}).
 The required alchemical transformations can be realized in several ways, each with distinct advantages and disadvantages.
@@
 In single topology calculations, the overall transformation is set up to involve as few additional atoms as possible, so benzene would be typically changed into benzyl alcohol by first changing one of the hydrogens into a carbon.
-The site of this transformation will also be the future home of two additional hydrogen atoms bound to the new carbon, so these must initially be present as non-interacting atoms called ``dummy atoms'', which retain their bonded interactions but do not interact with the rest of the system.
+The site of this transformation will also be the future home of two additional hydrogen atoms bound to the new carbon, so these must initially be present as non-interacting atoms called ``dummy atoms'' (labelled Du in Fig.~\ref{fig:fig_topology}), which retain their bonded interactions but do not interact with the rest of the system.
 Bond parameters as well as partial charges between the changing atoms are adjusted accordingly between the initial and final state; see below for further details.
-Thus, in a single topology calculation, atoms may change their type, ensuring minimal dummy atoms are created.
+In a single topology calculation, atoms may change their type, ensuring minimal dummy atoms are created.
 This is illustrated in the left arm of Fig.~\ref{fig:fig_topology}.
-
-In contrast, in a dual topology alchemical free energy calculation, \emph{no atoms are allowed to change type}~\cite{boresch1999role, shirts2012best}.
-This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxymethyl group, then passing through an intermediate state where some atoms are partially interacting---particularly, those atoms which are becoming dummy atoms or ceasing to be dummy atoms~\cite{mobley2014blind}.
-The transformation finally culminates in a state where benzyl alcohol is present along with the additional dummy atom which was previously a corresponding hydrogen of the benzene.
-Fig.~\ref{fig:fig_topology}'s right branch depicts how such a dual topology works.
-Note that the two groups that are switched on or off never see each other.
+At intermediate states, atoms behave as 'mixed' (M), either interpolating between two physical atoms, e.g., in the example, the hydrogen becomes a carbon, or between dummy and physical atoms.
 
@@
     \includegraphics[width=0.95\columnwidth]{figures/fig3_topol/Figure.pdf}
-    \caption{\textbf{Strategies for alchemical calculations: single coordinate single and dual topology vs.\ dual coordinate dual topology.} \textbf{Left}: A single coordinate single topology converts from one type of atom to an other.
+    \caption{\textbf{Strategies for alchemical calculations: single coordinate single and dual topology vs.\ dual coordinate dual topology.} \textbf{Left}: A single coordinate single topology converts from one type of atom to another.
 Dummy atoms (Du) are used when there is no corresponding maximum common substructure match between the two molecules for certain atoms, using a soft-core interaction to improve overlap between the dummy atoms and the "real" atoms.
-\textbf{Middle}: The single coordinate dual topology does not convert one species to another, but only converts between Du atoms and an interacting species, but usually uses soft-core potentials for this.
-The 'mixed' intermediate atoms (M) are used in both single coordinate approaches (single and dual topology).
-Only the way the transformation occurs and the end states differ.
-Following the arrow along the left and right illustrate the differences.
+At intermediate states, 'mixed' atoms (M) are present, either interpolating between physical atoms, e.g., a hydrogen to a carbon, or between physical and dummy atoms.
+\textbf{Middle}: The single coordinate dual topology does not convert one species to another, but only converts between Du atoms and an interacting species, typically using soft-core potentials for this.
+Here, 'mixed' intermediate atoms (M) always interpolate between physical and dummy atoms.
+Thus, 'mixed' atoms are used in both single coordinate approaches (single and dual topology), but the way the transformation occurs and the end states differ.
+      %Following the arrow along the left and right illustrate the differences.
 \textbf{Right}: In the dual coordinate dual topology approach both end states are present as separate molecules simultaneously.
@@
 
-\sbnote{This probably needs to be rewritten\ldots} As far as we are aware, the distinction between (single coordinate) single and dual topology approaches has been made primarily for two main reasons.
+In contrast, in a dual topology alchemical free energy calculation, \emph{no atoms are allowed to change type}~\cite{boresch1999role, shirts2012best}.
+This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxymethyl group, then passing through an intermediate state where some atoms are partially interacting---particularly, those atoms which are becoming dummy atoms or ceasing to be dummy atoms~\cite{mobley2014blind}.
+The transformation finally culminates in a state where benzyl alcohol is present along with the additional dummy atom which was previously a corresponding hydrogen of the benzene.
+Fig.~\ref{fig:fig_topology}'s right branch depicts how such a dual topology works.
+Note that the two groups that are switched on or off never see each other.
+The 'mixed' intermediate atoms (M) are used as well, but they strictly interpolate between physical and dummy atoms, or vice versa.
+
+%\sbnote{This probably needs to be rewritten\ldots}
+As far as we are aware, the distinction between (single coordinate) single and dual topology approaches has been made primarily for two main reasons.
 First, this choice affects the alchemical pathway followed, and thus may affect convergence properties---though we are not yet aware of a study of the relative efficiency and merits of these two approaches.
@@
 
-The chimeric constructs needed in single and dual topology approaches in the sense of Pearlman’s definition just described quickly become non-trivial to set up and require care to avoid artifacts \sbnote{See below; where do we place the dummy atom discussion?}.
+The chimeric constructs needed in single and dual topology approaches in the sense of Pearlman’s definition just described quickly become non-trivial to set up and require care to avoid artifacts (see below).
+%\sbnote{See below; where do we place the dummy atom discussion?}.
 Some of these difficulties might be avoided by having the initial and final state present as separate molecules with separate coordinate sets.
@@
 A recent implementation of this idea is Restraintmaker \cite{Ries_2022}, a tool to automatically assign the distance restraints needed to keep the two molecules / regions (one for each of the physical endstates) on top of each other.
+The idea is illustrated in the right branch of Fig.~\ref{fig:fig_topology}.
 Somewhat unfortunately, Axelsen and Li, as well as Ries et al refer to their approach as dual topology.
@@
 In SepTop, the two copies are held on top of each other via restraints with respect to the protein/receptor they bind to (as opposed to Restraintmaker, where the molecules are restrained to each other.
-In ATM, a restraint is used to keep the bound ligand 1 and the free ligand 2 always some minimum distance apart; during exchange moves, suitable rigid-body rotations/translations are applied to move the free ligand into a position/orientation as similar as possible to the one of the bound ligand.
+%In ATM, a restraint is used to keep the bound ligand 1 and the free ligand 2 always some minimum distance apart; during exchange moves, suitable rigid-body rotations/translations are applied to move the free ligand into a position/orientation as similar as possible to the one of the bound ligand.
+In ATM, restraints are used to keep the bound and free ligands at some distance apart and in the same orientation, so that during the translation move the free ligand is placed into a position and orientation as similar as possible to that of the bound ligand.
 Both copies are present simultaneously, but they do not occupy the same space.
@@
 For more on ring opening/closing, see also below.
-However, DC approaches are also extremely prone to exactly this type of problem: to keep the noninteracting molecule on top of the interacting one, typically multiple bonded restraints (e.g.
-harmonic bonds) are used.
+Nevertheless, DC approaches are susceptible to exactly this type of problem: to keep the noninteracting molecule on top of the interacting one, typically multiple bonded restraints (e.g.
+harmonic bonds) are used \cite{Axelsen_1998,Ries_2022}.
 These effectively, are equivalent to the formation of cyclic structures with potential erroneous contributions to the double free energy difference of interest.
+The type of restraints used in SepTop \cite{rocklin2013separated,Baumann_2023} and ATM \cite{Azimi_2022} aoid this problem.
 

### 2025-06-02 paper/manuscript.tex
binding site definition, restraints

--- 2025-05-30 paper/manuscript.tex
@@
 
-\subsection{Chemical potential}
+\subsection{The chemical potential}
 
@@
 
-\subsection{Solvation}
+\subsection{Solvation}\label{sec:theory-solvation}
 
@@
 
-\subsection{Solvent partitioning}
+\subsection{Solvent partitioning}\label{sec:theory-partitioning}
 
@@
 \subsection{Non-covalent molecular binding}
+\label{sec:non-covalent-molecular-binding}
 
@@
 \paragraph{Atom mapping}
-Following the choice of system representation and sampling strategy, it might be required to generate a atom mapping.
-This is the case especially for system representations using a single coordinate set (Type 1.
-& 2), however a dual coordinate set approach could use an atom mapping as well, depending on the sampling method, this will be explained in the next steps.
-Once a single coordinate set approach is selected is selected, a crucial next step is to identify the common atoms which will not be perturbed between the two molecules (respectivly end-states).
-Rigorously, this process typically comprises a MCSS search of the molecules involved to identify the common substructure.
-Specifically, with a single coordinate set approach in mind, atom types are allowed to change, so a permissive MCSS search can be done.
-This however requires either careful implementation of atom mass changes, bond length etc.
-or accepting the introduction of potential systematic error, that is in many cases negliable due to the cancelation of errors.
-\cite{ries2024Kartograf}
-
+Once a particular approach to the topology is selected, a crucial next step is to identify the common atoms which will not be perturbed.
+Rigorously, this process typically comprises a MCSS search of the molecules involved to identify the common substructure---though the parameters of the MCSS search will differ depending on whether single or dual topology calculations are planned.
+Specifically, with a single topology approach in mind, atom types are allowed to change, so a permissive MCSS search can be done, whereas with dual topology a more strict search is required.
 \begin{figure}
@@
 
-In general there are mutliple ways on how a MCSS can be calculated.
-One of the most prominent approaches is to estimate the MCSS with a 2D subgraph isomorphism solver(SIS) using the 2D topological stuructre information of molecules\cite{raymond2002maximum}.
-A large number of software tools can compute SIS MCSS matches using different cheminformatics packages.
-Some rely on RDKit~\cite{rdkit2019Dec}, such as pmx~\cite{gapsys2015pmx}, LOMAP~\cite{liu2013lead}, FESetup~\cite{loeffler2015fesetup}, ProtoCaller~\cite{suruzhon2020protocaller}, SMArt~\cite{petrov2021perturbation}, and partially BioSimSpace~\cite{hedges2019biosimspace}, while others such as fkckombu~\cite{kawabata20143d} are standalone tools.
+There are different tools that allow the generation of MCSS matches as well as single topology input.
+A large number of software tools can compute MCSS matches using different cheminformatics packages.
+Some rely on RDKit~\cite{rdkit2019Dec}, such as pmx~\cite{gapsys2015pmx}, LOMAP~\cite{liu2013lead}, FESetup~\cite{loeffler2015fesetup} and partially BioSimSpace~\cite{hedges2019biosimspace}, while others such as fkckombu~\cite{kawabata20143d} are standalone tools.
 Schr\"{o}dinger's FEP+ planning tool was originally based on a version of LOMAP, and it also uses MCSS matching as well as 3D considerations to plan the network of single topology calculations between molecules~\cite{wang2015accurate}.
 
-SIS based methods, showed to struggle with 3D properties of the molecules, that need to be mapped, like for example stereochemistry.
-Therefore some packages developed a post-mapping step, where they filtered for atom distances, and therefore added 3D information into the mapping.
-Still SIS MCSS searches can be relatively time consuming as the algorithmic complexity is classified as NP-complete, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial similarity assessment and then use MCSS only to identify final mappings for relative calculations~\cite{raymond2002maximum,klabunde2012mars,jones2009elucidating, Duesbury2018Comparison}.
-The SIS MCSS approach, though relatively standard, takes into account only topological similarity.
+MCSS searches can be relatively time consuming, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial similarity assessment and then use MCSS only to identify final mappings for relative calculations~\cite{raymond2002maximum,klabunde2012mars,jones2009elucidating}.
+The MCSS approach, though relatively standard, takes into account only topological similarity.
 It is possible that changes in binding mode could actually require a different choice of mapping, so in some cases mappings may need to be planned differently depending on 3D positioning of atoms in space.
-
-Here an alternative way to estimate the MCSS is to calculate a bipartite graph matching based on the 3D positions of two molecules.
-These approaches tend to be very fast (can have algorithmic complexity of $O(n)=N^3$) and geometric accurate in their mapping (e.g.
-stereochemistry or binding mode differing).
-The geometric approaches additionally allow mappings of differing binding modes of the same molecule, or can be even used for significant larger molecules like proteins, due to the speed increase.
-A disadvantage of these approaches is the strong dependency on the coordinate sets being well aligned.
-Another aspect is taking the topology changes of the molecules not into account, which can be done if necessary in a next step with a filtering approach, like checking for changes in ring hybridization or ring breaking (this problem is discusse below)|.
-Tools that can be used to estimate the MCSS this way are contained for example in pmx \cite{gapsys2015pmx} or Kartograf \cite{ries2024kartograf}.
-
-In any case Visual inspection prior to simulation is recommended to ensure that the mapping criteria correspond to the expected binding mode.
+Visual inspection prior to simulation is recommended to ensure that the mapping criteria correspond to the expected binding mode.
 If the mapping protocol returns simulations that correspond to different binding modes of a ligand within a perturbation map, this can cause large hysteresis.
 
-A next step for atom mappings, would be to support multistate methods (e.g.
-RE-EDS~\cite{sidler2016replica}, A-EDS~\cite{perthold2020toward} or Ladybugs~\cite{robo2022achieving}) with efficient mapping algorithms.
-There are protoype starting point for solving this problem, but to our knowledge, there is only few published work on this aspect.
-\cite{petrov2021Perturbation}
-
-Single coordinate set relative calculations and calculations based on substructure searches only work if in fact the ligands share a common substructure, e.g.
+Single topology relative calculations and calculations based on substructure searches only work if in fact the ligands share a common substructure, e.g.
 are part of a congeneric series, see Fig.~\ref{fig:fig_mcss}.
-If no common substructure is shared, then dual coordinate sets methods like linked-dual or separated topology free energy calculations are needed.
+If no common substructure is shared, then alternative dual or separated topology free energy calculations are needed.
 One would co-localize a pair of compounds in a binding site, exclude their interactions with one another, and compute the relative binding free energy by turning one molecule on from being dummy atoms while turning the other off.
-For such approaches it is crucial to find a good set of restraints, that keep the ligands colocalized over the time of the simulation .
-This problem is similar to the required restraints in ABFE Calculations (see \ref{ABFErest}).
-Few descriptions of possible approaches in RBFE calculations can be found for such approaches like the one suggested by \textit{Baumann et al.}~\cite{baumann2023broadening} based on boresch style restraints~\cite{boresch2003absolute}, with an elaborate automatic atom selection approach, for the seperated topology approach, or Restraintmaker~\cite{ries2022restraintmaker} for the linked-dualtopology approaches, which finds a small set of atoms to be linked via distance restraints, with a small distance to each other.
-However the implementation and testing of an fully automatized dual-topology pipeline for RBFE dual topology approaches is still to our knowledge a remaining research problem and was only tested with small datasets.
-\cite{jespers2019qligfep, Rieder2022Lev}
+To our knowledge no general pipeline for such calculations yet exists and this would likely remain a research problem.
+Using an absolute free energy approach instead seems more promising in such a case.
 
 \paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
-Some approaches have attempted to address this using a single coordinate sets and with soft bond approaches ~\cite{clark2019relative,wang2017accurate} or dual coordinate set approach \cite{jespers2019qligfep, ries2022relative} but a general solution is not yet in mainstream use.
-Still, FEP+ implements one solution\cite{wang2017accurate}.
+Some approaches have attempted to address this~\cite{clark2019relative} but a general solution is not yet in mainstream use. Still, FEP+ implements one solution.
 
@@
 
-\subsubsection{Absolute free energy calculations must handle the standard state and use restraints}
+\subsubsection{Absolute free energy calculations must handle the standard state and use binding site restraints}
+\label{sec:standardstate-restraints}
+
 \tocsubsubsectioncomment{@Emilio Gallicchio to comment on possible issues with theoretical discussion and possibly update in the context of revised theory section.
@@
 @Andrea Rizzi to reivew?}
-\label{sec:standardstate-restraints}
 
@@
 
-Absolute free energy calculations involve completely removing the interactions between the ligand or solute and its environment, taking it to a non-interacting state that may or may not retain intramolecular non-bonded interactions.
-This non-interacting state can then be shifted between environments---from the protein to water, or from one solution to another---without changing its free energy other than that due to the changing volume of the simulations, and then interactions can be restored in the new environment.
-
-Absolute free energies are by definition reported with respect to a specific reference or standard state, which effectively determines the arbitrary point at which the free energy is 0.
-The role of the standard state is particularly evident from the expression of the binding free energy between a receptor $R$ and ligand $L$
-\begin{equation} \label{eq:DGfromKAB}
-    \Delta G = -k_BT ~ \ln \left( c^{\circ} K_b \right)  = -k_BT ~ \ln\left( c^{\circ} \frac{[RL]}{[L][R]} \right) \, .
-\end{equation}
-Here, the reference state concentration $c^{\circ}$ converts the binding constant $K_b$ into a dimensionless quantity expressed in reference concentration units.
-It should be noted that ignoring the term $c^{\circ}$ is equivalent to assuming a reference concentration of 1~D$^{-1}$, where D are the units used to express $K_b$, and would thus cause the value of $\Delta G$ to vary with the choice of the units.
-It is convenient to define a standard state at a constant pressure of 1~atm and where each chemical species (i.e., R, L, and RL) in the reaction solvent has a concentration of $c^{\circ}$~=~1~M~=~1~molecule/1660~\r{A}$^3$ but do not interact with other molecules of $R$, $L$, or $RL$.
-
-\paragraph{Handling the standard state in absolute free energy calculations.}
-For solvation free energy calculations, handling the standard state is typically straightforward, and treating it correctly simply means ensuring that the non-interacting solute still occupies essentially the same volume as the solute in the interacting system.
-So typically in such cases no special care is required to ensure the correct standard state, as long as the \emph{experimental} data being analyzed uses the same standard state.
-If this is not the case, a simple entropic correction to the free energy of $k_B T\ln (V_f/V_i) = k_B T \ln (C_i/C_f)$ to the experimental data is needed.
-
-For binding, however, the situation is more complex and requires special care.
-Because the simulations are typically performed using restraints and at concentrations that are different from 1 M, the expression of the free energy requires the following correction~\cite{gilson1997statisticalthermodynamic} (see an example of such a thermodynamic cycle in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle})
-\begin{equation}\label{eq:restraint-correction}
-    \Delta G^{\circ}_{\mathrm{restr}} = -k_BT ~ \ln \left( c^{\circ}{V_L} \right) -k_BT ~ \ln \left( \frac{\xi_L}{8 \pi^2} \right),
-\end{equation}
-where $V_L$ and $\xi_L$ are respectively the volume of the translational and rotational degrees of freedom of the non-interacting ligand in the simulation box.
-When no restraints are used, the non-interacting ligand is free to translate and rotate in the simulation box (i.e., $V_L = V_{\mathrm{box}}$ and $\xi_L = 8\pi^2$), and the rotational term is zero.
-A sufficiently thorough exploration of the simulation box by the non-interacting ligand is, however, required for the formula to be valid.
-This is typically hard to achieve as the exploration process is governed by diffusion, and weak transient nonspecific binding will occur at other sites on the protein. 
-The addition of a restraint limits the volume available to the non-interacting ligand, thus speeding the convergence of the sampling.
-In addition, when enhanced sampling methods such as Hamiltonian $\lambda$ exchange are used (see Sec.~\ref{sec:sampling_schemes}), the use of a restraint is typically necessary as it keeps the ligand in the binding site in the interacting state (see also Sec.~\ref{sec:weak-binders}) and generally reduce the round-trip time of replicas.
-When restraints are employed, the values of $V_L$ and $\xi_L$ are restraint-dependent, but for commonly employed restraints, these can be usually easily computed analytically or numerically by solving the relevant integral.
-
-\paragraph{Several choices of restraints are possible.}\label{ABFErest}
-In practice, a variety of types of restraints are common, from simple harmonic distance restraints between the ligand and the protein~\cite{mobley2006use}, to flat-bottom restraints which work similarly but only exert a force if the ligand leaves a specific region~\cite{chen2007can}.
-Because these restraints do not limit the rotational degrees of freedom of the ligand, the rotational term entering the correction in Eq.~\ref{eq:restraint-correction} is zero.
-
-Alternatively, a set of restraints proposed by Boresch have also commonly been employed, where all six rigid-body degrees of freedom governing the orientation of the ligand relative to the receptor are restrained~\cite{boresch2003absolute, leitgeb2005alchemical}.
-Further restraints, such as on the overall ligand RMSD have also been used~\cite{woo2005calculation}.
-
-In principle, all of these forms will yield correct binding free energies in the limit of adequate sampling if their effects and connection to the standard state are correctly handled, but they have different strengths and weaknesses.
-For example, with more involved restraints, sampling at intermediate $\vec{\lambda}$ values will usually not need to be as extensive but more computational effort must go to computing the free energy to turn on the restraints.
-Additionally, such restraints would typically keep the ligand from exploring alternative binding modes.
-This  restriction may be undesirable when using Hamiltonian $\lambda$ exchange or expanded ensemble techniques where allowing the ligand to exchange binding modes when it is non-interacting could provide sampling benefits~\cite{wang2013identifying}.
-More specifically, flat-bottom restraints might allow a ligand to explore multiple binding sites, and harmonic restraints allow exploration of multiple binding modes within a site, while Boresch restraints only allow a single binding mode within a single site.
-See additional discussion of the possibility of multiple binding modes in Sec.~\ref{sec:multiple_binding_modes} below.
-
-Many choices of restraints involve selecting reference atoms.
-Again, in principle this choice is unimportant given adequate simulation time but practical considerations may be important.
-The choice is likely especially important with Boresch-style restraints, where some relative placements of reference atoms are likely to be numerically unstable; additionally, ligand reference atoms should likely be in a part of the molecule which defines the binding orientation well, rather than in a floppy solvent-exposed tail, for example.
+An absolute binding free energy calculation yields the standard free energy of formation of the molecular complex from the separated receptor and ligand species.
+Hence, as discussed in \ref{sec:non-covalent-molecular-binding}, it must depend on the standard state concentration $C^\circ$ (here $1 \, {\rm M} = 1 \, {\rm molecule}/1660$~\r{A}$^3$) and on the definition of the complex, that is on the specific binding pose for which the binding free energy is requested.
+Omitting the standard state concentration would yield an undefined result that would vary with the choice of concentration units.
+A similar vague estimate would result from omitting the binding pose definition that, for example, could equally describe binding to a specific site of the receptor or unspecific binding throughout the receptor surface depending on the extent of conformational sampling (see Sec.~\ref{sec:weak-binders}).
+
+In the statistical mechanics formulation of Gilson et al.~\cite{gilson1997statisticalthermodynamic}, described in Section \ref{sec:non-covalent-molecular-binding}, the standard state concentration appears explicitly in Eq.~(\ref{eq:DG0-binding-constant-def}) and the binding pose is defined by the indicator function $I(\zeta)$ that characterizes the partition function of the complex, Eq.~(\ref{eq:intra-zRL-def}).
+The role of the standard state and binding site definitions are particularly evident when expressing Eq.~(\ref{eq:DG0-binding-constant-def}) in terms of an ensemble average.
+To do so, we express the integrals in the numerator and denominator of Eq.~(\ref{eq:DG0-binding-constant-def}) over the same set of degrees of freedom.
+Note that the intramolecular configurational partition function of the complex, $z_{RL(v)}$, includes six more degrees of freedom--the position and orientation of the ligand relative to the receptor collectively described by the variable $\zeta$--than the product of the intramolecular partition functions, $z_{R(v)} z_{L(v)}$, of the separated receptor and ligand.
+To correct this discrepancy, we multiply and divide the right-hand side of Eq.~(\ref{eq:DG0-binding-constant-def})  by the quantity
+\begin{equation} \label{eq:Vsite-def}
+    V_{\rm site} \Omega = \int d\zeta I(\zeta) \, ,
+\end{equation}
+%\Omega is in my opinion a better symbol of the orientational extent than \xi_L. Lowercase greek letters are usually used
+%for dynamical variables rather than integrated quantities
+which measures the extent of the binding pose macrostate, where $V_{\rm site}$ corresponds to the spatial extent and $\Omega$ the orientational extent.
+The right-hand side of Eq.~(\ref{eq:Vsite-def}) combined with the integrals $z_{R(v)}$ and $z_{L(v)}$ yields a partition function similar to Eq.~(\ref{eq:intra-zRL-def}) but corresponding to a complex in which ligand and receptor are uncoupled~\cite{gallicchio2011recent} as if they were at infinite distance apart.
+The result is
+\begin{equation}\label{eq:DG0-binding-constant-def-average}
+K_b = C^\circ V_{\rm site} \frac{\Omega}{8 \pi^2} \langle e^{-\beta \Delta \Psi}  \rangle_0 
+\end{equation}
+where, similarly to Eq.~(\ref{eq:zratio-solv-average}),
+\begin{equation}\label{eq:pert-energy-def}
+\Delta \Psi(x_R, x_L, \zeta) =  \Psi_v(x_R, x_L, \zeta) - \left[  \Psi_v(x_R) + \Psi(x_L) \right]
+\end{equation}
+is the solvent-averaged binding energy of the complex when receptor and ligands are in intramolecular configurations $x_R$ and $x_L$.
+The ligand is in position and orientation $\zeta$ relative to the receptor (the uncoupled potential energy does not depend on  the position and orientation of the ligand relative to the receptor), and $\langle \ldots \rangle_0$ represents an ensemble average in the uncoupled ensemble.
+Finally, note that $\Omega = 8 \pi^2$ when the binding pose indicator function does not depend on the orientation of the ligand and Eq.~(\ref{eq:DG0-binding-constant-def-average}) simplifies to
+\begin{equation}\label{eq:DG0-binding-constant-def-average-noOmega}
+K_b = C^\circ V_{\rm site} \langle e^{-\beta \Delta \Psi}  \rangle_0 
+\end{equation}
+which, expressed in free energy units, yields
+\begin{eqnarray}\label{eq:DG0-def-Vsite}
+  \Delta G^\circ_b &=&  - k_B T \ln K_b = - k_B T \ln C^\circ V_{\rm site} - k_B T \ln \langle e^{-\beta \Delta \Psi}  \rangle_0 \nonumber \\
+{\rm \ }  &=& \Delta G^\circ_{b,{\rm ideal}} + \Delta G_{b,{\rm exc.}}
+\end{eqnarray}
+
+The term $\Delta G^\circ_{b,{\rm ideal}} = - k_B T \ln C^\circ V_{\rm site}$ in Eq.~(\ref{eq:DG0-def-Vsite}) is interpreted as the \emph{ideal} component of the standard binding free energy, that is the binding free energy that would be measured in the absence of ligand-receptor interactions ($\Delta \Psi = 0$) and when the ligand's concentration is $C^\circ$.
+The ideal term is responsible for the dependence of the standard binding free energy on the standard state concentration.
+The arguably inappropriate term ``standard state correction'' is sometimes used in the literature to refer to $\Delta G^\circ_{b,{\rm ideal}}$.
+As seen above, the ideal term is a fundamental aspect originating from the statistical mechanics theory of molecular binding rather than a mere correction to an inexact formulation.
+
+Conversely, the term $\Delta G_{b,{\rm exc.}} = - k_B T \ln \langle e^{-\beta \Delta \Psi}  \rangle_0 $ is the \emph{excess} component of the standard free energy of binding, which is the term that depends on intermolecular interactions and the specific chemical composition of the system.
+The excess component is the target of the alchemical molecular simulations discussed here, in contrast to the ideal term that is generally evaluated analytically~\cite{boresch2003absolute}.
+
+\paragraph{Handling the standard state in free energy calculations.}
+As discussed in Sections \ref{sec:theory-solvation} and \ref{sec:theory-partitioning}, the handling of the standard state in solvation and solvent partitioning free energy calculations is typically straightforward.
+Eq.~(\ref{eq:DG0-solvation-def}) identifies the ideal and excess components of the standard solvation free energy in terms of the standard gas pressure $P^\circ$, standard solution concentration $C^\circ$, and ensemble averages typically evaluated by alchemical free energy simulations.
+These relations provide simple recipes to relate computational estimates to experimental values obtained under different conditions.
+Moreover, typical experimental reporters (Henry's constants and partitioning coefficients) directly probe the excess term calculated by molecular simulations, which is independent of the standard state.
+
+As discussed above, however, the situation for binding is more complex and requires special care.
+The first important aspect is that binding free energy calculations must be performed with binding site restraints that implement the binding site indicator function $I(\zeta)$ by forcing the ligand to occupy the binding site even when it is decoupled from the receptor.
+Doing otherwise would fail to specify the bound macrostate of the complex for which we seek the free energy.
+Alchemical binding free energy simulations where the ligand leaves the binding site are not set up correctly and return undefined free energy estimates.
+Because it is a discontinuous function of the position and orientation of the ligand relative to the receptor, in molecular dynamics applications, the indicator function is often approximated by a flat-bottom restraint~\cite{chen2007can} (see below).
+Harmonic restraints are not as suitable in this respect because they favor a specific microstate at the bottom of the harmonic potential rather than specifying a region of conformational space.
+
+Note that $\Delta G^\circ_{b,{\rm ideal}}$ is independent of the chemical nature of the binding partners, and it is the same for all complexes sharing similar binding site definitions.
+The latter is the reason that the ideal term cancels out when taking the difference of absolute binding free energies when estimating relative binding free energies (RBFEs).
+However, while it does not appear explicitly in the definition, $\Delta G_{b,{\rm exc.}}$ depends implicitly on the definition of the binding site through the ensemble average in the uncoupled ensemble, which includes only ``bound'' configurations according to the definition of the indicator function $I(\zeta)$.
+Varying the binding site definition affects both the ideal and excess binding free energy terms such that their sum (the standard binding free energy) is often only weakly dependent on the binding site definition~\cite{gilson1997statisticalthermodynamic,gallicchio2011recent}.
+Conversely, great care should be exercised when comparing calculated excess binding free energy values corresponding to different binding site definitions.
+Similarly, the outcome of a binding free energy calculation conducted without a binding site definition, i.e.\ without binding site restraints, is, in principle, undefined.
+The latter is also true for relative binding free energies because while the ideal term cancels out, the difference in excess binding free energies depends on the binding pose definition.
+
+\paragraph{Several choices of binding site restraints are possible.}
+A variety of restraint types are commonly used in alchemical binding free energy calculations.
+Before going into their characteristics, we must acknowledge an important and often unrecognized distinction between the binding site restraints corresponding to the indicator function $I(\zeta)$ necessary for the implementation of the theory and optional auxiliary restraints and biasing potentials sometimes used to speed up the convergence of free energy calculations.
+Binding site restraints are an integral part of the molecular complex definition and are typically maintained throughout the alchemical process.
+In particular, they must be present unchanged at the unbound and bound states.
+In contrast, unless thermodynamic reweighting is applied in post-processing, auxiliary restraints cannot be present at the endstates to avoid affecting their free energies.
+Hence, they are typically turned on during an initial stage, starting from the uncoupled state, and are turned off in the final stage, near the bound state.
+In the alchemical binding free energy literature, often harmonic restraint potentials imposed in an initial stage and removed in a final stage are described as implementing standard state terms.
+However, because they utilize harmonic potentials that cannot correctly describe the binding pose macrostate (see above) and because they are turned on and then off, these are actually auxiliary restraints introduced to enhance convergence without affecting, in principle, binding free energy estimates.
+
+Binding site restraints implement the $I(\zeta)$ indicator function.
+They, therefore, can depend at most on the six rigid-body external coordinates of the ligand governing the position and orientation of the ligand relative to the receptor.
+Any other choice would affect the intramolecular conformational distribution of the ligand or the receptor and would  bias the binding free energy estimate.
+The external degrees of freedom introduced by Boresch et al.~\cite{boresch2003absolute,leitgeb2005alchemical} are a popular choice.
+The Boresch coordinates, as well as other similar approaches, involve selecting anchor atoms of the receptor and the ligand.
+In principle, the choice of anchor atoms is arbitrary, provided adequate simulation time is available, and as long as it leads to similar binding pose definitions.
+However, practical considerations may be important.
+For example, some choices of Boresch-style anchor atoms can be numerically unstable.
+Additionally, anchor atoms should likely be placed in a part of the molecule that defines the binding orientation well rather than in a floppy, solvent-exposed tail.
 Several automated methods have been proposed to select anchor points for Boresch restraints, which typically involve analysing a simulation of the fully-interacting receptor-ligand complex~\cite{alibay2022evaluating, alibay2021mdrestraintsgenerator, baumann2023broadening, chen2023enhancing, hedges2023suite}.
 
-More recently, restrictive restraint schemes have been proposed that avoid relying on a few critical reference atoms.
+Recently, more restrictive restraint schemes have been proposed that avoid relying on a few critical reference atoms.
 Instead, these aim to simplify restraint selection by using larger sets of atomic positions in the receptor and ligand.
 Fu et al.
-suggested restraining the six relative rigid-body degrees of freedom derived by finding the rotation of the ligand which minimises RMSD to a reference structure of the receptor-ligand complex (after accounting for rotation and translation of the receptor)~\cite{fu2017new}.
+suggested restraining the six relative rigid-body degrees of freedom derived by finding the rotation of the ligand, which minimises the RMSD to a reference structure of the receptor-ligand complex (after accounting for rotation and translation of the receptor)~\cite{fu2017new}.
 Salari et al.
 proposed restraining the ``distance-from-bound-configuration'' (DBC) variable, which is the RMSD of ligand coordinates in the frame of reference of the binding site~\cite{salari2018streamlined, ebrahimi2022symmetry}.
-This may allow the binding pose to be closely preserved as the ligand intermolecular interactions are removed, but has the disadvantage of coupling the internal and relative external degrees of freedom of the receptor and ligand, preventing easy computation of the standard state correction.
-This necessitates an additional step where the DBC restraints are released to a single harmonic restraint.
-Both restraint schemes are implemented in open-source workflows~\cite{fu2021bfee2, fu2022accurate, santiago-mcrae2023computing}.
+DBC may allow the binding pose to be closely preserved as the ligand intermolecular interactions are removed, but has the disadvantage of coupling the internal and relative external degrees of freedom of the receptor and ligand, making them unsuitable for describing the binding site macrostate and the corresponding standard state terms.
+However, these restraints are practical as auxiliary restraints where additional steps are used to introduce and release the DBC restraints.
+These restraint schemes are implemented in open-source workflows~\cite{fu2021bfee2, fu2022accurate, santiago-mcrae2023computing}.
+
+As discussed above, binding site restraints are commonly implemented using quadratic flat-bottom restraints, which impose a restoring harmonic  force similar to harmonic potentials but only if the ligand leaves a specific region~\cite{chen2007can}.
+Often, binding site restraints do not limit the rotational degrees of freedom of the ligand, and the corresponding standard state term is zero; see Eq.~(\ref{eq:DG0-def-Vsite}).
+
+Auxiliary restraints take a variety of forms, such as simple harmonic distance restraints between the ligand and the protein~\cite{mobley2006use} as well as flat-bottom restraints.
+The choice of degrees of freedom and restraining potentials for auxiliary restraints is not as restrictive as for those for binding site restraints.
+For example, the overall ligand RMSD has been used~\cite{woo2005calculation} for this purpose.
+Some auxiliary restraints are designed to reduce the amount of conformational sampling at intermediate $\vec{\lambda}$ values.
+However, more computational effort is required to compute the free energy when the restraints are turned on and off.
+Additionally, such restraints can limit the exploration of alternative binding modes.
+This  restriction may be undesirable when using Hamiltonian $\lambda$ exchange or expanded ensemble techniques where allowing the ligand to exchange binding modes when it is non-interacting could provide sampling benefits~\cite{wang2013identifying}.
+More specifically, flat-bottom restraints allow a ligand to explore multiple binding sites, and harmonic restraints allow the exploration of multiple binding modes within a site, while Boresch restraints only allow a single binding mode within a single site.
+See additional discussion of the possibility of multiple binding modes in Sec.~\ref{sec:multiple_binding_modes} below.
 

### 2025-06-02 paper/manuscript.tex

--- 2025-06-01 paper/manuscript.tex
@@
 
-Absolute free energy calculations involve completely removing the interactions between the ligand or solute and its environment, taking it to a non-interacting state that may or may not retain intramolecular non-bonded interactions.
-This non-interacting state can then be shifted between environments---from the protein to water, or from one solution to another---without changing its free energy other than that due to the changing volume of the simulations, and then interactions can be restored in the new environment.
-
-Absolute free energies are by definition reported with respect to a specific reference or standard state, which effectively determines the arbitrary point at which the free energy is 0.
-The role of the standard state is particularly evident from the expression of the binding free energy between a receptor $R$ and ligand $L$
-\begin{equation} \label{eq:DGfromKAB}
-    \Delta G = -k_BT ~ \ln \left( c^{\circ} K_b \right)  = -k_BT ~ \ln\left( c^{\circ} \frac{[RL]}{[L][R]} \right) \, .
-\end{equation}
-Here, the reference state concentration $c^{\circ}$ converts the binding constant $K_b$ into a dimensionless quantity expressed in reference concentration units.
-It should be noted that ignoring the term $c^{\circ}$ is equivalent to assuming a reference concentration of 1~D$^{-1}$, where D are the units used to express $K_b$, and would thus cause the value of $\Delta G$ to vary with the choice of the units.
-It is convenient to define a standard state at a constant pressure of 1~atm and where each chemical species (i.e., R, L, and RL) in the reaction solvent has a concentration of $c^{\circ}$~=~1~M~=~1~molecule/1660~\r{A}$^3$ but do not interact with other molecules of $R$, $L$, or $RL$.
-
-\paragraph{Handling the standard state in absolute free energy calculations.}
-For solvation free energy calculations, handling the standard state is typically straightforward, and treating it correctly simply means ensuring that the non-interacting solute still occupies essentially the same volume as the solute in the interacting system.
-So typically in such cases no special care is required to ensure the correct standard state, as long as the \emph{experimental} data being analyzed uses the same standard state.
-If this is not the case, a simple entropic correction to the free energy of $k_B T\ln (V_f/V_i) = k_B T \ln (C_i/C_f)$ to the experimental data is needed.
-
-For binding, however, the situation is more complex and requires special care.
-Because the simulations are typically performed using restraints and at concentrations that are different from 1 M, the expression of the free energy requires the following correction~\cite{gilson1997statisticalthermodynamic} (see an example of such a thermodynamic cycle in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle})
-\begin{equation}\label{eq:restraint-correction}
-    \Delta G^{\circ}_{\mathrm{restr}} = -k_BT ~ \ln \left( c^{\circ}{V_L} \right) -k_BT ~ \ln \left( \frac{\xi_L}{8 \pi^2} \right),
-\end{equation}
-where $V_L$ and $\xi_L$ are respectively the volume of the translational and rotational degrees of freedom of the non-interacting ligand in the simulation box.
-When no restraints are used, the non-interacting ligand is free to translate and rotate in the simulation box (i.e., $V_L = V_{\mathrm{box}}$ and $\xi_L = 8\pi^2$), and the rotational term is zero.
-A sufficiently thorough exploration of the simulation box by the non-interacting ligand is, however, required for the formula to be valid.
-This is typically hard to achieve as the exploration process is governed by diffusion, and weak transient nonspecific binding will occur at other sites on the protein. 
-The addition of a restraint limits the volume available to the non-interacting ligand, thus speeding the convergence of the sampling.
-In addition, when enhanced sampling methods such as Hamiltonian $\lambda$ exchange are used (see Sec.~\ref{sec:sampling_schemes}), the use of a restraint is typically necessary as it keeps the ligand in the binding site in the interacting state (see also Sec.~\ref{sec:weak-binders}) and generally reduce the round-trip time of replicas.
-When restraints are employed, the values of $V_L$ and $\xi_L$ are restraint-dependent, but for commonly employed restraints, these can be usually easily computed analytically or numerically by solving the relevant integral.
-
-\paragraph{Several choices of restraints are possible.}
-In practice, a variety of types of restraints are common, from simple harmonic distance restraints between the ligand and the protein~\cite{mobley2006use}, to flat-bottom restraints which work similarly but only exert a force if the ligand leaves a specific region~\cite{chen2007can}.
-Because these restraints do not limit the rotational degrees of freedom of the ligand, the rotational term entering the correction in Eq.~\ref{eq:restraint-correction} is zero.
-
-Alternatively, a set of restraints proposed by Boresch have also commonly been employed, where all six rigid-body degrees of freedom governing the orientation of the ligand relative to the receptor are restrained~\cite{boresch2003absolute, leitgeb2005alchemical}.
-Further restraints, such as on the overall ligand RMSD have also been used~\cite{woo2005calculation}.
-
-In principle, all of these forms will yield correct binding free energies in the limit of adequate sampling if their effects and connection to the standard state are correctly handled, but they have different strengths and weaknesses.
-For example, with more involved restraints, sampling at intermediate $\vec{\lambda}$ values will usually not need to be as extensive but more computational effort must go to computing the free energy to turn on the restraints.
-Additionally, such restraints would typically keep the ligand from exploring alternative binding modes.
-This  restriction may be undesirable when using Hamiltonian $\lambda$ exchange or expanded ensemble techniques where allowing the ligand to exchange binding modes when it is non-interacting could provide sampling benefits~\cite{wang2013identifying}.
-More specifically, flat-bottom restraints might allow a ligand to explore multiple binding sites, and harmonic restraints allow exploration of multiple binding modes within a site, while Boresch restraints only allow a single binding mode within a single site.
-See additional discussion of the possibility of multiple binding modes in Sec.~\ref{sec:multiple_binding_modes} below.
-
-Many choices of restraints involve selecting reference atoms.
-Again, in principle this choice is unimportant given adequate simulation time but practical considerations may be important.
-The choice is likely especially important with Boresch-style restraints, where some relative placements of reference atoms are likely to be numerically unstable; additionally, ligand reference atoms should likely be in a part of the molecule which defines the binding orientation well, rather than in a floppy solvent-exposed tail, for example.
+An absolute binding free energy calculation yields the standard free energy of formation of the molecular complex from the separated receptor and ligand species.
+Hence, as discussed in \ref{sec:non-covalent-molecular-binding}, it must depend on the standard state concentration $C^\circ$ (here $1 \, {\rm M} = 1 \, {\rm molecule}/1660$~\r{A}$^3$) and on the definition of the complex, that is on the specific binding pose for which the binding free energy is requested.
+Omitting the standard state concentration would yield an undefined result that would vary with the choice of concentration units.
+A similar vague estimate would result from omitting the binding pose definition that, for example, could equally describe binding to a specific site of the receptor or unspecific binding throughout the receptor surface depending on the extent of conformational sampling (see Sec.~\ref{sec:weak-binders}).
+
+In the statistical mechanics formulation of Gilson et al.~\cite{gilson1997statisticalthermodynamic}, described in Section \ref{sec:non-covalent-molecular-binding}, the standard state concentration appears explicitly in Eq.~(\ref{eq:DG0-binding-constant-def}) and the binding pose is defined by the indicator function $I(\zeta)$ that characterizes the partition function of the complex, Eq.~(\ref{eq:intra-zRL-def}).
+The role of the standard state and binding site definitions are particularly evident when expressing Eq.~(\ref{eq:DG0-binding-constant-def}) in terms of an ensemble average.
+To do so, we express the integrals in the numerator and denominator of Eq.~(\ref{eq:DG0-binding-constant-def}) over the same set of degrees of freedom.
+Note that the intramolecular configurational partition function of the complex, $z_{RL(v)}$, includes six more degrees of freedom--the position and orientation of the ligand relative to the receptor collectively described by the variable $\zeta$--than the product of the intramolecular partition functions, $z_{R(v)} z_{L(v)}$, of the separated receptor and ligand.
+To correct this discrepancy, we multiply and divide the right-hand side of Eq.~(\ref{eq:DG0-binding-constant-def})  by the quantity
+\begin{equation} \label{eq:Vsite-def}
+    V_{\rm site} \Omega = \int d\zeta I(\zeta) \, ,
+\end{equation}
+%\Omega is in my opinion a better symbol of the orientational extent than \xi_L. Lowercase greek letters are usually used
+%for dynamical variables rather than integrated quantities
+which measures the extent of the binding pose macrostate, where $V_{\rm site}$ corresponds to the spatial extent and $\Omega$ the orientational extent.
+The right-hand side of Eq.~(\ref{eq:Vsite-def}) combined with the integrals $z_{R(v)}$ and $z_{L(v)}$ yields a partition function similar to Eq.~(\ref{eq:intra-zRL-def}) but corresponding to a complex in which ligand and receptor are uncoupled~\cite{gallicchio2011recent} as if they were at infinite distance apart.
+The result is
+\begin{equation}\label{eq:DG0-binding-constant-def-average}
+K_b = C^\circ V_{\rm site} \frac{\Omega}{8 \pi^2} \langle e^{-\beta \Delta \Psi}  \rangle_0 
+\end{equation}
+where, similarly to Eq.~(\ref{eq:zratio-solv-average}),
+\begin{equation}\label{eq:pert-energy-def}
+\Delta \Psi(x_R, x_L, \zeta) =  \Psi_v(x_R, x_L, \zeta) - \left[  \Psi_v(x_R) + \Psi(x_L) \right]
+\end{equation}
+is the solvent-averaged binding energy of the complex when receptor and ligands are in intramolecular configurations $x_R$ and $x_L$.
+The ligand is in position and orientation $\zeta$ relative to the receptor (the uncoupled potential energy does not depend on  the position and orientation of the ligand relative to the receptor), and $\langle \ldots \rangle_0$ represents an ensemble average in the uncoupled ensemble.
+Finally, note that $\Omega = 8 \pi^2$ when the binding pose indicator function does not depend on the orientation of the ligand and Eq.~(\ref{eq:DG0-binding-constant-def-average}) simplifies to
+\begin{equation}\label{eq:DG0-binding-constant-def-average-noOmega}
+K_b = C^\circ V_{\rm site} \langle e^{-\beta \Delta \Psi}  \rangle_0 
+\end{equation}
+which, expressed in free energy units, yields
+\begin{eqnarray}\label{eq:DG0-def-Vsite}
+  \Delta G^\circ_b &=&  - k_B T \ln K_b = - k_B T \ln C^\circ V_{\rm site} - k_B T \ln \langle e^{-\beta \Delta \Psi}  \rangle_0 \nonumber \\
+{\rm \ }  &=& \Delta G^\circ_{b,{\rm ideal}} + \Delta G_{b,{\rm exc.}}
+\end{eqnarray}
+
+The term $\Delta G^\circ_{b,{\rm ideal}} = - k_B T \ln C^\circ V_{\rm site}$ in Eq.~(\ref{eq:DG0-def-Vsite}) is interpreted as the \emph{ideal} component of the standard binding free energy, that is the binding free energy that would be measured in the absence of ligand-receptor interactions ($\Delta \Psi = 0$) and when the ligand's concentration is $C^\circ$.
+The ideal term is responsible for the dependence of the standard binding free energy on the standard state concentration.
+The arguably inappropriate term ``standard state correction'' is sometimes used in the literature to refer to $\Delta G^\circ_{b,{\rm ideal}}$.
+As seen above, the ideal term is a fundamental aspect originating from the statistical mechanics theory of molecular binding rather than a mere correction to an inexact formulation.
+
+Conversely, the term $\Delta G_{b,{\rm exc.}} = - k_B T \ln \langle e^{-\beta \Delta \Psi}  \rangle_0 $ is the \emph{excess} component of the standard free energy of binding, which is the term that depends on intermolecular interactions and the specific chemical composition of the system.
+The excess component is the target of the alchemical molecular simulations discussed here, in contrast to the ideal term that is generally evaluated analytically~\cite{boresch2003absolute}.
+
+\paragraph{Handling the standard state in free energy calculations.}
+As discussed in Sections \ref{sec:theory-solvation} and \ref{sec:theory-partitioning}, the handling of the standard state in solvation and solvent partitioning free energy calculations is typically straightforward.
+Eq.~(\ref{eq:DG0-solvation-def}) identifies the ideal and excess components of the standard solvation free energy in terms of the standard gas pressure $P^\circ$, standard solution concentration $C^\circ$, and ensemble averages typically evaluated by alchemical free energy simulations.
+These relations provide simple recipes to relate computational estimates to experimental values obtained under different conditions.
+Moreover, typical experimental reporters (Henry's constants and partitioning coefficients) directly probe the excess term calculated by molecular simulations, which is independent of the standard state.
+
+As discussed above, however, the situation for binding is more complex and requires special care.
+The first important aspect is that binding free energy calculations must be performed with binding site restraints that implement the binding site indicator function $I(\zeta)$ by forcing the ligand to occupy the binding site even when it is decoupled from the receptor.
+Doing otherwise would fail to specify the bound macrostate of the complex for which we seek the free energy.
+Alchemical binding free energy simulations where the ligand leaves the binding site are not set up correctly and return undefined free energy estimates.
+Because it is a discontinuous function of the position and orientation of the ligand relative to the receptor, in molecular dynamics applications, the indicator function is often approximated by a flat-bottom restraint~\cite{chen2007can} (see below).
+Harmonic restraints are not as suitable in this respect because they favor a specific microstate at the bottom of the harmonic potential rather than specifying a region of conformational space.
+
+Note that $\Delta G^\circ_{b,{\rm ideal}}$ is independent of the chemical nature of the binding partners, and it is the same for all complexes sharing similar binding site definitions.
+The latter is the reason that the ideal term cancels out when taking the difference of absolute binding free energies when estimating relative binding free energies (RBFEs).
+However, while it does not appear explicitly in the definition, $\Delta G_{b,{\rm exc.}}$ depends implicitly on the definition of the binding site through the ensemble average in the uncoupled ensemble, which includes only ``bound'' configurations according to the definition of the indicator function $I(\zeta)$.
+Varying the binding site definition affects both the ideal and excess binding free energy terms such that their sum (the standard binding free energy) is often only weakly dependent on the binding site definition~\cite{gilson1997statisticalthermodynamic,gallicchio2011recent}.
+Conversely, great care should be exercised when comparing calculated excess binding free energy values corresponding to different binding site definitions.
+Similarly, the outcome of a binding free energy calculation conducted without a binding site definition, i.e.\ without binding site restraints, is, in principle, undefined.
+The latter is also true for relative binding free energies because while the ideal term cancels out, the difference in excess binding free energies depends on the binding pose definition.
+
+\paragraph{Several choices of binding site restraints are possible.}
+A variety of restraint types are commonly used in alchemical binding free energy calculations.
+Before going into their characteristics, we must acknowledge an important and often unrecognized distinction between the binding site restraints corresponding to the indicator function $I(\zeta)$ necessary for the implementation of the theory and optional auxiliary restraints and biasing potentials sometimes used to speed up the convergence of free energy calculations.
+Binding site restraints are an integral part of the molecular complex definition and are typically maintained throughout the alchemical process.
+In particular, they must be present unchanged at the unbound and bound states.
+In contrast, unless thermodynamic reweighting is applied in post-processing, auxiliary restraints cannot be present at the endstates to avoid affecting their free energies.
+Hence, they are typically turned on during an initial stage, starting from the uncoupled state, and are turned off in the final stage, near the bound state.
+In the alchemical binding free energy literature, often harmonic restraint potentials imposed in an initial stage and removed in a final stage are described as implementing standard state terms.
+However, because they utilize harmonic potentials that cannot correctly describe the binding pose macrostate (see above) and because they are turned on and then off, these are actually auxiliary restraints introduced to enhance convergence without affecting, in principle, binding free energy estimates.
+
+Binding site restraints implement the $I(\zeta)$ indicator function.
+They, therefore, can depend at most on the six rigid-body external coordinates of the ligand governing the position and orientation of the ligand relative to the receptor.
+Any other choice would affect the intramolecular conformational distribution of the ligand or the receptor and would  bias the binding free energy estimate.
+The external degrees of freedom introduced by Boresch et al.~\cite{boresch2003absolute,leitgeb2005alchemical} are a popular choice.
+The Boresch coordinates, as well as other similar approaches, involve selecting anchor atoms of the receptor and the ligand.
+In principle, the choice of anchor atoms is arbitrary, provided adequate simulation time is available, and as long as it leads to similar binding pose definitions.
+However, practical considerations may be important.
+For example, some choices of Boresch-style anchor atoms can be numerically unstable.
+Additionally, anchor atoms should likely be placed in a part of the molecule that defines the binding orientation well rather than in a floppy, solvent-exposed tail.
 Several automated methods have been proposed to select anchor points for Boresch restraints, which typically involve analysing a simulation of the fully-interacting receptor-ligand complex~\cite{alibay2022evaluating, alibay2021mdrestraintsgenerator, baumann2023broadening, chen2023enhancing, hedges2023suite}.
 
-More recently, restrictive restraint schemes have been proposed that avoid relying on a few critical reference atoms.
+Recently, more restrictive restraint schemes have been proposed that avoid relying on a few critical reference atoms.
 Instead, these aim to simplify restraint selection by using larger sets of atomic positions in the receptor and ligand.
 Fu et al.
-suggested restraining the six relative rigid-body degrees of freedom derived by finding the rotation of the ligand which minimises RMSD to a reference structure of the receptor-ligand complex (after accounting for rotation and translation of the receptor)~\cite{fu2017new}.
+suggested restraining the six relative rigid-body degrees of freedom derived by finding the rotation of the ligand, which minimises the RMSD to a reference structure of the receptor-ligand complex (after accounting for rotation and translation of the receptor)~\cite{fu2017new}.
 Salari et al.
 proposed restraining the ``distance-from-bound-configuration'' (DBC) variable, which is the RMSD of ligand coordinates in the frame of reference of the binding site~\cite{salari2018streamlined, ebrahimi2022symmetry}.
-This may allow the binding pose to be closely preserved as the ligand intermolecular interactions are removed, but has the disadvantage of coupling the internal and relative external degrees of freedom of the receptor and ligand, preventing easy computation of the standard state correction.
-This necessitates an additional step where the DBC restraints are released to a single harmonic restraint.
-Both restraint schemes are implemented in open-source workflows~\cite{fu2021bfee2, fu2022accurate, santiago-mcrae2023computing}.
+DBC may allow the binding pose to be closely preserved as the ligand intermolecular interactions are removed, but has the disadvantage of coupling the internal and relative external degrees of freedom of the receptor and ligand, making them unsuitable for describing the binding site macrostate and the corresponding standard state terms.
+However, these restraints are practical as auxiliary restraints where additional steps are used to introduce and release the DBC restraints.
+These restraint schemes are implemented in open-source workflows~\cite{fu2021bfee2, fu2022accurate, santiago-mcrae2023computing}.
+
+As discussed above, binding site restraints are commonly implemented using quadratic flat-bottom restraints, which impose a restoring harmonic  force similar to harmonic potentials but only if the ligand leaves a specific region~\cite{chen2007can}.
+Often, binding site restraints do not limit the rotational degrees of freedom of the ligand, and the corresponding standard state term is zero; see Eq.~(\ref{eq:DG0-def-Vsite}).
+
+Auxiliary restraints take a variety of forms, such as simple harmonic distance restraints between the ligand and the protein~\cite{mobley2006use} as well as flat-bottom restraints.
+The choice of degrees of freedom and restraining potentials for auxiliary restraints is not as restrictive as for those for binding site restraints.
+For example, the overall ligand RMSD has been used~\cite{woo2005calculation} for this purpose.
+Some auxiliary restraints are designed to reduce the amount of conformational sampling at intermediate $\vec{\lambda}$ values.
+However, more computational effort is required to compute the free energy when the restraints are turned on and off.
+Additionally, such restraints can limit the exploration of alternative binding modes.
+This  restriction may be undesirable when using Hamiltonian $\lambda$ exchange or expanded ensemble techniques where allowing the ligand to exchange binding modes when it is non-interacting could provide sampling benefits~\cite{wang2013identifying}.
+More specifically, flat-bottom restraints allow a ligand to explore multiple binding sites, and harmonic restraints allow the exploration of multiple binding modes within a site, while Boresch restraints only allow a single binding mode within a single site.
+See additional discussion of the possibility of multiple binding modes in Sec.~\ref{sec:multiple_binding_modes} below.
 

### 2025-06-06 paper/manuscript.tex
Update paper/manuscript.tex
Co-authored-by: Irfan Alibay <IAlibay@users.noreply.github.com>

--- 2025-06-02 paper/manuscript.tex
@@
 \author[1]{Jenke Scheen}
+\author[11]{David R. Slochower}
 \author[8*]{Michael R. Shirts}
@@
 \affil[10]{Cresset, Cambridgeshire, UK}
+\affil[11]{Vertex Pharmaceuticals, San Diego, CA, USA}
 
@@
 \orcid{Jenke Scheen}{0000-0001-9781-0445}
+\orcid{David R. Slochower}{0000-0003-3928-5050}
 \orcid{Michael R. Shirts}{0000-0003-3249-1097}
@@
 
-\subsection{The chemical potential}
+\subsection{Chemical potential}
 
@@
 
-\subsection{Solvation}\label{sec:theory-solvation}
+\subsection{Solvation}
 
@@
 
-\subsection{Solvent partitioning}\label{sec:theory-partitioning}
+\subsection{Solvent partitioning}
 
@@
 \subsection{Non-covalent molecular binding}
-\label{sec:non-covalent-molecular-binding}
 
@@
 Well-known protein systems with water mediated ligand interactions are for example: HSP90 which formed part of the D3R grand challenge 2015~\cite{mey2016blinded}, A2A~\cite{brucemacdonald2018ligand}, MUP~\cite{ross2015water}, ~\cite{deflorian2020accurate}, and others~\cite{michel2009energetics}.
+
+\paragraph{Membrane proteins require additional care}
+Integral membrane proteins require embedding in a suitable model membrane to remain stable during simulations.
+Lipids are usually not included in protein force fields and, although they may be covered by generic small molecule force fields, have often been parameterized separately (e.g., AMBER's Lipid21 \cite{AMBERLipid21} and the CHARMM lipid force field \cite{CHARMMLipidFF}).
+Membrane systems can be simulated in multiple thermodynamic ensembles, but the semi-isotropic scheme $NP_{z}P_{xy}T$ is frequently employed, allowing the pressure in the bilayer plane ($xy$) to scale independently \cite{LipidBestPracticesCOMS}.
+These systems require more extensive equilibration than soluble proteins.
+Some packages, such as Desmond, include a 7-step protocol to relax membrane systems.
+
+Alchemical transformations that involve net charge changes can provide difficulties for membrane systems; phospholipids carry functional groups with nonzero formal charge and cannot be treated as a solvent with a uniform dielectric in the same way as water.
+Several approaches are reviewed in \cite{wu2022correction-51c}.
+The authors recommend the use of a co-alchemical ion, specifically making an ion of equivalent charge appear alongside the disappearing charged molecule to maintain system neutrality.
+
+In the case of a lipid-facing binding site, there are two additional challenges.
+First, simulation time should be extended to ensure sufficient sampling of lipid configurational space.
+Lipids may either bind alongside small molecules, acting as cofactors, or be displaced from the binding site by a ligand.
+If lipids participate in binding through protein-lipid or ligand-lipid interactions, it is necessary to sample these intermolecular interactions through additional $\lambda$ windows, longer simulation times, or different alchemical protocols.
+Displacing a lipid will have its own free energy cost which may take a long time to converge if it requires diffusion and rearrangement of nearby lipids.
+Second, the free energy of binding for a small molecule may be strongly coupled to the lipid species.
+In some cases, the identity of the lipid may be known.
+But in other cases, it may be difficult to discern the identity of the lipid from X-ray or cryo-EM structures due to the flexible nature of lipid head groups and benchmarking is recommended to match experimental data.
 
@@
 
-\subsubsection{Absolute free energy calculations must handle the standard state and use binding site restraints}
-\label{sec:standardstate-restraints}
-
+\subsubsection{Absolute free energy calculations must handle the standard state and use restraints}
 \tocsubsubsectioncomment{@Emilio Gallicchio to comment on possible issues with theoretical discussion and possibly update in the context of revised theory section.
@@
 @Andrea Rizzi to reivew?}
+\label{sec:standardstate-restraints}
 
@@
 
-An absolute binding free energy calculation yields the standard free energy of formation of the molecular complex from the separated receptor and ligand species.
-Hence, as discussed in \ref{sec:non-covalent-molecular-binding}, it must depend on the standard state concentration $C^\circ$ (here $1 \, {\rm M} = 1 \, {\rm molecule}/1660$~\r{A}$^3$) and on the definition of the complex, that is on the specific binding pose for which the binding free energy is requested.
-Omitting the standard state concentration would yield an undefined result that would vary with the choice of concentration units.
-A similar vague estimate would result from omitting the binding pose definition that, for example, could equally describe binding to a specific site of the receptor or unspecific binding throughout the receptor surface depending on the extent of conformational sampling (see Sec.~\ref{sec:weak-binders}).
-
-In the statistical mechanics formulation of Gilson et al.~\cite{gilson1997statisticalthermodynamic}, described in Section \ref{sec:non-covalent-molecular-binding}, the standard state concentration appears explicitly in Eq.~(\ref{eq:DG0-binding-constant-def}) and the binding pose is defined by the indicator function $I(\zeta)$ that characterizes the partition function of the complex, Eq.~(\ref{eq:intra-zRL-def}).
-The role of the standard state and binding site definitions are particularly evident when expressing Eq.~(\ref{eq:DG0-binding-constant-def}) in terms of an ensemble average.
-To do so, we express the integrals in the numerator and denominator of Eq.~(\ref{eq:DG0-binding-constant-def}) over the same set of degrees of freedom.
-Note that the intramolecular configurational partition function of the complex, $z_{RL(v)}$, includes six more degrees of freedom--the position and orientation of the ligand relative to the receptor collectively described by the variable $\zeta$--than the product of the intramolecular partition functions, $z_{R(v)} z_{L(v)}$, of the separated receptor and ligand.
-To correct this discrepancy, we multiply and divide the right-hand side of Eq.~(\ref{eq:DG0-binding-constant-def})  by the quantity
-\begin{equation} \label{eq:Vsite-def}
-    V_{\rm site} \Omega = \int d\zeta I(\zeta) \, ,
-\end{equation}
-%\Omega is in my opinion a better symbol of the orientational extent than \xi_L. Lowercase greek letters are usually used
-%for dynamical variables rather than integrated quantities
-which measures the extent of the binding pose macrostate, where $V_{\rm site}$ corresponds to the spatial extent and $\Omega$ the orientational extent.
-The right-hand side of Eq.~(\ref{eq:Vsite-def}) combined with the integrals $z_{R(v)}$ and $z_{L(v)}$ yields a partition function similar to Eq.~(\ref{eq:intra-zRL-def}) but corresponding to a complex in which ligand and receptor are uncoupled~\cite{gallicchio2011recent} as if they were at infinite distance apart.
-The result is
-\begin{equation}\label{eq:DG0-binding-constant-def-average}
-K_b = C^\circ V_{\rm site} \frac{\Omega}{8 \pi^2} \langle e^{-\beta \Delta \Psi}  \rangle_0 
-\end{equation}
-where, similarly to Eq.~(\ref{eq:zratio-solv-average}),
-\begin{equation}\label{eq:pert-energy-def}
-\Delta \Psi(x_R, x_L, \zeta) =  \Psi_v(x_R, x_L, \zeta) - \left[  \Psi_v(x_R) + \Psi(x_L) \right]
-\end{equation}
-is the solvent-averaged binding energy of the complex when receptor and ligands are in intramolecular configurations $x_R$ and $x_L$.
-The ligand is in position and orientation $\zeta$ relative to the receptor (the uncoupled potential energy does not depend on  the position and orientation of the ligand relative to the receptor), and $\langle \ldots \rangle_0$ represents an ensemble average in the uncoupled ensemble.
-Finally, note that $\Omega = 8 \pi^2$ when the binding pose indicator function does not depend on the orientation of the ligand and Eq.~(\ref{eq:DG0-binding-constant-def-average}) simplifies to
-\begin{equation}\label{eq:DG0-binding-constant-def-average-noOmega}
-K_b = C^\circ V_{\rm site} \langle e^{-\beta \Delta \Psi}  \rangle_0 
-\end{equation}
-which, expressed in free energy units, yields
-\begin{eqnarray}\label{eq:DG0-def-Vsite}
-  \Delta G^\circ_b &=&  - k_B T \ln K_b = - k_B T \ln C^\circ V_{\rm site} - k_B T \ln \langle e^{-\beta \Delta \Psi}  \rangle_0 \nonumber \\
-{\rm \ }  &=& \Delta G^\circ_{b,{\rm ideal}} + \Delta G_{b,{\rm exc.}}
-\end{eqnarray}
-
-The term $\Delta G^\circ_{b,{\rm ideal}} = - k_B T \ln C^\circ V_{\rm site}$ in Eq.~(\ref{eq:DG0-def-Vsite}) is interpreted as the \emph{ideal} component of the standard binding free energy, that is the binding free energy that would be measured in the absence of ligand-receptor interactions ($\Delta \Psi = 0$) and when the ligand's concentration is $C^\circ$.
-The ideal term is responsible for the dependence of the standard binding free energy on the standard state concentration.
-The arguably inappropriate term ``standard state correction'' is sometimes used in the literature to refer to $\Delta G^\circ_{b,{\rm ideal}}$.
-As seen above, the ideal term is a fundamental aspect originating from the statistical mechanics theory of molecular binding rather than a mere correction to an inexact formulation.
-
-Conversely, the term $\Delta G_{b,{\rm exc.}} = - k_B T \ln \langle e^{-\beta \Delta \Psi}  \rangle_0 $ is the \emph{excess} component of the standard free energy of binding, which is the term that depends on intermolecular interactions and the specific chemical composition of the system.
-The excess component is the target of the alchemical molecular simulations discussed here, in contrast to the ideal term that is generally evaluated analytically~\cite{boresch2003absolute}.
-
-\paragraph{Handling the standard state in free energy calculations.}
-As discussed in Sections \ref{sec:theory-solvation} and \ref{sec:theory-partitioning}, the handling of the standard state in solvation and solvent partitioning free energy calculations is typically straightforward.
-Eq.~(\ref{eq:DG0-solvation-def}) identifies the ideal and excess components of the standard solvation free energy in terms of the standard gas pressure $P^\circ$, standard solution concentration $C^\circ$, and ensemble averages typically evaluated by alchemical free energy simulations.
-These relations provide simple recipes to relate computational estimates to experimental values obtained under different conditions.
-Moreover, typical experimental reporters (Henry's constants and partitioning coefficients) directly probe the excess term calculated by molecular simulations, which is independent of the standard state.
-
-As discussed above, however, the situation for binding is more complex and requires special care.
-The first important aspect is that binding free energy calculations must be performed with binding site restraints that implement the binding site indicator function $I(\zeta)$ by forcing the ligand to occupy the binding site even when it is decoupled from the receptor.
-Doing otherwise would fail to specify the bound macrostate of the complex for which we seek the free energy.
-Alchemical binding free energy simulations where the ligand leaves the binding site are not set up correctly and return undefined free energy estimates.
-Because it is a discontinuous function of the position and orientation of the ligand relative to the receptor, in molecular dynamics applications, the indicator function is often approximated by a flat-bottom restraint~\cite{chen2007can} (see below).
-Harmonic restraints are not as suitable in this respect because they favor a specific microstate at the bottom of the harmonic potential rather than specifying a region of conformational space.
-
-Note that $\Delta G^\circ_{b,{\rm ideal}}$ is independent of the chemical nature of the binding partners, and it is the same for all complexes sharing similar binding site definitions.
-The latter is the reason that the ideal term cancels out when taking the difference of absolute binding free energies when estimating relative binding free energies (RBFEs).
-However, while it does not appear explicitly in the definition, $\Delta G_{b,{\rm exc.}}$ depends implicitly on the definition of the binding site through the ensemble average in the uncoupled ensemble, which includes only ``bound'' configurations according to the definition of the indicator function $I(\zeta)$.
-Varying the binding site definition affects both the ideal and excess binding free energy terms such that their sum (the standard binding free energy) is often only weakly dependent on the binding site definition~\cite{gilson1997statisticalthermodynamic,gallicchio2011recent}.
-Conversely, great care should be exercised when comparing calculated excess binding free energy values corresponding to different binding site definitions.
-Similarly, the outcome of a binding free energy calculation conducted without a binding site definition, i.e.\ without binding site restraints, is, in principle, undefined.
-The latter is also true for relative binding free energies because while the ideal term cancels out, the difference in excess binding free energies depends on the binding pose definition.
-
-\paragraph{Several choices of binding site restraints are possible.}
-A variety of restraint types are commonly used in alchemical binding free energy calculations.
-Before going into their characteristics, we must acknowledge an important and often unrecognized distinction between the binding site restraints corresponding to the indicator function $I(\zeta)$ necessary for the implementation of the theory and optional auxiliary restraints and biasing potentials sometimes used to speed up the convergence of free energy calculations.
-Binding site restraints are an integral part of the molecular complex definition and are typically maintained throughout the alchemical process.
-In particular, they must be present unchanged at the unbound and bound states.
-In contrast, unless thermodynamic reweighting is applied in post-processing, auxiliary restraints cannot be present at the endstates to avoid affecting their free energies.
-Hence, they are typically turned on during an initial stage, starting from the uncoupled state, and are turned off in the final stage, near the bound state.
-In the alchemical binding free energy literature, often harmonic restraint potentials imposed in an initial stage and removed in a final stage are described as implementing standard state terms.
-However, because they utilize harmonic potentials that cannot correctly describe the binding pose macrostate (see above) and because they are turned on and then off, these are actually auxiliary restraints introduced to enhance convergence without affecting, in principle, binding free energy estimates.
-
-Binding site restraints implement the $I(\zeta)$ indicator function.
-They, therefore, can depend at most on the six rigid-body external coordinates of the ligand governing the position and orientation of the ligand relative to the receptor.
-Any other choice would affect the intramolecular conformational distribution of the ligand or the receptor and would  bias the binding free energy estimate.
-The external degrees of freedom introduced by Boresch et al.~\cite{boresch2003absolute,leitgeb2005alchemical} are a popular choice.
-The Boresch coordinates, as well as other similar approaches, involve selecting anchor atoms of the receptor and the ligand.
-In principle, the choice of anchor atoms is arbitrary, provided adequate simulation time is available, and as long as it leads to similar binding pose definitions.
-However, practical considerations may be important.
-For example, some choices of Boresch-style anchor atoms can be numerically unstable.
-Additionally, anchor atoms should likely be placed in a part of the molecule that defines the binding orientation well rather than in a floppy, solvent-exposed tail.
+Absolute free energy calculations involve completely removing the interactions between the ligand or solute and its environment, taking it to a non-interacting state that may or may not retain intramolecular non-bonded interactions.
+This non-interacting state can then be shifted between environments---from the protein to water, or from one solution to another---without changing its free energy other than that due to the changing volume of the simulations, and then interactions can be restored in the new environment.
+
+Absolute free energies are by definition reported with respect to a specific reference or standard state, which effectively determines the arbitrary point at which the free energy is 0.
+The role of the standard state is particularly evident from the expression of the binding free energy between a receptor $R$ and ligand $L$
+\begin{equation} \label{eq:DGfromKAB}
+    \Delta G = -k_BT ~ \ln \left( c^{\circ} K_b \right)  = -k_BT ~ \ln\left( c^{\circ} \frac{[RL]}{[L][R]} \right) \, .
+\end{equation}
+Here, the reference state concentration $c^{\circ}$ converts the binding constant $K_b$ into a dimensionless quantity expressed in reference concentration units.
+It should be noted that ignoring the term $c^{\circ}$ is equivalent to assuming a reference concentration of 1~D$^{-1}$, where D are the units used to express $K_b$, and would thus cause the value of $\Delta G$ to vary with the choice of the units.
+It is convenient to define a standard state at a constant pressure of 1~atm and where each chemical species (i.e., R, L, and RL) in the reaction solvent has a concentration of $c^{\circ}$~=~1~M~=~1~molecule/1660~\r{A}$^3$ but do not interact with other molecules of $R$, $L$, or $RL$.
+
+\paragraph{Handling the standard state in absolute free energy calculations.}
+For solvation free energy calculations, handling the standard state is typically straightforward, and treating it correctly simply means ensuring that the non-interacting solute still occupies essentially the same volume as the solute in the interacting system.
+So typically in such cases no special care is required to ensure the correct standard state, as long as the \emph{experimental} data being analyzed uses the same standard state.
+If this is not the case, a simple entropic correction to the free energy of $k_B T\ln (V_f/V_i) = k_B T \ln (C_i/C_f)$ to the experimental data is needed.
+
+For binding, however, the situation is more complex and requires special care.
+Because the simulations are typically performed using restraints and at concentrations that are different from 1 M, the expression of the free energy requires the following correction~\cite{gilson1997statisticalthermodynamic} (see an example of such a thermodynamic cycle in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle})
+\begin{equation}\label{eq:restraint-correction}
+    \Delta G^{\circ}_{\mathrm{restr}} = -k_BT ~ \ln \left( c^{\circ}{V_L} \right) -k_BT ~ \ln \left( \frac{\xi_L}{8 \pi^2} \right),
+\end{equation}
+where $V_L$ and $\xi_L$ are respectively the volume of the translational and rotational degrees of freedom of the non-interacting ligand in the simulation box.
+When no restraints are used, the non-interacting ligand is free to translate and rotate in the simulation box (i.e., $V_L = V_{\mathrm{box}}$ and $\xi_L = 8\pi^2$), and the rotational term is zero.
+A sufficiently thorough exploration of the simulation box by the non-interacting ligand is, however, required for the formula to be valid.
+This is typically hard to achieve as the exploration process is governed by diffusion, and weak transient nonspecific binding will occur at other sites on the protein. 
+The addition of a restraint limits the volume available to the non-interacting ligand, thus speeding the convergence of the sampling.
+In addition, when enhanced sampling methods such as Hamiltonian $\lambda$ exchange are used (see Sec.~\ref{sec:sampling_schemes}), the use of a restraint is typically necessary as it keeps the ligand in the binding site in the interacting state (see also Sec.~\ref{sec:weak-binders}) and generally reduce the round-trip time of replicas.
+When restraints are employed, the values of $V_L$ and $\xi_L$ are restraint-dependent, but for commonly employed restraints, these can be usually easily computed analytically or numerically by solving the relevant integral.
+
+\paragraph{Several choices of restraints are possible.}
+In practice, a variety of types of restraints are common, from simple harmonic distance restraints between the ligand and the protein~\cite{mobley2006use}, to flat-bottom restraints which work similarly but only exert a force if the ligand leaves a specific region~\cite{chen2007can}.
+Because these restraints do not limit the rotational degrees of freedom of the ligand, the rotational term entering the correction in Eq.~\ref{eq:restraint-correction} is zero.
+
+Alternatively, a set of restraints proposed by Boresch have also commonly been employed, where all six rigid-body degrees of freedom governing the orientation of the ligand relative to the receptor are restrained~\cite{boresch2003absolute, leitgeb2005alchemical}.
+Further restraints, such as on the overall ligand RMSD have also been used~\cite{woo2005calculation}.
+
+In principle, all of these forms will yield correct binding free energies in the limit of adequate sampling if their effects and connection to the standard state are correctly handled, but they have different strengths and weaknesses.
+For example, with more involved restraints, sampling at intermediate $\vec{\lambda}$ values will usually not need to be as extensive but more computational effort must go to computing the free energy to turn on the restraints.
+Additionally, such restraints would typically keep the ligand from exploring alternative binding modes.
+This  restriction may be undesirable when using Hamiltonian $\lambda$ exchange or expanded ensemble techniques where allowing the ligand to exchange binding modes when it is non-interacting could provide sampling benefits~\cite{wang2013identifying}.
+More specifically, flat-bottom restraints might allow a ligand to explore multiple binding sites, and harmonic restraints allow exploration of multiple binding modes within a site, while Boresch restraints only allow a single binding mode within a single site.
+See additional discussion of the possibility of multiple binding modes in Sec.~\ref{sec:multiple_binding_modes} below.
+
+Many choices of restraints involve selecting reference atoms.
+Again, in principle this choice is unimportant given adequate simulation time but practical considerations may be important.
+The choice is likely especially important with Boresch-style restraints, where some relative placements of reference atoms are likely to be numerically unstable; additionally, ligand reference atoms should likely be in a part of the molecule which defines the binding orientation well, rather than in a floppy solvent-exposed tail, for example.
 Several automated methods have been proposed to select anchor points for Boresch restraints, which typically involve analysing a simulation of the fully-interacting receptor-ligand complex~\cite{alibay2022evaluating, alibay2021mdrestraintsgenerator, baumann2023broadening, chen2023enhancing, hedges2023suite}.
 
-Recently, more restrictive restraint schemes have been proposed that avoid relying on a few critical reference atoms.
+More recently, restrictive restraint schemes have been proposed that avoid relying on a few critical reference atoms.
 Instead, these aim to simplify restraint selection by using larger sets of atomic positions in the receptor and ligand.
 Fu et al.
-suggested restraining the six relative rigid-body degrees of freedom derived by finding the rotation of the ligand, which minimises the RMSD to a reference structure of the receptor-ligand complex (after accounting for rotation and translation of the receptor)~\cite{fu2017new}.
+suggested restraining the six relative rigid-body degrees of freedom derived by finding the rotation of the ligand which minimises RMSD to a reference structure of the receptor-ligand complex (after accounting for rotation and translation of the receptor)~\cite{fu2017new}.
 Salari et al.
 proposed restraining the ``distance-from-bound-configuration'' (DBC) variable, which is the RMSD of ligand coordinates in the frame of reference of the binding site~\cite{salari2018streamlined, ebrahimi2022symmetry}.
-DBC may allow the binding pose to be closely preserved as the ligand intermolecular interactions are removed, but has the disadvantage of coupling the internal and relative external degrees of freedom of the receptor and ligand, making them unsuitable for describing the binding site macrostate and the corresponding standard state terms.
-However, these restraints are practical as auxiliary restraints where additional steps are used to introduce and release the DBC restraints.
-These restraint schemes are implemented in open-source workflows~\cite{fu2021bfee2, fu2022accurate, santiago-mcrae2023computing}.
-
-As discussed above, binding site restraints are commonly implemented using quadratic flat-bottom restraints, which impose a restoring harmonic  force similar to harmonic potentials but only if the ligand leaves a specific region~\cite{chen2007can}.
-Often, binding site restraints do not limit the rotational degrees of freedom of the ligand, and the corresponding standard state term is zero; see Eq.~(\ref{eq:DG0-def-Vsite}).
-
-Auxiliary restraints take a variety of forms, such as simple harmonic distance restraints between the ligand and the protein~\cite{mobley2006use} as well as flat-bottom restraints.
-The choice of degrees of freedom and restraining potentials for auxiliary restraints is not as restrictive as for those for binding site restraints.
-For example, the overall ligand RMSD has been used~\cite{woo2005calculation} for this purpose.
-Some auxiliary restraints are designed to reduce the amount of conformational sampling at intermediate $\vec{\lambda}$ values.
-However, more computational effort is required to compute the free energy when the restraints are turned on and off.
-Additionally, such restraints can limit the exploration of alternative binding modes.
-This  restriction may be undesirable when using Hamiltonian $\lambda$ exchange or expanded ensemble techniques where allowing the ligand to exchange binding modes when it is non-interacting could provide sampling benefits~\cite{wang2013identifying}.
-More specifically, flat-bottom restraints allow a ligand to explore multiple binding sites, and harmonic restraints allow the exploration of multiple binding modes within a site, while Boresch restraints only allow a single binding mode within a single site.
-See additional discussion of the possibility of multiple binding modes in Sec.~\ref{sec:multiple_binding_modes} below.
+This may allow the binding pose to be closely preserved as the ligand intermolecular interactions are removed, but has the disadvantage of coupling the internal and relative external degrees of freedom of the receptor and ligand, preventing easy computation of the standard state correction.
+This necessitates an additional step where the DBC restraints are released to a single harmonic restraint.
+Both restraint schemes are implemented in open-source workflows~\cite{fu2021bfee2, fu2022accurate, santiago-mcrae2023computing}.
 

### 2025-06-08 paper/manuscript.tex
comments on weak binders section

--- 2025-06-02 paper/manuscript.tex
@@
 }
+
+% aux. comments Emilio Gallicchio
+\newcommand{\egnote}[1]{%
+  {\bfseries{}[EG: }%
+  {\textcolor{green}{#1}}{\bfseries{}]}
+}
+
 \newcommand{\rref}{{\bfseries[REFs]}{}}
@@
 
-In the statistical mechanics formulation of Gilson et al.~\cite{gilson1997statisticalthermodynamic}, described in Section \ref{sec:non-covalent-molecular-binding}, the standard state concentration appears explicitly in Eq.~(\ref{eq:DG0-binding-constant-def}) and the binding pose is defined by the indicator function $I(\zeta)$ that characterizes the partition function of the complex, Eq.~(\ref{eq:intra-zRL-def}).
+In the statistical mechanics formulation of Gilson et al.~\cite{gilson1997statisticalthermodynamic}, described in Section \ref{sec:non-covalent-molecular-binding}, the standard state concentration appears explicitly in Eq.~(\ref{eq:DG0-binding-constant-def}) and the binding pose is defined by the indicator function $I(\zeta)$ that characterizes the partition function of the complex; Eq.~(\ref{eq:intra-zRL-def}).
 The role of the standard state and binding site definitions are particularly evident when expressing Eq.~(\ref{eq:DG0-binding-constant-def}) in terms of an ensemble average.
@@
 \end{equation}
-is the solvent-averaged binding energy of the complex when receptor and ligands are in intramolecular configurations $x_R$ and $x_L$.
-The ligand is in position and orientation $\zeta$ relative to the receptor (the uncoupled potential energy does not depend on  the position and orientation of the ligand relative to the receptor), and $\langle \ldots \rangle_0$ represents an ensemble average in the uncoupled ensemble.
+is the solvent-averaged binding energy of the complex when receptor and ligands are in intramolecular configurations $x_R$ and $x_L$, the ligand is in position and orientation $\zeta$ relative to the receptor (the uncoupled potential energy does not depend on  the position and orientation of the ligand relative to the receptor), and $\langle \ldots \rangle_0$ represents an ensemble average in the uncoupled ensemble.
 Finally, note that $\Omega = 8 \pi^2$ when the binding pose indicator function does not depend on the orientation of the ligand and Eq.~(\ref{eq:DG0-binding-constant-def-average}) simplifies to
@@
 The arguably inappropriate term ``standard state correction'' is sometimes used in the literature to refer to $\Delta G^\circ_{b,{\rm ideal}}$.
-As seen above, the ideal term is a fundamental aspect originating from the statistical mechanics theory of molecular binding rather than a mere correction to an inexact formulation.
+As seen above, the ideal term is a fundamental aspect of the statistical mechanics theory of molecular binding rather than a mere correction to an inexact formulation.
 
@@
 
-As discussed above, however, the situation for binding is more complex and requires special care.
+As discussed above, the situation for binding is more complex and requires special care.
 The first important aspect is that binding free energy calculations must be performed with binding site restraints that implement the binding site indicator function $I(\zeta)$ by forcing the ligand to occupy the binding site even when it is decoupled from the receptor.
 Doing otherwise would fail to specify the bound macrostate of the complex for which we seek the free energy.
-Alchemical binding free energy simulations where the ligand leaves the binding site are not set up correctly and return undefined free energy estimates.
-Because it is a discontinuous function of the position and orientation of the ligand relative to the receptor, in molecular dynamics applications, the indicator function is often approximated by a flat-bottom restraint~\cite{chen2007can} (see below).
+Moreover, the indicator function prescribes that the ligand remains in the binding site region throughout the alchemical process, including when it is uncoupled from the receptor.
+Consequently, alchemical binding free energy simulations where the ligand leaves the binding site are not set up correctly and return undefined free energy estimates.
+Because it is a discontinuous function of the position and orientation of the ligand relative to the receptor, molecular dynamics implementations often approximate the indicator function by means of flat-bottom restraints~\cite{chen2007can} (see below).
 Harmonic restraints are not as suitable in this respect because they favor a specific microstate at the bottom of the harmonic potential rather than specifying a region of conformational space.
@@
 Hence, they are typically turned on during an initial stage, starting from the uncoupled state, and are turned off in the final stage, near the bound state.
-In the alchemical binding free energy literature, often harmonic restraint potentials imposed in an initial stage and removed in a final stage are described as implementing standard state terms.
-However, because they utilize harmonic potentials that cannot correctly describe the binding pose macrostate (see above) and because they are turned on and then off, these are actually auxiliary restraints introduced to enhance convergence without affecting, in principle, binding free energy estimates.
+In the alchemical binding free energy literature, often harmonic restraint potentials imposed in an initial stage and removed in a final stage are described as implementing binding sites restraints.
+However, because they utilize harmonic potentials that cannot correctly describe the binding pose macrostate (see above) and because they are turned on and then off, these are instead auxiliary restraints introduced to enhance convergence without affecting, in principle, binding free energy estimates.
 
@@
 
-Recently, more restrictive restraint schemes have been proposed that avoid relying on a few critical reference atoms.
-Instead, these aim to simplify restraint selection by using larger sets of atomic positions in the receptor and ligand.
+Recently, degrees of freedom for binding site and auxiliary potentials based on larger sets of atomic positions of the receptor and ligand have been put forward.
 Fu et al.
@@
 Often, binding site restraints do not limit the rotational degrees of freedom of the ligand, and the corresponding standard state term is zero; see Eq.~(\ref{eq:DG0-def-Vsite}).
+However, it might be challenging to fully explore the bound macrostate at all alchemical states when it is only limited by receptor-ligand positional restraints.
+As further discussed below, auxiliary restraints are particularly helpful in the latter circumstance to limit the extent of conformational sampling necessary to achieve rapid convergence of binding free energy estimates.
 
@@
 For example, the overall ligand RMSD has been used~\cite{woo2005calculation} for this purpose.
-Some auxiliary restraints are designed to reduce the amount of conformational sampling at intermediate $\vec{\lambda}$ values.
+Auxiliary restraints are designed to reduce the amount of conformational sampling at intermediate $\vec{\lambda}$ values.
 However, more computational effort is required to compute the free energy when the restraints are turned on and off.
@@
 \subsubsection{Handling weak binders and high dissociation rates}\label{sec:weak-binders}
+\egnote{Is this section necessary?
+We could add a comment regarding weak binders in the preceding section clarifying how the theory covers weak binders (as well as non-binders) and the precautions the user should take to obtain accurate estimates (relative to the experiments).}
 In binding free energy calculations, only the conformations in which the receptor and ligand form a bound complex should be sampled from the bound states (Sec.~\ref{sec:theory}).
@@
 
-In absolute calculations, this need to explicitly define a binding site applies to the fully interacting state in the complex leg of the thermodynamic cycle (top-right state in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}), while in relative calculations the binding site must be defined at both end states of the complex leg (top- and bottom-right states in Fig.~\ref{fig:fig_binding_thermodynamic_cycle}).
+In absolute calculations, the need to explicitly define a binding site applies to the fully interacting state in the complex leg of the thermodynamic cycle (top-right state in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}),  while in relative calculations the binding site must be defined at both end states of the complex leg (top- and bottom-right states in Fig.~\ref{fig:fig_binding_thermodynamic_cycle}).
+\egnote{I do not agree with the latter.
+Binding site restraints apply to the coupled and uncoupled states.
+The same is true for RBFE for both ligands.
+(see above)}
 In principle, this requires defining which conformations are considered to be bound before running the calculation, but it is common practice to start the simulation with the ligand already placed in the binding site and rely on kinetic trapping to maintain the bound complex.
@@
 This is typical of weak binders such as fragments binding shallow pockets with $\mu$M-mM affinities~\cite{georgiou2017pushing,pan2017quantitative}.
-In the case of weak binders, using a flat-bottom or harmonic restraint between receptor and ligand in the bound state(s) can prevent dissociations~\cite{georgiou2017pushing,rizzi2019sampl6}.
+In the case of weak binders, \egnote{I would say for all ligands: strong, weak, as well as non-binders.} using a flat-bottom or harmonic restraint \egnote{Harmonic restraints are not suitable as binding site restraints, see above} between receptor and ligand in the bound state(s) can prevent dissociations~\cite{georgiou2017pushing,rizzi2019sampl6}.
 We stress that this type of restraint is normally avoided as it generally introduces bias in the free energy estimate, which is why the restraint is usually activated only in the intermediate states in absolute calculations.
-The bias can be corrected through reweighting schemes~\cite{rizzi2019sampl6}, but this post-processing step can be avoided if a flat-bottom restraint is used and the ligand is never sampled while hitting the potential wall during the simulation in the bound state as the numerical correction will be exactly zero.
+\egnote{The latter statement is incorrect.
+Binding site restraints are always necessary in principle.
+}
+The bias can be corrected through reweighting schemes~\cite{rizzi2019sampl6}, \egnote{Auxiliary restraints and biasing potentials require reweighting.
+But binding site restraints do not generally require any action in post-processing except for adding the $C^\circ V_{\rm site}$ term.
+} but this post-processing step can be avoided if a flat-bottom restraint is used and the ligand is never sampled while hitting the potential wall during the simulation in the bound state as the numerical correction will be exactly zero.
+\egnote{The ligand frequently hitting the boundary of the flat-bottom restraints while in the coupled state is indicative of an inappropriate definition of the binding site: the ligand is attempting to access a lower free energy binding pose.
+The free energy estimate is unlikely to agree with experiments but the calculation is otherwise valid.
+It reports the binding free energy in the binding pose specified by the region allowed by the flat-bottom binding site restraint.}
 It is important to note that the spring constant and/or radius parameters of the restraint effectively determine which conformations are considered to be bound.
 As a consequence, these parameters must be tuned to the system so that only the binding site is accessible to the ligand.
-Again, this step is particularly important for weak binders as their free energy of binding is known to be more sensitive to the definition of the binding site~\cite{gilson1997statisticalthermodynamic}.
+Again, this step is particularly important for weak binders \egnote{but necessary, in principle, for all binders} as their free energy of binding is known to be more sensitive to the definition of the binding site~\cite{gilson1997statisticalthermodynamic}.
 
@@
 In the latter case, dissociations can be averted simply by increasing the spring constant and/or reducing the radius of the restraint potential to prevent the exploration of ligand conformations outside the binding site in the decoupled state (bottom-right state in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}) that could be propagated to the bound state.
+\egnote{I disagree.
+Binding site restraints are never turned off, especially at the uncoupled state, which would result in the ligand wandering away from the receptor.
+Additional restraints can be applied to reduce the extent of conformational sampling in the intermediate alchemical steps.
+Conversely, energy barrier-flattening potentials can be applied to accelerate conformational sampling.
+These auxiliary potentials must be turned on and then turned off so that they are not active at the endstates.
+Alternatively, thermodynamic reweighting must be applied in post-processing to remove their effects on the binding free energy estimate.
+}
 

### 2025-06-08 paper/manuscript.tex

@@
 The first important aspect is that binding free energy calculations must be performed with binding site restraints that implement the binding site indicator function $I(\zeta)$ by forcing the ligand to occupy the binding site even when it is decoupled from the receptor.
-Doing otherwise would fail to specify the bound macrostate of the complex for which we seek the free energy.
-Moreover, the indicator function prescribes that the ligand remains in the binding site region throughout the alchemical process, including when it is uncoupled from the receptor.
-Consequently, alchemical binding free energy simulations where the ligand leaves the binding site are not set up correctly and return undefined free energy estimates.
+Doing otherwise would fail to specify the bound macrostate of the complex and return undefined binding free energy estimates.
+However, in practice, calculations lacking binding site restraints often return reliable binding free energy estimates corresponding to the limited sampled binding site region explored during molecular dynamics.
+Nevertheless, longer calculations could display persistent drift of the estimate as new regions of conformational space are sampled.
+A more serious consequence of neglecting binding site restraints is the possibility of the ligand leaving the binding site when it is not coupled to the receptor.
+Such occurrence indicates a serious violation of the theory.
+Alchemical binding free energy simulations where the ligand leaves the binding site are not set up correctly and can return grossly inaccurate binding free energy estimates.
+
 Because it is a discontinuous function of the position and orientation of the ligand relative to the receptor, molecular dynamics implementations often approximate the indicator function by means of flat-bottom restraints~\cite{chen2007can} (see below).

### 2025-06-09 paper/manuscript.tex
Add references per discussion

--- 2025-06-06 paper/manuscript.tex
@@
 These systems require more extensive equilibration than soluble proteins.
-Some packages, such as Desmond, include a 7-step protocol to relax membrane systems.
+Some packages, such as CHARMM or Desmond, recommend a multi-protocol to relax membrane systems \cite{https://doi.org/10.1002/jcc.23702,DesmondManual}.
 
@@
 
-In the case of a lipid-facing binding site, there are two additional challenges.
+In the case of a lipid-facing binding site, there are three additional challenges.
 First, simulation time should be extended to ensure sufficient sampling of lipid configurational space.
 Lipids may either bind alongside small molecules, acting as cofactors, or be displaced from the binding site by a ligand.
-If lipids participate in binding through protein-lipid or ligand-lipid interactions, it is necessary to sample these intermolecular interactions through additional $\lambda$ windows, longer simulation times, or different alchemical protocols.
+If lipids participate in binding through protein-lipid or igand-lipid interactions, it is necessary to sample these intermolecular interactions through additional $\lambda$ windows, longer simulation times, or different alchemical protocols.
 Displacing a lipid will have its own free energy cost which may take a long time to converge if it requires diffusion and rearrangement of nearby lipids.
-Second, the free energy of binding for a small molecule may be strongly coupled to the lipid species.
+Second, it may sometimes be necessary to consider the free energy cost of a ligand partitioning into the membrane \cite{doi:10.1021/acs.jcim.1c01147}.
+Third, the free energy of binding for a small molecule may be strongly coupled to the lipid species.
 In some cases, the identity of the lipid may be known.

### 2025-06-10 paper/manuscript.tex

--- 2025-06-08 paper/manuscript.tex
@@
   }
-
-% aux. comments Emilio Gallicchio
-\newcommand{\egnote}[1]{%
-  {\bfseries{}[EG: }%
-  {\textcolor{green}{#1}}{\bfseries{}]}
-}
-
 \newcommand{\rref}{{\bfseries[REFs]}{}}
@@
 \paragraph{Atom mapping}
-Once a particular approach to the topology is selected, a crucial next step is to identify the common atoms which will not be perturbed.
-Rigorously, this process typically comprises a MCSS search of the molecules involved to identify the common substructure---though the parameters of the MCSS search will differ depending on whether single or dual topology calculations are planned.
-Specifically, with a single topology approach in mind, atom types are allowed to change, so a permissive MCSS search can be done, whereas with dual topology a more strict search is required.
+Following the choice of system representation and sampling strategy, it might be required to generate a atom mapping.
+This is the case especially for system representations using a single coordinate set (Type 1.
+& 2), however a dual coordinate set approach could use an atom mapping as well, depending on the sampling method, this will be explained in the next steps.
+Once a single coordinate set approach is selected is selected, a crucial next step is to identify the common atoms which will not be perturbed between the two molecules (respectivly end-states).
+Rigorously, this process typically comprises a MCSS search of the molecules involved to identify the common substructure.
+Specifically, with a single coordinate set approach in mind, atom types are allowed to change, so a permissive MCSS search can be done.
+This however requires either careful implementation of atom mass changes, bond length etc.
+or accepting the introduction of potential systematic error, that is in many cases negliable due to the cancelation of errors.
+\cite{ries2024Kartograf}
+
 \begin{figure}
@@
 
-There are different tools that allow the generation of MCSS matches as well as single topology input.
-A large number of software tools can compute MCSS matches using different cheminformatics packages.
-Some rely on RDKit~\cite{rdkit2019Dec}, such as pmx~\cite{gapsys2015pmx}, LOMAP~\cite{liu2013lead}, FESetup~\cite{loeffler2015fesetup} and partially BioSimSpace~\cite{hedges2019biosimspace}, while others such as fkckombu~\cite{kawabata20143d} are standalone tools.
+In general there are mutliple ways on how a MCSS can be calculated.
+One of the most prominent approaches is to estimate the MCSS with a 2D subgraph isomorphism solver(SIS) using the 2D topological stuructre information of molecules\cite{raymond2002maximum}.
+A large number of software tools can compute SIS MCSS matches using different cheminformatics packages.
+Some rely on RDKit~\cite{rdkit2019Dec}, such as pmx~\cite{gapsys2015pmx}, LOMAP~\cite{liu2013lead}, FESetup~\cite{loeffler2015fesetup}, ProtoCaller~\cite{suruzhon2020protocaller}, SMArt~\cite{petrov2021perturbation}, and partially BioSimSpace~\cite{hedges2019biosimspace}, while others such as fkckombu~\cite{kawabata20143d} are standalone tools.
 Schr\"{o}dinger's FEP+ planning tool was originally based on a version of LOMAP, and it also uses MCSS matching as well as 3D considerations to plan the network of single topology calculations between molecules~\cite{wang2015accurate}.
 
-MCSS searches can be relatively time consuming, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial similarity assessment and then use MCSS only to identify final mappings for relative calculations~\cite{raymond2002maximum,klabunde2012mars,jones2009elucidating}.
-The MCSS approach, though relatively standard, takes into account only topological similarity.
+SIS based methods, showed to struggle with 3D properties of the molecules, that need to be mapped, like for example stereochemistry.
+Therefore some packages developed a post-mapping step, where they filtered for atom distances, and therefore added 3D information into the mapping.
+Still SIS MCSS searches can be relatively time consuming as the algorithmic complexity is classified as NP-complete, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial similarity assessment and then use MCSS only to identify final mappings for relative calculations~\cite{raymond2002maximum,klabunde2012mars,jones2009elucidating, Duesbury2018Comparison}.
+The SIS MCSS approach, though relatively standard, takes into account only topological similarity.
 It is possible that changes in binding mode could actually require a different choice of mapping, so in some cases mappings may need to be planned differently depending on 3D positioning of atoms in space.
-Visual inspection prior to simulation is recommended to ensure that the mapping criteria correspond to the expected binding mode.
+
+Here an alternative way to estimate the MCSS is to calculate a bipartite graph matching based on the 3D positions of two molecules.
+These approaches tend to be very fast (can have algorithmic complexity of $O(n)=N^3$) and geometric accurate in their mapping (e.g.
+stereochemistry or binding mode differing).
+The geometric approaches additionally allow mappings of differing binding modes of the same molecule, or can be even used for significant larger molecules like proteins, due to the speed increase.
+A disadvantage of these approaches is the strong dependency on the coordinate sets being well aligned.
+Another aspect is taking the topology changes of the molecules not into account, which can be done if necessary in a next step with a filtering approach, like checking for changes in ring hybridization or ring breaking (this problem is discusse below)|.
+Tools that can be used to estimate the MCSS this way are contained for example in pmx \cite{gapsys2015pmx} or Kartograf \cite{ries2024kartograf}.
+
+In any case Visual inspection prior to simulation is recommended to ensure that the mapping criteria correspond to the expected binding mode.
 If the mapping protocol returns simulations that correspond to different binding modes of a ligand within a perturbation map, this can cause large hysteresis.
 
-Single topology relative calculations and calculations based on substructure searches only work if in fact the ligands share a common substructure, e.g.
+A next step for atom mappings, would be to support multistate methods (e.g.
+RE-EDS~\cite{sidler2016replica}, A-EDS~\cite{perthold2020toward} or Ladybugs~\cite{robo2022achieving}) with efficient mapping algorithms.
+There are protoype starting point for solving this problem, but to our knowledge, there is only few published work on this aspect.
+\cite{petrov2021Perturbation}
+
+Single coordinate set relative calculations and calculations based on substructure searches only work if in fact the ligands share a common substructure, e.g.
 are part of a congeneric series, see Fig.~\ref{fig:fig_mcss}.
-If no common substructure is shared, then alternative dual or separated topology free energy calculations are needed.
+If no common substructure is shared, then dual coordinate sets methods like linked-dual or separated topology free energy calculations are needed.
 One would co-localize a pair of compounds in a binding site, exclude their interactions with one another, and compute the relative binding free energy by turning one molecule on from being dummy atoms while turning the other off.
-To our knowledge no general pipeline for such calculations yet exists and this would likely remain a research problem.
-Using an absolute free energy approach instead seems more promising in such a case.
+For such approaches it is crucial to find a good set of restraints, that keep the ligands colocalized over the time of the simulation .
+This problem is similar to the required restraints in ABFE Calculations (see \ref{ABFErest}).
+Few descriptions of possible approaches in RBFE calculations can be found for such approaches like the one suggested by \textit{Baumann et al.}~\cite{baumann2023broadening} based on boresch style restraints~\cite{boresch2003absolute}, with an elaborate automatic atom selection approach, for the seperated topology approach, or Restraintmaker~\cite{ries2022restraintmaker} for the linked-dualtopology approaches, which finds a small set of atoms to be linked via distance restraints, with a small distance to each other.
+However the implementation and testing of an fully automatized dual-topology pipeline for RBFE dual topology approaches is still to our knowledge a remaining research problem and was only tested with small datasets.
+\cite{jespers2019qligfep, Rieder2022Lev}
 
@@
 \paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
-Some approaches have attempted to address this~\cite{clark2019relative} but a general solution is not yet in mainstream use. Still, FEP+ implements one solution.
+Some approaches have attempted to address this using a single coordinate sets and with soft bond approaches ~\cite{clark2019relative,wang2017accurate} or dual coordinate set approach \cite{jespers2019qligfep, ries2022relative} but a general solution is not yet in mainstream use.
+Still, FEP+ implements one solution\cite{wang2017accurate}.
 
@@
 
-An absolute binding free energy calculation yields the standard free energy of formation of the molecular complex from the separated receptor and ligand species.
-Hence, as discussed in \ref{sec:non-covalent-molecular-binding}, it must depend on the standard state concentration $C^\circ$ (here $1 \, {\rm M} = 1 \, {\rm molecule}/1660$~\r{A}$^3$) and on the definition of the complex, that is on the specific binding pose for which the binding free energy is requested.
-Omitting the standard state concentration would yield an undefined result that would vary with the choice of concentration units.
-A similar vague estimate would result from omitting the binding pose definition that, for example, could equally describe binding to a specific site of the receptor or unspecific binding throughout the receptor surface depending on the extent of conformational sampling (see Sec.~\ref{sec:weak-binders}).
-
-In the statistical mechanics formulation of Gilson et al.~\cite{gilson1997statisticalthermodynamic}, described in Section \ref{sec:non-covalent-molecular-binding}, the standard state concentration appears explicitly in Eq.~(\ref{eq:DG0-binding-constant-def}) and the binding pose is defined by the indicator function $I(\zeta)$ that characterizes the partition function of the complex; Eq.~(\ref{eq:intra-zRL-def}).
-The role of the standard state and binding site definitions are particularly evident when expressing Eq.~(\ref{eq:DG0-binding-constant-def}) in terms of an ensemble average.
-To do so, we express the integrals in the numerator and denominator of Eq.~(\ref{eq:DG0-binding-constant-def}) over the same set of degrees of freedom.
-Note that the intramolecular configurational partition function of the complex, $z_{RL(v)}$, includes six more degrees of freedom--the position and orientation of the ligand relative to the receptor collectively described by the variable $\zeta$--than the product of the intramolecular partition functions, $z_{R(v)} z_{L(v)}$, of the separated receptor and ligand.
-To correct this discrepancy, we multiply and divide the right-hand side of Eq.~(\ref{eq:DG0-binding-constant-def})  by the quantity
-\begin{equation} \label{eq:Vsite-def}
-    V_{\rm site} \Omega = \int d\zeta I(\zeta) \, ,
-\end{equation}
-%\Omega is in my opinion a better symbol of the orientational extent than \xi_L. Lowercase greek letters are usually used
-%for dynamical variables rather than integrated quantities
-which measures the extent of the binding pose macrostate, where $V_{\rm site}$ corresponds to the spatial extent and $\Omega$ the orientational extent.
-The right-hand side of Eq.~(\ref{eq:Vsite-def}) combined with the integrals $z_{R(v)}$ and $z_{L(v)}$ yields a partition function similar to Eq.~(\ref{eq:intra-zRL-def}) but corresponding to a complex in which ligand and receptor are uncoupled~\cite{gallicchio2011recent} as if they were at infinite distance apart.
-The result is
-\begin{equation}\label{eq:DG0-binding-constant-def-average}
-K_b = C^\circ V_{\rm site} \frac{\Omega}{8 \pi^2} \langle e^{-\beta \Delta \Psi}  \rangle_0 
-\end{equation}
-where, similarly to Eq.~(\ref{eq:zratio-solv-average}),
-\begin{equation}\label{eq:pert-energy-def}
-\Delta \Psi(x_R, x_L, \zeta) =  \Psi_v(x_R, x_L, \zeta) - \left[  \Psi_v(x_R) + \Psi(x_L) \right]
-\end{equation}
-is the solvent-averaged binding energy of the complex when receptor and ligands are in intramolecular configurations $x_R$ and $x_L$, the ligand is in position and orientation $\zeta$ relative to the receptor (the uncoupled potential energy does not depend on  the position and orientation of the ligand relative to the receptor), and $\langle \ldots \rangle_0$ represents an ensemble average in the uncoupled ensemble.
-Finally, note that $\Omega = 8 \pi^2$ when the binding pose indicator function does not depend on the orientation of the ligand and Eq.~(\ref{eq:DG0-binding-constant-def-average}) simplifies to
-\begin{equation}\label{eq:DG0-binding-constant-def-average-noOmega}
-K_b = C^\circ V_{\rm site} \langle e^{-\beta \Delta \Psi}  \rangle_0 
-\end{equation}
-which, expressed in free energy units, yields
-\begin{eqnarray}\label{eq:DG0-def-Vsite}
-  \Delta G^\circ_b &=&  - k_B T \ln K_b = - k_B T \ln C^\circ V_{\rm site} - k_B T \ln \langle e^{-\beta \Delta \Psi}  \rangle_0 \nonumber \\
-{\rm \ }  &=& \Delta G^\circ_{b,{\rm ideal}} + \Delta G_{b,{\rm exc.}}
-\end{eqnarray}
-
-The term $\Delta G^\circ_{b,{\rm ideal}} = - k_B T \ln C^\circ V_{\rm site}$ in Eq.~(\ref{eq:DG0-def-Vsite}) is interpreted as the \emph{ideal} component of the standard binding free energy, that is the binding free energy that would be measured in the absence of ligand-receptor interactions ($\Delta \Psi = 0$) and when the ligand's concentration is $C^\circ$.
-The ideal term is responsible for the dependence of the standard binding free energy on the standard state concentration.
-The arguably inappropriate term ``standard state correction'' is sometimes used in the literature to refer to $\Delta G^\circ_{b,{\rm ideal}}$.
-As seen above, the ideal term is a fundamental aspect of the statistical mechanics theory of molecular binding rather than a mere correction to an inexact formulation.
-
-Conversely, the term $\Delta G_{b,{\rm exc.}} = - k_B T \ln \langle e^{-\beta \Delta \Psi}  \rangle_0 $ is the \emph{excess} component of the standard free energy of binding, which is the term that depends on intermolecular interactions and the specific chemical composition of the system.
-The excess component is the target of the alchemical molecular simulations discussed here, in contrast to the ideal term that is generally evaluated analytically~\cite{boresch2003absolute}.
-
-\paragraph{Handling the standard state in free energy calculations.}
-As discussed in Sections \ref{sec:theory-solvation} and \ref{sec:theory-partitioning}, the handling of the standard state in solvation and solvent partitioning free energy calculations is typically straightforward.
-Eq.~(\ref{eq:DG0-solvation-def}) identifies the ideal and excess components of the standard solvation free energy in terms of the standard gas pressure $P^\circ$, standard solution concentration $C^\circ$, and ensemble averages typically evaluated by alchemical free energy simulations.
-These relations provide simple recipes to relate computational estimates to experimental values obtained under different conditions.
-Moreover, typical experimental reporters (Henry's constants and partitioning coefficients) directly probe the excess term calculated by molecular simulations, which is independent of the standard state.
-
-As discussed above, the situation for binding is more complex and requires special care.
-The first important aspect is that binding free energy calculations must be performed with binding site restraints that implement the binding site indicator function $I(\zeta)$ by forcing the ligand to occupy the binding site even when it is decoupled from the receptor.
-Doing otherwise would fail to specify the bound macrostate of the complex for which we seek the free energy.
-Moreover, the indicator function prescribes that the ligand remains in the binding site region throughout the alchemical process, including when it is uncoupled from the receptor.
-Consequently, alchemical binding free energy simulations where the ligand leaves the binding site are not set up correctly and return undefined free energy estimates.
-Because it is a discontinuous function of the position and orientation of the ligand relative to the receptor, molecular dynamics implementations often approximate the indicator function by means of flat-bottom restraints~\cite{chen2007can} (see below).
-Harmonic restraints are not as suitable in this respect because they favor a specific microstate at the bottom of the harmonic potential rather than specifying a region of conformational space.
-
-Note that $\Delta G^\circ_{b,{\rm ideal}}$ is independent of the chemical nature of the binding partners, and it is the same for all complexes sharing similar binding site definitions.
-The latter is the reason that the ideal term cancels out when taking the difference of absolute binding free energies when estimating relative binding free energies (RBFEs).
-However, while it does not appear explicitly in the definition, $\Delta G_{b,{\rm exc.}}$ depends implicitly on the definition of the binding site through the ensemble average in the uncoupled ensemble, which includes only ``bound'' configurations according to the definition of the indicator function $I(\zeta)$.
-Varying the binding site definition affects both the ideal and excess binding free energy terms such that their sum (the standard binding free energy) is often only weakly dependent on the binding site definition~\cite{gilson1997statisticalthermodynamic,gallicchio2011recent}.
-Conversely, great care should be exercised when comparing calculated excess binding free energy values corresponding to different binding site definitions.
-Similarly, the outcome of a binding free energy calculation conducted without a binding site definition, i.e.\ without binding site restraints, is, in principle, undefined.
-The latter is also true for relative binding free energies because while the ideal term cancels out, the difference in excess binding free energies depends on the binding pose definition.
-
-\paragraph{Several choices of binding site restraints are possible.}
-A variety of restraint types are commonly used in alchemical binding free energy calculations.
-Before going into their characteristics, we must acknowledge an important and often unrecognized distinction between the binding site restraints corresponding to the indicator function $I(\zeta)$ necessary for the implementation of the theory and optional auxiliary restraints and biasing potentials sometimes used to speed up the convergence of free energy calculations.
-Binding site restraints are an integral part of the molecular complex definition and are typically maintained throughout the alchemical process.
-In particular, they must be present unchanged at the unbound and bound states.
-In contrast, unless thermodynamic reweighting is applied in post-processing, auxiliary restraints cannot be present at the endstates to avoid affecting their free energies.
-Hence, they are typically turned on during an initial stage, starting from the uncoupled state, and are turned off in the final stage, near the bound state.
-In the alchemical binding free energy literature, often harmonic restraint potentials imposed in an initial stage and removed in a final stage are described as implementing binding sites restraints.
-However, because they utilize harmonic potentials that cannot correctly describe the binding pose macrostate (see above) and because they are turned on and then off, these are instead auxiliary restraints introduced to enhance convergence without affecting, in principle, binding free energy estimates.
-
-Binding site restraints implement the $I(\zeta)$ indicator function.
-They, therefore, can depend at most on the six rigid-body external coordinates of the ligand governing the position and orientation of the ligand relative to the receptor.
-Any other choice would affect the intramolecular conformational distribution of the ligand or the receptor and would  bias the binding free energy estimate.
-The external degrees of freedom introduced by Boresch et al.~\cite{boresch2003absolute,leitgeb2005alchemical} are a popular choice.
-The Boresch coordinates, as well as other similar approaches, involve selecting anchor atoms of the receptor and the ligand.
-In principle, the choice of anchor atoms is arbitrary, provided adequate simulation time is available, and as long as it leads to similar binding pose definitions.
-However, practical considerations may be important.
-For example, some choices of Boresch-style anchor atoms can be numerically unstable.
-Additionally, anchor atoms should likely be placed in a part of the molecule that defines the binding orientation well rather than in a floppy, solvent-exposed tail.
+Absolute free energy calculations involve completely removing the interactions between the ligand or solute and its environment, taking it to a non-interacting state that may or may not retain intramolecular non-bonded interactions.
+This non-interacting state can then be shifted between environments---from the protein to water, or from one solution to another---without changing its free energy other than that due to the changing volume of the simulations, and then interactions can be restored in the new environment.
+
+Absolute free energies are by definition reported with respect to a specific reference or standard state, which effectively determines the arbitrary point at which the free energy is 0.
+The role of the standard state is particularly evident from the expression of the binding free energy between a receptor $R$ and ligand $L$
+\begin{equation} \label{eq:DGfromKAB}
+    \Delta G = -k_BT ~ \ln \left( c^{\circ} K_b \right)  = -k_BT ~ \ln\left( c^{\circ} \frac{[RL]}{[L][R]} \right) \, .
+\end{equation}
+Here, the reference state concentration $c^{\circ}$ converts the binding constant $K_b$ into a dimensionless quantity expressed in reference concentration units.
+It should be noted that ignoring the term $c^{\circ}$ is equivalent to assuming a reference concentration of 1~D$^{-1}$, where D are the units used to express $K_b$, and would thus cause the value of $\Delta G$ to vary with the choice of the units.
+It is convenient to define a standard state at a constant pressure of 1~atm and where each chemical species (i.e., R, L, and RL) in the reaction solvent has a concentration of $c^{\circ}$~=~1~M~=~1~molecule/1660~\r{A}$^3$ but do not interact with other molecules of $R$, $L$, or $RL$.
+
+\paragraph{Handling the standard state in absolute free energy calculations.}
+For solvation free energy calculations, handling the standard state is typically straightforward, and treating it correctly simply means ensuring that the non-interacting solute still occupies essentially the same volume as the solute in the interacting system.
+So typically in such cases no special care is required to ensure the correct standard state, as long as the \emph{experimental} data being analyzed uses the same standard state.
+If this is not the case, a simple entropic correction to the free energy of $k_B T\ln (V_f/V_i) = k_B T \ln (C_i/C_f)$ to the experimental data is needed.
+
+For binding, however, the situation is more complex and requires special care.
+Because the simulations are typically performed using restraints and at concentrations that are different from 1 M, the expression of the free energy requires the following correction~\cite{gilson1997statisticalthermodynamic} (see an example of such a thermodynamic cycle in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle})
+\begin{equation}\label{eq:restraint-correction}
+    \Delta G^{\circ}_{\mathrm{restr}} = -k_BT ~ \ln \left( c^{\circ}{V_L} \right) -k_BT ~ \ln \left( \frac{\xi_L}{8 \pi^2} \right),
+\end{equation}
+where $V_L$ and $\xi_L$ are respectively the volume of the translational and rotational degrees of freedom of the non-interacting ligand in the simulation box.
+When no restraints are used, the non-interacting ligand is free to translate and rotate in the simulation box (i.e., $V_L = V_{\mathrm{box}}$ and $\xi_L = 8\pi^2$), and the rotational term is zero.
+A sufficiently thorough exploration of the simulation box by the non-interacting ligand is, however, required for the formula to be valid.
+This is typically hard to achieve as the exploration process is governed by diffusion, and weak transient nonspecific binding will occur at other sites on the protein. 
+The addition of a restraint limits the volume available to the non-interacting ligand, thus speeding the convergence of the sampling.
+In addition, when enhanced sampling methods such as Hamiltonian $\lambda$ exchange are used (see Sec.~\ref{sec:sampling_schemes}), the use of a restraint is typically necessary as it keeps the ligand in the binding site in the interacting state (see also Sec.~\ref{sec:weak-binders}) and generally reduce the round-trip time of replicas.
+When restraints are employed, the values of $V_L$ and $\xi_L$ are restraint-dependent, but for commonly employed restraints, these can be usually easily computed analytically or numerically by solving the relevant integral.
+
+\paragraph{Several choices of restraints are possible.}\label{ABFErest}
+In practice, a variety of types of restraints are common, from simple harmonic distance restraints between the ligand and the protein~\cite{mobley2006use}, to flat-bottom restraints which work similarly but only exert a force if the ligand leaves a specific region~\cite{chen2007can}.
+Because these restraints do not limit the rotational degrees of freedom of the ligand, the rotational term entering the correction in Eq.~\ref{eq:restraint-correction} is zero.
+
+Alternatively, a set of restraints proposed by Boresch have also commonly been employed, where all six rigid-body degrees of freedom governing the orientation of the ligand relative to the receptor are restrained~\cite{boresch2003absolute, leitgeb2005alchemical}.
+Further restraints, such as on the overall ligand RMSD have also been used~\cite{woo2005calculation}.
+
+In principle, all of these forms will yield correct binding free energies in the limit of adequate sampling if their effects and connection to the standard state are correctly handled, but they have different strengths and weaknesses.
+For example, with more involved restraints, sampling at intermediate $\vec{\lambda}$ values will usually not need to be as extensive but more computational effort must go to computing the free energy to turn on the restraints.
+Additionally, such restraints would typically keep the ligand from exploring alternative binding modes.
+This  restriction may be undesirable when using Hamiltonian $\lambda$ exchange or expanded ensemble techniques where allowing the ligand to exchange binding modes when it is non-interacting could provide sampling benefits~\cite{wang2013identifying}.
+More specifically, flat-bottom restraints might allow a ligand to explore multiple binding sites, and harmonic restraints allow exploration of multiple binding modes within a site, while Boresch restraints only allow a single binding mode within a single site.
+See additional discussion of the possibility of multiple binding modes in Sec.~\ref{sec:multiple_binding_modes} below.
+
+Many choices of restraints involve selecting reference atoms.
+Again, in principle this choice is unimportant given adequate simulation time but practical considerations may be important.
+The choice is likely especially important with Boresch-style restraints, where some relative placements of reference atoms are likely to be numerically unstable; additionally, ligand reference atoms should likely be in a part of the molecule which defines the binding orientation well, rather than in a floppy solvent-exposed tail, for example.
 Several automated methods have been proposed to select anchor points for Boresch restraints, which typically involve analysing a simulation of the fully-interacting receptor-ligand complex~\cite{alibay2022evaluating, alibay2021mdrestraintsgenerator, baumann2023broadening, chen2023enhancing, hedges2023suite}.
 
-Recently, degrees of freedom for binding site and auxiliary potentials based on larger sets of atomic positions of the receptor and ligand have been put forward.
+More recently, restrictive restraint schemes have been proposed that avoid relying on a few critical reference atoms.
+Instead, these aim to simplify restraint selection by using larger sets of atomic positions in the receptor and ligand.
 Fu et al.
-suggested restraining the six relative rigid-body degrees of freedom derived by finding the rotation of the ligand, which minimises the RMSD to a reference structure of the receptor-ligand complex (after accounting for rotation and translation of the receptor)~\cite{fu2017new}.
+suggested restraining the six relative rigid-body degrees of freedom derived by finding the rotation of the ligand which minimises RMSD to a reference structure of the receptor-ligand complex (after accounting for rotation and translation of the receptor)~\cite{fu2017new}.
 Salari et al.
 proposed restraining the ``distance-from-bound-configuration'' (DBC) variable, which is the RMSD of ligand coordinates in the frame of reference of the binding site~\cite{salari2018streamlined, ebrahimi2022symmetry}.
-DBC may allow the binding pose to be closely preserved as the ligand intermolecular interactions are removed, but has the disadvantage of coupling the internal and relative external degrees of freedom of the receptor and ligand, making them unsuitable for describing the binding site macrostate and the corresponding standard state terms.
-However, these restraints are practical as auxiliary restraints where additional steps are used to introduce and release the DBC restraints.
-These restraint schemes are implemented in open-source workflows~\cite{fu2021bfee2, fu2022accurate, santiago-mcrae2023computing}.
-
-As discussed above, binding site restraints are commonly implemented using quadratic flat-bottom restraints, which impose a restoring harmonic  force similar to harmonic potentials but only if the ligand leaves a specific region~\cite{chen2007can}.
-Often, binding site restraints do not limit the rotational degrees of freedom of the ligand, and the corresponding standard state term is zero; see Eq.~(\ref{eq:DG0-def-Vsite}).
-However, it might be challenging to fully explore the bound macrostate at all alchemical states when it is only limited by receptor-ligand positional restraints.
-As further discussed below, auxiliary restraints are particularly helpful in the latter circumstance to limit the extent of conformational sampling necessary to achieve rapid convergence of binding free energy estimates.
-
-Auxiliary restraints take a variety of forms, such as simple harmonic distance restraints between the ligand and the protein~\cite{mobley2006use} as well as flat-bottom restraints.
-The choice of degrees of freedom and restraining potentials for auxiliary restraints is not as restrictive as for those for binding site restraints.
-For example, the overall ligand RMSD has been used~\cite{woo2005calculation} for this purpose.
-Auxiliary restraints are designed to reduce the amount of conformational sampling at intermediate $\vec{\lambda}$ values.
-However, more computational effort is required to compute the free energy when the restraints are turned on and off.
-Additionally, such restraints can limit the exploration of alternative binding modes.
-This  restriction may be undesirable when using Hamiltonian $\lambda$ exchange or expanded ensemble techniques where allowing the ligand to exchange binding modes when it is non-interacting could provide sampling benefits~\cite{wang2013identifying}.
-More specifically, flat-bottom restraints allow a ligand to explore multiple binding sites, and harmonic restraints allow the exploration of multiple binding modes within a site, while Boresch restraints only allow a single binding mode within a single site.
-See additional discussion of the possibility of multiple binding modes in Sec.~\ref{sec:multiple_binding_modes} below.
+This may allow the binding pose to be closely preserved as the ligand intermolecular interactions are removed, but has the disadvantage of coupling the internal and relative external degrees of freedom of the receptor and ligand, preventing easy computation of the standard state correction.
+This necessitates an additional step where the DBC restraints are released to a single harmonic restraint.
+Both restraint schemes are implemented in open-source workflows~\cite{fu2021bfee2, fu2022accurate, santiago-mcrae2023computing}.
 
@@
 \subsubsection{Handling weak binders and high dissociation rates}\label{sec:weak-binders}
-\egnote{Is this section necessary?
-We could add a comment regarding weak binders in the preceding section clarifying how the theory covers weak binders (as well as non-binders) and the precautions the user should take to obtain accurate estimates (relative to the experiments).}
 In binding free energy calculations, only the conformations in which the receptor and ligand form a bound complex should be sampled from the bound states (Sec.~\ref{sec:theory}).
@@
 
-In absolute calculations, the need to explicitly define a binding site applies to the fully interacting state in the complex leg of the thermodynamic cycle (top-right state in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}),  while in relative calculations the binding site must be defined at both end states of the complex leg (top- and bottom-right states in Fig.~\ref{fig:fig_binding_thermodynamic_cycle}).
-\egnote{I do not agree with the latter.
-Binding site restraints apply to the coupled and uncoupled states.
-The same is true for RBFE for both ligands.
-(see above)}
+In absolute calculations, this need to explicitly define a binding site applies to the fully interacting state in the complex leg of the thermodynamic cycle (top-right state in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}), while in relative calculations the binding site must be defined at both end states of the complex leg (top- and bottom-right states in Fig.~\ref{fig:fig_binding_thermodynamic_cycle}).
 In principle, this requires defining which conformations are considered to be bound before running the calculation, but it is common practice to start the simulation with the ligand already placed in the binding site and rely on kinetic trapping to maintain the bound complex.
@@
 This is typical of weak binders such as fragments binding shallow pockets with $\mu$M-mM affinities~\cite{georgiou2017pushing,pan2017quantitative}.
-In the case of weak binders, \egnote{I would say for all ligands: strong, weak, as well as non-binders.} using a flat-bottom or harmonic restraint \egnote{Harmonic restraints are not suitable as binding site restraints, see above} between receptor and ligand in the bound state(s) can prevent dissociations~\cite{georgiou2017pushing,rizzi2019sampl6}.
+In the case of weak binders, using a flat-bottom or harmonic restraint between receptor and ligand in the bound state(s) can prevent dissociations~\cite{georgiou2017pushing,rizzi2019sampl6}.
 We stress that this type of restraint is normally avoided as it generally introduces bias in the free energy estimate, which is why the restraint is usually activated only in the intermediate states in absolute calculations.
-\egnote{The latter statement is incorrect.
-Binding site restraints are always necessary in principle.
-}
-The bias can be corrected through reweighting schemes~\cite{rizzi2019sampl6}, \egnote{Auxiliary restraints and biasing potentials require reweighting.
-But binding site restraints do not generally require any action in post-processing except for adding the $C^\circ V_{\rm site}$ term.
-} but this post-processing step can be avoided if a flat-bottom restraint is used and the ligand is never sampled while hitting the potential wall during the simulation in the bound state as the numerical correction will be exactly zero.
-\egnote{The ligand frequently hitting the boundary of the flat-bottom restraints while in the coupled state is indicative of an inappropriate definition of the binding site: the ligand is attempting to access a lower free energy binding pose.
-The free energy estimate is unlikely to agree with experiments but the calculation is otherwise valid.
-It reports the binding free energy in the binding pose specified by the region allowed by the flat-bottom binding site restraint.}
+The bias can be corrected through reweighting schemes~\cite{rizzi2019sampl6}, but this post-processing step can be avoided if a flat-bottom restraint is used and the ligand is never sampled while hitting the potential wall during the simulation in the bound state as the numerical correction will be exactly zero.
 It is important to note that the spring constant and/or radius parameters of the restraint effectively determine which conformations are considered to be bound.
 As a consequence, these parameters must be tuned to the system so that only the binding site is accessible to the ligand.
-Again, this step is particularly important for weak binders \egnote{but necessary, in principle, for all binders} as their free energy of binding is known to be more sensitive to the definition of the binding site~\cite{gilson1997statisticalthermodynamic}.
+Again, this step is particularly important for weak binders as their free energy of binding is known to be more sensitive to the definition of the binding site~\cite{gilson1997statisticalthermodynamic}.
 
@@
 In the latter case, dissociations can be averted simply by increasing the spring constant and/or reducing the radius of the restraint potential to prevent the exploration of ligand conformations outside the binding site in the decoupled state (bottom-right state in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}) that could be propagated to the bound state.
-\egnote{I disagree.
-Binding site restraints are never turned off, especially at the uncoupled state, which would result in the ligand wandering away from the receptor.
-Additional restraints can be applied to reduce the extent of conformational sampling in the intermediate alchemical steps.
-Conversely, energy barrier-flattening potentials can be applied to accelerate conformational sampling.
-These auxiliary potentials must be turned on and then turned off so that they are not active at the endstates.
-Alternatively, thermodynamic reweighting must be applied in post-processing to remove their effects on the binding free energy estimate.
-}
 

### 2025-06-10 paper/manuscript.tex
Synchronising changes from Overleaf, including:
- Updates to the charge-changing perturbation discussion from @Gilli08, @oostenbrink, and @vgapsys, reviewed by @jmichel80
- Added discussion of the importance of conformational sampling by @wehs7661 - yet to be reviewed
- Added discussion of approaches for explicitly accounting for protonation/ tautomer changes by @egallicc, reviewed by @IAlibay
- Added discussion of non-equilibrium switching by @vgapsys - yet to be reviewed
- Added discussion of unsettled practices and current limitations by @jmichel80 - yet to be reviewed

--- 2025-06-09 paper/manuscript.tex
@@
 \author[3*]{John D. Chodera}
+\author[9]{Vytautas Gapsys}
 \author[9]{David F. Hahn}
@@
 \author[1]{Jenke Scheen}
-\author[11]{David R. Slochower}
 \author[8*]{Michael R. Shirts}
@@
 \affil[10]{Cresset, Cambridgeshire, UK}
-\affil[11]{Vertex Pharmaceuticals, San Diego, CA, USA}
 
@@
 \orcid{Jenke Scheen}{0000-0001-9781-0445}
-\orcid{David R. Slochower}{0000-0003-3928-5050}
 \orcid{Michael R. Shirts}{0000-0003-3249-1097}
@@
 \orcid{David F. Hahn}{0000-0003-2830-6880}
+\orcid{Vytautas Gapsys}{0000-0002-6761-7780}
 
@@
 
-In many cases, the quantity of interest is the change in binding affinity between a compound $A$ and a related compound $B$ (e.g., by modifying one of the drug scaffold's substituents, see (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{F})), which, by using Eq.~\ref{eq:binding-free-energy-from-bound-unbound-probability-ratio} and \ref{eq:bound-unbound-probability-ratio} is given by
+In many cases, the quantity of interest is the change in binding affinity between a compound $A$ and a related compound $B$ (e.g., by modifying one of the drug scaffold's substituents, see (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{E})), which, by using Eq.~\ref{eq:binding-free-energy-from-bound-unbound-probability-ratio} and \ref{eq:bound-unbound-probability-ratio} is given by
 \begin{equation}\label{eq:delta-delta-G-physical}
@@
 Well-known protein systems with water mediated ligand interactions are for example: HSP90 which formed part of the D3R grand challenge 2015~\cite{mey2016blinded}, A2A~\cite{brucemacdonald2018ligand}, MUP~\cite{ross2015water}, ~\cite{deflorian2020accurate}, and others~\cite{michel2009energetics}.
-
-\paragraph{Membrane proteins require additional care}
-Integral membrane proteins require embedding in a suitable model membrane to remain stable during simulations.
-Lipids are usually not included in protein force fields and, although they may be covered by generic small molecule force fields, have often been parameterized separately (e.g., AMBER's Lipid21 \cite{AMBERLipid21} and the CHARMM lipid force field \cite{CHARMMLipidFF}).
-Membrane systems can be simulated in multiple thermodynamic ensembles, but the semi-isotropic scheme $NP_{z}P_{xy}T$ is frequently employed, allowing the pressure in the bilayer plane ($xy$) to scale independently \cite{LipidBestPracticesCOMS}.
-These systems require more extensive equilibration than soluble proteins.
-Some packages, such as CHARMM or Desmond, recommend a multi-protocol to relax membrane systems \cite{https://doi.org/10.1002/jcc.23702,DesmondManual}.
-
-Alchemical transformations that involve net charge changes can provide difficulties for membrane systems; phospholipids carry functional groups with nonzero formal charge and cannot be treated as a solvent with a uniform dielectric in the same way as water.
-Several approaches are reviewed in \cite{wu2022correction-51c}.
-The authors recommend the use of a co-alchemical ion, specifically making an ion of equivalent charge appear alongside the disappearing charged molecule to maintain system neutrality.
-
-In the case of a lipid-facing binding site, there are three additional challenges.
-First, simulation time should be extended to ensure sufficient sampling of lipid configurational space.
-Lipids may either bind alongside small molecules, acting as cofactors, or be displaced from the binding site by a ligand.
-If lipids participate in binding through protein-lipid or igand-lipid interactions, it is necessary to sample these intermolecular interactions through additional $\lambda$ windows, longer simulation times, or different alchemical protocols.
-Displacing a lipid will have its own free energy cost which may take a long time to converge if it requires diffusion and rearrangement of nearby lipids.
-Second, it may sometimes be necessary to consider the free energy cost of a ligand partitioning into the membrane \cite{doi:10.1021/acs.jcim.1c01147}.
-Third, the free energy of binding for a small molecule may be strongly coupled to the lipid species.
-In some cases, the identity of the lipid may be known.
-But in other cases, it may be difficult to discern the identity of the lipid from X-ray or cryo-EM structures due to the flexible nature of lipid head groups and benchmarking is recommended to match experimental data.
 
@@
 If the net charge of the system changes as the alchemical variable changes during the calculation, this can pose major challenges.
-Specifically, finite-size effects can introduce significant charge-dependent artifacts into computed binding free energies, in part because typical schemes for long-range electrostatics (including PME and reaction field) do not handle free energy contributions from such changes effectively or as they would be handled in a hypothetical macroscopic bulk solution~\cite{lin2014overview, ohlknecht2020correcting, rocklin2013calculating}.
-
-There are two main potential solutions to avoid artifacts due to changes in net charge: avoiding changing the net charge, and correcting for the introduced artifacts.
-
+Specifically, finite-size effects can introduce significant charge-dependent artifacts into computed binding free energies, in part because typical schemes for long-range electrostatics, including particle mesh Ewald (PME) and reaction field, do not handle free energy contributions from such changes as they would be handled in a hypothetical macroscopic bulk solution.
+The outcome depends either on the size of the cutoff for reaction field methods or on the size of the simulation box when using lattice sum methods due to periodic boundary conditions~\cite{lin2014overview, ohlknecht2020correcting, rocklin2013calculating}.
+
+Lattice summation methods (e.g.
+PME) effectively neutralize the simulation box by applying an implicit homogeneously distributed background counter charge~\cite{figueirido1995finite,hummer1996free}.
+The background charge can be interpreted as a converged distribution of counterions and may be a reasonable choice for computing hydration free energies of monovalent ions (finite size effect corrections can be applied in these cases)~\cite{hummer1996free}.
+In inhomogeneous systems with different dielectrics, e.g.
+solvated protein or lipid bilayer in a water box, assuming uniform background countercharge distribution is unphysical, as the higher countercharge density should be located in the regions of higher dielectric~\cite{hub2014pme}.
+This, in turn, causes significant artifacts in sampled population densities of charged molecules, i.e.
+errors in calculated free energy differences.
+
+There are two main classes of approaches to address artefacts introduced by changes in net charge during alchemical free energy calculations: (1) avoiding changing the net charge altogether and (2) post-processing approaches to correct for artefacts using continuum electrostatics.
 Many relative free energy planning tools have been set up to avoid changing the net charge of the systems considered, including LOMAP~\cite{liu2013lead} and Schr\"{o}dinger's FEP+~\cite{wang2015accurate}.
-Absolute free energy calculations can also potentially avoid changing the charge of the system by making a charge perturbation of equal and opposite sign elsewhere in the system; for example, as a charged ligand is removed, a charged counterion of opposite sign could also be removed, or one of the same sign could be inserted.
-This is sometimes referred to as an "alchemical ion" approach for dealing with the needed charge change, and is also employed by the Yank free energy package~\cite{wang2013identifying}.
-Charge corrections have also been explored, and are potentially a viable solution to this problem~\cite{mey2018impact} where artifacts introduced by finite-size effects are corrected numerically~\cite{chen2018accurate, ohlknecht2020correcting}.
-However, application of such corrections typically remains less common than the use of a co-alchemical ion.
-A third approach has been proposed by Gapsys et al.~\cite{gapsys2015calculation} which uses a double-system/single box setup.
-
-When free energy calculations \emph{do} need to change the charge of a ligand or solute, the literature does not yet seem to indicate what approach should be preferable, so considerable care should be taken.
-We are not yet aware of a careful comparison of charge corrections versus other approaches such as decoupling an ion at the same time, so in our view the issue of proper handling of charge mutations in the context of alchemical calculations remains a research problem, and several papers give good guides for exploring the problem further~\cite{lin2014overview, ohlknecht2020correcting, rocklin2013calculating}.
+In absolute free energy calculations, it is often difficult to avoid a change in the net charge.
+One way to avoid charge changes is to ensure that, as a charged ligand is removed, a charged counterion of opposite sign is also removed, or one of the same sign is inserted.
+This is sometimes referred to as a “co-alchemical ion” approach to deal with the required charge change.
+This approach is easily supported in most free energy calculating programs, as it merely involves an additional simultaneous modification.
+There are special flavours of the co-alchemical ion approach such as proposed by Gapsys et al.~\cite{gapsys2015calculation}, which uses a double-system/single-box setup.
+Here, the two branches of a thermodynamic cycle as used in relative free energy calculations are placed in one simulation box, effectively circumventing a change in the overall net charge during a perturbation.
+Note that the co-alchemical approaches only avoid finite-size effects with lattice-sum methods such as PME.
+
+Charge-change correction schemes have been explored and offer potentially viable solutions for the problem~\cite{mey2018impact}, where artifacts introduced by finite-size effects are corrected a posteriori from Poisson-Boltzmann calculations ~\cite{chen2018accurate, ohlknecht2020correcting, rocklin2013calculating,Reif2014chargecorrection}.
+However, application of such corrections is typically inconvenient and computationally demanding.
+High grid resolutions for the Poisson-Boltzmann calculations are required to avoid numerical artifacts, and it may be necessary to average over multiple representative configurations to obtain reliable results.
+Furthermore, they rely on continuum electrostatic models, which cannot fully capture phenomena arising from discrete solvent effects.
+These effects can be quantified in the orientational-disorder limit such that analytical correction terms for this effect exist for solvent models with a single van der Waals interaction site.
+Readers interested in these conceptual and technical limitations are referred to detailed analyses by Hünenberger and co-workers~\cite{kastenholz2006solvfe, rocklin2013calculating}.
+Alternatively, Reif and Oostenbrink demonstrated in a proof-of-concept study that on-the-fly corrections to the physical forces can be effectively used to eliminate finite-size and approximate electrostatic artifacts~\cite{Reif2015onthefly}.
+
+Petrov et al.~\cite{Petrov2024guidelines} investigated the box-size dependencies of alchemical free energy calculations using PME together with the co-alchemical ion approach.
+They concluded that in neutral simulation boxes with a minimal distance to the box wall larger than 1 nm, finite-size artifacts could be made negligible.
+The following recommendations should be followed to minimize finite-size artifacts, when using lattice-sum electrostatics: (a) setup of free energy calculations such that there is no net charge change whenever possible, (b) ensuring an overall neutral simulation box, (c) use of a sufficiently large distance between the solute and the box wall, at least 1 nm, and (d) use of salt if possible to screen finite-size periodicity artifacts~\cite{chen2018accurate}.
+In cases where these recommendations are not practical for specific applications, such as small simulation boxes or when reaction field methods are preferred, charge-change corrections may be applied to reduce artifacts.
+Though both post-processing corrections and the co-alchemical ion approach come at a cost of increased uncertainties.
+The Poisson-Boltzmann calculations are approximate due to finite-grid sizes and are applied only to a subset of the ensemble.
+Co-alchemical ion approaches introduce additional degrees of freedom and increase sampling requirements.
 
@@
 
+\subsubsection{The importance of configurational sampling}
+\label{sec:configurational_sampling}
+As mentioned previously, free energy differences estimated from alchemical calculations should be independent of the initial system conformation, given sufficient configurational sampling.
+In cases where the system's slowest degrees of freedom are coupled with the alchemical variable, transitioning between alchemical states can indirectly enhance configurational sampling.
+This effect is particularly pronounced when certain alchemical states lack the energy barriers present at the fully coupled state (e.g.
+$\lambda=0$ in Fig.
+~\ref{fig:configurational_sampling}\textbf{A}), thereby allowing the system to bypass otherwise prohibitive configurational free energy barriers.
+For example, in a protein-ligand unbinding process where the kinetic barrier is predominantly associated with the disruption of intermolecular interactions or the separation of binding partners, simulating different intermediate states effectively samples configurations with varying center-of-mass distances, thus mitigating the sampling issue in the configurational space.
+
+Conversely, when the slowest degrees of freedom in the configurational space are largely orthogonal to the alchemical variable, the system can be stuck in its initial metastable state.
+Such entrapment leads to discrepancies in free energy estimates from simulations initiated from different conformations.
+This scenario, as illustrated in Fig.~\ref{fig:configurational_sampling}\textbf{B}, is not uncommon especially in flexible systems.
+For example, it can arise when the free energy difference of interest needs to account for metastable states separated by torsional barriers, such as the solvation free energy of a ligand having different rotamer states, the methylation free energy of a nucleobase in a duplex ~\cite{hsu2023alchemical}, and the binding free energy that involves flexible side chains near the binding site~\cite{la2022alchemical}.
+Similar sampling challenges have also been reflected in several recent studies, which highlights the sensitivity of alchemical calculations to initial conformations, whether derived from crystallographic structures~\cite{suruzhon2021sensitivity, baumann2024impact} or docking protocols~\cite{cappel2020impact}.
+
+Similarly, Fig.~\ref{fig:configurational_sampling}\textbf{C} highlights the importance of configurational sampling in alchemical calculations, where a free energy basin spans both the alchemical and configurational coordinates, potentially trapping the system in both dimensions.
+This can occur, for instance, when electrostatic interactions are turned off while van der Waals interactions remain, rendering the alchemically transformed molecule a ``greasy ball''.
+Notably, in cases where alchemical biases/weights are applied to facilitate even alchemical sampling across states, such as in an expanded ensemble simulation~\cite{lyubartsev1992new}, scenario Fig.~\ref{fig:configurational_sampling}\textbf{C} may evolve toward scenario Fig.~\ref{fig:configurational_sampling}\textbf{B} as the alchemical free energy profile is flattened out, while the basin in the configurational direction persists.
+
+There are multiple strategies to alleviate such sampling challenges in an alchemical calculation.
+A straightforward but computationally expensive method is to run longer simulations with multiple replicates to increase conformational heterogeneity~\cite{suruzhon2021sensitivity}.
+Another approach involves performing separate alchemical calculations from different metastable states, then computing the overall free energy difference as the negative log-sum-exp of the Boltzmann weights from each metastable state~\cite{hsu2023alchemical}.
+However, this method is only applicable to cases like Fig.~\ref{fig:configurational_sampling}\textbf{B}, where the system remains trapped in its initial metastable state throughout the calculation.
+A generally more effective strategy is to bias relevant configurational collective variables (CVs) during the alchemical simulation.
+This can be achieved by enhancing sampling within individual alchemical states and swapping conformations between alchemical states (preferably from different metastable states) via replica exchange protocols.
+More recently, several approaches have been proposed to simultaneously bias both configurational CV and alchemical variables, which can effectively address both scenarios Fig.~\ref{fig:configurational_sampling}\textbf{B} and \textbf{C}.
+Examples include alchemical metadynamics~\cite{hsu2023alchemical}, $\lambda$-ABF (adaptive biasing force)~\cite{lagardere2024lambda} and its combination~\cite{zhou2025zooming} with well-tempered metadynamics~\cite{barducci2008well}, to name a few.
+
+Notably, detecting and addressing configurational sampling challenges in alchemical calculations remains a non-trivial task.
+In relative free energy calculations, large cycle closure errors are commonly used as indicators of insufficient sampling, although small errors do not necessarily imply that sampling is adequate~\cite{suruzhon2021sensitivity}.
+In other types of calculations, diagnosing sampling deficiencies typically requires prior knowledge of relevant metastable states, as well as collective variables that distinguish them.
+Such collective variables are often based on physical intuition, experimental insights, or preliminary computational analyses.
+However, optimal collective variables are rarely known \textit{a priori}, complicating the systematic detection of sampling issues.
+Although recent ad-hoc machine learning methods for optimal CV identification have shown promise~\cite{gokdemir2025machine, sidky2020machine, bonati2021deep} and could in principle be straightforwardly integrated into alchemical sampling protocols, systematic studies explicitly demonstrating such integration have yet to appear.
+Consequently, widely accepted best practices about automated diagnostic protocols capable of identifying and resolving configurational sampling bottlenecks have not yet been established.
+
+\begin{figure}
+    % Figure to be updated
+    \includegraphics[width=0.88\columnwidth]{paper/figures/fig_configurational_sampling/FES_scenarios_path.pdf}
+    \caption{\textbf{Different scenarios of free energy surfaces.} In scenario \textbf{A}, the free energy barrier present at $\lambda=0$ is absent at other $\lambda$ states, so the system can go around the barrier by alternating sampling between alchemical and configurational space, e.g.
+via the path shown as the dashed line.
+In contrast, in scenario \textbf{B}, the free energy barrier extending all alchemical states prevents the system from sampling both metastable states at $\lambda=0$.
+In scenario \textbf{C}, the free energy basin extending in both alchemical and configurational directions can trap the system.}
+    \label{fig:configurational_sampling}
+\end{figure}
+
 \subsubsection{Which sampling scheme will work best for my problem?}
@@
 
-The last approach makes use of non-equilibrium simulations~\cite{aldeghi2018accurate}.
-In this approach, only end state $\vec{\lambda}$ replicas ($\vec{\lambda}$=0, $\vec{\lambda}=1$) are simulated at equilibrium; intermediate information is generated from non-equilibrium simulations that rapidly transition between end states.
-This approach is available in GROMACS and appears to be coming online in several other packages.
-A schematic of this approach is shown in Fig.~\ref{fig:fig_sampling_scheme}\textbf{D}.
+The non-equilibrium switching (Fig.~\ref{fig:fig_sampling_scheme}\textbf{D}) approach makes use of short out-of-equilibrium simulations~\cite{aldeghi2018accurate}.
+In this protocol, only end state $\vec{\lambda}$ replicas ($\vec{\lambda}$=0, $\vec{\lambda}=1$) are simulated at equilibrium; intermediate information is generated from non-equilibrium simulations that rapidly transition between end states.
+Over the course of each transition, Hamiltonian derivative with respect to $\lambda$ is recorded; subsequently the curves are integrated to yield work distributions for the forward and reverse processes.
+Crooks Fluctuation Theorem~\cite{crooks1999entropy} allows estimating free energy difference between the end states from the generated work distributions.~\cite{shirts2003equilibrium}
+
+The non-equilibrium (NEQ) approaches offer some advantages over the equilibrium sampling based methods.
+For example, the NEQ protocol allows equilibrium simulations of each end-state to be performed independently without pre-defining ligand network: the transitions between the ligands can be generated afterwards.
+This eliminates the redundant equilibrium sampling, e.g.
+in a starmap network, FEP based approaches would require sampling the central ligand for each edge, while NEQ would sample it once and connect to the other ligands afterwards.
+The NEQ protocol also provides flexibility in combining different structures in free energy calculations, e.g.
+simulations in states at $\lambda=0$ and $\lambda=1$ may be initialized with the independently generated ligand coordinates.
+Another example would be using true apo and holo protein structures as the end states to initialize simulations for absolute binding free energy calculations.
+
+The non-equilibrium approach is available in GROMACS and Amber.
+A closely related method, termed non-equilibrium cycling, is available in OpenMM.
+Among the commercial packages, it is implemented in OpenEye's Orion{\textregistered}~\cite{sorensen2024orion} platform.
+%A schematic of this approach is shown in Fig.~\ref{fig:fig_sampling_scheme}\textbf{D}.
 
@@
 Eq.~(\ref{eqn:observed-binding-constant-prot}) states that the observed binding constant is an average of the specific binding constants weighted by the populations, $p_0(AH)$ and  $p_0(A^-)$ of each ligand's chemical form in solution.
-In this specific example, if the $pKa$ of $AH$ is known and $\alpha = [A^-]/[AH] = 10^{pH-pKa} $ is ionization ratio of $A$ at a a given $pH$, the populations of the ionized and protonated forms are $p_0(A^-) = \alpha/(\alpha+1)$ and $p_0(AH) = 1/(\alpha+1)$, respectively.\cite{champion2024multistate,azimi2022application}
+In this specific example, if the $pKa$ of $AH$ is known and $\alpha = [A^-]/[AH] = 10^{pH-pKa} $ is the ionization ratio of $A$ at a a given $pH$, the populations of the ionized and protonated forms are $p_0(A^-) = \alpha/(\alpha+1)$ and $p_0(AH) = 1/(\alpha+1)$, respectively.\cite{champion2024multistate,azimi2022application}
 
@@
 \subsection{Unsettled practices}
+
+Alchemical Free Energy (AFE) calculations have primarily been applied to simulate the stoichiometric, non-reversible binding of organic molecules to well-structured proteins.
+However, several areas of drug discovery involve other classes of binding events.
+While the underlying theory of alchemical free energy calculations is generally applicable, additional work may be required to develop efficient protocols suited to systems beyond the traditional small molecule–protein complex.
+For example, there is growing interest in identifying small molecules that directly bind to RNA or DNA, with recent preliminary studies showing promising results using specialized simulation protocols.\cite{Abramyan2025,Rasouli2024} In contrast, there has been limited investigation into the applicability of AFE methods to small-molecule ligands targeting intrinsically disordered proteins (IDPs,\cite{Papadourakis2024} or proteins containing intrinsically disordered regions (IDRs) that undergo disorder-to-order transitions upon ligand binding.\cite{MendozaMartinez2022}
+The treatment of covalent binders has also received some attention.\cite{Luo2021,Zhang2019 } However, robust protocols for handling datasets of irreversible covalent binders containing diverse ‘warheads’ have not yet achieved widespread adoption.\cite{Yu2019}
 

### 2025-06-18 paper/manuscript.tex
Wording tweaks

--- 2025-06-10 paper/manuscript.tex
@@
 \author[3*]{John D. Chodera}
-\author[9]{Vytautas Gapsys}
 \author[9]{David F. Hahn}
@@
 \orcid{David F. Hahn}{0000-0003-2830-6880}
-\orcid{Vytautas Gapsys}{0000-0002-6761-7780}
 
@@
 
-In many cases, the quantity of interest is the change in binding affinity between a compound $A$ and a related compound $B$ (e.g., by modifying one of the drug scaffold's substituents, see (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{E})), which, by using Eq.~\ref{eq:binding-free-energy-from-bound-unbound-probability-ratio} and \ref{eq:bound-unbound-probability-ratio} is given by
+In many cases, the quantity of interest is the change in binding affinity between a compound $A$ and a related compound $B$ (e.g., by modifying one of the drug scaffold's substituents, see (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{F})), which, by using Eq.~\ref{eq:binding-free-energy-from-bound-unbound-probability-ratio} and \ref{eq:bound-unbound-probability-ratio} is given by
 \begin{equation}\label{eq:delta-delta-G-physical}
@@
 Well-known protein systems with water mediated ligand interactions are for example: HSP90 which formed part of the D3R grand challenge 2015~\cite{mey2016blinded}, A2A~\cite{brucemacdonald2018ligand}, MUP~\cite{ross2015water}, ~\cite{deflorian2020accurate}, and others~\cite{michel2009energetics}.
+
+\paragraph{Membrane proteins require additional care}
+Integral membrane proteins require embedding in a suitable model membrane to remain stable during simulations.
+Lipids are usually not included in protein force fields and, although they may be covered by generic small molecule force fields, have often been parameterized separately (e.g., AMBER's Lipid21 \cite{AMBERLipid21} and the CHARMM lipid force field \cite{CHARMMLipidFF}).
+Membrane systems can be simulated in multiple thermodynamic ensembles, but the semi-isotropic scheme $NP_{z}P_{xy}T$ is frequently employed, allowing the pressure in the bilayer plane ($xy$) to scale independently \cite{LipidBestPracticesCOMS}.
+These systems require more extensive equilibration than soluble proteins.
+Some workflows, such as CHARMM-GUI or Desmond membrane preparation, recommend a multi-protocol to relax membrane systems \cite{https://doi.org/10.1002/jcc.23702,DesmondManual}.
+
+Alchemical transformations that involve net charge changes can provide difficulties for membrane systems; phospholipids carry functional groups with nonzero formal charge and cannot be treated as a solvent with a uniform dielectric in the same way as water.
+Several approaches are reviewed in \cite{wu2022correction-51c}.
+The authors recommend the use of a co-alchemical ion, specifically making an ion of equivalent charge appear alongside the disappearing charged molecule to maintain system neutrality.
+
+In the case of a lipid-facing binding site, there are three additional challenges.
+First, simulation time should be extended to ensure sufficient sampling of lipid configurational space.
+Lipids may either bind alongside small molecules, acting as cofactors, or be displaced from the binding site by a ligand.
+If lipids participate in binding through protein-lipid or ligand-lipid interactions, it is necessary to sample these intermolecular interactions through additional $\lambda$ windows, longer simulation times, or different alchemical protocols.
+Displacing a lipid will have its own free energy cost which may take a long time to converge if it requires diffusion and rearrangement of nearby lipids.
+Second, considering the free energy cost of a ligand partitioning into the membrane may provide additional insights \cite{doi:10.1021/acs.jcim.1c01147}.
+Third, the free energy of binding for a small molecule may be strongly coupled to the lipid species.
+In some cases, the identity of the lipid may be known.
+But in other cases, it may be difficult to discern the identity of the lipid from X-ray or cryo-EM structures due to the flexible nature of lipid head groups and benchmarking is recommended to match experimental data.
 
@@
 If the net charge of the system changes as the alchemical variable changes during the  calculation, this can pose major challenges.
-Specifically, finite-size effects can introduce significant charge-dependent artifacts into computed binding free energies, in part because typical schemes for long-range electrostatics, including particle mesh Ewald (PME) and reaction field, do not handle free energy contributions from such changes as they would be handled in a hypothetical macroscopic bulk solution.
-The outcome depends either on the size of the cutoff for reaction field methods or on the size of the simulation box when using lattice sum methods due to periodic boundary conditions~\cite{lin2014overview, ohlknecht2020correcting, rocklin2013calculating}.
-
-Lattice summation methods (e.g.
-PME) effectively neutralize the simulation box by applying an implicit homogeneously distributed background counter charge~\cite{figueirido1995finite,hummer1996free}.
-The background charge can be interpreted as a converged distribution of counterions and may be a reasonable choice for computing hydration free energies of monovalent ions (finite size effect corrections can be applied in these cases)~\cite{hummer1996free}.
-In inhomogeneous systems with different dielectrics, e.g.
-solvated protein or lipid bilayer in a water box, assuming uniform background countercharge distribution is unphysical, as the higher countercharge density should be located in the regions of higher dielectric~\cite{hub2014pme}.
-This, in turn, causes significant artifacts in sampled population densities of charged molecules, i.e.
-errors in calculated free energy differences.
-
-There are two main classes of approaches to address artefacts introduced by changes in net charge during alchemical free energy calculations: (1) avoiding changing the net charge altogether and (2) post-processing approaches to correct for artefacts using continuum electrostatics.
+Specifically, finite-size effects can introduce significant charge-dependent artifacts into computed binding free energies, in part because typical schemes for long-range electrostatics (including PME and reaction field) do not handle free energy contributions from such changes effectively or as they would be handled in a hypothetical macroscopic bulk solution~\cite{lin2014overview, ohlknecht2020correcting, rocklin2013calculating}.
+
+There are two main potential solutions to avoid artifacts due to changes in net charge: avoiding changing the net charge, and correcting for the introduced artifacts.
+
 Many relative free energy planning tools have been set up to avoid changing the net charge of the systems considered, including LOMAP~\cite{liu2013lead} and Schr\"{o}dinger's FEP+~\cite{wang2015accurate}.
-In absolute free energy calculations, it is often difficult to avoid a change in the net charge.
-One way to avoid charge changes is to ensure that, as a charged ligand is removed, a charged counterion of opposite sign is also removed, or one of the same sign is inserted.
-This is sometimes referred to as a “co-alchemical ion” approach to deal with the required charge change.
-This approach is easily supported in most free energy calculating programs, as it merely involves an additional simultaneous modification.
-There are special flavours of the co-alchemical ion approach such as proposed by Gapsys et al.~\cite{gapsys2015calculation}, which uses a double-system/single-box setup.
-Here, the two branches of a thermodynamic cycle as used in relative free energy calculations are placed in one simulation box, effectively circumventing a change in the overall net charge during a perturbation.
-Note that the co-alchemical approaches only avoid finite-size effects with lattice-sum methods such as PME.
-
-Charge-change correction schemes have been explored and offer potentially viable solutions for the problem~\cite{mey2018impact}, where artifacts introduced by finite-size effects are corrected a posteriori from Poisson-Boltzmann calculations ~\cite{chen2018accurate, ohlknecht2020correcting, rocklin2013calculating,Reif2014chargecorrection}.
-However, application of such corrections is typically inconvenient and computationally demanding.
-High grid resolutions for the Poisson-Boltzmann calculations are required to avoid numerical artifacts, and it may be necessary to average over multiple representative configurations to obtain reliable results.
-Furthermore, they rely on continuum electrostatic models, which cannot fully capture phenomena arising from discrete solvent effects.
-These effects can be quantified in the orientational-disorder limit such that analytical correction terms for this effect exist for solvent models with a single van der Waals interaction site.
-Readers interested in these conceptual and technical limitations are referred to detailed analyses by Hünenberger and co-workers~\cite{kastenholz2006solvfe, rocklin2013calculating}.
-Alternatively, Reif and Oostenbrink demonstrated in a proof-of-concept study that on-the-fly corrections to the physical forces can be effectively used to eliminate finite-size and approximate electrostatic artifacts~\cite{Reif2015onthefly}.
-
-Petrov et al.~\cite{Petrov2024guidelines} investigated the box-size dependencies of alchemical free energy calculations using PME together with the co-alchemical ion approach.
-They concluded that in neutral simulation boxes with a minimal distance to the box wall larger than 1 nm, finite-size artifacts could be made negligible.
-The following recommendations should be followed to minimize finite-size artifacts, when using lattice-sum electrostatics: (a) setup of free energy calculations such that there is no net charge change whenever possible, (b) ensuring an overall neutral simulation box, (c) use of a sufficiently large distance between the solute and the box wall, at least 1 nm, and (d) use of salt if possible to screen finite-size periodicity artifacts~\cite{chen2018accurate}.
-In cases where these recommendations are not practical for specific applications, such as small simulation boxes or when reaction field methods are preferred, charge-change corrections may be applied to reduce artifacts.
-Though both post-processing corrections and the co-alchemical ion approach come at a cost of increased uncertainties.
-The Poisson-Boltzmann calculations are approximate due to finite-grid sizes and are applied only to a subset of the ensemble.
-Co-alchemical ion approaches introduce additional degrees of freedom and increase sampling requirements.
+Absolute free energy calculations can also potentially avoid changing the charge of the system by making a charge perturbation of equal and opposite sign elsewhere in the system; for example, as a charged ligand is removed, a charged counterion of opposite sign could also be removed, or one of the same sign could be inserted.
+This is sometimes referred to as an "alchemical ion" approach for dealing with the needed charge change, and is also employed by the Yank free energy package~\cite{wang2013identifying}.
+Charge corrections have also been explored, and are potentially a viable solution to this problem~\cite{mey2018impact} where artifacts introduced by finite-size effects are corrected numerically~\cite{chen2018accurate, ohlknecht2020correcting}.
+However, application of such corrections typically remains less common than the use of a co-alchemical ion.
+A third approach has been proposed by Gapsys et al.~\cite{gapsys2015calculation} which uses a double-system/single box setup.
+
+When free energy calculations \emph{do} need to change the charge of a ligand or solute, the literature does not yet seem to indicate what approach should be preferable, so considerable care should be taken.
+We are not yet aware of a careful comparison of charge corrections versus other approaches such as decoupling an ion at the same time, so in our view the issue of proper handling of charge mutations in the context of alchemical calculations remains a research problem, and several papers give good guides for exploring the problem further~\cite{lin2014overview, ohlknecht2020correcting, rocklin2013calculating}.
 
@@
 
-\subsubsection{The importance of configurational sampling}
-\label{sec:configurational_sampling}
-As mentioned previously, free energy differences estimated from alchemical calculations should be independent of the initial system conformation, given sufficient configurational sampling.
-In cases where the system's slowest degrees of freedom are coupled with the alchemical variable, transitioning between alchemical states can indirectly enhance configurational sampling.
-This effect is particularly pronounced when certain alchemical states lack the energy barriers present at the fully coupled state (e.g.
-$\lambda=0$ in Fig.
-~\ref{fig:configurational_sampling}\textbf{A}), thereby allowing the system to bypass otherwise prohibitive configurational free energy barriers.
-For example, in a protein-ligand unbinding process where the kinetic barrier is predominantly associated with the disruption of intermolecular interactions or the separation of binding partners, simulating different intermediate states effectively samples configurations with varying center-of-mass distances, thus mitigating the sampling issue in the configurational space.
-
-Conversely, when the slowest degrees of freedom in the configurational space are largely orthogonal to the alchemical variable, the system can be stuck in its initial metastable state.
-Such entrapment leads to discrepancies in free energy estimates from simulations initiated from different conformations.
-This scenario, as illustrated in Fig.~\ref{fig:configurational_sampling}\textbf{B}, is not uncommon especially in flexible systems.
-For example, it can arise when the free energy difference of interest needs to account for metastable states separated by torsional barriers, such as the solvation free energy of a ligand having different rotamer states, the methylation free energy of a nucleobase in a duplex ~\cite{hsu2023alchemical}, and the binding free energy that involves flexible side chains near the binding site~\cite{la2022alchemical}.
-Similar sampling challenges have also been reflected in several recent studies, which highlights the sensitivity of alchemical calculations to initial conformations, whether derived from crystallographic structures~\cite{suruzhon2021sensitivity, baumann2024impact} or docking protocols~\cite{cappel2020impact}.
-
-Similarly, Fig.~\ref{fig:configurational_sampling}\textbf{C} highlights the importance of configurational sampling in alchemical calculations, where a free energy basin spans both the alchemical and configurational coordinates, potentially trapping the system in both dimensions.
-This can occur, for instance, when electrostatic interactions are turned off while van der Waals interactions remain, rendering the alchemically transformed molecule a ``greasy ball''.
-Notably, in cases where alchemical biases/weights are applied to facilitate even alchemical sampling across states, such as in an expanded ensemble simulation~\cite{lyubartsev1992new}, scenario Fig.~\ref{fig:configurational_sampling}\textbf{C} may evolve toward scenario Fig.~\ref{fig:configurational_sampling}\textbf{B} as the alchemical free energy profile is flattened out, while the basin in the configurational direction persists.
-
-There are multiple strategies to alleviate such sampling challenges in an alchemical calculation.
-A straightforward but computationally expensive method is to run longer simulations with multiple replicates to increase conformational heterogeneity~\cite{suruzhon2021sensitivity}.
-Another approach involves performing separate alchemical calculations from different metastable states, then computing the overall free energy difference as the negative log-sum-exp of the Boltzmann weights from each metastable state~\cite{hsu2023alchemical}.
-However, this method is only applicable to cases like Fig.~\ref{fig:configurational_sampling}\textbf{B}, where the system remains trapped in its initial metastable state throughout the calculation.
-A generally more effective strategy is to bias relevant configurational collective variables (CVs) during the alchemical simulation.
-This can be achieved by enhancing sampling within individual alchemical states and swapping conformations between alchemical states (preferably from different metastable states) via replica exchange protocols.
-More recently, several approaches have been proposed to simultaneously bias both configurational CV and alchemical variables, which can effectively address both scenarios Fig.~\ref{fig:configurational_sampling}\textbf{B} and \textbf{C}.
-Examples include alchemical metadynamics~\cite{hsu2023alchemical}, $\lambda$-ABF (adaptive biasing force)~\cite{lagardere2024lambda} and its combination~\cite{zhou2025zooming} with well-tempered metadynamics~\cite{barducci2008well}, to name a few.
-
-Notably, detecting and addressing configurational sampling challenges in alchemical calculations remains a non-trivial task.
-In relative free energy calculations, large cycle closure errors are commonly used as indicators of insufficient sampling, although small errors do not necessarily imply that sampling is adequate~\cite{suruzhon2021sensitivity}.
-In other types of calculations, diagnosing sampling deficiencies typically requires prior knowledge of relevant metastable states, as well as collective variables that distinguish them.
-Such collective variables are often based on physical intuition, experimental insights, or preliminary computational analyses.
-However, optimal collective variables are rarely known \textit{a priori}, complicating the systematic detection of sampling issues.
-Although recent ad-hoc machine learning methods for optimal CV identification have shown promise~\cite{gokdemir2025machine, sidky2020machine, bonati2021deep} and could in principle be straightforwardly integrated into alchemical sampling protocols, systematic studies explicitly demonstrating such integration have yet to appear.
-Consequently, widely accepted best practices about automated diagnostic protocols capable of identifying and resolving configurational sampling bottlenecks have not yet been established.
-
-\begin{figure}
-    % Figure to be updated
-    \includegraphics[width=0.88\columnwidth]{paper/figures/fig_configurational_sampling/FES_scenarios_path.pdf}
-    \caption{\textbf{Different scenarios of free energy surfaces.} In scenario \textbf{A}, the free energy barrier present at $\lambda=0$ is absent at other $\lambda$ states, so the system can go around the barrier by alternating sampling between alchemical and configurational space, e.g.
-via the path shown as the dashed line.
-In contrast, in scenario \textbf{B}, the free energy barrier extending all alchemical states prevents the system from sampling both metastable states at $\lambda=0$.
-In scenario \textbf{C}, the free energy basin extending in both alchemical and configurational directions can trap the system.}
-    \label{fig:configurational_sampling}
-\end{figure}
-
 \subsubsection{Which sampling scheme will work best for my problem?}
@@
 
-The non-equilibrium switching (Fig.~\ref{fig:fig_sampling_scheme}\textbf{D}) approach makes use of short out-of-equilibrium simulations~\cite{aldeghi2018accurate}.
-In this protocol, only end state $\vec{\lambda}$ replicas ($\vec{\lambda}$=0, $\vec{\lambda}=1$) are simulated at equilibrium; intermediate information is generated from non-equilibrium simulations that rapidly transition between end states.
-Over the course of each transition, Hamiltonian derivative with respect to $\lambda$ is recorded; subsequently the curves are integrated to yield work distributions for the forward and reverse processes.
-Crooks Fluctuation Theorem~\cite{crooks1999entropy} allows estimating free energy difference between the end states from the generated work distributions.~\cite{shirts2003equilibrium}
-
-The non-equilibrium (NEQ) approaches offer some advantages over the equilibrium sampling based methods.
-For example, the NEQ protocol allows equilibrium simulations of each end-state to be performed independently without pre-defining ligand network: the transitions between the ligands can be generated afterwards.
-This eliminates the redundant equilibrium sampling, e.g.
-in a starmap network, FEP based approaches would require sampling the central ligand for each edge, while NEQ would sample it once and connect to the other ligands afterwards.
-The NEQ protocol also provides flexibility in combining different structures in free energy calculations, e.g.
-simulations in states at $\lambda=0$ and $\lambda=1$ may be initialized with the independently generated ligand coordinates.
-Another example would be using true apo and holo protein structures as the end states to initialize simulations for absolute binding free energy calculations.
-
-The non-equilibrium approach is available in GROMACS and Amber.
-A closely related method, termed non-equilibrium cycling, is available in OpenMM.
-Among the commercial packages, it is implemented in OpenEye's Orion{\textregistered}~\cite{sorensen2024orion} platform.
-%A schematic of this approach is shown in Fig.~\ref{fig:fig_sampling_scheme}\textbf{D}.
+The last approach makes use of non-equilibrium simulations~\cite{aldeghi2018accurate}.
+In this approach, only end state $\vec{\lambda}$ replicas ($\vec{\lambda}$=0, $\vec{\lambda}=1$) are simulated at equilibrium; intermediate information is generated from non-equilibrium simulations that rapidly transition between end states.
+This approach is available in GROMACS and appears to be coming online in several other packages.
+A schematic of this approach is shown in Fig.~\ref{fig:fig_sampling_scheme}\textbf{D}.
 
@@
 Eq.~(\ref{eqn:observed-binding-constant-prot}) states that the observed binding constant is an average of the specific binding constants weighted by the populations, $p_0(AH)$ and  $p_0(A^-)$ of each ligand's chemical form in solution.
-In this specific example, if the $pKa$ of $AH$ is known and $\alpha = [A^-]/[AH] = 10^{pH-pKa} $ is the ionization ratio of $A$ at a a given $pH$, the populations of the ionized and protonated forms are $p_0(A^-) = \alpha/(\alpha+1)$ and $p_0(AH) = 1/(\alpha+1)$, respectively.\cite{champion2024multistate,azimi2022application}
+In this specific example, if the $pKa$ of $AH$ is known and $\alpha = [A^-]/[AH] = 10^{pH-pKa} $ is ionization ratio of $A$ at a a given $pH$, the populations of the ionized and protonated forms are $p_0(A^-) = \alpha/(\alpha+1)$ and $p_0(AH) = 1/(\alpha+1)$, respectively.\cite{champion2024multistate,azimi2022application}
 
@@
 \subsection{Unsettled practices}
-
-Alchemical Free Energy (AFE) calculations have primarily been applied to simulate the stoichiometric, non-reversible binding of organic molecules to well-structured proteins.
-However, several areas of drug discovery involve other classes of binding events.
-While the underlying theory of alchemical free energy calculations is generally applicable, additional work may be required to develop efficient protocols suited to systems beyond the traditional small molecule–protein complex.
-For example, there is growing interest in identifying small molecules that directly bind to RNA or DNA, with recent preliminary studies showing promising results using specialized simulation protocols.\cite{Abramyan2025,Rasouli2024} In contrast, there has been limited investigation into the applicability of AFE methods to small-molecule ligands targeting intrinsically disordered proteins (IDPs,\cite{Papadourakis2024} or proteins containing intrinsically disordered regions (IDRs) that undergo disorder-to-order transitions upon ligand binding.\cite{MendozaMartinez2022}
-The treatment of covalent binders has also received some attention.\cite{Luo2021,Zhang2019 } However, robust protocols for handling datasets of irreversible covalent binders containing diverse ‘warheads’ have not yet achieved widespread adoption.\cite{Yu2019}
 

### 2025-06-22 paper/manuscript.tex
comments by Stefan Boresch and reworked section on weak binders

--- 2025-06-10 paper/manuscript.tex
@@
 %%% PREAMBLE
-\documentclass[9pt,bestpractices,pubversion]{livecoms}
+\documentclass[9pt,bestpractices]{livecoms}
 % Use the 'onehalfspacing' option for 1.5 line spacing
@@
 
-In general there are mutliple ways on how a MCSS can be calculated.
-One of the most prominent approaches is to estimate the MCSS with a 2D subgraph isomorphism solver(SIS) using the 2D topological stuructre information of molecules\cite{raymond2002maximum}.
-A large number of software tools can compute SIS MCSS matches using different cheminformatics packages.
-Some rely on RDKit~\cite{rdkit2019Dec}, such as pmx~\cite{gapsys2015pmx}, LOMAP~\cite{liu2013lead}, FESetup~\cite{loeffler2015fesetup}, ProtoCaller~\cite{suruzhon2020protocaller}, SMArt~\cite{petrov2021perturbation}, and partially BioSimSpace~\cite{hedges2019biosimspace}, while others such as fkckombu~\cite{kawabata20143d} are standalone tools.
+There are different tools that allow the generation of MCSS matches as well as single topology input.
+A large number of software tools can compute MCSS matches using different cheminformatics packages.
+Some rely on RDKit~\cite{rdkit2019Dec}, such as pmx~\cite{gapsys2015pmx}, LOMAP~\cite{liu2013lead}, FESetup~\cite{loeffler2015fesetup} and partially BioSimSpace~\cite{hedges2019biosimspace}, while others such as fkckombu~\cite{kawabata20143d} are standalone tools.
 Schr\"{o}dinger's FEP+ planning tool was originally based on a version of LOMAP, and it also uses MCSS matching as well as 3D considerations to plan the network of single topology calculations between molecules~\cite{wang2015accurate}.
 
-SIS based methods, showed to struggle with 3D properties of the molecules, that need to be mapped, like for example stereochemistry.
-Therefore some packages developed a post-mapping step, where they filtered for atom distances, and therefore added 3D information into the mapping.
-Still SIS MCSS searches can be relatively time consuming as the algorithmic complexity is classified as NP-complete, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial similarity assessment and then use MCSS only to identify final mappings for relative calculations~\cite{raymond2002maximum,klabunde2012mars,jones2009elucidating, Duesbury2018Comparison}.
-The SIS MCSS approach, though relatively standard, takes into account only topological similarity.
+MCSS searches can be relatively time consuming, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial similarity assessment and then use MCSS only to identify final mappings for relative calculations~\cite{raymond2002maximum,klabunde2012mars,jones2009elucidating}.
+The MCSS approach, though relatively standard, takes into account only topological similarity.
 It is possible that changes in binding mode could actually require a different choice of mapping, so in some cases mappings may need to be planned differently depending on 3D positioning of atoms in space.
-
-Here an alternative way to estimate the MCSS is to calculate a bipartite graph matching based on the 3D positions of two molecules.
-These approaches tend to be very fast (can have algorithmic complexity of $O(n)=N^3$) and geometric accurate in their mapping (e.g.
-stereochemistry or binding mode differing).
-The geometric approaches additionally allow mappings of differing binding modes of the same molecule, or can be even used for significant larger molecules like proteins, due to the speed increase.
-A disadvantage of these approaches is the strong dependency on the coordinate sets being well aligned.
-Another aspect is taking the topology changes of the molecules not into account, which can be done if necessary in a next step with a filtering approach, like checking for changes in ring hybridization or ring breaking (this problem is discusse below)|.
-Tools that can be used to estimate the MCSS this way are contained for example in pmx \cite{gapsys2015pmx} or Kartograf \cite{ries2024kartograf}.
-
-In any case Visual inspection prior to simulation is recommended to ensure that the mapping criteria correspond to the expected binding mode.
+Visual inspection prior to simulation is recommended to ensure that the mapping criteria correspond to the expected binding mode.
 If the mapping protocol returns simulations that correspond to different binding modes of a ligand within a perturbation map, this can cause large hysteresis.
 
-A next step for atom mappings, would be to support multistate methods (e.g.
-RE-EDS~\cite{sidler2016replica}, A-EDS~\cite{perthold2020toward} or Ladybugs~\cite{robo2022achieving}) with efficient mapping algorithms.
-There are protoype starting point for solving this problem, but to our knowledge, there is only few published work on this aspect.
-\cite{petrov2021Perturbation}
-
-Single coordinate set relative calculations and calculations based on substructure searches only work if in fact the ligands share a common substructure, e.g.
+Single topology relative calculations and calculations based on substructure searches only work if in fact the ligands share a common substructure, e.g.
 are part of a congeneric series, see Fig.~\ref{fig:fig_mcss}.
-If no common substructure is shared, then dual coordinate sets methods like linked-dual or separated topology free energy calculations are needed.
+If no common substructure is shared, then alternative dual or separated topology free energy calculations are needed.
 One would co-localize a pair of compounds in a binding site, exclude their interactions with one another, and compute the relative binding free energy by turning one molecule on from being dummy atoms while turning the other off.
-For such approaches it is crucial to find a good set of restraints, that keep the ligands colocalized over the time of the simulation .
-This problem is similar to the required restraints in ABFE Calculations (see \ref{ABFErest}).
-Few descriptions of possible approaches in RBFE calculations can be found for such approaches like the one suggested by \textit{Baumann et al.}~\cite{baumann2023broadening} based on boresch style restraints~\cite{boresch2003absolute}, with an elaborate automatic atom selection approach, for the seperated topology approach, or Restraintmaker~\cite{ries2022restraintmaker} for the linked-dualtopology approaches, which finds a small set of atoms to be linked via distance restraints, with a small distance to each other.
-However the implementation and testing of an fully automatized dual-topology pipeline for RBFE dual topology approaches is still to our knowledge a remaining research problem and was only tested with small datasets.
-\cite{jespers2019qligfep, Rieder2022Lev}
+To our knowledge no general pipeline for such calculations yet exists and this would likely remain a research problem.
+Using an absolute free energy approach instead seems more promising in such a case.
 
@@
 \paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
-Some approaches have attempted to address this using a single coordinate sets and with soft bond approaches ~\cite{clark2019relative,wang2017accurate} or dual coordinate set approach \cite{jespers2019qligfep, ries2022relative} but a general solution is not yet in mainstream use.
-Still, FEP+ implements one solution\cite{wang2017accurate}.
+Some approaches have attempted to address this~\cite{clark2019relative} but a general solution is not yet in mainstream use. Still, FEP+ implements one solution.
 
@@
 
-Absolute free energy calculations involve completely removing the interactions between the ligand or solute and its environment, taking it to a non-interacting state that may or may not retain intramolecular non-bonded interactions.
-This non-interacting state can then be shifted between environments---from the protein to water, or from one solution to another---without changing its free energy other than that due to the changing volume of the simulations, and then interactions can be restored in the new environment.
-
-Absolute free energies are by definition reported with respect to a specific reference or standard state, which effectively determines the arbitrary point at which the free energy is 0.
-The role of the standard state is particularly evident from the expression of the binding free energy between a receptor $R$ and ligand $L$
-\begin{equation} \label{eq:DGfromKAB}
-    \Delta G = -k_BT ~ \ln \left( c^{\circ} K_b \right)  = -k_BT ~ \ln\left( c^{\circ} \frac{[RL]}{[L][R]} \right) \, .
-\end{equation}
-Here, the reference state concentration $c^{\circ}$ converts the binding constant $K_b$ into a dimensionless quantity expressed in reference concentration units.
-It should be noted that ignoring the term $c^{\circ}$ is equivalent to assuming a reference concentration of 1~D$^{-1}$, where D are the units used to express $K_b$, and would thus cause the value of $\Delta G$ to vary with the choice of the units.
-It is convenient to define a standard state at a constant pressure of 1~atm and where each chemical species (i.e., R, L, and RL) in the reaction solvent has a concentration of $c^{\circ}$~=~1~M~=~1~molecule/1660~\r{A}$^3$ but do not interact with other molecules of $R$, $L$, or $RL$.
-
-\paragraph{Handling the standard state in absolute free energy calculations.}
-For solvation free energy calculations, handling the standard state is typically straightforward, and treating it correctly simply means ensuring that the non-interacting solute still occupies essentially the same volume as the solute in the interacting system.
-So typically in such cases no special care is required to ensure the correct standard state, as long as the \emph{experimental} data being analyzed uses the same standard state.
-If this is not the case, a simple entropic correction to the free energy of $k_B T\ln (V_f/V_i) = k_B T \ln (C_i/C_f)$ to the experimental data is needed.
-
-For binding, however, the situation is more complex and requires special care.
-Because the simulations are typically performed using restraints and at concentrations that are different from 1 M, the expression of the free energy requires the following correction~\cite{gilson1997statisticalthermodynamic} (see an example of such a thermodynamic cycle in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle})
-\begin{equation}\label{eq:restraint-correction}
-    \Delta G^{\circ}_{\mathrm{restr}} = -k_BT ~ \ln \left( c^{\circ}{V_L} \right) -k_BT ~ \ln \left( \frac{\xi_L}{8 \pi^2} \right),
-\end{equation}
-where $V_L$ and $\xi_L$ are respectively the volume of the translational and rotational degrees of freedom of the non-interacting ligand in the simulation box.
-When no restraints are used, the non-interacting ligand is free to translate and rotate in the simulation box (i.e., $V_L = V_{\mathrm{box}}$ and $\xi_L = 8\pi^2$), and the rotational term is zero.
-A sufficiently thorough exploration of the simulation box by the non-interacting ligand is, however, required for the formula to be valid.
-This is typically hard to achieve as the exploration process is governed by diffusion, and weak transient nonspecific binding will occur at other sites on the protein. 
-The addition of a restraint limits the volume available to the non-interacting ligand, thus speeding the convergence of the sampling.
-In addition, when enhanced sampling methods such as Hamiltonian $\lambda$ exchange are used (see Sec.~\ref{sec:sampling_schemes}), the use of a restraint is typically necessary as it keeps the ligand in the binding site in the interacting state (see also Sec.~\ref{sec:weak-binders}) and generally reduce the round-trip time of replicas.
-When restraints are employed, the values of $V_L$ and $\xi_L$ are restraint-dependent, but for commonly employed restraints, these can be usually easily computed analytically or numerically by solving the relevant integral.
-
-\paragraph{Several choices of restraints are possible.}\label{ABFErest}
-In practice, a variety of types of restraints are common, from simple harmonic distance restraints between the ligand and the protein~\cite{mobley2006use}, to flat-bottom restraints which work similarly but only exert a force if the ligand leaves a specific region~\cite{chen2007can}.
-Because these restraints do not limit the rotational degrees of freedom of the ligand, the rotational term entering the correction in Eq.~\ref{eq:restraint-correction} is zero.
-
-Alternatively, a set of restraints proposed by Boresch have also commonly been employed, where all six rigid-body degrees of freedom governing the orientation of the ligand relative to the receptor are restrained~\cite{boresch2003absolute, leitgeb2005alchemical}.
-Further restraints, such as on the overall ligand RMSD have also been used~\cite{woo2005calculation}.
-
-In principle, all of these forms will yield correct binding free energies in the limit of adequate sampling if their effects and connection to the standard state are correctly handled, but they have different strengths and weaknesses.
-For example, with more involved restraints, sampling at intermediate $\vec{\lambda}$ values will usually not need to be as extensive but more computational effort must go to computing the free energy to turn on the restraints.
-Additionally, such restraints would typically keep the ligand from exploring alternative binding modes.
+An absolute binding free energy calculation yields the standard free energy of formation of the molecular complex from the separated receptor and ligand species.
+Hence, as discussed in \ref{sec:non-covalent-molecular-binding}, it must depend on the standard state concentration $C^\circ$ (here $1 \, {\rm M} = 1 \, {\rm molecule}/1660$~\r{A}$^3$) and on the definition of the complex, that is on the specific binding pose for which the binding free energy is requested.
+Omitting the standard state concentration would yield an undefined result that would vary with the choice of concentration units.
+A similar vague estimate would result from omitting the binding pose definition that, for example, could equally describe binding to a specific site of the receptor or unspecific binding throughout the receptor surface depending on the extent of conformational sampling (see Sec.~\ref{sec:weak-binders}).
+
+In the statistical mechanics formulation of Gilson et al.~\cite{gilson1997statisticalthermodynamic}, described in Section \ref{sec:non-covalent-molecular-binding}, the standard state concentration appears explicitly in Eq.~(\ref{eq:DG0-binding-constant-def}) and the binding pose is defined by the indicator function $I(\zeta)$ that characterizes the partition function of the complex; Eq.~(\ref{eq:intra-zRL-def}).
+The role of the standard state and binding site definitions are particularly evident when expressing Eq.~(\ref{eq:DG0-binding-constant-def}) in terms of an ensemble average.
+To do so, we express the integrals in the numerator and denominator of Eq.~(\ref{eq:DG0-binding-constant-def}) over the same set of degrees of freedom.
+Note that the intramolecular configurational partition function of the complex, $z_{RL(v)}$, includes six more degrees of freedom--the position and orientation of the ligand relative to the receptor collectively described by the variable $\zeta$--than the product of the intramolecular partition functions, $z_{R(v)} z_{L(v)}$, of the separated receptor and ligand.
+To correct this discrepancy, we multiply and divide the right-hand side of Eq.~(\ref{eq:DG0-binding-constant-def})  by the quantity
+\begin{equation} \label{eq:Vsite-def}
+    V_{\rm site} \Omega = \int d\zeta I(\zeta) \, ,
+\end{equation}
+%\Omega is in my opinion a better symbol of the orientational extent than \xi_L. Lowercase greek letters are usually used
+%for dynamical variables rather than integrated quantities
+which measures the extent of the binding pose macrostate, where $V_{\rm site}$ corresponds to the spatial extent and $\Omega$ the orientational extent.
+The right-hand side of Eq.~(\ref{eq:Vsite-def}) combined with the integrals $z_{R(v)}$ and $z_{L(v)}$ yields a partition function similar to Eq.~(\ref{eq:intra-zRL-def}) but corresponding to a complex in which ligand and receptor are uncoupled~\cite{gallicchio2011recent} as if they were at infinite distance apart.
+The result is
+\begin{equation}\label{eq:DG0-binding-constant-def-average}
+K_b = C^\circ V_{\rm site} \frac{\Omega}{8 \pi^2} \langle e^{-\beta \Delta \Psi}  \rangle_0 
+\end{equation}
+where, similarly to Eq.~(\ref{eq:zratio-solv-average}),
+\begin{equation}\label{eq:pert-energy-def}
+\Delta \Psi(x_R, x_L, \zeta) =  \Psi_v(x_R, x_L, \zeta) - \left[  \Psi_v(x_R) + \Psi(x_L) \right]
+\end{equation}
+is the solvent-averaged binding energy of the complex when receptor and ligands are in intramolecular configurations $x_R$ and $x_L$, the ligand is in position and orientation $\zeta$ relative to the receptor (the uncoupled potential energy does not depend on  the position and orientation of the ligand relative to the receptor), and $\langle \ldots \rangle_0$ represents an ensemble average in the uncoupled ensemble.
+Finally, note that $\Omega = 8 \pi^2$ when the binding pose indicator function does not depend on the orientation of the ligand and Eq.~(\ref{eq:DG0-binding-constant-def-average}) simplifies to
+\begin{equation}\label{eq:DG0-binding-constant-def-average-noOmega}
+K_b = C^\circ V_{\rm site} \langle e^{-\beta \Delta \Psi}  \rangle_0 
+\end{equation}
+which, expressed in free energy units, yields
+\begin{eqnarray}\label{eq:DG0-def-Vsite}
+  \Delta G^\circ_b &=&  - k_B T \ln K_b = - k_B T \ln C^\circ V_{\rm site} - k_B T \ln \langle e^{-\beta \Delta \Psi}  \rangle_0 \nonumber \\
+{\rm \ }  &=& \Delta G^\circ_{b,{\rm ideal}} + \Delta G_{b,{\rm exc.}}
+\end{eqnarray}
+
+The term $\Delta G^\circ_{b,{\rm ideal}} = - k_B T \ln C^\circ V_{\rm site}$ in Eq.~(\ref{eq:DG0-def-Vsite}) is interpreted as the \emph{ideal} component of the standard binding free energy, that is the binding free energy that would be measured in the absence of ligand-receptor interactions ($\Delta \Psi = 0$) and when the ligand's concentration is $C^\circ$.
+The ideal term is responsible for the dependence of the standard binding free energy on the standard state concentration.
+The arguably inappropriate term ``standard state correction'' is sometimes used in the literature to refer to $\Delta G^\circ_{b,{\rm ideal}}$.
+As seen above, the ideal term is a fundamental aspect of the statistical mechanics theory of molecular binding rather than a mere correction to an inexact formulation.
+
+Conversely, the term $\Delta G_{b,{\rm exc.}} = - k_B T \ln \langle e^{-\beta \Delta \Psi}  \rangle_0 $ is the \emph{excess} component of the standard free energy of binding, which is the term that depends on intermolecular interactions and the specific chemical composition of the system.
+The excess component is the target of the alchemical molecular simulations discussed here, in contrast to the ideal term that is generally evaluated analytically~\cite{boresch2003absolute}.
+
+\paragraph{Handling the standard state in free energy calculations.}
+As discussed in Sections \ref{sec:theory-solvation} and \ref{sec:theory-partitioning}, the handling of the standard state in solvation and solvent partitioning free energy calculations is typically straightforward.
+Eq.~(\ref{eq:DG0-solvation-def}) identifies the ideal and excess components of the standard solvation free energy in terms of the standard gas pressure $P^\circ$, standard solution concentration $C^\circ$, and ensemble averages typically evaluated by alchemical free energy simulations.
+These relations provide simple recipes to relate computational estimates to experimental values obtained under different conditions.
+Moreover, typical experimental reporters (Henry's constants and partitioning coefficients) directly probe the excess term calculated by molecular simulations, which is independent of the standard state.
+
+As discussed above, the situation for binding is more complex and requires special care.
+The first important aspect is that binding free energy calculations must be performed with binding site restraints that implement the binding site indicator function $I(\zeta)$ by forcing the ligand to occupy the binding site even when it is decoupled from the receptor.
+Doing otherwise would fail to specify the bound macrostate of the complex and return undefined binding free energy estimates.
+However, in practice, calculations lacking binding site restraints often return reliable binding free energy estimates due to the kinetic trapping of the ligand in the binding site region during the relatively short molecular dynamics sampling typical of alchemical binding free energy simulations.
+Nevertheless, longer calculations could display persistent drift of the estimate as new regions of conformational space are sampled.
+A more serious consequence of neglecting binding site restraints is the possibility of the ligand leaving the binding site when it is not coupled to the receptor.
+Such occurrence indicates a serious violation of the theory.
+Alchemical binding free energy simulations where the ligand leaves the binding site are not set up correctly and can return grossly inaccurate binding free energy estimates.
+
+Because it is a discontinuous function of the position and orientation of the ligand relative to the receptor, molecular dynamics implementations often approximate the indicator function by means of flat-bottom restraints~\cite{chen2007can} (see below).
+Harmonic restraints are not as suitable in this respect because they favor a specific microstate at the bottom of the harmonic potential rather than specifying a region of conformational space.
+
+Note that $\Delta G^\circ_{b,{\rm ideal}}$ is independent of the chemical nature of the binding partners, and it is the same for all complexes sharing similar binding site definitions.
+The latter is the reason that the ideal term cancels out when taking the difference of absolute binding free energies when estimating relative binding free energies (RBFEs).
+However, while it does not appear explicitly in the definition, $\Delta G_{b,{\rm exc.}}$ depends implicitly on the definition of the binding site through the ensemble average in the uncoupled ensemble, which includes only ``bound'' configurations according to the definition of the indicator function $I(\zeta)$.
+Varying the binding site definition affects both the ideal and excess binding free energy terms such that their sum (the standard binding free energy) is often only weakly dependent on the binding site definition~\cite{gilson1997statisticalthermodynamic,gallicchio2011recent}.
+Conversely, great care should be exercised when comparing calculated excess binding free energy values corresponding to different binding site definitions.
+Similarly, the outcome of a binding free energy calculation conducted without a binding site definition, i.e.\ without binding site restraints, is, in principle, undefined.
+The latter is also true for relative binding free energies because while the ideal term cancels out, the difference in excess binding free energies depends on the binding pose definition.
+
+\paragraph{Several choices of restraints are possible in binding free energy calculations.}
+A variety of restraint types are commonly used in alchemical binding free energy calculations.
+Before going into their characteristics, we must acknowledge an important and often unrecognized distinction between the binding site restraints corresponding to the indicator function $I(\zeta)$ necessary for the implementation of the theory and optional auxiliary restraints and biasing potentials sometimes used to speed up the convergence of free energy calculations.
+Binding site restraints are an integral part of the molecular complex definition and are typically maintained throughout the alchemical process.
+In particular, they must be present unchanged at the unbound and bound states.
+In contrast, unless thermodynamic reweighting is applied in post-processing, auxiliary restraints cannot be present at the endstates to avoid affecting their free energies.
+Hence, typically the free energy cost of turning them on is computed at the uncoupled state, and the free energy gain of releasing them is calculated near the bound state.
+In the alchemical binding free energy literature, often harmonic restraint potentials imposed in an initial stage and removed in a final stage are described as implementing binding sites restraints.
+However, because they utilize harmonic potentials that cannot correctly describe the binding pose macrostate (see above) and because they are turned on and then off, these are instead auxiliary restraints introduced to enhance convergence without affecting, in principle, binding free energy estimates.
+
+Binding site restraints implement the $I(\zeta)$ indicator function.
+They, therefore, can depend at most on the six rigid-body external coordinates of the ligand governing the position and orientation of the ligand relative to the receptor.
+Any other choice would affect the intramolecular conformational distribution of the ligand or the receptor and would  bias the binding free energy estimate.
+The external degrees of freedom introduced by Boresch et al.~\cite{boresch2003absolute,leitgeb2005alchemical} are a popular choice.
+The Boresch coordinates, as well as other similar approaches, involve selecting anchor atoms of the receptor and the ligand.
+In principle, the choice of anchor atoms is arbitrary, provided adequate simulation time is available, and as long as it leads to similar binding pose definitions.
+However, practical considerations may be important.
+For example, some choices of Boresch-style anchor atoms can be numerically unstable.
+Additionally, anchor atoms should likely be placed in a part of the molecule that defines the binding orientation well rather than in a floppy, solvent-exposed tail.
+Several automated methods have been proposed to select anchor points for Boresch restraints, which typically involve analysing a simulation of the fully-interacting receptor-ligand complex~\cite{alibay2022evaluating, alibay2021mdrestraintsgenerator, baumann2023broadening, chen2023enhancing, hedges2023suite}.
+
+As discussed above, binding site restraints are commonly implemented using quadratic flat-bottom restraints, which impose a restoring harmonic  force similar to harmonic potentials but only if the ligand leaves a specific region~\cite{chen2007can}.
+Often, binding site restraints do not limit the rotational degrees of freedom of the ligand, and the corresponding standard state term is zero; see Eq.~(\ref{eq:DG0-def-Vsite}).
+However, it might be challenging to fully explore the bound macrostate at all alchemical states when it is only limited by receptor-ligand positional restraints.
+As further discussed below, auxiliary restraints are particularly helpful in the latter circumstance to limit the extent of conformational sampling necessary to achieve rapid convergence of binding free energy estimates.
+
+Most current binding free energy workflows do not use binding site restraints present at all stages of the simulation and implicitly assume that $V_{\rm site}$ in Eq.~(\ref{eq:Vsite-def}) is the volume of the simulation box $V$.
+In the absence of binding site restraints, only the protein-ligand interactions prevent the ligand from slipping out from the binding site and wandering throughout the simulation box within the simulation time frame.
+However, observations of ligands leaving the binding site become more likely as simulation times increase, especially when analyzing weak binders and non-binders (see below).
+Moreover, there is no obstacle for the ligand to diffuse away from the binding site at the decoupled state when protein-ligand interactions are turned off.
+
+In practice, because they often implement auxiliary restraints, binding free energy prediction workflows that omit binding site restraints rarely yield grossly erroneous predictions.
+Auxiliary restraints are introduced gradually as protein-ligand interactions are turned off or are introduced at the bound state when the ligand is naturally trapped in the binding site.
+Either way, the free energy of imposing them can be tracked while preventing the ligand from leaving the binding site at the decoupled state.
+Provided that the interacting ligand samples only bound configurations, i.e., it behaves as if binding site restraints were present, the current practice of not using them is likely to incur only minor or insignificant errors.
+
+The Boresch harmonic restraints~\cite{boresch2003absolute,baumann2023broadening} are a popular choice as auxiliary restraints because they are based on interatomic distances, bond angles, and dihedral angles degrees of freedom already implemented in molecular mechanics codes that do not interfere with intramolecular conformational distributions.
+Moreover, the free energy of releasing Boresch restraints within the simulation box volume $V$ is easily computed analytically~\cite{boresch2003absolute}.
+
+By reducing the conformational space to be sampled, auxiliary restraints are often crucial for obtaining converged binding free energy estimates, even when binding site restraints are employed.
+Auxiliary restraints take a variety of forms, such as distance restraints between the ligand and the protein~\cite{mobley2006use,clark2023comparison} as well as flat-bottom restraints.
+The choice of degrees of freedom and restraining potentials for auxiliary restraints is not as restrictive as for those for binding site restraints, provided that the free energy terms of restraining collective variables that couple protein and ligand degrees of freedom are adequately accounted for to avoid serious systematic errors~\cite{clark2023comparison}.
+For example, the overall ligand RMSD has been used~\cite{woo2005calculation} for this purpose.
+Auxiliary restraints are designed to reduce the amount of conformational sampling at intermediate $\vec{\lambda}$ values.
+However, more computational effort is required to compute the free energy when the restraints are turned on and off.
+Additionally, such restraints can limit the exploration of alternative binding modes.
 This restriction may be undesirable when using Hamiltonian $\lambda$ exchange or expanded ensemble techniques where allowing the ligand to exchange binding modes when it is non-interacting could provide sampling benefits~\cite{wang2013identifying}.
-More specifically, flat-bottom restraints might allow a ligand to explore multiple binding sites, and harmonic restraints allow exploration of multiple binding modes within a site, while Boresch restraints only allow a single binding mode within a single site.
+More specifically, flat-bottom restraints allow a ligand to explore multiple binding sites, and harmonic restraints allow the exploration of multiple binding modes within a site, while Boresch restraints only allow a single binding mode within a single site.
 See additional discussion of the possibility of multiple binding modes in Sec.~\ref{sec:multiple_binding_modes} below.
 
-Many choices of restraints involve selecting reference atoms.
-Again, in principle this choice is unimportant given adequate simulation time but practical considerations may be important.
-The choice is likely especially important with Boresch-style restraints, where some relative placements of reference atoms are likely to be numerically unstable; additionally, ligand reference atoms should likely be in a part of the molecule which defines the binding orientation well, rather than in a floppy solvent-exposed tail, for example.
-Several automated methods have been proposed to select anchor points for Boresch restraints, which typically involve analysing a simulation of the fully-interacting receptor-ligand complex~\cite{alibay2022evaluating, alibay2021mdrestraintsgenerator, baumann2023broadening, chen2023enhancing, hedges2023suite}.
-
-More recently, restrictive restraint schemes have been proposed that avoid relying on a few critical reference atoms.
-Instead, these aim to simplify restraint selection by using larger sets of atomic positions in the receptor and ligand.
+Recently, degrees of freedom for binding site and auxiliary potentials based on larger sets of atomic positions of the receptor and ligand have been put forward.
 Fu et al.
-suggested restraining the six relative rigid-body degrees of freedom derived by finding the rotation of the ligand which minimises RMSD to a reference structure of the receptor-ligand complex (after accounting for rotation and translation of the receptor)~\cite{fu2017new}.
+suggested restraining the six relative rigid-body degrees of freedom derived by finding the rotation of the ligand, which minimises the RMSD to a reference structure of the receptor-ligand complex (after accounting for rotation and translation of the receptor)~\cite{fu2017new}.
 Salari et al.
 proposed restraining the ``distance-from-bound-configuration'' (DBC) variable, which is the RMSD of ligand coordinates in the frame of reference of the binding site~\cite{salari2018streamlined, ebrahimi2022symmetry}.
-This may allow the binding pose to be closely preserved as the ligand intermolecular interactions are removed, but has the disadvantage of coupling the internal and relative external degrees of freedom of the receptor and ligand, preventing easy computation of the standard state correction.
-This necessitates an additional step where the DBC restraints are released to a single harmonic restraint.
-Both restraint schemes are implemented in open-source workflows~\cite{fu2021bfee2, fu2022accurate, santiago-mcrae2023computing}.
+DBC may allow the binding pose to be closely preserved as the ligand intermolecular interactions are removed, but has the disadvantage of coupling the internal and relative external degrees of freedom of the receptor and ligand, making them unsuitable for describing the binding site macrostate and the corresponding standard state terms.
+However, these restraints are practical as auxiliary restraints where additional steps are used to introduce and release the DBC restraints.
+These restraint schemes are implemented in open-source workflows~\cite{fu2021bfee2, fu2022accurate, santiago-mcrae2023computing, clark2023comparison}.
 
 \subsection{Absolute and relative calculations deal with some of the same issues}
-\subsubsection{Handling weak binders and high dissociation rates}\label{sec:weak-binders}
-In binding free energy calculations, only the conformations in which the receptor and ligand form a bound complex should be sampled from the bound states (Sec.~\ref{sec:theory}).
-Determining what the bound states actually are can be challenging for weakly bound ligands.
-For tightly bound ligands, virtually all reasonable definitions of the bound state will lead to be equivalent free energies, since the partition function will be dominated by a relatively small number of low-energy poses.
-
-For weak binders, this simplification breaks down.
-In fact, the correct bound state may depend on the type of experiment performed.
-For example, isothermal titration calorimetry (ITC) or surface plasmon resonance (SPR) measurements effectively define a binding state that includes all ligand comformations that are complexed with the protein, regardless of where on the protein they bind.
+
+\paragraph{Handling weak binders}\label{sec:weak-binders}
+According to the non-covalent statistical mechanics theory of bimolecular binding (Sec.~\ref{sec:theory}), only the conformations in which the receptor and ligand form a bound complex are sampled.
+Determining an appropriate definition of the bound macrostate can be challenging for weakly bound ligands.
+For tightly bound ligands, virtually all reasonable definitions of the bound state will lead to equivalent free energies since the partition function will be dominated by a relatively limited ensemble of low-energy poses.
+However, this simplification breaks down for weak binders.
+In fact, there might not be a perfect correspondence between the structural definition of the complex on which current alchemical models are based~\cite{gilson1997statisticalthermodynamic} (Sec.~\ref{sec:theory}) and experimental reporters~\cite{mihailescu2004theory}.
+For example, isothermal titration calorimetry (ITC) or surface plasmon resonance (SPR) measurements effectively define a binding state that encompasses all ligand conformations complexed with the protein, regardless of where they bind the protein.
 In contrast, fluorescence polarization competition assays measure binding to only a single location, where the ligand of interest displaces a competing binder.
@@
 
-In absolute calculations, this need to explicitly define a binding site applies to the fully interacting state in the complex leg of the thermodynamic cycle (top-right state in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}), while in relative calculations the binding site must be defined at both end states of the complex leg (top- and bottom-right states in Fig.~\ref{fig:fig_binding_thermodynamic_cycle}).
-In principle, this requires defining which conformations are considered to be bound before running the calculation, but it is common practice to start the simulation with the ligand already placed in the binding site and rely on kinetic trapping to maintain the bound complex.
-
-However, this strategy of using kinetic trapping to maintain the bound complex can fail when the dissociation rate of the ligand has the same or smaller order of magnitude than the length of the simulation.
-This is typical of weak binders such as fragments binding shallow pockets with $\mu$M-mM affinities~\cite{georgiou2017pushing,pan2017quantitative}.
-In the case of weak binders, using a flat-bottom or harmonic restraint between receptor and ligand in the bound state(s) can prevent dissociations~\cite{georgiou2017pushing,rizzi2019sampl6}.
-We stress that this type of restraint is normally avoided as it generally introduces bias in the free energy estimate, which is why the restraint is usually activated only in the intermediate states in absolute calculations.
-The bias can be corrected through reweighting schemes~\cite{rizzi2019sampl6}, but this post-processing step can be avoided if a flat-bottom restraint is used and the ligand is never sampled while hitting the potential wall during the simulation in the bound state as the numerical correction will be exactly zero.
-It is important to note that the spring constant and/or radius parameters of the restraint effectively determine which conformations are considered to be bound.
-As a consequence, these parameters must be tuned to the system so that only the binding site is accessible to the ligand.
-Again, this step is particularly important for weak binders as their free energy of binding is known to be more sensitive to the definition of the binding site~\cite{gilson1997statisticalthermodynamic}.
-
-In absolute calculations, this restraint can substitute or be added to the restraint used to handle the standard state correction (Sec.~\ref{sec:standardstate-restraints}).
-In the latter case, however, care must be taken when computing the standard state correction.
-When multiple restraining potentials are active in the non-interacting state, the correction can generally be computed only through numerical integration.
-Alternatively, one can adopt a protocol that removes the bound-state restraint in the non-interacting state.
-Finally, even for tight binders, dissociation events can be enhanced by methods such as Hamiltonian replica exchange~\cite{sugita2000multidimensional,chodera2011replica,wang2013identifying} and expanded ensemble~\cite{lyubartsev1992new,li2007simulated}, especially in absolute free energy calculations using harmonic or flat-bottom restraints.
-In the latter case, dissociations can be averted simply by increasing the spring constant and/or reducing the radius of the restraint potential to prevent the exploration of ligand conformations outside the binding site in the decoupled state (bottom-right state in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}) that could be propagated to the bound state.
+Apart from the difficulty of selecting an appropriate definition of the binding site region that relates to the experimental technique, the alchemical theory and the computational procedures are the same whether strong, weak, or non-binders are considered.
+The alchemical binding free energy calculation will report the binding free energy for the selected binding pose specified, for example, by the region allowed by flat-bottom binding site restraints.
+Changing the definition of the binding pose will yield a binding free energy for the new pose, which will generally differ from the first.
+
+Especially when simulating non-binders that are not tightly associated with the receptor, it is likely to observe the ligand exploring the boundaries of the allowed binding site region.
+In such occurrences, which can also occur when the allowed binding site region is too restrictive, and the ligand is attempting to access a lower free energy pose or when the ligand is initially docked in a high energy pose, it can be expected that the choice of the binding site restraints will have a significant effect on the binding free energy estimate.
+However, in practice, alchemical binding free energy models are typically employed to identify binders and optimize the affinity of strong binders.
+Hence, it is often not critical to quantitatively pinpoint the binding free energies of weak and non-binders, as well as their dependence on the binding site definition.
 

### 2025-06-22 paper/manuscript.tex

@@
 \paragraph{Atom mapping}
-Once a particular approach to the topology is selected, a crucial next step is to identify the common atoms which will not be perturbed.
-Rigorously, this process typically comprises a MCSS search of the molecules involved to identify the common substructure---though the parameters of the MCSS search will differ depending on whether single or dual topology calculations are planned.
-Specifically, with a single topology approach in mind, atom types are allowed to change, so a permissive MCSS search can be done, whereas with dual topology a more strict search is required.
+Following the choice of system representation and sampling strategy, it might be required to generate a atom mapping.
+This is the case especially for system representations using a single coordinate set (Type 1.
+\& 2), however a dual coordinate set approach could use an atom mapping as well, depending on the sampling method, this will be explained in the next steps.
+Once a single coordinate set approach is selected is selected, a crucial next step is to identify the common atoms which will not be perturbed between the two molecules (respectivly end-states).
+Rigorously, this process typically comprises a MCSS search of the molecules involved to identify the common substructure.
+Specifically, with a single coordinate set approach in mind, atom types are allowed to change, so a permissive MCSS search can be done.
+This however requires either careful implementation of atom mass changes, bond length etc.
+or accepting the introduction of potential systematic error, that is in many cases negliable due to the cancelation of errors.
+\cite{ries2024Kartograf}
+
 \begin{figure}

### 2025-06-25 paper/manuscript.tex
additional citations

@@
 
-Because it is a discontinuous function of the position and orientation of the ligand relative to the receptor, molecular dynamics implementations often approximate the indicator function by means of flat-bottom restraints~\cite{chen2007can} (see below).
+Because it is a discontinuous function of the position and orientation of the ligand relative to the receptor, molecular dynamics implementations often approximate the indicator function by means of flat-bottom restraints~\cite{chen2007can, wu2021alchemical} (see below).
 Harmonic restraints are not as suitable in this respect because they favor a specific microstate at the bottom of the harmonic potential rather than specifying a region of conformational space.
@@
 Additionally, anchor atoms should likely be placed in a part of the molecule that defines the binding orientation well rather than in a floppy, solvent-exposed tail.
-Several automated methods have been proposed to select anchor points for Boresch restraints, which typically involve analysing a simulation of the fully-interacting receptor-ligand complex~\cite{alibay2022evaluating, alibay2021mdrestraintsgenerator, baumann2023broadening, chen2023enhancing, hedges2023suite}.
+Several automated methods have been proposed to select anchor points for Boresch restraints, which typically involve analysing a simulation of the fully-interacting receptor-ligand complex~\cite{alibay2022evaluating, alibay2021mdrestraintsgenerator, baumann2023broadening, chen2023enhancing, hedges2023suite, wu2025optimizing}.
 
@@
 The Boresch harmonic restraints~\cite{boresch2003absolute,baumann2023broadening} are a popular choice as auxiliary restraints because they are based on interatomic distances, bond angles, and dihedral angles degrees of freedom already implemented in molecular mechanics codes that do not interfere with intramolecular conformational distributions.
-Moreover, the free energy of releasing Boresch restraints within the simulation box volume $V$ is easily computed analytically~\cite{boresch2003absolute}.
+Moreover, the free energy of releasing Boresch restraints within the simulation box volume $V$ is easily computed analytically~\cite{boresch2003absolute, boresch2024analytical}.
 
@@
 DBC may allow the binding pose to be closely preserved as the ligand intermolecular interactions are removed, but has the disadvantage of coupling the internal and relative external degrees of freedom of the receptor and ligand, making them unsuitable for describing the binding site macrostate and the corresponding standard state terms.
-However, these restraints are practical as auxiliary restraints where additional steps are used to introduce and release the DBC restraints.
-These restraint schemes are implemented in open-source workflows~\cite{fu2021bfee2, fu2022accurate, santiago-mcrae2023computing, clark2023comparison}.
+However, these restraints are practical as auxiliary restraints when additional steps are used to introduce and release them.
+Moreover, because they are based on flat-bottom potentials encompassing the desired binding poses, the free energy of releasing them in the bound state can be safely neglected in most cases.
+DBC restraint schemes are implemented in open-source workflows~\cite{fu2021bfee2, fu2022accurate, santiago-mcrae2023computing, clark2023comparison, hedges2023suite}.
 

### 2025-06-27 paper/manuscript.tex
clarify statement about restraints and open-source workflows

@@
 Moreover, because they are based on flat-bottom potentials encompassing the desired binding poses, the free energy of releasing them in the bound state can be safely neglected in most cases.
-DBC restraint schemes are implemented in open-source workflows~\cite{fu2021bfee2, fu2022accurate, santiago-mcrae2023computing, clark2023comparison, hedges2023suite}.
+The restraint schemes discussed here are implemented in open-source workflows~\cite{fu2021bfee2, fu2022accurate, santiago-mcrae2023computing, clark2023comparison, hedges2023suite}.
 

### 2025-06-30 paper/manuscript.tex

--- 2025-06-22 paper/manuscript.tex
@@
 %%% PREAMBLE
-\documentclass[9pt,bestpractices]{livecoms}
+\documentclass[9pt,bestpractices,pubversion]{livecoms}
 % Use the 'onehalfspacing' option for 1.5 line spacing
@@
 This is the case especially for system representations using a single coordinate set (Type 1.
-\& 2), however a dual coordinate set approach could use an atom mapping as well, depending on the sampling method, this will be explained in the next steps.
+& 2), however a dual coordinate set approach could use an atom mapping as well, depending on the sampling method, this will be explained in the next steps.
 Once a single coordinate set approach is selected is selected, a crucial next step is to identify the common atoms which will not be perturbed between the two molecules (respectivly end-states).
@@
 
-An absolute binding free energy calculation yields the standard free energy of formation of the molecular complex from the separated receptor and ligand species.
-Hence, as discussed in \ref{sec:non-covalent-molecular-binding}, it must depend on the standard state concentration $C^\circ$ (here $1 \, {\rm M} = 1 \, {\rm molecule}/1660$~\r{A}$^3$) and on the definition of the complex, that is on the specific binding pose for which the binding free energy is requested.
-Omitting the standard state concentration would yield an undefined result that would vary with the choice of concentration units.
-A similar vague estimate would result from omitting the binding pose definition that, for example, could equally describe binding to a specific site of the receptor or unspecific binding throughout the receptor surface depending on the extent of conformational sampling (see Sec.~\ref{sec:weak-binders}).
-
-In the statistical mechanics formulation of Gilson et al.~\cite{gilson1997statisticalthermodynamic}, described in Section \ref{sec:non-covalent-molecular-binding}, the standard state concentration appears explicitly in Eq.~(\ref{eq:DG0-binding-constant-def}) and the binding pose is defined by the indicator function $I(\zeta)$ that characterizes the partition function of the complex; Eq.~(\ref{eq:intra-zRL-def}).
-The role of the standard state and binding site definitions are particularly evident when expressing Eq.~(\ref{eq:DG0-binding-constant-def}) in terms of an ensemble average.
-To do so, we express the integrals in the numerator and denominator of Eq.~(\ref{eq:DG0-binding-constant-def}) over the same set of degrees of freedom.
-Note that the intramolecular configurational partition function of the complex, $z_{RL(v)}$, includes six more degrees of freedom--the position and orientation of the ligand relative to the receptor collectively described by the variable $\zeta$--than the product of the intramolecular partition functions, $z_{R(v)} z_{L(v)}$, of the separated receptor and ligand.
-To correct this discrepancy, we multiply and divide the right-hand side of Eq.~(\ref{eq:DG0-binding-constant-def})  by the quantity
-\begin{equation} \label{eq:Vsite-def}
-    V_{\rm site} \Omega = \int d\zeta I(\zeta) \, ,
-\end{equation}
-%\Omega is in my opinion a better symbol of the orientational extent than \xi_L. Lowercase greek letters are usually used
-%for dynamical variables rather than integrated quantities
-which measures the extent of the binding pose macrostate, where $V_{\rm site}$ corresponds to the spatial extent and $\Omega$ the orientational extent.
-The right-hand side of Eq.~(\ref{eq:Vsite-def}) combined with the integrals $z_{R(v)}$ and $z_{L(v)}$ yields a partition function similar to Eq.~(\ref{eq:intra-zRL-def}) but corresponding to a complex in which ligand and receptor are uncoupled~\cite{gallicchio2011recent} as if they were at infinite distance apart.
-The result is
-\begin{equation}\label{eq:DG0-binding-constant-def-average}
-K_b = C^\circ V_{\rm site} \frac{\Omega}{8 \pi^2} \langle e^{-\beta \Delta \Psi}  \rangle_0 
-\end{equation}
-where, similarly to Eq.~(\ref{eq:zratio-solv-average}),
-\begin{equation}\label{eq:pert-energy-def}
-\Delta \Psi(x_R, x_L, \zeta) =  \Psi_v(x_R, x_L, \zeta) - \left[  \Psi_v(x_R) + \Psi(x_L) \right]
-\end{equation}
-is the solvent-averaged binding energy of the complex when receptor and ligands are in intramolecular configurations $x_R$ and $x_L$, the ligand is in position and orientation $\zeta$ relative to the receptor (the uncoupled potential energy does not depend on  the position and orientation of the ligand relative to the receptor), and $\langle \ldots \rangle_0$ represents an ensemble average in the uncoupled ensemble.
-Finally, note that $\Omega = 8 \pi^2$ when the binding pose indicator function does not depend on the orientation of the ligand and Eq.~(\ref{eq:DG0-binding-constant-def-average}) simplifies to
-\begin{equation}\label{eq:DG0-binding-constant-def-average-noOmega}
-K_b = C^\circ V_{\rm site} \langle e^{-\beta \Delta \Psi}  \rangle_0 
-\end{equation}
-which, expressed in free energy units, yields
-\begin{eqnarray}\label{eq:DG0-def-Vsite}
-  \Delta G^\circ_b &=&  - k_B T \ln K_b = - k_B T \ln C^\circ V_{\rm site} - k_B T \ln \langle e^{-\beta \Delta \Psi}  \rangle_0 \nonumber \\
-{\rm \ }  &=& \Delta G^\circ_{b,{\rm ideal}} + \Delta G_{b,{\rm exc.}}
-\end{eqnarray}
-
-The term $\Delta G^\circ_{b,{\rm ideal}} = - k_B T \ln C^\circ V_{\rm site}$ in Eq.~(\ref{eq:DG0-def-Vsite}) is interpreted as the \emph{ideal} component of the standard binding free energy, that is the binding free energy that would be measured in the absence of ligand-receptor interactions ($\Delta \Psi = 0$) and when the ligand's concentration is $C^\circ$.
-The ideal term is responsible for the dependence of the standard binding free energy on the standard state concentration.
-The arguably inappropriate term ``standard state correction'' is sometimes used in the literature to refer to $\Delta G^\circ_{b,{\rm ideal}}$.
-As seen above, the ideal term is a fundamental aspect of the statistical mechanics theory of molecular binding rather than a mere correction to an inexact formulation.
-
-Conversely, the term $\Delta G_{b,{\rm exc.}} = - k_B T \ln \langle e^{-\beta \Delta \Psi}  \rangle_0 $ is the \emph{excess} component of the standard free energy of binding, which is the term that depends on intermolecular interactions and the specific chemical composition of the system.
-The excess component is the target of the alchemical molecular simulations discussed here, in contrast to the ideal term that is generally evaluated analytically~\cite{boresch2003absolute}.
-
-\paragraph{Handling the standard state in free energy calculations.}
-As discussed in Sections \ref{sec:theory-solvation} and \ref{sec:theory-partitioning}, the handling of the standard state in solvation and solvent partitioning free energy calculations is typically straightforward.
-Eq.~(\ref{eq:DG0-solvation-def}) identifies the ideal and excess components of the standard solvation free energy in terms of the standard gas pressure $P^\circ$, standard solution concentration $C^\circ$, and ensemble averages typically evaluated by alchemical free energy simulations.
-These relations provide simple recipes to relate computational estimates to experimental values obtained under different conditions.
-Moreover, typical experimental reporters (Henry's constants and partitioning coefficients) directly probe the excess term calculated by molecular simulations, which is independent of the standard state.
-
-As discussed above, the situation for binding is more complex and requires special care.
-The first important aspect is that binding free energy calculations must be performed with binding site restraints that implement the binding site indicator function $I(\zeta)$ by forcing the ligand to occupy the binding site even when it is decoupled from the receptor.
-Doing otherwise would fail to specify the bound macrostate of the complex and return undefined binding free energy estimates.
-However, in practice, calculations lacking binding site restraints often return reliable binding free energy estimates due to the kinetic trapping of the ligand in the binding site region during the relatively short molecular dynamics sampling typical of alchemical binding free energy simulations.
-Nevertheless, longer calculations could display persistent drift of the estimate as new regions of conformational space are sampled.
-A more serious consequence of neglecting binding site restraints is the possibility of the ligand leaving the binding site when it is not coupled to the receptor.
-Such occurrence indicates a serious violation of the theory.
-Alchemical binding free energy simulations where the ligand leaves the binding site are not set up correctly and can return grossly inaccurate binding free energy estimates.
-
-Because it is a discontinuous function of the position and orientation of the ligand relative to the receptor, molecular dynamics implementations often approximate the indicator function by means of flat-bottom restraints~\cite{chen2007can} (see below).
-Harmonic restraints are not as suitable in this respect because they favor a specific microstate at the bottom of the harmonic potential rather than specifying a region of conformational space.
-
-Note that $\Delta G^\circ_{b,{\rm ideal}}$ is independent of the chemical nature of the binding partners, and it is the same for all complexes sharing similar binding site definitions.
-The latter is the reason that the ideal term cancels out when taking the difference of absolute binding free energies when estimating relative binding free energies (RBFEs).
-However, while it does not appear explicitly in the definition, $\Delta G_{b,{\rm exc.}}$ depends implicitly on the definition of the binding site through the ensemble average in the uncoupled ensemble, which includes only ``bound'' configurations according to the definition of the indicator function $I(\zeta)$.
-Varying the binding site definition affects both the ideal and excess binding free energy terms such that their sum (the standard binding free energy) is often only weakly dependent on the binding site definition~\cite{gilson1997statisticalthermodynamic,gallicchio2011recent}.
-Conversely, great care should be exercised when comparing calculated excess binding free energy values corresponding to different binding site definitions.
-Similarly, the outcome of a binding free energy calculation conducted without a binding site definition, i.e.\ without binding site restraints, is, in principle, undefined.
-The latter is also true for relative binding free energies because while the ideal term cancels out, the difference in excess binding free energies depends on the binding pose definition.
-
-\paragraph{Several choices of restraints are possible in binding free energy calculations.}
-A variety of restraint types are commonly used in alchemical binding free energy calculations.
-Before going into their characteristics, we must acknowledge an important and often unrecognized distinction between the binding site restraints corresponding to the indicator function $I(\zeta)$ necessary for the implementation of the theory and optional auxiliary restraints and biasing potentials sometimes used to speed up the convergence of free energy calculations.
-Binding site restraints are an integral part of the molecular complex definition and are typically maintained throughout the alchemical process.
-In particular, they must be present unchanged at the unbound and bound states.
-In contrast, unless thermodynamic reweighting is applied in post-processing, auxiliary restraints cannot be present at the endstates to avoid affecting their free energies.
-Hence, typically the free energy cost of turning them on is computed at the uncoupled state, and the free energy gain of releasing them is calculated near the bound state.
-In the alchemical binding free energy literature, often harmonic restraint potentials imposed in an initial stage and removed in a final stage are described as implementing binding sites restraints.
-However, because they utilize harmonic potentials that cannot correctly describe the binding pose macrostate (see above) and because they are turned on and then off, these are instead auxiliary restraints introduced to enhance convergence without affecting, in principle, binding free energy estimates.
-
-Binding site restraints implement the $I(\zeta)$ indicator function.
-They, therefore, can depend at most on the six rigid-body external coordinates of the ligand governing the position and orientation of the ligand relative to the receptor.
-Any other choice would affect the intramolecular conformational distribution of the ligand or the receptor and would  bias the binding free energy estimate.
-The external degrees of freedom introduced by Boresch et al.~\cite{boresch2003absolute,leitgeb2005alchemical} are a popular choice.
-The Boresch coordinates, as well as other similar approaches, involve selecting anchor atoms of the receptor and the ligand.
-In principle, the choice of anchor atoms is arbitrary, provided adequate simulation time is available, and as long as it leads to similar binding pose definitions.
-However, practical considerations may be important.
-For example, some choices of Boresch-style anchor atoms can be numerically unstable.
-Additionally, anchor atoms should likely be placed in a part of the molecule that defines the binding orientation well rather than in a floppy, solvent-exposed tail.
+Absolute free energy calculations involve completely removing the interactions between the ligand or solute and its environment, taking it to a non-interacting state that may or may not retain intramolecular non-bonded interactions.
+This non-interacting state can then be shifted between environments---from the protein to water, or from one solution to another---without changing its free energy other than that due to the changing volume of the simulations, and then interactions can be restored in the new environment.
+
+Absolute free energies are by definition reported with respect to a specific reference or standard state, which effectively determines the arbitrary point at which the free energy is 0.
+The role of the standard state is particularly evident from the expression of the binding free energy between a receptor $R$ and ligand $L$
+\begin{equation} \label{eq:DGfromKAB}
+    \Delta G = -k_BT ~ \ln \left( c^{\circ} K_b \right)  = -k_BT ~ \ln\left( c^{\circ} \frac{[RL]}{[L][R]} \right) \, .
+\end{equation}
+Here, the reference state concentration $c^{\circ}$ converts the binding constant $K_b$ into a dimensionless quantity expressed in reference concentration units.
+It should be noted that ignoring the term $c^{\circ}$ is equivalent to assuming a reference concentration of 1~D$^{-1}$, where D are the units used to express $K_b$, and would thus cause the value of $\Delta G$ to vary with the choice of the units.
+It is convenient to define a standard state at a constant pressure of 1~atm and where each chemical species (i.e., R, L, and RL) in the reaction solvent has a concentration of $c^{\circ}$~=~1~M~=~1~molecule/1660~\r{A}$^3$ but do not interact with other molecules of $R$, $L$, or $RL$.
+
+\paragraph{Handling the standard state in absolute free energy calculations.}
+For solvation free energy calculations, handling the standard state is typically straightforward, and treating it correctly simply means ensuring that the non-interacting solute still occupies essentially the same volume as the solute in the interacting system.
+So typically in such cases no special care is required to ensure the correct standard state, as long as the \emph{experimental} data being analyzed uses the same standard state.
+If this is not the case, a simple entropic correction to the free energy of $k_B T\ln (V_f/V_i) = k_B T \ln (C_i/C_f)$ to the experimental data is needed.
+
+For binding, however, the situation is more complex and requires special care.
+Because the simulations are typically performed using restraints and at concentrations that are different from 1 M, the expression of the free energy requires the following correction~\cite{gilson1997statisticalthermodynamic} (see an example of such a thermodynamic cycle in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle})
+\begin{equation}\label{eq:restraint-correction}
+    \Delta G^{\circ}_{\mathrm{restr}} = -k_BT ~ \ln \left( c^{\circ}{V_L} \right) -k_BT ~ \ln \left( \frac{\xi_L}{8 \pi^2} \right),
+\end{equation}
+where $V_L$ and $\xi_L$ are respectively the volume of the translational and rotational degrees of freedom of the non-interacting ligand in the simulation box.
+When no restraints are used, the non-interacting ligand is free to translate and rotate in the simulation box (i.e., $V_L = V_{\mathrm{box}}$ and $\xi_L = 8\pi^2$), and the rotational term is zero.
+A sufficiently thorough exploration of the simulation box by the non-interacting ligand is, however, required for the formula to be valid.
+This is typically hard to achieve as the exploration process is governed by diffusion, and weak transient nonspecific binding will occur at other sites on the protein. 
+The addition of a restraint limits the volume available to the non-interacting ligand, thus speeding the convergence of the sampling.
+In addition, when enhanced sampling methods such as Hamiltonian $\lambda$ exchange are used (see Sec.~\ref{sec:sampling_schemes}), the use of a restraint is typically necessary as it keeps the ligand in the binding site in the interacting state (see also Sec.~\ref{sec:weak-binders}) and generally reduce the round-trip time of replicas.
+When restraints are employed, the values of $V_L$ and $\xi_L$ are restraint-dependent, but for commonly employed restraints, these can be usually easily computed analytically or numerically by solving the relevant integral.
+
+\paragraph{Several choices of restraints are possible.}\label{ABFErest}
+In practice, a variety of types of restraints are common, from simple harmonic distance restraints between the ligand and the protein~\cite{mobley2006use}, to flat-bottom restraints which work similarly but only exert a force if the ligand leaves a specific region~\cite{chen2007can}.
+Because these restraints do not limit the rotational degrees of freedom of the ligand, the rotational term entering the correction in Eq.~\ref{eq:restraint-correction} is zero.
+
+Alternatively, a set of restraints proposed by Boresch have also commonly been employed, where all six rigid-body degrees of freedom governing the orientation of the ligand relative to the receptor are restrained~\cite{boresch2003absolute, leitgeb2005alchemical}.
+Further restraints, such as on the overall ligand RMSD have also been used~\cite{woo2005calculation}.
+
+In principle, all of these forms will yield correct binding free energies in the limit of adequate sampling if their effects and connection to the standard state are correctly handled, but they have different strengths and weaknesses.
+For example, with more involved restraints, sampling at intermediate $\vec{\lambda}$ values will usually not need to be as extensive but more computational effort must go to computing the free energy to turn on the restraints.
+Additionally, such restraints would typically keep the ligand from exploring alternative binding modes.
+This  restriction may be undesirable when using Hamiltonian $\lambda$ exchange or expanded ensemble techniques where allowing the ligand to exchange binding modes when it is non-interacting could provide sampling benefits~\cite{wang2013identifying}.
+More specifically, flat-bottom restraints might allow a ligand to explore multiple binding sites, and harmonic restraints allow exploration of multiple binding modes within a site, while Boresch restraints only allow a single binding mode within a single site.
+See additional discussion of the possibility of multiple binding modes in Sec.~\ref{sec:multiple_binding_modes} below.
+
+Many choices of restraints involve selecting reference atoms.
+Again, in principle this choice is unimportant given adequate simulation time but practical considerations may be important.
+The choice is likely especially important with Boresch-style restraints, where some relative placements of reference atoms are likely to be numerically unstable; additionally, ligand reference atoms should likely be in a part of the molecule which defines the binding orientation well, rather than in a floppy solvent-exposed tail, for example.
 Several automated methods have been proposed to select anchor points for Boresch restraints, which typically involve analysing a simulation of the fully-interacting receptor-ligand complex~\cite{alibay2022evaluating, alibay2021mdrestraintsgenerator, baumann2023broadening, chen2023enhancing, hedges2023suite}.
 
-As discussed above, binding site restraints are commonly implemented using quadratic flat-bottom restraints, which impose a restoring harmonic  force similar to harmonic potentials but only if the ligand leaves a specific region~\cite{chen2007can}.
-Often, binding site restraints do not limit the rotational degrees of freedom of the ligand, and the corresponding standard state term is zero; see Eq.~(\ref{eq:DG0-def-Vsite}).
-However, it might be challenging to fully explore the bound macrostate at all alchemical states when it is only limited by receptor-ligand positional restraints.
-As further discussed below, auxiliary restraints are particularly helpful in the latter circumstance to limit the extent of conformational sampling necessary to achieve rapid convergence of binding free energy estimates.
-
-Most current binding free energy workflows do not use binding site restraints present at all stages of the simulation and implicitly assume that $V_{\rm site}$ in Eq.~(\ref{eq:Vsite-def}) is the volume of the simulation box $V$.
-In the absence of binding site restraints, only the protein-ligand interactions prevent the ligand from slipping out from the binding site and wandering throughout the simulation box within the simulation time frame.
-However, observations of ligands leaving the binding site become more likely as simulation times increase, especially when analyzing weak binders and non-binders (see below).
-Moreover, there is no obstacle for the ligand to diffuse away from the binding site at the decoupled state when protein-ligand interactions are turned off.
-
-In practice, because they often implement auxiliary restraints, binding free energy prediction workflows that omit binding site restraints rarely yield grossly erroneous predictions.
-Auxiliary restraints are introduced gradually as protein-ligand interactions are turned off or are introduced at the bound state when the ligand is naturally trapped in the binding site.
-Either way, the free energy of imposing them can be tracked while preventing the ligand from leaving the binding site at the decoupled state.
-Provided that the interacting ligand samples only bound configurations, i.e., it behaves as if binding site restraints were present, the current practice of not using them is likely to incur only minor or insignificant errors.
-
-The Boresch harmonic restraints~\cite{boresch2003absolute,baumann2023broadening} are a popular choice as auxiliary restraints because they are based on interatomic distances, bond angles, and dihedral angles degrees of freedom already implemented in molecular mechanics codes that do not interfere with intramolecular conformational distributions.
-Moreover, the free energy of releasing Boresch restraints within the simulation box volume $V$ is easily computed analytically~\cite{boresch2003absolute}.
-
-By reducing the conformational space to be sampled, auxiliary restraints are often crucial for obtaining converged binding free energy estimates, even when binding site restraints are employed.
-Auxiliary restraints take a variety of forms, such as distance restraints between the ligand and the protein~\cite{mobley2006use,clark2023comparison} as well as flat-bottom restraints.
-The choice of degrees of freedom and restraining potentials for auxiliary restraints is not as restrictive as for those for binding site restraints, provided that the free energy terms of restraining collective variables that couple protein and ligand degrees of freedom are adequately accounted for to avoid serious systematic errors~\cite{clark2023comparison}.
-For example, the overall ligand RMSD has been used~\cite{woo2005calculation} for this purpose.
-Auxiliary restraints are designed to reduce the amount of conformational sampling at intermediate $\vec{\lambda}$ values.
-However, more computational effort is required to compute the free energy when the restraints are turned on and off.
-Additionally, such restraints can limit the exploration of alternative binding modes.
-This restriction may be undesirable when using Hamiltonian $\lambda$ exchange or expanded ensemble techniques where allowing the ligand to exchange binding modes when it is non-interacting could provide sampling benefits~\cite{wang2013identifying}.
-More specifically, flat-bottom restraints allow a ligand to explore multiple binding sites, and harmonic restraints allow the exploration of multiple binding modes within a site, while Boresch restraints only allow a single binding mode within a single site.
-See additional discussion of the possibility of multiple binding modes in Sec.~\ref{sec:multiple_binding_modes} below.
-
-Recently, degrees of freedom for binding site and auxiliary potentials based on larger sets of atomic positions of the receptor and ligand have been put forward.
+More recently, restrictive restraint schemes have been proposed that avoid relying on a few critical reference atoms.
+Instead, these aim to simplify restraint selection by using larger sets of atomic positions in the receptor and ligand.
 Fu et al.
-suggested restraining the six relative rigid-body degrees of freedom derived by finding the rotation of the ligand, which minimises the RMSD to a reference structure of the receptor-ligand complex (after accounting for rotation and translation of the receptor)~\cite{fu2017new}.
+suggested restraining the six relative rigid-body degrees of freedom derived by finding the rotation of the ligand which minimises RMSD to a reference structure of the receptor-ligand complex (after accounting for rotation and translation of the receptor)~\cite{fu2017new}.
 Salari et al.
 proposed restraining the ``distance-from-bound-configuration'' (DBC) variable, which is the RMSD of ligand coordinates in the frame of reference of the binding site~\cite{salari2018streamlined, ebrahimi2022symmetry}.
-DBC may allow the binding pose to be closely preserved as the ligand intermolecular interactions are removed, but has the disadvantage of coupling the internal and relative external degrees of freedom of the receptor and ligand, making them unsuitable for describing the binding site macrostate and the corresponding standard state terms.
-However, these restraints are practical as auxiliary restraints where additional steps are used to introduce and release the DBC restraints.
-These restraint schemes are implemented in open-source workflows~\cite{fu2021bfee2, fu2022accurate, santiago-mcrae2023computing, clark2023comparison}.
+This may allow the binding pose to be closely preserved as the ligand intermolecular interactions are removed, but has the disadvantage of coupling the internal and relative external degrees of freedom of the receptor and ligand, preventing easy computation of the standard state correction.
+This necessitates an additional step where the DBC restraints are released to a single harmonic restraint.
+Both restraint schemes are implemented in open-source workflows~\cite{fu2021bfee2, fu2022accurate, santiago-mcrae2023computing}.
 
 \subsection{Absolute and relative calculations deal with some of the same issues}
-
-\paragraph{Handling weak binders}\label{sec:weak-binders}
-According to the non-covalent statistical mechanics theory of bimolecular binding (Sec.~\ref{sec:theory}), only the conformations in which the receptor and ligand form a bound complex are sampled.
-Determining an appropriate definition of the bound macrostate can be challenging for weakly bound ligands.
-For tightly bound ligands, virtually all reasonable definitions of the bound state will lead to equivalent free energies since the partition function will be dominated by a relatively limited ensemble of low-energy poses.
-However, this simplification breaks down for weak binders.
-In fact, there might not be a perfect correspondence between the structural definition of the complex on which current alchemical models are based~\cite{gilson1997statisticalthermodynamic} (Sec.~\ref{sec:theory}) and experimental reporters~\cite{mihailescu2004theory}.
-For example, isothermal titration calorimetry (ITC) or surface plasmon resonance (SPR) measurements effectively define a binding state that encompasses all ligand conformations complexed with the protein, regardless of where they bind the protein.
+\subsubsection{Handling weak binders and high dissociation rates}\label{sec:weak-binders}
+In binding free energy calculations, only the conformations in which the receptor and ligand form a bound complex should be sampled from the bound states (Sec.~\ref{sec:theory}).
+Determining what the bound states actually are can be challenging for weakly bound ligands.
+For tightly bound ligands, virtually all reasonable definitions of the bound state will lead to be equivalent free energies, since the partition function will be dominated by a relatively small number of low-energy poses.
+
+For weak binders, this simplification breaks down.
+In fact, the correct bound state may depend on the type of experiment performed.
+For example, isothermal titration calorimetry (ITC) or surface plasmon resonance (SPR) measurements effectively define a binding state that includes all ligand comformations that are complexed with the protein, regardless of where on the protein they bind.
 In contrast, fluorescence polarization competition assays measure binding to only a single location, where the ligand of interest displaces a competing binder.
@@
 
-Apart from the difficulty of selecting an appropriate definition of the binding site region that relates to the experimental technique, the alchemical theory and the computational procedures are the same whether strong, weak, or non-binders are considered.
-The alchemical binding free energy calculation will report the binding free energy for the selected binding pose specified, for example, by the region allowed by flat-bottom binding site restraints.
-Changing the definition of the binding pose will yield a binding free energy for the new pose, which will generally differ from the first.
-
-Especially when simulating non-binders that are not tightly associated with the receptor, it is likely to observe the ligand exploring the boundaries of the allowed binding site region.
-In such occurrences, which can also occur when the allowed binding site region is too restrictive, and the ligand is attempting to access a lower free energy pose or when the ligand is initially docked in a high energy pose, it can be expected that the choice of the binding site restraints will have a significant effect on the binding free energy estimate.
-However, in practice, alchemical binding free energy models are typically employed to identify binders and optimize the affinity of strong binders.
-Hence, it is often not critical to quantitatively pinpoint the binding free energies of weak and non-binders, as well as their dependence on the binding site definition.
+In absolute calculations, this need to explicitly define a binding site applies to the fully interacting state in the complex leg of the thermodynamic cycle (top-right state in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}), while in relative calculations the binding site must be defined at both end states of the complex leg (top- and bottom-right states in Fig.~\ref{fig:fig_binding_thermodynamic_cycle}).
+In principle, this requires defining which conformations are considered to be bound before running the calculation, but it is common practice to start the simulation with the ligand already placed in the binding site and rely on kinetic trapping to maintain the bound complex.
+
+However, this strategy of using kinetic trapping to maintain the bound complex can fail when the dissociation rate of the ligand has the same or smaller order of magnitude than the length of the simulation.
+This is typical of weak binders such as fragments binding shallow pockets with $\mu$M-mM affinities~\cite{georgiou2017pushing,pan2017quantitative}.
+In the case of weak binders, using a flat-bottom or harmonic restraint between receptor and ligand in the bound state(s) can prevent dissociations~\cite{georgiou2017pushing,rizzi2019sampl6}.
+We stress that this type of restraint is normally avoided as it generally introduces bias in the free energy estimate, which is why the restraint is usually activated only in the intermediate states in absolute calculations.
+The bias can be corrected through reweighting schemes~\cite{rizzi2019sampl6}, but this post-processing step can be avoided if a flat-bottom restraint is used and the ligand is never sampled while hitting the potential wall during the simulation in the bound state as the numerical correction will be exactly zero.
+It is important to note that the spring constant and/or radius parameters of the restraint effectively determine which conformations are considered to be bound.
+As a consequence, these parameters must be tuned to the system so that only the binding site is accessible to the ligand.
+Again, this step is particularly important for weak binders as their free energy of binding is known to be more sensitive to the definition of the binding site~\cite{gilson1997statisticalthermodynamic}.
+
+In absolute calculations, this restraint can substitute or be added to the restraint used to handle the standard state correction (Sec.~\ref{sec:standardstate-restraints}).
+In the latter case, however, care must be taken when computing the standard state correction.
+When multiple restraining potentials are active in the non-interacting state, the correction can generally be computed only through numerical integration.
+Alternatively, one can adopt a protocol that removes the bound-state restraint in the non-interacting state.
+Finally, even for tight binders, dissociation events can be enhanced by methods such as Hamiltonian replica exchange~\cite{sugita2000multidimensional,chodera2011replica,wang2013identifying} and expanded ensemble~\cite{lyubartsev1992new,li2007simulated}, especially in absolute free energy calculations using harmonic or flat-bottom restraints.
+In the latter case, dissociations can be averted simply by increasing the spring constant and/or reducing the radius of the restraint potential to prevent the exploration of ligand conformations outside the binding site in the decoupled state (bottom-right state in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}) that could be propagated to the bound state.
 

### 2025-07-01 paper/manuscript.tex
Update ABFE cycle figure

@@
 \begin{figure}
-    \includegraphics[width=0.95\linewidth]{figures/fig5_thermo_cyc_abs/Figure_1.pdf}
-    \caption{\textbf{Thermodynamic cycle required for an absolute free energy calculation -- absolute free energy of binding example} The fully interacting ligand in water (\textbf{A}), has its charges turned off to pass to (\textbf{B}) followed by turning of van der Waals terms, resulting in a non-interacting ligand in water in (\textbf{C}).
-Restraints are used on the fully interacting ligand in the binding site of a protein or host molecule (\textbf{D}).
-The next step is to turn off the charges again (\textbf{E}) followed by the van der Waals interactions resulting in a non-interacting complex state (\textbf{F}).
-Free energyes can be computed as $\Delta G_{bind} = (\Delta G^{\mathrm{elec}}_{\mathrm{solv}}+ \Delta G^{\mathrm{VdW}}_{\mathrm{solv}})-(\Delta G^{\mathrm{elec}}_{\mathrm{bound}}+ \Delta G^{\mathrm{VdW}}_{\mathrm{bound}}$).
-    }
+    \includegraphics[width=0.95\linewidth]{figures/fig5_thermo_cyc_abs/Figure.png}
+    \caption{\textbf{Thermodynamic cycle for an absolute binding free energy calculation with auxiliary restraints} The fully interacting ligand in water (\textbf{A}), has its charges turned off to pass to (\textbf{B}) followed by turning off van der Waals terms, resulting in a non-interacting ligand in water (\textbf{C}).
+The receptor-ligand complex (\textbf{D}) is defined by binding site restraints (dotted oval).
+Auxiliary restraints are often introduced (\textbf{E}) to improve convergence while the charges and van der Waals interactions are turned off (\textbf{F} and \textbf{G}).
+The free energy cost of releasing the auxiliary restraints (\textbf{G} to \textbf{H}) and transferring the non-interacting ligand from the standard state volume to the binding site (\textbf{C} to \textbf{H}) must be accounted for.
+The standard binding free energy is computed from this cycle as $\Delta G^{\circ}_{\mathrm{bind}} = (\Delta G^{\mathrm{elec}}_{\mathrm{solv}}+ \Delta G^{\mathrm{VdW}}_{\mathrm{solv}}) + \Delta G^{\circ}_{\mathrm{ideal}} -(\Delta G_{\mathrm{aux.restr.}}+\Delta G^{\mathrm{elec}}_{\mathrm{bound}}+ \Delta G^{\mathrm{VdW}}_{\mathrm{bound}}+\Delta G_{\mathrm{release}})$.
+Many variations of the route between the free (\textbf{A}) and bound (\textbf{D}) states can be used -- for example, state \textbf{H} is commonly skipped and the free energy cost of moving from \textbf{C} to \textbf{G} is calculated directly.}
     \label{fig:fig_absolute_thermodynamic_cycle}

### 2025-07-03 paper/manuscript.tex
Response to refs

--- 2025-06-30 paper/manuscript.tex
@@
 
-% aux. comments Stefan Boresch -> Stefan Boresch
-\newcommand{\sbnote}[1]{%
-  {\bfseries{}[SB: }%
-  {\textcolor{blue}{#1}}{\bfseries{}]}
-  }
-\newcommand{\rref}{{\bfseries[REFs]}{}}
-
 \makeatother
@@
 \author[2]{Huafeng Xu}
+
 %
@@
 Binding site water molecules may form water mediated protein-ligand interactions which can pose challenges whenever exchange with bulk water is slow compared to simulation timescales.
-This happens typically in buried binding sites.
-Overlaying multiple protein X-ray structures can identify conserved or additional water molecules that can be useful to include in calculations.
-In cases where water molecules are known to play an important role in the binding, software implementations that use water sampling facilitated by Grand Canonical Monte Carlo methods may be useful~\cite{michel2010prediction}.
-Other tools such as WaterMap or open source equivalents (SSTMap, GIST, and others) can be used to define water structure for systems with no experimental evidence of water sites~\cite{wang2011ligand}.
+This can happen in buried binding sites~\cite{laage2017water}.
+Overlaying multiple protein X-ray structures can identify conserved water molecules or those that are associated only with certain ligands in a series and are therefore important to include in calculations.
+In cases where water molecules are known to play an important role in ligand binding, for example if the presence of electron density suggests they are tightly bound (indicated by a high EDIA score~\cite{nittinger2015evidence,meyder2017estimating}), software implementations that use water sampling facilitated by Grand Canonical Monte Carlo methods may be useful~\cite{michel2010prediction}.
+Other tools such as WaterMap~\cite{abel2008role,young2007motifs} or open source equivalents (SSTMap~\cite{haider2018solvation}, GIST~\cite{ramsey2016solvation}, and others) can be used to define the hydration states for systems with no experimental evidence of water sites~\cite{wang2011ligand}.
 Well-known protein systems with water mediated ligand interactions are for example: HSP90 which formed part of the D3R grand challenge 2015~\cite{mey2016blinded}, A2A~\cite{brucemacdonald2018ligand}, MUP~\cite{ross2015water}, ~\cite{deflorian2020accurate}, and others~\cite{michel2009energetics}.
 
-\paragraph{Membrane proteins require additional care}
-Integral membrane proteins require embedding in a suitable model membrane to remain stable during simulations.
-Lipids are usually not included in protein force fields and, although they may be covered by generic small molecule force fields, have often been parameterized separately (e.g., AMBER's Lipid21 \cite{AMBERLipid21} and the CHARMM lipid force field \cite{CHARMMLipidFF}).
-Membrane systems can be simulated in multiple thermodynamic ensembles, but the semi-isotropic scheme $NP_{z}P_{xy}T$ is frequently employed, allowing the pressure in the bilayer plane ($xy$) to scale independently \cite{LipidBestPracticesCOMS}.
-These systems require more extensive equilibration than soluble proteins.
-Some workflows, such as CHARMM-GUI or Desmond membrane preparation, recommend a multi-protocol to relax membrane systems \cite{https://doi.org/10.1002/jcc.23702,DesmondManual}.
-
-Alchemical transformations that involve net charge changes can provide difficulties for membrane systems; phospholipids carry functional groups with nonzero formal charge and cannot be treated as a solvent with a uniform dielectric in the same way as water.
-Several approaches are reviewed in \cite{wu2022correction-51c}.
-The authors recommend the use of a co-alchemical ion, specifically making an ion of equivalent charge appear alongside the disappearing charged molecule to maintain system neutrality.
-
-In the case of a lipid-facing binding site, there are three additional challenges.
-First, simulation time should be extended to ensure sufficient sampling of lipid configurational space.
-Lipids may either bind alongside small molecules, acting as cofactors, or be displaced from the binding site by a ligand.
-If lipids participate in binding through protein-lipid or ligand-lipid interactions, it is necessary to sample these intermolecular interactions through additional $\lambda$ windows, longer simulation times, or different alchemical protocols.
-Displacing a lipid will have its own free energy cost which may take a long time to converge if it requires diffusion and rearrangement of nearby lipids.
-Second, considering the free energy cost of a ligand partitioning into the membrane may provide additional insights \cite{doi:10.1021/acs.jcim.1c01147}.
-Third, the free energy of binding for a small molecule may be strongly coupled to the lipid species.
-In some cases, the identity of the lipid may be known.
-But in other cases, it may be difficult to discern the identity of the lipid from X-ray or cryo-EM structures due to the flexible nature of lipid head groups and benchmarking is recommended to match experimental data.
+In the context of relative binding free energy calculations, determining the correct hydration pattern for the two endpoints may not be sufficient for an accurate and reproducible calculation.
+If the locations of the bound water molecules differs at the two endpoints - e.g.
+if a perturbation to the ligand displaces a bound water molecule - then the network needs to be able to adapt during the perturbation.
+Otherwise, it is likely that the estimated relative binding free energies will be dependent on the direction in which the perturbation is performed~\cite{ross2020enhancing}.
+
+In absolute binding free energy calculations, a ligand is fully decoupled from a binding site cavity which may otherwise be hydrated in the \textit{apo} state.
+In this case, the sampling of water molecules must be sufficient to hydrate the binding pocket as the ligand is decoupled (or vice versa for the reverse).
+When the binding site is exposed and water can move freely between the site and bulk solvent, then sampling is often not an issue.
+However, in the case of more buried sites or sites which form complex water networks, enhanced water sampling may be required to ensure an accurate binding affinity.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\section{\tocorange{What simulation protocol should I choose?}}
-\section{\tocorange{Choice of simulation protocol and implementation of alchemical transformations}}
-  \sbnote{Emilio suggested to make the title more general; this is my suggestion!}
+\section{\tocorange{What simulation protocol should I choose?}}
 \tocsectioncomment{Steering group: Possibly expand to include other methods group agrees on.}
@@
 
-\paragraph{Topologies} In regular biomolecular simulations one studies physical systems or, in other words, all entities in the system, e.g., protein, ligand and waters, represent real molecules.
-For the following, it is helpful to classify such simulation systems as having a {\em chemical} topology.
-In relative free energy calculations, however, one has to transform, e.g., ligand A into ligand B etc.
-(see \sbnote{appropriate cross-reference!}).
-The required alchemical transformations can be realized in several ways, each with distinct advantages and disadvantages.
-In the first relative free energy calculations, these transformations were achieved with the help of chimeric entities, capable of representing both ligand A or ligand B.
-The important point here is that there is a single set of coordinates for this chimeric ligand, but its ``topology'' is flexible and can, for the same set of coordinates, represent either end state.
-We will consider this class of approaches as using an {\em alchemical} topology with a single set of coordinates.
-The construction of these chimeric intermediates quickly gets complicated and is difficult to automate; therefore, approaches have been developed in which both end states (e.g., ligand A and B) are present simultaneously.
-Thus, we have dual coordinates, but chemical topologies for each of the entities.
-In the following, we outline these approaches in more detail, together with additional considerations how the various approaches differ in practical use.
-While we will point out the respective advantages and disadavantages, we stress that the user's option in practice often are limited by the choice of simulation software.
-
-\paragraph{Single coordinate approaches --- alchemical topologies}
-
-When operating with a single set of coordinates for the region undergoing the alchemical transformation, a key distinction between approaches concerns the atom mapping between end state molecules.
-%Often this is predetermined by the choice of simulation software.
-As first suggested by Pearlman, these have been referred to as \emph{dual topology} versus a \emph{single topology} \cite{Pearlman_1994}.
+\paragraph{Topologies} 
+A critical first step is to determine which approach to use for atom mapping between end state molecules.
+Often this is predetermined by the choice of simulation software.
+Typically it is possible to chose between a \emph{dual topology} versus a \emph{single topology}
+
 The distinction between single and dual topology can be illustrated by considering a hypothetical transformation from molecule A to molecule B, where both atoms share a common substructure but differ in their substituents; in particular, consider a transformation of benzene to benzyl alcohol shown in Fig.~\ref{fig:fig_topology}.
-The single and dual topology approaches as proposed by Pearlman correspond to the left and middle paths of Fig.~\ref{fig:fig_topology}.
-%\sbnote{Fig.~3 needs to be updated for dual coordinate methods and corrected, some bonds for the dummy atoms have the wrong anchor atom!}
-The common substructure between the two molecules is the benzene ring, though in practice the substructure may be selected to be larger depending on the mapping chosen, as we discuss below.
+In this case the common substructure between the two molecules is the benzene ring, though in practice substructure may be selected to be larger depending the mapping chosen, as we discuss below.
 
 In single topology calculations, the overall transformation is set up to involve as few additional atoms as possible, so benzene would be typically changed into benzyl alcohol by first changing one of the hydrogens into a carbon.
-The site of this transformation will also be the future home of two additional hydrogen atoms bound to the new carbon, so these must initially be present as non-interacting atoms called ``dummy atoms'' (labelled Du in Fig.~\ref{fig:fig_topology}), which retain their bonded interactions but do not interact with the rest of the system.
-Bond parameters as well as partial charges between the changing atoms are adjusted accordingly between the initial and final state; see below for further details.
-In a single topology calculation, atoms may change their type, ensuring minimal dummy atoms are created.
-This is illustrated in the left arm of Fig.~\ref{fig:fig_topology}.
-At intermediate states, atoms behave as 'mixed' (M), either interpolating between two physical atoms, e.g., in the example, the hydrogen becomes a carbon, or between dummy and physical atoms.
-
-\begin{figure}
-    \includegraphics[width=0.95\columnwidth]{figures/fig3_topol/Figure.pdf}
-    \caption{\textbf{Strategies for alchemical calculations: single coordinate single and dual topology vs.\ dual coordinate dual topology.} \textbf{Left}: A single coordinate single topology converts from one type of atom to another.
-Dummy atoms (Du) are used when there is no corresponding maximum common substructure match between the two molecules for certain atoms, using a soft-core interaction to improve overlap between the dummy atoms and the "real" atoms.
-At intermediate states, 'mixed' atoms (M) are present, either interpolating between physical atoms, e.g., a hydrogen to a carbon, or between physical and dummy atoms.
-\textbf{Middle}: The single coordinate dual topology does not convert one species to another, but only converts between Du atoms and an interacting species, typically using soft-core potentials for this.
-Here, 'mixed' intermediate atoms (M) always interpolate between physical and dummy atoms.
-Thus, 'mixed' atoms are used in both single coordinate approaches (single and dual topology), but the way the transformation occurs and the end states differ.
-      %Following the arrow along the left and right illustrate the differences.
-\textbf{Right}: In the dual coordinate dual topology approach both end states are present as separate molecules simultaneously.
-Initially, benzyl alcohol is not interacting with the environment, at intermediate states the environment sees a mixture of the interactions of the two solutes, and, finally, benzene is not interacting with the environment.
-The two solutes are kept on top of each other by suitable restraints, indicated by the black ``spring'', and  the two molecules never see each other.
-Figure adapted from the one used in V1 of this manuscript; adapted and extended by Sara Tkaczyk.
-      % \url{http://www.alchemistry.org/wiki/Constructing_a_Pathway_of_Intermediate_States}
-    }
-    \label{fig:fig_topology}
-\end{figure}
+The site of this transformation will also be the future home of two additional hydrogen atoms bound to the new carbon, so these must initially be present as non-interacting atoms called ``dummy atoms'', which retain their bonded interactions but do not interact with the rest of the system.
+Bond parameters as well as partial charges between the changing atoms are adjusted accordingly between the initial and final state.
+Thus, in a single topology calculation, atoms may change their type, ensuring minimal dummy atoms are created. This is illustrated in the left arm of Fig.~\ref{fig:fig_topology}.
 
@@
 Fig.~\ref{fig:fig_topology}'s right branch depicts how such a dual topology works.
-Note that the two groups that are switched on or off never see each other.
-The 'mixed' intermediate atoms (M) are used as well, but they strictly interpolate between physical and dummy atoms, or vice versa.
-
-%\sbnote{This probably needs to be rewritten\ldots}
-As far as we are aware, the distinction between (single coordinate) single and dual topology approaches has been made primarily for two main reasons.
+
+\begin{figure}
+    \includegraphics[width=0.95\columnwidth]{figures/fig3_topol/Figure.pdf}
+    \caption{\textbf{Two common topologies for alchemical calculations: single and dual topology.} \textbf{Left}: A single topology converts from one type of atom to an other.
+Dummy atoms (Du) are used when there is no corresponding maximum common substructure match between the two molecules for certain atoms, using a soft-core interaction to improve overlap between the dummy atoms and the "real" atoms.
+\textbf{Right}: The dual topology does not convert one species to another, but only converts between Du atoms and an interacting species, but usually uses soft-core potentials for this.
+The 'mixed' intermediate atoms (M) are used in both dual and single topology approaches.
+Only the way the transformation occurs and the end states differ.
+Following the arrow along the left and right illustrate the differences.
+Figure adapted from \url{http://www.alchemistry.org/wiki/Constructing_a_Pathway_of_Intermediate_States}}
+    \label{fig:fig_topology}
+\end{figure}
+
+As far as we are aware, the distinction between single and dual topology approaches is made primarily for two main reasons.
 First, this choice affects the alchemical pathway followed, and thus may affect convergence properties---though we are not yet aware of a study of the relative efficiency and merits of these two approaches.
 Additionally, historically, some simulation packages implemented only one approach and not the other, meaning that the distinction was functional.
-For an in-depth analyis of differences resulting from the use of a  single vs.\ a dual topology approach (single coordinate in both cases) see Ref.~\cite{Boresch_2002}.
-
-\paragraph{Dual coordinate approaches --- dual chemical topologies}
-
-The chimeric constructs needed in single and dual topology approaches in the sense of Pearlman’s definition just described quickly become non-trivial to set up and require care to avoid artifacts (see below).
-%\sbnote{See below; where do we place the dummy atom discussion?}.
-Some of these difficulties might be avoided by having the initial and final state present as separate molecules with separate coordinate sets.
-To the best of our knowledge, such an approach was first described by Axelsen and Li \cite{Axelsen_1998}.
-Both initial and final state are present simultaneously; the two copies never see each other and interactions with the environment are turned on/off as a function of $\lambda$.
-Harmonic restraints between corresponding atoms in the initial and final state copy prevent the two molecules (sets of atoms from drifting apart).
-%Unfortunately, Axelsen and Li referred to their ansatz also as dual-topology, though in practice it is quite different from Pearlman’s meaning of the term.
-A recent implementation of this idea is Restraintmaker \cite{Ries_2022}, a tool to automatically assign the distance restraints needed to keep the two molecules / regions (one for each of the physical endstates) on top of each other.
-The idea is illustrated in the right branch of Fig.~\ref{fig:fig_topology}.
-Somewhat unfortunately, Axelsen and Li, as well as Ries et al refer to their approach as dual topology.
-Ries et al suggested to use the term hybrid topology for (single coordinate) dual topology in the sense of Pearlman.
-We will attempt a disambiguation of the various nomenclatures below.
-
-Another variant of a dual coordinate approach is implemented in NAMD \cite{jiang2019computing}.
-Whereas Ries et al.\  use distance restraints to keep two molecules on top of each other, corresponding atoms in NAMD are held on top of each other by constraints; in practice, this means that these coordinates are forced to be identical.
-AMBER's free energy implementation recently switched to a similar approach \cite{Lee_2023}.
-%The authors refer to this approach as hybrid single-dual topology.
-While it technically uses two separate sets of coordinates, the constraint that corresponding atoms have identical coordinates makes it effectively a single coordinate method; hence, I consider NAMD’s approach to be a variant of Pearlman’s (single coordinate) dual topology.
-
-Dual coordinate approaches with chemical topologies encompass a third group of methods, where essentially two absolute free energy calculations are carried out in opposite directions at the same time.
-The so-called ``separated topologies'' (SepTop) method \cite{rocklin2013separated,Baumann_2023} and the ``alchemical transfer method'' (ATM) \cite{Azimi_2022} belong into this category.
-In SepTop, the two copies are held on top of each other via restraints with respect to the protein/receptor they bind to (as opposed to Restraintmaker, where the molecules are restrained to each other.
-%In ATM, a restraint is used to keep the bound ligand 1 and the free ligand 2 always some minimum distance apart; during exchange moves, suitable rigid-body rotations/translations are applied to move the free ligand into a position/orientation as similar as possible to the one of the bound ligand.
-In ATM, restraints are used to keep the bound and free ligands at some distance apart and in the same orientation, so that during the translation move the free ligand is placed into a position and orientation as similar as possible to that of the bound ligand.
-Both copies are present simultaneously, but they do not occupy the same space.
-Recently, a variant of $\lambda$-dynamics that uses chemical topologies (separate coordinates for each state) has been proposed \cite{Liesen_2024}.
-
-%Other methodologies also use both end states simultaneously; examples are the Separated Topologies (SepTop) approach [Baumann2023] and the ATM method [Azimi2022].
-In the former, the two copies are held on top of each other via restraints with respect to the protein/receptor they bind to.
-Ries et al.
-refer to such an approach as “separated dual topology”, whereas they label their approach as “linked dual topology”.
-In ATM, a restraint is used to keep the bound ligand 1 and the free ligand 2 always some minimum distance apart; during exchange moves, suitable rigid-body rotations/translations are applied to move the free ligand into a position/orientation as similar as possible to the one of the bound ligand.
-Both copies are present simultaneously, but they do not occupy the same space.
-
-%One final approach, so far in its infancy, has been called ``separated topologies'' and essentially consist of two absolute free energy calculations in opposite directions at the same time, turning one molecule's interactions with the environment off, while turning the other molecule's interaction on~\cite{jiang2019computing, rocklin2013separated}.
-
-\paragraph{Summary, comparison of single and dual coordinate methods, orthogonal classifications:}
-
-%Some additional terms have also been employed to talk about these different intermediate pathways .
+
+Some additional terms have also been employed to talk about these different intermediate pathways .
 Particularly, some studies refer to a ``hybrid'' topology approach to free energy calculations~\cite{gapsys2015pmx, gapsys2016accurate, gapsys2020large}, though this term may not yet have achieved widespread use.
@@
 
-Unfortunately, during the past years the use of the terms single and dual topology has proliferated, often with quite different meaning from the original intentions.
-The above presentation makes clear that the primary difference between approaches is the use of separate coordinates for each end state (chemical topology) or the use of an alchemical chimera with a single set of coordinates (alchemical topology).
-We therefore suggest the following classification
-\begin{itemize}
-\item Single coordinate (SC) --- Alchemical topology
-  \begin{itemize}
-  \item Single topology (SCST)
-    \item Dual topology (SCDT)
-  \end{itemize}
-  \item Dual coordinate --- Chemical topology (DC)
-  \end{itemize}
-For completeness, we list some additional terminology that has been used: Some studies refer to a ``hybrid'' topology approach to free energy calculations~\cite{gapsys2015pmx, gapsys2016accurate, gapsys2020large}, though in our view these calculations should be classified as single coordinate single topology \sbnote{Check whether this is correct!}.
-%this term may not yet have achieved widespread use.
-In this case, "hybrid" seems to indicate that the set up of these free energy calculations involves a hybrid of the two molecules, and much of what is done in these studies uses a single topology approach~\cite{gapsys2020large}.
-Ries et al.\ suggested terminology \cite{Ries_2022} in which single topology corresponds to SCST, hybrid topology to SCDT, and dual topology to DC.
-
-The practical implementation of the approaches to carry out alchemial transformations above leads to additional, often non-trivial differences within each methodology.
-In the description of DC approaches we already pointed out that the two sets of coordinates can be held on top of each other by constraints \cite{jiang2019computing,Lee_2023} or restraints.
-In the latter case, the restraints may couple the two set of coordinates directly \cite{Axelsen_1998,Ries_2022}, or the restraints may be defined relative to a common reference frame \cite{rocklin2013separated,Baumann_2023}.
-Finally, in ATM the two copies of the coordinates are held separate \cite{Azimi_2022}.
-
-Concerning SC approaches, Pearlman’s usage of the term single topology not only meant a chimeric construct as depicted in Fig.~\ref{fig:fig_topology}, but specifically implied that the coupling parameter $\lambda$ was applied to the force field parameters.
-E.g., a bond stretching energy term changing from C-H to C-C has the functional form $U_{bond}(\lambda)=K(\lambda)(r-r_0(\lambda))^2$.
-For $\lambda=0$, $K(\lambda)$ and $r_0(\lambda)$ correspond to the values appropriate for a C-H bond, and for $\lambda=1$ to those for a C-C bond.
-The other energy terms are treated analogously.
-The single topology paradigm as just summarized was the default mode of alchemical transformations in AMBER and GROMOS at that time.
-More recently, this implementation of single topology has been referred to as parameter interpolation or mixing \cite{Giese_2018}.
-
-The $\lambda$-dependent interpolation can also be accomplished by mixing the energies and forces of the initial ($\lambda=0$) and final state ($\lambda=1$), e.g., linearly according to $U(\lambda)=(1-\lambda) U_0+\lambda\,U_1$.
-Other λ-dependencies can be used, and soft-cores \rref may be used for Lennard-Jones and electrostatic interactions to avoid endpoint problems.
-Giese et al.
-\cite{Giese_2018} refer to this approach as energy mixing (interpolation).%; it requires two separate calls to the energy/force routines, but since only a fraction of interactions is affected by the alchemical changes, various optimizations are possible to keep the computational overhead manageable.
-An early example of energy mixing for SCST is the PERT free energy module of CHARMM (written by Bernie Brooks around 1990 and briefly described in Ref.~\cite{brooks_2009}.
-
-Any SCDT approach requires two energy/force calculations and, thus, uses energy mixing.
-Therefore, approaches that implement energy mixing together with an alchemical topology (a single set of coordinates) often support both SCST and SCDT.
-E.g., CHARMM's PERT module just mentioned could be used to set up either of the SC approaches sketched in Fig.~\ref{fig:fig_topology}.
-Finally, all DC approaches require two energy/force evaluations.
-
-Any energy mixing approach, whether in a SC or DC context, requires two energy/force evaluations.
-While various optimizations keep the computational overhead significantly below a factor of two,  parameter mixing, if applicable, is the compuationally most efficient approach.
-However, it is closely tied to SCST and the more complicated the alchemical transformation, the more challenging a strict SCST approach becomes.
-Also, the mixing of partial charges and the calculation of $\partial U/\partial\lambda$ for the reciprocal space particle-mesh Ewald (PME) summation is tricky.
-Giese et al.\ described how to do this \cite{Giese_2018}, but most energy mixing implementations calculate the PME term twice and mix the resulting energy and forces.
-
-\sbnote{I think this should get its own paragraph/subsection; to be expanded!} Software packages vary in their use of single or dual topology approaches; for example, AMBER TI uses a dual topology approach, while BioSimSpace uses a single topology approach.
+One final approach, so far in its infancy, has been called ``separated topologies'' and essentially consist of two absolute free energy calculations in opposite directions at the same time, turning one molecule's interactions with the environment off, while turning the other molecule's interaction on~\cite{jiang2019computing, rocklin2013separated}.
+
+Software packages vary in their use of single or dual topology approaches; for example, AMBER TI uses a dual topology approach, while BioSimSpace uses a single topology approach.
 Please make sure to check what approach is used with your software package of choice, or whether it supports your choice of approach (GROMACS and GROMOS, for example, support both).
@@
 
-In general there are mutliple ways on how a MCSS can be calculated.
-One of the most prominent approaches is to estimate the MCSS with a 2D subgraph isomorphism solver(SIS) using the 2D topological stuructre information of molecules\cite{raymond2002maximum}.
-A large number of software tools can compute SIS MCSS matches using different cheminformatics packages.
-Some rely on RDKit~\cite{rdkit2019Dec}, such as pmx~\cite{gapsys2015pmx}, LOMAP~\cite{liu2013lead}, FESetup~\cite{loeffler2015fesetup}, ProtoCaller~\cite{suruzhon2020protocaller}, SMArt~\cite{petrov2021perturbation}, and partially BioSimSpace~\cite{hedges2019biosimspace}, while others such as fkckombu~\cite{kawabata20143d} are standalone tools.
+There are different tools that allow the generation of MCSS matches as well as single topology input.
+A large number of software tools can compute MCSS matches using different cheminformatics packages.
+Some rely on RDKit~\cite{rdkit2019Dec}, such as pmx~\cite{gapsys2015pmx}, LOMAP~\cite{liu2013lead}, FESetup~\cite{loeffler2015fesetup} and partially BioSimSpace~\cite{hedges2019biosimspace}, while others such as fkckombu~\cite{kawabata20143d} are standalone tools.
 Schr\"{o}dinger's FEP+ planning tool was originally based on a version of LOMAP, and it also uses MCSS matching as well as 3D considerations to plan the network of single topology calculations between molecules~\cite{wang2015accurate}.
 
-SIS based methods, showed to struggle with 3D properties of the molecules, that need to be mapped, like for example stereochemistry.
-Therefore some packages developed a post-mapping step, where they filtered for atom distances, and therefore added 3D information into the mapping.
-Still SIS MCSS searches can be relatively time consuming as the algorithmic complexity is classified as NP-complete, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial similarity assessment and then use MCSS only to identify final mappings for relative calculations~\cite{raymond2002maximum,klabunde2012mars,jones2009elucidating, Duesbury2018Comparison}.
-The SIS MCSS approach, though relatively standard, takes into account only topological similarity.
+MCSS searches can be relatively time consuming, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial similarity assessment and then use MCSS only to identify final mappings for relative calculations~\cite{raymond2002maximum,klabunde2012mars,jones2009elucidating}.
+The MCSS approach, though relatively standard, takes into account only topological similarity.
 It is possible that changes in binding mode could actually require a different choice of mapping, so in some cases mappings may need to be planned differently depending on 3D positioning of atoms in space.
-
-Here an alternative way to estimate the MCSS is to calculate a bipartite graph matching based on the 3D positions of two molecules.
-These approaches tend to be very fast (can have algorithmic complexity of $O(n)=N^3$) and geometric accurate in their mapping (e.g.
-stereochemistry or binding mode differing).
-The geometric approaches additionally allow mappings of differing binding modes of the same molecule, or can be even used for significant larger molecules like proteins, due to the speed increase.
-A disadvantage of these approaches is the strong dependency on the coordinate sets being well aligned.
-Another aspect is taking the topology changes of the molecules not into account, which can be done if necessary in a next step with a filtering approach, like checking for changes in ring hybridization or ring breaking (this problem is discusse below)|.
-Tools that can be used to estimate the MCSS this way are contained for example in pmx \cite{gapsys2015pmx} or Kartograf \cite{ries2024kartograf}.
-
-In any case Visual inspection prior to simulation is recommended to ensure that the mapping criteria correspond to the expected binding mode.
+Visual inspection prior to simulation is recommended to ensure that the mapping criteria correspond to the expected binding mode.
 If the mapping protocol returns simulations that correspond to different binding modes of a ligand within a perturbation map, this can cause large hysteresis.
 
-A next step for atom mappings, would be to support multistate methods (e.g.
-RE-EDS~\cite{sidler2016replica}, A-EDS~\cite{perthold2020toward} or Ladybugs~\cite{robo2022achieving}) with efficient mapping algorithms.
-There are protoype starting point for solving this problem, but to our knowledge, there is only few published work on this aspect.
-\cite{petrov2021Perturbation}
-
-Single coordinate set relative calculations and calculations based on substructure searches only work if in fact the ligands share a common substructure, e.g.
+Single topology relative calculations and calculations based on substructure searches only work if in fact the ligands share a common substructure, e.g.
 are part of a congeneric series, see Fig.~\ref{fig:fig_mcss}.
-If no common substructure is shared, then dual coordinate sets methods like linked-dual or separated topology free energy calculations are needed.
+If no common substructure is shared, then alternative dual or separated topology free energy calculations are needed.
 One would co-localize a pair of compounds in a binding site, exclude their interactions with one another, and compute the relative binding free energy by turning one molecule on from being dummy atoms while turning the other off.
-For such approaches it is crucial to find a good set of restraints, that keep the ligands colocalized over the time of the simulation .
-This problem is similar to the required restraints in ABFE Calculations (see \ref{ABFErest}).
-Few descriptions of possible approaches in RBFE calculations can be found for such approaches like the one suggested by \textit{Baumann et al.}~\cite{baumann2023broadening} based on boresch style restraints~\cite{boresch2003absolute}, with an elaborate automatic atom selection approach, for the seperated topology approach, or Restraintmaker~\cite{ries2022restraintmaker} for the linked-dualtopology approaches, which finds a small set of atoms to be linked via distance restraints, with a small distance to each other.
-However the implementation and testing of an fully automatized dual-topology pipeline for RBFE dual topology approaches is still to our knowledge a remaining research problem and was only tested with small datasets.
-\cite{jespers2019qligfep, Rieder2022Lev}
-
-\paragraph{Dummy atoms}
-
-In practically all relative free energy simulations, regardless of the approach, SCST, SCDT, or DCDT, dummy atoms (non-interacting placeholder atoms) are required.
-We stress this point, as sometimes dummy atoms are viewed as something specific to SCST.
-In SC approaches, dummy atoms may even be required if the number of atoms between initial and final state of the system is identical, consider, e.g., tautomeric forms of a molecule.
-
-The need for and role of dummy atoms is clearly visible in Fig.~\ref{fig:fig_topology}: One or more dummy atoms are attached to the physical molecule.
-A crucial requirement is that these atoms, typically connected through bonded energy terms (bond stretching, angle bending, dihedral angles) do not affect equilibrium and dynamical properties of the physical system.
-However, since a physical molecule + dummy atoms attached to it can or must be viewed as a modified physical molecule, the partition function of the molecules with and without dummy atoms will NOT be identical.
-In other words, a single free energy difference between two molecules A and B will be different from that of A(+D) and B(+D), where the (+D) indicates the presence of one or more dummy atoms needed so that the alchemical transformation can be carried out.
-When a dual coordinate approach is used, the noninteracting part of the respective other endstate is tied to the physical one, and will contribute to the partition function.
-By construction, relative free energy simulatins involve two analogous alchemical transformations, e.g., A(+D) -> B(+D) in gas phase and aqueous solution (RSFEs), or in the bound and unbound state (RBFEs).
-If dummy atoms are interacting with the physical system in such a way that $\mathsf{Z = Z(R) \times Z(R\text{-}D) \times Z(D)}$, where Z is the total partition function, Z(R) the partition function of the physical system, Z(D) the partition function contribution resulting from interactions between only the dummy atoms, and Z(R-D) the contribution resulting from the bonded connections between the physical atoms and dummy atoms, then their contributions (Z(D) and Z(R-D)) factor from the total partition function.
-As dummy atoms do not interact with the environment, their contribution, therefore, is identical in the two alchemical transformations required in the two legs of the relative transformation and will cancel exactly from the double free energy difference of interest.
-
-While some early work on the correct treatment of dummy atoms, i.e., how to treat them so that their contribution to the partition function factors, has been available for years \cite{Boresch_2002,Shobana_2000,Wang_2012}, most workers seem to mostly turn off nonbonded interactions for dummy atoms, but leaving all bonded interactions intact \sbnote{To all: While I am quite sure that this is what most of us have been doing most of the time, I wonder whether this has been epxlicitly described anywhere?}.
-Since current force fields use multiple angle and torsion terms for each atom, there are often a large number of bonded energy terms connecting dummy atoms to the physical molecule.
-However, as already described in the early studies and repeated in a more recent study focusing on the correct treatment of dummy atoms \cite{Fleck_2021}, the desired factorization can only be guaranteed if any dummy atom connected to the physical molecule is held by exactly three (non-redundant) degrees of freedom.
-Naive application of the three-degree-of-freedom per dummy atom rule (for dummy atoms having at least one joint bonded term with the physical molecule), however, can result in simulations where the dummy part adopts unphysical configurations/conformations; Fleck et al.
-referred to this as “flapping”.
-Thus, depending on the details of the physical molecule -- dummy atom(s) junction, the three degree of freedoms need to be selected with care, or a carefully chosen additional redundant degree of freedom may be needed.
-In the latter case, the factorization of the partition function is not exact, but the overall error can still be made arbitrarily small.
-For the full details, we refer the reader to Ref.~\cite{Fleck_2021}.
-
-One situation involving dummy atoms that needs to be avoided is the creation of a cyclic structure involving dummy atoms, i.e., whenever the (group of) dummy atom(s) is attached to two different parts of the physical molecule.
-In this case, the factorization of the partition function is never possible \cite{Shobana_2000}.
-It follows that ring opening or closing is extremely challenging to carry out using SCST approaches \cite{liu2015ring}; these difficulties have in part motivated the development and adoption of dual coordinate approaches.
-For more on ring opening/closing, see also below.
-Nevertheless, DC approaches are susceptible to exactly this type of problem: to keep the noninteracting molecule on top of the interacting one, typically multiple bonded restraints (e.g.
-harmonic bonds) are used \cite{Axelsen_1998,Ries_2022}.
-These effectively, are equivalent to the formation of cyclic structures with potential erroneous contributions to the double free energy difference of interest.
-The type of restraints used in SepTop \cite{rocklin2013separated,Baumann_2023} and ATM \cite{Azimi_2022} aoid this problem.
-
-The influence from an incorrect treatment of dummy atoms (e.g., leaving all bonded terms intact) is difficult to estimate, though it may be small in many cases \cite{Fleck_2021}.
-Nevertheless, some guidelines should be followed:
-\begin{itemize}
-  \item Obviously, a system with and without dummy atoms should have the same energetic minimum.
-Note that in larger groups of dummy atoms, redundant bonded terms involving only dummy atoms will be essential to avoid flapping, and the energy of the optimized geometry may be different from that of the physical system without dummy atoms.
-Thus, a \underline{necessary} condition is as follows: Optimize the geometry of the physical system (no dummy atoms); the final energy is the reference value.
-Optimize the geometry of the same molecule in the presence of dummy atoms; the final energy may well be different from the reference energy.
-Now delete the dummy atoms and recompute the energy \underline{without} further minimization.
-The energy after the deletion of the dummy atoms must be identical to the reference energy.
-\item The above is only a \underline{necessary} criterion, and the dummy atoms may still influence the dynamical behavior of the physical molecule.
-Furthermore, overzealous removal of redundant energy terms may result in flapping of the dummy atoms.
-If one suspects the latter, then a short simulation in the gas phase should reveal flapping.
-Unfortunately, while straightforward in principle, both checks described are hard to automate, which is essential in large RFEs campaigns.
-\item The correct treatment of dummy atoms can be validated by closing a suitable thermodynamic cycle.
-E.g., Fleck et al.\ calculated ASFEs for each of their endstates, as well as the RSFE between them \cite{Fleck_2021}.
-In all cases when dummy atoms were treated according to best practices, the cycle closure error was well within the statistical uncertainty, which was well below 0.1 kcal/mol.
-A related, earlier study by Loeffler et al.
-also reported ASFEs and RSFEs.
-If one attempts to close the cycle based on the numbers reported there, cycle closure error was nonnegligible in some cases \cite{loeffler2018reproducibility}.
-\item While solvation free energies can nowadays be computed quite fast, their systematic calculation would nevertheless be too costly.
-The role of solvent can be replaced by some external restraint, distorting the average geometry of the physical molecule.
-If one computes the free energy cost of applying the restraint with and without dummy atoms, the two numbers must agree within statistical significance.
-While much faster than full solvation free energy calculations, applying a restraint in the gas phase requires careful thermalization of the simulation system, and automation is difficult.
-  \end{itemize}
-%Further work automating the correct setup and treatment of dummy atoms is necessary.
+To our knowledge no general pipeline for such calculations yet exists and this would likely remain a research problem.
+Using an absolute free energy approach instead seems more promising in such a case.
 
 \paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
-Some approaches have attempted to address this using a single coordinate sets and with soft bond approaches ~\cite{clark2019relative,wang2017accurate} or dual coordinate set approach \cite{jespers2019qligfep, ries2022relative} but a general solution is not yet in mainstream use.
-Still, FEP+ implements one solution\cite{wang2017accurate}.
+Some approaches have attempted to address this~\cite{clark2019relative} but a general solution is not yet in mainstream use. Still, FEP+ implements one solution.
 
@@
 
-\paragraph{Several choices of restraints are possible.}\label{ABFErest}
+\paragraph{Several choices of restraints are possible.}
 In practice, a variety of types of restraints are common, from simple harmonic distance restraints between the ligand and the protein~\cite{mobley2006use}, to flat-bottom restraints which work similarly but only exert a force if the ligand leaves a specific region~\cite{chen2007can}.
@@
 If the net charge of the system changes as the alchemical variable changes during the  calculation, this can pose major challenges.
-Specifically, finite-size effects can introduce significant charge-dependent artifacts into computed binding free energies, in part because typical schemes for long-range electrostatics, including particle mesh Ewald (PME) and reaction field, do not handle free energy contributions from such changes as they would be handled in a hypothetical macroscopic bulk solution.
-The outcome depends either on the size of the cutoff for reaction field methods or on the size of the simulation box when using lattice sum methods due to periodic boundary conditions~\cite{lin2014overview, ohlknecht2020correcting, rocklin2013calculating}.
-
-Lattice summation methods (e.g.
-PME) effectively neutralize the simulation box by applying an implicit homogeneously distributed background counter charge~\cite{figueirido1995finite,hummer1996free}.
-The background charge can be interpreted as a converged distribution of counterions and may be a reasonable choice for computing hydration free energies of monovalent ions (finite size effect corrections can be applied in these cases)~\cite{hummer1996free}.
-In inhomogeneous systems with different dielectrics, e.g.
-solvated protein or lipid bilayer in a water box, assuming uniform background countercharge distribution is unphysical, as the higher countercharge density should be located in the regions of higher dielectric~\cite{hub2014pme}.
-This, in turn, causes significant artifacts in sampled population densities of charged molecules, i.e.
-errors in calculated free energy differences.
-
-There are two main classes of approaches to address artefacts introduced by changes in net charge during alchemical free energy calculations: (1) avoiding changing the net charge altogether and (2) post-processing approaches to correct for artefacts using continuum electrostatics.
+Specifically, finite-size effects can introduce significant charge-dependent artifacts into computed binding free energies, in part because typical schemes for long-range electrostatics (including PME and reaction field) do not handle free energy contributions from such changes effectively or as they would be handled in a hypothetical macroscopic bulk solution~\cite{lin2014overview, ohlknecht2020correcting, rocklin2013calculating}.
+
+There are two main potential solutions to avoid artifacts due to changes in net charge: avoiding changing the net charge, and correcting for the introduced artifacts.
+
 Many relative free energy planning tools have been set up to avoid changing the net charge of the systems considered, including LOMAP~\cite{liu2013lead} and Schr\"{o}dinger's FEP+~\cite{wang2015accurate}.
-In absolute free energy calculations, it is often difficult to avoid a change in the net charge.
-One way to avoid charge changes is to ensure that, as a charged ligand is removed, a charged counterion of opposite sign is also removed, or one of the same sign is inserted.
-This is sometimes referred to as a “co-alchemical ion” approach to deal with the required charge change.
-This approach is easily supported in most free energy calculating programs, as it merely involves an additional simultaneous modification.
-There are special flavours of the co-alchemical ion approach such as proposed by Gapsys et al.~\cite{gapsys2015calculation}, which uses a double-system/single-box setup.
-Here, the two branches of a thermodynamic cycle as used in relative free energy calculations are placed in one simulation box, effectively circumventing a change in the overall net charge during a perturbation.
-Note that the co-alchemical approaches only avoid finite-size effects with lattice-sum methods such as PME.
-
-Charge-change correction schemes have been explored and offer potentially viable solutions for the problem~\cite{mey2018impact}, where artifacts introduced by finite-size effects are corrected a posteriori from Poisson-Boltzmann calculations ~\cite{chen2018accurate, ohlknecht2020correcting, rocklin2013calculating,Reif2014chargecorrection}.
-However, application of such corrections is typically inconvenient and computationally demanding.
-High grid resolutions for the Poisson-Boltzmann calculations are required to avoid numerical artifacts, and it may be necessary to average over multiple representative configurations to obtain reliable results.
-Furthermore, they rely on continuum electrostatic models, which cannot fully capture phenomena arising from discrete solvent effects.
-These effects can be quantified in the orientational-disorder limit such that analytical correction terms for this effect exist for solvent models with a single van der Waals interaction site.
-Readers interested in these conceptual and technical limitations are referred to detailed analyses by Hünenberger and co-workers~\cite{kastenholz2006solvfe, rocklin2013calculating}.
-Alternatively, Reif and Oostenbrink demonstrated in a proof-of-concept study that on-the-fly corrections to the physical forces can be effectively used to eliminate finite-size and approximate electrostatic artifacts~\cite{Reif2015onthefly}.
-
-Petrov et al.~\cite{Petrov2024guidelines} investigated the box-size dependencies of alchemical free energy calculations using PME together with the co-alchemical ion approach.
-They concluded that in neutral simulation boxes with a minimal distance to the box wall larger than 1 nm, finite-size artifacts could be made negligible.
-The following recommendations should be followed to minimize finite-size artifacts, when using lattice-sum electrostatics: (a) setup of free energy calculations such that there is no net charge change whenever possible, (b) ensuring an overall neutral simulation box, (c) use of a sufficiently large distance between the solute and the box wall, at least 1 nm, and (d) use of salt if possible to screen finite-size periodicity artifacts~\cite{chen2018accurate}.
-In cases where these recommendations are not practical for specific applications, such as small simulation boxes or when reaction field methods are preferred, charge-change corrections may be applied to reduce artifacts.
-Though both post-processing corrections and the co-alchemical ion approach come at a cost of increased uncertainties.
-The Poisson-Boltzmann calculations are approximate due to finite-grid sizes and are applied only to a subset of the ensemble.
-Co-alchemical ion approaches introduce additional degrees of freedom and increase sampling requirements.
+Absolute free energy calculations can also potentially avoid changing the charge of the system by making a charge perturbation of equal and opposite sign elsewhere in the system; for example, as a charged ligand is removed, a charged counterion of opposite sign could also be removed, or one of the same sign could be inserted.
+This is sometimes referred to as an "alchemical ion" approach for dealing with the needed charge change, and is also employed by the Yank free energy package~\cite{wang2013identifying}.
+Charge corrections have also been explored, and are potentially a viable solution to this problem~\cite{mey2018impact} where artifacts introduced by finite-size effects are corrected numerically~\cite{chen2018accurate, ohlknecht2020correcting}.
+However, application of such corrections typically remains less common than the use of a co-alchemical ion.
+A third approach has been proposed by Gapsys et al.~\cite{gapsys2015calculation} which uses a double-system/single box setup.
+
+When free energy calculations \emph{do} need to change the charge of a ligand or solute, the literature does not yet seem to indicate what approach should be preferable, so considerable care should be taken.
+We are not yet aware of a careful comparison of charge corrections versus other approaches such as decoupling an ion at the same time, so in our view the issue of proper handling of charge mutations in the context of alchemical calculations remains a research problem, and several papers give good guides for exploring the problem further~\cite{lin2014overview, ohlknecht2020correcting, rocklin2013calculating}.
+
+\subsubsection{Handling poorly sampled water networks}
+The location of any water molecules within the binding site of interest must be determined prior to carrying out a free energy calculation.
+Experimental structures, such as those determined by X-ray crystallography and cryo-EM, are often sufficient for this, provided the resolution is good enough.
+However, we recommend using grand canonical sampling methods\cite{melling2023enhanced, samways2020grand} as a means of verifying the experimental water placement and determining any potential differences owing to the classical forcefield used.
+Many other water placement methods are available, all varying in their methodology, computational cost and accuracy.
+Details of these methods can be found in the publication from Samways et al.\cite{samways2021water}.
+
+In relative binding free energy calculations, if the hydration patterns differ between the two ligand endpoints then careful consideration will need to be given to the adaptation of the water network during the free energy calculation.
+We recommend performing the free energy calculation in the conventional way, i.e.
+with no enhanced water sampling, to determine whether the water network can adapt to the ligand perturbation itself.
+This is likely to happen if the hydration site is not particularly occluded from bulk solvent and water molecules can move freely from bulk into the binding site.
+Similarly, performing calculations in both directions (i.e.
+using both endpoints as the starting structure) can aid understanding the effect of the water network on the estimated binding affinities.
+Replica exchange moves in lambda space, instigated from correctly equilibrated water distributions at the end points, may be sufficient to sample water distributions appropriately across a ligand perturbation.
+
+If a conventional sampling method, such as molecular dynamics or Monte Carlo, is unable to rearrange the water network with sufficient efficacy to accommodate the perturbation, then an enhanced sampling method will be required to achieve accurate and converged results.
+One such approach is to include an absolute free energy calculation of the water molecule into the thermodynamic cycle\cite{ross2020enhancing, hamelberg2004standard, barillari2007classification, yu2008free, ge2022abfewat} such that the free energies of both the ligand transformation, and the decoupling of any displaced water molecules, are calculated separately and combined to achieve the total free energy change for the process.
+This method can be difficult to use in practice as the water molecule needs to be carefully restrained while being decoupled, and care needs to be taken to ensure water does not diffuse back into the hydration site as the water is decoupled.
+
+Monte Carlo approaches such as the hybrid MC/MD approach proposed by Bergazin et al.\cite{bergazin2021enhancing, ben-shalom2021fast, ge2022enhancing} allow waters to move between bulk solvent and the binding site with greater frequency by proposing MC moves that translate the water molecules within a spherical region that encompasses both bulk solvent and the binding site.
+This avoids the difficulty of performing a decoupling calculation on the bound waters but can be time-consuming and risks computational time being wasted on move proposals that are not of interest.
+
+Grand canonical (GC) methods propose the addition and removal of water molecules to and from a binding site in accordance with a defined chemical potential.
+Several implementations of GC moves being coupled to RBFE calculations exist, such as in Schrodinger's FEP+ package, introduced by Ross et al.\cite{ross2020enhancing} with moves proposed both within the binding site (to ensure the correct hydration pattern) and also within bulk solvent (to enhance the sampling of the bulk density).
+A cavity-bias technique was used for the binding site moves.
+Grand canonical alchemical perturbation (GCAP)\cite{brucemacdonald2018ligand} is also available in ProtoMS\cite{woods2018protoms}, whereby 2-dimensional titrations are performed in both lambda and chemical potential space.
+However, given ProtoMS only allows MC sampling to be performed, the configurational sampling of the protein is restricted.
+
+More recently, an OpenMM-based implementation allows for GC moves to be carried out during both equilibrium and nonequilibrium (NES) RBFE calculations\cite{melling2025developing}.
+These moves are performed in the form of grand canonical nonequilibrium candidate Monte Carlo (GCNCMC) moves\cite{melling2023enhanced} and are run during the equilibration phase at each lambda window in the case of equilibrium FEP calculations, or during the nonequilibrium switch in the NES calculations.
+
+Although often water sampling is less of an issue in ABFE calculations, handling changes to water networks in buried pockets during ligand coupling/decoupling can be addressed in a similar fashion to RBFEs with grand canonical methods.
+
+% in ABFE calculations is less established, although in principle, the same rules apply.
+When conventional sampling of water is insufficient such that water does not efficiently bind during ligand decoupling, performing extra ABFE calculations for the removal or addition of water molecules to the thermodynamic cycle is one way of ensuring the contribution of these solvent effects is captured.
+Performing grand canonical water moves along side a more traditional ABFE calculation of a ligand will ensure that any changes to binding pocket hydration states are sampled \textit{in situ}.
+
+Employing the techniques described above with care should allow the water network to adapt during the perturbation of the ligand, ensuring more accurate and converged affinity estimations that are independent of the starting configuration and direction in which the calculation is performed.
 

### 2025-07-04 paper/manuscript.tex

--- 2025-06-30 paper/manuscript.tex
@@
 Binding site water molecules may form water mediated protein-ligand interactions which can pose challenges whenever exchange with bulk water is slow compared to simulation timescales.
-This happens typically in buried binding sites.
-Overlaying multiple protein X-ray structures can identify conserved or additional water molecules that can be useful to include in calculations.
-In cases where water molecules are known to play an important role in the binding, software implementations that use water sampling facilitated by Grand Canonical Monte Carlo methods may be useful~\cite{michel2010prediction}.
-Other tools such as WaterMap or open source equivalents (SSTMap, GIST, and others) can be used to define water structure for systems with no experimental evidence of water sites~\cite{wang2011ligand}.
+This can happen in buried binding sites~\cite{laage2017water}.
+Overlaying multiple protein X-ray structures can identify conserved water molecules or those that are associated only with certain ligands in a series and are therefore important to include in calculations.
+In cases where water molecules are known to play an important role in ligand binding, for example if the presence of electron density suggests they are tightly bound (indicated by a high EDIA score~\cite{nittinger2015evidence,meyder2017estimating}), software implementations that use water sampling facilitated by Grand Canonical Monte Carlo methods may be useful~\cite{michel2010prediction}.
+Other tools such as WaterMap~\cite{abel2008role,young2007motifs} or open source equivalents (SSTMap~\cite{haider2018solvation}, GIST~\cite{ramsey2016solvation}, and others) can be used to define the hydration states for systems with no experimental evidence of water sites~\cite{wang2011ligand}.
 Well-known protein systems with water mediated ligand interactions are for example: HSP90 which formed part of the D3R grand challenge 2015~\cite{mey2016blinded}, A2A~\cite{brucemacdonald2018ligand}, MUP~\cite{ross2015water}, ~\cite{deflorian2020accurate}, and others~\cite{michel2009energetics}.
+
+In the context of relative binding free energy calculations, determining the correct hydration pattern for the two endpoints may not be sufficient for an accurate and reproducible calculation.
+If the locations of the bound water molecules differs at the two endpoints - e.g.
+if a perturbation to the ligand displaces a bound water molecule - then the network needs to be able to adapt during the perturbation.
+Otherwise, it is likely that the estimated relative binding free energies will be dependent on the direction in which the perturbation is performed~\cite{ross2020enhancing}.
+
+In absolute binding free energy calculations, a ligand is fully decoupled from a binding site cavity which may otherwise be hydrated in the \textit{apo} state.
+In this case, the sampling of water molecules must be sufficient to hydrate the binding pocket as the ligand is decoupled (or vice versa for the reverse).
+When the binding site is exposed and water can move freely between the site and bulk solvent, then sampling is often not an issue.
+However, in the case of more buried sites or sites which form complex water networks, enhanced water sampling may be required to ensure an accurate binding affinity.
 
@@
 
-An absolute binding free energy calculation yields the standard free energy of formation of the molecular complex from the separated receptor and ligand species.
-Hence, as discussed in \ref{sec:non-covalent-molecular-binding}, it must depend on the standard state concentration $C^\circ$ (here $1 \, {\rm M} = 1 \, {\rm molecule}/1660$~\r{A}$^3$) and on the definition of the complex, that is on the specific binding pose for which the binding free energy is requested.
-Omitting the standard state concentration would yield an undefined result that would vary with the choice of concentration units.
-A similar vague estimate would result from omitting the binding pose definition that, for example, could equally describe binding to a specific site of the receptor or unspecific binding throughout the receptor surface depending on the extent of conformational sampling (see Sec.~\ref{sec:weak-binders}).
-
-In the statistical mechanics formulation of Gilson et al.~\cite{gilson1997statisticalthermodynamic}, described in Section \ref{sec:non-covalent-molecular-binding}, the standard state concentration appears explicitly in Eq.~(\ref{eq:DG0-binding-constant-def}) and the binding pose is defined by the indicator function $I(\zeta)$ that characterizes the partition function of the complex; Eq.~(\ref{eq:intra-zRL-def}).
-The role of the standard state and binding site definitions are particularly evident when expressing Eq.~(\ref{eq:DG0-binding-constant-def}) in terms of an ensemble average.
-To do so, we express the integrals in the numerator and denominator of Eq.~(\ref{eq:DG0-binding-constant-def}) over the same set of degrees of freedom.
-Note that the intramolecular configurational partition function of the complex, $z_{RL(v)}$, includes six more degrees of freedom--the position and orientation of the ligand relative to the receptor collectively described by the variable $\zeta$--than the product of the intramolecular partition functions, $z_{R(v)} z_{L(v)}$, of the separated receptor and ligand.
-To correct this discrepancy, we multiply and divide the right-hand side of Eq.~(\ref{eq:DG0-binding-constant-def})  by the quantity
-\begin{equation} \label{eq:Vsite-def}
-    V_{\rm site} \Omega = \int d\zeta I(\zeta) \, ,
-\end{equation}
-%\Omega is in my opinion a better symbol of the orientational extent than \xi_L. Lowercase greek letters are usually used
-%for dynamical variables rather than integrated quantities
-which measures the extent of the binding pose macrostate, where $V_{\rm site}$ corresponds to the spatial extent and $\Omega$ the orientational extent.
-The right-hand side of Eq.~(\ref{eq:Vsite-def}) combined with the integrals $z_{R(v)}$ and $z_{L(v)}$ yields a partition function similar to Eq.~(\ref{eq:intra-zRL-def}) but corresponding to a complex in which ligand and receptor are uncoupled~\cite{gallicchio2011recent} as if they were at infinite distance apart.
-The result is
-\begin{equation}\label{eq:DG0-binding-constant-def-average}
-K_b = C^\circ V_{\rm site} \frac{\Omega}{8 \pi^2} \langle e^{-\beta \Delta \Psi}  \rangle_0 
-\end{equation}
-where, similarly to Eq.~(\ref{eq:zratio-solv-average}),
-\begin{equation}\label{eq:pert-energy-def}
-\Delta \Psi(x_R, x_L, \zeta) =  \Psi_v(x_R, x_L, \zeta) - \left[  \Psi_v(x_R) + \Psi(x_L) \right]
-\end{equation}
-is the solvent-averaged binding energy of the complex when receptor and ligands are in intramolecular configurations $x_R$ and $x_L$, the ligand is in position and orientation $\zeta$ relative to the receptor (the uncoupled potential energy does not depend on  the position and orientation of the ligand relative to the receptor), and $\langle \ldots \rangle_0$ represents an ensemble average in the uncoupled ensemble.
-Finally, note that $\Omega = 8 \pi^2$ when the binding pose indicator function does not depend on the orientation of the ligand and Eq.~(\ref{eq:DG0-binding-constant-def-average}) simplifies to
-\begin{equation}\label{eq:DG0-binding-constant-def-average-noOmega}
-K_b = C^\circ V_{\rm site} \langle e^{-\beta \Delta \Psi}  \rangle_0 
-\end{equation}
-which, expressed in free energy units, yields
-\begin{eqnarray}\label{eq:DG0-def-Vsite}
-  \Delta G^\circ_b &=&  - k_B T \ln K_b = - k_B T \ln C^\circ V_{\rm site} - k_B T \ln \langle e^{-\beta \Delta \Psi}  \rangle_0 \nonumber \\
-{\rm \ }  &=& \Delta G^\circ_{b,{\rm ideal}} + \Delta G_{b,{\rm exc.}}
-\end{eqnarray}
-
-The term $\Delta G^\circ_{b,{\rm ideal}} = - k_B T \ln C^\circ V_{\rm site}$ in Eq.~(\ref{eq:DG0-def-Vsite}) is interpreted as the \emph{ideal} component of the standard binding free energy, that is the binding free energy that would be measured in the absence of ligand-receptor interactions ($\Delta \Psi = 0$) and when the ligand's concentration is $C^\circ$.
-The ideal term is responsible for the dependence of the standard binding free energy on the standard state concentration.
-The arguably inappropriate term ``standard state correction'' is sometimes used in the literature to refer to $\Delta G^\circ_{b,{\rm ideal}}$.
-As seen above, the ideal term is a fundamental aspect of the statistical mechanics theory of molecular binding rather than a mere correction to an inexact formulation.
-
-Conversely, the term $\Delta G_{b,{\rm exc.}} = - k_B T \ln \langle e^{-\beta \Delta \Psi}  \rangle_0 $ is the \emph{excess} component of the standard free energy of binding, which is the term that depends on intermolecular interactions and the specific chemical composition of the system.
-The excess component is the target of the alchemical molecular simulations discussed here, in contrast to the ideal term that is generally evaluated analytically~\cite{boresch2003absolute}.
-
-\paragraph{Handling the standard state in free energy calculations.}
-As discussed in Sections \ref{sec:theory-solvation} and \ref{sec:theory-partitioning}, the handling of the standard state in solvation and solvent partitioning free energy calculations is typically straightforward.
-Eq.~(\ref{eq:DG0-solvation-def}) identifies the ideal and excess components of the standard solvation free energy in terms of the standard gas pressure $P^\circ$, standard solution concentration $C^\circ$, and ensemble averages typically evaluated by alchemical free energy simulations.
-These relations provide simple recipes to relate computational estimates to experimental values obtained under different conditions.
-Moreover, typical experimental reporters (Henry's constants and partitioning coefficients) directly probe the excess term calculated by molecular simulations, which is independent of the standard state.
-
-As discussed above, the situation for binding is more complex and requires special care.
-The first important aspect is that binding free energy calculations must be performed with binding site restraints that implement the binding site indicator function $I(\zeta)$ by forcing the ligand to occupy the binding site even when it is decoupled from the receptor.
-Doing otherwise would fail to specify the bound macrostate of the complex and return undefined binding free energy estimates.
-However, in practice, calculations lacking binding site restraints often return reliable binding free energy estimates due to the kinetic trapping of the ligand in the binding site region during the relatively short molecular dynamics sampling typical of alchemical binding free energy simulations.
-Nevertheless, longer calculations could display persistent drift of the estimate as new regions of conformational space are sampled.
-A more serious consequence of neglecting binding site restraints is the possibility of the ligand leaving the binding site when it is not coupled to the receptor.
-Such occurrence indicates a serious violation of the theory.
-Alchemical binding free energy simulations where the ligand leaves the binding site are not set up correctly and can return grossly inaccurate binding free energy estimates.
-
-Because it is a discontinuous function of the position and orientation of the ligand relative to the receptor, molecular dynamics implementations often approximate the indicator function by means of flat-bottom restraints~\cite{chen2007can, wu2021alchemical} (see below).
-Harmonic restraints are not as suitable in this respect because they favor a specific microstate at the bottom of the harmonic potential rather than specifying a region of conformational space.
-
-Note that $\Delta G^\circ_{b,{\rm ideal}}$ is independent of the chemical nature of the binding partners, and it is the same for all complexes sharing similar binding site definitions.
-The latter is the reason that the ideal term cancels out when taking the difference of absolute binding free energies when estimating relative binding free energies (RBFEs).
-However, while it does not appear explicitly in the definition, $\Delta G_{b,{\rm exc.}}$ depends implicitly on the definition of the binding site through the ensemble average in the uncoupled ensemble, which includes only ``bound'' configurations according to the definition of the indicator function $I(\zeta)$.
-Varying the binding site definition affects both the ideal and excess binding free energy terms such that their sum (the standard binding free energy) is often only weakly dependent on the binding site definition~\cite{gilson1997statisticalthermodynamic,gallicchio2011recent}.
-Conversely, great care should be exercised when comparing calculated excess binding free energy values corresponding to different binding site definitions.
-Similarly, the outcome of a binding free energy calculation conducted without a binding site definition, i.e.\ without binding site restraints, is, in principle, undefined.
-The latter is also true for relative binding free energies because while the ideal term cancels out, the difference in excess binding free energies depends on the binding pose definition.
-
-\paragraph{Several choices of restraints are possible in binding free energy calculations.}
-A variety of restraint types are commonly used in alchemical binding free energy calculations.
-Before going into their characteristics, we must acknowledge an important and often unrecognized distinction between the binding site restraints corresponding to the indicator function $I(\zeta)$ necessary for the implementation of the theory and optional auxiliary restraints and biasing potentials sometimes used to speed up the convergence of free energy calculations.
-Binding site restraints are an integral part of the molecular complex definition and are typically maintained throughout the alchemical process.
-In particular, they must be present unchanged at the unbound and bound states.
-In contrast, unless thermodynamic reweighting is applied in post-processing, auxiliary restraints cannot be present at the endstates to avoid affecting their free energies.
-Hence, typically the free energy cost of turning them on is computed at the uncoupled state, and the free energy gain of releasing them is calculated near the bound state.
-In the alchemical binding free energy literature, often harmonic restraint potentials imposed in an initial stage and removed in a final stage are described as implementing binding sites restraints.
-However, because they utilize harmonic potentials that cannot correctly describe the binding pose macrostate (see above) and because they are turned on and then off, these are instead auxiliary restraints introduced to enhance convergence without affecting, in principle, binding free energy estimates.
-
-Binding site restraints implement the $I(\zeta)$ indicator function.
-They, therefore, can depend at most on the six rigid-body external coordinates of the ligand governing the position and orientation of the ligand relative to the receptor.
-Any other choice would affect the intramolecular conformational distribution of the ligand or the receptor and would  bias the binding free energy estimate.
-The external degrees of freedom introduced by Boresch et al.~\cite{boresch2003absolute,leitgeb2005alchemical} are a popular choice.
-The Boresch coordinates, as well as other similar approaches, involve selecting anchor atoms of the receptor and the ligand.
-In principle, the choice of anchor atoms is arbitrary, provided adequate simulation time is available, and as long as it leads to similar binding pose definitions.
-However, practical considerations may be important.
-For example, some choices of Boresch-style anchor atoms can be numerically unstable.
-Additionally, anchor atoms should likely be placed in a part of the molecule that defines the binding orientation well rather than in a floppy, solvent-exposed tail.
-Several automated methods have been proposed to select anchor points for Boresch restraints, which typically involve analysing a simulation of the fully-interacting receptor-ligand complex~\cite{alibay2022evaluating, alibay2021mdrestraintsgenerator, baumann2023broadening, chen2023enhancing, hedges2023suite, wu2025optimizing}.
-
-As discussed above, binding site restraints are commonly implemented using quadratic flat-bottom restraints, which impose a restoring harmonic  force similar to harmonic potentials but only if the ligand leaves a specific region~\cite{chen2007can}.
-Often, binding site restraints do not limit the rotational degrees of freedom of the ligand, and the corresponding standard state term is zero; see Eq.~(\ref{eq:DG0-def-Vsite}).
-However, it might be challenging to fully explore the bound macrostate at all alchemical states when it is only limited by receptor-ligand positional restraints.
-As further discussed below, auxiliary restraints are particularly helpful in the latter circumstance to limit the extent of conformational sampling necessary to achieve rapid convergence of binding free energy estimates.
-
-Most current binding free energy workflows do not use binding site restraints present at all stages of the simulation and implicitly assume that $V_{\rm site}$ in Eq.~(\ref{eq:Vsite-def}) is the volume of the simulation box $V$.
-In the absence of binding site restraints, only the protein-ligand interactions prevent the ligand from slipping out from the binding site and wandering throughout the simulation box within the simulation time frame.
-However, observations of ligands leaving the binding site become more likely as simulation times increase, especially when analyzing weak binders and non-binders (see below).
-Moreover, there is no obstacle for the ligand to diffuse away from the binding site at the decoupled state when protein-ligand interactions are turned off.
-
-In practice, because they often implement auxiliary restraints, binding free energy prediction workflows that omit binding site restraints rarely yield grossly erroneous predictions.
-Auxiliary restraints are introduced gradually as protein-ligand interactions are turned off or are introduced at the bound state when the ligand is naturally trapped in the binding site.
-Either way, the free energy of imposing them can be tracked while preventing the ligand from leaving the binding site at the decoupled state.
-Provided that the interacting ligand samples only bound configurations, i.e., it behaves as if binding site restraints were present, the current practice of not using them is likely to incur only minor or insignificant errors.
-
-The Boresch harmonic restraints~\cite{boresch2003absolute,baumann2023broadening} are a popular choice as auxiliary restraints because they are based on interatomic distances, bond angles, and dihedral angles degrees of freedom already implemented in molecular mechanics codes that do not interfere with intramolecular conformational distributions.
-Moreover, the free energy of releasing Boresch restraints within the simulation box volume $V$ is easily computed analytically~\cite{boresch2003absolute, boresch2024analytical}.
-
-By reducing the conformational space to be sampled, auxiliary restraints are often crucial for obtaining converged binding free energy estimates, even when binding site restraints are employed.
-Auxiliary restraints take a variety of forms, such as distance restraints between the ligand and the protein~\cite{mobley2006use,clark2023comparison} as well as flat-bottom restraints.
-The choice of degrees of freedom and restraining potentials for auxiliary restraints is not as restrictive as for those for binding site restraints, provided that the free energy terms of restraining collective variables that couple protein and ligand degrees of freedom are adequately accounted for to avoid serious systematic errors~\cite{clark2023comparison}.
-For example, the overall ligand RMSD has been used~\cite{woo2005calculation} for this purpose.
-Auxiliary restraints are designed to reduce the amount of conformational sampling at intermediate $\vec{\lambda}$ values.
-However, more computational effort is required to compute the free energy when the restraints are turned on and off.
-Additionally, such restraints can limit the exploration of alternative binding modes.
+Absolute free energy calculations involve completely removing the interactions between the ligand or solute and its environment, taking it to a non-interacting state that may or may not retain intramolecular non-bonded interactions.
+This non-interacting state can then be shifted between environments---from the protein to water, or from one solution to another---without changing its free energy other than that due to the changing volume of the simulations, and then interactions can be restored in the new environment.
+
+Absolute free energies are by definition reported with respect to a specific reference or standard state, which effectively determines the arbitrary point at which the free energy is 0.
+The role of the standard state is particularly evident from the expression of the binding free energy between a receptor $R$ and ligand $L$
+\begin{equation} \label{eq:DGfromKAB}
+    \Delta G = -k_BT ~ \ln \left( c^{\circ} K_b \right)  = -k_BT ~ \ln\left( c^{\circ} \frac{[RL]}{[L][R]} \right) \, .
+\end{equation}
+Here, the reference state concentration $c^{\circ}$ converts the binding constant $K_b$ into a dimensionless quantity expressed in reference concentration units.
+It should be noted that ignoring the term $c^{\circ}$ is equivalent to assuming a reference concentration of 1~D$^{-1}$, where D are the units used to express $K_b$, and would thus cause the value of $\Delta G$ to vary with the choice of the units.
+It is convenient to define a standard state at a constant pressure of 1~atm and where each chemical species (i.e., R, L, and RL) in the reaction solvent has a concentration of $c^{\circ}$~=~1~M~=~1~molecule/1660~\r{A}$^3$ but do not interact with other molecules of $R$, $L$, or $RL$.
+
+\paragraph{Handling the standard state in absolute free energy calculations.}
+For solvation free energy calculations, handling the standard state is typically straightforward, and treating it correctly simply means ensuring that the non-interacting solute still occupies essentially the same volume as the solute in the interacting system.
+So typically in such cases no special care is required to ensure the correct standard state, as long as the \emph{experimental} data being analyzed uses the same standard state.
+If this is not the case, a simple entropic correction to the free energy of $k_B T\ln (V_f/V_i) = k_B T \ln (C_i/C_f)$ to the experimental data is needed.
+
+For binding, however, the situation is more complex and requires special care.
+Because the simulations are typically performed using restraints and at concentrations that are different from 1 M, the expression of the free energy requires the following correction~\cite{gilson1997statisticalthermodynamic} (see an example of such a thermodynamic cycle in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle})
+\begin{equation}\label{eq:restraint-correction}
+    \Delta G^{\circ}_{\mathrm{restr}} = -k_BT ~ \ln \left( c^{\circ}{V_L} \right) -k_BT ~ \ln \left( \frac{\xi_L}{8 \pi^2} \right),
+\end{equation}
+where $V_L$ and $\xi_L$ are respectively the volume of the translational and rotational degrees of freedom of the non-interacting ligand in the simulation box.
+When no restraints are used, the non-interacting ligand is free to translate and rotate in the simulation box (i.e., $V_L = V_{\mathrm{box}}$ and $\xi_L = 8\pi^2$), and the rotational term is zero.
+A sufficiently thorough exploration of the simulation box by the non-interacting ligand is, however, required for the formula to be valid.
+This is typically hard to achieve as the exploration process is governed by diffusion, and weak transient nonspecific binding will occur at other sites on the protein. 
+The addition of a restraint limits the volume available to the non-interacting ligand, thus speeding the convergence of the sampling.
+In addition, when enhanced sampling methods such as Hamiltonian $\lambda$ exchange are used (see Sec.~\ref{sec:sampling_schemes}), the use of a restraint is typically necessary as it keeps the ligand in the binding site in the interacting state (see also Sec.~\ref{sec:weak-binders}) and generally reduce the round-trip time of replicas.
+When restraints are employed, the values of $V_L$ and $\xi_L$ are restraint-dependent, but for commonly employed restraints, these can be usually easily computed analytically or numerically by solving the relevant integral.
+
+\paragraph{Several choices of restraints are possible.}\label{ABFErest}
+In practice, a variety of types of restraints are common, from simple harmonic distance restraints between the ligand and the protein~\cite{mobley2006use}, to flat-bottom restraints which work similarly but only exert a force if the ligand leaves a specific region~\cite{chen2007can}.
+Because these restraints do not limit the rotational degrees of freedom of the ligand, the rotational term entering the correction in Eq.~\ref{eq:restraint-correction} is zero.
+
+Alternatively, a set of restraints proposed by Boresch have also commonly been employed, where all six rigid-body degrees of freedom governing the orientation of the ligand relative to the receptor are restrained~\cite{boresch2003absolute, leitgeb2005alchemical}.
+Further restraints, such as on the overall ligand RMSD have also been used~\cite{woo2005calculation}.
+
+In principle, all of these forms will yield correct binding free energies in the limit of adequate sampling if their effects and connection to the standard state are correctly handled, but they have different strengths and weaknesses.
+For example, with more involved restraints, sampling at intermediate $\vec{\lambda}$ values will usually not need to be as extensive but more computational effort must go to computing the free energy to turn on the restraints.
+Additionally, such restraints would typically keep the ligand from exploring alternative binding modes.
 This  restriction may be undesirable when using Hamiltonian $\lambda$ exchange or expanded ensemble techniques where allowing the ligand to exchange binding modes when it is non-interacting could provide sampling benefits~\cite{wang2013identifying}.
-More specifically, flat-bottom restraints allow a ligand to explore multiple binding sites, and harmonic restraints allow the exploration of multiple binding modes within a site, while Boresch restraints only allow a single binding mode within a single site.
+More specifically, flat-bottom restraints might allow a ligand to explore multiple binding sites, and harmonic restraints allow exploration of multiple binding modes within a site, while Boresch restraints only allow a single binding mode within a single site.
 See additional discussion of the possibility of multiple binding modes in Sec.~\ref{sec:multiple_binding_modes} below.
 
-Recently, degrees of freedom for binding site and auxiliary potentials based on larger sets of atomic positions of the receptor and ligand have been put forward.
+Many choices of restraints involve selecting reference atoms.
+Again, in principle this choice is unimportant given adequate simulation time but practical considerations may be important.
+The choice is likely especially important with Boresch-style restraints, where some relative placements of reference atoms are likely to be numerically unstable; additionally, ligand reference atoms should likely be in a part of the molecule which defines the binding orientation well, rather than in a floppy solvent-exposed tail, for example.
+Several automated methods have been proposed to select anchor points for Boresch restraints, which typically involve analysing a simulation of the fully-interacting receptor-ligand complex~\cite{alibay2022evaluating, alibay2021mdrestraintsgenerator, baumann2023broadening, chen2023enhancing, hedges2023suite}.
+
+More recently, restrictive restraint schemes have been proposed that avoid relying on a few critical reference atoms.
+Instead, these aim to simplify restraint selection by using larger sets of atomic positions in the receptor and ligand.
 Fu et al.
-suggested restraining the six relative rigid-body degrees of freedom derived by finding the rotation of the ligand, which minimises the RMSD to a reference structure of the receptor-ligand complex (after accounting for rotation and translation of the receptor)~\cite{fu2017new}.
+suggested restraining the six relative rigid-body degrees of freedom derived by finding the rotation of the ligand which minimises RMSD to a reference structure of the receptor-ligand complex (after accounting for rotation and translation of the receptor)~\cite{fu2017new}.
 Salari et al.
 proposed restraining the ``distance-from-bound-configuration'' (DBC) variable, which is the RMSD of ligand coordinates in the frame of reference of the binding site~\cite{salari2018streamlined, ebrahimi2022symmetry}.
-DBC may allow the binding pose to be closely preserved as the ligand intermolecular interactions are removed, but has the disadvantage of coupling the internal and relative external degrees of freedom of the receptor and ligand, making them unsuitable for describing the binding site macrostate and the corresponding standard state terms.
-However, these restraints are practical as auxiliary restraints when additional steps are used to introduce and release them.
-Moreover, because they are based on flat-bottom potentials encompassing the desired binding poses, the free energy of releasing them in the bound state can be safely neglected in most cases.
-The restraint schemes discussed here are implemented in open-source workflows~\cite{fu2021bfee2, fu2022accurate, santiago-mcrae2023computing, clark2023comparison, hedges2023suite}.
+This may allow the binding pose to be closely preserved as the ligand intermolecular interactions are removed, but has the disadvantage of coupling the internal and relative external degrees of freedom of the receptor and ligand, preventing easy computation of the standard state correction.
+This necessitates an additional step where the DBC restraints are released to a single harmonic restraint.
+Both restraint schemes are implemented in open-source workflows~\cite{fu2021bfee2, fu2022accurate, santiago-mcrae2023computing}.
 
 \subsection{Absolute and relative calculations deal with some of the same issues}
-
-\paragraph{Handling weak binders}\label{sec:weak-binders}
-According to the non-covalent statistical mechanics theory of bimolecular binding (Sec.~\ref{sec:theory}), only the conformations in which the receptor and ligand form a bound complex are sampled.
-Determining an appropriate definition of the bound macrostate can be challenging for weakly bound ligands.
-For tightly bound ligands, virtually all reasonable definitions of the bound state will lead to equivalent free energies since the partition function will be dominated by a relatively limited ensemble of low-energy poses.
-However, this simplification breaks down for weak binders.
-In fact, there might not be a perfect correspondence between the structural definition of the complex on which current alchemical models are based~\cite{gilson1997statisticalthermodynamic} (Sec.~\ref{sec:theory}) and experimental reporters~\cite{mihailescu2004theory}.
-For example, isothermal titration calorimetry (ITC) or surface plasmon resonance (SPR) measurements effectively define a binding state that encompasses all ligand conformations complexed with the protein, regardless of where they bind the protein.
+\subsubsection{Handling weak binders and high dissociation rates}\label{sec:weak-binders}
+In binding free energy calculations, only the conformations in which the receptor and ligand form a bound complex should be sampled from the bound states (Sec.~\ref{sec:theory}).
+Determining what the bound states actually are can be challenging for weakly bound ligands.
+For tightly bound ligands, virtually all reasonable definitions of the bound state will lead to be equivalent free energies, since the partition function will be dominated by a relatively small number of low-energy poses.
+
+For weak binders, this simplification breaks down.
+In fact, the correct bound state may depend on the type of experiment performed.
+For example, isothermal titration calorimetry (ITC) or surface plasmon resonance (SPR) measurements effectively define a binding state that includes all ligand comformations that are complexed with the protein, regardless of where on the protein they bind.
 In contrast, fluorescence polarization competition assays measure binding to only a single location, where the ligand of interest displaces a competing binder.
@@
 Co-alchemical ion approaches introduce additional degrees of freedom and increase sampling requirements.
+
+\subsubsection{Handling poorly sampled water networks}
+The location of any water molecules within the binding site of interest must be determined prior to carrying out a free energy calculation.
+Experimental structures, such as those determined by X-ray crystallography and cryo-EM, are often sufficient for this, provided the resolution is good enough.
+However, we recommend using grand canonical sampling methods\cite{melling2023enhanced, samways2020grand} as a means of verifying the experimental water placement and determining any potential differences owing to the classical forcefield used.
+Many other water placement methods are available, all varying in their methodology, computational cost and accuracy.
+Details of these methods can be found in the publication from Samways et al.\cite{samways2021water}.
+
+In relative binding free energy calculations, if the hydration patterns differ between the two ligand endpoints then careful consideration will need to be given to the adaptation of the water network during the free energy calculation.
+We recommend performing the free energy calculation in the conventional way, i.e.
+with no enhanced water sampling, to determine whether the water network can adapt to the ligand perturbation itself.
+This is likely to happen if the hydration site is not particularly occluded from bulk solvent and water molecules can move freely from bulk into the binding site.
+Similarly, performing calculations in both directions (i.e.
+using both endpoints as the starting structure) can aid understanding the effect of the water network on the estimated binding affinities.
+Replica exchange moves in lambda space, instigated from correctly equilibrated water distributions at the end points, may be sufficient to sample water distributions appropriately across a ligand perturbation.
+
+If a conventional sampling method, such as molecular dynamics or Monte Carlo, is unable to rearrange the water network with sufficient efficacy to accommodate the perturbation, then an enhanced sampling method will be required to achieve accurate and converged results.
+One such approach is to include an absolute free energy calculation of the water molecule into the thermodynamic cycle\cite{ross2020enhancing, hamelberg2004standard, barillari2007classification, yu2008free, ge2022abfewat} such that the free energies of both the ligand transformation, and the decoupling of any displaced water molecules, are calculated separately and combined to achieve the total free energy change for the process.
+This method can be difficult to use in practice as the water molecule needs to be carefully restrained while being decoupled, and care needs to be taken to ensure water does not diffuse back into the hydration site as the water is decoupled.
+
+Monte Carlo approaches such as the hybrid MC/MD approach proposed by Bergazin et al.\cite{bergazin2021enhancing, ben-shalom2021fast, ge2022enhancing} allow waters to move between bulk solvent and the binding site with greater frequency by proposing MC moves that translate the water molecules within a spherical region that encompasses both bulk solvent and the binding site.
+This avoids the difficulty of performing a decoupling calculation on the bound waters but can be time-consuming and risks computational time being wasted on move proposals that are not of interest.
+
+Grand canonical (GC) methods propose the addition and removal of water molecules to and from a binding site in accordance with a defined chemical potential.
+Several implementations of GC moves being coupled to RBFE calculations exist, such as in Schrodinger's FEP+ package, introduced by Ross et al.\cite{ross2020enhancing} with moves proposed both within the binding site (to ensure the correct hydration pattern) and also within bulk solvent (to enhance the sampling of the bulk density).
+A cavity-bias technique was used for the binding site moves.
+Grand canonical alchemical perturbation (GCAP)\cite{brucemacdonald2018ligand} is also available in ProtoMS\cite{woods2018protoms}, whereby 2-dimensional titrations are performed in both lambda and chemical potential space.
+However, given ProtoMS only allows MC sampling to be performed, the configurational sampling of the protein is restricted.
+
+More recently, an OpenMM-based implementation allows for GC moves to be carried out during both equilibrium and nonequilibrium (NES) RBFE calculations\cite{melling2025developing}.
+These moves are performed in the form of grand canonical nonequilibrium candidate Monte Carlo (GCNCMC) moves\cite{melling2023enhanced} and are run during the equilibration phase at each lambda window in the case of equilibrium FEP calculations, or during the nonequilibrium switch in the NES calculations.
+
+Although often water sampling is less of an issue in ABFE calculations, handling changes to water networks in buried pockets during ligand coupling/decoupling can be addressed in a similar fashion to RBFEs with grand canonical methods.
+
+% in ABFE calculations is less established, although in principle, the same rules apply.
+When conventional sampling of water is insufficient such that water does not efficiently bind during ligand decoupling, performing extra ABFE calculations for the removal or addition of water molecules to the thermodynamic cycle is one way of ensuring the contribution of these solvent effects is captured.
+Performing grand canonical water moves along side a more traditional ABFE calculation of a ligand will ensure that any changes to binding pocket hydration states are sampled \textit{in situ}.
+
+Employing the techniques described above with care should allow the water network to adapt during the perturbation of the ligand, ensuring more accurate and converged affinity estimations that are independent of the starting configuration and direction in which the calculation is performed.
 

### 2025-07-13 paper/manuscript.tex
Accentmarks fixes.
remove version info for now.
WPI for addressing subsampling issues.

@@
 However, this is not the only metric which can or should be used, as the uncertainty only captures the information about the sampled phase space, not necessarily the entirety of the phase space.  
-For example, convergence of relative free energy calculations in predictive simulations where the entire phase space is not known in advance, requires sampling the different kinetically stable states~\cite{mobley2012perspective}.
+For example, convergence of relative free energy calculations in predictive simulations where the entire phase space is not known in advance requires sampling the different kinetically stable states~\cite{mobley2012perspective}.
 This highlights the importance of choosing the correct thermodynamic path to ensure you sample the required thermodynamic states as discussed in Sec.~\ref{sec:important_path}.
@@
 
-There is one important thing to note about the uncertainty estimates when summing multiple free energies together to calculate an overall free energy estimate.
+There is one important thing to note about the uncertainty estimates when summing multiple free energy differnces together to calculate an overall estimate of a free energy difference.
 Although BAR itself gives a free energy estimate that is asymptotically correct and is much less biased than the uncertainty estimate for EXP, the uncertainties in $\Delta {f_{i-1\rightarrow i}}$ and $\Delta {f_{i\rightarrow i+1}}$ are not uncorrelated, because they both involve the energies $u_i(\vec{q})$.
-The variances of each of the free energies will \textit{not} propagate as variances usually do (in quadrature) into the variance of the overall free energy.
+The variances of each of the free energies cannot simply be summed together into the variance of the overall free energy.
 Instead, some other method for propagating the uncertainty, such as bootstrapping~\cite{grossfield2018best} must be used.
@@
 \paragraph{Computing the statistical inefficiency}
-Most estimators require an uncorrelated set of samples from the equilibrium distribution to produce (relatively) unbiased estimates of the free energy difference and its statistical uncertainty.
+Most estimators require an uncorrelated set of samples from the equilibrium distribution to produce (relatively) unbiased estimates of the statistical uncertainty, as most estimates of the uncertainty are derived assuming unbiased samples.
 To do this, the production region of the simulation is generally \emph{subsampled} with an interval approximately equal to or greater than the \emph{statistical inefficiency} $g \ge 1$ to produce a set of uncorrelated samples that can be fed to the estimator machinery~\cite{chodera2016simple},
@@
 
-If independent simulations are used, the alchemical state $\vec{\lambda}$ may have a significant impact on the correlation time, and these simulations should be subsampled independently using a separate estimate of the statistical inefficiency $g$ for each alchemical state.
-If coupled simulations are used (such as a Hamiltonian replica exchange simulation), the replicas should undergo equivalent random walks in alchemical space, and the replicas can be subsampled with the same $g$ to generate an equal number of uncorrelated samples at each alchemical state.
-Conveniently, the approach described above for automated equilibration detection produces an appropriate estimate of $g$ over the production region for automating this process.
+However, samples uncorrelated by this procedure are only approximately
+uncorrelated.  Recent careful investigation has shown that the lowest
+uncertainty results can be obtained by using \textit{all} of the data
+to obtain an estimate of the free energy itself, but using only
+subsampled data to calculate error estimates using the formulas or by
+using bootstrapping.  In fact, one can show that the standard
+deviation of the estimate determined by data that has been subsampled
+is the unbiased estimator of the free energy estimate using all the
+data, but is actually an underestimate of the uncertainty of estimate
+using just the subsampled data.
+
+We don't need to use all of the data to recover the remaining
+correlation.  In practice, if one subsamples the data with a
+correlation interval of $g/10$, where $g$ is the correlation
+time. Thus, if $g=3.0$ in units of timesteps, it is best to use all of
+the data.  If $g=100.0$, then one loses essentially no information by
+subsampling with $g=10.0$, and still saves significant time by not
+using all of the now-completely duplicative data.
+
+If independent simulations at each $\lambda$ are used, the alchemical
+state $\vec{\lambda}$ may have a significant impact on the correlation
+time, and these simulations should be subsampled independently using a
+separate estimate of the statistical inefficiency $g$ for each
+alchemical state. If coupled simulations are used (such as a
+Hamiltonian replica exchange simulation), the replicas should undergo
+equivalent random walks in alchemical space, and the replicas can be
+subsampled with the same $g$ to generate an equal number of
+uncorrelated samples at each alchemical state. Conveniently, the
+approach described above for automated equilibration detection
+produces an appropriate estimate of $g$ over the production region for
+automating this process.
+
+Our overall recommendations are therefore to determine the subsampling
+period $g$, subsample data at $g/10$ (i.e. more frequently than
+specified by $g$) to compute the free energy estimates, and then use
+fully subsampled data (with subsampling interval $g$) to estimate the
+uncertainties of the estimates.
 
@@
 Reliable estimation of the statistical inefficiency is difficult, and estimates will not generally be as precise (in a relative error sense) as averages.
-To ensure there is sufficient data available for reliable decorrelation and estimation of free energy differences, it is recommended that the effective number of uncorrelated samples $N_{\mathrm{eff}} \ge \approx 50$ if the BAR or MBAR estimators (discussed below in sec~\ref{subsec:estimators}) are used; the number may need to be much higher with alternate estimators.
+To ensure there is sufficient data available for reliable decorrelation and estimation of free energy differences, it is recommended that the effective number of uncorrelated samples $N_{\mathrm{eff}} \ge  \sim 50$ if the BAR or MBAR estimators (discussed below in sec~\ref{subsec:estimators}) are used; the number may need to be much higher with alternate estimators.
 
@@
 
-The computation of free energies using TI (Sec.n~\ref{subsec:estimators}) is straightforward and the trapezoidal rule is often recommended since it allows unequal spacing of $\vec{\lambda}$ states, which is required to minimize the variance in the free energy estimate, but in principle any good numerical integration method can be used.
-The determination of regions of high curvature when estimating the integral is helpful to determine regions of phase space where more sampling and/or more $\vec{\lambda}$ states are necessary to obtain the best approximation of the integral.
-Plotting $\vec{\lambda}$ with respect to the gradients at each of the $\vec{\lambda}$ values can be a helpful diagnostic.
-Additionally, computation of the overall variance of TI requires the calculation of the overall variance of integration, rather than each individual $\Delta$G$_{i,i+1}$ and assuming variances add independently.
-Therefore, $\mathrm{var}$($\Delta$f) = $\sum_{i=1}^{K}w_{k}^2 \mathrm{var}(\frac{du}{d\vec{\lambda}})_{k}$.
+Computating of free energies using TI (Sec.n~\ref{subsec:estimators}) is straightforward and the trapezoidal rule is often recommended.
+The trapezoid rule easily allows unequal spacing of $\vec{\lambda}$ states, which is often required to minimize the variance in the free energy estimate, but in principle any good numerical integration method can be used if care is taken.
+Determining of regions of high curvature when estimating the integral is helpful to determine regions of phase space where more sampling and/or more $\vec{\lambda}$ states are necessary to obtain the best approximation of the integral.
+Plotting $\vec{\lambda}$ with respect to the gradients at each of the $\vec{\lambda}$ values, which can be done automatically with tool such as alchemlyb,  can be a helpful diagnostic.
+Additionally, computating the overall variance of TI requires the calculation of the overall variance of the integration, rather than each individual $\Delta$G$_{i,i+1}$ and assuming variances add independently.
+Therefore, $\mathrm{var}$($\Delta$f) = $\sum_{i=1}^{K}w_{k}^2 \mathrm{var}(\frac{du}{d\vec{\lambda}})_{k}$, where the $w_k$ are the integration weights (see Sec~\eref{subsec:estimators} for more discussion of $w_k$).
 

### 2025-07-14 paper/manuscript.tex
added some more detail to subsampling.

@@
 It should be noted however, that uncertainty estimates have the same limitations as other metrics of convergence, as they are only an uncertainty based on the phase space sampled so far in a simulation, and cannot account for states not sampled, and it is worth considering that they will be an underestimation of the true uncertainty.
+
 A target statistical uncertainty should be chosen at the onset of the simulation to avoid excessively long simulations, or falling into the trap of running until the free energy estimate is "good enough," which is subjective and has no defined criteria.
@@
 
-MBAR can also be thought of as the Zwanzig estimator of the free energy to state $i$ where the sampled distribution is the \textit{mixture distribution} of all the other samples thrown together in one ``pot'', defined by $p_m(\vec{q}) = N^{-1} \sum_k N_k \exp(f_k-u_k\vec{q})$, which is the weighted average of all the individual normalized probability distributions from the simulations that are performed~\cite{shirts2017reweighting}.
+MBAR can also be thought of as the Zwanzig estimator of the free energy to state $i$ from a sampled distribution referred to as the \textit{mixture distribution}, where one throws all the samples from all simulations together in one ``pot''.
+Mathematically, this is defined by $p_m(\vec{q}) = N^{-1} \sum_k N_k \exp(f_k-u_k\vec{q})$, which is the average, weighted by the number of samples drawn from each distribution, of all the individual normalized probability distributions from the simulations that are performed~\cite{shirts2017reweighting}.
+
+All of the estimators above have uncertainty estimates associated with them.
+In the limit of sufficiently large data, perhaps several hundred or thousand independent points, exhaustive analyses of model data has shown that the error estimates of BAR, MBAR, and TI are all quite reliable when there is good overlap between intermediate states.
+For EXP, the number of samples at which they become reliable is significantly higher, especially if overlap between intermediate states is lower.
+However, in the limit of moderate or small amounts of data (tens or hundreds of independent samples) these errors become somewhat unreliable, often off by a factor of 2 from what would get if more data was collected.
+These estimator of uncertainty are all derived from propagation of error formulae from the estimators themselves.
+However, when the underlying data is noisy, this approximation is not very good.
+
+Some of the authors' experience has shown that significantly more reliable estimators at lower numbers of samples are obtained by performing bootstrap sampling on data that has been uncorrelated.
+We do not provide a full tutorial of bootstrapping here, but the basic idea is that one takes the $N_{sample}$ data points collected and draws, with replacement, $N_{sample}$ samples.
+This means some of the data points will be represented multiple times, and other data points none.
+This process is then repeated $N_{bootstrap}$ times, where $N$ should generally be at least 50 to up to 200 for good estimates of the standard error of the estimator.
+The data is then \textit{treated} as $N_{bootstrap}$ samples from the true distribution, and statistics such as the standard error are computed from this artificial set of $N_{bootstrap}$ experiments.
+In the limit of $N_{sample}$ large, it can be shown that the bootstrap estimate of the standard error converges to the true population standard error.
+One way this can be seen is that bootstrap estimators of the data are independent samples of the \textit{actual} distribution of the experiment performed, i.e., a bunch of $\delta$ functions at $N_{sample}$ points, rather than being independent samples from the true, usually continuous distribution.
+As more and more data is collected, our collection of $\delta$ functions becomes more and more like the true distribution.
+Another advantage of bootstrap sampling is that one can construct decent estimates of error bounds with arbitrary percentiles, and not just standard errors of the estimates, which assume Gaussian behavior of the underlying error, and that the error results are Gaussian as well.
+
+One important point is that the bootstrap samples use to estimate the standard error should be decorrelated, or else the bootstrap estimate of the standard error will be artificially small.
+More discussion of subsampling and decorrelation is included in section~\ref{sec:decorrelating-samples}.
+Another alternative to bootstrap sampling of individual points that better captures the correlations in the data is block bootstrapping~\cite{???}.
+Block bootstrapping is somewhat more complicated and less common to perform and analyze than bootstrap estimates of the error, so we will not go into details, but it can also provide more reliable estimates of uncertainty than the estimators above.
 
@@
 
+%Cooment: Finlay, can you add description/citation of RED?
+
 \paragraph{Subsampling data to generate uncorrelated samples}
@@
 
-However, samples uncorrelated by this procedure are only approximately
-uncorrelated.  Recent careful investigation has shown that the lowest
-uncertainty results can be obtained by using \textit{all} of the data
-to obtain an estimate of the free energy itself, but using only
-subsampled data to calculate error estimates using the formulas or by
-using bootstrapping.  In fact, one can show that the standard
-deviation of the estimate determined by data that has been subsampled
-is the unbiased estimator of the free energy estimate using all the
-data, but is actually an underestimate of the uncertainty of estimate
+However, samples uncorrelated by this procedure are only approximately uncorrelated.
+Recent careful investigation has shown that the lowest uncertainty results can be obtained by using \textit{all} of the data to obtain an estimate of the free energy itself, but using only subsampled data to calculate error estimates using the formulas or by using bootstrapping.
+In fact, one can show that the standard deviation of the estimate determined by data that has been subsampled is the unbiased estimator of the free energy estimate using all the data, but is actually an underestimate of the uncertainty of estimate
 using just the subsampled data.
 
-We don't need to use all of the data to recover the remaining
-correlation.  In practice, if one subsamples the data with a
-correlation interval of $g/10$, where $g$ is the correlation
-time. Thus, if $g=3.0$ in units of timesteps, it is best to use all of
-the data.  If $g=100.0$, then one loses essentially no information by
-subsampling with $g=10.0$, and still saves significant time by not
-using all of the now-completely duplicative data.
-
-If independent simulations at each $\lambda$ are used, the alchemical
-state $\vec{\lambda}$ may have a significant impact on the correlation
-time, and these simulations should be subsampled independently using a
-separate estimate of the statistical inefficiency $g$ for each
-alchemical state. If coupled simulations are used (such as a
-Hamiltonian replica exchange simulation), the replicas should undergo
-equivalent random walks in alchemical space, and the replicas can be
-subsampled with the same $g$ to generate an equal number of
-uncorrelated samples at each alchemical state. Conveniently, the
-approach described above for automated equilibration detection
-produces an appropriate estimate of $g$ over the production region for
-automating this process.
-
-Our overall recommendations are therefore to determine the subsampling
-period $g$, subsample data at $g/10$ (i.e. more frequently than
-specified by $g$) to compute the free energy estimates, and then use
-fully subsampled data (with subsampling interval $g$) to estimate the
-uncertainties of the estimates.
+We don't need to use all of the data to recover the remaining correlation.  In practice, if one subsamples the data with a
+correlation interval of $g/10$, where $g$ is the correlation time.
+Thus, if $g=3.0$ in units of timesteps, it is best to use all of the data.
+If $g=100.0$, then one loses essentially no information by subsampling with $g=10.0$, and still saves significant time by not using all of the now-completely duplicative data.
+
+If independent simulations at each $\lambda$ are used, the alchemical state $\vec{\lambda}$ may have a significant impact on the correlation time, and these simulations should be subsampled independently using a separate estimate of the statistical inefficiency $g$ for each alchemical state.
+If coupled simulations are used (such as a Hamiltonian replica exchange simulation), the replicas should undergo equivalent random walks in alchemical space, and the replicas can be subsampled with the same $g$ to generate an equal number of uncorrelated samples at each alchemical state.
+Conveniently, the approach described above for automated equilibration detection produces an appropriate estimate of $g$ over the production region for automating this process.
+
+Our overall recommendations are therefore to determine the subsampling period $g$, subsample data at $g/10$ (i.e.
+more frequently than specified by $g$) to compute the free energy estimates, and then use fully subsampled data (with subsampling interval $g$) to estimate the uncertainties of the estimates.
 
@@
         \item Set up your restraints correctly
-        \item Make sure you subsample the data in your free energy estimation protocol
+        \item Subsample the data to estimate the variance of your free energy estimation protocol (but not to estimate the values themselves)
         \item Apply the appropriate correction terms
@@
         \item Check whether dummy atoms were assigned correctly
-        \item Consider subsampling the data in your free energy estimation protocol
-        \item Apply the 
-        appropriate correction terms
+        \item Subsample the data to estimate the variance of your free energy estimation protocol (but not to estimate the values themselves)
+         \item Apply the appropriate correction terms
 \end{itemize}
@@
     \item Include error bars in all your visual analyses
+    \item Consider bootstrapping uncertainties rather than using free energy error estimators
 \end{itemize}

### 2025-07-18 paper/manuscript.tex

--- 2025-07-04 paper/manuscript.tex
@@
 This is the case especially for system representations using a single coordinate set (Type 1.
-& 2), however a dual coordinate set approach could use an atom mapping as well, depending on the sampling method, this will be explained in the next steps.
+\& 2), however a dual coordinate set approach could use an atom mapping as well, depending on the sampling method, this will be explained in the next steps.
 Once a single coordinate set approach is selected is selected, a crucial next step is to identify the common atoms which will not be perturbed between the two molecules (respectivly end-states).
@@
 @Andrea Rizzi to reivew?}
-\label{sec:standardstate-restraints}
 
 \begin{figure}
-    \includegraphics[width=0.95\linewidth]{figures/fig5_thermo_cyc_abs/Figure_1.pdf}
-    \caption{\textbf{Thermodynamic cycle required for an absolute free energy calculation -- absolute free energy of binding example} The fully interacting ligand in water (\textbf{A}), has its charges turned off to pass to (\textbf{B}) followed by turning of van der Waals terms, resulting in a non-interacting ligand in water in (\textbf{C}).
-Restraints are used on the fully interacting ligand in the binding site of a protein or host molecule (\textbf{D}).
-The next step is to turn off the charges again (\textbf{E}) followed by the van der Waals interactions resulting in a non-interacting complex state (\textbf{F}).
-Free energyes can be computed as $\Delta G_{bind} = (\Delta G^{\mathrm{elec}}_{\mathrm{solv}}+ \Delta G^{\mathrm{VdW}}_{\mathrm{solv}})-(\Delta G^{\mathrm{elec}}_{\mathrm{bound}}+ \Delta G^{\mathrm{VdW}}_{\mathrm{bound}}$).
-    }
+    \includegraphics[width=0.95\linewidth]{figures/fig5_thermo_cyc_abs/Figure.png}
+    \caption{\textbf{Thermodynamic cycle for an absolute binding free energy calculation with auxiliary restraints} The fully interacting ligand in water (\textbf{A}), has its charges turned off to pass to (\textbf{B}) followed by turning off van der Waals terms, resulting in a non-interacting ligand in water (\textbf{C}).
+The receptor-ligand complex (\textbf{D}) is defined by binding site restraints (dotted oval).
+Auxiliary restraints are often introduced (\textbf{E}) to improve convergence while the charges and van der Waals interactions are turned off (\textbf{F} and \textbf{G}).
+The free energy cost of releasing the auxiliary restraints (\textbf{G} to \textbf{H}) and transferring the non-interacting ligand from the standard state volume to the binding site (\textbf{C} to \textbf{H}) must be accounted for.
+The standard binding free energy is computed from this cycle as $\Delta G^{\circ}_{\mathrm{bind}} = (\Delta G^{\mathrm{elec}}_{\mathrm{solv}}+ \Delta G^{\mathrm{VdW}}_{\mathrm{solv}}) + \Delta G^{\circ}_{\mathrm{ideal}} -(\Delta G_{\mathrm{aux.restr.}}+\Delta G^{\mathrm{elec}}_{\mathrm{bound}}+ \Delta G^{\mathrm{VdW}}_{\mathrm{bound}}+\Delta G_{\mathrm{release}})$.
+Many variations of the route between the free (\textbf{A}) and bound (\textbf{D}) states can be used -- for example, state \textbf{H} is commonly skipped and the free energy cost of moving from \textbf{C} to \textbf{G} is calculated directly.}
     \label{fig:fig_absolute_thermodynamic_cycle}
@@
 
-Absolute free energy calculations involve completely removing the interactions between the ligand or solute and its environment, taking it to a non-interacting state that may or may not retain intramolecular non-bonded interactions.
-This non-interacting state can then be shifted between environments---from the protein to water, or from one solution to another---without changing its free energy other than that due to the changing volume of the simulations, and then interactions can be restored in the new environment.
-
-Absolute free energies are by definition reported with respect to a specific reference or standard state, which effectively determines the arbitrary point at which the free energy is 0.
-The role of the standard state is particularly evident from the expression of the binding free energy between a receptor $R$ and ligand $L$
-\begin{equation} \label{eq:DGfromKAB}
-    \Delta G = -k_BT ~ \ln \left( c^{\circ} K_b \right)  = -k_BT ~ \ln\left( c^{\circ} \frac{[RL]}{[L][R]} \right) \, .
-\end{equation}
-Here, the reference state concentration $c^{\circ}$ converts the binding constant $K_b$ into a dimensionless quantity expressed in reference concentration units.
-It should be noted that ignoring the term $c^{\circ}$ is equivalent to assuming a reference concentration of 1~D$^{-1}$, where D are the units used to express $K_b$, and would thus cause the value of $\Delta G$ to vary with the choice of the units.
-It is convenient to define a standard state at a constant pressure of 1~atm and where each chemical species (i.e., R, L, and RL) in the reaction solvent has a concentration of $c^{\circ}$~=~1~M~=~1~molecule/1660~\r{A}$^3$ but do not interact with other molecules of $R$, $L$, or $RL$.
-
-\paragraph{Handling the standard state in absolute free energy calculations.}
-For solvation free energy calculations, handling the standard state is typically straightforward, and treating it correctly simply means ensuring that the non-interacting solute still occupies essentially the same volume as the solute in the interacting system.
-So typically in such cases no special care is required to ensure the correct standard state, as long as the \emph{experimental} data being analyzed uses the same standard state.
-If this is not the case, a simple entropic correction to the free energy of $k_B T\ln (V_f/V_i) = k_B T \ln (C_i/C_f)$ to the experimental data is needed.
-
-For binding, however, the situation is more complex and requires special care.
-Because the simulations are typically performed using restraints and at concentrations that are different from 1 M, the expression of the free energy requires the following correction~\cite{gilson1997statisticalthermodynamic} (see an example of such a thermodynamic cycle in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle})
-\begin{equation}\label{eq:restraint-correction}
-    \Delta G^{\circ}_{\mathrm{restr}} = -k_BT ~ \ln \left( c^{\circ}{V_L} \right) -k_BT ~ \ln \left( \frac{\xi_L}{8 \pi^2} \right),
-\end{equation}
-where $V_L$ and $\xi_L$ are respectively the volume of the translational and rotational degrees of freedom of the non-interacting ligand in the simulation box.
-When no restraints are used, the non-interacting ligand is free to translate and rotate in the simulation box (i.e., $V_L = V_{\mathrm{box}}$ and $\xi_L = 8\pi^2$), and the rotational term is zero.
-A sufficiently thorough exploration of the simulation box by the non-interacting ligand is, however, required for the formula to be valid.
-This is typically hard to achieve as the exploration process is governed by diffusion, and weak transient nonspecific binding will occur at other sites on the protein. 
-The addition of a restraint limits the volume available to the non-interacting ligand, thus speeding the convergence of the sampling.
-In addition, when enhanced sampling methods such as Hamiltonian $\lambda$ exchange are used (see Sec.~\ref{sec:sampling_schemes}), the use of a restraint is typically necessary as it keeps the ligand in the binding site in the interacting state (see also Sec.~\ref{sec:weak-binders}) and generally reduce the round-trip time of replicas.
-When restraints are employed, the values of $V_L$ and $\xi_L$ are restraint-dependent, but for commonly employed restraints, these can be usually easily computed analytically or numerically by solving the relevant integral.
-
-\paragraph{Several choices of restraints are possible.}\label{ABFErest}
-In practice, a variety of types of restraints are common, from simple harmonic distance restraints between the ligand and the protein~\cite{mobley2006use}, to flat-bottom restraints which work similarly but only exert a force if the ligand leaves a specific region~\cite{chen2007can}.
-Because these restraints do not limit the rotational degrees of freedom of the ligand, the rotational term entering the correction in Eq.~\ref{eq:restraint-correction} is zero.
-
-Alternatively, a set of restraints proposed by Boresch have also commonly been employed, where all six rigid-body degrees of freedom governing the orientation of the ligand relative to the receptor are restrained~\cite{boresch2003absolute, leitgeb2005alchemical}.
-Further restraints, such as on the overall ligand RMSD have also been used~\cite{woo2005calculation}.
-
-In principle, all of these forms will yield correct binding free energies in the limit of adequate sampling if their effects and connection to the standard state are correctly handled, but they have different strengths and weaknesses.
-For example, with more involved restraints, sampling at intermediate $\vec{\lambda}$ values will usually not need to be as extensive but more computational effort must go to computing the free energy to turn on the restraints.
-Additionally, such restraints would typically keep the ligand from exploring alternative binding modes.
+An absolute binding free energy calculation yields the standard free energy of formation of the molecular complex from the separated receptor and ligand species.
+Hence, as discussed in \ref{sec:non-covalent-molecular-binding}, it must depend on the standard state concentration $C^\circ$ (here $1 \, {\rm M} = 1 \, {\rm molecule}/1660$~\r{A}$^3$) and on the definition of the complex, that is on the specific binding pose for which the binding free energy is requested.
+Omitting the standard state concentration would yield an undefined result that would vary with the choice of concentration units.
+A similar vague estimate would result from omitting the binding pose definition that, for example, could equally describe binding to a specific site of the receptor or unspecific binding throughout the receptor surface depending on the extent of conformational sampling (see Sec.~\ref{sec:weak-binders}).
+
+In the statistical mechanics formulation of Gilson et al.~\cite{gilson1997statisticalthermodynamic}, described in Section \ref{sec:non-covalent-molecular-binding}, the standard state concentration appears explicitly in Eq.~(\ref{eq:DG0-binding-constant-def}) and the binding pose is defined by the indicator function $I(\zeta)$ that characterizes the partition function of the complex; Eq.~(\ref{eq:intra-zRL-def}).
+The role of the standard state and binding site definitions are particularly evident when expressing Eq.~(\ref{eq:DG0-binding-constant-def}) in terms of an ensemble average.
+To do so, we express the integrals in the numerator and denominator of Eq.~(\ref{eq:DG0-binding-constant-def}) over the same set of degrees of freedom.
+Note that the intramolecular configurational partition function of the complex, $z_{RL(v)}$, includes six more degrees of freedom--the position and orientation of the ligand relative to the receptor collectively described by the variable $\zeta$--than the product of the intramolecular partition functions, $z_{R(v)} z_{L(v)}$, of the separated receptor and ligand.
+To correct this discrepancy, we multiply and divide the right-hand side of Eq.~(\ref{eq:DG0-binding-constant-def})  by the quantity
+\begin{equation} \label{eq:Vsite-def}
+    V_{\rm site} \Omega = \int d\zeta I(\zeta) \, ,
+\end{equation}
+%\Omega is in my opinion a better symbol of the orientational extent than \xi_L. Lowercase greek letters are usually used
+%for dynamical variables rather than integrated quantities
+which measures the extent of the binding pose macrostate, where $V_{\rm site}$ corresponds to the spatial extent and $\Omega$ the orientational extent.
+The right-hand side of Eq.~(\ref{eq:Vsite-def}) combined with the integrals $z_{R(v)}$ and $z_{L(v)}$ yields a partition function similar to Eq.~(\ref{eq:intra-zRL-def}) but corresponding to a complex in which ligand and receptor are uncoupled~\cite{gallicchio2011recent} as if they were at infinite distance apart.
+The result is
+\begin{equation}\label{eq:DG0-binding-constant-def-average}
+K_b = C^\circ V_{\rm site} \frac{\Omega}{8 \pi^2} \langle e^{-\beta \Delta \Psi}  \rangle_0 
+\end{equation}
+where, similarly to Eq.~(\ref{eq:zratio-solv-average}),
+\begin{equation}\label{eq:pert-energy-def}
+\Delta \Psi(x_R, x_L, \zeta) =  \Psi_v(x_R, x_L, \zeta) - \left[  \Psi_v(x_R) + \Psi(x_L) \right]
+\end{equation}
+is the solvent-averaged binding energy of the complex when receptor and ligands are in intramolecular configurations $x_R$ and $x_L$, the ligand is in position and orientation $\zeta$ relative to the receptor (the uncoupled potential energy does not depend on  the position and orientation of the ligand relative to the receptor), and $\langle \ldots \rangle_0$ represents an ensemble average in the uncoupled ensemble.
+Finally, note that $\Omega = 8 \pi^2$ when the binding pose indicator function does not depend on the orientation of the ligand and Eq.~(\ref{eq:DG0-binding-constant-def-average}) simplifies to
+\begin{equation}\label{eq:DG0-binding-constant-def-average-noOmega}
+K_b = C^\circ V_{\rm site} \langle e^{-\beta \Delta \Psi}  \rangle_0 
+\end{equation}
+which, expressed in free energy units, yields
+\begin{eqnarray}\label{eq:DG0-def-Vsite}
+  \Delta G^\circ_b &=&  - k_B T \ln K_b = - k_B T \ln C^\circ V_{\rm site} - k_B T \ln \langle e^{-\beta \Delta \Psi}  \rangle_0 \nonumber \\
+{\rm \ }  &=& \Delta G^\circ_{b,{\rm ideal}} + \Delta G_{b,{\rm exc.}}
+\end{eqnarray}
+
+The term $\Delta G^\circ_{b,{\rm ideal}} = - k_B T \ln C^\circ V_{\rm site}$ in Eq.~(\ref{eq:DG0-def-Vsite}) is interpreted as the \emph{ideal} component of the standard binding free energy, that is the binding free energy that would be measured in the absence of ligand-receptor interactions ($\Delta \Psi = 0$) and when the ligand's concentration is $C^\circ$.
+The ideal term is responsible for the dependence of the standard binding free energy on the standard state concentration.
+The arguably inappropriate term ``standard state correction'' is sometimes used in the literature to refer to $\Delta G^\circ_{b,{\rm ideal}}$.
+As seen above, the ideal term is a fundamental aspect of the statistical mechanics theory of molecular binding rather than a mere correction to an inexact formulation.
+
+Conversely, the term $\Delta G_{b,{\rm exc.}} = - k_B T \ln \langle e^{-\beta \Delta \Psi}  \rangle_0 $ is the \emph{excess} component of the standard free energy of binding, which is the term that depends on intermolecular interactions and the specific chemical composition of the system.
+The excess component is the target of the alchemical molecular simulations discussed here, in contrast to the ideal term that is generally evaluated analytically~\cite{boresch2003absolute}.
+
+\paragraph{Handling the standard state in free energy calculations.}
+As discussed in Sections \ref{sec:theory-solvation} and \ref{sec:theory-partitioning}, the handling of the standard state in solvation and solvent partitioning free energy calculations is typically straightforward.
+Eq.~(\ref{eq:DG0-solvation-def}) identifies the ideal and excess components of the standard solvation free energy in terms of the standard gas pressure $P^\circ$, standard solution concentration $C^\circ$, and ensemble averages typically evaluated by alchemical free energy simulations.
+These relations provide simple recipes to relate computational estimates to experimental values obtained under different conditions.
+Moreover, typical experimental reporters (Henry's constants and partitioning coefficients) directly probe the excess term calculated by molecular simulations, which is independent of the standard state.
+
+As discussed above, the situation for binding is more complex and requires special care.
+The first important aspect is that binding free energy calculations must be performed with binding site restraints that implement the binding site indicator function $I(\zeta)$ by forcing the ligand to occupy the binding site even when it is decoupled from the receptor.
+Doing otherwise would fail to specify the bound macrostate of the complex and return undefined binding free energy estimates.
+However, in practice, calculations lacking binding site restraints often return reliable binding free energy estimates due to the kinetic trapping of the ligand in the binding site region during the relatively short molecular dynamics sampling typical of alchemical binding free energy simulations.
+Nevertheless, longer calculations could display persistent drift of the estimate as new regions of conformational space are sampled.
+A more serious consequence of neglecting binding site restraints is the possibility of the ligand leaving the binding site when it is not coupled to the receptor.
+Such occurrence indicates a serious violation of the theory.
+Alchemical binding free energy simulations where the ligand leaves the binding site are not set up correctly and can return grossly inaccurate binding free energy estimates.
+
+Because it is a discontinuous function of the position and orientation of the ligand relative to the receptor, molecular dynamics implementations often approximate the indicator function by means of flat-bottom restraints~\cite{chen2007can, wu2021alchemical} (see below).
+Harmonic restraints are not as suitable in this respect because they favor a specific microstate at the bottom of the harmonic potential rather than specifying a region of conformational space.
+
+Note that $\Delta G^\circ_{b,{\rm ideal}}$ is independent of the chemical nature of the binding partners, and it is the same for all complexes sharing similar binding site definitions.
+The latter is the reason that the ideal term cancels out when taking the difference of absolute binding free energies when estimating relative binding free energies (RBFEs).
+However, while it does not appear explicitly in the definition, $\Delta G_{b,{\rm exc.}}$ depends implicitly on the definition of the binding site through the ensemble average in the uncoupled ensemble, which includes only ``bound'' configurations according to the definition of the indicator function $I(\zeta)$.
+Varying the binding site definition affects both the ideal and excess binding free energy terms such that their sum (the standard binding free energy) is often only weakly dependent on the binding site definition~\cite{gilson1997statisticalthermodynamic,gallicchio2011recent}.
+Conversely, great care should be exercised when comparing calculated excess binding free energy values corresponding to different binding site definitions.
+Similarly, the outcome of a binding free energy calculation conducted without a binding site definition, i.e.\ without binding site restraints, is, in principle, undefined.
+The latter is also true for relative binding free energies because while the ideal term cancels out, the difference in excess binding free energies depends on the binding pose definition.
+
+\paragraph{Several choices of restraints are possible in binding free energy calculations.}
+A variety of restraint types are commonly used in alchemical binding free energy calculations.
+Before going into their characteristics, we must acknowledge an important and often unrecognized distinction between the binding site restraints corresponding to the indicator function $I(\zeta)$ necessary for the implementation of the theory and optional auxiliary restraints and biasing potentials sometimes used to speed up the convergence of free energy calculations.
+Binding site restraints are an integral part of the molecular complex definition and are typically maintained throughout the alchemical process.
+In particular, they must be present unchanged at the unbound and bound states.
+In contrast, unless thermodynamic reweighting is applied in post-processing, auxiliary restraints cannot be present at the endstates to avoid affecting their free energies.
+Hence, typically the free energy cost of turning them on is computed at the uncoupled state, and the free energy gain of releasing them is calculated near the bound state.
+In the alchemical binding free energy literature, often harmonic restraint potentials imposed in an initial stage and removed in a final stage are described as implementing binding sites restraints.
+However, because they utilize harmonic potentials that cannot correctly describe the binding pose macrostate (see above) and because they are turned on and then off, these are instead auxiliary restraints introduced to enhance convergence without affecting, in principle, binding free energy estimates.
+
+Binding site restraints implement the $I(\zeta)$ indicator function.
+They, therefore, can depend at most on the six rigid-body external coordinates of the ligand governing the position and orientation of the ligand relative to the receptor.
+Any other choice would affect the intramolecular conformational distribution of the ligand or the receptor and would  bias the binding free energy estimate.
+The external degrees of freedom introduced by Boresch et al.~\cite{boresch2003absolute,leitgeb2005alchemical} are a popular choice.
+The Boresch coordinates, as well as other similar approaches, involve selecting anchor atoms of the receptor and the ligand.
+In principle, the choice of anchor atoms is arbitrary, provided adequate simulation time is available, and as long as it leads to similar binding pose definitions.
+However, practical considerations may be important.
+For example, some choices of Boresch-style anchor atoms can be numerically unstable.
+Additionally, anchor atoms should likely be placed in a part of the molecule that defines the binding orientation well rather than in a floppy, solvent-exposed tail.
+Several automated methods have been proposed to select anchor points for Boresch restraints, which typically involve analysing a simulation of the fully-interacting receptor-ligand complex~\cite{alibay2022evaluating, alibay2021mdrestraintsgenerator, baumann2023broadening, chen2023enhancing, hedges2023suite, wu2025optimizing}.
+
+As discussed above, binding site restraints are commonly implemented using quadratic flat-bottom restraints, which impose a restoring harmonic  force similar to harmonic potentials but only if the ligand leaves a specific region~\cite{chen2007can}.
+Often, binding site restraints do not limit the rotational degrees of freedom of the ligand, and the corresponding standard state term is zero; see Eq.~(\ref{eq:DG0-def-Vsite}).
+However, it might be challenging to fully explore the bound macrostate at all alchemical states when it is only limited by receptor-ligand positional restraints.
+As further discussed below, auxiliary restraints are particularly helpful in the latter circumstance to limit the extent of conformational sampling necessary to achieve rapid convergence of binding free energy estimates.
+
+Most current binding free energy workflows do not use binding site restraints present at all stages of the simulation and implicitly assume that $V_{\rm site}$ in Eq.~(\ref{eq:Vsite-def}) is the volume of the simulation box $V$.
+In the absence of binding site restraints, only the protein-ligand interactions prevent the ligand from slipping out from the binding site and wandering throughout the simulation box within the simulation time frame.
+However, observations of ligands leaving the binding site become more likely as simulation times increase, especially when analyzing weak binders and non-binders (see below).
+Moreover, there is no obstacle for the ligand to diffuse away from the binding site at the decoupled state when protein-ligand interactions are turned off.
+
+In practice, because they often implement auxiliary restraints, binding free energy prediction workflows that omit binding site restraints rarely yield grossly erroneous predictions.
+Auxiliary restraints are introduced gradually as protein-ligand interactions are turned off or are introduced at the bound state when the ligand is naturally trapped in the binding site.
+Either way, the free energy of imposing them can be tracked while preventing the ligand from leaving the binding site at the decoupled state.
+Provided that the interacting ligand samples only bound configurations, i.e., it behaves as if binding site restraints were present, the current practice of not using them is likely to incur only minor or insignificant errors.
+
+The Boresch harmonic restraints~\cite{boresch2003absolute,baumann2023broadening} are a popular choice as auxiliary restraints because they are based on interatomic distances, bond angles, and dihedral angles degrees of freedom already implemented in molecular mechanics codes that do not interfere with intramolecular conformational distributions.
+Moreover, the free energy of releasing Boresch restraints within the simulation box volume $V$ is easily computed analytically~\cite{boresch2003absolute, boresch2024analytical}.
+
+By reducing the conformational space to be sampled, auxiliary restraints are often crucial for obtaining converged binding free energy estimates, even when binding site restraints are employed.
+Auxiliary restraints take a variety of forms, such as distance restraints between the ligand and the protein~\cite{mobley2006use,clark2023comparison} as well as flat-bottom restraints.
+The choice of degrees of freedom and restraining potentials for auxiliary restraints is not as restrictive as for those for binding site restraints, provided that the free energy terms of restraining collective variables that couple protein and ligand degrees of freedom are adequately accounted for to avoid serious systematic errors~\cite{clark2023comparison}.
+For example, the overall ligand RMSD has been used~\cite{woo2005calculation} for this purpose.
+Auxiliary restraints are designed to reduce the amount of conformational sampling at intermediate $\vec{\lambda}$ values.
+However, more computational effort is required to compute the free energy when the restraints are turned on and off.
+Additionally, such restraints can limit the exploration of alternative binding modes.
 This restriction may be undesirable when using Hamiltonian $\lambda$ exchange or expanded ensemble techniques where allowing the ligand to exchange binding modes when it is non-interacting could provide sampling benefits~\cite{wang2013identifying}.
-More specifically, flat-bottom restraints might allow a ligand to explore multiple binding sites, and harmonic restraints allow exploration of multiple binding modes within a site, while Boresch restraints only allow a single binding mode within a single site.
+More specifically, flat-bottom restraints allow a ligand to explore multiple binding sites, and harmonic restraints allow the exploration of multiple binding modes within a site, while Boresch restraints only allow a single binding mode within a single site.
 See additional discussion of the possibility of multiple binding modes in Sec.~\ref{sec:multiple_binding_modes} below.
 
-Many choices of restraints involve selecting reference atoms.
-Again, in principle this choice is unimportant given adequate simulation time but practical considerations may be important.
-The choice is likely especially important with Boresch-style restraints, where some relative placements of reference atoms are likely to be numerically unstable; additionally, ligand reference atoms should likely be in a part of the molecule which defines the binding orientation well, rather than in a floppy solvent-exposed tail, for example.
-Several automated methods have been proposed to select anchor points for Boresch restraints, which typically involve analysing a simulation of the fully-interacting receptor-ligand complex~\cite{alibay2022evaluating, alibay2021mdrestraintsgenerator, baumann2023broadening, chen2023enhancing, hedges2023suite}.
-
-More recently, restrictive restraint schemes have been proposed that avoid relying on a few critical reference atoms.
-Instead, these aim to simplify restraint selection by using larger sets of atomic positions in the receptor and ligand.
+Recently, degrees of freedom for binding site and auxiliary potentials based on larger sets of atomic positions of the receptor and ligand have been put forward.
 Fu et al.
-suggested restraining the six relative rigid-body degrees of freedom derived by finding the rotation of the ligand which minimises RMSD to a reference structure of the receptor-ligand complex (after accounting for rotation and translation of the receptor)~\cite{fu2017new}.
+suggested restraining the six relative rigid-body degrees of freedom derived by finding the rotation of the ligand, which minimises the RMSD to a reference structure of the receptor-ligand complex (after accounting for rotation and translation of the receptor)~\cite{fu2017new}.
 Salari et al.
 proposed restraining the ``distance-from-bound-configuration'' (DBC) variable, which is the RMSD of ligand coordinates in the frame of reference of the binding site~\cite{salari2018streamlined, ebrahimi2022symmetry}.
-This may allow the binding pose to be closely preserved as the ligand intermolecular interactions are removed, but has the disadvantage of coupling the internal and relative external degrees of freedom of the receptor and ligand, preventing easy computation of the standard state correction.
-This necessitates an additional step where the DBC restraints are released to a single harmonic restraint.
-Both restraint schemes are implemented in open-source workflows~\cite{fu2021bfee2, fu2022accurate, santiago-mcrae2023computing}.
+DBC may allow the binding pose to be closely preserved as the ligand intermolecular interactions are removed, but has the disadvantage of coupling the internal and relative external degrees of freedom of the receptor and ligand, making them unsuitable for describing the binding site macrostate and the corresponding standard state terms.
+However, these restraints are practical as auxiliary restraints when additional steps are used to introduce and release them.
+Moreover, because they are based on flat-bottom potentials encompassing the desired binding poses, the free energy of releasing them in the bound state can be safely neglected in most cases.
+The restraint schemes discussed here are implemented in open-source workflows~\cite{fu2021bfee2, fu2022accurate, santiago-mcrae2023computing, clark2023comparison, hedges2023suite}.
 
 \subsection{Absolute and relative calculations deal with some of the same issues}
-\subsubsection{Handling weak binders and high dissociation rates}\label{sec:weak-binders}
-In binding free energy calculations, only the conformations in which the receptor and ligand form a bound complex should be sampled from the bound states (Sec.~\ref{sec:theory}).
-Determining what the bound states actually are can be challenging for weakly bound ligands.
-For tightly bound ligands, virtually all reasonable definitions of the bound state will lead to be equivalent free energies, since the partition function will be dominated by a relatively small number of low-energy poses.
-
-For weak binders, this simplification breaks down.
-In fact, the correct bound state may depend on the type of experiment performed.
-For example, isothermal titration calorimetry (ITC) or surface plasmon resonance (SPR) measurements effectively define a binding state that includes all ligand comformations that are complexed with the protein, regardless of where on the protein they bind.
+
+\paragraph{Handling weak binders}\label{sec:weak-binders}
+According to the non-covalent statistical mechanics theory of bimolecular binding (Sec.~\ref{sec:theory}), only the conformations in which the receptor and ligand form a bound complex are sampled.
+Determining an appropriate definition of the bound macrostate can be challenging for weakly bound ligands.
+For tightly bound ligands, virtually all reasonable definitions of the bound state will lead to equivalent free energies since the partition function will be dominated by a relatively limited ensemble of low-energy poses.
+However, this simplification breaks down for weak binders.
+In fact, there might not be a perfect correspondence between the structural definition of the complex on which current alchemical models are based~\cite{gilson1997statisticalthermodynamic} (Sec.~\ref{sec:theory}) and experimental reporters~\cite{mihailescu2004theory}.
+For example, isothermal titration calorimetry (ITC) or surface plasmon resonance (SPR) measurements effectively define a binding state that encompasses all ligand conformations complexed with the protein, regardless of where they bind the protein.
 In contrast, fluorescence polarization competition assays measure binding to only a single location, where the ligand of interest displaces a competing binder.

### 2025-07-18 paper/manuscript.tex
update subsampling.

--- 2025-07-14 paper/manuscript.tex
@@
 Another advantage of bootstrap sampling is that one can construct decent estimates of error bounds with arbitrary percentiles, and not just standard errors of the estimates, which assume Gaussian behavior of the underlying error, and that the error results are Gaussian as well.
+The bootstrap approach, on the other hand, captures non-Gaussian behavior in the data.
 
@@
 Another alternative to bootstrap sampling of individual points that better captures the correlations in the data is block bootstrapping~\cite{???}.
-Block bootstrapping is somewhat more complicated and less common to perform and analyze than bootstrap estimates of the error, so we will not go into details, but it can also provide more reliable estimates of uncertainty than the estimators above.
+Block bootstrapping is somewhat more complicated and less common to perform and analyze than bootstrap estimates of the error, so we refer readers to the citations for more description, but it can also provide more reliable estimates of uncertainty than the estimators above.
 
@@
 
+%MRS: Finlay, you can add information about RED here.
+
 \subsection{Decorrelating samples for analysis}
@@
 
-We recommend using the robust statistical inefficiency computation routines available within the MBAR~\cite{kylebeauchamp2019choderalab} and alchemlyb~\cite{daviddotson2019alchemistry} packages.
-
-%Cooment: Finlay, can you add description/citation of RED?
+We recommend using the robust statistical inefficiency computation routines available within the MBAR~\cite{kylebeauchamp2019choderalab} and alchemlyb~\cite{daviddotson2019alchemistry} packages, though we caution that any such algorithms will underestimate the correlation time with insufficient data.
+Such approaches also assume that the data doesn't have any long time scale drift over the course of the simulation, which is a clear sign of lack of equilibration.
+Generally, most methods will give reasonable correlation times when the simulation lasts 40--50 correlation times,
 
@@
 Recent careful investigation has shown that the lowest uncertainty results can be obtained by using \textit{all} of the data to obtain an estimate of the free energy itself, but using only subsampled data to calculate error estimates using the formulas or by using bootstrapping.
-In fact, one can show that the standard deviation of the estimate determined by data that has been subsampled is the unbiased estimator of the free energy estimate using all the data, but is actually an underestimate of the uncertainty of estimate
-using just the subsampled data.
+In fact, one can show that the standard deviation of the estimate determined by data that has been subsampled is the unbiased estimator of the free energy estimate using all the data, but is actually an underestimate of the uncertainty of estimate using just the subsampled data.
+%MRS: Finlay, is there a citation for the mathematical analysis you showed in the issue tracker?  This would be a good place to it.
 
 We don't need to use all of the data to recover the remaining correlation.  In practice, if one subsamples the data with a
-correlation interval of $g/10$, where $g$ is the correlation time.
+correlation interval of $g/10$, where $g$ is the correlation time, this will recover the missing correlation.
 Thus, if $g=3.0$ in units of timesteps, it is best to use all of the data.
-If $g=100.0$, then one loses essentially no information by subsampling with $g=10.0$, and still saves significant time by not using all of the now-completely duplicative data.
+If $g=100.0$, then one loses essentially no information by subsampling with $g=10.0$ in estimating the free energies, and still saves significant time by not using all of the now-completely duplicative data.
+Note that subsampling with correlation time $g$ should still be used for estimating the uncertainties, perhaps requiring two passes of the simulation analysis code.
 
@@
 
-Our overall recommendations are therefore to determine the subsampling period $g$, subsample data at $g/10$ (i.e.
+Our overall recommendations are therefore to determine the subsampling period $g$, either use all of the data or subsample with something like $g/10$ (i.e.
 more frequently than specified by $g$) to compute the free energy estimates, and then use fully subsampled data (with subsampling interval $g$) to estimate the uncertainties of the estimates.
+Uncertainties calculated with the full data set will be significant underestimates if $g > 1$.
 

### 2025-07-18 paper/manuscript.tex

--- 2025-07-18 paper/manuscript.tex
@@
 
-MBAR can also be thought of as the Zwanzig estimator of the free energy to state $i$ where the sampled distribution is the \textit{mixture distribution} of all the other samples thrown together in one ``pot'', defined by $p_m(\vec{q}) = N^{-1} \sum_k N_k \exp(f_k-u_k\vec{q})$, which is the weighted average of all the individual normalized probability distributions from the simulations that are performed~\cite{shirts2017reweighting}.
+MBAR can also be thought of as the Zwanzig estimator of the free energy to state $i$ from a sampled distribution referred to as the \textit{mixture distribution}, where one throws all the samples from all simulations together in one ``pot''.
+Mathematically, this is defined by $p_m(\vec{q}) = N^{-1} \sum_k N_k \exp(f_k-u_k\vec{q})$, which is the average, weighted by the number of samples drawn from each distribution, of all the individual normalized probability distributions from the simulations that are performed~\cite{shirts2017reweighting}.
+
+All of the estimators above have uncertainty estimates associated with them.
+In the limit of sufficiently large data, perhaps several hundred or thousand independent points, exhaustive analyses of model data has shown that the error estimates of BAR, MBAR, and TI are all quite reliable when there is good overlap between intermediate states.
+For EXP, the number of samples at which they become reliable is significantly higher, especially if overlap between intermediate states is lower.
+However, in the limit of moderate or small amounts of data (tens or hundreds of independent samples) these errors become somewhat unreliable, often off by a factor of 2 from what would get if more data was collected.
+These estimator of uncertainty are all derived from propagation of error formulae from the estimators themselves.
+However, when the underlying data is noisy, this approximation is not very good.
+
+Some of the authors' experience has shown that significantly more reliable estimators at lower numbers of samples are obtained by performing bootstrap sampling on data that has been uncorrelated.
+We do not provide a full tutorial of bootstrapping here, but the basic idea is that one takes the $N_{sample}$ data points collected and draws, with replacement, $N_{sample}$ samples.
+This means some of the data points will be represented multiple times, and other data points none.
+This process is then repeated $N_{bootstrap}$ times, where $N$ should generally be at least 50 to up to 200 for good estimates of the standard error of the estimator.
+The data is then \textit{treated} as $N_{bootstrap}$ samples from the true distribution, and statistics such as the standard error are computed from this artificial set of $N_{bootstrap}$ experiments.
+In the limit of $N_{sample}$ large, it can be shown that the bootstrap estimate of the standard error converges to the true population standard error.
+One way this can be seen is that bootstrap estimators of the data are independent samples of the \textit{actual} distribution of the experiment performed, i.e., a bunch of $\delta$ functions at $N_{sample}$ points, rather than being independent samples from the true, usually continuous distribution.
+As more and more data is collected, our collection of $\delta$ functions becomes more and more like the true distribution.
+Another advantage of bootstrap sampling is that one can construct decent estimates of error bounds with arbitrary percentiles, and not just standard errors of the estimates, which assume Gaussian behavior of the underlying error, and that the error results are Gaussian as well.
+The bootstrap approach, on the other hand, captures non-Gaussian behavior in the data.
+
+One important point is that the bootstrap samples use to estimate the standard error should be decorrelated, or else the bootstrap estimate of the standard error will be artificially small.
+More discussion of subsampling and decorrelation is included in section~\ref{sec:decorrelating-samples}.
+Another alternative to bootstrap sampling of individual points that better captures the correlations in the data is block bootstrapping~\cite{???}.
+Block bootstrapping is somewhat more complicated and less common to perform and analyze than bootstrap estimates of the error, so we refer readers to the citations for more description, but it can also provide more reliable estimates of uncertainty than the estimators above.
 
@@
 
-We recommend using the robust statistical inefficiency computation routines available within the MBAR~\cite{kylebeauchamp2019choderalab} and alchemlyb~\cite{daviddotson2019alchemistry} packages.
+We recommend using the robust statistical inefficiency computation routines available within the MBAR~\cite{kylebeauchamp2019choderalab} and alchemlyb~\cite{daviddotson2019alchemistry} packages, though we caution that any such algorithms will underestimate the correlation time with insufficient data.
+Such approaches also assume that the data doesn't have any long time scale drift over the course of the simulation, which is a clear sign of lack of equilibration.
+Generally, most methods will give reasonable correlation times when the simulation lasts 40--50 correlation times,
 
@@
 
-If independent simulations are used, the alchemical state $\vec{\lambda}$ may have a significant impact on the correlation time, and these simulations should be subsampled independently using a separate estimate of the statistical inefficiency $g$ for each alchemical state.
+However, samples uncorrelated by this procedure are only approximately uncorrelated.
+Recent careful investigation has shown that the lowest uncertainty results can be obtained by using \textit{all} of the data to obtain an estimate of the free energy itself, but using only subsampled data to calculate error estimates using the formulas or by using bootstrapping.
+In fact, one can show that the standard deviation of the estimate determined by data that has been subsampled is the unbiased estimator of the free energy estimate using all the data, but is actually an underestimate of the uncertainty of estimate using just the subsampled data.
+%MRS: Finlay, is there a citation for the mathematical analysis you showed in the issue tracker?  This would be a good place to it.
+
+We don't need to use all of the data to recover the remaining correlation.  In practice, if one subsamples the data with a
+correlation interval of $g/10$, where $g$ is the correlation time, this will recover the missing correlation.
+Thus, if $g=3.0$ in units of timesteps, it is best to use all of the data.
+If $g=100.0$, then one loses essentially no information by subsampling with $g=10.0$ in estimating the free energies, and still saves significant time by not using all of the now-completely duplicative data.
+Note that subsampling with correlation time $g$ should still be used for estimating the uncertainties, perhaps requiring two passes of the simulation analysis code.
+
+If independent simulations at each $\lambda$ are used, the alchemical state $\vec{\lambda}$ may have a significant impact on the correlation time, and these simulations should be subsampled independently using a separate estimate of the statistical inefficiency $g$ for each alchemical state.
 If coupled simulations are used (such as a Hamiltonian replica exchange simulation), the replicas should undergo equivalent random walks in alchemical space, and the replicas can be subsampled with the same $g$ to generate an equal number of uncorrelated samples at each alchemical state.
@@
 
+Our overall recommendations are therefore to determine the subsampling period $g$, either use all of the data or subsample with something like $g/10$ (i.e.
+more frequently than specified by $g$) to compute the free energy estimates, and then use fully subsampled data (with subsampling interval $g$) to estimate the uncertainties of the estimates.
+Uncertainties calculated with the full data set will be significant underestimates if $g > 1$.
+
 \paragraph{Cautions and considerations}
 Reliable estimation of the statistical inefficiency is difficult, and estimates will not generally be as precise (in a relative error sense) as averages.
-To ensure there is sufficient data available for reliable decorrelation and estimation of free energy differences, it is recommended that the effective number of uncorrelated samples $N_{\mathrm{eff}} \ge \approx 50$ if the BAR or MBAR estimators (discussed below in sec~\ref{subsec:estimators}) are used; the number may need to be much higher with alternate estimators.
+To ensure there is sufficient data available for reliable decorrelation and estimation of free energy differences, it is recommended that the effective number of uncorrelated samples $N_{\mathrm{eff}} \ge  \sim 50$ if the BAR or MBAR estimators (discussed below in sec~\ref{subsec:estimators}) are used; the number may need to be much higher with alternate estimators.
 

### 2025-07-18 paper/manuscript.tex
Synchronise changes made in Overleaf, incuding:
- Undated discussion of uncertainty estimation by Agastya Bhati and Peter Coveney (changes still required to address review)
- ATM section with comments from Darrin York addressed.
- Updated discussion of non-equilibrium switching by Vytas Gapsys.
- Updates to the "importance of conformational sampling" section made by Wei-Tse Hsu in response to comments from Emilio Gallicchio.
- Various other minor tweaks

--- 2025-07-18 paper/manuscript.tex
@@
 \author[2]{Bryce K. Allen}
+\author[11]{Agastya P. Bhati}
 \author[3]{Hannah E. Bruce Macdonald}
 \author[3*]{John D. Chodera}
+\author[11, 12, 13]{Peter V. Coveney}
 \author[9]{Vytautas Gapsys}
@@
 \author[2]{Huafeng Xu}
-
 %
@@
 \affil[10]{Cresset, Cambridgeshire, UK}
-\affil[11]{Vertex Pharmaceuticals, San Diego, CA, USA}
-
+\affil[11]{Centre for Computational Sciences, Department of Chemistry, University College London, London WC1H 0AJ, United Kingdom}
+\affil[12]{Advanced Research Computing Centre, University College London, London WC1H 9BT, United Kingdom}
+\affil[13]{Computational Science Laboratory, Institute for Informatics, Faculty of Science, University of Amsterdam, Amsterdam 1012, The Netherlands}
 %
@@
 \orcid{Vytautas Gapsys}{0000-0002-6761-7780}
+\orcid{Agastya P. Bhati}{0000-0003-4539-4819}
+\orcid{Peter V. Coveney}{0000-0002-8787-7256}
 
@@
 
-Non-equilibrium free energy techniques provide an alternate approach to this problem, driving $\lambda$ between states, but these are not our focus here.~\cite{jarzynski1997nonequilibrium,jarzynski1998equilibrium,crooks2000pathensemble, gapsys2020large}
+Non-equilibrium free energy techniques provide an alternate approach to this problem, driving $\lambda$ between states (see Sec.~\ref{sec:sampling_schemes}).~\cite{jarzynski1997nonequilibrium,jarzynski1998equilibrium,crooks2000pathensemble, gapsys2020large}
 
@@
 For more complex use cases such as protein-ligand binding free energies the repeatability is often substantially worse~\cite{rizzi2019sampl6}. 
-A good practice is to perform two or three runs of the same perturbation to assess precision with a given protocol, using different initial velocities.
+There is growing literature describing the theoretical underpinning behind such intrinsic non-reproducibility of MD simulation based predictions.
+Coveney \textit{et al.
+}~\cite{coveney2016} have noted that, for a complex dynamical system to reach equilibrium, the dynamics must at least be ``mixing'' which is an established result in ergodic theory.
+A system which is ``mixing'' is also ``ergodic'' by definition but the reverse is not true.
+This implies that MD trajectories are chaotic, in other words they are extremely sensitive to their starting conditions and hence one-off simulations can never be reproducible.
+Thus, it is a necessity to perform ensemble simulations in order to perform ensemble averaging to obtain thermodynamic quantities in a reliable manner.
+Coveney and Wan recently published a book on the probabilistic foundations of MD and how this approach naturally allows for the treatment of uncertainties that arise in the use of the methodology~\cite{MDbook2025}.
+In practice, in many cases the simplest and most effective way of performing ensemble simulations is to vary only the initial velocities as these yield reproducible (i.e.
+statistically robust) free energy predictions.
+Bhati \textit{et al.} developed the TIES method~\cite{bhati2017} which is based on the use of ensemble methods.
+It produces excellent results for large scale benchmarking datasets for both RBFEs~\cite{bhati2022} and ABFEs~\cite{bhati2025}.
+It should be noted that the size of ensemble, that is, the number of independent simulations, is an adjustable parameter whose optimal value is system-dependent and will also be affected by the desired level of precision.
+Further, this method is extremely flexible and adaptive.
+For example, one can simply add additional ensemble members as necessary or desired, without having to re-run anything done initially.
+It should also be noted that ensemble simulations are computationally more effective than long simulations.
+On a large supercomputer, all the sampling can be done concurrently obtaining the statistical results in the time it takes to run a single simulation.
+Studies have shown that, in general, an ensemble size of 5 and 10 members is a reasonable starting point for alchemical RBFE and ABFE calculations, respectively, but recommendations are provided to adjust them based on some rules of thumb~\cite{bhati2022, bhati2025}.
 The sample standard deviation will give a crude estimate of the reliability of the estimates, and whether the precision is sufficient for the problem at hand.
-When practical, a more stringent test is to use different input coordinates for each repeat run as well as different velocities.
+When practical, a more stringent test is to use different input coordinates for each replica as well as different velocities.
 
@@
 In the recent SAMPL6 SAMPLing challenge a larger variability of 0.3 to 1.0 kcal/mol was noted in the computed absolute binding free energies of host/guest systems even though the study sought to use identical input and simulation parameters~\cite{rizzi2019sampl6} and, in many cases, single-point energies were identical or nearly so.
-Further work is needed to ensure reproducibility of alchemical free energy calculations across different software implementations to guarantee that force-field development efforts lead to transferable potential energy functions.
+A recent study demonstrates reproducibility across software packages as well as free energy estimators, but only when ensemble simulations are employed~\cite{wade2022}.
+The authors reported results from two different MD engines (NAMD and OpenMM) as well as two free energy estimators (FEP and MBAR) for a set of 54 protein-ligand complexes which were found to be in good agreement when ensemble simulations were used to control uncertainties.
+To conclude, ensemble simulations can remove the aleatoric uncertainties in MD based predictions, however, further work is needed to ensure reproducibility across software packages arising due to systematic/parametric uncertainties.
 
@@
 Binding site water molecules may form water mediated protein-ligand interactions which can pose challenges whenever exchange with bulk water is slow compared to simulation timescales.
-This can happen in buried binding sites~\cite{laage2017water}.
-Overlaying multiple protein X-ray structures can identify conserved water molecules or those that are associated only with certain ligands in a series and are therefore important to include in calculations.
-In cases where water molecules are known to play an important role in ligand binding, for example if the presence of electron density suggests they are tightly bound (indicated by a high EDIA score~\cite{nittinger2015evidence,meyder2017estimating}), software implementations that use water sampling facilitated by Grand Canonical Monte Carlo methods may be useful~\cite{michel2010prediction}.
-Other tools such as WaterMap~\cite{abel2008role,young2007motifs} or open source equivalents (SSTMap~\cite{haider2018solvation}, GIST~\cite{ramsey2016solvation}, and others) can be used to define the hydration states for systems with no experimental evidence of water sites~\cite{wang2011ligand}.
+This happens typically in buried binding sites.
+Overlaying multiple protein X-ray structures can identify conserved or additional water molecules that can be useful to include in calculations.
+In cases where water molecules are known to play an important role in the binding, software implementations that use water sampling facilitated by Grand Canonical Monte Carlo methods may be useful~\cite{michel2010prediction}.
+Other tools such as WaterMap or open source equivalents (SSTMap, GIST, and others) can be used to define water structure for systems with no experimental evidence of water sites~\cite{wang2011ligand}.
 Well-known protein systems with water mediated ligand interactions are for example: HSP90 which formed part of the D3R grand challenge 2015~\cite{mey2016blinded}, A2A~\cite{brucemacdonald2018ligand}, MUP~\cite{ross2015water}, ~\cite{deflorian2020accurate}, and others~\cite{michel2009energetics}.
-
-In the context of relative binding free energy calculations, determining the correct hydration pattern for the two endpoints may not be sufficient for an accurate and reproducible calculation.
-If the locations of the bound water molecules differs at the two endpoints - e.g.
-if a perturbation to the ligand displaces a bound water molecule - then the network needs to be able to adapt during the perturbation.
-Otherwise, it is likely that the estimated relative binding free energies will be dependent on the direction in which the perturbation is performed~\cite{ross2020enhancing}.
-
-In absolute binding free energy calculations, a ligand is fully decoupled from a binding site cavity which may otherwise be hydrated in the \textit{apo} state.
-In this case, the sampling of water molecules must be sufficient to hydrate the binding pocket as the ligand is decoupled (or vice versa for the reverse).
-When the binding site is exposed and water can move freely between the site and bulk solvent, then sampling is often not an issue.
-However, in the case of more buried sites or sites which form complex water networks, enhanced water sampling may be required to ensure an accurate binding affinity.
-
-\paragraph{Membrane proteins require additional care}
-Integral membrane proteins require embedding in a suitable model membrane to remain stable during simulations.
-Lipids are usually not included in protein force fields and, although they may be covered by generic small molecule force fields, have often been parameterized separately (e.g., AMBER's Lipid21 \cite{AMBERLipid21} and the CHARMM lipid force field \cite{CHARMMLipidFF}).
-Membrane systems can be simulated in multiple thermodynamic ensembles, but the semi-isotropic scheme $NP_{z}P_{xy}T$ is frequently employed, allowing the pressure in the bilayer plane ($xy$) to scale independently \cite{LipidBestPracticesCOMS}.
-These systems require more extensive equilibration than soluble proteins.
-Some workflows, such as CHARMM-GUI or Desmond membrane preparation, recommend a multi-protocol to relax membrane systems \cite{https://doi.org/10.1002/jcc.23702,DesmondManual}.
-
-Alchemical transformations that involve net charge changes can provide difficulties for membrane systems; phospholipids carry functional groups with nonzero formal charge and cannot be treated as a solvent with a uniform dielectric in the same way as water.
-Several approaches are reviewed in \cite{wu2022correction-51c}.
-The authors recommend the use of a co-alchemical ion, specifically making an ion of equivalent charge appear alongside the disappearing charged molecule to maintain system neutrality.
-
-In the case of a lipid-facing binding site, there are three additional challenges.
-First, simulation time should be extended to ensure sufficient sampling of lipid configurational space.
-Lipids may either bind alongside small molecules, acting as cofactors, or be displaced from the binding site by a ligand.
-If lipids participate in binding through protein-lipid or ligand-lipid interactions, it is necessary to sample these intermolecular interactions through additional $\lambda$ windows, longer simulation times, or different alchemical protocols.
-Displacing a lipid will have its own free energy cost which may take a long time to converge if it requires diffusion and rearrangement of nearby lipids.
-Second, considering the free energy cost of a ligand partitioning into the membrane may provide additional insights \cite{doi:10.1021/acs.jcim.1c01147}.
-Third, the free energy of binding for a small molecule may be strongly coupled to the lipid species.
-In some cases, the identity of the lipid may be known.
-But in other cases, it may be difficult to discern the identity of the lipid from X-ray or cryo-EM structures due to the flexible nature of lipid head groups and benchmarking is recommended to match experimental data.
 
@@
 The transformation finally culminates in a state where benzyl alcohol is present along with the additional dummy atom which was previously a corresponding hydrogen of the benzene.
-Fig.~\ref{fig:fig_topology}'s right branch depicts how such a dual topology works.
+Fig.~\ref{fig:fig_topology}'s middle branch depicts how such a dual topology works.
 Note that the two groups that are switched on or off never see each other.
 The 'mixed' intermediate atoms (M) are used as well, but they strictly interpolate between physical and dummy atoms, or vice versa.
+Bhati \textit{et al.}~\cite{bhati2017, bieniek2021} presented a variant of the single coordinate dual topology approach which is used in the TIES method.
+Its distinguishing feature is the application of some filters to remove atoms with differences in atom types and partial charges from the structural MCSS identified to ensure chemical similarity between all atom pairs in the final MCSS across both end-points.
+This is followed by ensuring identical charges for all MCSS atoms by averaging between the two end-points.
+The resultant change in the sum of atomic charges for all MCSS atoms is uniformly distributed across all atoms in the corresponding alchemical region conserving the overall ligand charge for both end-points.
+The approach ensures that the final MCSS is both structurally and chemically identical within thresholds, allowing the use of a unique set of MCSS atoms for both ligands and dual topology for their alchemical parts.
 
@@
 Whereas Ries et al.\  use distance restraints to keep two molecules on top of each other, corresponding atoms in NAMD are held on top of each other by constraints; in practice, this means that these coordinates are forced to be identical.
-AMBER's free energy implementation recently switched to a similar approach \cite{Lee_2023}.
+AMBER's free energy implementation recently switched to a similar approach \cite{Lee_2023,York_ACSPhysChemAu_2023_v3_p478}.
+While the implementation technically uses two sets of coordinates for the entire transforming region, selected atoms that have a direct atom-atom mapping are constrained to have common coordinates (sometimes referred to as a "common core"), while others that are mapped into dummy atoms and are treated with softcore potentials~\cite{Lee_JChemTheoryComput_2020_v16_p5512,Tsai_JChemTheoryComput_2023_v19_p640} have separable coordinates.
+Thus the common core part of the dual topology are constrained to have the same coordinates to maximize phase space overlap, whereas the softcore region has separate coordinates.
+The two topologies do not see each other, and this property enables this approach to be leveraged in the alchemical enhanced sampling (ACES) method~\cite{Lee_2023,Tsai_JChemInfModel_2024_v64_p7046,Zhang_JChemTheoryComput_2024_v20_p3935}.
 %The authors refer to this approach as hybrid single-dual topology.
@@
 & 2), however a dual coordinate set approach could use an atom mapping as well, depending on the sampling method, this will be explained in the next steps.
-Once a single coordinate set approach is selected is selected, a crucial next step is to identify the common atoms which will not be perturbed between the two molecules (respectivly end-states).
+Once a single coordinate set approach is selected, a crucial next step is to identify the common atoms which will not be perturbed between the two molecules (respectivly end-states).
 Rigorously, this process typically comprises a MCSS search of the molecules involved to identify the common substructure.
@@
 This problem is similar to the required restraints in ABFE Calculations (see \ref{ABFErest}).
-Few descriptions of possible approaches in RBFE calculations can be found for such approaches like the one suggested by \textit{Baumann et al.}~\cite{baumann2023broadening} based on boresch style restraints~\cite{boresch2003absolute}, with an elaborate automatic atom selection approach, for the seperated topology approach, or Restraintmaker~\cite{ries2022restraintmaker} for the linked-dualtopology approaches, which finds a small set of atoms to be linked via distance restraints, with a small distance to each other.
+Few descriptions of possible approaches in RBFE calculations can be found for such approaches like the one suggested by \textit{Baumann et al.}~\cite{baumann2023broadening} based on Boresch style restraints~\cite{boresch2003absolute}, with an elaborate automatic atom selection approach, for the seperated topology approach, or Restraintmaker~\cite{ries2022restraintmaker} for the linked-DCDT approaches, which finds a small set of atoms to be linked via distance restraints, with a small distance to each other.
 However the implementation and testing of an fully automatized dual-topology pipeline for RBFE dual topology approaches is still to our knowledge a remaining research problem and was only tested with small datasets.
@@
 When a dual coordinate approach is used, the noninteracting part of the respective other endstate is tied to the physical one, and will contribute to the partition function.
-By construction, relative free energy simulatins involve two analogous alchemical transformations, e.g., A(+D) -> B(+D) in gas phase and aqueous solution (RSFEs), or in the bound and unbound state (RBFEs).
-If dummy atoms are interacting with the physical system in such a way that $\mathsf{Z = Z(R) \times Z(R\text{-}D) \times Z(D)}$, where Z is the total partition function, Z(R) the partition function of the physical system, Z(D) the partition function contribution resulting from interactions between only the dummy atoms, and Z(R-D) the contribution resulting from the bonded connections between the physical atoms and dummy atoms, then their contributions (Z(D) and Z(R-D)) factor from the total partition function.
+By construction, relative free energy simulations involve two analogous alchemical transformations, e.g., A(+D) -> B(+D) in gas phase and aqueous solution (RSFEs), or in the bound and unbound state (RBFEs).
+If dummy atoms  interact with the physical system in such a way that $\mathsf{Z = Z(R) \times Z(R\text{-}D) \times Z(D)}$, where Z is the total partition function, Z(R) the partition function of the physical system, Z(D) the partition function contribution resulting from interactions between only the dummy atoms, and Z(R-D) the contribution resulting from the bonded connections between the physical atoms and dummy atoms, then their contributions (Z(D) and Z(R-D)) factor from the total partition function.
 As dummy atoms do not interact with the environment, their contribution, therefore, is identical in the two alchemical transformations required in the two legs of the relative transformation and will cancel exactly from the double free energy difference of interest.
@@
 These effectively, are equivalent to the formation of cyclic structures with potential erroneous contributions to the double free energy difference of interest.
-The type of restraints used in SepTop \cite{rocklin2013separated,Baumann_2023} and ATM \cite{Azimi_2022} aoid this problem.
+The type of restraints used in SepTop \cite{rocklin2013separated,Baumann_2023} and ATM \cite{Azimi_2022} avoid this problem.
+Additionally, TIES~\cite{bhati2017, bieniek2021} circumvents this issue by design as it does not need any restraints.
 
@@
 \begin{itemize}
-  \item Obviously, a system with and without dummy atoms should have the same energetic minimum.
+  \item Obviously, a system with and without dummy atoms should have the same energetic minimum; practitioners should test this as follows.
 Note that in larger groups of dummy atoms, redundant bonded terms involving only dummy atoms will be essential to avoid flapping, and the energy of the optimized geometry may be different from that of the physical system without dummy atoms.
@@
 Still, FEP+ implements one solution\cite{wang2017accurate}.
+TIES also overcomes this issue by using dual topology and ensuring that MCSS atoms are structurally and chemically identical~\cite{bhati2017, bieniek2021}.
 
@@
 However, in single topology relative free energy calculations, the atoms involved might be mutated to other atom types---for example, in a mutation of methane to methanol, one hydrogen might become an oxygen atom.
-The bonds with such atoms might not have any constraints, or if all bond are constrained, would have constraints of different lengths.
+The bonds with such atoms might not have any constraints, or if all bonds are constrained, would have constraints of different lengths.
 Some molecular dynamics engines are not set up to recognize this change, or at least not to correctly include contributions to the free energy from changing constraints/constraint length, so results for a transformation can be erroneous.
@@
 \paragraph{Handling the standard state in absolute free energy calculations.}
-For solvation free energy calculations, handling the standard state is typically straightforward, and treating it correctly simply means ensuring that the non-interacting solute still occupies essentially the same volume as the solute in the interacting system.
+For solvation free energy calculations, handling the standard state is typically straightforward, and treating it correctly simply means ensuring that the non-interacting solute still occupies essentially the same volume as the solute in the interacting system, e.g.
+that the simulation box remains essentially the same size.
 So typically in such cases no special care is required to ensure the correct standard state, as long as the \emph{experimental} data being analyzed uses the same standard state.
@@
 A sufficiently thorough exploration of the simulation box by the non-interacting ligand is, however, required for the formula to be valid.
-This is typically hard to achieve as the exploration process is governed by diffusion, and weak transient nonspecific binding will occur at other sites on the protein. 
+This is typically hard to achieve as the exploration process is governed by diffusion, and weak transient nonspecific binding will occur at other sites on the protein when the ligand is weakly interacting, making convergence difficult.
 The addition of a restraint limits the volume available to the non-interacting ligand, thus speeding the convergence of the sampling.
-In addition, when enhanced sampling methods such as Hamiltonian $\lambda$ exchange are used (see Sec.~\ref{sec:sampling_schemes}), the use of a restraint is typically necessary as it keeps the ligand in the binding site in the interacting state (see also Sec.~\ref{sec:weak-binders}) and generally reduce the round-trip time of replicas.
+In addition, when enhanced sampling methods such as Hamiltonian $\lambda$ exchange are used (see Sec.~\ref{sec:sampling_schemes}), the use of a restraint is typically necessary as it keeps the ligand in the binding site in the interacting state (see also Sec.~\ref{sec:weak-binders}) and generally reduces the round-trip time of replicas.
 When restraints are employed, the values of $V_L$ and $\xi_L$ are restraint-dependent, but for commonly employed restraints, these can be usually easily computed analytically or numerically by solving the relevant integral.
@@
 Alternatively, a set of restraints proposed by Boresch have also commonly been employed, where all six rigid-body degrees of freedom governing the orientation of the ligand relative to the receptor are restrained~\cite{boresch2003absolute, leitgeb2005alchemical}.
-Further restraints, such as on the overall ligand RMSD have also been used~\cite{woo2005calculation}.
+Further restraints, such as on the overall ligand RMSD, have also been used~\cite{woo2005calculation}.
 
@@
 Co-alchemical ion approaches introduce additional degrees of freedom and increase sampling requirements.
-
-\subsubsection{Handling poorly sampled water networks}
-The location of any water molecules within the binding site of interest must be determined prior to carrying out a free energy calculation.
-Experimental structures, such as those determined by X-ray crystallography and cryo-EM, are often sufficient for this, provided the resolution is good enough.
-However, we recommend using grand canonical sampling methods\cite{melling2023enhanced, samways2020grand} as a means of verifying the experimental water placement and determining any potential differences owing to the classical forcefield used.
-Many other water placement methods are available, all varying in their methodology, computational cost and accuracy.
-Details of these methods can be found in the publication from Samways et al.\cite{samways2021water}.
-
-In relative binding free energy calculations, if the hydration patterns differ between the two ligand endpoints then careful consideration will need to be given to the adaptation of the water network during the free energy calculation.
-We recommend performing the free energy calculation in the conventional way, i.e.
-with no enhanced water sampling, to determine whether the water network can adapt to the ligand perturbation itself.
-This is likely to happen if the hydration site is not particularly occluded from bulk solvent and water molecules can move freely from bulk into the binding site.
-Similarly, performing calculations in both directions (i.e.
-using both endpoints as the starting structure) can aid understanding the effect of the water network on the estimated binding affinities.
-Replica exchange moves in lambda space, instigated from correctly equilibrated water distributions at the end points, may be sufficient to sample water distributions appropriately across a ligand perturbation.
-
-If a conventional sampling method, such as molecular dynamics or Monte Carlo, is unable to rearrange the water network with sufficient efficacy to accommodate the perturbation, then an enhanced sampling method will be required to achieve accurate and converged results.
-One such approach is to include an absolute free energy calculation of the water molecule into the thermodynamic cycle\cite{ross2020enhancing, hamelberg2004standard, barillari2007classification, yu2008free, ge2022abfewat} such that the free energies of both the ligand transformation, and the decoupling of any displaced water molecules, are calculated separately and combined to achieve the total free energy change for the process.
-This method can be difficult to use in practice as the water molecule needs to be carefully restrained while being decoupled, and care needs to be taken to ensure water does not diffuse back into the hydration site as the water is decoupled.
-
-Monte Carlo approaches such as the hybrid MC/MD approach proposed by Bergazin et al.\cite{bergazin2021enhancing, ben-shalom2021fast, ge2022enhancing} allow waters to move between bulk solvent and the binding site with greater frequency by proposing MC moves that translate the water molecules within a spherical region that encompasses both bulk solvent and the binding site.
-This avoids the difficulty of performing a decoupling calculation on the bound waters but can be time-consuming and risks computational time being wasted on move proposals that are not of interest.
-
-Grand canonical (GC) methods propose the addition and removal of water molecules to and from a binding site in accordance with a defined chemical potential.
-Several implementations of GC moves being coupled to RBFE calculations exist, such as in Schrodinger's FEP+ package, introduced by Ross et al.\cite{ross2020enhancing} with moves proposed both within the binding site (to ensure the correct hydration pattern) and also within bulk solvent (to enhance the sampling of the bulk density).
-A cavity-bias technique was used for the binding site moves.
-Grand canonical alchemical perturbation (GCAP)\cite{brucemacdonald2018ligand} is also available in ProtoMS\cite{woods2018protoms}, whereby 2-dimensional titrations are performed in both lambda and chemical potential space.
-However, given ProtoMS only allows MC sampling to be performed, the configurational sampling of the protein is restricted.
-
-More recently, an OpenMM-based implementation allows for GC moves to be carried out during both equilibrium and nonequilibrium (NES) RBFE calculations\cite{melling2025developing}.
-These moves are performed in the form of grand canonical nonequilibrium candidate Monte Carlo (GCNCMC) moves\cite{melling2023enhanced} and are run during the equilibration phase at each lambda window in the case of equilibrium FEP calculations, or during the nonequilibrium switch in the NES calculations.
-
-Although often water sampling is less of an issue in ABFE calculations, handling changes to water networks in buried pockets during ligand coupling/decoupling can be addressed in a similar fashion to RBFEs with grand canonical methods.
-
-% in ABFE calculations is less established, although in principle, the same rules apply.
-When conventional sampling of water is insufficient such that water does not efficiently bind during ligand decoupling, performing extra ABFE calculations for the removal or addition of water molecules to the thermodynamic cycle is one way of ensuring the contribution of these solvent effects is captured.
-Performing grand canonical water moves along side a more traditional ABFE calculation of a ligand will ensure that any changes to binding pocket hydration states are sampled \textit{in situ}.
-
-Employing the techniques described above with care should allow the water network to adapt during the perturbation of the ligand, ensuring more accurate and converged affinity estimations that are independent of the starting configuration and direction in which the calculation is performed.
 
@@
 This functional form has exactly the property we are looking for: it recovers the Lennard-Jones potential when $\vec{\lambda}=1$, and the at other endpoint ($\vec{\lambda}=0$), it is exactly zero for all $r_{ij}$ everywhere, and as $\vec{\lambda}$ goes to zero, the $\alpha(1-\vec{\lambda})$ term lowers the infinite energy in the core.
-There are several different variants of the same functional form~\cite{zacharias1994separationshifted, beutler1994avoiding,pham2011identifying}, but the one given in eq.~\ref{eq:softcore} is easy to understand and implement and fairly numerically stable.
+There are several different variants of the same functional form~\cite{zacharias1994separationshifted, beutler1994avoiding,pham2011identifying,Lee_JChemTheoryComput_2020_v16_p5512,Tsai_JChemTheoryComput_2023_v19_p640}, but the one given in eq.~\ref{eq:softcore} is easy to understand and implement and fairly numerically stable.
 This functional form is shown in \textbf{C} and \textbf{D} of Fig.~\ref{fig:fig_what_is_lambda}.
@@
 Although this is a simpler path to understand cognitively and can take advantage of the soft-core potential from Eq.~\ref{eq:softcore}, this can lead to more intermediate steps and thus be more computationally expensive.
-Other schemes, such as simultaneously changing electrostatic and Lennard-Jones interactions with electrostatic soft-core potentials~\cite{steinbrecher2007nonlinear}, as already discussed above, may be implemented with fewer intermediate states but could require fine-tuning of electrostatic and Lennard-Jones soft-core parameters to avoid numerical instabilities.
+Other schemes, such as simultaneously changing electrostatic and Lennard-Jones interactions with electrostatic soft-core potentials~\cite{steinbrecher2007nonlinear}, as already discussed above, may be implemented with fewer intermediate states but could require fine-tuning of electrostatic and Lennard-Jones soft-core parameters to avoid numerical instabilities~\cite{Tsai_JChemTheoryComput_2023_v19_p640}.
 At the time of writing, there has not been conclusive evidence to suggest the separate or simultaneous approach is in general better than the other, all factors considered, so discretion should be left up to the user as to what is viable from both hardware resources, and what the simulation software supports.
@@
 In cases where the system's slowest degrees of freedom are coupled with the alchemical variable, transitioning between alchemical states can indirectly enhance configurational sampling.
-This effect is particularly pronounced when certain alchemical states lack the energy barriers present at the fully coupled state (e.g.
+This effect is particularly pronounced when certain alchemical states, such as the decoupled state or loosely coupled states, lack the energy barriers often present at the fully coupled state (e.g.
 $\lambda=0$ in Fig.
@@
 For example, it can arise when the free energy difference of interest needs to account for metastable states separated by torsional barriers, such as the solvation free energy of a ligand having different rotamer states, the methylation free energy of a nucleobase in a duplex ~\cite{hsu2023alchemical}, and the binding free energy that involves flexible side chains near the binding site~\cite{la2022alchemical}.
-Similar sampling challenges have also been reflected in several recent studies, which highlights the sensitivity of alchemical calculations to initial conformations, whether derived from crystallographic structures~\cite{suruzhon2021sensitivity, baumann2024impact} or docking protocols~\cite{cappel2020impact}.
+Similar sampling challenges have also been reflected in several recent studies, which highlight the sensitivity of alchemical calculations to initial conformations, whether derived from crystallographic structures~\cite{suruzhon2021sensitivity, baumann2024impact} or docking protocols~\cite{cappel2020impact}.
 
@@
 Notably, in cases where alchemical biases/weights are applied to facilitate even alchemical sampling across states, such as in an expanded ensemble simulation~\cite{lyubartsev1992new}, scenario Fig.~\ref{fig:configurational_sampling}\textbf{C} may evolve toward scenario Fig.~\ref{fig:configurational_sampling}\textbf{B} as the alchemical free energy profile is flattened out, while the basin in the configurational direction persists.
-
-There are multiple strategies to alleviate such sampling challenges in an alchemical calculation.
-A straightforward but computationally expensive method is to run longer simulations with multiple replicates to increase conformational heterogeneity~\cite{suruzhon2021sensitivity}.
-Another approach involves performing separate alchemical calculations from different metastable states, then computing the overall free energy difference as the negative log-sum-exp of the Boltzmann weights from each metastable state~\cite{hsu2023alchemical}.
-However, this method is only applicable to cases like Fig.~\ref{fig:configurational_sampling}\textbf{B}, where the system remains trapped in its initial metastable state throughout the calculation.
-A generally more effective strategy is to bias relevant configurational collective variables (CVs) during the alchemical simulation.
-This can be achieved by enhancing sampling within individual alchemical states and swapping conformations between alchemical states (preferably from different metastable states) via replica exchange protocols.
-More recently, several approaches have been proposed to simultaneously bias both configurational CV and alchemical variables, which can effectively address both scenarios Fig.~\ref{fig:configurational_sampling}\textbf{B} and \textbf{C}.
-Examples include alchemical metadynamics~\cite{hsu2023alchemical}, $\lambda$-ABF (adaptive biasing force)~\cite{lagardere2024lambda} and its combination~\cite{zhou2025zooming} with well-tempered metadynamics~\cite{barducci2008well}, to name a few.
-
-Notably, detecting and addressing configurational sampling challenges in alchemical calculations remains a non-trivial task.
-In relative free energy calculations, large cycle closure errors are commonly used as indicators of insufficient sampling, although small errors do not necessarily imply that sampling is adequate~\cite{suruzhon2021sensitivity}.
-In other types of calculations, diagnosing sampling deficiencies typically requires prior knowledge of relevant metastable states, as well as collective variables that distinguish them.
-Such collective variables are often based on physical intuition, experimental insights, or preliminary computational analyses.
-However, optimal collective variables are rarely known \textit{a priori}, complicating the systematic detection of sampling issues.
-Although recent ad-hoc machine learning methods for optimal CV identification have shown promise~\cite{gokdemir2025machine, sidky2020machine, bonati2021deep} and could in principle be straightforwardly integrated into alchemical sampling protocols, systematic studies explicitly demonstrating such integration have yet to appear.
-Consequently, widely accepted best practices about automated diagnostic protocols capable of identifying and resolving configurational sampling bottlenecks have not yet been established.
 
@@
 
-\subsubsection{Which sampling scheme will work best for my problem?}
+There are multiple strategies to alleviate such sampling challenges in an alchemical calculation.
+A straightforward but computationally expensive method is to run longer simulations with multiple replicates to increase conformational heterogeneity~\cite{suruzhon2021sensitivity}.
+Another approach, known as integration over parts,\cite{jayachandran2006parallelized}, involves performing separate alchemical calculations limited to a series of binding macrostates and  then computing the overall free energy difference using the free energy combination formula~\cite{gallicchio2011recentadv,hsu2023alchemical}, i.e., computing the negative log-sum-exp of the Boltzmann weights from each macrostate.
+In the integration over parts approach, restraining potentials can be used to restrict the sampling to specific macrostates,\cite{khuttan2023taming}.
+However, in practice, restraining potentials are sometimes omitted when the states are known to be separated by large free energy barriers and do not interconvert within the simulation time scale, such as cases like Fig.~\ref{fig:configurational_sampling}\textbf{B}.
+A generally more effective strategy is to bias relevant configurational collective variables (CVs) during the alchemical simulation.
+This can be achieved by enhancing sampling within individual alchemical states and swapping conformations between alchemical states (preferably from states separated by large free energy barriers) via replica exchange protocols.
+More recently, several approaches have been proposed to simultaneously bias both configurational CV and alchemical variables, which can effectively address both scenarios Fig.~\ref{fig:configurational_sampling}\textbf{B} and \textbf{C}.
+Examples include alchemical metadynamics~\cite{hsu2023alchemical}, $\lambda$-ABF (adaptive biasing force)~\cite{lagardere2024lambda}, and optimized alchemical enhanced sampling (ACES)~\cite{Lee_JChemTheoryComput_2023_v19_p472}  and approaches that integrate well-tempered metadynamics~\cite{barducci2008well} with alchemical methods~\cite{zhou2025zooming, khuttan2024make}, to name a few.
+
+Notably, detecting and addressing configurational sampling challenges in alchemical calculations remains a non-trivial task.
+In relative free energy calculations, large cycle closure errors are commonly used as indicators of insufficient sampling, although small errors do not necessarily imply that sampling is adequate~\cite{suruzhon2021sensitivity}.
+In other types of calculations, diagnosing sampling deficiencies typically requires prior knowledge of relevant metastable states, as well as collective variables that distinguish them.
+Such collective variables are often based on physical intuition, experimental insights, or preliminary computational analyses.
+However, optimal collective variables are rarely known \textit{a priori}, complicating the systematic detection of sampling issues.
+Although recent ad-hoc machine learning methods for optimal CV identification have shown promise~\cite{gokdemir2025machine, sidky2020machine, bonati2021deep} and could in principle be straightforwardly integrated into alchemical sampling protocols, systematic studies explicitly demonstrating such integration have yet to appear.
+Consequently, widely accepted best practices about automated diagnostic protocols capable of identifying and resolving configurational sampling bottlenecks have not yet been established.
+
+\subsubsection{Which $\lambda$-sampling scheme will work best for my problem?}
 \label{sec:sampling_schemes}
@@
 The simplest approach involves running an independent simulation at each of the predefined $\vec{\lambda}$ values (see Fig.~\ref{fig:fig_sampling_scheme}\textbf{A}).
-This type of scheme is currently used for AMBER TI calculations~\cite{song2019using} and for Sire as implemented in BioSimSpace~\cite{hedges2019biosimspace}.
+This type of scheme is currently used for some AMBER TI calculations~\cite{song2019using} and for Sire as implemented in BioSimSpace~\cite{hedges2019biosimspace}.
 However, if these simulations can be run simultaneously with communication between them, a simple extension allows mixing between these replicas.
@@
 The basic idea of the replica exchange scheme is shown in Fig.~\ref{fig:fig_sampling_scheme}\textbf{B}.
-It is supported in various software packages that provide alchemical implementations, such as GROMACS~\cite{aldeghi2015accurate}, GROMOS~\cite{hritz2008hamiltonian,hritz2007optimization}, FEP+~\cite{wang2015accurate}, and NAMD~\cite{jiang2019computing}.
+It is supported in various software packages that provide alchemical implementations, such as GROMACS~\cite{aldeghi2015accurate}, GROMOS~\cite{hritz2008hamiltonian,hritz2007optimization},
+AMBER~\cite{Lee_JChemInfModel_2020_v60_p5595}, FEP+~\cite{wang2015accurate}, and NAMD~\cite{jiang2019computing}.
+A specific variant of Hamiltonian replica exchange is FEP/REST2~\cite{rest2, feprest} that involves ``heating'' a small portion of the protein-ligand system to enhance the sampling within the binding pocket.
+It has been promoted as a great way to enhance sampling and is employed by default in Schr\"{o}dinger's FEP+ package.
+However, it should be noted that several studies have questioned its reliability and reported that it degrades the quality of free energy predictions as compared to those based on standard MD simulations without any enhanced sampling~\cite{wan2020fep+, bhati2022}.
+Therefore, a blind application of this sampling scheme is not recommended and it must be used with caution if at all.
+A variant of this method that overcomes some of the reported issues with FEP/REST2 is called alchemical enhanced sampling (ACES)~\cite{Lee_JChemTheoryComput_2023_v19_p472}.
+ACES creates localized targeted "enhanced sampling states" that are connected to the real state end-points through HREMD.
+The method leverages the dual topology framework to produce a concerted “counter-diffusion” of alchemical states that leads to minimal rearrangement of the environment along the alchemical path~\cite{York_ACSPhysChemAu_2023_v3_p478}.
+The computational overhead of the method is negligible relative to more conventional alchemical free energy simulations with HREMD.
 
@@
 
-The non-equilibrium switching (Fig.~\ref{fig:fig_sampling_scheme}\textbf{D}) approach makes use of short out-of-equilibrium simulations~\cite{aldeghi2018accurate}.
+The non-equilibrium (NEQ) switching (Fig.~\ref{fig:fig_sampling_scheme}\textbf{D}) approach makes use of short out-of-equilibrium simulations~\cite{aldeghi2018accurate}.
 In this protocol, only end state $\vec{\lambda}$ replicas ($\vec{\lambda}$=0, $\vec{\lambda}=1$) are simulated at equilibrium; intermediate information is generated from non-equilibrium simulations that rapidly transition between end states.
@@
 
-The non-equilibrium (NEQ) approaches offer some advantages over the equilibrium sampling based methods.
-For example, the NEQ protocol allows equilibrium simulations of each end-state to be performed independently without pre-defining ligand network: the transitions between the ligands can be generated afterwards.
-This eliminates the redundant equilibrium sampling, e.g.
-in a starmap network, FEP based approaches would require sampling the central ligand for each edge, while NEQ would sample it once and connect to the other ligands afterwards.
-The NEQ protocol also provides flexibility in combining different structures in free energy calculations, e.g.
+The NEQ approaches offer some advantages over the equilibrium sampling based methods.
+For example, the NEQ protocol allows equilibrium simulations of each end-state to be performed independently without pre-defining a ligand network: the transitions between the ligands can be generated afterwards.
+This eliminates the redundant equilibrium sampling if one end state is present in multiple perturbations.
+Equilibrium sampling based approaches would require sampling the central ligand for each edge, while NEQ would sample it once and connect to the other ligands afterwards.
+The NEQ protocol also provides a simple approach in combining different structures in free energy calculations, e.g.
 simulations in states at $\lambda=0$ and $\lambda=1$ may be initialized with the independently generated ligand coordinates.
 Another example would be using true apo and holo protein structures as the end states to initialize simulations for absolute binding free energy calculations.
+Bhati \textit{et al.} recently published two studies directly comparing the equilibrium (EQ) and NEQ approaches at unprecedented scales for both RBFE and ABFE calculations using their ensemble simulation-based TIES method~\cite{wan2023eqvsneq, bhati2025}.
+Further, they systematically studied the effect of various parameters involved in setting up calculations for both approaches and recommended optimal protocols in each case for industrial applications.
+They highlighted several limitations of the NEQ approach that should be accounted for while performing such calculations to ensure reliability of predictions.
 
@@
 Single replica schemes and non-equilibrium schemes are not as established yet because of potential failure modes, but are very promising for use in the near future.
+Further, ensemble simulations are recommended irrespective of the choice of sampling scheme to ensure reliability of results~\cite{bhati2018, bhati2019, wan2020fep+, wade2022, wan2021uq, bhati2025, wan2023eqvsneq}.
 
@@
 In particular, if the rate of change in the free energy estimate is significant when this condition is met, the simulation may not be locally converged, and more sampling may be necessary to determine a stable free energy estimate which is no longer changing significantly over time.
+A few studies reported systematic variation of simulation lengths using ensemble simulations over an extended range to provide recommendations on appropriate values applicable for most protein-ligand systems.
+The authors suggest using 4 ns and 10 ns for RBFE~\cite{bhati2017, bhati2022, wan2023eqvsneq} and ABFE~\cite{bhati2025} calculations respectively.
+These values may be used as a general guideline.
 
 However, this is not the only metric which can or should be used, as the uncertainty only captures the information about the sampled phase space, not necessarily the entirety of the phase space.  
-For example, convergence of relative free energy calculations in predictive simulations where the entire phase space is not known in advance requires sampling the different kinetically stable states~\cite{mobley2012perspective}.
+For example, convergence of relative free energy calculations in predictive simulations where the entire phase space is not known in advance, requires sampling the different kinetically stable states~\cite{mobley2012perspective}.
 This highlights the importance of choosing the correct thermodynamic path to ensure you sample the required thermodynamic states as discussed in Sec.~\ref{sec:important_path}.
@@
 This can be done through the uncertainty estimator built into certain analysis tools such as MBAR, or through more general statistical tools like bootstrap sampling.
-It should be noted however, that uncertainty estimates have the same limitations as other metrics of convergence, as they are only an uncertainty based on the phase space sampled so far in a simulation, and cannot account for states not sampled, and it is worth considering that they will be an underestimation of the true uncertainty.
-
+It should be noted however, that uncertainty estimates have the same limitations as other metrics of convergence, as they are only an uncertainty based on the phase space sampled so far in a simulation, and cannot account for states not sampled, and it is worth considering that they will be an underestimation of the true uncertainty~\cite{wan2020fep+, wan2021uq, vassaux2021, gapsys2020large, wade2022, bhati2017, bhati2022, bhati2025, wan2023eqvsneq, bhati2018, bhati2019}.
 A target statistical uncertainty should be chosen at the onset of the simulation to avoid excessively long simulations, or falling into the trap of running until the free energy estimate is "good enough," which is subjective and has no defined criteria.
@@
 If more samples are needed, the simulations can be resumed, or, started again in different initial conditions.
+Bhati \textit{et al.} recommended such a scheme for large scale RBFE and ABFE calculations using TIES~\cite{bhati2022, bhati2025, wan2023eqvsneq, coveney2022corners}.
+They suggest a set of optimal parameters (ensemble size, simulation length, number of lambda-states, number and length of NEQ transitions, etc.) to be used initially.
+Then, only those systems which do not satisfy pre-defined criteria (uncertainties below 0.5 kcal/mol and overlap between forward and reverse work values in case of NEQ calculations) should be further evaluated using additional calculations (in the way suggested) until these criteria are met.
+Such a scheme optimizes the overall computational efficiency for the full dataset in large scale applications.
 
@@
 The uncertainty in the free energy, for example, can be estimated in multiple ways, e.g.
-through standard error propagation methods (including MBAR's estimator, which is based on the same principles as standard error propagation), through bootstrap methods, and through multiple independent runs.
+through standard error propagation methods (including MBAR's estimator, which is based on the same principles as standard error propagation), through bootstrap methods, and through ensemble simulations.
+However, as we have repeatedly pointed out, only ensemble simulations can capture the true uncertainty in MD based predictions.
+All other methods underestimate the true uncertainty as reported by several authors~\cite{wan2020fep+, wan2021uq, vassaux2021, gapsys2020large, wade2022, bhati2017, bhati2022, bhati2025, wan2023eqvsneq, bhati2018, bhati2019}.
 Independent of how the property is estimated, it is important to remember that results of any free energy analysis are \textit{estimations of the given property}, not the true underlying value of the property itself.
@@
 However, from various previous free energy studies using 1 fs integration timesteps it is not clear that a significant error would be introduced into a free energy estimate and may warrant some further investigation.
+
+\subsection{Alchemical Transfer Method for Absolute and Relative Binding Free Energy Calculations}
+
+\subsubsection{Introduction}
+
+The Alchemical Transfer Method (ATM)~\cite{wu2021alchemical,azimi2022relative} is an alchemical framework for estimating binding free energies that targets generality with a rigorous and straightforward implementation, with as few additions as possible to standard non-alchemical models of molecular systems.
+Additionally, the method targets complex ligand transformations in drug design, involving scaffold-hopping and charge-changing transformations that are not well supported by traditional algorithms.
+ATM's overarching philosophy is to deploy robust and rigorous free energy estimation algorithms with the simplest possible procedures for developers and end users.
+One critical design principle is the seamless support for advanced molecular potential energy functions, such as quantum-mechanical, polarizable, and neural network potentials, in free energy-based drug discovery applications.
+
+To reach its objectives, ATM represents the alchemical potential energy function through a coordinate transformation~\cite{azimi2022application,azimi2022relative}.
+Unlike traditional alchemical approaches that rely on parameter interpolation of interaction potentials, ATM employs an energy interpolation alchemical energy function~\cite{konig2021efficient} where the target potential is expressed as the reference potential after the ligand is translated from the solvent to the binding site (Figure~\ref{fig:atm-illustration}).
+The binding free energy calculation is performed directly, without solution and receptor legs, as well as electrostatic and non-electrostatic steps.
+The simulation system consists of a single solvent box prepared with conventional setup tools and without pairwise soft-core potentials.
+The leveraging of the natural capabilities of MD engines to apply coordinate transformations is a distinguishing feature of ATM.
+Alchemical transformations for binding implemented through coordinate transformations are computationally straightforward and applicable with any potential energy function, requiring only the evaluation of energy and derivatives before and after the coordinate transformation.
+Conversely, because it recalculates the potential energy and its gradients twice at each time-step, ATM's strategy suffers from the downside that it is as much as twice as slow per time-step as conventional parameter interpolation alchemical approaches.
+
+%mentions parameter vs. energy interpolation 
+%@article{konig2021efficient,
+%  title={Efficient Alchemical Intermediate States in Free Energy Calculations Using $\lambda$-Enveloping Distribution Sampling},
+%  author={K\"{o}nig, Gerhard and Ries, Benjamin and H\"{u}nenberger, Philippe H and Riniker, Sereina},
+%  journal={Journal of Chemical Theory and Computation},
+%  volume={17},
+%  number={9},
+%  pages={5805--5815},
+%  year={2021},
+%  publisher={ACS Publications}
+%}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.95\linewidth]
+    {paper/figures_atm/atm_scheme_abfe_rbfe.pdf}
+    \caption{\small Alchemical transfer method illustrated for (\textbf{A}) Absolute Binding Free Energy (ABFE) and (\textbf{B}) Relative Binding Free Energy (RBFE) calculations.
+In (A), the receptor (blue) is shown with the ligand in bulk solvent being translated to the active site along a vector $+h$.
+In (B), the RBFE setup is shown with two ligands: one in the binding site and the other in bulk solvent.
+A vector $h$ separates the ligands, and the perturbation energy is evaluated by virtually swapping their positions at each MD step.}
+    \label{fig:atm-illustration}
+\end{figure}
+
+\subsubsection{Methodology}
+
+Using the notation in Eq.~(\ref{eq:reduced-potential}), if $u(\vec{q})$ is the physical ($\lambda$-independent) reduced potential of the system when the ligand is at a location in the solvent away from the receptor and $\vec{h}$ is a displacement vector that brings the ligand atoms from the position in the solvent to the receptor binding site (Figure~\ref{fig:atm-illustration}), the ATM \emph{reduced binding energy function}
+\begin{equation}\label{eq:ATM-binding-energy}
+b(\vec{q}) = u(\vec{q}+\vec{h}) - u(\vec{q})
+\end{equation}
+is the change in the system's potential energy when the ligand is rigidly moved from the solvent to the receptor binding site. The ATM alchemical reduced potential is then written as
+\begin{equation}\label{eq:ATM-alchemical-potential}
+    u(\vec{q}; \vec{\lambda}) = u(\vec{q}) + w[b(\vec{q});\vec{\lambda}]
+\end{equation}
+where $w(b,\vec{\lambda})$ is the \emph{perturbation energy function}.
+The perturbation energy function can be any function as long as it is zero at the $\lambda$-state corresponding to the unbound state and it is equal to the binding energy $b(\vec{q})$ at the bound state:
+\begin{eqnarray}
+    w[b(\vec{q});\vec{\lambda}_{\rm U}] = 0           & \quad & \vec{\lambda}_{\rm U}: {\rm Unbound\ State} \\
+    w[b(\vec{q});\vec{\lambda}_{\rm B}] = b(\vec{q})  & \quad & \vec{\lambda}_{\rm B}: {\rm Bound\ State}
+\end{eqnarray}
+
+Current ATM implementations (see below) employ the softplus perturbation energy function\cite{pal2019perturbation}
+\begin{equation}\label{eq:ATM-softplus}
+w(b;\vec{\lambda}) = w[b_{\rm sc}(b);\vec{\lambda}] =  \frac{\lambda_2 - \lambda_1}{\alpha} \ln\left\{ 1 + e^{-\alpha [b_{\rm sc} - b_0]} \right\} + \lambda_2 b_{\rm sc}
+\end{equation}
+where the parameters ($\lambda_1>=0$, $\lambda_2>=0$, $\alpha>0$, and $b_0$) are functions of the $\lambda$-state, $\vec{\lambda}$.
+The softplus non-linear perturbation function satisfies the correct limits when $\lambda_1 = \lambda_2 = 0$ at the unbound state and $\lambda_1 = \lambda_2 = 1$ at the bound state.
+The quantity $b_{\rm sc}$ in Eq.~(\ref{eq:ATM-softplus}) is the soft-core binding energy function
+\begin{equation}
+  b_{\rm sc}(b)=
+\begin{cases}
+b & b \le b_c \\
+(b_{\rm max} - b_c ) f_{\rm sc}\left[\frac{b-b_c}{b_{\rm max}-b_c}\right] + b_c & b > b_c
+\end{cases}
+\label{eq:ATM-soft-core-general}
+\end{equation}
+with
+\begin{equation}
+f_{\rm sc}(y) = \frac{z(y)^{a}-1}{z(y)^{a}+1} \label{eq:rat-sc} \, ,
+\end{equation}
+and
+\begin{equation}
+    z(y)=1+2 y/a + 2 (y/a)^2
+    \label{eq:end-of-softplus}
+\end{equation}
+The soft-core perturbation energy function is designed to avoid singularities near the initial state of the alchemical transformation.\cite{pal2019perturbation,khuttan2021alchemical} The parameters $b_{\rm max}$, $b_c$, and $a$ are set to smoothly cap the perturbation energy to a maximum positive value without affecting it away from the singularity.
+The softplus function has the nice property that is linear for $b<<b_0$ and $b>>b_0$ but with different slopes,  $\lambda_1$ and $\lambda_2$, in these two regimes.
+The bilinear property of the softplus function allows for greater flexibility in creating smoother and more connected alchemical paths that lead to faster convergence~\cite{pal2019perturbation, khuttan2021alchemical}.
+
+The values of the binding energies sampled at regular intervals at each $\lambda$-state during the ATM alchemical simulation are the inputs for binding free energy estimation (see Section \label{sec:data_analysis}) through multi-state thermodynamic reweighting algorithms such as MBAR and UWHAM~\cite{shirts2008statistically,Tan2012}.
+One requirement for these algorithms is the evaluation of the perturbation energy $w(b,\vec{\lambda})$ for each sample $b$ at all $\lambda$-states.
+The analytical relationships Eq.~(\ref{eq:ATM-softplus})--(\ref{eq:end-of-softplus}) between the binding energy and the perturbation energy make this step straightforward with ATM.
+In contrast, conventional free energy perturbation methods that employ soft-core pair potentials lack analytical perturbation energy functions and require a more complex process involving calling the molecular dynamics engine to rescore simulation trajectories.
+The more straightforward free energy estimation process afforded by ATM saves computing and storage costs, and, since it is based solely on binding energies, allows for changing the perturbation energy function $w(b;\vec{\lambda})$ without running new simulations.
+The simplification of the free energy estimation process fits the general ATM philosophy of streamlining and generalizing alchemical binding free energy calculations as much as possible.
+
+{\bf The ATM Alchemical Intermediate.} Due to atomic clashes with the receptor, the reference potential $u(\vec{q})$ for the unbound state becomes ill-defined near the bound state.
+On the other hand, taking the bound state as the reference leads to numerical instabilities near the unbound state due to clashes with solvent atoms.
+The solution to this conundrum is to split the alchemical binding process into two legs that connect the unbound and bound states to the same alchemical intermediate state midway between the two end states (Figure~\ref{fig:atm_alchemical_intermediate})~\cite{wu2021alchemical,azimi2022relative}.
+In the alchemical leg (Leg 2) that connects the unbound state to the alchemical intermediate, the unbound potential is used as a reference potential.
+The reference potential for the leg connecting the bound state and the alchemical intermediate (Leg 1) is instead the potential energy function of the bound state.
+The transition from one alchemical potential function to another occurs transparently to the user in the ATM replica exchange implementations (see below).
+According to the ATM thermodynamic scheme (Figure~\ref{fig:atm_alchemical_intermediate}), the excess binding free energy is obtained as the difference between the free energy changes of the two legs:
+\begin{equation}
+\Delta G_{b, \text{exc.}} = \Delta G_{\text{Leg2}} - \Delta G_{\text{Leg1}}
+\end{equation}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.95\linewidth]
+    {paper/figures_atm/atm_alchemical_intermediate.pdf}
+    \caption{Illustration of ATM's thermodynamic path.
+At the thermodynamic intermediate (top, center), the ligand interacts at half the strength with the receptor and half strength with the solvent.
+The unbound state (bottom, right) is converted to the alchemical intermediate during Leg 2.
+Similarly, the bound state (bottom, left) is converted to the alchemical intermediate during Leg 1.
+The binding free energy is recovered by taking the difference between the free energies of Leg 2 and Leg 1.
+The receptor is depicted in blue, and the ligand is shown in gray.
+The faint gray color of the ligand in the alchemical intermediate represents its 50\% interaction strength with the receptor and the bulk of the solvent.}
+    \label{fig:atm_alchemical_intermediate}
+\end{figure}
+
+\subsubsection{ATM Applications: ABFE, RBFE, Relative Binding Selectivity, Macromolecular Binding}
+
+The method described so far, involving the translation of the ligand to the receptor binding site, calculates the standard, or absolute binding free energy (ABFE)~\cite{wu2021alchemical,azimi2022application}.
+A straightforward generalization of ATM's ligand translation transformation enables the calculation of the relative binding free energy of two ligands for the same receptor (RBFE) using a dual-coordinate approach~\cite{azimi2022relative}.
+In ATM's RBFE approach, one ligand is initially placed in the solvent, and the other is placed in the binding site.
+The alchemical potential is then expressed similarly to Eq.~(\ref{eq:ATM-alchemical-potential}) except that, as one ligand is translated into the binding site as in the ABFE protocol, the other ligand is translated in the opposite direction such that the position of the two ligands is swapped.
+Hence, ATM's ABFE and RBFE protocols differ only in the coordinate transformation: ligand translation for ABFE and ligand swapping for RBFE.
+
+ATM's RBFE protocol has recently been extended to the swapping of ligand positions between two receptors~\cite{azimi2024selectivity}.
+The resulting free energy measures the relative selectivity of the two ligands for the two receptors, that is, by how much the selectivity of a ligand for one receptor over another is affected by transforming it into the other.
+This information is critical for the design of drugs that are highly specific and not only strong binders of the target receptor.
+
+The types of coordinate transformations that can be implemented with ATM are not limited to translations.
+For example, coordinates swapping, where the coordinates of an atom are replaced with those of another, has been used to accelerate the RBFE convergence rate between congeneric ligands and to extend ATM to the study of the influence of residue mutations on the binding affinities of protein-peptide and protein-protein complexes~\cite{gallicchio2025mutants}.
+
+\subsubsection{ATM System Setup and Best Practices}
+
+Since it does not require alchemical topologies, preparing molecular complex systems for ATM is the same as for standard molecular dynamics.
+While we described the ABFE protocol in the binding direction, in practice, it is often more convenient to set up the system with the ligand bound to the receptor rather than placed in the solvent.
+If so, the ABFE alchemical calculation is performed in the dissociation direction by essentially reversing the two ATM legs in Figure~(\ref{fig:atm_alchemical_intermediate}).
+To prepare a system for RBFE, one ligand is placed in the binding site and another in the solvent, ensuring their distance is approximately equal to the displacement vector.
+In practice, the two ligands are prepared as superimposed in the binding site, and one of them is translated by the displacement vector to reach the desired location in the solvent.
+
+The ligand displacement vector, illustrated in Figure~\ref{fig:atm-illustration}, is a necessary input for ATM as it determines the ligands' positions in the solvent to represent their unbound states.
+While the dependence of ATM's predictions on this aspect of the method has not been studied systematically, in principle, it should not matter where a ligand is placed in the solvent as long as it is sufficiently far away from the receptor to call the resulting configuration \emph{unbound}.
+Three layers of water molecules, or approximately $10$ \AA, is probably the minimum distance between any two pairs of ligand and receptor atoms to ensure a reasonable degree of decorrelation between the two.
+In practice, the choice of the displacement vector is a trade-off between ensuring that the ligand is sufficiently removed from the receptor when unbound and the size of the solvent box required to contain the receptor and the displaced ligand.
+Some ATM implementations have automated this aspect of the system's setup.
+
+The receptor binding region is typically specified by a flat-bottom harmonic potential restraint between the centroids of a set of receptor and ligand atoms as outlined in Section \ref{sec:standardstate-restraints}.
+For RBFE, a second binding region restraint, identical to the first but displaced into the solvent, restrains the position of the second ligand.
+Although alignment restraints (see below) are more commonly used for this purpose.
+Similarly to other alchemical approaches, orientational restraints can be further deployed to specify the binding pose more precisely~\cite{khuttan2023taming}.
+
+As is common in dual-coordinate alchemical binding implementations~\cite{Ries_2022}, ATM RBFE calculations typically employ optional restraints to maintain the two ligands in approximate alignment.
+ATM's implementation consists of positional and orientational restraints between the internal coordinate frames of the two ligands defined by three reference atoms (Figure~\ref{fig:atm_lig_alignment})~\cite{azimi2022application}.
+A positional restraint keeps the origins of the two frames at a separation approximately equal to the displacement vector, and the axes of the frames are used to impose orientational restraints.
+The objective is to keep the unbound ligand in a position and orientation such that the translation into the binding site places its atoms near the corresponding atoms of the bound ligand.
+Robust alignment is ensured by selecting reference atoms from the rigid, common cores of the ligands.
+The first reference atoms are generally located near the center of the rigid core to define the origin of the coordinate frame.
+The second reference atoms are selected as far as possible from the first, still within the rigid core, to establish an axis that represents the core's orientation.
+Corresponding atoms in the second ligand are identified along the same axis and aligned by ATM, even if perfect overlap is not possible.
+The third reference atoms define a plane with the first two atoms, ensuring consistent rotational alignment (roll) of the rigid cores.
+These are chosen such that they remain approximately coplanar, facilitating close ligand alignment.
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.95\linewidth]{paper/figures_atm/atm_lig_alignment.pdf}
+    \caption{Illustrative example of the selection of the reference atoms for ATM's ligand alignment restraints between two ligands \textbf{A} and \textbf{B}.
+Atoms \textbf{$1_A$} and \textbf{$1_B$} are selected as the first reference points, positioned near the center of the rigid common core.
+These atoms define the origin of the coordinate frames, ensuring that both ligands remain at a fixed distance.
+Atoms \textbf{$2_A$} and \textbf{$2_B$} are chosen from within the common rigid core, positioned as far as possible from the first reference atom.
+The axis defined by $2_A$ and $2_B$ is kept approximately collinear.
+Finally, atoms \textbf{$3_A$} and \textbf{$3_B$} are chosen to maintain the planar alignment of the rigid cores, preserving the roll of the common structural framework.
+Reference atoms are preferably not selected from flexible regions of the ligands, as changes in these regions could alter the relative orientation between the ligands.
+    }
+    \label{fig:atm_lig_alignment}
+\end{figure}
+
+Additional optional restraints are often used in ATM to limit the conformational space to be sampled to improve convergence and reproducibility of binding free energy estimates.
+For example, the AToM-OpenMM~\cite{atom-openmm-github2025} protocol applies by default weak flat-bottom harmonic positional restraints to the C\textsubscript{$\alpha$} atoms of structurally rigid regions of the protein receptor backbone.
+Flexible loops near the binding site that undergo induced fit conformational reorganization are not usually restrained in this way to allow for receptor conformational reorganization.
+
+\subsubsection{ATM Implementations}
+
+There are now various ATM implementations~\cite{atom-openmm-github2025,femto-github2025,atm-acellera-github2025}.
+To our knowledge, currently they are all based on the ATMForce feature of the OpenMM molecular simulation package~\cite{eastman2023openmm}.
+ATMForce implements the ATM alchemical potential through two GPU contexts: the first evaluates the reference potential energy $u(\vec{q})$, and the second calculates the perturbed potential energy after the coordinate transformation.
+ATMForce retrieves energy and forces from each context and merges them into the alchemical potential energy function according to Eqs.~(\ref{eq:ATM-alchemical-potential})--(\ref{eq:end-of-softplus}).
+ATMForce is usually set up to consider only the potential energy terms that change upon the coordinate transformation, such as the non-bonded term for standard fixed charge force fields.
+Covalent terms (bond stretches, angle bending, and torsions) are typically unaffected by rigid translations of whole molecules and can be ignored when evaluating perturbation energies.
+
+The AToM-OpenMM Python package~\cite{atom-openmm-github2025,atom-openmm-documentation2025} is considered the reference ATM implementation.
+AToM-OpenMM provides a toolkit to prepare and run alchemical binding free energy calculations by interfacing AToMForce with the user's input.
+It supports systems with fixed charge force fields (Amber, CHARMM, OpenFF, and Espaloma) and implicit solvent models (HCT, OBC, AGBNP) supported by OpenMM on GPU (CUDA, HIP, OpenCL) and CPU platforms.
+It provides enhanced parallel conformational sampling with Asynchronous Hamiltonian replica exchange~\cite{Gallicchio2008,gallicchio2015asynchronous} over any number of local GPUs, as well as support for advanced features such as metadynamics, relative selectivity calculations with ligand swapping, and macromolecular RBFEs with coordinate swapping~\cite{gallicchio2025mutants}.
+
+The femto package\cite{femto-github2025}, developed by Psivant, offers a lightweight and highly automated interface for high-throughput execution and rapid prototyping of ATM ABFE and RBFE workflows.
+Additionally, Acellera's ATM implementation for the Quantumbind package~\cite{quantumbind-acellera-github2025} provides a commercially supported solution with graphical user interfaces and extended documentation, as well as support for neural-network quantum-mechanical potentials~\cite{sabanes2024enhancing}, enabling both novice and expert users to leverage ATM for routine and advanced free energy calculations.
+A CHARMM-GUI ATM module~\cite{zhang2024ligand} provides support for membrane-bound receptors such as GPCRs.
+Support for ATM is also available from the OpenBioSim infrastructure~\cite{openbiosim-diary-ATM2025}.
+These tools collectively enhance the scalability, usability, and reproducibility of ATM-based simulations.
+
+\subsubsection{ATM Validation Studies and Deployments}
+
+ATM has undergone rigorous early validation through the SAMPL (Statistical Assessment of Modeling of Proteins and Ligands) series of community blind challenges, which were designed to objectively assess computational methods for molecular solvation and binding.
+Notably, ATM was benchmarked in the SAMPL6, SAMPL8, and SAMPL9 blind challenges, consistently ranking among the top-performing methods for both absolute and relative binding free energy calculations \cite{wu2021alchemical,azimi2022application,khuttan2023taming}.
+Notably, ATM yielded SAMPL8 ABFE estimates in close agreement with the rigorous non-alchemical Potential of Mean Force implementation in GROMACS~\cite{azimi2022application}.
+Further large-scale validation studies were conducted in collaboration with leading drug discovery technology companies, including Psivant and Acellera, which focused on refining the method for relative binding free energy calculations using FEP benchmark systems~\cite{chen2024performance,sabanes2023validation}.
+
+Recent advancements have expanded the scope and capabilities of ATM.
+ATM has been integrated with metadynamics (metaD) in OpenMM, enabling it to support an arbitrary set of potentials without modifying the core molecular dynamics energy routines.
+For ligands that undergo substantial conformational reorganization upon binding, metadynamics-based sampling of internal degrees of freedom reduces the variance of free energy closure estimates for improving the robustness of RBFE estimates~\cite{khuttan2024make}.
+The AToM-OpenMM framework includes dedicated functionality to load external metadynamics potentials and apply reweighting or unbiasing schemes, enabling statistically rigorous recovery of free energy estimates from biased simulations.
+
+ATM has been evaluated for protein–ligand binding affinity prediction using hybrid neural network potentials and molecular mechanics (NNP/MM), showing improved accuracy over conventional force fields such as GAFF2 \cite{sabanes2024enhancing}.
+These models have been integrated into the ATM framework via the OpenMM-ML plugin and Acellera's implementation, enabling the use of advanced NNPs, such as ANI and AceForce, within standard ATM workflows.
+These implementations rely on PyTorch-based architectures for efficient GPU-accelerated evaluation of neural network potentials during simulation.
+In addition, ATM has been extended to enable receptor hopping and receptor swapping protocols for computing ligand selectivity between receptors.
+In receptor swapping, one ligand is transferred between two receptors while a second ligand moves in the opposite direction \cite{azimi2024selectivity}.
+Finally, the Alchemical Transfer Method with Coordinate Swapping (ATS) has been developed to compute relative binding free energies for large congeneric ligands and single-point mutant peptides within the ATM framework~\cite{gallicchio2025mutants}.
+
+These extensive efforts have firmly established ATM as a reliable and versatile tool, capable of addressing a broad spectrum of modeling modalities.
+Ongoing work by a growing community of developers and practitioners in academia and industry is increasing the robustness and expanding its range of applicability in computer-aided drug discovery and other fields.
 
@@
 
-There is one important thing to note about the uncertainty estimates when summing multiple free energy differnces together to calculate an overall estimate of a free energy difference.
+There is one important thing to note about the uncertainty estimates when summing multiple free energies together to calculate an overall free energy estimate.
 Although BAR itself gives a free energy estimate that is asymptotically correct and is much less biased than the uncertainty estimate for EXP, the uncertainties in $\Delta {f_{i-1\rightarrow i}}$ and $\Delta {f_{i\rightarrow i+1}}$ are not uncorrelated, because they both involve the energies $u_i(\vec{q})$.
-The variances of each of the free energies cannot simply be summed together into the variance of the overall free energy.
+The variances of each of the free energies will \textit{not} propagate as variances usually do (in quadrature) into the variance of the overall free energy.
 Instead, some other method for propagating the uncertainty, such as bootstrapping~\cite{grossfield2018best} must be used.
@@
 
-MBAR can also be thought of as the Zwanzig estimator of the free energy to state $i$ from a sampled distribution referred to as the \textit{mixture distribution}, where one throws all the samples from all simulations together in one ``pot''.
-Mathematically, this is defined by $p_m(\vec{q}) = N^{-1} \sum_k N_k \exp(f_k-u_k\vec{q})$, which is the average, weighted by the number of samples drawn from each distribution, of all the individual normalized probability distributions from the simulations that are performed~\cite{shirts2017reweighting}.
-
-All of the estimators above have uncertainty estimates associated with them.
-In the limit of sufficiently large data, perhaps several hundred or thousand independent points, exhaustive analyses of model data has shown that the error estimates of BAR, MBAR, and TI are all quite reliable when there is good overlap between intermediate states.
-For EXP, the number of samples at which they become reliable is significantly higher, especially if overlap between intermediate states is lower.
-However, in the limit of moderate or small amounts of data (tens or hundreds of independent samples) these errors become somewhat unreliable, often off by a factor of 2 from what would get if more data was collected.
-These estimator of uncertainty are all derived from propagation of error formulae from the estimators themselves.
-However, when the underlying data is noisy, this approximation is not very good.
-
-Some of the authors' experience has shown that significantly more reliable estimators at lower numbers of samples are obtained by performing bootstrap sampling on data that has been uncorrelated.
-We do not provide a full tutorial of bootstrapping here, but the basic idea is that one takes the $N_{sample}$ data points collected and draws, with replacement, $N_{sample}$ samples.
-This means some of the data points will be represented multiple times, and other data points none.
-This process is then repeated $N_{bootstrap}$ times, where $N$ should generally be at least 50 to up to 200 for good estimates of the standard error of the estimator.
-The data is then \textit{treated} as $N_{bootstrap}$ samples from the true distribution, and statistics such as the standard error are computed from this artificial set of $N_{bootstrap}$ experiments.
-In the limit of $N_{sample}$ large, it can be shown that the bootstrap estimate of the standard error converges to the true population standard error.
-One way this can be seen is that bootstrap estimators of the data are independent samples of the \textit{actual} distribution of the experiment performed, i.e., a bunch of $\delta$ functions at $N_{sample}$ points, rather than being independent samples from the true, usually continuous distribution.
-As more and more data is collected, our collection of $\delta$ functions becomes more and more like the true distribution.
-Another advantage of bootstrap sampling is that one can construct decent estimates of error bounds with arbitrary percentiles, and not just standard errors of the estimates, which assume Gaussian behavior of the underlying error, and that the error results are Gaussian as well.
-The bootstrap approach, on the other hand, captures non-Gaussian behavior in the data.
-
-One important point is that the bootstrap samples use to estimate the standard error should be decorrelated, or else the bootstrap estimate of the standard error will be artificially small.
-More discussion of subsampling and decorrelation is included in section~\ref{sec:decorrelating-samples}.
-Another alternative to bootstrap sampling of individual points that better captures the correlations in the data is block bootstrapping~\cite{???}.
-Block bootstrapping is somewhat more complicated and less common to perform and analyze than bootstrap estimates of the error, so we refer readers to the citations for more description, but it can also provide more reliable estimates of uncertainty than the estimators above.
+MBAR can also be thought of as the Zwanzig estimator of the free energy to state $i$ where the sampled distribution is the \textit{mixture distribution} of all the other samples thrown together in one ``pot'', defined by $p_m(\vec{q}) = N^{-1} \sum_k N_k \exp(f_k-u_k\vec{q})$, which is the weighted average of all the individual normalized probability distributions from the simulations that are performed~\cite{shirts2017reweighting}.
 
@@
 
-%MRS: Finlay, you can add information about RED here.
-
 \subsection{Decorrelating samples for analysis}
@@
 \paragraph{Computing the statistical inefficiency}
-Most estimators require an uncorrelated set of samples from the equilibrium distribution to produce (relatively) unbiased estimates of the statistical uncertainty, as most estimates of the uncertainty are derived assuming unbiased samples.
+The readers should note here that replicas within ensemble simulations are independently run and hence generate decorrelated samples by default.
+Therefore, the procedures in this section are only relevant for data from a single simulation.
+Further, Wade \textit{et al.} reported that the underestimation of true uncertainty using one-off simulations remains unresolved even on decorrelating the samples~\cite{wade2022}.
+
+Most estimators require an uncorrelated set of samples from the equilibrium distribution to produce (relatively) unbiased estimates of the free energy difference and its statistical uncertainty.
 To do this, the production region of the simulation is generally \emph{subsampled} with an interval approximately equal to or greater than the \emph{statistical inefficiency} $g \ge 1$ to produce a set of uncorrelated samples that can be fed to the estimator machinery~\cite{chodera2016simple},
@@
 
-We recommend using the robust statistical inefficiency computation routines available within the MBAR~\cite{kylebeauchamp2019choderalab} and alchemlyb~\cite{daviddotson2019alchemistry} packages, though we caution that any such algorithms will underestimate the correlation time with insufficient data.
-Such approaches also assume that the data doesn't have any long time scale drift over the course of the simulation, which is a clear sign of lack of equilibration.
-Generally, most methods will give reasonable correlation times when the simulation lasts 40--50 correlation times,
+We recommend using the robust statistical inefficiency computation routines available within the MBAR~\cite{kylebeauchamp2019choderalab} and alchemlyb~\cite{daviddotson2019alchemistry} packages.
 
@@
 
-However, samples uncorrelated by this procedure are only approximately uncorrelated.
-Recent careful investigation has shown that the lowest uncertainty results can be obtained by using \textit{all} of the data to obtain an estimate of the free energy itself, but using only subsampled data to calculate error estimates using the formulas or by using bootstrapping.
-In fact, one can show that the standard deviation of the estimate determined by data that has been subsampled is the unbiased estimator of the free energy estimate using all the data, but is actually an underestimate of the uncertainty of estimate using just the subsampled data.
-%MRS: Finlay, is there a citation for the mathematical analysis you showed in the issue tracker?  This would be a good place to it.
-
-We don't need to use all of the data to recover the remaining correlation.  In practice, if one subsamples the data with a
-correlation interval of $g/10$, where $g$ is the correlation time, this will recover the missing correlation.
-Thus, if $g=3.0$ in units of timesteps, it is best to use all of the data.
-If $g=100.0$, then one loses essentially no information by subsampling with $g=10.0$ in estimating the free energies, and still saves significant time by not using all of the now-completely duplicative data.
-Note that subsampling with correlation time $g$ should still be used for estimating the uncertainties, perhaps requiring two passes of the simulation analysis code.
-
-If independent simulations at each $\lambda$ are used, the alchemical state $\vec{\lambda}$ may have a significant impact on the correlation time, and these simulations should be subsampled independently using a separate estimate of the statistical inefficiency $g$ for each alchemical state.
+If independent simulations are used, the alchemical state $\vec{\lambda}$ may have a significant impact on the correlation time, and these simulations should be subsampled independently using a separate estimate of the statistical inefficiency $g$ for each alchemical state.
 If coupled simulations are used (such as a Hamiltonian replica exchange simulation), the replicas should undergo equivalent random walks in alchemical space, and the replicas can be subsampled with the same $g$ to generate an equal number of uncorrelated samples at each alchemical state.
@@
 
-Our overall recommendations are therefore to determine the subsampling period $g$, either use all of the data or subsample with something like $g/10$ (i.e.
-more frequently than specified by $g$) to compute the free energy estimates, and then use fully subsampled data (with subsampling interval $g$) to estimate the uncertainties of the estimates.
-Uncertainties calculated with the full data set will be significant underestimates if $g > 1$.
-
 \paragraph{Cautions and considerations}
 Reliable estimation of the statistical inefficiency is difficult, and estimates will not generally be as precise (in a relative error sense) as averages.
-To ensure there is sufficient data available for reliable decorrelation and estimation of free energy differences, it is recommended that the effective number of uncorrelated samples $N_{\mathrm{eff}} \ge  \sim 50$ if the BAR or MBAR estimators (discussed below in sec~\ref{subsec:estimators}) are used; the number may need to be much higher with alternate estimators.
+To ensure there is sufficient data available for reliable decorrelation and estimation of free energy differences, it is recommended that the effective number of uncorrelated samples $N_{\mathrm{eff}} \ge \approx 50$ if the BAR or MBAR estimators (discussed below in sec~\ref{subsec:estimators}) are used; the number may need to be much higher with alternate estimators.
 
@@
 In general, the quantification of different error metrics depends on both data generation and analysis methods used from the ones discussed above.
-
-Computating of free energies using TI (Sec.n~\ref{subsec:estimators}) is straightforward and the trapezoidal rule is often recommended.
-The trapezoid rule easily allows unequal spacing of $\vec{\lambda}$ states, which is often required to minimize the variance in the free energy estimate, but in principle any good numerical integration method can be used if care is taken.
-Determining of regions of high curvature when estimating the integral is helpful to determine regions of phase space where more sampling and/or more $\vec{\lambda}$ states are necessary to obtain the best approximation of the integral.
-Plotting $\vec{\lambda}$ with respect to the gradients at each of the $\vec{\lambda}$ values, which can be done automatically with tool such as alchemlyb,  can be a helpful diagnostic.
-Additionally, computating the overall variance of TI requires the calculation of the overall variance of the integration, rather than each individual $\Delta$G$_{i,i+1}$ and assuming variances add independently.
-Therefore, $\mathrm{var}$($\Delta$f) = $\sum_{i=1}^{K}w_{k}^2 \mathrm{var}(\frac{du}{d\vec{\lambda}})_{k}$, where the $w_k$ are the integration weights (see Sec~\eref{subsec:estimators} for more discussion of $w_k$).
+The user may decide to choose best practices based on their need for the calculations carried out.
+If only few free energies need to be calculated at very high accuracy a different strategy might be employed than if large scale estimates are needed.
+As such, considering overall computational cost or availability of resources with respect to accuracy desired from computations may influence the choice of sampling scheme.
+
+%Statistical mechanics provides a route to calculate macroscopic thermodynamic properties of a system from \textit{ensemble averaging} over microstates.
+Hitherto, MD practitioners have been applying the ergodic theorem that allows one to equate ensemble averaging to time averaging in the limit of time approaching infinity to predict thermodynamic properties using one-off simulations.
+However, the problem with this approach is that the simulation time required to satisfy this condition has been conveniently dictated by the extent of simulation time deemed accessible in a reasonable period of wall time on available resources.
+It is ignored that this limit is on the order of a Poincarè recurrence time, which is truly astronomical and well beyond the reach of today's supercomputers.
+In more rigorous terms, ergodic theory necessitates that, if a dynamical system is to approach equilibrium, it is not sufficient for it to be "ergodic" but it must also be ``mixing''~\cite{coveney2016, MDbook2025}.
+Rate processes and equilibration times are a property of the distribution function, not of individual trajectories.
+Consequently, neighbouring trajectories in the underlying phase space diverge exponentially which means that a single trajectory can never capture the true behaviour of a system.
+Rather, one must perform true ensemble averaging through running sufficiently large number of replicas, termed ``ensemble simulations''.
+This is manifest in the numerous published reports where authors have highlighted the non-reproducibility of results obtained from one-off simulations and that single simulations heavily underestimate the true uncertainty in predictions~\cite{wan2020fep+, wan2021uq, vassaux2021, gapsys2020large, wade2022, bhati2017, bhati2022, bhati2025, wan2023eqvsneq, bhati2018, bhati2019}.
+The uncertainty must be calculated as the standard error of mean for samples obtained from each replica in the ensemble.
+The ensemble size is a parameter that needs to be determined empirically for the system of interest and may need to be adjusted; however, some general recommendations (for both EQ and NEQ approaches) based on systematic analyses on a large and diverse dataset of protein-ligand complexes are available in a series of papers published by the Coveney group~\cite{bhati2017, bhati2018, bhati2019, bhati2022, wan2023eqvsneq, bhati2025}.
+Further, ensemble simulations are also necessary in case of enhanced sampling methods~\cite{bhati2018} as well as microsecond long simulations~\cite{bhati2023}.
+It is the influence of these reports that journals such as \textit{JCIM} and \textit{Communications Biology} have now mandated the use of ensemble simulations in computational studies~\cite{wan2023}.
+Coveney and Wan have recently published a book entitled \textit{Molecular Dynamics: Probability and Uncertainty}~\cite{MDbook2025} on this very topic.
+It must also be noted that ``replicas'' within an ensemble simulation are conceptually all members of a continuous distribution which are run concurrently, not in a serial mode.
+Therefore, the use of term ``repeats'' in this context is misleading and must be avoided.
+
+The computation of free energies using TI (Sec.n~\ref{subsec:estimators}) is straightforward and the trapezoidal rule is often recommended since it allows unequal spacing of $\vec{\lambda}$ states, which is required to minimize the variance in the free energy estimate, but in principle any good numerical integration method can be used.
+The determination of regions of high curvature when estimating the integral is helpful to determine regions of phase space where more sampling and/or more $\vec{\lambda}$ states are necessary to obtain the best approximation of the integral.
+Plotting $\vec{\lambda}$ with respect to the gradients at each of the $\vec{\lambda}$ values can be a helpful diagnostic.
+Additionally, computation of the overall variance of TI requires the calculation of the overall variance of integration, rather than each individual $\Delta$G$_{i,i+1}$ and assuming variances add independently.
+Therefore, $\mathrm{var}$($\Delta$f) = $\sum_{i=1}^{K}w_{k}^2 \mathrm{var}(\frac{du}{d\vec{\lambda}})_{k}$.
+
+The TIES method~\cite{bhati2017} employs a different scheme to obtain the overall variance of the integral.
+It assumes that the integrand is a Gaussian random process and hence the principle of stochastic integration must be employed to propagate the uncertainties in this integral.
+The following equation is used to compute the overall variance of the integral combining variance at each $\lambda$ state.
+$\mathrm{var}$($\Delta$f) = $\sum_{i=1}^{K}\Delta\lambda_{k}^2 \mathrm{var}(\frac{du}{d\vec{\lambda}})_{k}$.
 
@@
 The number of states and amount of sampling should be optimized to minimize the uncertainty in the MBAR free energy estimate, while balancing other key considerations such as computational expense. 
+However, this process based on one-off simulations grossly underestimates the true uncertainty and hence uncertainties for MBAR also must be obtained using ensemble simulations~\cite{bhati2018, wade2022}.
 
@@
 If different estimators agree the free energy estimate is more reliable than if there are differences between methods that are larger than 1 kcal/mol and would indicate poor convergence.
-
-Uncertainty can also be assessed for a particular perturbation by repeating calculations with slight changes in initial configurations, forcefield parameters, and different random seeds in the MD engine.
-The assessment of variability in free energy calculations due to repeating simulations has been previously reported~\cite{aldeghi2019accurate,paliwal2011benchmark,mey2016blinded,mey2018impact}, and large variance in free energies estimated from simulations with different random seeds should be flagged as issues with convergence.
 
@@
 Other factors, like system preparation (choice of protonation state, tautomer, counterion presence, salt concentration, etc.) can also substantially impact results~\cite{mobley2017predicting, mobley2017predictinga}, so unless modelers are confident they have these factors correct, sensitivity to these choices may also need to be examined.
+Incidentally, Vassaux \textit{et al.} performed a systematic study comparing the effect of both aleatoric as well as parametric uncertainties for protein-ligand free energy calculations using NAMD~\cite{vassaux2021}.
+They found that, indeed, ensemble simulations are necessary to control both sources of uncertainties arising in simulations, just as is anticipated in the field of uncertainty quantification.
 
@@
 Accordingly, the uncertainty of the free energy difference between the states $i$ and $j$ will be smaller when ${O}_{i,j,i\ne j}$ is larger (and thus the values in the main diagonal (${O}_{i,j,i=j}$) are smaller).
+Ensemble simulations generate wider distributions of energies and hence are expected to increase the overlap in neighbouring states~\cite{wade2022}.
 In order to obtain a reliable estimate of the free energy all neighbouring states must be connected, i.e.
@@
 
+ The precision of NEQ calculations is determined by the overlap between the forward and reverse work distributions corresponding to out-of-equilibrium transitions between the two end-points.
+Bhati \textit{et al.} recently studied the effect of such overlaps on the accuracy of predicted RBFEs~\cite{wan2023eqvsneq} as well as ABFEs~\cite{bhati2025} but found that overlap (or lack of it) does not capture the reliability of calculations very well.
+They recommended an alternative metric, ``distance'' defined as the difference between the inner extremes of the two work value distributions, which correlates quite well with the unsigned errors in free energy predictions and hence was found a better indicator of the reliability of NEQ calculations.
+
 \paragraph{Reversible binding simulations}
@@
 For small ligands with millimolar affinities, repeated binding to and unbinding from the protein can occur for a large number of times in a sufficiently long unbiased MD simulation (10-100 $\mu$s), and the equilibrium binding constants can be computed from the ratio of bound to unbound fractions of the simulation time.
+It is worth highlighting here that the necessity of ensemble simulations to capture the true uncertainty in predictions ensuring their reliability is equally applicable for long simulations.
+Bhati \textit{et al.} performed ensembles of 10 microsecond long simulations and reported large variability in several properties calculated from each replica~\cite{bhati2023}.
 The agreement between the binding free energy calculations and the reversible binding simulations---given the same system preparation and the same force field parameters---will strongly support the correctness of both calculations, as the same results are arrived at by two independent methods, and any discrepancy will suggest some systematic error in one, or both, of the two methods.
@@
 All of these estimators converge to the same results with sufficient sampling.
+Reports have shown that the use of ensemble simulations can ensure good agreement between different estimators as well as MD engines~\cite{wade2022}.
 Differences between them indicate poor overlap or errors in processing.
@@
 \section{\tocred{Unsettled Practices and Current Limitations}}
-\tocsectioncomment{@Everyone to contribute to section on unsettled practices/ current limitiations, as suggested by John Chodera.}
+\tocsectioncomment{@Everyone to contribute to section on unsettled practices/ current limitations, as suggested by John Chodera.}
 \label{sec:unsettled}
@@
 For example, there is growing interest in identifying small molecules that directly bind to RNA or DNA, with recent preliminary studies showing promising results using specialized simulation protocols.\cite{Abramyan2025,Rasouli2024} In contrast, there has been limited investigation into the applicability of AFE methods to small-molecule ligands targeting intrinsically disordered proteins (IDPs,\cite{Papadourakis2024} or proteins containing intrinsically disordered regions (IDRs) that undergo disorder-to-order transitions upon ligand binding.\cite{MendozaMartinez2022}
-The treatment of covalent binders has also received some attention.\cite{Luo2021,Zhang2019 } However, robust protocols for handling datasets of irreversible covalent binders containing diverse ‘warheads’ have not yet achieved widespread adoption.\cite{Yu2019}
+The treatment of covalent binders has also received some attention.\cite{Luo2021,Zhang2019 } However, robust protocols for handling datasets of irreversible covalent binders containing diverse ‘warheads’ have not yet achieved widespread adoption.\cite{Yu2019} A variety of other applications are less traditional and may be possible but raise new practical issues; for example, if multiple copies of a ligand bind at once in the same site, this could pose challenges, as could handling of binding and function prediction for molecular glues and related categories of molecules.
+
+Relative free energy calculations carry with them additional unsettled practices relating to the use and careful treatment of dummy atoms (see ``Dummy Atoms'' in Section~\label{sec:relative-fe-protocol}).
+In particular, as the work of~\cite{Fleck_2021} highlighted, valence terms for dummy atoms for dummy atoms must be treated far more carefully than previously appreciated in order for dummy atoms' contributions to the free energy to rigorously cancel.
+Less careful treatments may leave dummy atom terms only approximately canceling, or, worse, introducing uncharacterized errors in relative binding free energies.
+Indeed, such issues might have impeded previous attempts at reproducing free energy calculations across simulation engines~\cite{loeffler2018reproducibility}.
+
+While solutions to dummy atom challenges may in principle be known, most are not yet implemented in setup tools and free energy pipelines and have not yet been thoroughly characterized, so it seems likely that considerable science and engineering remains to be done in this space.
 
@@
 Lambda-dynamics-based free energy calculation can be carried out using the BLOCK module.
-	\item \href{https://ambermd.org/}{AMBER}, including its new pmemd.cuda version supports free energy calculations~\cite{salomon-ferrer2013overview}. 
+	\item \href{https://ambermd.org/}{AMBER}, including its GPU-accelerated free energy engine~\cite{Lee_JChemTheoryComput_2017_v13_p3077,Giese_JChemTheoryComput_2018_v14_p1564,Lee_JChemInfModel_2018_v58_p2043} that supports a wide range of alchemical free energy methods~\cite{Tsai_JChemTheoryComput_2023_v19_p640,Lee_2023,Zhang_JChemTheoryComput_2024_v20_p3935}, workflows~\cite{Ganguly_JChemInfModel_2022_v62_p6069} and analysis tools~\cite{Giese_JChemInfModel_2025_v65_p5273,Case_JChemInfModel_2023_v63_p6183}.
 	\item \href{http://www.gromos.net/}{GROMOS} offers an extensive and flexible molecular dynamics and simulations analysis suites with free energy calculation functionalities including customizable alchemical paths and various sampling protocols~\cite{schmid2012architecture, kunz2012new, eichenberger2011gromos}.
@@
 	\item \href{https://github.com/qusers/Q6}{Q} is MD code for performing FEP calculations using a variety of force fields~\cite{aaquist2017q6}. 
+    \item \href{http://www.ties-service.org}{TIES} enables the usage of NAMD and OpenMM for setting up alchemical free energy ensemble simulations \url{https://github.com/UCL-CCS/TIES_MD} \cite{bieniek2023}.
 	\end{itemize}
@@
 	\item \href{https://fesetup.readthedocs.io/en/latest/introduction.html}{FESetup} has been developed primarily to setup calculations in AMBER, GROMACS and SIRE~\cite{loeffler2015fesetup}.
+    \item \href{http://www.ties-service.org}{TIES} allows the setup of perturbation maps, hybrid topologies and structures as well as simulation-ready models for NAMD and OpenMM \url{https://github.com/UCL-CCS/TIES_MD} \cite{bieniek2023}.
 	\end{itemize}
@@
 	\item Free Energy Workflows: Sire-specific free energy map analysis using weighted path averages \url{https://github.com/michellab/freenrgworkflows}.
+    \item \href{http://www.ties-service.org}{TIES}: Perform statistical analysis of the output from alchemical ensemble simulations and report reliable predictions along with statistical uncertainties for a variety of free energy estimators \url{https://github.com/UCL-CCS/TIES_MD} \cite{bieniek2023}.
 	\end{itemize}
@@
         \item Set up your restraints correctly
-        \item Subsample the data to estimate the variance of your free energy estimation protocol (but not to estimate the values themselves)
+        \item Make sure you subsample the data in your free energy estimation protocol
         \item Apply the appropriate correction terms
@@
         \item Check whether dummy atoms were assigned correctly
-        \item Subsample the data to estimate the variance of your free energy estimation protocol (but not to estimate the values themselves)
-         \item Apply the appropriate correction terms
+        \item Consider subsampling the data in your free energy estimation protocol
+        \item Apply the 
+        appropriate correction terms
 \end{itemize}
@@
     \item Include error bars in all your visual analyses
-    \item Consider bootstrapping uncertainties rather than using free energy error estimators
 \end{itemize}
@@
 Marieke Schor proofread the manuscript. 
+We thank Sara Tkaczyk for creating Figure 3 (based on the earlier version by Jenke Scheen).
 
@@
 \makeorcid
-\bibliography{alchemical,manual,sb}
+\bibliography{alchemical,manual,sb,atm,uq}
 

### 2025-07-18 paper/manuscript.tex

--- 2025-07-18 paper/manuscript.tex
@@
 \author[1]{Jenke Scheen}
-\author[11]{David R. Slochower}
+\author[14]{David R. Slochower}
 \author[8*]{Michael R. Shirts}
@@
 \affil[10]{Cresset, Cambridgeshire, UK}
-\affil[11]{Vertex Pharmaceuticals, San Diego, CA, USA}
+\affil[11]{Centre for Computational Sciences, Department of Chemistry, University College London, London WC1H 0AJ, United Kingdom}
+\affil[12]{Advanced Research Computing Centre, University College London, London WC1H 9BT, United Kingdom}
+\affil[13]{Computational Science Laboratory, Institute for Informatics, Faculty of Science, University of Amsterdam, Amsterdam 1012, The Netherlands}
+\affil[14]{Vertex Pharmaceuticals, San Diego, CA, USA}
 
@@
 In particular, if the rate of change in the free energy estimate is significant when this condition is met, the simulation may not be locally converged, and more sampling may be necessary to determine a stable free energy estimate which is no longer changing significantly over time.
+A few studies reported systematic variation of simulation lengths using ensemble simulations over an extended range to provide recommendations on appropriate values applicable for most protein-ligand systems.
+The authors suggest using 4 ns and 10 ns for RBFE~\cite{bhati2017, bhati2022, wan2023eqvsneq} and ABFE~\cite{bhati2025} calculations respectively.
+These values may be used as a general guideline.
 
@@
 \paragraph{Computing the statistical inefficiency}
+The readers should note here that replicas within ensemble simulations are independently run and hence generate decorrelated samples by default.
+Therefore, the procedures in this section are only relevant for data from a single simulation.
+Further, Wade \textit{et al.} reported that the underestimation of true uncertainty using one-off simulations remains unresolved even on decorrelating the samples~\cite{wade2022}.
+
 Most estimators require an uncorrelated set of samples from the equilibrium distribution to produce (relatively) unbiased estimates of the statistical uncertainty, as most estimates of the uncertainty are derived assuming unbiased samples.
@@
 In general, the quantification of different error metrics depends on both data generation and analysis methods used from the ones discussed above.
+The user may decide to choose best practices based on their need for the calculations carried out.
+If only few free energies need to be calculated at very high accuracy a different strategy might be employed than if large scale estimates are needed.
+As such, considering overall computational cost or availability of resources with respect to accuracy desired from computations may influence the choice of sampling scheme.
+
+%Statistical mechanics provides a route to calculate macroscopic thermodynamic properties of a system from \textit{ensemble averaging} over microstates.
+Hitherto, MD practitioners have been applying the ergodic theorem that allows one to equate ensemble averaging to time averaging in the limit of time approaching infinity to predict thermodynamic properties using one-off simulations.
+However, the problem with this approach is that the simulation time required to satisfy this condition has been conveniently dictated by the extent of simulation time deemed accessible in a reasonable period of wall time on available resources.
+It is ignored that this limit is on the order of a Poincarè recurrence time, which is truly astronomical and well beyond the reach of today's supercomputers.
+In more rigorous terms, ergodic theory necessitates that, if a dynamical system is to approach equilibrium, it is not sufficient for it to be "ergodic" but it must also be ``mixing''~\cite{coveney2016, MDbook2025}.
+Rate processes and equilibration times are a property of the distribution function, not of individual trajectories.
+Consequently, neighbouring trajectories in the underlying phase space diverge exponentially which means that a single trajectory can never capture the true behaviour of a system.
+Rather, one must perform true ensemble averaging through running sufficiently large number of replicas, termed ``ensemble simulations''.
+This is manifest in the numerous published reports where authors have highlighted the non-reproducibility of results obtained from one-off simulations and that single simulations heavily underestimate the true uncertainty in predictions~\cite{wan2020fep+, wan2021uq, vassaux2021, gapsys2020large, wade2022, bhati2017, bhati2022, bhati2025, wan2023eqvsneq, bhati2018, bhati2019}.
+The uncertainty must be calculated as the standard error of mean for samples obtained from each replica in the ensemble.
+The ensemble size is a parameter that needs to be determined empirically for the system of interest and may need to be adjusted; however, some general recommendations (for both EQ and NEQ approaches) based on systematic analyses on a large and diverse dataset of protein-ligand complexes are available in a series of papers published by the Coveney group~\cite{bhati2017, bhati2018, bhati2019, bhati2022, wan2023eqvsneq, bhati2025}.
+Further, ensemble simulations are also necessary in case of enhanced sampling methods~\cite{bhati2018} as well as microsecond long simulations~\cite{bhati2023}.
+It is the influence of these reports that journals such as \textit{JCIM} and \textit{Communications Biology} have now mandated the use of ensemble simulations in computational studies~\cite{wan2023}.
+Coveney and Wan have recently published a book entitled \textit{Molecular Dynamics: Probability and Uncertainty}~\cite{MDbook2025} on this very topic.
+It must also be noted that ``replicas'' within an ensemble simulation are conceptually all members of a continuous distribution which are run concurrently, not in a serial mode.
+Therefore, the use of term ``repeats'' in this context is misleading and must be avoided.
 
@@
 
+The TIES method~\cite{bhati2017} employs a different scheme to obtain the overall variance of the integral.
+It assumes that the integrand is a Gaussian random process and hence the principle of stochastic integration must be employed to propagate the uncertainties in this integral.
+The following equation is used to compute the overall variance of the integral combining variance at each $\lambda$ state.
+$\mathrm{var}$($\Delta$f) = $\sum_{i=1}^{K}\Delta\lambda_{k}^2 \mathrm{var}(\frac{du}{d\vec{\lambda}})_{k}$.
+
 For alchemical changes that result in smooth, low curvature sets of $\expect{\frac{dU}{d\vec{\lambda}}}$, a relatively small number of $\vec{\lambda}$ states is necessary for sufficient accuracy and low variance in the free energy estimate.

### 2025-08-05 paper/manuscript.tex
Includes changes from @hannahbruce discussed in https://github.com/alchemistry/alchemical-best-practices/pull/162 which partially address https://github.com/alchemistry/alchemical-best-practices/issues/141.
Other tweaks also included.

@@
 \item non-equilibrium free energy calculations~\cite{gapsys2020large}
-\item Free energy calculations using QM/MM methods~\cite{beierlein2011simple,dybeck2016comparison,cave-ayland2015direct}.
-\item Free energy calculations using machine learning methods~\cite{rufa2020chemical, scheen2020hybrid, cole2020machine}
+\item Free energy calculations using QM/MM methods~\cite{beierlein2011simple,dybeck2016comparison,cave-ayland2015direct,Karwounopoulos2024}.
+\item Free energy calculations using machine learning methods~\cite{rufa2020chemical, scheen2020hybrid, cole2020machine,Schoeller2023}
 \end{itemize}
@@
 \paragraph{Have you considered stereoisomers and enantiomers?}
-Congeneric series can contain stereoisomers or enantiomers which can bind very differently, resulting in large errors if treated incorrectly.
-For racemates, the relative abundance of each stereoisomer is normally not known.
+Congeneric series can contain stereoisomers or enantiomers, which can bind very differently, resulting in large errors if treated incorrectly.
+For racemates, the relative abundance of each stereoisomer is normally not known, however when there is only one stereocenter in the molecule it can often be fair to assume that the substance is a 50\%:50\% mixture of the two.
+If one enantiomer is active (eutomer) and the other is inactive (distomer), this corresponds to the effective concentration of your bioactive molecule being half of what is expected in the assay, which corresponds to a ~0.2 pIC50 loss in potency relative to the pure eutomer being tested, which could be captured within the associated error calculations.\cite{h2011significance} If possible, both enantiomers could be considered computationally, however if a series contains multiple stereocenters, or multiple compounds within the series have stereocenters, this may be computationally restrictive.
+If compounds contain multiple undefined stereoecenters, the error and complication of understanding predictions may become prohibitive.
 Therefore, the experimental activity associated with just one stereoisomer/enantiomer is more uncertain.
-However, the modeling typically uses just the bioactive conformation that best fits the active site.
-Clearly this introduces potential for larger errors compared to experiment.
-Nevertheless, if all compounds in the congeneric series are racemic, originating from similar synthetic procedures with an expected similar abundance of stereoisomers, then the differences may cancel and the trend in calculated and observed binding energies may be robust.
+If enumerating unknown stereocenters is not tractable, modeling using just the conformation that best fits the active site is reasonable, however this clearly introduces potential for larger errors compared to experiment.
 Despite this, we can see that care and further testing is needed in this scenario, and the quality of the predictions may suffer.
 Additionally, unexpected changes in what stereoisomer binds experimentally, if they occur, could pose significant challenges for modelling efforts.
+
+\paragraph{Are there cofactors expected to be present?}
+Crystallographic structures can contain cofactors, which may be close to the ligand series of interest, and understanding if they are required or not for simulation can be an interdisciplinary effort between the modeller, crystallographer and bioligists to establish if they are a crystallographic artifact, or if they are likely to be present in the assay.
+In addition to both the crystallographic and assay conditions, what is most relevant \textit{in vivo} might be a third, equally difficult consideration.
+Following this, if a cofactor is considered to be included in the calculations, it is worth considering the appropriate forcefield for the cofactors.
+Cofactors, in particular metal ions can be challenging\cite{mobley2017predicting, guven2024protocols}.
+If the congeneric series of ligands is expected to interact with the cofactors differently, such as metal ions in different charge states, or to displace the cofactor for some of the molecules then this is likely beyond current modelling techniques.
+If there is still uncertainty around cofactor binding, running free energy calculations for a set of molecules both with and without the cofactor and comparing the relative rank-ordering of molecules could provide a sanity check.
 
@@
 \label{subsec:hydration}
-Preliminary considerations necessary for using free energy methods to compute partition coefficients are generally more straight forward.
+Preliminary considerations necessary for using free energy methods to compute partition coefficients are generally more straightforward.
 For example, a 3D minimised structure of a solute can be generated with a simple tool such as openBabel and solvated to prepare the input to compute a free energy of hydration~\cite{oboyle2011open}.
-However, in these cases a careful choice of forcefield for the organic solvent model, as well as water model is essential.
+However, in these cases, a careful choice of forcefield for the organic solvent model, as well as water model is essential.
 See for example~\cite{bosisio2016blinded,rustenburg2016measuring} for a good discussion of these choices.
@@
 In single topology calculations, the overall transformation is set up to involve as few additional atoms as possible, so benzene would be typically changed into benzyl alcohol by first changing one of the hydrogens into a carbon.
-The site of this transformation will also be the future home of two additional hydrogen atoms bound to the new carbon, so these must initially be present as non-interacting atoms called ``dummy atoms'' (labelled Du in Fig.~\ref{fig:fig_topology}), which retain their bonded interactions but do not interact with the rest of the system.
-Bond parameters as well as partial charges between the changing atoms are adjusted accordingly between the initial and final state; see below for further details.
+The site of this transformation will also be the future home of two additional hydrogen atoms and a hydroxyl group bound to the new carbon, so these must initially be present as non-interacting atoms called ``dummy atoms'' (labelled Du in Fig.~\ref{fig:fig_topology}), which retain their bonded interactions but do not interact with the rest of the system.
+Bond parameters as well as partial charges between the changing atoms are adjusted accordingly between the initial and final states; see below for further details.
 In a single topology calculation, atoms may change their type, ensuring minimal dummy atoms are created.
@@
 The so-called ``separated topologies'' (SepTop) method \cite{rocklin2013separated,Baumann_2023} and the ``alchemical transfer method'' (ATM) \cite{Azimi_2022} belong into this category.
-In SepTop, the two copies are held on top of each other via restraints with respect to the protein/receptor they bind to (as opposed to Restraintmaker, where the molecules are restrained to each other.
+In SepTop, the two copies are held on top of each other via restraints with respect to the protein/receptor they bind to (as opposed to Restraintmaker, where the molecules are restrained to each other).
 %In ATM, a restraint is used to keep the bound ligand 1 and the free ligand 2 always some minimum distance apart; during exchange moves, suitable rigid-body rotations/translations are applied to move the free ligand into a position/orientation as similar as possible to the one of the bound ligand.
@@
 For $\lambda=0$, $K(\lambda)$ and $r_0(\lambda)$ correspond to the values appropriate for a C-H bond, and for $\lambda=1$ to those for a C-C bond.
-The other energy terms are treated analogously.
+Other force field parameters are treated analogously.
 The single topology paradigm as just summarized was the default mode of alchemical transformations in AMBER and GROMOS at that time.
@@
 \cite{Giese_2018} refer to this approach as energy mixing (interpolation).%; it requires two separate calls to the energy/force routines, but since only a fraction of interactions is affected by the alchemical changes, various optimizations are possible to keep the computational overhead manageable.
-An early example of energy mixing for SCST is the PERT free energy module of CHARMM (written by Bernie Brooks around 1990 and briefly described in Ref.~\cite{brooks_2009}.
+An early example of energy mixing for SCST is the PERT free energy module of CHARMM (written by Bernie Brooks around 1990 and briefly described in Ref.~\cite{brooks_2009}).
 
@@
 \& 2), however a dual coordinate set approach could use an atom mapping as well, depending on the sampling method, this will be explained in the next steps.
-Once a single coordinate set approach is selected is selected, a crucial next step is to identify the common atoms which will not be perturbed between the two molecules (respectivly end-states).
+Once a single coordinate set approach is selected is selected, a crucial next step is to identify the common atoms which will not be perturbed between the two molecules (respectively end-states).
 Rigorously, this process typically comprises a MCSS search of the molecules involved to identify the common substructure.
@@
 This however requires either careful implementation of atom mass changes, bond length etc.
-or accepting the introduction of potential systematic error, that is in many cases negliable due to the cancelation of errors.
+or accepting the introduction of potential systematic error, that is in many cases negligible due to the cancellation of errors.
 \cite{ries2024Kartograf}
@@
 
-In general there are mutliple ways on how a MCSS can be calculated.
+In general, there are many ways on how an MCSS can be calculated.
 One of the most prominent approaches is to estimate the MCSS with a 2D subgraph isomorphism solver(SIS) using the 2D topological stuructre information of molecules\cite{raymond2002maximum}.
@@
 
-SIS based methods, showed to struggle with 3D properties of the molecules, that need to be mapped, like for example stereochemistry.
-Therefore some packages developed a post-mapping step, where they filtered for atom distances, and therefore added 3D information into the mapping.
+SIS based methods were shown to struggle with 3D properties of the molecules that need to be mapped, like for example stereochemistry.
+Therefore, some packages developed a post-mapping step, where they filtered for atom distances, and, therefore, added 3D information into the mapping.
 Still SIS MCSS searches can be relatively time consuming as the algorithmic complexity is classified as NP-complete, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial similarity assessment and then use MCSS only to identify final mappings for relative calculations~\cite{raymond2002maximum,klabunde2012mars,jones2009elucidating, Duesbury2018Comparison}.
@@
 A disadvantage of these approaches is the strong dependency on the coordinate sets being well aligned.
-Another aspect is taking the topology changes of the molecules not into account, which can be done if necessary in a next step with a filtering approach, like checking for changes in ring hybridization or ring breaking (this problem is discusse below)|.
+Another aspect is taking the topology changes of the molecules not into account, which can be done if necessary in a next step with a filtering approach, like checking for changes in ring hybridization or ring breaking (this problem is discussed below)|.
 Tools that can be used to estimate the MCSS this way are contained for example in pmx \cite{gapsys2015pmx} or Kartograf \cite{ries2024kartograf}.
 
-In any case Visual inspection prior to simulation is recommended to ensure that the mapping criteria correspond to the expected binding mode.
+In any case, visual inspection prior to simulation is recommended to ensure that the mapping criteria correspond to the expected binding mode.
 If the mapping protocol returns simulations that correspond to different binding modes of a ligand within a perturbation map, this can cause large hysteresis.
@@
 However, as already described in the early studies and repeated in a more recent study focusing on the correct treatment of dummy atoms \cite{Fleck_2021}, the desired factorization can only be guaranteed if any dummy atom connected to the physical molecule is held by exactly three (non-redundant) degrees of freedom.
-Naive application of the three-degree-of-freedom per dummy atom rule (for dummy atoms having at least one joint bonded term with the physical molecule), however, can result in simulations where the dummy part adopts unphysical configurations/conformations; Fleck et al.
-referred to this as “flapping”.
+Naive application of the three-degree-of-freedom per dummy atom rule (for dummy atoms having at least one joint bonded term with the physical molecule), however, can result in simulations where the dummy part adopts unphysical configurations/conformations; Fleck et al.\ referred to this as “flapping”.
 Thus, depending on the details of the physical molecule -- dummy atom(s) junction, the three degree of freedoms need to be selected with care, or a carefully chosen additional redundant degree of freedom may be needed.
@@
 
-\paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
+\paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms cancelling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
 Some approaches have attempted to address this using a single coordinate sets and with soft bond approaches ~\cite{clark2019relative,wang2017accurate} or dual coordinate set approach \cite{jespers2019qligfep, ries2022relative} but a general solution is not yet in mainstream use.
@@
 More recently, several approaches have been proposed to simultaneously bias both configurational CV and alchemical variables, which can effectively address both scenarios Fig.~\ref{fig:configurational_sampling}\textbf{B} and \textbf{C}.
-Examples include alchemical metadynamics~\cite{hsu2023alchemical}, $\lambda$-ABF (adaptive biasing force)~\cite{lagardere2024lambda}, and optimized alchemical enhanced sampling (ACES)~\cite{Lee_JChemTheoryComput_2023_v19_p472}  and approaches that integrate well-tempered metadynamics~\cite{barducci2008well} with alchemical methods~\cite{zhou2025zooming, khuttan2024make}, to name a few.
+Examples include alchemical metadynamics~\cite{hsu2023alchemical}, $\lambda$-ABF (adaptive biasing force)~\cite{lagardere2024lambda}, and optimized alchemical enhanced sampling (ACES)~\cite{lee2023aces}  and approaches that integrate well-tempered metadynamics~\cite{barducci2008well} with alchemical methods~\cite{zhou2025zooming, khuttan2024make}, to name a few.
 
@@
 In relative free energy calculations, large cycle closure errors are commonly used as indicators of insufficient sampling, although small errors do not necessarily imply that sampling is adequate~\cite{suruzhon2021sensitivity}.
+Measurement of the sensitivity of individual edges in the thermodynamic graph to the enforcement of cycle closure constraints can help to identify problematic alchemical transformations~\cite{giese2025transferability}.
 In other types of calculations, diagnosing sampling deficiencies typically requires prior knowledge of relevant metastable states, as well as collective variables that distinguish them.
@@
 In general, the quantification of different error metrics depends on both data generation and analysis methods used from the ones discussed above.
-The user may decide to choose best practices based on their need for the calculations carried out.
+The user may choose best practices based on their need for the calculations carried out.
 If only few free energies need to be calculated at very high accuracy a different strategy might be employed than if large scale estimates are needed.
-As such, considering overall computational cost or availability of resources with respect to accuracy desired from computations may influence the choice of sampling scheme.
+As such, considering overall computational cost or availability of resources with respect to accuracy desired from computations may influence the choice of sampling and analysis scheme.
 
@@
 While solutions to dummy atom challenges may in principle be known, most are not yet implemented in setup tools and free energy pipelines and have not yet been thoroughly characterized, so it seems likely that considerable science and engineering remains to be done in this space.
+
+More broadly, although several enhanced sampling strategies have been proposed to address configurational sampling bottlenecks in alchemical simulations, widely accepted best practices for detecting and resolving such challenges in a systematic and automated fashion remain elusive.
 

### 2025-08-05 paper/manuscript.tex
Remove ToDo notes and section colours

@@
 \usepackage{todonotes}
-
-% Define custom commands for colored TOC entries
-\newcommand{\tocgreen}[1]{\textcolor{green}{#1}}
-\newcommand{\tocorange}[1]{\textcolor{orange}{#1}}
-\newcommand{\tocred}[1]{\textcolor{red}{#1}}
-
-% Define custom commands for TOC entries with comments
-\makeatletter
-\newcommand{\tocsectioncomment}[1]{%
-  \addtocontents{toc}{%
-    {\leftskip \cftsecindent\relax
-     \advance\leftskip \cftsecnumwidth\relax
-     \rightskip \@tocrmarg\relax
-     \textit{#1}\protect\par}}%
-  \textit{#1}\par}
-
-% Command for subsection comments in TOC
-\newcommand{\tocsubsectioncomment}[1]{%
-  \addtocontents{toc}{%
-    {\leftskip \cftsubsecindent\relax
-     \advance\leftskip \cftsubsecnumwidth\relax
-     \rightskip \@tocrmarg\relax
-     \textit{#1}\protect\par}}%
-  \textit{#1}\par}
-
-% Command for subsubsection comments in TOC
-\newcommand{\tocsubsubsectioncomment}[1]{%
-  \addtocontents{toc}{%
-    {\leftskip \cftsubsecindent\relax
-     \advance\leftskip \cftsubsubsecnumwidth\relax
-     \rightskip \@tocrmarg\relax
-     \textit{#1}\protect\par}}%
-  \textit{#1}\par}
 
@@
 
-\todo[inline, color=green!20]{
-
-The original best practices guide will be split into two documents: the first will gives best practices for running a \emph{single} calculation (and cover theory in more depth/ detail system-specific setup etc.), while the second will give best practices for running \emph{multiple} calculations (with a focus on automation).
-The first will become version 2.0 of the current guide, while the second will be submitted as a new manuscript.
-
-This document outlines the contents of the first document for single calculations.
-}
-
-\todo[inline, color=green!20]{@Volunteer Update abstract to highlight the split of the document and focus on the single calculations.}
-
 Alchemical free energy calculations are a useful tool for predicting free energy differences associated with the transfer of molecules from one environment to another.
@@
 
-% \todo[inline, color=green!20]{JDC: Should we migrate this document to \url{https://github.com/alchemistry}? ASJSM: Let's migrate with the submitted version, but leave it here for now?}
 \end{abstract}
@@
 \newpage
-% Add a comment to the top of the contents page
-\addtocontents{toc}{\vskip2pt\noindent\textcolor{green}{\textbf{Green sections}: Minimal revisions required}\par\vskip2pt}
-\addtocontents{toc}{\vskip2pt\noindent\textcolor{orange}{\textbf{Orange sections}: Substantial revisions required}\par\vskip2pt}
-\addtocontents{toc}{\vskip2pt\noindent\textcolor{red}{\textbf{Red sections}: Needs written from scratch.}\par\vskip2pt}
 
@@
 
-\section{\tocgreen{What are alchemical free energy methods?}}
-\tocsectioncomment{Steering Meeting: @Jenke Scheen to think about how figure can be made the "central axis" of the paper.
-Possibly update/ create new full-page figure and expand so that it shows a ``dummy workflow'' for an alchemical simluation including all general stages (e.g setup, analysis with different estimators etc.), and guides readers to appropriate sections.}
+\section{What are alchemical free energy methods?}
 \label{sec:intro}
@@
 %%%%%%%%%%%%%%%%%%%%
-\section{\tocorange{Prerequisites and Scope}}
-\tocsectioncomment{Steering Meeting: @Volunteer to update reflecting new split between single and multiple calculations. @Michael Shirts to update best practices on multiple binding sites.}
+\section{Prerequisites and Scope}
 \label{sec:pre}
@@
 %%%%%%%%%%%%%%%%%%%%
-\section{\tocgreen{Statistical mechanics demonstrates why alchemical free energy calculations work}}
-%\section{\tocorange{Statistical mechanics theory of solvation, solvent partitioning, and binding}}
+\section{Statistical mechanics demonstrates why alchemical free energy calculations work}
+%\section{Statistical mechanics theory of solvation, solvent partitioning, and binding}
 \label{sec:theory}
@@
 %%%%%%%%%%%%
-\section{\tocorange{What can be expected from alchemical simulations?}}
-\tocsectioncomment{Steering Meeting: @Volunteer Shorten and move most material to second paper.
-Possibly add a walkthrough of a ``dummy calculation'' (which could be a particular realisation of a general workflow laid out in @Jenke Scheen's figure)?
-This could lead on to discussion of particular methods.
-@Toni Mey I could do this}
+\section{What can be expected from alchemical simulations?}
 \label{sec:step0}
@@
 %
-\section{\tocorange{Simulation prerequisites}}
-\tocsectioncomment{There was substantial interest in expanding this section (see initial suggestions/ discussion).}
-\tocsectioncomment{@Irfan Alibay/ @Volunteer to suggest additional points.}
-\tocsectioncomment{@Irfan Alibay/ @Emilio Gallicchio/ @Volunteer to expand on approaches for explicitly accounting for protonation/ tautomer changes.}
-\tocsectioncomment{@Hannah Bruce Macdonald/ @David Slochower/ @Vytas Gapsys/ @Volunteer to expand range of systems discussed, e.g. membrane proteins, cofactors, multimers, PPIs.}
+\section{Simulation prerequisites}
 \label{sec:prerequisites}
@@
 \subsection{Free energies of binding}
-\tocsubsectioncomment{{Steering Meeting: @Volunteer move paragraph ``Congeneric series often need alignment'' to second paper if we are assuming a single calculation?}}
 \label{subsec:binding}
@@
 \subsection{Free energies of hydration or partition coefficients}
-\tocsubsectioncomment{Steering Meeting: @Volunteer to expand coverage of free energies of hydration/ partition coefficients to make coverage of applications more equal.}
 \label{subsec:hydration}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\section{\tocorange{What simulation protocol should I choose?}}
-\section{\tocorange{Choice of simulation protocol and implementation of alchemical transformations}}
+%\section{What simulation protocol should I choose?}
+\section{Choice of simulation protocol and implementation of alchemical transformations}
   \sbnote{Emilio suggested to make the title more general; this is my suggestion!}
-\tocsectioncomment{Steering group: Possibly expand to include other methods group agrees on.}
-\tocsectioncomment{@Stefan Boresch to restructure as per his suggestion to remove single/dual topology section and begin with a more fundamental discussion which might more naturally include all options e.g.
-ATM.
-@Stefan Boresch to mention best practices for dummy atoms?
-This might benefit from following @Emilio Gallicchio's changes to make the theory section more general.}
-\tocsectioncomment{@Sheenam Khuttan / @Emilio Gallicchio to contribute section on ATM.}
-\tocsectioncomment{@Vytas Gapsys / @Volunteer to add best practices for non-equilibrium calculations.}
 \label{sec:simulation_protocol_choice}
@@
 \subsubsection{Choices unique to relative free energy calculations}
-\tocsubsubsectioncomment{Steering Group: @Volunteer to mainly move atom mapping section to second paper and assume that the change is small (and hence atom mapping is trivial).
-@Benjamin Ries to add discussion of 3D atom mapping in second paper.}
 \label{sec:relative-fe-protocol}
@@
 
-\tocsubsubsectioncomment{@Emilio Gallicchio to comment on possible issues with theoretical discussion and possibly update in the context of revised theory section.
-See PR from @Finlay Clark to improve discussion of restraint schemes and automated restraint selection.
-@Stefan Boresch may like to weigh in on analytical corrections/ general discussion.
-@Andrea Rizzi to reivew?}
-
 \begin{figure}
@@
 \subsubsection{Changes in net charge can be challenging/problematic.}
-\tocsubsubsectioncomment{@Chris Oostenbrink/ Vytas Gapsys to update.}
+
 If the net charge of the system changes as the alchemical variable changes during the calculation, this can pose major challenges.
@@
 \subsubsection{Multiple or uncertain binding modes may require considerable care}
-\tocsubsubsectioncomment{@Michael Shirts to update discussion of treatment of multiple binding poses.}
 \label{sec:multiple_binding_modes}
@@
 \subsection{General simulation setting choices that could affect free energy calculations.}
-\tocsubsectioncomment{@Volunteer: Add short discussion of recommended integrators? BAOAB?}
 There are many parameter choices that are common to standard MD simulations.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{\tocorange{Data analysis}}
+\section{Data analysis}
 \label{sec:data_analysis}
@@
 \subsection{Uncertainty estimation}
-\tocsubsectioncomment{To be split between single/ multiple papers. @Agastya P Bhati/ Peter Coveney to expand. Cycle closure etc. to be moved to second paper.}
-\tocsubsectioncomment{@Volunteer: Add discussion of bootstrapping vs t-based confidence intervals for repeats of single calculations (both bad options)?}
 \label{subsec:uncertainty}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{\tocred{Unsettled Practices and Current Limitations}}
-\tocsectioncomment{@Everyone to contribute to section on unsettled practices/ current limitations, as suggested by John Chodera.}
+\section{Unsettled Practices and Current Limitations}
 \label{sec:unsettled}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{\tocorange{Conclusions}}
-\tocsubsectioncomment{@Volunteer to update reflecting focus on single calculations.}
+\section{Conclusions}
 \label{sec:conclusion}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{\tocorange{Selection of available software packages}}
-\tocsectioncomment{@Volunteers to update.}
+\section{Selection of available software packages}
 
@@
 
-\section{\tocorange{Checklist}}
-\tocsectioncomment{@Volunteer to update checklists/ move relevant sections to second paper.}
+\section{Checklist}
 \label{sec:checklist}

### 2025-08-07 paper/manuscript.tex
Expand on free energies of hydration or partition coefficients by @megosato
These changes address https://github.com/alchemistry/alchemical-best-practices/issues/139# but are yet to be reviewed.

@@
 \author[11, 12, 13]{Peter V. Coveney}
+\author[15]{Jonathan W. Essex}
 \author[9]{Vytautas Gapsys}
@@
 \author[5]{Levi N. Naden}
+\author[15,16]{William G. Poole}
 \author[6]{Samarjeet Prasad}
@@
 \author[14]{David R. Slochower}
+\author[15]{Oliver J. Melling}
 \author[8*]{Michael R. Shirts}
@@
 \affil[14]{Vertex Pharmaceuticals, San Diego, CA, USA}
+\affil[15]{School of Chemistry and Chemical Engineering, University of Southampton, Southampton, SO17 1BJ, United Kingdom}
+\affil[16]{Astex Pharmaceuticals, 436 Cambridge Science Park, Milton Road, Cambridge, CB4 0QA, United Kingdom}
 
@@
 \orcid{Peter V. Coveney}{0000-0002-8787-7256}
+\orcid{William G. Poole}{0009-0003-2441-8794}
+\orcid{Oliver J. Melling}{0000-0001-6243-7433}
+\orcid{Jonathan W. Essex}{0000-0003-2639-2746}
 
@@
 For example, a 3D minimised structure of a solute can be generated with a simple tool such as openBabel and solvated to prepare the input to compute a free energy of hydration~\cite{oboyle2011open}.
-However, in these cases, a careful choice of forcefield for the organic solvent model, as well as water model is essential.
+Despite their relative simplicity, care should still be taken when preparing these systems to avoid introducing systematic errors.
+
+\paragraph{The choice of solvent forcefield impacts the calculation}
+A careful choice of forcefield for the organic solvent model, as well as water model is essential.
+A critical aspect of these calculations is the preparation ahead of simulation.
+The choice of force field for both the solute and solvents involved is critical.
+Accurate modeling of the solute's solvation behavior hinges on realistic representations of both water and the organic solvent (for partition coefficient calculations).
 See for example~\cite{bosisio2016blinded,rustenburg2016measuring} for a good discussion of these choices.
-And, while sampling problems might seem to be a non-issue for small molecules, this is not always the case; e.g.
-even the hydroxyl orientation on neutral carboxylic acids can occasionally pose a challenge~\cite{klimovich2010predicting, lim2019assessing}.
+The same level of scrutiny is needed for organic solvents such as octanol, cyclohexane, or others used in partitioning studies, where solvent model parameters may vary substantially between force fields and impact results accordingly.
+
+\paragraph{Charge models are non-trivial}
+Modeling the partial charges of the solute and solvent molecules is also highly important.
+In some fixed-charge models, the assigned partial charges depend on the input conformation of the molecule, meaning that poor or inconsistent conformer selection can introduce systematic errors in solvation free energy estimates \cite{jakalian2000, jambeck2003}.
+This is particularly true for methods such as RESP or AM1-BCC, where the geometry of the molecule at the time of charge derivation can influence the electrostatic potential fit \cite{jakalian2000, basma2001, zhang2011, dupradeau2010, osato2025}.
+Conformers with strong intramolecular interactions In these cases, the conformer(s) used for assigning the partial charges should be saved for reproducibility.
+Ensuring consistency between the charge derivation geometry and the initial simulation structure is therefore recommended, or alternatively, using ensemble-averaged charges or polarizable models where appropriate \cite{ren2002,lemkul2016}.
+
+\paragraph{Conformational sampling can impact solvation and partition free energies}
+Moreover, while sampling challenges are often perceived to be minor for small molecules, they can still be non-trivial.
+Even subtle degrees of freedom such as hydroxyl group orientations (as in neutral carboxylic acids) or intramolecular hydrogen bonding can influence the calculated free energy difference between two environments\cite{klimovich2010predicting, lim2019assessing}.
+Molecules with multiple rotatable bonds, internal hydrogen bonds, or conformational flexibility may exhibit environment-dependent populations that are difficult to sample adequately within the timescale of typical molecular dynamics simulations.
+In some cases, enhanced sampling techniques such as replica exchange or metadynamics may be beneficial, even in hydration or partition coefficient calculations.
+
+In partition coefficient calculations, the situation becomes even more complex, as two distinct environments must be adequately sampled: typically water and a nonpolar organic solvent.
+These systems not only differ in their dielectric properties and hydrogen bonding behavior but may also yield different dominant solute conformers.
+The partition free energy is sensitive to the relative population of conformers in each environment, so it is essential to ensure that sampling in each phase is independently converged.
+Some studies have employed restrained or confining potentials to prevent solutes from drifting to the interface (especially when using periodic systems with both phases present), while others model each phase separately and compute the free energy difference via thermodynamic cycles\cite{turchi2019,bannan2016,petris2021}.
+
+\paragraph{Validation through experimental comparisons}  
+Finally, experimental comparisons in these systems are often more direct than in binding free energies.
+Hydration and partition coefficients are frequently measured with high precision, enabling rigorous validation of computational protocols.
+This has led to their prominent use in blind challenges such as SAMPL, where participants predict hydration or partition free energies of small molecules using a variety of force fields, solvent models, and sampling strategies.
+These exercises have repeatedly shown that small changes in simulation parameters or protocols can result in significant deviations from experimental results, reinforcing the need for careful setup and validation even in seemingly "simple" solvation systems.
 
@@
 \textbf{BA}: Helped write the uncertainty estimation, stopping conditions, and output analysis sections and created figure ~\ref{fig:convergence_forward_reverse}.\\
+\textbf{APB} Contributed updates to the discussion of uncertainty quantification.\\
 \textbf{HBM} Contributed to Sec.~\ref{sec:plot_data} and Fig.~ \ref{fig:scatterplot_analysis} and helped edit the paper.\\
@@
 \\
+\textbf{PVC} Contributed updates to the discussion of uncertainty quantification.\\
+\textbf{JWE} Contributed to ``Conserved binding site waters can play an important role in binding free energies'' and ``Handling poorly sampled water networks'' in Sec. ~\ref{subsec:binding}.\\
 \textbf{DFH} Contributed to Sec.~\ref{sec:software} and helped edit the paper. \\
@@
 \textbf{LNN}: Helped write the simulation length, stopping conditions, and information saving section. Edited and reviewed alchemical path section.\\
+\textbf{WJP} Contributed to ``Conserved binding site waters can play an important role in binding free energies'' and ``Handling poorly sampled water networks'' in Sec. ~\ref{subsec:binding}.\\
 \textbf{SP} Wrote Sec.~\ref{sec:software}. \\
@@
 Wrote Sec.~\ref{sec:plot_data}, the checklist Sec.~\ref{sec:checklist}, and contributed to general formatting discussions and editing.\\
+\textbf{OJM} Contributed to ``Conserved binding site waters can play an important role in binding free energies'' and ``Handling poorly sampled water networks'' in Sec. ~\ref{subsec:binding}.\\
 \textbf{MRS}: Helped create figure~\ref{fig:fig_what_is_lambda}, wrote Sec.~\ref{sec:important_path} describing choices for alchemical pathways and parts of~\ref{sec:data_analysis} on the analysis for free energy calculations.
@@
 JDC is a current member of the Scientific Advisory Board of OpenEye Scientific Software and a consultant to Foresite Laboratories.
+JWE receives research funding from AZ, Astex Pharmaceuticals, Exscientia, GSK, dstl, UCB, Diamond Light Source. Astex Pharmaceuticals part funds WGP., where he is now employed.
 \section*{Funding Information}
@@
 The content is solely the responsibility of the authors and does not necessarily represent the official views of the National Institutes of Health.
+OJM, WGP, and JWE acknowledge funding from the EPSRC, Astex Pharmaceuticals, and Defence Threat Reduction Agency.
 

### 2025-08-07 paper/manuscript.tex
Update Figure 1 by @jenkescheen
This closes https://github.com/alchemistry/alchemical-best-practices/issues/133.

@@
 \author[2,7]{Andrea Rizzi}
-\author[1]{Jenke Scheen}
+\author[17]{Jenke Scheen}
 \author[14]{David R. Slochower}
@@
 \affil[16]{Astex Pharmaceuticals, 436 Cambridge Science Park, Milton Road, Cambridge, CB4 0QA, United Kingdom}
+\affil[17]{Open Molecular Software Foundation, Davis, CA 95618, USA}
 
@@
 
-\begin{figure}
-    \includegraphics[width=0.95\linewidth]{figures/fig1_what_is_alchemy/Figure.pdf}   
-    \caption{\textbf{Illustration of common types of free energies differences that can be calculated using alchemical free energy methods.} \textbf{A}: Change in free energy due to a conformational change of the molecule across a high barrier.
-\textbf{B}: Partition coefficient such as $\log P$ or $\log D$ depend on a change in free energy between different phases; here, as an example the partition coefficient between methanol and water is shown.
-\textbf{C}: Free energy difference associated with the insertion of a molecule into a membrane.
-\textbf{D}: Effect of mutations of protein or host residues on free energies of binding.
-\textbf{E}: Relative free energy of binding of one molecule with respect to another, here toluene and benzyl alcohol, \textbf{F}: absolute free energies of binding of a small molecule to a host (e.g.
+\begin{figure*}
+    \centering\includegraphics[width=0.85\linewidth]{figures/fig1_what_is_alchemy/Figure.pdf}   
+    \caption{\textbf{Illustration of common types of free energies differences that can be calculated using alchemical free energy methods and a typical workflow for executing a relative binding free energy calculation (yellow box).}
+    \textbf{A}: Partition coefficient such as $\log P$ or $\log D$ depend on a change in free energy between different phases; here, as an example the partition coefficient between methanol and water is shown.
+\textbf{B}: Change in free energy due to a conformational change of the molecule across a high barrier.
+\textbf{C}: Absolute free energies of binding of a small molecule to a host (e.g.
 protein).
+\textbf{D}: Free energy difference associated with the insertion of a molecule into a membrane.
+\textbf{E}: Relative free energy of binding of one molecule with respect to another, here toluene and benzyl alcohol.
+\textbf{F}: Effect of mutations of protein or host residues on free energies of binding.
+    \textbf{G.1}: the protein structure of the system is \textit{prepared} such that all erroneous patterns are resolved and proper protonation is applied.
+All query ligands are positioned into the binding site.
+\textbf{G.2}: $N=6$ compounds are predicted on in this example.
+\textbf{G.3}: $n=7$ transformations (edges) are chosen in this example.
+\textbf{G.4}: for each transformation, a $\lambda$ decoupling parameter is applied to guide the transformation.
+\textbf{G.5}: for each transformation and $\lambda$ window a simulation is run.
+\textbf{G.6}: for each transformation, the relative binding free energy ($\Delta\Delta$G$_{bind}$) is calculated given $\lambda$.
+\textbf{G.7}: Each transformation (edge) has a corresponding binding free energy estimation.
+\textbf{G.8}: a final calculation is done to estimate the absolute binding free energy of each compound using for example a maximum-likelihood estimator.
+\textbf{G.9}: Absolute binding free energy estimates are ready to be used for prioritization in synthesis.
 \label{fig:fig_what_is_alchemy}
     }
-\end{figure}
+\end{figure*}
 

### 2025-08-07 paper/manuscript.tex
Merge .bib files and deduplicate entries

@@
 or accepting the introduction of potential systematic error, that is in many cases negligible due to the cancellation of errors.
-\cite{ries2024Kartograf}
+\cite{ries2024kartograf}
 
@@
 A next step for atom mappings, would be to support multistate methods (e.g.
-RE-EDS~\cite{sidler2016replica}, A-EDS~\cite{perthold2020toward} or Ladybugs~\cite{robo2022achieving}) with efficient mapping algorithms.
+RE-EDS~\cite{sidler2016replica}, A-EDS~\cite{perthold2020toward} or Ladybugs~\cite{robo2023fast}) with efficient mapping algorithms.
 There are protoype starting point for solving this problem, but to our knowledge, there is only few published work on this aspect.
-\cite{petrov2021Perturbation}
+\cite{petrov2021perturbation}
 
@@
 Additionally, computating the overall variance of TI requires the calculation of the overall variance of the integration, rather than each individual $\Delta$G$_{i,i+1}$ and assuming variances add independently.
-Therefore, $\mathrm{var}$($\Delta$f) = $\sum_{i=1}^{K}w_{k}^2 \mathrm{var}(\frac{du}{d\vec{\lambda}})_{k}$, where the $w_k$ are the integration weights (see Sec~\eref{subsec:estimators} for more discussion of $w_k$).
+Therefore, $\mathrm{var}$($\Delta$f) = $\sum_{i=1}^{K}w_{k}^2 \mathrm{var}(\frac{du}{d\vec{\lambda}})_{k}$, where the $w_k$ are the integration weights (see Sec~\ref{subsec:estimators} for more discussion of $w_k$).
 
@@
 For example, there is growing interest in identifying small molecules that directly bind to RNA or DNA, with recent preliminary studies showing promising results using specialized simulation protocols.\cite{Abramyan2025,Rasouli2024} In contrast, there has been limited investigation into the applicability of AFE methods to small-molecule ligands targeting intrinsically disordered proteins (IDPs,\cite{Papadourakis2024} or proteins containing intrinsically disordered regions (IDRs) that undergo disorder-to-order transitions upon ligand binding.\cite{MendozaMartinez2022}
-The treatment of covalent binders has also received some attention.\cite{Luo2021,Zhang2019 } However, robust protocols for handling datasets of irreversible covalent binders containing diverse ‘warheads’ have not yet achieved widespread adoption.\cite{Yu2019} A variety of other applications are less traditional and may be possible but raise new practical issues; for example, if multiple copies of a ligand bind at once in the same site, this could pose challenges, as could handling of binding and function prediction for molecular glues and related categories of molecules.
+The treatment of covalent binders has also received some attention.\cite{luo2021,zhang2019} However, robust protocols for handling datasets of irreversible covalent binders containing diverse ‘warheads’ have not yet achieved widespread adoption.\cite{Yu2019} A variety of other applications are less traditional and may be possible but raise new practical issues; for example, if multiple copies of a ligand bind at once in the same site, this could pose challenges, as could handling of binding and function prediction for molecular glues and related categories of molecules.
 
@@
 \makeorcid
-\bibliography{alchemical,manual,sb,atm,uq}
+\bibliography{merged}
 

### 2025-08-11 paper/manuscript.tex
Update checklist and prerequisites and scope by weitse-hsu
Also minor tweaks by @JenkeScheen, @egallicc, and @fjclark

@@
 \begin{itemize}
-\item computing the free energy of a conformational change for a molecule with a high barrier to interconversion (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{A});
-\item computing partition ($\log P$) or distribution ($\log D$) coefficients between environments (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{B})~\cite{rustenburg2016measuring, bosisio2016blinded} 
-\item determining partitioning between compartments into membranes (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{C})~\cite{corey2019insights}. 
+\item computing the free energy of a conformational change for a molecule with a high barrier to interconversion (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{B});
+\item computing partition ($\log P$) or distribution ($\log D$) coefficients between environments (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{A})~\cite{rustenburg2016measuring, bosisio2016blinded} 
+\item determining partitioning between compartments into membranes (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{D})~\cite{corey2019insights}. 
 \end{itemize}
@@
 \begin{itemize}
-\item a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{D})\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate};
-\item the entire ligand can be alchemically transferred from protein to solvent in an absolute binding free energy calculation (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{E})~\cite{mobley2007predicting,aldeghi2015accurate,aldeghi2017predictions};
-\item small alchemical modifications can be made between chemically related ligands to estimate relative differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{F})~\cite{wang2015accurate,mey2016blinded,song2019using,gapsys2020large,kuhn2020assessment}.
+\item a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{F})\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate};
+\item the entire ligand can be alchemically transferred from protein to solvent in an absolute binding free energy calculation (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{C})~\cite{mobley2007predicting,aldeghi2015accurate,aldeghi2017predictions};
+\item small alchemical modifications can be made between chemically related ligands to estimate relative differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{E})~\cite{wang2015accurate,mey2016blinded,song2019using,gapsys2020large,kuhn2020assessment}.
 \end{itemize}
@@
 All query ligands are positioned into the binding site.
-\textbf{G.2}: $N=6$ compounds are predicted on in this example.
+\textbf{G.2}: $N=6$ compounds are predicted in this example.
 \textbf{G.3}: $n=7$ transformations (edges) are chosen in this example.
@@
 
-Some other background information may be needed depending on the nature of the alchemical project.
+Notably, this document---version 2.0 of the original guide--focuses specifically on best practices on single alchemical free energy calculations.
+A companion guide, Best Practices for Multiple Alchemical Free Energy Calculations, covers the considerations involved in high-throughput or large-scale campaigns, such as automation, perturbation network design, and campaign-level data management and analysis.
+Readers aiming to scale up to multiple calculations may find that guide more relevant to those aspects.
+
+In this guide, some other background information may be needed depending on the nature of the alchemical project.
 For example, often, if binding poses are not known, docking calculations can be used to generate an initial small molecule binding pose to start alchemical simulations.
@@
 $K_b^{\circ}$ --- binding  constant \\
-$c^{\circ}$ --- standard state concentration \\
+$C^{\circ}$ --- standard state concentration \\
+$P^{\circ}$ --- standard pressure \\
 $U$ --- potential energy\\
@@
 $\tau _{eq}$ --- integrated auto-correlation time\\
-$t_0$ --- equilibration time
+$t_0$ --- equilibration time \\
+$h$ --- Planck's constant \\
+$k_H$ --- Henry's constant \\
+$b$ --- reduced binding energy function \\
+$w(b, \vec{\lambda})$ --- perturbation energy function \\
+$\vec{h}$ --- a displacement vector in the ATM method
 \end{tcolorbox}
@@
 \end{equation}
-The objective of alchemical solvation free energy calculations is to evaluate Eq.~(\ref{eq:zratio-solv-dgexc}), by means of \emph{coupling} or \emph{decoupling} protocols.
+The objective of alchemical solvation free energy calculations is to evaluate Eq.~(\ref{eq:zratio-solv-dgexc}), by means of \emph{coupling} or \emph{decoupling} computational protocols discussed in the following sections.
 
@@
 \begin{equation}
-  \Delta G^\circ_{b} = \mu^\circ_{RL(v)}  - \mu^\circ_{R(v)} - \mu^\circ_{L(v)} = -k_B T \ln Kb 
+  \Delta G^\circ_{b} = \mu^\circ_{RL(v)}  - \mu^\circ_{R(v)} - \mu^\circ_{L(v)} = -k_B T \ln K_b 
   \label{eq:DG0-binding-def}
@@
 \begin{equation}
-  Kb = \frac{C^\circ}{8 \pi^2} \frac{z_{RL(v)}}{z_{R(v)} z_{L(v)}}
+  K_b = \frac{C^\circ}{8 \pi^2} \frac{z_{RL(v)}}{z_{R(v)} z_{L(v)}}
   \label{eq:DG0-binding-constant-def}
@@
 \& 2), however a dual coordinate set approach could use an atom mapping as well, depending on the sampling method, this will be explained in the next steps.
-Once a single coordinate set approach is selected is selected, a crucial next step is to identify the common atoms which will not be perturbed between the two molecules (respectively end-states).
+Once a single coordinate set approach is selected, a crucial next step is to identify the common atoms which will not be perturbed between the two molecules (respectively end-states).
 Rigorously, this process typically comprises a MCSS search of the molecules involved to identify the common substructure.
@@
 
-In general, there are many ways on how an MCSS can be calculated.
+In general, there are many ways to calculate an MCSS.
 One of the most prominent approaches is to estimate the MCSS with a 2D subgraph isomorphism solver(SIS) using the 2D topological stuructre information of molecules\cite{raymond2002maximum}.
@@
 \item WHAM is an approximation to MBAR, and there are no compelling reasons it should be used.
-If careful, it is not necessarily much worse than the other methods, but it always introduces some degree if binning error.
+If careful, it is not necessarily much worse than the other methods, but it always introduces some degree of binning error.
 \item Other variants, especially ones that adaptively determine the free energies can be useful in certain circumstances but beyond the scope of a Best Practices article.
@@
 \begin{itemize}
-    \item Do I understand the biology, chemistry and physics of my system?
-    \item Have I properly prepared my protein and ligand systems?
+    \item Is my problem suitable for an alchemical calculation?
+    \item Do I understand the biology, chemistry, and physics of my system (e.g., binding modes, protonation states at assay pH, or membrane environment)?
+    \item Have I properly prepared my protein and ligand systems (e.g., the use of high-quality starting structures, inclusion of binding site waters and cofactors, stereochemistry considerations)?
     \item Does my system contain any structures that require custom parameters?
@@
         \item Apply the appropriate correction terms
+        \item Ensure sufficient configurational sampling for slow degrees of freedom (e.g., check for kinetic trapping in metastable states).
 \end{itemize}
@@
         \item Check whether dummy atoms were assigned correctly
-        \item Subsample the data to estimate the variance of your free energy estimation protocol (but not to estimate the values themselves)
          \item Apply the appropriate correction terms
+         \item Ensure sufficient configurational sampling for slow degrees of freedom (e.g., check for kinetic trapping in metastable states)
 \end{itemize}
@@
 \begin{checklist}{How do I know which simulations are unreliable?}
-    \textbf{Situations suggesting your relative alchemical free energy calculations have not run properly (assuming absence of experimental affinities)}
+    \textbf{Situations suggesting your alchemical free energy calculations have not run properly (assuming absence of experimental affinities)}
         \begin{itemize}
@@
     \item Simulated systems have not converged - trajectories should be manually checked for consistency; other methods such as generating RMSD plots are also recommended
+    \item Simulated systems not sampling all relevant conformational states (e.g., different torsional states of a side chain in a binding pocket) that are expected to impact free energy values.
+    \item Statistically significant differences in free energy estimates from simulations starting from different initial states usually indicate insufficient configurational sampling
     \newline\newline\textit{Relative:}
@@
     \item Inspect the overlap matrices for lack of overlap
+    \item Check if all relevant conformational states are sufficiently sampled 
 \end{itemize}
@@
 \begin{itemize}
+    \item Decide which calculated values to assess, e.g., shifted or unshifted $\Delta G$,  $\Delta \Delta G$ of computed edges, or $\Delta \Delta G$ of all edges
+    \item Discard the equilibration region. Subsample the data to estimate the variance of your free energy estimation protocol (but not to estimate the values themselves)
+    \item Examine the phase space overlap and the convergence of your simulation
     \item Make sure you have run enough replicates to ensure statistical reliability (\textgreater3)
-    \item Compute both correlation and ranking coefficients and ranking statistics (e.g. r, \textrho, MUE and \texttau)
-    \item Include error bars in all your visual analyses
+    \item Compute both correlation and ranking coefficients and ranking statistics (e.g., r, \textrho, MUE and \texttau)
+    \item Include error bars in all your visual analyses, and state how they were calculated.
     \item Consider bootstrapping uncertainties rather than using free energy error estimators

### 2025-08-11 paper/manuscript.tex
Fix undefined citations

@@
 Conversely, the value of the binding free energy depends on the chosen structural definition of the bound complex.
-The relationship between the predictions of this theory and experimental reporters (thermochemical, spectroscopic, inhibition kinetics, and others) of complex formation has been discussed.\cite{gilson1997statisticalthermodynamic,mihailescu2004theory,gallicchio2011recentadv,simonson2016physical,gallicchio2021comppeptsci} The interpretation of experimental reporters of binding is a complex issue in itself.
+The relationship between the predictions of this theory and experimental reporters (thermochemical, spectroscopic, inhibition kinetics, and others) of complex formation has been discussed.\cite{gilson1997statisticalthermodynamic,mihailescu2004theory,gallicchio2011recent,simonson2016physical,gallicchio2021comppeptsci} The interpretation of experimental reporters of binding is a complex issue in itself.
 Still, the consensus is that this theory gives nearly equivalent results when the binding is strong and specific, and the definition of the bound complex covers the major bound poses while excluding obvious unbound configurations.
@@
 Whereas Ries et al.\  use distance restraints to keep two molecules on top of each other, corresponding atoms in NAMD are held on top of each other by constraints; in practice, this means that these coordinates are forced to be identical.
-AMBER's free energy implementation recently switched to a similar approach \cite{Lee_2023,York_ACSPhysChemAu_2023_v3_p478}.
+AMBER's free energy implementation recently switched to a similar approach \cite{lee2023aces,York_ACSPhysChemAu_2023_v3_p478}.
 While the implementation technically uses two sets of coordinates for the entire transforming region, selected atoms that have a direct atom-atom mapping are constrained to have common coordinates (sometimes referred to as a "common core"), while others that are mapped into dummy atoms and are treated with softcore potentials~\cite{Lee_JChemTheoryComput_2020_v16_p5512,Tsai_JChemTheoryComput_2023_v19_p640} have separable coordinates.
 Thus the common core part of the dual topology are constrained to have the same coordinates to maximize phase space overlap, whereas the softcore region has separate coordinates.
-The two topologies do not see each other, and this property enables this approach to be leveraged in the alchemical enhanced sampling (ACES) method~\cite{Lee_2023,Tsai_JChemInfModel_2024_v64_p7046,Zhang_JChemTheoryComput_2024_v20_p3935}.
+The two topologies do not see each other, and this property enables this approach to be leveraged in the alchemical enhanced sampling (ACES) method~\cite{lee2023aces,Tsai_JChemInfModel_2024_v64_p7046,Zhang_JChemTheoryComput_2024_v20_p3935}.
 %The authors refer to this approach as hybrid single-dual topology.
@@
 The practical implementation of the approaches to carry out alchemial transformations above leads to additional, often non-trivial differences within each methodology.
-In the description of DC approaches we already pointed out that the two sets of coordinates can be held on top of each other by constraints \cite{jiang2019computing,Lee_2023} or restraints.
+In the description of DC approaches we already pointed out that the two sets of coordinates can be held on top of each other by constraints \cite{jiang2019computing,lee2023aces} or restraints.
 In the latter case, the restraints may couple the two set of coordinates directly \cite{Axelsen_1998,Ries_2022}, or the restraints may be defined relative to a common reference frame \cite{rocklin2013separated,Baumann_2023}.
@@
 However the implementation and testing of an fully automatized dual-topology pipeline for RBFE dual topology approaches is still to our knowledge a remaining research problem and was only tested with small datasets.
-\cite{jespers2019qligfep, Rieder2022Lev}
+\cite{jespers2019qligfep, rieder2022lev}
 
@@
 A straightforward but computationally expensive method is to run longer simulations with multiple replicates to increase conformational heterogeneity~\cite{suruzhon2021sensitivity}.
-Another approach, known as integration over parts,\cite{jayachandran2006parallelized}, involves performing separate alchemical calculations limited to a series of binding macrostates and  then computing the overall free energy difference using the free energy combination formula~\cite{gallicchio2011recentadv,hsu2023alchemical}, i.e., computing the negative log-sum-exp of the Boltzmann weights from each macrostate.
+Another approach, known as integration over parts,\cite{jayachandran2006parallelized}, involves performing separate alchemical calculations limited to a series of binding macrostates and  then computing the overall free energy difference using the free energy combination formula~\cite{gallicchio2011recent,hsu2023alchemical}, i.e., computing the negative log-sum-exp of the Boltzmann weights from each macrostate.
 In the integration over parts approach, restraining potentials can be used to restrict the sampling to specific macrostates,\cite{khuttan2023taming}.
@@
 Therefore, a blind application of this sampling scheme is not recommended and it must be used with caution if at all.
-A variant of this method that overcomes some of the reported issues with FEP/REST2 is called alchemical enhanced sampling (ACES)~\cite{Lee_JChemTheoryComput_2023_v19_p472}.
+A variant of this method that overcomes some of the reported issues with FEP/REST2 is called alchemical enhanced sampling (ACES)~\cite{lee2023aces}.
 ACES creates localized targeted "enhanced sampling states" that are connected to the real state end-points through HREMD.
@@
 Lambda-dynamics-based free energy calculation can be carried out using the BLOCK module.
-	\item \href{https://ambermd.org/}{AMBER}, including its GPU-accelerated free energy engine~\cite{Lee_JChemTheoryComput_2017_v13_p3077,Giese_JChemTheoryComput_2018_v14_p1564,Lee_JChemInfModel_2018_v58_p2043} that supports a wide range of alchemical free energy methods~\cite{Tsai_JChemTheoryComput_2023_v19_p640,Lee_2023,Zhang_JChemTheoryComput_2024_v20_p3935}, workflows~\cite{Ganguly_JChemInfModel_2022_v62_p6069} and analysis tools~\cite{Giese_JChemInfModel_2025_v65_p5273,Case_JChemInfModel_2023_v63_p6183}.
+	\item \href{https://ambermd.org/}{AMBER}, including its GPU-accelerated free energy engine~\cite{Lee_JChemTheoryComput_2017_v13_p3077,Giese_JChemTheoryComput_2018_v14_p1564,Lee_JChemInfModel_2018_v58_p2043} that supports a wide range of alchemical free energy methods~\cite{Tsai_JChemTheoryComput_2023_v19_p640,lee2023aces,Zhang_JChemTheoryComput_2024_v20_p3935}, workflows~\cite{Ganguly_JChemInfModel_2022_v62_p6069} and analysis tools~\cite{Giese_JChemInfModel_2025_v65_p5273,Case_JChemInfModel_2023_v63_p6183}.
 	\item \href{http://www.gromos.net/}{GROMOS} offers an extensive and flexible molecular dynamics and simulations analysis suites with free energy calculation functionalities including customizable alchemical paths and various sampling protocols~\cite{schmid2012architecture, kunz2012new, eichenberger2011gromos}.

### 2025-08-13 paper/manuscript.tex
- Update checklists and scope by @weitse-hsu - Resolve some changes to the discussion of uncertainty by @ppxasjsm - Several minor tweaks and addition of SB and EG to author/ contribution list

@@
 \author[11]{Agastya P. Bhati}
+\author[18]{Stefan Boresch}
 \author[3]{Hannah E. Bruce Macdonald}
@@
 \author[15]{Jonathan W. Essex}
+\author[19]{Emilio Gallicchio}
 \author[9]{Vytautas Gapsys}
@@
 \affil[17]{Open Molecular Software Foundation, Davis, CA 95618, USA}
+\affil[18]{University of Vienna, Faculty of Chemistry, Institute of Computational Biological Chemistry, Wahringer Str. 17, A-1090 Vienna}
+\affil[19]{Department of Chemistry and Biochemistry, Brooklyn College of the City University of New York, New York, NY, 11210
+Ph.D. Program in Chemistry, The Graduate Center of the City University of New York, New York, NY, 10016
+Ph.D. Program in Biochemistry, The Graduate Center of the City University of New York, New York, NY, 10016}
 
@@
 \orcid{Bryce Allen}{0000-0002-0804-8127}
+\orcid{Stefan Boresch}{0000-0002-2793-6656}
 \orcid{Hannah E. Bruce Macdonald}{0000-0002-5562-6866}
@@
 \orcid{Maximilian Kuhn}{0000-0002-2811-3934}
+\orcid{Emilio Gallicchio}{0000-0002-2606-4913}
 \orcid{Julien Michel}{0000-0003-0360-1760}
@@
 
-For convenience we have also compiled a list of common acronyms and common symbols used throughout this paper.
+For convenience we have also compiled a list of common acronyms and common symbols (both in alphabetical order) used throughout this paper.
 \begin{tcolorbox}[title=Acronyms, colback=blue!10!white]
+    {\bf BAR} --- Bennett Acceptance Ratio\\
     {\bf CPU} --- Central Processing Unit\\
-     {\bf BAR} --- Bennett Acceptance Ratio\\
      {\bf FEP} --- Free Energy Perturbation\\
@@
      {\bf MD} --- Molecular Dynamics\\
+     {\bf MUE} --- Mean Unsigned Error\\
      {\bf RMSE} --- Root Mean Square Error\\
-     {\bf MUE} --- Mean Unsigned Error\\
      {\bf SAR} --- Structure-Activity Relationships\\
@@
 \begin{tcolorbox}[title=List of Symbols, colback=green!10!white]
+$A$ --- Helmholtz free energy (free energy in the canonical ensemble) or Helmholtz function, with Helmholtz free energy used in the text.\\
+$b$ --- reduced binding energy function \\
+$C^{\circ}$ --- standard state concentration \\
+$C_t$ --- discrete-time-normalized fluctuation auto-correlation function\\
+$f$ --- reduced (dimensionless) free energy \\
+$G$ --- Gibbs free energy (free energy in the isothermal isobaric ensemble), Gibbs function, or free enthalpy, though the most common term Gibbs free energy is used in the text\\
+$g$ --- statistical inefficiency\\
+$h$ --- Planck's constant \\
+$\vec{h}$ --- a displacement vector in the ATM method
+$k_B$ --- Boltzmann constant \\
+$k_H$ --- Henry's constant \\
+$K_b^{\circ}$ --- binding  constant \\
 $L$ and $R$ --- generic names for ligand and receptor\\
-$K_b^{\circ}$ --- binding  constant \\
-$C^{\circ}$ --- standard state concentration \\
+$\mathcal{O}$ --- overlap matrix\\
 $P^{\circ}$ --- standard pressure \\
+$p$ --- pressure \\
+$\vec{q}$ --- vector of a single configuration, i.e. $x$, $y$, $z$ coordinates of the simulation system\\
+$T$ --- temperature \\
+$t_0$ --- equilibration time \\
 $U$ --- potential energy\\
 $u$ --- reduced (dimensionless) potential describing a thermodynamic state \\
-$G$ --- Gibbs free energy (free energy in the isothermal isobaric ensemble), Gibbs function, or free enthalpy, though the most common term Gibbs free energy is used in the text\\
-$A$ --- Helmholtz free energy (free energy in the canonical ensemble) or Helmholtz function, with Helmholtz free energy used in the text.\\
-$f$ --- reduced (dimensionless) free energy \\
+$w(b, \vec{\lambda})$ --- perturbation energy function \\
+$Z$ --- partition function \\
+$\beta \equiv (k_B T)^{-1}$ --- inverse thermal energy \\
+$\Gamma$ --- configurational space accessible by simulations \\
 $\Delta \hat{f}$ --- estimate from an estimator for the reduced free energy difference between two states\\
-$\Gamma$ --- configurational space accessible by simulations \\
-$\vec{q}$ --- vector of a single configuration, i.e. $x$, $y$, $z$ coordinates of the simulation system\\
-$k_B$ --- Boltzmann constant \\
-$Z$ --- partition function \\
-$p$ --- pressure \\
+$\vec{\lambda}$ --- alchemical progress parameter, which may be multidimensional \\
 $\mu$ --- chemical potential (grand canonical ensemble)\\
-$T$ --- temperature \\
-$\beta \equiv (k_B T)^{-1}$ --- inverse thermal energy \\
-$\vec{\lambda}$ --- alchemical progress parameter, which may be multidimensional \\
-$g$ --- statistical inefficiency\\
-$\mathcal{O}$ --- overlap matrix\\
-$C_t$ --- discrete-time-normalized fluctuation auto-correlation function\\
 $\tau _{eq}$ --- integrated auto-correlation time\\
-$t_0$ --- equilibration time \\
-$h$ --- Planck's constant \\
-$k_H$ --- Henry's constant \\
-$b$ --- reduced binding energy function \\
-$w(b, \vec{\lambda})$ --- perturbation energy function \\
-$\vec{h}$ --- a displacement vector in the ATM method
 \end{tcolorbox}
@@
 An important consideration is how reproducible alchemical free energy calculations are in practice.
-In simple cases such as absolute hydration free energies of small organic molecules, or relative hydration free energy calculations between structurally similar small organic molecules, it should be possible to obtain highly precise estimates with a given software package, i.e.with a sample standard deviation under 0.01 kcal/mol~\cite{rizzi2019sampl6}.
+In simple cases such as absolute hydration free energies of small organic molecules, or relative hydration free energy calculations between structurally similar small organic molecules, it should be possible to obtain highly precise estimates with a given software package, i.e., with a sample standard deviation under 0.01 kcal/mol~\cite{rizzi2019sampl6}.
 For more complex use cases such as protein-ligand binding free energies the repeatability is often substantially worse~\cite{rizzi2019sampl6}. 
@@
 \subsection{Absolute and Relative Binding Free Energies with Multiple Chemical States}
+\label{sec:multiple-chemical-states}
 
@@
 \subsection{Alchemical Transfer Method for Absolute and Relative Binding Free Energy Calculations}
+\label{sec:atm}
 
@@
 Such approaches also assume that the data doesn't have any long time scale drift over the course of the simulation, which is a clear sign of lack of equilibration.
-Generally, most methods will give reasonable correlation times when the simulation lasts 40--50 correlation times,
+Generally, most methods will give reasonable correlation times when the simulation lasts 40--50 correlation times.
 
@@
 It is important to consider the variation in your computed free energies from your equilibrium simulations, in order to obtain an estimate of uncertainty of the obtained value for the free energies of interest.
-A recent Best Practices paper by Grossfield et al.,~\cite{grossfield2018best} provides substantial detail on how to estimate uncertainties from molecular simulations and is a good starting point for this topic.
-The uncertainty of a free energy estimate is limited however, as it is only an estimate of the configurations sampled and cannot contain any information from the phase space not sampled during a simulation.
+The Best Practices paper by Grossfield et al.,~\cite{grossfield2018best} provides substantial detail on how to estimate uncertainties from molecular simulations and is a good starting point for this topic.
+The uncertainty of a free energy estimate is limited, however, as it is only an estimate of the configurations sampled and cannot contain any information from the phase space not sampled during a simulation.
 As a consequence, the variance in free energy afforded by any estimator will always be an underestimate of the true uncertainty.
@@
 The user may choose best practices based on their need for the calculations carried out.
-If only few free energies need to be calculated at very high accuracy a different strategy might be employed than if large scale estimates are needed.
+If only few free energies need to be calculated at very high accuracy, a different strategy might be employed than if large-scale estimates are needed.
 As such, considering overall computational cost or availability of resources with respect to accuracy desired from computations may influence the choice of sampling and analysis scheme.
+Since sampling all of phase space is prohibitively expensive to satisfy the ergodic theorem, certain considerations are often made to to still assess uncertainties in computed free energy estimates.
+If a system exhibits a stable strong binding pose a short number of simulation repeats often offer adequate free energy estimates.
+However, situations with more flexible binding and poses will require more strategic sampling schemes to get estimate uncertainties.
 
@@
 
-Computating of free energies using TI (Sec.n~\ref{subsec:estimators}) is straightforward and the trapezoidal rule is often recommended.
+Computing free energies using TI (Sec.n~\ref{subsec:estimators}) is straightforward and the trapezoidal rule is often recommended.
 The trapezoid rule easily allows unequal spacing of $\vec{\lambda}$ states, which is often required to minimize the variance in the free energy estimate, but in principle any good numerical integration method can be used if care is taken.
 Determining of regions of high curvature when estimating the integral is helpful to determine regions of phase space where more sampling and/or more $\vec{\lambda}$ states are necessary to obtain the best approximation of the integral.
-Plotting $\vec{\lambda}$ with respect to the gradients at each of the $\vec{\lambda}$ values, which can be done automatically with tool such as alchemlyb,  can be a helpful diagnostic.
-Additionally, computating the overall variance of TI requires the calculation of the overall variance of the integration, rather than each individual $\Delta$G$_{i,i+1}$ and assuming variances add independently.
+Plotting $\vec{\lambda}$ with respect to the gradients at each of the $\vec{\lambda}$ values, which can be done automatically with tools such as alchemlyb,  can be a helpful diagnostic.
+Additionally, computing the overall variance of TI requires the calculation of the overall variance of the integration, rather than each individual $\Delta$G$_{i,i+1}$ and assuming variances add independently.
 Therefore, $\mathrm{var}$($\Delta$f) = $\sum_{i=1}^{K}w_{k}^2 \mathrm{var}(\frac{du}{d\vec{\lambda}})_{k}$, where the $w_k$ are the integration weights (see Sec~\ref{subsec:estimators} for more discussion of $w_k$).
@@
 Which torsions have low energy minima can best be found by comparing to the simulation in the solvent.
-There should be clear physical reasons that simulation in the complex has different torsional distributions that the ligand in the solvent.
+There should be clear physical reasons that simulation in the complex has different torsional distributions than the ligand in the solvent.
 \item \textbf{Free energy dependence on $\vec{\lambda}$} The free energy difference between states should vary relatively smoothly with $\vec{\lambda}$.
@@
 \textbf{APB} Contributed updates to the discussion of uncertainty quantification.\\
+\textbf{SB} Wrote the sections on topologies and dummy atoms in Sec.~\ref{sec:relative-fe-protocol} and contributed to the discussion and review of many sections, in particular Sec.~\ref{sec:standardstate-restraints}.\\
 \textbf{HBM} Contributed to Sec.~\ref{sec:plot_data} and Fig.~ \ref{fig:scatterplot_analysis} and helped edit the paper.\\
@@
 \textbf{JWE} Contributed to ``Conserved binding site waters can play an important role in binding free energies'' and ``Handling poorly sampled water networks'' in Sec. ~\ref{subsec:binding}.\\
+\textbf{EG} Wrote the updated Sec.~\ref{sec:theory} and Sec.~\ref{sec:multiple-chemical-states},  updated Sec.~\ref{sec:standardstate-restraints}, co-wrote Sec.~\ref{sec:atm} with SK, and and contributed to the discussion and review of many sections.\\
 \textbf{DFH} Contributed to Sec.~\ref{sec:software} and helped edit the paper. \\
@@
 DLM acknowledges support from the National Institutes of Health (R01GM108889, R01GM124270, and R01GM132386), and the National Science Foundation (CHE 1352608).
-JDC acknowledges support from the National Institutes of Health (NIH P30 CA008748, NIH R01 GM121505, R01 GM132386).
+JDC acknowledges support from the National Institutes of Health (NIH P30 CA008748, NIH R01 GM121505, R01 GM132386), as does EG (NIH 1R15GM151708).
 A complete funding history for the Chodera lab can be found at \url{http://choderalab.org/funding}.

### 2025-08-17 paper/manuscript.tex
Partially resolve changes to uncertainty estimation by @ppxasjsm and @fjclark

@@
 An important consideration is how reproducible alchemical free energy calculations are in practice.
-In simple cases such as absolute hydration free energies of small organic molecules, or relative hydration free energy calculations between structurally similar small organic molecules, it should be possible to obtain highly precise estimates with a given software package, i.e., with a sample standard deviation under 0.01 kcal/mol~\cite{rizzi2019sampl6}.
+In simple cases, such as absolute hydration free energies of small organic molecules, or relative hydration free energy calculations between structurally similar small organic molecules, it should be possible to obtain highly precise estimates with a given software package, i.e., with a sample standard deviation under 0.01 kcal/mol~\cite{rizzi2019sampl6}.
 For more complex use cases such as protein-ligand binding free energies the repeatability is often substantially worse~\cite{rizzi2019sampl6}.
-There is growing literature describing the theoretical underpinning behind such intrinsic non-reproducibility of MD simulation based predictions.
+
+%There is growing literature describing the theoretical underpinning behind such intrinsic non-reproducibility of MD simulation based predictions.
 Coveney \textit{et al.
@@
 The sample standard deviation will give a crude estimate of the reliability of the estimates, and whether the precision is sufficient for the problem at hand.
-When practical, a more stringent test is to use different input coordinates for each replica as well as different velocities.
+When practical, a more stringent test is to use different input coordinates for each repeat, or at least different starting velocities.
+Repeats or ensemble-based simulations will allow for a robust error statistic on single free energy calculations and are typically expected in the order of 1 kcal/mol or better.
+
+%To estimate ensemble averages and provide robust error statistics generally three independent repeats of simulations are recommended.
+In this case a repeat or replica are taken as synonyms.
+As such other approaches use even larger ensembles of simulations such as in the TIES method~\cite{bhati2017} developed by Bhati et al., where a minimum of 5 to 10 replicas are recommended to make up the ensemble for RBFE and ABFE calculations respectively~\cite{bhati2022, bhati2025}.
+Typically single free energy differences across multiple replicas are expected to be reproducible within 1 kcal/mol.
 
 Note that these types of statistical differences concern calculations carried out with a single software package, but simulation package variations can introduce additional discrepancies.
-Such issues of reproducibility of free energy calculations across different simulation packages have attracted attention recently~\cite{loeffler2018reproducibility,rizzi2019sampl6}.
+Such issues of reproducibility of free energy calculations across different simulation packages have attracted attention recently~\cite{loeffler2018reproducibility,rizzi2019sampl6, wade2022}.
 Greater variability is expected between packages due to methodological differences such as integrators, thermostats, barostats, treatment of long-range electrostatics, and potentially other factors.
@@
 In the recent SAMPL6 SAMPLing challenge a larger variability of 0.3 to 1.0 kcal/mol was noted in the computed absolute binding free energies of host/guest systems even though the study sought to use identical input and simulation parameters~\cite{rizzi2019sampl6} and, in many cases, single-point energies were identical or nearly so.
-A recent study demonstrates reproducibility across software packages as well as free energy estimators, but only when ensemble simulations are employed~\cite{wade2022}.
+% Add something on RBFE calculations across engines.
+% I think one of the points here that may be one for the other paper is that we are really just looking at a per edge DDG basis. 
+% - https://chemrxiv.org/engage/api-gateway/chemrxiv/assets/orp/resource/item/684b48f03ba0887c33ef29e7/original/modular-and-interoperable-workflows-for-benchmarking-alchemical-binding-free-energy-calculation-methodologies.pdf
+% - other references?
+
+%A recent study demonstrates reproducibility across software packages as well as free energy estimators, but only when ensemble simulations are employed~\cite{wade2022}.
 The authors reported results from two different MD engines (NAMD and OpenMM) as well as two free energy estimators (FEP and MBAR) for a set of 54 protein-ligand complexes which were found to be in good agreement when ensemble simulations were used to control uncertainties.
@@
 compounds might be easily and quickly available commercially for
-testing and free energy calculations could consume far more resources. Free energy calculations thus typically only
+testing and free energy calculations could consume far more resources. Free energy calculations typically only
 appeal when (slow or costly) synthesis would be required or experiments are otherwise cost-prohibitive.
@@
 An additional consideration arises for single topology relative free energy calculations.
-In this class of alchemical free energy calculations it is necessary to generate a molecular topology that may describe the initial and final states of the perturbation (see Fig.~\ref{fig:fig_topology}).
+In this class of alchemical free energy calculations, it is necessary to generate a molecular topology that may describe the initial and final states of the perturbation (see Fig.~\ref{fig:fig_topology}).
 In cases where the end states have high topological similarity and high structural overlap this is relatively straightforward and typically handled by use of MCSS calculations.
@@
 The 'mixed' intermediate atoms (M) are used as well, but they strictly interpolate between physical and dummy atoms, or vice versa.
-Bhati \textit{et al.}~\cite{bhati2017, bieniek2021} presented a variant of the single coordinate dual topology approach which is used in the TIES method.
-Its distinguishing feature is the application of some filters to remove atoms with differences in atom types and partial charges from the structural MCSS identified to ensure chemical similarity between all atom pairs in the final MCSS across both end-points.
-This is followed by ensuring identical charges for all MCSS atoms by averaging between the two end-points.
-The resultant change in the sum of atomic charges for all MCSS atoms is uniformly distributed across all atoms in the corresponding alchemical region conserving the overall ligand charge for both end-points.
-The approach ensures that the final MCSS is both structurally and chemically identical within thresholds, allowing the use of a unique set of MCSS atoms for both ligands and dual topology for their alchemical parts.
 
@@
 A large number of software tools can compute SIS MCSS matches using different cheminformatics packages.
-Some rely on RDKit~\cite{rdkit2019Dec}, such as pmx~\cite{gapsys2015pmx}, LOMAP~\cite{liu2013lead}, FESetup~\cite{loeffler2015fesetup}, ProtoCaller~\cite{suruzhon2020protocaller}, SMArt~\cite{petrov2021perturbation}, and partially BioSimSpace~\cite{hedges2019biosimspace}, while others such as fkckombu~\cite{kawabata20143d} are standalone tools.
+Some rely on RDKit~\cite{rdkit2019Dec}, such as pmx~\cite{gapsys2015pmx}, LOMAP~\cite{liu2013lead}, FESetup~\cite{loeffler2015fesetup}, ProtoCaller~\cite{suruzhon2020protocaller}, SMArt~\cite{petrov2021perturbation}, and partially BioSimSpace~\cite{hedges2019biosimspace}, while others such as fkcombu~\cite{kawabata20143d} are standalone tools.
 Schr\"{o}dinger's FEP+ planning tool was originally based on a version of LOMAP, and it also uses MCSS matching as well as 3D considerations to plan the network of single topology calculations between molecules~\cite{wang2015accurate}.
@@
 The type of restraints used in SepTop \cite{rocklin2013separated,Baumann_2023} and ATM \cite{Azimi_2022} avoid this problem.
-Additionally, TIES~\cite{bhati2017, bieniek2021} circumvents this issue by design as it does not need any restraints.
 
@@
 Still, FEP+ implements one solution\cite{wang2017accurate}.
-TIES also overcomes this issue by using dual topology and ensuring that MCSS atoms are structurally and chemically identical~\cite{bhati2017, bieniek2021}.
 
@@
 A straightforward but computationally expensive method is to run longer simulations with multiple replicates to increase conformational heterogeneity~\cite{suruzhon2021sensitivity}.
-Another approach, known as integration over parts,\cite{jayachandran2006parallelized}, involves performing separate alchemical calculations limited to a series of binding macrostates and  then computing the overall free energy difference using the free energy combination formula~\cite{gallicchio2011recent,hsu2023alchemical}, i.e., computing the negative log-sum-exp of the Boltzmann weights from each macrostate.
-In the integration over parts approach, restraining potentials can be used to restrict the sampling to specific macrostates,\cite{khuttan2023taming}.
+If using a large number of replicates these are often termed ensemble approaches and an example of such a sampling scheme is the TIES method~\cite{bhati2017} developed by Bhati et al., where a minimum of 5 to 10 replicas are recommended to make up the ensemble for RBFE and ABFE calculations respectively~\cite{bhati2022, bhati2025}.
+Another approach, known as integration over parts~\cite{jayachandran2006parallelized}, involves performing separate alchemical calculations limited to a series of binding macrostates and  then computing the overall free energy difference using the free energy combination formula~\cite{gallicchio2011recent,hsu2023alchemical}, i.e., computing the negative log-sum-exp of the Boltzmann weights from each macrostate.
+In the integration over parts approach, restraining potentials can be used to restrict the sampling to specific macrostates~\cite{khuttan2023taming}.
 However, in practice, restraining potentials are sometimes omitted when the states are known to be separated by large free energy barriers and do not interconvert within the simulation time scale, such as cases like Fig.~\ref{fig:configurational_sampling}\textbf{B}.
@@
 Another example would be using true apo and holo protein structures as the end states to initialize simulations for absolute binding free energy calculations.
-Bhati \textit{et al.} recently published two studies directly comparing the equilibrium (EQ) and NEQ approaches at unprecedented scales for both RBFE and ABFE calculations using their ensemble simulation-based TIES method~\cite{wan2023eqvsneq, bhati2025}.
+%Bhati \textit{et al.} recently published two studies directly comparing the equilibrium (EQ) and NEQ approaches at unprecedented scales for both RBFE and ABFE calculations using their ensemble simulation-based TIES method~\cite{wan2023eqvsneq, bhati2025}.
 Further, they systematically studied the effect of various parameters involved in setting up calculations for both approaches and recommended optimal protocols in each case for industrial applications.
@@
 In particular, if the rate of change in the free energy estimate is significant when this condition is met, the simulation may not be locally converged, and more sampling may be necessary to determine a stable free energy estimate which is no longer changing significantly over time.
-A few studies reported systematic variation of simulation lengths using ensemble simulations over an extended range to provide recommendations on appropriate values applicable for most protein-ligand systems.
+%A few studies reported systematic variation of simulation lengths using ensemble simulations over an extended range to provide recommendations on appropriate values applicable for most protein-ligand systems.
 The authors suggest using 4 ns and 10 ns for RBFE~\cite{bhati2017, bhati2022, wan2023eqvsneq} and ABFE~\cite{bhati2025} calculations respectively.
@@
 This could be a fixed value such as $0.20$ kcal/mol, or a functional quantity such as "below $0.5$ kcal/mol and $10\%$ of the free energy estimate." The user does not need to monitor this information in real-time and can choose to run simulations for fixed duration (either time or number of samples) and run analysis on the data collected thus far.
-If more samples are needed, the simulations can be resumed, or, started again in different initial conditions.
-Bhati \textit{et al.} recommended such a scheme for large scale RBFE and ABFE calculations using TIES~\cite{bhati2022, bhati2025, wan2023eqvsneq, coveney2022corners}.
+If more samples are needed, the simulations can be resumed, or, started again in different initial conditions enlarging the overall simulation ensemble.
+%Bhati \textit{et al.} recommended such a scheme for large scale RBFE and ABFE calculations using TIES~\cite{bhati2022, bhati2025, wan2023eqvsneq, coveney2022corners}.
 They suggest a set of optimal parameters (ensemble size, simulation length, number of lambda-states, number and length of NEQ transitions, etc.) to be used initially.
@@
 through standard error propagation methods (including MBAR's estimator, which is based on the same principles as standard error propagation), through bootstrap methods, and through ensemble simulations.
-However, as we have repeatedly pointed out, only ensemble simulations can capture the true uncertainty in MD based predictions.
+%However, as we have repeatedly pointed out, only ensemble simulations can capture the true uncertainty in MD based predictions.
 All other methods underestimate the true uncertainty as reported by several authors~\cite{wan2020fep+, wan2021uq, vassaux2021, gapsys2020large, wade2022, bhati2017, bhati2022, bhati2025, wan2023eqvsneq, bhati2018, bhati2019}.
@@
 As such, considering overall computational cost or availability of resources with respect to accuracy desired from computations may influence the choice of sampling and analysis scheme.
-Since sampling all of phase space is prohibitively expensive to satisfy the ergodic theorem, certain considerations are often made to to still assess uncertainties in computed free energy estimates.
-If a system exhibits a stable strong binding pose a short number of simulation repeats often offer adequate free energy estimates.
+Since sampling all of phase space is prohibitively expensive to satisfy the ergodic theorem, certain considerations are often made to still assess uncertainties in computed free energy estimates.
+If a system exhibits a stable strong binding pose a short number of simulation repeats, sometimes termed ensemble approaches (e.g.
+TIES~\cite{bhati2017}) often offer adequate free energy estimates.
 However, situations with more flexible binding and poses will require more strategic sampling schemes to get estimate uncertainties.
@@
 The number of states and amount of sampling should be optimized to minimize the uncertainty in the MBAR free energy estimate, while balancing other key considerations such as computational expense.
-However, this process based on one-off simulations grossly underestimates the true uncertainty and hence uncertainties for MBAR also must be obtained using ensemble simulations~\cite{bhati2018, wade2022}.
+A less accurate uncertainty estimate may be desirable over a reduced computational expense, increased sampling strategies such as using larger simulation ensembles will provide more accurate uncertainties but will drastically increase the computational cost.
+%However, this process based on one-off simulations grossly underestimates the true uncertainty and hence uncertainties for MBAR also must be obtained using ensemble simulations~\cite{bhati2018, wade2022}.
 
@@
 If different estimators agree the free energy estimate is more reliable than if there are differences between methods that are larger than 1 kcal/mol and would indicate poor convergence.
+
+Uncertainty can also be assessed for a particular perturbation by repeating calculations with slight changes in initial configurations, forcefield parameters, and different random seeds in the MD engine.
+The assessment of variability in free energy calculations due to repeating simulations has been previously reported~\cite{aldeghi2019accurate,paliwal2011benchmark,mey2016blinded,mey2018impact}, and large variance in free energies estimated from simulations with different random seeds should be flagged as issues with convergence.
 
@@
 Other factors, like system preparation (choice of protonation state, tautomer, counterion presence, salt concentration, etc.) can also substantially impact results~\cite{mobley2017predicting, mobley2017predictinga}, so unless modelers are confident they have these factors correct, sensitivity to these choices may also need to be examined.
-Incidentally, Vassaux \textit{et al.} performed a systematic study comparing the effect of both aleatoric as well as parametric uncertainties for protein-ligand free energy calculations using NAMD~\cite{vassaux2021}.
+%Incidentally, Vassaux \textit{et al.} performed a systematic study comparing the effect of both aleatoric as well as parametric uncertainties for protein-ligand free energy calculations using NAMD~\cite{vassaux2021}.
 They found that, indeed, ensemble simulations are necessary to control both sources of uncertainties arising in simulations, just as is anticipated in the field of uncertainty quantification.
@@
 Accordingly, the uncertainty of the free energy difference between the states $i$ and $j$ will be smaller when ${O}_{i,j,i\ne j}$ is larger (and thus the values in the main diagonal (${O}_{i,j,i=j}$) are smaller).
-Ensemble simulations generate wider distributions of energies and hence are expected to increase the overlap in neighbouring states~\cite{wade2022}.
+Ensemble simulations generate wider distributions of energies and hence are expected to increase the overlap between neighbouring states~\cite{wade2022}.
 In order to obtain a reliable estimate of the free energy all neighbouring states must be connected, i.e.
@@
  The precision of NEQ calculations is determined by the overlap between the forward and reverse work distributions corresponding to out-of-equilibrium transitions between the two end-points.
-Bhati \textit{et al.} recently studied the effect of such overlaps on the accuracy of predicted RBFEs~\cite{wan2023eqvsneq} as well as ABFEs~\cite{bhati2025} but found that overlap (or lack of it) does not capture the reliability of calculations very well.
-They recommended an alternative metric, ``distance'' defined as the difference between the inner extremes of the two work value distributions, which correlates quite well with the unsigned errors in free energy predictions and hence was found a better indicator of the reliability of NEQ calculations.
+To estimate the overlap, Wu & Kofke suggest using dissipated work values which can be shown to correspond to the relative entropies of the densities in the phase space.
+\cite{wu2005phase} Another approach to assess NEQ convergence relies on calculating the first two moments of the Fermi function in Bennett Acceptance Ratio estimator.
+\cite{hahn2010measuring} Since the second order estimate converges slower than the first, the overall convergence of the free energy estimate can be quantified as the normalized difference between the two moments of the Fermi function.
+In practice, analytical error estimate of the BAR estimator \cite{shirts2003equilibrium} gives an indication when the overlap is lacking: the estimated error rapidly grows with the increase in the dissipated work.
+\cite{shirts2005comparison} Other empirical measures for uncertainty have been used, e.g.
+Bhati \textit{et al.} observed that the difference between the inner extremes of the two work value distributions correlates with the unsigned errors in RBFE and ABFE predictions.~\cite{wan2023eqvsneq}
 
@@
 	\item Free Energy Workflows: Sire-specific free energy map analysis using weighted path averages \url{https://github.com/michellab/freenrgworkflows}.
-    \item \href{http://www.ties-service.org}{TIES}: Perform statistical analysis of the output from alchemical ensemble simulations and report reliable predictions along with statistical uncertainties for a variety of free energy estimators \url{https://github.com/UCL-CCS/TIES_MD} \cite{bieniek2023}.
+    \item \href{http://www.ties-service.org}{TIES}: Perform statistical analysis of the output of alchemical ensemble simulations and report reliable predictions along with statistical uncertainties for a variety of free energy estimators \url{https://github.com/UCL-CCS/TIES_MD} \cite{bieniek2023}.
 	\end{itemize}

### 2025-08-19 paper/manuscript.tex
Change TIES_MD GH repo links to TIES by @bieniekmateusz

@@
 Equilibrium sampling based approaches would require sampling the central ligand for each edge, while NEQ would sample it once and connect to the other ligands afterwards.
-The NEQ protocol also provides a simple approach in combining different structures in free energy calculations, e.g.
+The NEQ protocol also provides a simple way to combine different structures in free energy calculations, e.g.
 simulations in states at $\lambda=0$ and $\lambda=1$ may be initialized with the independently generated ligand coordinates.
@@
 	\item \href{https://fesetup.readthedocs.io/en/latest/introduction.html}{FESetup} has been developed primarily to setup calculations in AMBER, GROMACS and SIRE~\cite{loeffler2015fesetup}.
-    \item \href{http://www.ties-service.org}{TIES} allows the setup of perturbation maps, hybrid topologies and structures as well as simulation-ready models for NAMD and OpenMM \url{https://github.com/UCL-CCS/TIES_MD} \cite{bieniek2023}.
+    \item \href{http://www.ties-service.org}{TIES} allows the setup of perturbation maps, hybrid topologies and structures as well as simulation-ready models for NAMD and OpenMM \url{https://github.com/UCL-CCS/TIES} \cite{bieniek2023}.
 	\end{itemize}
@@
 	\item Free Energy Workflows: Sire-specific free energy map analysis using weighted path averages \url{https://github.com/michellab/freenrgworkflows}.
-    \item \href{http://www.ties-service.org}{TIES}: Perform statistical analysis of the output of alchemical ensemble simulations and report reliable predictions along with statistical uncertainties for a variety of free energy estimators \url{https://github.com/UCL-CCS/TIES_MD} \cite{bieniek2023}.
+    \item \href{http://www.ties-service.org}{TIES}: Perform statistical analysis of the output of alchemical ensemble simulations and report reliable predictions along with statistical uncertainties for a variety of free energy estimators \url{https://github.com/UCL-CCS/TIES} \cite{bieniek2023}.
 	\end{itemize}

### 2025-08-28 paper/manuscript.tex
Mainly responses to review of uncertainty quantification updates by @apbhati1. Also includes updates to authorship.

@@
 \author[9]{Vytautas Gapsys}
+\author[23,24]{Michael Gillhofer}
 \author[9]{David F. Hahn}
+\author[20]{Wei-Tse Hsu}
 \author[1,10]{Maximilian Kuhn}
+\author[15]{Oliver J. Melling}
 \author[1]{Julien Michel}
@@
 \author[5]{Levi N. Naden}
+\author[23,24]{Chris Oostenbrink}
+\author[4]{Meghan Osato}
 \author[15,16]{William G. Poole}
 \author[6]{Samarjeet Prasad}
+\author[21,22]{Benjamin Ries}
 \author[2,7]{Andrea Rizzi}
@@
 \author[14]{David R. Slochower}
-\author[15]{Oliver J. Melling}
 \author[8*]{Michael R. Shirts}
@@
 Ph.D. Program in Biochemistry, The Graduate Center of the City University of New York, New York, NY, 10016}
-
-%
+\affil[20]{Department of Biochemistry, University of Oxford, Oxford OX1 3QU, United Kingdom}
+\affil[21]{Boehringer Ingelheim Pharma GmbH \& Co KG, Medicinal Chemistry, Birkendorfer Str 65, 88397 Biberach an der Riss, Germany}
+\affil[22]{Open Free Energy, Open Molecular Software Foundation, Davis, CA, 95616, United State}
+\affil[23]{Institute of Molecular Modeling and Simulation, BOKU University, 1190 Vienna, Austria}
+\affil[24]{Christian Doppler Laboratory for Molecular Informatics in the Biosciences, BOKU University, 1190 Vienna, Austria}
+
 \corr{antonia.mey@ed.ac.uk}{ASJSM}
@@
 \orcid{Jonathan W. Essex}{0000-0003-2639-2746}
+\orcid{Wei-Tse Hsu}{0000-0001-6167-5480}
+\orcid{Benjamin Ries}{0000-0002-0945-8304}
+\orcid{Meghan Osato}{0000-0002-2732-457X}
+\orcid{Michael Gillhofer}{0000-0001-8081-6373}
+\orcid{Chris Oostenbrink}{0000-0002-4232-2556}
 
@@
 
-%There is growing literature describing the theoretical underpinning behind such intrinsic non-reproducibility of MD simulation based predictions.
-Coveney \textit{et al.
-}~\cite{coveney2016} have noted that, for a complex dynamical system to reach equilibrium, the dynamics must at least be ``mixing'' which is an established result in ergodic theory.
-A system which is ``mixing'' is also ``ergodic'' by definition but the reverse is not true.
-This implies that MD trajectories are chaotic, in other words they are extremely sensitive to their starting conditions and hence one-off simulations can never be reproducible.
+MD trajectories are extremely sensitive to their starting conditions and hence one-off simulations can never be reproducible~\cite{coveney2016}.
 Thus, it is a necessity to perform ensemble simulations in order to perform ensemble averaging to obtain thermodynamic quantities in a reliable manner.
@@
 statistically robust) free energy predictions.
-Bhati \textit{et al.} developed the TIES method~\cite{bhati2017} which is based on the use of ensemble methods.
-It produces excellent results for large scale benchmarking datasets for both RBFEs~\cite{bhati2022} and ABFEs~\cite{bhati2025}.
+Bhati \textit{et al.} developed the TIES method~\cite{bhati2017, bhati2022, bhati2025} which is based on the use of ensemble methods.
 It should be noted that the size of ensemble, that is, the number of independent simulations, is an adjustable parameter whose optimal value is system-dependent and will also be affected by the desired level of precision.
-Further, this method is extremely flexible and adaptive.
-For example, one can simply add additional ensemble members as necessary or desired, without having to re-run anything done initially.
-It should also be noted that ensemble simulations are computationally more effective than long simulations.
-On a large supercomputer, all the sampling can be done concurrently obtaining the statistical results in the time it takes to run a single simulation.
+Further, this method is extremely flexible and adaptive allowing one to reduce the computational costs when resources are limited~\cite{wan2023}.
 Studies have shown that, in general, an ensemble size of 5 and 10 members is a reasonable starting point for alchemical RBFE and ABFE calculations, respectively, but recommendations are provided to adjust them based on some rules of thumb~\cite{bhati2022, bhati2025}.
@@
 
-%A recent study demonstrates reproducibility across software packages as well as free energy estimators, but only when ensemble simulations are employed~\cite{wade2022}.
+A recent study demonstrates reproducibility across software packages as well as free energy estimators, but only when ensemble simulations are employed~\cite{wade2022}.
 The authors reported results from two different MD engines (NAMD and OpenMM) as well as two free energy estimators (FEP and MBAR) for a set of 54 protein-ligand complexes which were found to be in good agreement when ensemble simulations were used to control uncertainties.
@@
 
-\subsection{Free energies of hydration or partition coefficients}
+\subsection{Free energies of solvation or partition coefficients}
+As described in Section~\ref{sec:theory-solvation}, the theory and equations for computing solvation free energies are well established.
+Here, we focus instead on practical considerations for applying these methods to hydration and partition coefficient calculations.
+
 \label{subsec:hydration}
@@
 
-\paragraph{The choice of solvent forcefield impacts the calculation}
-A careful choice of forcefield for the organic solvent model, as well as water model is essential.
+\paragraph{The choice of solvent force field impacts the calculation}
+A careful choice of force field for the organic solvent model, as well as water model is essential.
 A critical aspect of these calculations is the preparation ahead of simulation.
 The choice of force field for both the solute and solvents involved is critical.
-Accurate modeling of the solute's solvation behavior hinges on realistic representations of both water and the organic solvent (for partition coefficient calculations).
+Accurate modeling of the solute's solvation behavior hinges on realistic representations of the solvent (for partition coefficient calculations).
 See for example~\cite{bosisio2016blinded,rustenburg2016measuring} for a good discussion of these choices.
-The same level of scrutiny is needed for organic solvents such as octanol, cyclohexane, or others used in partitioning studies, where solvent model parameters may vary substantially between force fields and impact results accordingly.
+Beyond water, The same level of scrutiny is needed for organic solvents such as octanol, cyclohexane, or others used in partitioning studies.
+In a study by Open Force Field, they found that the GAFF\cite{wang2004gaff} and Sage \cite{boothroyd2023sage} forcefields produced results in non-aqueous solutions.
+Care should be taken when choosing the forcefield as solvent model parameters may vary substantially between force fields and impact results accordingly.
 
@@
 In some fixed-charge models, the assigned partial charges depend on the input conformation of the molecule, meaning that poor or inconsistent conformer selection can introduce systematic errors in solvation free energy estimates \cite{jakalian2000, jambeck2003}.
-This is particularly true for methods such as RESP or AM1-BCC, where the geometry of the molecule at the time of charge derivation can influence the electrostatic potential fit \cite{jakalian2000, basma2001, zhang2011, dupradeau2010, osato2025}.
-Conformers with strong intramolecular interactions In these cases, the conformer(s) used for assigning the partial charges should be saved for reproducibility.
+This is particularly true for methods such as RESP\cite{bayly1993resp} or AM1-BCC\cite{jakalian2000}, where the geometry of the molecule at the time of charge derivation can influence the electrostatic potential fit \cite{basma2001, zhang2011, dupradeau2010, osato2025}.
+In these cases, the conformer(s) used for assigning the partial charges should be saved for reproducibility.
 Ensuring consistency between the charge derivation geometry and the initial simulation structure is therefore recommended, or alternatively, using ensemble-averaged charges or polarizable models where appropriate \cite{ren2002,lemkul2016}.
+
+\paragraph{Equilibration of the Solvent}
+Before computing solvation or partition free energies, it is important to ensure that the solvent itself is adequately equilibrated.
+While water typically equilibrates rapidly, especially when prepared using standard protocols, some organic solvents—particularly those with long hydrocarbon chains—can require substantially longer equilibration times\cite{fan2021sampl7}.
+Failing to properly equilibrate the solvent can introduce systematic errors in free energy calculations, as the solvent structure and density may not reflect the true thermodynamic ensemble of the target environment\cite{zhang2017solventmodel}.
 
@@
 Molecules with multiple rotatable bonds, internal hydrogen bonds, or conformational flexibility may exhibit environment-dependent populations that are difficult to sample adequately within the timescale of typical molecular dynamics simulations.
-In some cases, enhanced sampling techniques such as replica exchange or metadynamics may be beneficial, even in hydration or partition coefficient calculations.
+Typically, solvation free energies are computed as absolute transformations, meaning that as the size and flexibility of the ligand increase, the sampling burden grows and convergence of the free energy estimate becomes more challenging.
+In some cases, enhanced sampling techniques such as replica exchange\cite{rest2} or metadynamics\cite{hsu2023alchemical, khuttan2024make} may be beneficial, even in hydration or partition coefficient calculations.
 
@@
 The partition free energy is sensitive to the relative population of conformers in each environment, so it is essential to ensure that sampling in each phase is independently converged.
+It is also important to ensure adequate overlap in the common end states—if the solute is collapsed to two different gas-phase or non-interacting states, the thermodynamic cycle may not hold\cite{fan2021sampl7}.
 Some studies have employed restrained or confining potentials to prevent solutes from drifting to the interface (especially when using periodic systems with both phases present), while others model each phase separately and compute the free energy difference via thermodynamic cycles\cite{turchi2019,bannan2016,petris2021}.
@@
 Hydration and partition coefficients are frequently measured with high precision, enabling rigorous validation of computational protocols.
-This has led to their prominent use in blind challenges such as SAMPL, where participants predict hydration or partition free energies of small molecules using a variety of force fields, solvent models, and sampling strategies.
+Calculating solvation free energies for charged solutes, however, remains particularly challenging, as long-range electrostatics, finite-size artifacts, and treatment of counterions can significantly affect results.
+Moreover, direct experimental determination of ionic hydration free energies is generally not feasible; instead, such values are inferred by decomposing the solvation free energy of ion pairs \cite{mobley2014freesolv}.
+This has led to their prominent use in blind challenges such as SAMPL \cite{skillman2012sampl3, mobley2014sampl4, bergazin2021sampl7}, where participants predict hydration or partition free energies of small molecules using a variety of force fields, solvent models, and sampling strategies.
 These exercises have repeatedly shown that small changes in simulation parameters or protocols can result in significant deviations from experimental results, reinforcing the need for careful setup and validation even in seemingly "simple" solvation systems.
@@
 The 'mixed' intermediate atoms (M) are used as well, but they strictly interpolate between physical and dummy atoms, or vice versa.
+Bhati \textit{et al.}~\cite{bhati2017, bieniek2021} presented a variant of the single coordinate dual topology approach which is used in the TIES method.
+Its distinguishing feature is the application of some filters to remove atoms with differences in atom types and partial charges from the structural MCSS identified.
+This is followed by ensuring identical charges for all MCSS atoms.
+The approach ensures that the final MCSS is both structurally and chemically identical, allowing the use of a unique set of MCSS atoms for both ligands and dual topology for their alchemical parts.
 
@@
 
-\paragraph{Atom mapping}
+\paragraph{Atom mapping} \label{subsec: Atom Mapping}
 Following the choice of system representation and sampling strategy, it might be required to generate a atom mapping.
@@
 RE-EDS~\cite{sidler2016replica}, A-EDS~\cite{perthold2020toward} or Ladybugs~\cite{robo2023fast}) with efficient mapping algorithms.
-There are protoype starting point for solving this problem, but to our knowledge, there is only few published work on this aspect.
+There are starting points, suggesting protoype algorithms for solving this problem, but to our knowledge, there is only few published work on this aspect.
 \cite{petrov2021perturbation}
@@
 For such approaches it is crucial to find a good set of restraints, that keep the ligands colocalized over the time of the simulation .
-This problem is similar to the required restraints in ABFE Calculations (see \ref{ABFErest}).
+This problem is similar to the required restraints in ABFE Calculations (see \ref{sec:ABFE-standard-state} ).
 Few descriptions of possible approaches in RBFE calculations can be found for such approaches like the one suggested by \textit{Baumann et al.}~\cite{baumann2023broadening} based on Boresch style restraints~\cite{boresch2003absolute}, with an elaborate automatic atom selection approach, for the seperated topology approach, or Restraintmaker~\cite{ries2022restraintmaker} for the linked-DCDT approaches, which finds a small set of atoms to be linked via distance restraints, with a small distance to each other.
@@
 
-\paragraph{Dummy atoms}
+\paragraph{Dummy atoms} \label{subsec: Dummy Atoms}
 
@@
 The type of restraints used in SepTop \cite{rocklin2013separated,Baumann_2023} and ATM \cite{Azimi_2022} avoid this problem.
+Additionally, TIES~\cite{bhati2017} circumvents this issue by design as it does not need any restraints.
 
@@
 
-\paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms cancelling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
-Some approaches have attempted to address this using a single coordinate sets and with soft bond approaches ~\cite{clark2019relative,wang2017accurate} or dual coordinate set approach \cite{jespers2019qligfep, ries2022relative} but a general solution is not yet in mainstream use.
+\paragraph{Ring breaking and forming.} \label{subsec: Ring breaking and forming}
+Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms cancelling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
+Some approaches have attempted to address this using a single coordinate sets and with soft bond approaches ~\cite{clark2019relative,wang2017accurate,bhati2017} or dual coordinate set approach \cite{jespers2019qligfep, ries2022relative} but a general solution is not yet in mainstream use.
 Still, FEP+ implements one solution\cite{wang2017accurate}.
@@
 User manuals should carefully checked for how these effects are included if a constrained bond changes in length.
-We do note that if one performs calculations of the changing constraint in both solution and in the binding site, then the errors often cancel substantially, but the cancellation is not guaranteed.
 
@@
 
-\paragraph{Handling the standard state in free energy calculations.}
+\paragraph{Handling the standard state in free energy calculations.} \label{sec:ABFE-standard-state}
 As discussed in Sections \ref{sec:theory-solvation} and \ref{sec:theory-partitioning}, the handling of the standard state in solvation and solvent partitioning free energy calculations is typically straightforward.
@@
 \subsubsection{Changes in net charge can be challenging/problematic.}
-
+\label{subsec:net charge}
 If the net charge of the system changes as the alchemical variable changes during the calculation, this can pose major challenges.
@@
 Another example would be using true apo and holo protein structures as the end states to initialize simulations for absolute binding free energy calculations.
-%Bhati \textit{et al.} recently published two studies directly comparing the equilibrium (EQ) and NEQ approaches at unprecedented scales for both RBFE and ABFE calculations using their ensemble simulation-based TIES method~\cite{wan2023eqvsneq, bhati2025}.
+Bhati \textit{et al.} recently published two studies directly comparing the equilibrium (EQ) and NEQ approaches at unprecedented scales for both RBFE and ABFE calculations using their ensemble simulation-based TIES method~\cite{wan2023eqvsneq, bhati2025}.
 Further, they systematically studied the effect of various parameters involved in setting up calculations for both approaches and recommended optimal protocols in each case for industrial applications.
@@
 In particular, if the rate of change in the free energy estimate is significant when this condition is met, the simulation may not be locally converged, and more sampling may be necessary to determine a stable free energy estimate which is no longer changing significantly over time.
-%A few studies reported systematic variation of simulation lengths using ensemble simulations over an extended range to provide recommendations on appropriate values applicable for most protein-ligand systems.
+A few studies reported systematic variation of simulation lengths using ensemble simulations over an extended range to provide recommendations on appropriate values applicable for most protein-ligand systems.
 The authors suggest using 4 ns and 10 ns for RBFE~\cite{bhati2017, bhati2022, wan2023eqvsneq} and ABFE~\cite{bhati2025} calculations respectively.
-These values may be used as a general guideline.
+These values may be used as a general guideline in cases where the goal is to determine binding affinity for a given starting structure.
+However, these guidelines do not hold if one is interested in studying the kinetics and/or the starting structure available is not representative of the equilibrium structure.
 
@@
 If more samples are needed, the simulations can be resumed, or, started again in different initial conditions enlarging the overall simulation ensemble.
-%Bhati \textit{et al.} recommended such a scheme for large scale RBFE and ABFE calculations using TIES~\cite{bhati2022, bhati2025, wan2023eqvsneq, coveney2022corners}.
-They suggest a set of optimal parameters (ensemble size, simulation length, number of lambda-states, number and length of NEQ transitions, etc.) to be used initially.
-Then, only those systems which do not satisfy pre-defined criteria (uncertainties below 0.5 kcal/mol and overlap between forward and reverse work values in case of NEQ calculations) should be further evaluated using additional calculations (in the way suggested) until these criteria are met.
-Such a scheme optimizes the overall computational efficiency for the full dataset in large scale applications.
+Bhati \textit{et al.} recommended such a scheme for RBFE and ABFE calculations using TIES~\cite{bhati2022, bhati2025, wan2023eqvsneq, coveney2022corners}.
+They suggest a set of optimal parameters (ensemble size, simulation length, number of lambda-states, number and length of NEQ transitions, etc.) to be used initially and suggest pre-defined criteria (uncertainties below 0.5 kcal/mol and overlap between forward and reverse work values in case of NEQ calculations) to decide on the need of additional calculations (in the way suggested) until these criteria are met.
 
@@
 \paragraph{Computing the statistical inefficiency}
-The readers should note here that replicas within ensemble simulations are independently run and hence generate decorrelated samples by default.
-Therefore, the procedures in this section are only relevant for data from a single simulation.
-Further, Wade \textit{et al.} reported that the underestimation of true uncertainty using one-off simulations remains unresolved even on decorrelating the samples~\cite{wade2022}.
+The readers should note here that replicas within ensemble simulations are independently run (with randomly assigned starting velocities even when the starting structure is the same) and hence generate decorrelated samples by default.
+Further, Wade \textit{et al.} reported that the underestimation of true uncertainty using one-off simulations remains unresolved even on decorrelating the samples~\cite{wade2022} and hence ensemble simulations are strongly recommended.
+However, if only a single simulation is available, the procedures below can provide a lower-bound uncertainty estimate.
 
@@
 As such, considering overall computational cost or availability of resources with respect to accuracy desired from computations may influence the choice of sampling and analysis scheme.
-Since sampling all of phase space is prohibitively expensive to satisfy the ergodic theorem, certain considerations are often made to still assess uncertainties in computed free energy estimates.
+%Since sampling all of phase space is prohibitively expensive to satisfy the ergodic theorem, certain considerations are often made to still assess uncertainties in computed free energy estimates.
 If a system exhibits a stable strong binding pose a short number of simulation repeats, sometimes termed ensemble approaches (e.g.
@@
 
-%Statistical mechanics provides a route to calculate macroscopic thermodynamic properties of a system from \textit{ensemble averaging} over microstates.
+Statistical mechanics provides a route to calculate macroscopic thermodynamic properties of a system from \textit{ensemble averaging} over microstates.
 Hitherto, MD practitioners have been applying the ergodic theorem that allows one to equate ensemble averaging to time averaging in the limit of time approaching infinity to predict thermodynamic properties using one-off simulations.
-However, the problem with this approach is that the simulation time required to satisfy this condition has been conveniently dictated by the extent of simulation time deemed accessible in a reasonable period of wall time on available resources.
-It is ignored that this limit is on the order of a Poincarè recurrence time, which is truly astronomical and well beyond the reach of today's supercomputers.
+However, it is often ignored that this limit is on the order of a Poincarè recurrence time, which is truly astronomical and well beyond the reach of today's supercomputers.
 In more rigorous terms, ergodic theory necessitates that, if a dynamical system is to approach equilibrium, it is not sufficient for it to be "ergodic" but it must also be ``mixing''~\cite{coveney2016, MDbook2025}.
@@
 This is manifest in the numerous published reports where authors have highlighted the non-reproducibility of results obtained from one-off simulations and that single simulations heavily underestimate the true uncertainty in predictions~\cite{wan2020fep+, wan2021uq, vassaux2021, gapsys2020large, wade2022, bhati2017, bhati2022, bhati2025, wan2023eqvsneq, bhati2018, bhati2019}.
-The uncertainty must be calculated as the standard error of mean for samples obtained from each replica in the ensemble.
+The uncertainty is calculated as the standard error of mean for samples obtained from each replica in the ensemble.
 The ensemble size is a parameter that needs to be determined empirically for the system of interest and may need to be adjusted; however, some general recommendations (for both EQ and NEQ approaches) based on systematic analyses on a large and diverse dataset of protein-ligand complexes are available in a series of papers published by the Coveney group~\cite{bhati2017, bhati2018, bhati2019, bhati2022, wan2023eqvsneq, bhati2025}.
@@
 It is the influence of these reports that journals such as \textit{JCIM} and \textit{Communications Biology} have now mandated the use of ensemble simulations in computational studies~\cite{wan2023}.
-Coveney and Wan have recently published a book entitled \textit{Molecular Dynamics: Probability and Uncertainty}~\cite{MDbook2025} on this very topic.
-It must also be noted that ``replicas'' within an ensemble simulation are conceptually all members of a continuous distribution which are run concurrently, not in a serial mode.
-Therefore, the use of term ``repeats'' in this context is misleading and must be avoided.
 
@@
 The number of states and amount of sampling should be optimized to minimize the uncertainty in the MBAR free energy estimate, while balancing other key considerations such as computational expense. 
-A less accurate uncertainty estimate may be desirable over a reduced computational expense, increased sampling strategies such as using larger simulation ensembles will provide more accurate uncertainties but will drastically increase the computational cost.
-%However, this process based on one-off simulations grossly underestimates the true uncertainty and hence uncertainties for MBAR also must be obtained using ensemble simulations~\cite{bhati2018, wade2022}.
+%A less accurate uncertainty estimate may be desirable over a reduced computational expense, increased sampling strategies such as using larger simulation ensembles will provide more accurate uncertainties but will drastically increase the computational cost.
+However, this process based on one-off simulations underestimates the true uncertainty and hence uncertainties for MBAR also should be obtained using ensemble simulations~\cite{bhati2018, wade2022}.
 
@@
 Other factors, like system preparation (choice of protonation state, tautomer, counterion presence, salt concentration, etc.) can also substantially impact results~\cite{mobley2017predicting, mobley2017predictinga}, so unless modelers are confident they have these factors correct, sensitivity to these choices may also need to be examined.
-%Incidentally, Vassaux \textit{et al.} performed a systematic study comparing the effect of both aleatoric as well as parametric uncertainties for protein-ligand free energy calculations using NAMD~\cite{vassaux2021}.
+Incidentally, Vassaux \textit{et al.} performed a systematic study comparing the effect of both aleatoric as well as parametric uncertainties for protein-ligand free energy calculations using NAMD~\cite{vassaux2021}.
 They found that, indeed, ensemble simulations are necessary to control both sources of uncertainties arising in simulations, just as is anticipated in the field of uncertainty quantification.
@@
  The precision of NEQ calculations is determined by the overlap between the forward and reverse work distributions corresponding to out-of-equilibrium transitions between the two end-points.
-To estimate the overlap, Wu & Kofke suggest using dissipated work values which can be shown to correspond to the relative entropies of the densities in the phase space.
+To estimate the overlap, Wu \& Kofke suggest using dissipated work values which can be shown to correspond to the relative entropies of the densities in the phase space.
 \cite{wu2005phase} Another approach to assess NEQ convergence relies on calculating the first two moments of the Fermi function in Bennett Acceptance Ratio estimator.
@@
 In practice, analytical error estimate of the BAR estimator \cite{shirts2003equilibrium} gives an indication when the overlap is lacking: the estimated error rapidly grows with the increase in the dissipated work.
-\cite{shirts2005comparison} Other empirical measures for uncertainty have been used, e.g.
-Bhati \textit{et al.} observed that the difference between the inner extremes of the two work value distributions correlates with the unsigned errors in RBFE and ABFE predictions.~\cite{wan2023eqvsneq}
+\cite{shirts2005comparison} Bhati \textit{et al.} recently studied the effect of such overlaps on the accuracy of predicted RBFEs~\cite{wan2023eqvsneq} as well as ABFEs~\cite{bhati2025} but found that overlap (or lack of it) does not capture the reliability of calculations very well.
+They recommended an alternative metric, "distance" defined as the difference between the inner extremes of the two work value distributions, which correlates quite well with the unsigned errors in free energy predictions and hence was found a better indicator of the reliability of NEQ calculations.
 
@@
 \textbf{EG} Wrote the updated Sec.~\ref{sec:theory} and Sec.~\ref{sec:multiple-chemical-states},  updated Sec.~\ref{sec:standardstate-restraints}, co-wrote Sec.~\ref{sec:atm} with SK, and and contributed to the discussion and review of many sections.\\
+\textbf{MG} Contributed to ``Changes in net charge can be challenging/problematic'' in Sec. \ref{subsec:net charge}. \\
 \textbf{DFH} Contributed to Sec.~\ref{sec:software} and helped edit the paper. \\
+\textbf{W-TH}: Wrote Sec.~\ref{sec:configurational_sampling}, updated Sec.~\ref{sec:prerequisites},~\ref{sec:unsettled}, and~\ref{sec:checklist}, and created Fig.
+~\ref{fig:configurational_sampling}.\\
 \textbf{MK}: Contributed to Sec.~\ref{sec:data_analysis}, provided the data for figure~\ref{fig:overlap}, compiled the dataset for Sec.~\ref{sec:benchmark} and helped edit the paper.\\
@@
 \textbf{SP} Wrote Sec.~\ref{sec:software}. \\
+\textbf{BR}: Updated Section \ref{subsec: Atom Mapping}, \ref{subsec: Dummy Atoms}, \ref{subsec: Ring breaking and forming} and supported reviewing for Topologies.\\
 \textbf{AR}: Created figure~\ref{fig:freeenergytrajectories}, contributed to sections~\ref{sec:theory} and~\ref{sec:simulation_protocol_choice}, and helped edit the paper.\\
+\textbf{CO} Contributed to ``Changes in net charge can be challenging/problematic'' in Sec. \ref{subsec:net charge}. \\
 \textbf{JS}: Created Figs.~ \ref{fig:fig_what_is_alchemy},~\ref{fig:fig_binding_thermodynamic_cycle},~\ref{fig:fig_topology},~\ref{fig:fig_mcss},~\ref{fig:fig_absolute_thermodynamic_cycle},~\ref{fig:scatterplot_analysis}, and an initial draft of~\ref{fig:fig_types_of_networks}.
@@
 \textbf{GT}: Contributed to Sec.~\ref{sec:intro} and~\ref{sec:drugdiscovery}, and helped edit the paper.\\
+\textbf{MO} Contributed to Section \ref{subsec:hydration}
 \textbf{HX}: Contributed Sec.~\ref{subsec:accuracy}, to Sec.~\ref{sec:relative-fe-protocol}, and to Sec.~\ref{sec:are-they-good}.
@@
 OJM, WGP, and JWE acknowledge funding from the EPSRC, Astex Pharmaceuticals, and Defence Threat Reduction Agency.
+WTH acknowledges funding from Boehringer Ingelheim, which supports his postdoctoral position at the University of Oxford.
+CO and MG acknowledge the financial support by the Austrian Federal Ministry for Digital and Economic Affairs, the National Foundation for Research, Technology and Development, the Christian Doppler Research Association, BASF, and Boehringer Ingelheim.
 

### 2025-08-30 paper/manuscript.tex
Update to atom mapping section by @RiesBen in response to comments from @sboresch.

@@
 Once a single coordinate set approach is selected, a crucial next step is to identify the common atoms which will not be perturbed between the two molecules (respectively end-states).
-Rigorously, this process typically comprises a MCSS search of the molecules involved to identify the common substructure.
-Specifically, with a single coordinate set approach in mind, atom types are allowed to change, so a permissive MCSS search can be done.
-This however requires either careful implementation of atom mass changes, bond length etc.
-or accepting the introduction of potential systematic error, that is in many cases negligible due to the cancellation of errors.
-\cite{ries2024kartograf}
+Rigorously, this process typically comprises a MCSS search of the molecules involved to identify the common substructure.\cite{ries2024kartograf}
 

### 2025-09-08 paper/manuscript.tex
Responses to reviewer comments on uncertainty quantification changes by @apbhati1

@@
 Bhati \textit{et al.}~\cite{bhati2017, bieniek2021} presented a variant of the single coordinate dual topology approach which is used in the TIES method.
-Its distinguishing feature is the application of some filters to remove atoms with differences in atom types and partial charges from the structural MCSS identified.
-This is followed by ensuring identical charges for all MCSS atoms.
-The approach ensures that the final MCSS is both structurally and chemically identical, allowing the use of a unique set of MCSS atoms for both ligands and dual topology for their alchemical parts.
+Its distinguishing feature is the application of some filters to remove atoms with differences in atom types and partial charges from the structural MCSS identified to ensure chemical similarity between all atom pairs in the final MCSS across both end-points.
+This is followed by ensuring identical charges for all MCSS atoms by averaging between the two end-points.
+The resultant change in the sum of atomic charges for all MCSS atoms is uniformly distributed across all atoms in the corresponding alchemical region conserving the overall ligand charge for both end-points.
+The approach ensures that the final MCSS is both structurally and chemically identical within thresholds, allowing the use of a unique set of MCSS atoms for both ligands and dual topology for their alchemical parts.
 
@@
 Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms cancelling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
-Some approaches have attempted to address this using a single coordinate sets and with soft bond approaches ~\cite{clark2019relative,wang2017accurate,bhati2017} or dual coordinate set approach \cite{jespers2019qligfep, ries2022relative} but a general solution is not yet in mainstream use.
+Some approaches have attempted to address this using a single coordinate sets and with soft bond approaches ~\cite{clark2019relative,wang2017accurate} or dual coordinate set approach \cite{jespers2019qligfep, ries2022relative} but a general solution is not yet in mainstream use.
 Still, FEP+ implements one solution\cite{wang2017accurate}.
+TIES also overcomes this issue by using dual topology and ensuring that MCSS atoms are structurally and chemically identical~\cite{bhati2017}.
 
@@
 This can be done through the uncertainty estimator built into certain analysis tools such as MBAR, or through more general statistical tools like bootstrap sampling.
-It should be noted however, that uncertainty estimates have the same limitations as other metrics of convergence, as they are only an uncertainty based on the phase space sampled so far in a simulation, and cannot account for states not sampled, and it is worth considering that they will be an underestimation of the true uncertainty~\cite{wan2020fep+, wan2021uq, vassaux2021, gapsys2020large, wade2022, bhati2017, bhati2022, bhati2025, wan2023eqvsneq, bhati2018, bhati2019}.
+It should be noted however, that uncertainty estimates have the same limitations as other metrics of convergence, as they are only an uncertainty based on the phase space sampled so far in a simulation, and cannot account for states not sampled, and it is worth considering that they will be an underestimation of the true uncertainty~\cite{wan2020fep+, wan2021uq, vassaux2021, gapsys2020large, wade2022, bhati2017, bhati2022, bhati2025, wan2023eqvsneq, bhati2018, bhati2019, cournia2017relative, gapsys2021accurate, baumann2021, caves1998}.
 A target statistical uncertainty should be chosen at the onset of the simulation to avoid excessively long simulations, or falling into the trap of running until the free energy estimate is "good enough," which is subjective and has no defined criteria.
@@
 \paragraph{Computing the statistical inefficiency}
-The readers should note here that replicas within ensemble simulations are independently run (with randomly assigned starting velocities even when the starting structure is the same) and hence generate decorrelated samples by default.
-Further, Wade \textit{et al.} reported that the underestimation of true uncertainty using one-off simulations remains unresolved even on decorrelating the samples~\cite{wade2022} and hence ensemble simulations are strongly recommended.
-However, if only a single simulation is available, the procedures below can provide a lower-bound uncertainty estimate.
+Replicas in ensemble simulations are initialized with velocities drawn independently from the Maxwell–Boltzmann distribution and hence generate decorrelated samples by default.
+Ensemble approaches are therefore strongly preferred for uncertainty estimation, as single simulations remain constrained by the timescales they can access and may therefore underestimate the true uncertainty.
+Further, Wade \textit{et al.} reported that the underestimation of true uncertainty using one-off simulations remains unresolved even on decorrelating the samples~\cite{wade2022}.
+When only one simulation is available, the procedures described below provide a lower-bound estimate of the uncertainty.
 
@@
 Statistical mechanics provides a route to calculate macroscopic thermodynamic properties of a system from \textit{ensemble averaging} over microstates.
-Hitherto, MD practitioners have been applying the ergodic theorem that allows one to equate ensemble averaging to time averaging in the limit of time approaching infinity to predict thermodynamic properties using one-off simulations.
-However, it is often ignored that this limit is on the order of a Poincarè recurrence time, which is truly astronomical and well beyond the reach of today's supercomputers.
+Hitherto, MD practitioners have been assuming the validity of the ergodic hypothesis that allows one to equate ensemble averaging to time averaging in the limit of time approaching infinity to predict thermodynamic properties using one-off simulations.
+However, it is often ignored that this limit is on the order of a Poincaré recurrence time, which is truly astronomical and well beyond the reach of today's supercomputers.
 In more rigorous terms, ergodic theory necessitates that, if a dynamical system is to approach equilibrium, it is not sufficient for it to be "ergodic" but it must also be ``mixing''~\cite{coveney2016, MDbook2025}.
@@
 This is manifest in the numerous published reports where authors have highlighted the non-reproducibility of results obtained from one-off simulations and that single simulations heavily underestimate the true uncertainty in predictions~\cite{wan2020fep+, wan2021uq, vassaux2021, gapsys2020large, wade2022, bhati2017, bhati2022, bhati2025, wan2023eqvsneq, bhati2018, bhati2019}.
-The uncertainty is calculated as the standard error of mean for samples obtained from each replica in the ensemble.
+The uncertainty is calculated as the standard error of the mean for estimates obtained from each replica in the ensemble.
 The ensemble size is a parameter that needs to be determined empirically for the system of interest and may need to be adjusted; however, some general recommendations (for both EQ and NEQ approaches) based on systematic analyses on a large and diverse dataset of protein-ligand complexes are available in a series of papers published by the Coveney group~\cite{bhati2017, bhati2018, bhati2019, bhati2022, wan2023eqvsneq, bhati2025}.
@@
 The TIES method~\cite{bhati2017} employs a different scheme to obtain the overall variance of the integral.
-It assumes that the integrand is a Gaussian random process and hence the principle of stochastic integration must be employed to propagate the uncertainties in this integral.
+It assumes that the integrand is a Gaussian random process and hence the thermodynamic integral can be viewed as a so-called stochastic integral, which forms the basis of uncertainty propagation in this integral.
 The following equation is used to compute the overall variance of the integral combining variance at each $\lambda$ state.

### 2025-09-17 paper/manuscript.tex
Includes updates to software packages and tweaks to the setup of hydration free energies from @IAlibay, responses to reviewer comments from @apbhati, and updates to the author lists.

@@
 \author[1*]{Antonia S. J. S. Mey}
+\author[22]{Irfan Alibay}
 \author[2]{Bryce K. Allen}
@@
 \author[3*]{John D. Chodera}
+\author[26]{Finlay Clark}
 \author[11, 12, 13]{Peter V. Coveney}
@@
 \author[20]{Wei-Tse Hsu}
+\author[25]{Sheenam Khuttan}
 \author[1,10]{Maximilian Kuhn}
@@
 \affil[24]{Christian Doppler Laboratory for Molecular Informatics in the Biosciences, BOKU University, 1190 Vienna, Austria}
+\affil[25]{SandboxAQ, Palo Alto, California 94301, United States}
+\affil[26]{School of Natural and Environmental Sciences, Newcastle University, Newcastle upon Tyne NE1 7RU, United Kingdom}
 
@@
 \orcid{Chris Oostenbrink}{0000-0002-4232-2556}
+\orcid{Irfan Alibay}{0000-0001-5787-9130}
+\orcid{Sheenam Khuttan}{0000-0002-8873-6359}
+\orcid{Finlay Clark}{0000-0003-0474-5475}
 
@@
 A careful choice of force field for the organic solvent model, as well as water model is essential.
-A critical aspect of these calculations is the preparation ahead of simulation.
+A critical aspect of these calculations is the preparation of the system ahead of simulation.
 The choice of force field for both the solute and solvents involved is critical.
@@
 Beyond water, The same level of scrutiny is needed for organic solvents such as octanol, cyclohexane, or others used in partitioning studies.
-In a study by Open Force Field, they found that the GAFF\cite{wang2004gaff} and Sage \cite{boothroyd2023sage} forcefields produced results in non-aqueous solutions.
+In a study by Open Force Field, they found that the general AMBER force field (GAFF)\cite{wang2004gaff} and Open Force Field Sage \cite{boothroyd2023sage} force fields produced results in non-aqueous solutions.
 Care should be taken when choosing the forcefield as solvent model parameters may vary substantially between force fields and impact results accordingly.
@@
 In some fixed-charge models, the assigned partial charges depend on the input conformation of the molecule, meaning that poor or inconsistent conformer selection can introduce systematic errors in solvation free energy estimates \cite{jakalian2000, jambeck2003}.
-This is particularly true for methods such as RESP\cite{bayly1993resp} or AM1-BCC\cite{jakalian2000}, where the geometry of the molecule at the time of charge derivation can influence the electrostatic potential fit \cite{basma2001, zhang2011, dupradeau2010, osato2025}.
+This is particularly true for methods such as Restrained Electrostatic Potential (RESP)\cite{bayly1993resp} or AM1 population charges with bond-charge correction (AM1-BCC)\cite{jakalian2000}, where the geometry of the molecule at the time of charge derivation can influence the electrostatic potential fit \cite{basma2001, zhang2011, dupradeau2010, osato2025}.
 In these cases, the conformer(s) used for assigning the partial charges should be saved for reproducibility.
@@
 Bhati \textit{et al.}~\cite{bhati2017, bieniek2021} presented a variant of the single coordinate dual topology approach which is used in the TIES method.
-Its distinguishing feature is the application of some filters to remove atoms with differences in atom types and partial charges from the structural MCSS identified to ensure chemical similarity between all atom pairs in the final MCSS across both end-points.
-This is followed by ensuring identical charges for all MCSS atoms by averaging between the two end-points.
-The resultant change in the sum of atomic charges for all MCSS atoms is uniformly distributed across all atoms in the corresponding alchemical region conserving the overall ligand charge for both end-points.
-The approach ensures that the final MCSS is both structurally and chemically identical within thresholds, allowing the use of a unique set of MCSS atoms for both ligands and dual topology for their alchemical parts.
+It involves a unique post-processing of the initial MCSS to ensure that the final MCSS is both structurally and chemically identical and has a single set of coordinates while maintaining dual topology for the alchemical parts.
 
@@
 The type of restraints used in SepTop \cite{rocklin2013separated,Baumann_2023} and ATM \cite{Azimi_2022} avoid this problem.
-Additionally, TIES~\cite{bhati2017} circumvents this issue by design as it does not need any restraints.
+Additionally, SCDT approaches, such as TIES~\cite{bhati2017}, [MENTION OTHERS TOO], etc., circumvent this issue by design as they do not need any restraints to hold MCSS together between the two end-points.
 
@@
 Still, FEP+ implements one solution\cite{wang2017accurate}.
-TIES also overcomes this issue by using dual topology and ensuring that MCSS atoms are structurally and chemically identical~\cite{bhati2017}.
+TIES overcomes this issue by ensuring that MCSS atoms are structurally and chemically identical while allowing for partial rings~\cite{bhati2017, bieniek2021}.
 
@@
 However, it is often ignored that this limit is on the order of a Poincaré recurrence time, which is truly astronomical and well beyond the reach of today's supercomputers.
-In more rigorous terms, ergodic theory necessitates that, if a dynamical system is to approach equilibrium, it is not sufficient for it to be "ergodic" but it must also be ``mixing''~\cite{coveney2016, MDbook2025}.
-Rate processes and equilibration times are a property of the distribution function, not of individual trajectories.
+In more rigorous terms, ergodic theory necessitates that, if a dynamical system is to approach equilibrium, it is not sufficient for it to be ``ergodic'' but it must also be ``mixing'' (these terms refer to the different classes of dynamical systems provided by ergodic theory; please see the references for details)~\cite{coveney2016, MDbook2025}.
+Rate processes (a course of change within a system that occurs over time) and equilibration times are a property of the distribution function, not of individual trajectories.
 Consequently, neighbouring trajectories in the underlying phase space diverge exponentially which means that a single trajectory can never capture the true behaviour of a system.
@@
 \cite{shirts2005comparison} Bhati \textit{et al.} recently studied the effect of such overlaps on the accuracy of predicted RBFEs~\cite{wan2023eqvsneq} as well as ABFEs~\cite{bhati2025} but found that overlap (or lack of it) does not capture the reliability of calculations very well.
-They recommended an alternative metric, "distance" defined as the difference between the inner extremes of the two work value distributions, which correlates quite well with the unsigned errors in free energy predictions and hence was found a better indicator of the reliability of NEQ calculations.
+They recommended an alternative metric, ``distance'' defined as the difference between the inner extremes of the two work value distributions, which correlates quite well with the unsigned errors in free energy predictions and hence was found a better indicator of the reliability of NEQ calculations.
 
@@
 The PERT module can be used to define initial and final states and define the intermediate lambda points.
-FREN and BAR modules can be used to analyze the data after the MD run.
+FREN and BAR modules can be used to analyse the data after the MD run.
 Lambda-dynamics-based free energy calculation can be carried out using the BLOCK module.
+GPU optimized implementation of CHARMM's lambda dynamics is also implemented in the new package BLaDE \cite{hayes2021blade}.
+A number of energy interpolation schemes have been implemented in the new GPU package apoCHARMM \cite{prasad2025apocharmm}.
 	\item \href{https://ambermd.org/}{AMBER}, including its GPU-accelerated free energy engine~\cite{Lee_JChemTheoryComput_2017_v13_p3077,Giese_JChemTheoryComput_2018_v14_p1564,Lee_JChemInfModel_2018_v58_p2043} that supports a wide range of alchemical free energy methods~\cite{Tsai_JChemTheoryComput_2023_v19_p640,lee2023aces,Zhang_JChemTheoryComput_2024_v20_p3935}, workflows~\cite{Ganguly_JChemInfModel_2022_v62_p6069} and analysis tools~\cite{Giese_JChemInfModel_2025_v65_p5273,Case_JChemInfModel_2023_v63_p6183}.
@@
 It also offers a VMD based plugin for the computation of collective variables~\cite{bonomi2019promoting}.
-	\item \href{https://biosimspace.org/}{BioSimSpace} is a multiscale molecular simulation framework, written to allow computational modellers to quickly prototype and develop new algorithms for molecular simulation and molecular design~\cite{hedges2019biosimspace}.
-	\item \href{https://siremol.org/}{Sire} is a multiscale, molecular simulation framework that provides several applications, including SOMD, an MD/MC code for performing FEP calculations via an interface to OpenMM.
+	\item \href{https://biosimspace.org/}{BioSimSpace} is a multiscale molecular simulation framework, written to allow computational modellers to quickly prototype and develop new algorithms for molecular simulation and molecular design~\cite{hedges2019biosimspace, hedges2023suite}.
+	\item \href{https://siremol.org/}{Sire} is a multiscale, molecular simulation framework that provides several applications, including SOMD, an MD/MC code for performing FEP calculations via an interface to OpenMM~\cite{woods2024sire}.
 	\item \href{http://getyank.org/latest/index.html}{YANK} is a tool developed by John Chodera and group on the top of OpenMM MD package.
 It allows the users to write their inputs in easy-to-use YAML format.
+        \item \href{https://github.com/choderalab/openmmtools} provides tooling for creating alchemical simulations using OpenMM.
+These include tools for alchemically modifying OpenMM Systems, running multistate simulations and adding alchemical restraints.
 	\item \href{http://www.gromacs.org/}{GROMACS} is a molecular simulation package with a significant number of free energy methods implementations.
@@
     \item \href{http://www.ties-service.org}{TIES} enables the usage of NAMD and OpenMM for setting up alchemical free energy ensemble simulations \url{https://github.com/UCL-CCS/TIES_MD} \cite{bieniek2023}.
+    \item \href{https://openfree.energy/}{Open Free Energy} is a platform which offers and allows the creations of alchemical simulation workflows for both absolute and relative binding free energies.
+Current offerings, notably include a Perses-based \url{https://github.com/choderalab/perses/tree/main/perses} hybrid topology method using OpenMM.
 	\end{itemize}
@@
 	\item \href{http://pmx.mpibpc.mpg.de/instructions.html}{PMX}: Setup of perturbation maps, perturbed topologies and input coordinates for GROMACS simulations at \url{https://github.com/deGrootLab/pmx}.
-	\item \href{https://github.com/MobleyLab/Lomap}{Lomap/Lomap2} : Relative alchemical transformation graph planning for setting up perturbation networks~\cite{liu2013lead}.
+	\item \href{https://github.com/OpenFreeEnergy/Lomap}{Lomap/Lomap2} : Relative alchemical transformation graph planning for setting up perturbation networks~\cite{liu2013lead}.
+        \item \href{https://github.com/OpenFreeEnergy/kartograf} 
+    Kartograf: An geometry-based atom mapper for setting up relative free energies with a conserved core~\cite{ries2024kartograf}.
+        \item \href{https://github.com/OpenFreeEnergy/Konnektor}
+    Konnektor: Tooling to aid in the creation of alchemical perrturbation networks~\cite{}.
 	\item \href{http://www.charmm-gui.org/}{CHARMM-GUI} is a web-based tool for setting up a variety of MD simulations.
@@
 	\item \href{https://fesetup.readthedocs.io/en/latest/introduction.html}{FESetup} has been developed primarily to setup calculations in AMBER, GROMACS and SIRE~\cite{loeffler2015fesetup}.
+        \item 
+    \href{https://github.com/cbc-univie/transformato}{transformato} is a package for setting up single topology alchemical transformations to be run using CHARMM or OpenMM.
     \item \href{http://www.ties-service.org}{TIES} allows the setup of perturbation maps, hybrid topologies and structures as well as simulation-ready models for NAMD and OpenMM \url{https://github.com/UCL-CCS/TIES} \cite{bieniek2023}.
@@
 	\url{https://github.com/choderalab/pymbar} \cite{shirts2008statisticallya}.
-	\item Arsenic: Standardising alchemical free energy analysis \url{https://github.com/openforcefield/Arsenic}
-	\item Free Energy Workflows: Sire-specific free energy map analysis using weighted path averages \url{https://github.com/michellab/freenrgworkflows}.
-    \item \href{http://www.ties-service.org}{TIES}: Perform statistical analysis of the output of alchemical ensemble simulations and report reliable predictions along with statistical uncertainties for a variety of free energy estimators \url{https://github.com/UCL-CCS/TIES} \cite{bieniek2023}.
+	\item Cinnabar: Standardising alchemical free energy analysis.
+Includes tools for calculating network-wide analyses and deriving absolute free energies using a maximum-likelihood estimator \url{https://github.com/OpenFreeEnergy/cinnabar}.
+	\item Free Energy Workflows: Sire-specific free energy map analysis using weighted path averages 
+        \item \href{https://gitlab.com/RutgersLBSR/fe-toolkit} FE-Toolkit: collection of packages for free energy analysis, including network-wide MBAR analysis~/cite{}.
+    \url{https://github.com/michellab/freenrgworkflows}.
+    \item \href{http://www.ties-service.org}{TIES}: Perform statistical analysis of the output of alchemical ensemble simulations and report reliable predictions along with statistical uncertainties for a variety of free energy estimators \url{https://github.com/UCL-CCS/TIES_MD} \cite{bieniek2023}.
 	\end{itemize}
@@
 \textbf{BA}: Helped write the uncertainty estimation, stopping conditions, and output analysis sections and created figure ~\ref{fig:convergence_forward_reverse}.\\
+\textbf{IA} Reviewed and contributed to Sec. ~\ref{sec:prerequisites}, ~\ref{subsec:hydration}, and ~\ref{sec:multiple-chemical-states}. Updated Sec. ~\ref{sec:software}.
 \textbf{APB} Contributed updates to the discussion of uncertainty quantification.\\
@@
 \\
+\textbf{FC}: Coordinated the v2.0 updates to the article (TODO: be more specific?/ mention ABFE updates?). \\
 \textbf{PVC} Contributed updates to the discussion of uncertainty quantification.\\
@@
 ~\ref{fig:configurational_sampling}.\\
+\textbf{SK}: Contributed to Sec. \ref{sec:atm}.\\
 \textbf{MK}: Contributed to Sec.~\ref{sec:data_analysis}, provided the data for figure~\ref{fig:overlap}, compiled the dataset for Sec.~\ref{sec:benchmark} and helped edit the paper.\\
@@
 JM is a current member of the Scientific Advisory Board of Cresset. 
+SK is employed by SandboxAQ.
 MK is employed by Cresset who commercially distribute a software for performing alchemical free energy calculations. MRS is a Open Science Fellow and consultant for Silicon Therapeutics.
@@
 The content is solely the responsibility of the authors and does not necessarily represent the official views of the National Institutes of Health.
+FC acknowledges support from a UKRI Future Leaders Fellowship (Grant MR/T019654/1).
 OJM, WGP, and JWE acknowledge funding from the EPSRC, Astex Pharmaceuticals, and Defence Threat Reduction Agency.
@@
 CO and MG acknowledge the financial support by the Austrian Federal Ministry for Digital and Economic Affairs, the National Foundation for Research, Technology and Development, the Christian Doppler Research Association, BASF, and Boehringer Ingelheim.
+IA acknowledges funding from the Open Free Energy Fund. A list of financially contributing Open Free Energy members can be found at \url{https://openfree.energy/about/}.
 

### 2025-09-21 paper/manuscript.tex
Updates to free energies of hydration/ partition coefficients by @megosato in response to review
Changes to uncertainty quantification sections from @hbruce-beep while reviewing

@@
 
-MD trajectories are extremely sensitive to their starting conditions and hence one-off simulations can never be reproducible~\cite{coveney2016}.
-Thus, it is a necessity to perform ensemble simulations in order to perform ensemble averaging to obtain thermodynamic quantities in a reliable manner.
-Coveney and Wan recently published a book on the probabilistic foundations of MD and how this approach naturally allows for the treatment of uncertainties that arise in the use of the methodology~\cite{MDbook2025}.
-In practice, in many cases the simplest and most effective way of performing ensemble simulations is to vary only the initial velocities as these yield reproducible (i.e.
-statistically robust) free energy predictions.
-Bhati \textit{et al.} developed the TIES method~\cite{bhati2017, bhati2022, bhati2025} which is based on the use of ensemble methods.
-It should be noted that the size of ensemble, that is, the number of independent simulations, is an adjustable parameter whose optimal value is system-dependent and will also be affected by the desired level of precision.
-Further, this method is extremely flexible and adaptive allowing one to reduce the computational costs when resources are limited~\cite{wan2023}.
+MD trajectories are extremely sensitive to their starting conditions and hence one-off simulations can never be reproducible~\cite{coveney2016, MDbook2025}.
+Thus, it is recommended to perform ensemble simulations in order to perform ensemble averaging to obtain thermodynamic quantities in a reliable manner.
+Bhati \textit{et al.} developed the TIES method~\cite{bhati2017, bhati2022, bhati2025} which is based on the use of ensemble methods, where the ensemble size can be adjusted based on available computational resource and required precision~\cite{wan2023}.
 Studies have shown that, in general, an ensemble size of 5 and 10 members is a reasonable starting point for alchemical RBFE and ABFE calculations, respectively, but recommendations are provided to adjust them based on some rules of thumb~\cite{bhati2022, bhati2025}.
 The sample standard deviation will give a crude estimate of the reliability of the estimates, and whether the precision is sufficient for the problem at hand.
-When practical, a more stringent test is to use different input coordinates for each repeat, or at least different starting velocities.
-Repeats or ensemble-based simulations will allow for a robust error statistic on single free energy calculations and are typically expected in the order of 1 kcal/mol or better.
 
@@
 Note that these types of statistical differences concern calculations carried out with a single software package, but simulation package variations can introduce additional discrepancies.
-Such issues of reproducibility of free energy calculations across different simulation packages have attracted attention recently~\cite{loeffler2018reproducibility,rizzi2019sampl6, wade2022}.
+Such issues of reproducibility of free energy calculations across different simulation packages have attracted attention ~\cite{loeffler2018reproducibility,rizzi2019sampl6, wade2022, herz2025modular}.
 Greater variability is expected between packages due to methodological differences such as integrators, thermostats, barostats, treatment of long-range electrostatics, and potentially other factors.
@@
 Alchemical free energy protocols as discussed below (Sec.~\ref{sec:simulation_protocol_choice}) are defined for a specific type of free energy calculation, i.e.
-a free energy of binding or a free energy of hydration.
+a free energy of hydration or a free energy of binding.
 Different types of simulations require different choices for ligands, solvent, and host molecules (in the case of the estimation of free energies of binding).
+
+\subsection{Free energies of solvation or partition coefficients}
+As described in Section~\ref{sec:theory-solvation}, the theory and equations for computing solvation free energies are well established.
+Here, we focus instead on practical considerations for applying these methods to hydration and partition coefficient calculations.
+
+\label{subsec:hydration}
+Preliminary considerations necessary for using free energy methods to compute partition coefficients are generally more straightforward.
+For example, a 3D minimised structure of a solute can be generated with a simple tool such as openBabel and solvated to prepare the input to compute a free energy of hydration~\cite{oboyle2011open}.
+Despite their relative simplicity, care should still be taken when preparing these systems to avoid introducing systematic errors.
+
+\paragraph{The choice of solvent force field may impact the calculation}
+A careful choice of force field for the organic solvent model, as well as water model is essential.
+A critical aspect of these calculations is the preparation of the system ahead of simulation.
+The choice of force field for both the solute and solvents involved is critical.
+Accurate modeling of the solute's solvation behavior hinges on realistic representations of the solvent (for partition coefficient calculations).
+See for example~\cite{bosisio2016blinded,rustenburg2016measuring} for a good discussion of these choices.
+Beyond water, the same level of scrutiny is needed for organic solvents such as octanol, cyclohexane, or others used in partitioning studies.
+For example, in a study by Open Force Field, they found that the general AMBER force field (GAFF)\cite{wang2004gaff} and Open Force Field Sage \cite{boothroyd2023sage} force fields produced results in non-aqueous solutions.
+Care should be taken when choosing the forcefield as solvent model parameters may vary substantially between force fields \cite{jambeck2003, kadaoluwa_2021, mobley_2007charge} and impact results accordingly.
+
+\paragraph{Charge models are non-trivial}
+Modeling the partial charges of the solute and solvent molecules is also highly important.
+In some fixed-charge models, the assigned partial charges depend on the input conformation of the molecule, meaning that poor or inconsistent conformer selection can introduce systematic errors in solvation free energy estimates \cite{jakalian2000, jambeck2003}.
+This is particularly true for methods such as Restrained Electrostatic Potential (RESP)\cite{bayly1993resp} or AM1 population charges with bond-charge correction (AM1-BCC)\cite{jakalian2000}, where the geometry of the molecule at the time of charge derivation can influence the electrostatic potential fit \cite{basma2001, zhang2011, dupradeau2010, osato2025}.
+In these cases, the conformer(s) used for assigning the partial charges should be saved for reproducibility.
+Ensuring consistency between the charge derivation geometry and the initial simulation structure is therefore recommended, or alternatively, using ensemble-averaged charges or polarizable models where appropriate \cite{ren2002,lemkul2016}.
+
+\paragraph{Equilibration of the Solvent}
+Before computing solvation or partition free energies, it is important to ensure that the solvent itself is adequately equilibrated.
+While water typically equilibrates rapidly, especially when prepared using standard protocols, some organic solvents—particularly those with long hydrocarbon chains—can require substantially longer equilibration times\cite{fan2021sampl7}.
+Failing to properly equilibrate the solvent can introduce systematic errors in free energy calculations, as the solvent structure and density may not reflect the true thermodynamic ensemble of the target environment\cite{zhang2017solventmodel}.
+
+\paragraph{Validation through experimental comparisons}  
+Finally, experimental comparisons in these systems are often more direct than in binding free energies.
+Hydration and partition coefficients are frequently measured with high precision, enabling rigorous validation of computational protocols.
+Calculating solvation free energies for charged solutes, however, remains particularly challenging, as long-range electrostatics, finite-size artifacts, and treatment of counterions can significantly affect results.
+Moreover, direct experimental determination of ionic hydration free energies is generally not feasible; instead, such values are inferred by decomposing the solvation free energy of ion pairs \cite{mobley2014freesolv}.
+This has led to their prominent use in blind challenges such as SAMPL \cite{skillman2012sampl3, mobley2014sampl4, bergazin2021sampl7}, where participants predict hydration or partition free energies of small molecules using a variety of force fields, solvent models, and sampling strategies.
+These exercises have repeatedly shown that small changes in simulation parameters or protocols can result in significant deviations from experimental results, reinforcing the need for careful setup and validation even in seemingly "simple" solvation systems.
 
@@
 It is therefore generally important to carefully select input coordinates to obtain satisfactory results.
-The following questions may be relevant before diving into the simulation setup.
+The above considerations in Section \ref{subsec:hydration} on preparing solvation free energy and partition coefficient calculations also apply to binding free energy calculations.
+In addition to the above, the following questions may be relevant before diving into the binding free energy simulation setup.
 
@@
 One may have to account for the free energy cost of forming a binding site in the target receptor which may not be feasible on alchemical simulation timescales.
-
-\subsection{Free energies of solvation or partition coefficients}
-As described in Section~\ref{sec:theory-solvation}, the theory and equations for computing solvation free energies are well established.
-Here, we focus instead on practical considerations for applying these methods to hydration and partition coefficient calculations.
-
-\label{subsec:hydration}
-Preliminary considerations necessary for using free energy methods to compute partition coefficients are generally more straightforward.
-For example, a 3D minimised structure of a solute can be generated with a simple tool such as openBabel and solvated to prepare the input to compute a free energy of hydration~\cite{oboyle2011open}.
-Despite their relative simplicity, care should still be taken when preparing these systems to avoid introducing systematic errors.
-
-\paragraph{The choice of solvent force field impacts the calculation}
-A careful choice of force field for the organic solvent model, as well as water model is essential.
-A critical aspect of these calculations is the preparation of the system ahead of simulation.
-The choice of force field for both the solute and solvents involved is critical.
-Accurate modeling of the solute's solvation behavior hinges on realistic representations of the solvent (for partition coefficient calculations).
-See for example~\cite{bosisio2016blinded,rustenburg2016measuring} for a good discussion of these choices.
-Beyond water, The same level of scrutiny is needed for organic solvents such as octanol, cyclohexane, or others used in partitioning studies.
-In a study by Open Force Field, they found that the general AMBER force field (GAFF)\cite{wang2004gaff} and Open Force Field Sage \cite{boothroyd2023sage} force fields produced results in non-aqueous solutions.
-Care should be taken when choosing the forcefield as solvent model parameters may vary substantially between force fields and impact results accordingly.
-
-\paragraph{Charge models are non-trivial}
-Modeling the partial charges of the solute and solvent molecules is also highly important.
-In some fixed-charge models, the assigned partial charges depend on the input conformation of the molecule, meaning that poor or inconsistent conformer selection can introduce systematic errors in solvation free energy estimates \cite{jakalian2000, jambeck2003}.
-This is particularly true for methods such as Restrained Electrostatic Potential (RESP)\cite{bayly1993resp} or AM1 population charges with bond-charge correction (AM1-BCC)\cite{jakalian2000}, where the geometry of the molecule at the time of charge derivation can influence the electrostatic potential fit \cite{basma2001, zhang2011, dupradeau2010, osato2025}.
-In these cases, the conformer(s) used for assigning the partial charges should be saved for reproducibility.
-Ensuring consistency between the charge derivation geometry and the initial simulation structure is therefore recommended, or alternatively, using ensemble-averaged charges or polarizable models where appropriate \cite{ren2002,lemkul2016}.
-
-\paragraph{Equilibration of the Solvent}
-Before computing solvation or partition free energies, it is important to ensure that the solvent itself is adequately equilibrated.
-While water typically equilibrates rapidly, especially when prepared using standard protocols, some organic solvents—particularly those with long hydrocarbon chains—can require substantially longer equilibration times\cite{fan2021sampl7}.
-Failing to properly equilibrate the solvent can introduce systematic errors in free energy calculations, as the solvent structure and density may not reflect the true thermodynamic ensemble of the target environment\cite{zhang2017solventmodel}.
-
-\paragraph{Conformational sampling can impact solvation and partition free energies}
-Moreover, while sampling challenges are often perceived to be minor for small molecules, they can still be non-trivial.
-Even subtle degrees of freedom such as hydroxyl group orientations (as in neutral carboxylic acids) or intramolecular hydrogen bonding can influence the calculated free energy difference between two environments\cite{klimovich2010predicting, lim2019assessing}.
-Molecules with multiple rotatable bonds, internal hydrogen bonds, or conformational flexibility may exhibit environment-dependent populations that are difficult to sample adequately within the timescale of typical molecular dynamics simulations.
-Typically, solvation free energies are computed as absolute transformations, meaning that as the size and flexibility of the ligand increase, the sampling burden grows and convergence of the free energy estimate becomes more challenging.
-In some cases, enhanced sampling techniques such as replica exchange\cite{rest2} or metadynamics\cite{hsu2023alchemical, khuttan2024make} may be beneficial, even in hydration or partition coefficient calculations.
-
-In partition coefficient calculations, the situation becomes even more complex, as two distinct environments must be adequately sampled: typically water and a nonpolar organic solvent.
-These systems not only differ in their dielectric properties and hydrogen bonding behavior but may also yield different dominant solute conformers.
-The partition free energy is sensitive to the relative population of conformers in each environment, so it is essential to ensure that sampling in each phase is independently converged.
-It is also important to ensure adequate overlap in the common end states—if the solute is collapsed to two different gas-phase or non-interacting states, the thermodynamic cycle may not hold\cite{fan2021sampl7}.
-Some studies have employed restrained or confining potentials to prevent solutes from drifting to the interface (especially when using periodic systems with both phases present), while others model each phase separately and compute the free energy difference via thermodynamic cycles\cite{turchi2019,bannan2016,petris2021}.
-
-\paragraph{Validation through experimental comparisons}  
-Finally, experimental comparisons in these systems are often more direct than in binding free energies.
-Hydration and partition coefficients are frequently measured with high precision, enabling rigorous validation of computational protocols.
-Calculating solvation free energies for charged solutes, however, remains particularly challenging, as long-range electrostatics, finite-size artifacts, and treatment of counterions can significantly affect results.
-Moreover, direct experimental determination of ionic hydration free energies is generally not feasible; instead, such values are inferred by decomposing the solvation free energy of ion pairs \cite{mobley2014freesolv}.
-This has led to their prominent use in blind challenges such as SAMPL \cite{skillman2012sampl3, mobley2014sampl4, bergazin2021sampl7}, where participants predict hydration or partition free energies of small molecules using a variety of force fields, solvent models, and sampling strategies.
-These exercises have repeatedly shown that small changes in simulation parameters or protocols can result in significant deviations from experimental results, reinforcing the need for careful setup and validation even in seemingly "simple" solvation systems.
 
@@
 Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms cancelling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
-Some approaches have attempted to address this using a single coordinate sets and with soft bond approaches ~\cite{clark2019relative,wang2017accurate} or dual coordinate set approach \cite{jespers2019qligfep, ries2022relative} but a general solution is not yet in mainstream use.
+Some approaches have attempted to address this using a single coordinate sets and with soft bond approaches ~\cite{clark2019relative,wang2017accurate} or dual coordinate set approach \cite{jespers2019qligfep, ries2022relative, bhati2017, bieniek2021} but a general solution is not yet in mainstream use.
 Still, FEP+ implements one solution\cite{wang2017accurate}.
-TIES overcomes this issue by ensuring that MCSS atoms are structurally and chemically identical while allowing for partial rings~\cite{bhati2017, bieniek2021}.
+%TIES overcomes this issue by ensuring that MCSS atoms are structurally and chemically identical while allowing for partial rings~\cite{bhati2017, bieniek2021}.
 
@@
 
-\subsubsection{Which $\lambda$-sampling scheme will work best for my problem?}
+\paragraph{Conformational sampling can also solvation and binding free energies}
+While sampling challenges are often perceived to be minor for small molecules especially in hydration free energies, however, they can still be non-trivial.
+Even subtle degrees of freedom such as orientations of the hydroxyl group (as in neutral carboxylic acids) or intramolecular hydrogen bonding can influence the calculated free energy difference between two environments\cite{klimovich2010predicting, lim2019assessing}.
+Molecules with multiple rotatable bonds, internal hydrogen bonds, or conformational flexibility may exhibit environment-dependent populations that are difficult to sample adequately within the timescale of typical molecular dynamics simulations.
+Typically, solvation free energies are computed as absolute transformations, meaning that as the size and flexibility of the ligand increase, the sampling burden grows and convergence of the free energy estimate becomes more challenging.
+In some cases, enhanced sampling techniques such as replica exchange\cite{rest2} or metadynamics\cite{hsu2023alchemical, khuttan2024make} may be beneficial, even in hydration or partition coefficient calculations.
+
+In partition coefficient calculations, the situation becomes even more complex, as two distinct environments must be adequately sampled: typically water and a nonpolar organic solvent.
+These systems not only differ in their dielectric properties and hydrogen bonding behavior but may also yield different dominant solute conformers.
+The partition free energy is sensitive to the relative population of conformers in each environment, so it is essential to ensure that sampling in each phase is independently converged.
+It is also important to ensure adequate overlap in the common end states—if the solute is collapsed to two different gas-phase or non-interacting states, the thermodynamic cycle may not hold\cite{fan2021sampl7}.
+Some studies have employed restrained or confining potentials to prevent solutes from drifting to the interface (especially when using periodic systems with both phases present), while others model each phase separately and compute the free energy difference via thermodynamic cycles\cite{turchi2019,bannan2016,petris2021}.
+
+These challenges are magnified in binding free energy calculations, where sampling all relevant conformations within reasonable timescales becomes even more difficult.
+
+\subsubsection{What $\lambda$-sampling schemes are available?}
 \label{sec:sampling_schemes}
@@
 
-Currently, we recommend using Hamiltonian replica exchange type sampling schemes (Fig.~\ref{fig:fig_sampling_scheme} \textbf{B}).
+%Currently, we recommend using Hamiltonian replica exchange type sampling schemes (Fig.~\ref{fig:fig_sampling_scheme} \textbf{B}).
 If these are not available in the code of choice, running independent simulations at different $\vec{\lambda}$ values can be acceptable, especially when configurational sampling is fast (Fig.~\ref{fig:fig_sampling_scheme} \textbf{A}).
 Single replica schemes and non-equilibrium schemes are not as established yet because of potential failure modes, but are very promising for use in the near future.
-Further, ensemble simulations are recommended irrespective of the choice of sampling scheme to ensure reliability of results~\cite{bhati2018, bhati2019, wan2020fep+, wade2022, wan2021uq, bhati2025, wan2023eqvsneq}.
+%Further, ensemble simulations are recommended irrespective of the choice of sampling scheme to ensure reliability of results~\cite{bhati2018, bhati2019, wan2020fep+, wade2022, wan2021uq, bhati2025, wan2023eqvsneq}.
 
@@
 However, from various previous free energy studies using 1 fs integration timesteps it is not clear that a significant error would be introduced into a free energy estimate and may warrant some further investigation.
+
+\subsubsection{Charge models are non-trivial}
+Modeling the partial charges of the solute and solvent molecules is also highly important.
+In some fixed-charge models, the assigned partial charges depend on the input conformation of the molecule, meaning that poor or inconsistent conformer selection can introduce systematic errors in solvation free energy estimates \cite{jakalian2000, jambeck2003}.
+This is particularly true for methods such as Restrained Electrostatic Potential (RESP)\cite{bayly1993resp} or AM1 population charges with bond-charge correction (AM1-BCC)\cite{jakalian2000}, where the geometry of the molecule at the time of charge derivation can influence the electrostatic potential fit \cite{basma2001, zhang2011, dupradeau2010, osato2025}.
+In these cases, the conformer(s) used for assigning the partial charges should be saved for reproducibility.
+Ensuring consistency between the charge derivation geometry and the initial simulation structure is therefore recommended, or alternatively, using ensemble-averaged charges or polarizable models where appropriate \cite{ren2002,lemkul2016}.
 
@@
 Replicas in ensemble simulations are initialized with velocities drawn independently from the Maxwell–Boltzmann distribution and hence generate decorrelated samples by default.
-Ensemble approaches are therefore strongly preferred for uncertainty estimation, as single simulations remain constrained by the timescales they can access and may therefore underestimate the true uncertainty.
-Further, Wade \textit{et al.} reported that the underestimation of true uncertainty using one-off simulations remains unresolved even on decorrelating the samples~\cite{wade2022}.
 When only one simulation is available, the procedures described below provide a lower-bound estimate of the uncertainty.
@@
 
-Statistical mechanics provides a route to calculate macroscopic thermodynamic properties of a system from \textit{ensemble averaging} over microstates.
+%Statistical mechanics provides a route to calculate macroscopic thermodynamic properties of a system from \textit{ensemble averaging} over microstates.
 Hitherto, MD practitioners have been assuming the validity of the ergodic hypothesis that allows one to equate ensemble averaging to time averaging in the limit of time approaching infinity to predict thermodynamic properties using one-off simulations.
@@
 
-The TIES method~\cite{bhati2017} employs a different scheme to obtain the overall variance of the integral.
+%The TIES method~\cite{bhati2017} employs a different scheme to obtain the overall variance of the integral.
 It assumes that the integrand is a Gaussian random process and hence the thermodynamic integral can be viewed as a so-called stochastic integral, which forms the basis of uncertainty propagation in this integral.
@@
 $\mathrm{var}$($\Delta$f) = $\sum_{i=1}^{K}\Delta\lambda_{k}^2 \mathrm{var}(\frac{du}{d\vec{\lambda}})_{k}$.
+
+Beyond thermodynamic integration, the TIES method\cite{bhati2017} uses an Ito integral with Iso isomoetry for error propogation.
 
@@
 The number of states and amount of sampling should be optimized to minimize the uncertainty in the MBAR free energy estimate, while balancing other key considerations such as computational expense. 
-%A less accurate uncertainty estimate may be desirable over a reduced computational expense, increased sampling strategies such as using larger simulation ensembles will provide more accurate uncertainties but will drastically increase the computational cost.
+A less accurate uncertainty estimate may be desirable over a reduced computational expense, increased sampling strategies such as using larger simulation ensembles will provide more accurate uncertainties but will drastically increase the computational cost.
 However, this process based on one-off simulations underestimates the true uncertainty and hence uncertainties for MBAR also should be obtained using ensemble simulations~\cite{bhati2018, wade2022}.
@@
 Other factors, like system preparation (choice of protonation state, tautomer, counterion presence, salt concentration, etc.) can also substantially impact results~\cite{mobley2017predicting, mobley2017predictinga}, so unless modelers are confident they have these factors correct, sensitivity to these choices may also need to be examined.
-Incidentally, Vassaux \textit{et al.} performed a systematic study comparing the effect of both aleatoric as well as parametric uncertainties for protein-ligand free energy calculations using NAMD~\cite{vassaux2021}.
+Incidentally, Vassaux \textit{et al.} perfor med a systematic study comparing the effect of both aleatoric as well as parametric uncertainties for protein-ligand free energy calculations using NAMD~\cite{vassaux2021}.
 They found that, indeed, ensemble simulations are necessary to control both sources of uncertainties arising in simulations, just as is anticipated in the field of uncertainty quantification.

### 2025-09-28 paper/manuscript.tex
Extended theory to accommodate ATM etc

@@
 \usepackage{changepage}
+\usepackage{verbatim}
 
@@
 
+Next, we outline the application of the chemical potential statistical mechanics formalism for modeling key biophysical processes: solvation, solvent partitioning, and non-covalent molecular binding.
+Generally, turning statistical mechanics identities into computable quantities by molecular simulations involves expressing ratios of partition functions in terms of ensemble averages.
+In the present context, the quantity being averaged often turns out to be the exponential of a perturbation energy function (the \textit{Zwanzig relationship}~\cite{zwanzig1954hightemperature}), which is efficiently evaluated using the stratification algorithms and free energy estimators discussed in Sections \ref{subsec:formulations-abfe-rbfe} and \ref{subsec:estimators} rather than directly.
+Here we assume that an implementation is well-formed once it is expressed as a combination of exponential averages, assuming they can be evaluated in practice.
+As in the previous section, we use the solvent potential of mean force to lighten the notation when possible.
+The solvent degrees of freedom can be restored  in the resulting expressions using Eq.~(\ref{eq:solvent-pmf-def}).
+
 \subsection{Solvation}\label{sec:theory-solvation}
@@
 
-Eq.~(\ref{eq:DG0-binding-constant-def}) is the basis of a variety of alchemical computational implementations discussed in the following sections that have have been developed to study molecular binding by alchemical means.
+Eq.~(\ref{eq:DG0-binding-constant-def}) is the basis of a variety of alchemical computational implementations that have have been developed to study molecular binding by alchemical means.
+Some of these are discussed in Section \ref{subsec:formulations-abfe-rbfe}.
+Generally, to formulate the statistical mechanics expression of the binding constant Eq.~(\ref{eq:DG0-binding-constant-def}) as an ensemble average, we first need to ensure that the configurational integrals in the numerator and denominator are over the same set of degrees of freedom.
+Note that the intramolecular configurational partition function of the complex, $z_{RL(v)}$, includes six more degrees of freedom--the position and orientation of the ligand relative to the receptor collectively described by the variable $\zeta$--than the product of the intramolecular partition functions, $z_{R(v)} z_{L(v)}$, of the separated receptor and ligand.
+To correct this discrepancy, we multiply and divide the right-hand side of Eq.~(\ref{eq:DG0-binding-constant-def})  by the quantity
+\begin{equation} \label{eq:Vsite-def}
+    V_{\rm site} \Omega = \int d\zeta I(\zeta) \, ,
+\end{equation}
+which measures the extent of the binding pose macrostate, where $V_{\rm site}$ corresponds to the spatial extent and $\Omega$ the orientational extent.
+The right-hand side of Eq.~(\ref{eq:Vsite-def}) combined with the integrals $z_{R(v)}$ and $z_{L(v)}$ yields a partition function similar to Eq.~(\ref{eq:intra-zRL-def}) but corresponding to a complex in which ligand and receptor are uncoupled~\cite{gallicchio2011recent} as if they were at infinite distance apart.
+The result is
+\begin{equation}\label{eq:DG0-binding-constant-def-average}
+  K_b = C^\circ V_{\rm site} \frac{\Omega}{8 \pi^2} \langle e^{-\beta \Delta \Psi}  \rangle_0
+\end{equation}
+where, similarly to Eq.~(\ref{eq:zratio-solv-average}),
+\begin{equation}\label{eq:pert-energy-def}
+  \Delta \Psi(x_R, x_L, \zeta) =  \Psi_v(x_R, x_L, \zeta) - \left[  \Psi_v(x_R) + \Psi(x_L) \right]
+\end{equation}
+is the solvent-averaged binding energy of the complex when receptor and ligands are in intramolecular configurations $x_R$ and $x_L$, the ligand is in position and orientation $\zeta$ relative to the receptor (the uncoupled potential energy does not depend on  the position and orientation of the ligand relative to the receptor), and $\langle \ldots \rangle_0$ represents an ensemble average in the uncoupled ensemble.
+Finally, note that $\Omega = 8 \pi^2$ when the binding pose indicator function does not depend on the orientation of the ligand and Eq.~(\ref{eq:DG0-binding-constant-def-average}) simplifies to
+\begin{equation}\label{eq:DG0-binding-constant-def-average-noOmega}
+K_b = C^\circ V_{\rm site} \langle e^{-\beta \Delta \Psi}  \rangle_0 
+\end{equation}
+which, expressed in free energy units, yields
+\begin{eqnarray}\label{eq:DG0-def-Vsite}
+  \Delta G^\circ_b &=&  - k_B T \ln K_b = - k_B T \ln C^\circ V_{\rm site} - k_B T \ln \langle e^{-\beta \Delta \Psi}  \rangle_0 \nonumber \\
+{\rm \ }  &=& \Delta G^\circ_{b,{\rm ideal}} + \Delta G_{b,{\rm exc.}}
+\end{eqnarray}
+
+The term $\Delta G^\circ_{b,{\rm ideal}} = - k_B T \ln C^\circ V_{\rm site}$ in Eq.~(\ref{eq:DG0-def-Vsite}) is interpreted as the \emph{ideal} component of the standard binding free energy, the binding free energy that would be observed in the absence of molecular interactions.
+Conversely, the term $\Delta G_{b,{\rm exc.}} = - k_B T \ln \langle e^{-\beta \Delta \Psi}  \rangle_0 $ is the \emph{excess} component that depends on the specific molecular composition of the system and on the corresponding interactions.
+The excess component is the target of the alchemical molecular simulations discussed here, in contrast to the ideal term that is generally evaluated analytically~\cite{boresch2003absolute}.
+
+Eq~(\ref{eq:DG0-binding-constant-def-average-noOmega}) instructs to collect samples of the solvent-averaged binding energy $\Delta \Psi$ in the ensemble where ligand and receptor are not interacting.
+This can be done directly for simple solutes when the solvent is modeled implicitly.\cite{gallicchio2014virtual} However, with explicit solvation specialized techniques are required.
+We will discuss two: Double-Decoupling and Alchemical Transfer in section \ref{subsec:formulations-abfe-rbfe}.
 
@@
 
-\subsection{Alchemical free energy calculations yield predictions that do not require direct simulation of binding/unbinding events}
-
-In many cases, the quantity of interest is the change in binding affinity between a compound $A$ and a related compound $B$ (e.g., by modifying one of the drug scaffold's substituents, see (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{E})), which, by using Eq.~\ref{eq:binding-free-energy-from-bound-unbound-probability-ratio} and \ref{eq:bound-unbound-probability-ratio} is given by
-\begin{equation}\label{eq:delta-delta-G-physical}
-\begin{split}
-    \Delta \Delta G_{\mathrm{bind}, AB} &= \Delta G_{\mathrm{bind}, B} - \Delta G_{\mathrm{bind}, A} \\
-    &\approx -k_BT \left( \ln \frac{Z(RB)}{Z(R+B)} - \ln \frac{Z(RA)}{Z(R+A)} \right) \, .
-\end{split}
-\end{equation}
-Note that the terms involving the standard concentration cancel out when we assume that the volume is identical for $A$ and $B$.
-Predictions of $\Delta \Delta G_{\mathrm{bind}, AB}$ with non-alchemical methods generally require long simulations of both ligands, possibly through different binding pathways.
-Alchemical relative free energy calculations avoid the need to simulate binding and unbinding events by making use of the fact that the free energy is a state function and exploiting the thermodynamic cycle illustrated in Fig.~\ref{fig:fig_binding_thermodynamic_cycle}.
-This is apparent after rewriting Eq.~\ref{eq:delta-delta-G-physical} as
-\begin{equation}\label{eq:delta-delta-G-alchemical}
-\begin{split}
-    \Delta \Delta G_{\mathrm{bind}, AB} &\approx -k_BT \left( \ln \frac{Z(RB)}{Z(RA)} - \ln \frac{Z(R+B)}{Z(R+A)} \right) \\
-    &= -k_BT \left( \ln \frac{Z(RB)}{Z(RA)} - \ln \frac{Z(B)}{Z(A)} \right) \\
-    &= \Delta G_{\mathrm{bound}} - \Delta G_{\mathrm{unbound}} \, ,
-\end{split}
-\end{equation}
-where $\Delta G_{\mathrm{bound/unbound}}$ is the free energy of mutating $A$ to $B$ in the bound/unbound state.
-Eq.~\ref{eq:delta-delta-G-alchemical} and Fig.~\ref{fig:fig_binding_thermodynamic_cycle} tell us that the difference in free energy of binding between toluene ($A$) and benzyl alcohol ($B$) can be computed by running two independent calculations estimating the free energy cost of mutating $A$ into $B$ in the binding pocket ($\Delta G_{\mathrm{bound}}$) and in solvent ($\Delta G_{\mathrm{unbound}}$), saving us the need to simulate the physical binding process of the two compounds.
-In particular, the second line of Eq.~\ref{eq:delta-delta-G-alchemical} is a consequence of $\Delta G_{\mathrm{unbound}}$ being independent of the presence of the receptor in the simulation box as the definition of the unbound state assumes receptor and ligand to be at a sufficient distance for them to have no energetic interactions.
-Note that, when $A$ and $B$ have a different number of atoms, the factors $ \ln \frac{Z(RB)}{Z(RA)}$ and $\ln \frac{Z(B)}{Z(A)}$ in   Eq.~\ref{eq:delta-delta-G-alchemical} appear both to have factors with units of volume in the logarithms, but these factors exactly cancel between the terms.
+\subsection{Alchemical Binding Free Energy Methods Formulations}
+\label{subsec:formulations-abfe-rbfe}
+
+\subsubsection{Absolute binding free energy (ABFE)}
+
+In the alchemical literature, the term ``absolute binding free energy'' is another name for the standard binding free energy $\Delta G^\circ_b = - k_B T \ln K_b$.
+Here, we discuss two methods, Double-Decoupling (DDM) and Alchemical Transfer (ATM), used to evaluate it by alchemical computer simulations.
+
+\paragraph{Double Decoupling}
+
+The double-decoupling formula is derived from the statistical mechanics theory
+by inserting the definition of the solvent potential of mean force [Eq.~(\ref{eq:solvent-pmf-def})] in each of the configurational partition functions in  Eq.~(\ref{eq:DG0-binding-constant-def}).
+For example, after expanding the solvent potential of mean force, the configurational partition function of the complex becomes
+\begin{equation}\label{eq:partfunct-ratio-explsolv-1}
+z_{RL(v)} = \frac{Z_{N_v, RL}}{Z_{N_v}}
+\end{equation}
+where
+\begin{equation}\label{eq:partfunct-ratio-explsolv-2}
+Z_{N_v, RL} = \int d\vec{r}_v dx_R dx_L d\zeta I(\zeta) e^{-\beta U(x_R, x_L, \zeta, \vec{r}_v)}
+\end{equation}
+is the configurational partition function of the complex in a fixed position and orientation in $N_v$ solvent molecules, and $Z_{N_v}$ is the configurational partition function of the pure solvent.
+This process leads to an expression for the binding constant in terms of products of ratios of partition functions analogous to those encountered in Section \ref{sec:theory-solvation} to model solvation
+\begin{equation}
+  K_b = \frac{C^\circ}{8 \pi^2} \frac{Z_{N_v,RL}}{Z_{N_v,R} z_{L(g)}} \frac{Z_{N_v} z_{L(g)}}{Z_{N_v,L}}
+  \label{eq:ddm-ratios-1}
+\end{equation}
+where we multiplied and divided by the internal configurational partition function of the ligand in vacuum
+\begin{equation}
+z_{L(g)} = \int dx_L e^{-\beta U(x_L)}
+\end{equation}
+We recognize that the second ratio of partition functions in Eq.~(\ref{eq:ddm-ratios-1}) is the inverse of Eq.~(\ref{eq:zratio-solv-average}), hence it is related to the free energy $\Delta G_1$ of decoupling the ligand from the solvent
+\begin{equation}
+\frac{Z_{N_v} z_{L(g)}}{Z_{N_v,L}} = \langle e^{\beta \Delta U}  \rangle_{L(v)} = e^{-\beta \Delta G_1}
+\end{equation}
+where $\Delta U$ is the coupling energy between the ligand and the solvent and $\langle \ldots \rangle_{L(v)}$ is an average over the ligand solution ensemble [see Eq.~(\ref{eq:zratio-solv-dgexc})].
+The ratio of partition functions in Eq.~(\ref{eq:ddm-ratios-1}) involving the ligand-receptor complex is expressed as an ensemble average by multiplying it and dividing it by Eq.~(\ref{eq:Vsite-def}) in the same way that Eq.~(\ref{eq:DG0-binding-constant-def}) is cast as Eq.~(\ref{eq:DG0-binding-constant-def-average}).
+The result is the ratio of the partition functions of the complex in solvent, $Z_{N_v,RL}$, and the partition function $Z_{N_v,R + L(g)}$  of the same system when the ligand is decoupled from the
+receptor and the solvent but is restrained into the binding site region ($I(\zeta) = 1$)
+\begin{equation}
+\frac{Z_{N_v,RL}}{Z_{N_v,R+L} z_{L(g)}} = \frac{8 \pi^2 V_{\rm site} Z_{N_v,RL}}{Z_{N_v,R+L(g)}} = 8 \pi^2 V_{\rm site} \langle e^{\beta \Delta U}  \rangle_{RL(v)} = e^{\beta \Delta G_2}
+\end{equation}
+Collecting the terms, the double-decoupling standard binding free energy is finally expressed as
+\begin{equation}\label{eq:ddm-dg1-dg2}
+\Delta G^\circ_b = -k_B T \ln K_b = -k_B T \ln C^\circ V_{\rm site} + (\Delta G_1 - \Delta G_2)
+\end{equation}
+which includes the ideal term [see Eq.~(\ref{eq:DG0-def-Vsite})], and the difference of decoupling free energies of the ligand from the solvent and from the binding site of the receptor in solution.
+As illustrated in Figure \ref{fig:fig_absolute_thermodynamic_cycle}, the decoupling free energies are commonly computed in separate simulations using the alchemical techniques discussed in the following sections.
+
+\paragraph{Alchemical Transfer}
+
+In the alchemical transfer approach, the standard binding free energy is obtained by a coordinate translation that brings the ligand from the solvent to the receptor binding site~\cite{wu2021alchemical} rather than by decoupling it from the system, as in the double-decoupling method.
+The method derives from Eq.~(\ref{eq:DG0-binding-constant-def}) by inserting the solvent degrees of freedom similarly to the double-decoupling method, except that the product $z_{R(v)} z_{L(v)}$ of partition functions of the unbound state is represented collectively as one explicit solvent system in which the ligand is placed in a region spanned by an indicator function $I^\ast(\zeta)$ identical to the one used to define the binding site region but placed at some arbitrarily large distance from the receptor:
+\begin{equation}
+  z_{R(v)} z_{L(v)} =
+  \frac{
+    \int d\vec{r}_v dx_R dx_L d\zeta I^\ast(\zeta) e^{-\beta U(x_R, x_L, \zeta, \vec{r}_v)}
+  }{
+    8 \pi^2 V_{\rm site} Z_{N_v}
+  }
+\end{equation}
+where, as above, we multiplied and divided by the volume of the binding site, assuming for simplicity that the indicator function does not depend on the orientational degrees of freedom of the ligand.
+Combining this result with Eq.~(\ref{eq:partfunct-ratio-explsolv-1}) for the complex and canceling the common partition function of the pure solvent, leads with Eq.~(\ref{eq:DG0-binding-constant-def})  to an expression for the binding constant in terms of ratios of partition functions in explicit solvent that differ only by the location of the indicator function (in the binding site or in the solvent region):
+\begin{equation}\label{eq:atm-partfunc-ratio}
+  K_b = C^\circ V_{\rm site} \frac{
+    \int d\vec{r}_v dx_R dx_L d\zeta I(\zeta) e^{-\beta U(x_R, x_L, \zeta, \vec{r}_v)}
+  }{
+    \int d\vec{r}_v dx_R dx_L d\zeta I^\ast(\zeta) e^{-\beta U(x_R, x_L, \zeta, \vec{r}_v)}
+  }
+\end{equation}
+The idea of alchemical transfer, illustrated in Figure \ref{fig:atm-illustration}A, is to cast Eq.~(\ref{eq:atm-partfunc-ratio}) as an ansemble average by a change of coordinates noting that $I(\zeta) = I^\ast(\zeta - \vec{h})$ where $\vec{h}$ is the fixed vector distance between the binding site and solvent locations spanned by the two indicator functions.
+The symbol $\zeta$ represents the position and orientation of the ligand relative to the receptor, so here $\zeta - \vec{h}$ is the operation of shifting the position of the ligand by the displacement $\vec{h}$ without changing its orientation.
+Hence, by applying the change of variable $\zeta \rightarrow \zeta - \vec{h}$ to the integral in the numerator of Eq.~(\ref{eq:atm-partfunc-ratio}), we arrive at the following expression for the binding constant, equivalent to Eq.~(\ref{eq:DG0-binding-constant-def-average-noOmega}):
+\begin{equation}\label{eq:atm-average-abfe}
+K_b = C^\circ V_{\rm site} \langle e^{-\beta \Delta U} \rangle_0
+\end{equation}
+where
+\begin{equation}\label{eq:atm-deltaU-def}
+\Delta U(x) = U(x_R, x_L, \zeta + \vec{h}, \vec{r}_v) - U(x_R, x_L, \zeta, \vec{r}_v)
+\end{equation}
+is the perturbation energy of the method, and the ensemble average $\langle \ldots \rangle_0$ is over the unbound ensemble when the ligand is limited to the solvent region far away from the receptor.
+Eq.~(\ref{eq:atm-average-abfe}) instructs to simulate the unbounded system where the ligand is limited to a region in the solvent and to sample the change of potential energy $\Delta U$ resulting from transferring the ligand from the solvent to the receptor binding site by a fixed displacement of the ligand as a whole.
+
+While Eq.~(\ref{eq:atm-average-abfe}) is formally rigorous, as discussed in Section \ref{subsec:numerical-implementations} it is numerically necessary to divide the alchemical transfer ABFE calculation in explicit solvent into two legs (Figure \ref{fig:atm_alchemical_intermediate})~\cite{wu2021alchemical,azimi2022relative} that connect the unbound and bound states by a suitable alchemical intermediate state (Figure \ref{fig:atm_alchemical_intermediate}).
+Current implementations use an alchemical intermediate state midway between the bound and unbound states with the symmetric hybrid potential energy function
+\begin{equation}\label{eq:atm-intermediate-pot}
+U_{1/2}(x) = \frac12 [ U(x_R, x_L, \zeta + \vec{h}, \vec{r}_v) + U(x_R, x_L, \zeta, \vec{r}_v)  ] 
+\end{equation}
+To include the alchemical intermediate, Eq.~(\ref{eq:atm-partfunc-ratio}) is multiplied and divided by the partition function of the system with the hybrid potential energy function in Eq.~(\ref{eq:atm-intermediate-pot}), each serving as the sampling state for each leg.
+The expression for the binding constant becomes a ratio of ensemble averages
+\begin{equation}\label{eq:atm-average-abfe-intermediate}
+  K_b = C^\circ V_{\rm site} \frac{
+    \langle e^{-\beta \Delta U_1} \rangle_{1/2}
+  }{
+    \langle e^{-\beta \Delta U_2} \rangle_{1/2}
+  }
+\end{equation}
+where $\Delta U_1$ and $\Delta U_2$ are the perturbation energies of the bound and unbound states relative to the alchemical intermediate. For example,
+\begin{equation}
+\Delta U_1(x) = U(x_R, x_L, \zeta + \vec{h}, \vec{r}_v) - U_{1/2}(x)
+\end{equation}
+Setting,
+\begin{equation}
+\Delta G_1 = -k_B T \ln \langle e^{-\beta \Delta U_1} \rangle_{1/2}
+\end{equation}
+and similarly for the second leg, we recover Eq.~(\ref{eq:ddm-dg1-dg2}) obtained for double-decoupling.
+This shows that, while the thermodynamic path and the meaning of the $\Delta G_1$ and $\Delta G_2$ free energy components differ, the double-decoupling and alchemical transfer formulations both connect the unbound and bound states through an auxiliary intermediate state: the ligand decoupled state for double-decoupling, and the symmetric alchemical intermediate for alchemical transfer.
+However, alchemical transfer simulates both legs with a single simulation system while double-decoupling accomplishes them in separate simulations.
+
+\subsubsection{Relative binding free energy (RBFE)}
+
+In many cases, the quantity of interest is the change in binding affinity $\Delta \Delta G_b$ between a compound $A$ and a related compound $B$ (e.g., by modifying one of the drug scaffold's substituents, see (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{E})) for the same receptor.
+If the ligands are sufficiently similar, evaluating the Relative Binding Free Energy (RBFE) is often more convenient than taking the difference of their ABFEs.
+Formally, these approaches seek to compute the ratio $K_b(B)/K_b(A)$ of the binding constants of the two ligands rather than each binding constant individually.
+Here we present an RBFE formulation based on the double-decoupling ABFE approach, which we call double-transformation, and the RBFE formulation based on alchemical transfer.
+
+\paragraph{Double Transformation for RBFE estimation}
+
+To derive an RBFE formulation from double-decoupling, we write expressions for the binding constants of $A$ and $B$ using Eq.~(\ref{eq:ddm-ratios-1}) and taking their ratio.
+While doing so, the standard state, pure solvent, and gas-phase terms cancel, and we obtain
+\begin{equation}\label{eq:double-transformation-Kb-rbfe}
+\frac{K_b(B)}{K_b(A)} = \frac{Z_{N_v,RB}}{Z_{N_v,RA}} \frac{Z_{N_v,A}}{Z_{N_v,B}}
+\end{equation}
+where the first ratio of partition functions involves receptor-ligand complexes and the second only the ligands in solution.
+Eq.~(\ref{eq:double-transformation-Kb-rbfe}), suggests that the relative binding free energy, $\Delta \Delta G_b$, of $B$ relative to $A$ can be expressed as the difference of the free energies of ``mutating'' ligand $A$ into ligand $B$ in the presence and absence of the receptor (Figure \ref{fig:fig_binding_thermodynamic_cycle}):
+\begin{equation}\label{eq:double-transformation-dg-rbfe}
+\Delta \Delta G_b = -k_B T \ln \frac{Z_{N_v,A}}{Z_{N_v,B}} - (-k_B T \ln \frac{Z_{N_v,B}}{Z_{N_v,A}}) = \Delta G_{\rm bound} - \Delta G_{\rm unbound}
+\end{equation}
+In practice, each ratio of partition functions must first be turned into an ensemble average to evaluate it.
+Similar to the preceding case, this step requires manipulating the integrals so that they are carried over the same set of degrees of freedom.
+As discussed in Section \ref{sec:relative-fe-protocol}, this step involves using alchemical topologies capable of describing both ligands.
+For example, among other features, alchemical topologies include atoms common to both ligands and dummy atoms that exist in one but not in the other ligand (Figure \ref{fig:fig_topology}).
+In one limit, called dual coordinate dual topology in Section \ref{sec:relative-fe-protocol}, all atoms of both ligands are present, and the partition functions in Eq.~(\ref{eq:double-transformation-dg-rbfe}) differ in which ligand is coupled to the system.
+A key aspect of constructing a suitable alchemical topology is the requirement that the additional degrees of freedom do not affect the RBFE estimate.
 
@@
 \end{figure}
+
+\paragraph{Alchemical Transfer for RBFE Estimation}
+
+The RBFE alchemical transfer formulation is very similar to the ABFE formulation.
+The simulation system includes both ligands, one in solution and the other in the binding site, and the coordinate transformation involves displacing them in opposite directions so that they switch places.
+Formally,~\cite{azimi2022relative} the method derives from Eq.~(\ref{eq:DG0-binding-constant-def}) for each binding constant
+\begin{equation}
+\frac{K_b(B)}{K_b(A)} = \frac{z_{RB(v)} z_{A(v)}}{z_{B(v)} z_{RA(v)}}
+\end{equation}
+and noting that the numerator represents a system with $B$ bound and $A$ unbound, and in the denominator, the roles are switched.
+Next, the solvent PMF is used to indicate the solvent degrees of freedom explicitly, and the indicator functions $I^\ast(\zeta_A)$ and $I^\ast(\zeta_b)$ are introduced to keep the unbound ligands in the solvent, similarly to Eq.~(\ref{eq:atm-partfunc-ratio}).
+The result is
+\begin{equation}\label{eq:atm-partfunc-ratio-rbfe-1}
+  \frac{K_b(B)}{K_b(A)} = \frac{
+    \int d\vec{r}_v dx_R dx_A d\zeta_A dx_B d\zeta_B I^\ast(\zeta_A) I(\zeta_B)  e^{-\beta U(x_R, x_A, \zeta_A, x_B, \zeta_B, \vec{r}_v)}
+  }{
+    \int d\vec{r}_v dx_R dx_A d\zeta_A dx_B d\zeta_B I(\zeta_A) I^\ast(\zeta_B)  e^{-\beta U(x_R, x_A, \zeta_A, x_B, \zeta_B, \vec{r}_v)}
+  }
+\end{equation}
+where $x_A$ and $x_B$ are the internal coordinates of the ligands and $\zeta_A$, $\zeta_B$ are their external coordinates relative to the receptor.
+Next we apply the change of variables $\zeta_B \rightarrow \zeta_B + \vec{h}$ and $\zeta_A \rightarrow \zeta_A - \vec{h}$ to the partition function in the numerator to obtain
+\begin{equation}\label{eq:atm-partfunc-ratio-rbfe-2}
+  \frac{K_b(B)}{K_b(A)} = \frac{
+    \int d\vec{r}_v dx_R dx_A d\zeta_A dx_B d\zeta_B I(\zeta_A) I^\ast(\zeta_B)  e^{-\beta U(x_R, x_A, \zeta_A + \vec{h}, x_B, \zeta_B - \vec{h}, \vec{r}_v)}
+  }{
+    \int d\vec{r}_v dx_R dx_A d\zeta_A dx_B d\zeta_B I(\zeta_A) I^\ast(\zeta_B)  e^{-\beta U(x_R, x_A, \zeta_A, x_B, \zeta_B, \vec{r}_v)}
+  }
+\end{equation}
+which can be written as an ensemble average
+\begin{equation}\label{eq:atm-partfunc-ratio-rbfe-average}
+\frac{K_b(B)}{K_b(A)} = \langle e^{-\beta \Delta U}\rangle_{RA+B}
+\end{equation}
+where
+\begin{equation}\label{eq:atm-partfunc-ratio-rbfe-pert}
+\Delta U = U(x_R, x_A, \zeta_A + \vec{h}, x_B, \zeta_B - \vec{h}, \vec{r}_v) - U(x_R, x_A, \zeta_A, x_B, \zeta_B, \vec{r}_v)
+\end{equation}
+is the perturbation energy for displacing the two ligands in opposite directions, and the ensemble average $\langle \ldots \rangle_{RA+B}$ is carried out with $A$ bound and $B$ in the solvent.
+Except for the coordinate displacement, Eq.~(\ref{eq:atm-partfunc-ratio-rbfe-average}) is implemented similarly to the alchemical transfer ABFE protocol, including the use of the symmetric alchemical intermediate as in Eq.~(\ref{eq:atm-average-abfe-intermediate}).
+
+\subsection{Numerical Implementations of Alchemical Transformations}
+\label{subsec:numerical-implementations}
 
@@
 Dual coordinate approaches with chemical topologies encompass a third group of methods, where essentially two absolute free energy calculations are carried out in opposite directions at the same time.
-The so-called ``separated topologies'' (SepTop) method \cite{rocklin2013separated,Baumann_2023} and the ``alchemical transfer method'' (ATM) \cite{Azimi_2022} belong into this category.
+The so-called ``separated topologies'' (SepTop) method \cite{rocklin2013separated,Baumann_2023} and the ``alchemical transfer method'' (ATM) \cite{azimi2022relative} belong into this category.
 In SepTop, the two copies are held on top of each other via restraints with respect to the protein/receptor they bind to (as opposed to Restraintmaker, where the molecules are restrained to each other).
@@
 In the latter case, the restraints may couple the two set of coordinates directly \cite{Axelsen_1998,Ries_2022}, or the restraints may be defined relative to a common reference frame \cite{rocklin2013separated,Baumann_2023}.
-Finally, in ATM the two copies of the coordinates are held separate \cite{Azimi_2022}.
+Finally, in ATM the two copies of the coordinates are held separate \cite{azimi2022relative}.
 
@@
 For such approaches it is crucial to find a good set of restraints, that keep the ligands colocalized over the time of the simulation .
-This problem is similar to the required restraints in ABFE Calculations (see \ref{sec:ABFE-standard-state} ).
+This problem is similar to the required restraints in ABFE Calculations (see \ref{sec:standardstate-restraints} ).
 Few descriptions of possible approaches in RBFE calculations can be found for such approaches like the one suggested by \textit{Baumann et al.}~\cite{baumann2023broadening} based on Boresch style restraints~\cite{boresch2003absolute}, with an elaborate automatic atom selection approach, for the seperated topology approach, or Restraintmaker~\cite{ries2022restraintmaker} for the linked-DCDT approaches, which finds a small set of atoms to be linked via distance restraints, with a small distance to each other.
@@
 These effectively, are equivalent to the formation of cyclic structures with potential erroneous contributions to the double free energy difference of interest.
-The type of restraints used in SepTop \cite{rocklin2013separated,Baumann_2023} and ATM \cite{Azimi_2022} avoid this problem.
+The type of restraints used in SepTop \cite{rocklin2013separated,Baumann_2023} and ATM \cite{ azimi2022relative} avoid this problem.
 Additionally, SCDT approaches, such as TIES~\cite{bhati2017}, [MENTION OTHERS TOO], etc., circumvent this issue by design as they do not need any restraints to hold MCSS together between the two end-points.
@@
 
+%this was moved to the theory section, keeping it for now for reference
+\begin{comment}
 An absolute binding free energy calculation yields the standard free energy of formation of the molecular complex from the separated receptor and ligand species.
@@
 The excess component is the target of the alchemical molecular simulations discussed here, in contrast to the ideal term that is generally evaluated analytically~\cite{boresch2003absolute}.
-
-\paragraph{Handling the standard state in free energy calculations.} \label{sec:ABFE-standard-state}
+\end{comment}
+
 As discussed in Sections \ref{sec:theory-solvation} and \ref{sec:theory-partitioning}, the handling of the standard state in solvation and solvent partitioning free energy calculations is typically straightforward.
@@
     % Figure to be updated
-    \includegraphics[width=0.88\columnwidth]{paper/figures/fig_configurational_sampling/FES_scenarios_path.pdf}
+    \includegraphics[width=0.88\columnwidth]{figures/fig_configurational_sampling/FES_scenarios_path.pdf}
     \caption{\textbf{Different scenarios of free energy surfaces.} In scenario \textbf{A}, the free energy barrier present at $\lambda=0$ is absent at other $\lambda$ states, so the system can go around the barrier by alternating sampling between alchemical and configurational space, e.g.
@@
 \label{sec:atm}
+\egnote{This section will be removed and some of the content dispersed among the relevant sections of the paper}
 
@@
     \includegraphics[width=0.95\linewidth]
-    {paper/figures_atm/atm_scheme_abfe_rbfe.pdf}
+    {figures_atm/atm_scheme_abfe_rbfe.pdf}
     \caption{\small Alchemical transfer method illustrated for (\textbf{A}) Absolute Binding Free Energy (ABFE) and (\textbf{B}) Relative Binding Free Energy (RBFE) calculations.
@@
 
-The values of the binding energies sampled at regular intervals at each $\lambda$-state during the ATM alchemical simulation are the inputs for binding free energy estimation (see Section \label{sec:data_analysis}) through multi-state thermodynamic reweighting algorithms such as MBAR and UWHAM~\cite{shirts2008statistically,Tan2012}.
+The values of the binding energies sampled at regular intervals at each $\lambda$-state during the ATM alchemical simulation are the inputs for binding free energy estimation (see Section \ref{sec:data_analysis}) through multi-state thermodynamic reweighting algorithms such as MBAR and UWHAM~\cite{shirts2008statistically,Tan2012}.
 One requirement for these algorithms is the evaluation of the perturbation energy $w(b,\vec{\lambda})$ for each sample $b$ at all $\lambda$-states.
@@
     \includegraphics[width=0.95\linewidth]
-    {paper/figures_atm/atm_alchemical_intermediate.pdf}
+    {figures_atm/atm_alchemical_intermediate.pdf}
     \caption{Illustration of ATM's thermodynamic path.
@@
 The receptor is depicted in blue, and the ligand is shown in gray.
-The faint gray color of the ligand in the alchemical intermediate represents its 50\% interaction strength with the receptor and the bulk of the solvent.}
+The faint gray color of the ligand in the alchemical intermediate represents its 50\% interaction strength with the receptor and the bulk of the solvent.
+\egnote{To do: reverse the directions of the arrows}}
     \label{fig:atm_alchemical_intermediate}
@@
     \centering
-    \includegraphics[width=0.95\linewidth]{paper/figures_atm/atm_lig_alignment.pdf}
+    \includegraphics[width=0.95\linewidth]{figures_atm/atm_lig_alignment.pdf}
     \caption{Illustrative example of the selection of the reference atoms for ATM's ligand alignment restraints between two ligands \textbf{A} and \textbf{B}.
@@
 
-Relative free energy calculations carry with them additional unsettled practices relating to the use and careful treatment of dummy atoms (see ``Dummy Atoms'' in Section~\label{sec:relative-fe-protocol}).
+Relative free energy calculations carry with them additional unsettled practices relating to the use and careful treatment of dummy atoms (see ``Dummy Atoms'' in Section~\ref{sec:relative-fe-protocol}).
 In particular, as the work of~\cite{Fleck_2021} highlighted, valence terms for dummy atoms for dummy atoms must be treated far more carefully than previously appreciated in order for dummy atoms' contributions to the free energy to rigorously cancel.

### 2025-09-30 paper/manuscript.tex
Tweaks from @megosato

--- 2025-09-21 paper/manuscript.tex
@@
 Beyond water, the same level of scrutiny is needed for organic solvents such as octanol, cyclohexane, or others used in partitioning studies.
-For example, in a study by Open Force Field, they found that the general AMBER force field (GAFF)\cite{wang2004gaff} and Open Force Field Sage \cite{boothroyd2023sage} force fields produced results in non-aqueous solutions.
+For example, in a study by Open Force Field, they found that the general AMBER force field (GAFF)\cite{wang2004gaff} and Open Force Field Sage \cite{boothroyd2023sage} force fields produced different results in non-aqueous solutions.
 Care should be taken when choosing the forcefield as solvent model parameters may vary substantially between force fields \cite{jambeck2003, kadaoluwa_2021, mobley_2007charge} and impact results accordingly.
-
-\paragraph{Charge models are non-trivial}
-Modeling the partial charges of the solute and solvent molecules is also highly important.
-In some fixed-charge models, the assigned partial charges depend on the input conformation of the molecule, meaning that poor or inconsistent conformer selection can introduce systematic errors in solvation free energy estimates \cite{jakalian2000, jambeck2003}.
-This is particularly true for methods such as Restrained Electrostatic Potential (RESP)\cite{bayly1993resp} or AM1 population charges with bond-charge correction (AM1-BCC)\cite{jakalian2000}, where the geometry of the molecule at the time of charge derivation can influence the electrostatic potential fit \cite{basma2001, zhang2011, dupradeau2010, osato2025}.
-In these cases, the conformer(s) used for assigning the partial charges should be saved for reproducibility.
-Ensuring consistency between the charge derivation geometry and the initial simulation structure is therefore recommended, or alternatively, using ensemble-averaged charges or polarizable models where appropriate \cite{ren2002,lemkul2016}.
 
@@
 
-\paragraph{Conformational sampling can also solvation and binding free energies}
+\paragraph{Conformational sampling challenges can affect both solvation and binding free energies}
 While sampling challenges are often perceived to be minor for small molecules especially in hydration free energies, however, they can still be non-trivial.

### 2025-10-02 paper/manuscript.tex
Do not require ligands for ABFE have the same binding pose in the checklist section
Pointed out by @sboresch

@@
 
-MD trajectories are extremely sensitive to their starting conditions and hence one-off simulations can never be reproducible~\cite{coveney2016, MDbook2025}.
+MD trajectories are extremely sensitive to their starting conditions and hence one-off simulations can never be reproducible~\cite{coveney2016, MDbook2025, caves_locally_1998}.
 Thus, it is recommended to perform ensemble simulations in order to perform ensemble averaging to obtain thermodynamic quantities in a reliable manner.
@@
 \begin{itemize}
- \item Check your ligands have the same, biologically correct binding pose
+ \item Check your ligands have the biologically correct binding pose(s)
         \item Make sure your \textlambda-scheduling is  appropriate

### 2025-10-05 paper/manuscript.tex
Mainly updates to software package section from @IAlibay
Also minor tweaks from @apbhati

@@
 Alchemical free energy protocols as discussed below (Sec.~\ref{sec:simulation_protocol_choice}) are defined for a specific type of free energy calculation, i.e.
-a free energy of hydration or a free energy of binding.
+a free energy of solvation or a free energy of binding.
 Different types of simulations require different choices for ligands, solvent, and host molecules (in the case of the estimation of free energies of binding).
@@
 Preliminary considerations necessary for using free energy methods to compute partition coefficients are generally more straightforward.
-For example, a 3D minimised structure of a solute can be generated with a simple tool such as openBabel and solvated to prepare the input to compute a free energy of hydration~\cite{oboyle2011open}.
+For example, a 3D minimized structure of a solute can be generated with a simple tool such as Open Babel and solvated to prepare the input to compute a free energy of hydration~\cite{oboyle2011open}.
 Despite their relative simplicity, care should still be taken when preparing these systems to avoid introducing systematic errors.
@@
 The type of restraints used in SepTop \cite{rocklin2013separated,Baumann_2023} and ATM \cite{Azimi_2022} avoid this problem.
-Additionally, SCDT approaches, such as TIES~\cite{bhati2017}, [MENTION OTHERS TOO], etc., circumvent this issue by design as they do not need any restraints to hold MCSS together between the two end-points.
+Additionally, SCDT approaches, such as TIES~\cite{bhati2017}, circumvent this issue by design as they do not need any restraints to hold MCSS together between the two end-points.
 
@@
 There are multiple strategies to alleviate such sampling challenges in an alchemical calculation.
-A straightforward but computationally expensive method is to run longer simulations with multiple replicates to increase conformational heterogeneity~\cite{suruzhon2021sensitivity}.
-If using a large number of replicates these are often termed ensemble approaches and an example of such a sampling scheme is the TIES method~\cite{bhati2017} developed by Bhati et al., where a minimum of 5 to 10 replicas are recommended to make up the ensemble for RBFE and ABFE calculations respectively~\cite{bhati2022, bhati2025}.
+A straightforward but computationally expensive method is to run longer simulations with multiple replicas to increase conformational heterogeneity~\cite{suruzhon2021sensitivity}.
+If using a large number of replicas these are often termed ensemble approaches and an example of such a sampling scheme is the TIES method~\cite{bhati2017} developed by Bhati et al., where a minimum of 5 to 10 replicas are recommended to make up the ensemble for RBFE and ABFE calculations respectively~\cite{bhati2022, bhati2025}.
 Another approach, known as integration over parts~\cite{jayachandran2006parallelized}, involves performing separate alchemical calculations limited to a series of binding macrostates and  then computing the overall free energy difference using the free energy combination formula~\cite{gallicchio2011recent,hsu2023alchemical}, i.e., computing the negative log-sum-exp of the Boltzmann weights from each macrostate.
@@
 Other factors, like system preparation (choice of protonation state, tautomer, counterion presence, salt concentration, etc.) can also substantially impact results~\cite{mobley2017predicting, mobley2017predictinga}, so unless modelers are confident they have these factors correct, sensitivity to these choices may also need to be examined.
-Incidentally, Vassaux \textit{et al.} perfor med a systematic study comparing the effect of both aleatoric as well as parametric uncertainties for protein-ligand free energy calculations using NAMD~\cite{vassaux2021}.
+Incidentally, Vassaux \textit{et al.} performed a systematic study comparing the effect of both aleatoric as well as parametric uncertainties for protein-ligand free energy calculations using NAMD~\cite{vassaux2021}.
 They found that, indeed, ensemble simulations are necessary to control both sources of uncertainties arising in simulations, just as is anticipated in the field of uncertainty quantification.
@@
 	\item \href{https://siremol.org/}{Sire} is a multiscale, molecular simulation framework that provides several applications, including SOMD, an MD/MC code for performing FEP calculations via an interface to OpenMM~\cite{woods2024sire}.
-	\item \href{http://getyank.org/latest/index.html}{YANK} is a tool developed by John Chodera and group on the top of OpenMM MD package.
-It allows the users to write their inputs in easy-to-use YAML format.
-        \item \href{https://github.com/choderalab/openmmtools} provides tooling for creating alchemical simulations using OpenMM.
-These include tools for alchemically modifying OpenMM Systems, running multistate simulations and adding alchemical restraints.
+        \item \href{https://github.com/choderalab/openmmtools}{OpenMMTools} provides tooling for creating alchemical simulations using OpenMM.
+These include tools for alchemically modifying OpenMM Systems, running multistate simulations and adding alchemical restraints \cite{john_chodera_2025_17238654}.
 	\item \href{http://www.gromacs.org/}{GROMACS} is a molecular simulation package with a significant number of free energy methods implementations.
@@
     \item \href{https://openfree.energy/}{Open Free Energy} is a platform which offers and allows the creations of alchemical simulation workflows for both absolute and relative binding free energies.
-Current offerings, notably include a Perses-based \url{https://github.com/choderalab/perses/tree/main/perses} hybrid topology method using OpenMM.
+Current offerings, notably includes a \href{https://github.com/choderalab/perses/tree/main/perses}{Perses-based} hybrid topology method using OpenMM \cite{alibay_2025_17258732}.
+    \item \href{https://github.com/Gallicchio-Lab/AToM-OpenMM}{AToM-OpenMM} is a Python package for running simulations using the Alchemical Transfer Method in OpenMM \cite{azimi2022relative, wu2021alchemical,gallicchio2008,gallicchio2025mutants}.
+    \item \href{https://github.com/Psivant/femto/tree/main}{Femto} is a Python package for running conventional MD and free energy simulations.
+Notably it implements tooling for Separated Topologies\cite{baumann2023broadening} and the Alchemical Transfer Method\cite{azimi2022relative,wu2021alchemical,gallicchio2008,gallicchio2025mutants}.
 	\end{itemize}
@@
         \item \href{https://github.com/OpenFreeEnergy/Konnektor}
-    Konnektor: Tooling to aid in the creation of alchemical perrturbation networks~\cite{}.
+    Konnektor: Tooling to aid in the creation of alchemical perrturbation networks~\cite{ries2024konnektor}.
 	\item \href{http://www.charmm-gui.org/}{CHARMM-GUI} is a web-based tool for setting up a variety of MD simulations.
@@
         \item 
-    \href{https://github.com/cbc-univie/transformato}{transformato} is a package for setting up single topology alchemical transformations to be run using CHARMM or OpenMM.
+    \href{https://github.com/cbc-univie/transformato}{transformato} is a package for setting up single topology alchemical transformations to be run using CHARMM or OpenMM \cite{karwounopoulos2022relative, karwounopoulos2023calculations, wieder2022alchemical}.
     \item \href{http://www.ties-service.org}{TIES} allows the setup of perturbation maps, hybrid topologies and structures as well as simulation-ready models for NAMD and OpenMM \url{https://github.com/UCL-CCS/TIES} \cite{bieniek2023}.
@@
 	\url{https://github.com/choderalab/pymbar} \cite{shirts2008statisticallya}.
-	\item Cinnabar: Standardising alchemical free energy analysis.
-Includes tools for calculating network-wide analyses and deriving absolute free energies using a maximum-likelihood estimator \url{https://github.com/OpenFreeEnergy/cinnabar}.
+	\item \href{https://github.com/OpenFreeEnergy/cinnabar}{Cinnabar}: Standardising alchemical free energy analysis.
+Includes tools for calculating network-wide analyses and deriving absolute free energies using a maximum-likelihood estimator \cite{richard_gowers_2025_15678719}.
 	\item Free Energy Workflows: Sire-specific free energy map analysis using weighted path averages 
-        \item \href{https://gitlab.com/RutgersLBSR/fe-toolkit} FE-Toolkit: collection of packages for free energy analysis, including network-wide MBAR analysis~/cite{}.
+        \item \href{https://gitlab.com/RutgersLBSR/fe-toolkit}{FE-Toolkit}: collection of packages for free energy analysis, including network-wide MBAR analysis~\cite{giese2025fe}.
     \url{https://github.com/michellab/freenrgworkflows}.

### 2025-10-05 paper/manuscript.tex
Remove deprecated software

@@
 A large number of software tools can compute SIS MCSS matches using different cheminformatics packages.
-Some rely on RDKit~\cite{rdkit2019Dec}, such as pmx~\cite{gapsys2015pmx}, LOMAP~\cite{liu2013lead}, FESetup~\cite{loeffler2015fesetup}, ProtoCaller~\cite{suruzhon2020protocaller}, SMArt~\cite{petrov2021perturbation}, and partially BioSimSpace~\cite{hedges2019biosimspace}, while others such as fkcombu~\cite{kawabata20143d} are standalone tools.
+Some rely on RDKit~\cite{rdkit2019Dec}, such as pmx~\cite{gapsys2015pmx}, LOMAP~\cite{liu2013lead}, SMArt~\cite{petrov2021perturbation}, and partially BioSimSpace~\cite{hedges2019biosimspace}, while others such as fkcombu~\cite{kawabata20143d} are standalone tools.
 Schr\"{o}dinger's FEP+ planning tool was originally based on a version of LOMAP, and it also uses MCSS matching as well as 3D considerations to plan the network of single topology calculations between molecules~\cite{wang2015accurate}.
@@
 	\item \href{https://github.com/qusers/qligfep}{QligFEP} offers robust and fast setup of FEP calculations for the software package Q~\cite{jespers2019qligfep}.
-	\item \href{https://github.com/protocaller/ProtoCaller}{ProtoCaller}, a setup tool for the automation of Gromacs free energy calculations~\cite{suruzhon2020protocaller}.
-	\item \href{https://fesetup.readthedocs.io/en/latest/introduction.html}{FESetup} has been developed primarily to setup calculations in AMBER, GROMACS and SIRE~\cite{loeffler2015fesetup}.
         \item 
@@
 Includes tools for calculating network-wide analyses and deriving absolute free energies using a maximum-likelihood estimator \cite{richard_gowers_2025_15678719}.
-	\item Free Energy Workflows: Sire-specific free energy map analysis using weighted path averages 
         \item \href{https://gitlab.com/RutgersLBSR/fe-toolkit}{FE-Toolkit}: collection of packages for free energy analysis, including network-wide MBAR analysis~\cite{giese2025fe}.
-    \url{https://github.com/michellab/freenrgworkflows}.
     \item \href{http://www.ties-service.org}{TIES}: Perform statistical analysis of the output of alchemical ensemble simulations and report reliable predictions along with statistical uncertainties for a variety of free energy estimators \url{https://github.com/UCL-CCS/TIES_MD} \cite{bieniek2023}.

### 2025-10-08 paper/manuscript.tex
Tweak the equilibration detection section
Also includes a tweak to references by @sboresch

@@
 \item non-equilibrium free energy calculations~\cite{gapsys2020large}
-\item Free energy calculations using QM/MM methods~\cite{beierlein2011simple,dybeck2016comparison,cave-ayland2015direct,Karwounopoulos2024}.
-\item Free energy calculations using machine learning methods~\cite{rufa2020chemical, scheen2020hybrid, cole2020machine,Schoeller2023}
+\item Free energy calculations using QM/MM methods~\cite{beierlein2011simple,dybeck2016comparison,cave-ayland2015direct,Schoeller2023}.
+\item Free energy calculations using machine learning methods~\cite{rufa2020chemical, scheen2020hybrid, cole2020machine,Karwounopoulos2024}
 \end{itemize}
@@
 where $g_{[t_0,T]}$ is the \emph{statistical inefficiency} of a timeseries $a_t$, described in more detail below.
-Conveniently, this procedure also produces the information necessary to decorrelate the simulation data for estimating the free energy differences, a requisite next step in analysis.
 This approach is implemented within the MBAR~\cite{kylebeauchamp2019choderalab} and alchemlyb~\cite{daviddotson2019alchemistry} packages, and is highly recommended for standard practice.
+A very similar approach involves minimising the standard error of the production region~\cite{whiteEffectiveTruncationHeuristic1997, sprattHeuristicsStartupProblem1998, hoadAutomatingWarmupLength2010, clarkRobustAutomatedTruncation2025}.
 
@@
 If this is not possible, advanced analysis techniques that assume only local equilibrium (rather than global equilibrium) such as the TRAM estimators~\cite{mey2014xtram,wu2016multiensemble,nuske2017markov} may be more appropriate, but are beyond the scope of this paper.
-
-%MRS: Finlay, you can add information about RED here.
 
@@
 If coupled simulations are used (such as a Hamiltonian replica exchange simulation), the replicas should undergo equivalent random walks in alchemical space, and the replicas can be subsampled with the same $g$ to generate an equal number of uncorrelated samples at each alchemical state.
-Conveniently, the approach described above for automated equilibration detection produces an appropriate estimate of $g$ over the production region for automating this process.
+Conveniently, maximising the effective sample size for automated equilibration detection produces an appropriate estimate of $g$ over the production region for automating this process.
 

### 2025-10-28 paper/manuscript.tex
references for solvation

--- 2025-10-05 paper/manuscript.tex
@@
 
+Next, we outline the application of the chemical potential statistical mechanics formalism for modeling key biophysical processes: solvation, solvent partitioning, and non-covalent molecular binding.
+Generally, turning statistical mechanics identities into computable quantities by molecular simulations involves expressing ratios of partition functions in terms of ensemble averages.
+In the present context, the quantity being averaged often turns out to be the exponential of a perturbation energy function (the \textit{Zwanzig relationship}~\cite{zwanzig1954hightemperature}), which is efficiently evaluated using the stratification algorithms and free energy estimators discussed in Sections \ref{subsec:formulations-abfe-rbfe}, \ref{subsec:numerical-implementations},  and \ref{subsec:estimators} rather than directly.
+Here we assume that an implementation is well-formed once it is expressed as a combination of exponential averages, assuming they can be evaluated in practice.
+As in the previous section, we use the solvent potential of mean force to lighten the notation when possible.
+The solvent degrees of freedom can be restored  in the resulting expressions using Eq.~(\ref{eq:solvent-pmf-def}).
+
 \subsection{Solvation}\label{sec:theory-solvation}
@@
 The second term is the excess solvation free energy $\Delta G_{\mathrm{slv, exc.}}$ that can be evaluated by numerical alchemical calculations (see below).
-Hence Eq.~(\ref{eq:chemical-potential-standard}) can be used to compare calculated and experimental standard free energies of solvation.
+Hence Eq.~(\ref{eq:chemical-potential-standard}) can be used to compare calculated and experimental standard free energies of solvation.\cite{thompson2004new}
 
@@
 \end{equation}
-The relation above is used to connect alchemical calculations to Henry's constant values, often after converting them to the excess free energy scale~\cite{nicholls2008predicting,mobley2014freesolv}
+The relation above is used to connect alchemical calculations to Henry's constant values, often after converting them to the excess free energy scale~\cite{cabani1981group,nicholls2008predicting,mobley2014freesolv}
 \begin{equation}
@@
 \end{equation}
-The objective of alchemical solvation free energy calculations is to evaluate Eq.~(\ref{eq:zratio-solv-dgexc}), by means of \emph{coupling} or \emph{decoupling} computational protocols discussed in the following sections.
+The objective of alchemical solvation free energy calculations is to implement Eq.~(\ref{eq:zratio-solv-dgexc}), by means of \emph{coupling} or \emph{decoupling} computational protocols and alchemical intermediate states as  discussed in section \ref{subsec:numerical-implementations}.
 
@@
 
-MD trajectories are extremely sensitive to their starting conditions and hence one-off simulations can never be reproducible~\cite{coveney2016, MDbook2025, caves_locally_1998}.
+MD trajectories are extremely sensitive to their starting conditions and hence one-off simulations can never be reproducible~\cite{coveney2016, MDbook2025}.
 Thus, it is recommended to perform ensemble simulations in order to perform ensemble averaging to obtain thermodynamic quantities in a reliable manner.
@@
 Alchemical free energy protocols as discussed below (Sec.~\ref{sec:simulation_protocol_choice}) are defined for a specific type of free energy calculation, i.e.
-a free energy of solvation or a free energy of binding.
+a free energy of hydration or a free energy of binding.
 Different types of simulations require different choices for ligands, solvent, and host molecules (in the case of the estimation of free energies of binding).
@@
 Preliminary considerations necessary for using free energy methods to compute partition coefficients are generally more straightforward.
-For example, a 3D minimized structure of a solute can be generated with a simple tool such as Open Babel and solvated to prepare the input to compute a free energy of hydration~\cite{oboyle2011open}.
+For example, a 3D minimised structure of a solute can be generated with a simple tool such as openBabel and solvated to prepare the input to compute a free energy of hydration~\cite{oboyle2011open}.
 Despite their relative simplicity, care should still be taken when preparing these systems to avoid introducing systematic errors.
@@
 Beyond water, the same level of scrutiny is needed for organic solvents such as octanol, cyclohexane, or others used in partitioning studies.
-For example, in a study by Open Force Field, they found that the general AMBER force field (GAFF)\cite{wang2004gaff} and Open Force Field Sage \cite{boothroyd2023sage} force fields produced different results in non-aqueous solutions.
+For example, in a study by Open Force Field, they found that the general AMBER force field (GAFF)\cite{wang2004gaff} and Open Force Field Sage \cite{boothroyd2023sage} force fields produced results in non-aqueous solutions.
 Care should be taken when choosing the forcefield as solvent model parameters may vary substantially between force fields \cite{jambeck2003, kadaoluwa_2021, mobley_2007charge} and impact results accordingly.
+
+\paragraph{Charge models are non-trivial}
+Modeling the partial charges of the solute and solvent molecules is also highly important.
+In some fixed-charge models, the assigned partial charges depend on the input conformation of the molecule, meaning that poor or inconsistent conformer selection can introduce systematic errors in solvation free energy estimates \cite{jakalian2000, jambeck2003}.
+This is particularly true for methods such as Restrained Electrostatic Potential (RESP)\cite{bayly1993resp} or AM1 population charges with bond-charge correction (AM1-BCC)\cite{jakalian2000}, where the geometry of the molecule at the time of charge derivation can influence the electrostatic potential fit \cite{basma2001, zhang2011, dupradeau2010, osato2025}.
+In these cases, the conformer(s) used for assigning the partial charges should be saved for reproducibility.
+Ensuring consistency between the charge derivation geometry and the initial simulation structure is therefore recommended, or alternatively, using ensemble-averaged charges or polarizable models where appropriate \cite{ren2002,lemkul2016}.
 
@@
 These effectively, are equivalent to the formation of cyclic structures with potential erroneous contributions to the double free energy difference of interest.
-The type of restraints used in SepTop \cite{rocklin2013separated,Baumann_2023} and ATM \cite{Azimi_2022} avoid this problem.
-Additionally, SCDT approaches, such as TIES~\cite{bhati2017}, circumvent this issue by design as they do not need any restraints to hold MCSS together between the two end-points.
+The type of restraints used in SepTop \cite{rocklin2013separated,Baumann_2023} and ATM \cite{ azimi2022relative} avoid this problem.
+Additionally, SCDT approaches, such as TIES~\cite{bhati2017}, [MENTION OTHERS TOO], etc., circumvent this issue by design as they do not need any restraints to hold MCSS together between the two end-points.
 
@@
 There are multiple strategies to alleviate such sampling challenges in an alchemical calculation.
-A straightforward but computationally expensive method is to run longer simulations with multiple replicas to increase conformational heterogeneity~\cite{suruzhon2021sensitivity}.
-If using a large number of replicas these are often termed ensemble approaches and an example of such a sampling scheme is the TIES method~\cite{bhati2017} developed by Bhati et al., where a minimum of 5 to 10 replicas are recommended to make up the ensemble for RBFE and ABFE calculations respectively~\cite{bhati2022, bhati2025}.
+A straightforward but computationally expensive method is to run longer simulations with multiple replicates to increase conformational heterogeneity~\cite{suruzhon2021sensitivity}.
+If using a large number of replicates these are often termed ensemble approaches and an example of such a sampling scheme is the TIES method~\cite{bhati2017} developed by Bhati et al., where a minimum of 5 to 10 replicas are recommended to make up the ensemble for RBFE and ABFE calculations respectively~\cite{bhati2022, bhati2025}.
 Another approach, known as integration over parts~\cite{jayachandran2006parallelized}, involves performing separate alchemical calculations limited to a series of binding macrostates and  then computing the overall free energy difference using the free energy combination formula~\cite{gallicchio2011recent,hsu2023alchemical}, i.e., computing the negative log-sum-exp of the Boltzmann weights from each macrostate.
@@
 
-\paragraph{Conformational sampling challenges can affect both solvation and binding free energies}
+\paragraph{Conformational sampling can also solvation and binding free energies}
 While sampling challenges are often perceived to be minor for small molecules especially in hydration free energies, however, they can still be non-trivial.
@@
 	\item \href{https://siremol.org/}{Sire} is a multiscale, molecular simulation framework that provides several applications, including SOMD, an MD/MC code for performing FEP calculations via an interface to OpenMM~\cite{woods2024sire}.
-        \item \href{https://github.com/choderalab/openmmtools}{OpenMMTools} provides tooling for creating alchemical simulations using OpenMM.
-These include tools for alchemically modifying OpenMM Systems, running multistate simulations and adding alchemical restraints \cite{john_chodera_2025_17238654}.
+	\item \href{http://getyank.org/latest/index.html}{YANK} is a tool developed by John Chodera and group on the top of OpenMM MD package.
+It allows the users to write their inputs in easy-to-use YAML format.
+        \item \href{https://github.com/choderalab/openmmtools} provides tooling for creating alchemical simulations using OpenMM.
+These include tools for alchemically modifying OpenMM Systems, running multistate simulations and adding alchemical restraints.
 	\item \href{http://www.gromacs.org/}{GROMACS} is a molecular simulation package with a significant number of free energy methods implementations.
@@
     \item \href{https://openfree.energy/}{Open Free Energy} is a platform which offers and allows the creations of alchemical simulation workflows for both absolute and relative binding free energies.
-Current offerings, notably includes a \href{https://github.com/choderalab/perses/tree/main/perses}{Perses-based} hybrid topology method using OpenMM \cite{alibay_2025_17258732}.
-    \item \href{https://github.com/Gallicchio-Lab/AToM-OpenMM}{AToM-OpenMM} is a Python package for running simulations using the Alchemical Transfer Method in OpenMM \cite{azimi2022relative, wu2021alchemical,gallicchio2008,gallicchio2025mutants}.
-    \item \href{https://github.com/Psivant/femto/tree/main}{Femto} is a Python package for running conventional MD and free energy simulations.
-Notably it implements tooling for Separated Topologies\cite{baumann2023broadening} and the Alchemical Transfer Method\cite{azimi2022relative,wu2021alchemical,gallicchio2008,gallicchio2025mutants}.
+Current offerings, notably include a Perses-based \url{https://github.com/choderalab/perses/tree/main/perses} hybrid topology method using OpenMM.
 	\end{itemize}
@@
         \item \href{https://github.com/OpenFreeEnergy/Konnektor}
-    Konnektor: Tooling to aid in the creation of alchemical perrturbation networks~\cite{ries2024konnektor}.
+    Konnektor: Tooling to aid in the creation of alchemical perrturbation networks~\cite{}.
 	\item \href{http://www.charmm-gui.org/}{CHARMM-GUI} is a web-based tool for setting up a variety of MD simulations.
@@
         \item 
-    \href{https://github.com/cbc-univie/transformato}{transformato} is a package for setting up single topology alchemical transformations to be run using CHARMM or OpenMM \cite{karwounopoulos2022relative, karwounopoulos2023calculations, wieder2022alchemical}.
+    \href{https://github.com/cbc-univie/transformato}{transformato} is a package for setting up single topology alchemical transformations to be run using CHARMM or OpenMM.
     \item \href{http://www.ties-service.org}{TIES} allows the setup of perturbation maps, hybrid topologies and structures as well as simulation-ready models for NAMD and OpenMM \url{https://github.com/UCL-CCS/TIES} \cite{bieniek2023}.
@@
 	\url{https://github.com/choderalab/pymbar} \cite{shirts2008statisticallya}.
-	\item \href{https://github.com/OpenFreeEnergy/cinnabar}{Cinnabar}: Standardising alchemical free energy analysis.
-Includes tools for calculating network-wide analyses and deriving absolute free energies using a maximum-likelihood estimator \cite{richard_gowers_2025_15678719}.
+	\item Cinnabar: Standardising alchemical free energy analysis.
+Includes tools for calculating network-wide analyses and deriving absolute free energies using a maximum-likelihood estimator \url{https://github.com/OpenFreeEnergy/cinnabar}.
 	\item Free Energy Workflows: Sire-specific free energy map analysis using weighted path averages 
-        \item \href{https://gitlab.com/RutgersLBSR/fe-toolkit}{FE-Toolkit}: collection of packages for free energy analysis, including network-wide MBAR analysis~\cite{giese2025fe}.
+        \item \href{https://gitlab.com/RutgersLBSR/fe-toolkit} FE-Toolkit: collection of packages for free energy analysis, including network-wide MBAR analysis~/cite{}.
     \url{https://github.com/michellab/freenrgworkflows}.
@@
 \begin{itemize}
- \item Check your ligands have the biologically correct binding pose(s)
+ \item Check your ligands have the same, biologically correct binding pose
         \item Make sure your \textlambda-scheduling is  appropriate

### 2025-10-31 paper/manuscript.tex
removed ATM section, redistributed some content

@@
 
+\begin{figure}
+    \includegraphics[width=0.95\linewidth]{figures/fig5_thermo_cyc_abs/Figure.png}
+    \caption{\textbf{Thermodynamic cycle for an absolute binding free energy calculation with auxiliary restraints} The fully interacting ligand in water (\textbf{A}), has its charges turned off to pass to (\textbf{B}) followed by turning off van der Waals terms, resulting in a non-interacting ligand in water (\textbf{C}).
+The receptor-ligand complex (\textbf{D}) is defined by binding site restraints (dotted oval).
+Auxiliary restraints are often introduced (\textbf{E}) to improve convergence while the charges and van der Waals interactions are turned off (\textbf{F} and \textbf{G}).
+The free energy cost of releasing the auxiliary restraints (\textbf{G} to \textbf{H}) and transferring the non-interacting ligand from the standard state volume to the binding site (\textbf{C} to \textbf{H}) must be accounted for.
+The standard binding free energy is computed from this cycle as $\Delta G^{\circ}_{\mathrm{bind}} = (\Delta G^{\mathrm{elec}}_{\mathrm{solv}}+ \Delta G^{\mathrm{VdW}}_{\mathrm{solv}}) + \Delta G^{\circ}_{\mathrm{ideal}} -(\Delta G_{\mathrm{aux.restr.}}+\Delta G^{\mathrm{elec}}_{\mathrm{bound}}+ \Delta G^{\mathrm{VdW}}_{\mathrm{bound}}+\Delta G_{\mathrm{release}})$.
+Many variations of the route between the free (\textbf{A}) and bound (\textbf{D}) states can be used -- for example, state \textbf{H} is commonly skipped and the free energy cost of moving from \textbf{C} to \textbf{G} is calculated directly.}
+    \label{fig:fig_absolute_thermodynamic_cycle}
+\end{figure}
+
 \paragraph{Alchemical Transfer}
@@
 
-While Eq.~(\ref{eq:atm-average-abfe}) is formally rigorous, as discussed in Section \ref{subsec:numerical-implementations} it is numerically necessary to divide the alchemical transfer ABFE calculation in explicit solvent into two legs (Figure \ref{fig:atm_alchemical_intermediate})~\cite{wu2021alchemical,azimi2022relative} that connect the unbound and bound states by a suitable alchemical intermediate state (Figure \ref{fig:atm_alchemical_intermediate}).
+While Eq.~(\ref{eq:atm-average-abfe}) is formally rigorous, it is numerically necessary to divide the alchemical transfer ABFE calculation in explicit solvent into two legs~\cite{wu2021alchemical,azimi2022relative} that connect the unbound and bound states by a suitable alchemical intermediate state.
 Current implementations use an alchemical intermediate state midway between the bound and unbound states with the symmetric hybrid potential energy function
@@
 and similarly for the second leg, we recover Eq.~(\ref{eq:ddm-dg1-dg2}) obtained for double-decoupling.
-This shows that, while the thermodynamic path and the meaning of the $\Delta G_1$ and $\Delta G_2$ free energy components differ, the double-decoupling and alchemical transfer formulations both connect the unbound and bound states through an auxiliary intermediate state: the ligand decoupled state for double-decoupling, and the symmetric alchemical intermediate for alchemical transfer.
-However, alchemical transfer simulates both legs with a single simulation system while double-decoupling accomplishes them in separate simulations.
+This shows that, while the thermodynamic path and the meaning of the $\Delta G_1$ and $\Delta G_2$ free energy components differ, the double-decoupling and alchemical transfer formulations both connect the unbound and bound states through an auxiliary intermediate state: the ligand decoupled state for double-decoupling, and the symmetric alchemical intermediate for alchemical transfer.~\cite{azimi2025potential} However, alchemical transfer simulates both legs with a single simulation system, while double-decoupling accomplishes them in separate simulations.
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=0.95\linewidth]
+    {figures_atm/atm_scheme_abfe_rbfe.pdf}
+    \caption{\small Alchemical transfer method illustrated for (\textbf{A}) Absolute Binding Free Energy (ABFE) and (\textbf{B}) Relative Binding Free Energy (RBFE) calculations.
+In (A), the receptor (blue) is shown with the ligand in bulk solvent being translated to the active site along a vector $+h$.
+In (B), the RBFE setup is shown with two ligands: one in the binding site and the other in bulk solvent.
+A vector $h$ separates the ligands, and the perturbation energy is evaluated by virtually swapping their positions at each MD step.}
+    \label{fig:atm-illustration}
+\end{figure}
 
@@
 
-In practice, the mutation of $A$ to $B$ is carried out by introducing one or more parameters $\vec{\lambda}$ controlling the potential energy function $U(\vec{q};\vec{\lambda})$ such that the potential of compounds $A$ and $B$ is recovered at two particular values $\vec{\lambda}_A$ and $\vec{\lambda}_B$.
-Briefly, this is achieved by simulating a ``chimeric'' molecule composed of enough atoms to represent both $A$ and $B$.
+As discussed below, because of poor ensemble overlap, in most cases of interest, alchemical free energy calculations cannot be carried out directly by ensemble averages collected only at the reference or target states.
+Instead, in a process called stratification,~\cite{chipot2007free} we take advantage of the state function property of the free energy and break up the transformation into a series of manageable steps, distributed along an alchemical pathway, whose combination is equivalent to the desired transformation (see Section \ref{sec:important_path}).
+Intermediate alchemical states are typically defined in terms of one or more alchemical progress parameters $\vec{\lambda}$ controlling the potential energy function $U(\vec{q};\vec{\lambda})$ such that the potential energies of the initial and final states are recovered at two particular values $\vec{\lambda}_0$ and $\vec{\lambda}_1$.
+
+In the case of a relative binding free energy calculation, for example, this might be achieved by simulating a ``chimeric'' molecule composed of enough atoms to represent both $A$ and $B$.
 A subset of the energetic terms in $U(\vec{q};\vec{\lambda})$ is then modulated by $\vec{\lambda}$ so that at $\vec{\lambda}_A$, the atoms that form molecule $A$ are activated and those belonging exclusively to $B$ are non-interacting ``dummy atoms'', while the opposite occurs at $\vec{\lambda}_B$ (see Sec.~\ref{sec:relative-fe-protocol} for details).
@@
 In ATM, restraints are used to keep the bound and free ligands at some distance apart and in the same orientation, so that during the translation move the free ligand is placed into a position and orientation as similar as possible to that of the bound ligand.
-Both copies are present simultaneously, but they do not occupy the same space.
 Recently, a variant of $\lambda$-dynamics that uses chemical topologies (separate coordinates for each state) has been proposed \cite{Liesen_2024}.
@@
 
+Because they duplicate the common region, dual coordinate formulations are not generally applicable to small modifications of large molecules, such as single-point mutations of proteins.
+An exception, and an example of the variety of alchemical approaches, is the Alchemical Transfer with Coordinate Swapping method~\cite{gallicchio2025mutants}, which is a dual coordinate approach with similar convergence characteristics of single- and dual-topology formulations.
+It works by exchanging the coordinates of the corresponding atoms of the common substructure of the free and bound ligands and displacing the atoms of the variable substituents.
+The net effect is the switching of the positions of the two ligands (Figure \ref{fig:atm-illustration}\textbf{B}) by the transfer of only their variable regions.
+
 \paragraph{Atom mapping} \label{subsec: Atom Mapping}
@@
 These effectively, are equivalent to the formation of cyclic structures with potential erroneous contributions to the double free energy difference of interest.
-The type of restraints used in SepTop \cite{rocklin2013separated,Baumann_2023} and ATM \cite{ azimi2022relative} avoid this problem.
+The type of restraints used in SepTop \cite{rocklin2013separated,Baumann_2023} and ATM \cite{azimi2022relative} avoid this problem.
 Additionally, SCDT approaches, such as TIES~\cite{bhati2017}, [MENTION OTHERS TOO], etc., circumvent this issue by design as they do not need any restraints to hold MCSS together between the two end-points.
@@
 
-\begin{figure}
-    \includegraphics[width=0.95\linewidth]{figures/fig5_thermo_cyc_abs/Figure.png}
-    \caption{\textbf{Thermodynamic cycle for an absolute binding free energy calculation with auxiliary restraints} The fully interacting ligand in water (\textbf{A}), has its charges turned off to pass to (\textbf{B}) followed by turning off van der Waals terms, resulting in a non-interacting ligand in water (\textbf{C}).
-The receptor-ligand complex (\textbf{D}) is defined by binding site restraints (dotted oval).
-Auxiliary restraints are often introduced (\textbf{E}) to improve convergence while the charges and van der Waals interactions are turned off (\textbf{F} and \textbf{G}).
-The free energy cost of releasing the auxiliary restraints (\textbf{G} to \textbf{H}) and transferring the non-interacting ligand from the standard state volume to the binding site (\textbf{C} to \textbf{H}) must be accounted for.
-The standard binding free energy is computed from this cycle as $\Delta G^{\circ}_{\mathrm{bind}} = (\Delta G^{\mathrm{elec}}_{\mathrm{solv}}+ \Delta G^{\mathrm{VdW}}_{\mathrm{solv}}) + \Delta G^{\circ}_{\mathrm{ideal}} -(\Delta G_{\mathrm{aux.restr.}}+\Delta G^{\mathrm{elec}}_{\mathrm{bound}}+ \Delta G^{\mathrm{VdW}}_{\mathrm{bound}}+\Delta G_{\mathrm{release}})$.
-Many variations of the route between the free (\textbf{A}) and bound (\textbf{D}) states can be used -- for example, state \textbf{H} is commonly skipped and the free energy cost of moving from \textbf{C} to \textbf{G} is calculated directly.}
-    \label{fig:fig_absolute_thermodynamic_cycle}
-\end{figure}
-
 %this was moved to the theory section, keeping it for now for reference
@@
 
-Note that $\Delta G^\circ_{b,{\rm ideal}}$ is independent of the chemical nature of the binding partners, and it is the same for all complexes sharing similar binding site definitions.
-The latter is the reason that the ideal term cancels out when taking the difference of absolute binding free energies when estimating relative binding free energies (RBFEs).
-However, while it does not appear explicitly in the definition, $\Delta G_{b,{\rm exc.}}$ depends implicitly on the definition of the binding site through the ensemble average in the uncoupled ensemble, which includes only ``bound'' configurations according to the definition of the indicator function $I(\zeta)$.
+Note that the ideal binding free energy term $\Delta G^\circ_{b,{\rm ideal}}$ in Eq.~(\ref{eq:DG0-def-Vsite}) is independent of the chemical nature of the binding partners, and it is the same for all complexes sharing binding site definitions.
+The latter is why the ideal term cancels out when taking the difference of absolute binding free energies when estimating relative binding free energies (RBFEs).
+However, while it does not appear explicitly in its definition in Eq.~(\ref{eq:DG0-def-Vsite}), the excess component of the binding free energy $\Delta G_{b,{\rm exc.}}$ depends implicitly on the definition of the binding site through the ensemble average in the uncoupled ensemble, which includes only ``bound'' configurations according to the definition of the indicator function $I(\zeta)$.
 Varying the binding site definition affects both the ideal and excess binding free energy terms such that their sum (the standard binding free energy) is often only weakly dependent on the binding site definition~\cite{gilson1997statisticalthermodynamic,gallicchio2011recent}.
@@
 This approach is easily supported in most free energy calculating programs, as it merely involves an additional simultaneous modification.
-There are special flavours of the co-alchemical ion approach such as proposed by Gapsys et al.~\cite{gapsys2015calculation}, which uses a double-system/single-box setup.
-Here, the two branches of a thermodynamic cycle as used in relative free energy calculations are placed in one simulation box, effectively circumventing a change in the overall net charge during a perturbation.
+There are special flavours of the co-alchemical ion approach, such as proposed by Gapsys et al.~\cite{gapsys2015calculation}, which uses a double-system/single-box setup.
+Here, the two branches of a thermodynamic cycle used in relative free energy calculations are placed in one simulation box, effectively circumventing a change in the overall net charge during a perturbation.
 Note that the co-alchemical approaches only avoid finite-size effects with lattice-sum methods such as PME.
+The Alchemical Transfer approach,\cite{wu2021alchemical,azimi2022relative} discussed in Section \ref{subsec:formulations-abfe-rbfe}, is naturally free of charge-changing artefacts because it encodes alchemical binding processes by coordinate transformations that conserve the system's net charge.
 
@@
 
+The pathway implemented for the Alchemical Transfer Method (ATM)~\cite{pal2019perturbation,khuttan2021alchemical,wu2021alchemical,azimi2022relative} is an example of an alternative alchemical formulation for binding that does not require soft-core pair potentials and the splitting of electrostatic and non-electrostatic contributions.
+ATM employs an optimized softplus alchemical potential energy function~\cite{pal2019perturbation} and a soft-core binding energy function that smooths the variations of the perturbation energy $u(\vec{q}) = U_1(\vec{q}) - U_0(\vec{q})$ rather than each pair interaction.\cite{khuttan2021alchemical} The advantage of this approach is that it does not require modifications of the energy routines of the MD engine to implement soft-core pair potentials.
+Moreover, it has been shown to be applicable without electrostatic/LJ splitting to the transfer of molecules, equivalent to the disappearance and introduction of atoms in conventional single- and dual-topology methods.\cite{khuttan2021alchemical}
+
 A key additional consideration in choosing the alchemical pathway is the choice of spacing of intermediate states.
@@
 
-Some approaches have attempted to find alternative pathways to improve efficiency or find paths of low thermodynamic length~\cite{naden2014linear,naden2015linear,pham2012optimal}.
+Additional approaches have attempted to find alternative pathways to improve efficiency or find paths of low thermodynamic length~\cite{naden2014linear,naden2015linear,pham2012optimal}.
 For example, the enveloping distribution sampling (EDS) approach, and its multiple-replica and accelerated variant, works to improve efficiency by creating a single artificial intermediate state which simultaneously samples all end state phase spaces~\cite{perthold2018accelerated,sidler2017efficient, christ2007enveloping}.
@@
 It is supported in various software packages that provide alchemical implementations, such as GROMACS~\cite{aldeghi2015accurate}, GROMOS~\cite{hritz2008hamiltonian,hritz2007optimization},
-AMBER~\cite{Lee_JChemInfModel_2020_v60_p5595}, FEP+~\cite{wang2015accurate}, and NAMD~\cite{jiang2019computing}.
+AMBER~\cite{Lee_JChemInfModel_2020_v60_p5595}, FEP+~\cite{wang2015accurate}, NAMD~\cite{jiang2019computing}, and OpenMM~\cite{rufa2020chemical,azimi2022relative}.
 A specific variant of Hamiltonian replica exchange is FEP/REST2~\cite{rest2, feprest} that involves ``heating'' a small portion of the protein-ligand system to enhance the sampling within the binding pocket.
@@
 It is recommended to calculate the potential differences in code when possible to avoid extra overhead and possible errors produced by evaluating the energy of the configuration twice, and to reduce the amount of stored information.
-Although potential energy derivatives must usually be calculated in code there is one condition under which they can be easily computed in post-simulation analysis. 
-If the alchemical path you have chosen is a linear alchemical path, then $\frac{du}{d\vec{\lambda}} = u_0(\vec{q}) - u_1(\vec{q})$, which is the difference between the initial and final states, which are already calculated by the simulation and can be recorded easily without additional computational expense.
-However, because of the problems with linear paths already discussed in this paper, this simplification is rarely that useful.
+Although potential energy derivatives must usually be calculated in code, they can sometimes be easily computed in post-simulation analysis. 
+For example, if the alchemical path you have chosen is a linear alchemical path, then $\frac{du}{d\vec{\lambda}} = u_1(\vec{q}) - u_0(\vec{q}) = u(\vec{q})$, which is the perturbation energy $u$--the potential energy difference between the initial and final states.
+These values are already calculated by the simulation and can be recorded easily without additional computational expense.
+As discussed earlier, linear paths are inefficient.
+However, methods such as EDS and ATM, based on optimized non-linear energy interpolation alchemical energy functions $W_\lambda(u)$ that depend on atomic coordinates only through the perturbation function,~\cite{konig2021efficient,azimi2022relative} can achieve high computational efficiency by sampling and storing only perturbation energy values.
+The stored values can then be analyzed in post-processing to, for example, evaluate algebraically the $\Delta u_{k,j}$ quantities needed by multistate analysis (see above).
+
+For example, if the alchemical path you have chosen is a linear alchemical path, then $\frac{du}{d\vec{\lambda}} = u_1(\vec{q}) - u_0(\vec{q}) = u(\vec{q})$, which is the perturbation energy $u$--the potential energy difference between the initial and final states.
+These values are already calculated by the simulation and can be recorded easily without additional computational expense.
+As discussed earlier, linear paths are inefficient.
+However, methods such as EDS and ATM based on optimized non-linear energy interpolation alchemical energy functions $W_\lambda(u)$ that depend on atomic coordinates only through the perturbation function,~\cite{konig2021efficient,azimi2022relative} can achieve high computational efficiency by sampling and storing only perturbation energy values.
+The stored values can then be analyzed in post-processing to, for example, evaluate algebraically the $\Delta u_{k,j}$ quantities needed by multistate analysis (see above).
 
@@
 Ensuring consistency between the charge derivation geometry and the initial simulation structure is therefore recommended, or alternatively, using ensemble-averaged charges or polarizable models where appropriate \cite{ren2002,lemkul2016}.
-
-\subsection{Alchemical Transfer Method for Absolute and Relative Binding Free Energy Calculations}
-\label{sec:atm}
-\egnote{This section will be removed and some of the content dispersed among the relevant sections of the paper}
-
-\subsubsection{Introduction}
-
-The Alchemical Transfer Method (ATM)~\cite{wu2021alchemical,azimi2022relative} is an alchemical framework for estimating binding free energies that targets generality with a rigorous and straightforward implementation, with as few additions as possible to standard non-alchemical models of molecular systems.
-Additionally, the method targets complex ligand transformations in drug design, involving scaffold-hopping and charge-changing transformations that are not well supported by traditional algorithms.
-ATM's overarching philosophy is to deploy robust and rigorous free energy estimation algorithms with the simplest possible procedures for developers and end users.
-One critical design principle is the seamless support for advanced molecular potential energy functions, such as quantum-mechanical, polarizable, and neural network potentials, in free energy-based drug discovery applications.
-
-To reach its objectives, ATM represents the alchemical potential energy function through a coordinate transformation~\cite{azimi2022application,azimi2022relative}.
-Unlike traditional alchemical approaches that rely on parameter interpolation of interaction potentials, ATM employs an energy interpolation alchemical energy function~\cite{konig2021efficient} where the target potential is expressed as the reference potential after the ligand is translated from the solvent to the binding site (Figure~\ref{fig:atm-illustration}).
-The binding free energy calculation is performed directly, without solution and receptor legs, as well as electrostatic and non-electrostatic steps.
-The simulation system consists of a single solvent box prepared with conventional setup tools and without pairwise soft-core potentials.
-The leveraging of the natural capabilities of MD engines to apply coordinate transformations is a distinguishing feature of ATM.
-Alchemical transformations for binding implemented through coordinate transformations are computationally straightforward and applicable with any potential energy function, requiring only the evaluation of energy and derivatives before and after the coordinate transformation.
-Conversely, because it recalculates the potential energy and its gradients twice at each time-step, ATM's strategy suffers from the downside that it is as much as twice as slow per time-step as conventional parameter interpolation alchemical approaches.
-
-%mentions parameter vs. energy interpolation 
-%@article{konig2021efficient,
-%  title={Efficient Alchemical Intermediate States in Free Energy Calculations Using $\lambda$-Enveloping Distribution Sampling},
-%  author={K\"{o}nig, Gerhard and Ries, Benjamin and H\"{u}nenberger, Philippe H and Riniker, Sereina},
-%  journal={Journal of Chemical Theory and Computation},
-%  volume={17},
-%  number={9},
-%  pages={5805--5815},
-%  year={2021},
-%  publisher={ACS Publications}
-%}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.95\linewidth]
-    {figures_atm/atm_scheme_abfe_rbfe.pdf}
-    \caption{\small Alchemical transfer method illustrated for (\textbf{A}) Absolute Binding Free Energy (ABFE) and (\textbf{B}) Relative Binding Free Energy (RBFE) calculations.
-In (A), the receptor (blue) is shown with the ligand in bulk solvent being translated to the active site along a vector $+h$.
-In (B), the RBFE setup is shown with two ligands: one in the binding site and the other in bulk solvent.
-A vector $h$ separates the ligands, and the perturbation energy is evaluated by virtually swapping their positions at each MD step.}
-    \label{fig:atm-illustration}
-\end{figure}
-
-\subsubsection{Methodology}
-
-Using the notation in Eq.~(\ref{eq:reduced-potential}), if $u(\vec{q})$ is the physical ($\lambda$-independent) reduced potential of the system when the ligand is at a location in the solvent away from the receptor and $\vec{h}$ is a displacement vector that brings the ligand atoms from the position in the solvent to the receptor binding site (Figure~\ref{fig:atm-illustration}), the ATM \emph{reduced binding energy function}
-\begin{equation}\label{eq:ATM-binding-energy}
-b(\vec{q}) = u(\vec{q}+\vec{h}) - u(\vec{q})
-\end{equation}
-is the change in the system's potential energy when the ligand is rigidly moved from the solvent to the receptor binding site. The ATM alchemical reduced potential is then written as
-\begin{equation}\label{eq:ATM-alchemical-potential}
-    u(\vec{q}; \vec{\lambda}) = u(\vec{q}) + w[b(\vec{q});\vec{\lambda}]
-\end{equation}
-where $w(b,\vec{\lambda})$ is the \emph{perturbation energy function}.
-The perturbation energy function can be any function as long as it is zero at the $\lambda$-state corresponding to the unbound state and it is equal to the binding energy $b(\vec{q})$ at the bound state:
-\begin{eqnarray}
-    w[b(\vec{q});\vec{\lambda}_{\rm U}] = 0           & \quad & \vec{\lambda}_{\rm U}: {\rm Unbound\ State} \\
-    w[b(\vec{q});\vec{\lambda}_{\rm B}] = b(\vec{q})  & \quad & \vec{\lambda}_{\rm B}: {\rm Bound\ State}
-\end{eqnarray}
-
-Current ATM implementations (see below) employ the softplus perturbation energy function\cite{pal2019perturbation}
-\begin{equation}\label{eq:ATM-softplus}
-w(b;\vec{\lambda}) = w[b_{\rm sc}(b);\vec{\lambda}] =  \frac{\lambda_2 - \lambda_1}{\alpha} \ln\left\{ 1 + e^{-\alpha [b_{\rm sc} - b_0]} \right\} + \lambda_2 b_{\rm sc}
-\end{equation}
-where the parameters ($\lambda_1>=0$, $\lambda_2>=0$, $\alpha>0$, and $b_0$) are functions of the $\lambda$-state, $\vec{\lambda}$.
-The softplus non-linear perturbation function satisfies the correct limits when $\lambda_1 = \lambda_2 = 0$ at the unbound state and $\lambda_1 = \lambda_2 = 1$ at the bound state.
-The quantity $b_{\rm sc}$ in Eq.~(\ref{eq:ATM-softplus}) is the soft-core binding energy function
-\begin{equation}
-  b_{\rm sc}(b)=
-\begin{cases}
-b & b \le b_c \\
-(b_{\rm max} - b_c ) f_{\rm sc}\left[\frac{b-b_c}{b_{\rm max}-b_c}\right] + b_c & b > b_c
-\end{cases}
-\label{eq:ATM-soft-core-general}
-\end{equation}
-with
-\begin{equation}
-f_{\rm sc}(y) = \frac{z(y)^{a}-1}{z(y)^{a}+1} \label{eq:rat-sc} \, ,
-\end{equation}
-and
-\begin{equation}
-    z(y)=1+2 y/a + 2 (y/a)^2
-    \label{eq:end-of-softplus}
-\end{equation}
-The soft-core perturbation energy function is designed to avoid singularities near the initial state of the alchemical transformation.\cite{pal2019perturbation,khuttan2021alchemical} The parameters $b_{\rm max}$, $b_c$, and $a$ are set to smoothly cap the perturbation energy to a maximum positive value without affecting it away from the singularity.
-The softplus function has the nice property that is linear for $b<<b_0$ and $b>>b_0$ but with different slopes,  $\lambda_1$ and $\lambda_2$, in these two regimes.
-The bilinear property of the softplus function allows for greater flexibility in creating smoother and more connected alchemical paths that lead to faster convergence~\cite{pal2019perturbation, khuttan2021alchemical}.
-
-The values of the binding energies sampled at regular intervals at each $\lambda$-state during the ATM alchemical simulation are the inputs for binding free energy estimation (see Section \ref{sec:data_analysis}) through multi-state thermodynamic reweighting algorithms such as MBAR and UWHAM~\cite{shirts2008statistically,Tan2012}.
-One requirement for these algorithms is the evaluation of the perturbation energy $w(b,\vec{\lambda})$ for each sample $b$ at all $\lambda$-states.
-The analytical relationships Eq.~(\ref{eq:ATM-softplus})--(\ref{eq:end-of-softplus}) between the binding energy and the perturbation energy make this step straightforward with ATM.
-In contrast, conventional free energy perturbation methods that employ soft-core pair potentials lack analytical perturbation energy functions and require a more complex process involving calling the molecular dynamics engine to rescore simulation trajectories.
-The more straightforward free energy estimation process afforded by ATM saves computing and storage costs, and, since it is based solely on binding energies, allows for changing the perturbation energy function $w(b;\vec{\lambda})$ without running new simulations.
-The simplification of the free energy estimation process fits the general ATM philosophy of streamlining and generalizing alchemical binding free energy calculations as much as possible.
-
-{\bf The ATM Alchemical Intermediate.} Due to atomic clashes with the receptor, the reference potential $u(\vec{q})$ for the unbound state becomes ill-defined near the bound state.
-On the other hand, taking the bound state as the reference leads to numerical instabilities near the unbound state due to clashes with solvent atoms.
-The solution to this conundrum is to split the alchemical binding process into two legs that connect the unbound and bound states to the same alchemical intermediate state midway between the two end states (Figure~\ref{fig:atm_alchemical_intermediate})~\cite{wu2021alchemical,azimi2022relative}.
-In the alchemical leg (Leg 2) that connects the unbound state to the alchemical intermediate, the unbound potential is used as a reference potential.
-The reference potential for the leg connecting the bound state and the alchemical intermediate (Leg 1) is instead the potential energy function of the bound state.
-The transition from one alchemical potential function to another occurs transparently to the user in the ATM replica exchange implementations (see below).
-According to the ATM thermodynamic scheme (Figure~\ref{fig:atm_alchemical_intermediate}), the excess binding free energy is obtained as the difference between the free energy changes of the two legs:
-\begin{equation}
-\Delta G_{b, \text{exc.}} = \Delta G_{\text{Leg2}} - \Delta G_{\text{Leg1}}
-\end{equation}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.95\linewidth]
-    {figures_atm/atm_alchemical_intermediate.pdf}
-    \caption{Illustration of ATM's thermodynamic path.
-At the thermodynamic intermediate (top, center), the ligand interacts at half the strength with the receptor and half strength with the solvent.
-The unbound state (bottom, right) is converted to the alchemical intermediate during Leg 2.
-Similarly, the bound state (bottom, left) is converted to the alchemical intermediate during Leg 1.
-The binding free energy is recovered by taking the difference between the free energies of Leg 2 and Leg 1.
-The receptor is depicted in blue, and the ligand is shown in gray.
-The faint gray color of the ligand in the alchemical intermediate represents its 50\% interaction strength with the receptor and the bulk of the solvent.
-\egnote{To do: reverse the directions of the arrows}}
-    \label{fig:atm_alchemical_intermediate}
-\end{figure}
-
-\subsubsection{ATM Applications: ABFE, RBFE, Relative Binding Selectivity, Macromolecular Binding}
-
-The method described so far, involving the translation of the ligand to the receptor binding site, calculates the standard, or absolute binding free energy (ABFE)~\cite{wu2021alchemical,azimi2022application}.
-A straightforward generalization of ATM's ligand translation transformation enables the calculation of the relative binding free energy of two ligands for the same receptor (RBFE) using a dual-coordinate approach~\cite{azimi2022relative}.
-In ATM's RBFE approach, one ligand is initially placed in the solvent, and the other is placed in the binding site.
-The alchemical potential is then expressed similarly to Eq.~(\ref{eq:ATM-alchemical-potential}) except that, as one ligand is translated into the binding site as in the ABFE protocol, the other ligand is translated in the opposite direction such that the position of the two ligands is swapped.
-Hence, ATM's ABFE and RBFE protocols differ only in the coordinate transformation: ligand translation for ABFE and ligand swapping for RBFE.
-
-ATM's RBFE protocol has recently been extended to the swapping of ligand positions between two receptors~\cite{azimi2024selectivity}.
-The resulting free energy measures the relative selectivity of the two ligands for the two receptors, that is, by how much the selectivity of a ligand for one receptor over another is affected by transforming it into the other.
-This information is critical for the design of drugs that are highly specific and not only strong binders of the target receptor.
-
-The types of coordinate transformations that can be implemented with ATM are not limited to translations.
-For example, coordinates swapping, where the coordinates of an atom are replaced with those of another, has been used to accelerate the RBFE convergence rate between congeneric ligands and to extend ATM to the study of the influence of residue mutations on the binding affinities of protein-peptide and protein-protein complexes~\cite{gallicchio2025mutants}.
-
-\subsubsection{ATM System Setup and Best Practices}
-
-Since it does not require alchemical topologies, preparing molecular complex systems for ATM is the same as for standard molecular dynamics.
-While we described the ABFE protocol in the binding direction, in practice, it is often more convenient to set up the system with the ligand bound to the receptor rather than placed in the solvent.
-If so, the ABFE alchemical calculation is performed in the dissociation direction by essentially reversing the two ATM legs in Figure~(\ref{fig:atm_alchemical_intermediate}).
-To prepare a system for RBFE, one ligand is placed in the binding site and another in the solvent, ensuring their distance is approximately equal to the displacement vector.
-In practice, the two ligands are prepared as superimposed in the binding site, and one of them is translated by the displacement vector to reach the desired location in the solvent.
-
-The ligand displacement vector, illustrated in Figure~\ref{fig:atm-illustration}, is a necessary input for ATM as it determines the ligands' positions in the solvent to represent their unbound states.
-While the dependence of ATM's predictions on this aspect of the method has not been studied systematically, in principle, it should not matter where a ligand is placed in the solvent as long as it is sufficiently far away from the receptor to call the resulting configuration \emph{unbound}.
-Three layers of water molecules, or approximately $10$ \AA, is probably the minimum distance between any two pairs of ligand and receptor atoms to ensure a reasonable degree of decorrelation between the two.
-In practice, the choice of the displacement vector is a trade-off between ensuring that the ligand is sufficiently removed from the receptor when unbound and the size of the solvent box required to contain the receptor and the displaced ligand.
-Some ATM implementations have automated this aspect of the system's setup.
-
-The receptor binding region is typically specified by a flat-bottom harmonic potential restraint between the centroids of a set of receptor and ligand atoms as outlined in Section \ref{sec:standardstate-restraints}.
-For RBFE, a second binding region restraint, identical to the first but displaced into the solvent, restrains the position of the second ligand.
-Although alignment restraints (see below) are more commonly used for this purpose.
-Similarly to other alchemical approaches, orientational restraints can be further deployed to specify the binding pose more precisely~\cite{khuttan2023taming}.
-
-As is common in dual-coordinate alchemical binding implementations~\cite{Ries_2022}, ATM RBFE calculations typically employ optional restraints to maintain the two ligands in approximate alignment.
-ATM's implementation consists of positional and orientational restraints between the internal coordinate frames of the two ligands defined by three reference atoms (Figure~\ref{fig:atm_lig_alignment})~\cite{azimi2022application}.
-A positional restraint keeps the origins of the two frames at a separation approximately equal to the displacement vector, and the axes of the frames are used to impose orientational restraints.
-The objective is to keep the unbound ligand in a position and orientation such that the translation into the binding site places its atoms near the corresponding atoms of the bound ligand.
-Robust alignment is ensured by selecting reference atoms from the rigid, common cores of the ligands.
-The first reference atoms are generally located near the center of the rigid core to define the origin of the coordinate frame.
-The second reference atoms are selected as far as possible from the first, still within the rigid core, to establish an axis that represents the core's orientation.
-Corresponding atoms in the second ligand are identified along the same axis and aligned by ATM, even if perfect overlap is not possible.
-The third reference atoms define a plane with the first two atoms, ensuring consistent rotational alignment (roll) of the rigid cores.
-These are chosen such that they remain approximately coplanar, facilitating close ligand alignment.
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.95\linewidth]{figures_atm/atm_lig_alignment.pdf}
-    \caption{Illustrative example of the selection of the reference atoms for ATM's ligand alignment restraints between two ligands \textbf{A} and \textbf{B}.
-Atoms \textbf{$1_A$} and \textbf{$1_B$} are selected as the first reference points, positioned near the center of the rigid common core.
-These atoms define the origin of the coordinate frames, ensuring that both ligands remain at a fixed distance.
-Atoms \textbf{$2_A$} and \textbf{$2_B$} are chosen from within the common rigid core, positioned as far as possible from the first reference atom.
-The axis defined by $2_A$ and $2_B$ is kept approximately collinear.
-Finally, atoms \textbf{$3_A$} and \textbf{$3_B$} are chosen to maintain the planar alignment of the rigid cores, preserving the roll of the common structural framework.
-Reference atoms are preferably not selected from flexible regions of the ligands, as changes in these regions could alter the relative orientation between the ligands.
-    }
-    \label{fig:atm_lig_alignment}
-\end{figure}
-
-Additional optional restraints are often used in ATM to limit the conformational space to be sampled to improve convergence and reproducibility of binding free energy estimates.
-For example, the AToM-OpenMM~\cite{atom-openmm-github2025} protocol applies by default weak flat-bottom harmonic positional restraints to the C\textsubscript{$\alpha$} atoms of structurally rigid regions of the protein receptor backbone.
-Flexible loops near the binding site that undergo induced fit conformational reorganization are not usually restrained in this way to allow for receptor conformational reorganization.
-
-\subsubsection{ATM Implementations}
-
-There are now various ATM implementations~\cite{atom-openmm-github2025,femto-github2025,atm-acellera-github2025}.
-To our knowledge, currently they are all based on the ATMForce feature of the OpenMM molecular simulation package~\cite{eastman2023openmm}.
-ATMForce implements the ATM alchemical potential through two GPU contexts: the first evaluates the reference potential energy $u(\vec{q})$, and the second calculates the perturbed potential energy after the coordinate transformation.
-ATMForce retrieves energy and forces from each context and merges them into the alchemical potential energy function according to Eqs.~(\ref{eq:ATM-alchemical-potential})--(\ref{eq:end-of-softplus}).
-ATMForce is usually set up to consider only the potential energy terms that change upon the coordinate transformation, such as the non-bonded term for standard fixed charge force fields.
-Covalent terms (bond stretches, angle bending, and torsions) are typically unaffected by rigid translations of whole molecules and can be ignored when evaluating perturbation energies.
-
-The AToM-OpenMM Python package~\cite{atom-openmm-github2025,atom-openmm-documentation2025} is considered the reference ATM implementation.
-AToM-OpenMM provides a toolkit to prepare and run alchemical binding free energy calculations by interfacing AToMForce with the user's input.
-It supports systems with fixed charge force fields (Amber, CHARMM, OpenFF, and Espaloma) and implicit solvent models (HCT, OBC, AGBNP) supported by OpenMM on GPU (CUDA, HIP, OpenCL) and CPU platforms.
-It provides enhanced parallel conformational sampling with Asynchronous Hamiltonian replica exchange~\cite{Gallicchio2008,gallicchio2015asynchronous} over any number of local GPUs, as well as support for advanced features such as metadynamics, relative selectivity calculations with ligand swapping, and macromolecular RBFEs with coordinate swapping~\cite{gallicchio2025mutants}.
-
-The femto package\cite{femto-github2025}, developed by Psivant, offers a lightweight and highly automated interface for high-throughput execution and rapid prototyping of ATM ABFE and RBFE workflows.
-Additionally, Acellera's ATM implementation for the Quantumbind package~\cite{quantumbind-acellera-github2025} provides a commercially supported solution with graphical user interfaces and extended documentation, as well as support for neural-network quantum-mechanical potentials~\cite{sabanes2024enhancing}, enabling both novice and expert users to leverage ATM for routine and advanced free energy calculations.
-A CHARMM-GUI ATM module~\cite{zhang2024ligand} provides support for membrane-bound receptors such as GPCRs.
-Support for ATM is also available from the OpenBioSim infrastructure~\cite{openbiosim-diary-ATM2025}.
-These tools collectively enhance the scalability, usability, and reproducibility of ATM-based simulations.
-
-\subsubsection{ATM Validation Studies and Deployments}
-
-ATM has undergone rigorous early validation through the SAMPL (Statistical Assessment of Modeling of Proteins and Ligands) series of community blind challenges, which were designed to objectively assess computational methods for molecular solvation and binding.
-Notably, ATM was benchmarked in the SAMPL6, SAMPL8, and SAMPL9 blind challenges, consistently ranking among the top-performing methods for both absolute and relative binding free energy calculations \cite{wu2021alchemical,azimi2022application,khuttan2023taming}.
-Notably, ATM yielded SAMPL8 ABFE estimates in close agreement with the rigorous non-alchemical Potential of Mean Force implementation in GROMACS~\cite{azimi2022application}.
-Further large-scale validation studies were conducted in collaboration with leading drug discovery technology companies, including Psivant and Acellera, which focused on refining the method for relative binding free energy calculations using FEP benchmark systems~\cite{chen2024performance,sabanes2023validation}.
-
-Recent advancements have expanded the scope and capabilities of ATM.
-ATM has been integrated with metadynamics (metaD) in OpenMM, enabling it to support an arbitrary set of potentials without modifying the core molecular dynamics energy routines.
-For ligands that undergo substantial conformational reorganization upon binding, metadynamics-based sampling of internal degrees of freedom reduces the variance of free energy closure estimates for improving the robustness of RBFE estimates~\cite{khuttan2024make}.
-The AToM-OpenMM framework includes dedicated functionality to load external metadynamics potentials and apply reweighting or unbiasing schemes, enabling statistically rigorous recovery of free energy estimates from biased simulations.
-
-ATM has been evaluated for protein–ligand binding affinity prediction using hybrid neural network potentials and molecular mechanics (NNP/MM), showing improved accuracy over conventional force fields such as GAFF2 \cite{sabanes2024enhancing}.
-These models have been integrated into the ATM framework via the OpenMM-ML plugin and Acellera's implementation, enabling the use of advanced NNPs, such as ANI and AceForce, within standard ATM workflows.
-These implementations rely on PyTorch-based architectures for efficient GPU-accelerated evaluation of neural network potentials during simulation.
-In addition, ATM has been extended to enable receptor hopping and receptor swapping protocols for computing ligand selectivity between receptors.
-In receptor swapping, one ligand is transferred between two receptors while a second ligand moves in the opposite direction \cite{azimi2024selectivity}.
-Finally, the Alchemical Transfer Method with Coordinate Swapping (ATS) has been developed to compute relative binding free energies for large congeneric ligands and single-point mutant peptides within the ATM framework~\cite{gallicchio2025mutants}.
-
-These extensive efforts have firmly established ATM as a reliable and versatile tool, capable of addressing a broad spectrum of modeling modalities.
-Ongoing work by a growing community of developers and practitioners in academia and industry is increasing the robustness and expanding its range of applicability in computer-aided drug discovery and other fields.
 
@@
 \textbf{JWE} Contributed to ``Conserved binding site waters can play an important role in binding free energies'' and ``Handling poorly sampled water networks'' in Sec. ~\ref{subsec:binding}.\\
-\textbf{EG} Wrote the updated Sec.~\ref{sec:theory} and Sec.~\ref{sec:multiple-chemical-states},  updated Sec.~\ref{sec:standardstate-restraints}, co-wrote Sec.~\ref{sec:atm} with SK, and and contributed to the discussion and review of many sections.\\
+\textbf{EG} Wrote the updated Sec.~\ref{sec:theory} and Sec.~\ref{sec:multiple-chemical-states},  updated Sec.~\ref{sec:standardstate-restraints}, and and contributed to the discussion and review of many sections.\\
 \textbf{MG} Contributed to ``Changes in net charge can be challenging/problematic'' in Sec. \ref{subsec:net charge}. \\
@@
 ~\ref{fig:configurational_sampling}.\\
-\textbf{SK}: Contributed to Sec. \ref{sec:atm}.\\
+\textbf{SK}: Contributed to Sec. \ref{sec:theory} and \ref{sec:standardstate-restraints}  \\
 \textbf{MK}: Contributed to Sec.~\ref{sec:data_analysis}, provided the data for figure~\ref{fig:overlap}, compiled the dataset for Sec.~\ref{sec:benchmark} and helped edit the paper.\\

### 2025-11-05 paper/manuscript.tex
Fixed typo, duplication

@@
 \end{equation}
-The idea of alchemical transfer, illustrated in Figure \ref{fig:atm-illustration}A, is to cast Eq.~(\ref{eq:atm-partfunc-ratio}) as an ansemble average by a change of coordinates noting that $I(\zeta) = I^\ast(\zeta - \vec{h})$ where $\vec{h}$ is the fixed vector distance between the binding site and solvent locations spanned by the two indicator functions.
+The idea of alchemical transfer, illustrated in Figure \ref{fig:atm-illustration}A, is to cast Eq.~(\ref{eq:atm-partfunc-ratio}) as an ensemble average by a change of coordinates noting that $I(\zeta) = I^\ast(\zeta - \vec{h})$ where $\vec{h}$ is the fixed vector distance between the binding site and solvent locations spanned by the two indicator functions.
 The symbol $\zeta$ represents the position and orientation of the ligand relative to the receptor, so here $\zeta - \vec{h}$ is the operation of shifting the position of the ligand by the displacement $\vec{h}$ without changing its orientation.
@@
 However, methods such as EDS and ATM, based on optimized non-linear energy interpolation alchemical energy functions $W_\lambda(u)$ that depend on atomic coordinates only through the perturbation function,~\cite{konig2021efficient,azimi2022relative} can achieve high computational efficiency by sampling and storing only perturbation energy values.
-The stored values can then be analyzed in post-processing to, for example, evaluate algebraically the $\Delta u_{k,j}$ quantities needed by multistate analysis (see above).
-
-For example, if the alchemical path you have chosen is a linear alchemical path, then $\frac{du}{d\vec{\lambda}} = u_1(\vec{q}) - u_0(\vec{q}) = u(\vec{q})$, which is the perturbation energy $u$--the potential energy difference between the initial and final states.
-These values are already calculated by the simulation and can be recorded easily without additional computational expense.
-As discussed earlier, linear paths are inefficient.
-However, methods such as EDS and ATM based on optimized non-linear energy interpolation alchemical energy functions $W_\lambda(u)$ that depend on atomic coordinates only through the perturbation function,~\cite{konig2021efficient,azimi2022relative} can achieve high computational efficiency by sampling and storing only perturbation energy values.
 The stored values can then be analyzed in post-processing to, for example, evaluate algebraically the $\Delta u_{k,j}$ quantities needed by multistate analysis (see above).

### 2025-11-06 paper/manuscript.tex
Minor tweaks from @sboresch and @ppxasjsm

--- 2025-10-28 paper/manuscript.tex
@@
 \usepackage{changepage}
-\usepackage{verbatim}
 
@@
 
-Next, we outline the application of the chemical potential statistical mechanics formalism for modeling key biophysical processes: solvation, solvent partitioning, and non-covalent molecular binding.
-Generally, turning statistical mechanics identities into computable quantities by molecular simulations involves expressing ratios of partition functions in terms of ensemble averages.
-In the present context, the quantity being averaged often turns out to be the exponential of a perturbation energy function (the \textit{Zwanzig relationship}~\cite{zwanzig1954hightemperature}), which is efficiently evaluated using the stratification algorithms and free energy estimators discussed in Sections \ref{subsec:formulations-abfe-rbfe}, \ref{subsec:numerical-implementations},  and \ref{subsec:estimators} rather than directly.
-Here we assume that an implementation is well-formed once it is expressed as a combination of exponential averages, assuming they can be evaluated in practice.
-As in the previous section, we use the solvent potential of mean force to lighten the notation when possible.
-The solvent degrees of freedom can be restored  in the resulting expressions using Eq.~(\ref{eq:solvent-pmf-def}).
-
 \subsection{Solvation}\label{sec:theory-solvation}
@@
 The second term is the excess solvation free energy $\Delta G_{\mathrm{slv, exc.}}$ that can be evaluated by numerical alchemical calculations (see below).
-Hence Eq.~(\ref{eq:chemical-potential-standard}) can be used to compare calculated and experimental standard free energies of solvation.\cite{thompson2004new}
+Hence Eq.~(\ref{eq:chemical-potential-standard}) can be used to compare calculated and experimental standard free energies of solvation.
 
@@
 \end{equation}
-The relation above is used to connect alchemical calculations to Henry's constant values, often after converting them to the excess free energy scale~\cite{cabani1981group,nicholls2008predicting,mobley2014freesolv}
+The relation above is used to connect alchemical calculations to Henry's constant values, often after converting them to the excess free energy scale~\cite{nicholls2008predicting,mobley2014freesolv}
 \begin{equation}
@@
 \end{equation}
-The objective of alchemical solvation free energy calculations is to implement Eq.~(\ref{eq:zratio-solv-dgexc}), by means of \emph{coupling} or \emph{decoupling} computational protocols and alchemical intermediate states as  discussed in section \ref{subsec:numerical-implementations}.
+The objective of alchemical solvation free energy calculations is to evaluate Eq.~(\ref{eq:zratio-solv-dgexc}), by means of \emph{coupling} or \emph{decoupling} computational protocols discussed in the following sections.
 
@@
 
-Eq.~(\ref{eq:DG0-binding-constant-def}) is the basis of a variety of alchemical computational implementations that have have been developed to study molecular binding by alchemical means.
-Some of these are discussed in Section \ref{subsec:formulations-abfe-rbfe}.
-Generally, to formulate the statistical mechanics expression of the binding constant Eq.~(\ref{eq:DG0-binding-constant-def}) as an ensemble average, we first need to ensure that the configurational integrals in the numerator and denominator are over the same set of degrees of freedom.
-Note that the intramolecular configurational partition function of the complex, $z_{RL(v)}$, includes six more degrees of freedom--the position and orientation of the ligand relative to the receptor collectively described by the variable $\zeta$--than the product of the intramolecular partition functions, $z_{R(v)} z_{L(v)}$, of the separated receptor and ligand.
-To correct this discrepancy, we multiply and divide the right-hand side of Eq.~(\ref{eq:DG0-binding-constant-def})  by the quantity
-\begin{equation} \label{eq:Vsite-def}
-    V_{\rm site} \Omega = \int d\zeta I(\zeta) \, ,
-\end{equation}
-which measures the extent of the binding pose macrostate, where $V_{\rm site}$ corresponds to the spatial extent and $\Omega$ the orientational extent.
-The right-hand side of Eq.~(\ref{eq:Vsite-def}) combined with the integrals $z_{R(v)}$ and $z_{L(v)}$ yields a partition function similar to Eq.~(\ref{eq:intra-zRL-def}) but corresponding to a complex in which ligand and receptor are uncoupled~\cite{gallicchio2011recent} as if they were at infinite distance apart.
-The result is
-\begin{equation}\label{eq:DG0-binding-constant-def-average}
-  K_b = C^\circ V_{\rm site} \frac{\Omega}{8 \pi^2} \langle e^{-\beta \Delta \Psi}  \rangle_0
-\end{equation}
-where, similarly to Eq.~(\ref{eq:zratio-solv-average}),
-\begin{equation}\label{eq:pert-energy-def}
-  \Delta \Psi(x_R, x_L, \zeta) =  \Psi_v(x_R, x_L, \zeta) - \left[  \Psi_v(x_R) + \Psi(x_L) \right]
-\end{equation}
-is the solvent-averaged binding energy of the complex when receptor and ligands are in intramolecular configurations $x_R$ and $x_L$, the ligand is in position and orientation $\zeta$ relative to the receptor (the uncoupled potential energy does not depend on  the position and orientation of the ligand relative to the receptor), and $\langle \ldots \rangle_0$ represents an ensemble average in the uncoupled ensemble.
-Finally, note that $\Omega = 8 \pi^2$ when the binding pose indicator function does not depend on the orientation of the ligand and Eq.~(\ref{eq:DG0-binding-constant-def-average}) simplifies to
-\begin{equation}\label{eq:DG0-binding-constant-def-average-noOmega}
-K_b = C^\circ V_{\rm site} \langle e^{-\beta \Delta \Psi}  \rangle_0 
-\end{equation}
-which, expressed in free energy units, yields
-\begin{eqnarray}\label{eq:DG0-def-Vsite}
-  \Delta G^\circ_b &=&  - k_B T \ln K_b = - k_B T \ln C^\circ V_{\rm site} - k_B T \ln \langle e^{-\beta \Delta \Psi}  \rangle_0 \nonumber \\
-{\rm \ }  &=& \Delta G^\circ_{b,{\rm ideal}} + \Delta G_{b,{\rm exc.}}
-\end{eqnarray}
-
-The term $\Delta G^\circ_{b,{\rm ideal}} = - k_B T \ln C^\circ V_{\rm site}$ in Eq.~(\ref{eq:DG0-def-Vsite}) is interpreted as the \emph{ideal} component of the standard binding free energy, the binding free energy that would be observed in the absence of molecular interactions.
-Conversely, the term $\Delta G_{b,{\rm exc.}} = - k_B T \ln \langle e^{-\beta \Delta \Psi}  \rangle_0 $ is the \emph{excess} component that depends on the specific molecular composition of the system and on the corresponding interactions.
-The excess component is the target of the alchemical molecular simulations discussed here, in contrast to the ideal term that is generally evaluated analytically~\cite{boresch2003absolute}.
-
-Eq~(\ref{eq:DG0-binding-constant-def-average-noOmega}) instructs to collect samples of the solvent-averaged binding energy $\Delta \Psi$ in the ensemble where ligand and receptor are not interacting.
-This can be done directly for simple solutes when the solvent is modeled implicitly.\cite{gallicchio2014virtual} However, with explicit solvation specialized techniques are required.
-We will discuss two: Double-Decoupling and Alchemical Transfer in section \ref{subsec:formulations-abfe-rbfe}.
+Eq.~(\ref{eq:DG0-binding-constant-def}) is the basis of a variety of alchemical computational implementations discussed in the following sections that have have been developed to study molecular binding by alchemical means.
 
@@
 
-\subsection{Alchemical Binding Free Energy Methods Formulations}
-\label{subsec:formulations-abfe-rbfe}
-
-\subsubsection{Absolute binding free energy (ABFE)}
-
-In the alchemical literature, the term ``absolute binding free energy'' is another name for the standard binding free energy $\Delta G^\circ_b = - k_B T \ln K_b$.
-Here, we discuss two methods, Double-Decoupling (DDM) and Alchemical Transfer (ATM), used to evaluate it by alchemical computer simulations.
-
-\paragraph{Double Decoupling}
-
-The double-decoupling formula is derived from the statistical mechanics theory
-by inserting the definition of the solvent potential of mean force [Eq.~(\ref{eq:solvent-pmf-def})] in each of the configurational partition functions in  Eq.~(\ref{eq:DG0-binding-constant-def}).
-For example, after expanding the solvent potential of mean force, the configurational partition function of the complex becomes
-\begin{equation}\label{eq:partfunct-ratio-explsolv-1}
-z_{RL(v)} = \frac{Z_{N_v, RL}}{Z_{N_v}}
-\end{equation}
-where
-\begin{equation}\label{eq:partfunct-ratio-explsolv-2}
-Z_{N_v, RL} = \int d\vec{r}_v dx_R dx_L d\zeta I(\zeta) e^{-\beta U(x_R, x_L, \zeta, \vec{r}_v)}
-\end{equation}
-is the configurational partition function of the complex in a fixed position and orientation in $N_v$ solvent molecules, and $Z_{N_v}$ is the configurational partition function of the pure solvent.
-This process leads to an expression for the binding constant in terms of products of ratios of partition functions analogous to those encountered in Section \ref{sec:theory-solvation} to model solvation
-\begin{equation}
-  K_b = \frac{C^\circ}{8 \pi^2} \frac{Z_{N_v,RL}}{Z_{N_v,R} z_{L(g)}} \frac{Z_{N_v} z_{L(g)}}{Z_{N_v,L}}
-  \label{eq:ddm-ratios-1}
-\end{equation}
-where we multiplied and divided by the internal configurational partition function of the ligand in vacuum
-\begin{equation}
-z_{L(g)} = \int dx_L e^{-\beta U(x_L)}
-\end{equation}
-We recognize that the second ratio of partition functions in Eq.~(\ref{eq:ddm-ratios-1}) is the inverse of Eq.~(\ref{eq:zratio-solv-average}), hence it is related to the free energy $\Delta G_1$ of decoupling the ligand from the solvent
-\begin{equation}
-\frac{Z_{N_v} z_{L(g)}}{Z_{N_v,L}} = \langle e^{\beta \Delta U}  \rangle_{L(v)} = e^{-\beta \Delta G_1}
-\end{equation}
-where $\Delta U$ is the coupling energy between the ligand and the solvent and $\langle \ldots \rangle_{L(v)}$ is an average over the ligand solution ensemble [see Eq.~(\ref{eq:zratio-solv-dgexc})].
-The ratio of partition functions in Eq.~(\ref{eq:ddm-ratios-1}) involving the ligand-receptor complex is expressed as an ensemble average by multiplying it and dividing it by Eq.~(\ref{eq:Vsite-def}) in the same way that Eq.~(\ref{eq:DG0-binding-constant-def}) is cast as Eq.~(\ref{eq:DG0-binding-constant-def-average}).
-The result is the ratio of the partition functions of the complex in solvent, $Z_{N_v,RL}$, and the partition function $Z_{N_v,R + L(g)}$  of the same system when the ligand is decoupled from the
-receptor and the solvent but is restrained into the binding site region ($I(\zeta) = 1$)
-\begin{equation}
-\frac{Z_{N_v,RL}}{Z_{N_v,R+L} z_{L(g)}} = \frac{8 \pi^2 V_{\rm site} Z_{N_v,RL}}{Z_{N_v,R+L(g)}} = 8 \pi^2 V_{\rm site} \langle e^{\beta \Delta U}  \rangle_{RL(v)} = e^{\beta \Delta G_2}
-\end{equation}
-Collecting the terms, the double-decoupling standard binding free energy is finally expressed as
-\begin{equation}\label{eq:ddm-dg1-dg2}
-\Delta G^\circ_b = -k_B T \ln K_b = -k_B T \ln C^\circ V_{\rm site} + (\Delta G_1 - \Delta G_2)
-\end{equation}
-which includes the ideal term [see Eq.~(\ref{eq:DG0-def-Vsite})], and the difference of decoupling free energies of the ligand from the solvent and from the binding site of the receptor in solution.
-As illustrated in Figure \ref{fig:fig_absolute_thermodynamic_cycle}, the decoupling free energies are commonly computed in separate simulations using the alchemical techniques discussed in the following sections.
-
-\paragraph{Alchemical Transfer}
-
-In the alchemical transfer approach, the standard binding free energy is obtained by a coordinate translation that brings the ligand from the solvent to the receptor binding site~\cite{wu2021alchemical} rather than by decoupling it from the system, as in the double-decoupling method.
-The method derives from Eq.~(\ref{eq:DG0-binding-constant-def}) by inserting the solvent degrees of freedom similarly to the double-decoupling method, except that the product $z_{R(v)} z_{L(v)}$ of partition functions of the unbound state is represented collectively as one explicit solvent system in which the ligand is placed in a region spanned by an indicator function $I^\ast(\zeta)$ identical to the one used to define the binding site region but placed at some arbitrarily large distance from the receptor:
-\begin{equation}
-  z_{R(v)} z_{L(v)} =
-  \frac{
-    \int d\vec{r}_v dx_R dx_L d\zeta I^\ast(\zeta) e^{-\beta U(x_R, x_L, \zeta, \vec{r}_v)}
-  }{
-    8 \pi^2 V_{\rm site} Z_{N_v}
-  }
-\end{equation}
-where, as above, we multiplied and divided by the volume of the binding site, assuming for simplicity that the indicator function does not depend on the orientational degrees of freedom of the ligand.
-Combining this result with Eq.~(\ref{eq:partfunct-ratio-explsolv-1}) for the complex and canceling the common partition function of the pure solvent, leads with Eq.~(\ref{eq:DG0-binding-constant-def})  to an expression for the binding constant in terms of ratios of partition functions in explicit solvent that differ only by the location of the indicator function (in the binding site or in the solvent region):
-\begin{equation}\label{eq:atm-partfunc-ratio}
-  K_b = C^\circ V_{\rm site} \frac{
-    \int d\vec{r}_v dx_R dx_L d\zeta I(\zeta) e^{-\beta U(x_R, x_L, \zeta, \vec{r}_v)}
-  }{
-    \int d\vec{r}_v dx_R dx_L d\zeta I^\ast(\zeta) e^{-\beta U(x_R, x_L, \zeta, \vec{r}_v)}
-  }
-\end{equation}
-The idea of alchemical transfer, illustrated in Figure \ref{fig:atm-illustration}A, is to cast Eq.~(\ref{eq:atm-partfunc-ratio}) as an ansemble average by a change of coordinates noting that $I(\zeta) = I^\ast(\zeta - \vec{h})$ where $\vec{h}$ is the fixed vector distance between the binding site and solvent locations spanned by the two indicator functions.
-The symbol $\zeta$ represents the position and orientation of the ligand relative to the receptor, so here $\zeta - \vec{h}$ is the operation of shifting the position of the ligand by the displacement $\vec{h}$ without changing its orientation.
-Hence, by applying the change of variable $\zeta \rightarrow \zeta - \vec{h}$ to the integral in the numerator of Eq.~(\ref{eq:atm-partfunc-ratio}), we arrive at the following expression for the binding constant, equivalent to Eq.~(\ref{eq:DG0-binding-constant-def-average-noOmega}):
-\begin{equation}\label{eq:atm-average-abfe}
-K_b = C^\circ V_{\rm site} \langle e^{-\beta \Delta U} \rangle_0
-\end{equation}
-where
-\begin{equation}\label{eq:atm-deltaU-def}
-\Delta U(x) = U(x_R, x_L, \zeta + \vec{h}, \vec{r}_v) - U(x_R, x_L, \zeta, \vec{r}_v)
-\end{equation}
-is the perturbation energy of the method, and the ensemble average $\langle \ldots \rangle_0$ is over the unbound ensemble when the ligand is limited to the solvent region far away from the receptor.
-Eq.~(\ref{eq:atm-average-abfe}) instructs to simulate the unbounded system where the ligand is limited to a region in the solvent and to sample the change of potential energy $\Delta U$ resulting from transferring the ligand from the solvent to the receptor binding site by a fixed displacement of the ligand as a whole.
-
-While Eq.~(\ref{eq:atm-average-abfe}) is formally rigorous, as discussed in Section \ref{subsec:numerical-implementations} it is numerically necessary to divide the alchemical transfer ABFE calculation in explicit solvent into two legs (Figure \ref{fig:atm_alchemical_intermediate})~\cite{wu2021alchemical,azimi2022relative} that connect the unbound and bound states by a suitable alchemical intermediate state (Figure \ref{fig:atm_alchemical_intermediate}).
-Current implementations use an alchemical intermediate state midway between the bound and unbound states with the symmetric hybrid potential energy function
-\begin{equation}\label{eq:atm-intermediate-pot}
-U_{1/2}(x) = \frac12 [ U(x_R, x_L, \zeta + \vec{h}, \vec{r}_v) + U(x_R, x_L, \zeta, \vec{r}_v)  ] 
-\end{equation}
-To include the alchemical intermediate, Eq.~(\ref{eq:atm-partfunc-ratio}) is multiplied and divided by the partition function of the system with the hybrid potential energy function in Eq.~(\ref{eq:atm-intermediate-pot}), each serving as the sampling state for each leg.
-The expression for the binding constant becomes a ratio of ensemble averages
-\begin{equation}\label{eq:atm-average-abfe-intermediate}
-  K_b = C^\circ V_{\rm site} \frac{
-    \langle e^{-\beta \Delta U_1} \rangle_{1/2}
-  }{
-    \langle e^{-\beta \Delta U_2} \rangle_{1/2}
-  }
-\end{equation}
-where $\Delta U_1$ and $\Delta U_2$ are the perturbation energies of the bound and unbound states relative to the alchemical intermediate. For example,
-\begin{equation}
-\Delta U_1(x) = U(x_R, x_L, \zeta + \vec{h}, \vec{r}_v) - U_{1/2}(x)
-\end{equation}
-Setting,
-\begin{equation}
-\Delta G_1 = -k_B T \ln \langle e^{-\beta \Delta U_1} \rangle_{1/2}
-\end{equation}
-and similarly for the second leg, we recover Eq.~(\ref{eq:ddm-dg1-dg2}) obtained for double-decoupling.
-This shows that, while the thermodynamic path and the meaning of the $\Delta G_1$ and $\Delta G_2$ free energy components differ, the double-decoupling and alchemical transfer formulations both connect the unbound and bound states through an auxiliary intermediate state: the ligand decoupled state for double-decoupling, and the symmetric alchemical intermediate for alchemical transfer.
-However, alchemical transfer simulates both legs with a single simulation system while double-decoupling accomplishes them in separate simulations.
-
-\subsubsection{Relative binding free energy (RBFE)}
-
-In many cases, the quantity of interest is the change in binding affinity $\Delta \Delta G_b$ between a compound $A$ and a related compound $B$ (e.g., by modifying one of the drug scaffold's substituents, see (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{E})) for the same receptor.
-If the ligands are sufficiently similar, evaluating the Relative Binding Free Energy (RBFE) is often more convenient than taking the difference of their ABFEs.
-Formally, these approaches seek to compute the ratio $K_b(B)/K_b(A)$ of the binding constants of the two ligands rather than each binding constant individually.
-Here we present an RBFE formulation based on the double-decoupling ABFE approach, which we call double-transformation, and the RBFE formulation based on alchemical transfer.
-
-\paragraph{Double Transformation for RBFE estimation}
-
-To derive an RBFE formulation from double-decoupling, we write expressions for the binding constants of $A$ and $B$ using Eq.~(\ref{eq:ddm-ratios-1}) and taking their ratio.
-While doing so, the standard state, pure solvent, and gas-phase terms cancel, and we obtain
-\begin{equation}\label{eq:double-transformation-Kb-rbfe}
-\frac{K_b(B)}{K_b(A)} = \frac{Z_{N_v,RB}}{Z_{N_v,RA}} \frac{Z_{N_v,A}}{Z_{N_v,B}}
-\end{equation}
-where the first ratio of partition functions involves receptor-ligand complexes and the second only the ligands in solution.
-Eq.~(\ref{eq:double-transformation-Kb-rbfe}), suggests that the relative binding free energy, $\Delta \Delta G_b$, of $B$ relative to $A$ can be expressed as the difference of the free energies of ``mutating'' ligand $A$ into ligand $B$ in the presence and absence of the receptor (Figure \ref{fig:fig_binding_thermodynamic_cycle}):
-\begin{equation}\label{eq:double-transformation-dg-rbfe}
-\Delta \Delta G_b = -k_B T \ln \frac{Z_{N_v,A}}{Z_{N_v,B}} - (-k_B T \ln \frac{Z_{N_v,B}}{Z_{N_v,A}}) = \Delta G_{\rm bound} - \Delta G_{\rm unbound}
-\end{equation}
-In practice, each ratio of partition functions must first be turned into an ensemble average to evaluate it.
-Similar to the preceding case, this step requires manipulating the integrals so that they are carried over the same set of degrees of freedom.
-As discussed in Section \ref{sec:relative-fe-protocol}, this step involves using alchemical topologies capable of describing both ligands.
-For example, among other features, alchemical topologies include atoms common to both ligands and dummy atoms that exist in one but not in the other ligand (Figure \ref{fig:fig_topology}).
-In one limit, called dual coordinate dual topology in Section \ref{sec:relative-fe-protocol}, all atoms of both ligands are present, and the partition functions in Eq.~(\ref{eq:double-transformation-dg-rbfe}) differ in which ligand is coupled to the system.
-A key aspect of constructing a suitable alchemical topology is the requirement that the additional degrees of freedom do not affect the RBFE estimate.
+\subsection{Alchemical free energy calculations yield predictions that do not require direct simulation of binding/unbinding events}
+
+In many cases, the quantity of interest is the change in binding affinity between a compound $A$ and a related compound $B$ (e.g., by modifying one of the drug scaffold's substituents, see (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{E})), which, by using Eq.~\ref{eq:binding-free-energy-from-bound-unbound-probability-ratio} and \ref{eq:bound-unbound-probability-ratio} is given by
+\begin{equation}\label{eq:delta-delta-G-physical}
+\begin{split}
+    \Delta \Delta G_{\mathrm{bind}, AB} &= \Delta G_{\mathrm{bind}, B} - \Delta G_{\mathrm{bind}, A} \\
+    &\approx -k_BT \left( \ln \frac{Z(RB)}{Z(R+B)} - \ln \frac{Z(RA)}{Z(R+A)} \right) \, .
+\end{split}
+\end{equation}
+Note that the terms involving the standard concentration cancel out when we assume that the volume is identical for $A$ and $B$.
+Predictions of $\Delta \Delta G_{\mathrm{bind}, AB}$ with non-alchemical methods generally require long simulations of both ligands, possibly through different binding pathways.
+Alchemical relative free energy calculations avoid the need to simulate binding and unbinding events by making use of the fact that the free energy is a state function and exploiting the thermodynamic cycle illustrated in Fig.~\ref{fig:fig_binding_thermodynamic_cycle}.
+This is apparent after rewriting Eq.~\ref{eq:delta-delta-G-physical} as
+\begin{equation}\label{eq:delta-delta-G-alchemical}
+\begin{split}
+    \Delta \Delta G_{\mathrm{bind}, AB} &\approx -k_BT \left( \ln \frac{Z(RB)}{Z(RA)} - \ln \frac{Z(R+B)}{Z(R+A)} \right) \\
+    &= -k_BT \left( \ln \frac{Z(RB)}{Z(RA)} - \ln \frac{Z(B)}{Z(A)} \right) \\
+    &= \Delta G_{\mathrm{bound}} - \Delta G_{\mathrm{unbound}} \, ,
+\end{split}
+\end{equation}
+where $\Delta G_{\mathrm{bound/unbound}}$ is the free energy of mutating $A$ to $B$ in the bound/unbound state.
+Eq.~\ref{eq:delta-delta-G-alchemical} and Fig.~\ref{fig:fig_binding_thermodynamic_cycle} tell us that the difference in free energy of binding between toluene ($A$) and benzyl alcohol ($B$) can be computed by running two independent calculations estimating the free energy cost of mutating $A$ into $B$ in the binding pocket ($\Delta G_{\mathrm{bound}}$) and in solvent ($\Delta G_{\mathrm{unbound}}$), saving us the need to simulate the physical binding process of the two compounds.
+In particular, the second line of Eq.~\ref{eq:delta-delta-G-alchemical} is a consequence of $\Delta G_{\mathrm{unbound}}$ being independent of the presence of the receptor in the simulation box as the definition of the unbound state assumes receptor and ligand to be at a sufficient distance for them to have no energetic interactions.
+Note that, when $A$ and $B$ have a different number of atoms, the factors $ \ln \frac{Z(RB)}{Z(RA)}$ and $\ln \frac{Z(B)}{Z(A)}$ in   Eq.~\ref{eq:delta-delta-G-alchemical} appear both to have factors with units of volume in the logarithms, but these factors exactly cancel between the terms.
 
@@
 \end{figure}
-
-\paragraph{Alchemical Transfer for RBFE Estimation}
-
-The RBFE alchemical transfer formulation is very similar to the ABFE formulation.
-The simulation system includes both ligands, one in solution and the other in the binding site, and the coordinate transformation involves displacing them in opposite directions so that they switch places.
-Formally,~\cite{azimi2022relative} the method derives from Eq.~(\ref{eq:DG0-binding-constant-def}) for each binding constant
-\begin{equation}
-\frac{K_b(B)}{K_b(A)} = \frac{z_{RB(v)} z_{A(v)}}{z_{B(v)} z_{RA(v)}}
-\end{equation}
-and noting that the numerator represents a system with $B$ bound and $A$ unbound, and in the denominator, the roles are switched.
-Next, the solvent PMF is used to indicate the solvent degrees of freedom explicitly, and the indicator functions $I^\ast(\zeta_A)$ and $I^\ast(\zeta_b)$ are introduced to keep the unbound ligands in the solvent, similarly to Eq.~(\ref{eq:atm-partfunc-ratio}).
-The result is
-\begin{equation}\label{eq:atm-partfunc-ratio-rbfe-1}
-  \frac{K_b(B)}{K_b(A)} = \frac{
-    \int d\vec{r}_v dx_R dx_A d\zeta_A dx_B d\zeta_B I^\ast(\zeta_A) I(\zeta_B)  e^{-\beta U(x_R, x_A, \zeta_A, x_B, \zeta_B, \vec{r}_v)}
-  }{
-    \int d\vec{r}_v dx_R dx_A d\zeta_A dx_B d\zeta_B I(\zeta_A) I^\ast(\zeta_B)  e^{-\beta U(x_R, x_A, \zeta_A, x_B, \zeta_B, \vec{r}_v)}
-  }
-\end{equation}
-where $x_A$ and $x_B$ are the internal coordinates of the ligands and $\zeta_A$, $\zeta_B$ are their external coordinates relative to the receptor.
-Next we apply the change of variables $\zeta_B \rightarrow \zeta_B + \vec{h}$ and $\zeta_A \rightarrow \zeta_A - \vec{h}$ to the partition function in the numerator to obtain
-\begin{equation}\label{eq:atm-partfunc-ratio-rbfe-2}
-  \frac{K_b(B)}{K_b(A)} = \frac{
-    \int d\vec{r}_v dx_R dx_A d\zeta_A dx_B d\zeta_B I(\zeta_A) I^\ast(\zeta_B)  e^{-\beta U(x_R, x_A, \zeta_A + \vec{h}, x_B, \zeta_B - \vec{h}, \vec{r}_v)}
-  }{
-    \int d\vec{r}_v dx_R dx_A d\zeta_A dx_B d\zeta_B I(\zeta_A) I^\ast(\zeta_B)  e^{-\beta U(x_R, x_A, \zeta_A, x_B, \zeta_B, \vec{r}_v)}
-  }
-\end{equation}
-which can be written as an ensemble average
-\begin{equation}\label{eq:atm-partfunc-ratio-rbfe-average}
-\frac{K_b(B)}{K_b(A)} = \langle e^{-\beta \Delta U}\rangle_{RA+B}
-\end{equation}
-where
-\begin{equation}\label{eq:atm-partfunc-ratio-rbfe-pert}
-\Delta U = U(x_R, x_A, \zeta_A + \vec{h}, x_B, \zeta_B - \vec{h}, \vec{r}_v) - U(x_R, x_A, \zeta_A, x_B, \zeta_B, \vec{r}_v)
-\end{equation}
-is the perturbation energy for displacing the two ligands in opposite directions, and the ensemble average $\langle \ldots \rangle_{RA+B}$ is carried out with $A$ bound and $B$ in the solvent.
-Except for the coordinate displacement, Eq.~(\ref{eq:atm-partfunc-ratio-rbfe-average}) is implemented similarly to the alchemical transfer ABFE protocol, including the use of the symmetric alchemical intermediate as in Eq.~(\ref{eq:atm-average-abfe-intermediate}).
-
-\subsection{Numerical Implementations of Alchemical Transformations}
-\label{subsec:numerical-implementations}
 
@@
 Dual coordinate approaches with chemical topologies encompass a third group of methods, where essentially two absolute free energy calculations are carried out in opposite directions at the same time.
-The so-called ``separated topologies'' (SepTop) method \cite{rocklin2013separated,Baumann_2023} and the ``alchemical transfer method'' (ATM) \cite{azimi2022relative} belong into this category.
+The so-called ``separated topologies'' (SepTop) method \cite{rocklin2013separated,Baumann_2023} and the ``alchemical transfer method'' (ATM) \cite{Azimi_2022} belong into this category.
 In SepTop, the two copies are held on top of each other via restraints with respect to the protein/receptor they bind to (as opposed to Restraintmaker, where the molecules are restrained to each other).
@@
 In the latter case, the restraints may couple the two set of coordinates directly \cite{Axelsen_1998,Ries_2022}, or the restraints may be defined relative to a common reference frame \cite{rocklin2013separated,Baumann_2023}.
-Finally, in ATM the two copies of the coordinates are held separate \cite{azimi2022relative}.
+Finally, in ATM the two copies of the coordinates are held separate \cite{Azimi_2022}.
 
@@
 For such approaches it is crucial to find a good set of restraints, that keep the ligands colocalized over the time of the simulation .
-This problem is similar to the required restraints in ABFE Calculations (see \ref{sec:standardstate-restraints} ).
+This problem is similar to the required restraints in ABFE Calculations (see \ref{sec:ABFE-standard-state} ).
 Few descriptions of possible approaches in RBFE calculations can be found for such approaches like the one suggested by \textit{Baumann et al.}~\cite{baumann2023broadening} based on Boresch style restraints~\cite{boresch2003absolute}, with an elaborate automatic atom selection approach, for the seperated topology approach, or Restraintmaker~\cite{ries2022restraintmaker} for the linked-DCDT approaches, which finds a small set of atoms to be linked via distance restraints, with a small distance to each other.
@@
 These effectively, are equivalent to the formation of cyclic structures with potential erroneous contributions to the double free energy difference of interest.
-The type of restraints used in SepTop \cite{rocklin2013separated,Baumann_2023} and ATM \cite{ azimi2022relative} avoid this problem.
-Additionally, SCDT approaches, such as TIES~\cite{bhati2017}, [MENTION OTHERS TOO], etc., circumvent this issue by design as they do not need any restraints to hold MCSS together between the two end-points.
+The type of restraints used in SepTop \cite{rocklin2013separated,Baumann_2023} and ATM \cite{Azimi_2022} avoid this problem.
+Additionally, SCDT approaches, such as TIES~\cite{bhati2017}, circumvent this issue by design as they do not need any restraints to hold MCSS together between the two end-points.
 
@@
 
-%this was moved to the theory section, keeping it for now for reference
-\begin{comment}
 An absolute binding free energy calculation yields the standard free energy of formation of the molecular complex from the separated receptor and ligand species.
@@
 The excess component is the target of the alchemical molecular simulations discussed here, in contrast to the ideal term that is generally evaluated analytically~\cite{boresch2003absolute}.
-\end{comment}
-
+
+\paragraph{Handling the standard state in free energy calculations.} \label{sec:ABFE-standard-state}
 As discussed in Sections \ref{sec:theory-solvation} and \ref{sec:theory-partitioning}, the handling of the standard state in solvation and solvent partitioning free energy calculations is typically straightforward.
@@
     % Figure to be updated
-    \includegraphics[width=0.88\columnwidth]{figures/fig_configurational_sampling/FES_scenarios_path.pdf}
+    \includegraphics[width=0.88\columnwidth]{paper/figures/fig_configurational_sampling/FES_scenarios_path.pdf}
     \caption{\textbf{Different scenarios of free energy surfaces.} In scenario \textbf{A}, the free energy barrier present at $\lambda=0$ is absent at other $\lambda$ states, so the system can go around the barrier by alternating sampling between alchemical and configurational space, e.g.
@@
 \label{sec:atm}
-\egnote{This section will be removed and some of the content dispersed among the relevant sections of the paper}
 
@@
     \includegraphics[width=0.95\linewidth]
-    {figures_atm/atm_scheme_abfe_rbfe.pdf}
+    {paper/figures_atm/atm_scheme_abfe_rbfe.pdf}
     \caption{\small Alchemical transfer method illustrated for (\textbf{A}) Absolute Binding Free Energy (ABFE) and (\textbf{B}) Relative Binding Free Energy (RBFE) calculations.
@@
 
-The values of the binding energies sampled at regular intervals at each $\lambda$-state during the ATM alchemical simulation are the inputs for binding free energy estimation (see Section \ref{sec:data_analysis}) through multi-state thermodynamic reweighting algorithms such as MBAR and UWHAM~\cite{shirts2008statistically,Tan2012}.
+The values of the binding energies sampled at regular intervals at each $\lambda$-state during the ATM alchemical simulation are the inputs for binding free energy estimation (see Section \label{sec:data_analysis}) through multi-state thermodynamic reweighting algorithms such as MBAR and UWHAM~\cite{shirts2008statistically,Tan2012}.
 One requirement for these algorithms is the evaluation of the perturbation energy $w(b,\vec{\lambda})$ for each sample $b$ at all $\lambda$-states.
@@
     \includegraphics[width=0.95\linewidth]
-    {figures_atm/atm_alchemical_intermediate.pdf}
+    {paper/figures_atm/atm_alchemical_intermediate.pdf}
     \caption{Illustration of ATM's thermodynamic path.
@@
 The receptor is depicted in blue, and the ligand is shown in gray.
-The faint gray color of the ligand in the alchemical intermediate represents its 50\% interaction strength with the receptor and the bulk of the solvent.
-\egnote{To do: reverse the directions of the arrows}}
+The faint gray color of the ligand in the alchemical intermediate represents its 50\% interaction strength with the receptor and the bulk of the solvent.}
     \label{fig:atm_alchemical_intermediate}
@@
     \centering
-    \includegraphics[width=0.95\linewidth]{figures_atm/atm_lig_alignment.pdf}
+    \includegraphics[width=0.95\linewidth]{paper/figures_atm/atm_lig_alignment.pdf}
     \caption{Illustrative example of the selection of the reference atoms for ATM's ligand alignment restraints between two ligands \textbf{A} and \textbf{B}.
@@
 
-Beyond thermodynamic integration, the TIES method\cite{bhati2017} uses an Ito integral with Iso isomoetry for error propogation.
+Beyond thermodynamic integration, the TIES method\cite{bhati2017} uses an Ito integral with Iso isomoetry for error propagation.
 
@@
 
-Relative free energy calculations carry with them additional unsettled practices relating to the use and careful treatment of dummy atoms (see ``Dummy Atoms'' in Section~\ref{sec:relative-fe-protocol}).
+Relative free energy calculations carry with them additional unsettled practices relating to the use and careful treatment of dummy atoms (see ``Dummy Atoms'' in Section~\label{sec:relative-fe-protocol}).
 In particular, as the work of~\cite{Fleck_2021} highlighted, valence terms for dummy atoms for dummy atoms must be treated far more carefully than previously appreciated in order for dummy atoms' contributions to the free energy to rigorously cancel.
@@
 	\item \href{https://github.com/qusers/qligfep}{QligFEP} offers robust and fast setup of FEP calculations for the software package Q~\cite{jespers2019qligfep}.
-	\item \href{https://github.com/protocaller/ProtoCaller}{ProtoCaller}, a setup tool for the automation of Gromacs free energy calculations~\cite{suruzhon2020protocaller}.
-	\item \href{https://fesetup.readthedocs.io/en/latest/introduction.html}{FESetup} has been developed primarily to setup calculations in AMBER, GROMACS and SIRE~\cite{loeffler2015fesetup}.
         \item 
-    \href{https://github.com/cbc-univie/transformato}{transformato} is a package for setting up single topology alchemical transformations to be run using CHARMM or OpenMM.
+    \href{https://github.com/cbc-univie/transformato}{transformato} is a package for setting up single topology alchemical transformations to be run using CHARMM or OpenMM \cite{karwounopoulos2022relative, karwounopoulos2023calculations, wieder2022alchemical}.
     \item \href{http://www.ties-service.org}{TIES} allows the setup of perturbation maps, hybrid topologies and structures as well as simulation-ready models for NAMD and OpenMM \url{https://github.com/UCL-CCS/TIES} \cite{bieniek2023}.
@@
 	\url{https://github.com/choderalab/pymbar} \cite{shirts2008statisticallya}.
-	\item Cinnabar: Standardising alchemical free energy analysis.
-Includes tools for calculating network-wide analyses and deriving absolute free energies using a maximum-likelihood estimator \url{https://github.com/OpenFreeEnergy/cinnabar}.
-	\item Free Energy Workflows: Sire-specific free energy map analysis using weighted path averages 
-        \item \href{https://gitlab.com/RutgersLBSR/fe-toolkit} FE-Toolkit: collection of packages for free energy analysis, including network-wide MBAR analysis~/cite{}.
-    \url{https://github.com/michellab/freenrgworkflows}.
+	\item \href{https://github.com/OpenFreeEnergy/cinnabar}{Cinnabar}: Standardising alchemical free energy analysis.
+Includes tools for calculating network-wide analyses and deriving absolute free energies using a maximum-likelihood estimator \cite{richard_gowers_2025_15678719}.
+        \item \href{https://gitlab.com/RutgersLBSR/fe-toolkit}{FE-Toolkit}: collection of packages for free energy analysis, including network-wide MBAR analysis~\cite{giese2025fe}.
     \item \href{http://www.ties-service.org}{TIES}: Perform statistical analysis of the output of alchemical ensemble simulations and report reliable predictions along with statistical uncertainties for a variety of free energy estimators \url{https://github.com/UCL-CCS/TIES_MD} \cite{bieniek2023}.

### 2025-11-22 paper/manuscript.tex

--- 2025-11-05 paper/manuscript.tex
@@
 The type of restraints used in SepTop \cite{rocklin2013separated,Baumann_2023} and ATM \cite{azimi2022relative} avoid this problem.
-Additionally, SCDT approaches, such as TIES~\cite{bhati2017}, [MENTION OTHERS TOO], etc., circumvent this issue by design as they do not need any restraints to hold MCSS together between the two end-points.
+Additionally, SCDT approaches, such as TIES~\cite{bhati2017}, circumvent this issue by design as they do not need any restraints to hold MCSS together between the two end-points.
 
@@
 \textbf{JWE} Contributed to ``Conserved binding site waters can play an important role in binding free energies'' and ``Handling poorly sampled water networks'' in Sec. ~\ref{subsec:binding}.\\
-\textbf{EG} Wrote the updated Sec.~\ref{sec:theory} and Sec.~\ref{sec:multiple-chemical-states},  updated Sec.~\ref{sec:standardstate-restraints}, and and contributed to the discussion and review of many sections.\\
+\textbf{EG} Wrote the updated Sec.~\ref{sec:theory} and Sec.~\ref{sec:multiple-chemical-states},  updated Sec.~\ref{sec:standardstate-restraints}, and contributed to the discussion and review of many sections.\\
 \textbf{MG} Contributed to ``Changes in net charge can be challenging/problematic'' in Sec. \ref{subsec:net charge}. \\

### 2025-11-23 paper/manuscript.tex
Changes to uncertainty quantification section and suggestions for restructuring by @mrshirts

@@
 Alchemical free energy calculations are a useful tool for predicting free energy differences associated with the transfer of molecules from one environment to another.
-The hallmark of these methods is the use of "bridging" potential energy functions representing \emph{alchemical} intermediate states that cannot exist as real chemical species.
+The hallmark of these methods is the use of ``bridging'' potential energy functions representing \emph{alchemical} intermediate states that cannot exist as real chemical species.
 The data collected from these bridging alchemical thermodynamic states allows the efficient computation of transfer free energies (or differences in transfer free energies) with orders of magnitude less simulation time than simulating the transfer process directly.
@@
     \textbf{A}: Partition coefficient such as $\log P$ or $\log D$ depend on a change in free energy between different phases; here, as an example the partition coefficient between methanol and water is shown.
-\textbf{B}: Change in free energy due to a conformational change of the molecule across a high barrier.
+\textbf{B}: Change in free energy due to a conformational change of the molecule across a high barrier, such as the torsion reaching from the aromatic ring to the hydroxyl group.
 \textbf{C}: Absolute free energies of binding of a small molecule to a host (e.g.
@@
 This Best Practices guide focuses on providing a good starting point for new practitioners and a reference for experienced practitioners. 
- For this purpose we provide a convenient checklist (Sec.~\ref{sec:checklist}) to help ensure all calculations comply with currently-understood best practices for alchemical simulation and analysis.
+ For this purpose we provide a convenient checklist (Sec.~\ref{sec:checklist}) to help ensure all calculations can take advantage of currently-understood best practices for alchemical simulation and analysis.
 Where the best practices are currently not certain, we highlight areas where further research is needed to identify an unambiguous recommendation.
- This guide can also serve as a set of best practices to ensure simulation robustness and reproducibility which reviewers may wish to consider as they evaluate papers.
-
- We assume that novice practitioners have at least moderate experience with molecular simulation concepts and use of simulation packages. 
- Furthermore, basic familiarity with the principles of molecular mechanics, molecular dynamics simulations, statistical mechanics, and the biophysics of protein-ligand association are essential.
+ This guide can also serve as a set of best practices to ensure simulation robustness and reproducibility which reviewers may wish to consider as they evaluate papers including alchemical free energy calculations.
+
+In this best practices document, we assume that novice practitioners have at least moderate experience with molecular simulation concepts and use of simulation packages. 
+We also assume basic familiarity with the principles of molecular mechanics, molecular dynamics simulations, statistical mechanics, and the biophysics of protein-ligand association.
 If you feel unfamiliar with some of these concepts, good starting points can be found in these references~\cite{braun2019best, grossfield2018best, klimovich2015guidelines, shirts2012best}.
@@
     \item Is my problem suitable for an alchemical calculation? 
-     \item How do I select an appropriate alchemical protocol? 
+     \item How do I select an appropriate simulation protocol? 
      \item What software tools are available to perform alchemical calculations? 
@@
 In this guide, some other background information may be needed depending on the nature of the alchemical project.
-For example, often, if binding poses are not known, docking calculations can be used to generate an initial small molecule binding pose to start alchemical simulations.
-This will require some basic familiarity on how to perform docking to generate reasonable simulation starting points~\cite{grinter2014challenges}.
+For example, often, if binding poses are not known, docking calculations or co-folding models can be used to generate an initial small molecule binding pose to start alchemical simulations.
+This will require some basic familiarity on how to perform docking or using generative models.
+to generate reasonable simulation starting points~\cite{grinter2014challenges}.
 
@@
 In the remainder of this paper, we will cover topics that are key to the preparation (Sec.~\ref{sec:prerequisites}), choice and use of correct protocols (Sec.~\ref{sec:simulation_protocol_choice}), and finally the best practices that should be used in the analysis of alchemical calculations (Sec.~\ref{sec:data_analysis}).
-Particular focus will be given to aspects of the molecular simulations which are unique to alchemical calculations---these include the calculation of transfer free energies (hydration free energies, partition coefficients, etc.), and binding free energies (absolute and relative).
+We will give particular focus to aspects of the molecular simulations which are unique to alchemical calculations---these include the calculation of transfer free energies (hydration free energies, partition coefficients, etc.), and binding free energies (absolute and relative).
 We primarily focus on free energy calculations using simulations performed at \emph{equilibrium} in this Best Practices guide, as best practices for these are more developed, and non-equilibrium techniques may warrant their own guide as such practices evolve.
@@
 We restrict our analysis here to repetitions of the same calculation performed with precisely the same force field, as different force fields used to describe the same molecule can lead to wide differences in free energies in some cases.
-Even within this restriction,
-finite computing resources necessarily limit the generated number of uncorrelated samples of potential energy surfaces, and therefore alchemical free energy calculations only give free energy estimates to within finite precision.
-An important consideration is how reproducible alchemical free energy calculations are in practice.
-In simple cases, such as absolute hydration free energies of small organic molecules, or relative hydration free energy calculations between structurally similar small organic molecules, it should be possible to obtain highly precise estimates with a given software package, i.e., with a sample standard deviation under 0.01 kcal/mol~\cite{rizzi2019sampl6}.
+
+Simulations carried out with molecular dynamics or Monte Carlo techniques are fundamentally stochastic calculations, and two simulations can never match exactly.
+Even though Newton's equations of motion are deterministic, molecular systems are chaotic, and very small changes, like differences in rounding of floating point numbers due to different software versions or architectures, let alone different random number seeds used to initialize the velocities will lead to different results obtained with each repetition of a simulation.
+Any sort of agreement will only be up to this stochastic source of uncertainty, and one needs to have an understanding of how accurate one needs comparison to be, as well as a proper handling of statistics to decide if differences in simulations are statistically significant or not.
+Because free energy differences are calculated from ensemble averages, collecting more data will result in improved agreement in almost all cases, thanks to the central limit theorem, but the amount of simulation needed to converge to a desired level of statistical precision varies dramatically on the system.
+So the key consideration is how reproducible alchemical free energy calculations are in practice, and how large a difference between two measurements is required to be scientifically useful.
+
+There are two types of reproducibility that are important to examine.
+One is reproducibility between different research groups attempting to reproduce the same results, which is relevant when new users might be trying to see how well their calculations match with published calculations.
+The second is reproducibility between one researcher running the same calculation, up to differences in starting configurations, multiple times.
+This can be particularly relevant in determining the statistical uncertainty of the calculation, as discussed in Sec~\ref{subsec:uncertainty}.
+
+%Even within this restriction, finite computing resources necessarily limit the generated number of uncorrelated samples of potential energy surfaces, and therefore alchemical free energy calculations only give free energy estimates to within finite precision.
+
+The differences in free energies between different research groups, due to differences in choices of simulation parameters, simulation packages, and analysis treatments can be significant.
+Issues of reproducibility of free energy calculations across different simulation packages have recently attracted attention ~\cite{loeffler2018reproducibility,rizzi2019sampl6, wade2022, herz2025modular}.
+Greater variability is expected between simulations carried out between packages, as opposed to ones carried out with the same package, due to methodological differences such as integrators, thermostats, barostats, treatment of long-range electrostatics, and potentially other factors.
+For absolute and relative hydration free energies of small organic molecules a variability of ca.
+0.2 kcal/mol between popular simulation packages has been reported~\cite{loeffler2018reproducibility}.
+In the recent SAMPL6 SAMPLing challenge a larger variability of 0.3 to 1.0 kcal/mol was noted in the computed absolute binding free energies of host/guest systems even though the study sought to use identical input and simulation parameters~\cite{rizzi2019sampl6} and, in many cases, single-point energies were identical or nearly so.
 For more complex use cases such as protein-ligand binding free energies the repeatability is often substantially worse~\cite{rizzi2019sampl6}.
-
-MD trajectories are extremely sensitive to their starting conditions and hence one-off simulations can never be reproducible~\cite{coveney2016, MDbook2025, caves_locally_1998}.
+For benchmark simulations of some relative binding free energies, differences of {\color{red}Can we put some more citations here?}
+
+A different situation is the amount of reproducibility researchers should expect when performing repeated simulations of their own systems, perhaps with the same starting point.
+In simple cases, such as absolute hydration free energies of small organic molecules, or relative hydration free energy calculations between structurally similar small organic molecules, it should be possible to obtain highly precise estimates with a given software package, i.e., with a sample standard deviation under 0.01 kcal/mol possible~\cite{rizzi2019sampl6}, and agreement under 0.05 kcal/mol routine for simulations lasting 10's of ns~\cite{paliwal2011benchmark} {\color{red}This could use some more citations here from other studies.} However, for complex systems, the reproducibility can be signficantly worse, as two simulation of the same system can explore very different configurational ensembles.
+If the timescales of the slow scales of motion are comparable to or longer than the timescales of the simulation, then differences of up to multiple kcal/mol can be possible for absolute and even relative free energy calculations.
+
+In fact, the amount of reproducibility between repeated calculations is an important measure of whether a particular choices of simulation methods and approaches are appropriate.
+Large differences between results carried out with different initial velocities, or better yet started from points along a single initial long simulations can indicate that the sampling of the system is particularly hard and will require additional simulation and/or advances sampling methods.
+In fact, journals such as the Journal of Chemical Information and Modeling require that any results be confirmed by simulations run in at least triplicate.
+The question of how to combine results from ensembles of simulations for lowest bias and best reporting of statistical uncertainty are discussed in section~\ref{subsec:uncertainty}.
+
+%MRS: some good points here below, but this isn't the right place, since this is explicitly talking about reproducibility between simulations, i.e.
+if you run a simulation, how different do you expect it to be.
+This MOTIVATES a discusion of ensemble averaging in some other section for robustness, but should be  separate sections discussing ensembles vs.
+simgle trajectories, putting all these differences in the same place, rather than sprinkled throughout.
+Also, some of the claims are too broad as stated, and it included information about free energy calculation methods that doesn't fit here.
+
+%MD trajectories are extremely sensitive to their starting conditions and hence single simulationscan never be reproducible~\cite{coveney2016, MDbook2025, caves_locally_1998}.
 Thus, it is recommended to perform ensemble simulations in order to perform ensemble averaging to obtain thermodynamic quantities in a reliable manner.
@@
 Studies have shown that, in general, an ensemble size of 5 and 10 members is a reasonable starting point for alchemical RBFE and ABFE calculations, respectively, but recommendations are provided to adjust them based on some rules of thumb~\cite{bhati2022, bhati2025}.
-The sample standard deviation will give a crude estimate of the reliability of the estimates, and whether the precision is sufficient for the problem at hand.
+
+%MRS: Should be in a separate section. 
+%The sample standard deviation will give a crude estimate of the reliability of the estimates, and whether the precision is sufficient for the problem at hand.
 
@@
 
-Note that these types of statistical differences concern calculations carried out with a single software package, but simulation package variations can introduce additional discrepancies.
-Such issues of reproducibility of free energy calculations across different simulation packages have attracted attention ~\cite{loeffler2018reproducibility,rizzi2019sampl6, wade2022, herz2025modular}.
-Greater variability is expected between packages due to methodological differences such as integrators, thermostats, barostats, treatment of long-range electrostatics, and potentially other factors.
-For absolute and relative hydration free energies of small organic molecules a variability of ca.
-0.2 kcal/mol between popular simulation packages has been reported~\cite{loeffler2018reproducibility}.
-In the recent SAMPL6 SAMPLing challenge a larger variability of 0.3 to 1.0 kcal/mol was noted in the computed absolute binding free energies of host/guest systems even though the study sought to use identical input and simulation parameters~\cite{rizzi2019sampl6} and, in many cases, single-point energies were identical or nearly so.
 % Add something on RBFE calculations across engines.
@@
 
-\paragraph{Ring breaking and forming.} \label{subsec: Ring breaking and forming}
+\paragraph{Ring breaking and forming} \label{subsec: Ring breaking and forming}
 Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms cancelling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
@@
 simulations in states at $\lambda=0$ and $\lambda=1$ may be initialized with the independently generated ligand coordinates.
-Another example would be using true apo and holo protein structures as the end states to initialize simulations for absolute binding free energy calculations.
-Bhati \textit{et al.} recently published two studies directly comparing the equilibrium (EQ) and NEQ approaches at unprecedented scales for both RBFE and ABFE calculations using their ensemble simulation-based TIES method~\cite{wan2023eqvsneq, bhati2025}.
+
+%MRS: I think this below is a little too detailed for a general introduction of NEQ, as we should be talking about general strategies, not specific ones.
+Probably we should be putting in a few references for key NEQ uses so readers could follow up, however, THat would be useful to have here.
+
+%MRS: was this sentence the one with the violated assumptions?  please check, I am note sure. 
+%Another example would be using true apo and holo protein structures as the end states to initialize simulations for absolute binding free energy calculations.
+
+%Bhati \textit{et al.} recently published two studies directly comparing the equilibrium (EQ) and NEQ approaches at unprecedented scales for both RBFE and ABFE calculations using their ensemble simulation-based TIES method~\cite{wan2023eqvsneq, bhati2025}.
 Further, they systematically studied the effect of various parameters involved in setting up calculations for both approaches and recommended optimal protocols in each case for industrial applications.
@@
 Single replica schemes and non-equilibrium schemes are not as established yet because of potential failure modes, but are very promising for use in the near future.
+%MRS: doesn't below here. 
 %Further, ensemble simulations are recommended irrespective of the choice of sampling scheme to ensure reliability of results~\cite{bhati2018, bhati2019, wan2020fep+, wade2022, wan2021uq, bhati2025, wan2023eqvsneq}.
+
+%MRS: This should be broken up into two sections - 
+% The second of which is "what information should be saved - that should really be separated. 
+% The first one should be perhaps, " How much simulation data should I collect" and THIS is where we should put discussions of ensembles versus long siulations.
+I think the question "how long should I run?" is not the right questions, because as Agastya points out, how to break them up between MULTIPLE runs is an important discussion we should have.
 
@@
 In the limit of sufficiently large data, perhaps several hundred or thousand independent points, exhaustive analyses of model data has shown that the error estimates of BAR, MBAR, and TI are all quite reliable when there is good overlap between intermediate states.
-For EXP, the number of samples at which they become reliable is significantly higher, especially if overlap between intermediate states is lower.
-However, in the limit of moderate or small amounts of data (tens or hundreds of independent samples) these errors become somewhat unreliable, often off by a factor of 2 from what would get if more data was collected.
-These estimator of uncertainty are all derived from propagation of error formulae from the estimators themselves.
-However, when the underlying data is noisy, this approximation is not very good.
+For example, for a series of hydration free energy calculations of model systems of various sizes analytical estimates of the error, bootstrap estimates, and standard deviations over 100 independent simulations started at different locations along a long initial simulations all agreed within a 5---10\%.~\cite{paliwal2011benchmark} For EXP, the number of samples at which they become reliable is significantly higher, especially if overlap between intermediate states is lower.
+
+However, in the limit of moderate or small amounts of data (tens or hundreds of independent samples) these errors become somewhat unreliable, often off by a factor of 2 from what would get if more data was collected.~\cite{???} These estimators of uncertainty are all derived from propagation of error formulae from the estimators themselves.
+When the underlying data is noisy, this approximation is not very good.
 
@@
 Specifically, if the $\vec{\lambda}$s are chosen such that intermediate states have moderate overlap with their neighbors (i.e.
-between $i$ and $i+1$ and between $i$ and $i-1$, they will \textit{not} have significant overlap with their next nearest neighbors $i+2$ and $i-2$.
+between $i$ and $i+1$ and between $i$ and $i-1$), they will \textit{not} have significant overlap with their next nearest neighbors $i+2$ and $i-2$.
 Thus MBAR does not actually get significant information from these energy differences, so one might as well not even calculate them, and just perform BAR between nearest neighbors.~\cite{paliwal2011benchmark}
-\item TI usually gives similar values as MBAR implemented with sufficient numbers of intermediates, but quadrature errors that are hard to estimate beforehand  can occur if one is not careful.~\cite{paliwal2011benchmark}
+\item TI usually gives similar values as MBAR implemented with sufficient numbers of intermediates, but quadrature errors that are hard to estimate beforehand  can occur if one does not carefully select $\lambda$ values such that the overlap between states is sufficiently high.~\cite{paliwal2011benchmark}
 \item WHAM is an approximation to MBAR, and there are no compelling reasons it should be used.
-If careful, it is not necessarily much worse than the other methods, but it always introduces some degree of binning error.
+If bin widths are careful chosen and there are sufficient samples, it can given similar quality results to have the other methods, but it always introduces some degree of binning error, and determining when errors might occur can be challenging.
 \item Other variants, especially ones that adaptively determine the free energies can be useful in certain circumstances but beyond the scope of a Best Practices article.
@@
 
-To minimize this effect, an initial portion of the simulation is often discarded to \emph{equilibration}~\cite{braun2019best}, with the idea of removing the most heavily biased initial portion of simulation data but retaining the unbiased \emph{production} region that represents a stationary Markov chain process sampling from the desired equilibrium target distribution.
-Because the simulation time required for the atypical initial sampler state to relax toward equilibrium is a property of the specific system being simulated and the specific initial conditions selected, it is simplest to collect data for the whole process and use an automated algorithm to select how much data should be discarded to equilibration in a post-processing step.
+To minimize this effect, an initial portion of the simulation that is determined to be  \emph{equilibration}~\cite{braun2019best} should be omitted, with the idea of removing the most heavily biased initial portion of simulation data, but retaining the unbiased \emph{production} region that represents a stationary Markov chain process sampling from the desired equilibrium target distribution.
+Because the simulation time required for the atypical initial sampler state to relax toward equilibrium is a property of the specific system being simulated and the specific initial conditions selected, it is simplest to collect data for the whole process and use an automated algorithm to select how much data should be discarded to equilibration in a post-processing step, rather than attempting to figure out beforehand how long equilibration will take.
 
@@
 This approach is implemented within the MBAR~\cite{kylebeauchamp2019choderalab} and alchemlyb~\cite{daviddotson2019alchemistry} packages, and is highly recommended for standard practice.
-A very similar approach involves minimising the standard error of the production region~\cite{whiteEffectiveTruncationHeuristic1997, sprattHeuristicsStartupProblem1998, hoadAutomatingWarmupLength2010, clarkRobustAutomatedTruncation2025}.
+A very similar approach involves minimizing the standard error of the production region~\cite{whiteEffectiveTruncationHeuristic1997, sprattHeuristicsStartupProblem1998, hoadAutomatingWarmupLength2010, clarkRobustAutomatedTruncation2025}.
 
@@
 
-Note that the use of the effective reduced potential is not guaranteed to pick up on all slow relaxation processes that may be coupled to the alchemical free energy, but the simplicity of its computation means it is generally appropriate for most cases.
+Note that the use of the effective reduced potential is not guaranteed to detect all slow relaxation processes that may be coupled to the alchemical free energy, but the simplicity of its computation means it is generally appropriate for most cases.
 
 \paragraph{Cautions in automating equilibration detection}
-For simulations that are simply not long enough to contain a large number of samples from true equilibrium either because they are very short or contain slow processes, this procedure cannot completely remove the bias.
-In such cases, this approach simply selects the final portion of the the simulation, which may be contained in a single substate of configurational space, and may itself lead to biased estimates. 
-This situation can be detected if the equilibration boundary $t_0$ is a significant fraction of the total simulation length $T$, with a good rule of thumb being that $T \gtrsim 20 t_0$.
-If this is not possible, advanced analysis techniques that assume only local equilibrium (rather than global equilibrium) such as the TRAM estimators~\cite{mey2014xtram,wu2016multiensemble,nuske2017markov} may be more appropriate, but are beyond the scope of this paper.
+Procedures like those described above can fail to remove bias if the simulations are simply not long enough to contain a large number of samples from true equilibrium either because they are very short, or because one is simulating slow processes.
+In such cases, this approach often simply selects a small final portion of the simulation, which may be contained in a single substate of configurational space, which may itself lead to biased estimates.
+This situation can be detected if the equilibration boundary $t_0$ is a significant fraction of the total simulation length $T$, with a good rule of thumb being that under well-sampled situations $T \gtrsim 20 t_0$.
+If identifying an equilibration time that is only a fraction of the total time is not possible, advanced analysis techniques that assume only local equilibrium within kinetically separated metastable states, rather than global equilibrium over the entire phase space, such as the TRAM estimators~\cite{mey2014xtram,wu2016multiensemble,nuske2017markov} may be more appropriate, but are beyond the scope of this paper.
 
@@
 If coupled simulations are used (such as a Hamiltonian replica exchange simulation), the replicas should undergo equivalent random walks in alchemical space, and the replicas can be subsampled with the same $g$ to generate an equal number of uncorrelated samples at each alchemical state.
-Conveniently, maximising the effective sample size for automated equilibration detection produces an appropriate estimate of $g$ over the production region for automating this process.
+Conveniently, maximizing the effective sample size for automated equilibration detection produces an appropriate estimate of $g$ over the production region for automating this process.
 
@@
 Hitherto, MD practitioners have been assuming the validity of the ergodic hypothesis that allows one to equate ensemble averaging to time averaging in the limit of time approaching infinity to predict thermodynamic properties using one-off simulations.
-However, it is often ignored that this limit is on the order of a Poincaré recurrence time, which is truly astronomical and well beyond the reach of today's supercomputers.
+.
 In more rigorous terms, ergodic theory necessitates that, if a dynamical system is to approach equilibrium, it is not sufficient for it to be ``ergodic'' but it must also be ``mixing'' (these terms refer to the different classes of dynamical systems provided by ergodic theory; please see the references for details)~\cite{coveney2016, MDbook2025}.
@@
 
-Beyond thermodynamic integration, the TIES method\cite{bhati2017} uses an Ito integral with Iso isomoetry for error propogation.
+Beyond thermodynamic integration, the TIES method\cite{bhati2017} uses an It\^{o} integral with Iso isomoetry for error propogation.
 
@@
 In some cases, slow correlation time may not be expected and therefore not a feature known in advance.
-To this end, you should always look at all simulation data available and check convergence behaviour for each free energy estimate.
+To this end, you should always look at all simulation data available and check convergence behavior for each free energy estimate.
 Comparing the free energy trajectories as a function of the simulation time in the forward and reverse time direction is a useful convergence test~\cite{klimovich2015guidelines}.
@@
 Accordingly, the uncertainty of the free energy difference between the states $i$ and $j$ will be smaller when ${O}_{i,j,i\ne j}$ is larger (and thus the values in the main diagonal (${O}_{i,j,i=j}$) are smaller).
-Ensemble simulations generate wider distributions of energies and hence are expected to increase the overlap between neighbouring states~\cite{wade2022}.
-In order to obtain a reliable estimate of the free energy all neighbouring states must be connected, i.e.
+Ensemble simulations generate wider distributions of energies and hence are expected to increase the overlap between neighboring states~\cite{wade2022}.
+In order to obtain a reliable estimate of the free energy all neighboring states must be connected, i.e.
 there must be sufficient overlap between the samples of these states, such that ${O}_{i,j,i\ne j}\ge$ threshold).
@@
 \includegraphics[width=0.90\columnwidth]{figures/fig12_overlap/Figure.pdf}
-\caption{\label{fig:overlap} \textbf{Overlap matrices:} Visualising overlap matrices can help with assessing the quality of simulation data.
+\caption{\label{fig:overlap} \textbf{Overlap matrices:} Visualizing overlap matrices can help with assessing the quality of simulation data.
 (\textbf{A}) shows good overlap with all first off-diagonal entries well above 0.03, the suggested threshold, (\textbf{B}) is an example of mediocre overlap with good overlap at lower $\vec{\lambda}$ values and poor overlap at high $\vec{\lambda}$ values.
@@
 Fig.~\ref{fig:overlap}\textbf{A}, \textbf{B}, and \textbf{C} shows examples of good, mediocre, and poor overlap respectively.
-For Fig.~\ref{fig:overlap} \textbf{A}, the probability to find a sample from state $i$ in its neighbouring state $j$ is about 0.2 for all states adjacent to the main diagonal, and hence the overall connectivity is good.
+For Fig.~\ref{fig:overlap} \textbf{A}, the probability to find a sample from state $i$ in its neighboring state $j$ is about 0.2 for all states adjacent to the main diagonal, and hence the overall connectivity is good.
 In the case of Fig.~\ref{fig:overlap} \textbf{B}, the overlap is strongly diminishing in the lower right corner, raising concerns regarding the reliability of the free energy estimate obtained.
-For Fig.~\ref{fig:overlap} \textbf{C}, the state at $\vec{\lambda}$ index = 6 is connected to neither of its neighbouring states.
+For Fig.~\ref{fig:overlap} \textbf{C}, the state at $\vec{\lambda}$ index = 6 is connected to neither of its neighboring states.
 While this does not necessarily imply that the result for this perturbation is wrong, the energy estimate must at least be considered as highly unreliable.
@@
 \cite{hahn2010measuring} Since the second order estimate converges slower than the first, the overall convergence of the free energy estimate can be quantified as the normalized difference between the two moments of the Fermi function.
+
 In practice, analytical error estimate of the BAR estimator \cite{shirts2003equilibrium} gives an indication when the overlap is lacking: the estimated error rapidly grows with the increase in the dissipated work.
-\cite{shirts2005comparison} Bhati \textit{et al.} recently studied the effect of such overlaps on the accuracy of predicted RBFEs~\cite{wan2023eqvsneq} as well as ABFEs~\cite{bhati2025} but found that overlap (or lack of it) does not capture the reliability of calculations very well.
+\cite{shirts2005comparison} Bhati~\textit{et al.} recently studied the effect of such overlaps on the accuracy of predicted RBFEs~\cite{wan2023eqvsneq} as well as ABFEs~\cite{bhati2025} but found that overlap (or lack of it) does not capture the reliability of calculations very well.
 They recommended an alternative metric, ``distance'' defined as the difference between the inner extremes of the two work value distributions, which correlates quite well with the unsigned errors in free energy predictions and hence was found a better indicator of the reliability of NEQ calculations.
@@
 If the system is not mixing between states, then the states are insufficiently close for mixing, or else there are bottlenecks in the configurational sampling that limit the accuracy.
-\item \textbf{Behaviour of correlation times:} Correlation time that does not vary relatively smoothly as a function of $\vec{\lambda}$.
+\item \textbf{Behavior of correlation times:} Correlation time that does not vary relatively smoothly as a function of $\vec{\lambda}$.
 Discontinuities in correlation time with $\vec{\lambda}$ indicate that the system is sampling significantly different configurations with only small changes to the Hamiltonian changes.
@@
 \item \textbf{Dependence of the free energy on initial configuration of the system.} Ensemble average properties should not depend on the starting point.
+This is a particular instance of the utility of running multiple independent simulations; if there is statistically significant differences between the independently run simulations, there is likely a lack of sampling.
 \item \textbf{Torsional sampling} Torsions with multiple low-energy minima where some of these minima are visited rarely or not at all.
@@
 \item \textbf{Free energy dependence on $\vec{\lambda}$} The free energy difference between states should vary relatively smoothly with $\vec{\lambda}$.
-If it varies drastically, then either there need to be finer sampling in $\vec{\lambda}$ in this region, or there are sampling problems there.
-\item \textbf{Convergence of free energy} The free energy should clearly converge as a function of simulation time (Fig. ~\ref{fig:convergence_forward_reverse}).
+If it varies drastically, then either there needs to be finer sampling in $\vec{\lambda}$ in this region, or there are sampling problems present in at least these simulations (if not simulations at other $\lambda$.
+\item \textbf{Convergence of free energy} The free energy should clearly converge as a function of simulation time (Fig.
+~\ref{fig:convergence_forward_reverse}).
+Convergence of the free energy as a function of time is necessary, but not sufficient.
+It does not guarantee that simulations have run long enough, but lack of convergence means they certainly have not sampled the relevant configuration space.
 \item If using non-equilibrium methods, \textbf{is the result independent of the speed at which the non-equilibrium change is performed}?
@@
 %\item Other Diagnostics for non-equilibrium methods, robustness for switching time etc.?
-\item \textbf{Visualization of data} In general, inspect output data such as energies and visualize the simulation trajectories and assess if they match your expectations.
-Many issues can be spotted by a straight forward visualization.
+\item \textbf{Visualization of data} In general, always inspect output data such as kinetic or potential energies, pressures, and volumes as a function of time, and visualize the simulation trajectories and assess if they match your expectations.
+Many issues can be spotted by a straightforward visualization.
 \end{itemize}
@@
 \subsection{Current limitations}
+
+% notthing here?  It would actually be really nice to summarize this.
 

### 2025-12-01 paper/manuscript.tex
Fix typo Eq51 in double transformation RBFE section
Updates from Overleaf

@@
 
-Notably, this document---version 2.0 of the original guide--focuses specifically on best practices on single alchemical free energy calculations.
+Notably, this document---version 2.0 of the original guide---focuses specifically on best practices on single alchemical free energy calculations.
 A companion guide, Best Practices for Multiple Alchemical Free Energy Calculations, covers the considerations involved in high-throughput or large-scale campaigns, such as automation, perturbation network design, and campaign-level data management and analysis.
@@
 In this guide, some other background information may be needed depending on the nature of the alchemical project.
-For example, often, if binding poses are not known, docking calculations or co-folding models can be used to generate an initial small molecule binding pose to start alchemical simulations.
+For example, often, if binding poses are not known, docking calculations or co-folding models can be used to generate an initial small molecule binding pose to start alchemical simulations\cite{scheen2025leveraging, thaler2025boltz}.
 This will require some basic familiarity on how to perform docking or using generative models.
@@
 Large differences between results carried out with different initial velocities, or better yet started from points along a single initial long simulations can indicate that the sampling of the system is particularly hard and will require additional simulation and/or advances sampling methods.
-In fact, journals such as the Journal of Chemical Information and Modeling require that any results be confirmed by simulations run in at least triplicate.
+In fact, journals such as the Journal of Chemical Information and Modeling require that any results be confirmed by simulations run in at least triplicate~\cite{soares2023guidelines}.
 The question of how to combine results from ensembles of simulations for lowest bias and best reporting of statistical uncertainty are discussed in section~\ref{subsec:uncertainty}.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% What simulation protocol should I choose   %
+% Choice of simulation protocol and implementation of alchemical transformations   %
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\section{What simulation protocol should I choose?}
 \section{Choice of simulation protocol and implementation of alchemical transformations}
-  \sbnote{Emilio suggested to make the title more general; this is my suggestion!}
 \label{sec:simulation_protocol_choice}
@@
 \end{equation}
-is the specific equilibrium binding constant between the receptor and the protonated form of the ligand, and similarly for $K_b(A^-)$.
+is the specific equilibrium binding constant between the receptor and the protonated form of the ligand, and similarly the unprotonated form for $K_b(A^-)$.
 The specific $K_b$'s are the binding constants that would be obtained from ABFE calculations for those species.
@@
 To ensure there is sufficient data available for reliable decorrelation and estimation of free energy differences, it is recommended that the effective number of uncorrelated samples $N_{\mathrm{eff}} \ge  \sim 50$ if the BAR or MBAR estimators (discussed below in sec~\ref{subsec:estimators}) are used; the number may need to be much higher with alternate estimators.
+
+\subsection{Uncertainty estimation v2}
+
+%% Being edited on Nov 30. 2025
+
+%% MS: This is a very loose proposal of how one might address some of the issues on uncertainty quantification.
+It is only meant to help get the ball rolling, per the discussion in https://github.com/alchemistry/alchemical-best-practices/issues/150#issuecomment-3567254644.
+Otherwise, it can be mostly discarded.
+The important points in my view are:
+%
+% 1.
+clearly expressing the dependence of the density as a function (t, x_0), and what assumptions are made about its behavior; without this, I see no way of making statements about uncertainty quantification, as these all require some statistical theorem, like a Law of Large Numbers, or a Central Limit Theorem.
+% 2. a definition of an ensemble
+%
+% and to a lesser extent
+%
+% - expression for the sample variance in the case of i.i.d. or (conditionally iid) random variables,
+% - expectation of the sample variance in the case of correlated samples
+% - some definition of the asymptotic error
+% - some discussion about the two central limit theorems being used: the one for (c)iid RVs (e.g., Lindeberg CLT) and a temporal CLT for Markov processes.
+In the latter case, we would expect the error to decay like 1/sqrt(# timesteps), but this does not occur.
+I think we should highlight this.
+
+Consider a molecular dynamics trajectory $Z (t ; z_0)$ started at some initial configuration $z_0 = (x_0, v_0)$, where $x_0 \in \mathbb{R}^{3 N}$ denotes the initial positions of the system and $v_0 \in \mathbb{R}^{3 N}$ its initial velocities.
+If $z = (x_0, v_0)$ is drawn from some initial density $\rho_0 (z)$, the density time $t$ will be
+\begin{equation}
+  \rho_t (z' | z_0) = \int_{\mathbb{R}^{3 N} \times \mathbb{R}^{3 N}}
+  \rho_0 (\mathrm{d} z) P_t (z' | z) .
+\end{equation}
+Here, $P_t$ denotes the transition kernel, determined by the molecular dynamics integrator.
+The distribution of the velocities is usually ignored, since in the presence of a thermostat they are kept distributed roughly according to a Maxwell-Boltzmann distribution, so we only consider the $x$ component.
+In practice, the initial distribution $\rho_0 (x) \mathrm{d} x$ is seldom more than a Dirac-$\delta$, $\delta_{\{ x_0 \}} (\mathrm{d} x)$, or a mixture thereof, e.g.
+$\sum_{i = 1}^n \delta_{\{ x_i \}} / n$.
+Any "better" initial density would require some knowledge about the system at hand.
+We henceforth consider only the density at time $t$ starting at, denoted $\rho_t(x | x_0)$.
+
+It is difficult to make general statements about the behavior of $\rho_t (x | x_0)$.
+For systems not involving a protein, it may be possible to identify clear "plateaus'' of convergence" \footnote{The plateau can be mathematically defined as a quasi-stationary distribution, but should we say this?
+Could cite Art Voter at Los Alamos, maybe even Freidlin-Wentzell for the original quasipotential.}, i.e.
+intervals $[T_a, T_b]$, $T_a \ll T_b$, within which $\rho_t (x | x_0)$ does not vary much for $t \in [T_a, T_b]$, and is roughly invariant to $x_0 \in B$ in some appropriate "starting domain".
+Consider, for instance, the estimation of free energies of hydration for small molecules [cite Shirts paper on free energies of hydration].
+In this case, the free energy estimates behave in accordance with classical statistical theory, and the density $\rho_t (x | x_0)$ should not vary much along coordinates relevant to the free energy estimates.
+
+%  MS: The above paragraph is too vague, but some kind of discussion to this effect might be necessary to make sense of why some free energy calculations, such as solvation free energy calculations or computation of free energy landscapes for small molecules, have different statistical properties than more complicated calculations, like absolute binding free energy calculations.
+
+One should not confuse a plateau in such coordinates with having explored the entirety of the configuration space $\Omega$; most biomolecular systems, even ones involving only a single ligand, possess a whole hierarchy of timescales, the longest of which can easily exceed achievable simulation times.
+Importantly, there is no that $\rho_t (x | x_0)$ converges as $t \rightarrow \infty$ to some unique invariant distribution $\propto \exp (-\beta H (x))$, nor that $p_t (x | x_0)$ is invariant to $x_0$.
+
+% The following footnote feels out of place, but might address some comments by A. Bhati re: the Poincaré recurrence theorem.
+% \footnote{(Luckily, it is not necessary for simulations to visit every configuration $x \in \Omega$ in order to exhibit a regime of convergence.
+For instance, two configurations $x$ and $x'$ may be considered equivalent if they differ only by the permutation of water molecules.
+Nature too faces the same difficulty that not all states are accessible on timescales we can observe -- it will take an astronomically long time before all the gas particles in a room happen to be on one half-side of it -- but this does not prevent us from measuring the temperature of (most) rooms.)}
+
+For systems involving a protein-ligand complex, it is usually the case that $\rho_t (x | x_0)$ is sensitive to $x_0$ and additionally may not exhibit a clear plateau on any accessible time $T > 0$ for "most" $x_0$.
+In such a case, we can still draw independent samples from $\rho_t (\cdot | x_0)$ by simulating a trajectory up to time $t$ and taking the last frame, but this is much less valuable than knowing that $\rho_t (\cdot | x_0) \approx \rho_T (\cdot | x_0)$ for all $t \in [T_{\min}, T]$.\footnote{Should one define the mixing time of a Markov process?
+Cite Peres Markov Chains and Mixing Times, which is freely available?}
+
+In this section, we explore the various possible types of error analysis which can be performed, starting from the simplest, with end hope of making "$\rho_t (x | x_0)$ should not vary much'' a little more precise.
+We start with the following definition.
+
+%% MS: I'm not tied to any particular definition of ensemble, but we should probably make a concrete one:
+% - common x_0 ? 
+% - v ~ MaxwellBoltzmann(Temperature) 
+% - what are the individual trajectories called? "replicates" ? 
+We define an \textbf{ensemble} to be a set of $R \geqslant 1$ trajectories, all started at the same configuration $x_0$, with $v_0$ drawn from the Maxwell-Boltzmann distribution at the appropriate temperature, and run for a chemical time $T$.
+
+\subsubsection{Ensemble error estimates}
+The simplest  and most robust way of obtaining error bars on free energy estimates is by simulating an ensemble $\{ X^{(1)}, \ldots, X^{(R)} \}$ of $R$ molecular dynamics trajectories started at $x_0$ and producing trajectories of free energy differences $\{ \Delta G^{(1)}, \ldots, \Delta G^{(R)} \}$.
+The trajectories $\Delta
+G^{(i)} (\cdot)$ are identically distributed and conditionally independent given some $x_0$.
+For time $t > 0$, the variance of any $\Delta G^{(i)} (t)$ can be estimated with the sample variance $S^2_{(R - 1)} (t)$,
+\begin{equation}
+  S^2_R (t) = \frac{1}{R - 1} \sum_{i = 1}^R (\Delta G^{(i)} (t) -
+  \overline{\Delta G (t)})^2, \qquad \overline{\Delta G (t)} = \frac{1}{R}
+  \sum_{i = 1}^R \Delta G^{(i)} (t) .
+\end{equation}
+Note that, conditionally on $x_0$, we have
+\begin{equation}
+    \mathbb{E}_{x_0}[S_R^2(t)] = \mathrm{Var}_{x_0}(\Delta G^{(1)}(t))
+\end{equation}
+The above equation says that $S_R^2(t)$ is an unbiased estimator for the variance. As $R \rightarrow \infty$, we have $S^2_R (t) \rightarrow \mathrm{Var} _{x_0}(\Delta
+G^{(1)} (t)) = \sigma^2 (t ; x_0)$.
+
+The standard error
+\begin{equation}
+  \mathrm{SE}_R (t ; x_0) = \frac{S_R^2 (t)}{\sqrt{R}}
+\end{equation}
+is the gold standard for reporting the error of a set of free energies obtained from trajectories started from a common $x_0$. Note that $\mathrm{SE}_R
+(t ; x_0) \rightarrow 0$ as $R \rightarrow \infty$ no matter what $S_R^2 (t)$, so long as it is finite, which is true if the $\overline{\Delta G}$ are, say, bounded.
 
@@
 \textbf{LNN}: Helped write the simulation length, stopping conditions, and information saving section. Edited and reviewed alchemical path section.\\
+\textbf{MO} Contributed to Section \ref{subsec:hydration}.\\
+\textbf{CO} Contributed to ``Changes in net charge can be challenging/problematic'' in Sec. \ref{subsec:net charge}. \\
 \textbf{WJP} Contributed to ``Conserved binding site waters can play an important role in binding free energies'' and ``Handling poorly sampled water networks'' in Sec. ~\ref{subsec:binding}.\\
@@
 \textbf{AR}: Created figure~\ref{fig:freeenergytrajectories}, contributed to sections~\ref{sec:theory} and~\ref{sec:simulation_protocol_choice}, and helped edit the paper.\\
-\textbf{CO} Contributed to ``Changes in net charge can be challenging/problematic'' in Sec. \ref{subsec:net charge}. \\
 \textbf{JS}: Created Figs.~ \ref{fig:fig_what_is_alchemy},~\ref{fig:fig_binding_thermodynamic_cycle},~\ref{fig:fig_topology},~\ref{fig:fig_mcss},~\ref{fig:fig_absolute_thermodynamic_cycle},~\ref{fig:scatterplot_analysis}, and an initial draft of~\ref{fig:fig_types_of_networks}.
@@
 \textbf{GT}: Contributed to Sec.~\ref{sec:intro} and~\ref{sec:drugdiscovery}, and helped edit the paper.\\
-\textbf{MO} Contributed to Section \ref{subsec:hydration}
 \textbf{HX}: Contributed Sec.~\ref{subsec:accuracy}, to Sec.~\ref{sec:relative-fe-protocol}, and to Sec.~\ref{sec:are-they-good}.

### 2025-12-01 paper/manuscript.tex
Tweaks from @sheenam1509

@@
 \end{equation}
-where $k_B$ is Boltzmann's constant, $T$ is the absolute temperature, $C_u = N_u/V$ is the number concentration of the solute, $\Lambda_u = h/\sqrt{2 \pi M_u k_B T}$ is the thermal De-Broglie wavelength of the solute, where $h$ is Planck's constant and $M_u$ is the solute's total mass.
+where $k_B$ is Boltzmann's constant, $T$ is the absolute temperature, $C_u = N_u/V$ is the number concentration of the solute, $\Lambda_u = h/\sqrt{2 \pi M_u k_B T}$ is the thermal De Broglie wavelength of the solute, where $h$ is Planck's constant and $M_u$ is the solute's total mass.
 The quantity $q'_u$ in Eq.~(\ref{eq:chemical-potential-definition}) is the molecular partition function of the solute at rest in the solvent
@@
 where $\vec{r}_v$ represents the collection of Cartesian coordinates of the atoms of $N_v$ solvent molecules, $U(x_u, \vec{r}_v)$ is the potential energy of the solvent with one solute molecule in conformation $x_u$ placed in an arbitrary position in a volume $V$.
-When the potential energy function of the solution is separable into an intramolecular potential energy of the solute $U(x_u)$ and a solute-solvent non-bonded interaction energy, the potential of mean force can be written as the sum $\Psi_u(x_u) = U(x_u) + W_v(x_u)$ where $W_u(x_u)$ is the solvent potential of mean force, the solvation free energy of the solute fixed in configuration $x_u$.\cite{gilson2007calculation} However, the potential of mean force definition Eq.~(\ref{eq:solvent-pmf-def}) is valid in general.
+When the potential energy function of the solution is separable into an intramolecular potential energy of the solute $U(x_u)$ and a solute-solvent non-bonded interaction energy, the potential of mean force can be written as the sum $\Psi_v(x_u) = U(x_u) + W_v(x_u)$ where $W_u(x_u)$ is the solvent potential of mean force, the solvation free energy of the solute fixed in configuration $x_u$.\cite{gilson2007calculation} However, the potential of mean force definition Eq.~(\ref{eq:solvent-pmf-def}) is valid in general.
 Finally, $Z_{N_v}$ is the configurational partition function of the pure solvent
@@
 \end{equation}
-The standard state state of a dilute solution is defined as an ideally dilute solution at concentration $C^\circ$.~\cite{levine2009physicalchemistrybook6ed} Hence, Eq.~(\ref{eq:chemical-potential-standard}) is formally correct for real solutions as well.
+The standard state of a dilute solution is defined as an ideally dilute solution at concentration $C^\circ$.~\cite{levine2009physicalchemistrybook6ed} Hence, Eq.~(\ref{eq:chemical-potential-standard}) is formally correct for real solutions as well.
 
@@
 \begin{equation}\label{eq:double-transformation-dg-rbfe}
-\Delta \Delta G_b = -k_B T \ln \frac{Z_{N_v,A}}{Z_{N_v,B}} - (-k_B T \ln \frac{Z_{N_v,B}}{Z_{N_v,A}}) = \Delta G_{\rm bound} - \Delta G_{\rm unbound}
+\Delta \Delta G_b = -k_B T \ln \frac{Z_{N_v,RB}}{Z_{N_v,RA}} - (-k_B T \ln \frac{Z_{N_v,B}}{Z_{N_v,A}}) = \Delta G_{\rm bound} - \Delta G_{\rm unbound}
 \end{equation}

### 2025-12-02 paper/manuscript.tex
Tweaks to proposed uncertainty estimation v2 section from @snarski

@@
 
-%% Being edited on Nov 30. 2025
-
 %% MS: This is a very loose proposal of how one might address some of the issues on uncertainty quantification.
@@
 
-Consider a molecular dynamics trajectory $Z (t ; z_0)$ started at some initial configuration $z_0 = (x_0, v_0)$, where $x_0 \in \mathbb{R}^{3 N}$ denotes the initial positions of the system and $v_0 \in \mathbb{R}^{3 N}$ its initial velocities.
+Consider a molecular dynamics trajectory started at some initial configuration $z_0 = (x_0, v_0)$, where $x_0 \in \mathbb{R}^{3 N}$ denotes the initial positions of the system and $v_0 \in \mathbb{R}^{3 N}$ its initial velocities.
 If $z = (x_0, v_0)$ is drawn from some initial density $\rho_0 (z)$, the density time $t$ will be
@@
 It is difficult to make general statements about the behavior of $\rho_t (x | x_0)$.
-For systems not involving a protein, it may be possible to identify clear "plateaus'' of convergence" \footnote{The plateau can be mathematically defined as a quasi-stationary distribution, but should we say this?
+For systems not involving a protein, it may be possible to identify clear "plateaus'' of convergence \footnote{The relevant mathematical object here is a quasi-stationary distribution, but should we say this?
 Could cite Art Voter at Los Alamos, maybe even Freidlin-Wentzell for the original quasipotential.}, i.e.
@@
 
-One should not confuse a plateau in such coordinates with having explored the entirety of the configuration space $\Omega$; most biomolecular systems, even ones involving only a single ligand, possess a whole hierarchy of timescales, the longest of which can easily exceed achievable simulation times.
-Importantly, there is no that $\rho_t (x | x_0)$ converges as $t \rightarrow \infty$ to some unique invariant distribution $\propto \exp (-\beta H (x))$, nor that $p_t (x | x_0)$ is invariant to $x_0$.
-
-% The following footnote feels out of place, but might address some comments by A. Bhati re: the Poincaré recurrence theorem.
-% \footnote{(Luckily, it is not necessary for simulations to visit every configuration $x \in \Omega$ in order to exhibit a regime of convergence.
-For instance, two configurations $x$ and $x'$ may be considered equivalent if they differ only by the permutation of water molecules.
+One should not confuse a plateau in such coordinates with having explored the entirety of the configuration space $\Omega \subset \mathbb{R}^{3N}$; most biomolecular systems, even ones involving only a single ligand, possess a whole hierarchy of timescales, the longest of which can easily exceed achievable simulation times.
+Importantly, there is no claim that $\rho_t(x|x_0)$ must always converge to some unique density $\propto \exp(-\beta H(x))$, nor that the behavior of $\rho_t (x|x_0)$ is invariant to $x_0$.
+% Of course, it _could_ be the case for some systems, but in general we will have a so-called quasi-stationary distribution; i.e.
+we will consider the distribution conditional on not having exceeded an escape time.
+
+% The following footnote feels out of place, but might address some comments by A. Bhati re: the Poincaré recurrence times.
+% \footnote{(Luckily, it is not necessary for simulations to visit every configuration $x \in \Omega$ in order to exhibit a regime of convergence for the free energy estimates.
+For instance, two configurations $x$ and $x'$ may be considered equivalent for free energy estimation if they differ only by the permutation of water molecules.
 Nature too faces the same difficulty that not all states are accessible on timescales we can observe -- it will take an astronomically long time before all the gas particles in a room happen to be on one half-side of it -- but this does not prevent us from measuring the temperature of (most) rooms.)}
 
-For systems involving a protein-ligand complex, it is usually the case that $\rho_t (x | x_0)$ is sensitive to $x_0$ and additionally may not exhibit a clear plateau on any accessible time $T > 0$ for "most" $x_0$.
+For systems involving a protein-ligand complex, $\rho_t (x | x_0)$ is usually sensitive to $x_0$ and additionally may not exhibit a clear plateau on any computationally feasible time $t > 0$ for "most" $x_0$.
 In such a case, we can still draw independent samples from $\rho_t (\cdot | x_0)$ by simulating a trajectory up to time $t$ and taking the last frame, but this is much less valuable than knowing that $\rho_t (\cdot | x_0) \approx \rho_T (\cdot | x_0)$ for all $t \in [T_{\min}, T]$.\footnote{Should one define the mixing time of a Markov process?
@@
 
-In this section, we explore the various possible types of error analysis which can be performed, starting from the simplest, with end hope of making "$\rho_t (x | x_0)$ should not vary much'' a little more precise.
-We start with the following definition.
+In this section, we explore the various possible types of error analysis which can be performed, depending on the behavior of $\rho_t$. We begin with a definition.
 
@@
 \end{equation}
-The above equation says that $S_R^2(t)$ is an unbiased estimator for the variance. As $R \rightarrow \infty$, we have $S^2_R (t) \rightarrow \mathrm{Var} _{x_0}(\Delta
+The above equation says that $S_R^2(t)$ is an unbiased estimator for the variance. By a law of large numbers, as $R \rightarrow \infty$, we have $S^2_R (t) \rightarrow \mathrm{Var} _{x_0}(\Delta
 G^{(1)} (t)) = \sigma^2 (t ; x_0)$. 
+%% Should we say what kind of LLN (weak/strong) ? Do we provide a reference (standard: Durrett or Chung)?
 
@@
 (t ; x_0) \rightarrow 0$ as $R \rightarrow \infty$ no matter what $S_R^2 (t)$, so long as it is finite, which is true if the $\overline{\Delta G}$ are, say, bounded.
+Consequently, we can estimate the variance of the estimate to essentially arbitrary precision.
+Note, however, that $\Delta G^{(i)}(t)$ will not in general be normally distributed, and the variance may not be a good statistic to capture the error, as the distribution may be multi-modal and heavy-tailed.
+%% MS: The error estimation may further be compounded by a rare event problem, in which case simulating even R = 100 replicates may not be "enough", depending on the purposes. 
+%% Should we also state the central limit theorem?
+
+\subsubsection{Per-replicate error estimates}
+
+In the preceding case, we used minimal assumptions on the distribution of $\Delta G^{(i)}(t)$ to obtain estimates on $\sigma^2(t;x_0)$ by simulating conditionally independent replicates and using the sample variance.
+In this section, we impose much stricter assumptions on the distribution of $\Delta G^{(i)}(t)$, namely that they satisfy a (temporal) Central Limit Theorem (CLT):
+$$ \sqrt{t} ( \Delta G^{(i)}(t) - \Delta G^\star) \Rightarrow N(0, \sigma^2_{\mathrm{asym}}(x_0))$$
+where $\Delta G^\star$ denotes the true free energy difference, $\sigma_{\mathrm{asym}}^2(x_0)>0$ denotes the so-called "asymptotic variance", and $\Rightarrow$ denotes convergence in distribution.
+%% Should we define this?
+Might be a little too technical for the audience.
+Again, Durrett is one of the standard references.
+The CLT assumption implies that 
+$$\sigma^2(t;x_0)=\frac{\sigma^2_{\mathrm{asym}}(x_0)}{t}.$$
+In practice, it may not be the case that $\sigma^2(t;x_0)\propto t^{-1}$, nor even that $\sigma^2(t;x_0)$ is monotonically decreasing with $t>0$.
+In such a case, we are not in a regime where the CLT is satisfied.
+
+%% Here we can discuss time-blocking estimators, such as the batch-means estimator.
+Subsampling would also fit in nicely here.
+Finally, we could relate the sample variance at time $t$ to the mean of per-replicate errors at time $t$; the two should agree -- and they do for some simulations -- but in general the mean of the per-replicate errors is less than the empirical error obtained from the sample variance.
+
+%% We should also discuss the dependence on x_0.
 

### 2025-12-09 paper/manuscript.tex
Updates from Overleaf

@@
 The idea is illustrated in the right branch of Fig.~\ref{fig:fig_topology}.
-Somewhat unfortunately, Axelsen and Li, as well as Ries et al refer to their approach as dual topology.
+Somewhat unfortunately, Axelsen and Li, as well as Ries et al., refer to their approach as dual topology.
 Ries et al suggested to use the term hybrid topology for (single coordinate) dual topology in the sense of Pearlman.
@@
 intervals $[T_a, T_b]$, $T_a \ll T_b$, within which $\rho_t (x | x_0)$ does not vary much for $t \in [T_a, T_b]$, and is roughly invariant to $x_0 \in B$ in some appropriate "starting domain".
-Consider, for instance, the estimation of free energies of hydration for small molecules [cite Shirts paper on free energies of hydration].
+Consider, for instance, the estimation of free energies of hydration for small molecules [cite Shirts paper on free energies of hydration]~\cite{paliwal2011benchmark}.
 In this case, the free energy estimates behave in accordance with classical statistical theory, and the density $\rho_t (x | x_0)$ should not vary much along coordinates relevant to the free energy estimates.
@@
 % - what are the individual trajectories called? "replicates" ? 
-We define an \textbf{ensemble} to be a set of $R \geqslant 1$ trajectories, all started at the same configuration $x_0$, with $v_0$ drawn from the Maxwell-Boltzmann distribution at the appropriate temperature, and run for a chemical time $T$.
+We define an \textbf{ensemble} to be a set $\{ X^{(1)}, \ldots, X^{(R)} \}$ of $R \geqslant 1$ trajectories, all started at the same configuration $x_0$, with $v_0$ drawn from the Maxwell-Boltzmann distribution at the appropriate temperature, and run for a chemical time $T$.
 
 \subsubsection{Ensemble error estimates}
+
 The simplest  and most robust way of obtaining error bars on free energy estimates is by simulating an ensemble $\{ X^{(1)}, \ldots, X^{(R)} \}$ of $R$ molecular dynamics trajectories started at $x_0$ and producing trajectories of free energy differences $\{ \Delta G^{(1)}, \ldots, \Delta G^{(R)} \}$.
@@
 
-Alchemical Free Energy (AFE) calculations have primarily been applied to simulate the stoichiometric, non-reversible binding of organic molecules to well-structured proteins.
+Alchemical Free Energy (AFE) calculations have primarily been applied to simulate the stoichiometric, reversible binding of organic molecules to well-structured proteins.
 However, several areas of drug discovery involve other classes of binding events.

### 2025-12-09 paper/manuscript.tex

@@
 where the first ratio of partition functions involves receptor-ligand complexes and the second only the ligands in solution.
-Eq.~(\ref{eq:double-transformation-Kb-rbfe}), suggests that the relative binding free energy, $\Delta \Delta G_b$, of $B$ relative to $A$ can be expressed as the difference of the free energies of ``mutating'' ligand $A$ into ligand $B$ in the presence and absence of the receptor (Figure \ref{fig:fig_binding_thermodynamic_cycle}):
+Eq.~(\ref{eq:double-transformation-Kb-rbfe}), suggests that the relative binding free energy, $\Delta \Delta G_b$, of $B$ relative to $A$ can be expressed as the difference of the free energies, $\Delta G_{\rm bound}$ and $\Delta G_{\rm unbound}$, of ``mutating'' ligand $A$ into ligand $B$ in the presence and absence of the receptor, respectively (Figure \ref{fig:fig_binding_thermodynamic_cycle}):~\cite{tembre1984ligandreceptor,gilson1997statisticalthermodynamic,mobley2012perspective,cournia2017relative,jiang2019computing,lee_jcheminfmodel_2020_v60_p5595}
 \begin{equation}\label{eq:double-transformation-dg-rbfe}
@@
 \end{equation}
-In practice, each ratio of partition functions must first be turned into an ensemble average to evaluate it.
+
+In practice, each partition function ratio must first be converted into an ensemble average to be evaluated.
 Similar to the preceding case, this step requires manipulating the integrals so that they are carried over the same set of degrees of freedom.
-As discussed in Section \ref{sec:relative-fe-protocol}, this step involves using alchemical topologies capable of describing both ligands.
+As discussed in Section \ref{sec:relative-fe-protocol}, this step involves using alchemical topologies that can describe both ligands.
 For example, among other features, alchemical topologies include atoms common to both ligands and dummy atoms that exist in one but not in the other ligand (Figure \ref{fig:fig_topology}).
-In one limit, called dual coordinate dual topology in Section \ref{sec:relative-fe-protocol}, all atoms of both ligands are present, and the partition functions in Eq.~(\ref{eq:double-transformation-dg-rbfe}) differ in which ligand is coupled to the system.
 A key aspect of constructing a suitable alchemical topology is the requirement that the additional degrees of freedom do not affect the RBFE estimate.
+
+In the dual coordinate dual topology method (discussed in Section \ref{sec:relative-fe-protocol}), which is helpful for pairs of compounds that do not share the same scaffold, all atoms of both ligands are present but intermolecular interactions between the two ligands are turned off, and the partition functions in Eq.~(\ref{eq:double-transformation-dg-rbfe}) differ in which ligand is coupled to the system.
+An example of this formulation is the Separated Topologies (SepTop) approach~\cite{rocklin2013separated,baumann2023broadening}, which decouples one ligand as the coupling of the second is turned on.
+Additionally, the SepTop approach can estimate the relative binding free energies of pairs of ligands that do not share the same binding pose by employing different binding site indicator functions in the definitions of the partition functions of the complexes, $Z_{N_v,RA}$ and $Z_{N_v,RB}$, in Eq.~\ref{eq:double-transformation-dg-rbfe}.
 
@@
 
-The RBFE alchemical transfer formulation is very similar to the ABFE formulation.
-The simulation system includes both ligands, one in solution and the other in the binding site, and the coordinate transformation involves displacing them in opposite directions so that they switch places.
+The RBFE alchemical transfer formulation follows closely the ABFE formulation.
+Similar to the dual coordinate dual topology method, the simulation system includes both ligands, one in solution and the other in the binding site, and the coordinate transformation involves displacing them in opposite directions so that they switch places.
 Formally,~\cite{azimi2022relative} the method derives from Eq.~(\ref{eq:DG0-binding-constant-def}) for each binding constant

### 2025-12-09 paper/manuscript.tex
Minor tweaks

@@
 The idea is illustrated in the right branch of Fig.~\ref{fig:fig_topology}.
-Somewhat unfortunately, Axelsen and Li, as well as Ries et al., refer to their approach as dual topology.
+Somewhat unfortunately, Axelsen and Li, as well as Ries et al refer to their approach as dual topology.
 Ries et al suggested to use the term hybrid topology for (single coordinate) dual topology in the sense of Pearlman.
@@
 intervals $[T_a, T_b]$, $T_a \ll T_b$, within which $\rho_t (x | x_0)$ does not vary much for $t \in [T_a, T_b]$, and is roughly invariant to $x_0 \in B$ in some appropriate "starting domain".
-Consider, for instance, the estimation of free energies of hydration for small molecules [cite Shirts paper on free energies of hydration]~\cite{paliwal2011benchmark}.
+Consider, for instance, the estimation of free energies of hydration for small molecules [cite Shirts paper on free energies of hydration].
 In this case, the free energy estimates behave in accordance with classical statistical theory, and the density $\rho_t (x | x_0)$ should not vary much along coordinates relevant to the free energy estimates.
@@
 % - what are the individual trajectories called? "replicates" ? 
-We define an \textbf{ensemble} to be a set $\{ X^{(1)}, \ldots, X^{(R)} \}$ of $R \geqslant 1$ trajectories, all started at the same configuration $x_0$, with $v_0$ drawn from the Maxwell-Boltzmann distribution at the appropriate temperature, and run for a chemical time $T$.
+We define an \textbf{ensemble} to be a set of $R \geqslant 1$ trajectories, all started at the same configuration $x_0$, with $v_0$ drawn from the Maxwell-Boltzmann distribution at the appropriate temperature, and run for a chemical time $T$.
 
 \subsubsection{Ensemble error estimates}
-
 The simplest  and most robust way of obtaining error bars on free energy estimates is by simulating an ensemble $\{ X^{(1)}, \ldots, X^{(R)} \}$ of $R$ molecular dynamics trajectories started at $x_0$ and producing trajectories of free energy differences $\{ \Delta G^{(1)}, \ldots, \Delta G^{(R)} \}$.
@@
 
-Alchemical Free Energy (AFE) calculations have primarily been applied to simulate the stoichiometric, reversible binding of organic molecules to well-structured proteins.
+Alchemical Free Energy (AFE) calculations have primarily been applied to simulate the stoichiometric, non-reversible binding of organic molecules to well-structured proteins.
 However, several areas of drug discovery involve other classes of binding events.

### 2025-12-09 paper/manuscript.tex
Add NES in Orion to software section

@@
 It only runs on GPUs, using CUDA or OpenCL~\cite{kuhn2020assessment}.
-It is build on top of the open source software packages Sire and BioSimSpace (cf.
+It is built on top of the open source software packages Sire and BioSimSpace (cf.
 below).
+    \item \href{https://www.eyesopen.com/orion}{Orion}, OpenEye's cloud platform, offers relative binding free energy calculations using non-equilibrium switching (Section~\ref{sec:sampling_schemes})~\cite{sorensen2024orion}.
     \item The molecular operating environment (\href{https://www.chemcomp.com/Products.htm}{MOE}) offered by the Chemical Computing Group (CCG) has a tool for performing free energy calculations.

### 2025-12-09 paper/manuscript.tex
Fix citation key error

@@
 where the first ratio of partition functions involves receptor-ligand complexes and the second only the ligands in solution.
-Eq.~(\ref{eq:double-transformation-Kb-rbfe}), suggests that the relative binding free energy, $\Delta \Delta G_b$, of $B$ relative to $A$ can be expressed as the difference of the free energies, $\Delta G_{\rm bound}$ and $\Delta G_{\rm unbound}$, of ``mutating'' ligand $A$ into ligand $B$ in the presence and absence of the receptor, respectively (Figure \ref{fig:fig_binding_thermodynamic_cycle}):~\cite{tembre1984ligandreceptor,gilson1997statisticalthermodynamic,mobley2012perspective,cournia2017relative,jiang2019computing,lee_jcheminfmodel_2020_v60_p5595}
+Eq.~(\ref{eq:double-transformation-Kb-rbfe}), suggests that the relative binding free energy, $\Delta \Delta G_b$, of $B$ relative to $A$ can be expressed as the difference of the free energies, $\Delta G_{\rm bound}$ and $\Delta G_{\rm unbound}$, of ``mutating'' ligand $A$ into ligand $B$ in the presence and absence of the receptor, respectively (Figure \ref{fig:fig_binding_thermodynamic_cycle}):~\cite{tembre1984ligandreceptor,gilson1997statisticalthermodynamic,mobley2012perspective,cournia2017relative,jiang2019computing, Lee_JChemTheoryComput_2020_v16_p5512}
 \begin{equation}\label{eq:double-transformation-dg-rbfe}

### 2025-12-15 paper/manuscript.tex
Mention and cite Macuglia review

@@
  Early work used simple but statistically suboptimal estimators for this: free energy perturbation (FEP) used a simple (but highly biased) estimator based on the Zwanzig relation~\cite{zwanzig1954hightemperature} or numerical quadrature via thermodynamic integration (TI), for which the theory dates back the better part of a century but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamics, merz1989free}.
+Macuglia et al.
+review the evolution of biomolecular alchemical free energy calculations into practically useful tools \cite{macuglia2025dawn}.
 %
@@
 \subsubsection{Relative binding free energy (RBFE)}
+\label{sec:theory-rbfe}
 
@@
 Alchemical free energy calculations involving small molecules seem to achieve, in favorable cases, root mean square (RMS) errors relative to experiment around 1-2 kcal/mol depending on force field, system, and a variety of other factors such as simulation time, sampling method, and whether the calculations employed are absolute or relative.
-A small selection of example datasets and case studies can be found in Sec.~\ref{sec:benchmark} at the end of this document.
+Example datasets and case studies are available in the companion article on best practices for large-scale alchemical free energy calculations.
 However, the domain of applicability is a significant concern~\cite{sherborne2016collaborating, cournia2017relative}, especially for relative calculations, which typically require a high quality and usually experimental bound structure of a closely
@@
 In relative free energy calculations, however, one has to transform, e.g., ligand A into ligand B etc.
-(see \sbnote{appropriate cross-reference!}).
+(see Sec.~\ref{sec:theory-partitioning}).
 The required alchemical transformations can be realized in several ways, each with distinct advantages and disadvantages.

### 2025-12-23 paper/manuscript.tex
Tweaks from @mrshirts

@@
 The data collected from these bridging alchemical thermodynamic states allows the efficient computation of transfer free energies (or differences in transfer free energies) with orders of magnitude less simulation time than simulating the transfer process directly.
-While these methods are highly flexible, care must be taken in avoiding common pitfalls to ensure that computed free energy differences can be robust and reproducible for the chosen force field, and that appropriate corrections are included to permit direct comparison with experimental data.
-
-In this paper, we review current best practices for several popular application domains of  alchemical free energy calculations performed with equilibrium simulations, in particular relative and absolute small molecule binding free energy calculations to biomolecular targets.
+While these methods are highly flexible, significant care must be taken in avoiding common pitfalls to ensure that computed free energy differences can be robust and reproducible for the chosen force field, and that appropriate corrections are included to permit direct comparison with experimental data.
+
+In this paper, we review current best practices for several popular application domains of  alchemical free energy calculations performed with molecular dynamics simulations, with a particular focus on the intricacies and challenges of relative and absolute free energy calculations of binding of small molecules to biomolecular targets.
 
@@
 \label{sec:intro}
-Alchemical free energy calculations compute free energy differences associated with transfer processes, such as the binding of a small molecule to a receptor, the transfer of a small molecule from an aqueous to apolar phase~\cite{zwanzig1954hightemperature}, or the effects of protein side chain mutations on binding affinities or thermostabilities.
- These calculations use non-physical\footnote{Here, the non-physical nature of the transformation is referred to as "alchemical", a term coined by Tembre and McCammon in Ref.~\cite{tembre1984ligandreceptor}.} intermediate states in which the chemical identity of some portion of the system (such as a small molecule ligand or protein sidechain) is changed by modifying the potential governing the interactions with the environment for the atoms being modified, inserted, or deleted.
+Alchemical free energy calculations allow the estimation of free energy differences associated with transfer processes, such as the binding of a small molecule to a receptor, the transfer of a small molecule from an aqueous to apolar phase~\cite{zwanzig1954hightemperature}, or the effects of protein side chain mutations on binding affinities or thermostabilities.
+ These calculations use non-physical\footnote{Here, the non-physical nature of the transformation is referred to as "alchemical", a term coined by Tembre and McCammon in Ref.~\cite{tembre1984ligandreceptor}.} intermediate states in which the chemical identity of some portion of the system (such as a small molecule ligand or protein side chain) is changed by modifying the potential governing the interactions with the environment for the atoms being modified, inserted, or deleted.
 
 Fig.~\ref{fig:fig_what_is_alchemy} illustrates common free energy changes that may be difficult to compute with unbiased molecular dynamics methods, but are more tractable with alchemical methods.
-In alchemical simulations, the introduction of intermediate \textit{alchemical states} that bridge the high-probability regions of configuration space between two physical endstates of interest, permits the robust computation of free energy for large transformations.
+In alchemical simulations, the introduction of intermediate \textit{alchemical states} that bridge the high-probability regions of configuration space between two physical endstates of interest permits the robust computation of free energy for large transformations.
 Alchemical calculations can be used in a variety of scenarios, such as: 
@@
 \item computing the free energy of a conformational change for a molecule with a high barrier to interconversion (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{B});
-\item computing partition ($\log P$) or distribution ($\log D$) coefficients between environments (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{A})~\cite{rustenburg2016measuring, bosisio2016blinded} 
-\item determining partitioning between compartments into membranes (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{D})~\cite{corey2019insights}. 
+\item computing partition ($\log P$) or distribution ($\log D$) coefficients of a molecule moving between different environments (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{A})~\cite{rustenburg2016measuring, bosisio2016blinded}
+\item determining the free energy of partitioning between compartments into membranes (Fig.~\ref{fig:fig_what_is_alchemy}\textbf{D})~\cite{corey2019insights}. 
 \end{itemize}
 
-Furthermore, alchemical calculations are frequently used to estimate changes in free energies upon modifying a ligand or protein: 
+Furthermore, alchemical calculations are frequently used to estimate changes in free energies upon modifying a ligand or protein. Specifically:
 \begin{itemize}
@@
 
-After an alchemical calculation is performed, which generally involves multiple simulations at a variety of alchemical states, the data must be analyzed to compute an estimate of the free energy for the transformation of interest.
- Early work used simple but statistically suboptimal estimators for this: free energy perturbation (FEP) used a simple (but highly biased) estimator based on the Zwanzig relation~\cite{zwanzig1954hightemperature} or numerical quadrature via thermodynamic integration (TI), for which the theory dates back the better part of a century but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamics, merz1989free}.
+After an alchemical calculation is performed, which frequently involves multiple simulations at a variety of alchemical states, the data must be analyzed to compute an estimate of the free energy for the transformation of interest.
+Early work used simple but statistically suboptimal estimators for this: free energy perturbation (FEP) uses a simple (but highly biased) estimator based on the Zwanzig relation~\cite{zwanzig1954hightemperature} or numerical quadrature via thermodynamic integration (TI), for which the theory dates back the better part of a century but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamics, merz1989free}.
 Macuglia et al.
-review the evolution of biomolecular alchemical free energy calculations into practically useful tools \cite{macuglia2025dawn}.
+recently reviewed the evolution of biomolecular alchemical free energy calculations into practically useful tools \cite{macuglia2025dawn}.
 %
@@
 As the field of molecular simulation can now routinely access microsecond timescales with the aid of GPUs~\cite{salomon-ferrer2013routine}, and millisecond timescales appear to soon be within reach, accurate alchemical calculations on even more challenging problems will become reasonable to perform.
-In the meantime, today's users may find it difficult to get started with these complex calculations whilst also keeping up with the fast pace of change. 
-This Best Practices guide provides current recommendations and tips for users of all experience.
-Updates and suggestions are welcomed via our GitHub repository at \url{https://github.com/alchemistry/alchemical-best-practices}.
+In part because of the many advances that have been made in developing these important computational approaches, today's users may find it difficult to get started with these complex calculations while also keeping up with the fast pace of change.
+This ``Best Practices'' guide aims to provide current recommendations and tips for users of all experience levels, providing context with other new methods being developed.
+Updates and suggestions are highly encouraged via our GitHub repository at \url{https://github.com/alchemistry/alchemical-best-practices}.
 
@@
 
-This Best Practices guide focuses on providing a good starting point for new practitioners and a reference for experienced practitioners. 
+This ``Best Practices'' guide focuses on providing a good starting point for new practitioners and a reference for experienced practitioners. 
 For this purpose we provide a convenient checklist (Sec.~\ref{sec:checklist}) to help ensure all calculations can take advantage of currently-understood best practices for alchemical simulation and analysis.
@@
 
-In this best practices document, we assume that novice practitioners have at least moderate experience with molecular simulation concepts and use of simulation packages. 
-We also assume basic familiarity with the principles of molecular mechanics, molecular dynamics simulations, statistical mechanics, and the biophysics of protein-ligand association.
+In this document, we assume that novice practitioners have at least moderate experience with molecular simulation concepts and use of simulation packages. 
+We also assume a basic familiarity with the principles of molecular mechanics, molecular dynamics simulations, statistical mechanics, and the biophysics of protein-ligand association.
 If you feel unfamiliar with some of these concepts, good starting points can be found in these references~\cite{braun2019best, grossfield2018best, klimovich2015guidelines, shirts2012best}.
 
- While reading this Best Practices guide, it is important to bear in mind \emph{this is not a review} of all free energy calculation methods at the cutting edge of current research.
+ While reading this ``Best Practices'' guide, it is important to bear in mind \emph{this is not a review} of all free energy calculation methods at the cutting edge of current research.
 Instead this guide aims to answer the following questions:
@@
     \item Is my problem suitable for an alchemical calculation? 
+     \item How do I prepare my system to carry out an alchemical calculation?
      \item How do I select an appropriate simulation protocol? 
@@
 Notably, this document---version 2.0 of the original guide---focuses specifically on best practices on single alchemical free energy calculations.
-A companion guide, Best Practices for Multiple Alchemical Free Energy Calculations, covers the considerations involved in high-throughput or large-scale campaigns, such as automation, perturbation network design, and campaign-level data management and analysis.
-Readers aiming to scale up to multiple calculations may find that guide more relevant to those aspects.
-
-In this guide, some other background information may be needed depending on the nature of the alchemical project.
-For example, often, if binding poses are not known, docking calculations or co-folding models can be used to generate an initial small molecule binding pose to start alchemical simulations\cite{scheen2025leveraging, thaler2025boltz}.
-This will require some basic familiarity on how to perform docking or using generative models.
-to generate reasonable simulation starting points~\cite{grinter2014challenges}.
-
-As some of the theoretical background can seem daunting, we do, however, provide a guide to the essential theory behind alchemical free energy calculations in Sec.~\ref{sec:theory}.
+A companion guide, \textit{Best Practices for Multiple Alchemical Free Energy Calculations}, covers the considerations involved in high-throughput or large-scale campaigns, such as automation, perturbation network design, and campaign-level data management and analysis.
+Readers aiming to scale up to multiple calculations will find many important suggestions and ideas relevant to those problems in that document.
+
+In this guide, some additional background information may be needed depending on the nature of the alchemical project.
+For example, if binding poses are not known, docking calculations or co-folding models can be used to generate an initial small molecule binding pose to start alchemical simulations\cite{scheen2025leveraging, thaler2025boltz}.
+This will require some basic familiarity with how to perform docking or use generative models to generate reasonable simulation starting points~\cite{grinter2014challenges}.
+As some of the theoretical background can be daunting to beginners, we do, however, provide a guide to the essential theory behind alchemical free energy calculations in Sec.~\ref{sec:theory}.
+
 In the remainder of this paper, we will cover topics that are key to the preparation (Sec.~\ref{sec:prerequisites}), choice and use of correct protocols (Sec.~\ref{sec:simulation_protocol_choice}), and finally the best practices that should be used in the analysis of alchemical calculations (Sec.~\ref{sec:data_analysis}).
@@
 
-While we try to address as many methods and practices as possible, the field of free energy calculations is broad, and there are many advanced topics that are left to future Best Practices documents focusing on specific issues.
+While we try to address as many methods and practices as possible, the field of free energy calculations is broad, and there are many advanced topics that are left to future ``Best Practices documents'' focusing on specific issues.
 Below, we provide a non-exhaustive list of topics we have \emph{not} addressed, along with some references to provide starting points on these more advanced topics:
@@
 \item Free energy methods that extract the ligand using geometric order parameters and potential of mean force methods~\cite{heinzelmann2017attachpullrelease}
-\item forcefield dependence for protein, ligand, ions, co-solvents, and co-factors.
+\item Forcefield dependence for protein, ligand, ions, co-solvents, and co-factors.
 A number of different studies have looked at the influence of force fields and it is assumed the user has made an appropriate choice for the system under study~\cite{loeffler2018reproducibility, vassetti2019assessment, lopes2015current}.
-\item non-equilibrium free energy calculations~\cite{gapsys2020large}
+\item Non-equilibrium free energy calculations~\cite{gapsys2020large}
 \item Free energy calculations using QM/MM methods~\cite{beierlein2011simple,dybeck2016comparison,cave-ayland2015direct,Schoeller2023}.
@@
 $K_b^{\circ}$ --- binding  constant \\
-$L$ and $R$ --- generic names for ligand and receptor\\
-$\mathcal{O}$ --- overlap matrix\\
+$L$ and $R$ --- generic labels for ligand and receptor\\
+$\mathcal{O}$ --- overlap matrix between alchemical states\\
 $P^{\circ}$ --- standard pressure \\
@@
 $Z$ --- partition function \\
-$\beta \equiv (k_B T)^{-1}$ --- inverse thermal energy \\
+$\beta$ --- inverse thermal energy $(k_B T)^{-1}$ \\
 $\Gamma$ --- configurational space accessible by simulations \\
@@
 $\vec{\lambda}$ --- alchemical progress parameter, which may be multidimensional \\
-$\mu$ --- chemical potential (grand canonical ensemble)\\
+$\mu$ --- chemical potential\\
 $\tau _{eq}$ --- integrated auto-correlation time\\
@@
 
-In this section, we use a statistical mechanics theory of dilute solutions to derive free energy expressions that form the basis of alchemical computational protocols to model non-covalent molecular binding, solvation, and solvent partitioning equilibria.
+In this section, we use the statistical mechanics theory of dilute solutions to derive free energy expressions that form the basis of alchemical computational protocols to model non-covalent molecular binding, solvation, and solvent partitioning equilibria.
 The emphasis here is placed on bridging theoretical foundations and intuition.
@@
 \end{equation}
-where $k_B$ is Boltzmann's constant, $T$ is the absolute temperature, $C_u = N_u/V$ is the number concentration of the solute, $\Lambda_u = h/\sqrt{2 \pi M_u k_B T}$ is the thermal De Broglie wavelength of the solute, where $h$ is Planck's constant and $M_u$ is the solute's total mass.
+where $k_B$ is Boltzmann's constant, $T$ is the absolute temperature, $C_u = N_u/V$ is the number concentration of the solute, $\Lambda_u = h/\sqrt{2 \pi M_u k_B T}$ is the thermal de Broglie wavelength of the solute, where $h$ is Planck's constant and $M_u$ is the solute's total mass.
 The quantity $q'_u$ in Eq.~(\ref{eq:chemical-potential-definition}) is the molecular partition function of the solute at rest in the solvent
@@
 \end{equation}
-where $\beta = 1/(k_B T)$, $\lambda_i$ is the thermal De Broglie wavelength of the $i$-th atom of the solute, and $z_u$ is the intramolecular configurational partition function of the solute
+where $\beta = k_B T)^{-1}$, $\lambda_i$ is the thermal De Broglie wavelength of the $i$-th atom of the solute (not the alchemical parameter as used elsewhere in this document), and $z_u$ is the intramolecular configurational partition function of the solute
 \begin{equation}
@@
 where $\vec{r}_v$ represents the collection of Cartesian coordinates of the atoms of $N_v$ solvent molecules, $U(x_u, \vec{r}_v)$ is the potential energy of the solvent with one solute molecule in conformation $x_u$ placed in an arbitrary position in a volume $V$.
-When the potential energy function of the solution is separable into an intramolecular potential energy of the solute $U(x_u)$ and a solute-solvent non-bonded interaction energy, the potential of mean force can be written as the sum $\Psi_v(x_u) = U(x_u) + W_v(x_u)$ where $W_u(x_u)$ is the solvent potential of mean force, the solvation free energy of the solute fixed in configuration $x_u$.\cite{gilson2007calculation} However, the potential of mean force definition Eq.~(\ref{eq:solvent-pmf-def}) is valid in general.
+When the potential energy function of the solution is separable into an intramolecular potential energy of the solute $U(x_u)$ and a solute-solvent non-bonded interaction energy, the potential of mean force can be written as the sum $\Psi_v(x_u) = U(x_u) + W_v(x_u)$ where $W_u(x_u)$ is the solvent potential of mean force, the solvation free energy of the solute fixed in configuration $x_u$\cite{gilson2007calculation}.
+However, the potential of mean force definition Eq.~(\ref{eq:solvent-pmf-def}) is valid in general.
 Finally, $Z_{N_v}$ is the configurational partition function of the pure solvent
@@
 
-A few notes are in order.
-First, Eq.~(\ref{eq:chemical-potential-definition}) is derived using classical statistical mechanics theory.~\cite{hill1986statthermo} Classical mechanics is considered an excellent approximation for the room temperature processes studied here involving only changes in intermolecular interactions.
+A few notes of explanation are in order.
+First, Eq.~(\ref{eq:chemical-potential-definition}) is derived using classical statistical mechanics theory~\cite{hill1986statthermo}.
+Classical mechanics is considered an excellent approximation for the room temperature processes studied here involving only changes in intermolecular interactions.
 A quantum mechanical treatment is necessary for processes not covered here, such as covalent binding, that involve breaking and forming chemical bonds.
@@
 
-Eq.~(\ref{eq:chemical-potential-definition}) is derived by writing the classical canonical configurational partition function $Z(N_u = 1, N_v)$ of one solute molecule in a solvent of $N_v$ molecules in a volume $V$ and multiplying and dividing it by configurational partition function of the pure solvent [Eq.~(\ref{eq:Z-pure-solvent})], and separating out the integration over the external degrees of freedom of the solute (translations and rotations) giving an $8 \pi^2 V$ term to yield $Z(N_u = 1, N_v) = Z_{N_v} 8 \pi^2 V z_u$ where $z_u$ is defined by Eqs.~(\ref{eq:intra-zu-def}) and (\ref{eq:solvent-pmf-def}).
-The canonical configurational partition function of $N_u$ independent solute molecules in a solvent is then obtained by considering the product of the solute terms $Z(N_u, N_v) = Z_{N_v} (8 \pi^2 V z_u)^{N_u}/N_u!$,~\cite{guggenheim1952mixtures504,simonson2016physical} where the factorial term takes care of indistinguishability.
+Eq.~(\ref{eq:chemical-potential-definition}) is derived by writing the classical canonical configurational partition function $Z(N_u = 1, N_v)$ of one solute molecule in a solvent of $N_v$ molecules in a volume $V$ and multiplying and dividing it by configurational partition function of the pure solvent [Eq.~(\ref{eq:Z-pure-solvent})], and then separating out the integration over the external degrees of freedom of the solute (translations and rotations) giving an $8 \pi^2 V$ term to yield $Z(N_u = 1, N_v) = Z_{N_v} 8 \pi^2 V z_u$ where $z_u$ is defined by Eqs.~(\ref{eq:intra-zu-def}) and (\ref{eq:solvent-pmf-def}).
+The canonical configurational partition function of $N_u$ independent solute molecules in a solvent is then obtained by considering the product of the solute terms $Z(N_u, N_v) = Z_{N_v} (8 \pi^2 V z_u)^{N_u}/N_u!$,~\cite{guggenheim1952mixtures504,simonson2016physical} where the factorial term accounts for indistinguishability.
 Finally, the canonical partition function $Q(N_u, N_v)$ of the solution is obtained by including the appropriate terms coming from the momenta and differentiating the resulting Helmholtz free energy $A(N_u, N_v) = -k_B T \ln Q(N_u, Nv)$ with respect to $N_u$ to obtain Eq.~(\ref{eq:chemical-potential-definition}).
@@
 Since the chemical potential can be equivalently obtained from the constant-volume differentiation of the Helmholtz free energy or the constant-pressure differentiation of the Gibbs free energy, Eq.~(\ref{eq:chemical-potential-definition}) is correct.
-However, the volume dependence of the chemical potential has to be considered when modeling the Gibbs free energy change for constant-pressure processes where the system's volume changes significantly.~\cite{gilson1997statisticalthermodynamic} Terms due to volume variations are generally considered small for most processes of solvation, solvent partitioning, and molecular binding at atmospheric pressure and are often ignored in alchemical calculations.
+However, the volume dependence of the chemical potential has to be considered when modeling the Gibbs free energy change for constant-pressure processes where the system's volume changes significantly~\cite{gilson1997statisticalthermodynamic}.
+Terms due to volume variations at constant pressure are generally considered small for most processes of solvation, solvent partitioning, and molecular binding at atmospheric pressure and are often ignored in alchemical calculations.
+This is because when a single molecule is eliminated, errors will be of the magnitude of the fractional change of the system size.
+When one particle in 1000 disappears, this will be negligible with respect to other source of error.
 Hence, we will use Eq.~(\ref{eq:chemical-potential-definition}) to derive expressions for $\Delta G$'s.
-
-The standard chemical potential of a solute in an ideally dilute solution is obtained from Eq.~(\ref{eq:chemical-potential-definition}) by setting the concentration to the standard concentration $C^\circ$, generally $1$ mol/L equivalent to $1/1,661$ particles/\AA$^{3}$,
-\begin{equation}
-  \mu_u^\circ = -k_B T \ln \frac{q'_u}{C^\circ \Lambda_u^3} \, .
-  \label{eq:chemical-potential-standard}
-\end{equation}
-The standard state of a dilute solution is defined as an ideally dilute solution at concentration $C^\circ$.~\cite{levine2009physicalchemistrybook6ed} Hence, Eq.~(\ref{eq:chemical-potential-standard}) is formally correct for real solutions as well.
 
@@
 In the present context, the quantity being averaged often turns out to be the exponential of a perturbation energy function (the \textit{Zwanzig relationship}~\cite{zwanzig1954hightemperature}), which is efficiently evaluated using the stratification algorithms and free energy estimators discussed in Sections \ref{subsec:formulations-abfe-rbfe}, \ref{subsec:numerical-implementations},  and \ref{subsec:estimators} rather than directly.
-Here we assume that an implementation is well-formed once it is expressed as a combination of exponential averages, assuming they can be evaluated in practice.
-As in the previous section, we use the solvent potential of mean force to lighten the notation when possible.
+In this section, we assume that an implementation is well-formed once it is expressed as a combination of exponential averages, assuming they can be evaluated in practice.
+As in the previous section, we use the solvent potential of mean force to simplify the notation when possible.
 The solvent degrees of freedom can be restored  in the resulting expressions using Eq.~(\ref{eq:solvent-pmf-def}).
 
-\subsection{Solvation}\label{sec:theory-solvation}
+\subsection{Solvation free energy}\label{sec:theory-solvation}
+
+The standard chemical potential of a solute in an ideally dilute solution is obtained from Eq.~(\ref{eq:chemical-potential-definition}) by setting the concentration to the standard concentration $C^\circ$, generally $1$ mol/L equivalent to $1/1661.0$ particles/\AA$^{3}$,
+\begin{equation}
+  \mu_u^\circ = -k_B T \ln \frac{q'_u}{C^\circ \Lambda_u^3} \, .
+  \label{eq:chemical-potential-standard}
+\end{equation}
+The standard state of a dilute solution is defined as an ideally dilute solution at concentration $C^\circ$~\cite{levine2009physicalchemistrybook6ed}.
+Hence, Eq.~(\ref{eq:chemical-potential-standard}) is formally correct for real solutions as well.
 
@@
 \end{equation}
-where we used Eq.~(\ref{eq:chemical-potential-standard}) for the solution and  Eq.~(\ref{eq:chemical-potential-definition}) for the gas (assumed ideal) at the standard gas concentration $C = P^\circ/k_B T$ where $P^\circ = 1$ bar is the standard pressure, and Eq.~(\ref{eq:intra-qpu-def}) noting that all momentum terms cancel.
+where we use Eq.~(\ref{eq:chemical-potential-standard}) for the solution, Eq.~(\ref{eq:chemical-potential-definition}) for the gas (assumed ideal) at the standard gas concentration $C = P^\circ/k_B T$ where $P^\circ = 1$ bar is the standard pressure, and Eq.~(\ref{eq:intra-qpu-def}) noting that all momentum terms cancel.
 The first term in Eq.~(\ref{eq:DG0-solvation-def}) is an ideal term $\Delta G^\circ_{\mathrm{slv,ideal}}$ that accounts for the difference of standard states for solution and gases.
-The second term is the excess solvation free energy $\Delta G_{\mathrm{slv, exc.}}$ that can be evaluated by numerical alchemical calculations (see below).
-Hence Eq.~(\ref{eq:chemical-potential-standard}) can be used to compare calculated and experimental standard free energies of solvation.\cite{thompson2004new}
+The second term is the excess solvation free energy $\Delta G_{\mathrm{slv, exc.}}$ that can be evaluated by numerical alchemical calculations as described in this document.
+Hence Eq.~(\ref{eq:chemical-potential-standard}) can be used to compare calculated and experimental standard free energies of solvation\cite{thompson2004new}.
 
@@
 \end{equation}
-The relation above is used to connect alchemical calculations to Henry's constant values, often after converting them to the excess free energy scale~\cite{cabani1981group,nicholls2008predicting,mobley2014freesolv}
+The relation above is used to connect alchemical calculations to Henry's constant values, generally after converting them to the excess free energy scale~\cite{cabani1981group,nicholls2008predicting,mobley2014freesolv}
 \begin{equation}
@@
 \end{equation}
-Because it involves only intramolecular partition functions, Eq.~(\ref{eq:DG-solvation-excess}) states that the excess free energy of solvation corresponds to the process of moving the solute from a fixed position and orientation in the gas to a fixed position and orientation in the solvent.
-This is sometimes referred to as the solvation free energy in the Ben-Naim standard state.~\cite{bennaim1984solvation}
+Eq.~(\ref{eq:DG-solvation-excess}), which involves only intramolecular partition functions (as described above),  states that the excess free energy of solvation corresponds to the process of moving the solute from a fixed position and orientation in the gas to a fixed position and orientation in the solvent.
+This is sometimes referred to as the solvation free energy in the Ben-Naim standard state~\cite{bennaim1984solvation}.
 
@@
 The objective of alchemical solvation free energy calculations is to implement Eq.~(\ref{eq:zratio-solv-dgexc}), by means of \emph{coupling} or \emph{decoupling} computational protocols and alchemical intermediate states as  discussed in section \ref{subsec:numerical-implementations}.
-
-\subsection{Solvent partitioning}\label{sec:theory-partitioning}
+Subsequent sections will also discuss improved approaches to calculate this ratio of integrals.
+
+\subsection{Free energy of solvent partitioning}\label{sec:theory-partitioning}
 
@@
 \end{equation}
-where $P_{sv} = C_{A(v)}/C_{A(s)}$ is the partition coefficient of $A$ for the two solvents.~\cite{rustenburg2016measuring} Eq.~(\ref{eq:DG0-partitioning-def}) 
-immediately identifies the partition coefficient with the ratio of intramolecular configurational partition functions of $A$ in the two solvents:
+where $P_{sv} = C_{A(v)}/C_{A(s)}$ is the partition coefficient of $A$ for the two solvents~\cite{rustenburg2016measuring}. Eq.~(\ref{eq:DG0-partitioning-def}) 
+immediately identifies the partition coefficient with the ratio of intramolecular configurational partition functions of solute $A$ in the two solvents:
 \begin{equation}
@@
 The ratio of partition functions in Eq.~(\ref{eq:partition-coefficient}) can be treated in the same way as Eq.~(\ref{eq:zratio-solv-average}) to express it as the ensemble average in solvent $s$ of the change in potential energy $\Delta U$ for moving the solute to the other solvent $v$.
-However, in practice it has been more common to evaluate Eq.~(\ref{eq:DG0-partitioning-def}) as the difference of excess solvation free energies in the two solvents separately evaluated using Eq.~(\ref{eq:zratio-solv-dgexc}).~\cite{bosisio2016blinded} Formally, this approach is derived by multiplying and dividing Eq~(\ref{eq:partition-coefficient}) by $z_{A(\mathrm{g})}$ and expressing each terms as an ensemble average in the decoupled ensemble
+However, in practice it has been more common to evaluate Eq.~(\ref{eq:DG0-partitioning-def}) as the difference of excess solvation free energies in the two solvents separately evaluated using Eq.~(\ref{eq:zratio-solv-dgexc})~\cite{bosisio2016blinded}.
+Formally, this approach is derived by multiplying and dividing Eq~(\ref{eq:partition-coefficient}) by $z_{A(\mathrm{g})}$ and expressing each terms as an ensemble average in the decoupled ensemble
 \begin{equation}
@@
 \end{equation}
-is an example of a \emph{double-decoupling} alchemical computational protocol.~\cite{hermans1986freeenergy,gilson1997statisticalthermodynamic}
+is an example of a \emph{double-decoupling} alchemical computational protocol~\cite{hermans1986freeenergy,gilson1997statisticalthermodynamic}.
 
@@
 The intramolecular configurational partition function $z_{RL(v)}$ of the complex is similar but the integration is limited to a chosen range of configurations in which receptor and ligand are considered ``bound''.
-To this end, Gilson et al.~\cite{gilson1997statisticalthermodynamic} introduced an indicator function $I(\zeta)$ equal to one when the six degrees of freedom of the complex that defines the position and orientation of the ligand relative to the receptor, collectively denoted as $\zeta$, are within the specified range of the bound state and equal to zero otherwise:
+To this end, Gilson et al.~\cite{gilson1997statisticalthermodynamic}.
+introduced an indicator function $I(\zeta)$ equal to one when the six degrees of freedom of the complex that defines the position and orientation of the ligand relative to the receptor, collectively denoted as $\zeta$, are within the specified range of the bound state and equal to zero otherwise:
 \begin{equation}
@@
 Generally, to formulate the statistical mechanics expression of the binding constant Eq.~(\ref{eq:DG0-binding-constant-def}) as an ensemble average, we first need to ensure that the configurational integrals in the numerator and denominator are over the same set of degrees of freedom.
-Note that the intramolecular configurational partition function of the complex, $z_{RL(v)}$, includes six more degrees of freedom--the position and orientation of the ligand relative to the receptor collectively described by the variable $\zeta$--than the product of the intramolecular partition functions, $z_{R(v)} z_{L(v)}$, of the separated receptor and ligand.
+Note that the intramolecular configurational partition function of the complex, $z_{RL(v)}$, includes six more degrees of freedom---the position and orientation of the ligand relative to the receptor collectively described by the variable $\zeta$---than the product of the intramolecular partition functions, $z_{R(v)} z_{L(v)}$, of the separated receptor and ligand.
 To correct this discrepancy, we multiply and divide the right-hand side of Eq.~(\ref{eq:DG0-binding-constant-def})  by the quantity
@@
 
-Eq~(\ref{eq:DG0-binding-constant-def-average-noOmega}) instructs to collect samples of the solvent-averaged binding energy $\Delta \Psi$ in the ensemble where ligand and receptor are not interacting.
+Eq.~(\ref{eq:DG0-binding-constant-def-average-noOmega}) instructs to collect samples of the solvent-averaged binding energy $\Delta \Psi$ in the ensemble where ligand and receptor are not interacting.
 This can be done directly for simple solutes when the solvent is modeled implicitly.\cite{gallicchio2014virtual} However, with explicit solvation specialized techniques are required.
@@
 \end{equation}
-where $[RL]$, $[L]$, and $[R]$ are the equilibrium concentrations of complex, ligand, and receptor, respectively, by, for example, monitoring the relative population $P(RL)/P(R+L)$ of bound and unbound configurations in molecular dynamics simulations.~\cite{jong2011determining,pan2017quantitative}
+where $[RL]$, $[L]$, and $[R]$ are the equilibrium concentrations of complex, ligand, and receptor, respectively, by, for example, monitoring the relative population $P(RL)/P(R+L)$ of bound and unbound configurations in molecular dynamics simulations~\cite{jong2011determining,pan2017quantitative}.
 
@@
 and similarly for the second leg, we recover Eq.~(\ref{eq:ddm-dg1-dg2}) obtained for double-decoupling.
-This shows that, while the thermodynamic path and the meaning of the $\Delta G_1$ and $\Delta G_2$ free energy components differ, the double-decoupling and alchemical transfer formulations both connect the unbound and bound states through an auxiliary intermediate state: the ligand decoupled state for double-decoupling, and the symmetric alchemical intermediate for alchemical transfer.~\cite{azimi2025potential} However, alchemical transfer simulates both legs with a single simulation system, while double-decoupling accomplishes them in separate simulations.
+This shows that, while the thermodynamic path and the meaning of the $\Delta G_1$ and $\Delta G_2$ free energy components differ, the double-decoupling and alchemical transfer formulations both connect the unbound and bound states through an auxiliary intermediate state: the ligand decoupled state for double-decoupling, and the symmetric alchemical intermediate for alchemical transfer~\cite{azimi2025potential}.
+However, alchemical transfer simulates both legs with a single simulation system, while double-decoupling accomplishes them in separate simulations.
 
@@
 
-Non-equilibrium free energy techniques provide an alternate approach to this problem, driving $\lambda$ between states (see Sec.~\ref{sec:sampling_schemes}).~\cite{jarzynski1997nonequilibrium,jarzynski1998equilibrium,crooks2000pathensemble, gapsys2020large}
+Non-equilibrium free energy techniques provide an alternate approach to this problem, driving $\lambda$ between states (see Sec.~\ref{sec:sampling_schemes})~\cite{jarzynski1997nonequilibrium,jarzynski1998equilibrium,crooks2000pathensemble, gapsys2020large}.
 
@@
 Thus, it is recommended to perform ensemble simulations in order to perform ensemble averaging to obtain thermodynamic quantities in a reliable manner.
-Bhati \textit{et al.} developed the TIES method~\cite{bhati2017, bhati2022, bhati2025} which is based on the use of ensemble methods, where the ensemble size can be adjusted based on available computational resource and required precision~\cite{wan2023}.
+Bhati et al.
+developed the TIES method~\cite{bhati2017, bhati2022, bhati2025} which is based on the use of ensemble methods, where the ensemble size can be adjusted based on available computational resource and required precision~\cite{wan2023}.
 Studies have shown that, in general, an ensemble size of 5 and 10 members is a reasonable starting point for alchemical RBFE and ABFE calculations, respectively, but recommendations are provided to adjust them based on some rules of thumb~\cite{bhati2022, bhati2025}.
@@
 The 'mixed' intermediate atoms (M) are used as well, but they strictly interpolate between physical and dummy atoms, or vice versa.
-Bhati \textit{et al.}~\cite{bhati2017, bieniek2021} presented a variant of the single coordinate dual topology approach which is used in the TIES method.
+Bhati et al.~\cite{bhati2017, bieniek2021} presented a variant of the single coordinate dual topology approach which is used in the TIES method.
 It involves a unique post-processing of the initial MCSS to ensure that the final MCSS is both structurally and chemically identical and has a single set of coordinates while maintaining dual topology for the alchemical parts.
@@
 One issue which requires particular care is the use of constraints.
-Commonly, bonds involving hydrogen are constrained to a fixed length using algorithms such as SHAKE or LINCS, allowing the use of longer timesteps~\cite{krautler2001fast}.
+Commonly in molecular simulations, bonds involving hydrogen are constrained to a fixed length using algorithms such as SHAKE or LINCS, rather than being allowed to vibrate.
+This approach enables the use of longer timesteps~\cite{krautler2001fast}.
 However, in single topology relative free energy calculations, the atoms involved might be mutated to other atom types---for example, in a mutation of methane to methanol, one hydrogen might become an oxygen atom.
-The bonds with such atoms might not have any constraints, or if all bonds are constrained, would have constraints of different lengths.
+The bonds with such atoms might not have any constraints, or if all bonds are constrained, have constraints of different lengths.
 Some molecular dynamics engines are not set up to recognize this change, or at least not to correctly include contributions to the free energy from changing constraints/constraint length, so results for a transformation can be erroneous.
-At present the most general solution to this problem is simply to avoid the use of constraints (and thus use a smaller timestep if necessary, usually of around 1 fs) in any relative free energy calculation involving a transformation of a constrained bond.
+At present, the most general solution to this problem is simply to avoid the use of constraints (and thus use a smaller timestep if necessary, usually of around 1 fs) in any relative free energy calculation involving a transformation of a constrained bond.
 Individual software programs and settings can handle such issues.
@@
 Over the course of each transition, Hamiltonian derivative with respect to $\lambda$ is recorded; subsequently the curves are integrated to yield work distributions for the forward and reverse processes.
-Crooks Fluctuation Theorem~\cite{crooks1999entropy} allows estimating free energy difference between the end states from the generated work distributions.~\cite{shirts2003equilibrium}
+Crooks Fluctuation Theorem~\cite{crooks1999entropy} allows estimating free energy difference between the end states from the generated work distributions~\cite{shirts2003equilibrium}.
 
@@
 
-%Bhati \textit{et al.} recently published two studies directly comparing the equilibrium (EQ) and NEQ approaches at unprecedented scales for both RBFE and ABFE calculations using their ensemble simulation-based TIES method~\cite{wan2023eqvsneq, bhati2025}.
+%Bhati et al.
+recently published two studies directly comparing the equilibrium (EQ) and NEQ approaches at unprecedented scales for both RBFE and ABFE calculations using their ensemble simulation-based TIES method~\cite{wan2023eqvsneq, bhati2025}.
 Further, they systematically studied the effect of various parameters involved in setting up calculations for both approaches and recommended optimal protocols in each case for industrial applications.
@@
 If more samples are needed, the simulations can be resumed, or, started again in different initial conditions enlarging the overall simulation ensemble.
-Bhati \textit{et al.} recommended such a scheme for RBFE and ABFE calculations using TIES~\cite{bhati2022, bhati2025, wan2023eqvsneq, coveney2022corners}.
+Bhati et al.
+recommended such a scheme for RBFE and ABFE calculations using TIES~\cite{bhati2022, bhati2025, wan2023eqvsneq, coveney2022corners}.
 They suggest a set of optimal parameters (ensemble size, simulation length, number of lambda-states, number and length of NEQ transitions, etc.) to be used initially and suggest pre-defined criteria (uncertainties below 0.5 kcal/mol and overlap between forward and reverse work values in case of NEQ calculations) to decide on the need of additional calculations (in the way suggested) until these criteria are met.
@@
 In the limit of sufficiently large data, perhaps several hundred or thousand independent points, exhaustive analyses of model data has shown that the error estimates of BAR, MBAR, and TI are all quite reliable when there is good overlap between intermediate states.
-For example, for a series of hydration free energy calculations of model systems of various sizes analytical estimates of the error, bootstrap estimates, and standard deviations over 100 independent simulations started at different locations along a long initial simulations all agreed within a 5---10\%.~\cite{paliwal2011benchmark} For EXP, the number of samples at which they become reliable is significantly higher, especially if overlap between intermediate states is lower.
-
-However, in the limit of moderate or small amounts of data (tens or hundreds of independent samples) these errors become somewhat unreliable, often off by a factor of 2 from what would get if more data was collected.~\cite{???} These estimators of uncertainty are all derived from propagation of error formulae from the estimators themselves.
+For example, for a series of hydration free energy calculations of model systems of various sizes analytical estimates of the error, bootstrap estimates, and standard deviations over 100 independent simulations started at different locations along a long initial simulations all agreed within a 5---10\%~\cite{paliwal2011benchmark}.
+For EXP, the number of samples at which they become reliable is significantly higher, especially if overlap between intermediate states is lower.
+
+However, in the limit of moderate or small amounts of data (tens or hundreds of independent samples) these errors become somewhat unreliable, often off by a factor of 2 from what would get if more data was collected~\cite{???}.
+These estimators of uncertainty are all derived from propagation of error formulae from the estimators themselves.
 When the underlying data is noisy, this approximation is not very good.
@@
 between $i$ and $i+1$ and between $i$ and $i-1$), they will \textit{not} have significant overlap with their next nearest neighbors $i+2$ and $i-2$.
-Thus MBAR does not actually get significant information from these energy differences, so one might as well not even calculate them, and just perform BAR between nearest neighbors.~\cite{paliwal2011benchmark}
-\item TI usually gives similar values as MBAR implemented with sufficient numbers of intermediates, but quadrature errors that are hard to estimate beforehand  can occur if one does not carefully select $\lambda$ values such that the overlap between states is sufficiently high.~\cite{paliwal2011benchmark}
+Thus MBAR does not actually get significant information from these energy differences, so one might as well not even calculate them, and just perform BAR between nearest neighbors~\cite{paliwal2011benchmark}.
+\item TI usually gives similar values as MBAR implemented with sufficient numbers of intermediates, but quadrature errors that are hard to estimate beforehand  can occur if one does not carefully select $\lambda$ values such that the overlap between states is sufficiently high~\cite{paliwal2011benchmark}.
 \item WHAM is an approximation to MBAR, and there are no compelling reasons it should be used.
@@
 Other factors, like system preparation (choice of protonation state, tautomer, counterion presence, salt concentration, etc.) can also substantially impact results~\cite{mobley2017predicting, mobley2017predictinga}, so unless modelers are confident they have these factors correct, sensitivity to these choices may also need to be examined.
-Incidentally, Vassaux \textit{et al.} performed a systematic study comparing the effect of both aleatoric as well as parametric uncertainties for protein-ligand free energy calculations using NAMD~\cite{vassaux2021}.
+Incidentally, et al.
+performed a systematic study comparing the effect of both aleatoric as well as parametric uncertainties for protein-ligand free energy calculations using NAMD~\cite{vassaux2021}.
 They found that, indeed, ensemble simulations are necessary to control both sources of uncertainties arising in simulations, just as is anticipated in the field of uncertainty quantification.
@@
 In practice, analytical error estimate of the BAR estimator \cite{shirts2003equilibrium} gives an indication when the overlap is lacking: the estimated error rapidly grows with the increase in the dissipated work.
-\cite{shirts2005comparison} Bhati~\textit{et al.} recently studied the effect of such overlaps on the accuracy of predicted RBFEs~\cite{wan2023eqvsneq} as well as ABFEs~\cite{bhati2025} but found that overlap (or lack of it) does not capture the reliability of calculations very well.
+\cite{shirts2005comparison} Bhati et al.
+recently studied the effect of such overlaps on the accuracy of predicted RBFEs~\cite{wan2023eqvsneq} as well as ABFEs~\cite{bhati2025} but found that overlap (or lack of it) does not capture the reliability of calculations very well.
 They recommended an alternative metric, ``distance'' defined as the difference between the inner extremes of the two work value distributions, which correlates quite well with the unsigned errors in free energy predictions and hence was found a better indicator of the reliability of NEQ calculations.
@@
 It is worth highlighting here that the necessity of ensemble simulations to capture the true uncertainty in predictions ensuring their reliability is equally applicable for long simulations.
-Bhati \textit{et al.} performed ensembles of 10 microsecond long simulations and reported large variability in several properties calculated from each replica~\cite{bhati2023}.
+Bhati et al.
+performed ensembles of 10 microsecond long simulations and reported large variability in several properties calculated from each replica~\cite{bhati2023}.
 The agreement between the binding free energy calculations and the reversible binding simulations---given the same system preparation and the same force field parameters---will strongly support the correctness of both calculations, as the same results are arrived at by two independent methods, and any discrepancy will suggest some systematic error in one, or both, of the two methods.
